### 2013-06-16 doc/slides/gsoc_proposal/gsoc_proposal.tex
Compiled the GSoC proposal into a presentation

\documentclass{beamer}

\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usecolortheme{dove}

\begin{document}

\title{Typed Lua}
\subtitle{Google Summer of Code Proposal}
\author{André Murbach Maidl}
\institute[PUC-Rio]{LabLua\\PUC-Rio}
\date{}

\frame{\titlepage}

\section{Static Typing X Dynamic Typing}
\begin{frame}
\frametitle{Static Typing X Dynamic Typing}
\begin{itemize}
\item Static Typing
\begin{itemize}
\item Better documentation
\item Error detection at compile time
\item More efficient program execution
\end{itemize}
\item Dynamic Typing
\begin{itemize}
\item Faster adaptation to changing requirements
\item Quicker development
\end{itemize}
\end{itemize}
\end{frame}

\subsection{Multiple languages to build large softwares}
\begin{frame}
\frametitle{Multiple languages to build large softwares}
\begin{itemize}
\item Prototype using a dynamically typed language
\item Rewrite it in a statically typed language
\end{itemize}
\end{frame}

\section{Gradual Typing}
\begin{frame}
\frametitle{Gradual Typing}
\begin{itemize}
\item The choice between static and dynamic typing in the same language
\item Use type annotations where static typing is needed
\item Avoid type annotations where dynamic typing is sufficient
\item No rewriting after prototyping phase
\item The programmer can gradually migrate to static typing
\begin{itemize}
\item Specially in module boundaries
\end{itemize}
\end{itemize}
\end{frame}

\subsection{Gradual Typing and Dynamically Typed Languages}
\begin{frame}
\frametitle{Gradual Typing and Dynamically Typed Languages}
\begin{itemize}
\item Dart and TypeScript
\begin{itemize}
\item New languages derived from JavaScript
\item Both target JavaScript
\item Both are designed with a gradual typing system
\end{itemize}
\item Typed Scheme, Typed Racket, and core.typed
\begin{itemize}
\item Gradual typing extensions to Scheme, Racket, and Clojure
\end{itemize}
\item Perl
\item Python
\item Ruby
\end{itemize}
\end{frame}

\section{Typed Lua}
\begin{frame}
\frametitle{Typed Lua}
\begin{itemize}
\item A proposal to design a gradual type system for Lua
\begin{itemize}
\item Should be simple enough to preserve existent Lua idioms
\end{itemize}
\item Superset of Lua
\begin{itemize}
\item Type annotations
\item Compile-time type checking
\item Classes
\item Interfaces
\item Modules
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Typed Lua}
\begin{itemize}
\item Lua X OO programming and modules
\begin{itemize}
\item It is possible to write OO code in Lua
\item It is possible to split Lua code into modules
\item Does not set policies on how these features should behave
\item Due to its use as an embedded and extension language
\end{itemize}
\item Typed Lua intended usa is as an application language
\item So the policies for OO and modules should be part of the language
\item The gradual type system should also enforce these policies
\end{itemize}
\end{frame}

\subsection{Related Work}
\begin{frame}
\frametitle{Related Work}
\begin{itemize}
\item Some Lua projets implement parts of what is being proposed
\item However, none of them implement type checking along with OO programming
\item LuaInspect
\begin{itemize}
\item A tool that performs code analysis
\item For instance, flagging undeclared identifiers
\item It does not try to analyse OO code
\item It does not perform type checking
\end{itemize}
\item MetaLua and MoonScript
\begin{itemize}
\item Both support some sort of OO programming
\item However, both of them do not perform compile-time type checking
\end{itemize}
\item Tidal Lock
\begin{itemize}
\item A prototype of a gradual type system for Lua
\item It is not typing the common OO idioms used by Lua programmers
\end{itemize}
\end{itemize}
\end{frame}

\subsection{Conclusions}
\begin{frame}
\frametitle{Conclusions}
\begin{itemize}
\item We expect to deliver a tool that helps Lua community
\begin{itemize}
\item Allowing better documentation through a gradual type system
\item Enabling the definition of structured types using interfaces and classes
\item Offering compile-time type checking with OO programming style
\item Further integrate Typed Lua to text editors and IDEs
\item Allowing autocompletion and flagging type errors
\end{itemize}
\item Typed Lua will compile to Lua preserving existent idioms
\end{itemize}
\end{frame}

\end{document}

### 2013-06-17 doc/slides/gsoc_proposal/gsoc_proposal.tex
Added more information to the presentation about the GSoC proposal

@@
 \end{itemize}
-\item Typed Scheme, Typed Racket, and core.typed
+\item Typed Scheme, Typed Racket, core.typed, and StrongRuby
 \begin{itemize}
-\item Gradual typing extensions to Scheme, Racket, and Clojure
+\item Gradual typing extensions to Scheme, Racket, Clojure, and Ruby
 \end{itemize}
-\item Perl
+\item Perl 6
+\begin{itemize}
+\item Not compatible with Perl 5
+\end{itemize}
 \item Python
-\item Ruby
+\begin{itemize}
+\item gradual (package)
+\item Run-time type checking semantics of gradual typing
+\item Works only on Python 3
+\end{itemize}
 \end{itemize}
@@
 \end{itemize}
-\item Superset of Lua
+\item A superset of Lua
 \begin{itemize}
@@
 \end{itemize}
-\item Typed Lua intended usa is as an application language
+\item Typed Lua intended use is as an application language
 \item So the policies for OO and modules should be part of the language
@@
 \begin{itemize}
-\item Some Lua projets implement parts of what is being proposed
+\item Some Lua projects implement parts of what is being proposed
 \item However, none of them implement type checking along with OO programming
@@
 \item For instance, flagging undeclared identifiers
-\item It does not try to analyse OO code
+\item It does not try to analyze OO code
 \item It does not perform type checking

### 2013-06-18 doc/slides/gsoc_proposal/gsoc_proposal.tex
Added some statistics to the presentation about the GSoC proposal

@@
 
+\subsection{Some Statistics}
+\begin{frame}
+\frametitle{Some Statistics}
+\begin{itemize}
+\item Data from LuaRocks as of June 10th
+\item 2950 .lua files
+\item \textbf{luac} (version 5.2.2) compiles 2864 of the 2950 files
+\item We have 2069 files when we ignore files that are generated by
+MoonScript and documentation and test subdirectories
+\item 18864 functions are being declared
+\begin{itemize}
+\item 4127 (21.9\%) are actually methods
+\item 1586 (8.4\%) use \textbf{type} to inspect the tags of the arguments
+\item 771 (4.1\%) return \textbf{nil} plus something else
+\item 119 (0.6\%) return \textbf{false} plus something else
+\item 549 (2.9\%) use \textbf{setmetatable}
+\item 175 (0.9\%) use \textbf{getmetatable}
+\end{itemize}
+\item The table constructor is used 19049 times
+\begin{itemize}
+\item 4571 (24.00\%) times is the empty construct
+\item 14475 (75.99\%) times is a constructor with only static keys
+\item 3 (0.01\%) times is a constructor with only dynamic keys
+\item There is no constructor with static and dynamic keys
+\end{itemize}
+\end{itemize}
+\end{frame}
+
 \subsection{Conclusions}

### 2013-06-21 doc/slides/gsoc_proposal/gsoc_proposal.tex
Cleaning up the presentation about the GSoC proposal

@@
 \usepackage[utf8]{inputenc}
-\usecolortheme{dove}
+\usepackage{beamerthemesplit}
 
@@
 
-\section{Static Typing X Dynamic Typing}
+\section{Static Typing versus Dynamic Typing}
 \begin{frame}
-\frametitle{Static Typing X Dynamic Typing}
+\frametitle{Static Typing versus Dynamic Typing}
 \begin{itemize}
@@
 \item Avoid type annotations where dynamic typing is sufficient
-\item No rewriting after prototyping phase
+\end{itemize}
+\end{frame}
+
+\subsection{The levels of Gradual Typing}
+\begin{frame}
+\frametitle{The levels of Gradual Typing}
+\begin{itemize}
 \item The programmer can gradually migrate to static typing
-\begin{itemize}
 \item Specially in module boundaries
-\end{itemize}
+\begin{enumerate}
+\item The programmer write modules that use only dynamic typing.
+\item The programmer creates an interface to type a module that
+is written using only dynamic typing. This module is not
+type checked, but their code that use this module is.
+\item The programmer write modules that use only static typing.
+\end{enumerate}
 \end{itemize}
@@
 \item Both target JavaScript
-\item Both are designed with a gradual typing system
+\item Both are designed with a gradual type system
 \end{itemize}
-\item Typed Scheme, Typed Racket, core.typed, and StrongRuby
+\item Typed Racket and core.typed
 \begin{itemize}
-\item Gradual typing extensions to Scheme, Racket, Clojure, and Ruby
+\item Gradual typing extensions to Racket and Clojure
 \end{itemize}
@@
 \item gradual (package)
-\item Run-time type checking semantics of gradual typing
+\item Runtime type checking semantics of gradual typing
 \item Works only on Python 3
@@
 \begin{itemize}
-\item Should be simple enough to preserve existent Lua idioms
+\item Should be simple enough to preserve existing Lua idioms
 \end{itemize}
@@
 \begin{itemize}
-\item Lua X OO programming and modules
+\item Lua versus modules and OO programming
 \begin{itemize}
+\item It is possible to split Lua code into modules
 \item It is possible to write OO code in Lua
-\item It is possible to split Lua code into modules
-\item Does not set policies on how these features should behave
-\item Due to its use as an embedded and extension language
+\item Lua does not set policies on how these features should behave
+\item Probably due to its use as an embedded and extension language
 \end{itemize}
 \item Typed Lua intended use is as an application language
-\item So the policies for OO and modules should be part of the language
-\item The gradual type system should also enforce these policies
+\item So the policies for modules and OO programming should be
+part of the language
+\item The gradual type system helps to enforce these policies
 \end{itemize}
@@
 \item Some Lua projects implement parts of what is being proposed
-\item However, none of them implement type checking along with OO programming
-\item LuaInspect
+\item None implement type checking along with OO programming
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{LuaInspect}
 \begin{itemize}
@@
 \end{itemize}
-\item MetaLua and MoonScript
+\end{frame}
+
+\begin{frame}
+\frametitle{MetaLua and MoonScript}
 \begin{itemize}
 \item Both support some sort of OO programming
-\item However, both of them do not perform compile-time type checking
+\item Both do not perform type checking
 \end{itemize}
-\item Tidal Lock
+\end{frame}
+
+\begin{frame}
+\frametitle{Tidal Lock}
 \begin{itemize}
@@
 \item It is not typing the common OO idioms used by Lua programmers
-\end{itemize}
 \end{itemize}
@@
 \begin{itemize}
-\item Data from LuaRocks as of June 10th
-\item 2950 .lua files
+\item Data from LuaRocks repository as of June 10th
+\item LuaRocks repository contains 2950 .lua files
 \item \textbf{luac} (version 5.2.2) compiles 2864 of the 2950 files
@@
 MoonScript and documentation and test subdirectories
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Function declaration}
+\begin{itemize}
 \item 18864 functions are being declared
@@
 \item 119 (0.6\%) return \textbf{false} plus something else
+\end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Metatables}
+\begin{itemize}
+\item 18864 functions are being declared
+\begin{itemize}
 \item 549 (2.9\%) use \textbf{setmetatable}
@@
 \end{itemize}
-\item The table constructor is used 19049 times
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Table constructor}
+\begin{itemize}
+\item The table constructor is being used 19049 times
 \begin{itemize}
@@
 
-\subsection{Conclusions}
-\begin{frame}
-\frametitle{Conclusions}
-\begin{itemize}
-\item We expect to deliver a tool that helps Lua community
-\begin{itemize}
-\item Allowing better documentation through a gradual type system
-\item Enabling the definition of structured types using interfaces and classes
-\item Offering compile-time type checking with OO programming style
-\item Further integrate Typed Lua to text editors and IDEs
-\item Allowing autocompletion and flagging type errors
-\end{itemize}
-\item Typed Lua will compile to Lua preserving existent idioms
-\end{itemize}
-\end{frame}
-
 \end{document}

### 2013-06-24 doc/slides/gsoc_proposal/gsoc_proposal.tex
Added an example of gradual typing to the presentation about the GSoC proposal

@@
 \institute[PUC-Rio]{LabLua\\PUC-Rio}
-\date{}
+\date{June 27th 2013}
 
@@
 
+\begin{frame}[fragile]
+\frametitle{Gradual Typing Example}
+\begin{verbatim}
+local function div (x, y)
+  if y == 0 then
+    return nil,"division by zero"
+  end
+  return x / y
+end
+
+local function div (x:Number, y:Number) -> Any
+  if y == 0 then
+    return nil,"division by zero"
+  end
+  return x / y
+end
+\end{verbatim}
+\end{frame}
+
 \subsection{The levels of Gradual Typing}
@@
 is written using only dynamic typing. This module is not
-type checked, but their code that use this module is.
+type checked, but the code that uses this module is.
 \item The programmer write modules that use only static typing.
@@
 \end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Gradual Typing and Dynamically Typed Languages}
+\begin{itemize}
 \item Python
 \begin{itemize}
-\item gradual (package)
-\item Runtime type checking semantics of gradual typing
+\item py-gradual
+\begin{itemize}
+\item A package that provides runtime semantics of gradual typing
 \item Works only on Python 3
+\end{itemize}
+\item reticulated
+\begin{itemize}
+\item A tool that implements gradual typing
+\item Works on Python 2.7, 3.2, and 3.3
+\end{itemize}
+\end{itemize}
+\item Jython
+\begin{itemize}
+\item jython-gradual
+\begin{itemize}
+\item A fork that includes gradual typing
+\end{itemize}
 \end{itemize}
@@
 \item Some Lua projects implement parts of what is being proposed
-\item None implement type checking along with OO programming
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{LuaInspect}
-\begin{itemize}
-\item A tool that performs code analysis
-\item For instance, flagging undeclared identifiers
-\item It does not try to analyze OO code
-\item It does not perform type checking
+\item None implement compile-time type checking along with OO programming
 \end{itemize}
@@
 \item Both support some sort of OO programming
-\item Both do not perform type checking
+\item Both do not perform compile-time type checking
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{LuaInspect}
+\begin{itemize}
+\item A tool that uses MetaLua to perform code analysis
+\begin{itemize}
+\item Flags unknown global variables and table fields
+\item Checks number of function arguments against signatures
+\item Infers function return values
+\end{itemize}
+\item It does not try to analyze OO code
+\item It does not perform compile-time type checking
 \end{itemize}
@@
 
-\subsection{Some Statistics}
-\begin{frame}
-\frametitle{Some Statistics}
+\section{Statistics}
+\begin{frame}[fragile]
+\frametitle{Statistics}
 \begin{itemize}
-\item Data from LuaRocks repository as of June 10th
-\item LuaRocks repository contains 2950 .lua files
-\item \textbf{luac} (version 5.2.2) compiles 2864 of the 2950 files
-\item We have 2069 files when we ignore files that are generated by
-MoonScript and documentation and test subdirectories
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Function declaration}
+\item Data from LuaRocks repository as of June 22nd
+\item LuaRocks repository contains 3000 \verb'.lua' files
+\item \textbf{luac} (version 5.2.2) compiles 2913 of the 3000 files
+\item The statistics include 1966 of the 2913 files
 \begin{itemize}
-\item 18864 functions are being declared
-\begin{itemize}
-\item 4127 (21.9\%) are actually methods
-\item 1586 (8.4\%) use \textbf{type} to inspect the tags of the arguments
-\item 771 (4.1\%) return \textbf{nil} plus something else
-\item 119 (0.6\%) return \textbf{false} plus something else
+\item I did not include files that are generated by MoonScript
+\item I did not include documentation subdirectories
+\item I did not include test subdirectories
+\item I did not include test files in general
 \end{itemize}
@@
 
+\subsection{Function declaration}
 \begin{frame}
-\frametitle{Metatables}
+\frametitle{Function declaration}
 \begin{itemize}
-\item 18864 functions are being declared
+\item 17390 functions are being declared in the 1966 files
 \begin{itemize}
-\item 549 (2.9\%) use \textbf{setmetatable}
-\item 175 (0.9\%) use \textbf{getmetatable}
+\item 3961 (22.8\%) are actually methods
+\item 1512 (8.7\%) use \textbf{type} to inspect the tags of the arguments
+\item 773 (4.5\%) return \textbf{nil} plus something else
+\item 118 (0.7\%) return \textbf{false} plus something else
 \end{itemize}
@@
 
+\subsection{Metatables}
+\begin{frame}
+\frametitle{Metatables}
+\begin{itemize}
+\item 294 (15.0\%) of the 1966 files use \textbf{setmetatable}
+\item 99 (5.1\%) of the 196 files use \textbf{getmetatable}
+\end{itemize}
+\end{frame}
+
+\subsection{Table constructor}
 \begin{frame}
@@
 \begin{itemize}
-\item The table constructor is being used 19049 times
+\item The table constructor is being used 17369 times in the 1966 files
 \begin{itemize}
-\item 4571 (24.00\%) times is the empty construct
-\item 14475 (75.99\%) times is a constructor with only static keys
-\item 3 (0.01\%) times is a constructor with only dynamic keys
+\item 4391 (25.28\%) times is the empty construct
+\item 12975 (74.70\%) times is a constructor with only static keys
+\item 3 (0.02\%) times is a constructor with only dynamic keys
 \item There is no constructor with static and dynamic keys
@@
 
+\subsection{Module}
+\begin{frame}
+\frametitle{Module}
+\begin{itemize}
+\item 655 (33.3\%) of the 1966 files use \textbf{return} at the end of the file
+\item 602 (30.7\%) of the 1966 files use the function \textbf{module}
+\item The rest of the files are plain scripts
+\end{itemize}
+\end{frame}
+
 \end{document}

### 2013-06-27 doc/slides/gsoc_proposal/gsoc_proposal.tex
Minor improvements to the presentation about the GSoC proposal

@@
 
-\begin{frame}[fragile]
-\frametitle{Gradual Typing Example}
-\begin{verbatim}
-local function div (x, y)
-  if y == 0 then
-    return nil,"division by zero"
-  end
-  return x / y
-end
-
-local function div (x:Number, y:Number) -> Any
-  if y == 0 then
-    return nil,"division by zero"
-  end
-  return x / y
-end
-\end{verbatim}
-\end{frame}
-
-\subsection{The levels of Gradual Typing}
 \begin{frame}
-\frametitle{The levels of Gradual Typing}
+\frametitle{Gradual Typing}
 \begin{itemize}
@@
 \begin{enumerate}
-\item The programmer write modules that use only dynamic typing.
+\item The programmer writes modules that use only dynamic typing.
 \item The programmer creates an interface to type a module that
@@
 type checked, but the code that uses this module is.
-\item The programmer write modules that use only static typing.
+\item The programmer writes modules that use only static typing.
 \end{enumerate}
@@
 \begin{itemize}
-\item Should be simple enough to preserve existing Lua idioms
+\item Should be complex enough to preserve existing Lua idioms
 \end{itemize}
@@
 \item 294 (15.0\%) of the 1966 files use \textbf{setmetatable}
-\item 99 (5.1\%) of the 196 files use \textbf{getmetatable}
+\item 99 (5.1\%) of the 1966 files use \textbf{getmetatable}
 \end{itemize}
@@
 \begin{itemize}
-\item 4391 (25.28\%) times is the empty construct
-\item 12975 (74.70\%) times is a constructor with only static keys
-\item 3 (0.02\%) times is a constructor with only dynamic keys
+\item 4391 (25.28\%) times it is the empty construct
+\item 12975 (74.70\%) times it is a constructor with only static keys
+\item 3 (0.02\%) times it is a constructor with only dynamic keys
 \item There is no constructor with static and dynamic keys
@@
 
-\subsection{Module}
+\subsection{Modules}
 \begin{frame}
-\frametitle{Module}
+\frametitle{Modules}
 \begin{itemize}
-\item 655 (33.3\%) of the 1966 files use \textbf{return} at the end of the file
-\item 602 (30.7\%) of the 1966 files use the function \textbf{module}
-\item The rest of the files are plain scripts
+\item 616 (31.3\%) of the 1966 files use \textbf{return} at the end of the file
+\item 563 (28.7\%) of the 1966 files use the function \textbf{module}
+\item 39 (2.0\%) of the 1966 files use \textbf{return} and \textbf{module}
+\item The other 748 (38.0\%) files are plain scripts
+\item That is, 1218 (62.0\%) of the 1966 files are modules
 \end{itemize}

### 2013-09-10 typedlua/Type.hs
Starting to prototype the subtype relation with Haskell

@@ -1,0 +1,38 @@
+module Type where
+
+data Const = CNil
+           | CFalse
+           | CTrue
+           | CDouble
+           | CInteger
+           | CString
+           deriving (Eq,Show)
+
+data Base = BBoolean
+          | BNumber
+          | BString
+          deriving (Eq,Show)
+
+data Type = TConstant Const
+          | TBase Base
+          | TObject
+          | TAny
+          | TFunction SndClassType SndClassType
+          | TUnion Type Type
+          deriving (Show)
+
+data SndClassType = TFirstClass Type
+                  | TVarArg Type
+                  | TProd Type SndClassType
+                  deriving (Show)
+
+subtype :: Type -> Type -> Bool
+subtype t1 t2 =
+  subtype_top t1 t2
+
+subtype_top :: Type -> Type -> Bool
+subtype_top t1 t2 =
+  case t2 of TObject -> True
+             _ -> False
+
+

### 2013-09-10 typedlua/assert_Type.hs

@@ -1,0 +1,54 @@
+import Type
+import Control.Exception
+
+-- constant types for tests
+
+nil = TConstant CNil
+false = TConstant CFalse
+true = TConstant CTrue
+double = TConstant CDouble
+integer = TConstant CInteger
+str = TConstant CString
+
+-- base type for tests
+
+boolean = TBase BBoolean
+number = TBase BNumber
+string = TBase BString
+
+-- other types for tests
+
+object = TObject
+anyt = TAny
+
+f1 = TFunction (TFirstClass number) (TFirstClass number)
+
+u1 = TUnion number nil
+
+main :: IO ()
+main = do
+  print $ assert (subtype object object == True) "Object <: Object";
+  print $ assert (subtype nil object == True) "nil <: Object";
+  print $ assert (subtype false object == True) "false <: Object";
+  print $ assert (subtype true object == True) "true <: Object";
+  print $ assert (subtype double object == True) "1.1 <: Object";
+  print $ assert (subtype integer object == True) "1 <: Object";
+  print $ assert (subtype str object == True) "'w' <: Object";
+  print $ assert (subtype boolean object == True) "Boolean <: Object";
+  print $ assert (subtype number object == True) "Number <: Object";
+  print $ assert (subtype string object == True) "String <: Object";
+  print $ assert (subtype anyt object == True) "Any <: Object";
+  print $ assert (subtype f1 object == True) "(Number) -> Number <: Object";
+  print $ assert (subtype u1 object == True) "Number | nil <: Object";
+  print $ assert (subtype object nil == False) "Object <:/ nil";
+  print $ assert (subtype object false == False) "Object <:/ false";
+  print $ assert (subtype object true == False) "Object <:/ true";
+  print $ assert (subtype object double == False) "Object <:/ 1.1";
+  print $ assert (subtype object integer == False) "Object <:/ 1";
+  print $ assert (subtype object str == False) "Object <:/ 'w'";
+  print $ assert (subtype object boolean == False) "Object <:/ boolean";
+  print $ assert (subtype object number == False) "Object <:/ number";
+  print $ assert (subtype object string == False) "Object <:/ string";
+  print $ assert (subtype object anyt == False) "Object <:/ Any";
+  print $ assert (subtype object f1 == False) "Object <:/ (Number) -> Number";
+  print $ assert (subtype object u1 == False) "Object <:/ Number | nil";

### 2013-09-10 typedlua/Type.hs
Adding rules for constant, base, any, and union types in the prototype

@@ -28,11 +28,73 @@ data SndClassType = TFirstClass Type
 
 subtype :: Type -> Type -> Bool
 subtype t1 t2 =
-  subtype_top t1 t2
+  subtype_top t1 t2 ||
+  subtype_any t1 t2 ||
+  subtype_constants t1 t2 ||
+  subtype_base t1 t2 ||
+  subtype_union t1 t2
 
 subtype_top :: Type -> Type -> Bool
 subtype_top t1 t2 =
   case t2 of TObject -> True
              _ -> False
 
+subtype_any :: Type -> Type -> Bool
+subtype_any t1 t2 =
+  case t1 of
+    TAny -> case t2 of
+              TAny -> True
+              _ -> False
+    _ -> False
+
+subtype_constants :: Type -> Type -> Bool
+subtype_constants t1 t2 =
+  case t1 of
+    TConstant CNil -> case t2 of
+                        TConstant CNil -> True
+                        _ -> False
+    TConstant CFalse -> case t2 of
+                          TConstant CFalse -> True
+                          TBase BBoolean -> True
+                          _ -> False
+    TConstant CTrue -> case t2 of
+                         TConstant CTrue -> True
+                         TBase BBoolean -> True
+                         _ -> False
+    TConstant CDouble -> case t2 of
+                           TConstant CDouble -> True
+                           TBase BNumber -> True
+                           _ -> False
+    TConstant CInteger -> case t2 of
+                            TConstant CInteger -> True
+                            TConstant CDouble -> True
+                            TBase BNumber -> True
+                            _ -> False
+    TConstant CString -> case t2 of
+                           TConstant CString -> True
+                           TBase BString -> True
+                           _ -> False
+    _ -> False
+
+subtype_base :: Type -> Type -> Bool
+subtype_base t1 t2 =
+  case t1 of
+    TBase BBoolean -> case t2 of
+                        TBase BBoolean -> True
+                        _ -> False
+    TBase BNumber -> case t2 of
+                       TBase BNumber -> True
+                       _ -> False
+    TBase BString -> case t2 of
+                       TBase BString -> True
+                       _ -> False
+    _ -> False
+
+subtype_union :: Type -> Type -> Bool
+subtype_union t1 t2 =
+  case t1 of
+    TUnion t1' t1'' -> (subtype t1' t2) && (subtype t1'' t2)
+    t1 -> case t2 of
+            TUnion t2' t2'' -> (subtype t1 t2') || (subtype t1 t2'')
+            _ -> False
 

### 2013-09-10 typedlua/Type.hs
Adding function and snd class type rules in the prototype

@@ -32,7 +32,8 @@ subtype t1 t2 =
   subtype_any t1 t2 ||
   subtype_constants t1 t2 ||
   subtype_base t1 t2 ||
-  subtype_union t1 t2
+  subtype_union t1 t2 ||
+  subtype_function t1 t2
 
 subtype_top :: Type -> Type -> Bool
 subtype_top t1 t2 =
@@ -98,3 +99,28 @@ subtype_union t1 t2 =
             TUnion t2' t2'' -> (subtype t1 t2') || (subtype t1 t2'')
             _ -> False
 
+subtype_function :: Type -> Type -> Bool
+subtype_function t1 t2 =
+  case t1 of
+    TFunction s1 s2 -> case t2 of
+                         TFunction s3 s4 -> (subtype2 s3 s1) && (subtype2 s2 s4)
+                         _ -> False
+    _ -> False
+
+subtype2 :: SndClassType -> SndClassType -> Bool
+subtype2 a b =
+  case a of
+    TFirstClass t1 -> case b of
+                        TFirstClass t2 -> subtype t1 t2
+                        TVarArg t2 -> subtype t1 t2
+                        _ -> False
+    TVarArg t1 -> case b of
+                    TVarArg t2 -> subtype t1 t2
+                    TFirstClass t2 -> subtype t1 t2
+                    TProd t2 s2 -> (subtype t1 t2) && (subtype2 a s2)
+                    _ -> False
+    TProd t1 s1 -> case b of
+                     TProd t2 s2 -> (subtype t1 t2) && (subtype2 s1 s2)
+                     TVarArg t2 -> (subtype t1 t2) && (subtype2 s1 b)
+                     _ -> False
+    _ -> False

### 2013-09-10 typedlua/Type.hs
Removing Haskell prototype

@@ -1,126 +1,0 @@
-module Type where
-
-data Const = CNil
-           | CFalse
-           | CTrue
-           | CDouble
-           | CInteger
-           | CString
-           deriving (Eq,Show)
-
-data Base = BBoolean
-          | BNumber
-          | BString
-          deriving (Eq,Show)
-
-data Type = TConstant Const
-          | TBase Base
-          | TObject
-          | TAny
-          | TFunction SndClassType SndClassType
-          | TUnion Type Type
-          deriving (Show)
-
-data SndClassType = TFirstClass Type
-                  | TVarArg Type
-                  | TProd Type SndClassType
-                  deriving (Show)
-
-subtype :: Type -> Type -> Bool
-subtype t1 t2 =
-  subtype_top t1 t2 ||
-  subtype_any t1 t2 ||
-  subtype_constants t1 t2 ||
-  subtype_base t1 t2 ||
-  subtype_union t1 t2 ||
-  subtype_function t1 t2
-
-subtype_top :: Type -> Type -> Bool
-subtype_top t1 t2 =
-  case t2 of TObject -> True
-             _ -> False
-
-subtype_any :: Type -> Type -> Bool
-subtype_any t1 t2 =
-  case t1 of
-    TAny -> case t2 of
-              TAny -> True
-              _ -> False
-    _ -> False
-
-subtype_constants :: Type -> Type -> Bool
-subtype_constants t1 t2 =
-  case t1 of
-    TConstant CNil -> case t2 of
-                        TConstant CNil -> True
-                        _ -> False
-    TConstant CFalse -> case t2 of
-                          TConstant CFalse -> True
-                          TBase BBoolean -> True
-                          _ -> False
-    TConstant CTrue -> case t2 of
-                         TConstant CTrue -> True
-                         TBase BBoolean -> True
-                         _ -> False
-    TConstant CDouble -> case t2 of
-                           TConstant CDouble -> True
-                           TBase BNumber -> True
-                           _ -> False
-    TConstant CInteger -> case t2 of
-                            TConstant CInteger -> True
-                            TConstant CDouble -> True
-                            TBase BNumber -> True
-                            _ -> False
-    TConstant CString -> case t2 of
-                           TConstant CString -> True
-                           TBase BString -> True
-                           _ -> False
[…]

### 2013-09-10 typedlua/assert_Type.hs

@@ -1,54 +1,0 @@
-import Type
-import Control.Exception
-
--- constant types for tests
-
-nil = TConstant CNil
-false = TConstant CFalse
-true = TConstant CTrue
-double = TConstant CDouble
-integer = TConstant CInteger
-str = TConstant CString
-
--- base type for tests
-
-boolean = TBase BBoolean
-number = TBase BNumber
-string = TBase BString
-
--- other types for tests
-
-object = TObject
-anyt = TAny
-
-f1 = TFunction (TFirstClass number) (TFirstClass number)
-
-u1 = TUnion number nil
-
-main :: IO ()
-main = do
-  print $ assert (subtype object object == True) "Object <: Object";
-  print $ assert (subtype nil object == True) "nil <: Object";
-  print $ assert (subtype false object == True) "false <: Object";
-  print $ assert (subtype true object == True) "true <: Object";
-  print $ assert (subtype double object == True) "1.1 <: Object";
-  print $ assert (subtype integer object == True) "1 <: Object";
-  print $ assert (subtype str object == True) "'w' <: Object";
-  print $ assert (subtype boolean object == True) "Boolean <: Object";
-  print $ assert (subtype number object == True) "Number <: Object";
-  print $ assert (subtype string object == True) "String <: Object";
-  print $ assert (subtype anyt object == True) "Any <: Object";
-  print $ assert (subtype f1 object == True) "(Number) -> Number <: Object";
-  print $ assert (subtype u1 object == True) "Number | nil <: Object";
-  print $ assert (subtype object nil == False) "Object <:/ nil";
-  print $ assert (subtype object false == False) "Object <:/ false";
-  print $ assert (subtype object true == False) "Object <:/ true";
-  print $ assert (subtype object double == False) "Object <:/ 1.1";
-  print $ assert (subtype object integer == False) "Object <:/ 1";
-  print $ assert (subtype object str == False) "Object <:/ 'w'";
-  print $ assert (subtype object boolean == False) "Object <:/ boolean";
-  print $ assert (subtype object number == False) "Object <:/ number";
-  print $ assert (subtype object string == False) "Object <:/ string";
-  print $ assert (subtype object anyt == False) "Object <:/ Any";
-  print $ assert (subtype object f1 == False) "Object <:/ (Number) -> Number";
-  print $ assert (subtype object u1 == False) "Object <:/ Number | nil";

### 2014-02-02 doc/paper/typedlua.tex
Puting the text of the thesis proposal in a paper format

\documentclass[preprint]{sig-alternate}

\usepackage[utf8]{inputenc}
\usepackage{url}

\begin{document}

\conferenceinfo{Dyla}{'14, Edinburgh, UK}

\title{Typed Lua: An Optional Type System for Lua}

\numberofauthors{3}

\author{
\alignauthor
André Murbach Maidl\\
  \affaddr{PUC-Rio}\\
  \affaddr{Rio de Janeiro, Brazil}\\
  \email{amaidl@inf.puc-rio.br}
\alignauthor
Fabio Mascarenhas\\
  \affaddr{UFRJ}\\
  \affaddr{Rio de Janeiro, Brazil}\\
  \email{fabiom@dcc.ufrj.br}
\alignauthor
Roberto Ierusalimschy\\
  \affaddr{PUC-Rio}\\
  \affaddr{Rio de Janeiro, Brazil}\\
  \email{roberto@inf.puc-rio.br}
}

\date{June 12th 2014}

\maketitle

\begin{abstract}
\end{abstract}

\category{D.3.3}
         {Programming Languages}
         {Language Constructs and Features}
\category{D.2.3}
         {Programming Languages}
         {Coding Tools and Techniques}
         [Object-oriented programming]

\terms{Languages,Verification}

\keywords{Type Systems, Lua}

\section{Introduction} \label{sec:intro}

Dynamically typed languages such as Lua avoid static types in favor of
simplicity and flexibility, because the absence of static types means
that programmers do not need to bother about abstracting types that
should be validated by a type checker.
Instead, dynamically typed languages use run-time \textit{type tags}
to classify the values they compute, so their implementation can use
these tags to perform run-time (or dynamic) type checking
\cite{pierce2002tpl}.

This simplicity and flexibility allow programmers to write code that
might require a quite complex type system to statically type check,
though it may also hide bugs that will be caught only after deployment
if programmers do not properly test their code.
In contrast, static type checking helps programmers detect many 
bugs during the development phase.
Static types also provide a conceptual framework that helps
programmers define modules and interfaces that can be combined to
structure the development of programs.

Thus, early error detection and better program structure are two
advantages of static type checking that can lead programmers to
migrate their code from a dynamically typed to a statically
typed language, when their simple scripts become complex programs
\cite{tobin-hochstadt2006ims}.
Dynamically typed languages certainly help programmers during the
beginning of a project, because their simplicity and flexibility
allow quick development and make easier to change code according to
changing requirements.
However, programmers tend to migrate from dynamically typed to
statically typed code as soon as the project has consolidated its
requirements, because the robustness of static types helps
programmers link requirements to abstractions.
This migration usually involves different languages that have
different syntax and semantics, which usually require a complete
rewrite of existing programs instead of incremental evolution from
dynamic to static types.

Ideally, programming languages should offer programmers the
option to choose between static and dynamic typing:
\textit{optional type systems} \cite{bracha2004pluggable} and
\textit{gradual typing} \cite{siek2006gradual} are two similar
approaches for blending static and dynamic typing in the same
language.
The aim of both approaches is to offer programmers the option
to use type annotations where static typing is needed and avoid them
where dynamic typing is sufficient.
The difference between these two approaches is the way they treat
run-time semantics.
While optional type systems do not affect the run-time semantics,
gradual typing uses run-time checks to ensure that dynamically typed
code does not violate the invariants of statically typed code.

Programmers and researchers sometimes use the term gradual typing for
meaning incremental evolution of dynamically typed code to statically
typed code.
For this reason, gradual typing may also refer to optional type
systems and other approaches that blend static and dynamic typing to
help programmers incrementally migrate from dynamic to static typing
without having to switch to a different language, though all these
approaches differ in the way they handle static and dynamic typing
together.
We use the term gradual typing for referring to the work of Siek and
associates \cite{siek2006gradual}.

In this work we propose the design and implementation of Typed Lua:
an optional type system for Lua that should be complex enough to
preserve some of the Lua idioms that programmers are already used to,
but that should also include new constructs that help programmers
structure Lua programs.

Lua is a small scripting language, so designing an optional type
system for Lua should shed some light on optional type systems for
scripting languages in general.
Lua provides mechanisms for organizing a program in modules and
writing object-oriented programs, but it does not set policy on how
these features should behave, due to its use as an embedded and
extension language.
Thus, implementing an optional type system for Lua should offer Lua
programmers one way to obtain most of the advantages of static typing
without compromising the simplicity and flexibility of dynamic typing.

So far, Typed Lua should be a strict superset of Lua that
provides optional type annotations, compile-time type checking, and
class-based object oriented programming through the definition of
classes, interfaces, and modules.
In practice, we should implement Typed Lua as a programming
language that extends Lua syntax to add optional type annotations
and standard constructions for structuring Lua code.
The compiler should use static types to perform compile-time type
checking, but should also allow Lua code to coexist with Typed Lua
code, and generate Lua code that should run in unmodified Lua
implementations.
We have an ongoing implementation of the Typed Lua compiler that is
available online\footnote{https://github.com/andremm/typedlua}.

Typed Lua intended use is as an application language, and we view
that policies for organizing a program in modules and writing
object-oriented programs should be part of the language and
enforced by its optional type system.
An application language is a programming language that helps
programmers develop applications from scratch until these
applications evolve to complex systems rather than just scripts.

We also believe that Typed Lua will help programmers give a more
formal documentation to already existing Lua code, as static types
are also a useful source of documentation in languages that provide
type annotations, because type annotations are always validated by
the type checker and therefore never get outdated.
Thus, programmers can use Typed Lua to define axioms about the
interfaces and types of dynamically typed modules.
We will enforce this point by using Typed Lua to statically type
the interface of the Lua standard libraries and other commonly used
Lua libraries, so our compiler can check their usage by Typed Lua
code.

Typed Lua will perform a very limited form of local type inference
\cite{pierce2000lti}, as static typing does not necessarily mean
that programmers need to insert type annotations in the code.
Several statically typed languages such as Haskell provide some
amount of type inference that automatically deduces the types of
expressions.
Still, Typed Lua will need a small amount of type annotations due
to the nature of its optional type system.

Typed Lua will not deal with code optimization, although another
important advantage of static types is that they help the compiler
perform optimizations and generate more efficient code.
However, we believe that the formalization of our optional type
system will be precise enough to aid optimization in some Lua
implementations.

We will use some of the ideas of gradual typing to formalize Typed Lua.
Even though Typed Lua is an optional type system and thus will not
include run-time checks between dynamic and static regions of the
code, we believe that using the foundations of gradual typing to
formalize our optional type system will allow us to include run-time
checks in the future.

In Section \ref{sec:review} we review the literature about static and
dynamic typing in the same language, which includes the discussion
about the particularities between optional type systems and gradual
typing.
In Section \ref{sec:proposal} we describe the reasons why Typed Lua
includes features of object-oriented programming and the status of
its design and implementation.
In Section \ref{sec:con} we outline our expected contributions.

\section{Blending static and dynamic typing} \label{sec:review}

We begin this section presenting a little bit of the history about
combining static and dynamic typing in the same language.
After that, we introduce optional type systems and gradual typing.
We end this section discussing why optional type systems and two
other approaches are often called gradual typing.

\subsection{A little bit of history}

Common LISP \cite{steele1982ocl} introduced optional type annotations
in the early eighties, but not for static type checking.
Instead, programmers may choose to declare types of variables as
optimization hints to the compiler, that is, type declarations are
just one way to help the compiler to optimize code.
These annotations are unsafe because they can crash the program
when they are wrong.

\cite{abadi1989dts} extended the simply typed lambda-calculus with the
\verb'Dynamic' type and the \verb'dynamic' and \verb'typecase'
constructs, with the aim to safely integrate dynamic code in
statically typed languages.
The \verb'Dynamic' type is a pair \verb'(v,T)' where \verb'v' is a
value and \verb'T' is the tag that represents the type of \verb'v'.
The constructs \verb'dynamic' and \verb'typecase' are explicit
injection and projection operations, respectively.
That is, \verb'dynamic' builds values of type \verb'Dynamic' and
\verb'typecase' safely inspects the type of a \verb'Dynamic' value.
Thus, migrating code between dynamic and static type checking requires
changing type annotations and adding or removing \verb'dynamic' and
\verb'typecase' constructs throughout the code.

The \textit{quasi-static} type system proposed by \cite{thatte1990qst}
performs implicit coercions and run-time checks to replace the
\verb'dynamic' and \verb'typecase' constructs that were proposed by
\cite{abadi1989dts}.
To do that, quasi-static typing relies on subtyping with a top type
$\Omega$ that represents the dynamic type, and splits type checking
into two phases.
The first phase inserts implicit coercions from the dynamic type to
the expected type, while the second phase performs what Thatte calls
\textit{plausibility checking}, that is, it rewrites the program to
guarantee that sequences of up-casts and down-casts always have a
common subtype.

\textit{Soft typing} \cite{cartwright1991soft} is another approach
to combine static and dynamic typing in the same language.
The main goal of soft typing is to add static type checking to
dynamically typed languages without compromising their flexibility.
To do that, soft typing relies on type inference for
translating dynamically typed code to statically typed code.
The type checker inserts run-time checks around inconsistent code and
warns the programmer about the insertion of these run-time checks,
as they mean the existence of potential type errors.
However, the programmer is free to choose between inspecting the
run-time checks and simply running the code.
Soft typing gives this choice to the programmer because it is one of
its key ideas, that is, type inference and static type checking should
not prevent the programmer from running inconsistent code.
One advantage of soft typing is the fact that the compiler for
softly typed languages can use the translated code to generate
more efficient code, as the translated code statically type checks.
One disadvantage of soft typing is that it can be cumbersome when
the inferred types are meaningless large types that just confuse the
programmer.

\textit{Dynamic typing} \cite{henglein1994dts} is an approach
that optimizes code from dynamically typed languages by eliminating
unnecessary checks of tags.
This work describes how to translate dynamically typed code to
statically typed code that uses a \verb'Dynamic' type.
The translation is done through a coercion calculus that uses type
inference to insert the operations that are necessary to type check
the \verb'Dynamic' type during run-time.
Although soft typing and dynamic typing may seem similar, they are not.
They are not similar because soft typing targets statically type
checking of dynamically typed languages for detecting programming
errors while the dynamic typing targets the optimization of dynamically
typed code through the elimination of unnecessary run-time checks.
In other words, soft typing sees code optimization as a side effect
that comes with static type checking.

\cite{findler2002chf} proposed contracts for higher-order functions
and blame annotations for run-time checks.
Contracts perform dynamic checking instead of static checking,
but deferring all verifications to run-time can lead to defects
that are difficult to fix, because run-time errors can show a
stack trace where it is not clear to programmers if the cause of a
certain run-time error is in application code or library code.
Even if programmers identify that the source of a certain run-time
error is in library code, they still may have problems to identify
if this run-time error is due to a violation of library's contract,
or due to a bug, when the library is poorly documented.
In this approach, programmers can insert assertions in the form of
contracts that check the input and output of higher-order functions.
Thus, \cite{findler2002chf} show how programmers can insert
assertions in the form of contracts that check the input and output
of higher-order functions; and they also show how to add blame
annotations in the generated code as a way to track assertion
failures back to the source of the error.

BabyJ \cite{anderson2002babyj} is an object-oriented language
without inheritance that allows programmers to incrementally annotate
the code with more specific types.
Programmers can choose between using the dynamically typed version
of BabyJ, when they do not need types at all, and the statically
typed version of BabyJ, when they need to annotate the code.
In the statically typed BabyJ, programmers can use the
\textit{permissive type} $*$ to annotate the parts of the code that
still do not have a specific type or the parts of the code that should
have dynamic behavior.
The type system of BabyJ is nominal, so types are either class names
or the permissive type $*$.
However, the type system does not use type equality or subtyping,
but the relation $\approx$ between two types.
The relation $\approx$ holds when both types have the same name or
any of them is the permissive type $*$.
Even though the permissive type $*$ is similar to the dynamic type
from previous approaches, BabyJ does not provide any way to add
implicit or explicit run-time checks.

\cite{ou2004dtd} specified a language that combines static types
with dependent types.
To ensure safety, the compiler automatically inserts coercions
between dependent code and static code.
The coercions are run-time checks that ensure static code does not
crash dependent code during run-time.

\subsection{Optional Type Systems}

Optional type systems \cite{bracha2004pluggable} are an approach for
plugging static typing in dynamically typed languages through
optional type annotations that are used to perform compile-time type
checking, though not influencing the original run-time semantics of
the language.
This means that optional type systems can be unsound without affecting
language safety, because the run-time semantics should still catch
type errors independently of the static type checking.
More precisely, optional type systems are unsound if they do not catch
some type errors, but we say that languages with unsound optional type
systems are safe because the semantics of these languages catch during
run-time the type errors that the optional type system did not catch.
Although optional type systems maintain the simplicity and
flexibility of dynamically typed languages, their static type
checking cannot prove the absence of all type errors when the optional
type system is unsound.

Optional type systems have an interesting relation to the untyped
lambda calculus and the typed lambda calculus.
The semantic rules of the unytped lambda calculus are the same of
typed lambda calculus, and the type system is actually optional and
serves only for discarding programs that may have undesired behaviors
\cite{bracha2004pluggable}.

Optional type systems increase program evolution and
allow language designers to experiment with
\textit{pluggable type systems} \cite{bracha2004pluggable}.
Pluggable type systems are different optional type systems for
checking different aspects of the program.

According to \cite{bracha2004pluggable}, optional type systems should
not include mandatory type inference too, because mandatory type
inference may decrease the expressiveness of the type system.
Instead, optional type systems should implement type inference as
an independent mechanism that can be turned on and off according to
the programmers' needs.

Unfortunately, there is no formalization on optional type systems,
so each language ends up implementing its optional type system on
its own way.
Now we will briefly discuss some languages that provide an informal
description of their optional type systems rather than a formal one.

Strongtalk \cite{bracha1993strongtalk,bracha1996strongtalk} is
a version of Smalltalk that comes with an optional type system that
does not include any type inference.

Dart \cite{dart} and TypeScript \cite{typescript} are new
languages that are designed with an optional type system, and both
use JavaScript as their code generation target because their main
purpose is web development.
In fact, Dart is a new class-based object-oriented language with
optional type annotations while TypeScript is a strict superset of
JavaScript that provides optional type annotations and class-based
object-oriented programming.
Both Dart and TypeScript perform type inference to deduce static
types for unannotated code.

\subsection{Gradual Typing}

The main goal of gradual typing \cite{siek2006gradual} is to allow
programmers to choose between static and dynamic typing in the same
language.
To do that, \cite{siek2006gradual} extended the simply typed
lambda-calculus with the dynamic type $?$.
In gradual typing, type annotations are optional, so an untyped
variable is syntactic sugar for a variable whose declared type is
the dynamic type $?$.
Under these circumstances, we view gradual typing as a way to add
a dynamic type to statically typed languages.

The central idea of gradual typing is the \textit{consistency}
relation, written $\tau_{1} \sim \tau_{2}$, that allows implicit
conversions to and from the dynamic type, and disallows conversions
between inconsistent types \cite{siek2006gradual}.
For instance, $\mathbf{int} \sim \,?$ and $?\, \sim \mathbf{int}$,
but $\mathbf{int} \not\sim \mathbf{string}$ and
$\mathbf{string} \not\sim \mathbf{int}$. 
The consistency relation is neither commutative nor transitive.

Gradual typing is similar to the approaches proposed by
\cite{abadi1989dts} and \cite{thatte1990qst} by including a
dynamic type to a statically typed language.
However, these three approaches differ in the way they handle the
dynamic type.
While \cite{siek2006gradual} rely on the consistency relation,
\cite{abadi1989dts} rely on type equality with explicit projections
and injections, and \cite{thatte1990qst} relies on subtyping.
The subtyping relation actually is a pitfall on Thatte's quasi-static
typing, because the dynamic type is put as the top and the bottom of
the subtyping relation.
Therefore, down-casts combined with the transitivity of subtyping
accepts programs that should be rejected.

Later, \cite{siek2007objects} discovered that the consistency relation
is orthogonal to subtyping, so we can combine the two relations to
achieve both consistency and subtyping.
The consistent-subtyping relation,
written $\tau_{1} \lesssim \tau_{2}$,
is essential to design object oriented languages that have a
gradual type system.
Like the consistency relation, and unlike the subtyping relation,
the consistent-subtyping relation is not transitive.

Another important feature of gradual typing is the theoretic
foundation that it provides for inserting run-time checks that
prove dynamically typed code does not violate the invariants of
statically typed code, thus preserving type safety.
To do that, \cite{siek2006gradual} defined the run-time semantics
of gradual typing as a translation to an intermediate language with
explicit casts at the frontiers between statically and dynamically
typed code.
The semantics of these casts is based on the higher-order contracts
proposed by \cite{findler2002chf}.

However, \cite{herman2007sgt} showed that there is an efficiency
concern regarding the run-time checks, because there are two
ways that casts can lead to unbounded space consumption.
The first affects tail recursion while the second appears when
first-class functions or objects cross the border between
static code and dynamic code, that is, some programs can apply
repeated casts to the same function or object.
\cite{herman2007sgt} use the coercion calculus of
\cite{henglein1994dts} to solve the problem of space efficiency,
that is, they use coercions to express casts.
Their approach normalizes an arbitrary sequence of coercions to a
coercion of bounded size.

Another concern about casts is how to improve debugging support,
because a cast application can be delayed and the error related
to that cast application can appear far way from the real error.
\cite{wadler2009wpc} developed the \textit{blame calculus} as a way
to attack this issue, and \cite{ahmed2011bfa} extended the
blame calculus with polymorphism.
The blame calculus is an intermediate language to integrate
static and dynamic typing that uses the blame tracking approach
proposed by \cite{findler2002chf}.

On the one hand, the blame calculus solves the issue regarding
error reporting;
on the other hand, it has the space efficiency problem reported
by \cite{herman2007sgt}.
Thus, \cite{siek2009casts} extended the coercion calculus of
\cite{herman2007sgt} with blame tracking to achieve an
implementation of the blame calculus that is space efficient.
After that, \cite{siek2010blame} proposed a new solution that also
attacks both problems.
This new solution is based on a concept called \textit{threesome},
which is a way to split a cast between two parties into two casts
among three parties.
A cast has a source and a target type (a \textit{twosome}),
so we can split any cast into a downcast from the source to an
intermediate type that is followed by an upcast from the intermediate
type to the target type (a \textit{threesome}).

ActionScript \cite{moock2007as3} is the first language that
incorporated a gradual type system to its implementation and
Perl 6 \cite{tang2007pri} is also being designed with a
gradual type system, though there is little documentation about
the gradual type systems of these languages.
Gradual Python \cite{reticulated} is an attempt to develop a
gradual type system for Python that deals with mutable objects
\cite{siek2013mutable}.
Gradualtalk \cite{allende2013gts} is a gradually-typed Smalltalk
that introduces a new cast insertion strategy for gradually-typed
objects \cite{allende2013cis}.

\subsection{Approaches that are often called Gradual Typing}

Gradual typing is similar to optional type systems in that type
annotations are optional, and unannotated code is dynamically
typed, but unlike optional type systems, gradual typing affects
the run-time semantics to preserve type safety, and it is a way to
add a dynamic type to statically typed languages. 
More precisely, programming languages that include a gradual type
system implement the semantics of statically typed languages, so
the gradual type system inserts casts in the translated code to
guarantee that types are consistent before execution, while
programming languages that include an optional type system still
implement the semantics of dynamically typed languages, so all
the type checking also belongs to the semantics of each operation.

Still, we can view gradual typing as a way to formalize an optional
type system when the gradual type system does not insert run-time
checks.
BabyJ \cite{anderson2002babyj} is an example of object-oriented
language that uses the relation $\approx$ to implement its optional
type system, and this relation is similar to the consistency relation
of gradual typing.
Alore \cite{lehtosalo2011alore} is an example of object-oriented
language that uses subtyping along with the consistency relation to
implement its optional type system.
The authors of Alore defined a \textit{consistent-or-subtype}
relation that is different to the \textit{consistent-subtyping}
relation of \cite{siek2007objects}, but it is also written
$\tau_{1} \lesssim \tau_{2}$.
Their relation holds when $\tau_{1} \sim \tau_{2}$ or $\tau_{1} <: \tau_{2}$,
so $<:$ is transitive and $\sim$ is not.

Hence, optional type annotations for program evolution probably are
the reasons why optional type systems are commonly called
gradual type systems.
Typed Clojure \cite{bonnaire-sergeant2012typed-clojure} is an
optional type system for Clojure that now is adopting the
gradual typing slogan.

\cite{flanagan2006htc} introduced \textit{hybrid type checking},
an approach that combines static types and refinement types.
For instance, programmers can specify the refinement type
$\{x:Int \;|\; x \ge 0\}$ when they need a type for natural numbers.
The programmer can also choose between explicit or implicit casts.
When casts are not explicit, the type checker uses a theorem prover
to insert casts.
In our example of natural numbers, a cast would be inserted to check
whether an integer is greater than or equal to zero.

Sage \cite{gronski2006sage} is a programming language that
extended hybrid type checking with a dynamic type to
support dynamic and static typing in the same language.
Sage also offers optional type annotations in the gradual typing
style, that is, unannotated code is syntactic sugar for
code whose declared type is the dynamic type.

Thus, the inclusion of a dynamic type to hybrid type checking
along with optional type annotations, and the insertion of run-time
checks probably are the reasons why hybrid type checking is
also viewed as a form of gradual typing.

\cite{tobin-hochstadt2006ims} proposed another approach for gradually
migrating from dynamically typed to statically typed code,
and they coined the term \textit{from scripts to programs} for
referring to this kind of interlanguage migration.
In their approach, the migration from dynamically typed to
statically typed code happens module-by-module, so they designed
and implemented Typed Racket \cite{tobin-hochstadt2008ts} for
this purpose.
Typed Racket is a statically typed version of Racket
(a Scheme dialect) that allows the programmer write typed modules,
so Typed Racket modules can coexist with Racket modules,
which are untyped.

The approach used by \cite{tobin-hochstadt2008ts} to design and
implement Typed Racket probably is also called gradual typing
because it allows the programmer gradually migrate from untyped
scripts to typed programs.
However, note that Typed Racket is a statically typed language
and type annotations are mandatory.
So, what makes it gradual is a type system with a dynamic type
that handles the interaction between Racket and Typed Racket modules.

\section{Typed Lua} \label{sec:proposal}

We begin this section presenting statistics about the usage of Lua
features and idioms.
These statistics informed important design decisions on our optional
type system.
Then we introduce Typed Lua through examples, and describe a few Lua
projects that are similar to Typed Lua.

\subsection{Statistics about the usage of Lua}

We used the LuaRocks repository to build our statistics database;
LuaRocks \cite{hisham2013luarocks} is a package manager for Lua
modules.
We downloaded the 3928 \verb'.lua' files that were available in
the LuaRocks repository at February 1st 2014.
However, we ignored files that are not compatible with Lua 5.2,
the latest version of Lua.
We also ignored \textit{machine-generated} files and test files,
because these files may not represent idiomatic Lua code,
and might skew our statistics towards non-typical uses of Lua.
This left 2598 \verb'.lua' files from 262 different projects for
our statistics database;
we parsed these files and processed their abstract syntax tree
to gather the statistics that we show in this section.

The table constructor appears 23185 times.
In 36\% of the times it is a constructor that initializes a record;
in 29\% of the times it is a constructor that initializes a list;
in 8\% of the times it is a constructor that initializes a record
with a list part;
and in less than 1\% of the times (4 times) it is a constructor
that uses only the booleans \verb'true' and \verb'false' as indexes.
These numbers show us that in 73\% of the times it is a constructor
with literal keys only.
It also worth mentioning that 5\% of the initialization of lists
(or 2\% of the uses of the table constructor) use solely the vararg
expression.
Back to the statistics about the usage of the table constructor,
in 26\% of the times it is the empty constructor;
in 1\% of the times it is a constructor with non-literal keys
only, that is, a constructor that uses variables and function calls
to create the indexes of a table;
and in less than 1\% of the times (19 times) it is a constructor that
mixes literal keys and non-literal keys.

The indexing of tables appears 130448 times.
In 86\% of these times it is for reading a table field and
in 14\% of these times it is for writing into a table field.
While reading, 89\% of the reads use a string key,
4\% of the reads use a number key,
and less than 1\% of the reads (10 times) use a boolean key.
These numbers show us that 93\% of the reads use literals to index
a table field.
It also worth mentioning that 27\% of these reads that use literals
are function calls and 23\% of these reads that use literals are
method calls, that is, function calls that use the colon (\verb':')
syntactic sugar;
this means that 50\% of these reads that use literals are actually
function calls.
Still while reading, 7\% of the reads use non-literal
expressions to index a table field; 2\% of these reads that
use non-literals are function calls.
While writing, 69\% of the writes use a string key,
2\% of the writes use a number key,
and less than 1\% of the writes (1 time) uses a boolean key.
These numbers show us that 71\% of the writes use literals to index
a table field.
Still while writing, the other 29\% of the writes use non-literal
expressions to index a table field.

We also gathered data about how many files iterate over tables:
23\% of the files iterate over keys of any value, that is,
the function \verb'pairs' appears at least once in these files
(the median is twice per file);
21\% of the files iterate over integer keys, that is,
the function \verb'ipairs' appears at least once in these files
(the median is also twice per file);
and 10\% of the files use the numeric \verb'for' along with the length
operator (\verb'#') (the median is once per file).

Typed Lua will include a table type to represent Lua tables.
This table type should be able to represent lists, records,
and associative arrays.

We found a total of 24858 function declarations in our database
(the median is six per file).
We observed that 14\% of these declarations use the colon
(\verb':') syntactic sugar and 9\% of these declarations use
\verb'self' as their first parameter.
These numbers show us that 23\% of the functions are actually methods.

Regarding metatables, 63\% of the projects extend tables with
metatables, that is, they call \verb'setmetatable' at least once,
and 27\% of the projects access the metatable of a given table,
that is, they call \verb'getmetatable' at least once.

Typed Lua will include interfaces and classes as syntactic sugar
for table types that encode objects.
Proposing an optional type system for Lua that includes classes may
sound unusual at first glance, because Lua does not have standard
policies for object-oriented programming.
However, Lua provides mechanisms that allow programmers to abstract
their code in terms of objects, and our statistics confirm that
an expressive number of programmers are relying on these mechanisms to
use the object-oriented paradigm in Lua.
Typed Lua will include interfaces and classes as merely syntactic
mechanisms that help the compiler type check object-oriented code,
and they will not influence the semantics of Lua.

Regarding modules, 38\% of the files use the current way of defining
modules, that is, these files return a table that contains the
exported members of the module at the end of the file;
22\% of the files still use the deprecated way of defining modules,
that is, these files use the function \verb'module';
and 1\% of the files use both ways.
These numbers show us that 61\% of the files are modules while 39\%
of the files are plain scripts.
The number of plain scripts is quite high considering the origin of
our database.
However, we did not ignore sample scripts, which usually serve to
help the users of a given module on how to use this module, and
that is the reason why we have a quite high number of plain scripts.

Typed Lua will also use interfaces for defining the types of
userdata, threads, and modules.
Typed Lua will also support the deprecated style of module definition,
using global names as exported members of the module.

We observed that 9\% of the functions use \verb'type' to inspect the
tags of Lua values (the median is once per function).
We randomly selected 20 functions to sample how programmers are
using \verb'type', and we got the following data:
50\% of these functions use \verb'type' for asserting the tags of
their input parameters, that is, they raise an error when the tag of a
certain parameter does not match the expected tag;
40\% of these functions use \verb'type' for overloading, that is,
they execute different code according to the inspected tag;
and 10\% of these functions use \verb'type' inside a loop, that is,
they check the tags of the values of a given a table to decide
whether a computation should be executed using these values or not.

Typed Lua will include union types because the use of the \verb'type'
idiom shows that disjoint unions will help programmers define
data structures that can hold a value of different, but fixed types.
Typed Lua should also offer a mechanism for decomposing unions.
This mechanism should translate to this kind of idiom that uses
\verb'type', though it may be restricted to base types only.

We observed that 10\% of the functions explicitly return multiple
values, that is, these functions return two or more values.
We also observed that 50\% of these functions (or 5\% of the total)
return \verb'nil' plus something else to signal an unexpected
behavior while 10\% of these functions (or 1\% of the total) return
\verb'false' plus something else to signal an unexpected behavior.

Typed Lua will include function types to represent Lua functions
and tuple types to represent the signatures of Lua functions,
multiple return values, and multiple assignments.
Tuple types will require some special attention, because Typed Lua
should be able to adjust tuple types during compile-time, in a
similar way to what Lua does with function calls and multiple
assignments during run-time.

The last statistics that we collected is about variadic functions.
We observed that 8\% of the functions are variadic, that is,
their last parameter is the vararg expression (\verb'...').
Most of the times the vararg expression is used to initialize a
list or to call another function.

Typed Lua will include a \textit{vararg type} to handle variadic
functions and vararg expressions.

\subsection{Typed Lua through examples}

Our main goal is to allow programmers to combine static and dynamic
typing in Lua, so we propose Typed Lua as an extension to Lua 5.2
syntax that introduces optional type annotations, compile-time type
checking, and class-based object-oriented programming through the
definition of classes, interfaces, and modules.
Another important goal is to allow programmers to run any Lua code
without modifications, so that is the reason why these syntactic
extensions include new keywords that are not reserved words.

Typed Lua includes the following first-class types: \verb'any',
\verb'boolean', \verb'number', \verb'string', constant types,
union types, function types, table types, and recursive types.
The type \verb'any' is the dynamic type that allows dynamic code
coexist with static code.
Constant types mean that \verb'nil', \verb'false', \verb'true',
literal numbers, and literal strings are valid types.
Typed Lua also uses subtyping to define the relation among constant
types and the types \verb'boolean', \verb'number', and \verb'string'.
More precisely, Typed Lua uses subtyping to specify that
\verb'true' and \verb'false' are subtype of \verb'boolean',
any literal number is subtype of \verb'number', and any literal
string is subtype of \verb'string'.
We will discuss union types, function types, table types, and
recursive types along this section.

Typed Lua allows optional type annotations in variable and function
declarations.
The compiler assigns a more specific type to an unannotated
declaration whenever it is possible to use local type inference,
and it assigns the type \verb'any' to unannotated declarations
otherwise.
For instance, in function declarations, the compiler does not infer
a more specific type to unannotated input parameters.
Global type inference often is undecidable in a language with
subtyping and overloading \cite{wells1999typability}.

In the next example, we add type annotations in a function
declaration, but we avoid type annotations in a declaration of a local
variable:
\begin{verbatim}
    local function fact (n:number):number
      if n == 0 then
        return 1
      else
        return n * fact(n - 1)
      end
    end
    local x = 5
    print(fact(x))
\end{verbatim}

Even though we did not add type annotations in the declaration of
the local variable \verb'x', the compiler uses local type inference to
assign the type \verb'number' to \verb'x'.
The compiler does not use the constant type \verb'5' instead of the
most general type \verb'number' because this would prevent
programmers from assigning other numeric values to the variable
\verb'x'.

Typed Lua allows programmers to combine static and dynamic
typing in the same program, as we show in the following example:
\begin{verbatim}
    local function abs (n:number):number
      if n < 0 then
        return -n
      else
        return n
      end
    end
    local function dist (x, y)
      return abs(x - y)
    end
\end{verbatim}

The function \verb'dist' receives two parameters of type \verb'any'
and returns a \verb'number'.
The compiler assigns the dynamic type \verb'any' to the input
parameters of \verb'dist' because they do not have type annotations,
and the compiler does not use global type inference, as we mentioned
previously.
The compiler infers the return type of \verb'dist' because it is
not a recursive function, that is, Typed Lua does not infer the return
type of recursive functions.

We are using the consistent-subtyping relation of gradual
typing to check the interaction among the dynamic type and the other
types.
More precisely, we can call \verb'abs' inside \verb'dist' because
\verb'x - y' is of type \verb'any', and the type \verb'any' is
consistent with the type \verb'number'.

However, unlike gradual typing, we do not insert run-time checks that
inspect the interaction between dynamically typed code and statically
typed code during run-time.
In other words, Typed Lua does not guarantee that dynamically typed
code does not violate statically typed code.
Typed Lua does not insert run-time checks because it is an optional
type system, and optional type systems should not influence the
run-time semantics, as we saw in Section \ref{sec:review}.
In addition, run-time checks can decrease run-time performance
\cite{allende2013cis}.
In the previous example, we cannot guarantee that \verb'dist' is never
going to call \verb'abs' with a parameter that is not a number,
because in the semantics of Lua the minus operator can result a value
that is not a number.
Still, we can implement run-time checks in the future because we are
using the consistent-subtyping relation of gradual typing to
formalize our optional type system, and the type system is being
designed to be sound.

Typed Lua includes union types as a way to discriminate data
structures that can hold values of several different, but fixed types.
Next, we show an example:
\begin{verbatim}
    local x:number|nil         -- OK
    local y:number|string = 1  -- OK
    local z:number = x         -- Error
\end{verbatim}

The last line of this example gives a type error because we are
attempting to assign a value of type \verb'number' or \verb'nil'
to a variable that accepts only values of type \verb'number'.
Although we can solve this type error changing the annotation of
variable \verb'z' from \verb'number' to \verb'number|nil', there is
another way to solve this type error without changing the
annotation:
\begin{verbatim}
    local z:number = x or 0  -- OK
\end{verbatim}

Typed Lua has a typing rule for checking that $t$\verb'|nil or '$t$
always results the type $t$.

The syntactic construction $t$\verb'|nil' is so common that Typed Lua
includes the syntactic sugar $t$\verb'?' for it.
For instance, we can write \verb'local x:number?' instead of
\verb'local x:number|nil'.

Typed Lua includes a table type to represent Lua's tables.
Although Typed Lua has just one table type, it has three different
syntaxes for defining the type of a table.
The first syntax defines table types for hash tables;
it is written \verb'{'$t_{1}$\verb':'$t_{2}$\verb'}',
and maps to the table type \verb'{'$t_{1}$\verb':'$t_{2}$\verb'|nil}'.
This table type represents a hash table that maps values of type
$t_{1}$ to values of type $t_{2}$ or \verb'nil', as Lua returns
\verb'nil' when we use a non-existing key to index a table.
The second syntax defines table types for lists;
it is written \verb'{'$t$\verb'}',
and maps to the table type \verb'{number:'$t$\verb'|nil}'.
This table type is just a syntactic sugar to the definition of a
hash table that maps values of type \verb'number' to values of type
$t$ or \verb'nil'.
The third syntax defines table types for records;
it is written \verb'{'$s_{1}$\verb':'$t_{1}$\verb', ..., '$s_{n}$\verb':'$t_{n}$\verb'}',
and maps to the table type \verb'{'$s_{1}$\verb':'$t_{1}$\verb', ..., '$s_{n}$\verb':'$t_{n}$\verb'}'.
This table type represents a record that maps literals
$s_{i}$\verb', ..., '$s_{n}$ to values of types
$t_{i}$\verb', ..., '$t_{n}$, where $i$ may vary from one to $n$.
The commas separate the type of each record field when the declared
record has more than one field.

Next we show one example that types a hash table that maps values
of type \verb'string' to values of type \verb'number' or \verb'nil':
\begin{verbatim}
    local t:{string:number} = { foo = 1 }  -- OK
    local x:number = t.foo                 -- Error
    local y:number|nil = t.bar             -- OK
\end{verbatim}

The second line of this example gives a type error for the same
reason that the last line of the previous example gives a type error
too, that is, we are trying to assign a value that can be of two
different types to a variable that accepts values of only one type.
Although the field \verb'bar' does not exist in \verb't', the last
line of this example does not give a type error because the
annotated type matches the type of the values that can be stored in
\verb't'.

In the next example we show how we can use the table type to type a
list of strings:
\begin{verbatim}
    local days:{string} = { "Sunday", "Monday",
      "Tuesday", "Wednesday", "Thursday",
      "Friday", "Saturday" }  -- OK
\end{verbatim}

When we know that a list has fixed elements, we can leave the
variable declaration unannotated and let local type inference assign
a more specific table type to the list.
If we remove the annotation in this example, the compiler uses the
syntax of records to infer the following table type to \verb'days':
\begin{verbatim}
    { 1:string, 2:string, 3:string, 4:string,
      5:string, 6:string, 7:string }
\end{verbatim}

When we use the inferred table type while initializing a table, the
compiler can give an error message when we try to access an index that
is out of bounds.
We can also use this style for defining heterogeneous tuples.

Next we show one example of table type to type records:
\begin{verbatim}
    local person:{"firstname":string,
                  "lastname":string} =
      { firstname = "Lou", lastname = "Reed" } 
\end{verbatim}

We could have let type inference assign the annotated type to the
variable \verb'person', but we prefer using the annotated type to
exemplify the record syntax.

Typed Lua includes interfaces as syntactic sugar to the table type
that specifies records.
For instance, we can use the following interface as an alias to the
type of the record we showed in the previous example:
\begin{verbatim}
    interface Person
      firstname:string
      lastname:string
    end
\end{verbatim}

Now we can use the type \verb'Person' as an alias to a table type
that certainly has the fields \verb'firstname' and \verb'lastname'.
Next, we show an example of code that uses the type \verb'Person':
\begin{verbatim}
    local function greeter (person:Person):string
      return "Hello " .. person.firstname ..
             " " .. person.lastname
    end

    local user1 = { firstname = "Lewis",
                    middlename = "Allan",
                    lastname = "Reed" }
    local user2 = { firstname = "Lou" }
    local user3 = { lastname = "Reed",
                    firstname = "Lou" }
    local user4 = { "Lou", "Reed" }

    print(greeter(user1)) -- OK
    print(greeter(user2)) -- Error
    print(greeter(user3)) -- OK
    print(greeter(user4)) -- Error
\end{verbatim}

This example shows that our optional type system is structural rather
than nominal, that is, it checks the structure of types instead of
their names, and it uses subtyping and consistent-subtyping for
checking types.

Using interfaces we can also define recursive types such as linked
lists:

\begin{verbatim}
    interface Element
      info:number
      next:Element?
    end
\end{verbatim}

We need to explicitly annotate \verb'next' being \verb'Element'
or \verb'nil' because in our type system \verb'nil' is not the
bottom type, that is, \verb'nil' is subtype of itself only.
Typed Lua has this behavior because in Lua \verb'nil' represents
the absence of an useful value.

Besides interfaces, Typed Lua will also introduce classes.
We want to offer classes as a standard way for writing code that
Lua programmers are already using, but without tying them to a
specific model.
Our classes will not include static attributes and will also not offer
privacy rules.

Typed Lua will also introduce a specific syntax for defining modules.
For now, each file is a module.
We use all global names and their types to build the interface that
represents the type of the module, in case the module is not returned
at the end of the file.

Typed Lua also includes function types, as functions are first-class
values in Lua.
The syntax for defining a function type is \verb'('$s$\verb') -> ('$s$\verb')',
where $s$ is a second-class type.
A second-class type in Typed Lua can be a first-class type, a
tuple type, a variadic type, or the \verb'void' type.
We use tuple types to represent multiple parameters, multiple return
values, and multiple assignments.
We use the variadic type to represent variadic functions.
Even though we have specific types to represent multiple return values
and variadic functions, we are still working on these features.

The next example shows how we can annotate a \verb'map' function
that receives another function as its parameter and uses this
function to alter the elements of a list of numbers, which is
also a parameter of \verb'map':
\begin{verbatim}
    local function map (f:(number)->(number),
                        l:{number}):void
      for i=1, #l do
        l[i] = f(l[i]) or 0
      end
    end
\end{verbatim}

To conclude this section, we just started the design and
implementation of Typed Lua, and there is still several design
decisions to be taken on this subject.
For instance, our type system should handle the subtleties that
come with metatables.

\subsection{Similar Lua projects}

Metalua \cite{metalua} is a Lua compiler that supports compile-time
metaprogramming (CTMP).
CTMP is a kind of macro system that allows the programmers to interact
with the compiler \cite{fleutot2007contrasting}. 
Metalua extends Lua 5.1 syntax to include its macro system,
and allows programmers to define their own syntax.
Metalua can provide syntactical support for several object-oriented
styles, and can also provide syntax for turning simple type
annotations into run-time assertions, but without compile-time type
checking.

MoonScript \cite{moonscript} is a programming language that supports
class-based object-oriented programming, and compiles to idiomatic
Lua code.
However, it does not perform compile-time type checking.

LuaInspect \cite{luainspect} is a tool that uses MetaLua to perform
some code analysis, such as flagging unknown global variables and
table fields, checking the number of function arguments against
signatures, and inferring function return values, but it does not
try to analyze object-oriented code and does not perform compile-time
type checking.

Tidal Lock \cite{tidallock} is a prototype of another optional type
system for Lua, which is written in Metalua.
A remarkable feature of Tidal Lock is the way it handles tables,
because it allows the programmer to incrementally build the type of
a table.
That is, Tidal Lock allows the programmer to create an empty table to
build its type according to the keys that the programmer inserts in it,
so the programmer do not need to bother about declaring its type with
just one line of code.
We also plan to add this feature to Typed Lua.
Like Typed Lua, Tidal Lock also has local type inference.
Unlike Typed Lua, Tidal Lock is not typing the common object-oriented
idioms used by Lua programmers.

\section{Expected contributions} \label{sec:con}

Our main contribution is to document the process of designing and
implementing an optional type system for Lua, a major scripting
language.
Lua is a small scripting language that shares some features with other
scripting languages, so the thesis should help language designers
that wish to add optional and gradual type systems to other
scripting languages.

We also believe that offering a framework that Lua developers can
use to document, test, and better structure their code is also a
major contribution of our thesis.

Finally, our optional type system could also help some Lua
implementations to perform code optimization.
However, the fact that we are not currently providing guarantees on
the interaction between dynamic and static code is an obstacle for
performing safe optimizations.

\bibliographystyle{abbrv}
\bibliography{typedlua}

\end{document}

### 2014-02-06 doc/paper/typedlua.tex
Using the proposal text to update the paper text

@@
 \usepackage{url}
+
+\newcommand{\Any}{\mathbf{any}}
+\newcommand{\Nil}{\mathbf{nil}}
+\newcommand{\False}{\mathbf{false}}
+\newcommand{\True}{\mathbf{true}}
+\newcommand{\Boolean}{\mathbf{boolean}}
+\newcommand{\Number}{\mathbf{number}}
+\newcommand{\String}{\mathbf{string}}
+\newcommand{\Void}{\mathbf{void}}
+
+\def\dstart{\hbox to \hsize{\vrule depth 4pt\hrulefill\vrule depth 4pt}}
+\def\dend{\hbox to \hsize{\vrule height 4pt\hrulefill\vrule height 4pt}}
 
@@
 features and idioms.
-These statistics informed important design decisions on our optional
-type system.
-Then we introduce Typed Lua through examples, and describe a few Lua
-projects that are similar to Typed Lua.
+We collected statistics about how programmers use tables, functions,
+dynamic type checking, object-oriented programming, and modules.
+We shall see that these statistics informed important design decisions
+on our optional type system.
+We end this section introducing Typed Lua through examples, and
+describing a few Lua projects that are similar to Typed Lua.
 
@@
 
+To verify how programmers use tables, we measured how they
+initialize, index, and iterate tables.
+We present these statistics in the next three paragraphs to discuss
+their influence on our type system.
+
 The table constructor appears 23185 times.
-In 36\% of the times it is a constructor that initializes a record;
-in 29\% of the times it is a constructor that initializes a list;
-in 8\% of the times it is a constructor that initializes a record
-with a list part;
-and in less than 1\% of the times (4 times) it is a constructor
+In 36\% of the occurrences it is a constructor that initializes a
+record (e.g., \texttt{\{ x = 120, y = 121 \}});
+in 29\% of the occurrences it is a constructor that initializes a
+list (e.g., \texttt{\{ "one", "two", "three", "four" \}});
+in 8\% of the occurrences it is a constructor that initializes a
+record with a list part;
+and in less than 1\% of the occurrences (4 times) it is a constructor
 that uses only the booleans \verb'true' and \verb'false' as indexes.
-These numbers show us that in 73\% of the times it is a constructor
-with literal keys only.
-It also worth mentioning that 5\% of the initialization of lists
-(or 2\% of the uses of the table constructor) use solely the vararg
-expression.
-Back to the statistics about the usage of the table constructor,
-in 26\% of the times it is the empty constructor;
-in 1\% of the times it is a constructor with non-literal keys
+At all, in 73\% of the occurrences it is a constructor that uses
+only literal keys;
+in 26\% of the occurrences it is the empty constructor;
+in 1\% of the occurrences it is a constructor with non-literal keys
 only, that is, a constructor that uses variables and function calls
 to create the indexes of a table;
-and in less than 1\% of the times (19 times) it is a constructor that
-mixes literal keys and non-literal keys.
-
-The indexing of tables appears 130448 times.
-In 86\% of these times it is for reading a table field and
-in 14\% of these times it is for writing into a table field.
-While reading, 89\% of the reads use a string key,
-4\% of the reads use a number key,
-and less than 1\% of the reads (10 times) use a boolean key.
-These numbers show us that 93\% of the reads use literals to index
-a table field.
-It also worth mentioning that 27\% of these reads that use literals
-are function calls and 23\% of these reads that use literals are
-method calls, that is, function calls that use the colon (\verb':')
-syntactic sugar;
-this means that 50\% of these reads that use literals are actually
+and in less than 1\% of the occurrences (19 times) it is a constructor
+that mixes literal keys and non-literal keys.
+
+The indexing of tables appears 130448 times:
+86\% of them are for reading a table field while
+14\% of them are for writing into a table field.
+About the readings, 89\% of the reads use a literal string key,
+4\% of the reads use a literal number key,
+and less than 1\% of the reads (10 times) use a literal boolean key.
+At all, 93\% of the reads use literals to index a table while
+7\% of the reads use non-literal expressions to index a table.
+It also worth mentioning that 45\% of the reads are actually
 function calls.
-Still while reading, 7\% of the reads use non-literal
-expressions to index a table field; 2\% of these reads that
-use non-literals are function calls.
-While writing, 69\% of the writes use a string key,
-2\% of the writes use a number key,
-and less than 1\% of the writes (1 time) uses a boolean key.
-These numbers show us that 71\% of the writes use literals to index
-a table field.
-Still while writing, the other 29\% of the writes use non-literal
-expressions to index a table field.
-
-We also gathered data about how many files iterate over tables:
-23\% of the files iterate over keys of any value, that is,
-the function \verb'pairs' appears at least once in these files
-(the median is twice per file);
-21\% of the files iterate over integer keys, that is,
-the function \verb'ipairs' appears at least once in these files
+More precisely, 25\% of the reads use literals to call a function,
+20\% of the reads use literals to call a method, that is,
+a function call that uses the colon syntactic sugar, 
+and less than 1\% of the reads (195 times) use non-literal expressions
+to call a function.
+About the writings, 69\% of the writes use a literal string key,
+2\% of the writes use a literal number key,
+and less than 1\% of the writes (1 time) uses a literal boolean key.
+At all, 71\% of the writes use literals to index a table while
+29\% of the writes use non-literal expressions to index a table.
+
+We measured how many files have code that iterate over tables to
+observe how frequently iteration is used.
+We observed that 23\% of the files have code that iterate over keys
+of any value, that is, the call to \verb'pairs' appears at least once
+in these files (the median is twice per file);
+21\% of the files have code that iterate over integer keys, that is,
+the call to \verb'ipairs' appears at least once in these files
 (the median is also twice per file);
-and 10\% of the files use the numeric \verb'for' along with the length
-operator (\verb'#') (the median is once per file).
-
-Typed Lua will include a table type to represent Lua tables.
-This table type should be able to represent lists, records,
+and 10\% of the files have code that use the numeric \verb'for'
+along with the length operator (the median is once per file).
+
+The numbers about table initialization, indexing, and iteration
+show us that tables are mostly used to represent records, lists,
 and associative arrays.
+Therefore, Typed Lua should include a table type for handling
+these uses of Lua tables.
+Even though the statistics show that programmers initialize tables
+more often than they use the empty constructor to
+dynamically initialize tables, the statistics of the empty
+constructor are still expressive and indicate that Typed Lua should
+also include a way to handle this style of defining table types.
 
@@
 (the median is six per file).
-We observed that 14\% of these declarations use the colon
-(\verb':') syntactic sugar and 9\% of these declarations use
-\verb'self' as their first parameter.
-These numbers show us that 23\% of the functions are actually methods.
-
-Regarding metatables, 63\% of the projects extend tables with
+Next, we discuss how frequently programmers use dynamic type
+checking and multiple return values inside these functions.
+
+We observed that 9\% of the functions perform dynamic type checking
+on their input parameters, that is, these functions use \verb'type'
+to inspect the tags of Lua values (the median is once per function).
+We randomly selected 20 functions to sample how programmers are
+using \verb'type', and we got the following data:
+50\% of these functions use \verb'type' for asserting the tags of
+their input parameters, that is, they raise an error when the tag of a
+certain parameter does not match the expected tag, and
+50\% of these functions use \verb'type' for overloading, that is,
+they execute different code according to the inspected tag.
+
+These numbers show us that Typed Lua should include union types,
+because the use of the \verb'type' idiom shows that disjoint unions
+will help programmers define data structures that can hold a value of
+several different, but fixed types.
+Typed Lua should also use \verb'type' as a mechanism for decomposing
+unions, though it may be restricted to base types only.
+
+We observed that 10\% of the functions explicitly return multiple
+values.
+We also observed that 5\% of the functions return \verb'nil' plus
+something else, for signaling an unexpected behavior;
+and 1\% of the functions return \verb'false' plus something else,
+also for signaling an unexpected behavior.
+
+Typed Lua should include function types to represent Lua functions,
+and tuple types to represent the signatures of Lua functions,
+multiple return values, and multiple assignments.
+Tuple types will require some special attention, because Typed Lua
+should be able to adjust tuple types during compile-time, in a
+similar way to what Lua does with function calls and multiple
+assignments during run-time.
+In addition, the numbers show us that overloading on the return
+type will be useful to the type system.
+
+We also measured how frequently programmers use the object-oriented
+paradigm in Lua.
+We observed that 23\% of the function declarations are actually
+method declarations.
+More precisely, 14\% of them use the colon syntactic sugar while
+9\% of them use \verb'self' as their first parameter.
+We also observed that 63\% of the projects extend tables with
 metatables, that is, they call \verb'setmetatable' at least once,
@@
 that is, they call \verb'getmetatable' at least once.
-
-Typed Lua will include interfaces and classes as syntactic sugar
-for table types that encode objects.
+In fact, 45\% of the projects extend tables with metatables and
+declare methods:
+13\% using the colon syntactic sugar, 14\% using \verb'self', and
+18\% using both.
+
+Based on these observations, Typed Lua should include interfaces and
+classes as syntactic sugar for table types that encode objects.
 Proposing an optional type system for Lua that includes classes may
@@
 use the object-oriented paradigm in Lua.
-Typed Lua will include interfaces and classes as merely syntactic
+Typed Lua should include interfaces and classes as merely syntactic
 mechanisms that help the compiler type check object-oriented code,
-and they will not influence the semantics of Lua.
-
-Regarding modules, 38\% of the files use the current way of defining
+and they should not influence the semantics of Lua.
+
+We also measured how programmers are defining modules.
+We observed that 38\% of the files use the current way of defining
 modules, that is, these files return a table that contains the
@@
 22\% of the files still use the deprecated way of defining modules,
-that is, these files use the function \verb'module';
+that is, these files call the function \verb'module';
 and 1\% of the files use both ways.
-These numbers show us that 61\% of the files are modules while 39\%
-of the files are plain scripts.
+At all, 61\% of the files are modules while 39\% of the files are
+plain scripts.
 The number of plain scripts is quite high considering the origin of
@@
 
-Typed Lua will also use interfaces for defining the types of
-userdata, threads, and modules.
-Typed Lua will also support the deprecated style of module definition,
+Based on these observations, Typed Lua should also use interfaces as
+syntactic sugar for table types that define modules.
+Typed Lua should also support the deprecated style of module definition,
 using global names as exported members of the module.
 
-We observed that 9\% of the functions use \verb'type' to inspect the
-tags of Lua values (the median is once per function).
-We randomly selected 20 functions to sample how programmers are
-using \verb'type', and we got the following data:
-50\% of these functions use \verb'type' for asserting the tags of
-their input parameters, that is, they raise an error when the tag of a
-certain parameter does not match the expected tag;
-40\% of these functions use \verb'type' for overloading, that is,
-they execute different code according to the inspected tag;
-and 10\% of these functions use \verb'type' inside a loop, that is,
-they check the tags of the values of a given a table to decide
-whether a computation should be executed using these values or not.
-
-Typed Lua will include union types because the use of the \verb'type'
-idiom shows that disjoint unions will help programmers define
-data structures that can hold a value of different, but fixed types.
-Typed Lua should also offer a mechanism for decomposing unions.
-This mechanism should translate to this kind of idiom that uses
-\verb'type', though it may be restricted to base types only.
-
-We observed that 10\% of the functions explicitly return multiple
-values, that is, these functions return two or more values.
-We also observed that 50\% of these functions (or 5\% of the total)
-return \verb'nil' plus something else to signal an unexpected
-behavior while 10\% of these functions (or 1\% of the total) return
-\verb'false' plus something else to signal an unexpected behavior.
-
-Typed Lua will include function types to represent Lua functions
-and tuple types to represent the signatures of Lua functions,
-multiple return values, and multiple assignments.
-Tuple types will require some special attention, because Typed Lua
-should be able to adjust tuple types during compile-time, in a
-similar way to what Lua does with function calls and multiple
-assignments during run-time.
-
-The last statistics that we collected is about variadic functions.
+Besides using interfaces for encoding objects and defining modules,
+Typed Lua should also use interfaces for defining the types of
+userdata and threads.
+
+The last statistics that we collected were about variadic functions
+and vararg expressions.
 We observed that 8\% of the functions are variadic, that is,
-their last parameter is the vararg expression (\verb'...').
-Most of the times the vararg expression is used to initialize a
-list or to call another function.
-
-Typed Lua will include a \textit{vararg type} to handle variadic
+their last parameter is the vararg expression.
+We also observed that 5\% of the initialization of lists
+(or 2\% of the occurrences of the table constructor) use solely the
+vararg expression.
+Typed Lua should include a \textit{vararg type} to handle variadic
 functions and vararg expressions.
@@
 typing in Lua, so we propose Typed Lua as an extension to Lua 5.2
-syntax that introduces optional type annotations, compile-time type
-checking, and class-based object-oriented programming through the
-definition of classes, interfaces, and modules.
-Another important goal is to allow programmers to run any Lua code
-without modifications, so that is the reason why these syntactic
-extensions include new keywords that are not reserved words.
-
-Typed Lua includes the following first-class types: \verb'any',
-\verb'boolean', \verb'number', \verb'string', constant types,
+syntax that introduces optional type annotations, and class-based
+object-oriented programming through the definition of classes,
+interfaces, and modules.
+An important requirement of Typed Lua is that it should be backwards
+compatible with Lua, that is, it should allow programmers to run any
+Lua code without modifications.
+For this reason, these syntactic extensions include new keywords
+that are not reserved words.
+
+Figure \ref{fig:typelang} shows that Typed Lua includes includes
+first-class and second-class types.
+Typed Lua uses first-class types for representing Lua values and
+second-class types for defining the signatures of functions.
+First-class types include literal types, base types, the type $\Any$,
 union types, function types, table types, and recursive types.
-The type \verb'any' is the dynamic type that allows dynamic code
-coexist with static code.
-Constant types mean that \verb'nil', \verb'false', \verb'true',
-literal numbers, and literal strings are valid types.
-Typed Lua also uses subtyping to define the relation among constant
-types and the types \verb'boolean', \verb'number', and \verb'string'.
-More precisely, Typed Lua uses subtyping to specify that
-\verb'true' and \verb'false' are subtype of \verb'boolean',
-any literal number is subtype of \verb'number', and any literal
-string is subtype of \verb'string'.
-We will discuss union types, function types, table types, and
-recursive types along this section.
+Second-class types include the type $\Void$, first-class types,
+vararg types, and tuple types.
+Literal types are types that represent a single value
+(e.g., $\Nil$, $\False$, $\True$, literal numbers, and
+literal strings).
+Literal types are subtypes of their respective base types.
+More precisely, the literals $\False$ and $\True$ are subtypes of
+$\Boolean$, any literal number $n$ is subtype of $\Number$, and any
+literal string $s$ is subtype of $\String$.
+The type $\Any$ is the dynamic type that allows dynamic code coexist
+with static code.
+We will discuss union types, function types, table types, recursive
+types, and second-class types along this section.
+
+\begin{figure}[!ht]
+\textbf{Type Language}\\
+\dstart
+$$
+\begin{array}{rlr}
+T ::= & & \textit{first-class types}\\
+& \;\; L & \textit{literal types}\\
+& | \; B & \textit{base types}\\
+& | \; \Any & \textit{dynamic type}\\
+& | \; T \cup T & \textit{union type}\\
+& | \; S \rightarrow S & \textit{function type}\\
+& | \; \{T:T, ..., T:T\} & \textit{table type}\\
+& | \; X & \textit{type variable}\\
+& | \; \mu X.T & \textit{recursive type}\\
+L ::= & \;\; \Nil \; | \; \False \; | \; \True & \\
+& | \; {<}{\it numbe}r{>} \; | \; {<}{\it string}{>} & \\
+B ::= & \Boolean \; | \; \Number \; | \; \String & \\
+S ::= & \Void \; | \; V & \textit{second-class types}\\
+V ::= & \;\; T & \\
+& | \; T* & \textit{vararg type}\\
+& | \; T \times V & \textit{tuple type}
+\end{array}
+$$
+\dend
+\caption{Typed Lua types}
+\label{fig:typelang}
+\end{figure}
 
@@
 declarations.
-The compiler assigns a more specific type to an unannotated
-declaration whenever it is possible to use local type inference,
-and it assigns the type \verb'any' to unannotated declarations
-otherwise.
+Whenever it is possible to use local type inference, the compiler
+assigns a more specific type to an unannotated declaration.
+When it is not possible to use local type inference, the compiler
+assigns the type $\Any$ to an unannotated declaration.
 For instance, in function declarations, the compiler does not infer
-a more specific type to unannotated input parameters.
-Global type inference often is undecidable in a language with
-subtyping and overloading \cite{wells1999typability}.
-
-In the next example, we add type annotations in a function
-declaration, but we avoid type annotations in a declaration of a local
-variable:
+a more specific type to unannotated input parameters, because,
+in this case, global type inference often is undecidable in a
+language with subtyping and overloading \cite{wells1999typability}.
+
+In the following example we use type annotations in a function
+declaration, and we do not use type annotations in a declaration of a
+local variable:
 \begin{verbatim}
+
     local function fact (n:number):number
@@
 the local variable \verb'x', the compiler uses local type inference to
-assign the type \verb'number' to \verb'x'.
-The compiler does not use the constant type \verb'5' instead of the
-most general type \verb'number' because this would prevent
+assign the type $\Number$ to \verb'x'.
+The compiler does not use the literal type $5$ instead of the base
+type $\Number$ because this would prevent
 programmers from assigning other numeric values to the variable
 \verb'x'.
-
-Typed Lua allows programmers to combine static and dynamic
-typing in the same program, as we show in the following example:
+The inference that we implement in Typed Lua is quite simple, as it
+uses only the type of the local expression.
+
+Typed Lua allows programmers to combine annotated code with
+unannotated code, as we show in the following example:
 \begin{verbatim}
@@
     end
+
     local function dist (x, y)
@@
 
-The function \verb'dist' receives two parameters of type \verb'any'
-and returns a \verb'number'.
-The compiler assigns the dynamic type \verb'any' to the input
+The function \verb'dist' receives two parameters of type $\Any$
+and returns a value of type $\Number$.
+The compiler assigns the dynamic type $\Any$ to the input
 parameters of \verb'dist' because they do not have type annotations,
@@
 More precisely, we can call \verb'abs' inside \verb'dist' because
-\verb'x - y' is of type \verb'any', and the type \verb'any' is
-consistent with the type \verb'number'.
-
-However, unlike gradual typing, we do not insert run-time checks that
+\verb'x - y' is of type $\Any$, and the type $\Any$ is
+consistent with the type $\Number$.
+The type $\Any$ is not the union of all types, but it is a separate
+type.
+
+Unlike gradual typing, we do not insert run-time checks that
 inspect the interaction between dynamically typed code and statically
@@
 code does not violate statically typed code.
-Typed Lua does not insert run-time checks because it is an optional
-type system, and optional type systems should not influence the
-run-time semantics, as we saw in Section \ref{sec:review}.
-In addition, run-time checks can decrease run-time performance
-\cite{allende2013cis}.
+Typed Lua does not insert run-time checks because they can decrease
+run-time performance \cite{allende2013cis}.
+We believe that a careful evaluation of run-time checks should be done
+before inserting them on the type system.
 In the previous example, we cannot guarantee that \verb'dist' is never
 going to call \verb'abs' with a parameter that is not a number,
-because in the semantics of Lua the minus operator can result a value
-that is not a number.
+because in the semantics of Lua the minus operator can result in a
+value that is not a number.
 Still, we can implement run-time checks in the future because we are
@@
 \begin{verbatim}
-    local x:number|nil         -- OK
+    local x:number?            -- OK
     local y:number|string = 1  -- OK
@@
 
-The last line of this example gives a type error because we are
-attempting to assign a value of type \verb'number' or \verb'nil'
-to a variable that accepts only values of type \verb'number'.
+The annotation \verb'number?' is syntactic sugar for
+\verb'number|nil' that means the type $\Number\;\cup\;\Nil$.
+Typed Lua includes the syntactic sugar \verb't?' because the
+type $t\;\cup\;\Nil$ is very common in Lua code.
+
+The last line of the previous example gives a type error because we
+are attempting to assign a value of type $\Number\;\cup\;\Nil$
+to a variable that accepts only values of type $\Number$.
 Although we can solve this type error changing the annotation of
-variable \verb'z' from \verb'number' to \verb'number|nil', there is
-another way to solve this type error without changing the
-annotation:
+variable \verb'z' from \verb'number' to \verb'number?', there is
+another way that we can use to solve this type error without changing
+the annotation:
 \begin{verbatim}
@@
 
-Typed Lua has a typing rule for checking that $t$\verb'|nil or '$t$
-always results the type $t$.
-
-The syntactic construction $t$\verb'|nil' is so common that Typed Lua
-includes the syntactic sugar $t$\verb'?' for it.
-For instance, we can write \verb'local x:number?' instead of
-\verb'local x:number|nil'.
-
-Typed Lua includes a table type to represent Lua's tables.
-Although Typed Lua has just one table type, it has three different
-syntaxes for defining the type of a table.
+Typed Lua has a typing rule for checking that
+$t\;\cup\;\Nil\;\mathbf{or}\;t$ always has the type $t$.
+
+Typed Lua includes table types to represent Lua's tables.
+Although Typed Lua has just one table type constructor, it has three
+different syntaxes for defining the type of a table.
 The first syntax defines table types for hash tables;
-it is written \verb'{'$t_{1}$\verb':'$t_{2}$\verb'}',
-and maps to the table type \verb'{'$t_{1}$\verb':'$t_{2}$\verb'|nil}'.
+it is written \verb'{t1:t2}',
+and maps to the table type $\{t_{1}:t_{2}\;\cup\;\Nil\}$.
 This table type represents a hash table that maps values of type
-$t_{1}$ to values of type $t_{2}$ or \verb'nil', as Lua returns
-\verb'nil' when we use a non-existing key to index a table.
+$t_{1}$ to values of type $t_{2}\;\cup\;\Nil$, as Lua returns
+$\Nil$ when we use a non-existing key to index a table.
 The second syntax defines table types for lists;
-it is written \verb'{'$t$\verb'}',
-and maps to the table type \verb'{number:'$t$\verb'|nil}'.
+it is written \verb'{t}',
+and maps to the table type $\{\Number:t\;\cup\;\Nil\}$.
 This table type is just a syntactic sugar to the definition of a
-hash table that maps values of type \verb'number' to values of type
-$t$ or \verb'nil'.
+hash table that maps values of type $\Number$ to values of type
+$t\;\cup\;\Nil$.
 The third syntax defines table types for records;
-it is written \verb'{'$s_{1}$\verb':'$t_{1}$\verb', ..., '$s_{n}$\verb':'$t_{n}$\verb'}',
-and maps to the table type \verb'{'$s_{1}$\verb':'$t_{1}$\verb', ..., '$s_{n}$\verb':'$t_{n}$\verb'}'.
-This table type represents a record that maps literals
+it is written \verb'{s1:t1, ..., sn:tn}',
+and maps to the table type $\{s_{1}:t_{1}, ..., s_{n}:t_{n}\}$.
+This table type represents a record that maps the literals
 $s_{i}$\verb', ..., '$s_{n}$ to values of types
 $t_{i}$\verb', ..., '$t_{n}$, where $i$ may vary from one to $n$.
-The commas separate the type of each record field when the declared
-record has more than one field.
-
-Next we show one example that types a hash table that maps values
-of type \verb'string' to values of type \verb'number' or \verb'nil':
+
+Next, we show one example of table type to type a hash table:
 \begin{verbatim}
@@
     local x:number = t.foo                 -- Error
-    local y:number|nil = t.bar             -- OK
+    local y:number? = t.bar                -- OK
 \end{verbatim}
@@
 The second line of this example gives a type error for the same
-reason that the last line of the previous example gives a type error
+reason that the last line of the previous example gave a type error
 too, that is, we are trying to assign a value that can be of two
@@
 
-In the next example we show how we can use the table type to type a
-list of strings:
+Next, we show one example of table type to type a list:
 \begin{verbatim}
@@
 syntax of records to infer the following table type to \verb'days':
-\begin{verbatim}
-    { 1:string, 2:string, 3:string, 4:string,
-      5:string, 6:string, 7:string }
-\end{verbatim}
+\begin{align*}
+\{{1:\String},\;{2:\String},\;{3:\String},\;{4:\String},\;\\
+{5:\String},\;{6:\String},\;{7:\String}\}
+\end{align*}
+This inferred table type is a subtype of the type
+$\{\Number:\String\;\cup\;\Nil\}$, which is the type that we used to
+annotate the variable \verb'days'.
 
 When we use the inferred table type while initializing a table, the
-compiler can give an error message when we try to access an index that
+compiler gives an error message when we try to access an index that
 is out of bounds.
-We can also use this style for defining heterogeneous tuples.
-
-Next we show one example of table type to type records:
+
+It also worth mentioning that we can use the style of the inferred
+type to define heterogeneous tuples.
+
+Next, we show one example of table type to type a record:
 \begin{verbatim}
@@
 
-We could have let type inference assign the annotated type to the
-variable \verb'person', but we prefer using the annotated type to
-exemplify the record syntax.
+Type inference would infer the very same table type that we used to
+annotate the variable \verb'person'.
 
@@
 
-We need to explicitly annotate \verb'next' being \verb'Element'
-or \verb'nil' because in our type system \verb'nil' is not the
-bottom type, that is, \verb'nil' is subtype of itself only.
-Typed Lua has this behavior because in Lua \verb'nil' represents
+We need to explicitly annotate \verb'next' being of type
+${\tt Element}\;\cup\;\Nil$ because in our type system $\Nil$
+is not the bottom type, that is, $\Nil$ is subtype of itself only.
+Typed Lua has this behavior because in Lua $\Nil$ represents
 the absence of an useful value.
@@
 
-Typed Lua also includes function types, as functions are first-class
+Typed Lua includes function types because functions are first-class
 values in Lua.
-The syntax for defining a function type is \verb'('$s$\verb') -> ('$s$\verb')',
-where $s$ is a second-class type.
-A second-class type in Typed Lua can be a first-class type, a
-tuple type, a variadic type, or the \verb'void' type.
+The syntax for defining a function type is \verb'(s) -> (s)',
+where \verb's' is a second-class type.
+As we saw in Figure \ref{fig:typelang}, a second-class type in
+Typed Lua can be the type $\Void$, a first-class type, a variadic
+type, or a tuple type.
 We use tuple types to represent multiple parameters, multiple return
-values, and multiple assignments.
-We use the variadic type to represent variadic functions.
-Even though we have specific types to represent multiple return values
-and variadic functions, we are still working on these features.
+values, and multiple assignments, and we use the variadic type to
+represent variadic functions and the vararg expression.
+Even though we already included the tuple type and the vararg type,
+we are still working on these types.
 
@@
       for i=1, #l do
-        l[i] = f(l[i]) or 0
+        l[i] = f(l[i] or 0)
       end
@@
 a table.
-That is, Tidal Lock allows the programmer to create an empty table to
-build its type according to the keys that the programmer inserts in it,
-so the programmer do not need to bother about declaring its type with
-just one line of code.
+More precisely, Tidal Lock allows the programmer to create an empty
+table to build its type according to the keys that the programmer
+inserts in it, so the programmer do not need to bother about declaring
+its type with just one line of code.
 We also plan to add this feature to Typed Lua.

### 2014-02-26 doc/paper/typedlua.tex
Fixing paper bibliography

@@
 \usepackage[utf8]{inputenc}
+\usepackage[numbers]{natbib}
+\usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{url}
@@
 these tags to perform run-time (or dynamic) type checking
-\cite{pierce2002tpl}.
+\citep{pierce2002tpl}.
 
@@
 typed language, when their simple scripts become complex programs
-\cite{tobin-hochstadt2006ims}.
+\citep{tobin-hochstadt2006ims}.
 Dynamically typed languages certainly help programmers during the
 beginning of a project, because their simplicity and flexibility
-allow quick development and make easier to change code according to
+allow quick development and make it easier to change code according to
 changing requirements.
@@
 option to choose between static and dynamic typing:
-\textit{optional type systems} \cite{bracha2004pluggable} and
-\textit{gradual typing} \cite{siek2006gradual} are two similar
+\textit{optional type systems} \citep{bracha2004pluggable} and
+\textit{gradual typing} \citep{siek2006gradual} are two similar
 approaches for blending static and dynamic typing in the same
@@
 
-Programmers and researchers sometimes use the term gradual typing for
-meaning incremental evolution of dynamically typed code to statically
-typed code.
+Programmers and researchers sometimes use the term gradual typing
+to mean the incremental evolution of dynamically typed code to
+statically typed code.
 For this reason, gradual typing may also refer to optional type
@@
 together.
-We use the term gradual typing for referring to the work of Siek and
-associates \cite{siek2006gradual}.
+We use the term gradual typing for referring to the work of
+\citet{siek2006gradual}.
 
@@
 provides optional type annotations, compile-time type checking, and
-class-based object oriented programming through the definition of
+class-based object-oriented programming through the definition of
 classes, interfaces, and modules.
@@
 checking, but should also allow Lua code to coexist with Typed Lua
-code, and generate Lua code that should run in unmodified Lua
+code, and generate Lua code that runs in unmodified Lua
 implementations.
@@
 Typed Lua will perform a very limited form of local type inference
-\cite{pierce2000lti}, as static typing does not necessarily mean
+\citep{pierce2000lti}, as static typing does not necessarily mean
 that programmers need to insert type annotations in the code.
@@
 
-Common LISP \cite{steele1982ocl} introduced optional type annotations
+Common LISP \citep{steele1982ocl} introduced optional type annotations
 in the early eighties, but not for static type checking.
@@
 
-\cite{abadi1989dts} extended the simply typed lambda-calculus with the
-\verb'Dynamic' type and the \verb'dynamic' and \verb'typecase'
+\citet{abadi1989dts} extended the simply typed lambda calculus with the
+\texttt{Dynamic} type and the \texttt{dynamic} and \texttt{typecase}
 constructs, with the aim to safely integrate dynamic code in
 statically typed languages.
-The \verb'Dynamic' type is a pair \verb'(v,T)' where \verb'v' is a
-value and \verb'T' is the tag that represents the type of \verb'v'.
-The constructs \verb'dynamic' and \verb'typecase' are explicit
+The \texttt{Dynamic} type is a pair \texttt{(v,T)} where \texttt{v} is a
+value and \texttt{T} is the tag that represents the type of \texttt{v}.
+The constructs \texttt{dynamic} and \texttt{typecase} are explicit
 injection and projection operations, respectively.
-That is, \verb'dynamic' builds values of type \verb'Dynamic' and
-\verb'typecase' safely inspects the type of a \verb'Dynamic' value.
+That is, \texttt{dynamic} builds values of type \texttt{Dynamic} and
+\texttt{typecase} safely inspects the type of a \texttt{Dynamic} value.
 Thus, migrating code between dynamic and static type checking requires
-changing type annotations and adding or removing \verb'dynamic' and
-\verb'typecase' constructs throughout the code.
-
-The \textit{quasi-static} type system proposed by \cite{thatte1990qst}
+changing type annotations and adding or removing \texttt{dynamic} and
+\texttt{typecase} constructs throughout the code.
+
+The \textit{quasi-static} type system proposed by \citet{thatte1990qst}
 performs implicit coercions and run-time checks to replace the
-\verb'dynamic' and \verb'typecase' constructs that were proposed by
-\cite{abadi1989dts}.
+\texttt{dynamic} and \texttt{typecase} constructs that were proposed by
+\citet{abadi1989dts}.
 To do that, quasi-static typing relies on subtyping with a top type
@@
 \textit{plausibility checking}, that is, it rewrites the program to
-guarantee that sequences of up-casts and down-casts always have a
+guarantee that sequences of upcasts and downcasts always have a
 common subtype.
 
-\textit{Soft typing} \cite{cartwright1991soft} is another approach
+\textit{Soft typing} \citep{cartwright1991soft} is another approach
 to combine static and dynamic typing in the same language.
@@
 
-\textit{Dynamic typing} \cite{henglein1994dts} is an approach
+\textit{Dynamic typing} \citep{henglein1994dts} is an approach
 that optimizes code from dynamically typed languages by eliminating
 unnecessary checks of tags.
-This work describes how to translate dynamically typed code to
-statically typed code that uses a \verb'Dynamic' type.
+The paper describes how to translate dynamically typed code to
+statically typed code that uses a \texttt{Dynamic} type.
 The translation is done through a coercion calculus that uses type
 inference to insert the operations that are necessary to type check
-the \verb'Dynamic' type during run-time.
+the \texttt{Dynamic} type during run-time.
 Although soft typing and dynamic typing may seem similar, they are not.
@@
 checking of dynamically typed languages for detecting programming
-errors while the dynamic typing targets the optimization of dynamically
+errors while dynamic typing targets the optimization of dynamically
 typed code through the elimination of unnecessary run-time checks.
@@
 
-\cite{findler2002chf} proposed contracts for higher-order functions
+\citet{findler2002chf} proposed contracts for higher-order functions
 and blame annotations for run-time checks.
@@
 contracts that check the input and output of higher-order functions.
-Thus, \cite{findler2002chf} show how programmers can insert
+Thus, \citet{findler2002chf} show how programmers can insert
 assertions in the form of contracts that check the input and output
@@
 
-BabyJ \cite{anderson2002babyj} is an object-oriented language
+BabyJ \citep{anderson2002babyj} is an object-oriented language
 without inheritance that allows programmers to incrementally annotate
@@
 
-\cite{ou2004dtd} specified a language that combines static types
+\citet{ou2004dtd} specified a language that combines static types
 with dependent types.
@@
 
-Optional type systems \cite{bracha2004pluggable} are an approach for
+Optional type systems \citep{bracha2004pluggable} are an approach for
 plugging static typing in dynamically typed languages through
@@
 lambda calculus and the typed lambda calculus.
-The semantic rules of the unytped lambda calculus are the same of
+The semantic rules of the unytped lambda calculus are the same of the
 typed lambda calculus, and the type system is actually optional and
 serves only for discarding programs that may have undesired behaviors
-\cite{bracha2004pluggable}.
-
-Optional type systems increase program evolution and
+\citep{bracha2004pluggable}.
+
+Optional type systems make software evolution easier and
 allow language designers to experiment with
-\textit{pluggable type systems} \cite{bracha2004pluggable}.
+\textit{pluggable type systems} \citep{bracha2004pluggable}.
 Pluggable type systems are different optional type systems for
@@
 
-According to \cite{bracha2004pluggable}, optional type systems should
+According to \citet{bracha2004pluggable}, optional type systems should
 not include mandatory type inference too, because mandatory type
@@
 
-Strongtalk \cite{bracha1993strongtalk,bracha1996strongtalk} is
+Strongtalk \citep{bracha1993strongtalk,bracha1996strongtalk} is
 a version of Smalltalk that comes with an optional type system that
@@
 
-Dart \cite{dart} and TypeScript \cite{typescript} are new
+Dart \citep{dart} and TypeScript \citep{typescript} are new
 languages that are designed with an optional type system, and both
@@
 
-The main goal of gradual typing \cite{siek2006gradual} is to allow
+The main goal of gradual typing \citep{siek2006gradual} is to allow
 programmers to choose between static and dynamic typing in the same
 language.
-To do that, \cite{siek2006gradual} extended the simply typed
-lambda-calculus with the dynamic type $?$.
+To do that, \citet{siek2006gradual} extended the simply typed
+lambda calculus with the dynamic type $?$.
 In gradual typing, type annotations are optional, so an untyped
@@
 conversions to and from the dynamic type, and disallows conversions
-between inconsistent types \cite{siek2006gradual}.
+between inconsistent types \citep{siek2006gradual}.
 For instance, $\mathbf{int} \sim \,?$ and $?\, \sim \mathbf{int}$,
@@
 Gradual typing is similar to the approaches proposed by
-\cite{abadi1989dts} and \cite{thatte1990qst} by including a
+\citet{abadi1989dts} and \citet{thatte1990qst} by including a
 dynamic type to a statically typed language.
@@
 dynamic type.
-While \cite{siek2006gradual} rely on the consistency relation,
-\cite{abadi1989dts} rely on type equality with explicit projections
-and injections, and \cite{thatte1990qst} relies on subtyping.
+While \citet{siek2006gradual} rely on the consistency relation,
+\citet{abadi1989dts} rely on type equality with explicit projections
+and injections, and \citet{thatte1990qst} relies on subtyping.
 The subtyping relation actually is a pitfall on Thatte's quasi-static
@@
 the subtyping relation.
-Therefore, down-casts combined with the transitivity of subtyping
+Therefore, downcasts combined with the transitivity of subtyping
 accepts programs that should be rejected.
 
-Later, \cite{siek2007objects} discovered that the consistency relation
+Later, \citet{siek2007objects} discovered that the consistency relation
 is orthogonal to subtyping, so we can combine the two relations to
 achieve both consistency and subtyping.
-The consistent-subtyping relation,
+The \textit{consistent-subtyping} relation,
 written $\tau_{1} \lesssim \tau_{2}$,
-is essential to design object oriented languages that have a
+is essential to design object-oriented languages that have a
 gradual type system.
@@
 statically typed code, thus preserving type safety.
-To do that, \cite{siek2006gradual} defined the run-time semantics
+To do that, \citet{siek2006gradual} defined the run-time semantics
 of gradual typing as a translation to an intermediate language with
@@
 The semantics of these casts is based on the higher-order contracts
-proposed by \cite{findler2002chf}.
-
-However, \cite{herman2007sgt} showed that there is an efficiency
+proposed by \citet{findler2002chf}.
+
+However, \citet{herman2007sgt} showed that there is an efficiency
 concern regarding the run-time checks, because there are two
@@
 repeated casts to the same function or object.
-\cite{herman2007sgt} use the coercion calculus of
-\cite{henglein1994dts} to solve the problem of space efficiency,
+\citet{herman2007sgt} use the coercion calculus of
+\citet{henglein1994dts} to solve the problem of space efficiency,
 that is, they use coercions to express casts.
@@
 to that cast application can appear far way from the real error.
-\cite{wadler2009wpc} developed the \textit{blame calculus} as a way
-to attack this issue, and \cite{ahmed2011bfa} extended the
+\citet{wadler2009wpc} developed the \textit{blame calculus} as a way
+to attack this issue, and \citet{ahmed2011bfa} extended the
 blame calculus with polymorphism.
@@
 static and dynamic typing that uses the blame tracking approach
-proposed by \cite{findler2002chf}.
+proposed by \citet{findler2002chf}.
 
@@
 on the other hand, it has the space efficiency problem reported
-by \cite{herman2007sgt}.
-Thus, \cite{siek2009casts} extended the coercion calculus of
-\cite{herman2007sgt} with blame tracking to achieve an
+by \citet{herman2007sgt}.
+Thus, \citet{siek2009casts} extended the coercion calculus of
+\citet{herman2007sgt} with blame tracking to achieve an
 implementation of the blame calculus that is space efficient.
-After that, \cite{siek2010blame} proposed a new solution that also
+After that, \citet{siek2010blame} proposed a new solution that also
 attacks both problems.
@@
 
-ActionScript \cite{moock2007as3} is the first language that
+ActionScript \citep{moock2007as3} is the first language that
 incorporated a gradual type system to its implementation and
-Perl 6 \cite{tang2007pri} is also being designed with a
+Perl 6 \citep{tang2007pri} is also being designed with a
 gradual type system, though there is little documentation about
 the gradual type systems of these languages.
-Gradual Python \cite{reticulated} is an attempt to develop a
+Grace \citep{black2012grace,black2013sg} is a new object-oriented,
+gradually typed, educational language.
+In Grace, modules are gradually typed objects, that is, modules
+may have types and methods as attributes, and can have a state
+\citep{homer2013modules}.
+Gradual Python \citep{reticulated} is an attempt to develop a
 gradual type system for Python that deals with mutable objects
-\cite{siek2013mutable}.
-Gradualtalk \cite{allende2013gts} is a gradually-typed Smalltalk
+\citep{siek2013mutable}.
+Gradualtalk \citep{allende2013gts} is a gradually-typed Smalltalk
 that introduces a new cast insertion strategy for gradually-typed
-objects \cite{allende2013cis}.
+objects \citep{allende2013cis}.
 
@@
 annotations are optional, and unannotated code is dynamically
-typed, but unlike optional type systems, gradual typing affects
+typed, but unlike optional type systems, gradual typing changes 
 the run-time semantics to preserve type safety, and it is a way to
@@
 checks.
-BabyJ \cite{anderson2002babyj} is an example of object-oriented
-language that uses the relation $\approx$ to implement its optional
-type system, and this relation is similar to the consistency relation
-of gradual typing.
-Alore \cite{lehtosalo2011alore} is an example of object-oriented
-language that uses subtyping along with the consistency relation to
-implement its optional type system.
-The authors of Alore defined a \textit{consistent-or-subtype}
-relation that is different to the \textit{consistent-subtyping}
-relation of \cite{siek2007objects}, but it is also written
-$\tau_{1} \lesssim \tau_{2}$.
-Their relation holds when $\tau_{1} \sim \tau_{2}$ or $\tau_{1} <: \tau_{2}$,
-so $<:$ is transitive and $\sim$ is not.
-
-Hence, optional type annotations for program evolution probably are
+BabyJ \citep{anderson2002babyj} and Alore \citep{lehtosalo2011alore}
+are two examples of object-oriented languages that use gradual
+typing to formalize their optional type systems.
+The optional type systems of both BabyJ and Alore are nominal.
+BabyJ uses the relation $\approx$ that is similar to the consistency
+relation while Alore combines  subtyping along with the consistency
+relation to define a \textit{consistent-or-subtype} relation.
+The consistent-or-subtype relation is different from the
+consistent-subtyping relation of \citet{siek2007objects}, but it is
+also written $\tau_{1} \lesssim \tau_{2}$.
+The consistent-or-subtype relation holds when $\tau_{1} \sim \tau_{2}$
+or $\tau_{1} <: \tau_{2}$, so $<:$ is transitive and $\sim$ is not.
+Alore also extends its optional type system to include optional
+monitoring of run-time type errors in the gradual typing style.
+
+Hence, optional type annotations for software evolution probably are
 the reasons why optional type systems are commonly called
 gradual type systems.
-Typed Clojure \cite{bonnaire-sergeant2012typed-clojure} is an
+Typed Clojure \citep{bonnaire-sergeant2012typed-clojure} is an
 optional type system for Clojure that now is adopting the
@@
 
-\cite{flanagan2006htc} introduced \textit{hybrid type checking},
+\citet{flanagan2006htc} introduced \textit{hybrid type checking},
 an approach that combines static types and refinement types.
@@
 
-Sage \cite{gronski2006sage} is a programming language that
+Sage \citep{gronski2006sage} is a programming language that
 extended hybrid type checking with a dynamic type to
@@
 
-\cite{tobin-hochstadt2006ims} proposed another approach for gradually
+\citet{tobin-hochstadt2006ims} proposed another approach for gradually
 migrating from dynamically typed to statically typed code,
@@
 statically typed code happens module-by-module, so they designed
-and implemented Typed Racket \cite{tobin-hochstadt2008ts} for
+and implemented Typed Racket \citep{tobin-hochstadt2008ts} for
 this purpose.
 Typed Racket is a statically typed version of Racket
-(a Scheme dialect) that allows the programmer write typed modules,
+(a Scheme dialect) that allows the programmer to write typed modules,
 so Typed Racket modules can coexist with Racket modules,
@@
 
-The approach used by \cite{tobin-hochstadt2008ts} to design and
+The approach used by \citet{tobin-hochstadt2008ts} to design and
 implement Typed Racket probably is also called gradual typing
-because it allows the programmer gradually migrate from untyped
+because it allows the programmer to gradually migrate from untyped
 scripts to typed programs.
@@
 We used the LuaRocks repository to build our statistics database;
-LuaRocks \cite{hisham2013luarocks} is a package manager for Lua
+LuaRocks \citep{hisham2013luarocks} is a package manager for Lua
 modules.
-We downloaded the 3928 \verb'.lua' files that were available in
+We downloaded the 3928 \texttt{.lua} files that were available in
 the LuaRocks repository at February 1st 2014.
@@
 and might skew our statistics towards non-typical uses of Lua.
-This left 2598 \verb'.lua' files from 262 different projects for
+This left 2598 \texttt{.lua} files from 262 different projects for
 our statistics database;
@@
 and in less than 1\% of the occurrences (4 times) it is a constructor
-that uses only the booleans \verb'true' and \verb'false' as indexes.
+that uses only the booleans \texttt{true} and \texttt{false} as indexes.
 At all, in 73\% of the occurrences it is a constructor that uses
@@
 14\% of them are for writing into a table field.
-About the readings, 89\% of the reads use a literal string key,
+We can classify the indexing operations that are reads as follows:
+89\% of the reads use a literal string key,
 4\% of the reads use a literal number key,
@@
 to call a function.
-About the writings, 69\% of the writes use a literal string key,
+We can also classify the indexing operations that are writes as follows: 
+69\% of the writes use a literal string key,
 2\% of the writes use a literal number key,
@@
 
-We measured how many files have code that iterate over tables to
+We also measured how many files have code that iterate over tables to
 observe how frequently iteration is used.
 We observed that 23\% of the files have code that iterate over keys
-of any value, that is, the call to \verb'pairs' appears at least once
-in these files (the median is twice per file);
+of any value, that is, the call to \texttt{pairs} appears at least
+once in these files (the median is twice per file);
 21\% of the files have code that iterate over integer keys, that is,
-the call to \verb'ipairs' appears at least once in these files
+the call to \texttt{ipairs} appears at least once in these files
 (the median is also twice per file);
-and 10\% of the files have code that use the numeric \verb'for'
+and 10\% of the files have code that use the numeric \texttt{for}
 along with the length operator (the median is once per file).
@@
 We observed that 9\% of the functions perform dynamic type checking
-on their input parameters, that is, these functions use \verb'type'
+on their input parameters, that is, these functions use \texttt{type}
 to inspect the tags of Lua values (the median is once per function).
 We randomly selected 20 functions to sample how programmers are
-using \verb'type', and we got the following data:
-50\% of these functions use \verb'type' for asserting the tags of
+using \texttt{type}, and we got the following data:
+50\% of these functions use \texttt{type} for asserting the tags of
 their input parameters, that is, they raise an error when the tag of a
 certain parameter does not match the expected tag, and
-50\% of these functions use \verb'type' for overloading, that is,
+50\% of these functions use \texttt{type} for overloading, that is,
 they execute different code according to the inspected tag.
@@
 These numbers show us that Typed Lua should include union types,
-because the use of the \verb'type' idiom shows that disjoint unions
+because the use of the \texttt{type} idiom shows that disjoint unions
 will help programmers define data structures that can hold a value of
 several different, but fixed types.
-Typed Lua should also use \verb'type' as a mechanism for decomposing
+Typed Lua should also use \texttt{type} as a mechanism for decomposing
 unions, though it may be restricted to base types only.
@@
 values.
-We also observed that 5\% of the functions return \verb'nil' plus
+We also observed that 5\% of the functions return \texttt{nil} plus
 something else, for signaling an unexpected behavior;
-and 1\% of the functions return \verb'false' plus something else,
+and 1\% of the functions return \texttt{false} plus something else,
 also for signaling an unexpected behavior.
@@
 assignments during run-time.
-In addition, the numbers show us that overloading on the return
-type will be useful to the type system.
+In addition, the number of functions that return \texttt{nil} and
+\texttt{false} plus something else show us that overloading on the
+return type will be useful to the type system.
 
@@
 More precisely, 14\% of them use the colon syntactic sugar while
-9\% of them use \verb'self' as their first parameter.
+9\% of them use \texttt{self} as their first parameter.
 We also observed that 63\% of the projects extend tables with
-metatables, that is, they call \verb'setmetatable' at least once,
+metatables, that is, they call \texttt{setmetatable} at least once,
 and 27\% of the projects access the metatable of a given table,
-that is, they call \verb'getmetatable' at least once.
+that is, they call \texttt{getmetatable} at least once.
 In fact, 45\% of the projects extend tables with metatables and
 declare methods:
-13\% using the colon syntactic sugar, 14\% using \verb'self', and
+13\% using the colon syntactic sugar, 14\% using \texttt{self}, and
 18\% using both.
@@
 22\% of the files still use the deprecated way of defining modules,
-that is, these files call the function \verb'module';
+that is, these files call the function \texttt{module};
 and 1\% of the files use both ways.
@@
 plain scripts.
-The number of plain scripts is quite high considering the origin of
+The number of plain scripts is high considering the origin of
 our database.
@@
 help the users of a given module on how to use this module, and
-that is the reason why we have a quite high number of plain scripts.
+that is the reason why we have a high number of plain scripts.
 
@@
 syntactic sugar for table types that define modules.
-Typed Lua should also support the deprecated style of module definition,
-using global names as exported members of the module.
+Typed Lua should also support the deprecated style of module
+definition, using global names as exported members of the module.
 
@@
 Typed Lua should also use interfaces for defining the types of
-userdata and threads.
+userdata, as programmers often use userdata in an object-oriented
+style.
 
@@
 
-Figure \ref{fig:typelang} shows that Typed Lua includes includes
-first-class and second-class types.
+Figure \ref{fig:typelang} shows that Typed Lua includes first-class
+and second-class types.
 Typed Lua uses first-class types for representing Lua values and
@@
 & | \; \mu X.T & \textit{recursive type}\\
-L ::= & \;\; \Nil \; | \; \False \; | \; \True & \\
-& | \; {<}{\it numbe}r{>} \; | \; {<}{\it string}{>} & \\
+L ::= & \;\; \Nil \; | \; \False \; | \; \True \\
+& | \; {<}{\it number}{>} \; | \; {<}{\it string}{>} & \\
 B ::= & \Boolean \; | \; \Number \; | \; \String & \\
@@
 in this case, global type inference often is undecidable in a
-language with subtyping and overloading \cite{wells1999typability}.
+language with subtyping and overloading \citep{wells1999typability}.
 
@@
 Even though we did not add type annotations in the declaration of
-the local variable \verb'x', the compiler uses local type inference to
-assign the type $\Number$ to \verb'x'.
+the local variable \texttt{x}, the compiler uses local type inference to
+assign the type $\Number$ to \texttt{x}.
 The compiler does not use the literal type $5$ instead of the base
@@
 programmers from assigning other numeric values to the variable
-\verb'x'.
+\texttt{x}.
 The inference that we implement in Typed Lua is quite simple, as it
@@
 
-The function \verb'dist' receives two parameters of type $\Any$
+The function \texttt{dist} receives two parameters of type $\Any$
 and returns a value of type $\Number$.
 The compiler assigns the dynamic type $\Any$ to the input
-parameters of \verb'dist' because they do not have type annotations,
+parameters of \texttt{dist} because they do not have type annotations,
 and the compiler does not use global type inference, as we mentioned
 previously.
-The compiler infers the return type of \verb'dist' because it is
+The compiler infers the return type of \texttt{dist} because it is
 not a recursive function, that is, Typed Lua does not infer the return
@@
 types.
-More precisely, we can call \verb'abs' inside \verb'dist' because
-\verb'x - y' is of type $\Any$, and the type $\Any$ is
+More precisely, we can call \texttt{abs} inside \texttt{dist} because
+\texttt{x - y} is of type $\Any$, and the type $\Any$ is
 consistent with the type $\Number$.
@@
 Typed Lua does not insert run-time checks because they can decrease
-run-time performance \cite{allende2013cis}.
+run-time performance \citep{allende2013cis}.
 We believe that a careful evaluation of run-time checks should be done
 before inserting them on the type system.
-In the previous example, we cannot guarantee that \verb'dist' is never
-going to call \verb'abs' with a parameter that is not a number,
+In the previous example, we cannot guarantee that \texttt{dist} is never
+going to call \texttt{abs} with a parameter that is not a number,
 because in the semantics of Lua the minus operator can result in a
@@
 
-The annotation \verb'number?' is syntactic sugar for
-\verb'number|nil' that means the type $\Number\;\cup\;\Nil$.
-Typed Lua includes the syntactic sugar \verb't?' because the
+The annotation \texttt{number?} is syntactic sugar for
+\texttt{number|nil} that means the type $\Number\;\cup\;\Nil$.
+Typed Lua includes the syntactic sugar \texttt{t?} because the
 type $t\;\cup\;\Nil$ is very common in Lua code.
@@
 Although we can solve this type error changing the annotation of
-variable \verb'z' from \verb'number' to \verb'number?', there is
+variable \texttt{z} from \texttt{number} to \texttt{number?}, there is
 another way that we can use to solve this type error without changing
@@
 The first syntax defines table types for hash tables;
-it is written \verb'{t1:t2}',
+it is written \texttt{\{t1:t2\}},
 and maps to the table type $\{t_{1}:t_{2}\;\cup\;\Nil\}$.
@@
 The second syntax defines table types for lists;
-it is written \verb'{t}',
+it is written \texttt{\{t\}},
 and maps to the table type $\{\Number:t\;\cup\;\Nil\}$.
@@
 The third syntax defines table types for records;
-it is written \verb'{s1:t1, ..., sn:tn}',
+it is written \texttt{\{s1:t1, ..., sn:tn\}},
 and maps to the table type $\{s_{1}:t_{1}, ..., s_{n}:t_{n}\}$.
 This table type represents a record that maps the literals
-$s_{i}$\verb', ..., '$s_{n}$ to values of types
-$t_{i}$\verb', ..., '$t_{n}$, where $i$ may vary from one to $n$.
+$s_{i}$\texttt{, ..., }$s_{n}$ to values of types
+$t_{i}$\texttt{, ..., }$t_{n}$, where $i$ may vary from one to $n$.
 
@@
 The second line of this example gives a type error for the same
-reason that the last line of the previous example gave a type error
+reason that the last line of the previous example gives a type error
 too, that is, we are trying to assign a value that can be of two
 different types to a variable that accepts values of only one type.
-Although the field \verb'bar' does not exist in \verb't', the last
+Although the field \texttt{bar} does not exist in \texttt{t}, the last
 line of this example does not give a type error because the
 annotated type matches the type of the values that can be stored in
-\verb't'.
+\texttt{t}.
 
@@
 If we remove the annotation in this example, the compiler uses the
-syntax of records to infer the following table type to \verb'days':
+syntax of records to infer the following table type to \texttt{days}:
 \begin{align*}
@@
 $\{\Number:\String\;\cup\;\Nil\}$, which is the type that we used to
-annotate the variable \verb'days'.
+annotate the variable \texttt{days}.
 
@@
 Type inference would infer the very same table type that we used to
-annotate the variable \verb'person'.
-
-Typed Lua includes interfaces as syntactic sugar to the table type
-that specifies records.
+annotate the variable \texttt{person}.
+
+Typed Lua includes interfaces as syntactic sugar to table types
+that specify records.
 For instance, we can use the following interface as an alias to the
@@
 
-Now we can use the type \verb'Person' as an alias to a table type
-that certainly has the fields \verb'firstname' and \verb'lastname'.
-Next, we show an example of code that uses the type \verb'Person':
+Now we can use the type \texttt{Person} as an alias to a table type
+that certainly has the fields \texttt{firstname} and \texttt{lastname}.
+Next, we show an example of code that uses the type \texttt{Person}:
 \begin{verbatim}
@@
     local user2 = { firstname = "Lou" }
+
     local user3 = { lastname = "Reed",
@@
 
-We need to explicitly annotate \verb'next' being of type
-${\tt Element}\;\cup\;\Nil$ because in our type system $\Nil$
+We need to explicitly annotate \texttt{next} being of type
+\texttt{Element} $\cup\;\Nil$ because in our type system $\Nil$
 is not the bottom type, that is, $\Nil$ is subtype of itself only.
@@
 values in Lua.
-The syntax for defining a function type is \verb'(s) -> (s)',
-where \verb's' is a second-class type.
+The syntax for defining a function type is \texttt{(s) -> (s)},
+where \texttt{s} is a second-class type.
 As we saw in Figure \ref{fig:typelang}, a second-class type in
@@
 
-The next example shows how we can annotate a \verb'map' function
+The next example shows how we can annotate a \texttt{map} function
 that receives another function as its parameter and uses this
 function to alter the elements of a list of numbers, which is
-also a parameter of \verb'map':
+also a parameter of \texttt{map}:
 \begin{verbatim}
@@
 
-Metalua \cite{metalua} is a Lua compiler that supports compile-time
+Metalua \citep{metalua} is a Lua compiler that supports compile-time
 metaprogramming (CTMP).
 CTMP is a kind of macro system that allows the programmers to interact
-with the compiler \cite{fleutot2007contrasting}. 
+with the compiler \citep{fleutot2007contrasting}. 
 Metalua extends Lua 5.1 syntax to include its macro system,
@@
 styles, and can also provide syntax for turning simple type
-annotations into run-time assertions, but without compile-time type
-checking.
-
-MoonScript \cite{moonscript} is a programming language that supports
+annotations into run-time assertions.
+
+MoonScript \citep{moonscript} is a programming language that supports
 class-based object-oriented programming, and compiles to idiomatic
@@
 
-LuaInspect \cite{luainspect} is a tool that uses MetaLua to perform
+LuaInspect \citep{luainspect} is a tool that uses MetaLua to perform
 some code analysis, such as flagging unknown global variables and
@@
 
-Tidal Lock \cite{tidallock} is a prototype of another optional type
+Tidal Lock \citep{tidallock} is a prototype of another optional type
 system for Lua, which is written in Metalua.
-A remarkable feature of Tidal Lock is the way it handles tables,
+A remarkable feature of Tidal Lock is the way that it handles tables,
 because it allows the programmer to incrementally build the type of
@@
 table to build its type according to the keys that the programmer
-inserts in it, so the programmer do not need to bother about declaring
-its type with just one line of code.
-We also plan to add this feature to Typed Lua.
+inserts in it, so the programmer does not need to build the table in
+a single table constructor.
+Typed Lua has a simpler form of this feature, with a different
+formalization.
 Like Typed Lua, Tidal Lock also has local type inference.
@@
 
-\bibliographystyle{abbrv}
+\bibliographystyle{abbrvnat}
 \bibliography{typedlua}

### 2014-03-04 doc/paper/typedlua.tex
reorganization of sections, outline of section on the design of typed lua; removal of statistics

@@
 
-\section{Blending static and dynamic typing} \label{sec:review}
-
-We begin this section presenting a little bit of the history about
-combining static and dynamic typing in the same language.
-After that, we introduce optional type systems and gradual typing.
-We end this section discussing why optional type systems and two
-other approaches are often called gradual typing.
-
-\subsection{A little bit of history}
-
-Common LISP \citep{steele1982ocl} introduced optional type annotations
-in the early eighties, but not for static type checking.
-Instead, programmers may choose to declare types of variables as
-optimization hints to the compiler, that is, type declarations are
-just one way to help the compiler to optimize code.
-These annotations are unsafe because they can crash the program
-when they are wrong.
-
-\citet{abadi1989dts} extended the simply typed lambda calculus with the
-\texttt{Dynamic} type and the \texttt{dynamic} and \texttt{typecase}
-constructs, with the aim to safely integrate dynamic code in
-statically typed languages.
-The \texttt{Dynamic} type is a pair \texttt{(v,T)} where \texttt{v} is a
-value and \texttt{T} is the tag that represents the type of \texttt{v}.
-The constructs \texttt{dynamic} and \texttt{typecase} are explicit
-injection and projection operations, respectively.
-That is, \texttt{dynamic} builds values of type \texttt{Dynamic} and
-\texttt{typecase} safely inspects the type of a \texttt{Dynamic} value.
-Thus, migrating code between dynamic and static type checking requires
-changing type annotations and adding or removing \texttt{dynamic} and
-\texttt{typecase} constructs throughout the code.
-
-The \textit{quasi-static} type system proposed by \citet{thatte1990qst}
-performs implicit coercions and run-time checks to replace the
-\texttt{dynamic} and \texttt{typecase} constructs that were proposed by
-\citet{abadi1989dts}.
-To do that, quasi-static typing relies on subtyping with a top type
-$\Omega$ that represents the dynamic type, and splits type checking
-into two phases.
-The first phase inserts implicit coercions from the dynamic type to
-the expected type, while the second phase performs what Thatte calls
-\textit{plausibility checking}, that is, it rewrites the program to
-guarantee that sequences of upcasts and downcasts always have a
-common subtype.
-
-\textit{Soft typing} \citep{cartwright1991soft} is another approach
-to combine static and dynamic typing in the same language.
-The main goal of soft typing is to add static type checking to
-dynamically typed languages without compromising their flexibility.
-To do that, soft typing relies on type inference for
-translating dynamically typed code to statically typed code.
-The type checker inserts run-time checks around inconsistent code and
-warns the programmer about the insertion of these run-time checks,
-as they mean the existence of potential type errors.
-However, the programmer is free to choose between inspecting the
-run-time checks and simply running the code.
-Soft typing gives this choice to the programmer because it is one of
-its key ideas, that is, type inference and static type checking should
-not prevent the programmer from running inconsistent code.
-One advantage of soft typing is the fact that the compiler for
-softly typed languages can use the translated code to generate
-more efficient code, as the translated code statically type checks.
-One disadvantage of soft typing is that it can be cumbersome when
-the inferred types are meaningless large types that just confuse the
-programmer.
-
-\textit{Dynamic typing} \citep{henglein1994dts} is an approach
-that optimizes code from dynamically typed languages by eliminating
-unnecessary checks of tags.
-The paper describes how to translate dynamically typed code to
-statically typed code that uses a \texttt{Dynamic} type.
-The translation is done through a coercion calculus that uses type
-inference to insert the operations that are necessary to type check
-the \texttt{Dynamic} type during run-time.
-Although soft typing and dynamic typing may seem similar, they are not.
-They are not similar because soft typing targets statically type
-checking of dynamically typed languages for detecting programming
-errors while dynamic typing targets the optimization of dynamically
-typed code through the elimination of unnecessary run-time checks.
-In other words, soft typing sees code optimization as a side effect
-that comes with static type checking.
-
-\citet{findler2002chf} proposed contracts for higher-order functions
-and blame annotations for run-time checks.
-Contracts perform dynamic checking instead of static checking,
-but deferring all verifications to run-time can lead to defects
-that are difficult to fix, because run-time errors can show a
-stack trace where it is not clear to programmers if the cause of a
-certain run-time error is in application code or library code.
-Even if programmers identify that the source of a certain run-time
-error is in library code, they still may have problems to identify
-if this run-time error is due to a violation of library's contract,
-or due to a bug, when the library is poorly documented.
-In this approach, programmers can insert assertions in the form of
-contracts that check the input and output of higher-order functions.
-Thus, \citet{findler2002chf} show how programmers can insert
-assertions in the form of contracts that check the input and output
-of higher-order functions; and they also show how to add blame
-annotations in the generated code as a way to track assertion
-failures back to the source of the error.
-
-BabyJ \citep{anderson2002babyj} is an object-oriented language
-without inheritance that allows programmers to incrementally annotate
-the code with more specific types.
-Programmers can choose between using the dynamically typed version
-of BabyJ, when they do not need types at all, and the statically
-typed version of BabyJ, when they need to annotate the code.
-In the statically typed BabyJ, programmers can use the
-\textit{permissive type} $*$ to annotate the parts of the code that
-still do not have a specific type or the parts of the code that should
-have dynamic behavior.
-The type system of BabyJ is nominal, so types are either class names
-or the permissive type $*$.
-However, the type system does not use type equality or subtyping,
-but the relation $\approx$ between two types.
-The relation $\approx$ holds when both types have the same name or
-any of them is the permissive type $*$.
-Even though the permissive type $*$ is similar to the dynamic type
-from previous approaches, BabyJ does not provide any way to add
-implicit or explicit run-time checks.
-
-\citet{ou2004dtd} specified a language that combines static types
-with dependent types.
-To ensure safety, the compiler automatically inserts coercions
-between dependent code and static code.
-The coercions are run-time checks that ensure static code does not
-crash dependent code during run-time.
-
-\subsection{Optional Type Systems}
-
-Optional type systems \citep{bracha2004pluggable} are an approach for
-plugging static typing in dynamically typed languages through
-optional type annotations that are used to perform compile-time type
-checking, though not influencing the original run-time semantics of
-the language.
-This means that optional type systems can be unsound without affecting
-language safety, because the run-time semantics should still catch
-type errors independently of the static type checking.
-More precisely, optional type systems are unsound if they do not catch
-some type errors, but we say that languages with unsound optional type
-systems are safe because the semantics of these languages catch during
-run-time the type errors that the optional type system did not catch.
-Although optional type systems maintain the simplicity and
-flexibility of dynamically typed languages, their static type
-checking cannot prove the absence of all type errors when the optional
-type system is unsound.
-
-Optional type systems have an interesting relation to the untyped
-lambda calculus and the typed lambda calculus.
-The semantic rules of the unytped lambda calculus are the same of the
-typed lambda calculus, and the type system is actually optional and
-serves only for discarding programs that may have undesired behaviors
-\citep{bracha2004pluggable}.
-
-Optional type systems make software evolution easier and
-allow language designers to experiment with
-\textit{pluggable type systems} \citep{bracha2004pluggable}.
-Pluggable type systems are different optional type systems for
-checking different aspects of the program.
-
-According to \citet{bracha2004pluggable}, optional type systems should
-not include mandatory type inference too, because mandatory type
-inference may decrease the expressiveness of the type system.
-Instead, optional type systems should implement type inference as
-an independent mechanism that can be turned on and off according to
-the programmers' needs.
-
-Unfortunately, there is no formalization on optional type systems,
-so each language ends up implementing its optional type system on
-its own way.
-Now we will briefly discuss some languages that provide an informal
-description of their optional type systems rather than a formal one.
-
-Strongtalk \citep{bracha1993strongtalk,bracha1996strongtalk} is
-a version of Smalltalk that comes with an optional type system that
-does not include any type inference.
-
-Dart \citep{dart} and TypeScript \citep{typescript} are new
-languages that are designed with an optional type system, and both
-use JavaScript as their code generation target because their main
-purpose is web development.
-In fact, Dart is a new class-based object-oriented language with
-optional type annotations while TypeScript is a strict superset of
-JavaScript that provides optional type annotations and class-based
-object-oriented programming.
-Both Dart and TypeScript perform type inference to deduce static
-types for unannotated code.
-
-\subsection{Gradual Typing}
-
-The main goal of gradual typing \citep{siek2006gradual} is to allow
-programmers to choose between static and dynamic typing in the same
-language.
-To do that, \citet{siek2006gradual} extended the simply typed
-lambda calculus with the dynamic type $?$.
-In gradual typing, type annotations are optional, so an untyped
-variable is syntactic sugar for a variable whose declared type is
-the dynamic type $?$.
-Under these circumstances, we view gradual typing as a way to add
-a dynamic type to statically typed languages.
-
-The central idea of gradual typing is the \textit{consistency}
-relation, written $\tau_{1} \sim \tau_{2}$, that allows implicit
-conversions to and from the dynamic type, and disallows conversions
-between inconsistent types \citep{siek2006gradual}.
-For instance, $\mathbf{int} \sim \,?$ and $?\, \sim \mathbf{int}$,
-but $\mathbf{int} \not\sim \mathbf{string}$ and
-$\mathbf{string} \not\sim \mathbf{int}$. 
-The consistency relation is neither commutative nor transitive.
-
-Gradual typing is similar to the approaches proposed by
-\citet{abadi1989dts} and \citet{thatte1990qst} by including a
-dynamic type to a statically typed language.
-However, these three approaches differ in the way they handle the
-dynamic type.
-While \citet{siek2006gradual} rely on the consistency relation,
-\citet{abadi1989dts} rely on type equality with explicit projections
-and injections, and \citet{thatte1990qst} relies on subtyping.
-The subtyping relation actually is a pitfall on Thatte's quasi-static
-typing, because the dynamic type is put as the top and the bottom of
-the subtyping relation.
-Therefore, downcasts combined with the transitivity of subtyping
-accepts programs that should be rejected.
-
-Later, \citet{siek2007objects} discovered that the consistency relation
-is orthogonal to subtyping, so we can combine the two relations to
-achieve both consistency and subtyping.
-The \textit{consistent-subtyping} relation,
-written $\tau_{1} \lesssim \tau_{2}$,
-is essential to design object-oriented languages that have a
-gradual type system.
-Like the consistency relation, and unlike the subtyping relation,
-the consistent-subtyping relation is not transitive.
-
-Another important feature of gradual typing is the theoretic
-foundation that it provides for inserting run-time checks that
-prove dynamically typed code does not violate the invariants of
-statically typed code, thus preserving type safety.
-To do that, \citet{siek2006gradual} defined the run-time semantics
-of gradual typing as a translation to an intermediate language with
-explicit casts at the frontiers between statically and dynamically
-typed code.
-The semantics of these casts is based on the higher-order contracts
-proposed by \citet{findler2002chf}.
-
-However, \citet{herman2007sgt} showed that there is an efficiency
-concern regarding the run-time checks, because there are two
-ways that casts can lead to unbounded space consumption.
-The first affects tail recursion while the second appears when
-first-class functions or objects cross the border between
-static code and dynamic code, that is, some programs can apply
-repeated casts to the same function or object.
-\citet{herman2007sgt} use the coercion calculus of
-\citet{henglein1994dts} to solve the problem of space efficiency,
-that is, they use coercions to express casts.
-Their approach normalizes an arbitrary sequence of coercions to a
-coercion of bounded size.
-
-Another concern about casts is how to improve debugging support,
-because a cast application can be delayed and the error related
-to that cast application can appear far way from the real error.
-\citet{wadler2009wpc} developed the \textit{blame calculus} as a way
-to attack this issue, and \citet{ahmed2011bfa} extended the
-blame calculus with polymorphism.
-The blame calculus is an intermediate language to integrate
-static and dynamic typing that uses the blame tracking approach
-proposed by \citet{findler2002chf}.
-
-On the one hand, the blame calculus solves the issue regarding
-error reporting;
-on the other hand, it has the space efficiency problem reported
-by \citet{herman2007sgt}.
-Thus, \citet{siek2009casts} extended the coercion calculus of
-\citet{herman2007sgt} with blame tracking to achieve an
-implementation of the blame calculus that is space efficient.
-After that, \citet{siek2010blame} proposed a new solution that also
-attacks both problems.
-This new solution is based on a concept called \textit{threesome},
-which is a way to split a cast between two parties into two casts
-among three parties.
-A cast has a source and a target type (a \textit{twosome}),
-so we can split any cast into a downcast from the source to an
-intermediate type that is followed by an upcast from the intermediate
-type to the target type (a \textit{threesome}).
-
-ActionScript \citep{moock2007as3} is the first language that
-incorporated a gradual type system to its implementation and
-Perl 6 \citep{tang2007pri} is also being designed with a
-gradual type system, though there is little documentation about
-the gradual type systems of these languages.
-Grace \citep{black2012grace,black2013sg} is a new object-oriented,
-gradually typed, educational language.
-In Grace, modules are gradually typed objects, that is, modules
-may have types and methods as attributes, and can have a state
-\citep{homer2013modules}.
-Gradual Python \citep{reticulated} is an attempt to develop a
-gradual type system for Python that deals with mutable objects
-\citep{siek2013mutable}.
-Gradualtalk \citep{allende2013gts} is a gradually-typed Smalltalk
-that introduces a new cast insertion strategy for gradually-typed
-objects \citep{allende2013cis}.
-
-\subsection{Approaches that are often called Gradual Typing}
-
-Gradual typing is similar to optional type systems in that type
-annotations are optional, and unannotated code is dynamically
-typed, but unlike optional type systems, gradual typing changes 
-the run-time semantics to preserve type safety, and it is a way to
-add a dynamic type to statically typed languages. 
-More precisely, programming languages that include a gradual type
-system implement the semantics of statically typed languages, so
-the gradual type system inserts casts in the translated code to
-guarantee that types are consistent before execution, while
-programming languages that include an optional type system still
-implement the semantics of dynamically typed languages, so all
-the type checking also belongs to the semantics of each operation.
-
-Still, we can view gradual typing as a way to formalize an optional
-type system when the gradual type system does not insert run-time
-checks.
-BabyJ \citep{anderson2002babyj} and Alore \citep{lehtosalo2011alore}
-are two examples of object-oriented languages that use gradual
-typing to formalize their optional type systems.
-The optional type systems of both BabyJ and Alore are nominal.
-BabyJ uses the relation $\approx$ that is similar to the consistency
-relation while Alore combines  subtyping along with the consistency
-relation to define a \textit{consistent-or-subtype} relation.
-The consistent-or-subtype relation is different from the
-consistent-subtyping relation of \citet{siek2007objects}, but it is
-also written $\tau_{1} \lesssim \tau_{2}$.
-The consistent-or-subtype relation holds when $\tau_{1} \sim \tau_{2}$
-or $\tau_{1} <: \tau_{2}$, so $<:$ is transitive and $\sim$ is not.
-Alore also extends its optional type system to include optional
-monitoring of run-time type errors in the gradual typing style.
-
-Hence, optional type annotations for software evolution probably are
-the reasons why optional type systems are commonly called
-gradual type systems.
-Typed Clojure \citep{bonnaire-sergeant2012typed-clojure} is an
-optional type system for Clojure that now is adopting the
-gradual typing slogan.
-
-\citet{flanagan2006htc} introduced \textit{hybrid type checking},
-an approach that combines static types and refinement types.
-For instance, programmers can specify the refinement type
-$\{x:Int \;|\; x \ge 0\}$ when they need a type for natural numbers.
-The programmer can also choose between explicit or implicit casts.
-When casts are not explicit, the type checker uses a theorem prover
-to insert casts.
-In our example of natural numbers, a cast would be inserted to check
-whether an integer is greater than or equal to zero.
-
-Sage \citep{gronski2006sage} is a programming language that
-extended hybrid type checking with a dynamic type to
-support dynamic and static typing in the same language.
-Sage also offers optional type annotations in the gradual typing
-style, that is, unannotated code is syntactic sugar for
-code whose declared type is the dynamic type.
-
-Thus, the inclusion of a dynamic type to hybrid type checking
-along with optional type annotations, and the insertion of run-time
-checks probably are the reasons why hybrid type checking is
-also viewed as a form of gradual typing.
-
-\citet{tobin-hochstadt2006ims} proposed another approach for gradually
-migrating from dynamically typed to statically typed code,
-and they coined the term \textit{from scripts to programs} for
-referring to this kind of interlanguage migration.
-In their approach, the migration from dynamically typed to
-statically typed code happens module-by-module, so they designed
-and implemented Typed Racket \citep{tobin-hochstadt2008ts} for
-this purpose.
-Typed Racket is a statically typed version of Racket
-(a Scheme dialect) that allows the programmer to write typed modules,
-so Typed Racket modules can coexist with Racket modules,
-which are untyped.
-
-The approach used by \citet{tobin-hochstadt2008ts} to design and
-implement Typed Racket probably is also called gradual typing
-because it allows the programmer to gradually migrate from untyped
-scripts to typed programs.
-However, note that Typed Racket is a statically typed language
-and type annotations are mandatory.
-So, what makes it gradual is a type system with a dynamic type
-that handles the interaction between Racket and Typed Racket modules.
-
 \section{Typed Lua} \label{sec:proposal}
@@
 
-\subsection{Statistics about the usage of Lua}
-
-We used the LuaRocks repository to build our statistics database;
-LuaRocks \citep{hisham2013luarocks} is a package manager for Lua
-modules.
-We downloaded the 3928 \texttt{.lua} files that were available in
-the LuaRocks repository at February 1st 2014.
-However, we ignored files that are not compatible with Lua 5.2,
-the latest version of Lua.
-We also ignored \textit{machine-generated} files and test files,
-because these files may not represent idiomatic Lua code,
-and might skew our statistics towards non-typical uses of Lua.
-This left 2598 \texttt{.lua} files from 262 different projects for
-our statistics database;
-we parsed these files and processed their abstract syntax tree
-to gather the statistics that we show in this section.
-
-To verify how programmers use tables, we measured how they
-initialize, index, and iterate tables.
-We present these statistics in the next three paragraphs to discuss
-their influence on our type system.
-
-The table constructor appears 23185 times.
-In 36\% of the occurrences it is a constructor that initializes a
-record (e.g., \texttt{\{ x = 120, y = 121 \}});
-in 29\% of the occurrences it is a constructor that initializes a
-list (e.g., \texttt{\{ "one", "two", "three", "four" \}});
-in 8\% of the occurrences it is a constructor that initializes a
-record with a list part;
-and in less than 1\% of the occurrences (4 times) it is a constructor
-that uses only the booleans \texttt{true} and \texttt{false} as indexes.
-At all, in 73\% of the occurrences it is a constructor that uses
-only literal keys;
-in 26\% of the occurrences it is the empty constructor;
-in 1\% of the occurrences it is a constructor with non-literal keys
-only, that is, a constructor that uses variables and function calls
-to create the indexes of a table;
-and in less than 1\% of the occurrences (19 times) it is a constructor
-that mixes literal keys and non-literal keys.
-
-The indexing of tables appears 130448 times:
-86\% of them are for reading a table field while
-14\% of them are for writing into a table field.
-We can classify the indexing operations that are reads as follows:
-89\% of the reads use a literal string key,
-4\% of the reads use a literal number key,
-and less than 1\% of the reads (10 times) use a literal boolean key.
-At all, 93\% of the reads use literals to index a table while
-7\% of the reads use non-literal expressions to index a table.
-It also worth mentioning that 45\% of the reads are actually
-function calls.
-More precisely, 25\% of the reads use literals to call a function,
-20\% of the reads use literals to call a method, that is,
-a function call that uses the colon syntactic sugar, 
-and less than 1\% of the reads (195 times) use non-literal expressions
-to call a function.
-We can also classify the indexing operations that are writes as follows: 
-69\% of the writes use a literal string key,
-2\% of the writes use a literal number key,
-and less than 1\% of the writes (1 time) uses a literal boolean key.
-At all, 71\% of the writes use literals to index a table while
-29\% of the writes use non-literal expressions to index a table.
-
-We also measured how many files have code that iterate over tables to
-observe how frequently iteration is used.
-We observed that 23\% of the files have code that iterate over keys
-of any value, that is, the call to \texttt{pairs} appears at least
-once in these files (the median is twice per file);
-21\% of the files have code that iterate over integer keys, that is,
-the call to \texttt{ipairs} appears at least once in these files
-(the median is also twice per file);
-and 10\% of the files have code that use the numeric \texttt{for}
-along with the length operator (the median is once per file).
-
-The numbers about table initialization, indexing, and iteration
-show us that tables are mostly used to represent records, lists,
-and associative arrays.
-Therefore, Typed Lua should include a table type for handling
-these uses of Lua tables.
-Even though the statistics show that programmers initialize tables
-more often than they use the empty constructor to
-dynamically initialize tables, the statistics of the empty
-constructor are still expressive and indicate that Typed Lua should
-also include a way to handle this style of defining table types.
-
-We found a total of 24858 function declarations in our database
-(the median is six per file).
-Next, we discuss how frequently programmers use dynamic type
-checking and multiple return values inside these functions.
-
-We observed that 9\% of the functions perform dynamic type checking
-on their input parameters, that is, these functions use \texttt{type}
-to inspect the tags of Lua values (the median is once per function).
-We randomly selected 20 functions to sample how programmers are
-using \texttt{type}, and we got the following data:
-50\% of these functions use \texttt{type} for asserting the tags of
-their input parameters, that is, they raise an error when the tag of a
-certain parameter does not match the expected tag, and
-50\% of these functions use \texttt{type} for overloading, that is,
-they execute different code according to the inspected tag.
-
-These numbers show us that Typed Lua should include union types,
-because the use of the \texttt{type} idiom shows that disjoint unions
-will help programmers define data structures that can hold a value of
-several different, but fixed types.
-Typed Lua should also use \texttt{type} as a mechanism for decomposing
-unions, though it may be restricted to base types only.
-
-We observed that 10\% of the functions explicitly return multiple
-values.
-We also observed that 5\% of the functions return \texttt{nil} plus
-something else, for signaling an unexpected behavior;
-and 1\% of the functions return \texttt{false} plus something else,
-also for signaling an unexpected behavior.
-
-Typed Lua should include function types to represent Lua functions,
-and tuple types to represent the signatures of Lua functions,
-multiple return values, and multiple assignments.
-Tuple types will require some special attention, because Typed Lua
-should be able to adjust tuple types during compile-time, in a
-similar way to what Lua does with function calls and multiple
-assignments during run-time.
-In addition, the number of functions that return \texttt{nil} and
-\texttt{false} plus something else show us that overloading on the
-return type will be useful to the type system.
-
-We also measured how frequently programmers use the object-oriented
-paradigm in Lua.
-We observed that 23\% of the function declarations are actually
-method declarations.
-More precisely, 14\% of them use the colon syntactic sugar while
-9\% of them use \texttt{self} as their first parameter.
-We also observed that 63\% of the projects extend tables with
-metatables, that is, they call \texttt{setmetatable} at least once,
-and 27\% of the projects access the metatable of a given table,
-that is, they call \texttt{getmetatable} at least once.
-In fact, 45\% of the projects extend tables with metatables and
-declare methods:
-13\% using the colon syntactic sugar, 14\% using \texttt{self}, and
-18\% using both.
-
-Based on these observations, Typed Lua should include interfaces and
-classes as syntactic sugar for table types that encode objects.
-Proposing an optional type system for Lua that includes classes may
-sound unusual at first glance, because Lua does not have standard
-policies for object-oriented programming.
-However, Lua provides mechanisms that allow programmers to abstract
-their code in terms of objects, and our statistics confirm that
-an expressive number of programmers are relying on these mechanisms to
-use the object-oriented paradigm in Lua.
-Typed Lua should include interfaces and classes as merely syntactic
-mechanisms that help the compiler type check object-oriented code,
-and they should not influence the semantics of Lua.
-
-We also measured how programmers are defining modules.
-We observed that 38\% of the files use the current way of defining
-modules, that is, these files return a table that contains the
-exported members of the module at the end of the file;
-22\% of the files still use the deprecated way of defining modules,
-that is, these files call the function \texttt{module};
-and 1\% of the files use both ways.
-At all, 61\% of the files are modules while 39\% of the files are
-plain scripts.
-The number of plain scripts is high considering the origin of
-our database.
-However, we did not ignore sample scripts, which usually serve to
-help the users of a given module on how to use this module, and
-that is the reason why we have a high number of plain scripts.
-
-Based on these observations, Typed Lua should also use interfaces as
-syntactic sugar for table types that define modules.
-Typed Lua should also support the deprecated style of module
-definition, using global names as exported members of the module.
-
-Besides using interfaces for encoding objects and defining modules,
-Typed Lua should also use interfaces for defining the types of
-userdata, as programmers often use userdata in an object-oriented
-style.
-
-The last statistics that we collected were about variadic functions
-and vararg expressions.
-We observed that 8\% of the functions are variadic, that is,
-their last parameter is the vararg expression.
-We also observed that 5\% of the initialization of lists
-(or 2\% of the occurrences of the table constructor) use solely the
-vararg expression.
-Typed Lua should include a \textit{vararg type} to handle variadic
-functions and vararg expressions.
-
-\subsection{Typed Lua through examples}
+\subsection{Atomic Types and Functions}
 
@@
 \begin{verbatim}
-
     local function fact (n:number):number
@@
 
+Typed Lua includes function types because functions are first-class
+values in Lua.
+The syntax for defining a function type is \texttt{(s) -> (s)},
+where \texttt{s} is a second-class type.
+As we saw in Figure \ref{fig:typelang}, a second-class type in
+Typed Lua can be the type $\Void$, a first-class type, a variadic
+type, or a tuple type.
+We use tuple types to represent multiple parameters, multiple return
+values, and multiple assignments, and we use the variadic type to
+represent variadic functions and the vararg expression.
+Even though we already included the tuple type and the vararg type,
+we are still working on these types.
+
+The next example shows how we can annotate a \texttt{map} function
+that receives another function as its parameter and uses this
+function to alter the elements of a list of numbers, which is
+also a parameter of \texttt{map}:
+\begin{verbatim}
+    local function map (f:(number)->(number),
+                        l:{number}):void
+      for i=1, #l do
+        l[i] = f(l[i] or 0)
+      end
+    end
+\end{verbatim}
+
+\subsection{Unions}
+
 Typed Lua includes union types as a way to discriminate data
@@
 $t\;\cup\;\Nil\;\mathbf{or}\;t$ always has the type $t$.
+
+\subsection{Tables and Interfaces}
 
@@
 
+\subsection{Modules}
+
+Typed Lua will also introduce a specific syntax for defining modules.
+For now, each file is a module.
+We use all global names and their types to build the interface that
+represents the type of the module, in case the module is not returned
+at the end of the file.
+
+\subsection{Objects and Classes}
+
 Besides interfaces, Typed Lua will also introduce classes.
@@
 
-Typed Lua will also introduce a specific syntax for defining modules.
-For now, each file is a module.
-We use all global names and their types to build the interface that
-represents the type of the module, in case the module is not returned
-at the end of the file.
-
-Typed Lua includes function types because functions are first-class
-values in Lua.
-The syntax for defining a function type is \texttt{(s) -> (s)},
-where \texttt{s} is a second-class type.
-As we saw in Figure \ref{fig:typelang}, a second-class type in
-Typed Lua can be the type $\Void$, a first-class type, a variadic
-type, or a tuple type.
-We use tuple types to represent multiple parameters, multiple return
-values, and multiple assignments, and we use the variadic type to
-represent variadic functions and the vararg expression.
-Even though we already included the tuple type and the vararg type,
-we are still working on these types.
-
-The next example shows how we can annotate a \texttt{map} function
-that receives another function as its parameter and uses this
-function to alter the elements of a list of numbers, which is
-also a parameter of \texttt{map}:
-\begin{verbatim}
-    local function map (f:(number)->(number),
-                        l:{number}):void
-      for i=1, #l do
-        l[i] = f(l[i] or 0)
-      end
-    end
-\end{verbatim}
-
-To conclude this section, we just started the design and
-implementation of Typed Lua, and there is still several design
-decisions to be taken on this subject.
-For instance, our type system should handle the subtleties that
-come with metatables.
+\section{Related Work}
+ \label{sec:review}
+
+We begin this section presenting a little bit of the history about
+combining static and dynamic typing in the same language.
+After that, we introduce optional type systems and gradual typing.
+We end this section discussing why optional type systems and two
+other approaches are often called gradual typing.
+
+\subsection{A little bit of history}
+
+Common LISP \citep{steele1982ocl} introduced optional type annotations
+in the early eighties, but not for static type checking.
+Instead, programmers may choose to declare types of variables as
+optimization hints to the compiler, that is, type declarations are
+just one way to help the compiler to optimize code.
+These annotations are unsafe because they can crash the program
+when they are wrong.
+
+\citet{abadi1989dts} extended the simply typed lambda calculus with the
+\texttt{Dynamic} type and the \texttt{dynamic} and \texttt{typecase}
+constructs, with the aim to safely integrate dynamic code in
+statically typed languages.
+The \texttt{Dynamic} type is a pair \texttt{(v,T)} where \texttt{v} is a
+value and \texttt{T} is the tag that represents the type of \texttt{v}.
+The constructs \texttt{dynamic} and \texttt{typecase} are explicit
+injection and projection operations, respectively.
+That is, \texttt{dynamic} builds values of type \texttt{Dynamic} and
+\texttt{typecase} safely inspects the type of a \texttt{Dynamic} value.
+Thus, migrating code between dynamic and static type checking requires
+changing type annotations and adding or removing \texttt{dynamic} and
+\texttt{typecase} constructs throughout the code.
+
+The \textit{quasi-static} type system proposed by \citet{thatte1990qst}
+performs implicit coercions and run-time checks to replace the
+\texttt{dynamic} and \texttt{typecase} constructs that were proposed by
+\citet{abadi1989dts}.
+To do that, quasi-static typing relies on subtyping with a top type
+$\Omega$ that represents the dynamic type, and splits type checking
+into two phases.
+The first phase inserts implicit coercions from the dynamic type to
+the expected type, while the second phase performs what Thatte calls
+\textit{plausibility checking}, that is, it rewrites the program to
+guarantee that sequences of upcasts and downcasts always have a
+common subtype.
+
+\textit{Soft typing} \citep{cartwright1991soft} is another approach
+to combine static and dynamic typing in the same language.
+The main goal of soft typing is to add static type checking to
+dynamically typed languages without compromising their flexibility.
+To do that, soft typing relies on type inference for
+translating dynamically typed code to statically typed code.
+The type checker inserts run-time checks around inconsistent code and
+warns the programmer about the insertion of these run-time checks,
+as they mean the existence of potential type errors.
+However, the programmer is free to choose between inspecting the
+run-time checks and simply running the code.
+Soft typing gives this choice to the programmer because it is one of
+its key ideas, that is, type inference and static type checking should
+not prevent the programmer from running inconsistent code.
+One advantage of soft typing is the fact that the compiler for
+softly typed languages can use the translated code to generate
+more efficient code, as the translated code statically type checks.
+One disadvantage of soft typing is that it can be cumbersome when
+the inferred types are meaningless large types that just confuse the
+programmer.
+
+\textit{Dynamic typing} \citep{henglein1994dts} is an approach
+that optimizes code from dynamically typed languages by eliminating
+unnecessary checks of tags.
+The paper describes how to translate dynamically typed code to
+statically typed code that uses a \texttt{Dynamic} type.
+The translation is done through a coercion calculus that uses type
+inference to insert the operations that are necessary to type check
+the \texttt{Dynamic} type during run-time.
+Although soft typing and dynamic typing may seem similar, they are not.
+They are not similar because soft typing targets statically type
+checking of dynamically typed languages for detecting programming
+errors while dynamic typing targets the optimization of dynamically
+typed code through the elimination of unnecessary run-time checks.
+In other words, soft typing sees code optimization as a side effect
+that comes with static type checking.
+
+\citet{findler2002chf} proposed contracts for higher-order functions
+and blame annotations for run-time checks.
+Contracts perform dynamic checking instead of static checking,
+but deferring all verifications to run-time can lead to defects
+that are difficult to fix, because run-time errors can show a
+stack trace where it is not clear to programmers if the cause of a
+certain run-time error is in application code or library code.
+Even if programmers identify that the source of a certain run-time
+error is in library code, they still may have problems to identify
+if this run-time error is due to a violation of library's contract,
+or due to a bug, when the library is poorly documented.
+In this approach, programmers can insert assertions in the form of
+contracts that check the input and output of higher-order functions.
+Thus, \citet{findler2002chf} show how programmers can insert
+assertions in the form of contracts that check the input and output
+of higher-order functions; and they also show how to add blame
+annotations in the generated code as a way to track assertion
+failures back to the source of the error.
+
+BabyJ \citep{anderson2002babyj} is an object-oriented language
+without inheritance that allows programmers to incrementally annotate
+the code with more specific types.
+Programmers can choose between using the dynamically typed version
+of BabyJ, when they do not need types at all, and the statically
+typed version of BabyJ, when they need to annotate the code.
+In the statically typed BabyJ, programmers can use the
+\textit{permissive type} $*$ to annotate the parts of the code that
+still do not have a specific type or the parts of the code that should
+have dynamic behavior.
+The type system of BabyJ is nominal, so types are either class names
+or the permissive type $*$.
+However, the type system does not use type equality or subtyping,
+but the relation $\approx$ between two types.
+The relation $\approx$ holds when both types have the same name or
+any of them is the permissive type $*$.
+Even though the permissive type $*$ is similar to the dynamic type
+from previous approaches, BabyJ does not provide any way to add
+implicit or explicit run-time checks.
+
+\citet{ou2004dtd} specified a language that combines static types
+with dependent types.
+To ensure safety, the compiler automatically inserts coercions
+between dependent code and static code.
+The coercions are run-time checks that ensure static code does not
+crash dependent code during run-time.
+
+\subsection{Optional Type Systems}
+
+Optional type systems \citep{bracha2004pluggable} are an approach for
+plugging static typing in dynamically typed languages through
+optional type annotations that are used to perform compile-time type
+checking, though not influencing the original run-time semantics of
+the language.
+This means that optional type systems can be unsound without affecting
+language safety, because the run-time semantics should still catch
+type errors independently of the static type checking.
+More precisely, optional type systems are unsound if they do not catch
+some type errors, but we say that languages with unsound optional type
+systems are safe because the semantics of these languages catch during
+run-time the type errors that the optional type system did not catch.
+Although optional type systems maintain the simplicity and
+flexibility of dynamically typed languages, their static type
+checking cannot prove the absence of all type errors when the optional
+type system is unsound.
+
+Optional type systems have an interesting relation to the untyped
+lambda calculus and the typed lambda calculus.
+The semantic rules of the unytped lambda calculus are the same of the
+typed lambda calculus, and the type system is actually optional and
+serves only for discarding programs that may have undesired behaviors
+\citep{bracha2004pluggable}.
+
+Optional type systems make software evolution easier and
+allow language designers to experiment with
+\textit{pluggable type systems} \citep{bracha2004pluggable}.
+Pluggable type systems are different optional type systems for
+checking different aspects of the program.
+
+According to \citet{bracha2004pluggable}, optional type systems should
+not include mandatory type inference too, because mandatory type
+inference may decrease the expressiveness of the type system.
+Instead, optional type systems should implement type inference as
+an independent mechanism that can be turned on and off according to
+the programmers' needs.
+
+Unfortunately, there is no formalization on optional type systems,
+so each language ends up implementing its optional type system on
+its own way.
+Now we will briefly discuss some languages that provide an informal
+description of their optional type systems rather than a formal one.
+
+Strongtalk \citep{bracha1993strongtalk,bracha1996strongtalk} is
+a version of Smalltalk that comes with an optional type system that
+does not include any type inference.
+
+Dart \citep{dart} and TypeScript \citep{typescript} are new
+languages that are designed with an optional type system, and both
+use JavaScript as their code generation target because their main
+purpose is web development.
+In fact, Dart is a new class-based object-oriented language with
+optional type annotations while TypeScript is a strict superset of
+JavaScript that provides optional type annotations and class-based
+object-oriented programming.
+Both Dart and TypeScript perform type inference to deduce static
+types for unannotated code.
+
+\subsection{Gradual Typing}
+
+The main goal of gradual typing \citep{siek2006gradual} is to allow
+programmers to choose between static and dynamic typing in the same
+language.
+To do that, \citet{siek2006gradual} extended the simply typed
+lambda calculus with the dynamic type $?$.
+In gradual typing, type annotations are optional, so an untyped
+variable is syntactic sugar for a variable whose declared type is
+the dynamic type $?$.
+Under these circumstances, we view gradual typing as a way to add
+a dynamic type to statically typed languages.
+
+The central idea of gradual typing is the \textit{consistency}
+relation, written $\tau_{1} \sim \tau_{2}$, that allows implicit
+conversions to and from the dynamic type, and disallows conversions
+between inconsistent types \citep{siek2006gradual}.
+For instance, $\mathbf{int} \sim \,?$ and $?\, \sim \mathbf{int}$,
+but $\mathbf{int} \not\sim \mathbf{string}$ and
+$\mathbf{string} \not\sim \mathbf{int}$. 
+The consistency relation is neither commutative nor transitive.
+
+Gradual typing is similar to the approaches proposed by
+\citet{abadi1989dts} and \citet{thatte1990qst} by including a
+dynamic type to a statically typed language.
+However, these three approaches differ in the way they handle the
+dynamic type.
+While \citet{siek2006gradual} rely on the consistency relation,
+\citet{abadi1989dts} rely on type equality with explicit projections
+and injections, and \citet{thatte1990qst} relies on subtyping.
+The subtyping relation actually is a pitfall on Thatte's quasi-static
+typing, because the dynamic type is put as the top and the bottom of
+the subtyping relation.
+Therefore, downcasts combined with the transitivity of subtyping
+accepts programs that should be rejected.
+
+Later, \citet{siek2007objects} discovered that the consistency relation
+is orthogonal to subtyping, so we can combine the two relations to
+achieve both consistency and subtyping.
+The \textit{consistent-subtyping} relation,
+written $\tau_{1} \lesssim \tau_{2}$,
+is essential to design object-oriented languages that have a
+gradual type system.
+Like the consistency relation, and unlike the subtyping relation,
+the consistent-subtyping relation is not transitive.
+
+Another important feature of gradual typing is the theoretic
+foundation that it provides for inserting run-time checks that
+prove dynamically typed code does not violate the invariants of
+statically typed code, thus preserving type safety.
+To do that, \citet{siek2006gradual} defined the run-time semantics
+of gradual typing as a translation to an intermediate language with
+explicit casts at the frontiers between statically and dynamically
+typed code.
+The semantics of these casts is based on the higher-order contracts
+proposed by \citet{findler2002chf}.
+
+However, \citet{herman2007sgt} showed that there is an efficiency
+concern regarding the run-time checks, because there are two
+ways that casts can lead to unbounded space consumption.
+The first affects tail recursion while the second appears when
+first-class functions or objects cross the border between
+static code and dynamic code, that is, some programs can apply
+repeated casts to the same function or object.
+\citet{herman2007sgt} use the coercion calculus of
+\citet{henglein1994dts} to solve the problem of space efficiency,
+that is, they use coercions to express casts.
+Their approach normalizes an arbitrary sequence of coercions to a
+coercion of bounded size.
+
+Another concern about casts is how to improve debugging support,
+because a cast application can be delayed and the error related
+to that cast application can appear far way from the real error.
+\citet{wadler2009wpc} developed the \textit{blame calculus} as a way
+to attack this issue, and \citet{ahmed2011bfa} extended the
+blame calculus with polymorphism.
+The blame calculus is an intermediate language to integrate
+static and dynamic typing that uses the blame tracking approach
+proposed by \citet{findler2002chf}.
+
+On the one hand, the blame calculus solves the issue regarding
+error reporting;
+on the other hand, it has the space efficiency problem reported
+by \citet{herman2007sgt}.
+Thus, \citet{siek2009casts} extended the coercion calculus of
+\citet{herman2007sgt} with blame tracking to achieve an
+implementation of the blame calculus that is space efficient.
+After that, \citet{siek2010blame} proposed a new solution that also
+attacks both problems.
+This new solution is based on a concept called \textit{threesome},
+which is a way to split a cast between two parties into two casts
+among three parties.
+A cast has a source and a target type (a \textit{twosome}),
+so we can split any cast into a downcast from the source to an
+intermediate type that is followed by an upcast from the intermediate
+type to the target type (a \textit{threesome}).
+
+ActionScript \citep{moock2007as3} is the first language that
+incorporated a gradual type system to its implementation and
+Perl 6 \citep{tang2007pri} is also being designed with a
+gradual type system, though there is little documentation about
+the gradual type systems of these languages.
+Grace \citep{black2012grace,black2013sg} is a new object-oriented,
+gradually typed, educational language.
+In Grace, modules are gradually typed objects, that is, modules
+may have types and methods as attributes, and can have a state
+\citep{homer2013modules}.
+Gradual Python \citep{reticulated} is an attempt to develop a
+gradual type system for Python that deals with mutable objects
+\citep{siek2013mutable}.
+Gradualtalk \citep{allende2013gts} is a gradually-typed Smalltalk
+that introduces a new cast insertion strategy for gradually-typed
+objects \citep{allende2013cis}.
+
+\subsection{Approaches that are often called Gradual Typing}
+
+Gradual typing is similar to optional type systems in that type
+annotations are optional, and unannotated code is dynamically
+typed, but unlike optional type systems, gradual typing changes 
+the run-time semantics to preserve type safety, and it is a way to
+add a dynamic type to statically typed languages. 
+More precisely, programming languages that include a gradual type
+system implement the semantics of statically typed languages, so
+the gradual type system inserts casts in the translated code to
+guarantee that types are consistent before execution, while
+programming languages that include an optional type system still
+implement the semantics of dynamically typed languages, so all
+the type checking also belongs to the semantics of each operation.
+
+Still, we can view gradual typing as a way to formalize an optional
+type system when the gradual type system does not insert run-time
+checks.
+BabyJ \citep{anderson2002babyj} and Alore \citep{lehtosalo2011alore}
+are two examples of object-oriented languages that use gradual
+typing to formalize their optional type systems.
+The optional type systems of both BabyJ and Alore are nominal.
+BabyJ uses the relation $\approx$ that is similar to the consistency
+relation while Alore combines  subtyping along with the consistency
+relation to define a \textit{consistent-or-subtype} relation.
+The consistent-or-subtype relation is different from the
+consistent-subtyping relation of \citet{siek2007objects}, but it is
+also written $\tau_{1} \lesssim \tau_{2}$.
+The consistent-or-subtype relation holds when $\tau_{1} \sim \tau_{2}$
+or $\tau_{1} <: \tau_{2}$, so $<:$ is transitive and $\sim$ is not.
+Alore also extends its optional type system to include optional
+monitoring of run-time type errors in the gradual typing style.
+
+Hence, optional type annotations for software evolution probably are
+the reasons why optional type systems are commonly called
+gradual type systems.
+Typed Clojure \citep{bonnaire-sergeant2012typed-clojure} is an
+optional type system for Clojure that now is adopting the
+gradual typing slogan.
+
+\citet{flanagan2006htc} introduced \textit{hybrid type checking},
+an approach that combines static types and refinement types.
+For instance, programmers can specify the refinement type
+$\{x:Int \;|\; x \ge 0\}$ when they need a type for natural numbers.
+The programmer can also choose between explicit or implicit casts.
+When casts are not explicit, the type checker uses a theorem prover
+to insert casts.
+In our example of natural numbers, a cast would be inserted to check
+whether an integer is greater than or equal to zero.
+
+Sage \citep{gronski2006sage} is a programming language that
+extended hybrid type checking with a dynamic type to
+support dynamic and static typing in the same language.
+Sage also offers optional type annotations in the gradual typing
+style, that is, unannotated code is syntactic sugar for
+code whose declared type is the dynamic type.
+
+Thus, the inclusion of a dynamic type to hybrid type checking
+along with optional type annotations, and the insertion of run-time
+checks probably are the reasons why hybrid type checking is
+also viewed as a form of gradual typing.
+
+\citet{tobin-hochstadt2006ims} proposed another approach for gradually
+migrating from dynamically typed to statically typed code,
+and they coined the term \textit{from scripts to programs} for
+referring to this kind of interlanguage migration.
+In their approach, the migration from dynamically typed to
+statically typed code happens module-by-module, so they designed
+and implemented Typed Racket \citep{tobin-hochstadt2008ts} for
+this purpose.
+Typed Racket is a statically typed version of Racket
+(a Scheme dialect) that allows the programmer to write typed modules,
+so Typed Racket modules can coexist with Racket modules,
+which are untyped.
+
+The approach used by \citet{tobin-hochstadt2008ts} to design and
+implement Typed Racket probably is also called gradual typing
+because it allows the programmer to gradually migrate from untyped
+scripts to typed programs.
+However, note that Typed Racket is a statically typed language
+and type annotations are mandatory.
+So, what makes it gradual is a type system with a dynamic type
+that handles the interaction between Racket and Typed Racket modules.
 
@@
 
-\section{Expected contributions} \label{sec:con}
-
-Our main contribution is to document the process of designing and
-implementing an optional type system for Lua, a major scripting
-language.
-Lua is a small scripting language that shares some features with other
-scripting languages, so the thesis should help language designers
-that wish to add optional and gradual type systems to other
-scripting languages.
-
-We also believe that offering a framework that Lua developers can
-use to document, test, and better structure their code is also a
-major contribution of our thesis.
-
-Finally, our optional type system could also help some Lua
-implementations to perform code optimization.
-However, the fact that we are not currently providing guarantees on
-the interaction between dynamic and static code is an obstacle for
-performing safe optimizations.
+\section{Conclusion}
 

### 2014-03-04 doc/paper/typedlua.tex
rewrite of the introduction

@@
 Dynamically typed languages such as Lua avoid static types in favor of
-simplicity and flexibility, because the absence of static types means
-that programmers do not need to bother about abstracting types that
-should be validated by a type checker.
-Instead, dynamically typed languages use run-time \textit{type tags}
-to classify the values they compute, so their implementation can use
-these tags to perform run-time (or dynamic) type checking
-\citep{pierce2002tpl}.
+run-time {\em type tags} that classify the values they compute, and
+their implementations use these tags to perform run-time (or dynamic)
+checking and guarantee that only valid operations are
+performed~\citep{pierce2002tpl}.
 
-This simplicity and flexibility allow programmers to write code that
-might require a quite complex type system to statically type check,
-though it may also hide bugs that will be caught only after deployment
-if programmers do not properly test their code.
+The absence of static types means that programmers do not need to
+bother about abstracting types that might require a complex type
+system and type checker to validate, leading to simpler and more flexible
+languages and implementations. But this absence may also hide bugs
+that will be caught only after deployment if programmers do not properly
+test their code.
+
 In contrast, static type checking helps programmers detect many 
-bugs during the development phase.
-Static types also provide a conceptual framework that helps
-programmers define modules and interfaces that can be combined to
-structure the development of programs.
+bugs during the development phase. Static types also provide a
+conceptual framework that helps programmers define modules
+and interfaces that can be combined to structure the development
+of large programs.
 
-Thus, early error detection and better program structure are two
-advantages of static type checking that can lead programmers to
-migrate their code from a dynamically typed to a statically
-typed language, when their simple scripts become complex programs
-\citep{tobin-hochstadt2006ims}.
-Dynamically typed languages certainly help programmers during the
-beginning of a project, because their simplicity and flexibility
-allow quick development and make it easier to change code according to
-changing requirements.
-However, programmers tend to migrate from dynamically typed to
-statically typed code as soon as the project has consolidated its
-requirements, because the robustness of static types helps
-programmers link requirements to abstractions.
-This migration usually involves different languages that have
-different syntax and semantics, which usually require a complete
-rewrite of existing programs instead of incremental evolution from
-dynamic to static types.
+The early error detection and better program structure afforded by
+static type checking can lead programmers to migrate their code from
+a dynamically typed to a statically typed language, once their simple
+scripts become complex programs~\citep{tobin-hochstadt2006ims}.
+As this migration involves languages with different syntax and
+semantics, it requires a complete rewrite of existing programs intead
+of incremental evolution from dynamic to static types.
 
@@
 option to choose between static and dynamic typing:
-\textit{optional type systems} \citep{bracha2004pluggable} and
-\textit{gradual typing} \citep{siek2006gradual} are two similar
-approaches for blending static and dynamic typing in the same
-language.
-The aim of both approaches is to offer programmers the option
-to use type annotations where static typing is needed and avoid them
-where dynamic typing is sufficient.
-The difference between these two approaches is the way they treat
-run-time semantics.
-While optional type systems do not affect the run-time semantics,
+\textit{optional type systems}~\citep{bracha2004pluggable} and
+\textit{gradual typing}~\citep{siek2006gradual} are two approaches
+that offer programmers the option to use type annotations where static
+typing is needed, incrementally migrating a system from dynamic
+to static types. The difference between these two approaches is the
+way they treat run-time semantics: while optional type systems
+do not affect the run-time semantics,
 gradual typing uses run-time checks to ensure that dynamically typed
@@
 
-Programmers and researchers sometimes use the term gradual typing
-to mean the incremental evolution of dynamically typed code to
-statically typed code.
-For this reason, gradual typing may also refer to optional type
-systems and other approaches that blend static and dynamic typing to
-help programmers incrementally migrate from dynamic to static typing
-without having to switch to a different language, though all these
-approaches differ in the way they handle static and dynamic typing
-together.
-We use the term gradual typing for referring to the work of
-\citet{siek2006gradual}.
+In this work we outline the design of Typed Lua:
+an optional type system for Lua that is complex enough to
+preserve some of the idioms that Lua programmers are already used to,
+while adding new constructs that help programmers structure Lua
+programs.
 
-In this work we propose the design and implementation of Typed Lua:
-an optional type system for Lua that should be complex enough to
-preserve some of the Lua idioms that programmers are already used to,
-but that should also include new constructs that help programmers
-structure Lua programs.
+Lua is a small imperative scripting language that prefers to provide
+mechanisms instead of fixed policies for structuring programs,
+due to its history as an embedded extension language. Even features
+such as a module system and object orientation are a matter of 
+convention instead of built-in features.
 
-Lua is a small scripting language, so designing an optional type
-system for Lua should shed some light on optional type systems for
-scripting languages in general.
-Lua provides mechanisms for organizing a program in modules and
-writing object-oriented programs, but it does not set policy on how
-these features should behave, due to its use as an embedded and
-extension language.
-Thus, implementing an optional type system for Lua should offer Lua
-programmers one way to obtain most of the advantages of static typing
-without compromising the simplicity and flexibility of dynamic typing.
+The lack of standard policies
+is a challenge for the design of a static type system for the Lua
+language, and our approach was to extract the most common
+idioms from a large corpus of Lua libraries as a guide to what
+Typed Lua should focus on, and setting fixed policies for organizing
+a program in modules and using class-based object orientation.
 
-So far, Typed Lua should be a strict superset of Lua that
-provides optional type annotations, compile-time type checking, and
-class-based object-oriented programming through the definition of
-classes, interfaces, and modules.
-In practice, we should implement Typed Lua as a programming
-language that extends Lua syntax to add optional type annotations
-and standard constructions for structuring Lua code.
-The compiler should use static types to perform compile-time type
-checking, but should also allow Lua code to coexist with Typed Lua
-code, and generate Lua code that runs in unmodified Lua
-implementations.
-We have an ongoing implementation of the Typed Lua compiler that is
-available online\footnote{https://github.com/andremm/typedlua}.
+Typed Lua allows statically typed Lua code to coexist and interact with 
+dynamically typed code. The Typed Lua compiler warns the programmer
+about type errors, but always generates Lua code that runs in unmodified
+Lua implementations. The programmer can enjoy some of the benefits
+of static types even without converting existing Lua modules to typed Lua:
+a dynamically typed module can export a statically typed interface, and
+statically typed users of the module will have their use of the module
+checked by the compiler.
 
-Typed Lua intended use is as an application language, and we view
-that policies for organizing a program in modules and writing
-object-oriented programs should be part of the language and
-enforced by its optional type system.
-An application language is a programming language that helps
-programmers develop applications from scratch until these
-applications evolve to complex systems rather than just scripts.
+Unlike gradual type systems, Typed Lua does not insert run-time
+checks between dynamically and statically typed parts of the program.
+Unlike some optional type systems, the statically typed subset of
+Typed Lua is sound by design, so a later version of the Typed Lua
+compiler can swicth to gradual instead of just optional typing.
 
-We also believe that Typed Lua will help programmers give a more
-formal documentation to already existing Lua code, as static types
-are also a useful source of documentation in languages that provide
-type annotations, because type annotations are always validated by
-the type checker and therefore never get outdated.
-Thus, programmers can use Typed Lua to define axioms about the
-interfaces and types of dynamically typed modules.
-We will enforce this point by using Typed Lua to statically type
-the interface of the Lua standard libraries and other commonly used
-Lua libraries, so our compiler can check their usage by Typed Lua
-code.
-
-Typed Lua will perform a very limited form of local type inference
-\citep{pierce2000lti}, as static typing does not necessarily mean
-that programmers need to insert type annotations in the code.
-Several statically typed languages such as Haskell provide some
-amount of type inference that automatically deduces the types of
-expressions.
-Still, Typed Lua will need a small amount of type annotations due
-to the nature of its optional type system.
-
-Typed Lua will not deal with code optimization, although another
-important advantage of static types is that they help the compiler
-perform optimizations and generate more efficient code.
-However, we believe that the formalization of our optional type
-system will be precise enough to aid optimization in some Lua
-implementations.
-
-We will use some of the ideas of gradual typing to formalize Typed Lua.
-Even though Typed Lua is an optional type system and thus will not
-include run-time checks between dynamic and static regions of the
-code, we believe that using the foundations of gradual typing to
-formalize our optional type system will allow us to include run-time
-checks in the future.
-
-In Section \ref{sec:review} we review the literature about static and
-dynamic typing in the same language, which includes the discussion
-about the particularities between optional type systems and gradual
-typing.
-In Section \ref{sec:proposal} we describe the reasons why Typed Lua
-includes features of object-oriented programming and the status of
-its design and implementation.
-In Section \ref{sec:con} we outline our expected contributions.
+In Section~\ref{sec:proposal} we outline the key parts in
+the design of Typed Lua. In Section~\ref{sec:review} we review
+related work on mixing static and dynamic typing in the same language.
+Finally, Section~\ref{sec:con} gives some concluding remarks and
+our plans for the future of Typed Lua.
 
@@
 
-\section{Conclusion}
+\section{Conclusion} \label{sec:con}
 

### 2014-03-04 doc/paper/typedlua.tex
Cosmetic changes to the paper

--- 2014-02-26 doc/paper/typedlua.tex
@@
 warns the programmer about the insertion of these run-time checks,
-as they mean the existence of potential type errors.
+as they indicate the existence of potential type errors.
 However, the programmer is free to choose between inspecting the
@@
 
-BabyJ \citep{anderson2002babyj} is an object-oriented language
+BabyJ \citep{anderson2003babyj} is an object-oriented language
 without inheritance that allows programmers to incrementally annotate
@@
 checks.
-BabyJ \citep{anderson2002babyj} and Alore \citep{lehtosalo2011alore}
+BabyJ \citep{anderson2003babyj} and Alore \citep{lehtosalo2011alore}
 are two examples of object-oriented languages that use gradual
@@
 
-\subsection{Similar Lua projects}
+\subsection{Related Lua projects}
 
@@
 Like Typed Lua, Tidal Lock also has local type inference.
-Unlike Typed Lua, Tidal Lock is not typing the common object-oriented
+Unlike Typed Lua, Tidal Lock does not type the common object-oriented
 idioms used by Lua programmers.

### 2014-03-05 doc/paper/typedlua.tex
start of rewrite of section 2: prologue, and subsection on atomic types and functions

--- 2014-03-04 doc/paper/typedlua.tex
@@
 \newcommand{\Any}{\mathbf{any}}
+\newcommand{\Top}{\mathbf{value}}
 \newcommand{\Nil}{\mathbf{nil}}
@@
 
-We begin this section presenting statistics about the usage of Lua
-features and idioms.
-We collected statistics about how programmers use tables, functions,
-dynamic type checking, object-oriented programming, and modules.
-We shall see that these statistics informed important design decisions
-on our optional type system.
-We end this section introducing Typed Lua through examples, and
-describing a few Lua projects that are similar to Typed Lua.
+Lua is an imperative scripting language with first-class functions
+(with proper lexical scoping) where the main data structure is the
+{\em table}, an associative array that can play the part of arrays,
+records, maps, objects, etc., with syntactic sugar and
+metaprogramming through operator overloading built into
+the language. Unlike other scripting languages, Lua has very
+limited coercion among different data types.
+
+The primary use of Lua has always been as an embedded language
+for configuration and extension of other applications, and the result
+is a fragmented ecosystem of libraries, and different ideas among
+Lua programmers on how they should use the language features and
+how they should structure programs.
+
+The design of Typed Lua is informed by a (mostly automated) survey
+of Lua idioms used in a large corpus of Lua libraries, instead of
+relying just on the semantics of the language. We cover the main parts
+of the design, along with how they relate to Lua, in the rest of this section,
+and discuss some of what was left out in the current version of the design
+in the conclusion of the paper.
 
@@
 
-Our main goal is to allow programmers to combine static and dynamic
-typing in Lua, so we propose Typed Lua as an extension to Lua 5.2
-syntax that introduces optional type annotations, and class-based
-object-oriented programming through the definition of classes,
-interfaces, and modules.
-An important requirement of Typed Lua is that it should be backwards
-compatible with Lua, that is, it should allow programmers to run any
-Lua code without modifications.
-For this reason, these syntactic extensions include new keywords
-that are not reserved words.
+Lua values can have one of eight {\em tags}: {\em nil}, {\em boolean},
+{\em number}, {\em string}, {\em function}, {\em table}, {\em userdata},
+and {\em thread}. We will see how Typed Lua assigns types to
+values of the first five in this section.
 
-Figure \ref{fig:typelang} shows that Typed Lua includes first-class
-and second-class types.
-Typed Lua uses first-class types for representing Lua values and
-second-class types for defining the signatures of functions.
-First-class types include literal types, base types, the type $\Any$,
-union types, function types, table types, and recursive types.
-Second-class types include the type $\Void$, first-class types,
-vararg types, and tuple types.
-Literal types are types that represent a single value
-(e.g., $\Nil$, $\False$, $\True$, literal numbers, and
-literal strings).
-Literal types are subtypes of their respective base types.
-More precisely, the literals $\False$ and $\True$ are subtypes of
-$\Boolean$, any literal number $n$ is subtype of $\Number$, and any
-literal string $s$ is subtype of $\String$.
-The type $\Any$ is the dynamic type that allows dynamic code coexist
-with static code.
-We will discuss union types, function types, table types, recursive
-types, and second-class types along this section.
+Figure~\ref{fig:typelang} gives the abstract syntax of Typed Lua
+types. Only {\em first-class types} correspond to actual Lua
+values; {\em second-class types} correspond to expression lists,
+and Typed Lua uses them to type multiple assignment and function
+application. Types are ordered by a subtype relationship, where
+any first-class type is a subtype of $\Top$.
+
+The {\em dynamic type} $\Any$ allows dynamically typed code to
+interoperate with statically typed code; it is a subtype of $\Top$, but
+neither a supertype nor a subtype of any other type. We relate $\Any$
+to other types with the {\em consistency} and {\em consistent-subtype}
+relationships used by gradual type systems. In practice, we can
+pass a value of the dynamic type anytime we want a value of some
+other type, and can pass any value where a value of the dynamic type
+is expected, but these operations are tracked by the type system,
+and the programmer can choose to be warned about them.
 
@@
 \begin{array}{rlr}
-T ::= & & \textit{first-class types}\\
-& \;\; L & \textit{literal types}\\
+\multicolumn{3}{c}{\textbf{First-class types}} \\
+T ::= & \;\; L & \textit{literal types}\\
 & | \; B & \textit{base types}\\
+& | \; \Top & \textit{top type}\\
 & | \; \Any & \textit{dynamic type}\\
-& | \; T \cup T & \textit{union type}\\
-& | \; S \rightarrow S & \textit{function type}\\
-& | \; \{T:T, ..., T:T\} & \textit{table type}\\
-& | \; X & \textit{type variable}\\
-& | \; \mu X.T & \textit{recursive type}\\
-L ::= & \;\; \Nil \; | \; \False \; | \; \True \\
-& | \; {<}{\it number}{>} \; | \; {<}{\it string}{>} & \\
-B ::= & \Boolean \; | \; \Number \; | \; \String & \\
-S ::= & \Void \; | \; V & \textit{second-class types}\\
+& | \; T \cup T & \textit{disjoint union types}\\
+& | \; S \rightarrow S & \textit{function types}\\
+& | \; \{T:T, ..., T:T\} & \textit{table types}\\
+& | \; X & \textit{type variables}\\
+& | \; \mu X.T & \textit{recursive types}\\
+& | \; X_i & \textit{projection types}\\
+L ::= & \multicolumn{2}{l}{\False \; | \; \True \; | \; {<}{\it number}{>} \; | \; {<}{\it string}{>} } \\
+B ::= & \multicolumn{2}{l}{\Nil \; | \; \Boolean \; | \; \Number \; | \; \String} \\
+\multicolumn{3}{c}{} \\
+\multicolumn{3}{c}{\textbf{Second-class types}} \\
+S ::= &  \multicolumn{2}{l}{\Void \; |\; V \; | \; S \cup S} \\
 V ::= & \;\; T & \\
-& | \; T* & \textit{vararg type}\\
-& | \; T \times V & \textit{tuple type}
+& | \; T* & \textit{vararg types}\\
+& | \; T \times V & \textit{tuple types}
 \end{array}
@@
 \dend
-\caption{Typed Lua types}
+\caption{Abstract syntax of Typed Lua types}
 \label{fig:typelang}
@@
 
-Typed Lua allows optional type annotations in variable and function
-declarations.
-Whenever it is possible to use local type inference, the compiler
-assigns a more specific type to an unannotated declaration.
-When it is not possible to use local type inference, the compiler
-assigns the type $\Any$ to an unannotated declaration.
-For instance, in function declarations, the compiler does not infer
-a more specific type to unannotated input parameters, because,
-in this case, global type inference often is undecidable in a
-language with subtyping and overloading \citep{wells1999typability}.
+Typed Lua allows optional type annotations in variable function
+declarations. It assigns the dynamic type to any parameter that
+does not have an type annotation, but assigns more precise
+types to unnanotated variables, based on the type of the
+expression that gives the initial value of the variable.
 
-In the following example we use type annotations in a function
-declaration, and we do not use type annotations in a declaration of a
+In the following example, we use type annotations in a function
+declaration but do not use type annotations in the declaration of a
 local variable:
 \begin{verbatim}
-    local function fact (n:number):number
+    local function factorial(n: number): number
       if n == 0 then
@@
       else
-        return n * fact(n - 1)
+        return n * factorial(n - 1)
       end
@@
     local x = 5
-    print(fact(x))
+    print(factorial(x))
 \end{verbatim}
 
-Even though we did not add type annotations in the declaration of
-the local variable \texttt{x}, the compiler uses local type inference to
-assign the type $\Number$ to \texttt{x}.
-The compiler does not use the literal type $5$ instead of the base
-type $\Number$ because this would prevent
-programmers from assigning other numeric values to the variable
-\texttt{x}.
-The inference that we implement in Typed Lua is quite simple, as it
-uses only the type of the local expression.
-
-Typed Lua allows programmers to combine annotated code with
+The compiler assigns the type $\Number$ to the local variable \texttt{x},
+and this example compiles without warnings. Typed Lua allows
+programmers to combine annotated code with
 unannotated code, as we show in the following example:
 \begin{verbatim}
-    local function abs (n:number):number
+    local function abs(n: number)
       if n < 0 then
@@
 
-    local function dist (x, y)
+    local function distance(x, y)
       return abs(x - y)
@@
 
-The function \texttt{dist} receives two parameters of type $\Any$
-and returns a value of type $\Number$.
 The compiler assigns the dynamic type $\Any$ to the input
-parameters of \texttt{dist} because they do not have type annotations,
-and the compiler does not use global type inference, as we mentioned
-previously.
-The compiler infers the return type of \texttt{dist} because it is
-not a recursive function, that is, Typed Lua does not infer the return
-type of recursive functions.
+parameters of \texttt{distance} because they do not have type annotations.
+Subtracting a value of type $\Any$ from another also yields a value of
+type $\Any$ (Lua has operator overloading, so the minus operation
+is not guaranteed to return a number in this case), but
+consistent subtyping lets us pass a value of type $\Any$ to a
+function that expects a $\Number$.
 
-We are using the consistent-subtyping relation of gradual
-typing to check the interaction among the dynamic type and the other
-types.
-More precisely, we can call \texttt{abs} inside \texttt{dist} because
-\texttt{x - y} is of type $\Any$, and the type $\Any$ is
-consistent with the type $\Number$.
-The type $\Any$ is not the union of all types, but it is a separate
-type.
+Even though the return types
+of both {\tt abs} and {\tt distance} are not given, the compiler is able to
+infer a return type of $\Number$ to both functions, as they are
+local and not recursive.
 
-Unlike gradual typing, we do not insert run-time checks that
-inspect the interaction between dynamically typed code and statically
-typed code during run-time.
-In other words, Typed Lua does not guarantee that dynamically typed
-code does not violate statically typed code.
-Typed Lua does not insert run-time checks because they can decrease
-run-time performance \citep{allende2013cis}.
-We believe that a careful evaluation of run-time checks should be done
-before inserting them on the type system.
-In the previous example, we cannot guarantee that \texttt{dist} is never
-going to call \texttt{abs} with a parameter that is not a number,
-because in the semantics of Lua the minus operator can result in a
-value that is not a number.
-Still, we can implement run-time checks in the future because we are
-using the consistent-subtyping relation of gradual typing to
-formalize our optional type system, and the type system is being
-designed to be sound.
+Lua has first-class functions, but they have some peculiarities. First,
+the number of arguments passed to a function does not need to
+match the function's arity; Lua silently drops extra arguments after
+evaluating them, or passes {\tt nil} in place of any missing arguments.
+Second, functions can return any number of values, and the number
+of values returned may not be statically known. Third, Lua also has
+multiple assignment, and the semantics of argument passing match
+those of multiple assignment (or vice-versa); calling a function is like
+doing a multiple assigment where the left side is the parameter list
+and the right side is the argument list.
 
-Typed Lua includes function types because functions are first-class
-values in Lua.
-The syntax for defining a function type is \texttt{(s) -> (s)},
-where \texttt{s} is a second-class type.
+Typed Lua uses {\em second-class types} to encode the peculiarities
+of argument passing, multiple returns, and multiple assignment. We call
+them second-class because these types do not correspond to actual
+values and cannot be assigned to variables or parameters: they are an
+artifact of the interaction between the type system and the semantics of Lua.
+
 As we saw in Figure \ref{fig:typelang}, a second-class type in
-Typed Lua can be the type $\Void$, a first-class type, a variadic
-type, or a tuple type.
-We use tuple types to represent multiple parameters, multiple return
-values, and multiple assignments, and we use the variadic type to
-represent variadic functions and the vararg expression.
-Even though we already included the tuple type and the vararg type,
-we are still working on these types.
+Typed Lua can be the type $\Void$ (the empty tuple),
+a tuple of first-class types optionally ending in a variadic type,
+or a union of these tuples. A variadic type $T*$ is a generator for a
+sequence of values of type $T \cup \Nil$. Unions of tuples play
+an important part in functions that are overloaded on the return type,
+together with {\em projection types}. Both are explained in the next
+section.
 
-The next example shows how we can annotate a \texttt{map} function
-that receives another function as its parameter and uses this
-function to alter the elements of a list of numbers, which is
-also a parameter of \texttt{map}:
+Typed Lua always adds a
+variadic tail to the parts of a function type if none is specified, to match
+the semantics of Lua function calls. In the examples above, the types
+of {\tt factorial} and {\tt abs} are actually $\Number \times \Top *
+ \rightarrow \Number \times \Nil *$, and the type of {\tt distance} is
+$\Any \times \Any \times \Top * \rightarrow \Number \times \Nil *$.
+
+If we call {\tt abs} with extra arguments, Typed Lua silently ignores
+them, as the type signature lets {\tt abs} receive any number of extra
+arguments. If we call {\tt abs} in the right side of an assignment to
+more than one lvalue, Typed Lua checks if the first lvalue has a type
+consistent with $\Number$, and any other lvalues need to have a type
+consistent with $\Nil$.
+
+A variadic type can only appear in the tail position of a tuple,
+because Lua takes only the first value of any expression that appears
+in an expression list that is not in tail position. The following example
+shows the interaction between multiple returns and expression lists:
+
 \begin{verbatim}
-    local function map (f:(number)->(number),
-                        l:{number}):void
-      for i=1, #l do
-        l[i] = f(l[i] or 0)
+    local function multiple()
+      return 2, "foo"
     end
+
+    local function sum(x: number, y: number)
+      return x + y
     end
+
+    local x, y, z = multiple(), multiple()
+    print(sum(multiple(), multiple())
 \end{verbatim}
+
+Function {\tt multiple} is
+$\Top * \rightarrow \Number \times \String \times \Nil *$,
+and {\tt sum} is $\Number \times \Number \times \Top * \rightarrow
+\Number \times \Nil *$. In the right side of the multiple assignment,
+only the first value produced by the first call to multiple gets used, so
+the type of the right side is $\Number \times \Number \times \String \times \Nil*$,
+and the types assigned to {\tt x}, {\tt y}, and {\tt z} are respectively $\Number$,
+$\Number$, and $\String$. This also means that the call to {\tt sum} compiles
+without errors, as the first two components of the tuple are consistent with
+the types of the parameters, and the other components are consistent with
+$\Top$.
 

### 2014-03-06 doc/paper/typedlua.tex
rewrote section on union types

@@
 
-Typed Lua includes union types as a way to discriminate data
-structures that can hold values of several different, but fixed types.
-Next, we show an example:
-\begin{verbatim}
-    local x:number?            -- OK
-    local y:number|string = 1  -- OK
-    local z:number = x         -- Error
-\end{verbatim}
-
-The annotation \texttt{number?} is syntactic sugar for
-\texttt{number|nil} that means the type $\Number\;\cup\;\Nil$.
-Typed Lua includes the syntactic sugar \texttt{t?} because the
-type $t\;\cup\;\Nil$ is very common in Lua code.
-
-The last line of the previous example gives a type error because we
-are attempting to assign a value of type $\Number\;\cup\;\Nil$
-to a variable that accepts only values of type $\Number$.
-Although we can solve this type error changing the annotation of
-variable \texttt{z} from \texttt{number} to \texttt{number?}, there is
-another way that we can use to solve this type error without changing
-the annotation:
-\begin{verbatim}
-    local z:number = x or 0  -- OK
-\end{verbatim}
-
-Typed Lua has a typing rule for checking that
-$t\;\cup\;\Nil\;\mathbf{or}\;t$ always has the type $t$.
+Typed Lua uses union types to encode some common Lua idioms:
+optional values, overloading based on the tags of input parameters,
+and overloading on the return type of the functions.
+
+Optional values are unions of some type and $\Nil$, and are so
+common that Typed Lua uses the {\tt t?} syntax for these unions.
+They appear any time a function has optional parameters, and
+any time the program reads a value from an array or map.
+
+\begin{verbatim}
+    local function message(name: string,
+                           greeting: string?)
+      local greeting = greeting or "Hello "
+      return greeting .. name
+    end
+
+    print(message("Lua"))
+    print(message("Lua", "Hi"))
+\end{verbatim}
+
+In this example, the second parameter is optional but, in the first
+line of the function, we declare a new variable that is guaranteed to
+have type $\String$ instead of $\String \cup \Nil$. In Lua, any value
+except {\tt nil} and {\tt false} are "truthy", so the short-circuiting
+{\tt or} operator is a common way of giving a default value to an
+optional parameter. Typed Lua encodes this idiom with a typing rule: if
+the left side of {\tt or} has type $T \cup Nil$ and the right side
+has type $T$ then the {\tt or} expression has type $T$.
+
+Declaring a new {\tt greeting} variable that shadows the parameter
+is not necessary:
+
+\begin{verbatim}
+    local function message(name: string, 
+                           greeting: string?)
+      greeting = greeting or "Hello "
+      return greeting .. name
+    end
+\end{verbatim}
+
+Typed Lua lets assignment change the type of a local variable in
+cases that respect a simple heuristic: the new type must be a subtype
+of the previous one, and only applies for the remainder of the
+current scope.
+
+Overloaded functions use the {\tt type} function to inspect
+the tag of their parameters, and perform different actions depending
+on what those tags are. The simplest case overloads on just a
+single parameter:
+
+\begin{verbatim}
+    local function overload(s1: string, 
+                            s2: string|number)
+      if type(s2) == "string" then
+        return s1 .. s2
+      else
+        -- string.rep: (string, number) -> string
+        return string.rep(s1, s2)
+      end
+    end
+\end{verbatim}
+
+Typed Lua has a small set of type predicates that, when used
+over a local variable in a condition, constrain the type of that
+variable. The function above uses the {\tt type(X) == "string"}
+predicate which constrains the type of $X$ from $T \cup \String$ to
+$\String$ when the predicate is true and $T$ otherwise. This is
+a simplified form of {\em flow typing}~\cite{guha:flow}.
+
+The type predicates can only discriminate based on tags, so they
+are limited on the kinds of unions that they can discriminate. It
+is possible to discriminate a union that combines a table type with
+a base type, or a table type with a function type, or a two base types,
+but it is not possible to discriminate between two different function
+types.
+
+Functions that overload their return types to signal the ocurrence
+of errors are another common Lua idiom. In this idiom, a function
+returns its normal set of return values in case of success but,
+if anything fails, returns {\tt nil} as the first value, followed
+by an error message or other data describing the error, as in the
+following example:
+
+\begin{verbatim}
+    local function idiv(d1: number, d2: number):
+          (number, number)|(nil, string)
+      if d2 == 0 then
+        return nil, "division by zero"
+      else
+        local r = d1 % d2
+        local q = (d1 - r)/d2
+        return q, r
+      end
+    end
+\end{verbatim}
+
+There is also special syntax for this idiom: we could
+annotate the return type of {\tt idiv} with {\tt (number, number)?}
+to denote the same union\footnote{The parentheses are always
+necessary here: {\tt number?} is {\tt number|nil}, while {\tt (number)}
+is {\tt (number)|(nil, string)}.}.
+
+The full type of {\tt idiv} is $\Number \times \Number \times
+\Top * \rightarrow (\Number \times \Number \times \Nil *) \cup
+(\Nil \times \String \times \Nil *)$. A typical client of this
+function would use it as follows:
+
+\begin{verbatim}
+    local q, r = idiv(n1, n2)
+    -- q is number|nil, r is number|string
+    if q then
+      -- q and r are numbers
+    else
+      -- r is a string
+    end
+\end{verbatim}
+
+When Typed Lua encounters a union of tuples in the right side
+of an declaration, it stores the the union in a special type
+environment with a fresh name and assigns {\em projection types} 
+to the variables in the left side of the declaration. If the type
+variable is $X$, variable {\tt q} gets type $X_1$ and variable
+{\tt r} gets type $X_2$.
+
+If we need to check a projection type
+against some other type, we take the union of the corresponding
+component in each tuple. But if we a variable with a projection
+type appears in a type predicate, the predicate discriminates
+against all tuples in the union. In the example above,
+$X$ is $(\Number \times \Number \times \Nil *) \cup
+(\Nil \times \String \times \Nil *)$ outside of the {\tt if}
+statement, but $\Number \times \Number \times \Nil *$ in the
+{\tt then} block and $\Nil \times \String \times \Nil *$ in
+the {\tt else} block.
+
+Notice that we could also discriminate using {\tt r}, using
+{\tt type(r) == "number"} as our predicate, with the same
+result. The first form is more succint, and more idiomatic.
+We can also use projection types to write overloaded functions
+where the type of a parameter depends on the type of another
+parameter.
+
+Currently, a limitation of our overloading mechanisms is that
+the return type cannot depend on the input types; we cannot
+write a function that is guaranteed to return a number if
+passed a number and guaranteed to return a string if passed
+a string, for example.
 

### 2014-03-10 doc/paper/typedlua.tex
correction on projection types; most of the rewrite of section 2.3 done

@@
 
+Assigning to a variable with a projection type is forbidden,
+unless the union has been discriminated down to a single tuple,
+Unrestricted assignment to these variables would be unsound,
+as it could break the dependency relation between the types
+in each tuple that is part of the union.
+
 Currently, a limitation of our overloading mechanisms is that
@@
 
-Typed Lua includes table types to represent Lua's tables.
-Although Typed Lua has just one table type constructor, it has three
-different syntaxes for defining the type of a table.
-The first syntax defines table types for hash tables;
-it is written \texttt{\{t1:t2\}},
-and maps to the table type $\{t_{1}:t_{2}\;\cup\;\Nil\}$.
-This table type represents a hash table that maps values of type
-$t_{1}$ to values of type $t_{2}\;\cup\;\Nil$, as Lua returns
-$\Nil$ when we use a non-existing key to index a table.
-The second syntax defines table types for lists;
-it is written \texttt{\{t\}},
-and maps to the table type $\{\Number:t\;\cup\;\Nil\}$.
-This table type is just a syntactic sugar to the definition of a
-hash table that maps values of type $\Number$ to values of type
-$t\;\cup\;\Nil$.
-The third syntax defines table types for records;
-it is written \texttt{\{s1:t1, ..., sn:tn\}},
-and maps to the table type $\{s_{1}:t_{1}, ..., s_{n}:t_{n}\}$.
-This table type represents a record that maps the literals
-$s_{i}$\texttt{, ..., }$s_{n}$ to values of types
-$t_{i}$\texttt{, ..., }$t_{n}$, where $i$ may vary from one to $n$.
-
-Next, we show one example of table type to type a hash table:
-\begin{verbatim}
-    local t:{string:number} = { foo = 1 }  -- OK
-    local x:number = t.foo                 -- Error
-    local y:number? = t.bar                -- OK
-\end{verbatim}
-
-The second line of this example gives a type error for the same
-reason that the last line of the previous example gives a type error
-too, that is, we are trying to assign a value that can be of two
-different types to a variable that accepts values of only one type.
-Although the field \texttt{bar} does not exist in \texttt{t}, the last
-line of this example does not give a type error because the
-annotated type matches the type of the values that can be stored in
-\texttt{t}.
-
-Next, we show one example of table type to type a list:
-\begin{verbatim}
-    local days:{string} = { "Sunday", "Monday",
+Tables are the main mechanism that Lua has to build data
+structures. They are associative arrays where any value
+(except {\tt nil}) can be a key, but with language support
+for efficiently using tables as tuples, arrays (dense or sparse),
+records, modules, and objects. In this section,
+we show how Typed Lua encodes tables as arrays, records, tuples,
+and plain maps in its type system.
+
+Typed Lua uses the same framework to represent the different
+uses that a Lua table has: {\em table types}. A table type
+$\{ t_{1}:u_{1}, \ldots, t_{n}:u_{n}\}$ represents a map
+from values of type $t_i$ to values of type $u_i$.
+
+The concrete syntax of Typed Lua has syntax for
+common table types. One syntax defines table types for
+maps: it is written \texttt{\{ t: u \}},
+and maps to the table type $\{t:u\}$.
+This table type represents a map that maps values of type
+$t$ to values of type $u \cup \Nil$.
+Another syntax defines table types for arrays:
+it is written \texttt{\{ t \}}, and maps to the table type
+$\{\Number:t\}$. A third syntax sugar defines
+table types for records:
+it is written \texttt{\{ s1: t1, ..., sn: tn \}}, where
+each $s_i$ is a literal number, string, or boolean, 
+and maps to the table type $\{s_{1}:t_{1}, ..., s_{n}:t_{n}\}$,
+where each $s_i$ is the corresponding literal type.
+
+The example below shows how we can define a map from
+strings to numbers; the dot syntax for field access
+in Lua is actually syntactic sugar for indexing a table
+with a string literal:
+
+\begin{verbatim}
+    local t: { string: number } = { foo = 1 }
+    local x: number = t.foo         -- x gets 1
+    local y: number = t["bar"]      -- runtime error
+\end{verbatim}
+
+When accessing a map, there is always the possibility that
+the key is not there. In Lua, accessing a non-existing key
+returns {\tt nil}. Typed Lua is stricter, and raises a runtime
+error in this case. To get a map with the behavior of standard
+Lua tables, the programmer can use an union:
+
+\begin{verbatim}
+    local t: { string: number? } = { foo = 1 }
+    local x: number = t.foo         -- error
+    local y: number = t.bar or 0    -- y gets 0
+    local z: number? = t["bar"]     -- z gets nil
+\end{verbatim}
+
+Now the Typed Lua compiler will complain about the assignment
+on line two. The following example shows how we can declare
+an array:
+\begin{verbatim}
+    local days: { string } = { "Sunday", "Monday",
       "Tuesday", "Wednesday", "Thursday",
-      "Friday", "Saturday" }  -- OK
-\end{verbatim}
-
-When we know that a list has fixed elements, we can leave the
-variable declaration unannotated and let local type inference assign
-a more specific table type to the list.
-If we remove the annotation in this example, the compiler uses the
-syntax of records to infer the following table type to \texttt{days}:
+      "Friday", "Saturday" }
+    local x = days[1]          -- x gets "Sunday"
+    local y = days[8]          -- runtime error
+\end{verbatim}
+
+Notice that we have the same strictness with missing
+elements, unless the type of the elements has {\tt nil}
+as a possible value.
+
+If we want to declare a tuple, we can leave the
+variable declaration unannotated and let Typed Lua assign
+a more specific table type to the variable.
+If we remove the annotation in the previous example, 
+the compiler assigns the following table type to \texttt{days}:
 \begin{align*}
@@
 \end{align*}
-This inferred table type is a subtype of the type
-$\{\Number:\String\;\cup\;\Nil\}$, which is the type that we used to
-annotate the variable \texttt{days}.
-
-When we use the inferred table type while initializing a table, the
-compiler gives an error message when we try to access an index that
-is out of bounds.
-
-It also worth mentioning that we can use the style of the inferred
-type to define heterogeneous tuples.
-
-Next, we show one example of table type to type a record:
-\begin{verbatim}
-    local person:{"firstname":string,
-                  "lastname":string} =
+
+This type is not a subtype of $\{\Number:\String\}$, nor
+is $\{\Number:\String\}$ a subtype of $\{\Number:
+\String\cup\Nil\}$, because in both cases the subtype
+relationship would be unsound. In the first case,
+a table type with the same fields as the type above,
+plus $8: \Number$, is a subtype of the table type
+above, so would also be a subtype of $\{\Number:\String\}$,
+which is clearly unsound. In the second case, covariance
+in the type of mutable fields is also unsound, for the
+same reason as the unsoundness of array covariance.
+
+While the record type above, $\{\Number:\String\}$, and
+$\{\Number: \String\cup\Nil\}$ are disjoint, all three
+types are valid for the table constructed in the example.
+Typed Lua actually assigns different types to a table constructor
+expression depending on the context where it is used.
+
+Finally, the next example shows how we can declare a record:
+
+\begin{verbatim}
+    local person: { "firstname": string,
+                    "lastname": string } =
       { firstname = "Lou", lastname = "Reed" } 
@@
 
-Type inference would infer the very same table type that we used to
-annotate the variable \texttt{person}.
-
-Typed Lua includes interfaces as syntactic sugar to table types
-that specify records.
-For instance, we can use the following interface as an alias to the
-type of the record we showed in the previous example:
-\begin{verbatim}
-    interface Person
-      firstname:string
-      lastname:string
-    end
-\end{verbatim}
-
-Now we can use the type \texttt{Person} as an alias to a table type
-that certainly has the fields \texttt{firstname} and \texttt{lastname}.
-Next, we show an example of code that uses the type \texttt{Person}:
-\begin{verbatim}
-    local function greeter (person:Person):string
-      return "Hello " .. person.firstname ..
+We could leave the type annotation out, and Typed Lua would
+assign the same type to {\tt person}.
+
+As records get bigger, and types of record fields get more
+complicated, writing table types can be unwieldy, so Typed
+Lua has {\em interfaces} as syntactic sugar for record types:
+
+\begin{verbatim}
+    local interface Person
+      firstname: string
+      lastname: string
+    end
+\end{verbatim}
+
+The declaration above declares {\tt Person} as an alias to
+the record type $\{$``firstname": $\String$,
+ ``lastname": $\String\}$ in
+the remainder of the current scope. We can now use {\tt Person} in
+type declarations:
+
+\begin{verbatim}
+    local function greet(person: Person)
+      return "Hello, " .. person.firstname ..
              " " .. person.lastname
@@
     local user2 = { firstname = "Lou" }
-
     local user3 = { lastname = "Reed",
@@
 
-    print(greeter(user1)) -- OK
+    print(greeter(user1)) -- Hello, Lewis Reed
     print(greeter(user2)) -- Error
-    print(greeter(user3)) -- OK
+    print(greeter(user3)) -- Hello, Lou Reed
     print(greeter(user4)) -- Error
@@
 
-This example shows that our optional type system is structural rather
-than nominal, that is, it checks the structure of types instead of
-their names, and it uses subtyping and consistent-subtyping for
-checking types.
-
-Using interfaces we can also define recursive types such as linked
-lists:
-
-\begin{verbatim}
-    interface Element
-      info:number
-      next:Element?
-    end
-\end{verbatim}
-
-We need to explicitly annotate \texttt{next} being of type
-\texttt{Element} $\cup\;\Nil$ because in our type system $\Nil$
-is not the bottom type, that is, $\Nil$ is subtype of itself only.
-Typed Lua has this behavior because in Lua $\Nil$ represents
-the absence of an useful value.
+If our record type has fields that can be {\tt nil}, we need
+to use an explicit type declaration when declaring a
+variable of this record type, as the following example shows:
+
+\begin{verbatim}
+    local interface Person
+      firstname: string
+      middlename: string?
+      lastname: string
+    end
+
+    local user1: Person = { firstname = "Lewis",
+                            middlename = "Allan",
+                            lastname = "Reed" }
+    local user2: Person = { lastname = "Reed",
+                            firstname = "Lou" }
+\end{verbatim}
+
+We need an explicit type declaration because neither of
+the types that the Typed Lua compiler assigns to the
+table constructors above is a subtype of $\{$``firstname":
+$\String$, ``middlename": $\String \cup \Nil$, ``lastname":
+$\String\}$, the type that {\tt Person} describes.
+
+We can also use interfaces to define recursive types:
+
+\begin{verbatim}
+    local interface Element
+      info: number
+      next: Element?
+    end
+\end{verbatim}
+
+It is common in Lua programs to build a record incrementally,
+starting with an empty table, as in the following example:
+
+\begin{verbatim}
+    local person = {}
+    person.firstname = "Lou"
+    person.lastname = "Reed"
+\end{verbatim}
+
+Ideally, we want the type of {\tt person} to change as the
+table gets built, from $\{\}$ to $\{$``firstname": $\String\}$
+and finally to $\{$``firstname": $\String$, ``lastname":
+ $\String\}$. This is tricker than the type change introduced
+by assignment that we saw in Section 2.2, as what is changing
+is not just the type of the variable {\tt person} but the
+type of the value that {\tt person} points to.
 

### 2014-03-10 doc/paper/typedlua.tex
Updating expected contributions

--- 2014-03-05 doc/paper/typedlua.tex
@@
 \newcommand{\Any}{\mathbf{any}}
-\newcommand{\Top}{\mathbf{value}}
 \newcommand{\Nil}{\mathbf{nil}}
@@
 Dynamically typed languages such as Lua avoid static types in favor of
-run-time {\em type tags} that classify the values they compute, and
-their implementations use these tags to perform run-time (or dynamic)
-checking and guarantee that only valid operations are
-performed~\citep{pierce2002tpl}.
-
-The absence of static types means that programmers do not need to
-bother about abstracting types that might require a complex type
-system and type checker to validate, leading to simpler and more flexible
-languages and implementations. But this absence may also hide bugs
-that will be caught only after deployment if programmers do not properly
-test their code.
-
+simplicity and flexibility, because the absence of static types means
+that programmers do not need to bother about abstracting types that
+should be validated by a type checker.
+Instead, dynamically typed languages use run-time \textit{type tags}
+to classify the values they compute, so their implementation can use
+these tags to perform run-time (or dynamic) type checking
+\citep{pierce2002tpl}.
+
+This simplicity and flexibility allow programmers to write code that
+might require a quite complex type system to statically type check,
+though it may also hide bugs that will be caught only after deployment
+if programmers do not properly test their code.
 In contrast, static type checking helps programmers detect many 
-bugs during the development phase. Static types also provide a
-conceptual framework that helps programmers define modules
-and interfaces that can be combined to structure the development
-of large programs.
-
-The early error detection and better program structure afforded by
-static type checking can lead programmers to migrate their code from
-a dynamically typed to a statically typed language, once their simple
-scripts become complex programs~\citep{tobin-hochstadt2006ims}.
-As this migration involves languages with different syntax and
-semantics, it requires a complete rewrite of existing programs intead
-of incremental evolution from dynamic to static types.
+bugs during the development phase.
+Static types also provide a conceptual framework that helps
+programmers define modules and interfaces that can be combined to
+structure the development of programs.
+
+Thus, early error detection and better program structure are two
+advantages of static type checking that can lead programmers to
+migrate their code from a dynamically typed to a statically
+typed language, when their simple scripts become complex programs
+\citep{tobin-hochstadt2006ims}.
+Dynamically typed languages certainly help programmers during the
+beginning of a project, because their simplicity and flexibility
+allow quick development and make it easier to change code according to
+changing requirements.
+However, programmers tend to migrate from dynamically typed to
+statically typed code as soon as the project has consolidated its
+requirements, because the robustness of static types helps
+programmers link requirements to abstractions.
+This migration usually involves different languages that have
+different syntax and semantics, which usually require a complete
+rewrite of existing programs instead of incremental evolution from
+dynamic to static types.
 
@@
 option to choose between static and dynamic typing:
-\textit{optional type systems}~\citep{bracha2004pluggable} and
-\textit{gradual typing}~\citep{siek2006gradual} are two approaches
-that offer programmers the option to use type annotations where static
-typing is needed, incrementally migrating a system from dynamic
-to static types. The difference between these two approaches is the
-way they treat run-time semantics: while optional type systems
-do not affect the run-time semantics,
+\textit{optional type systems} \citep{bracha2004pluggable} and
+\textit{gradual typing} \citep{siek2006gradual} are two similar
+approaches for blending static and dynamic typing in the same
+language.
+The aim of both approaches is to offer programmers the option
+to use type annotations where static typing is needed and avoid them
+where dynamic typing is sufficient.
+The difference between these two approaches is the way they treat
+run-time semantics.
+While optional type systems do not affect the run-time semantics,
 gradual typing uses run-time checks to ensure that dynamically typed
@@
 
-In this work we outline the design of Typed Lua:
-an optional type system for Lua that is complex enough to
-preserve some of the idioms that Lua programmers are already used to,
-while adding new constructs that help programmers structure Lua
-programs.
-
-Lua is a small imperative scripting language that prefers to provide
-mechanisms instead of fixed policies for structuring programs,
-due to its history as an embedded extension language. Even features
-such as a module system and object orientation are a matter of 
-convention instead of built-in features.
-
-The lack of standard policies
-is a challenge for the design of a static type system for the Lua
-language, and our approach was to extract the most common
-idioms from a large corpus of Lua libraries as a guide to what
-Typed Lua should focus on, and setting fixed policies for organizing
-a program in modules and using class-based object orientation.
-
-Typed Lua allows statically typed Lua code to coexist and interact with 
-dynamically typed code. The Typed Lua compiler warns the programmer
-about type errors, but always generates Lua code that runs in unmodified
-Lua implementations. The programmer can enjoy some of the benefits
-of static types even without converting existing Lua modules to typed Lua:
-a dynamically typed module can export a statically typed interface, and
-statically typed users of the module will have their use of the module
-checked by the compiler.
-
-Unlike gradual type systems, Typed Lua does not insert run-time
-checks between dynamically and statically typed parts of the program.
-Unlike some optional type systems, the statically typed subset of
-Typed Lua is sound by design, so a later version of the Typed Lua
-compiler can swicth to gradual instead of just optional typing.
-
-In Section~\ref{sec:proposal} we outline the key parts in
-the design of Typed Lua. In Section~\ref{sec:review} we review
-related work on mixing static and dynamic typing in the same language.
-Finally, Section~\ref{sec:con} gives some concluding remarks and
-our plans for the future of Typed Lua.
+Programmers and researchers sometimes use the term gradual typing
+to mean the incremental evolution of dynamically typed code to
+statically typed code.
+For this reason, gradual typing may also refer to optional type
+systems and other approaches that blend static and dynamic typing to
+help programmers incrementally migrate from dynamic to static typing
+without having to switch to a different language, though all these
+approaches differ in the way they handle static and dynamic typing
+together.
+We use the term gradual typing for referring to the work of
+\citet{siek2006gradual}.
+
+In this work we propose the design and implementation of Typed Lua:
+an optional type system for Lua that should be complex enough to
+preserve some of the Lua idioms that programmers are already used to,
+but that should also include new constructs that help programmers
+structure Lua programs.
+
+Lua is a small scripting language, so designing an optional type
+system for Lua should shed some light on optional type systems for
+scripting languages in general.
+Lua provides mechanisms for organizing a program in modules and
+writing object-oriented programs, but it does not set policy on how
+these features should behave, due to its use as an embedded and
+extension language.
+Thus, implementing an optional type system for Lua should offer Lua
+programmers one way to obtain most of the advantages of static typing
+without compromising the simplicity and flexibility of dynamic typing.
+
+So far, Typed Lua should be a strict superset of Lua that
+provides optional type annotations, compile-time type checking, and
+class-based object-oriented programming through the definition of
+classes, interfaces, and modules.
+In practice, we should implement Typed Lua as a programming
+language that extends Lua syntax to add optional type annotations
+and standard constructions for structuring Lua code.
+The compiler should use static types to perform compile-time type
+checking, but should also allow Lua code to coexist with Typed Lua
+code, and generate Lua code that runs in unmodified Lua
+implementations.
+We have an ongoing implementation of the Typed Lua compiler that is
+available online\footnote{https://github.com/andremm/typedlua}.
+
+Typed Lua intended use is as an application language, and we view
+that policies for organizing a program in modules and writing
+object-oriented programs should be part of the language and
+enforced by its optional type system.
+An application language is a programming language that helps
+programmers develop applications from scratch until these
+applications evolve to complex systems rather than just scripts.
+
+We also believe that Typed Lua will help programmers give a more
+formal documentation to already existing Lua code, as static types
+are also a useful source of documentation in languages that provide
+type annotations, because type annotations are always validated by
+the type checker and therefore never get outdated.
+Thus, programmers can use Typed Lua to define axioms about the
+interfaces and types of dynamically typed modules.
+We will enforce this point by using Typed Lua to statically type
+the interface of the Lua standard libraries and other commonly used
+Lua libraries, so our compiler can check their usage by Typed Lua
+code.
+
+Typed Lua will perform a very limited form of local type inference
+\citep{pierce2000lti}, as static typing does not necessarily mean
+that programmers need to insert type annotations in the code.
+Several statically typed languages such as Haskell provide some
+amount of type inference that automatically deduces the types of
+expressions.
+Still, Typed Lua will need a small amount of type annotations due
+to the nature of its optional type system.
+
+Typed Lua will not deal with code optimization, although another
+important advantage of static types is that they help the compiler
+perform optimizations and generate more efficient code.
+However, we believe that the formalization of our optional type
+system will be precise enough to aid optimization in some Lua
+implementations.
+
+We will use some of the ideas of gradual typing to formalize Typed Lua.
+Even though Typed Lua is an optional type system and thus will not
+include run-time checks between dynamic and static regions of the
+code, we believe that using the foundations of gradual typing to
+formalize our optional type system will allow us to include run-time
+checks in the future.
+
+In Section \ref{sec:review} we review the literature about static and
+dynamic typing in the same language, which includes the discussion
+about the particularities between optional type systems and gradual
+typing.
+In Section \ref{sec:proposal} we describe the reasons why Typed Lua
+includes features of object-oriented programming and the status of
+its design and implementation.
+In Section \ref{sec:con} we outline our expected contributions.
+
+\section{Blending static and dynamic typing} \label{sec:review}
+
+We begin this section presenting a little bit of the history about
+combining static and dynamic typing in the same language.
+After that, we introduce optional type systems and gradual typing.
+We end this section discussing why optional type systems and two
+other approaches are often called gradual typing.
+
+\subsection{A little bit of history}
+
+Common LISP \citep{steele1982ocl} introduced optional type annotations
+in the early eighties, but not for static type checking.
+Instead, programmers may choose to declare types of variables as
+optimization hints to the compiler, that is, type declarations are
+just one way to help the compiler to optimize code.
+These annotations are unsafe because they can crash the program
+when they are wrong.
+
+\citet{abadi1989dts} extended the simply typed lambda calculus with the
+\texttt{Dynamic} type and the \texttt{dynamic} and \texttt{typecase}
+constructs, with the aim to safely integrate dynamic code in
+statically typed languages.
+The \texttt{Dynamic} type is a pair \texttt{(v,T)} where \texttt{v} is a
+value and \texttt{T} is the tag that represents the type of \texttt{v}.
+The constructs \texttt{dynamic} and \texttt{typecase} are explicit
+injection and projection operations, respectively.
+That is, \texttt{dynamic} builds values of type \texttt{Dynamic} and
+\texttt{typecase} safely inspects the type of a \texttt{Dynamic} value.
+Thus, migrating code between dynamic and static type checking requires
+changing type annotations and adding or removing \texttt{dynamic} and
+\texttt{typecase} constructs throughout the code.
+
+The \textit{quasi-static} type system proposed by \citet{thatte1990qst}
+performs implicit coercions and run-time checks to replace the
+\texttt{dynamic} and \texttt{typecase} constructs that were proposed by
+\citet{abadi1989dts}.
+To do that, quasi-static typing relies on subtyping with a top type
+$\Omega$ that represents the dynamic type, and splits type checking
+into two phases.
+The first phase inserts implicit coercions from the dynamic type to
+the expected type, while the second phase performs what Thatte calls
+\textit{plausibility checking}, that is, it rewrites the program to
+guarantee that sequences of upcasts and downcasts always have a
+common subtype.
+
+\textit{Soft typing} \citep{cartwright1991soft} is another approach
+to combine static and dynamic typing in the same language.
+The main goal of soft typing is to add static type checking to
+dynamically typed languages without compromising their flexibility.
+To do that, soft typing relies on type inference for
+translating dynamically typed code to statically typed code.
+The type checker inserts run-time checks around inconsistent code and
+warns the programmer about the insertion of these run-time checks,
+as they indicate the existence of potential type errors.
+However, the programmer is free to choose between inspecting the
+run-time checks and simply running the code.
+Soft typing gives this choice to the programmer because it is one of
+its key ideas, that is, type inference and static type checking should
+not prevent the programmer from running inconsistent code.
+One advantage of soft typing is the fact that the compiler for
+softly typed languages can use the translated code to generate
+more efficient code, as the translated code statically type checks.
+One disadvantage of soft typing is that it can be cumbersome when
+the inferred types are meaningless large types that just confuse the
+programmer.
+
+\textit{Dynamic typing} \citep{henglein1994dts} is an approach
+that optimizes code from dynamically typed languages by eliminating
+unnecessary checks of tags.
+The paper describes how to translate dynamically typed code to
+statically typed code that uses a \texttt{Dynamic} type.
+The translation is done through a coercion calculus that uses type
+inference to insert the operations that are necessary to type check
+the \texttt{Dynamic} type during run-time.
+Although soft typing and dynamic typing may seem similar, they are not.
+They are not similar because soft typing targets statically type
+checking of dynamically typed languages for detecting programming
+errors while dynamic typing targets the optimization of dynamically
+typed code through the elimination of unnecessary run-time checks.
+In other words, soft typing sees code optimization as a side effect
+that comes with static type checking.
+
+\citet{findler2002chf} proposed contracts for higher-order functions
+and blame annotations for run-time checks.
+Contracts perform dynamic checking instead of static checking,
+but deferring all verifications to run-time can lead to defects
+that are difficult to fix, because run-time errors can show a
+stack trace where it is not clear to programmers if the cause of a
+certain run-time error is in application code or library code.
+Even if programmers identify that the source of a certain run-time
+error is in library code, they still may have problems to identify
+if this run-time error is due to a violation of library's contract,
+or due to a bug, when the library is poorly documented.
+In this approach, programmers can insert assertions in the form of
+contracts that check the input and output of higher-order functions.
+Thus, \citet{findler2002chf} show how programmers can insert
+assertions in the form of contracts that check the input and output
+of higher-order functions; and they also show how to add blame
+annotations in the generated code as a way to track assertion
+failures back to the source of the error.
+
+BabyJ \citep{anderson2003babyj} is an object-oriented language
+without inheritance that allows programmers to incrementally annotate
+the code with more specific types.
+Programmers can choose between using the dynamically typed version
+of BabyJ, when they do not need types at all, and the statically
+typed version of BabyJ, when they need to annotate the code.
+In the statically typed BabyJ, programmers can use the
+\textit{permissive type} $*$ to annotate the parts of the code that
+still do not have a specific type or the parts of the code that should
+have dynamic behavior.
+The type system of BabyJ is nominal, so types are either class names
+or the permissive type $*$.
+However, the type system does not use type equality or subtyping,
+but the relation $\approx$ between two types.
+The relation $\approx$ holds when both types have the same name or
+any of them is the permissive type $*$.
+Even though the permissive type $*$ is similar to the dynamic type
+from previous approaches, BabyJ does not provide any way to add
+implicit or explicit run-time checks.
+
+\citet{ou2004dtd} specified a language that combines static types
+with dependent types.
+To ensure safety, the compiler automatically inserts coercions
+between dependent code and static code.
+The coercions are run-time checks that ensure static code does not
+crash dependent code during run-time.
+
+\subsection{Optional Type Systems}
+
+Optional type systems \citep{bracha2004pluggable} are an approach for
+plugging static typing in dynamically typed languages through
+optional type annotations that are used to perform compile-time type
+checking, though not influencing the original run-time semantics of
+the language.
+This means that optional type systems can be unsound without affecting
+language safety, because the run-time semantics should still catch
+type errors independently of the static type checking.
+More precisely, optional type systems are unsound if they do not catch
+some type errors, but we say that languages with unsound optional type
+systems are safe because the semantics of these languages catch during
+run-time the type errors that the optional type system did not catch.
+Although optional type systems maintain the simplicity and
+flexibility of dynamically typed languages, their static type
+checking cannot prove the absence of all type errors when the optional
+type system is unsound.
+
+Optional type systems have an interesting relation to the untyped
+lambda calculus and the typed lambda calculus.
+The semantic rules of the unytped lambda calculus are the same of the
+typed lambda calculus, and the type system is actually optional and
+serves only for discarding programs that may have undesired behaviors
+\citep{bracha2004pluggable}.
+
+Optional type systems make software evolution easier and
+allow language designers to experiment with
+\textit{pluggable type systems} \citep{bracha2004pluggable}.
+Pluggable type systems are different optional type systems for
+checking different aspects of the program.
+
+According to \citet{bracha2004pluggable}, optional type systems should
+not include mandatory type inference too, because mandatory type
+inference may decrease the expressiveness of the type system.
+Instead, optional type systems should implement type inference as
+an independent mechanism that can be turned on and off according to
+the programmers' needs.
+
+Unfortunately, there is no formalization on optional type systems,
+so each language ends up implementing its optional type system on
+its own way.
+Now we will briefly discuss some languages that provide an informal
+description of their optional type systems rather than a formal one.
+
+Strongtalk \citep{bracha1993strongtalk,bracha1996strongtalk} is
+a version of Smalltalk that comes with an optional type system that
+does not include any type inference.
+
+Dart \citep{dart} and TypeScript \citep{typescript} are new
+languages that are designed with an optional type system, and both
+use JavaScript as their code generation target because their main
+purpose is web development.
+In fact, Dart is a new class-based object-oriented language with
+optional type annotations while TypeScript is a strict superset of
+JavaScript that provides optional type annotations and class-based
+object-oriented programming.
+Both Dart and TypeScript perform type inference to deduce static
+types for unannotated code.
+
+\subsection{Gradual Typing}
+
+The main goal of gradual typing \citep{siek2006gradual} is to allow
+programmers to choose between static and dynamic typing in the same
+language.
+To do that, \citet{siek2006gradual} extended the simply typed
+lambda calculus with the dynamic type $?$.
+In gradual typing, type annotations are optional, so an untyped
+variable is syntactic sugar for a variable whose declared type is
+the dynamic type $?$.
+Under these circumstances, we view gradual typing as a way to add
+a dynamic type to statically typed languages.
+
+The central idea of gradual typing is the \textit{consistency}
+relation, written $\tau_{1} \sim \tau_{2}$, that allows implicit
+conversions to and from the dynamic type, and disallows conversions
+between inconsistent types \citep{siek2006gradual}.
+For instance, $\mathbf{int} \sim \,?$ and $?\, \sim \mathbf{int}$,
+but $\mathbf{int} \not\sim \mathbf{string}$ and
+$\mathbf{string} \not\sim \mathbf{int}$. 
+The consistency relation is neither commutative nor transitive.
+
+Gradual typing is similar to the approaches proposed by
+\citet{abadi1989dts} and \citet{thatte1990qst} by including a
+dynamic type to a statically typed language.
+However, these three approaches differ in the way they handle the
+dynamic type.
+While \citet{siek2006gradual} rely on the consistency relation,
+\citet{abadi1989dts} rely on type equality with explicit projections
+and injections, and \citet{thatte1990qst} relies on subtyping.
+The subtyping relation actually is a pitfall on Thatte's quasi-static
+typing, because the dynamic type is put as the top and the bottom of
+the subtyping relation.
+Therefore, downcasts combined with the transitivity of subtyping
+accepts programs that should be rejected.
+
+Later, \citet{siek2007objects} discovered that the consistency relation
+is orthogonal to subtyping, so we can combine the two relations to
+achieve both consistency and subtyping.
+The \textit{consistent-subtyping} relation,
+written $\tau_{1} \lesssim \tau_{2}$,
+is essential to design object-oriented languages that have a
+gradual type system.
+Like the consistency relation, and unlike the subtyping relation,
+the consistent-subtyping relation is not transitive.
+
+Another important feature of gradual typing is the theoretic
+foundation that it provides for inserting run-time checks that
+prove dynamically typed code does not violate the invariants of
+statically typed code, thus preserving type safety.
+To do that, \citet{siek2006gradual} defined the run-time semantics
+of gradual typing as a translation to an intermediate language with
+explicit casts at the frontiers between statically and dynamically
+typed code.
+The semantics of these casts is based on the higher-order contracts
+proposed by \citet{findler2002chf}.
+
+However, \citet{herman2007sgt} showed that there is an efficiency
+concern regarding the run-time checks, because there are two
+ways that casts can lead to unbounded space consumption.
+The first affects tail recursion while the second appears when
+first-class functions or objects cross the border between
+static code and dynamic code, that is, some programs can apply
+repeated casts to the same function or object.
+\citet{herman2007sgt} use the coercion calculus of
+\citet{henglein1994dts} to solve the problem of space efficiency,
+that is, they use coercions to express casts.
+Their approach normalizes an arbitrary sequence of coercions to a
+coercion of bounded size.
+
+Another concern about casts is how to improve debugging support,
+because a cast application can be delayed and the error related
+to that cast application can appear far way from the real error.
+\citet{wadler2009wpc} developed the \textit{blame calculus} as a way
+to attack this issue, and \citet{ahmed2011bfa} extended the
+blame calculus with polymorphism.
+The blame calculus is an intermediate language to integrate
+static and dynamic typing that uses the blame tracking approach
+proposed by \citet{findler2002chf}.
+
+On the one hand, the blame calculus solves the issue regarding
+error reporting;
+on the other hand, it has the space efficiency problem reported
+by \citet{herman2007sgt}.
+Thus, \citet{siek2009casts} extended the coercion calculus of
+\citet{herman2007sgt} with blame tracking to achieve an
+implementation of the blame calculus that is space efficient.
+After that, \citet{siek2010blame} proposed a new solution that also
+attacks both problems.
+This new solution is based on a concept called \textit{threesome},
+which is a way to split a cast between two parties into two casts
+among three parties.
+A cast has a source and a target type (a \textit{twosome}),
+so we can split any cast into a downcast from the source to an
+intermediate type that is followed by an upcast from the intermediate
+type to the target type (a \textit{threesome}).
+
+ActionScript \citep{moock2007as3} is the first language that
+incorporated a gradual type system to its implementation and
+Perl 6 \citep{tang2007pri} is also being designed with a
+gradual type system, though there is little documentation about
+the gradual type systems of these languages.
+Grace \citep{black2012grace,black2013sg} is a new object-oriented,
+gradually typed, educational language.
+In Grace, modules are gradually typed objects, that is, modules
+may have types and methods as attributes, and can have a state
+\citep{homer2013modules}.
+Gradual Python \citep{reticulated} is an attempt to develop a
+gradual type system for Python that deals with mutable objects
+\citep{siek2013mutable}.
+Gradualtalk \citep{allende2013gts} is a gradually-typed Smalltalk
+that introduces a new cast insertion strategy for gradually-typed
+objects \citep{allende2013cis}.
+
+\subsection{Approaches that are often called Gradual Typing}
+
+Gradual typing is similar to optional type systems in that type
+annotations are optional, and unannotated code is dynamically
+typed, but unlike optional type systems, gradual typing changes 
+the run-time semantics to preserve type safety, and it is a way to
+add a dynamic type to statically typed languages. 
+More precisely, programming languages that include a gradual type
+system implement the semantics of statically typed languages, so
+the gradual type system inserts casts in the translated code to
+guarantee that types are consistent before execution, while
+programming languages that include an optional type system still
+implement the semantics of dynamically typed languages, so all
+the type checking also belongs to the semantics of each operation.
+
+Still, we can view gradual typing as a way to formalize an optional
+type system when the gradual type system does not insert run-time
+checks.
+BabyJ \citep{anderson2003babyj} and Alore \citep{lehtosalo2011alore}
+are two examples of object-oriented languages that use gradual
+typing to formalize their optional type systems.
+The optional type systems of both BabyJ and Alore are nominal.
+BabyJ uses the relation $\approx$ that is similar to the consistency
+relation while Alore combines  subtyping along with the consistency
+relation to define a \textit{consistent-or-subtype} relation.
+The consistent-or-subtype relation is different from the
+consistent-subtyping relation of \citet{siek2007objects}, but it is
+also written $\tau_{1} \lesssim \tau_{2}$.
+The consistent-or-subtype relation holds when $\tau_{1} \sim \tau_{2}$
+or $\tau_{1} <: \tau_{2}$, so $<:$ is transitive and $\sim$ is not.
+Alore also extends its optional type system to include optional
+monitoring of run-time type errors in the gradual typing style.
+
+Hence, optional type annotations for software evolution probably are
+the reasons why optional type systems are commonly called
+gradual type systems.
+Typed Clojure \citep{bonnaire-sergeant2012typed-clojure} is an
+optional type system for Clojure that now is adopting the
+gradual typing slogan.
+
+\citet{flanagan2006htc} introduced \textit{hybrid type checking},
+an approach that combines static types and refinement types.
+For instance, programmers can specify the refinement type
+$\{x:Int \;|\; x \ge 0\}$ when they need a type for natural numbers.
+The programmer can also choose between explicit or implicit casts.
+When casts are not explicit, the type checker uses a theorem prover
+to insert casts.
+In our example of natural numbers, a cast would be inserted to check
+whether an integer is greater than or equal to zero.
+
+Sage \citep{gronski2006sage} is a programming language that
+extended hybrid type checking with a dynamic type to
+support dynamic and static typing in the same language.
+Sage also offers optional type annotations in the gradual typing
+style, that is, unannotated code is syntactic sugar for
+code whose declared type is the dynamic type.
+
+Thus, the inclusion of a dynamic type to hybrid type checking
+along with optional type annotations, and the insertion of run-time
+checks probably are the reasons why hybrid type checking is
+also viewed as a form of gradual typing.
+
+\citet{tobin-hochstadt2006ims} proposed another approach for gradually
+migrating from dynamically typed to statically typed code,
+and they coined the term \textit{from scripts to programs} for
+referring to this kind of interlanguage migration.
+In their approach, the migration from dynamically typed to
+statically typed code happens module-by-module, so they designed
+and implemented Typed Racket \citep{tobin-hochstadt2008ts} for
+this purpose.
+Typed Racket is a statically typed version of Racket
+(a Scheme dialect) that allows the programmer to write typed modules,
+so Typed Racket modules can coexist with Racket modules,
+which are untyped.
+
+The approach used by \citet{tobin-hochstadt2008ts} to design and
+implement Typed Racket probably is also called gradual typing
+because it allows the programmer to gradually migrate from untyped
+scripts to typed programs.
+However, note that Typed Racket is a statically typed language
+and type annotations are mandatory.
+So, what makes it gradual is a type system with a dynamic type
+that handles the interaction between Racket and Typed Racket modules.
 
@@
 
-Lua is an imperative scripting language with first-class functions
-(with proper lexical scoping) where the main data structure is the
-{\em table}, an associative array that can play the part of arrays,
-records, maps, objects, etc., with syntactic sugar and
-metaprogramming through operator overloading built into
-the language. Unlike other scripting languages, Lua has very
-limited coercion among different data types.
-
-The primary use of Lua has always been as an embedded language
-for configuration and extension of other applications, and the result
-is a fragmented ecosystem of libraries, and different ideas among
-Lua programmers on how they should use the language features and
-how they should structure programs.
-
-The design of Typed Lua is informed by a (mostly automated) survey
-of Lua idioms used in a large corpus of Lua libraries, instead of
-relying just on the semantics of the language. We cover the main parts
-of the design, along with how they relate to Lua, in the rest of this section,
-and discuss some of what was left out in the current version of the design
-in the conclusion of the paper.
-
-\subsection{Atomic Types and Functions}
-
-Lua values can have one of eight {\em tags}: {\em nil}, {\em boolean},
-{\em number}, {\em string}, {\em function}, {\em table}, {\em userdata},
-and {\em thread}. We will see how Typed Lua assigns types to
-values of the first five in this section.
-
-Figure~\ref{fig:typelang} gives the abstract syntax of Typed Lua
-types. Only {\em first-class types} correspond to actual Lua
-values; {\em second-class types} correspond to expression lists,
-and Typed Lua uses them to type multiple assignment and function
-application. Types are ordered by a subtype relationship, where
-any first-class type is a subtype of $\Top$.
-
-The {\em dynamic type} $\Any$ allows dynamically typed code to
-interoperate with statically typed code; it is a subtype of $\Top$, but
-neither a supertype nor a subtype of any other type. We relate $\Any$
-to other types with the {\em consistency} and {\em consistent-subtype}
-relationships used by gradual type systems. In practice, we can
-pass a value of the dynamic type anytime we want a value of some
-other type, and can pass any value where a value of the dynamic type
-is expected, but these operations are tracked by the type system,
-and the programmer can choose to be warned about them.
+We begin this section presenting statistics about the usage of Lua
+features and idioms.
+We collected statistics about how programmers use tables, functions,
+dynamic type checking, object-oriented programming, and modules.
+We shall see that these statistics informed important design decisions
+on our optional type system.
+We end this section introducing Typed Lua through examples, and
+describing a few Lua projects that are similar to Typed Lua.
+
+\subsection{Statistics about the usage of Lua}
+
+We used the LuaRocks repository to build our statistics database;
+LuaRocks \citep{hisham2013luarocks} is a package manager for Lua
+modules.
+We downloaded the 3928 \texttt{.lua} files that were available in
+the LuaRocks repository at February 1st 2014.
+However, we ignored files that are not compatible with Lua 5.2,
+the latest version of Lua.
+We also ignored \textit{machine-generated} files and test files,
+because these files may not represent idiomatic Lua code,
+and might skew our statistics towards non-typical uses of Lua.
+This left 2598 \texttt{.lua} files from 262 different projects for
+our statistics database;
+we parsed these files and processed their abstract syntax tree
+to gather the statistics that we show in this section.
+
+To verify how programmers use tables, we measured how they
+initialize, index, and iterate tables.
+We present these statistics in the next three paragraphs to discuss
+their influence on our type system.
+
+The table constructor appears 23185 times.
+In 36\% of the occurrences it is a constructor that initializes a
+record (e.g., \texttt{\{ x = 120, y = 121 \}});
+in 29\% of the occurrences it is a constructor that initializes a
+list (e.g., \texttt{\{ "one", "two", "three", "four" \}});
+in 8\% of the occurrences it is a constructor that initializes a
+record with a list part;
+and in less than 1\% of the occurrences (4 times) it is a constructor
+that uses only the booleans \texttt{true} and \texttt{false} as indexes.
+At all, in 73\% of the occurrences it is a constructor that uses
+only literal keys;
+in 26\% of the occurrences it is the empty constructor;
+in 1\% of the occurrences it is a constructor with non-literal keys
+only, that is, a constructor that uses variables and function calls
+to create the indexes of a table;
+and in less than 1\% of the occurrences (19 times) it is a constructor
+that mixes literal keys and non-literal keys.
+
+The indexing of tables appears 130448 times:
+86\% of them are for reading a table field while
+14\% of them are for writing into a table field.
+We can classify the indexing operations that are reads as follows:
+89\% of the reads use a literal string key,
+4\% of the reads use a literal number key,
+and less than 1\% of the reads (10 times) use a literal boolean key.
+At all, 93\% of the reads use literals to index a table while
+7\% of the reads use non-literal expressions to index a table.
+It also worth mentioning that 45\% of the reads are actually
+function calls.
+More precisely, 25\% of the reads use literals to call a function,
+20\% of the reads use literals to call a method, that is,
+a function call that uses the colon syntactic sugar, 
+and less than 1\% of the reads (195 times) use non-literal expressions
+to call a function.
+We can also classify the indexing operations that are writes as follows: 
+69\% of the writes use a literal string key,
+2\% of the writes use a literal number key,
+and less than 1\% of the writes (1 time) uses a literal boolean key.
+At all, 71\% of the writes use literals to index a table while
+29\% of the writes use non-literal expressions to index a table.
+
+We also measured how many files have code that iterate over tables to
+observe how frequently iteration is used.
+We observed that 23\% of the files have code that iterate over keys
+of any value, that is, the call to \texttt{pairs} appears at least
+once in these files (the median is twice per file);
+21\% of the files have code that iterate over integer keys, that is,
+the call to \texttt{ipairs} appears at least once in these files
+(the median is also twice per file);
+and 10\% of the files have code that use the numeric \texttt{for}
+along with the length operator (the median is once per file).
+
+The numbers about table initialization, indexing, and iteration
+show us that tables are mostly used to represent records, lists,
+and associative arrays.
+Therefore, Typed Lua should include a table type for handling
+these uses of Lua tables.
+Even though the statistics show that programmers initialize tables
+more often than they use the empty constructor to
+dynamically initialize tables, the statistics of the empty
+constructor are still expressive and indicate that Typed Lua should
+also include a way to handle this style of defining table types.
+
+We found a total of 24858 function declarations in our database
+(the median is six per file).
+Next, we discuss how frequently programmers use dynamic type
+checking and multiple return values inside these functions.
+
+We observed that 9\% of the functions perform dynamic type checking
+on their input parameters, that is, these functions use \texttt{type}
+to inspect the tags of Lua values (the median is once per function).
+We randomly selected 20 functions to sample how programmers are
+using \texttt{type}, and we got the following data:
+50\% of these functions use \texttt{type} for asserting the tags of
+their input parameters, that is, they raise an error when the tag of a
+certain parameter does not match the expected tag, and
+50\% of these functions use \texttt{type} for overloading, that is,
+they execute different code according to the inspected tag.
+
+These numbers show us that Typed Lua should include union types,
+because the use of the \texttt{type} idiom shows that disjoint unions
+will help programmers define data structures that can hold a value of
+several different, but fixed types.
+Typed Lua should also use \texttt{type} as a mechanism for decomposing
+unions, though it may be restricted to base types only.
+
+We observed that 10\% of the functions explicitly return multiple
+values.
+We also observed that 5\% of the functions return \texttt{nil} plus
+something else, for signaling an unexpected behavior;
+and 1\% of the functions return \texttt{false} plus something else,
+also for signaling an unexpected behavior.
+
+Typed Lua should include function types to represent Lua functions,
+and tuple types to represent the signatures of Lua functions,
+multiple return values, and multiple assignments.
+Tuple types will require some special attention, because Typed Lua
+should be able to adjust tuple types during compile-time, in a
+similar way to what Lua does with function calls and multiple
+assignments during run-time.
+In addition, the number of functions that return \texttt{nil} and
+\texttt{false} plus something else show us that overloading on the
+return type will be useful to the type system.
+
+We also measured how frequently programmers use the object-oriented
+paradigm in Lua.
+We observed that 23\% of the function declarations are actually
+method declarations.
+More precisely, 14\% of them use the colon syntactic sugar while
+9\% of them use \texttt{self} as their first parameter.
+We also observed that 63\% of the projects extend tables with
+metatables, that is, they call \texttt{setmetatable} at least once,
+and 27\% of the projects access the metatable of a given table,
+that is, they call \texttt{getmetatable} at least once.
+In fact, 45\% of the projects extend tables with metatables and
+declare methods:
+13\% using the colon syntactic sugar, 14\% using \texttt{self}, and
+18\% using both.
+
+Based on these observations, Typed Lua should include interfaces and
+classes as syntactic sugar for table types that encode objects.
+Proposing an optional type system for Lua that includes classes may
+sound unusual at first glance, because Lua does not have standard
+policies for object-oriented programming.
+However, Lua provides mechanisms that allow programmers to abstract
+their code in terms of objects, and our statistics confirm that
+an expressive number of programmers are relying on these mechanisms to
+use the object-oriented paradigm in Lua.
+Typed Lua should include interfaces and classes as merely syntactic
+mechanisms that help the compiler type check object-oriented code,
+and they should not influence the semantics of Lua.
+
+We also measured how programmers are defining modules.
+We observed that 38\% of the files use the current way of defining
+modules, that is, these files return a table that contains the
+exported members of the module at the end of the file;
+22\% of the files still use the deprecated way of defining modules,
+that is, these files call the function \texttt{module};
+and 1\% of the files use both ways.
+At all, 61\% of the files are modules while 39\% of the files are
+plain scripts.
+The number of plain scripts is high considering the origin of
+our database.
+However, we did not ignore sample scripts, which usually serve to
+help the users of a given module on how to use this module, and
+that is the reason why we have a high number of plain scripts.
+
+Based on these observations, Typed Lua should also use interfaces as
+syntactic sugar for table types that define modules.
+Typed Lua should also support the deprecated style of module
+definition, using global names as exported members of the module.
+
+Besides using interfaces for encoding objects and defining modules,
+Typed Lua should also use interfaces for defining the types of
+userdata, as programmers often use userdata in an object-oriented
+style.
+
+The last statistics that we collected were about variadic functions
+and vararg expressions.
+We observed that 8\% of the functions are variadic, that is,
+their last parameter is the vararg expression.
+We also observed that 5\% of the initialization of lists
+(or 2\% of the occurrences of the table constructor) use solely the
+vararg expression.
+Typed Lua should include a \textit{vararg type} to handle variadic
+functions and vararg expressions.
+
+\subsection{Typed Lua through examples}
+
+Our main goal is to allow programmers to combine static and dynamic
+typing in Lua, so we propose Typed Lua as an extension to Lua 5.2
+syntax that introduces optional type annotations, and class-based
+object-oriented programming through the definition of classes,
+interfaces, and modules.
+An important requirement of Typed Lua is that it should be backwards
+compatible with Lua, that is, it should allow programmers to run any
+Lua code without modifications.
+For this reason, these syntactic extensions include new keywords
+that are not reserved words.
+
+Figure \ref{fig:typelang} shows that Typed Lua includes first-class
+and second-class types.
+Typed Lua uses first-class types for representing Lua values and
+second-class types for defining the signatures of functions.
+First-class types include literal types, base types, the type $\Any$,
+union types, function types, table types, and recursive types.
+Second-class types include the type $\Void$, first-class types,
+vararg types, and tuple types.
+Literal types are types that represent a single value
+(e.g., $\Nil$, $\False$, $\True$, literal numbers, and
+literal strings).
+Literal types are subtypes of their respective base types.
+More precisely, the literals $\False$ and $\True$ are subtypes of
+$\Boolean$, any literal number $n$ is subtype of $\Number$, and any
+literal string $s$ is subtype of $\String$.
+The type $\Any$ is the dynamic type that allows dynamic code coexist
+with static code.
+We will discuss union types, function types, table types, recursive
+types, and second-class types along this section.
 
@@
 \begin{array}{rlr}
-\multicolumn{3}{c}{\textbf{First-class types}} \\
-T ::= & \;\; L & \textit{literal types}\\
+T ::= & & \textit{first-class types}\\
+& \;\; L & \textit{literal types}\\
 & | \; B & \textit{base types}\\
-& | \; \Top & \textit{top type}\\
 & | \; \Any & \textit{dynamic type}\\
-& | \; T \cup T & \textit{disjoint union types}\\
-& | \; S \rightarrow S & \textit{function types}\\
-& | \; \{T:T, ..., T:T\} & \textit{table types}\\
-& | \; X & \textit{type variables}\\
-& | \; \mu X.T & \textit{recursive types}\\
-& | \; X_i & \textit{projection types}\\
-L ::= & \multicolumn{2}{l}{\False \; | \; \True \; | \; {<}{\it number}{>} \; | \; {<}{\it string}{>} } \\
-B ::= & \multicolumn{2}{l}{\Nil \; | \; \Boolean \; | \; \Number \; | \; \String} \\
-\multicolumn{3}{c}{} \\
-\multicolumn{3}{c}{\textbf{Second-class types}} \\
-S ::= &  \multicolumn{2}{l}{\Void \; |\; V \; | \; S \cup S} \\
+& | \; T \cup T & \textit{union type}\\
+& | \; S \rightarrow S & \textit{function type}\\
+& | \; \{T:T, ..., T:T\} & \textit{table type}\\
+& | \; X & \textit{type variable}\\
+& | \; \mu X.T & \textit{recursive type}\\
+L ::= & \;\; \Nil \; | \; \False \; | \; \True \\
+& | \; {<}{\it number}{>} \; | \; {<}{\it string}{>} & \\
+B ::= & \Boolean \; | \; \Number \; | \; \String & \\
+S ::= & \Void \; | \; V & \textit{second-class types}\\
 V ::= & \;\; T & \\
-& | \; T* & \textit{vararg types}\\
-& | \; T \times V & \textit{tuple types}
+& | \; T* & \textit{vararg type}\\
+& | \; T \times V & \textit{tuple type}
 \end{array}
@@
 \dend
-\caption{Abstract syntax of Typed Lua types}
+\caption{Typed Lua types}
 \label{fig:typelang}
@@
 
-Typed Lua allows optional type annotations in variable function
-declarations. It assigns the dynamic type to any parameter that
-does not have an type annotation, but assigns more precise
-types to unnanotated variables, based on the type of the
-expression that gives the initial value of the variable.
-
-In the following example, we use type annotations in a function
-declaration but do not use type annotations in the declaration of a
+Typed Lua allows optional type annotations in variable and function
+declarations.
+Whenever it is possible to use local type inference, the compiler
+assigns a more specific type to an unannotated declaration.
+When it is not possible to use local type inference, the compiler
+assigns the type $\Any$ to an unannotated declaration.
+For instance, in function declarations, the compiler does not infer
+a more specific type to unannotated input parameters, because,
+in this case, global type inference often is undecidable in a
+language with subtyping and overloading \citep{wells1999typability}.
+
+In the following example we use type annotations in a function
+declaration, and we do not use type annotations in a declaration of a
 local variable:
 \begin{verbatim}
-    local function factorial(n: number): number
+
+    local function fact (n:number):number
       if n == 0 then
@@
       else
-        return n * factorial(n - 1)
+        return n * fact(n - 1)
       end
@@
     local x = 5
-    print(factorial(x))
+    print(fact(x))
 \end{verbatim}
 
-The compiler assigns the type $\Number$ to the local variable \texttt{x},
-and this example compiles without warnings. Typed Lua allows
-programmers to combine annotated code with
+Even though we did not add type annotations in the declaration of
+the local variable \texttt{x}, the compiler uses local type inference to
+assign the type $\Number$ to \texttt{x}.
+The compiler does not use the literal type $5$ instead of the base
+type $\Number$ because this would prevent
+programmers from assigning other numeric values to the variable
+\texttt{x}.
+The inference that we implement in Typed Lua is quite simple, as it
+uses only the type of the local expression.
+
+Typed Lua allows programmers to combine annotated code with
 unannotated code, as we show in the following example:
 \begin{verbatim}
-    local function abs(n: number)
+    local function abs (n:number):number
       if n < 0 then
@@
 
-    local function distance(x, y)
+    local function dist (x, y)
       return abs(x - y)
@@
 
+The function \texttt{dist} receives two parameters of type $\Any$
+and returns a value of type $\Number$.
 The compiler assigns the dynamic type $\Any$ to the input
-parameters of \texttt{distance} because they do not have type annotations.
-Subtracting a value of type $\Any$ from another also yields a value of
-type $\Any$ (Lua has operator overloading, so the minus operation
-is not guaranteed to return a number in this case), but
-consistent subtyping lets us pass a value of type $\Any$ to a
-function that expects a $\Number$.
-
-Even though the return types
-of both {\tt abs} and {\tt distance} are not given, the compiler is able to
-infer a return type of $\Number$ to both functions, as they are
-local and not recursive.
-
-Lua has first-class functions, but they have some peculiarities. First,
-the number of arguments passed to a function does not need to
-match the function's arity; Lua silently drops extra arguments after
-evaluating them, or passes {\tt nil} in place of any missing arguments.
-Second, functions can return any number of values, and the number
-of values returned may not be statically known. Third, Lua also has
-multiple assignment, and the semantics of argument passing match
-those of multiple assignment (or vice-versa); calling a function is like
-doing a multiple assigment where the left side is the parameter list
-and the right side is the argument list.
-
-Typed Lua uses {\em second-class types} to encode the peculiarities
-of argument passing, multiple returns, and multiple assignment. We call
-them second-class because these types do not correspond to actual
-values and cannot be assigned to variables or parameters: they are an
-artifact of the interaction between the type system and the semantics of Lua.
-
-As we saw in Figure \ref{fig:typelang}, a second-class type in
-Typed Lua can be the type $\Void$ (the empty tuple),
-a tuple of first-class types optionally ending in a variadic type,
-or a union of these tuples. A variadic type $T*$ is a generator for a
-sequence of values of type $T \cup \Nil$. Unions of tuples play
-an important part in functions that are overloaded on the return type,
-together with {\em projection types}. Both are explained in the next
-section.
-
-Typed Lua always adds a
-variadic tail to the parts of a function type if none is specified, to match
-the semantics of Lua function calls. In the examples above, the types
-of {\tt factorial} and {\tt abs} are actually $\Number \times \Top *
- \rightarrow \Number \times \Nil *$, and the type of {\tt distance} is
-$\Any \times \Any \times \Top * \rightarrow \Number \times \Nil *$.
-
-If we call {\tt abs} with extra arguments, Typed Lua silently ignores
-them, as the type signature lets {\tt abs} receive any number of extra
-arguments. If we call {\tt abs} in the right side of an assignment to
-more than one lvalue, Typed Lua checks if the first lvalue has a type
-consistent with $\Number$, and any other lvalues need to have a type
-consistent with $\Nil$.
-
-A variadic type can only appear in the tail position of a tuple,
-because Lua takes only the first value of any expression that appears
-in an expression list that is not in tail position. The following example
-shows the interaction between multiple returns and expression lists:
-
-\begin{verbatim}
-    local function multiple()
-      return 2, "foo"
-    end
-
-    local function sum(x: number, y: number)
-      return x + y
-    end
-
-    local x, y, z = multiple(), multiple()
-    print(sum(multiple(), multiple())
-\end{verbatim}
-
-Function {\tt multiple} is
-$\Top * \rightarrow \Number \times \String \times \Nil *$,
-and {\tt sum} is $\Number \times \Number \times \Top * \rightarrow
-\Number \times \Nil *$. In the right side of the multiple assignment,
-only the first value produced by the first call to multiple gets used, so
-the type of the right side is $\Number \times \Number \times \String \times \Nil*$,
-and the types assigned to {\tt x}, {\tt y}, and {\tt z} are respectively $\Number$,
-$\Number$, and $\String$. This also means that the call to {\tt sum} compiles
-without errors, as the first two components of the tuple are consistent with
-the types of the parameters, and the other components are consistent with
-$\Top$.
-
-\subsection{Unions}
+parameters of \texttt{dist} because they do not have type annotations,
+and the compiler does not use global type inference, as we mentioned
+previously.
+The compiler infers the return type of \texttt{dist} because it is
+not a recursive function, that is, Typed Lua does not infer the return
+type of recursive functions.
+
+We are using the consistent-subtyping relation of gradual
+typing to check the interaction among the dynamic type and the other
+types.
+More precisely, we can call \texttt{abs} inside \texttt{dist} because
+\texttt{x - y} is of type $\Any$, and the type $\Any$ is
+consistent with the type $\Number$.
+The type $\Any$ is not the union of all types, but it is a separate
+type.
+
+Unlike gradual typing, we do not insert run-time checks that
+inspect the interaction between dynamically typed code and statically
+typed code during run-time.
+In other words, Typed Lua does not guarantee that dynamically typed
+code does not violate statically typed code.
+Typed Lua does not insert run-time checks because they can decrease
+run-time performance \citep{allende2013cis}.
+We believe that a careful evaluation of run-time checks should be done
+before inserting them on the type system.
+In the previous example, we cannot guarantee that \texttt{dist} is never
+going to call \texttt{abs} with a parameter that is not a number,
+because in the semantics of Lua the minus operator can result in a
+value that is not a number.
+Still, we can implement run-time checks in the future because we are
+using the consistent-subtyping relation of gradual typing to
+formalize our optional type system, and the type system is being
+designed to be sound.
 
@@
 $t\;\cup\;\Nil\;\mathbf{or}\;t$ always has the type $t$.
-
-\subsection{Tables and Interfaces}
 
@@
 We need to explicitly annotate \texttt{next} being of type
-\texttt{Element} $\cup\;\Nil$ because in our type system $\Nil$
+\texttt{Element} $\;\cup\;\Nil$ because in our type system $\Nil$
 is not the bottom type, that is, $\Nil$ is subtype of itself only.
@@
 the absence of an useful value.
-
-\subsection{Modules}
-
-Typed Lua will also introduce a specific syntax for defining modules.
-For now, each file is a module.
-We use all global names and their types to build the interface that
-represents the type of the module, in case the module is not returned
-at the end of the file.
-
-\subsection{Objects and Classes}
 
@@
 
-\section{Related Work}
- \label{sec:review}
-
-We begin this section presenting a little bit of the history about
-combining static and dynamic typing in the same language.
-After that, we introduce optional type systems and gradual typing.
-We end this section discussing why optional type systems and two
-other approaches are often called gradual typing.
-
-\subsection{A little bit of history}
-
-Common LISP \citep{steele1982ocl} introduced optional type annotations
-in the early eighties, but not for static type checking.
-Instead, programmers may choose to declare types of variables as
-optimization hints to the compiler, that is, type declarations are
-just one way to help the compiler to optimize code.
-These annotations are unsafe because they can crash the program
-when they are wrong.
-
-\citet{abadi1989dts} extended the simply typed lambda calculus with the
-\texttt{Dynamic} type and the \texttt{dynamic} and \texttt{typecase}
-constructs, with the aim to safely integrate dynamic code in
-statically typed languages.
-The \texttt{Dynamic} type is a pair \texttt{(v,T)} where \texttt{v} is a
-value and \texttt{T} is the tag that represents the type of \texttt{v}.
-The constructs \texttt{dynamic} and \texttt{typecase} are explicit
-injection and projection operations, respectively.
-That is, \texttt{dynamic} builds values of type \texttt{Dynamic} and
-\texttt{typecase} safely inspects the type of a \texttt{Dynamic} value.
-Thus, migrating code between dynamic and static type checking requires
-changing type annotations and adding or removing \texttt{dynamic} and
-\texttt{typecase} constructs throughout the code.
-
-The \textit{quasi-static} type system proposed by \citet{thatte1990qst}
-performs implicit coercions and run-time checks to replace the
-\texttt{dynamic} and \texttt{typecase} constructs that were proposed by
-\citet{abadi1989dts}.
-To do that, quasi-static typing relies on subtyping with a top type
-$\Omega$ that represents the dynamic type, and splits type checking
-into two phases.
-The first phase inserts implicit coercions from the dynamic type to
-the expected type, while the second phase performs what Thatte calls
-\textit{plausibility checking}, that is, it rewrites the program to
-guarantee that sequences of upcasts and downcasts always have a
-common subtype.
-
-\textit{Soft typing} \citep{cartwright1991soft} is another approach
-to combine static and dynamic typing in the same language.
-The main goal of soft typing is to add static type checking to
-dynamically typed languages without compromising their flexibility.
-To do that, soft typing relies on type inference for
-translating dynamically typed code to statically typed code.
-The type checker inserts run-time checks around inconsistent code and
-warns the programmer about the insertion of these run-time checks,
-as they mean the existence of potential type errors.
-However, the programmer is free to choose between inspecting the
-run-time checks and simply running the code.
-Soft typing gives this choice to the programmer because it is one of
-its key ideas, that is, type inference and static type checking should
-not prevent the programmer from running inconsistent code.
-One advantage of soft typing is the fact that the compiler for
-softly typed languages can use the translated code to generate
-more efficient code, as the translated code statically type checks.
-One disadvantage of soft typing is that it can be cumbersome when
-the inferred types are meaningless large types that just confuse the
-programmer.
-
-\textit{Dynamic typing} \citep{henglein1994dts} is an approach
-that optimizes code from dynamically typed languages by eliminating
-unnecessary checks of tags.
-The paper describes how to translate dynamically typed code to
-statically typed code that uses a \texttt{Dynamic} type.
-The translation is done through a coercion calculus that uses type
-inference to insert the operations that are necessary to type check
-the \texttt{Dynamic} type during run-time.
-Although soft typing and dynamic typing may seem similar, they are not.
-They are not similar because soft typing targets statically type
-checking of dynamically typed languages for detecting programming
-errors while dynamic typing targets the optimization of dynamically
-typed code through the elimination of unnecessary run-time checks.
-In other words, soft typing sees code optimization as a side effect
-that comes with static type checking.
-
-\citet{findler2002chf} proposed contracts for higher-order functions
-and blame annotations for run-time checks.
-Contracts perform dynamic checking instead of static checking,
-but deferring all verifications to run-time can lead to defects
-that are difficult to fix, because run-time errors can show a
-stack trace where it is not clear to programmers if the cause of a
-certain run-time error is in application code or library code.
-Even if programmers identify that the source of a certain run-time
-error is in library code, they still may have problems to identify
-if this run-time error is due to a violation of library's contract,
-or due to a bug, when the library is poorly documented.
-In this approach, programmers can insert assertions in the form of
-contracts that check the input and output of higher-order functions.
-Thus, \citet{findler2002chf} show how programmers can insert
-assertions in the form of contracts that check the input and output
-of higher-order functions; and they also show how to add blame
-annotations in the generated code as a way to track assertion
-failures back to the source of the error.
-
-BabyJ \citep{anderson2002babyj} is an object-oriented language
-without inheritance that allows programmers to incrementally annotate
-the code with more specific types.
-Programmers can choose between using the dynamically typed version
-of BabyJ, when they do not need types at all, and the statically
-typed version of BabyJ, when they need to annotate the code.
-In the statically typed BabyJ, programmers can use the
-\textit{permissive type} $*$ to annotate the parts of the code that
-still do not have a specific type or the parts of the code that should
-have dynamic behavior.
-The type system of BabyJ is nominal, so types are either class names
-or the permissive type $*$.
-However, the type system does not use type equality or subtyping,
-but the relation $\approx$ between two types.
-The relation $\approx$ holds when both types have the same name or
-any of them is the permissive type $*$.
-Even though the permissive type $*$ is similar to the dynamic type
-from previous approaches, BabyJ does not provide any way to add
-implicit or explicit run-time checks.
-
-\citet{ou2004dtd} specified a language that combines static types
-with dependent types.
-To ensure safety, the compiler automatically inserts coercions
-between dependent code and static code.
-The coercions are run-time checks that ensure static code does not
-crash dependent code during run-time.
-
-\subsection{Optional Type Systems}
-
-Optional type systems \citep{bracha2004pluggable} are an approach for
-plugging static typing in dynamically typed languages through
-optional type annotations that are used to perform compile-time type
-checking, though not influencing the original run-time semantics of
-the language.
-This means that optional type systems can be unsound without affecting
-language safety, because the run-time semantics should still catch
-type errors independently of the static type checking.
-More precisely, optional type systems are unsound if they do not catch
-some type errors, but we say that languages with unsound optional type
-systems are safe because the semantics of these languages catch during
-run-time the type errors that the optional type system did not catch.
-Although optional type systems maintain the simplicity and
-flexibility of dynamically typed languages, their static type
-checking cannot prove the absence of all type errors when the optional
-type system is unsound.
-
-Optional type systems have an interesting relation to the untyped
-lambda calculus and the typed lambda calculus.
-The semantic rules of the unytped lambda calculus are the same of the
-typed lambda calculus, and the type system is actually optional and
-serves only for discarding programs that may have undesired behaviors
-\citep{bracha2004pluggable}.
-
-Optional type systems make software evolution easier and
-allow language designers to experiment with
-\textit{pluggable type systems} \citep{bracha2004pluggable}.
-Pluggable type systems are different optional type systems for
-checking different aspects of the program.
-
-According to \citet{bracha2004pluggable}, optional type systems should
-not include mandatory type inference too, because mandatory type
-inference may decrease the expressiveness of the type system.
-Instead, optional type systems should implement type inference as
-an independent mechanism that can be turned on and off according to
-the programmers' needs.
-
-Unfortunately, there is no formalization on optional type systems,
-so each language ends up implementing its optional type system on
-its own way.
-Now we will briefly discuss some languages that provide an informal
-description of their optional type systems rather than a formal one.
-
-Strongtalk \citep{bracha1993strongtalk,bracha1996strongtalk} is
-a version of Smalltalk that comes with an optional type system that
-does not include any type inference.
-
-Dart \citep{dart} and TypeScript \citep{typescript} are new
-languages that are designed with an optional type system, and both
-use JavaScript as their code generation target because their main
-purpose is web development.
-In fact, Dart is a new class-based object-oriented language with
-optional type annotations while TypeScript is a strict superset of
-JavaScript that provides optional type annotations and class-based
-object-oriented programming.
-Both Dart and TypeScript perform type inference to deduce static
-types for unannotated code.
-
-\subsection{Gradual Typing}
-
-The main goal of gradual typing \citep{siek2006gradual} is to allow
-programmers to choose between static and dynamic typing in the same
-language.
-To do that, \citet{siek2006gradual} extended the simply typed
-lambda calculus with the dynamic type $?$.
-In gradual typing, type annotations are optional, so an untyped
-variable is syntactic sugar for a variable whose declared type is
-the dynamic type $?$.
-Under these circumstances, we view gradual typing as a way to add
-a dynamic type to statically typed languages.
-
-The central idea of gradual typing is the \textit{consistency}
-relation, written $\tau_{1} \sim \tau_{2}$, that allows implicit
-conversions to and from the dynamic type, and disallows conversions
-between inconsistent types \citep{siek2006gradual}.
-For instance, $\mathbf{int} \sim \,?$ and $?\, \sim \mathbf{int}$,
-but $\mathbf{int} \not\sim \mathbf{string}$ and
-$\mathbf{string} \not\sim \mathbf{int}$. 
-The consistency relation is neither commutative nor transitive.
-
-Gradual typing is similar to the approaches proposed by
-\citet{abadi1989dts} and \citet{thatte1990qst} by including a
-dynamic type to a statically typed language.
-However, these three approaches differ in the way they handle the
-dynamic type.
-While \citet{siek2006gradual} rely on the consistency relation,
-\citet{abadi1989dts} rely on type equality with explicit projections
-and injections, and \citet{thatte1990qst} relies on subtyping.
-The subtyping relation actually is a pitfall on Thatte's quasi-static
-typing, because the dynamic type is put as the top and the bottom of
-the subtyping relation.
-Therefore, downcasts combined with the transitivity of subtyping
-accepts programs that should be rejected.
-
-Later, \citet{siek2007objects} discovered that the consistency relation
-is orthogonal to subtyping, so we can combine the two relations to
-achieve both consistency and subtyping.
-The \textit{consistent-subtyping} relation,
-written $\tau_{1} \lesssim \tau_{2}$,
-is essential to design object-oriented languages that have a
-gradual type system.
-Like the consistency relation, and unlike the subtyping relation,
-the consistent-subtyping relation is not transitive.
-
-Another important feature of gradual typing is the theoretic
-foundation that it provides for inserting run-time checks that
-prove dynamically typed code does not violate the invariants of
-statically typed code, thus preserving type safety.
-To do that, \citet{siek2006gradual} defined the run-time semantics
-of gradual typing as a translation to an intermediate language with
-explicit casts at the frontiers between statically and dynamically
-typed code.
-The semantics of these casts is based on the higher-order contracts
-proposed by \citet{findler2002chf}.
-
-However, \citet{herman2007sgt} showed that there is an efficiency
-concern regarding the run-time checks, because there are two
-ways that casts can lead to unbounded space consumption.
-The first affects tail recursion while the second appears when
-first-class functions or objects cross the border between
-static code and dynamic code, that is, some programs can apply
-repeated casts to the same function or object.
-\citet{herman2007sgt} use the coercion calculus of
-\citet{henglein1994dts} to solve the problem of space efficiency,
-that is, they use coercions to express casts.
-Their approach normalizes an arbitrary sequence of coercions to a
-coercion of bounded size.
-
-Another concern about casts is how to improve debugging support,
-because a cast application can be delayed and the error related
-to that cast application can appear far way from the real error.
-\citet{wadler2009wpc} developed the \textit{blame calculus} as a way
-to attack this issue, and \citet{ahmed2011bfa} extended the
-blame calculus with polymorphism.
-The blame calculus is an intermediate language to integrate
-static and dynamic typing that uses the blame tracking approach
-proposed by \citet{findler2002chf}.
-
-On the one hand, the blame calculus solves the issue regarding
-error reporting;
-on the other hand, it has the space efficiency problem reported
-by \citet{herman2007sgt}.
-Thus, \citet{siek2009casts} extended the coercion calculus of
-\citet{herman2007sgt} with blame tracking to achieve an
-implementation of the blame calculus that is space efficient.
-After that, \citet{siek2010blame} proposed a new solution that also
-attacks both problems.
-This new solution is based on a concept called \textit{threesome},
-which is a way to split a cast between two parties into two casts
-among three parties.
-A cast has a source and a target type (a \textit{twosome}),
-so we can split any cast into a downcast from the source to an
-intermediate type that is followed by an upcast from the intermediate
-type to the target type (a \textit{threesome}).
-
-ActionScript \citep{moock2007as3} is the first language that
-incorporated a gradual type system to its implementation and
-Perl 6 \citep{tang2007pri} is also being designed with a
-gradual type system, though there is little documentation about
-the gradual type systems of these languages.
-Grace \citep{black2012grace,black2013sg} is a new object-oriented,
-gradually typed, educational language.
-In Grace, modules are gradually typed objects, that is, modules
-may have types and methods as attributes, and can have a state
-\citep{homer2013modules}.
-Gradual Python \citep{reticulated} is an attempt to develop a
-gradual type system for Python that deals with mutable objects
-\citep{siek2013mutable}.
-Gradualtalk \citep{allende2013gts} is a gradually-typed Smalltalk
-that introduces a new cast insertion strategy for gradually-typed
-objects \citep{allende2013cis}.
-
-\subsection{Approaches that are often called Gradual Typing}
-
-Gradual typing is similar to optional type systems in that type
-annotations are optional, and unannotated code is dynamically
-typed, but unlike optional type systems, gradual typing changes 
-the run-time semantics to preserve type safety, and it is a way to
-add a dynamic type to statically typed languages. 
-More precisely, programming languages that include a gradual type
-system implement the semantics of statically typed languages, so
-the gradual type system inserts casts in the translated code to
-guarantee that types are consistent before execution, while
-programming languages that include an optional type system still
-implement the semantics of dynamically typed languages, so all
-the type checking also belongs to the semantics of each operation.
-
-Still, we can view gradual typing as a way to formalize an optional
-type system when the gradual type system does not insert run-time
-checks.
-BabyJ \citep{anderson2002babyj} and Alore \citep{lehtosalo2011alore}
-are two examples of object-oriented languages that use gradual
-typing to formalize their optional type systems.
-The optional type systems of both BabyJ and Alore are nominal.
-BabyJ uses the relation $\approx$ that is similar to the consistency
-relation while Alore combines  subtyping along with the consistency
-relation to define a \textit{consistent-or-subtype} relation.
-The consistent-or-subtype relation is different from the
-consistent-subtyping relation of \citet{siek2007objects}, but it is
-also written $\tau_{1} \lesssim \tau_{2}$.
-The consistent-or-subtype relation holds when $\tau_{1} \sim \tau_{2}$
-or $\tau_{1} <: \tau_{2}$, so $<:$ is transitive and $\sim$ is not.
-Alore also extends its optional type system to include optional
-monitoring of run-time type errors in the gradual typing style.
-
-Hence, optional type annotations for software evolution probably are
-the reasons why optional type systems are commonly called
-gradual type systems.
-Typed Clojure \citep{bonnaire-sergeant2012typed-clojure} is an
-optional type system for Clojure that now is adopting the
-gradual typing slogan.
-
-\citet{flanagan2006htc} introduced \textit{hybrid type checking},
-an approach that combines static types and refinement types.
-For instance, programmers can specify the refinement type
-$\{x:Int \;|\; x \ge 0\}$ when they need a type for natural numbers.
-The programmer can also choose between explicit or implicit casts.
-When casts are not explicit, the type checker uses a theorem prover
-to insert casts.
-In our example of natural numbers, a cast would be inserted to check
-whether an integer is greater than or equal to zero.
-
-Sage \citep{gronski2006sage} is a programming language that
-extended hybrid type checking with a dynamic type to
-support dynamic and static typing in the same language.
-Sage also offers optional type annotations in the gradual typing
-style, that is, unannotated code is syntactic sugar for
-code whose declared type is the dynamic type.
-
-Thus, the inclusion of a dynamic type to hybrid type checking
-along with optional type annotations, and the insertion of run-time
-checks probably are the reasons why hybrid type checking is
-also viewed as a form of gradual typing.
-
-\citet{tobin-hochstadt2006ims} proposed another approach for gradually
-migrating from dynamically typed to statically typed code,
-and they coined the term \textit{from scripts to programs} for
-referring to this kind of interlanguage migration.
-In their approach, the migration from dynamically typed to
-statically typed code happens module-by-module, so they designed
-and implemented Typed Racket \citep{tobin-hochstadt2008ts} for
-this purpose.
-Typed Racket is a statically typed version of Racket
-(a Scheme dialect) that allows the programmer to write typed modules,
-so Typed Racket modules can coexist with Racket modules,
-which are untyped.
-
-The approach used by \citet{tobin-hochstadt2008ts} to design and
-implement Typed Racket probably is also called gradual typing
-because it allows the programmer to gradually migrate from untyped
-scripts to typed programs.
-However, note that Typed Racket is a statically typed language
-and type annotations are mandatory.
-So, what makes it gradual is a type system with a dynamic type
-that handles the interaction between Racket and Typed Racket modules.
-
-\subsection{Similar Lua projects}
+Typed Lua will also introduce a specific syntax for defining modules.
+For now, each file is a module.
+We use all global names and their types to build the interface that
+represents the type of the module, in case the module is not returned
+at the end of the file.
+
+Typed Lua includes function types because functions are first-class
+values in Lua.
+The syntax for defining a function type is \texttt{(s) -> (s)},
+where \texttt{s} is a second-class type.
+As we saw in Figure \ref{fig:typelang}, a second-class type in
+Typed Lua can be the type $\Void$, a first-class type, a variadic
+type, or a tuple type.
+We use tuple types to represent multiple parameters, multiple return
+values, and multiple assignments, and we use the variadic type to
+represent variadic functions and the vararg expression.
+Even though we already included the tuple type and the vararg type,
+we are still working on these types.
+
+The next example shows how we can annotate a \texttt{map} function
+that receives another function as its parameter and uses this
+function to alter the elements of a list of numbers, which is
+also a parameter of \texttt{map}:
+\begin{verbatim}
+    local function map (f:(number)->(number),
+                        l:{number}):void
+      for i=1, #l do
+        l[i] = f(l[i] or 0)
+      end
+    end
+\end{verbatim}
+
+To conclude this section, we just started the design and
+implementation of Typed Lua, and there is still several design
+decisions to be taken on this subject.
+For instance, our type system should handle the subtleties that
+come with metatables.
+
+\subsection{Related Lua projects}
 
@@
 Like Typed Lua, Tidal Lock also has local type inference.
-Unlike Typed Lua, Tidal Lock is not typing the common object-oriented
+Unlike Typed Lua, Tidal Lock does not type the common object-oriented
 idioms used by Lua programmers.
 
-\section{Conclusion} \label{sec:con}
+\section{Expected contributions} \label{sec:con}
+
+Our main contribution is the design of a complete optional type system
+for Lua, a procedural scripting language.
+TypeScript is a complete optional type system for JavaScript,
+but it does not include union types, and it uses arrays to represent
+variadic functions and multiple return values. 
+Gradualtalk is a complete gradual type system for Smalltalk,
+but Smalltalk is an object-oriented scripting language that does
+not handle multiple return values, multiple assignments, and
+variadic functions.
+Gradualtalk also mixes nominal and structural typing because of
+the object-oriented nature of Smalltalk.
+Typed Racket is a complete static type system for Racket, a functional
+scripting language.
+Typed Racket makes it easier to migrate from untyped to typed code,
+but the migration happens module-by-module and type annotations are
+mandatory on typed modules.
+There are research on the design of optional and gradual type systems
+for JavaScript, Perl, and Python, but they are either experimental or
+limited in their coverage.
+Furthermore, Typed Lua includes table types that use types as keys,
+instead of identifiers, and the fact that Typed Lua also includes
+literal types makes its table types subsume the usual idea of
+record types.
+
+Although each language is different, we believe that several parts of
+the design of Typed Lua should be relevant to other scripting languages.
+Most notably, JavaScript shares with Lua the use of tests of type tags
+as a way of encoding optional and overloaded parameters, and the
+proliferation of class-based object systems built on top of the
+language's delegation mechanisms.
+
+Typed Lua should also be a major contribution to the Lua community,
+because it offers a framework that programmers can use to document,
+test, and better structure their applications.
+For libraries where a full conversion to static type checking should
+prove unfeasible or too much work, the community can use Typed Lua
+just to document the external interfaces of the libraries,
+giving the benefits of static type checking to the users of these
+libraries.
+
+Finally, Typed Lua can be a base for further research and development
+of static type systems for the Lua language, such as the use of static
+analysis of Lua API calls for checking that a native library is
+exporting the interface to Lua code that it declares, the use of
+effect types for typing Lua coroutines, and the use of static types
+for code optimization.
 

### 2014-03-10 doc/paper/typedlua.tex
finished rewrite of section 2.3

--- 2014-03-10 doc/paper/typedlua.tex
@@
 Typed Lua lets assignment change the type of a local variable in
-cases that respect a simple heuristic: the new type must be a subtype
-of the previous one, and only applies for the remainder of the
-current scope.
+cases that respect a simple heuristic: the new type must be a
+subtype of the previous one, and the variable must be local
+to the current function. The new type only applies for the
+remainder of the current scope.
 
 Overloaded functions use the {\tt type} function to inspect
-the tag of their parameters, and perform different actions depending
-on what those tags are. The simplest case overloads on just a
-single parameter:
+the tag of their parameters, and perform different actions
+depending on what those tags are. The simplest case overloads
+on just a single parameter:
 
@@
 is not just the type of the variable {\tt person} but the
-type of the value that {\tt person} points to.
+type of the value that {\tt person} points to. This is safe
+in the example above, but not in the example below:
+
+\begin{verbatim}
+    local bogus = { firstname = 1 }
+    local person: {} = bogus
+    person.firstname = "Lou"
+    person.lastname = "Reed"
+\end{verbatim}
+
+The assignment on line two is perfectly legal, as the type
+of {\tt bogus} is a subtype of $\{\}$. But changing the type
+of {\tt person} would be unsound: {\tt person.firstname} is
+now a $\String$, but {\tt bogus.firstname} is still typed
+as a $\Number$. We do not even need to declare a type for
+aliasing to be a problem:
+
+\begin{verbatim}
+    local person = {}
+    local bogus = person
+    bogus.firstname = 1
+    person.firstname = "Lou"
+    person.lastname = "Reed"
+\end{verbatim}
+
+Taken individually, the changes to the type of the two
+variables look ok, but aliasing makes one of them unsound.
+The location of the change also matters, as the next example
+shows:
+
+\begin{verbatim}
+    local person = {}
+    local bogus = { firstname = 1 }
+    do
+      person.firstname = 1
+      bogus = person
+    end
+    do
+      person.firstname = "Lou"
+    end
+    -- bogus.firstname is now "Lou"
+\end{verbatim}
+
+The initial type of {\tt person} in all of these examples is
+$\{\}$, but the {\em origin} of this type judgment matters
+on whether it is sound to allow a change to the type of
+{\tt person} or not, even if the change is always towards
+a subtype of the current type.
+
+Typed Lua tags a variable with a table type as either
+{\em open} or {\em closed}. If a variable gets its type
+from a table constructor then it is open, otherwise it is
+always closed. The type of an open variable may change
+by field assignment, subject to two restrictions: the
+variable must be local to the current block, and the new type
+must be a subtype of the old type.
+
+Using a variable with an open type can also trigger a
+type change, if the type of the missing fields has
+{\tt nil} as a possible value. This lets the programmer
+incrementally create an instance of an interface with
+an optional type:
+
+\begin{verbatim}
+    local interface Person
+      firstname: string
+      middlename: string?
+      lastname: string
+    end
+
+    local user = {}
+    user.firstname = "Lou"
+    user.lastname = "Reed"
+    local person: Person = user
+\end{verbatim}
+
+Table types are the foundation for modules and objects in
+Typed Lua. Type changes triggered by field assignment are
+also an important part of Typed Lua's support for the
+idiomatic definition of Lua modules, which are the
+subject of the next section.
 

### 2014-03-12 doc/paper/typedlua.tex
constant fields in the grammar of table types

@@
 \newcommand{\Void}{\mathbf{void}}
+\newcommand{\Const}{\mathbf{const}}
 
@@
 & | \; S \rightarrow S & \textit{function types}\\
-& | \; \{T:T, ..., T:T\} & \textit{table types}\\
+& | \; \{F, ..., F\} & \textit{table types}\\
 & | \; X & \textit{type variables}\\
@@
 B ::= & \multicolumn{2}{l}{\Nil \; | \; \Boolean \; | \; \Number \; | \; \String} \\
+F ::= & T:T \; | \; \Const \; T:T & \textit{field types} \\
 \multicolumn{3}{c}{} \\

### 2014-03-12 doc/paper/typedlua.tex
reorganization breaking the subsections of section 2 into individual sections; rewrite of section on modules

@@
 
-Lua is a small imperative scripting language that prefers to provide
-mechanisms instead of fixed policies for structuring programs,
-due to its history as an embedded extension language. Even features
+Lua is a small imperative scripting language with first-class
+functions (with proper lexical scoping) where the main data
+structure is the {\em table}, an associative array that can
+play the part of arrays, records, maps, objects, etc.
+with syntactic sugar and metaprogramming through operator overloading built into
+the language. Unlike other scripting languages, Lua has very
+limited coercion among different data types.
+
+The primary use of Lua has always been as an embedded language
+for configuration and extension of other applications.
+Lua prefers to provide mechanisms instead of fixed policies
+for structuring programs, and even features
 such as a module system and object orientation are a matter of 
-convention instead of built-in features.
+convention instead of built into the language.
+The result is a fragmented ecosystem of libraries, and
+different ideas among Lua programmers on how they should use
+the language features and how they should structure programs.
 
@@
 is a challenge for the design of a static type system for the Lua
-language, and our approach was to extract the most common
-idioms from a large corpus of Lua libraries as a guide to what
-Typed Lua should focus on, and setting fixed policies for organizing
-a program in modules and using class-based object orientation.
-
-Typed Lua allows statically typed Lua code to coexist and interact with 
-dynamically typed code. The Typed Lua compiler warns the programmer
-about type errors, but always generates Lua code that runs in unmodified
-Lua implementations. The programmer can enjoy some of the benefits
-of static types even without converting existing Lua modules to typed Lua:
-a dynamically typed module can export a statically typed interface, and
-statically typed users of the module will have their use of the module
-checked by the compiler.
+language. The design of Typed Lua is informed by a (mostly
+automated) survey of Lua idioms used in a large corpus of Lua
+libraries, instead of relying just on the semantics of the
+language.
+
+Typed Lua allows statically typed Lua code to coexist and
+interact with  ynamically typed code. The Typed Lua compiler
+warns the programmer about type errors, but always generates
+Lua code that runs in unmodified Lua implementations. The
+programmer can enjoy some of the benefits of static types even
+without converting existing Lua modules to typed Lua:
+a dynamically typed module can export a statically typed
+interface, and statically typed users of the module will have
+their use of the module checked by the compiler.
 
 Unlike gradual type systems, Typed Lua does not insert run-time
-checks between dynamically and statically typed parts of the program.
-Unlike some optional type systems, the statically typed subset of
-Typed Lua is sound by design, so a later version of the Typed Lua
-compiler can swicth to gradual instead of just optional typing.
-
-In Section~\ref{sec:proposal} we outline the key parts in
-the design of Typed Lua. In Section~\ref{sec:review} we review
-related work on mixing static and dynamic typing in the same language.
-Finally, Section~\ref{sec:con} gives some concluding remarks and
-our plans for the future of Typed Lua.
-
-\section{Typed Lua} \label{sec:proposal}
-
-Lua is an imperative scripting language with first-class functions
-(with proper lexical scoping) where the main data structure is the
-{\em table}, an associative array that can play the part of arrays,
-records, maps, objects, etc., with syntactic sugar and
-metaprogramming through operator overloading built into
-the language. Unlike other scripting languages, Lua has very
-limited coercion among different data types.
-
-The primary use of Lua has always been as an embedded language
-for configuration and extension of other applications, and the result
-is a fragmented ecosystem of libraries, and different ideas among
-Lua programmers on how they should use the language features and
-how they should structure programs.
-
-The design of Typed Lua is informed by a (mostly automated) survey
-of Lua idioms used in a large corpus of Lua libraries, instead of
-relying just on the semantics of the language. We cover the main parts
-of the design, along with how they relate to Lua, in the rest of this section,
-and discuss some of what was left out in the current version of the design
-in the conclusion of the paper.
-
-\subsection{Atomic Types and Functions}
+checks between dynamically and statically typed parts of the
+program. Unlike some optional type systems, the statically
+typed subset of Typed Lua is sound by design, so a later
+version of the Typed Lua compiler can swicth to gradual
+instead of just optional typing.
+
+We cover the main parts of the design, along with
+how they relate to Lua, in Sections~\ref{sec:atomic}
+through~\ref{sec:classes}. Section~\ref{sec:review} reviews
+related work on mixing static and dynamic typing in the same
+language. Finally, Section~\ref{sec:con} gives some
+concluding remarks and future extensions for Typed Lua.
+
+\section{Atomic Types and Functions}
+\label{sec:atomic}
 
@@
 
-\subsection{Unions}
+\section{Unions}
+\label{sec:unions}
 
@@
 
-\subsection{Tables and Interfaces}
+\section{Tables and Interfaces}
+\label{sec:tables}
 
@@
 
-\subsection{Modules}
-
-Typed Lua will also introduce a specific syntax for defining modules.
-For now, each file is a module.
-We use all global names and their types to build the interface that
-represents the type of the module, in case the module is not returned
-at the end of the file.
-
-\subsection{Objects and Classes}
+\section{Modules}
+\label{sec:modules}
+
+Lua's module system, like other parts of the language, is
+a matter of convention. When Lua first needs to load a
+module, it executes the module's source file as a function;
+the value that this function returns is the module, and Lua
+caches it for future loads. While a module can be any
+Lua value, most modules are tables where the fields
+of the table are functions and other values that the
+module exports.
+
+The modules that we surveyed build this table using
+three distinct styles. In the first style, the module's
+source file ends
+with a {\tt return} statement that uses a table constructor
+with the exported members. In the second style, the
+module declares an empty table in the beginning of its
+source file, adds exported members to this table
+throughout the module, and returns this table at the end.
+
+These two styles are straightforward for Typed Lua, which
+can just take the type of the first value that the module
+returns and use it as the type of the module.
+
+In the third style, which has been deprecated in the current
+version of Lua, a module begins with a call to the {\tt module}
+function. This function installs a fresh table as the
+global environment for the rest of the module, so any
+assignments to global variables are field assignments to
+this table. The {\tt module} function also sets an {\tt \_M}
+field in this table as a circular reference to the table
+itself, so the module can end with {\tt return \_M}, but
+this explicit return is not necessary.
+
+While this style has been deprecated, our survey indicated
+that around a third of Lua modules in a popular module
+repository still use this style, so Typed Lua also
+supports this style: it treats accesses to global
+variables as field accesses to an open table
+in the top-level scope.
+
+\section{Objects and Classes}
+\label{sec:classes}
 

### 2014-03-13 doc/paper/typedlua.tex
started section on objects

@@
 
+Lua's built-in support for object oriented programming is
+minimal. The basic mechanism is the {\tt :} syntactic
+sugar for method calls and method declarations.
+The Lua compiler translates {\tt obj:method(args)}
+to an operation that evaluates {\tt obj}, looks
+for a field named ``method" in the result, then calls
+it with the result of evaluating {\tt obj} as the
+first argument, followed by the result of evaluating
+the argument list in the original expression.
+
 Besides interfaces, Typed Lua will also introduce classes.

### 2014-03-14 doc/paper/typedlua.tex
remove nil in map type
rewrite of section on objects and classes

@@
 This table type represents a map that maps values of type
-$t$ to values of type $u \cup \Nil$.
+$t$ to values of type $u$.
 Another syntax defines table types for arrays:
@@
 
-Besides interfaces, Typed Lua will also introduce classes.
-We want to offer classes as a standard way for writing code that
-Lua programmers are already using, but without tying them to a
-specific model.
-Our classes will not include static attributes and will also not offer
-privacy rules.
+We can use recursive table types to represent objects,
+and Typed Lua has syntactic sugar to make defining these
+types easier:
+
+\begin{verbatim}
+    interface Shape
+      x, y: number
+      const move: (dx: number, dy: number) => ()
+    end
+\end{verbatim}
+
+The double arrow in the type of the two methods is
+syntactic sugar for having a first parameter named
+{\tt self} with the same type as the interface. The
+{\tt const} qualifier is necessary for covariance
+in the types of the methods, and to make subtyping
+among object types work.
+
+Type-checking a method call is straightforward, but
+there is still the matter of how to construct a value with
+the object type above. The following example shows one
+way:
+
+\begin{verbatim}
+    local shape = { x = 0, y = 0 }
+    const function shape:move(dx: number, 
+                              dy: number)
+      self.x = self.x + dx
+      self.y = self.y + dy
+    end
+\end{verbatim}
+
+The {\tt :} syntactic sugar that the example uses also comes
+from Lua, and assigns a function to the field with a first
+parameter named {\tt self}, plus any other parameters.
+
+Lua has a mechanism for Self-like delegation of missing
+fields in a table. After we do {\tt setmetatable(t1,
+{ \_\_index = t2 })}, Lua looks up in {\tt t2} any missing
+fields of {\tt t1}. Lua programmers often use this mechanism
+to simulate classes, as in the following example:
+
+\begin{verbatim}
+    local Shape = { x = 0, y = 0 }
+    const function Shape:new(x: number, y: number)
+      local s = setmetatable({},
+                             { __index = self })
+      s.x = x
+      s.y = y
+      return shape
+    end
+    const function Shape:move(dx: number,
+                              dy: number)
+      self.x = self.x + dx
+      self.y = self.y + dy
+    end
+    local shape1 = Shape:new(0, 5)
+    local shape2 = Shape:new(10, 10)
+\end{verbatim}
+
+Typed Lua assigns the type
+$\mathbf{Shape} \times \Number \times \Number \times
+\Top* \rightarrow \mathbf{Shape} \times \Nil*$ to
+{\tt new}. In a {\tt setmetatable} expression,
+if the type $t_1$ of the first operand is open and a supertype
+of the type $t_2$ of the second operand's {\_\_index} field,
+it changes the type of the first operand to $t_1$, and
+it remains open.
+
+As the result of {\tt setmetatable} can be an open
+table type, we can simulate single inheritance, and
+override methods:
+
+\begin{verbatim}
+    local Circle = setmetatable({},
+                                { __index = Shape })
+    Circle.radius = 0
+    const function Circle:new(x: number,
+                              y: number,
+                              radius: number)
+      local c = setmetatable(Shape:new(x, y),
+                             { __index = self })
+      c.radius = radius
+      return c
+    end
+    const function Circle:area()
+      return math.pi * self.radius * self.radius
+    end
+\end{verbatim}
+
+Notice how we can call {\tt Shape}'s constructor inside
+the overriden constructor. A limitation of this class
+system is that the overriden constructors must be a
+subtype of the original constructor.
+
+If we erase all type and {\tt const} annotations, the
+two examples above are valid Lua code, with the same
+semantics as the Typed Lua Code.
+
+The current version of Typed Lua does not have a polymorphic
+type system, so programmers currently cannot hide the
+calls to {\tt setmetatable} behind nicer abstractions, as
+some Lua libraries do. A few Lua programs also use
+other features of {\tt setmetatable} which are currently
+not typeable, such as operator overloading.
 
@@
 
-We begin this section presenting a little bit of the history about
-combining static and dynamic typing in the same language.
-After that, we introduce optional type systems and gradual typing.
-We end this section discussing why optional type systems and two
-other approaches are often called gradual typing.
+We begin this section presenting a little bit of the history
+about combining static and dynamic typing in the same language.
+After that, we introduce optional type systems and gradual
+typing. We end this section discussing why optional type
+systems and two other approaches are often called gradual typing.
 

### 2014-03-14 doc/paper/typedlua.tex
started rewrite of related work section

@@
 In the following example, we use type annotations in a function
-declaration but do not use type annotations in the declaration of a
-local variable:
+declaration but do not use type annotations in the declaration
+of a local variable:
 \begin{verbatim}
@@
     local shape1 = Shape:new(0, 5)
-    local shape2 = Shape:new(10, 10)
-\end{verbatim}
+    local shape2: Shape = Shape:new(10, 10)
+\end{verbatim}
+
+In the last line of the example, notice how we can refer
+to {\tt Shape} in the type annotation, as a shortcut
+to the recursive table type that Typed Lua has assigned
+to this variable.
 
@@
 
-Notice how we can call {\tt Shape}'s constructor inside
+In the first line of the redefinition of {\tt new},
+notice how we can call {\tt Shape}'s constructor inside
 the overriden constructor. A limitation of this class
@@
 
-We begin this section presenting a little bit of the history
-about combining static and dynamic typing in the same language.
-After that, we introduce optional type systems and gradual
-typing. We end this section discussing why optional type
-systems and two other approaches are often called gradual typing.
-
-\subsection{A little bit of history}
-
-Common LISP \citep{steele1982ocl} introduced optional type annotations
-in the early eighties, but not for static type checking.
-Instead, programmers may choose to declare types of variables as
-optimization hints to the compiler, that is, type declarations are
-just one way to help the compiler to optimize code.
-These annotations are unsafe because they can crash the program
-when they are wrong.
-
-\citet{abadi1989dts} extended the simply typed lambda calculus with the
-\texttt{Dynamic} type and the \texttt{dynamic} and \texttt{typecase}
-constructs, with the aim to safely integrate dynamic code in
-statically typed languages.
-The \texttt{Dynamic} type is a pair \texttt{(v,T)} where \texttt{v} is a
-value and \texttt{T} is the tag that represents the type of \texttt{v}.
-The constructs \texttt{dynamic} and \texttt{typecase} are explicit
-injection and projection operations, respectively.
-That is, \texttt{dynamic} builds values of type \texttt{Dynamic} and
-\texttt{typecase} safely inspects the type of a \texttt{Dynamic} value.
-Thus, migrating code between dynamic and static type checking requires
-changing type annotations and adding or removing \texttt{dynamic} and
-\texttt{typecase} constructs throughout the code.
-
-The \textit{quasi-static} type system proposed by \citet{thatte1990qst}
-performs implicit coercions and run-time checks to replace the
-\texttt{dynamic} and \texttt{typecase} constructs that were proposed by
-\citet{abadi1989dts}.
-To do that, quasi-static typing relies on subtyping with a top type
-$\Omega$ that represents the dynamic type, and splits type checking
-into two phases.
-The first phase inserts implicit coercions from the dynamic type to
-the expected type, while the second phase performs what Thatte calls
-\textit{plausibility checking}, that is, it rewrites the program to
-guarantee that sequences of upcasts and downcasts always have a
-common subtype.
-
-\textit{Soft typing} \citep{cartwright1991soft} is another approach
-to combine static and dynamic typing in the same language.
-The main goal of soft typing is to add static type checking to
-dynamically typed languages without compromising their flexibility.
-To do that, soft typing relies on type inference for
-translating dynamically typed code to statically typed code.
-The type checker inserts run-time checks around inconsistent code and
-warns the programmer about the insertion of these run-time checks,
-as they mean the existence of potential type errors.
-However, the programmer is free to choose between inspecting the
-run-time checks and simply running the code.
-Soft typing gives this choice to the programmer because it is one of
-its key ideas, that is, type inference and static type checking should
-not prevent the programmer from running inconsistent code.
-One advantage of soft typing is the fact that the compiler for
-softly typed languages can use the translated code to generate
-more efficient code, as the translated code statically type checks.
-One disadvantage of soft typing is that it can be cumbersome when
-the inferred types are meaningless large types that just confuse the
-programmer.
-
-\textit{Dynamic typing} \citep{henglein1994dts} is an approach
-that optimizes code from dynamically typed languages by eliminating
-unnecessary checks of tags.
-The paper describes how to translate dynamically typed code to
-statically typed code that uses a \texttt{Dynamic} type.
-The translation is done through a coercion calculus that uses type
-inference to insert the operations that are necessary to type check
-the \texttt{Dynamic} type during run-time.
-Although soft typing and dynamic typing may seem similar, they are not.
-They are not similar because soft typing targets statically type
-checking of dynamically typed languages for detecting programming
-errors while dynamic typing targets the optimization of dynamically
-typed code through the elimination of unnecessary run-time checks.
-In other words, soft typing sees code optimization as a side effect
-that comes with static type checking.
-
-\citet{findler2002chf} proposed contracts for higher-order functions
-and blame annotations for run-time checks.
-Contracts perform dynamic checking instead of static checking,
-but deferring all verifications to run-time can lead to defects
-that are difficult to fix, because run-time errors can show a
-stack trace where it is not clear to programmers if the cause of a
-certain run-time error is in application code or library code.
-Even if programmers identify that the source of a certain run-time
-error is in library code, they still may have problems to identify
-if this run-time error is due to a violation of library's contract,
-or due to a bug, when the library is poorly documented.
-In this approach, programmers can insert assertions in the form of
-contracts that check the input and output of higher-order functions.
-Thus, \citet{findler2002chf} show how programmers can insert
-assertions in the form of contracts that check the input and output
-of higher-order functions; and they also show how to add blame
-annotations in the generated code as a way to track assertion
-failures back to the source of the error.
-
-BabyJ \citep{anderson2002babyj} is an object-oriented language
-without inheritance that allows programmers to incrementally annotate
-the code with more specific types.
-Programmers can choose between using the dynamically typed version
-of BabyJ, when they do not need types at all, and the statically
-typed version of BabyJ, when they need to annotate the code.
-In the statically typed BabyJ, programmers can use the
-\textit{permissive type} $*$ to annotate the parts of the code that
-still do not have a specific type or the parts of the code that should
-have dynamic behavior.
-The type system of BabyJ is nominal, so types are either class names
-or the permissive type $*$.
-However, the type system does not use type equality or subtyping,
-but the relation $\approx$ between two types.
-The relation $\approx$ holds when both types have the same name or
-any of them is the permissive type $*$.
-Even though the permissive type $*$ is similar to the dynamic type
-from previous approaches, BabyJ does not provide any way to add
-implicit or explicit run-time checks.
-
-\citet{ou2004dtd} specified a language that combines static types
-with dependent types.
-To ensure safety, the compiler automatically inserts coercions
-between dependent code and static code.
-The coercions are run-time checks that ensure static code does not
-crash dependent code during run-time.
-
-\subsection{Optional Type Systems}
-
-Optional type systems \citep{bracha2004pluggable} are an approach for
-plugging static typing in dynamically typed languages through
-optional type annotations that are used to perform compile-time type
-checking, though not influencing the original run-time semantics of
-the language.
-This means that optional type systems can be unsound without affecting
-language safety, because the run-time semantics should still catch
-type errors independently of the static type checking.
-More precisely, optional type systems are unsound if they do not catch
-some type errors, but we say that languages with unsound optional type
-systems are safe because the semantics of these languages catch during
-run-time the type errors that the optional type system did not catch.
-Although optional type systems maintain the simplicity and
-flexibility of dynamically typed languages, their static type
-checking cannot prove the absence of all type errors when the optional
-type system is unsound.
-
-Optional type systems have an interesting relation to the untyped
-lambda calculus and the typed lambda calculus.
-The semantic rules of the unytped lambda calculus are the same of the
-typed lambda calculus, and the type system is actually optional and
-serves only for discarding programs that may have undesired behaviors
-\citep{bracha2004pluggable}.
-
-Optional type systems make software evolution easier and
-allow language designers to experiment with
-\textit{pluggable type systems} \citep{bracha2004pluggable}.
-Pluggable type systems are different optional type systems for
-checking different aspects of the program.
-
-According to \citet{bracha2004pluggable}, optional type systems should
-not include mandatory type inference too, because mandatory type
-inference may decrease the expressiveness of the type system.
-Instead, optional type systems should implement type inference as
-an independent mechanism that can be turned on and off according to
-the programmers' needs.
-
-Unfortunately, there is no formalization on optional type systems,
-so each language ends up implementing its optional type system on
-its own way.
-Now we will briefly discuss some languages that provide an informal
-description of their optional type systems rather than a formal one.
+Common LISP introduced optional type annotations in the early
+eighties~\citep{steele1982ocl}, but they were optimization
+hints to the compiler instead of types for static checking.
+These annotations were unsafe, and could crash the program
+when wrong.
+
+\citet{abadi1989dts} used tagged pairs and explicit
+injection and projection operations (coercions) to embed dynamic
+typing in the simply-typed lambda calculus. Dynamically-typed
+values had a {\tt Dynamic} static type. \citet{thatte1990qst}
+removes the necessity of explicit coercions with a system
+that automatically inserts coercions and checks them for
+correctness.
+
+\textit{Soft typing}~\citep{cartwright1991soft} starts
+with a dynamically-typed language, and layers a static
+type system with a sophisticated global type inference
+algorithm on top, to try to find errors in programs
+without needing to rewrite them. In cases where an error
+may or may not be present, it warns the programmer and
+inserts a runtime check. One problem with the soft
+typing approach was the complexity of the inferred
+types, leading to errors that were difficult to understand
+and fix.
+
+\textit{Dynamic typing}~\citep{henglein1994dts} is
+another approach for optimizing dynamically-typed programs.
+First the program is translated to a program that uses
+a {\tt Dynamic} type and explicit coercions and runtime checks,
+then a static analysis removes some  of these coercions and
+checks.
+
+Instead of trying to add static checking to a dynamic
+language, \citet{findler2002chf} enhances
+the dynamic checks with the possibility of {\em contracts}
+that give assertions about the input and output of (possibly
+higher-order) functions. In case of higher-order functions,
+the actual failing check can be far away from the actual
+source of the error, so contracts can also add {\em blame
+annotations} to values as a way to trace failures back to
+the source.
 
 Strongtalk \citep{bracha1993strongtalk,bracha1996strongtalk} is
-a version of Smalltalk that comes with an optional type system that
-does not include any type inference.
-
-Dart \citep{dart} and TypeScript \citep{typescript} are new
-languages that are designed with an optional type system, and both
-use JavaScript as their code generation target because their main
-purpose is web development.
-In fact, Dart is a new class-based object-oriented language with
-optional type annotations while TypeScript is a strict superset of
-JavaScript that provides optional type annotations and class-based
-object-oriented programming.
-Both Dart and TypeScript perform type inference to deduce static
-types for unannotated code.
-
-\subsection{Gradual Typing}
+an optionally-typed version of Smalltalk. It has a
+polymorphic structural type system that programmers can use
+to annotate Smalltalk programs, but type annotations can be
+left out; unnanotated terms are dynamically typed, and can
+be cast to any static type. The interaction of the dynamic
+type with the rest of the type system is unsound, so
+Strongtalk uses the dynamically-checked semantics of
+Smalltalk when executing programs, even if the programs
+are statically typed.
+
+Pluggable type systems~\citep{bracha2004pluggable} generalize
+the idea of Strongtalk, to have type systems that can be
+layered on top of a dynamic language without influencing
+its runtime semantics. These systems can be unsound in
+themselves, or in their interaction with the dynamically
+typed part of the language, without sacrificing runtime
+safety, as the semantics of the language catch any
+runtime errors caused by an unsound type system.
+
+Dart~\citep{dart} and TypeScript~\citep{typescript} are
+two recent examples of languages with optional type systems
+in this style. Dart is an object-oriented language with
+a semantics that is similar to Smalltalk's, while TypeScript
+is an object-oriented extension of JavaScript. Dart has a
+nominal type system, while TypeScript has a structural one,
+but both type systems are designed on purpose with unsound
+parts (such as covariant arrays in case of Dart, and
+covariant function return types in case of TypeScript) to
+increase programmer convenience. The interaction of
+statically and dynamically typed code is also unsound.
+
+...........HERE........
 

### 2014-03-15 doc/paper/typedlua.tex
finished rewrite of related work section

@@
 
-...........HERE........
-
-The main goal of gradual typing \citep{siek2006gradual} is to allow
-programmers to choose between static and dynamic typing in the same
-language.
-To do that, \citet{siek2006gradual} extended the simply typed
-lambda calculus with the dynamic type $?$.
-In gradual typing, type annotations are optional, so an untyped
-variable is syntactic sugar for a variable whose declared type is
-the dynamic type $?$.
-Under these circumstances, we view gradual typing as a way to add
-a dynamic type to statically typed languages.
-
-The central idea of gradual typing is the \textit{consistency}
-relation, written $\tau_{1} \sim \tau_{2}$, that allows implicit
-conversions to and from the dynamic type, and disallows conversions
-between inconsistent types \citep{siek2006gradual}.
-For instance, $\mathbf{int} \sim \,?$ and $?\, \sim \mathbf{int}$,
-but $\mathbf{int} \not\sim \mathbf{string}$ and
-$\mathbf{string} \not\sim \mathbf{int}$. 
-The consistency relation is neither commutative nor transitive.
-
-Gradual typing is similar to the approaches proposed by
-\citet{abadi1989dts} and \citet{thatte1990qst} by including a
-dynamic type to a statically typed language.
-However, these three approaches differ in the way they handle the
-dynamic type.
-While \citet{siek2006gradual} rely on the consistency relation,
-\citet{abadi1989dts} rely on type equality with explicit projections
-and injections, and \citet{thatte1990qst} relies on subtyping.
-The subtyping relation actually is a pitfall on Thatte's quasi-static
-typing, because the dynamic type is put as the top and the bottom of
-the subtyping relation.
-Therefore, downcasts combined with the transitivity of subtyping
-accepts programs that should be rejected.
-
-Later, \citet{siek2007objects} discovered that the consistency relation
-is orthogonal to subtyping, so we can combine the two relations to
-achieve both consistency and subtyping.
-The \textit{consistent-subtyping} relation,
-written $\tau_{1} \lesssim \tau_{2}$,
-is essential to design object-oriented languages that have a
-gradual type system.
-Like the consistency relation, and unlike the subtyping relation,
-the consistent-subtyping relation is not transitive.
-
-Another important feature of gradual typing is the theoretic
-foundation that it provides for inserting run-time checks that
-prove dynamically typed code does not violate the invariants of
-statically typed code, thus preserving type safety.
-To do that, \citet{siek2006gradual} defined the run-time semantics
-of gradual typing as a translation to an intermediate language with
-explicit casts at the frontiers between statically and dynamically
-typed code.
-The semantics of these casts is based on the higher-order contracts
-proposed by \citet{findler2002chf}.
-
-However, \citet{herman2007sgt} showed that there is an efficiency
-concern regarding the run-time checks, because there are two
-ways that casts can lead to unbounded space consumption.
-The first affects tail recursion while the second appears when
-first-class functions or objects cross the border between
-static code and dynamic code, that is, some programs can apply
-repeated casts to the same function or object.
-\citet{herman2007sgt} use the coercion calculus of
-\citet{henglein1994dts} to solve the problem of space efficiency,
-that is, they use coercions to express casts.
-Their approach normalizes an arbitrary sequence of coercions to a
-coercion of bounded size.
-
-Another concern about casts is how to improve debugging support,
-because a cast application can be delayed and the error related
-to that cast application can appear far way from the real error.
-\citet{wadler2009wpc} developed the \textit{blame calculus} as a way
-to attack this issue, and \citet{ahmed2011bfa} extended the
-blame calculus with polymorphism.
-The blame calculus is an intermediate language to integrate
-static and dynamic typing that uses the blame tracking approach
-proposed by \citet{findler2002chf}.
-
-On the one hand, the blame calculus solves the issue regarding
-error reporting;
-on the other hand, it has the space efficiency problem reported
-by \citet{herman2007sgt}.
-Thus, \citet{siek2009casts} extended the coercion calculus of
-\citet{herman2007sgt} with blame tracking to achieve an
-implementation of the blame calculus that is space efficient.
-After that, \citet{siek2010blame} proposed a new solution that also
-attacks both problems.
-This new solution is based on a concept called \textit{threesome},
-which is a way to split a cast between two parties into two casts
-among three parties.
-A cast has a source and a target type (a \textit{twosome}),
-so we can split any cast into a downcast from the source to an
-intermediate type that is followed by an upcast from the intermediate
-type to the target type (a \textit{threesome}).
-
-ActionScript \citep{moock2007as3} is the first language that
-incorporated a gradual type system to its implementation and
-Perl 6 \citep{tang2007pri} is also being designed with a
-gradual type system, though there is little documentation about
-the gradual type systems of these languages.
-Grace \citep{black2012grace,black2013sg} is a new object-oriented,
-gradually typed, educational language.
-In Grace, modules are gradually typed objects, that is, modules
-may have types and methods as attributes, and can have a state
-\citep{homer2013modules}.
-Gradual Python \citep{reticulated} is an attempt to develop a
-gradual type system for Python that deals with mutable objects
-\citep{siek2013mutable}.
-Gradualtalk \citep{allende2013gts} is a gradually-typed Smalltalk
-that introduces a new cast insertion strategy for gradually-typed
-objects \citep{allende2013cis}.
-
-\subsection{Approaches that are often called Gradual Typing}
-
-Gradual typing is similar to optional type systems in that type
-annotations are optional, and unannotated code is dynamically
-typed, but unlike optional type systems, gradual typing changes 
-the run-time semantics to preserve type safety, and it is a way to
-add a dynamic type to statically typed languages. 
-More precisely, programming languages that include a gradual type
-system implement the semantics of statically typed languages, so
-the gradual type system inserts casts in the translated code to
-guarantee that types are consistent before execution, while
-programming languages that include an optional type system still
-implement the semantics of dynamically typed languages, so all
-the type checking also belongs to the semantics of each operation.
-
-Still, we can view gradual typing as a way to formalize an optional
-type system when the gradual type system does not insert run-time
-checks.
-BabyJ \citep{anderson2002babyj} and Alore \citep{lehtosalo2011alore}
-are two examples of object-oriented languages that use gradual
-typing to formalize their optional type systems.
-The optional type systems of both BabyJ and Alore are nominal.
-BabyJ uses the relation $\approx$ that is similar to the consistency
-relation while Alore combines  subtyping along with the consistency
-relation to define a \textit{consistent-or-subtype} relation.
-The consistent-or-subtype relation is different from the
-consistent-subtyping relation of \citet{siek2007objects}, but it is
-also written $\tau_{1} \lesssim \tau_{2}$.
-The consistent-or-subtype relation holds when $\tau_{1} \sim \tau_{2}$
-or $\tau_{1} <: \tau_{2}$, so $<:$ is transitive and $\sim$ is not.
-Alore also extends its optional type system to include optional
-monitoring of run-time type errors in the gradual typing style.
-
-Hence, optional type annotations for software evolution probably are
-the reasons why optional type systems are commonly called
-gradual type systems.
-Typed Clojure \citep{bonnaire-sergeant2012typed-clojure} is an
-optional type system for Clojure that now is adopting the
-gradual typing slogan.
-
-\citet{flanagan2006htc} introduced \textit{hybrid type checking},
-an approach that combines static types and refinement types.
-For instance, programmers can specify the refinement type
-$\{x:Int \;|\; x \ge 0\}$ when they need a type for natural numbers.
-The programmer can also choose between explicit or implicit casts.
-When casts are not explicit, the type checker uses a theorem prover
-to insert casts.
-In our example of natural numbers, a cast would be inserted to check
-whether an integer is greater than or equal to zero.
-
-Sage \citep{gronski2006sage} is a programming language that
-extended hybrid type checking with a dynamic type to
-support dynamic and static typing in the same language.
-Sage also offers optional type annotations in the gradual typing
-style, that is, unannotated code is syntactic sugar for
-code whose declared type is the dynamic type.
-
-Thus, the inclusion of a dynamic type to hybrid type checking
-along with optional type annotations, and the insertion of run-time
-checks probably are the reasons why hybrid type checking is
-also viewed as a form of gradual typing.
-
-\citet{tobin-hochstadt2006ims} proposed another approach for gradually
-migrating from dynamically typed to statically typed code,
-and they coined the term \textit{from scripts to programs} for
-referring to this kind of interlanguage migration.
-In their approach, the migration from dynamically typed to
-statically typed code happens module-by-module, so they designed
-and implemented Typed Racket \citep{tobin-hochstadt2008ts} for
-this purpose.
-Typed Racket is a statically typed version of Racket
-(a Scheme dialect) that allows the programmer to write typed modules,
-so Typed Racket modules can coexist with Racket modules,
-which are untyped.
-
-The approach used by \citet{tobin-hochstadt2008ts} to design and
-implement Typed Racket probably is also called gradual typing
-because it allows the programmer to gradually migrate from untyped
-scripts to typed programs.
-However, note that Typed Racket is a statically typed language
-and type annotations are mandatory.
-So, what makes it gradual is a type system with a dynamic type
-that handles the interaction between Racket and Typed Racket modules.
-
-\subsection{Similar Lua projects}
-
-Metalua \citep{metalua} is a Lua compiler that supports compile-time
-metaprogramming (CTMP).
-CTMP is a kind of macro system that allows the programmers to interact
-with the compiler \citep{fleutot2007contrasting}. 
-Metalua extends Lua 5.1 syntax to include its macro system,
-and allows programmers to define their own syntax.
-Metalua can provide syntactical support for several object-oriented
-styles, and can also provide syntax for turning simple type
-annotations into run-time assertions.
-
-MoonScript \citep{moonscript} is a programming language that supports
-class-based object-oriented programming, and compiles to idiomatic
-Lua code.
-However, it does not perform compile-time type checking.
-
-LuaInspect \citep{luainspect} is a tool that uses MetaLua to perform
-some code analysis, such as flagging unknown global variables and
-table fields, checking the number of function arguments against
-signatures, and inferring function return values, but it does not
-try to analyze object-oriented code and does not perform compile-time
-type checking.
-
-Tidal Lock \citep{tidallock} is a prototype of another optional type
-system for Lua, which is written in Metalua.
-A remarkable feature of Tidal Lock is the way that it handles tables,
-because it allows the programmer to incrementally build the type of
-a table.
-More precisely, Tidal Lock allows the programmer to create an empty
-table to build its type according to the keys that the programmer
-inserts in it, so the programmer does not need to build the table in
-a single table constructor.
-Typed Lua has a simpler form of this feature, with a different
-formalization.
-Like Typed Lua, Tidal Lock also has local type inference.
-Unlike Typed Lua, Tidal Lock is not typing the common object-oriented
-idioms used by Lua programmers.
+\citep{tobin-hochstadt2006ims} shows how programs
+in the untyped lambda calculus can be incrementally
+translated to the simply typed lambda calculus, using
+contracts to guarantee that the untyped part cannot
+cause errors in the typed part, as a model on how
+scripts in a dynamically typed language can be
+incrementally translated to a statically typed
+language. This approach has been realized in the
+Typed Scheme (later Typed Racket) language~\citet{tobin-hochstadt2008ts}.
+
+Gradual typing~\citep{siek2006gradual} combines the
+optional type annotations of optional and pluggable
+type systems with higher-order contracts. The gradual
+type system is the simply typed lambda calculus enriched
+with a {\em dynamic type} $?$, where a value with the
+dynamic type can assume any type, and vice-versa.
+A dynamic value that assumes a static type generates
+a runtime check for a first-order value, or a wrapper
+for a higher-order value. A static value that assumes
+the dynamic type is tagged. The whole system is sound:
+any runtime errors in a well-typed program must happen
+in the dynamic parts.
+
+While both gradual typing and the approach of
+\citep{tobin-hochstadt2006ims} have the goal of
+having a sound interaction of dynamically and statically
+typed code, they differ in the granularity, with the
+gradual typing providing a finer-grained transition from
+dynamically typed to statically typed code.
+
+Gradual typing has been combined with subtyping in
+a simple object calculus~\citet{siek2007objects}, with
+a nominal, polymorphic type system~\citet{ina:gradual},
+with a row polymorphism~\citet{tobin:gradual},
+and with the polymorphic lambda calculus~\citet{ahmed2011bfa}.
+Gradual typing also adopted blame tracking from
+higher-order
+contracts~\citet{siek2010blame,ahmed2001bfa,wadler2009wpc}.
+
+While Typed Racket is the first fully-featured programming
+language with a sound mixture of dynamic and static typing,
+Gradualtalk~\citep{allende2013gts} is the first 
+fully-featured language with a fine-grained gradual type
+system. In Gradualtalk's case, the extra runtime checks
+needed by gradual typing impose a big runtime
+cost~\citep{allende2013cis}, and the programmer has the
+option of turning off these checks, and downgrading
+Gradualtalk to an optional type system.
+
+Tidal Lock \citep{tidallock} is a prototype of another optional
+type system for Lua. Compared with Typed Lua, Tidal Lock
+has richer record types, and a more robust system of
+incremental evolution of these record types, but it lacks
+unions and support for expressing objects and classes.
 

### 2014-03-16 doc/paper/typedlua.tex
fixing small typos, and giving more references to related work

@@
 Dynamically typed languages such as Lua avoid static types in favor of
-run-time {\em type tags} that classify the values they compute, and
-their implementations use these tags to perform run-time (or dynamic)
+runtime {\em type tags} that classify the values they compute, and
+their implementations use these tags to perform runtime (or dynamic)
 checking and guarantee that only valid operations are
@@
 to static types. The difference between these two approaches is the
-way they treat run-time semantics: while optional type systems
-do not affect the run-time semantics,
-gradual typing uses run-time checks to ensure that dynamically typed
+way they treat runtime semantics: while optional type systems
+do not affect the runtime semantics,
+gradual typing uses runtime checks to ensure that dynamically typed
 code does not violate the invariants of statically typed code.
@@
 Typed Lua allows statically typed Lua code to coexist and
-interact with  ynamically typed code. The Typed Lua compiler
+interact with dynamically typed code. The Typed Lua compiler
 warns the programmer about type errors, but always generates
@@
 programmer can enjoy some of the benefits of static types even
-without converting existing Lua modules to typed Lua:
+without converting existing Lua modules to Typed Lua:
 a dynamically typed module can export a statically typed
@@
 
-Unlike gradual type systems, Typed Lua does not insert run-time
+Unlike gradual type systems, Typed Lua does not insert runtime
 checks between dynamically and statically typed parts of the
@@
 The {\em dynamic type} $\Any$ allows dynamically typed code to
-interoperate with statically typed code; it is a subtype of $\Top$, but
-neither a supertype nor a subtype of any other type. We relate $\Any$
-to other types with the {\em consistency} and {\em consistent-subtype}
-relationships used by gradual type systems. In practice, we can
+interoperate with statically typed code; it is a subtype of 
+$\Top$, but
+neither a supertype nor a subtype of any other type. We relate 
+$\Any$
+to other types with the {\em consistency} and {\em consistent-
+subtype}
+relationships used by gradual type systems~\citep{siek2006gradual,siek2007objects}.
+ In practice, we can
 pass a value of the dynamic type anytime we want a value of some
@@
 \multicolumn{3}{c}{\textbf{Second-class types}} \\
-S ::= &  \multicolumn{2}{l}{\Void \; |\; V \; | \; S \cup S} \\
+S ::= &  \multicolumn{2}{l}{\; V \; | \; V \cup S} \\
 V ::= & \;\; T & \\
@@
 declarations. It assigns the dynamic type to any parameter that
-does not have an type annotation, but assigns more precise
+does not have a type annotation, but assigns more precise
 types to unnanotated variables, based on the type of the
@@
 
-As we saw in Figure \ref{fig:typelang}, a second-class type in
-Typed Lua can be the type $\Void$ (the empty tuple),
-a tuple of first-class types optionally ending in a variadic type,
-or a union of these tuples. A variadic type $T*$ is a generator for a
+As we can see in Figure \ref{fig:typelang}, a second-class
+type in Typed Lua can be a tuple of first-class types
+optionally ending in a variadic type, or a union of these
+tuples. A variadic type $T*$ is a generator for a
 sequence of values of type $T \cup \Nil$. Unions of tuples play
-an important part in functions that are overloaded on the return type,
-together with {\em projection types}. Both are explained in the next
+an important part in functions that are overloaded on the
+return type, together with {\em projection types}. Both are
+explained in the next
 section.
@@
 
-In this example, the second parameter is optional but, in the first
-line of the function, we declare a new variable that is guaranteed to
-have type $\String$ instead of $\String \cup \Nil$. In Lua, any value
-except {\tt nil} and {\tt false} are "truthy", so the short-circuiting
-{\tt or} operator is a common way of giving a default value to an
-optional parameter. Typed Lua encodes this idiom with a typing rule: if
-the left side of {\tt or} has type $T \cup Nil$ and the right side
-has type $T$ then the {\tt or} expression has type $T$.
-
-Declaring a new {\tt greeting} variable that shadows the parameter
-is not necessary:
+In this example, the second parameter is optional but, in the
+first line of the function, we declare a new variable that is
+guaranteed to have type $\String$ instead of $\String \cup \Nil$.
+In Lua, any value except {\tt nil} and {\tt false} are ``truthy",
+so the short-circuiting {\tt or} operator is a common way of
+giving a default value to an optional parameter. Typed Lua
+encodes this idiom with a typing rule: if
+the left side of {\tt or} has type $T \cup \Nil$ and the right side has type $T$ then the {\tt or} expression has type $T$.
+
+Declaring a new {\tt greeting} variable that shadows the
+parameter is not necessary:
 
@@
 cases that respect a simple heuristic: the new type must be a
-subtype of the previous one, and the variable must be local
-to the current function. The new type only applies for the
-remainder of the current scope.
+subtype of the previous one, the variable must be local
+to the current function, and the variable must not have been
+assigned to in another function. The new type only applies for
+the remainder of the current scope. In the case of {\tt
+greeting}, the assignment on line three changes its type to
+$\String$.
 
@@
 to denote the same union\footnote{The parentheses are always
-necessary here: {\tt number?} is {\tt number|nil}, while {\tt (number)}
+necessary here: {\tt number?} is {\tt number|nil}, while {\tt (number)?}
 is {\tt (number)|(nil, string)}.}.
@@
 against some other type, we take the union of the corresponding
-component in each tuple. But if we a variable with a projection
+component in each tuple. But if a variable with a projection
 type appears in a type predicate, the predicate discriminates
@@
 
-Notice that we could also discriminate using {\tt r}, using
+Notice that we could also discriminate {\tt r} using
 {\tt type(r) == "number"} as our predicate, with the same
@@
 by field assignment, subject to two restrictions: the
-variable must be local to the current block, and the new type
-must be a subtype of the old type.
+variable must be local to the current block, the new type
+must be a subtype of the old type, and the variable
+cannot have been assigned to in another function.
 
@@
 
-Lua's built-in support for object oriented programming is
+Lua's built-in support for object-oriented programming is
 minimal. The basic mechanism is the {\tt :} syntactic
@@
       s.y = y
-      return shape
+      return s
     end
@@
                               y: number,
-                              radius: number)
+                              radius: value)
       local c = setmetatable(Shape:new(x, y),
                              { __index = self })
-      c.radius = radius
+      c.radius = tonumber(radius)
       return c
@@
 the overriden constructor. A limitation of this class
-system is that the overriden constructors must be a
-subtype of the original constructor.
+system is that the overriden constructor must be a
+subtype of the original constructor, so the type of
+{\tt radius} has to be very permissive.
 
@@
 
-\textit{Dynamic typing}~\citep{henglein1994dts} is
+While the goal of soft typing is catching errors,
+\textit{dynamic typing}~\citep{henglein1994dts} is
 another approach for optimizing dynamically-typed programs.
@@
 
-Tidal Lock \citep{tidallock} is a prototype of another optional
+Grace~\citet{black2013sg} is an object-oriented language
+with optional typing. Grace is not an dynamically typed
+language that has been extended with an optional type system,
+but a language that has been designed from the ground up
+to have both static and dynamic typing. \citet{homer2013modules}
+explores some useful patterns that derive from Grace's
+use of objects as modules and its brand of optional
+structural typing, which can also be expressed with
+Typed Lua's modules as tables.
+
+The main feature of Typed Racket's type system is
+{\em ocurrence typing}~\citep{tobin:occur}, where
+the type of a variable can be refined through tests
+using predicates that check the runtime type of
+the variable. As these types of checks are common in
+other languages, related systems have
+appeared~\citep{guha:flow,winther:flow,pearce:whiley}.
+
+Tidal Lock~\citep{tidallock} is a prototype of another optional
 type system for Lua. Compared with Typed Lua, Tidal Lock

### 2014-03-16 doc/paper/typedlua.tex
final touches for submission to dyla'14

@@
   \affaddr{Rio de Janeiro, Brazil}\\
-  \email{fabiom@dcc.ufrj.br}
+  \email{mascarenhas@ufrj.br}
 \alignauthor
@@
 \begin{abstract}
+Dynamically typed languages trade flexibility and ease of
+use for safety, while statically typed languages prioritize
+the early detection of bugs, and provide a better framework
+for structure large programs. The idea of optional typing
+is to combine the two approaches in the same
+language: the programmer can begin development with
+dynamic types, and migrate to static types as the
+program matures. The challenge is designing a type system
+that feels natural to the programmer that is used to
+programming in a dynamic language.
+
+This paper presents the initial design of Typed Lua, an
+optionally-typed extension of the Lua scripting language.
+The design of Typed Lua's type system has a novel combination
+of features that preserves some of the idioms that Lua
+programmers are used to, while bringing static type
+safety to them.
 \end{abstract}
 
-\category{D.3.3}
-         {Programming Languages}
-         {Language Constructs and Features}
-\category{D.2.3}
-         {Programming Languages}
-         {Coding Tools and Techniques}
-         [Object-oriented programming]
-
-\terms{Languages,Verification}
-
-\keywords{Type Systems, Lua}
+\keywords{Lua programming language, type systems, optional typing, gradual typing}
 
@@
 As this migration involves languages with different syntax and
-semantics, it requires a complete rewrite of existing programs intead
+semantics, it requires a complete rewrite of existing programs instead
 of incremental evolution from dynamic to static types.
@@
 
-In this work we outline the design of Typed Lua:
+This paper presents the initial design of Typed Lua:
 an optional type system for Lua that is complex enough to
-preserve some of the idioms that Lua programmers are already used to,
-while adding new constructs that help programmers structure Lua
-programs.
+preserve some of the idioms that Lua programmers are already
+used to, while adding new constructs that help programmers
+structure Lua programs.
 
@@
 {\em number}, {\em string}, {\em function}, {\em table}, {\em userdata},
-and {\em thread}. We will see how Typed Lua assigns types to
-values of the first five in this section.
+and {\em thread}. In this section, we will see how Typed Lua assigns types to
+values of the first five.
 
@@
 does not have a type annotation, but assigns more precise
-types to unnanotated variables, based on the type of the
+types to unannotated variables, based on the type of the
 expression that gives the initial value of the variable.
@@
     end
-
     local function distance(x, y)
@@
 those of multiple assignment (or vice-versa); calling a function is like
-doing a multiple assigment where the left side is the parameter list
+doing a multiple assignment where the left side is the parameter list
 and the right side is the argument list.
@@
     end
-
     local function sum(x: number, y: number)
@@
     end
-
     local x, y, z = multiple(), multiple()
@@
 
-Functions that overload their return types to signal the ocurrence
+Functions that overload their return types to signal the occurrence
 of errors are another common Lua idiom. In this idiom, a function
@@
 {\tt type(r) == "number"} as our predicate, with the same
-result. The first form is more succint, and more idiomatic.
+result. The first form is more succinct, and more idiomatic.
 We can also use projection types to write overloaded functions
@@
     local t: { string: number? } = { foo = 1 }
-    local x: number = t.foo         -- error
+    local x: number = t.foo       -- compile error
     local y: number = t.bar or 0  -- y gets 0
@@
     end
-
     local user1 = { firstname = "Lewis",
@@
     local user4 = { "Lou", "Reed" }
-
     print(greeter(user1)) -- Hello, Lewis Reed
@@
     end
-
     local user1: Person = { firstname = "Lewis",
@@
 Taken individually, the changes to the type of the two
-variables look ok, but aliasing makes one of them unsound.
+variables look fine, but aliasing makes one of them unsound.
 The location of the change also matters, as the next example
@@
     end
-
     local user = {}
@@
 notice how we can call {\tt Shape}'s constructor inside
-the overriden constructor. A limitation of this class
-system is that the overriden constructor must be a
+the overridden constructor. A limitation of this class
+system is that the overridden constructor must be a
 subtype of the original constructor, so the type of
@@
 to annotate Smalltalk programs, but type annotations can be
-left out; unnanotated terms are dynamically typed, and can
+left out; unannotated terms are dynamically typed, and can
 be cast to any static type. The interaction of the dynamic
@@
 The main feature of Typed Racket's type system is
-{\em ocurrence typing}~\citep{tobin:occur}, where
+{\em occurrence typing}~\citep{tobin:occur}, where
 the type of a variable can be refined through tests
@@
 
-\section{Conclusion} \label{sec:con}
+\section{Conclusion}
+\label{sec:con}
+
+This paper presented the initial design of Typed Lua,
+an optional type system for the Lua scripting language.
+While Lua shares several characteristics with other
+dynamic languages such as JavaScript, so Typed Lua's
+type system has several parts in common with optional
+and gradual type systems for dynamic languages,
+Lua also has some uncommon features.
+
+These language features demanded features in the type
+system that are not present in the type systems of these
+other languages: functions with flexible arity and
+their interaction with multiple assignment, functions
+that are overloaded on the {\em number} of values
+they return, and how to support the idiomatic way of
+handling their return values, and incremental evolution
+of record and object types.
+
+Typed Lua is a work in progress: while the design covers
+most of the Lua language as it is actually used by Lua
+programmers, there are missing parts. A major omission
+of the current version is the lack of polymorphic function
+and table types, which will be present in the next version.
+The {\tt setmetatable} primitive that we briefly saw in
+Section~6 has several other uses that are currently
+outside of the scope of the type system: operator
+overloading, proxies, changing the behavior of built-in
+types, multiple inheritance, etc. Finally, Lua has
+one-shot delimited continuations~\citep{james2011yield}
+in the form of {\em couroutines}~\citep{moura:coro},
+which are also currently ignored by the type system.
+
+Unlike some of the other optional type systems,
+the type system of Typed Lua does not have deliberately
+unsound parts, but we still do not have proofs that the
+novel parts of our type system are sound.
+
+Finally, while we our survey of Lua modules gives some
+confidence that the design of Typed Lua is useful for
+the majority of Lua programs, we have to validate the
+design by using Typed Lua to to type a representative 
+sample of Lua modules.
 

### 2014-04-07 doc/thesis/conclusions.tex
Using the proposal text to add thesis files

Our main contribution is the design of a complete optional type system
for Lua, a procedural scripting language.
TypeScript is a complete optional type system for JavaScript,
but it does not include union types, and it uses arrays to represent
variadic functions and multiple return values.
Gradualtalk is a complete gradual type system for Smalltalk,
but Smalltalk is an object-oriented language that does
not handle multiple return values, multiple assignments, and
variadic functions.
Gradualtalk also mixes nominal and structural typing because of
the object-oriented nature of Smalltalk.
Typed Racket is a complete static type system for Racket, a functional
language.
Typed Racket makes it easier to migrate from untyped to typed code,
but the migration happens module-by-module and type annotations are
mandatory on typed modules.
There are research on the design of optional and gradual type systems
for JavaScript, Perl, and Python, but they are either experimental or
limited in their coverage.
Furthermore, Typed Lua includes table types that use types as keys,
instead of identifiers, and the fact that Typed Lua also includes
literal types makes its table types subsume the usual idea of
record types.

Although each language is different, we believe that several parts of
the design of Typed Lua should be relevant to other scripting languages.
Most notably, JavaScript shares with Lua the use of tests of type tags
as a way of encoding optional and overloaded parameters, and the
proliferation of class-based object systems built on top of the
language's delegation mechanisms.

Typed Lua should also be a major contribution to the Lua community,
because it offers a framework that programmers can use to document,
test, and better structure their applications.
For libraries where a full conversion to static type checking should
prove unfeasible or too much work, the community can use Typed Lua
just to document the external interfaces of the libraries,
giving the benefits of static type checking to the users of these
libraries.

Finally, Typed Lua can be a base for further research and development
of static type systems for the Lua language, such as the use of static
analysis of Lua API calls for checking that a native library is
exporting the interface to Lua code that it declares, the use of
effect types for typing Lua coroutines, and the use of static types
for code optimization.

### 2014-04-07 doc/thesis/introduction.tex
Using the proposal text to add thesis files

Dynamically typed languages such as Lua avoid static types in favor of
simplicity and flexibility, because the absence of static types means
that programmers do not need to bother about abstracting types that
should be validated by a type checker.
Instead, dynamically typed languages use run-time \textit{type tags}
to classify the values they compute, so their implementation can use
these tags to perform run-time (or dynamic) type checking
\citep{pierce2002tpl}.

This simplicity and flexibility allow programmers to write code that
might require a quite complex type system to statically type check,
though it may also hide bugs that will be caught only after deployment
if programmers do not properly test their code.
In contrast, static type checking helps programmers detect many 
bugs during the development phase.
Static types also provide a conceptual framework that helps
programmers define modules and interfaces that can be combined to
structure the development of programs.

Thus, early error detection and better program structure are two
advantages of static type checking that can lead programmers to
migrate their code from a dynamically typed to a statically
typed language, when their simple scripts become complex programs
\citep{tobin-hochstadt2006ims}.
Dynamically typed languages certainly help programmers during the
beginning of a project, because their simplicity and flexibility
allow quick development and make it easier to change code according to
changing requirements.
However, programmers tend to migrate from dynamically typed to
statically typed code as soon as the project has consolidated its
requirements, because the robustness of static types helps
programmers link requirements to abstractions.
This migration usually involves different languages that have
different syntax and semantics, which usually require a complete
rewrite of existing programs instead of incremental evolution from
dynamic to static types.

Ideally, programming languages should offer programmers the
option to choose between static and dynamic typing:
\textit{optional type systems} \citep{bracha2004pluggable} and
\textit{gradual typing} \citep{siek2006gradual} are two similar
approaches for blending static and dynamic typing in the same
language.
The aim of both approaches is to offer programmers the option
to use type annotations where static typing is needed and avoid them
where dynamic typing is sufficient.
The difference between these two approaches is the way they treat
run-time semantics.
While optional type systems do not affect the run-time semantics,
gradual typing uses run-time checks to ensure that dynamically typed
code does not violate the invariants of statically typed code.

Programmers and researchers sometimes use the term gradual typing
to mean the incremental evolution of dynamically typed code to
statically typed code.
For this reason, gradual typing may also refer to optional type
systems and other approaches that blend static and dynamic typing to
help programmers incrementally migrate from dynamic to static typing
without having to switch to a different language, though all these
approaches differ in the way they handle static and dynamic typing
together.
We use the term gradual typing for referring to the work of
\citet{siek2006gradual}.

In this work we propose the design and implementation of Typed Lua:
an optional type system for Lua that should be complex enough to
preserve some of the Lua idioms that programmers are already used to,
but that should also include new constructs that help programmers
structure Lua programs.

Lua is a small scripting language, so designing an optional type
system for Lua should shed some light on optional type systems for
scripting languages in general.
Lua provides mechanisms for organizing a program in modules and
writing object-oriented programs, but it does not set policy on how
these features should behave, due to its use as an embedded and
extension language.
Thus, implementing an optional type system for Lua should offer Lua
programmers one way to obtain most of the advantages of static typing
without compromising the simplicity and flexibility of dynamic typing.

So far, Typed Lua should be a strict superset of Lua that
provides optional type annotations, compile-time type checking, and
class-based object-oriented programming through the definition of
classes, interfaces, and modules.
In practice, we should implement Typed Lua as a programming
language that extends Lua syntax to add optional type annotations
and standard constructions for structuring Lua code.
The compiler should use static types to perform compile-time type
checking, but should also allow Lua code to coexist with Typed Lua
code, and generate Lua code that runs in unmodified Lua
implementations.
We have an ongoing implementation of the Typed Lua compiler that is
available online\footnote{https://github.com/andremm/typedlua}.

Typed Lua intended use is as an application language, and we view
that policies for organizing a program in modules and writing
object-oriented programs should be part of the language and
enforced by its optional type system.
An application language is a programming language that helps
programmers develop applications from scratch until these
applications evolve to complex systems rather than just scripts.

We also believe that Typed Lua will help programmers give a more
formal documentation to already existing Lua code, as static types
are also a useful source of documentation in languages that provide
type annotations, because type annotations are always validated by
the type checker and therefore never get outdated.
Thus, programmers can use Typed Lua to define axioms about the
interfaces and types of dynamically typed modules.
We will enforce this point by using Typed Lua to statically type
the interface of the Lua standard libraries and other commonly used
Lua libraries, so our compiler can check their usage by Typed Lua
code.

Typed Lua will perform a very limited form of local type inference
\citep{pierce2000lti}, as static typing does not necessarily mean
that programmers need to insert type annotations in the code.
Several statically typed languages such as Haskell provide some
amount of type inference that automatically deduces the types of
expressions.
Still, Typed Lua will need a small amount of type annotations due
to the nature of its optional type system.

Typed Lua will not deal with code optimization, although another
important advantage of static types is that they help the compiler
perform optimizations and generate more efficient code.
However, we believe that the formalization of our optional type
system will be precise enough to aid optimization in some Lua
implementations.

We will use some of the ideas of gradual typing to formalize Typed Lua.
Even though Typed Lua is an optional type system and thus will not
include run-time checks between dynamic and static regions of the
code, we believe that using the foundations of gradual typing to
formalize our optional type system will allow us to include run-time
checks in the future.

%In Section \ref{sec:review} we review the literature about static and
%dynamic typing in the same language, which includes the discussion
%about the particularities between optional type systems and gradual
%typing.
%In Section \ref{sec:proposal} we describe the reasons why Typed Lua
%includes features of object-oriented programming and the status of
%its design and implementation.
%In Section \ref{sec:con} we outline our expected contributions.
%In Section \ref{sec:plan} we present the work plan for thesis completion.

### 2014-04-07 doc/thesis/review.tex
Using the proposal text to add thesis files

We begin this section presenting a little bit of the history about
combining static and dynamic typing in the same language.
After that, we introduce optional type systems and gradual typing.
We end this section discussing why optional type systems and two
other approaches are often called gradual typing.

\section{A little bit of history}

Common LISP \citep{steele1982ocl} introduced optional type annotations
in the early eighties, but not for static type checking.
Instead, programmers may choose to declare types of variables as
optimization hints to the compiler, that is, type declarations are
just one way to help the compiler to optimize code.
These annotations are unsafe because they can crash the program
when they are wrong.

\citet{abadi1989dts} extended the simply typed lambda calculus with the
\texttt{Dynamic} type and the \texttt{dynamic} and \texttt{typecase}
constructs, with the aim to safely integrate dynamic code in
statically typed languages.
The \texttt{Dynamic} type is a pair \texttt{(v,T)} where \texttt{v} is a
value and \texttt{T} is the tag that represents the type of \texttt{v}.
The constructs \texttt{dynamic} and \texttt{typecase} are explicit
injection and projection operations, respectively.
That is, \texttt{dynamic} builds values of type \texttt{Dynamic} and
\texttt{typecase} safely inspects the type of a \texttt{Dynamic} value.
Thus, migrating code between dynamic and static type checking requires
changing type annotations and adding or removing \texttt{dynamic} and
\texttt{typecase} constructs throughout the code.

The \textit{quasi-static} type system proposed by \citet{thatte1990qst}
performs implicit coercions and run-time checks to replace the
\texttt{dynamic} and \texttt{typecase} constructs that were proposed by
\citet{abadi1989dts}.
To do that, quasi-static typing relies on subtyping with a top type
$\Omega$ that represents the dynamic type, and splits type checking
into two phases.
The first phase inserts implicit coercions from the dynamic type to
the expected type, while the second phase performs what Thatte calls
\textit{plausibility checking}, that is, it rewrites the program to
guarantee that sequences of upcasts and downcasts always have a
common subtype.

\textit{Soft typing} \citep{cartwright1991soft} is another approach
to combine static and dynamic typing in the same language.
The main goal of soft typing is to add static type checking to
dynamically typed languages without compromising their flexibility.
To do that, soft typing relies on type inference for
translating dynamically typed code to statically typed code.
The type checker inserts run-time checks around inconsistent code and
warns the programmer about the insertion of these run-time checks,
as they indicate the existence of potential type errors.
However, the programmer is free to choose between inspecting the
run-time checks and simply running the code.
Soft typing gives this choice to the programmer because it is one of
its key ideas, that is, type inference and static type checking should
not prevent the programmer from running inconsistent code.
One advantage of soft typing is the fact that the compiler for
softly typed languages can use the translated code to generate
more efficient code, as the translated code statically type checks.
One disadvantage of soft typing is that it can be cumbersome when
the inferred types are meaningless large types that just confuse the
programmer.

\textit{Dynamic typing} \citep{henglein1994dts} is an approach
that optimizes code from dynamically typed languages by eliminating
unnecessary checks of tags.
The paper describes how to translate dynamically typed code to
statically typed code that uses a \texttt{Dynamic} type.
The translation is done through a coercion calculus that uses type
inference to insert the operations that are necessary to type check
the \texttt{Dynamic} type during run-time.
Although soft typing and dynamic typing may seem similar, they are not.
They are not similar because soft typing targets statically type
checking of dynamically typed languages for detecting programming
errors while dynamic typing targets the optimization of dynamically
typed code through the elimination of unnecessary run-time checks.
In other words, soft typing sees code optimization as a side effect
that comes with static type checking.

\citet{findler2002chf} proposed contracts for higher-order functions
and blame annotations for run-time checks.
Contracts perform dynamic checking instead of static checking,
but deferring all verifications to run-time can lead to defects
that are difficult to fix, because run-time errors can show a
stack trace where it is not clear to programmers if the cause of a
certain run-time error is in application code or library code.
Even if programmers identify that the source of a certain run-time
error is in library code, they still may have problems to identify
if this run-time error is due to a violation of library's contract,
or due to a bug, when the library is poorly documented.
In this approach, programmers can insert assertions in the form of
contracts that check the input and output of higher-order functions.
Thus, \citet{findler2002chf} show how programmers can insert
assertions in the form of contracts that check the input and output
of higher-order functions; and they also show how to add blame
annotations in the generated code as a way to track assertion
failures back to the source of the error.

BabyJ \citep{anderson2003babyj} is an object-oriented language
without inheritance that allows programmers to incrementally annotate
the code with more specific types.
Programmers can choose between using the dynamically typed version
of BabyJ, when they do not need types at all, and the statically
typed version of BabyJ, when they need to annotate the code.
In the statically typed BabyJ, programmers can use the
\textit{permissive type} $*$ to annotate the parts of the code that
still do not have a specific type or the parts of the code that should
have dynamic behavior.
The type system of BabyJ is nominal, so types are either class names
or the permissive type $*$.
However, the type system does not use type equality or subtyping,
but the relation $\approx$ between two types.
The relation $\approx$ holds when both types have the same name or
any of them is the permissive type $*$.
Even though the permissive type $*$ is similar to the dynamic type
from previous approaches, BabyJ does not provide any way to add
implicit or explicit run-time checks.

\citet{ou2004dtd} specified a language that combines static types
with dependent types.
To ensure safety, the compiler automatically inserts coercions
between dependent code and static code.
The coercions are run-time checks that ensure static code does not
crash dependent code during run-time.

\section{Optional Type Systems}

Optional type systems \citep{bracha2004pluggable} are an approach for
plugging static typing in dynamically typed languages through
optional type annotations that are used to perform compile-time type
checking, though not influencing the original run-time semantics of
the language.
This means that optional type systems can be unsound without affecting
language safety, because the run-time semantics should still catch
type errors independently of the static type checking.
More precisely, optional type systems are unsound if they do not catch
some type errors, but we say that languages with unsound optional type
systems are safe because the semantics of these languages catch during
run-time the type errors that the optional type system did not catch.
Although optional type systems maintain the simplicity and
flexibility of dynamically typed languages, their static type
checking cannot prove the absence of all type errors when the optional
type system is unsound.

Optional type systems have an interesting relation to the untyped
lambda calculus and the typed lambda calculus.
The semantic rules of the unytped lambda calculus are the same of the
typed lambda calculus, and the type system is actually optional and
serves only for discarding programs that may have undesired behaviors
\citep{bracha2004pluggable}.

Optional type systems make software evolution easier and
allow language designers to experiment with
\textit{pluggable type systems} \citep{bracha2004pluggable}.
Pluggable type systems are different optional type systems for
checking different aspects of the program.

According to \citet{bracha2004pluggable}, optional type systems should
not include mandatory type inference too, because mandatory type
inference may decrease the expressiveness of the type system.
Instead, optional type systems should implement type inference as
an independent mechanism that can be turned on and off according to
the programmers' needs.

Unfortunately, there is no formalization on optional type systems,
so each language ends up implementing its optional type system on
its own way.
Now we will briefly discuss some languages that provide an informal
description of their optional type systems rather than a formal one.

Strongtalk \citep{bracha1993strongtalk,bracha1996strongtalk} is
a version of Smalltalk that comes with an optional type system that
does not include any type inference.

Dart \citep{dart} and TypeScript \citep{typescript} are new
languages that are designed with an optional type system, and both
use JavaScript as their code generation target because their main
purpose is web development.
In fact, Dart is a new class-based object-oriented language with
optional type annotations while TypeScript is a strict superset of
JavaScript that provides optional type annotations and class-based
object-oriented programming.
Both Dart and TypeScript perform type inference to deduce static
types for unannotated code.

\section{Gradual Typing}

The main goal of gradual typing \citep{siek2006gradual} is to allow
programmers to choose between static and dynamic typing in the same
language.
To do that, \citet{siek2006gradual} extended the simply typed
lambda calculus with the dynamic type $?$.
In gradual typing, type annotations are optional, so an untyped
variable is syntactic sugar for a variable whose declared type is
the dynamic type $?$.
Under these circumstances, we view gradual typing as a way to add
a dynamic type to statically typed languages.

The central idea of gradual typing is the \textit{consistency}
relation, written $\tau_{1} \sim \tau_{2}$, that allows implicit
conversions to and from the dynamic type, and disallows conversions
between inconsistent types \citep{siek2006gradual}.
For instance, $\mathbf{int} \sim \,?$ and $?\, \sim \mathbf{int}$,
but $\mathbf{int} \not\sim \mathbf{string}$ and
$\mathbf{string} \not\sim \mathbf{int}$. 
The consistency relation is neither commutative nor transitive.

Gradual typing is similar to the approaches proposed by
\citet{abadi1989dts} and \citet{thatte1990qst} by including a
dynamic type to a statically typed language.
However, these three approaches differ in the way they handle the
dynamic type.
While \citet{siek2006gradual} rely on the consistency relation,
\citet{abadi1989dts} rely on type equality with explicit projections
and injections, and \citet{thatte1990qst} relies on subtyping.
The subtyping relation actually is a pitfall on Thatte's quasi-static
typing, because the dynamic type is put as the top and the bottom of
the subtyping relation.
Therefore, downcasts combined with the transitivity of subtyping
accepts programs that should be rejected.

Later, \citet{siek2007objects} discovered that the consistency relation
is orthogonal to subtyping, so we can combine the two relations to
achieve both consistency and subtyping.
The \textit{consistent-subtyping} relation,
written $\tau_{1} \lesssim \tau_{2}$,
is essential to design object-oriented languages that have a
gradual type system.
Like the consistency relation, and unlike the subtyping relation,
the consistent-subtyping relation is not transitive.

Another important feature of gradual typing is the theoretic
foundation that it provides for inserting run-time checks that
prove dynamically typed code does not violate the invariants of
statically typed code, thus preserving type safety.
To do that, \citet{siek2006gradual} defined the run-time semantics
of gradual typing as a translation to an intermediate language with
explicit casts at the frontiers between statically and dynamically
typed code.
The semantics of these casts is based on the higher-order contracts
proposed by \citet{findler2002chf}.

However, \citet{herman2007sgt} showed that there is an efficiency
concern regarding the run-time checks, because there are two
ways that casts can lead to unbounded space consumption.
The first affects tail recursion while the second appears when
first-class functions or objects cross the border between
static code and dynamic code, that is, some programs can apply
repeated casts to the same function or object.
\citet{herman2007sgt} use the coercion calculus of
\citet{henglein1994dts} to solve the problem of space efficiency,
that is, they use coercions to express casts.
Their approach normalizes an arbitrary sequence of coercions to a
coercion of bounded size.

Another concern about casts is how to improve debugging support,
because a cast application can be delayed and the error related
to that cast application can appear far way from the real error.
\citet{wadler2009wpc} developed the \textit{blame calculus} as a way
to attack this issue, and \citet{ahmed2011bfa} extended the
blame calculus with polymorphism.
The blame calculus is an intermediate language to integrate
static and dynamic typing that uses the blame tracking approach
proposed by \citet{findler2002chf}.

On the one hand, the blame calculus solves the issue regarding
error reporting;
on the other hand, it has the space efficiency problem reported
by \citet{herman2007sgt}.
Thus, \citet{siek2009casts} extended the coercion calculus of
\citet{herman2007sgt} with blame tracking to achieve an
implementation of the blame calculus that is space efficient.
After that, \citet{siek2010blame} proposed a new solution that also
attacks both problems.
This new solution is based on a concept called \textit{threesome},
which is a way to split a cast between two parties into two casts
among three parties.
A cast has a source and a target type (a \textit{twosome}),
so we can split any cast into a downcast from the source to an
intermediate type that is followed by an upcast from the intermediate
type to the target type (a \textit{threesome}).

ActionScript \citep{moock2007as3} is the first language that
incorporated a gradual type system to its implementation and
Perl 6 \citep{tang2007pri} is also being designed with a
gradual type system, though there is little documentation about
the gradual type systems of these languages.
Grace \citep{black2012grace,black2013sg} is a new object-oriented,
gradually typed, educational language.
In Grace, modules are gradually typed objects, that is, modules
may have types and methods as attributes, and can have a state
\citep{homer2013modules}.
Gradual Python \citep{reticulated} is an attempt to develop a
gradual type system for Python that deals with mutable objects
\citep{siek2013mutable}.
Gradualtalk \citep{allende2013gts} is a gradually-typed Smalltalk
that introduces a new cast insertion strategy for gradually-typed
objects \citep{allende2013cis}.

\section{Approaches that are often called Gradual Typing}

Gradual typing is similar to optional type systems in that type
annotations are optional, and unannotated code is dynamically
typed, but unlike optional type systems, gradual typing changes 
the run-time semantics to preserve type safety, and it is a way to
add a dynamic type to statically typed languages. 
More precisely, programming languages that include a gradual type
system implement the semantics of statically typed languages, so
the gradual type system inserts casts in the translated code to
guarantee that types are consistent before execution, while
programming languages that include an optional type system still
implement the semantics of dynamically typed languages, so all
the type checking also belongs to the semantics of each operation.

Still, we can view gradual typing as a way to formalize an optional
type system when the gradual type system does not insert run-time
checks.
BabyJ \citep{anderson2003babyj} and Alore \citep{lehtosalo2011alore}
are two examples of object-oriented languages that use gradual
typing to formalize their optional type systems.
The optional type systems of both BabyJ and Alore are nominal.
BabyJ uses the relation $\approx$ that is similar to the consistency
relation while Alore combines  subtyping along with the consistency
relation to define a \textit{consistent-or-subtype} relation.
The consistent-or-subtype relation is different from the
consistent-subtyping relation of \citet{siek2007objects}, but it is
also written $\tau_{1} \lesssim \tau_{2}$.
The consistent-or-subtype relation holds when $\tau_{1} \sim \tau_{2}$
or $\tau_{1} <: \tau_{2}$, so $<:$ is transitive and $\sim$ is not.
Alore also extends its optional type system to include optional
monitoring of run-time type errors in the gradual typing style.

Hence, optional type annotations for software evolution probably are
the reasons why optional type systems are commonly called
gradual type systems.
Typed Clojure \citep{bonnaire-sergeant2012typed-clojure} is an
optional type system for Clojure that now is adopting the
gradual typing slogan.

\citet{flanagan2006htc} introduced \textit{hybrid type checking},
an approach that combines static types and refinement types.
For instance, programmers can specify the refinement type
$\{x:Int \;|\; x \ge 0\}$ when they need a type for natural numbers.
The programmer can also choose between explicit or implicit casts.
When casts are not explicit, the type checker uses a theorem prover
to insert casts.
In our example of natural numbers, a cast would be inserted to check
whether an integer is greater than or equal to zero.

Sage \citep{gronski2006sage} is a programming language that
extended hybrid type checking with a dynamic type to
support dynamic and static typing in the same language.
Sage also offers optional type annotations in the gradual typing
style, that is, unannotated code is syntactic sugar for
code whose declared type is the dynamic type.

Thus, the inclusion of a dynamic type to hybrid type checking
along with optional type annotations, and the insertion of run-time
checks probably are the reasons why hybrid type checking is
also viewed as a form of gradual typing.

\citet{tobin-hochstadt2006ims} proposed another approach for gradually
migrating from dynamically typed to statically typed code,
and they coined the term \textit{from scripts to programs} for
referring to this kind of interlanguage migration.
In their approach, the migration from dynamically typed to
statically typed code happens module-by-module, so they designed
and implemented Typed Racket \citep{tobin-hochstadt2008ts} for
this purpose.
Typed Racket is a statically typed version of Racket
(a Scheme dialect) that allows the programmer to write typed modules,
so Typed Racket modules can coexist with Racket modules,
which are untyped.

The approach used by \citet{tobin-hochstadt2008ts} to design and
implement Typed Racket probably is also called gradual typing
because it allows the programmer to gradually migrate from untyped
scripts to typed programs.
However, note that Typed Racket is a statically typed language
and type annotations are mandatory.
So, what makes it gradual is a type system with a dynamic type
that handles the interaction between Racket and Typed Racket modules.

### 2014-04-07 doc/thesis/typedlua.tex
Using the proposal text to add thesis files

We begin this section presenting statistics about the usage of Lua
features and idioms.
We collected statistics about how programmers use tables, functions,
dynamic type checking, object-oriented programming, and modules.
We shall see that these statistics informed important design decisions
on our optional type system.
We end this section introducing Typed Lua through examples, and
describing a few Lua projects that are similar to Typed Lua.

\section{Statistics about the usage of Lua}

We used the LuaRocks repository to build our statistics database;
LuaRocks \citep{hisham2013luarocks} is a package manager for Lua
modules.
We downloaded the 3928 \texttt{.lua} files that were available in
the LuaRocks repository at February 1st 2014.
However, we ignored files that are not compatible with Lua 5.2,
the latest version of Lua.
We also ignored \textit{machine-generated} files and test files,
because these files may not represent idiomatic Lua code,
and might skew our statistics towards non-typical uses of Lua.
This left 2598 \texttt{.lua} files from 262 different projects for
our statistics database;
we parsed these files and processed their abstract syntax tree
to gather the statistics that we show in this section.

To verify how programmers use tables, we measured how they
initialize, index, and iterate tables.
We present these statistics in the next three paragraphs to discuss
their influence on our type system.

The table constructor appears 23185 times.
In 36\% of the occurrences it is a constructor that initializes a
record (e.g., \texttt{\{ x = 120, y = 121 \}});
in 29\% of the occurrences it is a constructor that initializes a
list (e.g., \texttt{\{ "one", "two", "three", "four" \}});
in 8\% of the occurrences it is a constructor that initializes a
record with a list part;
and in less than 1\% of the occurrences (4 times) it is a constructor
that uses only the booleans \texttt{true} and \texttt{false} as indexes.
At all, in 73\% of the occurrences it is a constructor that uses
only literal keys;
in 26\% of the occurrences it is the empty constructor;
in 1\% of the occurrences it is a constructor with non-literal keys
only, that is, a constructor that uses variables and function calls
to create the indexes of a table;
and in less than 1\% of the occurrences (19 times) it is a constructor
that mixes literal keys and non-literal keys.

The indexing of tables appears 130448 times:
86\% of them are for reading a table field while
14\% of them are for writing into a table field.
We can classify the indexing operations that are reads as follows:
89\% of the reads use a literal string key,
4\% of the reads use a literal number key,
and less than 1\% of the reads (10 times) use a literal boolean key.
At all, 93\% of the reads use literals to index a table while
7\% of the reads use non-literal expressions to index a table.
It also worth mentioning that 45\% of the reads are actually
function calls.
More precisely, 25\% of the reads use literals to call a function,
20\% of the reads use literals to call a method, that is,
a function call that uses the colon syntactic sugar, 
and less than 1\% of the reads (195 times) use non-literal expressions
to call a function.
We can also classify the indexing operations that are writes as follows: 
69\% of the writes use a literal string key,
2\% of the writes use a literal number key,
and less than 1\% of the writes (1 time) uses a literal boolean key.
At all, 71\% of the writes use literals to index a table while
29\% of the writes use non-literal expressions to index a table.

We also measured how many files have code that iterate over tables to
observe how frequently iteration is used.
We observed that 23\% of the files have code that iterate over keys
of any value, that is, the call to \texttt{pairs} appears at least
once in these files (the median is twice per file);
21\% of the files have code that iterate over integer keys, that is,
the call to \texttt{ipairs} appears at least once in these files
(the median is also twice per file);
and 10\% of the files have code that use the numeric \texttt{for}
along with the length operator (the median is once per file).

The numbers about table initialization, indexing, and iteration
show us that tables are mostly used to represent records, lists,
and associative arrays.
Therefore, Typed Lua should include a table type for handling
these uses of Lua tables.
Even though the statistics show that programmers initialize tables
more often than they use the empty constructor to
dynamically initialize tables, the statistics of the empty
constructor are still expressive and indicate that Typed Lua should
also include a way to handle this style of defining table types.

We found a total of 24858 function declarations in our database
(the median is six per file).
Next, we discuss how frequently programmers use dynamic type
checking and multiple return values inside these functions.

We observed that 9\% of the functions perform dynamic type checking
on their input parameters, that is, these functions use \texttt{type}
to inspect the tags of Lua values (the median is once per function).
We randomly selected 20 functions to sample how programmers are
using \texttt{type}, and we got the following data:
50\% of these functions use \texttt{type} for asserting the tags of
their input parameters, that is, they raise an error when the tag of a
certain parameter does not match the expected tag, and
50\% of these functions use \texttt{type} for overloading, that is,
they execute different code according to the inspected tag.

These numbers show us that Typed Lua should include union types,
because the use of the \texttt{type} idiom shows that disjoint unions
will help programmers define data structures that can hold a value of
several different, but fixed types.
Typed Lua should also use \texttt{type} as a mechanism for decomposing
unions, though it may be restricted to base types only.

We observed that 10\% of the functions explicitly return multiple
values.
We also observed that 5\% of the functions return \texttt{nil} plus
something else, for signaling an unexpected behavior;
and 1\% of the functions return \texttt{false} plus something else,
also for signaling an unexpected behavior.

Typed Lua should include function types to represent Lua functions,
and tuple types to represent the signatures of Lua functions,
multiple return values, and multiple assignments.
Tuple types will require some special attention, because Typed Lua
should be able to adjust tuple types during compile-time, in a
similar way to what Lua does with function calls and multiple
assignments during run-time.
In addition, the number of functions that return \texttt{nil} and
\texttt{false} plus something else show us that overloading on the
return type will be useful to the type system.

We also measured how frequently programmers use the object-oriented
paradigm in Lua.
We observed that 23\% of the function declarations are actually
method declarations.
More precisely, 14\% of them use the colon syntactic sugar while
9\% of them use \texttt{self} as their first parameter.
We also observed that 63\% of the projects extend tables with
metatables, that is, they call \texttt{setmetatable} at least once,
and 27\% of the projects access the metatable of a given table,
that is, they call \texttt{getmetatable} at least once.
In fact, 45\% of the projects extend tables with metatables and
declare methods:
13\% using the colon syntactic sugar, 14\% using \texttt{self}, and
18\% using both.

Based on these observations, Typed Lua should include interfaces and
classes as syntactic sugar for table types that encode objects.
Proposing an optional type system for Lua that includes classes may
sound unusual at first glance, because Lua does not have standard
policies for object-oriented programming.
However, Lua provides mechanisms that allow programmers to abstract
their code in terms of objects, and our statistics confirm that
an expressive number of programmers are relying on these mechanisms to
use the object-oriented paradigm in Lua.
Typed Lua should include interfaces and classes as merely syntactic
mechanisms that help the compiler type check object-oriented code,
and they should not influence the semantics of Lua.

We also measured how programmers are defining modules.
We observed that 38\% of the files use the current way of defining
modules, that is, these files return a table that contains the
exported members of the module at the end of the file;
22\% of the files still use the deprecated way of defining modules,
that is, these files call the function \texttt{module};
and 1\% of the files use both ways.
At all, 61\% of the files are modules while 39\% of the files are
plain scripts.
The number of plain scripts is high considering the origin of
our database.
However, we did not ignore sample scripts, which usually serve to
help the users of a given module on how to use this module, and
that is the reason why we have a high number of plain scripts.

Based on these observations, Typed Lua should also use interfaces as
syntactic sugar for table types that define modules.
Typed Lua should also support the deprecated style of module
definition, using global names as exported members of the module.

Besides using interfaces for encoding objects and defining modules,
Typed Lua should also use interfaces for defining the types of
userdata, as programmers often use userdata in an object-oriented
style.

The last statistics that we collected were about variadic functions
and vararg expressions.
We observed that 8\% of the functions are variadic, that is,
their last parameter is the vararg expression.
We also observed that 5\% of the initialization of lists
(or 2\% of the occurrences of the table constructor) use solely the
vararg expression.
Typed Lua should include a \textit{vararg type} to handle variadic
functions and vararg expressions.

\section{Typed Lua through examples}

Our main goal is to allow programmers to combine static and dynamic
typing in Lua, so we propose Typed Lua as an extension to Lua 5.2
syntax that introduces optional type annotations, and class-based
object-oriented programming through the definition of classes,
interfaces, and modules.
An important requirement of Typed Lua is that it should be backwards
compatible with Lua, that is, it should allow programmers to run any
Lua code without modifications.
For this reason, these syntactic extensions include new keywords
that are not reserved words.

Figure \ref{fig:typelang} shows that Typed Lua includes first-class
and second-class types.
Typed Lua uses first-class types for representing Lua values and
second-class types for defining the signatures of functions.
First-class types include literal types, base types, the type $\Any$,
union types, function types, table types, and recursive types.
Second-class types include the type $\Void$, first-class types,
vararg types, and tuple types.
Literal types are types that represent a single value
(e.g., $\Nil$, $\False$, $\True$, literal numbers, and
literal strings).
Literal types are subtypes of their respective base types.
More precisely, the literals $\False$ and $\True$ are subtypes of
$\Boolean$, any literal number $n$ is subtype of $\Number$, and any
literal string $s$ is subtype of $\String$.
The type $\Any$ is the dynamic type that allows dynamic code coexist
with static code.
We will discuss union types, function types, table types, recursive
types, and second-class types along this section.

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
T ::= & & \textit{first-class types}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Any & \textit{dynamic type}\\
& | \; T \cup T & \textit{union type}\\
& | \; S \rightarrow S & \textit{function type}\\
& | \; \{T:T, ..., T:T\} & \textit{table type}\\
& | \; X & \textit{type variable}\\
& | \; \mu X.T & \textit{recursive type}\\
L ::= & \Nil \; | \; \False \; | \; \True \; | \;
  {<}{\it number}{>} \; | \; {<}{\it string}{>} & \\
B ::= & \Boolean \; | \; \Number \; | \; \String & \\
S ::= & \Void \; | \; V & \textit{second-class types}\\
V ::= & \;\; T & \\
& | \; T* & \textit{vararg type}\\
& | \; T \times V & \textit{tuple type}
\end{array}
$$
\dend
\caption{Typed Lua types}
\label{fig:typelang}
\end{figure}

Typed Lua allows optional type annotations in variable and function
declarations.
Whenever it is possible to use local type inference, the compiler
assigns a more specific type to an unannotated declaration.
When it is not possible to use local type inference, the compiler
assigns the type $\Any$ to an unannotated declaration.
For instance, in function declarations, the compiler does not infer
a more specific type to unannotated input parameters, because,
in this case, global type inference often is undecidable in a
language with subtyping and overloading \citep{wells1999typability}.

In the following example we use type annotations in a function
declaration, and we do not use type annotations in a declaration of a
local variable:
\begin{verbatim}
    local function fact (n:number):number
      if n == 0 then
        return 1
      else
        return n * fact(n - 1)
      end
    end
    local x = 5
    print(fact(x))
\end{verbatim}

Even though we did not add type annotations in the declaration of
the local variable \texttt{x}, the compiler uses local type inference to
assign the type $\Number$ to \texttt{x}.
The compiler does not use the literal type $5$ instead of the base
type $\Number$ because this would prevent
programmers from assigning other numeric values to the variable
\texttt{x}.
The inference that we implement in Typed Lua is quite simple, as it
uses only the type of the local expression.

Typed Lua allows programmers to combine annotated code with
unannotated code, as we show in the following example:
\begin{verbatim}
    local function abs (n:number):number
      if n < 0 then
        return -n
      else
        return n
      end
    end

    local function dist (x, y)
      return abs(x - y)
    end
\end{verbatim}

The function \texttt{dist} receives two parameters of type $\Any$
and returns a value of type $\Number$.
The compiler assigns the dynamic type $\Any$ to the input
parameters of \texttt{dist} because they do not have type annotations,
and the compiler does not use global type inference, as we mentioned
previously.
The compiler infers the return type of \texttt{dist} because it is
not a recursive function, that is, Typed Lua does not infer the return
type of recursive functions.

We are using the consistent-subtyping relation of gradual
typing to check the interaction among the dynamic type and the other
types.
More precisely, we can call \texttt{abs} inside \texttt{dist} because
\texttt{x - y} is of type $\Any$, and the type $\Any$ is
consistent with the type $\Number$.
The type $\Any$ is not the union of all types, but it is a separate
type.

Unlike gradual typing, we do not insert run-time checks that
inspect the interaction between dynamically typed code and statically
typed code during run-time.
In other words, Typed Lua does not guarantee that dynamically typed
code does not violate statically typed code.
Typed Lua does not insert run-time checks because they can decrease
run-time performance \citep{allende2013cis}.
We believe that a careful evaluation of run-time checks should be done
before inserting them on the type system.
In the previous example, we cannot guarantee that \texttt{dist} is never
going to call \texttt{abs} with a parameter that is not a number,
because in the semantics of Lua the minus operator can result in a
value that is not a number.
Still, we can implement run-time checks in the future because we are
using the consistent-subtyping relation of gradual typing to
formalize our optional type system, and the type system is being
designed to be sound.

Typed Lua includes union types as a way to discriminate data
structures that can hold values of several different, but fixed types.
Next, we show an example:
\begin{verbatim}
    local x:number?            -- OK
    local y:number|string = 1  -- OK
    local z:number = x         -- Error
\end{verbatim}

The annotation \texttt{number?} is syntactic sugar for
\texttt{number|nil} that means the type $\Number\;\cup\;\Nil$.
Typed Lua includes the syntactic sugar \texttt{t?} because the
type $t\;\cup\;\Nil$ is very common in Lua code.

The last line of the previous example gives a type error because we
are attempting to assign a value of type $\Number\;\cup\;\Nil$
to a variable that accepts only values of type $\Number$.
Although we can solve this type error changing the annotation of
variable \texttt{z} from \texttt{number} to \texttt{number?}, there is
another way that we can use to solve this type error without changing
the annotation:
\begin{verbatim}
    local z:number = x or 0  -- OK
\end{verbatim}

Typed Lua has a typing rule for checking that
$t\;\cup\;\Nil\;\mathbf{or}\;t$ always has the type $t$.

Typed Lua includes table types to represent Lua's tables.
Although Typed Lua has just one table type constructor, it has three
different syntaxes for defining the type of a table.
The first syntax defines table types for hash tables;
it is written \texttt{\{t1:t2\}},
and maps to the table type $\{t_{1}:t_{2}\;\cup\;\Nil\}$.
This table type represents a hash table that maps values of type
$t_{1}$ to values of type $t_{2}\;\cup\;\Nil$, as Lua returns
$\Nil$ when we use a non-existing key to index a table.
The second syntax defines table types for lists;
it is written \texttt{\{t\}},
and maps to the table type $\{\Number:t\;\cup\;\Nil\}$.
This table type is just a syntactic sugar to the definition of a
hash table that maps values of type $\Number$ to values of type
$t\;\cup\;\Nil$.
The third syntax defines table types for records;
it is written \texttt{\{s1:t1, ..., sn:tn\}},
and maps to the table type $\{s_{1}:t_{1}, ..., s_{n}:t_{n}\}$.
This table type represents a record that maps the literals
$s_{i}$\texttt{, ..., }$s_{n}$ to values of types
$t_{i}$\texttt{, ..., }$t_{n}$, where $i$ may vary from one to $n$.

Next, we show one example of table type to type a hash table:
\begin{verbatim}
    local t:{string:number} = { foo = 1 }  -- OK
    local x:number = t.foo                 -- Error
    local y:number? = t.bar                -- OK
\end{verbatim}

The second line of this example gives a type error for the same
reason that the last line of the previous example gives a type error
too, that is, we are trying to assign a value that can be of two
different types to a variable that accepts values of only one type.
Although the field \texttt{bar} does not exist in \texttt{t}, the last
line of this example does not give a type error because the
annotated type matches the type of the values that can be stored in
\texttt{t}.

Next, we show one example of table type to type a list:
\begin{verbatim}
    local days:{string} = { "Sunday", "Monday", "Tuesday",
                            "Wednesday", "Thursday", "Friday",
                            "Saturday" }  -- OK
\end{verbatim}

When we know that a list has fixed elements, we can leave the
variable declaration unannotated and let local type inference assign
a more specific table type to the list.
If we remove the annotation in this example, the compiler uses the
syntax of records to infer the following table type to \texttt{days}:
\begin{align*}
\{{1:\String},\;{2:\String},\;{3:\String},\;{4:\String},\;\\
{5:\String},\;{6:\String},\;{7:\String}\}
\end{align*}
This inferred table type is a subtype of the type
$\{\Number:\String\;\cup\;\Nil\}$, which is the type that we used to
annotate the variable \texttt{days}.

When we use the inferred table type while initializing a table, the
compiler gives an error message when we try to access an index that
is out of bounds.

It also worth mentioning that we can use the style of the inferred
type to define heterogeneous tuples.

Next, we show one example of table type to type a record:
\begin{verbatim}
    local person:{"firstname":string, "lastname":string} =
      { firstname = "Lou", lastname = "Reed" } 
\end{verbatim}

Type inference would infer the very same table type that we used to
annotate the variable \texttt{person}.

Typed Lua includes interfaces as syntactic sugar to table types
that specify records.
For instance, we can use the following interface as an alias to the
type of the record we showed in the previous example:
\begin{verbatim}
    interface Person
      firstname:string
      lastname:string
    end
\end{verbatim}

Now we can use the type \texttt{Person} as an alias to a table type
that certainly has the fields \texttt{firstname} and \texttt{lastname}.
Next, we show an example of code that uses the type \texttt{Person}:
\begin{verbatim}
    local function greeter (person:Person):string
      return "Hello " .. person.firstname ..
             " " .. person.lastname
    end

    local user1 = { firstname = "Lewis",
                    middlename = "Allan",
                    lastname = "Reed" }
    local user2 = { firstname = "Lou" }
    local user3 = { lastname = "Reed",
                    firstname = "Lou" }
    local user4 = { "Lou", "Reed" }

    print(greeter(user1)) -- OK
    print(greeter(user2)) -- Error
    print(greeter(user3)) -- OK
    print(greeter(user4)) -- Error
\end{verbatim}

This example shows that our optional type system is structural rather
than nominal, that is, it checks the structure of types instead of
their names, and it uses subtyping and consistent-subtyping for
checking types.

Using interfaces we can also define recursive types such as linked
lists:

\begin{verbatim}
    interface Element
      info:number
      next:Element?
    end
\end{verbatim}

We need to explicitly annotate \texttt{next} being of type
\texttt{Element}$\;\cup\;\Nil$ because in our type system $\Nil$
is not the bottom type, that is, $\Nil$ is subtype of itself only.
Typed Lua has this behavior because in Lua $\Nil$ represents
the absence of an useful value.

Besides interfaces, Typed Lua will also introduce classes.
We want to offer classes as a standard way for writing code that
Lua programmers are already using, but without tying them to a
specific model.
Our classes will not include static attributes and will also not offer
privacy rules.

Typed Lua will also introduce a specific syntax for defining modules.
For now, each file is a module.
We use all global names and their types to build the interface that
represents the type of the module, in case the module is not returned
at the end of the file.

Typed Lua includes function types because functions are first-class
values in Lua.
The syntax for defining a function type is \texttt{(s) -> (s)},
where \texttt{s} is a second-class type.
As we saw in Figure \ref{fig:typelang}, a second-class type in
Typed Lua can be the type $\Void$, a first-class type, a variadic
type, or a tuple type.
We use tuple types to represent multiple parameters, multiple return
values, and multiple assignments, and we use the variadic type to
represent variadic functions and the vararg expression.
Even though we already included the tuple type and the vararg type,
we are still working on these types.

The next example shows how we can annotate a \texttt{map} function
that receives another function as its parameter and uses this
function to alter the elements of a list of numbers, which is
also a parameter of \texttt{map}:
\begin{verbatim}
    local function map (f:(number)->(number), l:{number}):void
      for i=1, #l do
        l[i] = f(l[i] or 0)
      end
    end
\end{verbatim}

To conclude this section, we just started the design and
implementation of Typed Lua, and there is still several design
decisions to be taken on this subject.
For instance, our type system should handle the subtleties that
come with metatables.

\section{Related Lua projects}

Metalua \citep{metalua} is a Lua compiler that supports compile-time
metaprogramming (CTMP).
CTMP is a kind of macro system that allows the programmers to interact
with the compiler \citep{fleutot2007contrasting}. 
Metalua extends Lua 5.1 syntax to include its macro system,
and allows programmers to define their own syntax.
Metalua can provide syntactical support for several object-oriented
styles, and can also provide syntax for turning simple type
annotations into run-time assertions.

MoonScript \citep{moonscript} is a programming language that supports
class-based object-oriented programming, and compiles to idiomatic
Lua code.
However, it does not perform compile-time type checking.

LuaInspect \citep{luainspect} is a tool that uses MetaLua to perform
some code analysis, such as flagging unknown global variables and
table fields, checking the number of function arguments against
signatures, and inferring function return values, but it does not
try to analyze object-oriented code and does not perform compile-time
type checking.

Tidal Lock \citep{tidallock} is a prototype of another optional type
system for Lua, which is written in Metalua.
A remarkable feature of Tidal Lock is the way that it handles tables,
because it allows the programmer to incrementally build the type of
a table.
More precisely, Tidal Lock allows the programmer to create an empty
table to build its type according to the keys that the programmer
inserts in it, so the programmer does not need to build the table in
a single table constructor.
Typed Lua has a simpler form of this feature, with a different
formalization.
Like Typed Lua, Tidal Lock also has local type inference.
Unlike Typed Lua, Tidal Lock does not type the common object-oriented
idioms used by Lua programmers.

### 2014-04-07 doc/thesis/introduction.tex
Minor text adaptations to the introduction

@@
 should be validated by a type checker.
-Instead, dynamically typed languages use run-time \textit{type tags}
+Instead, dynamically typed languages use run-time \emph{type tags}
 to classify the values they compute, so their implementation can use
@@
 option to choose between static and dynamic typing:
-\textit{optional type systems} \citep{bracha2004pluggable} and
-\textit{gradual typing} \citep{siek2006gradual} are two similar
+\emph{optional type systems} \citep{bracha2004pluggable} and
+\emph{gradual typing} \citep{siek2006gradual} are two similar
 approaches for blending static and dynamic typing in the same
@@
 The aim of both approaches is to offer programmers the option
-to use type annotations where static typing is needed and avoid them
-where dynamic typing is sufficient.
+to use type annotations where static typing is needed and to not use
+them where dynamic typing is sufficient.
 The difference between these two approaches is the way they treat
@@
 
-In this work we propose the design and implementation of Typed Lua:
-an optional type system for Lua that should be complex enough to
+In this work we present the design and implementation of Typed Lua:
+an optional type system for Lua that is complex enough to
 preserve some of the Lua idioms that programmers are already used to,
-but that should also include new constructs that help programmers
+but that also includes new constructs that help programmers
 structure Lua programs.
@@
 
-So far, Typed Lua should be a strict superset of Lua that
+So far, Typed Lua is a strict superset of Lua that
 provides optional type annotations, compile-time type checking, and
@@
 classes, interfaces, and modules.
-In practice, we should implement Typed Lua as a programming
-language that extends Lua syntax to add optional type annotations
-and standard constructions for structuring Lua code.
-The compiler should use static types to perform compile-time type
-checking, but should also allow Lua code to coexist with Typed Lua
-code, and generate Lua code that runs in unmodified Lua
-implementations.
+In practice, we implement Typed Lua as a programming language that
+extends Lua syntax to add optional type annotations and standard
+constructions for structuring Lua code.
+The compiler uses static types to perform compile-time type
+checking, but also allows Lua code to coexist with Typed Lua code,
+and generates Lua code that runs in unmodified Lua implementations.
 We have an ongoing implementation of the Typed Lua compiler that is
@@
 interfaces and types of dynamically typed modules.
-We will enforce this point by using Typed Lua to statically type
+We enforce this point by using Typed Lua to statically type
 the interface of the Lua standard libraries and other commonly used
@@
 
-Typed Lua will perform a very limited form of local type inference
+Typed Lua performs a very limited form of local type inference
 \citep{pierce2000lti}, as static typing does not necessarily mean
@@
 expressions.
-Still, Typed Lua will need a small amount of type annotations due
+Still, Typed Lua needs a small amount of type annotations due
 to the nature of its optional type system.
 
-Typed Lua will not deal with code optimization, although another
+Typed Lua does not deal with code optimization, although another
 important advantage of static types is that they help the compiler
@@
 However, we believe that the formalization of our optional type
-system will be precise enough to aid optimization in some Lua
-implementations.
+system is precise enough to aid optimization in some Lua implementations.
 
-We will use some of the ideas of gradual typing to formalize Typed Lua.
-Even though Typed Lua is an optional type system and thus will not
+We use some of the ideas of gradual typing to formalize Typed Lua.
+Even though Typed Lua is an optional type system and thus does not
 include run-time checks between dynamic and static regions of the
@@
 
-%In Section \ref{sec:review} we review the literature about static and
-%dynamic typing in the same language, which includes the discussion
-%about the particularities between optional type systems and gradual
-%typing.
-%In Section \ref{sec:proposal} we describe the reasons why Typed Lua
-%includes features of object-oriented programming and the status of
-%its design and implementation.
-%In Section \ref{sec:con} we outline our expected contributions.
-%In Section \ref{sec:plan} we present the work plan for thesis completion.
+In Chapter \ref{chap:review} we review the literature about static and
+dynamic typing in the same language, which includes the discussion
+about the particularities between optional type systems and gradual
+typing.
+In Chapter \ref{chap:typedlua} we present the design and implementation
+of Typed Lua along with the reasons why it includes features of
+object-oriented programming.
+In Chapter \ref{chap:cases} we present some case studies.
+In Chapter \ref{chap:conc} we outline our contributions.

### 2014-04-10 doc/thesis/review.tex
Minor fixes to the literature review

@@
-We begin this section presenting a little bit of the history about
+We begin this chapter presenting a little bit of the history about
 combining static and dynamic typing in the same language.
@@
 
-The \textit{quasi-static} type system proposed by \citet{thatte1990qst}
+The \emph{quasi-static} type system proposed by \citet{thatte1990qst}
 performs implicit coercions and run-time checks to replace the
@@
 the expected type, while the second phase performs what Thatte calls
-\textit{plausibility checking}, that is, it rewrites the program to
+\emph{plausibility checking}, that is, it rewrites the program to
 guarantee that sequences of upcasts and downcasts always have a
@@
 
-\textit{Soft typing} \citep{cartwright1991soft} is another approach
+\emph{Soft typing} \citep{cartwright1991soft} is another approach
 to combine static and dynamic typing in the same language.
@@
 However, the programmer is free to choose between inspecting the
-run-time checks and simply running the code.
+run-time checks or simply running the code.
 Soft typing gives this choice to the programmer because it is one of
@@
 
-\textit{Dynamic typing} \citep{henglein1994dts} is an approach
+\emph{Dynamic typing} \citep{henglein1994dts} is an approach
 that optimizes code from dynamically typed languages by eliminating
@@
 and blame annotations for run-time checks.
-Contracts perform dynamic checking instead of static checking,
+Contracts perform dynamic type checking instead of static type checking,
 but deferring all verifications to run-time can lead to defects
@@
 In this approach, programmers can insert assertions in the form of
-contracts that check the input and output of higher-order functions.
-Thus, \citet{findler2002chf} show how programmers can insert
-assertions in the form of contracts that check the input and output
-of higher-order functions; and they also show how to add blame
-annotations in the generated code as a way to track assertion
-failures back to the source of the error.
+contracts that check the input and output of higher-order functions;
+and the compiler add blame annotations in the generated code as a way
+to track assertion failures back to the source of the error.
 
@@
 In the statically typed BabyJ, programmers can use the
-\textit{permissive type} $*$ to annotate the parts of the code that
+\emph{permissive type} $*$ to annotate the parts of the code that
 still do not have a specific type or the parts of the code that should
@@
 the language.
-This means that optional type systems can be unsound without affecting
-language safety, because the run-time semantics should still catch
-type errors independently of the static type checking.
-More precisely, optional type systems are unsound if they do not catch
-some type errors, but we say that languages with unsound optional type
-systems are safe because the semantics of these languages catch during
-run-time the type errors that the optional type system did not catch.
-Although optional type systems maintain the simplicity and
-flexibility of dynamically typed languages, their static type
-checking cannot prove the absence of all type errors when the optional
-type system is unsound.
-
-Optional type systems have an interesting relation to the untyped
-lambda calculus and the typed lambda calculus.
-The semantic rules of the unytped lambda calculus are the same of the
-typed lambda calculus, and the type system is actually optional and
-serves only for discarding programs that may have undesired behaviors
-\citep{bracha2004pluggable}.
-
-Optional type systems make software evolution easier and
-allow language designers to experiment with
-\textit{pluggable type systems} \citep{bracha2004pluggable}.
-Pluggable type systems are different optional type systems for
-checking different aspects of the program.
-
-According to \citet{bracha2004pluggable}, optional type systems should
-not include mandatory type inference too, because mandatory type
-inference may decrease the expressiveness of the type system.
-Instead, optional type systems should implement type inference as
-an independent mechanism that can be turned on and off according to
-the programmers' needs.
-
-Unfortunately, there is no formalization on optional type systems,
-so each language ends up implementing its optional type system on
-its own way.
-Now we will briefly discuss some languages that provide an informal
-description of their optional type systems rather than a formal one.
+That is, the run-time semantics should still catch type errors
+independently of the static type checking.
+For instance, we can view the typed lambda calculus as an optional
+type system for the untyped lambda calculus, because both have the
+same semantic rules and the type system serves only for discarding
+programs that may have undesired behaviors \citep{bracha2004pluggable}.
 
 Strongtalk \citep{bracha1993strongtalk,bracha1996strongtalk} is
-a version of Smalltalk that comes with an optional type system that
-does not include any type inference.
+a version of Smalltalk that comes with an optional type system.
+It has a polymorphic type system that programmers can use to annotate
+Smalltalk code or leave type annotations out.
+Strongtalk assigns a dynamic type to unannotated expressions and allows
+programmers to cast unannotated expressions to any static type.
+This means that the interaction of the dynamic type with the rest of
+the type system is unsound, so Strongtalk uses the original run-time
+semantics of Smalltalk when executing programs, even if programs are
+statically typed.
+
+\emph{Pluggable type systems} \citep{bracha2004pluggable} generalize
+the idea of optional type systems that Strongtalk put in practice,
+to have different optional type systems that can be layered on top
+of a dynamically typed language without affecting its original run-time
+semantics.
+Although these systems can be unsound on their interaction with the
+dynamically typed part of the language or even by design, their
+unsoundness does not affect run-time safety, as the language run-time
+semantics still cathes any run-time errors caused by an unsound
+type system.
 
@@
 In fact, Dart is a new class-based object-oriented language with
-optional type annotations while TypeScript is a strict superset of
+optional type annotations and a semantics that resembles the
+semantics of Smalltalk, while TypeScript is a strict superset of
 JavaScript that provides optional type annotations and class-based
 object-oriented programming.
-Both Dart and TypeScript perform type inference to deduce static
-types for unannotated code.
+Dart has a nominal type system, while TypeScript has a structural
+one, but both are unsound by design.
+For instance, Dart has covariant arrays, while TypeScript has
+covariant function return types;
+without counting the interaction between statically and dynamically
+typed code that is also unsound.
+
+Unfortunately, there is no common formalization for optional type
+systems, so each language ends up implementing its optional type
+system on its own way.
+Strongtalk, Dart, and TypeScript provide an informal description of
+their optional type systems rather than a formal one.
+In the next section we shall see that we can use some aspects
+of gradual typing \citep{siek2006gradual,siek2007objects} to
+formalize optional type systems.
 
@@
 
-The central idea of gradual typing is the \textit{consistency}
+The central idea of gradual typing is the \emph{consistency}
 relation, written $\tau_{1} \sim \tau_{2}$, that allows implicit
@@
 between inconsistent types \citep{siek2006gradual}.
-For instance, $\mathbf{int} \sim \,?$ and $?\, \sim \mathbf{int}$,
-but $\mathbf{int} \not\sim \mathbf{string}$ and
+For instance, $\mathbf{int} \sim \;?$, $? \sim \mathbf{int}$,
+$\mathbf{string} \sim \;?$, and $? \sim \mathbf{string}$,
+but $\mathbf{int} \not\sim \mathbf{string}$, and
 $\mathbf{string} \not\sim \mathbf{int}$.
@@
 achieve both consistency and subtyping.
-The \textit{consistent-subtyping} relation,
+The \emph{consistent-subtyping} relation,
 written $\tau_{1} \lesssim \tau_{2}$,
@@
 to that cast application can appear far way from the real error.
-\citet{wadler2009wpc} developed the \textit{blame calculus} as a way
+\citet{wadler2009wpc} developed the \emph{blame calculus} as a way
 to attack this issue, and \citet{ahmed2011bfa} extended the
@@
 attacks both problems.
-This new solution is based on a concept called \textit{threesome},
+This new solution is based on a concept called \emph{threesome},
 which is a way to split a cast between two parties into two casts
 among three parties.
-A cast has a source and a target type (a \textit{twosome}),
+A cast has a source and a target type (a \emph{twosome}),
 so we can split any cast into a downcast from the source to an
 intermediate type that is followed by an upcast from the intermediate
-type to the target type (a \textit{threesome}).
+type to the target type (a \emph{threesome}).
 
@@
 relation while Alore combines  subtyping along with the consistency
-relation to define a \textit{consistent-or-subtype} relation.
+relation to define a \emph{consistent-or-subtype} relation.
 The consistent-or-subtype relation is different from the
@@
 The consistent-or-subtype relation holds when $\tau_{1} \sim \tau_{2}$
-or $\tau_{1} <: \tau_{2}$, so $<:$ is transitive and $\sim$ is not.
+or $\tau_{1} <: \tau_{2}$, where $<:$ is transitive and $\sim$ is not.
 Alore also extends its optional type system to include optional
@@
 
-\citet{flanagan2006htc} introduced \textit{hybrid type checking},
-an approach that combines static types and refinement types.
+\citet{flanagan2006htc} introduced \emph{hybrid type checking},
+an approach that combines static types and \emph{refinement} types.
 For instance, programmers can specify the refinement type
@@
 migrating from dynamically typed to statically typed code,
-and they coined the term \textit{from scripts to programs} for
+and they coined the term \emph{from scripts to programs} for
 referring to this kind of interlanguage migration.

### 2014-04-22 doc/thesis/typedlua.tex
Cosmetic changes to Typed Lua chapter

@@
-We begin this section presenting statistics about the usage of Lua
+We begin this chapter presenting statistics about the usage of Lua
 features and idioms.
@@
 on our optional type system.
-We end this section introducing Typed Lua through examples, and
+We end this chapter introducing Typed Lua through examples, and
 describing a few Lua projects that are similar to Typed Lua.
@@
 the latest version of Lua.
-We also ignored \textit{machine-generated} files and test files,
+We also ignored \emph{machine-generated} files and test files,
 because these files may not represent idiomatic Lua code,
@@
 vararg expression.
-Typed Lua should include a \textit{vararg type} to handle variadic
+Typed Lua should include a \emph{vararg type} to handle variadic
 functions and vararg expressions.
@@
 \begin{array}{rlr}
-T ::= & & \textit{first-class types}\\
-& \;\; L & \textit{literal types}\\
-& | \; B & \textit{base types}\\
-& | \; \Any & \textit{dynamic type}\\
-& | \; T \cup T & \textit{union type}\\
-& | \; S \rightarrow S & \textit{function type}\\
-& | \; \{T:T, ..., T:T\} & \textit{table type}\\
-& | \; X & \textit{type variable}\\
-& | \; \mu X.T & \textit{recursive type}\\
+T ::= & & \emph{first-class types}\\
+& \;\; L & \emph{literal types}\\
+& | \; B & \emph{base types}\\
+& | \; \Any & \emph{dynamic type}\\
+& | \; T \cup T & \emph{union type}\\
+& | \; S \rightarrow S & \emph{function type}\\
+& | \; \{T:T, ..., T:T\} & \emph{table type}\\
+& | \; X & \emph{type variable}\\
+& | \; \mu X.T & \emph{recursive type}\\
 L ::= & \Nil \; | \; \False \; | \; \True \; | \;
@@
 B ::= & \Boolean \; | \; \Number \; | \; \String & \\
-S ::= & \Void \; | \; V & \textit{second-class types}\\
+S ::= & \Void \; | \; V & \emph{second-class types}\\
 V ::= & \;\; T & \\
-& | \; T* & \textit{vararg type}\\
-& | \; T \times V & \textit{tuple type}
+& | \; T* & \emph{vararg type}\\
+& | \; T \times V & \emph{tuple type}
 \end{array}

### 2014-04-30 doc/thesis/cases.tex
Starting to type Lua standard libraries

\section{Lua standard libraries}

\subsection{Basic functions}

\begin{enumerate}
\item assert (v [, message]) :
$\Top \times
(\String \cup \Nil) \times
\TopStar \rightarrow
\NilStar \cup
(\Top \times
(\String \cup \Nil) \times
\TopStar)$ 
\\
When \textbf{v} is $\Nil$ or $\False$, \textbf{assert}
shows the error \textbf{message}.
Otherwise, \textbf{assert} returns \textbf{v} and all the extra
parameters.
\item collectgarbage ([opt [, arg]]) :
$(\String \cup \Nil) \times
(\Number \cup \Nil) \times
\TopStar \rightarrow
(\Number \times \Nil \times \NilStar) \cup
(\Number \times \Number \times \NilStar) \cup
(\Boolean \times \Nil \times \NilStar)$
\\
The return type depends on the type of \textbf{opt}:
\begin{itemize}
\item $\Nil \times \TopStar \rightarrow \Number \times \NilStar$
\item $``collect" \times \TopStar \rightarrow \Number \times \NilStar$
\item $``stop" \times \TopStar \rightarrow \Number \times \NilStar$
\item $``restart" \times \TopStar \rightarrow \Number \times \NilStar$
\item $``count" \times \TopStar \rightarrow \Number \times \Number \times \NilStar$
\item $``step" \times \TopStar \rightarrow \Boolean \times \NilStar$
\item $``setpause" \times \TopStar \rightarrow \Number \times \NilStar$
\item $``setstepmul" \times \TopStar \rightarrow \Number \times \NilStar$
\item $``isrunning" \times \TopStar \rightarrow \Boolean \times \NilStar$
\item $``generational" \times \TopStar \rightarrow \Number \times \NilStar$
\item $``incremental" \times \TopStar \rightarrow \Number \times \NilStar$
\end{itemize}
\item dofile ([filename]) :
$(\String \cup \Nil) \times
\TopStar \rightarrow
\AnyStar
$
\item error (message [, level]) :
$\String \times
(0 \cup 1 \cup 2 \cup \Nil) \times
\TopStar \rightarrow
\NilStar$
\item \_G : $\{ \String:\Top \}$
\item getmetatable (object) :
$\mu X.\{ \_\_metatable:\Top \} \times
\TopStar \rightarrow
(\Nil \cup \{ \_\_metatable:\Top \}) \times
\NilStar$
\item ipairs (t) :
$\{ \Number:\Any \} \times
\TopStar \rightarrow
\Number \times
\Any \times
\NilStar$
\begin{itemize}
\item What happens when \textbf{t} has a metamethod \textbf{\_\_ipairs}?
\item Will $\{ ``tag":\String,\; \Number:\Any \}$ type check?
\item In fact, \textbf{ipairs} and \textbf{pairs} return three values.
Need to check this.
\end{itemize}
\item load (ld [, source [, mode [, env]]]) :
$(\String \cup (\TopStar \rightarrow \TopStar)) \times
(\String \cup \Nil) \times
(``b" \cup ``t" \cup ``bt" \cup \Nil) \times
(\Top \cup \Nil) \times
\TopStar \rightarrow
((\TopStar \rightarrow \TopStar) \times \NilStar) \cup
(\Nil \times \String \times \NilStar)$
\item loadfile ([filename [, mode [, env]]]) :
$(\String \cup \Nil) \times
(``b" \cup ``t" \cup ``bt" \cup \Nil) \times
(\Top \cup \Nil) \times
\TopStar \rightarrow
((\TopStar \rightarrow \TopStar) \times \NilStar) \cup
(\Nil \times \String \times \NilStar)$
\item next (table [, index]) :
$\{ \Any:\Any \} \times
(\Any \cup \Nil) \times
\TopStar \rightarrow
(\Nil \times \NilStar) \cup
(\Any \times \Any \times \NilStar)$
\item pairs (t) :
$\{ \Any:\Any \} \times
\TopStar \rightarrow
\Any \times 
\Any \times
\NilStar$
\item pcall (f [, arg1, ···]) :
$(\TopStar \rightarrow \TopStar) \times
\TopStar \rightarrow
(\Boolean \times \AnyStar) \cup
(\Boolean \times \String \times \NilStar)$
\item print (···) : $\TopStar \rightarrow \NilStar$
\item rawequal (v1, v2) :
$\Top \times
\Top \times
\TopStar \rightarrow
\Boolean \times
\NilStar$
\item rawget (table, index) :
$\{ \Top:\Top \} \times
\Top \times
\TopStar \rightarrow
\Any \times
\NilStar$
\item rawlen (v) :
$(\{ \Top:\Top \} \cup \String) \times
\TopStar \rightarrow
\Number \times
\NilStar$
\item rawset (table, index, value) :
$\{ \Top:\Top \} \times
\Top \times
\Top \times
\TopStar \rightarrow
\{ \Any:\Any \} \times
\NilStar$
\\
In fact, \textbf{index} cannot be $\Nil$ or \texttt{NaN}.
\item select (index, ···) :
$\Number \times
\AnyStar \rightarrow
\AnyStar$
\\
When \textbf{index} is the string ``\texttt{\#}", the typing is:
$\String \times \AnyStar \rightarrow \Number \times \NilStar$
\item setmetatable (table, metatable) :
$\{ \Top:\Top \} \times
(\Nil \cup \{ \_\_index:\mu X.\{ \Top:\Top \} \}) \times
\TopStar \rightarrow
\{ \_\_index:\mu X.\{ \Top:\Top \} \} \times
\NilStar$
\\
Even though \textbf{metatable} can be $\Nil$, it is not optional.
\item tonumber (e [, base]) :
$\String \times
(\Number \cup \Nil) \times
\TopStar \rightarrow
\Number \times
\NilStar$
\item tostring (v) :
$\Top \times
\TopStar \rightarrow
\String \times
\NilStar$
\item type (v) :
$\Top \times
\TopStar \rightarrow
(``nil" \cup
``number" \cup
``string" \cup
``boolean" \cup
``table" \cup
``function" \cup
``thread" \cup
``userdata") \times
\NilStar$
\item \_VERSION : $\String$
\item xpcall (f, msgh [, arg1, ···]) :
$(\TopStar \rightarrow \TopStar) \times
\Top \times
\TopStar \rightarrow
(\Boolean \times \AnyStar) \cup
(\Boolean \times \String \times \NilStar)$
\end{enumerate}

\subsection{Coroutine manipulation}

\subsection{Modules}

\subsection{String manipulation}

\begin{enumerate}
\item string.byte (s [, i [, j]]) :
$\String \times
(\Number \cup \Nil) \times
(\Number \cup \Nil) \times
\TopStar \rightarrow
\Number \times
\Number{*}$
\item string.char (···) :
$\Number{*} \rightarrow
\String \times
\NilStar$
\item string.dump (function) :
$(\TopStar \rightarrow \TopStar) \times
\TopStar \rightarrow
\String \times
\NilStar$
\item string.find (s, pattern [, init [, plain]]) :
$\String \times
\String \times
(\Number \cup \Nil) \times
(\Boolean \cup \Nil) \times
\TopStar \rightarrow
(\Number \times \Number \times \NilStar) \cup
(\Nil \times \NilStar)$
\item string.format (formatstring, ···) :
$\String \times
\TopStar \rightarrow
\String \times
\NilStar$
\item string.gmatch (s, pattern) :
$\String \times
\String \times
\TopStar \rightarrow
\String{*}$
\\
This function actually returns an iterator,
so I still need to confirm its return type.
\item string.gsub (s, pattern, repl [, n]) :
$\String \times
\String \times
(\String \cup \{ \Top:\Top \} \cup (\TopStar \rightarrow \TopStar)) \times
(\Number \cup \Nil) \times
\TopStar \rightarrow
\String \times
\Number \times
\NilStar$
\item string.len (s) :
$\String \times
\TopStar \rightarrow
\Number \times
\NilStar$
\item string.lower (s) :
$\String \times
\TopStar \rightarrow
\String \times
\NilStar$
\item string.match (s, pattern [, init]) :
$\String \times
\String \times
(\Number \cup \Nil) \times
\TopStar \rightarrow
(\String \times \String{*}) \cup
(\Nil \times \NilStar)$
\item string.rep (s, n [, sep]) :
$\String \times
\Number \times
(\String \cup \Nil) \times
\TopStar \rightarrow
\String \times
\NilStar$
\item string.reverse (s) :
$\String \times
\TopStar \rightarrow
\String \times
\NilStar$
\item string.sub (s, i [, j]) :
$\String \times
\Number \times
(\Number \cup \Nil) \times
\TopStar \rightarrow
\String \times
\NilStar$
\item string.upper (s) :
$\String \times
\TopStar \rightarrow
\String \times
\NilStar$
\end{enumerate}

\subsection{Table manipulation}

\begin{enumerate}
\item table.concat (list [, sep [, i [, j]]]) :
$\{ \Number:\String \cup \Number \} \times
(\String \cup \Nil) \times
(\Number \cup \Nil) \times
(\Number \cup \Nil) \times
\TopStar \rightarrow
\String \times
\NilStar$
\item table.insert (list, [pos,] value) :
$\{ \Number:\Top \} \times
(\Number \times \Top) \cup
(\Top \times \Nil) \times
\TopStar \rightarrow
\NilStar$
\\
This function does not accept/discard extra parameters.
Perhaps due to the fact that the parameter \textbf{pos} is optional.
If \textbf{value} is $\TopStar$, the function inserts \textbf{pos}
in the \textbf{list}, that is, the input type is
$\{ \Number:\Top \} \times \Top \times \TopStar$.
If \textbf{value} is $\Nil$, the function does not insert \textbf{pos}
in the \textbf{list}, that is, the input type is
$\{ \Number:\Top \} \times \Number \times \Top \times \TopStar$,
but the element is not inserted in the list because $\Top$ is actually $\Nil$.
\item table.pack (···) :
$\TopStar \rightarrow
\{ \Number:\Top \} \times
\NilStar$
\\
Even though \textbf{table.pack} gets any number of parameters of
any type, the $\Nil$ type generates a hole in the returned table.
\item table.remove (list [, pos]) :
$\{ \Number:\Top \} \times
(\Number \cup \Nil) \times
\TopStar \rightarrow
\Top \times
\NilStar
$
\\
The return type depends on the type of the element that is removed
from the list.
\item table.sort (list [, comp]) :
$\{ \Number:\Top \} \times
((\TopStar \rightarrow \TopStar) \cup \Nil) \times
\TopStar \rightarrow
\NilStar$
\item table.unpack (list [, i [, j]]) :
$\{ \Number:\Top \} \times
(\Number \cup \Nil) \times
(\Number \cup \Nil) \times
\TopStar \rightarrow
\TopStar$
\\
The return type depends on the type of the elements that are stored in
the list.
\end{enumerate}

\subsection{Mathematical functions}

\begin{enumerate}
\item math.abs (x) :
$\Number \times \TopStar \rightarrow \Number \times \NilStar$
\item math.acos (x) :
$\Number \times \TopStar \rightarrow \Number \times \NilStar$
\item math.asin (x) :
$\Number \times \TopStar \rightarrow \Number \times \NilStar$
\item math.atan (x) :
$\Number \times \TopStar \rightarrow \Number \times \NilStar$
\item math.atan2 (y, x) :
$\Number \times \Number \times \TopStar \rightarrow \Number \times \NilStar$
\item math.ceil (x) :
$\Number \times \TopStar \rightarrow \Number \times \NilStar$
\item math.cos (x) :
$\Number \times \TopStar \rightarrow \Number \times \NilStar$
\item math.cosh (x) :
$\Number \times \TopStar \rightarrow \Number \times \NilStar$  
\item math.deg (x) :
$\Number \times \TopStar \rightarrow \Number \times \NilStar$
\item math.exp (x) :
$\Number \times \TopStar \rightarrow \Number \times \NilStar$
\item math.floor (x) :
$\Number \times \TopStar \rightarrow \Number \times \NilStar$
\item math.fmod (x, y) :
$\Number \times \Number \times \TopStar \rightarrow \Number \times \NilStar$
\item math.frexp (x) :
$\Number \times \TopStar \rightarrow \Number \times \Number \times \NilStar$
\item math.huge : $\Number$
\item math.ldexp (m, e) :
$\Number \times \Number \times \TopStar \rightarrow \Number \times \NilStar$
\item math.log (x [, base]) :
$\Number \times
(\Number \cup \Nil) \times
\TopStar \rightarrow
\Number \times
\NilStar$
\item math.max (x, ···) :
$\Number \times \Number{*} \rightarrow \Number \times \NilStar$
\item math.min (x, ···) :
$\Number \times \Number{*} \rightarrow \Number \times \NilStar$
\item math.modf (x) :
$\Number \times \TopStar \rightarrow \Number \times \Number \times \NilStar$
\item math.pi : $\Number$
\item math.pow (x, y) :
$\Number \times \Number \times \TopStar \rightarrow \Number \times \NilStar$
\item math.rad (x) :
$\Number \times \TopStar \rightarrow \Number \times \NilStar$
\item math.random ([m [, n]]) :
$(\Number \cup \Nil) \times
(\Number \cup \Nil) \rightarrow
\Number \times
\NilStar$
\\
This function does not accept $\Nil$ as its only parameter,
although the parameter passing is optional.
This function does not accept/discard extra parameters.
\item math.randomseed (x) :
$\Number \times \TopStar \rightarrow \NilStar$
\item math.sin (x) :
$\Number \times \TopStar \rightarrow \Number \times \NilStar$
\item math.sinh (x) :
$\Number \times \TopStar \rightarrow \Number \times \NilStar$
\item math.sqrt (x) :
$\Number \times \TopStar \rightarrow \Number \times \NilStar$
\item math.tan (x) :
$\Number \times \TopStar \rightarrow \Number \times \NilStar$
\item math.tanh (x) :
$\Number \times \TopStar \rightarrow \Number \times \NilStar$
\end{enumerate}

\subsection{Bitwise operations}

\subsection{Input and Output facilities}

\subsection{Operating System facilities}

\subsection{The debug library}

### 2014-05-05 doc/thesis/cases.tex
Finished the porting of typing Lua standard libraries

@@
 
+\begin{enumerate}
+\item coroutine.create (f) :
+$(\TopStar \rightarrow \TopStar) \times \TopStar \rightarrow Thread \times \NilStar$
+\item coroutine.resume (co [, val1, ···]) :
+$Thread \times
+\TopStar \rightarrow
+(\True \times \AnyStar) \cup
+(\False \times \String \times \NilStar)$
+\item coroutine.running () :
+$\TopStar \rightarrow Thread \times \Boolean \times \NilStar$
+\item coroutine.status (co) :
+$Thread \times \TopStar \rightarrow \String \times \NilStar$
+\item coroutine.wrap (f) :
+$(\TopStar \rightarrow \TopStar) \times
+\TopStar \rightarrow
+(\TopStar \rightarrow \TopStar) \times
+\NilStar$ 
+\item coroutine.yield (···) :
+$\TopStar \rightarrow \NilStar$
+\end{enumerate}
+
 \subsection{Modules}
+
+\begin{enumerate}
+\item require (modname) :
+$\String \times \TopStar \rightarrow \{\Any:\Any\} \times \NilStar$
+\\
+The function \textbf{require} returns exactly the same value that is
+returned by loading \textbf{modname}.
+However, if the returned value is $\Nil$, this value is
+converted to $\True$.
+\item package.config : $\String$
+\item package.cpath : $\String$
+\item package.loaded : $\{\String:\{\Any:\Any\}\}$
+\item package.loadlib (libname, funcname) :
+$\String \times
+\String \times
+\TopStar \rightarrow
+(\True \times \NilStar) \cup
+(\Nil \times \String \times \NilStar)$
+\item package.path : $\String$
+\item package.preload : $\{\String:\{\Any:\Any\}\}$
+\item package.searchers : $\{\Number:\TopStar \rightarrow \TopStar\}$
+\item package.searchpath (name, path [, sep [, rep]]) :
+$\String \times
+\String \times
+(\String \cup \Nil) \times
+(\String \cup \Nil) \times
+\TopStar \rightarrow
+(\String \times \NilStar) \cup
+(\Nil \times \String \times \NilStar)$
+\end{enumerate}
 
@@
 
+\begin{enumerate}
+\item bit32.arshift (x, disp) :
+$\Number \times \Number \times \TopStar \rightarrow \Number \times \NilStar$
+\item bit32.band (···) :
+$\Number{*} \rightarrow \Number \times \NilStar$
+\item bit32.bnot (x) :
+$\Number \times \TopStar \rightarrow \Number \times \NilStar$
+\item bit32.bor (···) :
+$\Number{*} \rightarrow \Number \times \NilStar$
+\item bit32.btest (···) :
+$\Number{*} \rightarrow \Number \times \NilStar$
+\item bit32.bxor (···) :
+$\Number{*} \rightarrow \Number \times \NilStar$
+\item bit32.extract (n, field [, width]) :
+$\Number \times
+\Number \times
+(\Number \cup \Nil) \times
+\TopStar \rightarrow
+\Number \times
+\NilStar$
+\item bit32.replace (n, v, field [, width]) :
+$\Number \times
+\Number \times
+\Number \times
+(\Number \cup \Nil) \times
+\TopStar \rightarrow
+\Number \times
+\NilStar$
+\item bit32.lrotate (x, disp) :
+$\Number \times \Number \times \TopStar \rightarrow \Number \times \NilStar$
+\item bit32.lshift (x, disp) :
+$\Number \times \Number \times \TopStar \rightarrow \Number \times \NilStar$
+\item bit32.rrotate (x, disp) :
+$\Number \times \Number \times \TopStar \rightarrow \Number \times \NilStar$
+\item bit32.rshift (x, disp) :
+$\Number \times \Number \times \TopStar \rightarrow \Number \times \NilStar$
+\end{enumerate}
+
 \subsection{Input and Output facilities}
+
+\begin{enumerate}
+\item io.close ([file]) :
+$(Userdata \cup \Nil) \times
+\TopStar \rightarrow
+(\Boolean \times \NilStar) \cup
+(\Nil \times \String \times \NilStar) \cup
+((\True \cup \Nil) \times \String \times \Number \times \NilStar)$
+\\
+Returns $((\True \cup \Nil) \times \String \times \Number \times \NilStar)$
+when the file handle was created with \textbf{io.popen}.
+\item io.flush () :
+$\TopStar \rightarrow \Boolean \times \NilStar$
+\item io.input ([file]) :
+$(\String \cup Userdata \cup \Nil) \times
+\TopStar \rightarrow
+Userdata \times
+\NilStar$
+\item io.lines ([filename ···]) :
+$(\String \cup Userdata \cup \Nil) \times
+\TopStar \rightarrow
+(\String \times \NilStar) \cup
+(\Nil \times \NilStar)$
+\\
+In fact, this function returns an iterator. Need to check the return type.
+\item io.open (filename [, mode]) :
+$\String \times
+(``r" \cup ``w" \cup ``a" \cup
+``r+" \cup ``w+" \cup ``a+" \cup
+``rb" \cup ``wb" \cup ``ab" \cup
+``r+b" \cup ``w+b" \cup ``a+b" \cup
+\Nil) \times
+\TopStar \rightarrow
+(Userdata \times \NilStar) \cup
+(\Nil \times \String \times \Number \times \NilStar)$
+\item io.output ([file]) :
+$(\String \cup Userdata \cup \Nil) \times
+\TopStar \rightarrow
+Userdata \times
+\NilStar$
+\item io.popen (prog [, mode]) :
+$\String \times
+(``r" \cup ``w" \cup \Nil) \times
+\TopStar \rightarrow
+(Userdata \times \NilStar) \cup
+(\Nil \times \String \times \Number \times \NilStar)$
+\item io.read (···) :
+$(``*n" \cup ``*a" \cup ``*l" \cup ``*L" \cup \Number){*} \rightarrow
+(\String \cup \Number){*}
+$
+\item io.tmpfile () :
+$\TopStar \rightarrow Userdata \times \NilStar$
+\item io.type (obj) :
+$\Top \times \TopStar \rightarrow (``file" \cup ``closed\;file" \cup \Nil) \times \NilStar$
+\item io.write (···) :
+$(\String \cup \Number){*} \rightarrow
+(Userdata \times \NilStar) \cup
+(\Nil \times \String \times \Number \times \NilStar)$
+\item file:close () : ?
+\item file:flush () : ?
+\item file:lines (···) : ?
+\item file:read (···) : ?
+\item file:seek ([whence [, offset]]) : ?
+\item file:setvbuf (mode [, size]) : ?
+\item file:write (···) : ?
+\end{enumerate}
 
@@
 
+\begin{enumerate}
+\item os.clock () :
+$\TopStar \rightarrow \Number \times \NilStar$
+\item os.date ([format [, time]]) :
+$(\String \cup \Nil) \times
+(\Number \cup \Nil) \times
+\TopStar \rightarrow
+(\String \times \NilStar) \cup
+(\{``year":\Number,\;
+   ``month":\Number,\;
+   ``day":\Number,\;
+   ``hour":\Number,\;
+   ``min":\Number,\;
+   ``sec":\Number,\;
+   ``wday":\Number,\;
+   ``yday":\Number,\;
+   ``isdst":\Boolean \} \times \NilStar)$
+\\
+Returns a table when \textbf{format} contains ``\textbf{*t}''.
+\item os.difftime (t2, t1) :
+$\Number \times
+(\Number \cup \Nil) \times
+\TopStar \rightarrow
+\Number \times
+\NilStar$
+\\
+Returns \textbf{t2} when \textbf{t1} is $\Nil$.
+\item os.execute ([command]) :
+$(\String \cup \Nil) \times
+\TopStar \rightarrow
+(\Boolean \times \NilStar) \cup
+((\True \cup \Nil) \times \String \times \Number \times \NilStar)$
+\\
+Returns just a Boolean when \textbf{command} is $\Nil$.
+This Boolean means whether there is a shell available.
+Returns $\True \times \String \times \Number \times \NilStar$ when
+\textbf{command} terminated successfully.
+Returns $\Nil \times \String \times \Number \times \NilStar$ when
+\textbf{command} did not terminate sucessfully.
+\item os.exit ([code [, close]]) :
+$(\Number \cup \Boolean \cup \Nil) \times
+\Top \times
+\TopStar \rightarrow
+\NilStar$
+\item os.getenv (varname) :
+$\String \times
+\TopStar \rightarrow
+(\String \times \NilStar) \cup
+(\Nil \times \NilStar)$
+\item os.remove (filename) :
+$\String \times
+\TopStar \rightarrow
+(\True \times \NilStar) \cup
+(\Nil \times \String \times \Number \NilStar)$
+\item os.rename (oldname, newname) :
+$\String \times
+\String \times
+\TopStar \rightarrow
+(\True \times \NilStar) \cup
+(\Nil \times \String \times \Number \NilStar)$
+\item os.setlocale (locale [, category]) :
+$(\String \cup \Nil) \times
+(``all" \cup
+``collate" \cup
+``ctype" \cup
+``monetary" \cup
+``numeric" \cup
+``time" \cup
+\Nil) \times
+\TopStar \rightarrow
+(\String \times \NilStar) \cup
+(\Nil \times \NilStar)$
+\item os.time ([table]) :
+$(\{``year":\Number,\;
+   ``month":\Number,\;
+   ``day":\Number,\;
+   ``hour":(\Number \cup \Nil),\;
+   ``min":(\Number \cup \Nil),\;
+   ``sec":(\Number \cup \Nil),\;
+   ``wday":(\Number \cup \Nil),\;
+   ``yday":(\Number \cup \Nil),\;
+   ``isdst":(\Boolean \cup \Nil) \} \cup \Nil) \times
+\TopStar \rightarrow
+\Number \times
+\NilStar$
+\item os.tmpname () :
+$\TopStar \rightarrow \String \times \NilStar$
+\end{enumerate}
+
 \subsection{The debug library}
+
+\begin{enumerate}
+\item debug.debug () : ?
+\item debug.gethook ([thread]) : ?
+\item debug.getinfo ([thread,] f [, what]) : ?
+\item debug.getlocal ([thread,] f, local) : ?
+\item debug.getmetatable (value) : ?
+\item debug.getregistry () : ?
+\item debug.getupvalue (f, up) : ?
+\item debug.getuservalue (u) : ?
+\item debug.sethook ([thread,] hook, mask [, count]) : ?
+\item debug.setlocal ([thread,] level, local, value) : ?
+\item debug.setmetatable (value, table) : ?
+\item debug.setupvalue (f, up, value) : ?
+\item debug.setuservalue (udata, value) : ?
+\item debug.traceback ([thread,] [message [, level]]) : ?
+\item debug.upvalueid (f, n) : ?
+\item debug.upvaluejoin (f1, n1, f2, n2) : ?
+\end{enumerate}

### 2014-05-14 doc/thesis/review.tex
Adding some figures to the section about gradual typing

@@
 their optional type systems rather than a formal one.
-In the next section we shall see that we can use some aspects
+In the next section we shall see that we can use some features
 of gradual typing \citep{siek2006gradual,siek2007objects} to
@@
 
+\begin{figure}[!ht]
+\textbf{Syntax of the gradually-typed lambda calculus}\\
+\dstart
+$$
+\begin{array}{rlr}
+T ::= & & \emph{types}\\
+& \;\; \Number & \emph{base type number}\\
+& | \; \String & \emph{base type string}\\
+& | \; ? & \emph{dynamic type}\\
+& | \; T \rightarrow T & \emph{function types}\\
+e ::= & & \emph{expressions}\\
+& \;\; c & \emph{constants}\\
+& | \; x & \emph{variables}\\
+& | \; \lambda x{:}T{.}e & \emph{abstractions}\\
+& | \; e_{1} e_{2} & \emph{application}
+\end{array}
+$$
+\dend
+\caption{Syntax of the gradually-typed lambda calculus}
+\label{fig:gtlc}
+\end{figure}
+
 The central idea of gradual typing is the \emph{consistency}
@@
 The consistency relation is neither commutative nor transitive.
+
+\begin{figure}[!ht]
+\textbf{Type consistency}\\
+\dstart
+$$
+\begin{matrix}
+\Number \sim \Number \mylabel{C-NUM} \;\;\; \String \sim \String \mylabel{C-STR}
+\\ \\
+T \sim \;? \mylabel{C-ANY1} \;\;\; ? \sim T \mylabel{C-ANY2}
+\\ \\
+\dfrac{T_{3} \sim T_{1} \;\;\; T_{2} \sim T_{4}}
+      {T_{1} \rightarrow T_{2} \sim T_{3} \rightarrow T_{4}} \mylabel{C-FUN}
+\end{matrix}
+$$
+\dend
+\caption{The consistency relation}
+\label{fig:consistency}
+\end{figure}
 
@@
 
+\begin{figure}[!ht]
+\textbf{Gradual Type System}\\
+\dstart
+$$
+\begin{matrix}
+\dfrac{\env(x) = T}
+      {\env \vdash x:T} \mylabel{T-VAR}
+\;\;\;
+\dfrac{type(c) = T}
+      {\env \vdash c:T} \mylabel{T-CONST}
+\\ \\
+\dfrac{\env, x:T_{1} \vdash e:T_{2}}
+      {\env \vdash \lambda x:T_{1}.e:T_{1} \rightarrow T_{2}} \mylabel{T-ABS}
+\;\;\;
+\dfrac{\env \vdash e_{1}:\;? \;\;\;
+       \env \vdash e_{2}:T}
+      {\env \vdash e_{1} e_{2}:\;?} \mylabel{T-APP1}
+\\ \\
+\dfrac{\env \vdash e_{1}:T_{1} \rightarrow T_{2} \;\;\;
+       \env \vdash e_{2}:T_{3} \;\;\;
+       T_{3} \sim T_{1}}
+      {\env \vdash e_{1} e_{2}:T_{2}} \mylabel{T-APP2}
+\end{matrix}
+$$
+\dend
+\caption{Typing rules of gradually-typed lambda calculus}
+\label{fig:gts}
+\end{figure}
+
 Later, \citet{siek2007objects} discovered that the consistency relation
@@
 the consistent-subtyping relation is not transitive.
+
+\begin{figure}[!ht]
+\textbf{Consistent-Subtyping}\\
+\dstart
+$$
+\begin{matrix}
+\Number \lesssim \Number \mylabel{C-NUM} \;\;\; \String \lesssim \String \mylabel{C-STR}
+\\ \\
+T \lesssim \;? \mylabel{C-ANY1} \;\;\; ? \lesssim T \mylabel{C-ANY2}
+\\ \\
+\dfrac{T_{3} \lesssim T_{1} \;\;\; T_{2} \lesssim T_{4}}
+      {T_{1} \rightarrow T_{2} \sim T_{3} \rightarrow T_{4}} \mylabel{C-FUN}
+\;\;\;
+\dfrac{\forall i \in 1..n \;\;\; T_{i} \lesssim T_{i}'}
+      {[l_{i}:T_{i}^{\;i\in1..n+m}] \lesssim [l_{i}:T_{i}'^{\;i\in1..n}]} \mylabel{C-OBJ}
+\end{matrix}
+$$
+\dend
+\caption{The consistent-subtyping relation}
+\label{fig:consistent-subtyping}
+\end{figure}
 

### 2014-05-14 doc/thesis/review.tex
Trying to improve a little bit the text about gradual typing

@@
 unsoundness does not affect run-time safety, as the language run-time
-semantics still cathes any run-time errors caused by an unsound
+semantics still catches any run-time errors caused by an unsound
 type system.
@@
 To do that, \citet{siek2006gradual} extended the simply typed
-lambda calculus with the dynamic type $?$.
+lambda calculus with the dynamic type $?$, as we shall see in
+Figure \ref{fig:gtlc}.
 In gradual typing, type annotations are optional, so an untyped
 variable is syntactic sugar for a variable whose declared type is
-the dynamic type $?$.
+the dynamic type $?$, that is, $\lambda x.e \equiv \lambda x{:}?.e$.
 Under these circumstances, we view gradual typing as a way to add
@@
 \begin{figure}[!ht]
-\textbf{Syntax of the gradually-typed lambda calculus}\\
 \dstart
@@
 The central idea of gradual typing is the \emph{consistency}
-relation, written $\tau_{1} \sim \tau_{2}$, that allows implicit
-conversions to and from the dynamic type, and disallows conversions
-between inconsistent types \citep{siek2006gradual}.
-For instance, $\mathbf{int} \sim \;?$, $? \sim \mathbf{int}$,
-$\mathbf{string} \sim \;?$, and $? \sim \mathbf{string}$,
-but $\mathbf{int} \not\sim \mathbf{string}$, and
-$\mathbf{string} \not\sim \mathbf{int}$.
-The consistency relation is neither commutative nor transitive.
+relation, written $T_{1} \sim T_{2}$.
+The consistency relation allows implicit conversions to and from the
+dynamic type, and disallows conversions between inconsistent types
+\citep{siek2006gradual}.
+For instance, $\Number \sim \;?$, $? \sim \Number$,
+$\String \sim \;?$, and $? \sim \String$,
+but $\Number \not\sim \String$, and
+$\String \not\sim \Number$.
+The consistency relation is both reflexive and symmetric, but
+it is neither commutative nor transitive.
 
 \begin{figure}[!ht]
-\textbf{Type consistency}\\
 \dstart
@@
 
-Gradual typing is similar to the approaches proposed by
-\citet{abadi1989dts} and \citet{thatte1990qst} by including a
-dynamic type to a statically typed language.
-However, these three approaches differ in the way they handle the
-dynamic type.
-While \citet{siek2006gradual} rely on the consistency relation,
-\citet{abadi1989dts} rely on type equality with explicit projections
-and injections, and \citet{thatte1990qst} relies on subtyping.
-The subtyping relation actually is a pitfall on Thatte's quasi-static
-typing, because the dynamic type is put as the top and the bottom of
-the subtyping relation.
-Therefore, downcasts combined with the transitivity of subtyping
-accepts programs that should be rejected.
+Figure \ref{fig:consistency} defines the consistency relation.
+Rules \textsc{C-NUM} and \textsc{C-STR} are reflexive. 
+Rules \textsc{C-ANY1} and \textsc{C-ANY2} are the rules that allow
+conversions to and from the dynamic type $?$.
+The rule \textsc{C-FUN} resembles the subtyping of functions
+because it is contravariant on the argument type and covariant on the
+return type.
+
+Figure \ref{fig:gts} uses the consistency relation in the typing rules
+of the gradual type system of the simply typed lambda calculus extended
+with the dynamic type $?$.
+The environment $\env$ is a function from variables to types, and
+the directive $type$ is a function from constants to types.
+The rule \textsc{T-VAR} uses the environment function $\env$ to get the
+type of a variable $x$.
+The rule \textsc{T-CONST} uses the directive $type$ to get the type of
+a constant $c$.
+The rule \textsc{T-ABS} evaluates the expression $e$ with an environment
+$\env$ that binds the variable $x$ to the type $T_{1}$, and the resulting
+type is the the function type $T_{1} \rightarrow T_{2}$.
+The rule \textsc{T-APP1} handles function calls where the type of the
+function is dynamically typed; in this case, the argument type may have
+any type and the resulting type has the dynamic type.
+The rule \textsc{T-APP2} handles function calls where the type of the
+function is statically typed; in this case, the argument type should
+be consistent to the argument type of the function's signature.
 
 \begin{figure}[!ht]
-\textbf{Gradual Type System}\\
 \dstart
@@
 \dend
-\caption{Typing rules of gradually-typed lambda calculus}
+\caption{Gradual type system gradually-typed lambda calculus}
 \label{fig:gts}
@@
 
-Later, \citet{siek2007objects} discovered that the consistency relation
+Gradual typing is similar to the approaches proposed by
+\citet{abadi1989dts} and \citet{thatte1990qst} by including a
+dynamic type to a statically typed language.
+However, these three approaches differ in the way they handle the
+dynamic type.
+While \citet{siek2006gradual} rely on the consistency relation,
+\citet{abadi1989dts} rely on type equality with explicit projections
+and injections, and \citet{thatte1990qst} relies on subtyping.
+
+The subtyping relation $\subtype$ actually is a pitfall on Thatte's
+quasi-static typing, because quasi-static typing sets the dynamic type
+as the top and the bottom of the subtying relation:
+$T \subtype \;?$ and $? \subtype T$.
+Subtyping is transitive, so we have that
+\[
+\frac{\Number \subtype \;? \;\;\;
+      ? \subtype \String}
+     {\Number \subtype \String}
+\]
+Therefore, downcasts combined with the transitivity of subtyping
+accepts programs that should be rejected.
+
+Later, \citet{siek2007objects} reported that the consistency relation
 is orthogonal to subtyping, so we can combine the two relations to
@@
 The \emph{consistent-subtyping} relation,
-written $\tau_{1} \lesssim \tau_{2}$,
-is essential to design object-oriented languages that have a
+written $T_{1} \lesssim T_{2}$,
+is the foundation to design object-oriented languages that have a
 gradual type system.
@@
 the consistent-subtyping relation is not transitive.
+Therefore, $\Number \lesssim \;?$, $? \lesssim \Number$,
+$\String \lesssim \;?$, and $? \lesssim \String$,
+but $\Number \not\lesssim \String$, and
+$\String \not\lesssim \Number$.
 
-\begin{figure}[!ht]
-\textbf{Consistent-Subtyping}\\
-\dstart
-$$
-\begin{matrix}
-\Number \lesssim \Number \mylabel{C-NUM} \;\;\; \String \lesssim \String \mylabel{C-STR}
-\\ \\
-T \lesssim \;? \mylabel{C-ANY1} \;\;\; ? \lesssim T \mylabel{C-ANY2}
-\\ \\
-\dfrac{T_{3} \lesssim T_{1} \;\;\; T_{2} \lesssim T_{4}}
-      {T_{1} \rightarrow T_{2} \sim T_{3} \rightarrow T_{4}} \mylabel{C-FUN}
-\;\;\;
-\dfrac{\forall i \in 1..n \;\;\; T_{i} \lesssim T_{i}'}
-      {[l_{i}:T_{i}^{\;i\in1..n+m}] \lesssim [l_{i}:T_{i}'^{\;i\in1..n}]} \mylabel{C-OBJ}
-\end{matrix}
-$$
-\dend
-\caption{The consistent-subtyping relation}
-\label{fig:consistent-subtyping}
-\end{figure}
+So far, gradual typing looks like a mere formalization to optional
+type systems, as a gradual type system uses the consistency or
+consistent-subtyping relation to statically check the interaction
+between statically and dynamically typed code, and not influencing
+the run-time semantics.
 
-Another important feature of gradual typing is the theoretic
-foundation that it provides for inserting run-time checks that
+However, another important feature of gradual typing is the theoretic
+foundation that is provides for inserting run-time checks that
 prove dynamically typed code does not violate the invariants of
 statically typed code, thus preserving type safety.
-To do that, \citet{siek2006gradual} defined the run-time semantics
-of gradual typing as a translation to an intermediate language with
-explicit casts at the frontiers between statically and dynamically
-typed code.
+To do that, \citet{siek2006gradual,siek2007objects} defined the
+run-time semantics of gradual typing as a translation to an
+intermediate language with explicit casts at the frontiers between
+statically and dynamically typed code.
 The semantics of these casts is based on the higher-order contracts
@@
 consistent-subtyping relation of \citet{siek2007objects}, but it is
-also written $\tau_{1} \lesssim \tau_{2}$.
-The consistent-or-subtype relation holds when $\tau_{1} \sim \tau_{2}$
-or $\tau_{1} <: \tau_{2}$, where $<:$ is transitive and $\sim$ is not.
+also written $T_{1} \lesssim T_{2}$.
+The consistent-or-subtype relation holds when $T_{1} \sim T_{2}$
+or $T_{1} <: T_{2}$, where $<:$ is transitive and $\sim$ is not.
 Alore also extends its optional type system to include optional

### 2014-07-16 doc/thesis/cases.tex
Starting to port the current typing of the Lua Standard Libraries to its description file

@@
+Categories:
+\begin{enumerate}
+\item Easy to type, because the current type system is enough.
+\item Not easy to type, because it requires polymorphism.
+\item Hard to type, because neither the current type system nor
+      polymorphism are enough.
+\end{enumerate}
+
 \section{Lua standard libraries}
@@
 \item assert (v [, message]) :
-$\Top \times
-(\String \cup \Nil) \times
-\TopStar \rightarrow
-\NilStar \cup
-(\Top \times
-(\String \cup \Nil) \times
-\TopStar)$ 
+\begin{verbatim}
+(any, string?) -> (any, string?) | (nil)
+\end{verbatim}
+Category 2?
 \\
@@
 \item collectgarbage ([opt [, arg]]) :
-$(\String \cup \Nil) \times
-(\Number \cup \Nil) \times
-\TopStar \rightarrow
-(\Number \times \Nil \times \NilStar) \cup
-(\Number \times \Number \times \NilStar) \cup
-(\Boolean \times \Nil \times \NilStar)$
+\begin{verbatim}
+(string?, number?) -> (number) | (number, number) | (boolean)
+\end{verbatim}
+Category 3.
 \\
 The return type depends on the type of \textbf{opt}:
-\begin{itemize}
-\item $\Nil \times \TopStar \rightarrow \Number \times \NilStar$
-\item $``collect" \times \TopStar \rightarrow \Number \times \NilStar$
-\item $``stop" \times \TopStar \rightarrow \Number \times \NilStar$
-\item $``restart" \times \TopStar \rightarrow \Number \times \NilStar$
-\item $``count" \times \TopStar \rightarrow \Number \times \Number \times \NilStar$
-\item $``step" \times \TopStar \rightarrow \Boolean \times \NilStar$
-\item $``setpause" \times \TopStar \rightarrow \Number \times \NilStar$
-\item $``setstepmul" \times \TopStar \rightarrow \Number \times \NilStar$
-\item $``isrunning" \times \TopStar \rightarrow \Boolean \times \NilStar$
-\item $``generational" \times \TopStar \rightarrow \Number \times \NilStar$
-\item $``incremental" \times \TopStar \rightarrow \Number \times \NilStar$
-\end{itemize}
+\begin{verbatim}
+() -> (number)
+("collect") -> (number)
+("stop") -> (number)
+("restart") -> (number)
+("count") -> (number, number)
+("step") -> (boolean)
+("setpause") -> (number)
+("setstepmul") -> (number)
+("isrunning") -> (boolean)
+("generational") -> (number)
+("incremental") -> (number)
+\end{verbatim}
 \item dofile ([filename]) :
-$(\String \cup \Nil) \times
-\TopStar \rightarrow
-\AnyStar
-$
+\begin{verbatim}
+(string?) -> (any*)
+\end{verbatim}
+Category 3?
 \item error (message [, level]) :
-$\String \times
-(0 \cup 1 \cup 2 \cup \Nil) \times
-\TopStar \rightarrow
-\NilStar$
-\item \_G : $\{ \String:\Top \}$
+\begin{verbatim}
+(string, number?) -> ()
+\end{verbatim}
+Category 1.
+\item \_G :
+\begin{verbatim}
+{string:any}
+\end{verbatim}
+Category 3?
 \item getmetatable (object) :
-$\mu X.\{ \_\_metatable:\Top \} \times
-\TopStar \rightarrow
-(\Nil \cup \{ \_\_metatable:\Top \}) \times
-\NilStar$
+\begin{verbatim}
+(any) -> (any?)
+\end{verbatim}
+Category 3.
 \item ipairs (t) :
-$\{ \Number:\Any \} \times
-\TopStar \rightarrow
-\Number \times
-\Any \times
-\NilStar$
+\begin{verbatim}
+({any}) -> ((any*) -> (any*)
+\end{verbatim}
+Category 3.
+\\
 \begin{itemize}
 \item What happens when \textbf{t} has a metamethod \textbf{\_\_ipairs}?
-\item Will $\{ ``tag":\String,\; \Number:\Any \}$ type check?
+\item Will \texttt{\{"tag":string, number:any\}} type check?
 \item In fact, \textbf{ipairs} and \textbf{pairs} return three values.
@@
 \item load (ld [, source [, mode [, env]]]) :
-$(\String \cup (\TopStar \rightarrow \TopStar)) \times
-(\String \cup \Nil) \times
-(``b" \cup ``t" \cup ``bt" \cup \Nil) \times
-(\Top \cup \Nil) \times
-\TopStar \rightarrow
-((\TopStar \rightarrow \TopStar) \times \NilStar) \cup
-(\Nil \times \String \times \NilStar)$
+\begin{verbatim}
+(string|(any*) -> (any*), string?, string?, any?) ->
+((any*) -> (any*))?
+\end{verbatim}
+Category 3?
+\\
+I guess the proper type should be:
+\begin{verbatim}
+(string|(any*) -> (any*)) |
+(string|(any*) -> (any*), string) |
+(string|(any*) -> (any*), string, string, any?) ->
+((any*) -> (any*))?
+\end{verbatim}
 \item loadfile ([filename [, mode [, env]]]) :
-$(\String \cup \Nil) \times
-(``b" \cup ``t" \cup ``bt" \cup \Nil) \times
-(\Top \cup \Nil) \times
-\TopStar \rightarrow
-((\TopStar \rightarrow \TopStar) \times \NilStar) \cup
-(\Nil \times \String \times \NilStar)$
+\begin{verbatim}
+(string?, string?, any?) -> ((any*) -> (any*))?
+\end{verbatim}
+Category 3?
+\\
+I guess the proper type should be:
+\begin{verbatim}
+() | (string) | (string, string) | (string, string, any?) ->
+((any*) -> (any*))?
+\end{verbatim}
 \item next (table [, index]) :
-$\{ \Any:\Any \} \times
-(\Any \cup \Nil) \times
-\TopStar \rightarrow
-(\Nil \times \NilStar) \cup
-(\Any \times \Any \times \NilStar)$
+\begin{verbatim}
+({any:any}, any?) -> (any, any) | (nil)
+\end{verbatim}
+Category 2.
 \item pairs (t) :
-$\{ \Any:\Any \} \times
-\TopStar \rightarrow
-\Any \times 
-\Any \times
-\NilStar$
+\begin{verbatim}
+({any}) -> ((any*) -> (any*)
+\end{verbatim}
+Category 3.
 \item pcall (f [, arg1, ···]) :
-$(\TopStar \rightarrow \TopStar) \times
-\TopStar \rightarrow
-(\Boolean \times \AnyStar) \cup
-(\Boolean \times \String \times \NilStar)$
-\item print (···) : $\TopStar \rightarrow \NilStar$
+\begin{verbatim}
+((any*) -> (any*), any*) -> (true, any*) | (false, string)
+\end{verbatim}
+Category 3?
+\item print (···) :
+\begin{verbatim}
+(value*) -> ()
+\end{verbatim}
+Category 1.
 \item rawequal (v1, v2) :
-$\Top \times
-\Top \times
-\TopStar \rightarrow
-\Boolean \times
-\NilStar$
+\begin{verbatim}
+(value, value) -> (boolean)
+\end{verbatim}
+Category 2.
 \item rawget (table, index) :
-$\{ \Top:\Top \} \times
-\Top \times
-\TopStar \rightarrow
-\Any \times
-\NilStar$
+\begin{verbatim}
+({}, any) -> (any)
+\end{verbatim}
+Category 2.
 \item rawlen (v) :
-$(\{ \Top:\Top \} \cup \String) \times
-\TopStar \rightarrow
-\Number \times
-\NilStar$
+\begin{verbatim}
+({}|string) -> (number)
+\end{verbatim}
+Category 1.
 \item rawset (table, index, value) :
-$\{ \Top:\Top \} \times
-\Top \times
-\Top \times
-\TopStar \rightarrow
-\{ \Any:\Any \} \times
-\NilStar$
+\begin{verbatim}
+({any:any}, any, any) -> ({any:any})
+\end{verbatim}
+Category 3.
 \\
@@
 \item select (index, ···) :
-$\Number \times
-\AnyStar \rightarrow
-\AnyStar$
+\begin{verbatim}
+(number, any*) -> (any*)
+\end{verbatim}
+Category 3.
 \\
 When \textbf{index} is the string ``\texttt{\#}", the typing is:
-$\String \times \AnyStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(string, any*) -> (number)
+\end{verbatim}
 \item setmetatable (table, metatable) :
-$\{ \Top:\Top \} \times
-(\Nil \cup \{ \_\_index:\mu X.\{ \Top:\Top \} \}) \times
-\TopStar \rightarrow
-\{ \_\_index:\mu X.\{ \Top:\Top \} \} \times
-\NilStar$
+\begin{verbatim}
+({any:any}, {any:any}|nil) -> ({any:any}) 
+\end{verbatim}
+Category 3.
 \\
 Even though \textbf{metatable} can be $\Nil$, it is not optional.
+This function is a primitive in the type system.
 \item tonumber (e [, base]) :
-$\String \times
-(\Number \cup \Nil) \times
-\TopStar \rightarrow
-\Number \times
-\NilStar$
+\begin{verbatim}
+(value, number|string?) -> (number) | (nil)
+\end{verbatim}
+Category 3.
+\\
+The type of \textbf{base} depends on the type of \textbf{e}:
+\begin{verbatim}
+(value) -> (number) | (nil)
+(string, number|string) -> (number) | (nil)
+\end{verbatim}
 \item tostring (v) :
-$\Top \times
-\TopStar \rightarrow
-\String \times
-\NilStar$
+\begin{verbatim}
+(value) -> (string)
+\end{verbatim}
+Category 1.
 \item type (v) :
-$\Top \times
-\TopStar \rightarrow
-(``nil" \cup
-``number" \cup
-``string" \cup
-``boolean" \cup
-``table" \cup
-``function" \cup
-``thread" \cup
-``userdata") \times
-\NilStar$
-\item \_VERSION : $\String$
+\begin{verbatim}
+(value) ->
+("nil"|"number"|"string"|"boolean"|"table"|"function"|"thread"|"userdata")
+\end{verbatim}
+Category 1.
+\\
+This function is a primitive in the type system.
+\item \_VERSION :
+\begin{verbatim}
+string
+\end{verbatim}
+Category 1.
 \item xpcall (f, msgh [, arg1, ···]) :
-$(\TopStar \rightarrow \TopStar) \times
-\Top \times
-\TopStar \rightarrow
-(\Boolean \times \AnyStar) \cup
-(\Boolean \times \String \times \NilStar)$
+\begin{verbatim}
+((any*) -> (any*), any, any*) -> (true, any*) | (false, string)
+\end{verbatim}
+Category 3?
 \end{enumerate}
@@
 
-\subsection{String manipulation}
+\subsection{String Manipulation}
 
@@
 \item string.byte (s [, i [, j]]) :
-$\String \times
-(\Number \cup \Nil) \times
-(\Number \cup \Nil) \times
-\TopStar \rightarrow
-\Number \times
-\Number{*}$
+\begin{verbatim}
+(string, number?, number?) -> (number*)
+\end{verbatim}
+Category 1.
+\\
+I guess the proper type should be:
+\begin{verbatim}
+(string) |
+(string, number, number?) -> (number*)
+\end{verbatim}
 \item string.char (···) :
-$\Number{*} \rightarrow
-\String \times
-\NilStar$
+\begin{verbatim}
+(number*) -> (string)
+\end{verbatim}
+Category 1.
 \item string.dump (function) :
-$(\TopStar \rightarrow \TopStar) \times
-\TopStar \rightarrow
-\String \times
-\NilStar$
+\begin{verbatim}
+((value*) -> (value*)) -> (string)
+\end{verbatim}
+Category 3?
 \item string.find (s, pattern [, init [, plain]]) :
-$\String \times
-\String \times
-(\Number \cup \Nil) \times
-(\Boolean \cup \Nil) \times
-\TopStar \rightarrow
-(\Number \times \Number \times \NilStar) \cup
-(\Nil \times \NilStar)$
+\begin{verbatim}
+(string, string, number?, true?) -> (number, number) | (nil)
+\end{verbatim}
+Category 1.
+\\
+I guess the proper type should be:
+\begin{verbatim}
+(string, string) |
+(string, string, number, true?) -> (number, number) | (nil)
+\end{verbatim}
 \item string.format (formatstring, ···) :
-$\String \times
-\TopStar \rightarrow
-\String \times
-\NilStar$
+\begin{verbatim}
+(string, any*) -> (string)
+\end{verbatim}
+Category 1.
 \item string.gmatch (s, pattern) :
-$\String \times
-\String \times
-\TopStar \rightarrow
-\String{*}$
+\begin{verbatim}
+(string, string) -> ((any*) -> (any*))
+\end{verbatim}
+Category 3.
 \\
@@
 \item string.gsub (s, pattern, repl [, n]) :
-$\String \times
-\String \times
-(\String \cup \{ \Top:\Top \} \cup (\TopStar \rightarrow \TopStar)) \times
-(\Number \cup \Nil) \times
-\TopStar \rightarrow
-\String \times
-\Number \times
-\NilStar$
+\begin{verbatim}
+(string, string, string|{}|(any*) -> (any*), number?) ->
+(string, number)
+\end{verbatim}
+Category 1.
 \item string.len (s) :
-$\String \times
-\TopStar \rightarrow
-\Number \times
-\NilStar$
+\begin{verbatim}
+(string) -> (number)
+\end{verbatim}
+Category 1.
 \item string.lower (s) :
-$\String \times
-\TopStar \rightarrow
-\String \times
-\NilStar$
+\begin{verbatim}
+(string) -> (string)
+\end{verbatim}
+Category 1.
 \item string.match (s, pattern [, init]) :
-$\String \times
-\String \times
-(\Number \cup \Nil) \times
-\TopStar \rightarrow
-(\String \times \String{*}) \cup
-(\Nil \times \NilStar)$
+\begin{verbatim}
+(string, string, number?) -> (string, string*) | (nil)
+\end{verbatim}
+Category 1.
 \item string.rep (s, n [, sep]) :
-$\String \times
-\Number \times
-(\String \cup \Nil) \times
-\TopStar \rightarrow
-\String \times
-\NilStar$
+\begin{verbatim}
+(string, number, string?) -> (string)
+\end{verbatim}
+Category 1.
 \item string.reverse (s) :
-$\String \times
-\TopStar \rightarrow
-\String \times
-\NilStar$
+\begin{verbatim}
+(string) -> (string)
+\end{verbatim}
+Category 1.
 \item string.sub (s, i [, j]) :
-$\String \times
-\Number \times
-(\Number \cup \Nil) \times
-\TopStar \rightarrow
-\String \times
-\NilStar$
+\begin{verbatim}
+(string, number, number?) -> (string)
+\end{verbatim}
+Category 1.
 \item string.upper (s) :
-$\String \times
-\TopStar \rightarrow
-\String \times
-\NilStar$
+\begin{verbatim}
+(string) -> (string)
+\end{verbatim}
+Category 1.
 \end{enumerate}
 
-\subsection{Table manipulation}
+\subsection{Table Manipulation}
 
@@
 \item table.concat (list [, sep [, i [, j]]]) :
-$\{ \Number:\String \cup \Number \} \times
-(\String \cup \Nil) \times
-(\Number \cup \Nil) \times
-(\Number \cup \Nil) \times
-\TopStar \rightarrow
-\String \times
-\NilStar$
+\begin{verbatim}
+({string|number}, string?, number?, number?) -> (string)
+\end{verbatim}
+Category 1.
+\\
+I guess the proper type should be:
+\begin{verbatim}
+({string|number}) |
+({string|number}, string) |
+({string|number}, string, number?) -> (string)
+\end{verbatim}
 \item table.insert (list, [pos,] value) :
-$\{ \Number:\Top \} \times
-(\Number \times \Top) \cup
-(\Top \times \Nil) \times
-\TopStar \rightarrow
-\NilStar$
+\begin{verbatim}
+({any}, any) -> ()
+\end{verbatim}
+Category 2.
 \\
@@
 in the \textbf{list}, that is, the input type is
-$\{ \Number:\Top \} \times \Top \times \TopStar$.
+\texttt{(\{any\}, any) -> ()}.
 If \textbf{value} is $\Nil$, the function does not insert \textbf{pos}
 in the \textbf{list}, that is, the input type is
-$\{ \Number:\Top \} \times \Number \times \Top \times \TopStar$,
-but the element is not inserted in the list because $\Top$ is actually $\Nil$.
+\texttt{(\{any\}, number, any) -> ()},
+but the element is not inserted in the list because \textbf{value}
+is actually $\Nil$.
+I guess the proper type should be:
+\begin{verbatim}
+({any}, any) | ({any}, number, any) -> ()
+\end{verbatim}
 \item table.pack (···) :
-$\TopStar \rightarrow
-\{ \Number:\Top \} \times
-\NilStar$
+\begin{verbatim}
+(any*) -> ({number:any, "n":number})
+\end{verbatim}
+Category 2.
 \\
@@
 \item table.remove (list [, pos]) :
-$\{ \Number:\Top \} \times
-(\Number \cup \Nil) \times
-\TopStar \rightarrow
-\Top \times
-\NilStar
-$
+\begin{verbatim}
+({any}, number?) -> (any)
+\end{verbatim}
+Category 2.
 \\
@@
 \item table.sort (list [, comp]) :
-$\{ \Number:\Top \} \times
-((\TopStar \rightarrow \TopStar) \cup \Nil) \times
-\TopStar \rightarrow
-\NilStar$
+\begin{verbatim}
+({any}, any?) -> ()
+\end{verbatim}
+Category 2.
+\\
+The proper type should be:
+\begin{verbatim}
+({any}, (any, any) -> (boolean)?) -> ()
+\end{verbatim}
+But there is a syntactic conflict for defining optional function types
+that should be fixed in the parser.
 \item table.unpack (list [, i [, j]]) :
-$\{ \Number:\Top \} \times
-(\Number \cup \Nil) \times
-(\Number \cup \Nil) \times
-\TopStar \rightarrow
-\TopStar$
+\begin{verbatim}
+({any}, number?, number?) -> (any*)
+\end{verbatim}
+Category 2.
 \\
@@
 the list.
+I guess the proper type should be:
+\begin{verbatim}
+({any}) | ({any}, number, number?) -> (any*)
+\end{verbatim}
 \end{enumerate}
 
-\subsection{Mathematical functions}
+\subsection{Mathematical Functions}
 
@@
 \item math.abs (x) :
-$\Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.acos (x) :
-$\Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.asin (x) :
-$\Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.atan (x) :
-$\Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.atan2 (y, x) :
-$\Number \times \Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number, number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.ceil (x) :
-$\Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.cos (x) :
-$\Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.cosh (x) :
-$\Number \times \TopStar \rightarrow \Number \times \NilStar$  
+\begin{verbatim}
+(number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.deg (x) :
-$\Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.exp (x) :
-$\Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.floor (x) :
-$\Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.fmod (x, y) :
-$\Number \times \Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number, number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.frexp (x) :
-$\Number \times \TopStar \rightarrow \Number \times \Number \times \NilStar$
-\item math.huge : $\Number$
+\begin{verbatim}
+(number) -> (number, number)
+\end{verbatim}
+Category 1.
+\item math.huge :
+\begin{verbatim}
+number
+\end{verbatim}
+Category 1.
 \item math.ldexp (m, e) :
-$\Number \times \Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number, number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.log (x [, base]) :
-$\Number \times
-(\Number \cup \Nil) \times
-\TopStar \rightarrow
-\Number \times
-\NilStar$
+\begin{verbatim}
+(number, number?) -> (number)
+\end{verbatim}
+Category 1.
 \item math.max (x, ···) :
-$\Number \times \Number{*} \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number, number*) -> (number)
+\end{verbatim}
+Category 1.
 \item math.min (x, ···) :
-$\Number \times \Number{*} \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number, number*) -> (number)
+\end{verbatim}
+Category 1.
 \item math.modf (x) :
-$\Number \times \TopStar \rightarrow \Number \times \Number \times \NilStar$
-\item math.pi : $\Number$
+\begin{verbatim}
+(number) -> (number, number)
+\end{verbatim}
+Category 1.
+\item math.pi :
+\begin{verbatim}
+number
+\end{verbatim}
+Category 1.
 \item math.pow (x, y) :
-$\Number \times \Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number, number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.rad (x) :
-$\Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.random ([m [, n]]) :
-$(\Number \cup \Nil) \times
-(\Number \cup \Nil) \rightarrow
-\Number \times
-\NilStar$
+\begin{verbatim}
+(number?, number?) -> (number)
+\end{verbatim}
+Category 1.
 \\
@@
 This function does not accept/discard extra parameters.
+I guess the proper type should be:
+\begin{verbatim}
+() | (number, number?) -> (number)
+\end{verbatim}
 \item math.randomseed (x) :
-$\Number \times \TopStar \rightarrow \NilStar$
+\begin{verbatim}
+(number) -> ()
+\end{verbatim}
+Category 1.
 \item math.sin (x) :
-$\Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.sinh (x) :
-$\Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.sqrt (x) :
-$\Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.tan (x) :
-$\Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number) -> (number)
+\end{verbatim}
+Category 1.
 \item math.tanh (x) :
-$\Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number) -> (number)
+\end{verbatim}
+Category 1.
 \end{enumerate}
 
-\subsection{Bitwise operations}
+\subsection{Bitwise Operations}
 
@@
 \item bit32.arshift (x, disp) :
-$\Number \times \Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number, number) -> (number)
+\end{verbatim}
+Category 1.
 \item bit32.band (···) :
-$\Number{*} \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number*) -> (number)
+\end{verbatim}
+Category 1.
 \item bit32.bnot (x) :
-$\Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number) -> (number)
+\end{verbatim}
+Category 1.
 \item bit32.bor (···) :
-$\Number{*} \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number*) -> (number)
+\end{verbatim}
+Category 1.
 \item bit32.btest (···) :
-$\Number{*} \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number*) -> (number)
+\end{verbatim}
+Category 1.
 \item bit32.bxor (···) :
-$\Number{*} \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number*) -> (number)
+\end{verbatim}
+Category 1.
 \item bit32.extract (n, field [, width]) :
-$\Number \times
-\Number \times
-(\Number \cup \Nil) \times
-\TopStar \rightarrow
-\Number \times
-\NilStar$
+\begin{verbatim}
+(number, number, number?) -> (number)
+\end{verbatim}
+Category 1.
+\\
+Should it be?
+\begin{verbatim}
+(number, 0|...|31, 1|...|32|nil) -> (number)
+\end{verbatim}
 \item bit32.replace (n, v, field [, width]) :
-$\Number \times
-\Number \times
-\Number \times
-(\Number \cup \Nil) \times
-\TopStar \rightarrow
-\Number \times
-\NilStar$
+\begin{verbatim}
+(number, number, number, number?) -> (number)
+\end{verbatim}
+Category 1.
+\\
+Should it be?
+\begin{verbatim}
+(number, number, 0|...|31, 1|...|32|nil) -> (number)
+\end{verbatim}
 \item bit32.lrotate (x, disp) :
-$\Number \times \Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number, number) -> (number)
+\end{verbatim}
+Category 1.
 \item bit32.lshift (x, disp) :
-$\Number \times \Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number, number) -> (number)
+\end{verbatim}
+Category 1.
 \item bit32.rrotate (x, disp) :
-$\Number \times \Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number, number) -> (number)
+\end{verbatim}
+Category 1.
 \item bit32.rshift (x, disp) :
-$\Number \times \Number \times \TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+(number, number) -> (number)
+\end{verbatim}
+Category 1.
 \end{enumerate}

### 2014-07-17 doc/thesis/cases.tex
Finishing to port the current typing of the Lua Standard Libraries to its description file

@@
 
-\subsection{Coroutine manipulation}
+\subsection{Coroutine Manipulation}
 
@@
 \item coroutine.create (f) :
-$(\TopStar \rightarrow \TopStar) \times \TopStar \rightarrow Thread \times \NilStar$
+\begin{verbatim}
+((any*) -> (any*)) -> (any)
+\end{verbatim}
+Category 3.
 \item coroutine.resume (co [, val1, ···]) :
-$Thread \times
-\TopStar \rightarrow
-(\True \times \AnyStar) \cup
-(\False \times \String \times \NilStar)$
+\begin{verbatim}
+(any, any*) -> (true, any*) | (false, string)
+\end{verbatim}
+Category 3.
 \item coroutine.running () :
-$\TopStar \rightarrow Thread \times \Boolean \times \NilStar$
+\begin{verbatim}
+() -> (any, boolean)
+\end{verbatim}
+Category 3.
 \item coroutine.status (co) :
-$Thread \times \TopStar \rightarrow \String \times \NilStar$
+\begin{verbatim}
+(any) -> ("running"|"suspended"|"normal"|"dead")
+\end{verbatim}
+Category 3.
 \item coroutine.wrap (f) :
-$(\TopStar \rightarrow \TopStar) \times
-\TopStar \rightarrow
-(\TopStar \rightarrow \TopStar) \times
-\NilStar$ 
+\begin{verbatim}
+((any*) -> (true, any*) | (false, string)) ->
+((any*) -> (true, any*) | (false, string))
+\end{verbatim}
+Category 3.
 \item coroutine.yield (···) :
-$\TopStar \rightarrow \NilStar$
+\begin{verbatim}
+(any*) -> ()
+\end{verbatim}
+Category 1?
 \end{enumerate}
@@
 \item require (modname) :
-$\String \times \TopStar \rightarrow \{\Any:\Any\} \times \NilStar$
+\begin{verbatim}
+(string) -> ({any:any})
+\end{verbatim}
+Category 1.
 \\
@@
 converted to $\True$.
-\item package.config : $\String$
-\item package.cpath : $\String$
-\item package.loaded : $\{\String:\{\Any:\Any\}\}$
+This function is a primitive in the type system.
+\item package.config :
+\begin{verbatim}
+string
+\end{verbatim}
+Category 1.
+\item package.cpath :
+\begin{verbatim}
+string
+\end{verbatim}
+Category 1.
+\item package.loaded :
+\begin{verbatim}
+{string:{any:any}}
+\end{verbatim}
+Category 2?
 \item package.loadlib (libname, funcname) :
-$\String \times
-\String \times
-\TopStar \rightarrow
-(\True \times \NilStar) \cup
-(\Nil \times \String \times \NilStar)$
-\item package.path : $\String$
-\item package.preload : $\{\String:\{\Any:\Any\}\}$
-\item package.searchers : $\{\Number:\TopStar \rightarrow \TopStar\}$
+\begin{verbatim}
+(string, string) -> (true)?
+\end{verbatim}
+Category 1?
+\item package.path :
+\begin{verbatim}
+string
+\end{verbatim}
+Category 1.
+\item package.preload :
+\begin{verbatim}
+{string:{any:any}}
+\end{verbatim}
+Category 2?
+\item package.searchers :
+\begin{verbatim}
+{(any*) -> (any*)}
+\end{verbatim}
+Category 2?
 \item package.searchpath (name, path [, sep [, rep]]) :
-$\String \times
-\String \times
-(\String \cup \Nil) \times
-(\String \cup \Nil) \times
-\TopStar \rightarrow
-(\String \times \NilStar) \cup
-(\Nil \times \String \times \NilStar)$
+\begin{verbatim}
+(string, string, string?, string?) -> (string)?
+\end{verbatim}
+Category 1.
+\\
+I guess the proper type should be:
+\begin{verbatim}
+(string, string) |
+(string, string, string) |
+(string, string, string, string?) -> (string)?
+\end{verbatim}
 \end{enumerate}
@@
 
-\subsection{Input and Output facilities}
+\subsection{Input and Output Facilities}
 
@@
 \item io.close ([file]) :
-$(Userdata \cup \Nil) \times
-\TopStar \rightarrow
-(\Boolean \times \NilStar) \cup
-(\Nil \times \String \times \NilStar) \cup
-((\True \cup \Nil) \times \String \times \Number \times \NilStar)$
+\begin{verbatim}
+(any?) -> (boolean) | (nil, string) | (true?, string, number)
+\end{verbatim}
+Category 1?
 \\
-Returns $((\True \cup \Nil) \times \String \times \Number \times \NilStar)$
-when the file handle was created with \textbf{io.popen}.
+Returns \texttt{(true?, string, number)} when the file handle
+was created with \textbf{io.popen}.
+The parameter \textbf{file} is a \textit{userdata} and I need
+to fix that on the type system.
 \item io.flush () :
-$\TopStar \rightarrow \Boolean \times \NilStar$
+\begin{verbatim}
+() -> (boolean)
+\end{verbatim}
+Category 1.
 \item io.input ([file]) :
-$(\String \cup Userdata \cup \Nil) \times
-\TopStar \rightarrow
-Userdata \times
-\NilStar$
+\begin{verbatim}
+(string|any|nil) -> (any)
+\end{verbatim}
+Category 1?
+Here the type \texttt{any} refers to a \textit{userdata} and
+I need to fix that on the type system.
 \item io.lines ([filename ···]) :
-$(\String \cup Userdata \cup \Nil) \times
-\TopStar \rightarrow
-(\String \times \NilStar) \cup
-(\Nil \times \NilStar)$
+\begin{verbatim}
+(string|any|nil) -> ((any*) -> (any*))
+\end{verbatim}
+Category 1?
 \\
 In fact, this function returns an iterator. Need to check the return type.
+Here the type \texttt{any} refers to a \textit{userdata} and
+I need to fix that on the type system.
 \item io.open (filename [, mode]) :
-$\String \times
-(``r" \cup ``w" \cup ``a" \cup
-``r+" \cup ``w+" \cup ``a+" \cup
-``rb" \cup ``wb" \cup ``ab" \cup
-``r+b" \cup ``w+b" \cup ``a+b" \cup
-\Nil) \times
-\TopStar \rightarrow
-(Userdata \times \NilStar) \cup
-(\Nil \times \String \times \Number \times \NilStar)$
+\begin{verbatim}
+(string,
+"r"|"w"|"a"|"r+"|"w+"|"a+"|"rb"|"wb"|"ab"|"r+b"|"w+b"|"w+b"|nil) ->
+(any) | (nil, string, number)
+\end{verbatim}
+Category 1?
+\\
+Here the type \texttt{any} refers to a \textit{userdata} and
+I need to fix that on the type system.
 \item io.output ([file]) :
-$(\String \cup Userdata \cup \Nil) \times
-\TopStar \rightarrow
-Userdata \times
-\NilStar$
+\begin{verbatim}
+(string|any|nil) -> (any)
+\end{verbatim}
+Category 1?
+\\
+Here the type \texttt{any} refers to a \textit{userdata} and
+I need to fix that on the type system.
 \item io.popen (prog [, mode]) :
-$\String \times
-(``r" \cup ``w" \cup \Nil) \times
-\TopStar \rightarrow
-(Userdata \times \NilStar) \cup
-(\Nil \times \String \times \Number \times \NilStar)$
+\begin{verbatim}
+(string, "r"|"w"|nil) -> (any) | (nil, string, number)
+\end{verbatim}
+Category 1?
+\\
+Here the type \texttt{any} refers to a \textit{userdata} and
+I need to fix that on the type system.
 \item io.read (···) :
-$(``*n" \cup ``*a" \cup ``*l" \cup ``*L" \cup \Number){*} \rightarrow
-(\String \cup \Number){*}
-$
+\begin{verbatim}
+("*n"|"*a"|"*l"|"*L"|number*) -> (string|number*)
+\end{verbatim}
+Category 1.
 \item io.tmpfile () :
-$\TopStar \rightarrow Userdata \times \NilStar$
+\begin{verbatim}
+() -> (any)
+\end{verbatim}
+Category 1?
+\\
+Here the type \texttt{any} refers to a \textit{userdata} and
+I need to fix that on the type system.
 \item io.type (obj) :
-$\Top \times \TopStar \rightarrow (``file" \cup ``closed\;file" \cup \Nil) \times \NilStar$
+\begin{verbatim}
+(value) -> ("file"|"closed file"|nil)
+\end{verbatim}
+Category 1.
 \item io.write (···) :
-$(\String \cup \Number){*} \rightarrow
-(Userdata \times \NilStar) \cup
-(\Nil \times \String \times \Number \times \NilStar)$
+\begin{verbatim}
+(string|number*) -> (any) | (nil, string, number)
+\end{verbatim}
+Category 1?
+\\
+Here the type \texttt{any} refers to a \textit{userdata} and
+I need to fix that on the type system.
 \item file:close () : ?
@@
 
-\subsection{Operating System facilities}
+\subsection{Operating System Facilities}
 
@@
 \item os.clock () :
-$\TopStar \rightarrow \Number \times \NilStar$
+\begin{verbatim}
+() -> (number)
+\end{verbatim}
+Category 1.
 \item os.date ([format [, time]]) :
-$(\String \cup \Nil) \times
-(\Number \cup \Nil) \times
-\TopStar \rightarrow
-(\String \times \NilStar) \cup
-(\{``year":\Number,\;
-   ``month":\Number,\;
-   ``day":\Number,\;
-   ``hour":\Number,\;
-   ``min":\Number,\;
-   ``sec":\Number,\;
-   ``wday":\Number,\;
-   ``yday":\Number,\;
-   ``isdst":\Boolean \} \times \NilStar)$
+\begin{verbatim}
+(string?, number?) -> (string) |
+                      ({"year":number, "month":number, "day":number,
+                        "hour":number, "min":number, "sec":number,
+                        "wday":number, "yday":number, "isdst":boolean})
+\end{verbatim}
+Category 1.
 \\
 Returns a table when \textbf{format} contains ``\textbf{*t}''.
+I guess the proper type should be:
+\begin{verbatim}
+() | (string) | (string, number?) -> (string) |
+                      ({"year":number, "month":number, "day":number,
+                        "hour":number, "min":number, "sec":number,
+                        "wday":number, "yday":number, "isdst":boolean})
+\end{verbatim}
 \item os.difftime (t2, t1) :
-$\Number \times
-(\Number \cup \Nil) \times
-\TopStar \rightarrow
-\Number \times
-\NilStar$
+\begin{verbatim}
+(number, number?) -> (number)
+\end{verbatim}
+Category 1.
 \\
@@
 \item os.execute ([command]) :
-$(\String \cup \Nil) \times
-\TopStar \rightarrow
-(\Boolean \times \NilStar) \cup
-((\True \cup \Nil) \times \String \times \Number \times \NilStar)$
+\begin{verbatim}
+(string?) -> (boolean) | (true?, string, number)
+\end{verbatim}
+Category 3.
 \\
@@
 This Boolean means whether there is a shell available.
-Returns $\True \times \String \times \Number \times \NilStar$ when
+Returns \texttt{(true, string, number)} when
 \textbf{command} terminated successfully.
-Returns $\Nil \times \String \times \Number \times \NilStar$ when
-\textbf{command} did not terminate sucessfully.
+Returns \texttt{(nil, string, number)} when
+\textbf{command} did not terminate successfully.
 \item os.exit ([code [, close]]) :
-$(\Number \cup \Boolean \cup \Nil) \times
-\Top \times
-\TopStar \rightarrow
-\NilStar$
+\begin{verbatim}
+(boolean|number?, value?) -> ()
+\end{verbatim}
+Category 1.
+\\
+I guess the proper type should be:
+\begin{verbatim}
+() | (boolean|number) | (boolean|number, value?) -> ()
+\end{verbatim}
 \item os.getenv (varname) :
-$\String \times
-\TopStar \rightarrow
-(\String \times \NilStar) \cup
-(\Nil \times \NilStar)$
+\begin{verbatim}
+(string) -> (string)
+\end{verbatim}
+Category 1.
 \item os.remove (filename) :
-$\String \times
-\TopStar \rightarrow
-(\True \times \NilStar) \cup
-(\Nil \times \String \times \Number \NilStar)$
+\begin{verbatim}
+(string) -> (true) | (nil, string, number)
+\end{verbatim}
+Category 1.
 \item os.rename (oldname, newname) :
-$\String \times
-\String \times
-\TopStar \rightarrow
-(\True \times \NilStar) \cup
-(\Nil \times \String \times \Number \NilStar)$
+\begin{verbatim}
+(string) -> (true) | (nil, string, number)
+\end{verbatim}
+Category 1.
 \item os.setlocale (locale [, category]) :
-$(\String \cup \Nil) \times
-(``all" \cup
-``collate" \cup
-``ctype" \cup
-``monetary" \cup
-``numeric" \cup
-``time" \cup
-\Nil) \times
-\TopStar \rightarrow
-(\String \times \NilStar) \cup
-(\Nil \times \NilStar)$
+\begin{verbatim}
+(string?, "all"|"collate"|"ctype"|"monetary"|"numeric"|"time"|nil) ->
+(string) | (nil)
+\end{verbatim}
+Category 3?
+\\
+The return type depends on the input type:
+\begin{verbatim}
+() -> (string)
+(string, "all"|"collate"|"ctype"|"monetary"|"numeric"|"time"|nil) ->
+(string) | (nil)
+\end{verbatim}
 \item os.time ([table]) :
-$(\{``year":\Number,\;
-   ``month":\Number,\;
-   ``day":\Number,\;
-   ``hour":(\Number \cup \Nil),\;
-   ``min":(\Number \cup \Nil),\;
-   ``sec":(\Number \cup \Nil),\;
-   ``wday":(\Number \cup \Nil),\;
-   ``yday":(\Number \cup \Nil),\;
-   ``isdst":(\Boolean \cup \Nil) \} \cup \Nil) \times
-\TopStar \rightarrow
-\Number \times
-\NilStar$
+\begin{verbatim}
+({"year":number, "month":number, "day":number,
+  "hour":number?, "min":number?, "sec":number?,
+  "wday":number?, "yday":number?, "isdst":boolean?}?) -> (number)
+\end{verbatim}
+Category 1.
 \item os.tmpname () :
-$\TopStar \rightarrow \String \times \NilStar$
+\begin{verbatim}
+() -> (string)
+\end{verbatim}
+Category 1.
 \end{enumerate}
 
-\subsection{The debug library}
+\subsection{The Debug Library}
 

### 2014-07-24 doc/thesis/cases.tex
Removing the typyping of the debug library
Fixing the type of table.sort

@@
 \begin{verbatim}
-({any}, any?) -> ()
+({any}, nil|(any, any) -> (boolean)) -> ()
 \end{verbatim}
@@
 \\
-The proper type should be:
+Do we really need grouping?
 \begin{verbatim}
-({any}, (any, any) -> (boolean)?) -> ()
+({any}, [(any, any) -> (boolean)]?) -> ()
 \end{verbatim}
-But there is a syntactic conflict for defining optional function types
-that should be fixed in the parser.
+Grouping would be useful for defining types like this:
+\begin{verbatim}
+([(number) -> (number)] | (string) -> (string)) -> ({string})
+\end{verbatim}
+But we do not have intersection types anyway.
 \item table.unpack (list [, i [, j]]) :
@@
 \begin{verbatim}
-(string) -> (string)
+(string) -> (string?)
 \end{verbatim}
@@
 \begin{verbatim}
-(string) -> (true) | (nil, string, number)
+(string, string) -> (true) | (nil, string, number)
 \end{verbatim}
@@
 
-\begin{enumerate}
-\item debug.debug () : ?
-\item debug.gethook ([thread]) : ?
-\item debug.getinfo ([thread,] f [, what]) : ?
-\item debug.getlocal ([thread,] f, local) : ?
-\item debug.getmetatable (value) : ?
-\item debug.getregistry () : ?
-\item debug.getupvalue (f, up) : ?
-\item debug.getuservalue (u) : ?
-\item debug.sethook ([thread,] hook, mask [, count]) : ?
-\item debug.setlocal ([thread,] level, local, value) : ?
-\item debug.setmetatable (value, table) : ?
-\item debug.setupvalue (f, up, value) : ?
-\item debug.setuservalue (udata, value) : ?
-\item debug.traceback ([thread,] [message [, level]]) : ?
-\item debug.upvalueid (f, n) : ?
-\item debug.upvaluejoin (f1, n1, f2, n2) : ?
-\end{enumerate}
+We will not type the debug libraty.

### 2014-07-26 doc/thesis/introduction.tex
Starting the rewriting of the Typed Lua chapter

@@
 In this work we present the design and implementation of Typed Lua:
-an optional type system for Lua that is complex enough to
+an optional type system for Lua that is rich enough to
 preserve some of the Lua idioms that programmers are already used to,

### 2014-07-26 doc/thesis/typedlua.tex
Starting the rewriting of the Typed Lua chapter

@@
 
-\section{Typed Lua through examples}
+\section{Typed Lua}
 
-Our main goal is to allow programmers to combine static and dynamic
-typing in Lua, so we propose Typed Lua as an extension to Lua 5.2
+The main goal of Typed Lua is to allow programmers to combine static
+and dynamic typing in Lua, so Typed Lua is an extension to Lua 5.2
 syntax that introduces optional type annotations, and class-based
@@
 interfaces, and modules.
-An important requirement of Typed Lua is that it should be backwards
-compatible with Lua, that is, it should allow programmers to run any
-Lua code without modifications.
-For this reason, these syntactic extensions include new keywords
-that are not reserved words.
+An important feature of Typed Lua is its backwards compatibility with
+Lua, that is, unannotated Typed Lua code is valid Lua code.
+For this reason, the syntactic extensions introduced by Typed Lua
+include new keywords that are not reserved words.
+Appendix \ref{app:syntax} presents the complete syntax of Typed Lua.
 
-Figure \ref{fig:typelang} shows that Typed Lua includes first-class
-and second-class types.
-Typed Lua uses first-class types for representing Lua values and
-second-class types for defining the signatures of functions.
-First-class types include literal types, base types, the type $\Any$,
-union types, function types, table types, and recursive types.
-Second-class types include the type $\Void$, first-class types,
-vararg types, and tuple types.
-Literal types are types that represent a single value
-(e.g., $\Nil$, $\False$, $\True$, literal numbers, and
-literal strings).
-Literal types are subtypes of their respective base types.
-More precisely, the literals $\False$ and $\True$ are subtypes of
-$\Boolean$, any literal number $n$ is subtype of $\Number$, and any
-literal string $s$ is subtype of $\String$.
-The type $\Any$ is the dynamic type that allows dynamic code coexist
-with static code.
-We will discuss union types, function types, table types, recursive
-types, and second-class types along this section.
+\subsection{Type Language}
+
+Lua values can have one of eight tags:
+\emph{nil}, \emph{boolean}, \emph{number}, \emph{string},
+\emph{function}, \emph{table}, \emph{userdata}, and \emph{thread}.
+In this section we present the types that Typed Lua introduces
+to represent these tags.
 
@@
 \begin{array}{rlr}
-T ::= & & \emph{first-class types}\\
-& \;\; L & \emph{literal types}\\
-& | \; B & \emph{base types}\\
-& | \; \Any & \emph{dynamic type}\\
-& | \; T \cup T & \emph{union type}\\
-& | \; S \rightarrow S & \emph{function type}\\
-& | \; \{T:T, ..., T:T\} & \emph{table type}\\
-& | \; X & \emph{type variable}\\
-& | \; \mu X.T & \emph{recursive type}\\
-L ::= & \Nil \; | \; \False \; | \; \True \; | \;
-  {<}{\it number}{>} \; | \; {<}{\it string}{>} & \\
+\multicolumn{3}{c}{\textbf{First-level types}}\\
+T ::= & \;\; L & \textit{literal types}\\
+& | \; B & \textit{base types}\\
+& | \; \Nil & \textit{nil type}\\
+& | \; \Value & \textit{top type}\\
+& | \; \Any & \textit{dynamic type}\\
+& | \; \Self & \textit{self type}\\
+& | \; T \cup T & \textit{disjoint union types}\\
+& | \; P \rightarrow R & \textit{function types}\\
+& | \; \{ \}_{o|c} & \textit{empty table type}\\
+& | \; \{F, ..., F\}_{o|c} & \textit{table types}\\
+& | \; X & \textit{type variables}\\
+& | \; \mu X.T & \textit{recursive types}\\
+& | \; X_{i} & \textit{projection types}\\
+L ::= & \False \; | \; \True \; | \; {<}{\it number}{>} \; | \; {<}{\it string}{>} & \\
 B ::= & \Boolean \; | \; \Number \; | \; \String & \\
-S ::= & \Void \; | \; V & \emph{second-class types}\\
-V ::= & \;\; T & \\
-& | \; T* & \emph{vararg type}\\
-& | \; T \times V & \emph{tuple type}
+F ::= & K:T \; | \; \Const \; K:T & \textit{field types}\\ 
+K ::= & L \; | \; B \; | \; \Value \; | \; \Any & \textit{key types}\\
+\multicolumn{3}{c}{}\\
+\multicolumn{3}{c}{\textbf{Second-level types}}\\
+S ::= & \;\; T &\\
+& | \; T* & \textit{vararg types}\\
+& | \; T \times S & \textit{tuple types}\\
+P ::= & \Void \; | \; S & \textit{parameters list type}\\
+R ::= & \Void \; | \; S \; | \; S \cup R & \textit{return types}\\
 \end{array}
@@
 \end{figure}
+
+Figure \ref{fig:typelang} presents the abstract syntax of the
+Typed Lua types.
+We split Typed Lua types into two categories:
+\emph{first-level types} and \emph{second-level types}.
+We use first-level types to represent Lua values, and
+second-level types to type expression lists, multiple assignments,
+and function applications.
+We will all Typed Lua types along this section.
+
+\subsection{Subtyping and Consistent-Subtyping}
+
+\subsection{Atomic types and Functions}
 
@@
 
+\subsection{Unions}
+
 Typed Lua includes union types as a way to discriminate data
@@
 
+\subsection{Modules}
+
+\subsection{Objects and Classes}
+
 \section{Related Lua projects}

### 2014-07-28 doc/thesis/typedlua.tex
Just moving the subtyping rules to the thesis files without any formatting

@@
 \emph{first-level types} and \emph{second-level types}.
-We use first-level types to represent Lua values, and
+We use first-level types to represent Lua values and
 second-level types to type expression lists, multiple assignments,
 and function applications.
-We will all Typed Lua types along this section.
+We will explain all Typed Lua types in more detail along this section.
 
 \subsection{Subtyping and Consistent-Subtyping}
+
+\[
+\senv \vdash L \subtype L
+\mylabel{S-LITERAL}
+\]
+
+\[
+\senv \vdash \False \subtype \Boolean
+\mylabel{S-FALSE}
+\]
+
+\[
+\senv \vdash \True \subtype \Boolean
+\mylabel{S-TRUE}
+\]
+
+\[
+\senv \vdash {<}{\it number}{>} \subtype \Number
+\mylabel{S-NUMBER}
+\]
+
+\[
+\senv \vdash {<}{\it string}{>} \subtype \String
+\mylabel{S-STRING}
+\]
+
+\[
+\senv \vdash B \subtype B
+\mylabel{S-BASE}
+\]
+
+\[
+\senv \vdash \Nil \subtype \Nil
+\mylabel{S-NIL}
+\]
+
+\[
+\senv \vdash T \subtype \Value
+\mylabel{S-VALUE}
+\]
+
+\[
+\senv \vdash \Any \subtype \Any
+\mylabel{S-ANY}
+\]
+
+\[
+\senv \vdash \Self \subtype \Self
+\mylabel{S-SELF}
+\]
+
+\[
+\frac{\senv \vdash T_{1} \subtype T \;\;\;
+      \senv \vdash T_{2} \subtype T}
+     {\senv \vdash T_{1} \cup T_{2} \subtype T}
+\mylabel{S-UNION1}
+\]
+
+\[
+\frac{\senv \vdash T \subtype T_{1}}
+     {\senv \vdash T \subtype T_{1} \cup T_{2}}
+\mylabel{S-UNION2}
+\]
+
+\[
+\frac{\senv \vdash T \subtype T_{2}}
+     {\senv \vdash T \subtype T_{1} \cup T_{2}}
+\mylabel{S-UNION3}
+\]
+
+\[
+\frac{\senv \vdash P_{2} \subtype P_{1} \;\;\;
+      \senv \vdash R_{1} \subtype R_{2}}
+     {\senv \vdash P_{1} \rightarrow R_{1} \subtype P_{2} \rightarrow R_{2}}
+\mylabel{S-FUNCTION}
+\]
+
+\[
+\frac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
+      \senv \vdash F_{j} \subtype F_{i}}
+     {\senv \vdash \{F_{1}, ..., F_{m}\}_{c} \subtype \{F_{1}, ..., F_{n}\}}
+\mylabel{S-TABLE1}
+\]
+
+\[
+\frac{\forall i \in 1..m \; \senv \vdash F_{i}' \subtype F_{i}}
+     {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \subtype \{F_{1}', ..., F_{m}', ..., F_{n}'\}}
+\mylabel{S-TABLE2}
+\]
+
+\[
+\frac{\forall i \in 1..m \; \senv \vdash F \subtype F_{i}}
+     {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \subtype \{F\}} 
+\mylabel{S-TABLE3}
+\]
+
+\[
+\frac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
+      \senv \vdash T_{1} \subtype T_{2} \;\;\;
+      \senv \vdash T_{2} \subtype T_{1}}
+     {\senv \vdash K_{1}:T_{1} \subtype K_{2}:T_{2}}
+\mylabel{S-FIELD1}
+\]
+
+\[
+\frac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
+      \senv \vdash T_{1} \subtype T_{2}}
+     {\senv \vdash \Const \; K_{1}:T_{1} \subtype \Const \; K_{2}:T_{2}}
+\mylabel{S-FIELD2}
+\]
+
+\[
+\frac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
+      \senv \vdash T_{1} \subtype T_{2}}
+     {\senv \vdash K_{1}:T_{1} \subtype \Const \; K_{2}:T_{2}}
+\mylabel{S-FIELD3}
+\]
+
+\[
+\frac{X_{1} \subtype X_{2} \in \senv}
+     {\senv \vdash X_{1} \subtype X_{2}}
+\mylabel{S-VARIABLE}
+\]
+
+\[
+\frac{\senv, X_{1} \subtype X_{2} \vdash T_{1} \subtype T_{2}}
+     {\senv \vdash \mu X_{1}.T_{1} \subtype \mu X_{2}.T_{2}}
+\mylabel{S-RECURSIVE1}
+\]
+
+\[
+\frac{X \subtype X \not\in \senv \;\;\;
+      \senv, X \subtype X \vdash T_{1} \subtype T_{2}}
+     {\senv \vdash \mu X.T_{1} \subtype T_{2}}
+\mylabel{S-RECURSIVE2}
+\]
+
+\[
+\frac{X \subtype X \in \senv \;\;\;
+      \senv \vdash X \subtype T_{2}}
+     {\senv \vdash \mu X.T_{1} \subtype T_{2}}
+\mylabel{S-RECURSIVE3}
+\]
+
+\[
+\frac{X \subtype X \not\in \senv \;\;\;
+      \senv, X \subtype X \vdash T_{1} \subtype T_{2}}
+     {\senv \vdash T_{1} \subtype \mu X.T_{2}}
+\mylabel{S-RECURSIVE4}
+\]
+
+\[
+\frac{X \subtype X \in \senv \;\;\;
+      \senv \vdash T_{1} \subtype X}
+     {\senv \vdash T_{1} \subtype \mu X.T_{2}}
+\mylabel{S-RECURSIVE5}
+\]
+
+\[
+\frac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \cup \Nil}
+     {\senv \vdash T_{1}* \subtype T_{2}*}
+\mylabel{S-VARARG}
+\]
+
+\[
+\frac{\senv \vdash T_{1} \subtype T_{2} \;\;\;
+      \senv \vdash S_{1} \subtype S_{2}}
+     {\senv \vdash T_{1} \times S_{1} \subtype T_{2} \times S_{2}}
+\mylabel{S-TUPLE1}
+\]
+
+\[
+\frac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \;\;\;
+      \senv \vdash T_{1}* \subtype S_{2}}
+     {\senv \vdash T_{1}* \subtype T_{2} \times S_{2}}
+\mylabel{S-TUPLE2}
+\]
+
+\[
+\frac{\senv \vdash T_{1} \subtype T_{2} \cup \Nil \;\;\;
+      \senv \vdash S_{1} \subtype T_{2}*}
+     {\senv \vdash T_{1} \times S_{1} \subtype T_{2}*}
+\mylabel{S-TUPLE3}
+\]
+
+\[
+\senv \vdash \Void \subtype \Void
+\mylabel{S-VOID}
+\]
+
+The consistent-subtyping rules are exactly the same as the subtyping rules,
+except those that handle the dynamic type:
+\[
+\senv \vdash T \lesssim \Any
+\mylabel{C-ANY1}
+\]
+
+\[
+\senv \vdash \Any \lesssim T
+\mylabel{C-ANY2}
+\]
 

### 2014-08-05 doc/thesis/typedlua.tex
Formatting the subtyping and consistent-subtyping rules

--- 2014-07-26 doc/thesis/typedlua.tex
@@
 \section{Statistics about the usage of Lua}
+\label{sec:statistics}
 
@@
 
-\section{Typed Lua}
-
-The main goal of Typed Lua is to allow programmers to combine static
-and dynamic typing in Lua, so Typed Lua is an extension to Lua 5.2
-syntax that introduces optional type annotations, and class-based
-object-oriented programming through the definition of classes,
-interfaces, and modules.
-An important feature of Typed Lua is its backwards compatibility with
-Lua, that is, unannotated Typed Lua code is valid Lua code.
-For this reason, the syntactic extensions introduced by Typed Lua
-include new keywords that are not reserved words.
-Appendix \ref{app:syntax} presents the complete syntax of Typed Lua.
-
-\subsection{Type Language}
+\section{Type Language}
+\label{sec:types}
 
@@
 \emph{first-level types} and \emph{second-level types}.
-We use first-level types to represent Lua values, and
+We use first-level types to represent Lua values and
 second-level types to type expression lists, multiple assignments,
 and function applications.
-We will all Typed Lua types along this section.
+We will explain all Typed Lua types in more detail along this section.
 
-\subsection{Subtyping and Consistent-Subtyping}
+\section{Typed Lua through examples}
+\label{sec:examples}
+
+The main goal of Typed Lua is to allow programmers to combine static
+and dynamic typing in Lua, so Typed Lua is an extension to Lua 5.2
+syntax that introduces optional type annotations, and class-based
+object-oriented programming through the definition of classes,
+interfaces, and modules.
+An important feature of Typed Lua is its backwards compatibility with
+Lua, that is, unannotated Typed Lua code is valid Lua code.
+For this reason, the syntactic extensions introduced by Typed Lua
+include new keywords that are not reserved words.
+Appendix \ref{app:syntax} presents the complete syntax of Typed Lua.
 
@@
 
+\section{The design of Typed Lua}
+\label{sec:design}
+
+\subsection{Subtyping and Consistent-Subtyping}
+
+\begin{figure}[!ht]
+\textbf{Subtyping}\\
+\dstart
+\begin{footnotesize}
+$$
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-LITERAL}\\
+\senv \vdash L \subtype L
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FALSE}\\
+\senv \vdash \False \subtype \Boolean
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-TRUE}\\
+\senv \vdash \True \subtype \Boolean
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-NUMBER}\\
+\senv \vdash {<}{\it number}{>} \subtype \Number
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-STRING}\\
+\senv \vdash {<}{\it string}{>} \subtype \String
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-BASE}\\
+\senv \vdash B \subtype B
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-NIL}\\
+\senv \vdash \Nil \subtype \Nil
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-VALUE}\\
+\senv \vdash T \subtype \Value
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-ANY}\\
+\senv \vdash \Any \subtype \Any
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-SELF}\\
+\senv \vdash \Self \subtype \Self
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-UNION1}\\
+\dfrac{\senv \vdash T_{1} \subtype T \;\;\;
+       \senv \vdash T_{2} \subtype T}
+      {\senv \vdash T_{1} \cup T_{2} \subtype T}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-UNION2}\\
+\dfrac{\senv \vdash T \subtype T_{1}}
+      {\senv \vdash T \subtype T_{1} \cup T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-UNION3}\\
+\dfrac{\senv \vdash T \subtype T_{2}}
+      {\senv \vdash T \subtype T_{1} \cup T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-FUNCTION}\\
+\dfrac{\senv \vdash P_{2} \subtype P_{1} \;\;\;
+       \senv \vdash R_{1} \subtype R_{2}}
+      {\senv \vdash P_{1} \rightarrow R_{1} \subtype P_{2} \rightarrow R_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-TABLE1}\\
+\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
+       \senv \vdash F_{j} \subtype F_{i}}
+      {\senv \vdash \{F_{1}, ..., F_{m}\}_{c} \subtype \{F_{1}, ..., F_{n}\}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-TABLE2}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash F_{i}' \subtype F_{i}}
+      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \subtype \{F_{1}', ..., F_{m}', ..., F_{n}'\}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-TABLE3}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash F \subtype F_{i}}
+      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \subtype \{F\}} 
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-FIELD1}\\
+\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
+       \senv \vdash T_{1} \subtype T_{2} \;\;\;
+       \senv \vdash T_{2} \subtype T_{1}}
+      {\senv \vdash K_{1}:T_{1} \subtype K_{2}:T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FIELD2}\\
+\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
+       \senv \vdash T_{1} \subtype T_{2}}
+      {\senv \vdash \Const \; K_{1}:T_{1} \subtype \Const \; K_{2}:T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-FIELD3}\\
+\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
+       \senv \vdash T_{1} \subtype T_{2}}
+      {\senv \vdash K_{1}:T_{1} \subtype \Const \; K_{2}:T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-VARIABLE}\\
+\dfrac{X_{1} \subtype X_{2} \in \senv}
+      {\senv \vdash X_{1} \subtype X_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-RECURSIVE1}\\
+\dfrac{\senv, X_{1} \subtype X_{2} \vdash T_{1} \subtype T_{2}}
+      {\senv \vdash \mu X_{1}.T_{1} \subtype \mu X_{2}.T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-RECURSIVE2}\\
+\dfrac{X \subtype X \not\in \senv \;\;\;
+      \senv, X \subtype X \vdash T_{1} \subtype T_{2}}
+      {\senv \vdash \mu X.T_{1} \subtype T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-RECURSIVE3}\\
+\dfrac{X \subtype X \in \senv \;\;\;
+       \senv \vdash X \subtype T_{2}}
+      {\senv \vdash \mu X.T_{1} \subtype T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-RECURSIVE4}\\
+\dfrac{X \subtype X \not\in \senv \;\;\;
+       \senv, X \subtype X \vdash T_{1} \subtype T_{2}}
+      {\senv \vdash T_{1} \subtype \mu X.T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-RECURSIVE5}\\
+\dfrac{X \subtype X \in \senv \;\;\;
+       \senv \vdash T_{1} \subtype X}
+      {\senv \vdash T_{1} \subtype \mu X.T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-VOID}\\
+\senv \vdash \Void \subtype \Void
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-VARARG}\\
+\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \cup \Nil}
+      {\senv \vdash T_{1}* \subtype T_{2}*}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-TUPLE1}\\
+\dfrac{\senv \vdash T_{1} \subtype T_{2} \;\;\;
+       \senv \vdash S_{1} \subtype S_{2}}
+      {\senv \vdash T_{1} \times S_{1} \subtype T_{2} \times S_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-TUPLE2}\\
+\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \;\;\;
+       \senv \vdash T_{1}* \subtype S_{2}}
+      {\senv \vdash T_{1}* \subtype T_{2} \times S_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-TUPLE3}\\
+\dfrac{\senv \vdash T_{1} \subtype T_{2} \cup \Nil \;\;\;
+       \senv \vdash S_{1} \subtype T_{2}*}
+      {\senv \vdash T_{1} \times S_{1} \subtype T_{2}*}
+\end{array}
+\end{array}
+$$
+\end{footnotesize}
+\dend
+\caption{The subtyping relation}
+\label{fig:subtyping}
+\end{figure}
+
+The consistent-subtyping rules are exactly the same as the subtyping rules,
+except those that handle the dynamic type.
+
+\begin{figure}[!ht]
+\textbf{Consistent-Subtyping}\\
+\dstart
+\begin{footnotesize}
+$$
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{C-LITERAL}\\
+\senv \vdash L \lesssim L
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-FALSE}\\
+\senv \vdash \False \lesssim \Boolean
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-TRUE}\\
+\senv \vdash \True \lesssim \Boolean
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-NUMBER}\\
+\senv \vdash {<}{\it number}{>} \lesssim \Number
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-STRING}\\
+\senv \vdash {<}{\it string}{>} \lesssim \String
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-BASE}\\
+\senv \vdash B \lesssim B
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-NIL}\\
+\senv \vdash \Nil \lesssim \Nil
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-VALUE}\\
+\senv \vdash T \lesssim \Value
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-ANY1}\\
+\senv \vdash T \lesssim \Any
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-ANY2}\\
+\senv \vdash \Any \lesssim T
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-SELF}\\
+\senv \vdash \Self \lesssim \Self
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-UNION1}\\
+\dfrac{\senv \vdash T_{1} \lesssim T \;\;\;
+       \senv \vdash T_{2} \lesssim T}
+      {\senv \vdash T_{1} \cup T_{2} \lesssim T}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-UNION2}\\
+\dfrac{\senv \vdash T \lesssim T_{1}}
+      {\senv \vdash T \lesssim T_{1} \cup T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-UNION3}\\
+\dfrac{\senv \vdash T \lesssim T_{2}}
+      {\senv \vdash T \lesssim T_{1} \cup T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-FUNCTION}\\
+\dfrac{\senv \vdash P_{2} \lesssim P_{1} \;\;\;
+       \senv \vdash R_{1} \lesssim R_{2}}
+      {\senv \vdash P_{1} \rightarrow R_{1} \lesssim P_{2} \rightarrow R_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-TABLE1}\\
+\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
+       \senv \vdash F_{j} \lesssim F_{i}}
+      {\senv \vdash \{F_{1}, ..., F_{m}\}_{c} \lesssim \{F_{1}, ..., F_{n}\}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-TABLE2}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash F_{i}' \lesssim F_{i}}
+      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \lesssim \{F_{1}', ..., F_{m}', ..., F_{n}'\}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-TABLE3}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash F \lesssim F_{i}}
+      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \lesssim \{F\}} 
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-FIELD1}\\
+\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
+       \senv \vdash T_{1} \lesssim T_{2} \;\;\;
+       \senv \vdash T_{2} \lesssim T_{1}}
+      {\senv \vdash K_{1}:T_{1} \lesssim K_{2}:T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-FIELD2}\\
+\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
+       \senv \vdash T_{1} \lesssim T_{2}}
+      {\senv \vdash \Const \; K_{1}:T_{1} \lesssim \Const \; K_{2}:T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-FIELD3}\\
+\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
+       \senv \vdash T_{1} \lesssim T_{2}}
+      {\senv \vdash K_{1}:T_{1} \lesssim \Const \; K_{2}:T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-VARIABLE}\\
+\dfrac{X_{1} \lesssim X_{2} \in \senv}
+      {\senv \vdash X_{1} \lesssim X_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-RECURSIVE1}\\
+\dfrac{\senv, X_{1} \lesssim X_{2} \vdash T_{1} \lesssim T_{2}}
+      {\senv \vdash \mu X_{1}.T_{1} \lesssim \mu X_{2}.T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-RECURSIVE2}\\
+\dfrac{X \lesssim X \not\in \senv \;\;\;
+      \senv, X \lesssim X \vdash T_{1} \lesssim T_{2}}
+      {\senv \vdash \mu X.T_{1} \lesssim T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-RECURSIVE3}\\
+\dfrac{X \lesssim X \in \senv \;\;\;
+       \senv \vdash X \lesssim T_{2}}
+      {\senv \vdash \mu X.T_{1} \lesssim T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-RECURSIVE4}\\
+\dfrac{X \lesssim X \not\in \senv \;\;\;
+       \senv, X \lesssim X \vdash T_{1} \lesssim T_{2}}
+      {\senv \vdash T_{1} \lesssim \mu X.T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-RECURSIVE5}\\
+\dfrac{X \lesssim X \in \senv \;\;\;
+       \senv \vdash T_{1} \lesssim X}
+      {\senv \vdash T_{1} \lesssim \mu X.T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-VOID}\\
+\senv \vdash \Void \lesssim \Void
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-VARARG}\\
+\dfrac{\senv \vdash T_{1} \cup \Nil \lesssim T_{2} \cup \Nil}
+      {\senv \vdash T_{1}* \lesssim T_{2}*}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-TUPLE1}\\
+\dfrac{\senv \vdash T_{1} \lesssim T_{2} \;\;\;
+       \senv \vdash S_{1} \lesssim S_{2}}
+      {\senv \vdash T_{1} \times S_{1} \lesssim T_{2} \times S_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-TUPLE2}\\
+\dfrac{\senv \vdash T_{1} \cup \Nil \lesssim T_{2} \;\;\;
+       \senv \vdash T_{1}* \lesssim S_{2}}
+      {\senv \vdash T_{1}* \lesssim T_{2} \times S_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-TUPLE3}\\
+\dfrac{\senv \vdash T_{1} \lesssim T_{2} \cup \Nil \;\;\;
+       \senv \vdash S_{1} \lesssim T_{2}*}
+      {\senv \vdash T_{1} \times S_{1} \lesssim T_{2}*}
+\end{array}
+\end{array}
+$$
+\end{footnotesize}
+\dend
+\caption{The consistent-subtyping relation}
+\label{fig:consistent_subtyping}
+\end{figure}
+
 \section{Related Lua projects}
+\label{sec:related}
 

### 2014-08-06 doc/thesis/cases.tex
Starting the rewriting of the Typed Lua chapter
Fixing _ENV to include setmetatable, type, and require
Trying another formatting style to the section about Typed Lua and sketching the file userdata

@@
 \begin{verbatim}
-(string) -> ({any:any})
+(string) -> (any)
 \end{verbatim}
@@
 Perhaps due to the fact that the parameter \textbf{pos} is optional.
-If \textbf{value} is $\TopStar$, the function inserts \textbf{pos}
+If \textbf{value} is $\Value{*}$, the function inserts \textbf{pos}
 in the \textbf{list}, that is, the input type is
@@
 
+\begin{verbatim}
+userdata file
+  close : (self) -> (boolean) | (nil, string) | (true?, string, number),
+  flush : (self) -> (boolean),
+  lines : (self, string|number*) -> (() -> (any*)),
+  read : (self, string|number*) -> (string*),
+  seek : (self, string?, number?) -> (number?),
+  setvbuf : (self, string, number?) -> (true) | (nil, string, number),
+  write : (self, string|number*) -> (self?)
+end
+\end{verbatim}
+
 \begin{enumerate}
@@
 \begin{verbatim}
-(any?) -> (boolean) | (nil, string) | (true?, string, number)
+(file?) -> (boolean) | (nil, string) | (true?, string, number)
 \end{verbatim}
@@
 was created with \textbf{io.popen}.
-The parameter \textbf{file} is a \textit{userdata} and I need
-to fix that on the type system.
 \item io.flush () :
@@
 \begin{verbatim}
-(string|any|nil) -> (any)
+(string|file|nil) -> (file)
 \end{verbatim}
 Category 1?
-Here the type \texttt{any} refers to a \textit{userdata} and
-I need to fix that on the type system.
 \item io.lines ([filename ···]) :
 \begin{verbatim}
-(string|any|nil) -> ((any*) -> (any*))
+(string|file|nil) -> ((any*) -> (any*))
 \end{verbatim}
@@
 In fact, this function returns an iterator. Need to check the return type.
-Here the type \texttt{any} refers to a \textit{userdata} and
-I need to fix that on the type system.
 \item io.open (filename [, mode]) :
@@
 "r"|"w"|"a"|"r+"|"w+"|"a+"|"rb"|"wb"|"ab"|"r+b"|"w+b"|"w+b"|nil) ->
-(any) | (nil, string, number)
+(file) | (nil, string, number)
 \end{verbatim}
 Category 1?
-\\
-Here the type \texttt{any} refers to a \textit{userdata} and
-I need to fix that on the type system.
 \item io.output ([file]) :
 \begin{verbatim}
-(string|any|nil) -> (any)
+(string|file|nil) -> (file)
 \end{verbatim}
 Category 1?
-\\
-Here the type \texttt{any} refers to a \textit{userdata} and
-I need to fix that on the type system.
 \item io.popen (prog [, mode]) :
 \begin{verbatim}
-(string, "r"|"w"|nil) -> (any) | (nil, string, number)
+(string, "r"|"w"|nil) -> (file) | (nil, string, number)
 \end{verbatim}
 Category 1?
-\\
-Here the type \texttt{any} refers to a \textit{userdata} and
-I need to fix that on the type system.
 \item io.read (···) :
@@
 \begin{verbatim}
-() -> (any)
+() -> (file)
 \end{verbatim}
 Category 1?
-\\
-Here the type \texttt{any} refers to a \textit{userdata} and
-I need to fix that on the type system.
 \item io.type (obj) :
@@
 \begin{verbatim}
-(string|number*) -> (any) | (nil, string, number)
+(string|number*) -> (file) | (nil, string, number)
 \end{verbatim}
 Category 1?
-\\
-Here the type \texttt{any} refers to a \textit{userdata} and
-I need to fix that on the type system.
-\item file:close () : ?
-\item file:flush () : ?
-\item file:lines (···) : ?
-\item file:read (···) : ?
-\item file:seek ([whence [, offset]]) : ?
-\item file:setvbuf (mode [, size]) : ?
-\item file:write (···) : ?
 \end{enumerate}

### 2014-08-06 doc/thesis/typedlua.tex
Trying another formatting style to the section about Typed Lua and sketching the file userdata

@@
 
-\section{Type Language}
+\section{Typed Lua}
+\label{sec:typedlua}
+
+Typed Lua allows statically typed Lua code to coexist and interact
+with dynamically typed Lua code.
+The Typed Lua compiler checks for type errors, but alwayas removes
+the annotations, generating Lua code that runs in unmodified Lua
+implementations.
+The programmer can enjoy some of the benefits of static types even
+without converting existing Lua modules to Typed Lua:
+a dynamically typed module can export a statically typed interface,
+and statically typed users of the module will have their use of the
+module checked by the compiler.
+
+Unlike gradual typing, Typed Lua does not insert run-time checks
+between dynamically typed and statically typed parts of the program.
+Unlike some optional type systems, the statically typed subset of
+Typed Lua aims to be sound, so a later version of the Typed Lua
+compiler can switch to gradual typing instead of just optional typing.
+
+In this section we cover the design of Typed Lua:
+In Section \ref{sec:types} we present the types that are available
+in Typed Lua along with the subtyping and consistent-subtyping
+relations.
+In Section \ref{sec:examples} we present the design of Typed Lua
+through code examples.
+In Section \ref{sec:rules} we present the design of Typed Lua
+through typing rules.
+
+\subsection{Type Language}
 \label{sec:types}
@@
 and function applications.
-We will explain all Typed Lua types in more detail along this section.
-
-\section{Typed Lua through examples}
+
+Figure \ref{fig:subtyping} presents the subtyping relation among
+Typed Lua types.
+Any first-level type is a subtype of $\Value$.
+The rest of the subtyping relationship is standard:
+union types are supertypes of their respective literal types,
+function types are related by contravariance on the input
+and covariance on the output, table types have width subtyping,
+with depth subtyping on $\Const$ fields, tuple and vararg types
+are covariant.
+
+\begin{figure}[!ht]
+\textbf{Subtyping}\\
+\dstart
+\begin{footnotesize}
+$$
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-LITERAL}\\
+\senv \vdash L \subtype L
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FALSE}\\
+\senv \vdash \False \subtype \Boolean
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-TRUE}\\
+\senv \vdash \True \subtype \Boolean
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-NUMBER}\\
+\senv \vdash {<}{\it number}{>} \subtype \Number
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-STRING}\\
+\senv \vdash {<}{\it string}{>} \subtype \String
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-BASE}\\
+\senv \vdash B \subtype B
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-NIL}\\
+\senv \vdash \Nil \subtype \Nil
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-VALUE}\\
+\senv \vdash T \subtype \Value
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-ANY}\\
+\senv \vdash \Any \subtype \Any
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-SELF}\\
+\senv \vdash \Self \subtype \Self
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-UNION1}\\
+\dfrac{\senv \vdash T_{1} \subtype T \;\;\;
+       \senv \vdash T_{2} \subtype T}
+      {\senv \vdash T_{1} \cup T_{2} \subtype T}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-UNION2}\\
+\dfrac{\senv \vdash T \subtype T_{1}}
+      {\senv \vdash T \subtype T_{1} \cup T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-UNION3}\\
+\dfrac{\senv \vdash T \subtype T_{2}}
+      {\senv \vdash T \subtype T_{1} \cup T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-FUNCTION}\\
+\dfrac{\senv \vdash P_{2} \subtype P_{1} \;\;\;
+       \senv \vdash R_{1} \subtype R_{2}}
+      {\senv \vdash P_{1} \rightarrow R_{1} \subtype P_{2} \rightarrow R_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-TABLE1}\\
+\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
+       \senv \vdash F_{j} \subtype F_{i}}
+      {\senv \vdash \{F_{1}, ..., F_{m}\}_{c} \subtype \{F_{1}, ..., F_{n}\}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-TABLE2}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash F_{i}' \subtype F_{i}}
+      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \subtype \{F_{1}', ..., F_{m}', ..., F_{n}'\}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-TABLE3}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash F \subtype F_{i}}
+      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \subtype \{F\}} 
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-FIELD1}\\
+\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
+       \senv \vdash T_{1} \subtype T_{2} \;\;\;
+       \senv \vdash T_{2} \subtype T_{1}}
+      {\senv \vdash K_{1}:T_{1} \subtype K_{2}:T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FIELD2}\\
+\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
+       \senv \vdash T_{1} \subtype T_{2}}
+      {\senv \vdash \Const \; K_{1}:T_{1} \subtype \Const \; K_{2}:T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-FIELD3}\\
+\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
+       \senv \vdash T_{1} \subtype T_{2}}
+      {\senv \vdash K_{1}:T_{1} \subtype \Const \; K_{2}:T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-VARIABLE}\\
+\dfrac{X_{1} \subtype X_{2} \in \senv}
+      {\senv \vdash X_{1} \subtype X_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-RECURSIVE1}\\
+\dfrac{\senv, X_{1} \subtype X_{2} \vdash T_{1} \subtype T_{2}}
+      {\senv \vdash \mu X_{1}.T_{1} \subtype \mu X_{2}.T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-RECURSIVE2}\\
+\dfrac{X \subtype X \not\in \senv \;\;\;
+      \senv, X \subtype X \vdash T_{1} \subtype T_{2}}
+      {\senv \vdash \mu X.T_{1} \subtype T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-RECURSIVE3}\\
+\dfrac{X \subtype X \in \senv \;\;\;
+       \senv \vdash X \subtype T_{2}}
+      {\senv \vdash \mu X.T_{1} \subtype T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-RECURSIVE4}\\
+\dfrac{X \subtype X \not\in \senv \;\;\;
+       \senv, X \subtype X \vdash T_{1} \subtype T_{2}}
+      {\senv \vdash T_{1} \subtype \mu X.T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-RECURSIVE5}\\
+\dfrac{X \subtype X \in \senv \;\;\;
+       \senv \vdash T_{1} \subtype X}
+      {\senv \vdash T_{1} \subtype \mu X.T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-VOID}\\
+\senv \vdash \Void \subtype \Void
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-VARARG}\\
+\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \cup \Nil}
+      {\senv \vdash T_{1}* \subtype T_{2}*}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-TUPLE1}\\
+\dfrac{\senv \vdash T_{1} \subtype T_{2} \;\;\;
+       \senv \vdash S_{1} \subtype S_{2}}
+      {\senv \vdash T_{1} \times S_{1} \subtype T_{2} \times S_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-TUPLE2}\\
+\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \;\;\;
+       \senv \vdash T_{1}* \subtype S_{2}}
+      {\senv \vdash T_{1}* \subtype T_{2} \times S_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-TUPLE3}\\
+\dfrac{\senv \vdash T_{1} \subtype T_{2} \cup \Nil \;\;\;
+       \senv \vdash S_{1} \subtype T_{2}*}
+      {\senv \vdash T_{1} \times S_{1} \subtype T_{2}*}
+\end{array}
+\end{array}
+$$
+\end{footnotesize}
+\dend
+\caption{The subtyping relation}
+\label{fig:subtyping}
+\end{figure}
+
+Figure \ref{fig:consistent_subtyping} presents the consistent-subtyping
+relation that allows the dynamic type $\Any$ interoperate between
+dynamically typed and statically typed code.
+The dynamic type $\Any$ is a subtype of $\Value$, but neither a
+supertype nor a subtype of any other type.
+Our consistent-subtyping relationship follows the standards defined
+by the gradual typing of objects \citep{siek2007objects,siek2013mutable}.
+In practice, we can pass a value of the dynamic type anytime we want
+a value of some other type, and can pass any value where where a
+value of the dynamic type is expected, but these operations are tracked
+by the type system, and the programmer can choose to be warned about them.
+
+\begin{figure}[!ht]
+\textbf{Consistent-Subtyping}\\
+\dstart
+\begin{footnotesize}
+$$
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{C-LITERAL}\\
+\senv \vdash L \lesssim L
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-FALSE}\\
+\senv \vdash \False \lesssim \Boolean
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-TRUE}\\
+\senv \vdash \True \lesssim \Boolean
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-NUMBER}\\
+\senv \vdash {<}{\it number}{>} \lesssim \Number
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-STRING}\\
+\senv \vdash {<}{\it string}{>} \lesssim \String
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-BASE}\\
+\senv \vdash B \lesssim B
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-NIL}\\
+\senv \vdash \Nil \lesssim \Nil
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-VALUE}\\
+\senv \vdash T \lesssim \Value
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-ANY1}\\
+\senv \vdash T \lesssim \Any
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-ANY2}\\
+\senv \vdash \Any \lesssim T
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-SELF}\\
+\senv \vdash \Self \lesssim \Self
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-UNION1}\\
+\dfrac{\senv \vdash T_{1} \lesssim T \;\;\;
+       \senv \vdash T_{2} \lesssim T}
+      {\senv \vdash T_{1} \cup T_{2} \lesssim T}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-UNION2}\\
+\dfrac{\senv \vdash T \lesssim T_{1}}
+      {\senv \vdash T \lesssim T_{1} \cup T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-UNION3}\\
+\dfrac{\senv \vdash T \lesssim T_{2}}
+      {\senv \vdash T \lesssim T_{1} \cup T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-FUNCTION}\\
+\dfrac{\senv \vdash P_{2} \lesssim P_{1} \;\;\;
+       \senv \vdash R_{1} \lesssim R_{2}}
+      {\senv \vdash P_{1} \rightarrow R_{1} \lesssim P_{2} \rightarrow R_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-TABLE1}\\
+\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
+       \senv \vdash F_{j} \lesssim F_{i}}
+      {\senv \vdash \{F_{1}, ..., F_{m}\}_{c} \lesssim \{F_{1}, ..., F_{n}\}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-TABLE2}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash F_{i}' \lesssim F_{i}}
+      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \lesssim \{F_{1}', ..., F_{m}', ..., F_{n}'\}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-TABLE3}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash F \lesssim F_{i}}
+      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \lesssim \{F\}} 
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-FIELD1}\\
+\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
+       \senv \vdash T_{1} \lesssim T_{2} \;\;\;
+       \senv \vdash T_{2} \lesssim T_{1}}
+      {\senv \vdash K_{1}:T_{1} \lesssim K_{2}:T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-FIELD2}\\
+\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
+       \senv \vdash T_{1} \lesssim T_{2}}
+      {\senv \vdash \Const \; K_{1}:T_{1} \lesssim \Const \; K_{2}:T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-FIELD3}\\
+\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
+       \senv \vdash T_{1} \lesssim T_{2}}
+      {\senv \vdash K_{1}:T_{1} \lesssim \Const \; K_{2}:T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-VARIABLE}\\
+\dfrac{X_{1} \lesssim X_{2} \in \senv}
+      {\senv \vdash X_{1} \lesssim X_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-RECURSIVE1}\\
+\dfrac{\senv, X_{1} \lesssim X_{2} \vdash T_{1} \lesssim T_{2}}
+      {\senv \vdash \mu X_{1}.T_{1} \lesssim \mu X_{2}.T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-RECURSIVE2}\\
+\dfrac{X \lesssim X \not\in \senv \;\;\;
+      \senv, X \lesssim X \vdash T_{1} \lesssim T_{2}}
+      {\senv \vdash \mu X.T_{1} \lesssim T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-RECURSIVE3}\\
+\dfrac{X \lesssim X \in \senv \;\;\;
+       \senv \vdash X \lesssim T_{2}}
+      {\senv \vdash \mu X.T_{1} \lesssim T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-RECURSIVE4}\\
+\dfrac{X \lesssim X \not\in \senv \;\;\;
+       \senv, X \lesssim X \vdash T_{1} \lesssim T_{2}}
+      {\senv \vdash T_{1} \lesssim \mu X.T_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-RECURSIVE5}\\
+\dfrac{X \lesssim X \in \senv \;\;\;
+       \senv \vdash T_{1} \lesssim X}
+      {\senv \vdash T_{1} \lesssim \mu X.T_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-VOID}\\
+\senv \vdash \Void \lesssim \Void
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-VARARG}\\
+\dfrac{\senv \vdash T_{1} \cup \Nil \lesssim T_{2} \cup \Nil}
+      {\senv \vdash T_{1}* \lesssim T_{2}*}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-TUPLE1}\\
+\dfrac{\senv \vdash T_{1} \lesssim T_{2} \;\;\;
+       \senv \vdash S_{1} \lesssim S_{2}}
+      {\senv \vdash T_{1} \times S_{1} \lesssim T_{2} \times S_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{C-TUPLE2}\\
+\dfrac{\senv \vdash T_{1} \cup \Nil \lesssim T_{2} \;\;\;
+       \senv \vdash T_{1}* \lesssim S_{2}}
+      {\senv \vdash T_{1}* \lesssim T_{2} \times S_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-TUPLE3}\\
+\dfrac{\senv \vdash T_{1} \lesssim T_{2} \cup \Nil \;\;\;
+       \senv \vdash S_{1} \lesssim T_{2}*}
+      {\senv \vdash T_{1} \times S_{1} \lesssim T_{2}*}
+\end{array}
+\end{array}
+$$
+\end{footnotesize}
+\dend
+\caption{The consistent-subtyping relation}
+\label{fig:consistent_subtyping}
+\end{figure}
+
+\subsection{Typed Lua through examples}
 \label{sec:examples}
@@
 include new keywords that are not reserved words.
-Appendix \ref{app:syntax} presents the complete syntax of Typed Lua.
+Appendix \ref{app:syntax} presents the complete syntax of Typed Lua,
+which we cover in this section through code examples.
 
@@
 
-\section{The design of Typed Lua}
-\label{sec:design}
-
-\subsection{Subtyping and Consistent-Subtyping}
-
-\begin{figure}[!ht]
-\textbf{Subtyping}\\
-\dstart
-\begin{footnotesize}
-$$
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-LITERAL}\\
-\senv \vdash L \subtype L
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FALSE}\\
-\senv \vdash \False \subtype \Boolean
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TRUE}\\
-\senv \vdash \True \subtype \Boolean
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-NUMBER}\\
-\senv \vdash {<}{\it number}{>} \subtype \Number
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-STRING}\\
-\senv \vdash {<}{\it string}{>} \subtype \String
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-BASE}\\
-\senv \vdash B \subtype B
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-NIL}\\
-\senv \vdash \Nil \subtype \Nil
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-VALUE}\\
-\senv \vdash T \subtype \Value
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-ANY}\\
-\senv \vdash \Any \subtype \Any
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-SELF}\\
-\senv \vdash \Self \subtype \Self
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-UNION1}\\
-\dfrac{\senv \vdash T_{1} \subtype T \;\;\;
-       \senv \vdash T_{2} \subtype T}
-      {\senv \vdash T_{1} \cup T_{2} \subtype T}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-UNION2}\\
-\dfrac{\senv \vdash T \subtype T_{1}}
-      {\senv \vdash T \subtype T_{1} \cup T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-UNION3}\\
-\dfrac{\senv \vdash T \subtype T_{2}}
-      {\senv \vdash T \subtype T_{1} \cup T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-FUNCTION}\\
-\dfrac{\senv \vdash P_{2} \subtype P_{1} \;\;\;
-       \senv \vdash R_{1} \subtype R_{2}}
-      {\senv \vdash P_{1} \rightarrow R_{1} \subtype P_{2} \rightarrow R_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TABLE1}\\
-\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
-       \senv \vdash F_{j} \subtype F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{c} \subtype \{F_{1}, ..., F_{n}\}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-TABLE2}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash F_{i}' \subtype F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \subtype \{F_{1}', ..., F_{m}', ..., F_{n}'\}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TABLE3}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash F \subtype F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \subtype \{F\}} 
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-FIELD1}\\
-\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
-       \senv \vdash T_{1} \subtype T_{2} \;\;\;
-       \senv \vdash T_{2} \subtype T_{1}}
-      {\senv \vdash K_{1}:T_{1} \subtype K_{2}:T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FIELD2}\\
-\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
-       \senv \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash \Const \; K_{1}:T_{1} \subtype \Const \; K_{2}:T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-FIELD3}\\
-\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
-       \senv \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash K_{1}:T_{1} \subtype \Const \; K_{2}:T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-VARIABLE}\\
-\dfrac{X_{1} \subtype X_{2} \in \senv}
-      {\senv \vdash X_{1} \subtype X_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-RECURSIVE1}\\
-\dfrac{\senv, X_{1} \subtype X_{2} \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash \mu X_{1}.T_{1} \subtype \mu X_{2}.T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-RECURSIVE2}\\
-\dfrac{X \subtype X \not\in \senv \;\;\;
-      \senv, X \subtype X \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash \mu X.T_{1} \subtype T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-RECURSIVE3}\\
-\dfrac{X \subtype X \in \senv \;\;\;
-       \senv \vdash X \subtype T_{2}}
-      {\senv \vdash \mu X.T_{1} \subtype T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-RECURSIVE4}\\
-\dfrac{X \subtype X \not\in \senv \;\;\;
-       \senv, X \subtype X \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash T_{1} \subtype \mu X.T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-RECURSIVE5}\\
-\dfrac{X \subtype X \in \senv \;\;\;
-       \senv \vdash T_{1} \subtype X}
-      {\senv \vdash T_{1} \subtype \mu X.T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-VOID}\\
-\senv \vdash \Void \subtype \Void
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-VARARG}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \cup \Nil}
-      {\senv \vdash T_{1}* \subtype T_{2}*}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TUPLE1}\\
-\dfrac{\senv \vdash T_{1} \subtype T_{2} \;\;\;
-       \senv \vdash S_{1} \subtype S_{2}}
-      {\senv \vdash T_{1} \times S_{1} \subtype T_{2} \times S_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-TUPLE2}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \;\;\;
-       \senv \vdash T_{1}* \subtype S_{2}}
-      {\senv \vdash T_{1}* \subtype T_{2} \times S_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TUPLE3}\\
-\dfrac{\senv \vdash T_{1} \subtype T_{2} \cup \Nil \;\;\;
-       \senv \vdash S_{1} \subtype T_{2}*}
-      {\senv \vdash T_{1} \times S_{1} \subtype T_{2}*}
-\end{array}
-\end{array}
-$$
-\end{footnotesize}
-\dend
-\caption{The subtyping relation}
-\label{fig:subtyping}
-\end{figure}
-
-The consistent-subtyping rules are exactly the same as the subtyping rules,
-except those that handle the dynamic type.
-
-\begin{figure}[!ht]
-\textbf{Consistent-Subtyping}\\
-\dstart
-\begin{footnotesize}
-$$
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{C-LITERAL}\\
-\senv \vdash L \lesssim L
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-FALSE}\\
-\senv \vdash \False \lesssim \Boolean
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TRUE}\\
-\senv \vdash \True \lesssim \Boolean
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-NUMBER}\\
-\senv \vdash {<}{\it number}{>} \lesssim \Number
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-STRING}\\
-\senv \vdash {<}{\it string}{>} \lesssim \String
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-BASE}\\
-\senv \vdash B \lesssim B
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-NIL}\\
-\senv \vdash \Nil \lesssim \Nil
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-VALUE}\\
-\senv \vdash T \lesssim \Value
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-ANY1}\\
-\senv \vdash T \lesssim \Any
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-ANY2}\\
-\senv \vdash \Any \lesssim T
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-SELF}\\
-\senv \vdash \Self \lesssim \Self
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-UNION1}\\
-\dfrac{\senv \vdash T_{1} \lesssim T \;\;\;
-       \senv \vdash T_{2} \lesssim T}
-      {\senv \vdash T_{1} \cup T_{2} \lesssim T}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-UNION2}\\
-\dfrac{\senv \vdash T \lesssim T_{1}}
-      {\senv \vdash T \lesssim T_{1} \cup T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-UNION3}\\
-\dfrac{\senv \vdash T \lesssim T_{2}}
-      {\senv \vdash T \lesssim T_{1} \cup T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-FUNCTION}\\
-\dfrac{\senv \vdash P_{2} \lesssim P_{1} \;\;\;
-       \senv \vdash R_{1} \lesssim R_{2}}
-      {\senv \vdash P_{1} \rightarrow R_{1} \lesssim P_{2} \rightarrow R_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TABLE1}\\
-\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
-       \senv \vdash F_{j} \lesssim F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{c} \lesssim \{F_{1}, ..., F_{n}\}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-TABLE2}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash F_{i}' \lesssim F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \lesssim \{F_{1}', ..., F_{m}', ..., F_{n}'\}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TABLE3}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash F \lesssim F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \lesssim \{F\}} 
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-FIELD1}\\
-\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
-       \senv \vdash T_{1} \lesssim T_{2} \;\;\;
-       \senv \vdash T_{2} \lesssim T_{1}}
-      {\senv \vdash K_{1}:T_{1} \lesssim K_{2}:T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-FIELD2}\\
-\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
-       \senv \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash \Const \; K_{1}:T_{1} \lesssim \Const \; K_{2}:T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-FIELD3}\\
-\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
-       \senv \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash K_{1}:T_{1} \lesssim \Const \; K_{2}:T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-VARIABLE}\\
-\dfrac{X_{1} \lesssim X_{2} \in \senv}
-      {\senv \vdash X_{1} \lesssim X_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-RECURSIVE1}\\
-\dfrac{\senv, X_{1} \lesssim X_{2} \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash \mu X_{1}.T_{1} \lesssim \mu X_{2}.T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-RECURSIVE2}\\
-\dfrac{X \lesssim X \not\in \senv \;\;\;
-      \senv, X \lesssim X \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash \mu X.T_{1} \lesssim T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-RECURSIVE3}\\
-\dfrac{X \lesssim X \in \senv \;\;\;
-       \senv \vdash X \lesssim T_{2}}
-      {\senv \vdash \mu X.T_{1} \lesssim T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-RECURSIVE4}\\
-\dfrac{X \lesssim X \not\in \senv \;\;\;
-       \senv, X \lesssim X \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash T_{1} \lesssim \mu X.T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-RECURSIVE5}\\
-\dfrac{X \lesssim X \in \senv \;\;\;
-       \senv \vdash T_{1} \lesssim X}
-      {\senv \vdash T_{1} \lesssim \mu X.T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-VOID}\\
-\senv \vdash \Void \lesssim \Void
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-VARARG}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \lesssim T_{2} \cup \Nil}
-      {\senv \vdash T_{1}* \lesssim T_{2}*}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TUPLE1}\\
-\dfrac{\senv \vdash T_{1} \lesssim T_{2} \;\;\;
-       \senv \vdash S_{1} \lesssim S_{2}}
-      {\senv \vdash T_{1} \times S_{1} \lesssim T_{2} \times S_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-TUPLE2}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \lesssim T_{2} \;\;\;
-       \senv \vdash T_{1}* \lesssim S_{2}}
-      {\senv \vdash T_{1}* \lesssim T_{2} \times S_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TUPLE3}\\
-\dfrac{\senv \vdash T_{1} \lesssim T_{2} \cup \Nil \;\;\;
-       \senv \vdash S_{1} \lesssim T_{2}*}
-      {\senv \vdash T_{1} \times S_{1} \lesssim T_{2}*}
-\end{array}
-\end{array}
-$$
-\end{footnotesize}
-\dend
-\caption{The consistent-subtyping relation}
-\label{fig:consistent_subtyping}
-\end{figure}
+\subsection{The typing rules of Typed Lua}
+\label{sec:rules}
 

### 2014-08-07 doc/thesis/typedlua.tex
Moving typing rules to the thesis and removing old documentation files
Rewriting the begining of the section about Typed Lua

@@
 
-Typed Lua allows statically typed Lua code to coexist and interact
-with dynamically typed Lua code.
-The Typed Lua compiler checks for type errors, but alwayas removes
-the annotations, generating Lua code that runs in unmodified Lua
-implementations.
-The programmer can enjoy some of the benefits of static types even
-without converting existing Lua modules to Typed Lua:
-a dynamically typed module can export a statically typed interface,
-and statically typed users of the module will have their use of the
-module checked by the compiler.
-
-Unlike gradual typing, Typed Lua does not insert run-time checks
-between dynamically typed and statically typed parts of the program.
-Unlike some optional type systems, the statically typed subset of
-Typed Lua aims to be sound, so a later version of the Typed Lua
-compiler can switch to gradual typing instead of just optional typing.
-
-In this section we cover the design of Typed Lua:
-In Section \ref{sec:types} we present the types that are available
-in Typed Lua along with the subtyping and consistent-subtyping
-relations.
-In Section \ref{sec:examples} we present the design of Typed Lua
-through code examples.
-In Section \ref{sec:rules} we present the design of Typed Lua
-through typing rules.
+Typed Lua is an optional type system for Lua, and its main goal is to
+allow programmers to combine static and dynamic typing in Lua.
+In practice, Typed Lua extends the syntax of Lua to introduce
+optional type annotations and class-based object-oriented
+programming through the definition of classes, interfaces, and
+modules.
+In addition, Typed Lua performs local type inference \citep{pierce2000lti}
+to detect more precise types for unannotated expressions.
+Even though the compiler warns the programmer about type errors,
+it always removes the type annotations to generate Lua code that
+runs in unmodified Lua implementations.
+
+An important feature of Typed Lua is its backwards compatibility with Lua,
+that is, unannotated Typed Lua code is valid Lua code.
+For this reason, the syntactic extensions introduced by Typed Lua
+include new keywords that are not reserved words.
+Appendix \ref{app:syntax} presents the complete syntax of Typed Lua.
+
+Another important feature of Typed Lua is its interface between
+statically and dynamically typed modules. 
+This means that the programmer can enjoy some of the benefits of
+static types even without converting existing Lua modules to Typed Lua,
+as a dynamically typed module can export a statically typed interface,
+and statically typed users of the module have their use of the module
+checked by the compiler.
+
+Although we use a gradual type system to formalize our optional type system,
+Typed Lua does not insert run-time checks in the gradual typing style.
+In gradual typing, run-time checks inspect the interaction between
+dynamically typed and statically typed code to guarantee that dynamically
+typed code does not violate statically typed code during run-time.
+A good reason for not inserting run-time checks at this moment is that they
+can decrease run-time performance \citep{allende2013cis}, and we believe that
+a careful evaluation of run-time checks should be done before inserting them
+on the type system.
+
+Unlike some optional type systems, we are designing Typed Lua aiming soundness
+to make it possible to switch Typed Lua from optional typing to gradual typing
+in the future, as a sound type system is the first step to make Typed Lua
+insert run-time checks after static type checking.
+
+In this section we show that Typed Lua is rich enough to preserve some Lua idioms.
+First, we present the abstract syntax of Typed Lua types along with the
+relationship among types.
+Then we use some code examples to present an overview about Typed Lua and
+its syntactical extensions.
+After that we present the formalization of Typed Lua.
 
@@
 
-The main goal of Typed Lua is to allow programmers to combine static
-and dynamic typing in Lua, so Typed Lua is an extension to Lua 5.2
-syntax that introduces optional type annotations, and class-based
-object-oriented programming through the definition of classes,
-interfaces, and modules.
-An important feature of Typed Lua is its backwards compatibility with
-Lua, that is, unannotated Typed Lua code is valid Lua code.
-For this reason, the syntactic extensions introduced by Typed Lua
-include new keywords that are not reserved words.
-Appendix \ref{app:syntax} presents the complete syntax of Typed Lua,
-which we cover in this section through code examples.
+In this section we use some examples of Typed Lua code to show how
+they relate to Lua.
+This examples shall give an informal overview of the type system.
+It worth mentioning that we used our implementation to produce and
+test all the examples that we present along this section.
 
@@
 
-Unlike gradual typing, we do not insert run-time checks that
-inspect the interaction between dynamically typed code and statically
-typed code during run-time.
-In other words, Typed Lua does not guarantee that dynamically typed
-code does not violate statically typed code.
-Typed Lua does not insert run-time checks because they can decrease
-run-time performance \citep{allende2013cis}.
-We believe that a careful evaluation of run-time checks should be done
-before inserting them on the type system.
 In the previous example, we cannot guarantee that \texttt{dist} is never
@@
 value that is not a number.
-Still, we can implement run-time checks in the future because we are
-using the consistent-subtyping relation of gradual typing to
-formalize our optional type system, and the type system is being
-designed to be sound.
 
@@
 
+\begin{figure}[!ht]
+\textbf{Abstract Syntax}\\
+\dstart
+$$
+\begin{array}{rl}
+s ::= & \;\; \mathbf{skip} \; | \;
+s_{1} \; ; \; s_{2} \; | \;
+\vec{l} = \vec{e}  \; | \;
+\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
+\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
+& | \; \mathbf{local} \; \vec{n}{:}\vec{T} = \vec{e} \; \mathbf{in} \; s \; | \;
+\mathbf{rec} \; n{:}T = f \; \mathbf{in} \; s \; | \;
+\mathbf{return} \; \vec{e} \; | \;
+a\\
+e ::= & \;\; \mathbf{nil} \; | \;
+\mathbf{false} \; | \;
+\mathbf{true} \; | \;
+{<}{\it number}{>} \; | \;
+{<}{\it string}{>} \; | \;
+{...} \; | \;
+f \; | \;
+\{ \; \vec{c}\; \} \\
+& | \; e_{1} + e_{2} \; | \;
+e_{1} \; {..} \; e_{2} \; | \;
+e_{1} == e_{2} \; | \;
+e_{1} < e_{2} \; | \;
+e_{1} \; \mathbf{and} \; e_{2} \; | \;
+e_{1} \; \mathbf{or} \; e_{2} \\
+& | \; \mathbf{not} \; e \; | \;
+- e \; | \;
+\# \; e \; | \;
+a \; | \;
+l \\
+l ::= & \; n \; | \;
+e_{1}[e_{2}]\\
+a ::= & \;\; e_{1}(\vec{e_{2}}) \; | \;
+e_{1}{:}n(\vec{e_{2}})\\
+f ::= & \; \mathbf{fun} \; (){:}R \; s \; | \;
+\mathbf{fun} \; ({...}{:}T){:}R \; s \; | \;
+\mathbf{fun} \; (\vec{n}{:}\vec{T}){:}R \; s \; | \;
+\mathbf{fun} \; (\vec{n}{:}\vec{T},{...}{:}T){:}R \; s\\
+c ::= & \; [e_{1}] = e_{2}\\
+n ::= & \; {<}{\it name}{>}\\
+\end{array}
+$$
+\dend
+\caption{Typed Lua abstract syntax}
+\label{fig:syntax}
+\end{figure}
+
 \section{Related Lua projects}

### 2014-08-11 doc/thesis/typedlua.tex
Rewriting the section about atomic types and functions

@@
 \begin{verbatim}
-    local function fact (n:number):number
+    local function factorial (n:number):number
       if n == 0 then
@@
     local x = 5
-    print(fact(x))
+    print(factorial(x))
 \end{verbatim}
@@
 the local variable \texttt{x}, the compiler uses local type inference to
-assign the type $\Number$ to \texttt{x}.
+assign the type $\Number$ to \texttt{x}, and this example compiles
+without any warnings.
 The compiler does not use the literal type $5$ instead of the base
@@
 previously.
-The compiler infers the return type of \texttt{dist} because it is
-not a recursive function, that is, Typed Lua does not infer the return
-type of recursive functions.
+Even though we did not annotated the return type of \texttt{dist},
+the compiler is able to infer its return type because it is local
+and not recursive.
 
@@
 
-In the previous example, we cannot guarantee that \texttt{dist} is never
+In this example, Typed Lua cannot guarantee that \texttt{dist} is never
 going to call \texttt{abs} with a parameter that is not a number,
@@
 value that is not a number.
+This is a typical example where run-time checks would ensure safety
+between the interaction of dynamically typed and statically typed code.
+
+Lua has first-class functions, and they have three peculiarities:
+flexible arity, multiple return values, and multiple assignment.
+In the first case, the number of arguments of a function call does
+not need to match the arity of the function declaration, as Lua silently
+drops extra arguments after evaluating them, or uses \texttt{nil} to
+replace missing arguments.
+In the second case, Lua function can return any number of values,
+and this number of returned values may not be statically known.
+In the third case, Lua has multiple assignment, and the semantics
+of argument passing is the same of the multiple assignment
+(or vice-versa), that is, calling a function is like doing a
+multiple assignment where the left hand side is the parameter list
+and the right hand side is the argument list.
+
+Typed Lua uses \emph{second-level types} to encode functions with
+flexible arity, multiple returns, and multiple assignment.
+We call them second-level because these types do not correspond
+to actual Lua values and we cannot use them to type variables or
+parameters.
+They are just an artifact of the interaction between the type system
+and the semantics of Lua.
+
+As we can see in Figure \ref{fig:typelang}, a second-level type is
+a tuple of first-level types optionally ending in a variadic type,
+or a union of these tuples.
+A variadic type $T{*}$ is a generator for a sequence of values of
+type $T \cup \Nil$.
+Unions of tuples play an important role in functions that are
+overloaded on the return type, together with \emph{projection types}.
+We will explain union types and projection types in more detail
+in the next section.
+
+To match the semantics of Lua function calls, the default operation
+mode of Typed Lua adds a variadic tail to the parts of a function
+type whenever the programmer does not specify one.
+In the previous examples, the type of both \texttt{factorial} and
+\texttt{abs} is
+$\Number \times \Value{*} \rightarrow \Number \times \Nil{*}$,
+and the type of \texttt{dist} is
+$\Any \times \Any \times \Value{*} \rightarrow \Number \times \Nil{*}$.
+If we call \texttt{abs} with extra arguments, Typed Lua silently ignores
+them, as the type signature lets \texttt{abs} receive any number of
+extra arguments.
+If we call \texttt{abs} with fewer arguments, Typed Lua uses $\Nil$ to
+replace the missing arguments and in the case of \texttt{abs} generates
+a warning, as $\Nil$ is neither a subtype of $\Number$ nor consistent
+with $\Number$.
+
+Typed Lua also provides a strict operation mode, and programmers can
+use this mode if they want to catch arity mismatch of function calls.
+In the strict mode, Typed Lua does not add a variadic tail to the
+parts of a function type unless the programmer explicitly declares one.
+In the previous examples, the type of both \texttt{factorial} and
+\texttt{abs} is $\Number \rightarrow \Number$, and the type of
+\texttt{dist} is $\Any \times \Any \rightarrow \Number$.
+If we call \texttt{abs} with either extra or missing arguments,
+Typed Lua generates warnings about arity mismatch.
+
+A variadic type can only appear in the tail position of a tuple,
+because Lua takes only the first value of any expression that appears
+in an expression list that is not in tail position.
+The following example shows the interaction between multiple returns
+and expression lists:
+\begin{verbatim}
+local function multiple ()
+  return 2, "foo"
+end
+local function sum (x:number, y:number)
+  return x + y
+end
+local x, y, z = multiple(), multiple()
+print(sum(multiple(), multiple()))
+\end{verbatim}
+
+In the default mode, the type of \texttt{multiple} is
+$\Value{*} \rightarrow \Number \times \String \times \Nil{*}$,
+and the type of \texttt{sum} is
+$\Number \times \Number \times \Value{*} \rightarrow \Number \times \Nil{*}$.
+In the right hand side of the multiple assignment, only the first
+value produced by the first call to multiple gets used,
+so the type of the right hand side is
+$\Number \times \Number \times \String \times \Nil{*}$,
+and the type assigned to \texttt{x}, \texttt{y}, and \texttt{z}
+are respectively $\Number$, $\Number$, and $\String$.
+The same tuple is assigned to the call to \texttt{sum} and it
+compiles without errors, as the first two components of the tuple
+are consistent with the types of the parameters, and the other
+components are consistent with \texttt{value}.
+
+In the strict mode, the type of \texttt{multiple} is
+$\Void \rightarrow \Number \times \String$,
+and the type of \texttt{sum} is $\Number \times \Number \rightarrow \Number$.
+Like in the default mode, only the first value produced by the
+first call to multiple gets used in the right hand side of the
+multiple assignment;
+and the type of the right hand side is also
+$\Number \times \Number \times \String \times \Nil{*}$,
+because the strict mode affects only the type of function declarations
+and multiple assignments that appear in function calls.
+Unlike in the default mode, the type of the call to \texttt{sum} is
+$\Number \times \Number \times \String$, and does not include the
+variadic tail to catch arity mismatch.
+Thus, this example compiles with error because the call to \texttt{sum}
+passes an unexpected extra argument of type $\String$.
 

### 2014-08-12 doc/thesis/typedlua.tex
Rewriting the section about union types

@@
 
-\subsection{Atomic types and Functions}
+\subsubsection{Atomic types and Functions}
 
@@
 
-\subsection{Unions}
-
-Typed Lua includes union types as a way to discriminate data
-structures that can hold values of several different, but fixed types.
+\subsubsection{Unions}
+
+Typed Lua includes union types to encode three common Lua idioms:
+the use of optional values, the overloading based on the tags of
+input parameters, and the overloading on the return type of functions.
+
+Optional values are unions of some type $t$ and $\Nil$, and Typed Lua
+includes the syntactic sugar \texttt{t?} to represent them because
+they appear quite often.
+The concrete syntax \texttt{t?} is syntactic sugar for \texttt{t|nil}
+that maps to the type $T\;\cup\;\Nil$.
+Optional values can appear when a function has optional parameters
+and when the program reads a value from an array or a map.
+The following example shows a function that has an optional parameter:
+\begin{verbatim}
+    local function message (name:string, greeting:string?)
+      local greeting = greeting or "Hello "
+      return greeting .. name
+    end
+
+    print(message("Typed Lua"))         -- prints Typed Lua
+    print(message("Typed Lua", "Hi "))  -- prints Hi Typed Lua
+\end{verbatim}
+
+Although the parameter \texttt{greeting} is optional, and has type
+$\String\;\cup\;\Nil$, the concatenation does not generate a type error
+because we used the short-circuiting \texttt{or} operator to declare a
+new variable \texttt{greeting} that is guaranteed to have type $\String$. 
+In Lua, only the values \texttt{nil} and \texttt{false} represent a
+false condition, so programmers often use the \texttt{or} operator as a
+common idiom for giving a default value to an optional parameter. 
+Typed Lua uses a typing rule to encode this idiom: 
+if the left hand side of \texttt{or} has type $T\;\cup\;\Nil$ and
+the right hand side has type $T$ then the \texttt{or} expression has
+type $T$.
+
+In fact, we do need to declare a new variable \texttt{greeting} that
+shadows the optional parameter:
+\begin{verbatim}
+    local function message (name:string, greeting:string?)
+      greeting = greeting or "Hello "
+      return greeting .. name
+    end
+\end{verbatim}
+
+Typed Lua allows the assignment $x\;=\;x\;\mathbf{or}\;e$ to refine
+the type of $x$ from $T\;\cup\;\Nil$ to $T$ if it matches three restrictions: 
+the type of $e$ is a subtype of $T$, the variable $x$ is local to the
+current function, and it is not being assigned in another function.
+The refinement only affects the type of $x$ in the remainder of the
+current scope.
+In the case of \texttt{greeting}, the assignment refines its type
+from $\String\;\cup\;\Nil$ to $\String$.
+
+Lua programmers often overload the input parameters of functions,
+and use the \texttt{type} function to inspect the tag of the
+input parameters to take different actions depending on what
+those tags are.
+The simplest case overloads on just a single parameter:
+\begin{verbatim}
+    local function overload (s1:string, s2:string|number)
+      if type(s2) == "string" then
+        return s1 .. s2
+      else
+        -- string.rep : (string, number, string) -> (string)
+        return string.rep(s1, s2)
+      end
+    end
+\end{verbatim}
+
+Typed Lua has a small set of \texttt{type} predicates that allows
+programmers to constrain the type of a local variable inside a condition.
+This example uses the \texttt{type(x) == "string"} predicate that
+constrains the type of $x$ from $T\;\cup\;\String$ to $\String$ when
+the predicate is true and $T$ otherwise.
+This is a simplified form of \emph{flow typing} \citep{guha2011tlc,tobin-hochstadt2010ltu}.
+As with \texttt{or}, the variable must be local to the function,
+and it is not being assigned in another function.
+
+The \texttt{type} predicates can only discriminate based on tags,
+so they are limited on the kinds of unions that they can discriminate.
+For instance, the predicates can discriminate a union that combines
+a table type with a base type, or a table type with a function type,
+or two base types, but they cannot discriminate a union that combines
+two different table types, or two different table types.
+
+Lua programmers also overload the return type of functions,
+usually for signaling the occurence of errors.
+In this idiom, a function returns its normal set of return values
+when it successfully finished its execution,
+and usually returns \texttt{nil} as the first value plus an
+error message or other data that describes the error
+when something failed during its execution.
 Next, we show an example:
 \begin{verbatim}
-    local x:number?            -- OK
-    local y:number|string = 1  -- OK
-    local z:number = x         -- Error
-\end{verbatim}
-
-The annotation \texttt{number?} is syntactic sugar for
-\texttt{number|nil} that means the type $\Number\;\cup\;\Nil$.
-Typed Lua includes the syntactic sugar \texttt{t?} because the
-type $t\;\cup\;\Nil$ is very common in Lua code.
-
-The last line of the previous example gives a type error because we
-are attempting to assign a value of type $\Number\;\cup\;\Nil$
-to a variable that accepts only values of type $\Number$.
-Although we can solve this type error changing the annotation of
-variable \texttt{z} from \texttt{number} to \texttt{number?}, there is
-another way that we can use to solve this type error without changing
-the annotation:
-\begin{verbatim}
-    local z:number = x or 0  -- OK
-\end{verbatim}
-
-Typed Lua has a typing rule for checking that
-$t\;\cup\;\Nil\;\mathbf{or}\;t$ always has the type $t$.
-
-\subsection{Tables and Interfaces}
+    local function idiv(dividend:number, divisor:number):
+          (number, number)|(nil, string)
+      if divisor == 0 then
+        return nil, "division by zero"
+      else
+        local r = dividend % divisor
+        local q = (dividend - r) / divisor
+        return q, r
+      end 
+    end
+\end{verbatim}
+
+Typed Lua also includes a syntactic sugar for this idiom:
+we can annotate the return type of \texttt{idiv} with
+\texttt{(number, number)?} to denote the same union.
+The parentheses are always necessary in this case because
+\texttt{t?} is syntactic sugar for \texttt{t|nil},
+while \texttt{(t)?} is syntactic sugar for \texttt{(t)|(nil, string)}.
+
+The type of idiv, in the default mode, is
+$\Number \times \Number \times \Value{*} \rightarrow
+(\Number \times \Number \times \Nil{*}) \cup
+(\Nil \times \String \times \Nil{*})$
+A typical client of this function would use it as follows:
+\begin{verbatim}
+    local q, r = idiv(a, b)
+    if q then
+      print(a == b * q + r)
+    else
+      print("ERROR: " .. r)
+    end
+\end{verbatim}
+
+When Typed Lua finds a union of tuples in the right hand side of
+a declaration, it stores this union in a special type environment
+with a fresh name and assigns \emph{projection types} to the
+variables that appear unannotated in the left hand side of the declaration.
+In our example, Typed Lua assigns the tuple
+$(\Number \times \Number \times \Nil{*}) \cup (\Nil \times \String \times \Nil{*})$
+to a type variable $X$.
+Thus, variable \texttt{q} gets the type $X_{1}$
+that is equivalent to $\Number\;\cup\;\Nil$, 
+and variable \texttt{r} gets the type $X_{2}$
+that is equivalent to $\Number\;\cup\;\String$.
+
+Projection types do not appear in the subtyping and consistent-subtyping
+relations because Typed Lua actually replaces them with the union of the
+corresponding component of each tuple.
+If a variable with a projection type appears in a \texttt{type} predicate,
+it discriminates against all tuples in the union.
+In our example, $X$ is
+$(\Number \times \Number \times \Nil{*}) \cup (\Nil \times \String \times \Nil{*})$
+outside of the \texttt{if} statement,
+but $\Number \times \Number \times \Nil{*}$ inside the \texttt{then} block,
+and $\Nil \times \String \times \Nil{*}$ inside the \texttt{else} block.
+Thus, variable \texttt{q} has type $\Number$ and variable \texttt{r} also
+has type $\Number$ inside the \texttt{then} block;
+and variable \texttt{q} has type $\Nil$ while variable \texttt{r}
+has type $\String$ inside the \texttt{else} block.
+
+We could also have used \texttt{type(q) == "number"} or
+even \texttt{type(r) == "number"} as the predicate of our example,
+as both predicates would produce the same result.
+However, the form that appears in our example is much more succint and idiomatic.
+We can also use projection types to write overloaded functions where
+the type of an input parameter depends on the type of another input parameter.
+
+Typed Lua does not allow assignments to variables that hold a projection type,
+unless the union has been discriminated down to a single value.
+Unrestricted assignment to these variables would be unsound,
+as it could break the dependency relation among the types in each tuple
+that is part of the union.
+
+The overloading mechanims of Typed Lua have a limitation:
+the return type cannot depend on the input type.
+Next, we show an example:
+\begin{verbatim}
+    local function limitation (x:number|string, y:number|string)
+      if type(x) == "number" then
+        if type(y) == "number" then
+          return x + y
+        end
+      else
+        if type(y) == "string" then
+          return x .. y
+        end
+      end
+    end
+\end{verbatim}
+
+This example shows that we cannot write a function that is
+guaranteed to return a number when we pass two numbers and
+guaranteed to return a string when we pass two strings.
+Even though we can use intersection types to express the type
+of this function as
+$(\Number \times \Number \rightarrow \Number) \cap (\String \times \String \rightarrow \String)$,
+and thefore solving the limitation,
+intersection types require more sophisticated flow typing to
+check whether a function has this type,
+and we are still working on this problem.
+
+\subsubsection{Tables and Interfaces}
 
@@
 
-\subsection{Modules}
-
-\subsection{Objects and Classes}
+\subsubsection{Modules}
+
+\subsubsection{Objects and Classes}
 

### 2014-08-13 doc/thesis/typedlua.tex
Rewriting the section about table types and interfaces

@@
 
-Typed Lua includes table types to represent Lua's tables.
-Although Typed Lua has just one table type constructor, it has three
-different syntaxes for defining the type of a table.
-The first syntax defines table types for hash tables;
-it is written \texttt{\{t1:t2\}},
-and maps to the table type $\{t_{1}:t_{2}\;\cup\;\Nil\}$.
-This table type represents a hash table that maps values of type
-$t_{1}$ to values of type $t_{2}\;\cup\;\Nil$, as Lua returns
-$\Nil$ when we use a non-existing key to index a table.
-The second syntax defines table types for lists;
+Tables are the only mechanism that Lua has to build data structures;
+they are associative arrays where any value (except \texttt{nil})
+can be used as a key.
+Programmers can use tables to represent tuples, arrays (dense or sparse),
+records, graphs, modules, objects, etc.
+Lua has syntactic sugar for indexing tables as records:
+\texttt{t.k} is syntactic sugar for \texttt{t["k"]}, and means that
+we are accessing field \texttt{k} from table \texttt{t}. 
+In this section, we show how Typed Lua encodes tables as maps,
+arrays, and records.
+
+Table types are the only framework that Typed Lua has to represent
+the different uses that a Lua table has.
+A table type $\{K_{1}:T_{1}, ..., K_{n}:T_{n}\}$ represents a map
+from values of type $K_{i}$ to values of type $T_{i}$.
+
+Although Typed Lua has just one framework to represent tables,
+it has three different syntaxes for defining the type of a table.
+The first syntax defines table types for maps:
+it is written \texttt{\{k:t\}},
+and maps to the table type $\{K:T\;\cup\;\Nil\}$.
+This table type represents a map that maps values of type
+$K$ to values of type $T\;\cup\;\Nil$,
+as Lua returns $\Nil$ when we use a non-existing key to index a table.
+The second syntax defines table types for arrays:
 it is written \texttt{\{t\}},
-and maps to the table type $\{\Number:t\;\cup\;\Nil\}$.
+and maps to the table type $\{\Number:T\;\cup\;\Nil\}$.
 This table type is just a syntactic sugar to the definition of a
-hash table that maps values of type $\Number$ to values of type
-$t\;\cup\;\Nil$.
-The third syntax defines table types for records;
-it is written \texttt{\{s1:t1, ..., sn:tn\}},
-and maps to the table type $\{s_{1}:t_{1}, ..., s_{n}:t_{n}\}$.
-This table type represents a record that maps the literals
-$s_{i}$\texttt{, ..., }$s_{n}$ to values of types
-$t_{i}$\texttt{, ..., }$t_{n}$, where $i$ may vary from one to $n$.
-
-Next, we show one example of table type to type a hash table:
-\begin{verbatim}
-    local t:{string:number} = { foo = 1 }  -- OK
-    local x:number = t.foo                 -- Error
-    local y:number? = t.bar                -- OK
-\end{verbatim}
-
-The second line of this example gives a type error for the same
-reason that the last line of the previous example gives a type error
-too, that is, we are trying to assign a value that can be of two
-different types to a variable that accepts values of only one type.
-Although the field \texttt{bar} does not exist in \texttt{t}, the last
+map that maps values of type $\Number$ to values of type $T\;\cup\;\Nil$.
+The third syntax defines table types for records:
+it is written \texttt{\{k1:t1, ..., kn:tn\}},
+and maps to the table type $\{K_{1}:T_{1}, ..., K_{n}:T_{n}\}$.
+This table type represents a record that maps the literal types
+$K_{i}$\texttt{, ..., }$K_{n}$ to values of types
+$T_{i}$\texttt{, ..., }$T_{n}$, where $i$ may vary from one to $n$.
+
+Next, we show one example of table type to type a map from strings to numbers:
+\begin{verbatim}
+    local t:{string:number} = { foo = 1 } 
+    local x:number = t.foo               -- compile-time error
+    local y:number? = t.bar              -- y gets nil
+    local z:number = t["bar"] or 0       -- z gets 0 
+\end{verbatim}
+
+The second line of this example gives a type error because we are
+attempting to assign a value of type $\Number\;\cup\;\Nil$ to a
+variable that accepts only values of type $\Number$.
+Although the field \texttt{bar} does not exist in \texttt{t}, the third
 line of this example does not give a type error because the
@@
 \texttt{t}.
-
-Next, we show one example of table type to type a list:
+The last line shows that the \texttt{or} idiom is also useful to
+give a default value to a missing table field.
+
+Next, we show one example of table type to type an array:
 \begin{verbatim}
@@
                             "Wednesday", "Thursday", "Friday",
-                            "Saturday" }  -- OK
-\end{verbatim}
+                            "Saturday" }
+    local x = days[1]                    -- x gets "Sunday"
+    local y = days[8]                    -- y gets nil
+\end{verbatim}
+
+In this example, the type of both variables \texttt{x} and \texttt{y}
+is $\String\;\cup\;\Nil$.
+An incovenience aspect of making the types of maps and arrays always
+include the type $\Nil$ is to overload the programmers,
+as they need to use \texttt{or} or \texttt{if} to narrow the type of
+the elements they are accessing.
 
@@
 variable declaration unannotated and let local type inference assign
-a more specific table type to the list.
+a more specific table type to the variable.
 If we remove the annotation in this example, the compiler uses the
@@
 \end{align*}
-This inferred table type is a subtype of the type
-$\{\Number:\String\;\cup\;\Nil\}$, which is the type that we used to
-annotate the variable \texttt{days}.
+% text about subtyping
 
@@
 
-It also worth mentioning that we can use the style of the inferred
-type to define heterogeneous tuples.
+We can also use the style of the inferred type to define heterogeneous
+tuples:
+\begin{verbatim}
+    local album = { "Transformer", 1972, "Lou Reed" }
+\end{verbatim}
+
+The next example shows how we can annotate a \texttt{map} function
+that receives another function as its parameter and uses this
+function to alter the elements of a list of numbers, which is
+also a parameter of \texttt{map}:
+\begin{verbatim}
+    local function map (f:(number)->(number), l:{number})
+      for i = 1, #l do
+        l[i] = f(l[i] or 0)
+      end
+    end
+\end{verbatim}
 
@@
 
-Typed Lua includes interfaces as syntactic sugar to table types
-that specify records.
+Typed Lua includes \emph{interfaces} as syntactic sugar to table types
+that specify records, as writing table types can be unwieldy when
+records get bigger and types of record fields get more complicated.
 For instance, we can use the following interface as an alias to the
@@
 \begin{verbatim}
-    interface Person
+    local interface Person
       firstname:string
@@
 
-Now we can use the type \texttt{Person} as an alias to a table type
-that certainly has the fields \texttt{firstname} and \texttt{lastname}.
-Next, we show an example of code that uses the type \texttt{Person}:
+The declaration of the interface \texttt{Person} is an alias to the table type
+$\{``firstname":\String, ``lastname":\String\}$ in the remainder
+of the current scope.
+Now we can use \texttt{Person} in type declarations:
 \begin{verbatim}
@@
 
-    print(greeter(user1)) -- OK
-    print(greeter(user2)) -- Error
-    print(greeter(user3)) -- OK
-    print(greeter(user4)) -- Error
+    print(greeter(user1)) -- Hello, Lewis Allan
+    print(greeter(user2)) -- compile-time error
+    print(greeter(user3)) -- Hello, Lou Reed
+    print(greeter(user4)) -- compile-time error
 \end{verbatim}
@@
 
-Using interfaces we can also define recursive types such as linked
-lists:
-
-\begin{verbatim}
-    interface Element
+If our record type has optinal fields, that is, fields that can be \texttt{nil},
+we need to use an explicit type declaration when declaring a
+variable of this record type, as we show in the following example:
+\begin{verbatim}
+    local interface Person
+      firstname:string
+      middlename:string?
+      lastname:string
+    end
+    local user1:Person = { firstname = "Lewis",
+                           middlename = "Allan" or nil,
+                           lastname = "Reed" }
+    local user2:Person = { lastname = "Reed",
+                           firstname = "Lou" }
+\end{verbatim}
+
+We need an explicit type declaration because neither of the types
+that the Typed Lua compiler assigns to the above table constructors
+is a subtype of
+$\{``firstname":\String, ``middlename":\String\cup\Nil, ``lastname":\String\}$,
+the type that \texttt{Person} describes.
+While declaring \texttt{user1}, we use \texttt{or} to cast the type of
+\texttt{middlename} because of the contravariance in the type of
+mutable fields.
+
+We can also use interfaces to define recursive types such as liked lists:
+
+\begin{verbatim}
+    local interface Element
       info:number
@@
 
+The declaration of the interface \texttt{Element} is an alias to the
+recursive type
+$\mu Element.\{``info":\Number, ``next":Element \cup \Nil\}$ in the
+remainder of the current scope.
+
 We need to explicitly annotate \texttt{next} being of type
-\texttt{Element}$\;\cup\;\Nil$ because in our type system $\Nil$
+$Element\;\cup\;\Nil$ because in our type system $\Nil$
 is not the bottom type, that is, $\Nil$ is subtype of itself only.
@@
 the absence of an useful value.
+Thus, making $\Nil$ the bottom type would severaly reduce the amount
+of type safety in the system.
+
+Now we can use \texttt{Element} to annotate a function that
+inserts an element at the begining of a list:
+\begin{verbatim}
+    local function insert (e:Element?, v:number):Element 
+      return { info = v, next = e }
+    end
+\end{verbatim}
+
+We need an explicit type declaration on the return type because
+the Typed Lua compiler cannot infer recursive types, though
+it has subtyping rules to check that the inferred return type
+matches the annotation.
+
+Now we can write a program that declares a list, inserts some
+elements in this list, and then traverses the list printing
+each element of the list:
+\begin{verbatim}
+    local l:Element?
+
+    l = insert(l, 4)
+    l = insert(l, 3)
+    l = insert(l, 2)
+    l = insert(l, 1)
+
+    while l do
+      if l then print(l.info) end
+      l = l.next                  -- compile-time error
+    end
+\end{verbatim}
+
+However, this example has one compile-time error because we
+are attempting to index the variable \texttt{l},
+which can be \texttt{nil}.
+Notice that we would have the same compile-time error while
+printing the list element if we had not narrowed the type of
+\texttt{l}.
+To eliminate this compile-time error we have to write a
+function that narrows the type of \texttt{l} before using it:
+\begin{verbatim}
+    local function get_next (l:Element?)
+      if l then return l.next else return nil end
+    end
+\end{verbatim}
+
+Now we can use \texttt{get\string_next} to rewrite the \texttt{while}
+from the previous example, thus eliminating the compile-time error:
+\begin{verbatim}
+    while l do
+      if l then print(l.info) end
+      l = get_next(l)
+    end
+\end{verbatim}
+
+\subsubsection{Modules}
+
+Typed Lua will also introduce a specific syntax for defining modules.
+For now, each file is a module.
+We use all global names and their types to build the interface that
+represents the type of the module, in case the module is not returned
+at the end of the file.
+
+\subsubsection{Objects and Classes}
 
@@
 privacy rules.
-
-Typed Lua will also introduce a specific syntax for defining modules.
-For now, each file is a module.
-We use all global names and their types to build the interface that
-represents the type of the module, in case the module is not returned
-at the end of the file.
-
-Typed Lua includes function types because functions are first-class
-values in Lua.
-The syntax for defining a function type is \texttt{(s) -> (s)},
-where \texttt{s} is a second-class type.
-As we saw in Figure \ref{fig:typelang}, a second-class type in
-Typed Lua can be the type $\Void$, a first-class type, a variadic
-type, or a tuple type.
-We use tuple types to represent multiple parameters, multiple return
-values, and multiple assignments, and we use the variadic type to
-represent variadic functions and the vararg expression.
-Even though we already included the tuple type and the vararg type,
-we are still working on these types.
-
-The next example shows how we can annotate a \texttt{map} function
-that receives another function as its parameter and uses this
-function to alter the elements of a list of numbers, which is
-also a parameter of \texttt{map}:
-\begin{verbatim}
-    local function map (f:(number)->(number), l:{number}):void
-      for i=1, #l do
-        l[i] = f(l[i] or 0)
-      end
-    end
-\end{verbatim}
-
-To conclude this section, we just started the design and
-implementation of Typed Lua, and there is still several design
-decisions to be taken on this subject.
-For instance, our type system should handle the subtleties that
-come with metatables.
-
-\subsubsection{Modules}
-
-\subsubsection{Objects and Classes}
 
@@
 
+Rule for checking if a table type is well formed:
+\[
+\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
+\]
+
 \section{Related Lua projects}

### 2014-08-14 doc/thesis/typedlua.tex
Adding text about the refinement of table types

@@
 \begin{verbatim}
-    local function greeter (person:Person):string
-      return "Hello " .. person.firstname ..
+    local function byebye (person:Person):string
+      return "Goodbye " .. person.firstname ..
              " " .. person.lastname
@@
 
-    print(greeter(user1)) -- Hello, Lewis Allan
+    print(greeter(user1)) -- Goodbye Lewis Allan
     print(greeter(user2)) -- compile-time error
-    print(greeter(user3)) -- Hello, Lou Reed
+    print(greeter(user3)) -- Goodbye Lou Reed
     print(greeter(user4)) -- compile-time error
@@
 \end{verbatim}
+
+Lua programmers often build records incrementally, that is,
+they usually declares a local variable with an empty table,
+and then use assignment to add fields to this table:
+\begin{verbatim}
+    local person = {}
+    person.firstname = "Lou"
+    person.lastname = "Reed"
+    print("bye bye " .. person.firstname)  -- bye bye Lou
+\end{verbatim}
+
+In this example, we want to refine the type of \texttt{person} as
+we build the table:
+starting with $\{\}$, and then refining to $\{``firstname":\String\}$,
+and finally reaching $\{``firstname":\String, ``lastname":\String\}$.
+This type change is trickier than the one that we introduced for narrowing
+union types, as we are not just allowing the programmer to change
+the type of the variable \texttt{person}, but we are actually allowing
+the programmer to change the type of the value that \texttt{person} points to.
+This type change is safe in this example, but it is not in the following one:
+\begin{verbatim}
+    local person = {}
+    local bogus = person
+    person.firstname = "Lou"
+    person.lastname = "Reed"
+    bogus.firstname = true               -- compile-time error
+    print("bye bye " .. person.firstname)
+\end{verbatim}
+
+In this example, the fifth line is unsound because it allows the
+programmer to change the type of the value that is stored in
+\texttt{person.firstname}, and makes the last line crash during
+run-time.
+On the first line, we assign an empty table to \texttt{person}.
+On the second line, we assign the type of \texttt{person} to
+\texttt{bogus}.
+On the third and fourth lines we change the type of \texttt{person}
+as we did in the previous example.
+On the fifth line we try to change the type of \texttt{bogus}
+from $\{\}$ to $\{``firstname":\Boolean\}$, but if we allow this
+type change we also allow changing the value that is stored in
+\texttt{person.firstname}, regardless of its type.
+Taking the changes individually looks fine, but the truth is that
+aliasing makes one of them unsound.
+
+The location of the change also matters, as the next example shows:
+\begin{verbatim}
+    local person = { lastname = "Reed" }
+    local function spoil ()
+      person.firstname = nil             -- compile-time error
+    end
+    person.firstname = "Lou"
+    spoil()
+    print("bye bye " .. person.firstname)
+\end{verbatim}
+
+In this example, the third line is unsound because it allows the
+programmer to change the type of \texttt{person} outside of the scope
+that it was declared.
+The call to \texttt{spoil} erases the field \texttt{firstname} from
+\texttt{person}, and makes the last line crash during run-time.
+
+It is also not sound to allow the type of the fields to change,
+as the next example shows:
+\begin{verbatim}
+    local person = { firstname = "Lewis", lastname = "Reed" }
+    person.middlename = "Allan"
+    person.firstname = 1964              -- compile-time error
+    person.lastname = 2013               -- compile-time error
+    print("bye bye " .. person.firstname)
+\end{verbatim}
+
+In this example, both third and fourth line are unsound because
+they are trying to change the type of fields that already exist.
+
+Typed Lua tags a variable that holds a table type as either
+\emph{open} or \emph{closed}.
+If a variable gets its type from a table constructor then it
+is open, otherwise it is always closed.
+
+Typed Lua uses three restrictions to decide whether it is
+sound to allow field assignment to change the type of an
+open variable or not:
+the variable must be local to the current block,
+the new type must be a subtype of the old one, and
+the variable cannot have been assigned in another function.
+
+We can create aliases to open variables, but they are always closed.
+Any mutation on these aliases is not a problem, as the type of
+the original reference can only change towards a subtype.
+For mutable fields this means that the type of the field cannot
+change after it is added to the type of the table.
+
+Table types along with their refinement through field assignment
+are the foundation for defining modules and objects in Typed Lua.
+In the next section, we show that the refinement of table types
+through field assignment is crucial for making Typed Lua
+support idiomatic definition of Lua modules.
 

### 2014-08-14 doc/thesis/typedlua.tex
Adding text about modules

@@
     person.middlename = "Allan"
-    person.firstname = 1964              -- compile-time error
+    person.firstname = 1942              -- compile-time error
     person.lastname = 2013               -- compile-time error
@@
 
-Typed Lua will also introduce a specific syntax for defining modules.
-For now, each file is a module.
-We use all global names and their types to build the interface that
-represents the type of the module, in case the module is not returned
-at the end of the file.
+Lua does not set policies on how programmers should define modules,
+but it provides mechanisms for organizing a program in modules.
+To load a module, Lua first checks whether the module is already
+loaded or not.
+When the module is not loaded, Lua executes its soure file as a
+function, and the value returned by this function is the module.
+Thus, Lua modules can be any Lua value.
+However, our survey from Section \ref{sec:statistics} confirms that
+the convention among Lua programmers is to use tables for defining modules. 
+In this case, the fields of the table are the functions and other
+values that the module exports.
+
+The simplest way for defining a module in Lua is to declare an
+empty table at the begin of the source file,
+populate this table with the members that the module should export,
+and return this table at the end of the source file. 
+The following example shows how Typed Lua handles this case:
+\begin{verbatim}
+    local mymath = {}
+    local RADIANS_PER_DEGREE = 3.14 / 180.0
+    mymath.deg = function (x:number)
+      return x / RADIANS_PER_DEGREE
+    end
+    mymath.rad = function (x:number)
+      return x * RADIANS_PER_DEGREE
+    end
+    mymath.pow = function (x:number, y:number)
+      return x ^ y
+    end
+    return mymath
+\end{verbatim}
+
+In this example, Typed Lua uses the refinement of table types to
+allow the programmer to incrementally build the type of the
+module \texttt{mymath}.
+Like in Lua, the variable \texttt{RADIANS\string_PER\string_DEGREE}
+is private because we declared it as local to the module. 
+The \texttt{return} at the end of the source file gives the
+final type of the module \texttt{mymath}:
+\begin{align*}
+\{ & ``deg":\Number \times \Value{*} \rightarrow \Number \times \Nil{*},\\
+& ``rad":\Number \times \Value{*} \rightarrow \Number \times \Nil{*},\\
+& ``pow":\Number \times \Number \times \Value{*} \rightarrow \Number \times \Nil{*}\}
+\end{align*}
+
+Another way of defining modules in Lua is to declare only locals
+in the source file and return a table constructor that includes
+the members that the module should export at the end of the source file.
+The following example shows how Typed Lua handles this case:
+\begin{verbatim}
+    local RADIANS_PER_DEGREE = 3.14 / 180.0
+    local function deg (x:number)
+      return x / RADIANS_PER_DEGREE
+    end
+    local function rad (x:number)
+      return x * RADIANS_PER_DEGREE
+    end
+    local function pow (x:number, y:number)
+      return x ^ y
+    end
+    return {
+      deg = deg,
+      rad = rad,
+      pow = pow,
+    }
+\end{verbatim}
+
+In this example, the \texttt{return} at the end of the source file
+also gives the final type of the module \texttt{mymath}, which is
+the same of the previous example.
+
+There is still a third way of defining modules in Lua through the
+\texttt{module} function, which has been deprecated.
+In this style, the first line of the module calls \texttt{module}.
+This function installs a fresh table as the global environment for
+the rest of the module, so any assignments to global variables are
+field assignments to this table.
+The \texttt{module} function also sets an \texttt{\string_M} field
+in this table as a circular reference to the table itself,
+so the module can end with \texttt{return \string_M},
+but this explicit return is not necessary.
+
+Even though this style has been deprecated, our survey from
+Section \ref{sec:statistics} indicated that around a third of the
+Lua modules that are in LuaRocks still use this style.
+Thus, Typed Lua also supports this style:
+\begin{verbatim}
+    module "mymath"
+    local RADIANS_PER_DEGREE = 3.14 / 180.0
+    deg = function (x:number)
+      return x / RADIANS_PER_DEGREE
+    end
+    rad = function (x:number)
+      return x * RADIANS_PER_DEGREE
+    end
+    pow = function (x:number, y:number)
+      return x ^ y
+    end
+\end{verbatim}
+
+In this example, Typed Lua treats accesses to global variables as
+field accesses to an open table in the top-level scope,
+and at the end of file this open table gives the final type of the
+module \texttt{mymath}, which is also the same of the previous two
+examples.
+
+After we define the module \texttt{mymath}, regardless of adopted style,
+users can use it in the standard way:
+\begin{verbatim}
+    local m = require "mymath"
+    print(m.pow(2, 3))                   -- 8
+    print(m.pow(2, "foo"))               -- compile-time error
+\end{verbatim}
+
+And now the users of the module \texttt{mymath} have the policies that
+Typed Lua provides.
+The call to \texttt{require} assigns the type of the module \texttt{mymath}
+to \texttt{m}, so Typed Lua can catch misuses of the module.
+
+The way Typed Lua handle modules using table types is also
+relevant for supporting object-oriented programming, as we discuss in
+the next section.
 

### 2014-08-14 doc/thesis/typedlua.tex
Adding text about object-oriented programming

@@
 
-\subsubsection{Objects and Classes}
-
-Besides interfaces, Typed Lua will also introduce classes.
-We want to offer classes as a standard way for writing code that
-Lua programmers are already using, but without tying them to a
-specific model.
-Our classes will not include static attributes and will also not offer
-privacy rules.
+\subsubsection{Object-Oriented Programming}
+
+%Besides interfaces, Typed Lua will also introduce classes.
+%We want to offer classes as a standard way for writing code that
+%Lua programmers are already using, but without tying them to a
+%specific model.
+%Our classes will not include static attributes and will also not offer
+%privacy rules.
+
+Lua provides minimal support for object-oriented programming.
+The basic mechanism is the \texttt{:} syntactic sugar for method
+calls and method definitions.
+The Lua interpreter translates \texttt{obj:method(args)} into
+an operation that first evaluates \texttt{obj}, if it is a table
+Lua checks if there is field \texttt{method} in this table,
+and then calls it with the result of evaluating \texttt{obj} as
+the first argument, followed by the result of evaluating the
+argument list in the original expression.
+
+Typed Lua uses table types along with the \emph{self type} to
+represent objects, as we show in the following example:
+\begin{verbatim}
+    local Shape = { x = 0, y = 0 }
+    const function Shape:move (dx:number, dy:number)
+      self.x = self.x + dx
+      self.y = self.y + dy
+    end
+\end{verbatim}
+
+This example introduces some features that we must explain
+before we show the type of \texttt{Shape}.
+In Lua, the \texttt{:} syntactic sugar assigns a function to the
+field with a first parameter named \texttt{self}, plus any other
+parameters.
+Typed Lua does the same, and assigns the type $\Self$ to the
+implicit parameter \texttt{self}.
+Typed Lua also adds the \texttt{const} annotation because the
+$\Const$ qualifier is necessary for covariance among object types work.
+Under this circumstances, the type of \texttt{Shape} is an object of
+the following type:
+\begin{align*}
+\{ & ``x":\Number, ``y":\Number, \\
+& \Const \; ``move":\Self \times \Number \times \Number \times \Value{*} \rightarrow \Nil{*}\}
+\end{align*}
+
+While \texttt{:} is syntactic sugar in plain Lua, Typed Lua uses it
+to check method calls, binding any occurence of the $\Self$ type
+in the type of the method to the receiver.
+Indexing a method and not immediately calling it with the correct
+receiver is a compile-time error:
+\begin{verbatim}
+    Shape.move(Shape, 10, 10)  -- Shape.x = 10 and Shape.y = 10
+    Shape:move(5, 20)          -- Shape.x = 5 and Shape.y = 20
+    local p = Shape.move       -- compile-time error
+\end{verbatim}
+
+Lua has a mechanism for self-like (or JavaScript-like) delegation of
+missing fields in a table.
+After the call \texttt{setmetatable(t1, \{ \string_\string_index = t2 \})},
+Lua looks up in \texttt{t2} for any missing fields of \texttt{t1}.
+Lua programmers often use this mechanism to simulate classes.
+
+Typed Lua combines this mechanism along with interfaces to allow
+programmers to define classes in a similar fashion to the definition
+of modules, as we show in the following example:
+\begin{verbatim}
+    interface Shape
+      x:number
+      y:number
+      const new:(self, number, number) -> (self)
+      const move:(self, number, number) -> ()
+    end
+    local Shape = { x = 0, y = 0 }
+    const function Shape:new (x:number, y:number)
+      local s = setmetatable({}, { __index = self })
+      s.x = x
+      s.y = y
+      return s
+    end
+    const function Shape:move (dx:number, dy:number)
+      self.x = self.x + dx
+      self.y = self.y + dy
+    end
+    return Shape
+\end{verbatim}
+
+In Typed Lua, \texttt{setmetatable} is a strict primitive.
+Typed Lua only allows calling \texttt{setmetatable} with
+two operands.
+The first operand can be any table \texttt{t1}, but the second operand
+must be a table of the form \texttt{\{ \string_\string_index = t2 \}},
+where \texttt{t2} can be any table.
+
+The reason for being so strict with \texttt{setmetatable} is because
+it is the only mechanism that Typed Lua has to create classes
+as prototype objects.
+In a \texttt{setmetatable} expression, if the type of \texttt{t2}
+is a subtype of the type of \texttt{t1}, it changes \texttt{t1} to
+have the same type of \texttt{t2}.
+
+In our example, we use \texttt{setmetatable} to define a constructor
+\texttt{new} to the class \texttt{Shape}.
+The type of \texttt{new} is
+$\Self \times \Number \times \Number \times \Value{*} \rightarrow \Self \times \Nil{*}$,
+because we use \texttt{setmetatable} to declare \texttt{s} having
+the same type of \texttt{self}.
+
+In the same example, we also declare the interface \texttt{Shape}.
+This allows us to export the type of the class \texttt{Shape},
+which is the same type that is given by the variable \texttt{Shape}:
+\begin{align*}
+\{ & ``x":\Number, ``y":\Number,\\
+& \Const \; ``new":\Self \times \Number \times \Number \times \Value{*} \rightarrow \Self \times \Nil{*},\\
+& \Const \; ``move":\Self \times \Number \times \Number \times \Value{*} \rightarrow \Nil{*}\}
+\end{align*}
+
+In Typed Lua, exporting interfaces allow us to use \texttt{require}
+for creating prototype objects that work as classes,
+along with a shortcut to the table type of the class.
+This allows us to use this shortcut on type annotatoins,
+as we show in the following example:
+\begin{verbatim}
+    local Shape = require "shape"
+    local shape1 = Shape:new(0, 5)
+    local shape2:Shape = Shape:new(10, 10)
+\end{verbatim}
+
+Typed Lua also uses \texttt{setmetatable} to simulate single inheritance,
+as we show in the following example:
+
+\begin{verbatim}
+    local Shape = require "shape"
+    interface Circle
+      x:number
+      y:number
+      radius:number
+      const new:(self, number, number, value) -> (self)
+      const move:(self, number, number) -> ()
+      const area:(self) -> (number)
+    end
+    local Circle = setmetatable({}, { __index = Shape })
+    Circle.radius = 0
+    const function Circle:new (x:number, y:number, radius:value)
+      local c = setmetatable(Shape:new(x, y), { __index = self })
+      c.radius = tonumber(radius) or 0
+      return c
+    end
+    const function Circle:area ()
+      return math.pi * self.radius * self.radius
+    end
+    return Circle
+\end{verbatim}
+
+A call to \texttt{setmetatable} where \texttt{t1} is a table
+constructor assigned to a fresh local variable, and \texttt{t2}
+is not \texttt{self} gives this variable an open type.
+That is the reason why we can add new methods and override
+existing ones in \texttt{Circle}.
+
+In this example, we also call \texttt{setmetatable} to redefine \texttt{new}.
+Notice how we can call the constructor of \texttt{Shape} inside
+the overridden constructor.
+A limitation of this class system is that the overridden constructor
+must be a subtype of the original constructor,
+so the type of \texttt{radius} has to be quite permissive.
+
+In the same example, we also declare the interface \texttt{Circle},
+but with a syntactical overhead to manually include the new and
+overridden methods.
+The interface \texttt{Circle} exports the following type:
+\begin{align*}
+\{ & ``x":\Number, ``y":\Number, ``radius":\Number,\\
+& \Const \; ``new":\Self \times \Number \times \Number \times \Value \times \Value{*} \rightarrow \Self \times \Nil{*},\\
+& \Const \; ``move":\Self \times \Number \times \Number \times \Value{*} \rightarrow \Nil{*},\\
+& \Const \; ``area":\Self \times \Value{*} \rightarrow \Number \times \Nil{*}\}
+\end{align*}
+
+This table type is a subtype of the table type exported by the interface \texttt{Shape},
+and we can use both shortcuts on type annotations, as we show in the following example:
+\begin{verbatim}
+    local Circle = require "circle"
+    local circle1 = Circle:new(0, 5, 10)
+    local circle2:Circle = Circle:new(10, 10, 10)
+    local circle3:Shape = circle1
+    local circle4:Shape = circle2
+    print(circle2:area())                -- 314.15926535898
+    print(circle3:area())                -- compile-time error
+\end{verbatim}
+
+In all examples, if we erase all type and \texttt{const} annotations,
+and type declarations, they become valid Lua code, which run with the
+same semantics as Typed Lua code.
+
+The current version of Typed Lua has some limitations regarding
+the use of \texttt{setmetatable}.
+One limitation is that Typed Lua does not have polymorphism,
+so programmers cannot hide the calls to \texttt{setmetatable} behind
+nicer abstractions, as some Lua libraries do.
+Another limitation is that Typed Lua does not support operator overloading,
+so programmers cannot use \texttt{setmetatable} to change the behavior
+of predefined operations, as some Lua libraries do.
 

### 2014-08-14 doc/thesis/typedlua.tex
Forgot to mention privacy rules and multiple inheritance in the OO section

@@
 
-%Besides interfaces, Typed Lua will also introduce classes.
-%We want to offer classes as a standard way for writing code that
-%Lua programmers are already using, but without tying them to a
-%specific model.
-%Our classes will not include static attributes and will also not offer
-%privacy rules.
-
 Lua provides minimal support for object-oriented programming.
@@
 The current version of Typed Lua has some limitations regarding
-the use of \texttt{setmetatable}.
+the use of \texttt{setmetatable} that are on plans for future work.
 One limitation is that Typed Lua does not have polymorphism,
@@
 of predefined operations, as some Lua libraries do.
+
+Our classes system also does not support multiple inheritance
+and does not offer privacy rules, but these limitations are not
+on plans for future work anyway.
 

### 2014-08-15 doc/thesis/statistics.tex
Reorganizing the chapters of the thesis

--- 2014-08-14 doc/thesis/typedlua.tex
@@
-We begin this chapter presenting statistics about the usage of Lua
+In this chapter we present statistics about the usage of Lua
 features and idioms.
@@
 on our optional type system.
-We end this chapter introducing Typed Lua through examples, and
-describing a few Lua projects that are similar to Typed Lua.
-
-\section{Statistics about the usage of Lua}
-\label{sec:statistics}
 
@@
 functions and vararg expressions.
-
-\section{Typed Lua}
-\label{sec:typedlua}
-
-Typed Lua is an optional type system for Lua, and its main goal is to
-allow programmers to combine static and dynamic typing in Lua.
-In practice, Typed Lua extends the syntax of Lua to introduce
-optional type annotations and class-based object-oriented
-programming through the definition of classes, interfaces, and
-modules.
-In addition, Typed Lua performs local type inference \citep{pierce2000lti}
-to detect more precise types for unannotated expressions.
-Even though the compiler warns the programmer about type errors,
-it always removes the type annotations to generate Lua code that
-runs in unmodified Lua implementations.
-
-An important feature of Typed Lua is its backwards compatibility with Lua,
-that is, unannotated Typed Lua code is valid Lua code.
-For this reason, the syntactic extensions introduced by Typed Lua
-include new keywords that are not reserved words.
-Appendix \ref{app:syntax} presents the complete syntax of Typed Lua.
-
-Another important feature of Typed Lua is its interface between
-statically and dynamically typed modules. 
-This means that the programmer can enjoy some of the benefits of
-static types even without converting existing Lua modules to Typed Lua,
-as a dynamically typed module can export a statically typed interface,
-and statically typed users of the module have their use of the module
-checked by the compiler.
-
-Although we use a gradual type system to formalize our optional type system,
-Typed Lua does not insert run-time checks in the gradual typing style.
-In gradual typing, run-time checks inspect the interaction between
-dynamically typed and statically typed code to guarantee that dynamically
-typed code does not violate statically typed code during run-time.
-A good reason for not inserting run-time checks at this moment is that they
-can decrease run-time performance \citep{allende2013cis}, and we believe that
-a careful evaluation of run-time checks should be done before inserting them
-on the type system.
-
-Unlike some optional type systems, we are designing Typed Lua aiming soundness
-to make it possible to switch Typed Lua from optional typing to gradual typing
-in the future, as a sound type system is the first step to make Typed Lua
-insert run-time checks after static type checking.
-
-In this section we show that Typed Lua is rich enough to preserve some Lua idioms.
-First, we present the abstract syntax of Typed Lua types along with the
-relationship among types.
-Then we use some code examples to present an overview about Typed Lua and
-its syntactical extensions.
-After that we present the formalization of Typed Lua.
-
-\subsection{Type Language}
-\label{sec:types}
-
-Lua values can have one of eight tags:
-\emph{nil}, \emph{boolean}, \emph{number}, \emph{string},
-\emph{function}, \emph{table}, \emph{userdata}, and \emph{thread}.
-In this section we present the types that Typed Lua introduces
-to represent these tags.
-
-\begin{figure}[!ht]
-\textbf{Type Language}\\
-\dstart
-$$
-\begin{array}{rlr}
-\multicolumn{3}{c}{\textbf{First-level types}}\\
-T ::= & \;\; L & \textit{literal types}\\
-& | \; B & \textit{base types}\\
-& | \; \Nil & \textit{nil type}\\
-& | \; \Value & \textit{top type}\\
-& | \; \Any & \textit{dynamic type}\\
-& | \; \Self & \textit{self type}\\
-& | \; T \cup T & \textit{disjoint union types}\\
-& | \; P \rightarrow R & \textit{function types}\\
-& | \; \{ \}_{o|c} & \textit{empty table type}\\
-& | \; \{F, ..., F\}_{o|c} & \textit{table types}\\
-& | \; X & \textit{type variables}\\
-& | \; \mu X.T & \textit{recursive types}\\
-& | \; X_{i} & \textit{projection types}\\
-L ::= & \False \; | \; \True \; | \; {<}{\it number}{>} \; | \; {<}{\it string}{>} & \\
-B ::= & \Boolean \; | \; \Number \; | \; \String & \\
-F ::= & K:T \; | \; \Const \; K:T & \textit{field types}\\ 
-K ::= & L \; | \; B \; | \; \Value \; | \; \Any & \textit{key types}\\
-\multicolumn{3}{c}{}\\
-\multicolumn{3}{c}{\textbf{Second-level types}}\\
-S ::= & \;\; T &\\
-& | \; T* & \textit{vararg types}\\
-& | \; T \times S & \textit{tuple types}\\
-P ::= & \Void \; | \; S & \textit{parameters list type}\\
-R ::= & \Void \; | \; S \; | \; S \cup R & \textit{return types}\\
-\end{array}
-$$
-\dend
-\caption{Typed Lua types}
-\label{fig:typelang}
-\end{figure}
-
-Figure \ref{fig:typelang} presents the abstract syntax of the
-Typed Lua types.
-We split Typed Lua types into two categories:
-\emph{first-level types} and \emph{second-level types}.
-We use first-level types to represent Lua values and
-second-level types to type expression lists, multiple assignments,
-and function applications.
-
-Figure \ref{fig:subtyping} presents the subtyping relation among
-Typed Lua types.
-Any first-level type is a subtype of $\Value$.
-The rest of the subtyping relationship is standard:
-union types are supertypes of their respective literal types,
-function types are related by contravariance on the input
-and covariance on the output, table types have width subtyping,
-with depth subtyping on $\Const$ fields, tuple and vararg types
-are covariant.
-
-\begin{figure}[!ht]
-\textbf{Subtyping}\\
-\dstart
-\begin{footnotesize}
-$$
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-LITERAL}\\
-\senv \vdash L \subtype L
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FALSE}\\
-\senv \vdash \False \subtype \Boolean
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TRUE}\\
-\senv \vdash \True \subtype \Boolean
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-NUMBER}\\
-\senv \vdash {<}{\it number}{>} \subtype \Number
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-STRING}\\
-\senv \vdash {<}{\it string}{>} \subtype \String
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-BASE}\\
-\senv \vdash B \subtype B
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-NIL}\\
-\senv \vdash \Nil \subtype \Nil
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-VALUE}\\
-\senv \vdash T \subtype \Value
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-ANY}\\
-\senv \vdash \Any \subtype \Any
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-SELF}\\
-\senv \vdash \Self \subtype \Self
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-UNION1}\\
-\dfrac{\senv \vdash T_{1} \subtype T \;\;\;
-       \senv \vdash T_{2} \subtype T}
-      {\senv \vdash T_{1} \cup T_{2} \subtype T}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-UNION2}\\
-\dfrac{\senv \vdash T \subtype T_{1}}
-      {\senv \vdash T \subtype T_{1} \cup T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-UNION3}\\
-\dfrac{\senv \vdash T \subtype T_{2}}
-      {\senv \vdash T \subtype T_{1} \cup T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-FUNCTION}\\
-\dfrac{\senv \vdash P_{2} \subtype P_{1} \;\;\;
-       \senv \vdash R_{1} \subtype R_{2}}
-      {\senv \vdash P_{1} \rightarrow R_{1} \subtype P_{2} \rightarrow R_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TABLE1}\\
-\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
-       \senv \vdash F_{j} \subtype F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{c} \subtype \{F_{1}, ..., F_{n}\}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-TABLE2}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash F_{i}' \subtype F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \subtype \{F_{1}', ..., F_{m}', ..., F_{n}'\}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TABLE3}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash F \subtype F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \subtype \{F\}} 
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-FIELD1}\\
-\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
-       \senv \vdash T_{1} \subtype T_{2} \;\;\;
-       \senv \vdash T_{2} \subtype T_{1}}
-      {\senv \vdash K_{1}:T_{1} \subtype K_{2}:T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FIELD2}\\
-\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
-       \senv \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash \Const \; K_{1}:T_{1} \subtype \Const \; K_{2}:T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-FIELD3}\\
-\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
-       \senv \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash K_{1}:T_{1} \subtype \Const \; K_{2}:T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-VARIABLE}\\
-\dfrac{X_{1} \subtype X_{2} \in \senv}
-      {\senv \vdash X_{1} \subtype X_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-RECURSIVE1}\\
-\dfrac{\senv, X_{1} \subtype X_{2} \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash \mu X_{1}.T_{1} \subtype \mu X_{2}.T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-RECURSIVE2}\\
-\dfrac{X \subtype X \not\in \senv \;\;\;
-      \senv, X \subtype X \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash \mu X.T_{1} \subtype T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-RECURSIVE3}\\
-\dfrac{X \subtype X \in \senv \;\;\;
-       \senv \vdash X \subtype T_{2}}
-      {\senv \vdash \mu X.T_{1} \subtype T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-RECURSIVE4}\\
-\dfrac{X \subtype X \not\in \senv \;\;\;
-       \senv, X \subtype X \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash T_{1} \subtype \mu X.T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-RECURSIVE5}\\
-\dfrac{X \subtype X \in \senv \;\;\;
-       \senv \vdash T_{1} \subtype X}
-      {\senv \vdash T_{1} \subtype \mu X.T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-VOID}\\
-\senv \vdash \Void \subtype \Void
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-VARARG}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \cup \Nil}
-      {\senv \vdash T_{1}* \subtype T_{2}*}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TUPLE1}\\
-\dfrac{\senv \vdash T_{1} \subtype T_{2} \;\;\;
-       \senv \vdash S_{1} \subtype S_{2}}
-      {\senv \vdash T_{1} \times S_{1} \subtype T_{2} \times S_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-TUPLE2}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \;\;\;
-       \senv \vdash T_{1}* \subtype S_{2}}
-      {\senv \vdash T_{1}* \subtype T_{2} \times S_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TUPLE3}\\
-\dfrac{\senv \vdash T_{1} \subtype T_{2} \cup \Nil \;\;\;
-       \senv \vdash S_{1} \subtype T_{2}*}
-      {\senv \vdash T_{1} \times S_{1} \subtype T_{2}*}
-\end{array}
-\end{array}
-$$
-\end{footnotesize}
-\dend
-\caption{The subtyping relation}
-\label{fig:subtyping}
-\end{figure}
-
-Figure \ref{fig:consistent_subtyping} presents the consistent-subtyping
-relation that allows the dynamic type $\Any$ interoperate between
-dynamically typed and statically typed code.
-The dynamic type $\Any$ is a subtype of $\Value$, but neither a
-supertype nor a subtype of any other type.
-Our consistent-subtyping relationship follows the standards defined
-by the gradual typing of objects \citep{siek2007objects,siek2013mutable}.
-In practice, we can pass a value of the dynamic type anytime we want
-a value of some other type, and can pass any value where where a
-value of the dynamic type is expected, but these operations are tracked
-by the type system, and the programmer can choose to be warned about them.
-
-\begin{figure}[!ht]
-\textbf{Consistent-Subtyping}\\
-\dstart
-\begin{footnotesize}
-$$
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{C-LITERAL}\\
-\senv \vdash L \lesssim L
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-FALSE}\\
-\senv \vdash \False \lesssim \Boolean
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TRUE}\\
-\senv \vdash \True \lesssim \Boolean
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-NUMBER}\\
-\senv \vdash {<}{\it number}{>} \lesssim \Number
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-STRING}\\
-\senv \vdash {<}{\it string}{>} \lesssim \String
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-BASE}\\
-\senv \vdash B \lesssim B
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-NIL}\\
-\senv \vdash \Nil \lesssim \Nil
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-VALUE}\\
-\senv \vdash T \lesssim \Value
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-ANY1}\\
-\senv \vdash T \lesssim \Any
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-ANY2}\\
-\senv \vdash \Any \lesssim T
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-SELF}\\
-\senv \vdash \Self \lesssim \Self
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-UNION1}\\
-\dfrac{\senv \vdash T_{1} \lesssim T \;\;\;
-       \senv \vdash T_{2} \lesssim T}
-      {\senv \vdash T_{1} \cup T_{2} \lesssim T}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-UNION2}\\
-\dfrac{\senv \vdash T \lesssim T_{1}}
-      {\senv \vdash T \lesssim T_{1} \cup T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-UNION3}\\
-\dfrac{\senv \vdash T \lesssim T_{2}}
-      {\senv \vdash T \lesssim T_{1} \cup T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-FUNCTION}\\
-\dfrac{\senv \vdash P_{2} \lesssim P_{1} \;\;\;
-       \senv \vdash R_{1} \lesssim R_{2}}
-      {\senv \vdash P_{1} \rightarrow R_{1} \lesssim P_{2} \rightarrow R_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TABLE1}\\
-\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
-       \senv \vdash F_{j} \lesssim F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{c} \lesssim \{F_{1}, ..., F_{n}\}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-TABLE2}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash F_{i}' \lesssim F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \lesssim \{F_{1}', ..., F_{m}', ..., F_{n}'\}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TABLE3}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash F \lesssim F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \lesssim \{F\}} 
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-FIELD1}\\
-\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
-       \senv \vdash T_{1} \lesssim T_{2} \;\;\;
-       \senv \vdash T_{2} \lesssim T_{1}}
-      {\senv \vdash K_{1}:T_{1} \lesssim K_{2}:T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-FIELD2}\\
-\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
-       \senv \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash \Const \; K_{1}:T_{1} \lesssim \Const \; K_{2}:T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-FIELD3}\\
-\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
-       \senv \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash K_{1}:T_{1} \lesssim \Const \; K_{2}:T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-VARIABLE}\\
-\dfrac{X_{1} \lesssim X_{2} \in \senv}
-      {\senv \vdash X_{1} \lesssim X_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-RECURSIVE1}\\
-\dfrac{\senv, X_{1} \lesssim X_{2} \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash \mu X_{1}.T_{1} \lesssim \mu X_{2}.T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-RECURSIVE2}\\
-\dfrac{X \lesssim X \not\in \senv \;\;\;
-      \senv, X \lesssim X \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash \mu X.T_{1} \lesssim T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-RECURSIVE3}\\
-\dfrac{X \lesssim X \in \senv \;\;\;
-       \senv \vdash X \lesssim T_{2}}
-      {\senv \vdash \mu X.T_{1} \lesssim T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-RECURSIVE4}\\
-\dfrac{X \lesssim X \not\in \senv \;\;\;
-       \senv, X \lesssim X \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash T_{1} \lesssim \mu X.T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-RECURSIVE5}\\
-\dfrac{X \lesssim X \in \senv \;\;\;
-       \senv \vdash T_{1} \lesssim X}
-      {\senv \vdash T_{1} \lesssim \mu X.T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-VOID}\\
-\senv \vdash \Void \lesssim \Void
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-VARARG}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \lesssim T_{2} \cup \Nil}
-      {\senv \vdash T_{1}* \lesssim T_{2}*}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TUPLE1}\\
-\dfrac{\senv \vdash T_{1} \lesssim T_{2} \;\;\;
-       \senv \vdash S_{1} \lesssim S_{2}}
-      {\senv \vdash T_{1} \times S_{1} \lesssim T_{2} \times S_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-TUPLE2}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \lesssim T_{2} \;\;\;
-       \senv \vdash T_{1}* \lesssim S_{2}}
-      {\senv \vdash T_{1}* \lesssim T_{2} \times S_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TUPLE3}\\
-\dfrac{\senv \vdash T_{1} \lesssim T_{2} \cup \Nil \;\;\;
-       \senv \vdash S_{1} \lesssim T_{2}*}
-      {\senv \vdash T_{1} \times S_{1} \lesssim T_{2}*}
-\end{array}
-\end{array}
-$$
-\end{footnotesize}
-\dend
-\caption{The consistent-subtyping relation}
-\label{fig:consistent_subtyping}
-\end{figure}
-
-\subsection{Typed Lua through examples}
-\label{sec:examples}
-
-In this section we use some examples of Typed Lua code to show how
-they relate to Lua.
-This examples shall give an informal overview of the type system.
-It worth mentioning that we used our implementation to produce and
-test all the examples that we present along this section.
-
-\subsubsection{Atomic types and Functions}
-
-Typed Lua allows optional type annotations in variable and function
-declarations.
-Whenever it is possible to use local type inference, the compiler
-assigns a more specific type to an unannotated declaration.
-When it is not possible to use local type inference, the compiler
-assigns the type $\Any$ to an unannotated declaration.
-For instance, in function declarations, the compiler does not infer
-a more specific type to unannotated input parameters, because,
-in this case, global type inference often is undecidable in a
-language with subtyping and overloading \citep{wells1999typability}.
-
-In the following example we use type annotations in a function
-declaration, and we do not use type annotations in a declaration of a
-local variable:
-\begin{verbatim}
-    local function factorial (n:number):number
-      if n == 0 then
-        return 1
-      else
-        return n * factorial(n - 1)
-      end
-    end
-    local x = 5
-    print(factorial(x))
-\end{verbatim}
-
-Even though we did not add type annotations in the declaration of
-the local variable \texttt{x}, the compiler uses local type inference to
-assign the type $\Number$ to \texttt{x}, and this example compiles
-without any warnings.
-The compiler does not use the literal type $5$ instead of the base
-type $\Number$ because this would prevent
-programmers from assigning other numeric values to the variable
-\texttt{x}.
-The inference that we implement in Typed Lua is quite simple, as it
-uses only the type of the local expression.
-
-Typed Lua allows programmers to combine annotated code with
-unannotated code, as we show in the following example:
-\begin{verbatim}
-    local function abs (n:number):number
-      if n < 0 then
-        return -n
-      else
-        return n
-      end
-    end
-
-    local function dist (x, y)
-      return abs(x - y)
-    end
-\end{verbatim}
-
-The function \texttt{dist} receives two parameters of type $\Any$
-and returns a value of type $\Number$.
-The compiler assigns the dynamic type $\Any$ to the input
-parameters of \texttt{dist} because they do not have type annotations,
-and the compiler does not use global type inference, as we mentioned
-previously.
-Even though we did not annotated the return type of \texttt{dist},
-the compiler is able to infer its return type because it is local
-and not recursive.
-
-We are using the consistent-subtyping relation of gradual
-typing to check the interaction among the dynamic type and the other
-types.
-More precisely, we can call \texttt{abs} inside \texttt{dist} because
-\texttt{x - y} is of type $\Any$, and the type $\Any$ is
-consistent with the type $\Number$.
-The type $\Any$ is not the union of all types, but it is a separate
-type.
-
-In this example, Typed Lua cannot guarantee that \texttt{dist} is never
-going to call \texttt{abs} with a parameter that is not a number,
-because in the semantics of Lua the minus operator can result in a
-value that is not a number.
-This is a typical example where run-time checks would ensure safety
-between the interaction of dynamically typed and statically typed code.
-
-Lua has first-class functions, and they have three peculiarities:
-flexible arity, multiple return values, and multiple assignment.
-In the first case, the number of arguments of a function call does
-not need to match the arity of the function declaration, as Lua silently
-drops extra arguments after evaluating them, or uses \texttt{nil} to
-replace missing arguments.
-In the second case, Lua function can return any number of values,
-and this number of returned values may not be statically known.
-In the third case, Lua has multiple assignment, and the semantics
-of argument passing is the same of the multiple assignment
-(or vice-versa), that is, calling a function is like doing a
-multiple assignment where the left hand side is the parameter list
-and the right hand side is the argument list.
-
-Typed Lua uses \emph{second-level types} to encode functions with
-flexible arity, multiple returns, and multiple assignment.
-We call them second-level because these types do not correspond
-to actual Lua values and we cannot use them to type variables or
-parameters.
-They are just an artifact of the interaction between the type system
-and the semantics of Lua.
-
-As we can see in Figure \ref{fig:typelang}, a second-level type is
-a tuple of first-level types optionally ending in a variadic type,
-or a union of these tuples.
-A variadic type $T{*}$ is a generator for a sequence of values of
-type $T \cup \Nil$.
-Unions of tuples play an important role in functions that are
-overloaded on the return type, together with \emph{projection types}.
-We will explain union types and projection types in more detail
-in the next section.
-
-To match the semantics of Lua function calls, the default operation
-mode of Typed Lua adds a variadic tail to the parts of a function
-type whenever the programmer does not specify one.
-In the previous examples, the type of both \texttt{factorial} and
-\texttt{abs} is
-$\Number \times \Value{*} \rightarrow \Number \times \Nil{*}$,
-and the type of \texttt{dist} is
-$\Any \times \Any \times \Value{*} \rightarrow \Number \times \Nil{*}$.
-If we call \texttt{abs} with extra arguments, Typed Lua silently ignores
-them, as the type signature lets \texttt{abs} receive any number of
-extra arguments.
-If we call \texttt{abs} with fewer arguments, Typed Lua uses $\Nil$ to
-replace the missing arguments and in the case of \texttt{abs} generates
-a warning, as $\Nil$ is neither a subtype of $\Number$ nor consistent
-with $\Number$.
-
-Typed Lua also provides a strict operation mode, and programmers can
-use this mode if they want to catch arity mismatch of function calls.
-In the strict mode, Typed Lua does not add a variadic tail to the
-parts of a function type unless the programmer explicitly declares one.
-In the previous examples, the type of both \texttt{factorial} and
-\texttt{abs} is $\Number \rightarrow \Number$, and the type of
-\texttt{dist} is $\Any \times \Any \rightarrow \Number$.
-If we call \texttt{abs} with either extra or missing arguments,
-Typed Lua generates warnings about arity mismatch.
-
-A variadic type can only appear in the tail position of a tuple,
-because Lua takes only the first value of any expression that appears
-in an expression list that is not in tail position.
-The following example shows the interaction between multiple returns
-and expression lists:
-\begin{verbatim}
-    local function multiple ()
-      return 2, "foo"
-    end
-    local function sum (x:number, y:number)
-      return x + y
-    end
-    local x, y, z = multiple(), multiple()
-    print(sum(multiple(), multiple()))
-\end{verbatim}
-
-In the default mode, the type of \texttt{multiple} is
-$\Value{*} \rightarrow \Number \times \String \times \Nil{*}$,
-and the type of \texttt{sum} is
-$\Number \times \Number \times \Value{*} \rightarrow \Number \times \Nil{*}$.
-In the right hand side of the multiple assignment, only the first
-value produced by the first call to multiple gets used,
-so the type of the right hand side is
-$\Number \times \Number \times \String \times \Nil{*}$,
-and the type assigned to \texttt{x}, \texttt{y}, and \texttt{z}
-are respectively $\Number$, $\Number$, and $\String$.
-The same tuple is assigned to the call to \texttt{sum} and it
-compiles without errors, as the first two components of the tuple
-are consistent with the types of the parameters, and the other
-components are consistent with \texttt{value}.
-
-In the strict mode, the type of \texttt{multiple} is
-$\Void \rightarrow \Number \times \String$,
-and the type of \texttt{sum} is $\Number \times \Number \rightarrow \Number$.
-Like in the default mode, only the first value produced by the
-first call to multiple gets used in the right hand side of the
-multiple assignment;
-and the type of the right hand side is also
-$\Number \times \Number \times \String \times \Nil{*}$,
-because the strict mode affects only the type of function declarations
-and multiple assignments that appear in function calls.
-Unlike in the default mode, the type of the call to \texttt{sum} is
-$\Number \times \Number \times \String$, and does not include the
-variadic tail to catch arity mismatch.
-Thus, this example compiles with error because the call to \texttt{sum}
-passes an unexpected extra argument of type $\String$.
-
-\subsubsection{Unions}
-
-Typed Lua includes union types to encode three common Lua idioms:
-the use of optional values, the overloading based on the tags of
-input parameters, and the overloading on the return type of functions.
-
-Optional values are unions of some type $t$ and $\Nil$, and Typed Lua
-includes the syntactic sugar \texttt{t?} to represent them because
-they appear quite often.
-The concrete syntax \texttt{t?} is syntactic sugar for \texttt{t|nil}
-that maps to the type $T\;\cup\;\Nil$.
-Optional values can appear when a function has optional parameters
-and when the program reads a value from an array or a map.
-The following example shows a function that has an optional parameter:
-\begin{verbatim}
-    local function message (name:string, greeting:string?)
-      local greeting = greeting or "Hello "
-      return greeting .. name
-    end
-
-    print(message("Typed Lua"))         -- prints Typed Lua
-    print(message("Typed Lua", "Hi "))  -- prints Hi Typed Lua
-\end{verbatim}
-
-Although the parameter \texttt{greeting} is optional, and has type
-$\String\;\cup\;\Nil$, the concatenation does not generate a type error
-because we used the short-circuiting \texttt{or} operator to declare a
-new variable \texttt{greeting} that is guaranteed to have type $\String$. 
-In Lua, only the values \texttt{nil} and \texttt{false} represent a
-false condition, so programmers often use the \texttt{or} operator as a
-common idiom for giving a default value to an optional parameter. 
-Typed Lua uses a typing rule to encode this idiom: 
-if the left hand side of \texttt{or} has type $T\;\cup\;\Nil$ and
-the right hand side has type $T$ then the \texttt{or} expression has
-type $T$.
-
-In fact, we do need to declare a new variable \texttt{greeting} that
-shadows the optional parameter:
-\begin{verbatim}
-    local function message (name:string, greeting:string?)
-      greeting = greeting or "Hello "
-      return greeting .. name
-    end
-\end{verbatim}
-
-Typed Lua allows the assignment $x\;=\;x\;\mathbf{or}\;e$ to refine
-the type of $x$ from $T\;\cup\;\Nil$ to $T$ if it matches three restrictions: 
-the type of $e$ is a subtype of $T$, the variable $x$ is local to the
-current function, and it is not being assigned in another function.
-The refinement only affects the type of $x$ in the remainder of the
-current scope.
-In the case of \texttt{greeting}, the assignment refines its type
-from $\String\;\cup\;\Nil$ to $\String$.
-
-Lua programmers often overload the input parameters of functions,
-and use the \texttt{type} function to inspect the tag of the
-input parameters to take different actions depending on what
-those tags are.
-The simplest case overloads on just a single parameter:
-\begin{verbatim}
-    local function overload (s1:string, s2:string|number)
-      if type(s2) == "string" then
-        return s1 .. s2
-      else
-        -- string.rep : (string, number, string) -> (string)
-        return string.rep(s1, s2)
-      end
-    end
-\end{verbatim}
-
-Typed Lua has a small set of \texttt{type} predicates that allows
-programmers to constrain the type of a local variable inside a condition.
-This example uses the \texttt{type(x) == "string"} predicate that
-constrains the type of $x$ from $T\;\cup\;\String$ to $\String$ when
-the predicate is true and $T$ otherwise.
-This is a simplified form of \emph{flow typing} \citep{guha2011tlc,tobin-hochstadt2010ltu}.
-As with \texttt{or}, the variable must be local to the function,
-and it is not being assigned in another function.
-
-The \texttt{type} predicates can only discriminate based on tags,
-so they are limited on the kinds of unions that they can discriminate.
-For instance, the predicates can discriminate a union that combines
-a table type with a base type, or a table type with a function type,
-or two base types, but they cannot discriminate a union that combines
-two different table types, or two different table types.
-
-Lua programmers also overload the return type of functions,
-usually for signaling the occurence of errors.
-In this idiom, a function returns its normal set of return values
-when it successfully finished its execution,
-and usually returns \texttt{nil} as the first value plus an
-error message or other data that describes the error
-when something failed during its execution.
-Next, we show an example:
-\begin{verbatim}
-    local function idiv(dividend:number, divisor:number):
-          (number, number)|(nil, string)
-      if divisor == 0 then
-        return nil, "division by zero"
-      else
-        local r = dividend % divisor
-        local q = (dividend - r) / divisor
-        return q, r
-      end 
-    end
-\end{verbatim}
-
-Typed Lua also includes a syntactic sugar for this idiom:
-we can annotate the return type of \texttt{idiv} with
-\texttt{(number, number)?} to denote the same union.
-The parentheses are always necessary in this case because
-\texttt{t?} is syntactic sugar for \texttt{t|nil},
-while \texttt{(t)?} is syntactic sugar for \texttt{(t)|(nil, string)}.
-
-The type of idiv, in the default mode, is
-$\Number \times \Number \times \Value{*} \rightarrow
-(\Number \times \Number \times \Nil{*}) \cup
-(\Nil \times \String \times \Nil{*})$
-A typical client of this function would use it as follows:
-\begin{verbatim}
-    local q, r = idiv(a, b)
-    if q then
-      print(a == b * q + r)
-    else
-      print("ERROR: " .. r)
-    end
-\end{verbatim}
-
-When Typed Lua finds a union of tuples in the right hand side of
-a declaration, it stores this union in a special type environment
-with a fresh name and assigns \emph{projection types} to the
-variables that appear unannotated in the left hand side of the declaration.
-In our example, Typed Lua assigns the tuple
-$(\Number \times \Number \times \Nil{*}) \cup (\Nil \times \String \times \Nil{*})$
-to a type variable $X$.
-Thus, variable \texttt{q} gets the type $X_{1}$
-that is equivalent to $\Number\;\cup\;\Nil$, 
-and variable \texttt{r} gets the type $X_{2}$
-that is equivalent to $\Number\;\cup\;\String$.
-
-Projection types do not appear in the subtyping and consistent-subtyping
-relations because Typed Lua actually replaces them with the union of the
-corresponding component of each tuple.
-If a variable with a projection type appears in a \texttt{type} predicate,
-it discriminates against all tuples in the union.
-In our example, $X$ is
-$(\Number \times \Number \times \Nil{*}) \cup (\Nil \times \String \times \Nil{*})$
-outside of the \texttt{if} statement,
-but $\Number \times \Number \times \Nil{*}$ inside the \texttt{then} block,
-and $\Nil \times \String \times \Nil{*}$ inside the \texttt{else} block.
-Thus, variable \texttt{q} has type $\Number$ and variable \texttt{r} also
-has type $\Number$ inside the \texttt{then} block;
-and variable \texttt{q} has type $\Nil$ while variable \texttt{r}
-has type $\String$ inside the \texttt{else} block.
-
-We could also have used \texttt{type(q) == "number"} or
-even \texttt{type(r) == "number"} as the predicate of our example,
-as both predicates would produce the same result.
-However, the form that appears in our example is much more succint and idiomatic.
-We can also use projection types to write overloaded functions where
-the type of an input parameter depends on the type of another input parameter.
-
-Typed Lua does not allow assignments to variables that hold a projection type,
-unless the union has been discriminated down to a single value.
-Unrestricted assignment to these variables would be unsound,
-as it could break the dependency relation among the types in each tuple
-that is part of the union.
-
-The overloading mechanims of Typed Lua have a limitation:
-the return type cannot depend on the input type.
-Next, we show an example:
-\begin{verbatim}
-    local function limitation (x:number|string, y:number|string)
-      if type(x) == "number" then
-        if type(y) == "number" then
-          return x + y
-        end
-      else
-        if type(y) == "string" then
-          return x .. y
-        end
-      end
-    end
-\end{verbatim}
-
-This example shows that we cannot write a function that is
-guaranteed to return a number when we pass two numbers and
-guaranteed to return a string when we pass two strings.
-Even though we can use intersection types to express the type
-of this function as
-$(\Number \times \Number \rightarrow \Number) \cap (\String \times \String \rightarrow \String)$,
-and thefore solving the limitation,
-intersection types require more sophisticated flow typing to
-check whether a function has this type,
-and we are still working on this problem.
-
-\subsubsection{Tables and Interfaces}
-
-Tables are the only mechanism that Lua has to build data structures;
-they are associative arrays where any value (except \texttt{nil})
-can be used as a key.
-Programmers can use tables to represent tuples, arrays (dense or sparse),
-records, graphs, modules, objects, etc.
-Lua has syntactic sugar for indexing tables as records:
-\texttt{t.k} is syntactic sugar for \texttt{t["k"]}, and means that
-we are accessing field \texttt{k} from table \texttt{t}. 
-In this section, we show how Typed Lua encodes tables as maps,
-arrays, and records.
-
-Table types are the only framework that Typed Lua has to represent
-the different uses that a Lua table has.
-A table type $\{K_{1}:T_{1}, ..., K_{n}:T_{n}\}$ represents a map
-from values of type $K_{i}$ to values of type $T_{i}$.
-
-Although Typed Lua has just one framework to represent tables,
-it has three different syntaxes for defining the type of a table.
-The first syntax defines table types for maps:
-it is written \texttt{\{k:t\}},
-and maps to the table type $\{K:T\;\cup\;\Nil\}$.
-This table type represents a map that maps values of type
-$K$ to values of type $T\;\cup\;\Nil$,
-as Lua returns $\Nil$ when we use a non-existing key to index a table.
-The second syntax defines table types for arrays:
-it is written \texttt{\{t\}},
-and maps to the table type $\{\Number:T\;\cup\;\Nil\}$.
-This table type is just a syntactic sugar to the definition of a
-map that maps values of type $\Number$ to values of type $T\;\cup\;\Nil$.
-The third syntax defines table types for records:
-it is written \texttt{\{k1:t1, ..., kn:tn\}},
-and maps to the table type $\{K_{1}:T_{1}, ..., K_{n}:T_{n}\}$.
-This table type represents a record that maps the literal types
-$K_{i}$\texttt{, ..., }$K_{n}$ to values of types
-$T_{i}$\texttt{, ..., }$T_{n}$, where $i$ may vary from one to $n$.
-
-Next, we show one example of table type to type a map from strings to numbers:
-\begin{verbatim}
-    local t:{string:number} = { foo = 1 } 
-    local x:number = t.foo               -- compile-time error
-    local y:number? = t.bar              -- y gets nil
-    local z:number = t["bar"] or 0       -- z gets 0 
-\end{verbatim}
-
-The second line of this example gives a type error because we are
-attempting to assign a value of type $\Number\;\cup\;\Nil$ to a
-variable that accepts only values of type $\Number$.
-Although the field \texttt{bar} does not exist in \texttt{t}, the third
-line of this example does not give a type error because the
-annotated type matches the type of the values that can be stored in
-\texttt{t}.
-The last line shows that the \texttt{or} idiom is also useful to
-give a default value to a missing table field.
-
-Next, we show one example of table type to type an array:
-\begin{verbatim}
-    local days:{string} = { "Sunday", "Monday", "Tuesday",
-                            "Wednesday", "Thursday", "Friday",
-                            "Saturday" }
-    local x = days[1]                    -- x gets "Sunday"
-    local y = days[8]                    -- y gets nil
-\end{verbatim}
-
-In this example, the type of both variables \texttt{x} and \texttt{y}
-is $\String\;\cup\;\Nil$.
-An incovenience aspect of making the types of maps and arrays always
-include the type $\Nil$ is to overload the programmers,
-as they need to use \texttt{or} or \texttt{if} to narrow the type of
-the elements they are accessing.
-
-When we know that a list has fixed elements, we can leave the
-variable declaration unannotated and let local type inference assign
-a more specific table type to the variable.
-If we remove the annotation in this example, the compiler uses the
-syntax of records to infer the following table type to \texttt{days}:
-\begin{align*}
-\{{1:\String},\;{2:\String},\;{3:\String},\;{4:\String},\;\\
-{5:\String},\;{6:\String},\;{7:\String}\}
-\end{align*}
-% text about subtyping
-
-When we use the inferred table type while initializing a table, the
-compiler gives an error message when we try to access an index that
-is out of bounds.
-
-We can also use the style of the inferred type to define heterogeneous
-tuples:
-\begin{verbatim}
-    local album = { "Transformer", 1972, "Lou Reed" }
-\end{verbatim}
-
-The next example shows how we can annotate a \texttt{map} function
-that receives another function as its parameter and uses this
-function to alter the elements of a list of numbers, which is
-also a parameter of \texttt{map}:
-\begin{verbatim}
-    local function map (f:(number)->(number), l:{number})
-      for i = 1, #l do
-        l[i] = f(l[i] or 0)
-      end
-    end
-\end{verbatim}
-
-Next, we show one example of table type to type a record:
-\begin{verbatim}
-    local person:{"firstname":string, "lastname":string} =
-      { firstname = "Lou", lastname = "Reed" } 
-\end{verbatim}
-
-Type inference would infer the very same table type that we used to
-annotate the variable \texttt{person}.
-
-Typed Lua includes \emph{interfaces} as syntactic sugar to table types
-that specify records, as writing table types can be unwieldy when
-records get bigger and types of record fields get more complicated.
-For instance, we can use the following interface as an alias to the
-type of the record we showed in the previous example:
-\begin{verbatim}
-    local interface Person
-      firstname:string
-      lastname:string
-    end
-\end{verbatim}
-
-The declaration of the interface \texttt{Person} is an alias to the table type
-$\{``firstname":\String, ``lastname":\String\}$ in the remainder
-of the current scope.
-Now we can use \texttt{Person} in type declarations:
-\begin{verbatim}
-    local function byebye (person:Person):string
-      return "Goodbye " .. person.firstname ..
-             " " .. person.lastname
-    end
-
-    local user1 = { firstname = "Lewis",
-                    middlename = "Allan",
-                    lastname = "Reed" }
-    local user2 = { firstname = "Lou" }
-    local user3 = { lastname = "Reed",
-                    firstname = "Lou" }
-    local user4 = { "Lou", "Reed" }
-
-    print(greeter(user1)) -- Goodbye Lewis Allan
-    print(greeter(user2)) -- compile-time error
-    print(greeter(user3)) -- Goodbye Lou Reed
-    print(greeter(user4)) -- compile-time error
-\end{verbatim}
-
-This example shows that our optional type system is structural rather
-than nominal, that is, it checks the structure of types instead of
-their names, and it uses subtyping and consistent-subtyping for
-checking types.
-
-If our record type has optinal fields, that is, fields that can be \texttt{nil},
-we need to use an explicit type declaration when declaring a
-variable of this record type, as we show in the following example:
-\begin{verbatim}
-    local interface Person
-      firstname:string
-      middlename:string?
-      lastname:string
-    end
-    local user1:Person = { firstname = "Lewis",
-                           middlename = "Allan" or nil,
-                           lastname = "Reed" }
-    local user2:Person = { lastname = "Reed",
-                           firstname = "Lou" }
-\end{verbatim}
-
-We need an explicit type declaration because neither of the types
-that the Typed Lua compiler assigns to the above table constructors
-is a subtype of
-$\{``firstname":\String, ``middlename":\String\cup\Nil, ``lastname":\String\}$,
-the type that \texttt{Person} describes.
-While declaring \texttt{user1}, we use \texttt{or} to cast the type of
-\texttt{middlename} because of the contravariance in the type of
-mutable fields.
-
-We can also use interfaces to define recursive types such as liked lists:
-
-\begin{verbatim}
-    local interface Element
-      info:number
-      next:Element?
-    end
-\end{verbatim}
-
-The declaration of the interface \texttt{Element} is an alias to the
-recursive type
-$\mu Element.\{``info":\Number, ``next":Element \cup \Nil\}$ in the
-remainder of the current scope.
-
-We need to explicitly annotate \texttt{next} being of type
-$Element\;\cup\;\Nil$ because in our type system $\Nil$
-is not the bottom type, that is, $\Nil$ is subtype of itself only.
-Typed Lua has this behavior because in Lua $\Nil$ represents
-the absence of an useful value.
-Thus, making $\Nil$ the bottom type would severaly reduce the amount
-of type safety in the system.
-
-Now we can use \texttt{Element} to annotate a function that
-inserts an element at the begining of a list:
-\begin{verbatim}
-    local function insert (e:Element?, v:number):Element 
-      return { info = v, next = e }
-    end
-\end{verbatim}
-
-We need an explicit type declaration on the return type because
-the Typed Lua compiler cannot infer recursive types, though
-it has subtyping rules to check that the inferred return type
-matches the annotation.
-
-Now we can write a program that declares a list, inserts some
-elements in this list, and then traverses the list printing
-each element of the list:
-\begin{verbatim}
-    local l:Element?
-
-    l = insert(l, 4)
-    l = insert(l, 3)
-    l = insert(l, 2)
-    l = insert(l, 1)
-
-    while l do
-      if l then print(l.info) end
-      l = l.next                  -- compile-time error
-    end
-\end{verbatim}
-
-However, this example has one compile-time error because we
-are attempting to index the variable \texttt{l},
-which can be \texttt{nil}.
-Notice that we would have the same compile-time error while
-printing the list element if we had not narrowed the type of
-\texttt{l}.
-To eliminate this compile-time error we have to write a
-function that narrows the type of \texttt{l} before using it:
-\begin{verbatim}
-    local function get_next (l:Element?)
-      if l then return l.next else return nil end
-    end
-\end{verbatim}
-
-Now we can use \texttt{get\string_next} to rewrite the \texttt{while}
-from the previous example, thus eliminating the compile-time error:
-\begin{verbatim}
-    while l do
-      if l then print(l.info) end
-      l = get_next(l)
-    end
-\end{verbatim}
-
-Lua programmers often build records incrementally, that is,
-they usually declares a local variable with an empty table,
-and then use assignment to add fields to this table:
-\begin{verbatim}
-    local person = {}
-    person.firstname = "Lou"
-    person.lastname = "Reed"
-    print("bye bye " .. person.firstname)  -- bye bye Lou
-\end{verbatim}
-
-In this example, we want to refine the type of \texttt{person} as
-we build the table:
-starting with $\{\}$, and then refining to $\{``firstname":\String\}$,
-and finally reaching $\{``firstname":\String, ``lastname":\String\}$.
-This type change is trickier than the one that we introduced for narrowing
-union types, as we are not just allowing the programmer to change
-the type of the variable \texttt{person}, but we are actually allowing
-the programmer to change the type of the value that \texttt{person} points to.
-This type change is safe in this example, but it is not in the following one:
-\begin{verbatim}
-    local person = {}
-    local bogus = person
-    person.firstname = "Lou"
-    person.lastname = "Reed"
-    bogus.firstname = true               -- compile-time error
-    print("bye bye " .. person.firstname)
-\end{verbatim}
-
-In this example, the fifth line is unsound because it allows the
-programmer to change the type of the value that is stored in
-\texttt{person.firstname}, and makes the last line crash during
-run-time.
-On the first line, we assign an empty table to \texttt{person}.
-On the second line, we assign the type of \texttt{person} to
-\texttt{bogus}.
-On the third and fourth lines we change the type of \texttt{person}
-as we did in the previous example.
-On the fifth line we try to change the type of \texttt{bogus}
-from $\{\}$ to $\{``firstname":\Boolean\}$, but if we allow this
-type change we also allow changing the value that is stored in
-\texttt{person.firstname}, regardless of its type.
-Taking the changes individually looks fine, but the truth is that
-aliasing makes one of them unsound.
-
-The location of the change also matters, as the next example shows:
-\begin{verbatim}
-    local person = { lastname = "Reed" }
-    local function spoil ()
-      person.firstname = nil             -- compile-time error
-    end
-    person.firstname = "Lou"
-    spoil()
-    print("bye bye " .. person.firstname)
-\end{verbatim}
-
-In this example, the third line is unsound because it allows the
-programmer to change the type of \texttt{person} outside of the scope
-that it was declared.
-The call to \texttt{spoil} erases the field \texttt{firstname} from
-\texttt{person}, and makes the last line crash during run-time.
-
-It is also not sound to allow the type of the fields to change,
-as the next example shows:
-\begin{verbatim}
-    local person = { firstname = "Lewis", lastname = "Reed" }
-    person.middlename = "Allan"
-    person.firstname = 1942              -- compile-time error
-    person.lastname = 2013               -- compile-time error
-    print("bye bye " .. person.firstname)
-\end{verbatim}
-
-In this example, both third and fourth line are unsound because
-they are trying to change the type of fields that already exist.
-
-Typed Lua tags a variable that holds a table type as either
-\emph{open} or \emph{closed}.
-If a variable gets its type from a table constructor then it
-is open, otherwise it is always closed.
-
-Typed Lua uses three restrictions to decide whether it is
-sound to allow field assignment to change the type of an
-open variable or not:
-the variable must be local to the current block,
-the new type must be a subtype of the old one, and
-the variable cannot have been assigned in another function.
-
-We can create aliases to open variables, but they are always closed.
-Any mutation on these aliases is not a problem, as the type of
-the original reference can only change towards a subtype.
-For mutable fields this means that the type of the field cannot
-change after it is added to the type of the table.
-
-Table types along with their refinement through field assignment
-are the foundation for defining modules and objects in Typed Lua.
-In the next section, we show that the refinement of table types
-through field assignment is crucial for making Typed Lua
-support idiomatic definition of Lua modules.
-
-\subsubsection{Modules}
-
-Lua does not set policies on how programmers should define modules,
-but it provides mechanisms for organizing a program in modules.
-To load a module, Lua first checks whether the module is already
-loaded or not.
-When the module is not loaded, Lua executes its soure file as a
-function, and the value returned by this function is the module.
-Thus, Lua modules can be any Lua value.
-However, our survey from Section \ref{sec:statistics} confirms that
-the convention among Lua programmers is to use tables for defining modules. 
-In this case, the fields of the table are the functions and other
-values that the module exports.
-
-The simplest way for defining a module in Lua is to declare an
-empty table at the begin of the source file,
-populate this table with the members that the module should export,
-and return this table at the end of the source file. 
-The following example shows how Typed Lua handles this case:
-\begin{verbatim}
-    local mymath = {}
-    local RADIANS_PER_DEGREE = 3.14 / 180.0
-    mymath.deg = function (x:number)
-      return x / RADIANS_PER_DEGREE
-    end
-    mymath.rad = function (x:number)
-      return x * RADIANS_PER_DEGREE
-    end
-    mymath.pow = function (x:number, y:number)
-      return x ^ y
-    end
-    return mymath
-\end{verbatim}
-
-In this example, Typed Lua uses the refinement of table types to
-allow the programmer to incrementally build the type of the
-module \texttt{mymath}.
-Like in Lua, the variable \texttt{RADIANS\string_PER\string_DEGREE}
-is private because we declared it as local to the module. 
-The \texttt{return} at the end of the source file gives the
-final type of the module \texttt{mymath}:
-\begin{align*}
-\{ & ``deg":\Number \times \Value{*} \rightarrow \Number \times \Nil{*},\\
-& ``rad":\Number \times \Value{*} \rightarrow \Number \times \Nil{*},\\
-& ``pow":\Number \times \Number \times \Value{*} \rightarrow \Number \times \Nil{*}\}
-\end{align*}
-
-Another way of defining modules in Lua is to declare only locals
-in the source file and return a table constructor that includes
-the members that the module should export at the end of the source file.
-The following example shows how Typed Lua handles this case:
-\begin{verbatim}
-    local RADIANS_PER_DEGREE = 3.14 / 180.0
-    local function deg (x:number)
-      return x / RADIANS_PER_DEGREE
-    end
-    local function rad (x:number)
-      return x * RADIANS_PER_DEGREE
-    end
-    local function pow (x:number, y:number)
-      return x ^ y
-    end
-    return {
-      deg = deg,
-      rad = rad,
-      pow = pow,
-    }
-\end{verbatim}
-
-In this example, the \texttt{return} at the end of the source file
-also gives the final type of the module \texttt{mymath}, which is
-the same of the previous example.
-
-There is still a third way of defining modules in Lua through the
-\texttt{module} function, which has been deprecated.
-In this style, the first line of the module calls \texttt{module}.
-This function installs a fresh table as the global environment for
-the rest of the module, so any assignments to global variables are
-field assignments to this table.
-The \texttt{module} function also sets an \texttt{\string_M} field
-in this table as a circular reference to the table itself,
-so the module can end with \texttt{return \string_M},
-but this explicit return is not necessary.
-
-Even though this style has been deprecated, our survey from
-Section \ref{sec:statistics} indicated that around a third of the
-Lua modules that are in LuaRocks still use this style.
-Thus, Typed Lua also supports this style:
-\begin{verbatim}
-    module "mymath"
-    local RADIANS_PER_DEGREE = 3.14 / 180.0
-    deg = function (x:number)
-      return x / RADIANS_PER_DEGREE
-    end
-    rad = function (x:number)
-      return x * RADIANS_PER_DEGREE
-    end
-    pow = function (x:number, y:number)
-      return x ^ y
-    end
-\end{verbatim}
-
-In this example, Typed Lua treats accesses to global variables as
-field accesses to an open table in the top-level scope,
-and at the end of file this open table gives the final type of the
-module \texttt{mymath}, which is also the same of the previous two
-examples.
-
-After we define the module \texttt{mymath}, regardless of adopted style,
-users can use it in the standard way:
-\begin{verbatim}
-    local m = require "mymath"
-    print(m.pow(2, 3))                   -- 8
-    print(m.pow(2, "foo"))               -- compile-time error
-\end{verbatim}
-
-And now the users of the module \texttt{mymath} have the policies that
-Typed Lua provides.
-The call to \texttt{require} assigns the type of the module \texttt{mymath}
-to \texttt{m}, so Typed Lua can catch misuses of the module.
-
-The way Typed Lua handle modules using table types is also
-relevant for supporting object-oriented programming, as we discuss in
-the next section.
-
-\subsubsection{Objects and Classes}
-
-Besides interfaces, Typed Lua will also introduce classes.
-We want to offer classes as a standard way for writing code that
-Lua programmers are already using, but without tying them to a
-specific model.
-Our classes will not include static attributes and will also not offer
-privacy rules.
-
-\subsection{The typing rules of Typed Lua}
-\label{sec:rules}
-
-\begin{figure}[!ht]
-\textbf{Abstract Syntax}\\
-\dstart
-$$
-\begin{array}{rl}
-s ::= & \;\; \mathbf{skip} \; | \;
-s_{1} \; ; \; s_{2} \; | \;
-\vec{l} = \vec{e}  \; | \;
-\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
-\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
-& | \; \mathbf{local} \; \vec{n}{:}\vec{T} = \vec{e} \; \mathbf{in} \; s \; | \;
-\mathbf{rec} \; n{:}T = f \; \mathbf{in} \; s \; | \;
-\mathbf{return} \; \vec{e} \; | \;
-a\\
-e ::= & \;\; \mathbf{nil} \; | \;
-\mathbf{false} \; | \;
-\mathbf{true} \; | \;
-{<}{\it number}{>} \; | \;
-{<}{\it string}{>} \; | \;
-{...} \; | \;
-f \; | \;
-\{ \; \vec{c}\; \} \\
-& | \; e_{1} + e_{2} \; | \;
-e_{1} \; {..} \; e_{2} \; | \;
-e_{1} == e_{2} \; | \;
-e_{1} < e_{2} \; | \;
-e_{1} \; \mathbf{and} \; e_{2} \; | \;
-e_{1} \; \mathbf{or} \; e_{2} \\
-& | \; \mathbf{not} \; e \; | \;
-- e \; | \;
-\# \; e \; | \;
-a \; | \;
-l \\
-l ::= & \; n \; | \;
-e_{1}[e_{2}]\\
-a ::= & \;\; e_{1}(\vec{e_{2}}) \; | \;
-e_{1}{:}n(\vec{e_{2}})\\
-f ::= & \; \mathbf{fun} \; (){:}R \; s \; | \;
-\mathbf{fun} \; ({...}{:}T){:}R \; s \; | \;
-\mathbf{fun} \; (\vec{n}{:}\vec{T}){:}R \; s \; | \;
-\mathbf{fun} \; (\vec{n}{:}\vec{T},{...}{:}T){:}R \; s\\
-c ::= & \; [e_{1}] = e_{2}\\
-n ::= & \; {<}{\it name}{>}\\
-\end{array}
-$$
-\dend
-\caption{Typed Lua abstract syntax}
-\label{fig:syntax}
-\end{figure}
-
-Rule for checking if a table type is well formed:
-\[
-\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
-\]
-
-\section{Related Lua projects}
-\label{sec:related}
-
-Metalua \citep{metalua} is a Lua compiler that supports compile-time
-metaprogramming (CTMP).
-CTMP is a kind of macro system that allows the programmers to interact
-with the compiler \citep{fleutot2007contrasting}. 
-Metalua extends Lua 5.1 syntax to include its macro system,
-and allows programmers to define their own syntax.
-Metalua can provide syntactical support for several object-oriented
-styles, and can also provide syntax for turning simple type
-annotations into run-time assertions.
-
-MoonScript \citep{moonscript} is a programming language that supports
-class-based object-oriented programming, and compiles to idiomatic
-Lua code.
-However, it does not perform compile-time type checking.
-
-LuaInspect \citep{luainspect} is a tool that uses MetaLua to perform
-some code analysis, such as flagging unknown global variables and
-table fields, checking the number of function arguments against
-signatures, and inferring function return values, but it does not
-try to analyze object-oriented code and does not perform compile-time
-type checking.
-
-Tidal Lock \citep{tidallock} is a prototype of another optional type
-system for Lua, which is written in Metalua.
-A remarkable feature of Tidal Lock is the way that it handles tables,
-because it allows the programmer to incrementally build the type of
-a table.
-More precisely, Tidal Lock allows the programmer to create an empty
-table to build its type according to the keys that the programmer
-inserts in it, so the programmer does not need to build the table in
-a single table constructor.
-Typed Lua has a simpler form of this feature, with a different
-formalization.
-Like Typed Lua, Tidal Lock also has local type inference.
-Unlike Typed Lua, Tidal Lock does not type the common object-oriented
-idioms used by Lua programmers.

### 2014-08-15 doc/thesis/system.tex
Reorganizing the chapters of the thesis

--- 2014-08-14 doc/thesis/typedlua.tex
@@
-We begin this chapter presenting statistics about the usage of Lua
-features and idioms.
-We collected statistics about how programmers use tables, functions,
-dynamic type checking, object-oriented programming, and modules.
-We shall see that these statistics informed important design decisions
-on our optional type system.
-We end this chapter introducing Typed Lua through examples, and
-describing a few Lua projects that are similar to Typed Lua.
-
-\section{Statistics about the usage of Lua}
-\label{sec:statistics}
-
-We used the LuaRocks repository to build our statistics database;
-LuaRocks \citep{hisham2013luarocks} is a package manager for Lua
-modules.
-We downloaded the 3928 \texttt{.lua} files that were available in
-the LuaRocks repository at February 1st 2014.
-However, we ignored files that are not compatible with Lua 5.2,
-the latest version of Lua.
-We also ignored \emph{machine-generated} files and test files,
-because these files may not represent idiomatic Lua code,
-and might skew our statistics towards non-typical uses of Lua.
-This left 2598 \texttt{.lua} files from 262 different projects for
-our statistics database;
-we parsed these files and processed their abstract syntax tree
-to gather the statistics that we show in this section.
-
-To verify how programmers use tables, we measured how they
-initialize, index, and iterate tables.
-We present these statistics in the next three paragraphs to discuss
-their influence on our type system.
-
-The table constructor appears 23185 times.
-In 36\% of the occurrences it is a constructor that initializes a
-record (e.g., \texttt{\{ x = 120, y = 121 \}});
-in 29\% of the occurrences it is a constructor that initializes a
-list (e.g., \texttt{\{ "one", "two", "three", "four" \}});
-in 8\% of the occurrences it is a constructor that initializes a
-record with a list part;
-and in less than 1\% of the occurrences (4 times) it is a constructor
-that uses only the booleans \texttt{true} and \texttt{false} as indexes.
-At all, in 73\% of the occurrences it is a constructor that uses
-only literal keys;
-in 26\% of the occurrences it is the empty constructor;
-in 1\% of the occurrences it is a constructor with non-literal keys
-only, that is, a constructor that uses variables and function calls
-to create the indexes of a table;
-and in less than 1\% of the occurrences (19 times) it is a constructor
-that mixes literal keys and non-literal keys.
-
-The indexing of tables appears 130448 times:
-86\% of them are for reading a table field while
-14\% of them are for writing into a table field.
-We can classify the indexing operations that are reads as follows:
-89\% of the reads use a literal string key,
-4\% of the reads use a literal number key,
-and less than 1\% of the reads (10 times) use a literal boolean key.
-At all, 93\% of the reads use literals to index a table while
-7\% of the reads use non-literal expressions to index a table.
-It also worth mentioning that 45\% of the reads are actually
-function calls.
-More precisely, 25\% of the reads use literals to call a function,
-20\% of the reads use literals to call a method, that is,
-a function call that uses the colon syntactic sugar, 
-and less than 1\% of the reads (195 times) use non-literal expressions
-to call a function.
-We can also classify the indexing operations that are writes as follows: 
-69\% of the writes use a literal string key,
-2\% of the writes use a literal number key,
-and less than 1\% of the writes (1 time) uses a literal boolean key.
-At all, 71\% of the writes use literals to index a table while
-29\% of the writes use non-literal expressions to index a table.
-
-We also measured how many files have code that iterate over tables to
-observe how frequently iteration is used.
-We observed that 23\% of the files have code that iterate over keys
-of any value, that is, the call to \texttt{pairs} appears at least
-once in these files (the median is twice per file);
-21\% of the files have code that iterate over integer keys, that is,
-the call to \texttt{ipairs} appears at least once in these files
-(the median is also twice per file);
-and 10\% of the files have code that use the numeric \texttt{for}
-along with the length operator (the median is once per file).
-
-The numbers about table initialization, indexing, and iteration
-show us that tables are mostly used to represent records, lists,
-and associative arrays.
-Therefore, Typed Lua should include a table type for handling
-these uses of Lua tables.
-Even though the statistics show that programmers initialize tables
-more often than they use the empty constructor to
-dynamically initialize tables, the statistics of the empty
-constructor are still expressive and indicate that Typed Lua should
-also include a way to handle this style of defining table types.
-
-We found a total of 24858 function declarations in our database
-(the median is six per file).
-Next, we discuss how frequently programmers use dynamic type
-checking and multiple return values inside these functions.
-
-We observed that 9\% of the functions perform dynamic type checking
-on their input parameters, that is, these functions use \texttt{type}
-to inspect the tags of Lua values (the median is once per function).
-We randomly selected 20 functions to sample how programmers are
-using \texttt{type}, and we got the following data:
-50\% of these functions use \texttt{type} for asserting the tags of
-their input parameters, that is, they raise an error when the tag of a
-certain parameter does not match the expected tag, and
-50\% of these functions use \texttt{type} for overloading, that is,
-they execute different code according to the inspected tag.
-
-These numbers show us that Typed Lua should include union types,
-because the use of the \texttt{type} idiom shows that disjoint unions
-will help programmers define data structures that can hold a value of
-several different, but fixed types.
-Typed Lua should also use \texttt{type} as a mechanism for decomposing
-unions, though it may be restricted to base types only.
-
-We observed that 10\% of the functions explicitly return multiple
-values.
-We also observed that 5\% of the functions return \texttt{nil} plus
-something else, for signaling an unexpected behavior;
-and 1\% of the functions return \texttt{false} plus something else,
-also for signaling an unexpected behavior.
-
-Typed Lua should include function types to represent Lua functions,
-and tuple types to represent the signatures of Lua functions,
-multiple return values, and multiple assignments.
-Tuple types will require some special attention, because Typed Lua
-should be able to adjust tuple types during compile-time, in a
-similar way to what Lua does with function calls and multiple
-assignments during run-time.
-In addition, the number of functions that return \texttt{nil} and
-\texttt{false} plus something else show us that overloading on the
-return type will be useful to the type system.
-
-We also measured how frequently programmers use the object-oriented
-paradigm in Lua.
-We observed that 23\% of the function declarations are actually
-method declarations.
-More precisely, 14\% of them use the colon syntactic sugar while
-9\% of them use \texttt{self} as their first parameter.
-We also observed that 63\% of the projects extend tables with
-metatables, that is, they call \texttt{setmetatable} at least once,
-and 27\% of the projects access the metatable of a given table,
-that is, they call \texttt{getmetatable} at least once.
-In fact, 45\% of the projects extend tables with metatables and
-declare methods:
-13\% using the colon syntactic sugar, 14\% using \texttt{self}, and
-18\% using both.
-
-Based on these observations, Typed Lua should include interfaces and
-classes as syntactic sugar for table types that encode objects.
-Proposing an optional type system for Lua that includes classes may
-sound unusual at first glance, because Lua does not have standard
-policies for object-oriented programming.
-However, Lua provides mechanisms that allow programmers to abstract
-their code in terms of objects, and our statistics confirm that
-an expressive number of programmers are relying on these mechanisms to
-use the object-oriented paradigm in Lua.
-Typed Lua should include interfaces and classes as merely syntactic
-mechanisms that help the compiler type check object-oriented code,
-and they should not influence the semantics of Lua.
-
-We also measured how programmers are defining modules.
-We observed that 38\% of the files use the current way of defining
-modules, that is, these files return a table that contains the
-exported members of the module at the end of the file;
-22\% of the files still use the deprecated way of defining modules,
-that is, these files call the function \texttt{module};
-and 1\% of the files use both ways.
-At all, 61\% of the files are modules while 39\% of the files are
-plain scripts.
-The number of plain scripts is high considering the origin of
-our database.
-However, we did not ignore sample scripts, which usually serve to
-help the users of a given module on how to use this module, and
-that is the reason why we have a high number of plain scripts.
-
-Based on these observations, Typed Lua should also use interfaces as
-syntactic sugar for table types that define modules.
-Typed Lua should also support the deprecated style of module
-definition, using global names as exported members of the module.
-
-Besides using interfaces for encoding objects and defining modules,
-Typed Lua should also use interfaces for defining the types of
-userdata, as programmers often use userdata in an object-oriented
-style.
-
-The last statistics that we collected were about variadic functions
-and vararg expressions.
-We observed that 8\% of the functions are variadic, that is,
-their last parameter is the vararg expression.
-We also observed that 5\% of the initialization of lists
-(or 2\% of the occurrences of the table constructor) use solely the
-vararg expression.
-Typed Lua should include a \emph{vararg type} to handle variadic
-functions and vararg expressions.
-
-\section{Typed Lua}
-\label{sec:typedlua}
-
-Typed Lua is an optional type system for Lua, and its main goal is to
-allow programmers to combine static and dynamic typing in Lua.
-In practice, Typed Lua extends the syntax of Lua to introduce
-optional type annotations and class-based object-oriented
-programming through the definition of classes, interfaces, and
-modules.
-In addition, Typed Lua performs local type inference \citep{pierce2000lti}
-to detect more precise types for unannotated expressions.
-Even though the compiler warns the programmer about type errors,
-it always removes the type annotations to generate Lua code that
-runs in unmodified Lua implementations.
-
-An important feature of Typed Lua is its backwards compatibility with Lua,
-that is, unannotated Typed Lua code is valid Lua code.
-For this reason, the syntactic extensions introduced by Typed Lua
-include new keywords that are not reserved words.
-Appendix \ref{app:syntax} presents the complete syntax of Typed Lua.
-
-Another important feature of Typed Lua is its interface between
-statically and dynamically typed modules. 
-This means that the programmer can enjoy some of the benefits of
-static types even without converting existing Lua modules to Typed Lua,
-as a dynamically typed module can export a statically typed interface,
-and statically typed users of the module have their use of the module
-checked by the compiler.
-
-Although we use a gradual type system to formalize our optional type system,
-Typed Lua does not insert run-time checks in the gradual typing style.
-In gradual typing, run-time checks inspect the interaction between
-dynamically typed and statically typed code to guarantee that dynamically
-typed code does not violate statically typed code during run-time.
-A good reason for not inserting run-time checks at this moment is that they
-can decrease run-time performance \citep{allende2013cis}, and we believe that
-a careful evaluation of run-time checks should be done before inserting them
-on the type system.
-
-Unlike some optional type systems, we are designing Typed Lua aiming soundness
-to make it possible to switch Typed Lua from optional typing to gradual typing
-in the future, as a sound type system is the first step to make Typed Lua
-insert run-time checks after static type checking.
-
-In this section we show that Typed Lua is rich enough to preserve some Lua idioms.
-First, we present the abstract syntax of Typed Lua types along with the
-relationship among types.
-Then we use some code examples to present an overview about Typed Lua and
-its syntactical extensions.
-After that we present the formalization of Typed Lua.
-
-\subsection{Type Language}
-\label{sec:types}
-
-Lua values can have one of eight tags:
-\emph{nil}, \emph{boolean}, \emph{number}, \emph{string},
-\emph{function}, \emph{table}, \emph{userdata}, and \emph{thread}.
-In this section we present the types that Typed Lua introduces
-to represent these tags.
-
 \begin{figure}[!ht]
@@
 \dend
-\caption{Typed Lua types}
+\caption{The abstract syntax of Typed Lua types}
 \label{fig:typelang}
@@
 
-Figure \ref{fig:subtyping} presents the subtyping relation among
-Typed Lua types.
-Any first-level type is a subtype of $\Value$.
-The rest of the subtyping relationship is standard:
-union types are supertypes of their respective literal types,
-function types are related by contravariance on the input
-and covariance on the output, table types have width subtyping,
-with depth subtyping on $\Const$ fields, tuple and vararg types
-are covariant.
+Rule for checking if a table type is well formed:
+\[
+\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
+\]
 
@@
 
-Figure \ref{fig:consistent_subtyping} presents the consistent-subtyping
-relation that allows the dynamic type $\Any$ interoperate between
-dynamically typed and statically typed code.
-The dynamic type $\Any$ is a subtype of $\Value$, but neither a
-supertype nor a subtype of any other type.
-Our consistent-subtyping relationship follows the standards defined
-by the gradual typing of objects \citep{siek2007objects,siek2013mutable}.
-In practice, we can pass a value of the dynamic type anytime we want
-a value of some other type, and can pass any value where where a
-value of the dynamic type is expected, but these operations are tracked
-by the type system, and the programmer can choose to be warned about them.
+\begin{figure}[!ht]
+\textbf{Abstract Syntax}\\
+\dstart
+$$
+\begin{array}{rl}
+s ::= & \;\; \mathbf{skip} \; | \;
+s_{1} \; ; \; s_{2} \; | \;
+\vec{l} = \vec{e}  \; | \;
+\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
+\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
+& | \; \mathbf{local} \; \vec{n}{:}\vec{T} = \vec{e} \; \mathbf{in} \; s \; | \;
+\mathbf{rec} \; n{:}T = f \; \mathbf{in} \; s \; | \;
+\mathbf{return} \; \vec{e} \; | \;
+a\\
+e ::= & \;\; \mathbf{nil} \; | \;
+\mathbf{false} \; | \;
+\mathbf{true} \; | \;
+{<}{\it number}{>} \; | \;
+{<}{\it string}{>} \; | \;
+{...} \; | \;
+f \; | \;
+\{ \; \vec{c}\; \} \\
+& | \; e_{1} + e_{2} \; | \;
+e_{1} \; {..} \; e_{2} \; | \;
+e_{1} == e_{2} \; | \;
+e_{1} < e_{2} \; | \;
+e_{1} \; \mathbf{and} \; e_{2} \; | \;
+e_{1} \; \mathbf{or} \; e_{2} \\
+& | \; \mathbf{not} \; e \; | \;
+- e \; | \;
+\# \; e \; | \;
+a \; | \;
+l \\
+l ::= & \; n \; | \;
+e_{1}[e_{2}]\\
+a ::= & \;\; e_{1}(\vec{e_{2}}) \; | \;
+e_{1}{:}n(\vec{e_{2}})\\
+f ::= & \; \mathbf{fun} \; (){:}R \; s \; | \;
+\mathbf{fun} \; ({...}{:}T){:}R \; s \; | \;
+\mathbf{fun} \; (\vec{n}{:}\vec{T}){:}R \; s \; | \;
+\mathbf{fun} \; (\vec{n}{:}\vec{T},{...}{:}T){:}R \; s\\
+c ::= & \; [e_{1}] = e_{2}\\
+n ::= & \; {<}{\it name}{>}\\
+\end{array}
+$$
+\dend
+\caption{Typed Lua abstract syntax}
+\label{fig:syntax}
+\end{figure}
 
@@
 \end{figure}
-
-\subsection{Typed Lua through examples}
-\label{sec:examples}
-
-In this section we use some examples of Typed Lua code to show how
-they relate to Lua.
-This examples shall give an informal overview of the type system.
-It worth mentioning that we used our implementation to produce and
-test all the examples that we present along this section.
-
-\subsubsection{Atomic types and Functions}
-
-Typed Lua allows optional type annotations in variable and function
-declarations.
-Whenever it is possible to use local type inference, the compiler
-assigns a more specific type to an unannotated declaration.
-When it is not possible to use local type inference, the compiler
-assigns the type $\Any$ to an unannotated declaration.
-For instance, in function declarations, the compiler does not infer
-a more specific type to unannotated input parameters, because,
-in this case, global type inference often is undecidable in a
-language with subtyping and overloading \citep{wells1999typability}.
-
-In the following example we use type annotations in a function
-declaration, and we do not use type annotations in a declaration of a
-local variable:
-\begin{verbatim}
-    local function factorial (n:number):number
-      if n == 0 then
-        return 1
-      else
-        return n * factorial(n - 1)
-      end
-    end
-    local x = 5
-    print(factorial(x))
-\end{verbatim}
-
-Even though we did not add type annotations in the declaration of
-the local variable \texttt{x}, the compiler uses local type inference to
-assign the type $\Number$ to \texttt{x}, and this example compiles
-without any warnings.
-The compiler does not use the literal type $5$ instead of the base
-type $\Number$ because this would prevent
-programmers from assigning other numeric values to the variable
-\texttt{x}.
-The inference that we implement in Typed Lua is quite simple, as it
-uses only the type of the local expression.
-
-Typed Lua allows programmers to combine annotated code with
-unannotated code, as we show in the following example:
-\begin{verbatim}
-    local function abs (n:number):number
-      if n < 0 then
-        return -n
-      else
-        return n
-      end
-    end
-
-    local function dist (x, y)
-      return abs(x - y)
-    end
-\end{verbatim}
-
-The function \texttt{dist} receives two parameters of type $\Any$
-and returns a value of type $\Number$.
-The compiler assigns the dynamic type $\Any$ to the input
-parameters of \texttt{dist} because they do not have type annotations,
-and the compiler does not use global type inference, as we mentioned
-previously.
-Even though we did not annotated the return type of \texttt{dist},
-the compiler is able to infer its return type because it is local
-and not recursive.
-
-We are using the consistent-subtyping relation of gradual
-typing to check the interaction among the dynamic type and the other
-types.
-More precisely, we can call \texttt{abs} inside \texttt{dist} because
-\texttt{x - y} is of type $\Any$, and the type $\Any$ is
-consistent with the type $\Number$.
-The type $\Any$ is not the union of all types, but it is a separate
-type.
-
-In this example, Typed Lua cannot guarantee that \texttt{dist} is never
-going to call \texttt{abs} with a parameter that is not a number,
-because in the semantics of Lua the minus operator can result in a
-value that is not a number.
-This is a typical example where run-time checks would ensure safety
-between the interaction of dynamically typed and statically typed code.
-
-Lua has first-class functions, and they have three peculiarities:
-flexible arity, multiple return values, and multiple assignment.
-In the first case, the number of arguments of a function call does
-not need to match the arity of the function declaration, as Lua silently
-drops extra arguments after evaluating them, or uses \texttt{nil} to
-replace missing arguments.
-In the second case, Lua function can return any number of values,
-and this number of returned values may not be statically known.
-In the third case, Lua has multiple assignment, and the semantics
-of argument passing is the same of the multiple assignment
-(or vice-versa), that is, calling a function is like doing a
-multiple assignment where the left hand side is the parameter list
-and the right hand side is the argument list.
-
-Typed Lua uses \emph{second-level types} to encode functions with
-flexible arity, multiple returns, and multiple assignment.
-We call them second-level because these types do not correspond
-to actual Lua values and we cannot use them to type variables or
-parameters.
-They are just an artifact of the interaction between the type system
-and the semantics of Lua.
-
-As we can see in Figure \ref{fig:typelang}, a second-level type is
-a tuple of first-level types optionally ending in a variadic type,
-or a union of these tuples.
-A variadic type $T{*}$ is a generator for a sequence of values of
-type $T \cup \Nil$.
-Unions of tuples play an important role in functions that are
-overloaded on the return type, together with \emph{projection types}.
-We will explain union types and projection types in more detail
-in the next section.
-
-To match the semantics of Lua function calls, the default operation
-mode of Typed Lua adds a variadic tail to the parts of a function
-type whenever the programmer does not specify one.
-In the previous examples, the type of both \texttt{factorial} and
-\texttt{abs} is
-$\Number \times \Value{*} \rightarrow \Number \times \Nil{*}$,
-and the type of \texttt{dist} is
-$\Any \times \Any \times \Value{*} \rightarrow \Number \times \Nil{*}$.
-If we call \texttt{abs} with extra arguments, Typed Lua silently ignores
-them, as the type signature lets \texttt{abs} receive any number of
-extra arguments.
-If we call \texttt{abs} with fewer arguments, Typed Lua uses $\Nil$ to
-replace the missing arguments and in the case of \texttt{abs} generates
-a warning, as $\Nil$ is neither a subtype of $\Number$ nor consistent
-with $\Number$.
-
-Typed Lua also provides a strict operation mode, and programmers can
-use this mode if they want to catch arity mismatch of function calls.
-In the strict mode, Typed Lua does not add a variadic tail to the
-parts of a function type unless the programmer explicitly declares one.
-In the previous examples, the type of both \texttt{factorial} and
-\texttt{abs} is $\Number \rightarrow \Number$, and the type of
-\texttt{dist} is $\Any \times \Any \rightarrow \Number$.
-If we call \texttt{abs} with either extra or missing arguments,
-Typed Lua generates warnings about arity mismatch.
-
-A variadic type can only appear in the tail position of a tuple,
-because Lua takes only the first value of any expression that appears
-in an expression list that is not in tail position.
-The following example shows the interaction between multiple returns
-and expression lists:
-\begin{verbatim}
-    local function multiple ()
-      return 2, "foo"
-    end
-    local function sum (x:number, y:number)
-      return x + y
-    end
-    local x, y, z = multiple(), multiple()
-    print(sum(multiple(), multiple()))
-\end{verbatim}
-
-In the default mode, the type of \texttt{multiple} is
-$\Value{*} \rightarrow \Number \times \String \times \Nil{*}$,
-and the type of \texttt{sum} is
-$\Number \times \Number \times \Value{*} \rightarrow \Number \times \Nil{*}$.
-In the right hand side of the multiple assignment, only the first
-value produced by the first call to multiple gets used,
-so the type of the right hand side is
-$\Number \times \Number \times \String \times \Nil{*}$,
-and the type assigned to \texttt{x}, \texttt{y}, and \texttt{z}
-are respectively $\Number$, $\Number$, and $\String$.
-The same tuple is assigned to the call to \texttt{sum} and it
-compiles without errors, as the first two components of the tuple
-are consistent with the types of the parameters, and the other
-components are consistent with \texttt{value}.
-
-In the strict mode, the type of \texttt{multiple} is
-$\Void \rightarrow \Number \times \String$,
-and the type of \texttt{sum} is $\Number \times \Number \rightarrow \Number$.
-Like in the default mode, only the first value produced by the
-first call to multiple gets used in the right hand side of the
-multiple assignment;
-and the type of the right hand side is also
-$\Number \times \Number \times \String \times \Nil{*}$,
-because the strict mode affects only the type of function declarations
-and multiple assignments that appear in function calls.
-Unlike in the default mode, the type of the call to \texttt{sum} is
-$\Number \times \Number \times \String$, and does not include the
-variadic tail to catch arity mismatch.
-Thus, this example compiles with error because the call to \texttt{sum}
-passes an unexpected extra argument of type $\String$.
-
-\subsubsection{Unions}
-
-Typed Lua includes union types to encode three common Lua idioms:
-the use of optional values, the overloading based on the tags of
-input parameters, and the overloading on the return type of functions.
-
-Optional values are unions of some type $t$ and $\Nil$, and Typed Lua
-includes the syntactic sugar \texttt{t?} to represent them because
-they appear quite often.
-The concrete syntax \texttt{t?} is syntactic sugar for \texttt{t|nil}
-that maps to the type $T\;\cup\;\Nil$.
-Optional values can appear when a function has optional parameters
-and when the program reads a value from an array or a map.
-The following example shows a function that has an optional parameter:
-\begin{verbatim}
-    local function message (name:string, greeting:string?)
-      local greeting = greeting or "Hello "
-      return greeting .. name
-    end
-
-    print(message("Typed Lua"))         -- prints Typed Lua
-    print(message("Typed Lua", "Hi "))  -- prints Hi Typed Lua
-\end{verbatim}
-
-Although the parameter \texttt{greeting} is optional, and has type
-$\String\;\cup\;\Nil$, the concatenation does not generate a type error
-because we used the short-circuiting \texttt{or} operator to declare a
-new variable \texttt{greeting} that is guaranteed to have type $\String$. 
-In Lua, only the values \texttt{nil} and \texttt{false} represent a
-false condition, so programmers often use the \texttt{or} operator as a
-common idiom for giving a default value to an optional parameter. 
-Typed Lua uses a typing rule to encode this idiom: 
-if the left hand side of \texttt{or} has type $T\;\cup\;\Nil$ and
-the right hand side has type $T$ then the \texttt{or} expression has
-type $T$.
-
-In fact, we do need to declare a new variable \texttt{greeting} that
-shadows the optional parameter:
-\begin{verbatim}
-    local function message (name:string, greeting:string?)
-      greeting = greeting or "Hello "
-      return greeting .. name
-    end
-\end{verbatim}
-
-Typed Lua allows the assignment $x\;=\;x\;\mathbf{or}\;e$ to refine
-the type of $x$ from $T\;\cup\;\Nil$ to $T$ if it matches three restrictions: 
-the type of $e$ is a subtype of $T$, the variable $x$ is local to the
-current function, and it is not being assigned in another function.
-The refinement only affects the type of $x$ in the remainder of the
-current scope.
-In the case of \texttt{greeting}, the assignment refines its type
-from $\String\;\cup\;\Nil$ to $\String$.
-
-Lua programmers often overload the input parameters of functions,
-and use the \texttt{type} function to inspect the tag of the
-input parameters to take different actions depending on what
-those tags are.
-The simplest case overloads on just a single parameter:
-\begin{verbatim}
-    local function overload (s1:string, s2:string|number)
-      if type(s2) == "string" then
-        return s1 .. s2
-      else
-        -- string.rep : (string, number, string) -> (string)
-        return string.rep(s1, s2)
-      end
-    end
-\end{verbatim}
-
-Typed Lua has a small set of \texttt{type} predicates that allows
-programmers to constrain the type of a local variable inside a condition.
-This example uses the \texttt{type(x) == "string"} predicate that
-constrains the type of $x$ from $T\;\cup\;\String$ to $\String$ when
-the predicate is true and $T$ otherwise.
-This is a simplified form of \emph{flow typing} \citep{guha2011tlc,tobin-hochstadt2010ltu}.
-As with \texttt{or}, the variable must be local to the function,
-and it is not being assigned in another function.
-
-The \texttt{type} predicates can only discriminate based on tags,
-so they are limited on the kinds of unions that they can discriminate.
-For instance, the predicates can discriminate a union that combines
-a table type with a base type, or a table type with a function type,
-or two base types, but they cannot discriminate a union that combines
-two different table types, or two different table types.
-
-Lua programmers also overload the return type of functions,
-usually for signaling the occurence of errors.
-In this idiom, a function returns its normal set of return values
-when it successfully finished its execution,
-and usually returns \texttt{nil} as the first value plus an
-error message or other data that describes the error
-when something failed during its execution.
-Next, we show an example:
-\begin{verbatim}
-    local function idiv(dividend:number, divisor:number):
-          (number, number)|(nil, string)
-      if divisor == 0 then
-        return nil, "division by zero"
-      else
-        local r = dividend % divisor
-        local q = (dividend - r) / divisor
-        return q, r
-      end 
-    end
-\end{verbatim}
-
-Typed Lua also includes a syntactic sugar for this idiom:
-we can annotate the return type of \texttt{idiv} with
-\texttt{(number, number)?} to denote the same union.
-The parentheses are always necessary in this case because
-\texttt{t?} is syntactic sugar for \texttt{t|nil},
-while \texttt{(t)?} is syntactic sugar for \texttt{(t)|(nil, string)}.
-
-The type of idiv, in the default mode, is
-$\Number \times \Number \times \Value{*} \rightarrow
-(\Number \times \Number \times \Nil{*}) \cup
-(\Nil \times \String \times \Nil{*})$
-A typical client of this function would use it as follows:
-\begin{verbatim}
-    local q, r = idiv(a, b)
-    if q then
-      print(a == b * q + r)
-    else
-      print("ERROR: " .. r)
-    end
-\end{verbatim}
-
-When Typed Lua finds a union of tuples in the right hand side of
-a declaration, it stores this union in a special type environment
-with a fresh name and assigns \emph{projection types} to the
-variables that appear unannotated in the left hand side of the declaration.
-In our example, Typed Lua assigns the tuple
-$(\Number \times \Number \times \Nil{*}) \cup (\Nil \times \String \times \Nil{*})$
-to a type variable $X$.
-Thus, variable \texttt{q} gets the type $X_{1}$
-that is equivalent to $\Number\;\cup\;\Nil$, 
-and variable \texttt{r} gets the type $X_{2}$
-that is equivalent to $\Number\;\cup\;\String$.
-
-Projection types do not appear in the subtyping and consistent-subtyping
-relations because Typed Lua actually replaces them with the union of the
-corresponding component of each tuple.
-If a variable with a projection type appears in a \texttt{type} predicate,
-it discriminates against all tuples in the union.
-In our example, $X$ is
-$(\Number \times \Number \times \Nil{*}) \cup (\Nil \times \String \times \Nil{*})$
-outside of the \texttt{if} statement,
-but $\Number \times \Number \times \Nil{*}$ inside the \texttt{then} block,
-and $\Nil \times \String \times \Nil{*}$ inside the \texttt{else} block.
-Thus, variable \texttt{q} has type $\Number$ and variable \texttt{r} also
-has type $\Number$ inside the \texttt{then} block;
-and variable \texttt{q} has type $\Nil$ while variable \texttt{r}
-has type $\String$ inside the \texttt{else} block.
-
-We could also have used \texttt{type(q) == "number"} or
-even \texttt{type(r) == "number"} as the predicate of our example,
-as both predicates would produce the same result.
-However, the form that appears in our example is much more succint and idiomatic.
-We can also use projection types to write overloaded functions where
-the type of an input parameter depends on the type of another input parameter.
-
-Typed Lua does not allow assignments to variables that hold a projection type,
-unless the union has been discriminated down to a single value.
-Unrestricted assignment to these variables would be unsound,
-as it could break the dependency relation among the types in each tuple
-that is part of the union.
-
-The overloading mechanims of Typed Lua have a limitation:
-the return type cannot depend on the input type.
-Next, we show an example:
-\begin{verbatim}
-    local function limitation (x:number|string, y:number|string)
-      if type(x) == "number" then
-        if type(y) == "number" then
-          return x + y
-        end
-      else
-        if type(y) == "string" then
-          return x .. y
-        end
-      end
-    end
-\end{verbatim}
-
-This example shows that we cannot write a function that is
-guaranteed to return a number when we pass two numbers and
-guaranteed to return a string when we pass two strings.
-Even though we can use intersection types to express the type
-of this function as
-$(\Number \times \Number \rightarrow \Number) \cap (\String \times \String \rightarrow \String)$,
-and thefore solving the limitation,
-intersection types require more sophisticated flow typing to
-check whether a function has this type,
-and we are still working on this problem.
-
-\subsubsection{Tables and Interfaces}
-
-Tables are the only mechanism that Lua has to build data structures;
-they are associative arrays where any value (except \texttt{nil})
-can be used as a key.
-Programmers can use tables to represent tuples, arrays (dense or sparse),
-records, graphs, modules, objects, etc.
-Lua has syntactic sugar for indexing tables as records:
-\texttt{t.k} is syntactic sugar for \texttt{t["k"]}, and means that
-we are accessing field \texttt{k} from table \texttt{t}. 
-In this section, we show how Typed Lua encodes tables as maps,
-arrays, and records.
-
-Table types are the only framework that Typed Lua has to represent
-the different uses that a Lua table has.
-A table type $\{K_{1}:T_{1}, ..., K_{n}:T_{n}\}$ represents a map
-from values of type $K_{i}$ to values of type $T_{i}$.
-
-Although Typed Lua has just one framework to represent tables,
-it has three different syntaxes for defining the type of a table.
-The first syntax defines table types for maps:
-it is written \texttt{\{k:t\}},
-and maps to the table type $\{K:T\;\cup\;\Nil\}$.
-This table type represents a map that maps values of type
-$K$ to values of type $T\;\cup\;\Nil$,
-as Lua returns $\Nil$ when we use a non-existing key to index a table.
-The second syntax defines table types for arrays:
-it is written \texttt{\{t\}},
-and maps to the table type $\{\Number:T\;\cup\;\Nil\}$.
-This table type is just a syntactic sugar to the definition of a
-map that maps values of type $\Number$ to values of type $T\;\cup\;\Nil$.
-The third syntax defines table types for records:
-it is written \texttt{\{k1:t1, ..., kn:tn\}},
-and maps to the table type $\{K_{1}:T_{1}, ..., K_{n}:T_{n}\}$.
-This table type represents a record that maps the literal types
-$K_{i}$\texttt{, ..., }$K_{n}$ to values of types
-$T_{i}$\texttt{, ..., }$T_{n}$, where $i$ may vary from one to $n$.
-
-Next, we show one example of table type to type a map from strings to numbers:
-\begin{verbatim}
-    local t:{string:number} = { foo = 1 } 
-    local x:number = t.foo               -- compile-time error
-    local y:number? = t.bar              -- y gets nil
-    local z:number = t["bar"] or 0       -- z gets 0 
-\end{verbatim}
-
-The second line of this example gives a type error because we are
-attempting to assign a value of type $\Number\;\cup\;\Nil$ to a
-variable that accepts only values of type $\Number$.
-Although the field \texttt{bar} does not exist in \texttt{t}, the third
-line of this example does not give a type error because the
-annotated type matches the type of the values that can be stored in
-\texttt{t}.
-The last line shows that the \texttt{or} idiom is also useful to
-give a default value to a missing table field.
-
-Next, we show one example of table type to type an array:
-\begin{verbatim}
-    local days:{string} = { "Sunday", "Monday", "Tuesday",
-                            "Wednesday", "Thursday", "Friday",
-                            "Saturday" }
-    local x = days[1]                    -- x gets "Sunday"
-    local y = days[8]                    -- y gets nil
-\end{verbatim}
-
-In this example, the type of both variables \texttt{x} and \texttt{y}
-is $\String\;\cup\;\Nil$.
-An incovenience aspect of making the types of maps and arrays always
-include the type $\Nil$ is to overload the programmers,
-as they need to use \texttt{or} or \texttt{if} to narrow the type of
-the elements they are accessing.
-
-When we know that a list has fixed elements, we can leave the
-variable declaration unannotated and let local type inference assign
-a more specific table type to the variable.
-If we remove the annotation in this example, the compiler uses the
-syntax of records to infer the following table type to \texttt{days}:
-\begin{align*}
-\{{1:\String},\;{2:\String},\;{3:\String},\;{4:\String},\;\\
-{5:\String},\;{6:\String},\;{7:\String}\}
-\end{align*}
-% text about subtyping
-
-When we use the inferred table type while initializing a table, the
-compiler gives an error message when we try to access an index that
-is out of bounds.
-
-We can also use the style of the inferred type to define heterogeneous
-tuples:
-\begin{verbatim}
-    local album = { "Transformer", 1972, "Lou Reed" }
-\end{verbatim}
-
-The next example shows how we can annotate a \texttt{map} function
-that receives another function as its parameter and uses this
-function to alter the elements of a list of numbers, which is
-also a parameter of \texttt{map}:
-\begin{verbatim}
-    local function map (f:(number)->(number), l:{number})
-      for i = 1, #l do
-        l[i] = f(l[i] or 0)
-      end
-    end
-\end{verbatim}
-
-Next, we show one example of table type to type a record:
-\begin{verbatim}
-    local person:{"firstname":string, "lastname":string} =
-      { firstname = "Lou", lastname = "Reed" } 
-\end{verbatim}
-
-Type inference would infer the very same table type that we used to
-annotate the variable \texttt{person}.
-
-Typed Lua includes \emph{interfaces} as syntactic sugar to table types
-that specify records, as writing table types can be unwieldy when
-records get bigger and types of record fields get more complicated.
-For instance, we can use the following interface as an alias to the
-type of the record we showed in the previous example:
-\begin{verbatim}
-    local interface Person
-      firstname:string
-      lastname:string
-    end
-\end{verbatim}
-
-The declaration of the interface \texttt{Person} is an alias to the table type
-$\{``firstname":\String, ``lastname":\String\}$ in the remainder
-of the current scope.
-Now we can use \texttt{Person} in type declarations:
-\begin{verbatim}
-    local function byebye (person:Person):string
-      return "Goodbye " .. person.firstname ..
-             " " .. person.lastname
-    end
-
-    local user1 = { firstname = "Lewis",
-                    middlename = "Allan",
-                    lastname = "Reed" }
-    local user2 = { firstname = "Lou" }
-    local user3 = { lastname = "Reed",
-                    firstname = "Lou" }
-    local user4 = { "Lou", "Reed" }
-
-    print(greeter(user1)) -- Goodbye Lewis Allan
-    print(greeter(user2)) -- compile-time error
-    print(greeter(user3)) -- Goodbye Lou Reed
-    print(greeter(user4)) -- compile-time error
-\end{verbatim}
-
-This example shows that our optional type system is structural rather
-than nominal, that is, it checks the structure of types instead of
-their names, and it uses subtyping and consistent-subtyping for
-checking types.
-
-If our record type has optinal fields, that is, fields that can be \texttt{nil},
-we need to use an explicit type declaration when declaring a
-variable of this record type, as we show in the following example:
-\begin{verbatim}
-    local interface Person
-      firstname:string
-      middlename:string?
-      lastname:string
-    end
-    local user1:Person = { firstname = "Lewis",
-                           middlename = "Allan" or nil,
-                           lastname = "Reed" }
-    local user2:Person = { lastname = "Reed",
-                           firstname = "Lou" }
-\end{verbatim}
-
-We need an explicit type declaration because neither of the types
-that the Typed Lua compiler assigns to the above table constructors
-is a subtype of
-$\{``firstname":\String, ``middlename":\String\cup\Nil, ``lastname":\String\}$,
-the type that \texttt{Person} describes.
-While declaring \texttt{user1}, we use \texttt{or} to cast the type of
-\texttt{middlename} because of the contravariance in the type of
-mutable fields.
-
-We can also use interfaces to define recursive types such as liked lists:
-
-\begin{verbatim}
-    local interface Element
-      info:number
-      next:Element?
-    end
-\end{verbatim}
-
-The declaration of the interface \texttt{Element} is an alias to the
-recursive type
-$\mu Element.\{``info":\Number, ``next":Element \cup \Nil\}$ in the
-remainder of the current scope.
-
-We need to explicitly annotate \texttt{next} being of type
-$Element\;\cup\;\Nil$ because in our type system $\Nil$
-is not the bottom type, that is, $\Nil$ is subtype of itself only.
-Typed Lua has this behavior because in Lua $\Nil$ represents
-the absence of an useful value.
-Thus, making $\Nil$ the bottom type would severaly reduce the amount
-of type safety in the system.
-
-Now we can use \texttt{Element} to annotate a function that
-inserts an element at the begining of a list:
-\begin{verbatim}
-    local function insert (e:Element?, v:number):Element 
-      return { info = v, next = e }
-    end
-\end{verbatim}
-
-We need an explicit type declaration on the return type because
-the Typed Lua compiler cannot infer recursive types, though
-it has subtyping rules to check that the inferred return type
-matches the annotation.
-
-Now we can write a program that declares a list, inserts some
-elements in this list, and then traverses the list printing
-each element of the list:
-\begin{verbatim}
-    local l:Element?
-
-    l = insert(l, 4)
-    l = insert(l, 3)
-    l = insert(l, 2)
-    l = insert(l, 1)
-
-    while l do
-      if l then print(l.info) end
-      l = l.next                  -- compile-time error
-    end
-\end{verbatim}
-
-However, this example has one compile-time error because we
-are attempting to index the variable \texttt{l},
-which can be \texttt{nil}.
-Notice that we would have the same compile-time error while
-printing the list element if we had not narrowed the type of
-\texttt{l}.
-To eliminate this compile-time error we have to write a
-function that narrows the type of \texttt{l} before using it:
-\begin{verbatim}
-    local function get_next (l:Element?)
-      if l then return l.next else return nil end
-    end
-\end{verbatim}
-
-Now we can use \texttt{get\string_next} to rewrite the \texttt{while}
-from the previous example, thus eliminating the compile-time error:
-\begin{verbatim}
-    while l do
-      if l then print(l.info) end
-      l = get_next(l)
-    end
-\end{verbatim}
-
-Lua programmers often build records incrementally, that is,
-they usually declares a local variable with an empty table,
-and then use assignment to add fields to this table:
-\begin{verbatim}
-    local person = {}
-    person.firstname = "Lou"
-    person.lastname = "Reed"
-    print("bye bye " .. person.firstname)  -- bye bye Lou
-\end{verbatim}
-
-In this example, we want to refine the type of \texttt{person} as
-we build the table:
-starting with $\{\}$, and then refining to $\{``firstname":\String\}$,
-and finally reaching $\{``firstname":\String, ``lastname":\String\}$.
-This type change is trickier than the one that we introduced for narrowing
-union types, as we are not just allowing the programmer to change
-the type of the variable \texttt{person}, but we are actually allowing
-the programmer to change the type of the value that \texttt{person} points to.
-This type change is safe in this example, but it is not in the following one:
-\begin{verbatim}
-    local person = {}
-    local bogus = person
-    person.firstname = "Lou"
-    person.lastname = "Reed"
-    bogus.firstname = true               -- compile-time error
-    print("bye bye " .. person.firstname)
-\end{verbatim}
-
-In this example, the fifth line is unsound because it allows the
-programmer to change the type of the value that is stored in
-\texttt{person.firstname}, and makes the last line crash during
-run-time.
-On the first line, we assign an empty table to \texttt{person}.
-On the second line, we assign the type of \texttt{person} to
-\texttt{bogus}.
-On the third and fourth lines we change the type of \texttt{person}
-as we did in the previous example.
-On the fifth line we try to change the type of \texttt{bogus}
-from $\{\}$ to $\{``firstname":\Boolean\}$, but if we allow this
-type change we also allow changing the value that is stored in
-\texttt{person.firstname}, regardless of its type.
-Taking the changes individually looks fine, but the truth is that
-aliasing makes one of them unsound.
-
-The location of the change also matters, as the next example shows:
-\begin{verbatim}
-    local person = { lastname = "Reed" }
-    local function spoil ()
-      person.firstname = nil             -- compile-time error
-    end
-    person.firstname = "Lou"
-    spoil()
-    print("bye bye " .. person.firstname)
-\end{verbatim}
-
-In this example, the third line is unsound because it allows the
-programmer to change the type of \texttt{person} outside of the scope
-that it was declared.
-The call to \texttt{spoil} erases the field \texttt{firstname} from
-\texttt{person}, and makes the last line crash during run-time.
-
-It is also not sound to allow the type of the fields to change,
-as the next example shows:
-\begin{verbatim}
-    local person = { firstname = "Lewis", lastname = "Reed" }
-    person.middlename = "Allan"
-    person.firstname = 1942              -- compile-time error
-    person.lastname = 2013               -- compile-time error
-    print("bye bye " .. person.firstname)
-\end{verbatim}
-
-In this example, both third and fourth line are unsound because
-they are trying to change the type of fields that already exist.
-
-Typed Lua tags a variable that holds a table type as either
-\emph{open} or \emph{closed}.
-If a variable gets its type from a table constructor then it
-is open, otherwise it is always closed.
-
-Typed Lua uses three restrictions to decide whether it is
-sound to allow field assignment to change the type of an
-open variable or not:
-the variable must be local to the current block,
-the new type must be a subtype of the old one, and
-the variable cannot have been assigned in another function.
-
-We can create aliases to open variables, but they are always closed.
-Any mutation on these aliases is not a problem, as the type of
-the original reference can only change towards a subtype.
-For mutable fields this means that the type of the field cannot
-change after it is added to the type of the table.
-
-Table types along with their refinement through field assignment
-are the foundation for defining modules and objects in Typed Lua.
-In the next section, we show that the refinement of table types
-through field assignment is crucial for making Typed Lua
-support idiomatic definition of Lua modules.
-
-\subsubsection{Modules}
-
-Lua does not set policies on how programmers should define modules,
-but it provides mechanisms for organizing a program in modules.
-To load a module, Lua first checks whether the module is already
-loaded or not.
-When the module is not loaded, Lua executes its soure file as a
-function, and the value returned by this function is the module.
-Thus, Lua modules can be any Lua value.
-However, our survey from Section \ref{sec:statistics} confirms that
-the convention among Lua programmers is to use tables for defining modules. 
-In this case, the fields of the table are the functions and other
-values that the module exports.
-
-The simplest way for defining a module in Lua is to declare an
-empty table at the begin of the source file,
-populate this table with the members that the module should export,
-and return this table at the end of the source file. 
-The following example shows how Typed Lua handles this case:
-\begin{verbatim}
-    local mymath = {}
-    local RADIANS_PER_DEGREE = 3.14 / 180.0
-    mymath.deg = function (x:number)
-      return x / RADIANS_PER_DEGREE
-    end
-    mymath.rad = function (x:number)
-      return x * RADIANS_PER_DEGREE
-    end
-    mymath.pow = function (x:number, y:number)
-      return x ^ y
-    end
-    return mymath
-\end{verbatim}
-
-In this example, Typed Lua uses the refinement of table types to
-allow the programmer to incrementally build the type of the
-module \texttt{mymath}.
-Like in Lua, the variable \texttt{RADIANS\string_PER\string_DEGREE}
-is private because we declared it as local to the module. 
-The \texttt{return} at the end of the source file gives the
-final type of the module \texttt{mymath}:
-\begin{align*}
-\{ & ``deg":\Number \times \Value{*} \rightarrow \Number \times \Nil{*},\\
-& ``rad":\Number \times \Value{*} \rightarrow \Number \times \Nil{*},\\
-& ``pow":\Number \times \Number \times \Value{*} \rightarrow \Number \times \Nil{*}\}
-\end{align*}
-
-Another way of defining modules in Lua is to declare only locals
-in the source file and return a table constructor that includes
-the members that the module should export at the end of the source file.
-The following example shows how Typed Lua handles this case:
-\begin{verbatim}
-    local RADIANS_PER_DEGREE = 3.14 / 180.0
-    local function deg (x:number)
-      return x / RADIANS_PER_DEGREE
-    end
-    local function rad (x:number)
-      return x * RADIANS_PER_DEGREE
-    end
-    local function pow (x:number, y:number)
-      return x ^ y
-    end
-    return {
-      deg = deg,
-      rad = rad,
-      pow = pow,
-    }
-\end{verbatim}
-
-In this example, the \texttt{return} at the end of the source file
-also gives the final type of the module \texttt{mymath}, which is
-the same of the previous example.
-
-There is still a third way of defining modules in Lua through the
-\texttt{module} function, which has been deprecated.
-In this style, the first line of the module calls \texttt{module}.
-This function installs a fresh table as the global environment for
-the rest of the module, so any assignments to global variables are
-field assignments to this table.
-The \texttt{module} function also sets an \texttt{\string_M} field
-in this table as a circular reference to the table itself,
-so the module can end with \texttt{return \string_M},
-but this explicit return is not necessary.
-
-Even though this style has been deprecated, our survey from
-Section \ref{sec:statistics} indicated that around a third of the
-Lua modules that are in LuaRocks still use this style.
-Thus, Typed Lua also supports this style:
-\begin{verbatim}
-    module "mymath"
-    local RADIANS_PER_DEGREE = 3.14 / 180.0
-    deg = function (x:number)
-      return x / RADIANS_PER_DEGREE
-    end
-    rad = function (x:number)
-      return x * RADIANS_PER_DEGREE
-    end
-    pow = function (x:number, y:number)
-      return x ^ y
-    end
-\end{verbatim}
-
-In this example, Typed Lua treats accesses to global variables as
-field accesses to an open table in the top-level scope,
-and at the end of file this open table gives the final type of the
-module \texttt{mymath}, which is also the same of the previous two
-examples.
-
-After we define the module \texttt{mymath}, regardless of adopted style,
-users can use it in the standard way:
-\begin{verbatim}
-    local m = require "mymath"
-    print(m.pow(2, 3))                   -- 8
-    print(m.pow(2, "foo"))               -- compile-time error
-\end{verbatim}
-
-And now the users of the module \texttt{mymath} have the policies that
-Typed Lua provides.
-The call to \texttt{require} assigns the type of the module \texttt{mymath}
-to \texttt{m}, so Typed Lua can catch misuses of the module.
-
-The way Typed Lua handle modules using table types is also
-relevant for supporting object-oriented programming, as we discuss in
-the next section.
-
-\subsubsection{Objects and Classes}
-
-Besides interfaces, Typed Lua will also introduce classes.
-We want to offer classes as a standard way for writing code that
-Lua programmers are already using, but without tying them to a
-specific model.
-Our classes will not include static attributes and will also not offer
-privacy rules.
-
-\subsection{The typing rules of Typed Lua}
-\label{sec:rules}
-
-\begin{figure}[!ht]
-\textbf{Abstract Syntax}\\
-\dstart
-$$
-\begin{array}{rl}
-s ::= & \;\; \mathbf{skip} \; | \;
-s_{1} \; ; \; s_{2} \; | \;
-\vec{l} = \vec{e}  \; | \;
-\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
-\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
-& | \; \mathbf{local} \; \vec{n}{:}\vec{T} = \vec{e} \; \mathbf{in} \; s \; | \;
-\mathbf{rec} \; n{:}T = f \; \mathbf{in} \; s \; | \;
-\mathbf{return} \; \vec{e} \; | \;
-a\\
-e ::= & \;\; \mathbf{nil} \; | \;
-\mathbf{false} \; | \;
-\mathbf{true} \; | \;
-{<}{\it number}{>} \; | \;
-{<}{\it string}{>} \; | \;
-{...} \; | \;
-f \; | \;
-\{ \; \vec{c}\; \} \\
-& | \; e_{1} + e_{2} \; | \;
-e_{1} \; {..} \; e_{2} \; | \;
-e_{1} == e_{2} \; | \;
-e_{1} < e_{2} \; | \;
-e_{1} \; \mathbf{and} \; e_{2} \; | \;
-e_{1} \; \mathbf{or} \; e_{2} \\
-& | \; \mathbf{not} \; e \; | \;
-- e \; | \;
-\# \; e \; | \;
-a \; | \;
-l \\
-l ::= & \; n \; | \;
-e_{1}[e_{2}]\\
-a ::= & \;\; e_{1}(\vec{e_{2}}) \; | \;
-e_{1}{:}n(\vec{e_{2}})\\
-f ::= & \; \mathbf{fun} \; (){:}R \; s \; | \;
-\mathbf{fun} \; ({...}{:}T){:}R \; s \; | \;
-\mathbf{fun} \; (\vec{n}{:}\vec{T}){:}R \; s \; | \;
-\mathbf{fun} \; (\vec{n}{:}\vec{T},{...}{:}T){:}R \; s\\
-c ::= & \; [e_{1}] = e_{2}\\
-n ::= & \; {<}{\it name}{>}\\
-\end{array}
-$$
-\dend
-\caption{Typed Lua abstract syntax}
-\label{fig:syntax}
-\end{figure}
-
-Rule for checking if a table type is well formed:
-\[
-\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
-\]
-
-\section{Related Lua projects}
-\label{sec:related}
-
-Metalua \citep{metalua} is a Lua compiler that supports compile-time
-metaprogramming (CTMP).
-CTMP is a kind of macro system that allows the programmers to interact
-with the compiler \citep{fleutot2007contrasting}. 
-Metalua extends Lua 5.1 syntax to include its macro system,
-and allows programmers to define their own syntax.
-Metalua can provide syntactical support for several object-oriented
-styles, and can also provide syntax for turning simple type
-annotations into run-time assertions.
-
-MoonScript \citep{moonscript} is a programming language that supports
-class-based object-oriented programming, and compiles to idiomatic
-Lua code.
-However, it does not perform compile-time type checking.
-
-LuaInspect \citep{luainspect} is a tool that uses MetaLua to perform
-some code analysis, such as flagging unknown global variables and
-table fields, checking the number of function arguments against
-signatures, and inferring function return values, but it does not
-try to analyze object-oriented code and does not perform compile-time
-type checking.
-
-Tidal Lock \citep{tidallock} is a prototype of another optional type
-system for Lua, which is written in Metalua.
-A remarkable feature of Tidal Lock is the way that it handles tables,
-because it allows the programmer to incrementally build the type of
-a table.
-More precisely, Tidal Lock allows the programmer to create an empty
-table to build its type according to the keys that the programmer
-inserts in it, so the programmer does not need to build the table in
-a single table constructor.
-Typed Lua has a simpler form of this feature, with a different
-formalization.
-Like Typed Lua, Tidal Lock also has local type inference.
-Unlike Typed Lua, Tidal Lock does not type the common object-oriented
-idioms used by Lua programmers.

### 2014-08-15 doc/thesis/typedlua.tex
Reorganizing the chapters of the thesis

@@
-We begin this chapter presenting statistics about the usage of Lua
-features and idioms.
-We collected statistics about how programmers use tables, functions,
-dynamic type checking, object-oriented programming, and modules.
-We shall see that these statistics informed important design decisions
-on our optional type system.
-We end this chapter introducing Typed Lua through examples, and
-describing a few Lua projects that are similar to Typed Lua.
-
-\section{Statistics about the usage of Lua}
-\label{sec:statistics}
-
-We used the LuaRocks repository to build our statistics database;
-LuaRocks \citep{hisham2013luarocks} is a package manager for Lua
-modules.
-We downloaded the 3928 \texttt{.lua} files that were available in
-the LuaRocks repository at February 1st 2014.
-However, we ignored files that are not compatible with Lua 5.2,
-the latest version of Lua.
-We also ignored \emph{machine-generated} files and test files,
-because these files may not represent idiomatic Lua code,
-and might skew our statistics towards non-typical uses of Lua.
-This left 2598 \texttt{.lua} files from 262 different projects for
-our statistics database;
-we parsed these files and processed their abstract syntax tree
-to gather the statistics that we show in this section.
-
-To verify how programmers use tables, we measured how they
-initialize, index, and iterate tables.
-We present these statistics in the next three paragraphs to discuss
-their influence on our type system.
-
-The table constructor appears 23185 times.
-In 36\% of the occurrences it is a constructor that initializes a
-record (e.g., \texttt{\{ x = 120, y = 121 \}});
-in 29\% of the occurrences it is a constructor that initializes a
-list (e.g., \texttt{\{ "one", "two", "three", "four" \}});
-in 8\% of the occurrences it is a constructor that initializes a
-record with a list part;
-and in less than 1\% of the occurrences (4 times) it is a constructor
-that uses only the booleans \texttt{true} and \texttt{false} as indexes.
-At all, in 73\% of the occurrences it is a constructor that uses
-only literal keys;
-in 26\% of the occurrences it is the empty constructor;
-in 1\% of the occurrences it is a constructor with non-literal keys
-only, that is, a constructor that uses variables and function calls
-to create the indexes of a table;
-and in less than 1\% of the occurrences (19 times) it is a constructor
-that mixes literal keys and non-literal keys.
-
-The indexing of tables appears 130448 times:
-86\% of them are for reading a table field while
-14\% of them are for writing into a table field.
-We can classify the indexing operations that are reads as follows:
-89\% of the reads use a literal string key,
-4\% of the reads use a literal number key,
-and less than 1\% of the reads (10 times) use a literal boolean key.
-At all, 93\% of the reads use literals to index a table while
-7\% of the reads use non-literal expressions to index a table.
-It also worth mentioning that 45\% of the reads are actually
-function calls.
-More precisely, 25\% of the reads use literals to call a function,
-20\% of the reads use literals to call a method, that is,
-a function call that uses the colon syntactic sugar, 
-and less than 1\% of the reads (195 times) use non-literal expressions
-to call a function.
-We can also classify the indexing operations that are writes as follows: 
-69\% of the writes use a literal string key,
-2\% of the writes use a literal number key,
-and less than 1\% of the writes (1 time) uses a literal boolean key.
-At all, 71\% of the writes use literals to index a table while
-29\% of the writes use non-literal expressions to index a table.
-
-We also measured how many files have code that iterate over tables to
-observe how frequently iteration is used.
-We observed that 23\% of the files have code that iterate over keys
-of any value, that is, the call to \texttt{pairs} appears at least
-once in these files (the median is twice per file);
-21\% of the files have code that iterate over integer keys, that is,
-the call to \texttt{ipairs} appears at least once in these files
-(the median is also twice per file);
-and 10\% of the files have code that use the numeric \texttt{for}
-along with the length operator (the median is once per file).
-
-The numbers about table initialization, indexing, and iteration
-show us that tables are mostly used to represent records, lists,
-and associative arrays.
-Therefore, Typed Lua should include a table type for handling
-these uses of Lua tables.
-Even though the statistics show that programmers initialize tables
-more often than they use the empty constructor to
-dynamically initialize tables, the statistics of the empty
-constructor are still expressive and indicate that Typed Lua should
-also include a way to handle this style of defining table types.
-
-We found a total of 24858 function declarations in our database
-(the median is six per file).
-Next, we discuss how frequently programmers use dynamic type
-checking and multiple return values inside these functions.
-
-We observed that 9\% of the functions perform dynamic type checking
-on their input parameters, that is, these functions use \texttt{type}
-to inspect the tags of Lua values (the median is once per function).
-We randomly selected 20 functions to sample how programmers are
-using \texttt{type}, and we got the following data:
-50\% of these functions use \texttt{type} for asserting the tags of
-their input parameters, that is, they raise an error when the tag of a
-certain parameter does not match the expected tag, and
-50\% of these functions use \texttt{type} for overloading, that is,
-they execute different code according to the inspected tag.
-
-These numbers show us that Typed Lua should include union types,
-because the use of the \texttt{type} idiom shows that disjoint unions
-will help programmers define data structures that can hold a value of
-several different, but fixed types.
-Typed Lua should also use \texttt{type} as a mechanism for decomposing
-unions, though it may be restricted to base types only.
-
-We observed that 10\% of the functions explicitly return multiple
-values.
-We also observed that 5\% of the functions return \texttt{nil} plus
-something else, for signaling an unexpected behavior;
-and 1\% of the functions return \texttt{false} plus something else,
-also for signaling an unexpected behavior.
-
-Typed Lua should include function types to represent Lua functions,
-and tuple types to represent the signatures of Lua functions,
-multiple return values, and multiple assignments.
-Tuple types will require some special attention, because Typed Lua
-should be able to adjust tuple types during compile-time, in a
-similar way to what Lua does with function calls and multiple
-assignments during run-time.
-In addition, the number of functions that return \texttt{nil} and
-\texttt{false} plus something else show us that overloading on the
-return type will be useful to the type system.
-
-We also measured how frequently programmers use the object-oriented
-paradigm in Lua.
-We observed that 23\% of the function declarations are actually
-method declarations.
-More precisely, 14\% of them use the colon syntactic sugar while
-9\% of them use \texttt{self} as their first parameter.
-We also observed that 63\% of the projects extend tables with
-metatables, that is, they call \texttt{setmetatable} at least once,
-and 27\% of the projects access the metatable of a given table,
-that is, they call \texttt{getmetatable} at least once.
-In fact, 45\% of the projects extend tables with metatables and
-declare methods:
-13\% using the colon syntactic sugar, 14\% using \texttt{self}, and
-18\% using both.
-
-Based on these observations, Typed Lua should include interfaces and
-classes as syntactic sugar for table types that encode objects.
-Proposing an optional type system for Lua that includes classes may
-sound unusual at first glance, because Lua does not have standard
-policies for object-oriented programming.
-However, Lua provides mechanisms that allow programmers to abstract
-their code in terms of objects, and our statistics confirm that
-an expressive number of programmers are relying on these mechanisms to
-use the object-oriented paradigm in Lua.
-Typed Lua should include interfaces and classes as merely syntactic
-mechanisms that help the compiler type check object-oriented code,
-and they should not influence the semantics of Lua.
-
-We also measured how programmers are defining modules.
-We observed that 38\% of the files use the current way of defining
-modules, that is, these files return a table that contains the
-exported members of the module at the end of the file;
-22\% of the files still use the deprecated way of defining modules,
-that is, these files call the function \texttt{module};
-and 1\% of the files use both ways.
-At all, 61\% of the files are modules while 39\% of the files are
-plain scripts.
-The number of plain scripts is high considering the origin of
-our database.
-However, we did not ignore sample scripts, which usually serve to
-help the users of a given module on how to use this module, and
-that is the reason why we have a high number of plain scripts.
-
-Based on these observations, Typed Lua should also use interfaces as
-syntactic sugar for table types that define modules.
-Typed Lua should also support the deprecated style of module
-definition, using global names as exported members of the module.
-
-Besides using interfaces for encoding objects and defining modules,
-Typed Lua should also use interfaces for defining the types of
-userdata, as programmers often use userdata in an object-oriented
-style.
-
-The last statistics that we collected were about variadic functions
-and vararg expressions.
-We observed that 8\% of the functions are variadic, that is,
-their last parameter is the vararg expression.
-We also observed that 5\% of the initialization of lists
-(or 2\% of the occurrences of the table constructor) use solely the
-vararg expression.
-Typed Lua should include a \emph{vararg type} to handle variadic
-functions and vararg expressions.
-
-\section{Typed Lua}
-\label{sec:typedlua}
-
 Typed Lua is an optional type system for Lua, and its main goal is to
@@
 
-Although we use a gradual type system to formalize our optional type system,
-Typed Lua does not insert run-time checks in the gradual typing style.
+We use the consistent-subtyping relation of gradual typing
+\citep{siek2007objects,siek2013mutable} to formalize Typed Lua,
+though it does not insert run-time checks in the gradual typing style.
 In gradual typing, run-time checks inspect the interaction between
@@
 
-Unlike some optional type systems, we are designing Typed Lua aiming soundness
-to make it possible to switch Typed Lua from optional typing to gradual typing
-in the future, as a sound type system is the first step to make Typed Lua
-insert run-time checks after static type checking.
-
-In this section we show that Typed Lua is rich enough to preserve some Lua idioms.
-First, we present the abstract syntax of Typed Lua types along with the
-relationship among types.
-Then we use some code examples to present an overview about Typed Lua and
-its syntactical extensions.
-After that we present the formalization of Typed Lua.
-
-\subsection{Type Language}
+Unlike Dart \citep{dart} and TypeScript \citep{typescript},
+we are designing Typed Lua aiming soundness to make it possible to switch
+Typed Lua from optional typing to gradual typing in the future,
+as a sound type system is a prerequisite to make Typed Lua insert
+run-time checks after static type checking.
+
+In this chapter we use some examples of Typed Lua code to show how
+they relate to Lua.
+This examples shall give an informal overview of our optional type system.
+We use typing rules to present the formalization of our optional type system
+in Chapter \ref{chap:system}.
+It worth mentioning that we used our implementation to produce and
+test all the examples that we present along this section.
+
+\section{Types}
 \label{sec:types}
@@
 \begin{figure}[!ht]
-\textbf{Type Language}\\
+\textbf{Types}\\
 \dstart
-$$
-\begin{array}{rlr}
-\multicolumn{3}{c}{\textbf{First-level types}}\\
-T ::= & \;\; L & \textit{literal types}\\
-& | \; B & \textit{base types}\\
-& | \; \Nil & \textit{nil type}\\
-& | \; \Value & \textit{top type}\\
-& | \; \Any & \textit{dynamic type}\\
-& | \; \Self & \textit{self type}\\
-& | \; T \cup T & \textit{disjoint union types}\\
-& | \; P \rightarrow R & \textit{function types}\\
-& | \; \{ \}_{o|c} & \textit{empty table type}\\
-& | \; \{F, ..., F\}_{o|c} & \textit{table types}\\
-& | \; X & \textit{type variables}\\
-& | \; \mu X.T & \textit{recursive types}\\
-& | \; X_{i} & \textit{projection types}\\
-L ::= & \False \; | \; \True \; | \; {<}{\it number}{>} \; | \; {<}{\it string}{>} & \\
-B ::= & \Boolean \; | \; \Number \; | \; \String & \\
-F ::= & K:T \; | \; \Const \; K:T & \textit{field types}\\ 
-K ::= & L \; | \; B \; | \; \Value \; | \; \Any & \textit{key types}\\
-\multicolumn{3}{c}{}\\
-\multicolumn{3}{c}{\textbf{Second-level types}}\\
-S ::= & \;\; T &\\
-& | \; T* & \textit{vararg types}\\
-& | \; T \times S & \textit{tuple types}\\
-P ::= & \Void \; | \; S & \textit{parameters list type}\\
-R ::= & \Void \; | \; S \; | \; S \cup R & \textit{return types}\\
-\end{array}
-$$
+\begin{align*}
+\textit{type} & ::= \; \textit{primarytype} \; [\texttt{`?'}]\\
+\textit{primarytype} & ::= \; \textit{literaltype} \; | \;
+  \textit{basetype} \; | \;
+  \textbf{nil} \; | \;
+  \textbf{value} \; | \;
+  \textbf{any} \; | \;
+  \textbf{self} \; | \;
+  \textit{Name}\\
+& | \; \textit{functiontype} \; | \;
+  \textit{tabletype} \; | \;
+  \textit{primarytype} \; \texttt{`|'} \; \textit{primarytype}\\
+\textit{literaltype} & ::= \; \textbf{false} \; | \;
+  \textbf{true} \; | \;
+  \textit{Number} \; | \;
+  \textit{String}\\
+\textit{basetype} & ::= \; \textbf{boolean} \; | \;
+  \textbf{number} \; | \;
+  \textbf{string}\\
+\textit{functiontype} & ::= \; \textit{tupletype} \; \texttt{`->'} \; \textit{rettypelist}\\
+\textit{methodtype} & ::= \; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettypelist}\\
+\textit{tabletype} & ::= \; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
+\textit{tupletype} & ::= \; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
+\textit{typelist} & ::= \; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
+\textit{rettypelist} & ::= \; \textit{unionlist} \; [\texttt{`?'}]\\
+\textit{unionlist} & ::= \; \textit{tupletype} \; | \;
+  \textit{unionlist} \; \texttt{`|'} \; \textit{unionlist}\\
+\textit{tabletypebody} & ::= \; \textit{type} \; | \; \textit{fieldtypelist}\\
+\textit{fieldtypelist} & ::= \; [\textbf{const}] \; \textit{fieldtype} \; \{\texttt{`,'} \; [\textbf{const}] \; \textit{fieldtype}\}\\ 
+\textit{fieldtype} & ::= \; \textit{keytype} \; \texttt{`:'} \; \textit{type}\\
+\textit{keytype} & ::= \; \textit{literaltype} \; | \;
+  \textit{basetype} \; | \;
+  \textbf{value} \; | \;
+  \textbf{any}
+\end{align*}
 \dend
-\caption{Typed Lua types}
-\label{fig:typelang}
+\caption{The concrete syntax of Typed Lua types}
+\label{fig:types}
 \end{figure}
 
-Figure \ref{fig:typelang} presents the abstract syntax of the
+Figure \ref{fig:types} presents the concrete syntax of the
 Typed Lua types.
@@
 
-Figure \ref{fig:subtyping} presents the subtyping relation among
-Typed Lua types.
+Typed Lua uses subtyping to order types.
 Any first-level type is a subtype of $\Value$.
@@
 
-\begin{figure}[!ht]
-\textbf{Subtyping}\\
-\dstart
-\begin{footnotesize}
-$$
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-LITERAL}\\
-\senv \vdash L \subtype L
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FALSE}\\
-\senv \vdash \False \subtype \Boolean
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TRUE}\\
-\senv \vdash \True \subtype \Boolean
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-NUMBER}\\
-\senv \vdash {<}{\it number}{>} \subtype \Number
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-STRING}\\
-\senv \vdash {<}{\it string}{>} \subtype \String
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-BASE}\\
-\senv \vdash B \subtype B
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-NIL}\\
-\senv \vdash \Nil \subtype \Nil
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-VALUE}\\
-\senv \vdash T \subtype \Value
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-ANY}\\
-\senv \vdash \Any \subtype \Any
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-SELF}\\
-\senv \vdash \Self \subtype \Self
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-UNION1}\\
-\dfrac{\senv \vdash T_{1} \subtype T \;\;\;
-       \senv \vdash T_{2} \subtype T}
-      {\senv \vdash T_{1} \cup T_{2} \subtype T}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-UNION2}\\
-\dfrac{\senv \vdash T \subtype T_{1}}
-      {\senv \vdash T \subtype T_{1} \cup T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-UNION3}\\
-\dfrac{\senv \vdash T \subtype T_{2}}
-      {\senv \vdash T \subtype T_{1} \cup T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-FUNCTION}\\
-\dfrac{\senv \vdash P_{2} \subtype P_{1} \;\;\;
-       \senv \vdash R_{1} \subtype R_{2}}
-      {\senv \vdash P_{1} \rightarrow R_{1} \subtype P_{2} \rightarrow R_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TABLE1}\\
-\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
-       \senv \vdash F_{j} \subtype F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{c} \subtype \{F_{1}, ..., F_{n}\}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-TABLE2}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash F_{i}' \subtype F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \subtype \{F_{1}', ..., F_{m}', ..., F_{n}'\}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TABLE3}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash F \subtype F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \subtype \{F\}} 
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-FIELD1}\\
-\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
-       \senv \vdash T_{1} \subtype T_{2} \;\;\;
-       \senv \vdash T_{2} \subtype T_{1}}
-      {\senv \vdash K_{1}:T_{1} \subtype K_{2}:T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FIELD2}\\
-\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
-       \senv \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash \Const \; K_{1}:T_{1} \subtype \Const \; K_{2}:T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-FIELD3}\\
-\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
-       \senv \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash K_{1}:T_{1} \subtype \Const \; K_{2}:T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-VARIABLE}\\
-\dfrac{X_{1} \subtype X_{2} \in \senv}
-      {\senv \vdash X_{1} \subtype X_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-RECURSIVE1}\\
-\dfrac{\senv, X_{1} \subtype X_{2} \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash \mu X_{1}.T_{1} \subtype \mu X_{2}.T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-RECURSIVE2}\\
-\dfrac{X \subtype X \not\in \senv \;\;\;
-      \senv, X \subtype X \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash \mu X.T_{1} \subtype T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-RECURSIVE3}\\
-\dfrac{X \subtype X \in \senv \;\;\;
-       \senv \vdash X \subtype T_{2}}
-      {\senv \vdash \mu X.T_{1} \subtype T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-RECURSIVE4}\\
-\dfrac{X \subtype X \not\in \senv \;\;\;
-       \senv, X \subtype X \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash T_{1} \subtype \mu X.T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-RECURSIVE5}\\
-\dfrac{X \subtype X \in \senv \;\;\;
-       \senv \vdash T_{1} \subtype X}
-      {\senv \vdash T_{1} \subtype \mu X.T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-VOID}\\
-\senv \vdash \Void \subtype \Void
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-VARARG}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \cup \Nil}
-      {\senv \vdash T_{1}* \subtype T_{2}*}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TUPLE1}\\
-\dfrac{\senv \vdash T_{1} \subtype T_{2} \;\;\;
-       \senv \vdash S_{1} \subtype S_{2}}
-      {\senv \vdash T_{1} \times S_{1} \subtype T_{2} \times S_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-TUPLE2}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \;\;\;
-       \senv \vdash T_{1}* \subtype S_{2}}
-      {\senv \vdash T_{1}* \subtype T_{2} \times S_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TUPLE3}\\
-\dfrac{\senv \vdash T_{1} \subtype T_{2} \cup \Nil \;\;\;
-       \senv \vdash S_{1} \subtype T_{2}*}
-      {\senv \vdash T_{1} \times S_{1} \subtype T_{2}*}
-\end{array}
-\end{array}
-$$
-\end{footnotesize}
-\dend
-\caption{The subtyping relation}
-\label{fig:subtyping}
-\end{figure}
-
-Figure \ref{fig:consistent_subtyping} presents the consistent-subtyping
-relation that allows the dynamic type $\Any$ interoperate between
-dynamically typed and statically typed code.
+Typed Lua uses consistent-subtyping to allow the dynamic type $\Any$
+to interoperate between dynamically typed and statically typed code.
 The dynamic type $\Any$ is a subtype of $\Value$, but neither a
@@
 
-\begin{figure}[!ht]
-\textbf{Consistent-Subtyping}\\
-\dstart
-\begin{footnotesize}
-$$
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{C-LITERAL}\\
-\senv \vdash L \lesssim L
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-FALSE}\\
-\senv \vdash \False \lesssim \Boolean
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TRUE}\\
-\senv \vdash \True \lesssim \Boolean
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-NUMBER}\\
-\senv \vdash {<}{\it number}{>} \lesssim \Number
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-STRING}\\
-\senv \vdash {<}{\it string}{>} \lesssim \String
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-BASE}\\
-\senv \vdash B \lesssim B
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-NIL}\\
-\senv \vdash \Nil \lesssim \Nil
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-VALUE}\\
-\senv \vdash T \lesssim \Value
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-ANY1}\\
-\senv \vdash T \lesssim \Any
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-ANY2}\\
-\senv \vdash \Any \lesssim T
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-SELF}\\
-\senv \vdash \Self \lesssim \Self
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-UNION1}\\
-\dfrac{\senv \vdash T_{1} \lesssim T \;\;\;
-       \senv \vdash T_{2} \lesssim T}
-      {\senv \vdash T_{1} \cup T_{2} \lesssim T}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-UNION2}\\
-\dfrac{\senv \vdash T \lesssim T_{1}}
-      {\senv \vdash T \lesssim T_{1} \cup T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-UNION3}\\
-\dfrac{\senv \vdash T \lesssim T_{2}}
-      {\senv \vdash T \lesssim T_{1} \cup T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-FUNCTION}\\
-\dfrac{\senv \vdash P_{2} \lesssim P_{1} \;\;\;
-       \senv \vdash R_{1} \lesssim R_{2}}
-      {\senv \vdash P_{1} \rightarrow R_{1} \lesssim P_{2} \rightarrow R_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TABLE1}\\
-\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
-       \senv \vdash F_{j} \lesssim F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{c} \lesssim \{F_{1}, ..., F_{n}\}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-TABLE2}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash F_{i}' \lesssim F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \lesssim \{F_{1}', ..., F_{m}', ..., F_{n}'\}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TABLE3}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash F \lesssim F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \lesssim \{F\}} 
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-FIELD1}\\
-\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
-       \senv \vdash T_{1} \lesssim T_{2} \;\;\;
-       \senv \vdash T_{2} \lesssim T_{1}}
-      {\senv \vdash K_{1}:T_{1} \lesssim K_{2}:T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-FIELD2}\\
-\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
-       \senv \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash \Const \; K_{1}:T_{1} \lesssim \Const \; K_{2}:T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-FIELD3}\\
-\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
-       \senv \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash K_{1}:T_{1} \lesssim \Const \; K_{2}:T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-VARIABLE}\\
-\dfrac{X_{1} \lesssim X_{2} \in \senv}
-      {\senv \vdash X_{1} \lesssim X_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-RECURSIVE1}\\
-\dfrac{\senv, X_{1} \lesssim X_{2} \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash \mu X_{1}.T_{1} \lesssim \mu X_{2}.T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-RECURSIVE2}\\
-\dfrac{X \lesssim X \not\in \senv \;\;\;
-      \senv, X \lesssim X \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash \mu X.T_{1} \lesssim T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-RECURSIVE3}\\
-\dfrac{X \lesssim X \in \senv \;\;\;
-       \senv \vdash X \lesssim T_{2}}
-      {\senv \vdash \mu X.T_{1} \lesssim T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-RECURSIVE4}\\
-\dfrac{X \lesssim X \not\in \senv \;\;\;
-       \senv, X \lesssim X \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash T_{1} \lesssim \mu X.T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-RECURSIVE5}\\
-\dfrac{X \lesssim X \in \senv \;\;\;
-       \senv \vdash T_{1} \lesssim X}
-      {\senv \vdash T_{1} \lesssim \mu X.T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-VOID}\\
-\senv \vdash \Void \lesssim \Void
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-VARARG}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \lesssim T_{2} \cup \Nil}
-      {\senv \vdash T_{1}* \lesssim T_{2}*}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TUPLE1}\\
-\dfrac{\senv \vdash T_{1} \lesssim T_{2} \;\;\;
-       \senv \vdash S_{1} \lesssim S_{2}}
-      {\senv \vdash T_{1} \times S_{1} \lesssim T_{2} \times S_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-TUPLE2}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \lesssim T_{2} \;\;\;
-       \senv \vdash T_{1}* \lesssim S_{2}}
-      {\senv \vdash T_{1}* \lesssim T_{2} \times S_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TUPLE3}\\
-\dfrac{\senv \vdash T_{1} \lesssim T_{2} \cup \Nil \;\;\;
-       \senv \vdash S_{1} \lesssim T_{2}*}
-      {\senv \vdash T_{1} \times S_{1} \lesssim T_{2}*}
-\end{array}
-\end{array}
-$$
-\end{footnotesize}
-\dend
-\caption{The consistent-subtyping relation}
-\label{fig:consistent_subtyping}
-\end{figure}
-
-\subsection{Typed Lua through examples}
-\label{sec:examples}
-
-In this section we use some examples of Typed Lua code to show how
-they relate to Lua.
-This examples shall give an informal overview of the type system.
-It worth mentioning that we used our implementation to produce and
-test all the examples that we present along this section.
-
-\subsubsection{Atomic types and Functions}
+\section{Atomic types and Functions}
 
@@
 
-\subsubsection{Unions}
+\section{Unions}
 
@@
 
-\subsubsection{Tables and Interfaces}
+\section{Tables and Interfaces}
 
@@
 
-\subsubsection{Modules}
+\section{Modules}
 
@@
 Thus, Lua modules can be any Lua value.
-However, our survey from Section \ref{sec:statistics} confirms that
+However, our survey from Chapter \ref{chapter:statistics} confirms that
 the convention among Lua programmers is to use tables for defining modules. 
@@
 Even though this style has been deprecated, our survey from
-Section \ref{sec:statistics} indicated that around a third of the
+Chapter \ref{chap:statistics} indicated that around a third of the
 Lua modules that are in LuaRocks still use this style.
@@
 
-\subsubsection{Object-Oriented Programming}
+\section{Object-Oriented Programming}
 
@@
 
-\subsection{The typing rules of Typed Lua}
-\label{sec:rules}
-
-\begin{figure}[!ht]
-\textbf{Abstract Syntax}\\
-\dstart
-$$
-\begin{array}{rl}
-s ::= & \;\; \mathbf{skip} \; | \;
-s_{1} \; ; \; s_{2} \; | \;
-\vec{l} = \vec{e}  \; | \;
-\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
-\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
-& | \; \mathbf{local} \; \vec{n}{:}\vec{T} = \vec{e} \; \mathbf{in} \; s \; | \;
-\mathbf{rec} \; n{:}T = f \; \mathbf{in} \; s \; | \;
-\mathbf{return} \; \vec{e} \; | \;
-a\\
-e ::= & \;\; \mathbf{nil} \; | \;
-\mathbf{false} \; | \;
-\mathbf{true} \; | \;
-{<}{\it number}{>} \; | \;
-{<}{\it string}{>} \; | \;
-{...} \; | \;
-f \; | \;
-\{ \; \vec{c}\; \} \\
-& | \; e_{1} + e_{2} \; | \;
-e_{1} \; {..} \; e_{2} \; | \;
-e_{1} == e_{2} \; | \;
-e_{1} < e_{2} \; | \;
-e_{1} \; \mathbf{and} \; e_{2} \; | \;
-e_{1} \; \mathbf{or} \; e_{2} \\
-& | \; \mathbf{not} \; e \; | \;
-- e \; | \;
-\# \; e \; | \;
-a \; | \;
-l \\
-l ::= & \; n \; | \;
-e_{1}[e_{2}]\\
-a ::= & \;\; e_{1}(\vec{e_{2}}) \; | \;
-e_{1}{:}n(\vec{e_{2}})\\
-f ::= & \; \mathbf{fun} \; (){:}R \; s \; | \;
-\mathbf{fun} \; ({...}{:}T){:}R \; s \; | \;
-\mathbf{fun} \; (\vec{n}{:}\vec{T}){:}R \; s \; | \;
-\mathbf{fun} \; (\vec{n}{:}\vec{T},{...}{:}T){:}R \; s\\
-c ::= & \; [e_{1}] = e_{2}\\
-n ::= & \; {<}{\it name}{>}\\
-\end{array}
-$$
-\dend
-\caption{Typed Lua abstract syntax}
-\label{fig:syntax}
-\end{figure}
-
-Rule for checking if a table type is well formed:
-\[
-\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
-\]
-
 \section{Related Lua projects}

### 2014-08-18 doc/thesis/typedlua.tex
Fixing Chapter 3 to show types using concrete syntax instead of abstract syntax

@@
 
-\section{Types}
-\label{sec:types}
+\section{Optional type annotations}
 
@@
 \emph{function}, \emph{table}, \emph{userdata}, and \emph{thread}.
-In this section we present the types that Typed Lua introduces
-to represent these tags.
+Typed Lua includes types for the first six.
+Typed Lua also includes a syntactical extension that programmers can use
+to define the types of \emph{userdata}, and uses it to define the type of \emph{thread}. 
+In this section, we present the Typed Lua types that may appear on annotations.
+We explain all Typed Lua types and syntactical extensions along this chapter.
 
 \begin{figure}[!ht]
-\textbf{Types}\\
+\textbf{Type annotations}\\
 \dstart
@@
   \textbf{string}\\
-\textit{functiontype} & ::= \; \textit{tupletype} \; \texttt{`->'} \; \textit{rettypelist}\\
-\textit{methodtype} & ::= \; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettypelist}\\
+\textit{functiontype} & ::= \; \textit{tupletype} \; \texttt{`->'} \; \textit{rettupletype}\\
+\textit{methodtype} & ::= \; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettupletype}\\
 \textit{tabletype} & ::= \; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
@@
 \textit{typelist} & ::= \; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
-\textit{rettypelist} & ::= \; \textit{unionlist} \; [\texttt{`?'}]\\
+\textit{rettupletype} & ::= \; \textit{unionlist} \; [\texttt{`?'}]\\
 \textit{unionlist} & ::= \; \textit{tupletype} \; | \;
@@
   \textbf{value} \; | \;
-  \textbf{any}
+  \textbf{any}\\
+\textit{rettype} & ::= \; \textit{type} \; | \; \textit{rettupletype}
 \end{align*}
@@
 Figure \ref{fig:types} presents the concrete syntax of the
-Typed Lua types.
-We split Typed Lua types into two categories:
+Typed Lua types in extended BNF.
+We classify Typed Lua types into two categories:
 \emph{first-level types} and \emph{second-level types}.
-We use first-level types to represent Lua values and
-second-level types to type expression lists, multiple assignments,
-and function applications.
+First-level types consist of \emph{type} and represent Lua values,
+while second-level types consist of either \emph{tupletype} or
+\emph{rettype} and represent the type of expression lists,
+multiple assignments, and function applications.
+First-level types include literal types, base types, the type \texttt{nil},
+the top type \texttt{value}, the dynamic type \texttt{any},
+the self type \texttt{self}, named types, function types, table types,
+and union types.
+Second-level types include the type void \texttt{()}, first-level types,
+vararg types, tuple types, and union of tuple types.
 
 Typed Lua uses subtyping to order types.
-Any first-level type is a subtype of $\Value$.
+Any first-level type is a subtype of \texttt{value}.
 The rest of the subtyping relationship is standard:
-union types are supertypes of their respective literal types,
+union types are supertypes of their parts,
+\texttt{boolean}, \texttt{number}, and \texttt{string} are supertypes
+of their respective literal types,
 function types are related by contravariance on the input
 and covariance on the output, table types have width subtyping,
-with depth subtyping on $\Const$ fields, tuple and vararg types
+with depth subtyping on \texttt{const} fields, tuple and vararg types
 are covariant.
 
-Typed Lua uses consistent-subtyping to allow the dynamic type $\Any$
-to interoperate between dynamically typed and statically typed code.
-The dynamic type $\Any$ is a subtype of $\Value$, but neither a
-supertype nor a subtype of any other type.
+Typed Lua uses consistent-subtyping to check the interaction among the
+dynamic type \texttt{any} and other types.
+The dynamic type \texttt{any} is a subtype of \texttt{value}, but neither
+a supertype nor a subtype of any other type.
 Our consistent-subtyping relationship follows the standards defined
@@
 In practice, we can pass a value of the dynamic type anytime we want
-a value of some other type, and can pass any value where where a
-value of the dynamic type is expected, but these operations are tracked
-by the type system, and the programmer can choose to be warned about them.
-
-\section{Atomic types and Functions}
+a value of some other type, and can pass any value where a
+value of the dynamic type is expected, but the compiler tracks these
+operations, and the programmer can choose to be warned about them.
 
@@
 When it is not possible to use local type inference, the compiler
-assigns the type $\Any$ to an unannotated declaration.
+assigns the type \texttt{any} to an unannotated declaration.
 For instance, in function declarations, the compiler does not infer
@@
 
-In the following example we use type annotations in a function
+In the following example, we use type annotations in a function
 declaration, and we do not use type annotations in a declaration of a
@@
 the local variable \texttt{x}, the compiler uses local type inference to
-assign the type $\Number$ to \texttt{x}, and this example compiles
-without any warnings.
-The compiler does not use the literal type $5$ instead of the base
-type $\Number$ because this would prevent
+assign the type \texttt{number} to \texttt{x}, and this example compiles
+without any errors.
+The compiler does not use the literal type \texttt{5},
+instead of the base type \texttt{number}, because this would prevent
 programmers from assigning other numeric values to the variable
@@
 
-Typed Lua allows programmers to combine annotated code with
-unannotated code, as we show in the following example:
-\begin{verbatim}
-    local function abs (n:number):number
+Typed Lua allows programmers to combine statically typed code with
+dynamically typed code, as we show in the following example:
+\begin{verbatim}
+    local function absolute (n:number):number
       if n < 0 then
@@
 
-    local function dist (x, y)
+    local function distance (x, y)
       return abs(x - y)
@@
 
-The function \texttt{dist} receives two parameters of type $\Any$
-and returns a value of type $\Number$.
-The compiler assigns the dynamic type $\Any$ to the input
-parameters of \texttt{dist} because they do not have type annotations,
+The function \texttt{distance} receives two parameters of type \texttt{any}
+and returns a value of type \texttt{number}.
+The compiler assigns the dynamic type \texttt{any} to the input
+parameters of \texttt{distance} because they do not have type annotations,
 and the compiler does not use global type inference, as we mentioned
 previously.
-Even though we did not annotated the return type of \texttt{dist},
+Even though we did not annotated the return type of \texttt{distance},
 the compiler is able to infer its return type because it is local
@@
 
-We are using the consistent-subtyping relation of gradual
-typing to check the interaction among the dynamic type and the other
-types.
-More precisely, we can call \texttt{abs} inside \texttt{dist} because
-\texttt{x - y} is of type $\Any$, and the type $\Any$ is
-consistent with the type $\Number$.
-The type $\Any$ is not the union of all types, but it is a separate
-type.
-
-In this example, Typed Lua cannot guarantee that \texttt{dist} is never
-going to call \texttt{abs} with a parameter that is not a number,
+In this example, Typed Lua cannot guarantee that \texttt{distance} is never
+going to call \texttt{absolute} with a parameter that is not a number,
 because in the semantics of Lua the minus operator can result in a
 value that is not a number.
+More precisely, we can call \texttt{absolute} inside \texttt{distance}
+because the expression \texttt{x - y} has type \texttt{any},
+and it is consistent with type \texttt{number}.
+Still, the dynamic type \texttt{any} may be hiding a value of a type
+that is not a number, making the dynamically typed code crash the
+statically typed code.
 This is a typical example where run-time checks would ensure safety
@@
 
-Lua has first-class functions, and they have three peculiarities:
-flexible arity, multiple return values, and multiple assignment.
-In the first case, the number of arguments of a function call does
+\section{Functions}
+
+Lua has first-class functions, but they have some peculiarities.
+First, the number of arguments of a function call does
 not need to match the arity of the function declaration, as Lua silently
@@
 replace missing arguments.
-In the second case, Lua function can return any number of values,
+Second, functions can return any number of values,
 and this number of returned values may not be statically known.
-In the third case, Lua has multiple assignment, and the semantics
+Third, Lua has multiple assignment, and the semantics
 of argument passing is the same of the multiple assignment
@@
 
-Typed Lua uses \emph{second-level types} to encode functions with
-flexible arity, multiple returns, and multiple assignment.
+Typed Lua uses second-level types to encode function types,
+and to preserve these peculiarities.
 We call them second-level because these types do not correspond
@@
 
-As we can see in Figure \ref{fig:typelang}, a second-level type is
+As we can see in Figure \ref{fig:types}, second-level types consist
+of either \emph{tupletype} or \emph{rettupletype}.
+A second-level type is the type void \texttt{()},
 a tuple of first-level types optionally ending in a variadic type,
 or a union of these tuples.
-A variadic type $T{*}$ is a generator for a sequence of values of
-type $T \cup \Nil$.
-Unions of tuples play an important role in functions that are
-overloaded on the return type, together with \emph{projection types}.
+A variadic type \texttt{t*} is a generator for a sequence of values
+of the union type \texttt{t|nil}.
+Union of tuples may appear only on the return type of functions,
+and together with \emph{projection types} play an important role
+in functions that are overloaded on the return type.
 We will explain union types and projection types in more detail
@@
 In the previous examples, the type of both \texttt{factorial} and
-\texttt{abs} is
-$\Number \times \Value{*} \rightarrow \Number \times \Nil{*}$,
-and the type of \texttt{dist} is
-$\Any \times \Any \times \Value{*} \rightarrow \Number \times \Nil{*}$.
-If we call \texttt{abs} with extra arguments, Typed Lua silently ignores
-them, as the type signature lets \texttt{abs} receive any number of
+\texttt{absolute} is \texttt{(number, value*) -> (number, nil*)},
+and the type of \texttt{distance} is \texttt{(any, any, value*) -> (number, nil*)}.
+If we call \texttt{absolute} with extra arguments, Typed Lua silently ignores
+them, as the type signature lets \texttt{absolute} receive any number of
 extra arguments.
-If we call \texttt{abs} with fewer arguments, Typed Lua uses $\Nil$ to
-replace the missing arguments and in the case of \texttt{abs} generates
-a warning, as $\Nil$ is neither a subtype of $\Number$ nor consistent
-with $\Number$.
+If we call \texttt{absolute} with fewer arguments, Typed Lua uses \texttt{nil} to
+replace the missing arguments and in the case of \texttt{absolute} generates
+an error, as \texttt{nil} is neither a subtype of \texttt{number} nor consistent
+with \texttt{number}.
 
@@
 In the previous examples, the type of both \texttt{factorial} and
-\texttt{abs} is $\Number \rightarrow \Number$, and the type of
-\texttt{dist} is $\Any \times \Any \rightarrow \Number$.
-If we call \texttt{abs} with either extra or missing arguments,
-Typed Lua generates warnings about arity mismatch.
+\texttt{absolute} is \texttt{(number) -> (number)},
+and the type of \texttt{distance} is \texttt{(any, any) -> (number)}.
+If we call \texttt{absolute} with either extra or missing arguments,
+Typed Lua generates errors about arity mismatch.
 
@@
 In the default mode, the type of \texttt{multiple} is
-$\Value{*} \rightarrow \Number \times \String \times \Nil{*}$,
-and the type of \texttt{sum} is
-$\Number \times \Number \times \Value{*} \rightarrow \Number \times \Nil{*}$.
+\texttt{(value*) -> (number, string, nil*)},
+and the type of \texttt{sum} is \texttt{(number, number, value*) -> (number, nil*)}.
 In the right hand side of the multiple assignment, only the first
@@
 so the type of the right hand side is
-$\Number \times \Number \times \String \times \Nil{*}$,
+\texttt{(number, number, string, nil*)},
 and the type assigned to \texttt{x}, \texttt{y}, and \texttt{z}
-are respectively $\Number$, $\Number$, and $\String$.
+are respectively \texttt{number}, \texttt{number}, and \texttt{string}.
 The same tuple is assigned to the call to \texttt{sum} and it
@@
 
-In the strict mode, the type of \texttt{multiple} is
-$\Void \rightarrow \Number \times \String$,
-and the type of \texttt{sum} is $\Number \times \Number \rightarrow \Number$.
+In the strict mode, the type of \texttt{multiple} is \texttt{() -> (number, string)},
+and the type of \texttt{sum} is \texttt{(number, number) -> (number)}.
 Like in the default mode, only the first value produced by the
@@
 and the type of the right hand side is also
-$\Number \times \Number \times \String \times \Nil{*}$,
+\texttt{(number, number, string, nil*)},
 because the strict mode affects only the type of function declarations
@@
 Unlike in the default mode, the type of the call to \texttt{sum} is
-$\Number \times \Number \times \String$, and does not include the
+\texttt{(number, number, string)}, and does not include the
 variadic tail to catch arity mismatch.
 Thus, this example compiles with error because the call to \texttt{sum}
-passes an unexpected extra argument of type $\String$.
+passes an unexpected extra argument of type \texttt{string}.
 
@@
 
-Optional values are unions of some type $t$ and $\Nil$, and Typed Lua
-includes the syntactic sugar \texttt{t?} to represent them because
-they appear quite often.
-The concrete syntax \texttt{t?} is syntactic sugar for \texttt{t|nil}
-that maps to the type $T\;\cup\;\Nil$.
+Optional values are unions of some type \texttt{t} and \texttt{nil},
+and Typed Lua includes the syntactic sugar \texttt{t?} to represent
+them because they appear quite often.
+The concrete syntax \texttt{t?} is syntactic sugar for \texttt{t|nil}.
 Optional values can appear when a function has optional parameters
@@
 
-    print(message("Typed Lua"))         -- prints Typed Lua
-    print(message("Typed Lua", "Hi "))  -- prints Hi Typed Lua
+    print(message("Typed Lua"))         -- Typed Lua
+    print(message("Typed Lua", "Hi "))  -- Hi Typed Lua
 \end{verbatim}
@@
 Although the parameter \texttt{greeting} is optional, and has type
-$\String\;\cup\;\Nil$, the concatenation does not generate a type error
+\texttt{string|nil}, the concatenation does not generate a type error
 because we used the short-circuiting \texttt{or} operator to declare a
-new variable \texttt{greeting} that is guaranteed to have type $\String$. 
+new variable \texttt{greeting} that is guaranteed to have type \texttt{string}. 
 In Lua, only the values \texttt{nil} and \texttt{false} represent a
@@
 Typed Lua uses a typing rule to encode this idiom: 
-if the left hand side of \texttt{or} has type $T\;\cup\;\Nil$ and
-the right hand side has type $T$ then the \texttt{or} expression has
-type $T$.
-
-In fact, we do need to declare a new variable \texttt{greeting} that
+if the left hand side of \texttt{or} has type \texttt{t|nil} and
+the right hand side has type \texttt{t} then the \texttt{or} expression has
+type \texttt{t}.
+
+In fact, we do not need to declare a new variable \texttt{greeting} that
 shadows the optional parameter:
@@
 
-Typed Lua allows the assignment $x\;=\;x\;\mathbf{or}\;e$ to refine
-the type of $x$ from $T\;\cup\;\Nil$ to $T$ if it matches three restrictions: 
-the type of $e$ is a subtype of $T$, the variable $x$ is local to the
-current function, and it is not being assigned in another function.
-The refinement only affects the type of $x$ in the remainder of the
-current scope.
+Typed Lua allows the assignment \texttt{x = x or e} to refine
+the type of \texttt{x} from \texttt{t|nil} to \texttt{t}
+if it matches the following restrictions: 
+the type of \texttt{e} is a subtype of \texttt{t},
+and the variable \texttt{x} is local to the current function
+and it is not being assigned in another function.
+The refinement only affects the type of \texttt{x} in the remainder
+of the current scope.
 In the case of \texttt{greeting}, the assignment refines its type
-from $\String\;\cup\;\Nil$ to $\String$.
+from \texttt{string|nil} to \texttt{string}.
 
@@
 This example uses the \texttt{type(x) == "string"} predicate that
-constrains the type of $x$ from $T\;\cup\;\String$ to $\String$ when
-the predicate is true and $T$ otherwise.
+constrains the type of \texttt{x} from \texttt{t|string} to
+\texttt{string} when the predicate is true and \texttt{t} otherwise.
 This is a simplified form of \emph{flow typing} \citep{guha2011tlc,tobin-hochstadt2010ltu}.
-As with \texttt{or}, the variable must be local to the function,
+As with \texttt{or}, the variable must be local to the function
 and it is not being assigned in another function.
@@
 or two base types, but they cannot discriminate a union that combines
-two different table types, or two different table types.
+two different table types, or two different function types.
 
@@
 \texttt{(number, number)?} to denote the same union.
-The parentheses are always necessary in this case because
+The parentheses are always necessary in this case, because
 \texttt{t?} is syntactic sugar for \texttt{t|nil},
@@
 
-The type of idiv, in the default mode, is
-$\Number \times \Number \times \Value{*} \rightarrow
-(\Number \times \Number \times \Nil{*}) \cup
-(\Nil \times \String \times \Nil{*})$
 A typical client of this function would use it as follows:
@@
 variables that appear unannotated in the left hand side of the declaration.
-In our example, Typed Lua assigns the tuple
-$(\Number \times \Number \times \Nil{*}) \cup (\Nil \times \String \times \Nil{*})$
-to a type variable $X$.
+In our example, Typed Lua assigns the union of tuples
+\texttt{(number, number, nil*)|(nil, string, nil*)} to a type variable $X$.
 Thus, variable \texttt{q} gets the type $X_{1}$
-that is equivalent to $\Number\;\cup\;\Nil$, 
+that is equivalent to \texttt{number|nil},
 and variable \texttt{r} gets the type $X_{2}$
-that is equivalent to $\Number\;\cup\;\String$.
+that is equivalent to \texttt{number|string}.
 
@@
 it discriminates against all tuples in the union.
-In our example, $X$ is
-$(\Number \times \Number \times \Nil{*}) \cup (\Nil \times \String \times \Nil{*})$
+In our example, $X$ is \texttt{(number, number, nil*)|(nil, string, nil*)}
 outside of the \texttt{if} statement,
-but $\Number \times \Number \times \Nil{*}$ inside the \texttt{then} block,
-and $\Nil \times \String \times \Nil{*}$ inside the \texttt{else} block.
-Thus, variable \texttt{q} has type $\Number$ and variable \texttt{r} also
-has type $\Number$ inside the \texttt{then} block;
-and variable \texttt{q} has type $\Nil$ while variable \texttt{r}
-has type $\String$ inside the \texttt{else} block.
+but \texttt{(number, number, nil*)} inside the \texttt{then} block,
+and \texttt{(nil, string, nil*)} inside the \texttt{else} block.
+Thus, variable \texttt{q} has type \texttt{number} and variable \texttt{r} also
+has type \texttt{number} inside the \texttt{then} block;
+and variable \texttt{q} has type \texttt{nil} while variable \texttt{r}
+has type \texttt{string} inside the \texttt{else} block.
 
@@
 However, the form that appears in our example is much more succint and idiomatic.
-We can also use projection types to write overloaded functions where
-the type of an input parameter depends on the type of another input parameter.
 
@@
 of this function as
-$(\Number \times \Number \rightarrow \Number) \cap (\String \times \String \rightarrow \String)$,
-and thefore solving the limitation,
+\texttt{(number, number, value*) -> (number, nil*) \string&
+(string, string, value*) -> (string, nil*)},
+and thefore solving this limitation,
 intersection types require more sophisticated flow typing to
@@
 \texttt{t.k} is syntactic sugar for \texttt{t["k"]}, and means that
-we are accessing field \texttt{k} from table \texttt{t}. 
+we are accessing the field \texttt{k} from table \texttt{t}. 
 In this section, we show how Typed Lua encodes tables as maps,
@@
 the different uses that a Lua table has.
-A table type $\{K_{1}:T_{1}, ..., K_{n}:T_{n}\}$ represents a map
-from values of type $K_{i}$ to values of type $T_{i}$.
+A table type $\texttt{\{k}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
+\texttt{, ..., k}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}$ represents a map
+from values of type $\texttt{k}_{\texttt{i}}$ to values of type
+$\texttt{t}_{\texttt{i}}$.
 
 Although Typed Lua has just one framework to represent tables,
-it has three different syntaxes for defining the type of a table.
+it has three different syntaxes for defining the type of a table:
+\begin{align}
+\label{syn:table1}
+\texttt{\{k:t\}} & \equiv \texttt{\{k:t|nil\}}\\
+\label{syn:table2}
+\texttt{\{t\}} & \equiv \texttt{\{number:t|nil\}}\\
+\label{syn:table3}
+\texttt{\{k}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
+\texttt{, ..., k}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}} & \equiv
+\texttt{\{k}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
+\texttt{, ..., k}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}
+\end{align}
 The first syntax defines table types for maps:
 it is written \texttt{\{k:t\}},
-and maps to the table type $\{K:T\;\cup\;\Nil\}$.
+and maps to the table type \texttt{\{k:t|nil\}}.
 This table type represents a map that maps values of type
-$K$ to values of type $T\;\cup\;\Nil$,
-as Lua returns $\Nil$ when we use a non-existing key to index a table.
+\texttt{k} to values of type \texttt{t|nil},
+as Lua returns \texttt{nil} when we use a non-existing key to index a table.
 The second syntax defines table types for arrays:
 it is written \texttt{\{t\}},
-and maps to the table type $\{\Number:T\;\cup\;\Nil\}$.
+and maps to the table type \texttt{\{number:t|nil\}}.
 This table type is just a syntactic sugar to the definition of a
-map that maps values of type $\Number$ to values of type $T\;\cup\;\Nil$.
-The third syntax defines table types for records:
-it is written \texttt{\{k1:t1, ..., kn:tn\}},
-and maps to the table type $\{K_{1}:T_{1}, ..., K_{n}:T_{n}\}$.
+map that maps values of type \texttt{number} to values of type \texttt{t|nil}.
+The third syntax defines table types for records: it is written
+$\texttt{\{k}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
+\texttt{, ..., k}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}$,
+and maps to the same table type.
 This table type represents a record that maps the literal types
-$K_{i}$\texttt{, ..., }$K_{n}$ to values of types
-$T_{i}$\texttt{, ..., }$T_{n}$, where $i$ may vary from one to $n$.
+$\texttt{k}_{\texttt{i}}$\texttt{, ..., }$\texttt{k}_{\texttt{n}}$
+to values of types $\texttt{t}_{\texttt{i}}$\texttt{, ..., }$\texttt{t}_{\texttt{n}}$.
 
@@
 The second line of this example gives a type error because we are
-attempting to assign a value of type $\Number\;\cup\;\Nil$ to a
-variable that accepts only values of type $\Number$.
+attempting to assign a value of type \texttt{number|nil} to a
+variable that accepts only values of type \texttt{number}.
 Although the field \texttt{bar} does not exist in \texttt{t}, the third
@@
 In this example, the type of both variables \texttt{x} and \texttt{y}
-is $\String\;\cup\;\Nil$.
+is \texttt{string|nil}.
 An incovenience aspect of making the types of maps and arrays always
-include the type $\Nil$ is to overload the programmers,
+include the type \texttt{nil} is to overload the programmers,
 as they need to use \texttt{or} or \texttt{if} to narrow the type of
@@
 syntax of records to infer the following table type to \texttt{days}:
-\begin{align*}
-\{{1:\String},\;{2:\String},\;{3:\String},\;{4:\String},\;\\
-{5:\String},\;{6:\String},\;{7:\String}\}
-\end{align*}
-% text about subtyping
+\texttt{\{1:string, 2:string, 3:string, 4:string, 5:string, 6:string, 7:string\}}.
 
@@
 Typed Lua includes \emph{interfaces} as syntactic sugar to table types
-that specify records, as writing table types can be unwieldy when
+that specify records, as writing table types can be unwidely when
 records get bigger and types of record fields get more complicated.
@@
 
-The declaration of the interface \texttt{Person} is an alias to the table type
-$\{``firstname":\String, ``lastname":\String\}$ in the remainder
-of the current scope.
+The declaration of the interface \texttt{Person} is an alias to the
+table type \texttt{\{"firstname":string, "lastname":string\}} in the
+remainder of the current scope.
 Now we can use \texttt{Person} in type declarations:
@@
 
-    print(greeter(user1)) -- Goodbye Lewis Allan
+    print(greeter(user1)) -- Goodbye Lewis Reed 
     print(greeter(user2)) -- compile-time error
@@
 is a subtype of
-$\{``firstname":\String, ``middlename":\String\cup\Nil, ``lastname":\String\}$,
+\texttt{\{"firstname":string, "middlename":string|nil, "lastname":string\}},
 the type that \texttt{Person} describes.
@@
 
-The declaration of the interface \texttt{Element} is an alias to the
-recursive type
-$\mu Element.\{``info":\Number, ``next":Element \cup \Nil\}$ in the
-remainder of the current scope.
-
 We need to explicitly annotate \texttt{next} being of type
-$Element\;\cup\;\Nil$ because in our type system $\Nil$
-is not the bottom type, that is, $\Nil$ is subtype of itself only.
-Typed Lua has this behavior because in Lua $\Nil$ represents
+\texttt{Element|nil} because in our type system \texttt{nil}
+is not the bottom type, that is, \texttt{nil} is subtype of itself only.
+Typed Lua has this behavior because in Lua \texttt{nil} represents
 the absence of an useful value.
-Thus, making $\Nil$ the bottom type would severaly reduce the amount
+Thus, making \texttt{nil} the bottom type would severaly reduce the amount
 of type safety in the system.
@@
 we build the table:
-starting with $\{\}$, and then refining to $\{``firstname":\String\}$,
-and finally reaching $\{``firstname":\String, ``lastname":\String\}$.
+starting with \texttt{\{\}}, and then refining to \texttt{\{"firstname":string\}},
+and finally reaching \texttt{\{"firstname":string, "lastname":string\}}.
 This type change is trickier than the one that we introduced for narrowing
@@
 On the fifth line we try to change the type of \texttt{bogus}
-from $\{\}$ to $\{``firstname":\Boolean\}$, but if we allow this
-type change we also allow changing the value that is stored in
-\texttt{person.firstname}, regardless of its type.
+from \texttt{\{\}} to \texttt{\{"firstname":boolean\}},
+but if we allow this type change we also allow changing the value
+that is stored in \texttt{person.firstname}, regardless of its type.
 Taking the changes individually looks fine, but the truth is that
@@
 Thus, Lua modules can be any Lua value.
-However, our survey from Chapter \ref{chapter:statistics} confirms that
+However, our survey from Chapter \ref{chap:statistics} confirms that
 the convention among Lua programmers is to use tables for defining modules. 
@@
     local mymath = {}
-    local RADIANS_PER_DEGREE = 3.14 / 180.0
+    local RADIANS_PER_DEGREE = 3.14159 / 180.0
     mymath.deg = function (x:number)
@@
 final type of the module \texttt{mymath}:
-\begin{align*}
-\{ & ``deg":\Number \times \Value{*} \rightarrow \Number \times \Nil{*},\\
-& ``rad":\Number \times \Value{*} \rightarrow \Number \times \Nil{*},\\
-& ``pow":\Number \times \Number \times \Value{*} \rightarrow \Number \times \Nil{*}\}
-\end{align*}
+\begin{verbatim}
+    { "deg":(number) -> (number),
+      "rad":(number) -> (number),
+      "pow":(number, number) -> (number) }
+\end{verbatim}
 
@@
 \begin{verbatim}
-    local RADIANS_PER_DEGREE = 3.14 / 180.0
+    local RADIANS_PER_DEGREE = 3.14159 / 180.0
     local function deg (x:number)
@@
     module "mymath"
-    local RADIANS_PER_DEGREE = 3.14 / 180.0
+    local RADIANS_PER_DEGREE = 3.14159 / 180.0
     deg = function (x:number)
@@
 
-Typed Lua uses table types along with the \emph{self type} to
+Typed Lua uses table types along with the self type to
 represent objects, as we show in the following example:
@@
 parameters.
-Typed Lua does the same, and assigns the type $\Self$ to the
+Typed Lua does the same, and assigns the type \texttt{self} to the
 implicit parameter \texttt{self}.
 Typed Lua also adds the \texttt{const} annotation because the
-$\Const$ qualifier is necessary for covariance among object types work.
+\texttt{const} qualifier is necessary for covariance among object types work.
 Under this circumstances, the type of \texttt{Shape} is an object of
 the following type:
-\begin{align*}
-\{ & ``x":\Number, ``y":\Number, \\
-& \Const \; ``move":\Self \times \Number \times \Number \times \Value{*} \rightarrow \Nil{*}\}
-\end{align*}
+\begin{verbatim}
+    { "x":number, "y":number,
+      const "move":(self, number, number) -> () }
+\end{verbatim}
 
 While \texttt{:} is syntactic sugar in plain Lua, Typed Lua uses it
-to check method calls, binding any occurence of the $\Self$ type
+to check method calls, binding any occurence of the type \texttt{self}
 in the type of the method to the receiver.
@@
     interface Shape
-      x:number
-      y:number
-      const new:(self, number, number) -> (self)
-      const move:(self, number, number) -> ()
+      x, y:number
+      const new:(number, number) => (self)
+      const move:(number, number) => ()
     end
@@
 \texttt{new} to the class \texttt{Shape}.
-The type of \texttt{new} is
-$\Self \times \Number \times \Number \times \Value{*} \rightarrow \Self \times \Nil{*}$,
+The type of \texttt{new} is \texttt{(self, number, number) -> (self)},
 because we use \texttt{setmetatable} to declare \texttt{s} having
@@
 In the same example, we also declare the interface \texttt{Shape}.
-This allows us to export the type of the class \texttt{Shape},
+Notice that we used double arrow instead of single arrow inside the
+definition of the interface \texttt{Shape}.
+The double arrow is syntactic sugar for defining the type of methods,
+as it includes an implicit first parameter of type \texttt{self}.
+The interface definition allows us to export the type of the class \texttt{Shape},
 which is the same type that is given by the variable \texttt{Shape}:
-\begin{align*}
-\{ & ``x":\Number, ``y":\Number,\\
-& \Const \; ``new":\Self \times \Number \times \Number \times \Value{*} \rightarrow \Self \times \Nil{*},\\
-& \Const \; ``move":\Self \times \Number \times \Number \times \Value{*} \rightarrow \Nil{*}\}
-\end{align*}
+\begin{verbatim}
+    { "x":number, "y":number,
+      const "new":(self, number, number) -> (self),
+      const "move":(self, number, number) -> () }
+\end{verbatim}
 
@@
     interface Circle
-      x:number
-      y:number
-      radius:number
-      const new:(self, number, number, value) -> (self)
-      const move:(self, number, number) -> ()
-      const area:(self) -> (number)
+      x, y, radius:number
+      const new:(number, number, value) => (self)
+      const move:(number, number) => ()
+      const area:() => (number)
     end
@@
 The interface \texttt{Circle} exports the following type:
-\begin{align*}
-\{ & ``x":\Number, ``y":\Number, ``radius":\Number,\\
-& \Const \; ``new":\Self \times \Number \times \Number \times \Value \times \Value{*} \rightarrow \Self \times \Nil{*},\\
-& \Const \; ``move":\Self \times \Number \times \Number \times \Value{*} \rightarrow \Nil{*},\\
-& \Const \; ``area":\Self \times \Value{*} \rightarrow \Number \times \Nil{*}\}
-\end{align*}
+\begin{verbatim}
+    { "x":number, "y":number, "radius":number,
+      const "new":(self, number, number, value) -> (self),
+      const "move":(self, number, number) -> (),
+      const "radius":() -> (number) }
+\end{verbatim}
 
@@
 \section{Related Lua projects}
-\label{sec:related}
 

### 2014-08-18 doc/thesis/typedlua.tex
Adding text about description files

@@
 
+\section{Description files}
+
+The Typed Lua compiler also handles description files.
+The programmer can use description files for two purposes:
+export a statically typed interface for an existing Lua
+module, without coverting it to Typed Lua,
+or export a statically typed interface for a Lua module
+that is written in C.
+
+\begin{figure}[!ht]
+\textbf{The complete syntax of description files}\\
+\dstart
+\begin{align*}
+\textit{description} & ::= \; \textit{desclist}\\
+\textit{desclist} & ::= \; \textit{descitem} \; \{ \textit{descitem} \}\\
+\textit{descitem} & ::= \; \textit{typedid} \; | \;
+  \textit{interface} \; | \;
+  \textit{userdata}\\
+\textit{typedid} & ::= \; \textit{id} \; \texttt{`:'} \; \textit{type}\\ 
+\textit{interface} & ::= \; \textbf{interface} \; \textit{typedec}\\
+\textit{userdata} & ::= \; \textbf{userdata} \; \textit{typedec}\\
+\textit{typedec} & ::= \; \textit{Name} \; { \; \textit{decitem} \; } \; \textbf{end}\\
+\textit{decitem} & ::= \; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
+\textit{idtype} & ::= \; \textit{type} \; | \; \textit{methodtype}\\
+\textit{idlist} & ::= \; \textit{id} \; \{ \; \texttt{`,'} \; \textit{id} \; \}\\
+\textit{id} & ::= \; [\textbf{const}] \; \textit{Name}
+\end{align*}
+\dend
+\caption{The concrete syntax of Typed Lua description files}
+\label{fig:tldsyntax}
+\end{figure}
+
+The Typed Lua compiler searches for description files when it cannot
+not find the respective Typed Lua file that is the argument of
+\texttt{require}, as we can also use description files to give and
+export the type of modules.
+
+The following example shows the description file for \texttt{lmd5},
+a MD5 digest library for Lua that is written in C:
+\begin{verbatim}
+    userdata md5_context
+      __tostring : (self) -> (string)
+      clone : (self) -> (self)
+      digest : (self|string, value?) -> (string)
+      new : () -> (self)
+      reset : (self) -> (self)
+      update : (self, string*) -> (self)
+      version : string
+    end
+
+    __tostring : (md5_context) -> (string)
+    clone : (md5_context) -> (md5_context)
+    digest : (md5_context|string, value) -> (string)
+    new : () -> (md5_context)
+    reset : (md5_context) -> (md5_context)
+    update : (md5_context, string*) -> (md5_context)
+    version : string
+\end{verbatim}
+
+This description file exports the type \texttt{md5\string_context}
+through the \texttt{userdata} declaration, and the table type
+that represents the module.
+Now we can use the Typed Lua compiler to check for type errors on
+our use of the library:
+\begin{verbatim}
+    local m = require "md5"
+    local x = m.new()
+    local y = x:clone()
+    local z = m.clone("foo")             -- compile-time error
+    print(x:digest() == m.digest(x))
+\end{verbatim}
+
+In this example there is a type error because function \texttt{clone}
+expects a value of type \texttt{md5\string_context} instead of a value
+of type \texttt{string}.
+
+The description files are the mechanism that we used to include the
+typing of the Lua standard libraries inside Typed Lua.
+
 \section{Related Lua projects}

### 2014-08-20 doc/thesis/system.tex
Reorganizing the sections of Chapter 5 to start writing it
Cosmetic changes to chapters 4 and 5

@@
+\section{Type Language}
+
 \begin{figure}[!ht]
@@
 & | \; X_{i} & \textit{projection types}\\
-L ::= & \False \; | \; \True \; | \; {<}{\it number}{>} \; | \; {<}{\it string}{>} & \\
+L ::= & \False \; | \; \True \; | \; {\it number} \; | \; {\it string} & \\
 B ::= & \Boolean \; | \; \Number \; | \; \String & \\
@@
 
-Rule for checking if a table type is well formed:
-\[
-\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
-\]
+\section{Subtyping}
 
@@
 \mylabel{S-NUMBER}\\
-\senv \vdash {<}{\it number}{>} \subtype \Number
+\senv \vdash {\it number} \subtype \Number
 \end{array}
@@
 \mylabel{S-STRING}\\
-\senv \vdash {<}{\it string}{>} \subtype \String
+\senv \vdash {\it string} \subtype \String
 \end{array}
@@
 
-\begin{figure}[!ht]
-\textbf{Abstract Syntax}\\
-\dstart
-$$
-\begin{array}{rl}
-s ::= & \;\; \mathbf{skip} \; | \;
-s_{1} \; ; \; s_{2} \; | \;
-\vec{l} = \vec{e}  \; | \;
-\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
-\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
-& | \; \mathbf{local} \; \vec{n}{:}\vec{T} = \vec{e} \; \mathbf{in} \; s \; | \;
-\mathbf{rec} \; n{:}T = f \; \mathbf{in} \; s \; | \;
-\mathbf{return} \; \vec{e} \; | \;
-a\\
-e ::= & \;\; \mathbf{nil} \; | \;
-\mathbf{false} \; | \;
-\mathbf{true} \; | \;
-{<}{\it number}{>} \; | \;
-{<}{\it string}{>} \; | \;
-{...} \; | \;
-f \; | \;
-\{ \; \vec{c}\; \} \\
-& | \; e_{1} + e_{2} \; | \;
-e_{1} \; {..} \; e_{2} \; | \;
-e_{1} == e_{2} \; | \;
-e_{1} < e_{2} \; | \;
-e_{1} \; \mathbf{and} \; e_{2} \; | \;
-e_{1} \; \mathbf{or} \; e_{2} \\
-& | \; \mathbf{not} \; e \; | \;
-- e \; | \;
-\# \; e \; | \;
-a \; | \;
-l \\
-l ::= & \; n \; | \;
-e_{1}[e_{2}]\\
-a ::= & \;\; e_{1}(\vec{e_{2}}) \; | \;
-e_{1}{:}n(\vec{e_{2}})\\
-f ::= & \; \mathbf{fun} \; (){:}R \; s \; | \;
-\mathbf{fun} \; ({...}{:}T){:}R \; s \; | \;
-\mathbf{fun} \; (\vec{n}{:}\vec{T}){:}R \; s \; | \;
-\mathbf{fun} \; (\vec{n}{:}\vec{T},{...}{:}T){:}R \; s\\
-c ::= & \; [e_{1}] = e_{2}\\
-n ::= & \; {<}{\it name}{>}\\
-\end{array}
-$$
-\dend
-\caption{Typed Lua abstract syntax}
-\label{fig:syntax}
-\end{figure}
+\section{Consistent-Subtyping}
 
@@
 \mylabel{C-NUMBER}\\
-\senv \vdash {<}{\it number}{>} \lesssim \Number
+\senv \vdash {\it number} \lesssim \Number
 \end{array}
@@
 \mylabel{C-STRING}\\
-\senv \vdash {<}{\it string}{>} \lesssim \String
+\senv \vdash {\it string} \lesssim \String
 \end{array}
@@
 \end{figure}
+
+\section{Typing rules}
+
+\begin{figure}[!ht]
+\textbf{Abstract Syntax}\\
+\dstart
+$$
+\begin{array}{rl}
+s ::= & \;\; \mathbf{skip} \; | \;
+s_{1} \; ; \; s_{2} \; | \;
+\vec{l} = \vec{e}  \; | \;
+\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
+\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
+& | \; \mathbf{local} \; \vec{n}{:}\vec{T} = \vec{e} \; \mathbf{in} \; s \; | \;
+\mathbf{rec} \; n{:}T = f \; \mathbf{in} \; s \; | \;
+\mathbf{return} \; \vec{e} \; | \;
+a\\
+e ::= & \;\; \mathbf{nil} \; | \;
+\mathbf{false} \; | \;
+\mathbf{true} \; | \;
+{\it number} \; | \;
+{\it string} \; | \;
+{...} \; | \;
+f \; | \;
+\{ \; \vec{c}\; \} \\
+& | \; e_{1} + e_{2} \; | \;
+e_{1} \; {..} \; e_{2} \; | \;
+e_{1} == e_{2} \; | \;
+e_{1} < e_{2} \; | \;
+e_{1} \; \mathbf{and} \; e_{2} \; | \;
+e_{1} \; \mathbf{or} \; e_{2} \\
+& | \; \mathbf{not} \; e \; | \;
+- e \; | \;
+\# \; e \; | \;
+a \; | \;
+l \; | \;
+(e) \; | \;
+{<}t{>} \; e\\
+l ::= & \; n \; | \;
+e_{1}[e_{2}]\\
+a ::= & \;\; e_{1}(\vec{e_{2}}) \; | \;
+e_{1}{:}n(\vec{e_{2}})\\
+f ::= & \; \mathbf{fun} \; (){:}R \; s \; | \;
+\mathbf{fun} \; ({...}{:}T){:}R \; s \; | \;
+\mathbf{fun} \; (\vec{n}{:}\vec{T}){:}R \; s \; | \;
+\mathbf{fun} \; (\vec{n}{:}\vec{T},{...}{:}T){:}R \; s\\
+c ::= & \; [e_{1}] = e_{2}\\
+n ::= & \; {\it name}\\
+\end{array}
+$$
+\dend
+\caption{Typed Lua abstract syntax}
+\label{fig:syntax}
+\end{figure}
+
+Rule for checking if a table type is well formed:
+\[
+\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
+\]

### 2014-08-20 doc/thesis/typedlua.tex
Cosmetic changes to chapters 4 and 5

@@
 include new keywords that are not reserved words.
-Appendix \ref{app:syntax} presents the complete syntax of Typed Lua.
+Appendix \ref{app:syntax} presents the complete syntax of Typed Lua
+in extended BNF.
 
@@
 \begin{figure}[!ht]
-\textbf{Type annotations}\\
+\textbf{Types}\\
 \dstart
@@
 and union types.
-Second-level types include the type void \texttt{()}, first-level types,
+Second-level types include the type void \texttt{()},
 vararg types, tuple types, and union of tuple types.
@@
 
-To match the semantics of Lua function calls, the default operation
+To match the semantics of Lua function calls, the \emph{default} operation
 mode of Typed Lua adds a variadic tail to the parts of a function
@@
 
-Typed Lua also provides a strict operation mode, and programmers can
+Typed Lua also provides a \emph{strict} operation mode, and programmers can
 use this mode if they want to catch arity mismatch of function calls.
@@
 false condition, so programmers often use the \texttt{or} operator as a
-common idiom for giving a default value to an optional parameter. 
+common idiom to assign a default value to an optional parameter. 
 Typed Lua uses a typing rule to encode this idiom: 
@@
 Lua programmers also overload the return type of functions,
-usually for signaling the occurence of errors.
+usually for signaling the occurrence of errors.
 In this idiom, a function returns its normal set of return values
@@
 \begin{verbatim}
-    local function idiv(dividend:number, divisor:number):
+    local function idiv (dividend:number, divisor:number):
           (number, number)|(nil, string)
@@
 as both predicates would produce the same result.
-However, the form that appears in our example is much more succint and idiomatic.
+However, the form that appears in our example is much more succinct and idiomatic.
 
@@
 
-The overloading mechanims of Typed Lua have a limitation:
+The overloading mechanism of Typed Lua have a limitation:
 the return type cannot depend on the input type.
@@
 guaranteed to return a string when we pass two strings.
-Even though we can use intersection types to express the type
+Even though we could use intersection types to express the type
 of this function as
-\texttt{(number, number, value*) -> (number, nil*) \string&
-(string, string, value*) -> (string, nil*)},
-and thefore solving this limitation,
+\texttt{(number, number) -> (number) \string&
+(string, string) -> (string)},
+and therefore solve this limitation,
 intersection types require more sophisticated flow typing to
 check whether a function has this type,
-and we are still working on this problem.
+and we still need to work on this problem.
 
@@
 is \texttt{string|nil}.
-An incovenience aspect of making the types of maps and arrays always
+An inconvenience aspect of making the types of maps and arrays always
 include the type \texttt{nil} is to overload the programmers,
@@
 Typed Lua includes \emph{interfaces} as syntactic sugar to table types
-that specify records, as writing table types can be unwidely when
+that specify records, as writing table types can be unwieldy when
 records get bigger and types of record fields get more complicated.
@@
 
-If our record type has optinal fields, that is, fields that can be \texttt{nil},
+If our record type has optional fields, that is, fields that can be \texttt{nil},
 we need to use an explicit type declaration when declaring a
@@
 the absence of an useful value.
-Thus, making \texttt{nil} the bottom type would severaly reduce the amount
+Thus, making \texttt{nil} the bottom type would severally reduce the amount
 of type safety in the system.
@@
 Now we can use \texttt{Element} to annotate a function that
-inserts an element at the begining of a list:
+inserts an element at the beginning of a list:
 \begin{verbatim}
@@
 loaded or not.
-When the module is not loaded, Lua executes its soure file as a
+When the module is not loaded, Lua executes its source file as a
 function, and the value returned by this function is the module.
@@
 Another way of defining modules in Lua is to declare only locals
-in the source file and return a table constructor that includes
+in the source file, and return a table constructor that includes
 the members that the module should export at the end of the source file.
@@
 
-The way Typed Lua handle modules using table types is also
+The way Typed Lua handles modules using table types is also
 relevant for supporting object-oriented programming, as we discuss in
@@
 While \texttt{:} is syntactic sugar in plain Lua, Typed Lua uses it
-to check method calls, binding any occurence of the type \texttt{self}
+to check method calls, binding any occurrence of the type \texttt{self}
 in the type of the method to the receiver.
@@
 In the same example, we also declare the interface \texttt{Shape}.
-Notice that we used double arrow instead of single arrow inside the
+Notice that we use double arrow instead of single arrow inside the
 definition of the interface \texttt{Shape}.
@@
 
-In Typed Lua, exporting interfaces allow us to use \texttt{require}
+In Typed Lua, this style of classes definition allow us to use \texttt{require}
 for creating prototype objects that work as classes,
 along with a shortcut to the table type of the class.
-This allows us to use this shortcut on type annotatoins,
+This allows us to use this shortcut on type annotations,
 as we show in the following example:
@@
 export a statically typed interface for an existing Lua
-module, without coverting it to Typed Lua,
+module, without converting it to Typed Lua,
 or export a statically typed interface for a Lua module
@@
 
-The Typed Lua compiler searches for description files when it cannot
+Figure \ref{fig:tldsyntax} shows the complete syntax of Typed Lua
+description files in extended BNF.
+A Typed Lua description file defines the table type that represents
+a certain module and the type names that are exported
+along with this table type.
+We can export a type name through the declaration of an interface or
+userdata.
+As we saw, an interface declaration is just a shortcut to a table type.
+An userdata declaration is a shortcut to a table type that includes
+a \emph{brand} that is its name.
+The Typed Lua compiler uses this brand to combine structural with
+nominal type checking, so two different userdata are not subtype
+of each other because they do not share the same brand.
+The Typed Lua compiler searches for a description file when it cannot
 not find the respective Typed Lua file that is the argument of
-\texttt{require}, as we can also use description files to give and
-export the type of modules.
+\texttt{require}.
 
@@
 a single table constructor.
-Typed Lua has a simpler form of this feature, with a different
-formalization.
+This feature inspired us to introduce it into Typed Lua,
+but in a simpler way and with different formalization.
 Like Typed Lua, Tidal Lock also has local type inference.
 Unlike Typed Lua, Tidal Lock does not type the common object-oriented
-idioms used by Lua programmers.
+idioms used by Lua programmers, though it also uses subtyping.

### 2014-08-25 doc/thesis/typedlua.tex
Fixing the syntax of table types in the section about tables and interfaces

@@
 \label{syn:table1}
-\texttt{\{k:t\}} & \equiv \texttt{\{k:t|nil\}}\\
+\texttt{`\{'} \; \textit{keytype} \; \texttt{`:'} \; \textit{type} \; \texttt{`\}'}\\
 \label{syn:table2}
-\texttt{\{t\}} & \equiv \texttt{\{number:t|nil\}}\\
+\texttt{`\{'} \; \textit{type} \; \texttt{`\}'}\\
 \label{syn:table3}
-\texttt{\{k}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
-\texttt{, ..., k}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}} & \equiv
-\texttt{\{k}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
-\texttt{, ..., k}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}
+\texttt{`\{'} \; \textit{literaltype} \; \texttt{`:'} \; \textit{type} \;
+\{\texttt{`,'} \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}\} \;
+[\texttt{`,'} \; \textit{type}] \; \texttt{`\}'}
 \end{align}
-The first syntax defines table types for maps:
+The syntax \ref{syn:table1} defines table types for maps:
 it is written \texttt{\{k:t\}},
@@
 as Lua returns \texttt{nil} when we use a non-existing key to index a table.
-The second syntax defines table types for arrays:
+The syntax \ref{syn:table2} defines table types for arrays:
 it is written \texttt{\{t\}},
@@
 map that maps values of type \texttt{number} to values of type \texttt{t|nil}.
-The third syntax defines table types for records: it is written
-$\texttt{\{k}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
-\texttt{, ..., k}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}$,
+The syntax \ref{syn:table3} defines table types for records with an
+optional array part: it is written
+$\texttt{\{l}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
+\texttt{, ..., l}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}$,
 and maps to the same table type.
 This table type represents a record that maps the literal types
-$\texttt{k}_{\texttt{i}}$\texttt{, ..., }$\texttt{k}_{\texttt{n}}$
+$\texttt{l}_{\texttt{i}}$\texttt{, ..., }$\texttt{l}_{\texttt{n}}$
 to values of types $\texttt{t}_{\texttt{i}}$\texttt{, ..., }$\texttt{t}_{\texttt{n}}$.
@@
 \begin{verbatim}
-    local album = { "Transformer", 1972, "Lou Reed" }
+    local album:{1:string, 2:number, 3:string} =
+      { "Transformer", 1972, "Lou Reed" }
 \end{verbatim}
@@
     local user1:Person = { firstname = "Lewis",
-                           middlename = "Allan" or nil,
+                           middlename = "Allan",
                            lastname = "Reed" }
@@
 the type that \texttt{Person} describes.
-While declaring \texttt{user1}, we use \texttt{or} to cast the type of
-\texttt{middlename} because of the contravariance in the type of
-mutable fields.
 
@@
 
-The following example shows the description file for \texttt{lmd5},
+The following example shows the description file for \texttt{lmd5} \citep{lmd5},
 a MD5 digest library for Lua that is written in C:

### 2014-08-29 doc/thesis/evaluation.tex
Adding evaluation data of Lua Standard Libraries and MD5

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
\textbf{Case study} & \textbf{easy} & \textbf{over} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
Lua Standard Libraries & 79 & 8 & 12 & 24 & 123 & 71\% \\
\hline
MD5 & 7 & 0 & 0 & 0 & 7 & 100\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation of Typed Lua by library}
\end{table}

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{over} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
\multirow{9}{*}{Lua Standard Libraries}
& base & 8 & 2 & 5 & 11 & 26 & 38\% \\
\cline{2-8}
& coroutine & 0 & 0 & 0 & 6 & 6 & 0\% \\
\cline{2-8}
& package & 4 & 1 & 2 & 1 & 8 & 62\% \\
\cline{2-8}
& string & 11 & 2 & 0 & 1 & 14 & 93\% \\
\cline{2-8}
& table & 0 & 1 & 5 & 0 & 6 & 17\% \\
\cline{2-8}
& math & 28 & 0 & 0 & 1 & 29 & 97\% \\
\cline{2-8}
& bit32 & 12 & 0 & 0 & 0 & 12 & 100\% \\
\cline{2-8}
& io & 9 & 0 & 0 & 2 & 11 & 82\% \\
\cline{2-8}
& os & 7 & 2 & 0 & 2 & 11 & 82\% \\
\hline
\multirow{1}{*}{MD5}
& md5 & 7 & 0 & 0 & 0 & 7 & 100\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation of Typed Lua by module}
\end{table}

### 2014-09-02 doc/thesis/evaluation.tex
Removing over category from evaluation results

@@
+We performed some case studies on existing Lua libraries
+to evaluate the design of our type system.
+For each library, we used Typed Lua to either rewrite its modules
+or to write statically typed interfaces to its modules through
+Typed Lua's description files.
+In this chapter we present our evaluation results and discuss
+some interesting cases.
+
+Tables \ref{tab:evalbylib} and \ref{tab:evalbymod} sumarizes our
+evaluation results:
+the first presents the results agruped by library while
+the second presents the results agruped by module.
+We split the members that we typed into three categories:
+\emph{easy}, \emph{poly}, and \emph{hard}.
+The \emph{easy} category represents the members that we could give
+a precise static type for them.
+The \emph{poly} category represents the members that we had to
+rely on the dynamic type, and we believe we could give a precise
+static type for them if our type system had parametric polymorphism.
+The \emph{hard} category represents the members that we could not
+give a precise static type for them, relying on the dynamic type
+in most of the cases, but with no hopes that parametric polymorphism
+would help us to give a precise static type for them.
+The last column of each table contains the actual evaluation result
+of each case study, that is basically the percentage of members
+that are under the \emph{easy} category.
+
 \begin{table}[!ht]
 \begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|c|}
+\begin{tabular}{|l|c|c|c|c|c|c|}
 \hline
-\textbf{Case study} & \textbf{easy} & \textbf{over} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
 \hline
-Lua Standard Libraries & 79 & 8 & 12 & 24 & 123 & 71\% \\
+Lua Standard Libraries & 87 & 11 & 25 & 123 & 71\% \\
 \hline
-MD5 & 7 & 0 & 0 & 0 & 7 & 100\% \\
+MD5 & 7 & 0 & 0 & 7 & 100\% \\
 \hline
@@
 \caption{Evaluation of Typed Lua by library}
+\label{tab:evalbylib}
 \end{table}
@@
 \begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|c|}
+\begin{tabular}{|l|c|c|c|c|c|c|}
 \hline
-\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{over} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
 \hline
 \multirow{9}{*}{Lua Standard Libraries}
-& base & 8 & 2 & 5 & 11 & 26 & 38\% \\
-\cline{2-8}
-& coroutine & 0 & 0 & 0 & 6 & 6 & 0\% \\
-\cline{2-8}
-& package & 4 & 1 & 2 & 1 & 8 & 62\% \\
-\cline{2-8}
-& string & 11 & 2 & 0 & 1 & 14 & 93\% \\
-\cline{2-8}
-& table & 0 & 1 & 5 & 0 & 6 & 17\% \\
-\cline{2-8}
-& math & 28 & 0 & 0 & 1 & 29 & 97\% \\
-\cline{2-8}
-& bit32 & 12 & 0 & 0 & 0 & 12 & 100\% \\
-\cline{2-8}
-& io & 9 & 0 & 0 & 2 & 11 & 82\% \\
-\cline{2-8}
-& os & 7 & 2 & 0 & 2 & 11 & 82\% \\
+& base & 9 & 5 & 12 & 26 & 35\% \\
+\cline{2-7}
+& coroutine & 0 & 0 & 6 & 6 & 0\% \\
+\cline{2-7}
+& package & 5 & 2 & 1 & 8 & 62\% \\
+\cline{2-7}
+& string & 13 & 0 & 1 & 14 & 93\% \\
+\cline{2-7}
+& table & 1 & 4 & 1 & 6 & 17\% \\
+\cline{2-7}
+& math & 28 & 0 & 1 & 29 & 97\% \\
+\cline{2-7}
+& bit32 & 12 & 0 & 0 & 12 & 100\% \\
+\cline{2-7}
+& io & 9 & 0 & 2 & 11 & 82\% \\
+\cline{2-7}
+& os & 10 & 0 & 1 & 11 & 91\% \\
 \hline
 \multirow{1}{*}{MD5}
-& md5 & 7 & 0 & 0 & 0 & 7 & 100\% \\
+& md5 & 7 & 0 & 0 & 7 & 100\% \\
 \hline
@@
 \caption{Evaluation of Typed Lua by module}
+\label{tab:evalbymod}
 \end{table}

### 2014-09-04 doc/thesis/system.tex
Reviewing the typing rules of expressions

@@
 & | \; P \rightarrow R & \textit{function types}\\
-& | \; \{ \}_{o|c} & \textit{empty table type}\\
-& | \; \{F, ..., F\}_{o|c} & \textit{table types}\\
+& | \; \{ \}_{c|o|u} & \textit{empty table type}\\
+& | \; \{F, ..., F\}_{c|o|u} & \textit{table types}\\
 & | \; X & \textit{type variables}\\
@@
 f \; | \;
-\{ \; \vec{c}\; \} \\
+\{ \; \vec{c} \; \} \\
 & | \; e_{1} + e_{2} \; | \;
@@
 e_{1}[e_{2}]\\
-a ::= & \;\; e_{1}(\vec{e_{2}}) \; | \;
+a ::= & \; e_{1}(\vec{e_{2}}) \; | \;
 e_{1}{:}n(\vec{e_{2}})\\

### 2014-09-29 doc/thesis/typedlua.tex
Fixing evaluation to include methods from userdata

@@
       clone : (self) -> (self)
-      digest : (self|string, value?) -> (string)
-      new : () -> (self)
+      digest : (self, value) -> (string)
+      new : (self) -> (self)
       reset : (self) -> (self)
       update : (self, string*) -> (self)
-      version : string
     end

### 2014-09-30 doc/thesis/system.tex
Trying to start writing about the type system

@@
-\section{Type Language}
+In the previous chapter we presented an informal overview of Typed Lua.
+We showed that programmers can use Typed Lua to combine static and dynamic
+typing in the same code.
+
+In this chapter we present the formalization of Typed Lua.
+We use typing rules to show how we implemented the features that we
+discussed in the previous chapter.
+
+\section{Types}
 
@@
 \multicolumn{3}{c}{\textbf{First-level types}}\\
-T ::= & \;\; L & \textit{literal types}\\
-& | \; B & \textit{base types}\\
+t ::= & \;\; l & \textit{literal types}\\
+& | \; b & \textit{base types}\\
 & | \; \Nil & \textit{nil type}\\
@@
 & | \; \Self & \textit{self type}\\
-& | \; T \cup T & \textit{disjoint union types}\\
-& | \; P \rightarrow R & \textit{function types}\\
-& | \; \{ \}_{c|o|u} & \textit{empty table type}\\
-& | \; \{F, ..., F\}_{c|o|u} & \textit{table types}\\
-& | \; X & \textit{type variables}\\
-& | \; \mu X.T & \textit{recursive types}\\
-& | \; X_{i} & \textit{projection types}\\
-L ::= & \False \; | \; \True \; | \; {\it number} \; | \; {\it string} & \\
-B ::= & \Boolean \; | \; \Number \; | \; \String & \\
-F ::= & K:T \; | \; \Const \; K:T & \textit{field types}\\ 
-K ::= & L \; | \; B \; | \; \Value \; | \; \Any & \textit{key types}\\
+& | \; t \cup t & \textit{disjoint union types}\\
+& | \; p \rightarrow r & \textit{function types}\\
+& | \; \{f_{i}, ..., f_{n}\}_{c|o|u} & \textit{table types}\\
+& | \; x & \textit{type variables}\\
+& | \; \mu x.t & \textit{recursive types}\\
+& | \; x_{i} & \textit{projection types}\\
+l ::= & \False \; | \; \True \; | \; {\it number} \; | \; {\it string} & \\
+b ::= & \Boolean \; | \; \Number \; | \; \String & \\
+f ::= & k:t \; | \; \Const \; k:t & \textit{field types}\\ 
+k ::= & l \; | \; b \; | \; \Any & \textit{key types}\\
 \multicolumn{3}{c}{}\\
 \multicolumn{3}{c}{\textbf{Second-level types}}\\
-S ::= & \;\; T &\\
-& | \; T* & \textit{vararg types}\\
-& | \; T \times S & \textit{tuple types}\\
-P ::= & \Void \; | \; S & \textit{parameters list type}\\
-R ::= & \Void \; | \; S \; | \; S \cup R & \textit{return types}\\
+s ::= & \;\; t &\\
+& | \; t* & \textit{variadic types}\\
+& | \; t \times s & \textit{tuple types}\\
+p ::= & \Void \; | \; s & \textit{parameters list type}\\
+r ::= & \Void \; | \; s \; | \; s \cup r & \textit{return types}\\
 \end{array}
@@
 Typed Lua types.
-We split Typed Lua types into two categories:
+Typed Lua splits types into two categories:
 \emph{first-level types} and \emph{second-level types}.
-We use first-level types to represent Lua values and
-second-level types to type expression lists, multiple assignments,
-and function applications.
-
-\section{Subtyping}
+Typed Lua uses first-level types to represent Lua values and
+second-level types to represent expression lists, and it also
+uses them to type multiple assignments and function applications.
+First-level types include literal types, base types, the type $\Nil$,
+the top type $\Value$, the dynamic type $\Any$, the type $\Self$,
+union types, function types, table types, recursive types, and
+projection types.
+Second-level types include the type $\Void$, tuples of first-level
+types optionally ending in a variadic type, and union of these tuples.
 
@@
 \end{figure}
-
-\section{Consistent-Subtyping}
 
@@
 \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
-& | \; \mathbf{local} \; \vec{n}{:}\vec{T} = \vec{e} \; \mathbf{in} \; s \; | \;
-\mathbf{rec} \; n{:}T = f \; \mathbf{in} \; s \; | \;
+& | \; \mathbf{local} \; \vec{n}{:}\vec{t} = \vec{e} \; \mathbf{in} \; s \; | \;
+\mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
 \mathbf{return} \; \vec{e} \; | \;
@@
 e_{1}{:}n(\vec{e_{2}})\\
-f ::= & \; \mathbf{fun} \; (){:}R \; s \; | \;
-\mathbf{fun} \; ({...}{:}T){:}R \; s \; | \;
-\mathbf{fun} \; (\vec{n}{:}\vec{T}){:}R \; s \; | \;
-\mathbf{fun} \; (\vec{n}{:}\vec{T},{...}{:}T){:}R \; s\\
+f ::= & \; \mathbf{fun} \; (){:}r \; s \; | \;
+\mathbf{fun} \; ({...}{:}t){:}r \; s \; | \;
+\mathbf{fun} \; (\vec{n}{:}\vec{t}){:}r \; s \; | \;
+\mathbf{fun} \; (\vec{n}{:}\vec{t},{...}{:}t){:}r \; s\\
 c ::= & \; [e_{1}] = e_{2}\\

### 2014-10-07 doc/thesis/review.tex
Moving statistics from a separated chapter to section 2

@@
 that handles the interaction between Racket and Typed Racket modules.
+
+\section{Statistics about the usage of Lua}
+\label{sec:statistics}
+
+In this section we present statistics about the usage of Lua
+features and idioms.
+We collected statistics about how programmers use tables, functions,
+dynamic type checking, object-oriented programming, and modules.
+We shall see that these statistics informed important design decisions
+on our optional type system.
+
+We used the LuaRocks repository to build our statistics database;
+LuaRocks \citep{hisham2013luarocks} is a package manager for Lua
+modules.
+We downloaded the 3928 \texttt{.lua} files that were available in
+the LuaRocks repository at February 1st 2014.
+However, we ignored files that are not compatible with Lua 5.2,
+the latest version of Lua.
+We also ignored \emph{machine-generated} files and test files,
+because these files may not represent idiomatic Lua code,
+and might skew our statistics towards non-typical uses of Lua.
+This left 2598 \texttt{.lua} files from 262 different projects for
+our statistics database;
+we parsed these files and processed their abstract syntax tree
+to gather the statistics that we show in this section.
+
+To verify how programmers use tables, we measured how they
+initialize, index, and iterate tables.
+We present these statistics in the next three paragraphs to discuss
+their influence on our type system.
+
+The table constructor appears 23185 times.
+In 36\% of the occurrences it is a constructor that initializes a
+record (e.g., \texttt{\{ x = 120, y = 121 \}});
+in 29\% of the occurrences it is a constructor that initializes a
+list (e.g., \texttt{\{ "one", "two", "three", "four" \}});
+in 8\% of the occurrences it is a constructor that initializes a
+record with a list part;
+and in less than 1\% of the occurrences (4 times) it is a constructor
+that uses only the booleans \texttt{true} and \texttt{false} as indexes.
+At all, in 73\% of the occurrences it is a constructor that uses
+only literal keys;
+in 26\% of the occurrences it is the empty constructor;
+in 1\% of the occurrences it is a constructor with non-literal keys
+only, that is, a constructor that uses variables and function calls
+to create the indexes of a table;
+and in less than 1\% of the occurrences (19 times) it is a constructor
+that mixes literal keys and non-literal keys.
+
+The indexing of tables appears 130448 times:
+86\% of them are for reading a table field while
+14\% of them are for writing into a table field.
+We can classify the indexing operations that are reads as follows:
+89\% of the reads use a literal string key,
+4\% of the reads use a literal number key,
+and less than 1\% of the reads (10 times) use a literal boolean key.
+At all, 93\% of the reads use literals to index a table while
+7\% of the reads use non-literal expressions to index a table.
+It also worth mentioning that 45\% of the reads are actually
+function calls.
+More precisely, 25\% of the reads use literals to call a function,
+20\% of the reads use literals to call a method, that is,
+a function call that uses the colon syntactic sugar, 
+and less than 1\% of the reads (195 times) use non-literal expressions
+to call a function.
+We can also classify the indexing operations that are writes as follows: 
+69\% of the writes use a literal string key,
+2\% of the writes use a literal number key,
+and less than 1\% of the writes (1 time) uses a literal boolean key.
+At all, 71\% of the writes use literals to index a table while
+29\% of the writes use non-literal expressions to index a table.
+
+We also measured how many files have code that iterate over tables to
+observe how frequently iteration is used.
+We observed that 23\% of the files have code that iterate over keys
+of any value, that is, the call to \texttt{pairs} appears at least
+once in these files (the median is twice per file);
+21\% of the files have code that iterate over integer keys, that is,
+the call to \texttt{ipairs} appears at least once in these files
+(the median is also twice per file);
+and 10\% of the files have code that use the numeric \texttt{for}
+along with the length operator (the median is once per file).
+
+The numbers about table initialization, indexing, and iteration
+show us that tables are mostly used to represent records, lists,
+and associative arrays.
+Therefore, Typed Lua should include a table type for handling
+these uses of Lua tables.
+Even though the statistics show that programmers initialize tables
+more often than they use the empty constructor to
+dynamically initialize tables, the statistics of the empty
+constructor are still expressive and indicate that Typed Lua should
+also include a way to handle this style of defining table types.
+
+We found a total of 24858 function declarations in our database
+(the median is six per file).
+Next, we discuss how frequently programmers use dynamic type
+checking and multiple return values inside these functions.
+
+We observed that 9\% of the functions perform dynamic type checking
+on their input parameters, that is, these functions use \texttt{type}
+to inspect the tags of Lua values (the median is once per function).
+We randomly selected 20 functions to sample how programmers are
+using \texttt{type}, and we got the following data:
+50\% of these functions use \texttt{type} for asserting the tags of
+their input parameters, that is, they raise an error when the tag of a
+certain parameter does not match the expected tag, and
+50\% of these functions use \texttt{type} for overloading, that is,
+they execute different code according to the inspected tag.
+
+These numbers show us that Typed Lua should include union types,
+because the use of the \texttt{type} idiom shows that disjoint unions
+will help programmers define data structures that can hold a value of
+several different, but fixed types.
+Typed Lua should also use \texttt{type} as a mechanism for decomposing
+unions, though it may be restricted to base types only.
+
+We observed that 10\% of the functions explicitly return multiple
+values.
+We also observed that 5\% of the functions return \texttt{nil} plus
+something else, for signaling an unexpected behavior;
+and 1\% of the functions return \texttt{false} plus something else,
+also for signaling an unexpected behavior.
+
+Typed Lua should include function types to represent Lua functions,
+and tuple types to represent the signatures of Lua functions,
+multiple return values, and multiple assignments.
+Tuple types will require some special attention, because Typed Lua
+should be able to adjust tuple types during compile-time, in a
+similar way to what Lua does with function calls and multiple
+assignments during run-time.
+In addition, the number of functions that return \texttt{nil} and
+\texttt{false} plus something else show us that overloading on the
+return type will be useful to the type system.
+
+We also measured how frequently programmers use the object-oriented
+paradigm in Lua.
+We observed that 23\% of the function declarations are actually
+method declarations.
+More precisely, 14\% of them use the colon syntactic sugar while
+9\% of them use \texttt{self} as their first parameter.
+We also observed that 63\% of the projects extend tables with
+metatables, that is, they call \texttt{setmetatable} at least once,
+and 27\% of the projects access the metatable of a given table,
+that is, they call \texttt{getmetatable} at least once.
+In fact, 45\% of the projects extend tables with metatables and
+declare methods:
+13\% using the colon syntactic sugar, 14\% using \texttt{self}, and
+18\% using both.
+
+Based on these observations, Typed Lua should include interfaces and
+classes as syntactic sugar for table types that encode objects.
+Proposing an optional type system for Lua that includes classes may
+sound unusual at first glance, because Lua does not have standard
+policies for object-oriented programming.
+However, Lua provides mechanisms that allow programmers to abstract
+their code in terms of objects, and our statistics confirm that
+an expressive number of programmers are relying on these mechanisms to
+use the object-oriented paradigm in Lua.
+Typed Lua should include interfaces and classes as merely syntactic
+mechanisms that help the compiler type check object-oriented code,
+and they should not influence the semantics of Lua.
+
+We also measured how programmers are defining modules.
+We observed that 38\% of the files use the current way of defining
+modules, that is, these files return a table that contains the
+exported members of the module at the end of the file;
+22\% of the files still use the deprecated way of defining modules,
+that is, these files call the function \texttt{module};
+and 1\% of the files use both ways.
+At all, 61\% of the files are modules while 39\% of the files are
+plain scripts.
+The number of plain scripts is high considering the origin of
+our database.
+However, we did not ignore sample scripts, which usually serve to
+help the users of a given module on how to use this module, and
+that is the reason why we have a high number of plain scripts.
+
+Based on these observations, Typed Lua should also use interfaces as
+syntactic sugar for table types that define modules.
+Typed Lua should also support the deprecated style of module
+definition, using global names as exported members of the module.
+
+Besides using interfaces for encoding objects and defining modules,
+Typed Lua should also use interfaces for defining the types of
+userdata, as programmers often use userdata in an object-oriented
+style.
+
+The last statistics that we collected were about variadic functions
+and vararg expressions.
+We observed that 8\% of the functions are variadic, that is,
+their last parameter is the vararg expression.
+We also observed that 5\% of the initialization of lists
+(or 2\% of the occurrences of the table constructor) use solely the
+vararg expression.
+Typed Lua should include a \emph{vararg type} to handle variadic
+functions and vararg expressions.

### 2014-10-07 doc/thesis/typedlua.tex
Moving statistics from a separated chapter to section 2

@@
 Thus, Lua modules can be any Lua value.
-However, our survey from Chapter \ref{chap:statistics} confirms that
+However, our survey from Section \ref{sec:statistics} confirms that
 the convention among Lua programmers is to use tables for defining modules. 
@@
 Even though this style has been deprecated, our survey from
-Chapter \ref{chap:statistics} indicated that around a third of the
+Section \ref{sec:statistics} indicated that around a third of the
 Lua modules that are in LuaRocks still use this style.

### 2014-10-09 doc/thesis/system.tex
Adding first versions of the typing rules for statements
Starting to write about subtyping

@@
 We showed that programmers can use Typed Lua to combine static and dynamic
-typing in the same code.
+typing in the same code, allowing them to incrementally migrate from
+dynamic to static typing.
+This is a benefit to programmers that use dynamically typed languages
+to build large applications, as static types detect many bugs
+during the development phase, and also provide better documentation.
 
 In this chapter we present the formalization of Typed Lua.
-We use typing rules to show how we implemented the features that we
-discussed in the previous chapter.
+Besides its practical contribution, Typed Lua also has some interesting
+contributions on the field of optional type systems for scripting
+languages that we explain in this chapter.
+Namely, it is a simple type system that covers several Lua idioms
+and features, such as the refinement of tables, multiple return values,
+and optional parameters.
+We use typing rules to explain how we implemented the features of Typed Lua.
 
 \section{Types}
+
+Our type system uses subtyping \citep{abadi1996to,cardelli1984smi} to order
+types and consistent-subtyping \citep{siek2007objects,siek2013mutable}
+to allow the interaction between statically and dynamically typed code.
+We will explain these two relations along this section.
+Firt, we will explain the subtyping relationship while we introduce
+each type that is present on our type system.
+Then, we will explain the differences between subtyping and
+consistent-subtyping, showing that they are orthogonal relations.
 
@@
 types optionally ending in a variadic type, and union of these tuples.
+
+Literal types represent the type of boolean, number, and string
+constants.
+We shall see that literal types are important to type tables as
+records.
+The following subtyping rule shows that each literal type
+is a subtype of itself:
+\[
+\begin{array}{c}
+\mylabel{S-LITERAL}\\
+\senv \vdash l \subtype l
+\end{array}
+\]
+
+The variable $\senv$ is a set of pairs of recursion variables that
+records the pairs of valid recursive types.
+
+The base types $\Boolean$, $\Number$, and $\String$ represent the
+type of values that during run-time Lua tags as booleans, numbers,
+and strings, respectively.
+Literal types are subtypes of their respective base types, and
+each base type is a subtype of itself, as the following subtyping
+rules show:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-FALSE}\\
+\senv \vdash \False \subtype \Boolean
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-TRUE}\\
+\senv \vdash \True \subtype \Boolean
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-NUMBER}\\
+\senv \vdash {\it number} \subtype \Number
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-STRING}\\
+\senv \vdash {\it string} \subtype \String
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-BASE}\\
+\senv \vdash b \subtype b
+\end{array}
+\end{array}
+\]
+
+The type $\Nil$ represents both the type of the constant \texttt{nil}
+and the type of values that during run-time Lua tags as nil.
+It is a separated type because in our type system $\Nil$ is
+not the bottom type.
+Typed Lua has this behavior because in Lua the tag \texttt{nil}
+represents the absence of an useful value.
+Thus, making $\Nil$ the bottom type would severally reduce the
+amount of type safety in our type system.
+The next subtyping rule shows this behavior:
+\[
+\begin{array}{c}
+\mylabel{S-NIL}\\
+\senv \vdash \Nil \subtype \Nil
+\end{array}
+\]
+
+The type $\Value$ is the top type, so any first-level type is a
+subtype of $\Value$.
+We shall see that this type along with variadic types help to
+preserve the semantics of Lua to drop extra values on multiple
+assignments, and function calls.
+Following we show its subtyping rule:
+\[
+\begin{array}{c}
+\mylabel{S-VALUE}\\
+\senv \vdash t \subtype \Value
+\end{array}
+\]
+
+The dynamic type $\Any$ is the type that allows programmers to mix
+static and dynamic typing.
+The type $\Any$ is neither the bottom nor the top type, but a
+separated typed that is a subtype only of itself:
+\[
+\begin{array}{c}
+\mylabel{S-ANY}\\
+\senv \vdash \Any \subtype \Any
+\end{array}
+\]
+
+If we set the type $\Any$ as both bottom and top types of our type
+system, then all programs would type check without errors.
+This would happen due to the transitivity of subtyping, that is,
+we would be able to down cast any type to $\Any$ and then up cast
+$\Any$ to any type.
+We shall see later in this section that we use consistent-subtyping
+to allow the dynamic type to interact with other first-level types,
+and thus allow dynamically typed code coexist with statically typed
+code.
+
+Typed Lua uses the type $\Self$ to represent the type of the \texttt{self}
+parameter in object-oriented Lua code.
+It is only a subtype of itself:
+\[
+\begin{array}{c}
+\mylabel{S-SELF}\\
+\senv \vdash \Self \subtype \Self
+\end{array}
+\]
+
+Union types represent types that can hold a value of several
+different, but fixed types.
+Notice that union types are disjoint, that is, all types in
+the union are not subtypes or supertypes among themselfs.
+Next we show the subtyping rules for union types:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-UNION1}\\
+\dfrac{\senv \vdash t_{1} \subtype t \;\;\;
+       \senv \vdash t_{2} \subtype t}
+      {\senv \vdash t_{1} \cup t_{2} \subtype t}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-UNION2}\\
+\dfrac{\senv \vdash t \subtype t_{1}}
+      {\senv \vdash t \subtype t_{1} \cup t_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-UNION3}\\
+\dfrac{\senv \vdash t \subtype t_{2}}
+      {\senv \vdash t \subtype t_{1} \cup t_{2}}
+\end{array}
+\end{array}
+\]
+
+The first rule shows that a union type $t_{1} \cup t_{2}$
+is subtype of $t$ if both $t_{1}$ and $t_{2}$ are subtypes
+of $t$;
+while the other rules show that a type $t$ is subtype
+of a union type $t_{1} \cup t_{2}$ if $t$ is subtype of
+$t_{1}$ or if $t$ is subtype of $t_{2}$.
+
+To introduce function types we first need to introduce
+second-level types, as they appear on both sides of function types.
+Basically, second-level types are tuples of first-level types
+that can optionally end with a variadic type.
+Our type system needs second-level types because tuples are
+not first-class values in Lua, though they may appear on
+argument passing, multiple returns, and multiple assignments.
+A variadic type $t*$ is a generator for a sequence of values of
+type $t \cup \Nil$, and it represents the type of the vararg
+expression.
+Next we show their subtyping rules:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-VARARG}\\
+\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2} \cup \Nil}
+      {\senv \vdash t_{1}* \subtype t_{2}*}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-TUPLE1}\\
+\dfrac{\senv \vdash t_{1} \subtype t_{2} \;\;\;
+       \senv \vdash s_{1} \subtype s_{2}}
+      {\senv \vdash t_{1} \times s_{1} \subtype t_{2} \times s_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-TUPLE2}\\
+\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2} \;\;\;
+       \senv \vdash t_{1}* \subtype s_{2}}
+      {\senv \vdash t_{1}* \subtype t_{2} \times s_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-TUPLE3}\\
+\dfrac{\senv \vdash t_{1} \subtype t_{2} \cup \Nil \;\;\;
+       \senv \vdash s_{1} \subtype t_{2}*}
+      {\senv \vdash t_{1} \times s_{1} \subtype t_{2}*}
+\end{array}
+\end{array}
+\]
+
+The subtyping rules for variadic and tuple types show that
+these types are covariant.
+The subtyping rule for variadic types is straightforward:
+it shows that a variadic type $t_{1}*$ is subtype of another
+variadic type $t_{2}*$ if $t_{1} \cup \Nil$ is subtype of
+$t_{2} \cup \Nil$.
+The first rule for tuple types shows that a tuple type
+$t_{1} \times s_{1}$ is subtype of another tuple type
+$t_{2} \times s_{2}$ if $t_{1}$ is subtype of $t_{2}$
+and $s_{1}$ is subtype of $s_{2}$.
+As we mentioned, a tuple can optionally end with a variadic
+type, and the second and third rules for tuple types handle
+this case.
+The second rule shows that a variadic type is subtype of
+a tuple whenever the type of the value that it generates is subtype
+of all the elements of the tuple.
+The third rule shows that a tuple is subtype of a variadic
+type whenever all the elements of the tuple are subtype of the
+type of the value that it generates.
+We shall see that these subtyping relations along with the
+types $\Value$ and $\Nil$ allow our type system to reflect
+the semantics of Lua on discarding extra parameters and
+replacing missing parameters.
 
@@
 \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
-& | \; \mathbf{local} \; \vec{n}{:}\vec{t} = \vec{e} \; \mathbf{in} \; s \; | \;
+& | \; \mathbf{local} \; \vec{n{:}t} = \vec{e} \; \mathbf{in} \; s \; | \;
 \mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
@@
 \mathbf{fun} \; ({...}{:}t){:}r \; s \; | \;
-\mathbf{fun} \; (\vec{n}{:}\vec{t}){:}r \; s \; | \;
-\mathbf{fun} \; (\vec{n}{:}\vec{t},{...}{:}t){:}r \; s\\
+\mathbf{fun} \; (\vec{n{:}t}){:}r \; s \; | \;
+\mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; s\\
 c ::= & \; [e_{1}] = e_{2}\\

### 2014-10-13 doc/thesis/system.tex
Adding integer to formal type system
Fixing the type checking of multiple assignments

@@
 
-\begin{figure}[!ht]
-\textbf{Subtyping}\\
-\dstart
-\begin{footnotesize}
-$$
+In our type system function types have the form $p \rightarrow r$,
+where $p$ is the type of the parameters list and $r$ is the
+return type.
+The type of the parameters list can be $\Void$ or a tuple type,
+while the return type can be $\Void$, a tuple type, or a union of tuples.
+The type $\Void$ means an empty tuple.
+Our type system includes union of tuples on the return type because
+in Lua it is common to overload the return type to denote error,
+as we mentioned on Section \ref{sec:statistics}.
+
+The type $\Void$ and union of tuple types are also second-level types.
+The type $\Void$ is subtype only of itself.
+The subtyping rules for union of tuple types are similar to the
+subtyping rules for union of first-level types.
+Next we show the subtyping rules for these second-level types:
+\[
 \begin{array}{c}
 \begin{array}{c}
-\mylabel{S-LITERAL}\\
-\senv \vdash L \subtype L
+\mylabel{S-VOID}\\
+\senv \vdash \Void \subtype \Void
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-UNION4}\\
+\dfrac{\senv \vdash s_{1} \subtype s \;\;\;
+       \senv \vdash s_{2} \subtype s}
+      {\senv \vdash s_{1} \sqcup s_{2} \subtype s}
 \end{array}
@@
 \begin{array}{c}
-\mylabel{S-FALSE}\\
-\senv \vdash \False \subtype \Boolean
+\mylabel{S-UNION5}\\
+\dfrac{\senv \vdash s \subtype s_{1}}
+      {\senv \vdash s \subtype s_{1} \sqcup s_{2}}
 \end{array}
@@
 \begin{array}{c}
-\mylabel{S-TRUE}\\
-\senv \vdash \True \subtype \Boolean
+\mylabel{S-UNION6}\\
+\dfrac{\senv \vdash s \subtype s_{2}}
+      {\senv \vdash s \subtype s_{1} \sqcup s_{2}}
 \end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-NUMBER}\\
-\senv \vdash {\it number} \subtype \Number
 \end{array}
-\;
-\begin{array}{c}
-\mylabel{S-STRING}\\
-\senv \vdash {\it string} \subtype \String
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-BASE}\\
-\senv \vdash B \subtype B
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-NIL}\\
-\senv \vdash \Nil \subtype \Nil
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-VALUE}\\
-\senv \vdash T \subtype \Value
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-ANY}\\
-\senv \vdash \Any \subtype \Any
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-SELF}\\
-\senv \vdash \Self \subtype \Self
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-UNION1}\\
-\dfrac{\senv \vdash T_{1} \subtype T \;\;\;
-       \senv \vdash T_{2} \subtype T}
-      {\senv \vdash T_{1} \cup T_{2} \subtype T}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-UNION2}\\
-\dfrac{\senv \vdash T \subtype T_{1}}
-      {\senv \vdash T \subtype T_{1} \cup T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-UNION3}\\
-\dfrac{\senv \vdash T \subtype T_{2}}
-      {\senv \vdash T \subtype T_{1} \cup T_{2}}
-\end{array}
-\\ \\
+\]
+
+The subtyping rule for function types is standard, that is, it is
+contravariant on the type of the parameters list and
+covariant on the return type:
+\[
 \begin{array}{c}
 \mylabel{S-FUNCTION}\\
-\dfrac{\senv \vdash P_{2} \subtype P_{1} \;\;\;
-       \senv \vdash R_{1} \subtype R_{2}}
-      {\senv \vdash P_{1} \rightarrow R_{1} \subtype P_{2} \rightarrow R_{2}}
+\dfrac{\senv \vdash p_{2} \subtype p_{1} \;\;\;
+       \senv \vdash r_{1} \subtype r_{2}}
+      {\senv \vdash p_{1} \rightarrow r_{1} \subtype p_{2} \rightarrow r_{2}}
 \end{array}
-\;
+\]
+
+\[
+\begin{array}{c}
 \begin{array}{c}
@@
 \dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
-       \senv \vdash F_{j} \subtype F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{c} \subtype \{F_{1}, ..., F_{n}\}}
+       \senv \vdash f_{j} \subtype f_{i}}
+      {\senv \vdash \{f_{1}, ..., f_{m}\}_{c} \subtype \{f_{1}, ..., f_{n}\}}
 \end{array}
@@
 \mylabel{S-TABLE2}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash F_{i}' \subtype F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \subtype \{F_{1}', ..., F_{m}', ..., F_{n}'\}}
+\dfrac{\forall i \in 1..m \; \senv \vdash f_{i}' \subtype f_{i}}
+      {\senv \vdash \{f_{1}, ..., f_{m}\}_{o} \subtype \{f_{1}', ..., f_{m}', ..., f_{n}'\}}
 \end{array}
@@
 \mylabel{S-TABLE3}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash F \subtype F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \subtype \{F\}} 
+\dfrac{\forall i \in 1..m \; \senv \vdash f \subtype f_{i}}
+      {\senv \vdash \{f_{1}, ..., f_{m}\}_{o} \subtype \{f\}} 
 \end{array}
@@
 \mylabel{S-FIELD1}\\
-\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
-       \senv \vdash T_{1} \subtype T_{2} \;\;\;
-       \senv \vdash T_{2} \subtype T_{1}}
-      {\senv \vdash K_{1}:T_{1} \subtype K_{2}:T_{2}}
+\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
+       \senv \vdash t_{1} \subtype t_{2} \;\;\;
+       \senv \vdash t_{2} \subtype t_{1}}
+      {\senv \vdash k_{1}:t_{1} \subtype k_{2}:t_{2}}
 \end{array}
@@
 \mylabel{S-FIELD2}\\
-\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
-       \senv \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash \Const \; K_{1}:T_{1} \subtype \Const \; K_{2}:T_{2}}
+\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
+       \senv \vdash t_{1} \subtype t_{2}}
+      {\senv \vdash \Const \; k_{1}:t_{1} \subtype \Const \; k_{2}:t_{2}}
 \end{array}
@@
 \mylabel{S-FIELD3}\\
-\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
-       \senv \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash K_{1}:T_{1} \subtype \Const \; K_{2}:T_{2}}
+\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
+       \senv \vdash t_{1} \subtype t_{2}}
+      {\senv \vdash k_{1}:t_{1} \subtype \Const \; k_{2}:t_{2}}
+\end{array}
+\end{array}
+\]
+
+Our type system also includes recursive types: $\mu x.t$,
+where $t$ is a function of table type with $x$ appearing anywhere
+a function of table type could appear.
+For instance, $\mu x.\{``info":\Number, ``next":x \cup \Nil\}$
+represents the type of single linked lists of numbers.
+We introduce the following subtyping rules for recursive types:
+
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-VARIABLE}\\
+\dfrac{x_{1} \subtype x_{2} \in \senv}
+      {\senv \vdash x_{1} \subtype x_{2}}
 \end{array}
@@
 \begin{array}{c}
-\mylabel{S-VARIABLE}\\
-\dfrac{X_{1} \subtype X_{2} \in \senv}
-      {\senv \vdash X_{1} \subtype X_{2}}
+\mylabel{S-AMBER}\\
+\dfrac{\senv, x_{1} \subtype x_{2} \vdash t_{1} \subtype t_{2}}
+      {\senv \vdash \mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-AMBERL}\\
+\dfrac{\senv \vdash t_{1}[x \mapsto \mu x.t_{1}] \subtype t_{2}}
+      {\senv \vdash \mu x.t_{1} \subtype t_{2}}
 \end{array}
@@
 \begin{array}{c}
-\mylabel{S-RECURSIVE1}\\
-\dfrac{\senv, X_{1} \subtype X_{2} \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash \mu X_{1}.T_{1} \subtype \mu X_{2}.T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-RECURSIVE2}\\
-\dfrac{X \subtype X \not\in \senv \;\;\;
-      \senv, X \subtype X \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash \mu X.T_{1} \subtype T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-RECURSIVE3}\\
-\dfrac{X \subtype X \in \senv \;\;\;
-       \senv \vdash X \subtype T_{2}}
-      {\senv \vdash \mu X.T_{1} \subtype T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-RECURSIVE4}\\
-\dfrac{X \subtype X \not\in \senv \;\;\;
-       \senv, X \subtype X \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash T_{1} \subtype \mu X.T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-RECURSIVE5}\\
-\dfrac{X \subtype X \in \senv \;\;\;
-       \senv \vdash T_{1} \subtype X}
-      {\senv \vdash T_{1} \subtype \mu X.T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-VOID}\\
-\senv \vdash \Void \subtype \Void
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-VARARG}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \cup \Nil}
-      {\senv \vdash T_{1}* \subtype T_{2}*}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TUPLE1}\\
-\dfrac{\senv \vdash T_{1} \subtype T_{2} \;\;\;
-       \senv \vdash S_{1} \subtype S_{2}}
-      {\senv \vdash T_{1} \times S_{1} \subtype T_{2} \times S_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-TUPLE2}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \;\;\;
-       \senv \vdash T_{1}* \subtype S_{2}}
-      {\senv \vdash T_{1}* \subtype T_{2} \times S_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TUPLE3}\\
-\dfrac{\senv \vdash T_{1} \subtype T_{2} \cup \Nil \;\;\;
-       \senv \vdash S_{1} \subtype T_{2}*}
-      {\senv \vdash T_{1} \times S_{1} \subtype T_{2}*}
+\mylabel{S-AMBERR}\\
+\dfrac{\senv \vdash t_{1} \subtype t_{2}[x \mapsto \mu x.t_{2}]}
+      {\senv \vdash t_{1} \subtype \mu x.t_{2}}
 \end{array}
 \end{array}
-$$
-\end{footnotesize}
-\dend
-\caption{The subtyping relation}
-\label{fig:subtyping}
-\end{figure}
+\]
 
-\begin{figure}[!ht]
-\textbf{Consistent-Subtyping}\\
-\dstart
-\begin{footnotesize}
-$$
+Our type system also introduces projection types.
+We shall see in the next section that our type system uses
+projection types to handle union of typle types when they
+appear on the right-hand side of the declaration of local variables.
+For this reason, they do not have subtyping rules.
+
+\[
 \begin{array}{c}
 \begin{array}{c}
-\mylabel{C-LITERAL}\\
-\senv \vdash L \lesssim L
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-FALSE}\\
-\senv \vdash \False \lesssim \Boolean
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TRUE}\\
-\senv \vdash \True \lesssim \Boolean
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-NUMBER}\\
-\senv \vdash {\it number} \lesssim \Number
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-STRING}\\
-\senv \vdash {\it string} \lesssim \String
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-BASE}\\
-\senv \vdash B \lesssim B
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-NIL}\\
-\senv \vdash \Nil \lesssim \Nil
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-VALUE}\\
-\senv \vdash T \lesssim \Value
-\end{array}
-\;
-\begin{array}{c}
 \mylabel{C-ANY1}\\
-\senv \vdash T \lesssim \Any
+\senv \vdash t \lesssim \Any
 \end{array}
@@
 \mylabel{C-ANY2}\\
-\senv \vdash \Any \lesssim T
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-SELF}\\
-\senv \vdash \Self \lesssim \Self
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-UNION1}\\
-\dfrac{\senv \vdash T_{1} \lesssim T \;\;\;
-       \senv \vdash T_{2} \lesssim T}
-      {\senv \vdash T_{1} \cup T_{2} \lesssim T}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-UNION2}\\
-\dfrac{\senv \vdash T \lesssim T_{1}}
-      {\senv \vdash T \lesssim T_{1} \cup T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-UNION3}\\
-\dfrac{\senv \vdash T \lesssim T_{2}}
-      {\senv \vdash T \lesssim T_{1} \cup T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-FUNCTION}\\
-\dfrac{\senv \vdash P_{2} \lesssim P_{1} \;\;\;
-       \senv \vdash R_{1} \lesssim R_{2}}
-      {\senv \vdash P_{1} \rightarrow R_{1} \lesssim P_{2} \rightarrow R_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TABLE1}\\
-\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
-       \senv \vdash F_{j} \lesssim F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{c} \lesssim \{F_{1}, ..., F_{n}\}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-TABLE2}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash F_{i}' \lesssim F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \lesssim \{F_{1}', ..., F_{m}', ..., F_{n}'\}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TABLE3}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash F \lesssim F_{i}}
-      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \lesssim \{F\}} 
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-FIELD1}\\
-\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
-       \senv \vdash T_{1} \lesssim T_{2} \;\;\;
-       \senv \vdash T_{2} \lesssim T_{1}}
-      {\senv \vdash K_{1}:T_{1} \lesssim K_{2}:T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-FIELD2}\\
-\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
-       \senv \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash \Const \; K_{1}:T_{1} \lesssim \Const \; K_{2}:T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-FIELD3}\\
-\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
-       \senv \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash K_{1}:T_{1} \lesssim \Const \; K_{2}:T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-VARIABLE}\\
-\dfrac{X_{1} \lesssim X_{2} \in \senv}
-      {\senv \vdash X_{1} \lesssim X_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-RECURSIVE1}\\
-\dfrac{\senv, X_{1} \lesssim X_{2} \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash \mu X_{1}.T_{1} \lesssim \mu X_{2}.T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-RECURSIVE2}\\
-\dfrac{X \lesssim X \not\in \senv \;\;\;
-      \senv, X \lesssim X \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash \mu X.T_{1} \lesssim T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-RECURSIVE3}\\
-\dfrac{X \lesssim X \in \senv \;\;\;
-       \senv \vdash X \lesssim T_{2}}
-      {\senv \vdash \mu X.T_{1} \lesssim T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-RECURSIVE4}\\
-\dfrac{X \lesssim X \not\in \senv \;\;\;
-       \senv, X \lesssim X \vdash T_{1} \lesssim T_{2}}
-      {\senv \vdash T_{1} \lesssim \mu X.T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-RECURSIVE5}\\
-\dfrac{X \lesssim X \in \senv \;\;\;
-       \senv \vdash T_{1} \lesssim X}
-      {\senv \vdash T_{1} \lesssim \mu X.T_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-VOID}\\
-\senv \vdash \Void \lesssim \Void
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-VARARG}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \lesssim T_{2} \cup \Nil}
-      {\senv \vdash T_{1}* \lesssim T_{2}*}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TUPLE1}\\
-\dfrac{\senv \vdash T_{1} \lesssim T_{2} \;\;\;
-       \senv \vdash S_{1} \lesssim S_{2}}
-      {\senv \vdash T_{1} \times S_{1} \lesssim T_{2} \times S_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{C-TUPLE2}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \lesssim T_{2} \;\;\;
-       \senv \vdash T_{1}* \lesssim S_{2}}
-      {\senv \vdash T_{1}* \lesssim T_{2} \times S_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-TUPLE3}\\
-\dfrac{\senv \vdash T_{1} \lesssim T_{2} \cup \Nil \;\;\;
-       \senv \vdash S_{1} \lesssim T_{2}*}
-      {\senv \vdash T_{1} \times S_{1} \lesssim T_{2}*}
+\senv \vdash \Any \lesssim t
 \end{array}
 \end{array}
-$$
-\end{footnotesize}
-\dend
-\caption{The consistent-subtyping relation}
-\label{fig:consistent_subtyping}
-\end{figure}
+\]
 

### 2014-10-13 doc/thesis/system.tex
Adding more text about subtyping

@@
 We showed that programmers can use Typed Lua to combine static and dynamic
-typing in the same code, allowing them to incrementally migrate from
+typing in the same code, and it allows them to incrementally migrate from
 dynamic to static typing.
@@
 contributions on the field of optional type systems for scripting
-languages that we explain in this chapter.
+languages.
 Namely, it is a simple type system that covers several Lua idioms
@@
 
-Our type system uses subtyping \citep{abadi1996to,cardelli1984smi} to order
+Our type system uses subtyping \citep{cardelli1984smi,abadi1996to} to order
 types and consistent-subtyping \citep{siek2007objects,siek2013mutable}
 to allow the interaction between statically and dynamically typed code.
-We will explain these two relations along this section.
-Firt, we will explain the subtyping relationship while we introduce
-each type that is present on our type system.
-Then, we will explain the differences between subtyping and
-consistent-subtyping, showing that they are orthogonal relations.
+We will explain the subtyping and consistent-subtyping rules along this
+section while we introduce each type from our type system.
+We will focus the discussion on the definition of subtyping because,
+as we mentioned in Chapter \ref{chap:review}, we achieve the definition of
+consistent-subtyping combining consistency and subtyping.
+The differences between subtyping and consistent-subtyping are the way
+they handle the dynamic type, and the fact that subtyping is transitive,
+but consistent-subtyping is not.
 
@@
 & | \; p \rightarrow r & \textit{function types}\\
-& | \; \{f_{i}, ..., f_{n}\}_{c|o|u} & \textit{table types}\\
+& | \; \{v_{i}, ..., v_{n}\}_{closed|open|unique} & \textit{table types}\\
 & | \; x & \textit{type variables}\\
 & | \; \mu x.t & \textit{recursive types}\\
-& | \; x_{i} & \textit{projection types}\\
-l ::= & \False \; | \; \True \; | \; {\it number} \; | \; {\it string} & \\
-b ::= & \Boolean \; | \; \Number \; | \; \String & \\
-f ::= & k:t \; | \; \Const \; k:t & \textit{field types}\\ 
+& | \; \pi_{i} & \textit{projection types}\\
+l ::= & \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} & \\
+b ::= & \Boolean \; | \; \Integer \; | \; \Number \; | \; \String & \\
+v ::= & k:t \; | \; \Const \; k:t & \textit{field types}\\ 
 k ::= & l \; | \; b \; | \; \Any & \textit{key types}\\
@@
 p ::= & \Void \; | \; s & \textit{parameters list type}\\
-r ::= & \Void \; | \; s \; | \; s \cup r & \textit{return types}\\
+r ::= & \Void \; | \; s \; | \; s \sqcup r & \textit{return types}\\
 \end{array}
@@
 
-Literal types represent the type of boolean, number, and string
-constants.
-We shall see that literal types are important to type tables as
-records.
-The following subtyping rule shows that each literal type
-is a subtype of itself:
+Literal types represent the type of literal constants.
+They can be the boolean values $\False$ and $\True$,
+any integer value, any floating point value, and any string value.
+We shall see that literal types are important to type tables as records.
+
+Our type system also includes the following base types: $\Boolean$,
+$\Integer$, $\Number$, and $\String$.
+The base types $\Boolean$ and $\String$ represent the values that
+during run-time Lua tags as \texttt{boolean} and \texttt{string},
+respectively;
+while the base types $\Integer$ and $\Number$ represent the values
+that during run-time Lua tags as \texttt{number}.
+We introduced two different base types to represent Lua numbers
+because Lua introduced integers recently.
+
+The subtyping rules for literal types include a reflexive rule
+and the rules for defining that literal types are subtypes of
+their respective base types;
+the subtyping rules for base types include a reflexive rule
+and another rule for defining that $\Integer$ is subtype of
+$\Number$.
+We defined these subtyping rules as follows:
 \[
+\begin{array}{c}
 \begin{array}{c}
@@
 \end{array}
-\]
-
-The variable $\senv$ is a set of pairs of recursion variables that
-records the pairs of valid recursive types.
-
-The base types $\Boolean$, $\Number$, and $\String$ represent the
-type of values that during run-time Lua tags as booleans, numbers,
-and strings, respectively.
-Literal types are subtypes of their respective base types, and
-each base type is a subtype of itself, as the following subtyping
-rules show:
-\[
-\begin{array}{c}
+\;
 \begin{array}{c}
@@
 \begin{array}{c}
-\mylabel{S-NUMBER}\\
-\senv \vdash {\it number} \subtype \Number
+\mylabel{S-INT1}\\
+\senv \vdash {\it int} \subtype \Integer
 \end{array}
 \;
+\begin{array}{c}
+\mylabel{S-INT2}\\
+\senv \vdash {\it int} \subtype \Number
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FLOAT}\\
+\senv \vdash {\it float} \subtype \Number
+\end{array}
+\\ \\
 \begin{array}{c}
@@
 \end{array}
-\;
 \begin{array}{c}
@@
 \end{array}
+\;
+\begin{array}{c}
+\mylabel{S-NUMBER}\\
+\senv \vdash \Integer \subtype \Number
+\end{array}
 \end{array}
@@
 
-The type $\Nil$ represents both the type of the constant \texttt{nil}
-and the type of values that during run-time Lua tags as nil.
+The variable $\senv$ is a set of pairs of recursion variables.
+We shall notice that we need this set to record the pairs of valid
+recursive types.
+
+Since subtyping is reflexive and transitive,
+we could have ommited the reflexive rules \textsc{S-LITERAL} and
+\textsc{S-BASE}, as well as the rule \textsc{S-INT2}.
+More precisely, we could have used subsumption rules that define
+the reflexiveness and transitiveness of subtyping.
+In the case of \textsc{S-INT2}, we would not need this rule because
+we could use the transitive subsumption rule to derive that
+\[
+\dfrac{\senv \vdash 1 \subtype \Integer \;\;\;
+       \senv \vdash \Integer \subtype \Number}
+      {\senv \vdash 1 \subtype \Number}
+\]
+
+However, we are using the subtyping rules as the template for defining
+the consistent-subtyping rules, and consistent-subtyping is not
+transitive.
+More precisely, we want the subtyping and consistent-subtyping rules
+to differ only in the way they handle the dynamic type.
+Thus, we define the subtyping rules using an algorithmic
+way that is close to the implementation, as this way allows us to use
+subtyping to formalize consistent-subtyping.
+
+The type $\Nil$ represents the type of values that during run-time
+Lua tags as \texttt{nil}.
 It is a separated type because in our type system $\Nil$ is
@@
 
-If we set the type $\Any$ as both bottom and top types of our type
-system, then all programs would type check without errors.
+Even though the dynamic type $\Any$ is neutral to subtyping,
+it is not to consistent-subtyping:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{C-ANY1}\\
+\senv \vdash t \lesssim \Any
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-ANY2}\\
+\senv \vdash \Any \lesssim t
+\end{array}
+\end{array}
+\]
+
+The rules that handle the dynamic type are the only rules that
+differ between subtyping and consistent-subtyping.
+If we set the type $\Any$ as both bottom and top types of our
+subtyping relation, then all programs would type check without errors.
 This would happen due to the transitivity of subtyping, that is,
-we would be able to down cast any type to $\Any$ and then up cast
+we would be able to down-cast any type to $\Any$ and then up-cast
 $\Any$ to any type.
-We shall see later in this section that we use consistent-subtyping
-to allow the dynamic type to interact with other first-level types,
-and thus allow dynamically typed code coexist with statically typed
-code.
+While the rule \textsc{S-ANY} is neutral and makes transitivity
+not be a problem, the rules \textsc{C-ANY1} and \textsc{C-ANY2}
+are the rules that allow the dynamic type to interact with other
+first-level types, and thus allow dynamically typed code coexist
+with statically typed code.
+Consistent-subtyping cannot be transitive because of these rules,
+as transitivity combined with them would make all programs compile
+without type errors.
 
-Typed Lua uses the type $\Self$ to represent the type of the \texttt{self}
-parameter in object-oriented Lua code.
-It is only a subtype of itself:
+Typed Lua uses the type $\Self$ to represent \emph{objects}.
+As we mentioned in Chapter \ref{chap:typedlua}, we need the type
+$\Self$ to prevent programmers from indexing a method without
+calling it with the correct receiver.
+The type $\Self$ is only a subtype of itself:
 \[
@@
 different, but fixed types.
-Notice that union types are disjoint, that is, all types in
-the union are not subtypes or supertypes among themselfs.
 Next we show the subtyping rules for union types:
@@
 of $t$;
-while the other rules show that a type $t$ is subtype
+and the other rules show that a type $t$ is subtype
 of a union type $t_{1} \cup t_{2}$ if $t$ is subtype of
 $t_{1}$ or if $t$ is subtype of $t_{2}$.
+
+Union types are disjoint, that is, the types in the union are not
+consistent-subtype of any other type in the same union.
+For instance, the union type $\Boolean \;\cup\; \Any$ results the
+type $\Any$, because $\Boolean$ is consistent-subtype of $\Any$.
+The union type $\Number \;\cup\; \Nil \;\cup\; 1$ results the union
+type $\Number \;\cup\; \Nil$ because $1 \subtype \Number$.
 
@@
 in Lua it is common to overload the return type to denote error,
-as we mentioned on Section \ref{sec:statistics}.
+as we mentioned in Section \ref{sec:statistics}.
 
@@
 
+Table types have only one syntactical form, but they also have one of
+these three tags: \emph{closed}, \emph{open}, and \emph{unique}.
+The tag \emph{closed} is the tag of the table types that the type
+annotations describe.
+The tag \emph{open} is the tag of the table types that we can
+add new fields through field assignment.
+The tag \emph{unique} is the tag of the table types that represent
+the type of the table constructors.
+The syntactical form of table types is $\{ v_{i}, ..., v_{n} \}$,
+where $i$ varies from 1 to $n$ when $n > 0$,
+but $i$ can also be 0 when $n = 0$.
+
 \[
@@
 \dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
-       \senv \vdash f_{j} \subtype f_{i}}
-      {\senv \vdash \{f_{1}, ..., f_{m}\}_{c} \subtype \{f_{1}, ..., f_{n}\}}
+       \senv \vdash v_{j} \subtype v_{i}}
+      {\senv \vdash \{v_{1}, ..., v_{m}\}_{closed} \subtype \{v_{1}, ..., v_{n}\}_{closed}}
 \end{array}
@@
 \mylabel{S-TABLE2}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash f_{i}' \subtype f_{i}}
-      {\senv \vdash \{f_{1}, ..., f_{m}\}_{o} \subtype \{f_{1}', ..., f_{m}', ..., f_{n}'\}}
+\dfrac{\forall i \in 1..m \; \senv \vdash v_{i}' \subtype v_{i}}
+      {\senv \vdash \{v_{1}, ..., v_{m}\}_{open} \subtype \{v_{1}', ..., v_{m}', ..., v_{n}'\}_{open|closed}}
 \end{array}
-\;
+\\ \\
 \begin{array}{c}
 \mylabel{S-TABLE3}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash f \subtype f_{i}}
-      {\senv \vdash \{f_{1}, ..., f_{m}\}_{o} \subtype \{f\}} 
+\dfrac{\forall i \in 1..m \; \senv \vdash v \subtype v_{i}}
+      {\senv \vdash \{v_{1}, ..., v_{m}\}_{open} \subtype \{v\}_{open|closed}} 
 \end{array}
 \\ \\
+\begin{array}{c}
+\mylabel{S-TABLE4}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash v_{i}' \subtype_{n} v_{i}}
+      {\senv \vdash \{v_{1}, ..., v_{m}\}_{unique} \subtype \{v_{1}', ..., v_{m}', ..., v_{n}'\}_{unique|open|closed}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-TABLE5}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash v \subtype_{n} v_{i}}
+      {\senv \vdash \{v_{1}, ..., v_{m}\}_{unique} \subtype \{v\}_{unique|open|closed}} 
+\end{array}
+\end{array}
+\]
+
+\[
+\begin{array}{c}
 \begin{array}{c}
@@
 \end{array}
-\;
+\\ \\
 \begin{array}{c}
@@
 \end{array}
-\\ \\
+\;
 \begin{array}{c}
@@
 
-Our type system also includes recursive types: $\mu x.t$,
-where $t$ is a function of table type with $x$ appearing anywhere
-a function of table type could appear.
-For instance, $\mu x.\{``info":\Number, ``next":x \cup \Nil\}$
-represents the type of single linked lists of numbers.
-We introduce the following subtyping rules for recursive types:
-
 \[
@@
 \begin{array}{c}
-\mylabel{S-VARIABLE}\\
+\mylabel{S-FIELD4}\\
+\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
+       \senv \vdash t_{1} \subtype t_{2}}
+      {\senv \vdash k_{1}:t_{1} \subtype_{n} k_{2}:t_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-FIELD5}\\
+\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
+       \senv \vdash t_{1} \subtype t_{2}}
+      {\senv \vdash \Const \; k_{1}:t_{1} \subtype_{n} \Const \; k_{2}:t_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FIELD6}\\
+\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
+       \senv \vdash t_{1} \subtype t_{2}}
+      {\senv \vdash k_{1}:t_{1} \subtype_{n} \Const \; k_{2}:t_{2}}
+\end{array}
+\end{array}
+\]
+
+Table types can be ambiguous.
+For instance, $\{1:\Number, \Number:\String\}$ is an ambiguous
+table type because the type of the value stored by key $1$ can be
+$\Number$ or $\String$, as $1 \subtype 1$ and
+$1 \subtype \Number$.
+Table types can get even more ambiguous when we use the dynamic
+type in the type of table keys.
+For instance, the table type $\{1:\Number, \Number:\String, \Any:\Boolean\}$
+is ambiguous because the type of the value stored by key $1$ can be
+$\Number$, or $\String$, or $\Boolean$, as $1 \subtype 1$,
+$1 \subtype \Number$, and $1 \lesssim \Any$.
+The type of the value stored by a key of type $\Number$,
+which is not the literal type $1$, can be $\Number$ or $\Boolean$,
+as $\Number \subtype \Number$, and $\Number \lesssim \Any$.
+To avoid ambiguous table types we use the following rule to check
+whether a table type is well formed:
+\[
+\forall i \not\exists j \; i \not= j \wedge k_{i} \lesssim k_{j}
+\]
+
+Our type system also includes recursive types: $\mu x.t$,
+where $t$ is a function or table type with $x$ appearing anywhere
+a function or table type could appear.
+For instance, $\mu x.\{``info":\Number, ``next":x \;\cup\; \Nil\}$
+represents the type of single linked lists of numbers.
+
+We use the \emph{Amber rule} from the Amber language
+\citep{cardelli1986amber} to define the subtyping rule for
+recursive types:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-ASSUMPTION}\\
 \dfrac{x_{1} \subtype x_{2} \in \senv}
@@
 \mylabel{S-AMBER}\\
-\dfrac{\senv, x_{1} \subtype x_{2} \vdash t_{1} \subtype t_{2}}
+\dfrac{\senv[x_{1} \subtype x_{2}] \vdash t_{1} \subtype t_{2}}
       {\senv \vdash \mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-AMBERL}\\
-\dfrac{\senv \vdash t_{1}[x \mapsto \mu x.t_{1}] \subtype t_{2}}
-      {\senv \vdash \mu x.t_{1} \subtype t_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-AMBERR}\\
-\dfrac{\senv \vdash t_{1} \subtype t_{2}[x \mapsto \mu x.t_{2}]}
-      {\senv \vdash t_{1} \subtype \mu x.t_{2}}
 \end{array}
@@
 
-Our type system also introduces projection types.
-We shall see in the next section that our type system uses
-projection types to handle union of typle types when they
-appear on the right-hand side of the declaration of local variables.
-For this reason, they do not have subtyping rules.
+The rule \textsc{S-AMBER} needs the rule \textsc{S-ASSUMPTION}
+to show that $\mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}$.
+The rule \textsc{S-ASSUMPTION} checks if the assumption
+$x_{1} \subtype x_{2}$ exists in the set of assumptions $\senv$.
+The rule \textsc{S-AMBER} extends the set of assumptions
+with the assumption $x_{1} \subtype x_{2}$ to check whether
+$t_{1} \subtype t_{2}$ or not.
 
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{C-ANY1}\\
-\senv \vdash t \lesssim \Any
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-ANY2}\\
-\senv \vdash \Any \lesssim t
-\end{array}
-\end{array}
-\]
+Our type system also introduces projection types, but they do not
+appear on the subtyping and consistent-subtyping rules.
+They are just a mechanism that the type system uses for handling
+union of tuple types, when they appear on the right-hand side of
+the declaration of local variables.
+That is the reason why we do not need to define subtyping and
+consistent-subtyping rules for them.
+We will see projection types in more detail in the next section.
 
@@
 \end{figure}
-
-Rule for checking if a table type is well formed:
-\[
-\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
-\]

### 2014-10-14 doc/thesis/system.tex
Adding text about the tags of table types

@@
 Since subtyping is reflexive and transitive,
-we could have ommited the reflexive rules \textsc{S-LITERAL} and
+we could have omitted the reflexive rules \textsc{S-LITERAL} and
 \textsc{S-BASE}, as well as the rule \textsc{S-INT2}.
@@
 
+\emph{/* Explain how Typed Lua refine union types */}
+
 To introduce function types we first need to introduce
@@
 
-Table types have only one syntactical form, but they also have one of
-these three tags: \emph{closed}, \emph{open}, and \emph{unique}.
-The tag \emph{closed} is the tag of the table types that the type
-annotations describe.
-The tag \emph{open} is the tag of the table types that we can
-add new fields through field assignment.
-The tag \emph{unique} is the tag of the table types that represent
-the type of the table constructors.
-The syntactical form of table types is $\{ v_{i}, ..., v_{n} \}$,
-where $i$ varies from 1 to $n$ when $n > 0$,
-but $i$ can also be 0 when $n = 0$.
+Our type system includes table types to represent Lua tables.
+The syntactical form of table types is $\{ v_{i}, ..., v_{n} \}_{tag}$.
+When $n = 0$, then $i = 0$ and it represents the type of an empty table.
+When $n > 0$, then $i$ varies from 1 to $n$, and each $v_{i}$
+represents the type of a table field.
+Table fields represent maps from keys to values, and we restricted
+the type of keys to literal types, base types, and the dynamic type;
+but we can use any first-level type to represent the type of values.
+We made this restriction to the type of keys because the statistics
+that we discussed in Section \ref{sec:statistics} showed that most
+of the tables are records, lists, and hashes.
+The dynamic type is an option when we need a more loosen table type.
+
+We also use the tags \emph{closed}, \emph{open}, and \emph{unique}
+to classify table types, though they have only one syntactical form.
+The first tag represents the table types that the type annotations,
+interface, and userdata declarations describe.
+The second tag represents the table types that we can use field assignment
+to add new fields to them.
+The third tag represents the table types of table constructors.
+We have different subtyping rules for table types according to their
+tags.
+We will discuss these subtyping rules in the next paragraphs to
+explain why we needed different tags.
+
+The following subtyping rules show that \emph{closed} table types
+have width subtyping, and depth subtyping on $\Const$ fields:
 
@@
 \\ \\
-\begin{array}{c}
-\mylabel{S-TABLE2}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash v_{i}' \subtype v_{i}}
-      {\senv \vdash \{v_{1}, ..., v_{m}\}_{open} \subtype \{v_{1}', ..., v_{m}', ..., v_{n}'\}_{open|closed}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-TABLE3}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash v \subtype v_{i}}
-      {\senv \vdash \{v_{1}, ..., v_{m}\}_{open} \subtype \{v\}_{open|closed}} 
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-TABLE4}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash v_{i}' \subtype_{n} v_{i}}
-      {\senv \vdash \{v_{1}, ..., v_{m}\}_{unique} \subtype \{v_{1}', ..., v_{m}', ..., v_{n}'\}_{unique|open|closed}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-TABLE5}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash v \subtype_{n} v_{i}}
-      {\senv \vdash \{v_{1}, ..., v_{m}\}_{unique} \subtype \{v\}_{unique|open|closed}} 
-\end{array}
-\end{array}
-\]
-
-\[
-\begin{array}{c}
 \begin{array}{c}
@@
 
+First, we had only \emph{closed} table types.
+However, programmers often define tables that include optional fields,
+and the subtyping rules we introduced do not let programmers to use
+the table constructor to initialize table types that include optional
+fields.
+For instance, consider the table type we introduced in
+Chapter 3:
+\begin{align*}
+\{ & ``firstname":\String, \\
+   & ``middlename":\String \cup \Nil,\\
+   & ``lastname":\String \}_{closed}
+\end{align*}
+Following we show a table type that Typed Lua could assign to a
+table constructor that does not include the field $``middlename"$,
+but such table type is not subtype of the above table type:
+\begin{align*}
+\{ ``firstname":\String, ``lastname":\String \}_{closed}
+\end{align*}
+
+We introduced \emph{open} table types to allow the refinement
+of table types, and also to allow programmers to omit optional
+fields in the initialization of table types.
+To do that, we defined new subtyping rules that allow \emph{open}
+table types to omit fields when $\Nil$ is subtype of the of type
+of the value of the omitted key.
+The following subtyping rules show this behavior:
+
 \[
 \begin{array}{c}
+\begin{array}{c}
+\mylabel{S-TABLE2}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash v_{i}' \subtype v_{i}}
+      {\senv \vdash \{v_{1}, ..., v_{m}\}_{open} \subtype \{v_{1}', ..., v_{m}', ..., v_{n}'\}_{open|closed}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-TABLE3}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash v \subtype v_{i}}
+      {\senv \vdash \{v_{1}, ..., v_{m}\}_{open} \subtype \{v\}_{open|closed}} 
+\end{array}
+\end{array}
+\]
+
+However, we still had the that
+\begin{align*}
+\{ & ``firstname":\String, \\
+   & ``middlename":\String,\\
+   & ``lastname":\String \}_{open} \not \subtype \\
+\{ & ``firstname":\String, \\
+   & ``middlename":\String \cup \Nil,\\
+   & ``lastname":\String \}_{closed}
+\end{align*}
+
+We introduced \emph{unique} table types to solve this problem.
+We use \emph{unique} table types only to denote the type of the
+table constructor.
+The following subtyping rules allow us to use table constructors
+that initialize optional fields:
+
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-TABLE4}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash v_{i}' \subtype_{n} v_{i}}
+      {\senv \vdash \{v_{1}, ..., v_{m}\}_{unique} \subtype \{v_{1}', ..., v_{m}', ..., v_{n}'\}_{unique|open|closed}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-TABLE5}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash v \subtype_{n} v_{i}}
+      {\senv \vdash \{v_{1}, ..., v_{m}\}_{unique} \subtype \{v\}_{unique|open|closed}} 
+\end{array}
+\\ \\
 \begin{array}{c}
@@
 
+Notice that we introduced a special subtyping rule to check the
+subtyping of \emph{unique} table fields.
+The difference is that \emph{unique} table types have width and
+depth subtyping on all fields.
+
+Even though \emph{unique} table types solve the problem of using
+table constructors to initialize tables that contain optional fields,
+we still have the problem when we try to use a local variable that
+holds an \emph{open} table to initialize another table that contains
+optional fields.
+
+\emph{/* Do we really need closed, open, and unique? */}
+
+\emph{/* Also, can we make these subtyping rules for tables more clear? */}
+
 Table types can be ambiguous.
@@
 as $\Number \subtype \Number$, and $\Number \lesssim \Any$.
+
 To avoid ambiguous table types we use the following rule to check
@@
 \]
+
+\emph{/* Can we use this well formed rule to allow $k$ to be $t$? */}
 
@@
 $t_{1} \subtype t_{2}$ or not.
+
+\emph{/* Where should we handle the following case? */}
+\begin{align*}
+\{ & ``info":\Number, \\
+   & ``next":\mu x.\{``info":\Number, ``next":x \;\cup\; \Nil\} \cup \Nil \} \subtype \\
+& \mu x.\{``info":\Number, ``next":x \;\cup\; \Nil\}
+\end{align*}
+
+\emph{/* Should we also have the following subtyping rules? */}
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-AMBERL}\\
+\dfrac{\senv \vdash [x \mapsto \mu x.t_{1}]t_{1} \subtype t_{2}}
+      {\senv \vdash \mu x.t_{1} \subtype t_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-AMBERR}\\
+\dfrac{\senv \vdash t_{1} \subtype [x \mapsto \mu x.t_{2}]t_{2}}
+      {\senv \vdash t_{1} \subtype \mu x.t_{2}}
+\end{array}
+\end{array}
+\]
+
+\emph{/* Or having the unfolding typing rule is enough? */}
 

### 2014-10-15 doc/thesis/system.tex
Fixing strings in the generated code

@@
 & | \; \pi_{i} & \textit{projection types}\\
-l ::= & \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} & \\
-b ::= & \Boolean \; | \; \Integer \; | \; \Number \; | \; \String & \\
-v ::= & k:t \; | \; \Const \; k:t & \textit{field types}\\ 
-k ::= & l \; | \; b \; | \; \Any & \textit{key types}\\
+l ::= & \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} & \\
+b ::= & \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String & \\
+v ::= & \;\; k:t \; | \; \Const \; k:t & \textit{field types}\\ 
+k ::= & \;\; l \; | \; b \; | \; \Any & \textit{key types}\\
 \multicolumn{3}{c}{}\\
@@
 & | \; t \times s & \textit{tuple types}\\
-p ::= & \Void \; | \; s & \textit{parameters list type}\\
-r ::= & \Void \; | \; s \; | \; s \sqcup r & \textit{return types}\\
+p ::= & \;\; \Void \; | \; s & \textit{parameters list type}\\
+r ::= & \;\; \Void \; | \; s \; | \; s \sqcup r & \textit{return types}\\
 \end{array}
@@
 
+In this section we will use a reduced core of Typed Lua to present
+its typing rules in an easier way.
+It is a reduced core of the Lua language with explicit type annotations,
+without syntactic sugar, with only if and while control flow statements,
+and making variable scope explicit.
+
 \begin{figure}[!ht]
@@
 s_{1} \; ; \; s_{2} \; | \;
-\vec{l} = \vec{e}  \; | \;
+\vec{l} = el  \; | \;
 \mathbf{while} \; e \; \mathbf{do} \; s \; | \;
 \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
-& | \; \mathbf{local} \; \vec{n{:}t} = \vec{e} \; \mathbf{in} \; s \; | \;
+& | \; \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s \; | \;
 \mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
-\mathbf{return} \; \vec{e} \; | \;
+\mathbf{return} \; el \; | \;
 a\\
@@
 \mathbf{true} \; | \;
-{\it number} \; | \;
+{\it int} \; | \;
+{\it float} \; | \;
 {\it string} \; | \;
@@
 f \; | \;
-\{ \; \vec{c} \; \} \\
+\{ \; cl \; \} \\
 & | \; e_{1} + e_{2} \; | \;
@@
 l \; | \;
-(e) \; | \;
-{<}t{>} \; e\\
-l ::= & \; n \; | \;
+{<}t{>} \; n\\
+l ::= & \;\; n \; | \;
 e_{1}[e_{2}]\\
-a ::= & \; e_{1}(\vec{e_{2}}) \; | \;
+el ::= & \;\; \mathbf{nothing} \; | \;
+\vec{e} \; | \;
+me \; | \;
+\vec{e}, me\\
+cl ::= & \;\; \mathbf{nothing} \; | \;
+\vec{c} \; | \;
+\vec{e} \; | \;
+me \; | \;
+\vec{c}, \vec{e} \; | \;
+\vec{c}, me \; | \;
+\vec{c}, \vec{e}, me \; | \;
+\vec{e}, me\\
+me ::= & \;\; a \; | \;
+{...}\\
+a ::= & \;\; e_{1}(\vec{e_{2}}) \; | \;
 e_{1}{:}n(\vec{e_{2}})\\
-f ::= & \; \mathbf{fun} \; (){:}r \; s \; | \;
-\mathbf{fun} \; ({...}{:}t){:}r \; s \; | \;
-\mathbf{fun} \; (\vec{n{:}t}){:}r \; s \; | \;
-\mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; s\\
-c ::= & \; [e_{1}] = e_{2}\\
-n ::= & \; {\it name}\\
+f ::= & \;\; \mathbf{fun} \; (){:}r \; s \; ; \; \mathbf{return} \; el \; | \;
+\mathbf{fun} \; ({...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
+& | \; \mathbf{fun} \; (\vec{n{:}t}){:}r \; s \; ; \; \mathbf{return} \; el \; | \;
+\mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
+c ::= & \;\; [e_{1}] = e_{2}\\
+n ::= & \;\; {\it name}\\
 \end{array}
@@
 \end{figure}
+
+Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
+It splits the syntactic categories as follows:
+$s$ are statements, $e$ are expressions, $l$ are left-hand values,
+$el$ are expression lists, $c$ are constructor fields, $cl$ are lists of
+constructor fields, $me$ are multiple expressions, $a$ are function
+applications, $f$ are function declarations, $n$ are variable names,
+$t$ are first-level types, and $r$ are second-level types.
+The notation $\vec{n{:}t}$ denotes the non-empty list
+$n_{1}{:}t_{1}, ..., n_{n}{:}t_{n}$.
+
+We will present the typing rules as a deduction system for the typing
+relations $\env_{1} \vdash s : \env_{2}$ and $\env_{1} \vdash e : t, \env_{2}$. 
+We will use the first one for typing statements, and the sencond one
+for typing expressions. 
+The first relation means that given the type environment $\env_{1}$,
+typing the statement $s$ produces the new type environment $\env_{2}$. 
+The sencond relation means that given the type environment $\env_{1}$,
+the expression $e$ has type $t$, and produces the new type environment $\env_{2}$.
+We need two typing relations because only expressions have types,
+though both expressions and statements can modify the environment.
+The environment maps variables to types, and we will use $\env_{1}[n \mapsto t]$
+for meaning that we extended the environment $\env_{1}$ with the
+variable $n$ that maps to type $t$, and this mapping does not affect
+the other mappings that extisted before.

### 2014-10-21 doc/thesis/system.tex
Improving the text about subtyping rules

@@
 and optional parameters.
-We use typing rules to explain how we implemented the features of Typed Lua.
+We will use typing rules to explain how we implemented the features of Typed Lua.
 
@@
 We will focus the discussion on the definition of subtyping because,
-as we mentioned in Chapter \ref{chap:review}, we achieve the definition of
-consistent-subtyping combining consistency and subtyping.
+as we mentioned in Chapter \ref{chap:review}, we can combine the
+consistency and subtyping relations to achieve consistent-subtyping.
 The differences between subtyping and consistent-subtyping are the way
@@
 & | \; p \rightarrow r & \textit{function types}\\
-& | \; \{v_{i}, ..., v_{n}\}_{closed|open|unique} & \textit{table types}\\
+& | \; \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{closed|open|unique} & \textit{table types}\\
 & | \; x & \textit{type variables}\\
@@
 b ::= & \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String & \\
-v ::= & \;\; k:t \; | \; \Const \; k:t & \textit{field types}\\ 
 k ::= & \;\; l \; | \; b \; | \; \Any & \textit{key types}\\
+v ::= & \;\; t \; | \; \Const \; t & \textit{field types}\\ 
 \multicolumn{3}{c}{}\\
@@
 
-Literal types represent the type of literal constants.
+Literal types represent the type of literal values.
 They can be the boolean values $\False$ and $\True$,
 any integer value, any floating point value, and any string value.
-We shall see that literal types are important to type tables as records.
+We shall see that literal types are important to represet table types
+as records.
 
@@
 respectively;
-while the base types $\Integer$ and $\Number$ represent the values
+the base types $\Integer$ and $\Number$ represent the values
 that during run-time Lua tags as \texttt{number}.
 We introduced two different base types to represent Lua numbers
-because Lua introduced integers recently.
+because Lua introduced integers in version 5.3.
 
-The subtyping rules for literal types include a reflexive rule
+The subtyping rules for literal types include a reflexive rule,
 and the rules for defining that literal types are subtypes of
 their respective base types;
-the subtyping rules for base types include a reflexive rule
+the subtyping rules for base types include a reflexive rule,
 and another rule for defining that $\Integer$ is subtype of
@@
 
-Since subtyping is reflexive and transitive,
-we could have omitted the reflexive rules \textsc{S-LITERAL} and
-\textsc{S-BASE}, as well as the rule \textsc{S-INT2}.
-More precisely, we could have used subsumption rules that define
-the reflexiveness and transitiveness of subtyping.
-In the case of \textsc{S-INT2}, we would not need this rule because
-we could use the transitive subsumption rule to derive that
+We could have omitted the rules \textsc{S-LITERAL}, \textsc{S-BASE},
+and \textsc{S-INT2}, because the first two are reflexive and we
+could achieve the third one trhough transitivity.
+More precisely, we could have defined subsumption rules that
+specify the reflexive and trasitive behavior of subtyping,
+and use these rules instead of the ones we defined.
+For instance, we would not need \textsc{S-INT2} because we could
+use the transisitive subsumption rule to derive that
 \[
@@
 
-The type $\Value$ is the top type, so any first-level type is a
+The type $\Value$ is the top type, so any first-level type is
 subtype of $\Value$.
 We shall see that this type along with variadic types help to
-preserve the semantics of Lua to drop extra values on multiple
-assignments, and function calls.
+drop extra values on multiple assignments, and function calls,
+thus preserving the semantics of Lua in these cases.
 Following we show its subtyping rule:
@@
 The type $\Any$ is neither the bottom nor the top type, but a
-separated typed that is a subtype only of itself:
+separated typed that is subtype only of itself:
 \[
@@
 with statically typed code.
-Consistent-subtyping cannot be transitive because of these rules,
-as transitivity combined with them would make all programs compile
-without type errors.
+Because of these two rules consistent-subtyping cannot be transitive.
 
@@
 For instance, the union type $\Boolean \;\cup\; \Any$ results the
-type $\Any$, because $\Boolean$ is consistent-subtype of $\Any$.
-The union type $\Number \;\cup\; \Nil \;\cup\; 1$ results the union
-type $\Number \;\cup\; \Nil$ because $1 \subtype \Number$.
+type $\Any$, because $\Boolean \lesssim \Any$;
+and the union type $\Number \;\cup\; \Nil \;\cup\; 1$ results the union
+type $\Number \;\cup\; \Nil$ because $1 \lesssim \Number$.
 
-\emph{/* Explain how Typed Lua refine union types */}
+\emph{/* Do we need to show how Typed Lua simplifies union types? */}
 
@@
 second-level types, as they appear on both sides of function types.
-Basically, second-level types are tuples of first-level types
+Second-level types are tuples of first-level types
 that can optionally end with a variadic type.
@@
 and $s_{1}$ is subtype of $s_{2}$.
-As we mentioned, a tuple can optionally end with a variadic
-type, and the second and third rules for tuple types handle
-this case.
-The second rule shows that a variadic type is subtype of
-a tuple whenever the type of the value that it generates is subtype
-of all the elements of the tuple.
-The third rule shows that a tuple is subtype of a variadic
-type whenever all the elements of the tuple are subtype of the
-type of the value that it generates.
+A tuple type can optionally end with a variadic type, and
+the rules \textsc{S-TUPLE2} and \textsc{S-TUPLE3} handle this case.
+The rule \textsc{S-TUPLE2} shows that a variadic type is subtype of
+a tuple type whenever the type of the value that it generates is
+subtype of all the elements of the tuple.
+The rule \textsc{S-TUPLE3} shows that a tuple type is subtype of a
+variadic type whenever all the elements of the tuple are subtype of
+the type of the value that it generates.
 We shall see that these subtyping relations along with the
@@
 The type of the parameters list can be $\Void$ or a tuple type,
-while the return type can be $\Void$, a tuple type, or a union of tuples.
+while the return type can be $\Void$, a tuple type, or a union of
+tuple types.
 The type $\Void$ means an empty tuple.
@@
 Our type system includes table types to represent Lua tables.
-The syntactical form of table types is $\{ v_{i}, ..., v_{n} \}_{tag}$.
+The syntactical form of table types is $\{ k_{i}{:}v_{i}, ..., k_{n}{:}v_{n} \}_{tag}$.
 When $n = 0$, then $i = 0$ and it represents the type of an empty table.
-When $n > 0$, then $i$ varies from 1 to $n$, and each $v_{i}$
-represents the type of a table field.
-Table fields represent maps from keys to values, and we restricted
-the type of keys to literal types, base types, and the dynamic type;
-but we can use any first-level type to represent the type of values.
-We made this restriction to the type of keys because the statistics
+When $n > 0$, then $i$ varies from 1 to $n$, each $k_{i}$ represents
+the type of a table key, and each $v_{i}$ represents the type of the
+value that the table key $k_{i}$ maps to.
+We restricted the type of keys to literal types, base types, and the dynamic type,
+but we can use any first-level type to represent the type of values they map to.
+We made this restriction to the type of the keys because the statistics
 that we discussed in Section \ref{sec:statistics} showed that most
@@
 
-The following subtyping rules show that \emph{closed} table types
-have width subtyping, and depth subtyping on $\Const$ fields:
+The following subtyping rule shows that \emph{closed} table types
+have width subtyping, and depth subtyping on $\Const$ fields;
+it resembles the subtyping rule for record types.
 
@@
 \begin{array}{c}
-\begin{array}{c}
 \mylabel{S-TABLE1}\\
 \dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
-       \senv \vdash v_{j} \subtype v_{i}}
-      {\senv \vdash \{v_{1}, ..., v_{m}\}_{closed} \subtype \{v_{1}, ..., v_{n}\}_{closed}}
+       \senv \vdash k_{j} \subtype k_{i}' \;\;\;
+       \senv \vdash k_{i}' \subtype k_{j} \;\;\;
+       \senv \vdash v_{j} \subtype_{c} v_{i}'}
+      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{closed} \subtype \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
 \end{array}
-\\ \\
+\]
+
+The rule \textsc{S-TABLE1} introduced the relation $\subtype_{c}$.
+It is just an auxiliary relation that defines the subtyping of the
+type of the values stored in table fields.
+We needed an auxiliary relation because the subtyping of the type
+of the values stored in the table fields change according to the
+tag of the table type.
+For closed tables, subtyping handles mutable fields invariantly,
+and imutable fields covariantly.
+The following rules define this behavior.
+\[
+\begin{array}{c}
 \begin{array}{c}
 \mylabel{S-FIELD1}\\
-\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
-       \senv \vdash t_{1} \subtype t_{2} \;\;\;
-       \senv \vdash t_{2} \subtype t_{1}}
-      {\senv \vdash k_{1}:t_{1} \subtype k_{2}:t_{2}}
+\dfrac{\senv \vdash v_{1} \subtype v_{2} \;\;\;
+       \senv \vdash v_{2} \subtype v_{1}}
+      {\senv \vdash v_{1} \subtype_{c} v_{2}}
 \end{array}
@@
 \mylabel{S-FIELD2}\\
-\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
-       \senv \vdash t_{1} \subtype t_{2}}
-      {\senv \vdash \Const \; k_{1}:t_{1} \subtype \Const \; k_{2}:t_{2}}
+\dfrac{\senv \vdash v_{1} \subtype v_{2}}
+      {\senv \vdash \Const \; v_{1} \subtype_{c} \Const \; v_{2}}
 \end{array}
@@
 \mylabel{S-FIELD3}\\
-\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
-       \senv \vdash t_{1} \subtype t_{2}}
-      {\senv \vdash k_{1}:t_{1} \subtype \Const \; k_{2}:t_{2}}
+\dfrac{\senv \vdash v_{1} \subtype v_{2}}
+      {\senv \vdash v_{1} \subtype_{c} \Const \; v_{2}}
 \end{array}
@@
 
-First, we had only \emph{closed} table types.
-However, programmers often define tables that include optional fields,
-and the subtyping rules we introduced do not let programmers to use
-the table constructor to initialize table types that include optional
-fields.
-For instance, consider the table type we introduced in
-Chapter 3:
+There are two limitations on \emph{closed} table types that led us to
+introduce \emph{open} and \emph{unique} table types.
+First, they do not allow programmers to use the table constructor to
+initialize table types that include optional fields.
+For instance,
 \begin{align*}
-\{ & ``firstname":\String, \\
-   & ``middlename":\String \cup \Nil,\\
-   & ``lastname":\String \}_{closed}
+\{ & ``firstname":\String, ``lastname":\String \}_{closed} \not \subtype\\
+\{ & ``firstname":\String,\\
+  &  ``middlename":\String \cup \Nil, ``lastname":\String \}_{closed}
 \end{align*}
-Following we show a table type that Typed Lua could assign to a
-table constructor that does not include the field $``middlename"$,
-but such table type is not subtype of the above table type:
+Second, they do not allow programmers to use the table construtor to
+initialize table types that represent hashes.
+For instance,
 \begin{align*}
-\{ ``firstname":\String, ``lastname":\String \}_{closed}
+\{ & ``x":\Number, ``y":\Number, ``z":\Number \}_{closed} \not \subtype\\
+\{ & \String:\Number \cup \Nil\}_{closed}
 \end{align*}
@@
 fields in the initialization of table types.
-To do that, we defined new subtyping rules that allow \emph{open}
-table types to omit fields when $\Nil$ is subtype of the of type
-of the value of the omitted key.
-The following subtyping rules show this behavior:
+To do that, we defined a subtyping rule that allow \emph{open}
+table types to omit fields when $\Nil$ is subtype of the type of
+the value of the ommited key.
+The following subtyping rule show this behavior:
+\[
+\begin{array}{c}
+\mylabel{S-TABLE2}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
+       \senv \vdash v_{i} \subtype_{c} v_{i}' \;\;\;
+       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
+      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
+                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open}}
+\end{array}
+\]
 
+The rule \textsc{S-TABLE2} introduced the relation $\subtype_{o}$.
+It is an auxiliary relation that defines the subtyping relation
+on optional fields, and we defined it as follows:
 \[
@@
 \begin{array}{c}
-\mylabel{S-TABLE2}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash v_{i}' \subtype v_{i}}
-      {\senv \vdash \{v_{1}, ..., v_{m}\}_{open} \subtype \{v_{1}', ..., v_{m}', ..., v_{n}'\}_{open|closed}}
+\mylabel{S-FIELD4}\\
+\dfrac{\senv \vdash \Nil \subtype v}
+      {\senv \vdash \Nil \subtype_{o} v}
 \end{array}
-\\ \\
+\;
 \begin{array}{c}
-\mylabel{S-TABLE3}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash v \subtype v_{i}}
-      {\senv \vdash \{v_{1}, ..., v_{m}\}_{open} \subtype \{v\}_{open|closed}} 
+\mylabel{S-FIELD5}\\
+\dfrac{\senv \vdash \Nil \subtype v}
+      {\senv \vdash \Nil \subtype_{o} \Const \; v}
 \end{array}
@@
 
-However, we still had the that
+Although \emph{open} table types let programmers omit optional fields,
+they do not let programmers initialize optional fields that are
+mutable, unless they use the same type of the field in the initialization.
+For instance,
 \begin{align*}
+\{ & ``firstname":\String,\\
+   & ``middlename":\String, ``lastname":\String \}_{open} \not \subtype \\
 \{ & ``firstname":\String, \\
-   & ``middlename":\String,\\
-   & ``lastname":\String \}_{open} \not \subtype \\
-\{ & ``firstname":\String, \\
-   & ``middlename":\String \cup \Nil,\\
-   & ``lastname":\String \}_{closed}
+   & ``middlename":\String \cup \Nil, ``lastname":\String \}_{closed}
 \end{align*}
@@
 We introduced \emph{unique} table types to solve this problem.
-We use \emph{unique} table types only to denote the type of the
+We use \emph{unique} table types to denote the type of the
 table constructor.
-The following subtyping rules allow us to use table constructors
-that initialize optional fields:
+The following subtyping rule allow programmers to use table constructors
+to initialize and omit optional fields:
+\[
+\begin{array}{c}
+\mylabel{S-TABLE3}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
+       \senv \vdash v_{i} \subtype_{u} v_{i}' \;\;\;
+       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
+      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
+                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open|unique}}
+\end{array}
+\]
 
+The rule \textsc{S-TABLE3} introduced the relation $\subtype_{u}$.
+It is an auxiliary relation that makes \emph{unique} table types
+have width and depth subtyping on all fields, and we defined it
+as follows:
 \[
@@
 \begin{array}{c}
-\mylabel{S-TABLE4}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash v_{i}' \subtype_{n} v_{i}}
-      {\senv \vdash \{v_{1}, ..., v_{m}\}_{unique} \subtype \{v_{1}', ..., v_{m}', ..., v_{n}'\}_{unique|open|closed}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-TABLE5}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash v \subtype_{n} v_{i}}
-      {\senv \vdash \{v_{1}, ..., v_{m}\}_{unique} \subtype \{v\}_{unique|open|closed}} 
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-FIELD4}\\
-\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
-       \senv \vdash t_{1} \subtype t_{2}}
-      {\senv \vdash k_{1}:t_{1} \subtype_{n} k_{2}:t_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-FIELD5}\\
-\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
-       \senv \vdash t_{1} \subtype t_{2}}
-      {\senv \vdash \Const \; k_{1}:t_{1} \subtype_{n} \Const \; k_{2}:t_{2}}
+\mylabel{S-FIELD6}\\
+\dfrac{\senv \vdash v_{1} \subtype v_{2}}
+      {\senv \vdash v_{1} \subtype_{u} v_{2}}
 \end{array}
@@
 \begin{array}{c}
-\mylabel{S-FIELD6}\\
-\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
-       \senv \vdash t_{1} \subtype t_{2}}
-      {\senv \vdash k_{1}:t_{1} \subtype_{n} \Const \; k_{2}:t_{2}}
+\mylabel{S-FIELD7}\\
+\dfrac{\senv \vdash v_{1} \subtype v_{2}}
+      {\senv \vdash \Const \; v_{1} \subtype_{u} \Const \; v_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FIELD8}\\
+\dfrac{\senv \vdash v_{1} \subtype v_{2}}
+      {\senv \vdash v_{1} \subtype_{u} \Const \; v_{2}}
 \end{array}
@@
 
-Notice that we introduced a special subtyping rule to check the
-subtyping of \emph{unique} table fields.
-The difference is that \emph{unique} table types have width and
-depth subtyping on all fields.
-
-Even though \emph{unique} table types solve the problem of using
-table constructors to initialize tables that contain optional fields,
-we still have the problem when we try to use a local variable that
-holds an \emph{open} table to initialize another table that contains
-optional fields.
-
-\emph{/* Do we really need closed, open, and unique? */}
-
-\emph{/* Also, can we make these subtyping rules for tables more clear? */}
-
 Table types can be ambiguous.
-For instance, $\{1:\Number, \Number:\String\}$ is an ambiguous
+For instance, $\{1:\Number, \Integer:\String\}$ is an ambiguous
 table type because the type of the value stored by key $1$ can be
 $\Number$ or $\String$, as $1 \subtype 1$ and
-$1 \subtype \Number$.
+$1 \subtype \Integer$.
 Table types can get even more ambiguous when we use the dynamic
 type in the type of table keys.
-For instance, the table type $\{1:\Number, \Number:\String, \Any:\Boolean\}$
+For instance, the table type $\{1:\Number, \Integer:\String, \Any:\Boolean\}$
 is ambiguous because the type of the value stored by key $1$ can be
 $\Number$, or $\String$, or $\Boolean$, as $1 \subtype 1$,
-$1 \subtype \Number$, and $1 \lesssim \Any$.
-The type of the value stored by a key of type $\Number$,
+$1 \subtype \Integer$, and $1 \lesssim \Any$.
+Moreover, the type of the value stored by a key of type $\Integer$,
 which is not the literal type $1$, can be $\Number$ or $\Boolean$,
-as $\Number \subtype \Number$, and $\Number \lesssim \Any$.
+as $\Integer \subtype \Integer$, and $\Integer \lesssim \Any$.
 
@@
 a function or table type could appear.
-For instance, $\mu x.\{``info":\Number, ``next":x \;\cup\; \Nil\}$
-represents the type of single linked lists of numbers.
+For instance, $\mu x.\{``info":\Number, ``next":x \;\cup\; \Nil\}_{closed}$
+represents the type of single linked list of numbers.
 
@@
 
-\emph{/* Where should we handle the following case? */}
-\begin{align*}
-\{ & ``info":\Number, \\
-   & ``next":\mu x.\{``info":\Number, ``next":x \;\cup\; \Nil\} \cup \Nil \} \subtype \\
-& \mu x.\{``info":\Number, ``next":x \;\cup\; \Nil\}
-\end{align*}
-
-\emph{/* Should we also have the following subtyping rules? */}
+Our type system does not infer recursive types, so we defined two
+extra subtyping rules that handle the cases where they may
+appear inside a table or function type:
 \[
@@
 
-\emph{/* Or having the unfolding typing rule is enough? */}
+For instance, the rule \textsc{S-AMBERR} allows our type system
+check whether the type of a table constructor is subtype of
+a recursive type or not, as we show in the following example:
+\begin{align*}
+\{ & ``info":\Number, \\
+   & ``next":\mu x.\{``info":\Number, ``next":x \;\cup\; \Nil\}_{closed} \cup \Nil \}_{unique} \subtype \\
+& \mu x.\{``info":\Number, ``next":x \;\cup\; \Nil\}_{closed}
+\end{align*}
 

### 2014-10-22 doc/thesis/evaluation.tex
Typing the core of luasocket
First version of the luasocket description files
Fixing evaluation to include methods from userdata
Improving the text about subtyping rules
Reorganizing the chapter about evaluation and writing the section about the MD5 case study

@@
 Typed Lua's description files.
-In this chapter we present our evaluation results and discuss
-some interesting cases.
-
-Tables \ref{tab:evalbylib} and \ref{tab:evalbymod} sumarizes our
-evaluation results:
-the first presents the results agruped by library while
-the second presents the results agruped by module.
-We split the members that we typed into three categories:
-\emph{easy}, \emph{poly}, and \emph{hard}.
-The \emph{easy} category represents the members that we could give
-a precise static type for them.
-The \emph{poly} category represents the members that we had to
-rely on the dynamic type, and we believe we could give a precise
-static type for them if our type system had parametric polymorphism.
-The \emph{hard} category represents the members that we could not
-give a precise static type for them, relying on the dynamic type
-in most of the cases, but with no hopes that parametric polymorphism
-would help us to give a precise static type for them.
-The last column of each table contains the actual evaluation result
-of each case study, that is basically the percentage of members
-that are under the \emph{easy} category.
+In this chapter we present our evaluation results, discuss
+some interesting cases, and compare our type system to
+related work.
 
@@
 \hline
-Lua Standard Libraries & 87 & 11 & 25 & 123 & 71\% \\
+Lua Standard Libraries & 92 & 11 & 27 & 130 & 71\% \\
 \hline
-MD5 & 7 & 0 & 0 & 7 & 100\% \\
+MD5 & 13 & 0 & 0 & 13 & 100\% \\
+\hline
+Lua Socket & 85 & 22 & 19 & 126 & 67\% \\
 \hline
@@
 \end{center}
-\caption{Evaluation of Typed Lua by library}
-\label{tab:evalbylib}
+\caption{Evaluation results for each case study}
+\label{tab:evalbycase}
 \end{table}
+
+Table \ref{tab:evalbycase} sumarizes our evaluation results for each
+case study that we used Typed Lua for typing their members.
+We split the members that we typed into three categories:
+\emph{easy}, \emph{poly}, and \emph{hard}.
+The \emph{easy} category represents the members that we could give
+a precise static type for them.
+The \emph{poly} category represents the members that we believe that
+parametric polymorphism would help give a precise static type for them,
+and we had to rely on the dynamic type to type them because our
+type system does not support parametric polymorphism.
+The \emph{hard} category represents the members that even parametric
+polymorphism would not help give a precise static type for them,
+and we also had to rely on the dynamic type to type them;
+for instance, this category includes functions that require
+intersection types to describe their precise static type.
+The last column of the table shows the percentage of members that
+are under the \emph{easy} category for each case study.
+This percentage is our evaluation of Typed Lua, as it
+represents how much static typing we could introduce to each one of
+our case studies.
+
+Before comparing our type system to related work, we will discuss
+each case study in more detail.
+For each case study, we will split the evaluation results according
+to the modules that each one of them include.
+This shall allow us to better discuss the contributions and limitations
+of our type system.
+
+\section{Lua Standard Libraries}
 
@@
 \cline{2-7}
-& io & 9 & 0 & 2 & 11 & 82\% \\
+& io & 14 & 0 & 4 & 18 & 78\% \\
 \cline{2-7}
@@
 \hline
+\end{tabular}
+\end{center}
+\caption{Evaluation results for Lua Standard Libraries}
+\label{tab:evallsl}
+\end{table}
+
+\section{MD5}
+
+The MD5 library \citep{lmd5} is the case study that we used to
+introduce Typed Lua's description files.
+We chose this case study because it is simple and contains just
+one module, being a good example for introducing how we can use
+Typed Lua's description files for giving a statically typed
+interface to a Lua library that is written in C, and to 
+introduce how we can type an \emph{userdata}.
+Table \ref{tab:evalmd5} sumarizes the evaluation results for MD5.
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|c|}
+\hline
+\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\hline
 \multirow{1}{*}{MD5}
-& md5 & 7 & 0 & 0 & 7 & 100\% \\
+& md5 & 13 & 0 & 0 & 13 & 100\% \\
 \hline
@@
 \end{center}
-\caption{Evaluation of Typed Lua by module}
-\label{tab:evalbymod}
+\caption{Evaluation results for MD5}
+\label{tab:evalmd5}
 \end{table}
+
+Even tough the MD5 library is quite simple and it was quite easy to type,
+we found a little discrepancy between its documentation and its static
+typing.
+The documentation suggested that the typing of the function \texttt{update}
+should be \texttt{(md5\string_context, string) -> (md5\string_context)},
+and a comment in the source code suggested that it should be
+should be \texttt{(md5\string_context, string, string*) -> (md5\string_context)}.
+However, while testing it and reading its source code, we noticed that
+its actual type is \texttt{(md5\string_context, string*) -> (md5\string_context)},
+that is, we can pass zero or more strings to \texttt{update}.
+
+This case study shows a connection between documentation and software testing.
+Even though type annotations are a good source of documentation, this case
+study shows that identifying what functions discard extra parameters can be
+not so obvious for someone that is typing an external library.
+Reading the source code and the library's test script was essential to confirm
+the typing of \texttt{update}.
+
+\section{Lua Socket}
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|c|}
+\hline
+\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\hline
+\multirow{7}{*}{Lua Socket}
+& socket & 50 & 3 & 7 & 60 & 83\% \\
+\cline{2-7}
+& ftp & 6 & 1 & 1 & 8 & 75\% \\
+\cline{2-7}
+& http & 3 & 0 & 2 & 5 & 60\% \\
+\cline{2-7}
+& ltn12 & 10 & 10 & 0 & 20 & 50\% \\
+\cline{2-7}
+& mime & 2 & 8 & 7 & 17 & 12\% \\
+\cline{2-7}
+& smtp & 6 & 0 & 2 & 8 & 75\% \\
+\cline{2-7}
+& url & 8 & 0 & 0 & 8 & 100\% \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Evaluation results for Lua Socket}
+\label{tab:evalsocket}
+\end{table}
+
+\section{Related Work}

### 2014-10-23 doc/thesis/evaluation.tex
Adding text about typing the Lua Standard Libraries

@@
 \hline
-Lua Standard Libraries & 92 & 11 & 27 & 130 & 71\% \\
+Lua Standard Libraries & 92 & 4 & 34 & 130 & 71\% \\
 \hline
@@
 
+The Lua Standard Libraries \citep{luamanual} were our first case study.
+We started to think about how we would type them at the same time that
+we started to design our type system, as they could give us some hints
+on our type system.
+And they did: optional parameters and overloading on the return type
+are two Lua features that our type system should handle to allow us
+typing some of the functions that the standard libraries provide.
+
+All libraries are separated C modules, and we used Typed Lua's description
+files to give a statically typed interface to each module.
+The \texttt{debug} module is the only one that we did not include in our
+evaluation results, because it provides several functions that violate
+basic assumptions about Lua code \citep{luamanual}.
+For instance, we can use its function \texttt{setlocal} to change the value
+of a local variable that is outside of the scope.
+Table \ref{tab:evallsl} sumarizes the evaluation results for the Lua Standard Libraries.
+
 \begin{table}[!ht]
@@
 \multirow{9}{*}{Lua Standard Libraries}
-& base & 9 & 5 & 12 & 26 & 35\% \\
+& base & 8 & 0 & 18 & 26 & 31\% \\
 \cline{2-7}
@@
 \cline{2-7}
-& package & 5 & 2 & 1 & 8 & 62\% \\
+& package & 5 & 0 & 3 & 8 & 62\% \\
 \cline{2-7}
-& string & 13 & 0 & 1 & 14 & 93\% \\
+& string & 14 & 0 & 0 & 14 & 100\% \\
 \cline{2-7}
@@
 \end{table}
+
+We could give precise static types to most of the modules,
+but some of them are still hard to type with the current type
+system.
+The only modules that we could type all their members are
+\texttt{string} and \texttt{bit32}.
+Now we will discuss in more detail the particularities that
+we found on each module, and the limitations that we found
+on our type system.
+
+The \texttt{base} module was quite hard to type because it
+includes several functions that have dynamic behavior.
+For instance, the function \texttt{next} traverses all
+fields of a table, but the order that the indices are
+enumerated is not specified.
+The function \texttt{pairs} is also difficult to type, as
+our type system cannot garantee that it will no try to
+access a field that does not exist.
+The functions \texttt{rawget} and \texttt{rawset} are also
+examples of functions that their behavior depend on the
+C implementation.
+
+There are some functions that have a dynamic behavior and
+the return type depend on the input type.
+For instance, \texttt{assert} and \texttt{select}.
+
+Moreover, the \texttt{base} module also include some functions
+that the type of one input parameter depends on the type of
+another.
+For instance, there are two static types that we can
+assign to the function \texttt{tonumber}:
+\texttt{(value) -> (number?)} and
+\texttt{(string, number) -> (number?)}.
+More precisely, the first argument of \texttt{tostring} can be
+a value of any type if it is the only argument, but it must
+be a value of type string if there is a second argument,
+which must be a value of type number.
+
+We could not type the \texttt{coroutine} module because our
+type system does not include the type \emph{thread}.
+Lua has one-shot delimited continuations \citep{james2011yield}
+in the form of \emph{coroutines} \citep{moura2009rc}, and
+effect systems \citep{nielson1999type} are an approach that we
+could use to describe control transfers with continuations.
+However, for now coroutines are out of the scope of our type
+system, and we use an empty \emph{userdata} declaration
+to represent the type \emph{thread}.
+
+We could type most the members of the \texttt{package} module,
+but we could not type any member that is a table.
+More precisely, we could not type the table \texttt{loaded}
+that stores loaded modules, and the tables \texttt{preload}
+and \texttt{searchers} that store module \emph{loaders}.
+They are difficult to type because their type depend on the
+modules a program loads.
+
+The \texttt{table} module is specially difficult to type because
+all of its members require parametric polymorphism.
+All the functions of this module either receive or return a list
+of elements, and parametric polymorphism would help us to describe
+them with a generic type.
+
+However, the lack of parametric polymorphism did not prevent us from
+giving a precise type to \texttt{concat}, as it operates over lists
+where all elements are strings or numbers.
+
+Even if our type system had parametric polymorphism, it would
+be difficult to type \texttt{insert}, as its type depends on
+the calling arity.
+We can call \texttt{insert} passing two or three parameters.
+The first parameter is always a list.
+If we call it with two arguments, then the second parameter
+is the value to be inserted in the end of the list.
+If we call it with three arguments, then the third parameter
+is the value to be inserted in the list, and the second
+parameter is the position where it should be inserted.
+This function also does not follow the semantics of Lua on
+discarding extra arguments.
+Lua generates a run-time error whenever we pass more than three
+arguments to \texttt{insert}, even if the first three parameters
+match its definition.
+
+Even though the \texttt{math} module is straightforward to type,
+it includes a special case that is hard to type: the function \texttt{random}.
+This function is difficult to type because its type also depends
+on the calling arity.
+We can call \texttt{random} passing zero, one, or three parameters.
+If we pass no parameter, then it returns a float number.
+If we pass one or two parameters, then all the parameters should be
+integers and it returns an integer number.
+Another problem with this function is that it also does not follow
+the semantics of Lua on discarding extra arguments, and generates
+a run-time error whenever we pass more than three arguments.
+There is also a problem on its documentation, as it suggests that
+the two integers are optional parameters, but, for \texttt{random},
+optional parameters behave in a different way.
+Usually, Lua functions replace optional parameters with a default value
+when they are \texttt{nil}, but \texttt{random} generates a run-time
+error instead.
+
+We could type most members from \texttt{io} module, but we could not
+precisely type \texttt{close} and \texttt{lines}.
+The function and the method \texttt{close} are difficult to type
+because their return type depend whether the file handle that is
+being closed was created with \texttt{popen} or not.
+If it was created with \texttt{popen}, then \texttt{close}
+returns \texttt{(boolean) | (nil, string, number)};
+otwerwise \texttt{close} returns \texttt{(boolean) | (nil, string)}.
+The function and the method \texttt{lines} are difficult to type
+because they number or string depending on the type of the argument.
+
+The function \texttt{execute} is the only member from \texttt{os}
+module that we could not give a precise type.
+It is difficult to type because its return type depend on the
+input type.
+Its type can be either \texttt{() -> (boolean)} or
+\texttt{(string) -> (boolean?, string, number)}.
+
+To increase the static typing of the Lua Standard Libraries
+we need to include effect types, parametric polymorphism,
+and intersection types.
+The first will allow us to type coroutines, the second will
+improve static typing of table types, and the third is
+important to describe functions that return different types
+according to their input types.
 

### 2014-10-29 doc/thesis/evaluation.tex
Improving the text about the typing of Lua Standard Libraries

--- 2014-10-22 doc/thesis/evaluation.tex
@@
 \hline
-Lua Standard Libraries & 92 & 11 & 27 & 130 & 71\% \\
+Lua Standard Libraries & 94 & 5 & 34 & 133 & 71\% \\
 \hline
@@
 \hline
-Lua Socket & 85 & 22 & 19 & 126 & 67\% \\
+LuaSocket & 85 & 22 & 19 & 126 & 67\% \\
 \hline
@@
 \multirow{9}{*}{Lua Standard Libraries}
-& base & 9 & 5 & 12 & 26 & 35\% \\
+& base & 9 & 1 & 16 & 26 & 35\% \\
 \cline{2-7}
@@
 \cline{2-7}
-& io & 14 & 0 & 4 & 18 & 78\% \\
+& io & 15 & 0 & 6 & 21 & 71\% \\
 \cline{2-7}
@@
 \end{table}
+
+We could give precise static types to most of the modules,
+but some of them are still hard to type with the current type
+system.
+Now, we will discuss why some modules are difficult to type,
+and what kind of limitations they represent on our type system.
+
+The \texttt{base} module was quite hard to type because it
+includes several functions that have dynamic behavior.
+For instance, the function \texttt{next} traverses all the
+fields of a table, but the order that the indices are
+enumerated is not specified.
+For this reason, it is also difficult to type the function
+\texttt{pairs}.
+The functions \texttt{getmetatable},  \texttt{rawget}, and
+\texttt{rawset} are also examples of functions that their
+behavior depend on the C implementation.
+
+There are also some functions that have a dynamic behavior and
+their return type depend on their input type.
+For instance, the functions \texttt{assert} and \texttt{select}
+can return all the values that they received as their arguments.
+
+We could give a more precise type to \texttt{ipairs} if our
+type system had parametric polymorphism, as this function
+returns an iterator function that operates over a list of
+elements.
+
+Furthermore, the \texttt{base} module also includes two very
+special cases: the functions \texttt{tonumber} and \texttt{collectgarbage}.
+
+The function \texttt{tonumber} is a special case because the
+type of one input parameter depends on the type of another.
+For instance, there are two static types that we can
+assign to the function \texttt{tonumber}:
+\texttt{(value) -> (number?)} and
+\texttt{(string, number) -> (number?)}.
+More precisely, the first argument of \texttt{tonumber} can be
+a value of any type if it is the only argument, but it must
+be a value of type string if there is a second argument,
+which must be a value of type number.
+
+The function \texttt{collectgarbage} is a special case because
+its return type depends on the value of the first parameter,
+which should be a string.
+This function can return one number, one boolean, or two numbers
+depending on the literal string that was given as its first argument.
+
+We could not type the \texttt{coroutine} module because our
+type system does not include the type \emph{thread}.
+Lua has one-shot delimited continuations \citep{james2011yield}
+in the form of \emph{coroutines} \citep{moura2009rc}, and
+effect systems \citep{nielson1999type} are an approach that we
+could use to describe control transfers with continuations.
+However, for now, coroutines are out of the scope of our type
+system, and we use an empty \emph{userdata} declaration
+to represent the type \emph{thread}.
+
+We could type most of the members of the \texttt{package} module,
+but we could not type the following tables: \texttt{package.loaded},
+\texttt{package.preload}, and \texttt{package.loaders}.
+The first stores loaded modules, while the others store module \emph{loaders}.
+They are difficult to type because their types are dynamic and
+depend on the modules a program loads.
+
+We could type all the members of the \texttt{string} module.
+Even though we can use the \texttt{string} functions in object-oriented style,
+we did not use an \emph{userdata} declaration to handle this feature,
+as it would complicate the subtyping rules because the type \texttt{string}
+would become a table type instead of a base type.
+It was simpler and more starightforward to handle this feature as a
+special case in the implementation.
+For this reason, we did not include the typing of the \texttt{string} methods
+in our evaluation results.
+
+The \texttt{table} module is specially difficult to type because
+all of its members require parametric polymorphism.
+All the functions of this module either receive or return a list
+of elements, and parametric polymorphism would help us to describe
+them with a generic type.
+
+However, the lack of parametric polymorphism did not prevent us from
+giving a precise type to \texttt{table.concat}, as it operates over lists
+where all elements are strings or numbers, and we can use the table type
+\texttt{\{string|number\}} to express this type on our type system.
+
+Even if our type system had parametric polymorphism, it would
+be difficult to type \texttt{table.insert}, as its type depends on
+the calling arity.
+We can call \texttt{table.insert} passing two or three parameters.
+The first parameter is always a list.
+If we call it with two arguments, then the second parameter
+is the value to be inserted at the end of the list.
+If we call it with three arguments, then the third parameter
+is the value to be inserted in the list, and the second
+parameter is the position where it should be inserted.
+This function also does not follow the semantics of Lua on
+discarding extra arguments, and generates a run-time error whenever
+we pass more than three arguments, even if the first three arguments
+match its signature.
+
+Even though the \texttt{math} module is straightforward to type,
+it includes a special case that is hard to type: the function \texttt{math.random}.
+This function is difficult to type because its type also depends
+on the calling arity.
+We can call \texttt{math.random} passing zero, one, or two parameters.
+If we pass no parameter, then it returns a random floating point
+number between \texttt{0} and \texttt{1}.
+If we pass one or two parameters, then all the parameters should be
+integers and it returns an integer number.
+Another problem with this function is that it also does not follow
+the semantics of Lua on discarding extra arguments, and generates
+a run-time error whenever we pass more than three arguments.
+There is also a problem on its documentation, as it suggests that
+the two integers are optional parameters, but, for \texttt{math.random},
+optional parameters behave in a different way.
+Usually, Lua functions replace optional parameters with a default value
+when they are \texttt{nil}, but \texttt{math.random} generates a run-time
+error instead.
+
+The \texttt{bit32} module was quite straightforward to type,
+as its members are only bitwise operations over numbers.
+
+The \texttt{io} module provides operations for manipulating files,
+and these operations can use implicit or explicit file descriptors.
+The implicit operations are functions in the \texttt{io} table,
+while the explicit operations are methods of a file descriptor.
+We used an \emph{userdata} declaration to introduce the type
+\texttt{file} for representing the type of a file descriptor
+and its methods.
+For this reason, we included the typing of the \texttt{file}
+methods in our evaluation results.
+
+We could type most of members of the \texttt{io} module,
+but we could not precisely type the functions \texttt{io.close},
+\texttt{io.read}, and \texttt{io.lines}.
+We also could not type the methods \texttt{file:close},
+\texttt{file:read}, and \texttt{file:lines}.
+
+The function \texttt{io.close} is difficult to type
+because its return type depends on whether the file handle that is
+being closed was created with \texttt{popen} or not.
+If it was created with \texttt{popen}, then \texttt{io.close}
+has type \texttt{(file?) -> (boolean?, string, number)};
+otwerwise \texttt{io.close} has type \texttt{(file?) -> (boolean)?}.
+The method \texttt{file:close} is difficult to type for the same reason.
+
+The functions \texttt{io.read} and \texttt{io.lines} are difficult
+to type because their return types depend on their input types.
+The type of \texttt{io.read} can be either \texttt{() -> (string*)}
+or \texttt{(string|number*) -> (string|number*)}, and
+the type of \texttt{io.lines} can be either \texttt{(string?) -> (string*)}
+or \texttt{(string, string|number*) -> (string|number*)}.
+Returning \texttt{(string|number*)} is also painful to programmers, as
+they need to constrain the value of the return type among three
+different types before cocatenating it, for example.
+The methods \texttt{file:read} and \texttt{file:lines} are difficult
+to type for the same reason.
+
+The function \texttt{os.execute} is the only member of the \texttt{os}
+module that we could not give a precise type.
+This function is difficult to type because its return type depend on
+the input type.
+Its type can be either \texttt{() -> (boolean)} or
+\texttt{(string) -> (boolean?, string, number)}.
+
+The evaluation results show that our type system should include effect
+types, parametric polymorphism, and intersection types, as these
+features would help to increase the static typing of the Lua Standard Libraries.
+Effect types would allow us to type coroutines.
+Parametric polymorphism would allow us to define generic function and table types.
+Intersection types would allow us to define function types that have
+different return types according to their input types.
 
@@
 
-Even tough the MD5 library is quite simple and it was quite easy to type,
-we found a little discrepancy between its documentation and its static
-typing.
+Even tough it was quite straightforward to type the MD5 library,
+we found a little difference between its documentation and its static typing.
 The documentation suggested that the typing of the function \texttt{update}
 should be \texttt{(md5\string_context, string) -> (md5\string_context)},
-and a comment in the source code suggested that it should be
-should be \texttt{(md5\string_context, string, string*) -> (md5\string_context)}.
+while a comment in the source code suggested that it should be
+\texttt{(md5\string_context, string, string*) -> (md5\string_context)}.
 However, while testing it and reading its source code, we noticed that
@@
 
-This case study shows a connection between documentation and software testing.
-Even though type annotations are a good source of documentation, this case
-study shows that identifying what functions discard extra parameters can be
-not so obvious for someone that is typing an external library.
+This case study shows that type annotations are a good source of documentation,
+tough it might be not so obious to identify which functions discard extra
+parameters when we are typing an external library.
 Reading the source code and the library's test script was essential to confirm
@@
 
-\section{Lua Socket}
+\section{LuaSocket}
+
+LuaSocket \citep{luasocket} is a library that adds network support to Lua,
+and it is split into two parts: a core that is written in C and a set of
+Lua modules.
+The C core provides TCP and UDP support, while the Lua modules provide
+support for SMTP, HTTP, and FTP client protocols, MIME encoding,
+URL manipulation, and LTN012 filters \citep{nehab2008ltn012}.
+We chose LuaSocket as a case study because it is the most popular Lua library.
+We used Typed Lua's description files to type both parts.
+This is the only case study that we test description files to statically type
+a Lua module instead of rewriting it.
+Table \ref{tab:evalsocket} sumarizes the evaluation results for LuaSocket.
 
@@
 \hline
-\multirow{7}{*}{Lua Socket}
+\multirow{7}{*}{LuaSocket}
 & socket & 50 & 3 & 7 & 60 & 83\% \\
@@
 \end{center}
-\caption{Evaluation results for Lua Socket}
+\caption{Evaluation results for LuaSocket}
 \label{tab:evalsocket}
@@
 
+We could type most of the members in the \texttt{socket} module,
+which is the C core.
+
 \section{Related Work}

### 2014-10-29 doc/thesis/evaluation.tex
Minor improvements to the text about typing MD5 library

@@
 
-The MD5 library \citep{lmd5} is the case study that we used to
-introduce Typed Lua's description files.
-We chose this case study because it is simple and contains just
-one module, being a good example for introducing how we can use
-Typed Lua's description files for giving a statically typed
-interface to a Lua library that is written in C, and to 
-introduce how we can type an \emph{userdata}.
+We chose the MD5 library \citep{lmd5} as a case study because it is
+simple and contains just one module.
+We needed a simple case study to introduce Typed Lua's description
+files and \texttt{userdata} declarations.
+These Typed Lua's mechanims allow programmers to give statically typed
+interfaces to Lua libraries that are written in C.
 Table \ref{tab:evalmd5} sumarizes the evaluation results for MD5.
@@
 
-This case study shows that type annotations are a good source of documentation,
-tough it might be not so obious to identify which functions discard extra
-parameters when we are typing an external library.
-Reading the source code and the library's test script was essential to confirm
-the typing of \texttt{update}.
+This case study shows that it might be not so obvious to identify which
+functions discard extra arguments when we are typing an external library.
+In this case study, reading its source code and test script was essential
+to confirm the typing of \texttt{update}.
 
@@
 \section{Related Work}
+
+\begin{itemize}
+\item Tidal Lock
+\item Typed Clojure
+\item Gradual Python
+\item Gradualtalk
+\item Typed Racket
+\item Dart
+\item TypeScript
+\end{itemize}

### 2014-10-30 doc/thesis/review.tex
Minor improvements to the section about gradual typing

@@
 variable is syntactic sugar for a variable whose declared type is
-the dynamic type $?$, that is, $\lambda x.e \equiv \lambda x{:}?.e$.
+the dynamic type $?$, that is, $\lambda x.e$ is equivalent to $\lambda x{:}?.e$.
 Under these circumstances, we view gradual typing as a way to add
@@
 $$
-\begin{matrix}
-\Number \sim \Number \mylabel{C-NUM} \;\;\; \String \sim \String \mylabel{C-STR}
+\begin{array}{c}
+\begin{array}{c}
+T \sim T \mylabel{C-REFL}
+\end{array}
+\;
+\begin{array}{c}
+T \sim \;? \mylabel{C-DYNR}
+\end{array}
+\;
+\begin{array}{c}
+? \sim T \mylabel{C-DYNL}
+\end{array}
 \\ \\
-T \sim \;? \mylabel{C-ANY1} \;\;\; ? \sim T \mylabel{C-ANY2}
-\\ \\
+\begin{array}{c}
 \dfrac{T_{3} \sim T_{1} \;\;\; T_{2} \sim T_{4}}
-      {T_{1} \rightarrow T_{2} \sim T_{3} \rightarrow T_{4}} \mylabel{C-FUN}
-\end{matrix}
+      {T_{1} \rightarrow T_{2} \sim T_{3} \rightarrow T_{4}} \mylabel{C-FUNC}
+\end{array}
+\end{array}
 $$
@@
 Figure \ref{fig:consistency} defines the consistency relation.
-Rules \textsc{C-NUM} and \textsc{C-STR} are reflexive. 
-Rules \textsc{C-ANY1} and \textsc{C-ANY2} are the rules that allow
+The rule \textsc{C-REFL} is the reflexive rule.
+Rules \textsc{C-DYNR} and \textsc{C-DYNL} are the rules that allow
 conversions to and from the dynamic type $?$.
-The rule \textsc{C-FUN} resembles the subtyping of functions
+The rule \textsc{C-FUNC} resembles the subtyping of functions
 because it is contravariant on the argument type and covariant on the
@@
 
-Later, \citet{siek2007objects} reported that the consistency relation
-is orthogonal to subtyping, so we can combine the two relations to
-achieve both consistency and subtyping.
-The \emph{consistent-subtyping} relation,
-written $T_{1} \lesssim T_{2}$,
-is the foundation to design object-oriented languages that have a
-gradual type system.
+Later, \citet{siek2007objects} reported that the consistency relation is
+orthogonal to the subtyping relation, so we can combine them to achieve
+the \emph{consistent-subtyping} relation, written $T_{1} \lesssim T_{2}$.
+This relation is essential for designing gradual type systems for
+object-oriented languages.
 Like the consistency relation, and unlike the subtyping relation,
@@
 $\String \not\lesssim \Number$.
+
+Now, we will show how we can combine consistency and subtyping
+to compose a consistent-subtyping relation for the simply typed
+lambda calculus extended with the dynamic type $?$.
+
+\begin{figure}[!ht]
+\dstart
+$$
+\begin{array}{c}
+\begin{array}{c}
+\Number \subtype \Number \mylabel{S-NUM}
+\end{array}
+\;
+\begin{array}{c}
+\String \subtype \String \mylabel{S-STR}
+\end{array}
+\\ \\
+\begin{array}{c}
+? \subtype \;? \mylabel{S-ANY}
+\end{array}
+\;
+\begin{array}{c}
+\dfrac{T_{3} \subtype T_{1} \;\;\; T_{2} \subtype T_{4}}
+      {T_{1} \rightarrow T_{2} \subtype T_{3} \rightarrow T_{4}} \mylabel{S-FUN}
+\end{array}
+\end{array}
+$$
+\dend
+\caption{The subtyping relation}
+\label{fig:subtyping}
+\end{figure}
+
+Figure \ref{fig:subtyping} presents the subtyping relation for the simply
+typed lambda calculus extended with the dynamic type $?$.
+Even though we could have used the reflexive rule $T \subtype T$ to express
+the rules \textsc{S-NUM}, \textsc{S-STR}, and \textsc{S-ANY},
+we did not combine these rules in a single one to make explicit the
+neutrality of the dynamic type $?$ to the subtyping rules.
+The dynamic type $?$ must be neutral to subtyping to avoid the pitfall
+from Thatte's quasi-static typing.
+The rule \textsc{S-FUN} defines the subtyping relation for function types,
+which are contravariant on the argument type and covariant on the return type.
+
+\begin{figure}[!ht]
+\dstart
+$$
+\begin{array}{c}
+\begin{array}{c}
+\Number \lesssim \Number \mylabel{C-NUM}
+\end{array}
+\;
+\begin{array}{c}
+\String \lesssim \String \mylabel{C-STR}
+\end{array}
+\\ \\
+\begin{array}{c}
+T \lesssim \;? \mylabel{C-ANY1}
+\end{array}
+\;
+\begin{array}{c}
+? \lesssim T \mylabel{C-ANY2}
+\end{array}
+\\ \\
+\begin{array}{c}
+\dfrac{T_{3} \lesssim T_{1} \;\;\; T_{2} \lesssim T_{4}}
+      {T_{1} \rightarrow T_{2} \lesssim T_{3} \rightarrow T_{4}} \mylabel{C-FUN}
+\end{array}
+\end{array}
+$$
+\dend
+\caption{The consistent-subtyping relation}
+\label{fig:consistent-subtyping}
+\end{figure}
+
+Figure \ref{fig:consistent-subtyping} combines consistency and subtyping to
+compose the consistent-subtyping relation for the simply typed
+lambda calculus extended with the dynamic type $?$.
+When we combine consistency and subtyping, we are making subtyping handle
+what casts are safe among static types, and we are making consistency
+handle the casts that involve the dynamic type $?$.
+The consistent-subtyping relation is not transitive, and that is the
+reason why the dynamic type $?$ is not neutral to this relation.
 

### 2014-11-02 doc/thesis/introduction.tex
Minor improvements to the introduction

@@
 migrate their code from a dynamically typed to a statically
-typed language, when their simple scripts become complex programs
+typed language, when their simple scripts evolve into complex programs
 \citep{tobin-hochstadt2006ims}.
@@
 
-In this work we present the design and implementation of Typed Lua:
+In this work we present the design and evaluation of Typed Lua:
 an optional type system for Lua that is rich enough to
@@
 Lua is a small scripting language, so designing an optional type
-system for Lua should shed some light on optional type systems for
+system for Lua may shed some light on optional type systems for
 scripting languages in general.
@@
 extension language.
-Thus, implementing an optional type system for Lua should offer Lua
+Thus, implementing an optional type system for Lua offers Lua
 programmers one way to obtain most of the advantages of static typing
@@
 and generates Lua code that runs in unmodified Lua implementations.
-We have an ongoing implementation of the Typed Lua compiler that is
+We have an implementation of the Typed Lua compiler that is
 available online\footnote{https://github.com/andremm/typedlua}.
@@
 
-We also believe that Typed Lua will help programmers give a more
+We also believe that Typed Lua help programmers give a more
 formal documentation to already existing Lua code, as static types
@@
 typing.
-In Chapter \ref{chap:typedlua} we present the design and implementation
-of Typed Lua along with the reasons why it includes features of
-object-oriented programming.
-In Chapter \ref{chap:cases} we present some case studies.
+In Chapter \ref{chap:typedlua} we use code examples to present the
+design of Typed Lua.
+In Chapther \ref{chap:system} we use typing rules to present the
+formalization of Typed Lua.
+In Chapter \ref{chap:evaluation} we discuss some case studies that
+we used to evaluate our design.
 In Chapter \ref{chap:conc} we outline our contributions.

### 2014-11-03 doc/thesis/evaluation.tex
Minor improvements to the introduction
Adding text about typing LuaSocket

@@
 \hline
-LuaSocket & 85 & 22 & 19 & 126 & 67\% \\
+LuaSocket & 109 & 1 & 13 & 123 & 89\% \\
 \hline
@@
 support for SMTP, HTTP, and FTP client protocols, MIME encoding,
-URL manipulation, and LTN012 filters \citep{nehab2008ltn012}.
+URL manipulation, and LTN12 filters \citep{nehab2008ltn012}.
 We chose LuaSocket as a case study because it is the most popular Lua library.
@@
 \multirow{7}{*}{LuaSocket}
-& socket & 50 & 3 & 7 & 60 & 83\% \\
+& socket & 50 & 0 & 10 & 60 & 83\% \\
 \cline{2-7}
-& ftp & 6 & 1 & 1 & 8 & 75\% \\
+& ftp & 5 & 0 & 1 & 6 & 83\% \\
 \cline{2-7}
-& http & 3 & 0 & 2 & 5 & 60\% \\
+& http & 4 & 0 & 1 & 5 & 80\% \\
 \cline{2-7}
-& ltn12 & 10 & 10 & 0 & 20 & 50\% \\
+& smtp & 6 & 0 & 1 & 7 & 86\% \\
 \cline{2-7}
-& mime & 2 & 8 & 7 & 17 & 12\% \\
+& mime & 17 & 0 & 0 & 17 & 100\% \\
 \cline{2-7}
-& smtp & 6 & 0 & 2 & 8 & 75\% \\
+& ltn12 & 19 & 1 & 0 & 20 & 95\% \\
 \cline{2-7}
@@
 which is the C core.
+However, this module includes some functions such as
+\texttt{socket.try}, \texttt{socket.protect}, and \texttt{socket.skip}
+that have a dynamic behavior.
+This means that the return type of these functions depend on the
+type of the values that we use to call them.
+
+We could type most of the members of the modules \texttt{ftp},
+\texttt{http}, and \texttt{smtp}, but we could not precisely type
+the functions \texttt{ftp.get}, \texttt{http.request}, and
+\texttt{smtp.message}.
+
+The function \texttt{ftp.get} downloads data from a given URL,
+and it is difficult to type because its return type depend on
+the type of its parameters.
+Its type is either \texttt{(string) -> (string)?} or
+\texttt{(url\string_argument|url\string_path) -> (number)?}.
+The types \texttt{url\string_argument} and \texttt{url\string_path}
+are table types that describe the URL that should be used to
+download data.
+Both table types include the same fields, except one:
+while the first includes the mandatory field \texttt{argument},
+of type \texttt{string}, the latter includes the mandatory field
+\texttt{path}, also of type \texttt{string}.
+
+The function \texttt{http.request} downloads data from a given URL,
+and it is difficult to type because its return type depend on
+the type of its parameters.
+Its type is either
+\texttt{(string, string?) -> (string, number, \{string : string\}, number)?} or
+\texttt{(url\string_request) -> (number, number, \{string : string\}, number)?}.
+The type \texttt{url\string_request} is a table type containing
+details about the downloading URL.
+
+The function \texttt{smtp.message} creates a function that sends
+an SMTP message body, and it is difficult to type because of the
+type of the message body, which is recursive and makes harder
+its initialization.
+
+The modules \texttt{mime} and \texttt{ltn12} have a strong connection.
+The \texttt{mime} module offers low-level and high-level filters
+that apply and remove some text encodings.
+The low-level filters are written in C, while the high-level filters
+use the function \texttt{ltn12.filter.cycle} along with the low level
+filters to create standard filters.
+Even though we could type all the members of the \texttt{mime} module,
+the function \texttt{ltn12.filter.cycle} is the only member of the
+\texttt{ltn12} module that we could not give a precise type.
+This function is difficult to type because the \texttt{mime} low-level
+filters do not have a standard API.
+
+The \texttt{url} module provides functions that manipulate URLs.
+It is straightforward to type, although it defines a table type for
+parsed URLs that is not precise enough. 
+This type is problematic because it is a record type where all
+fields are optional, and makes any table type that does not include
+the optional fields statically type check.
+
+This case study shows that abusing of dynamic typing usually impacts
+the users of a module, as they have to keep on checking each return
+value before using it to prevent API misuses.
 
@@
 \begin{itemize}
-\item Tidal Lock
-\item Typed Clojure
-\item Gradual Python
-\item Gradualtalk
-\item Typed Racket
-\item Dart
-\item TypeScript
+\item \citep{tidallock}
+\item \citep{bonnaire-sergeant2012typed-clojure}
+\item \citep{vitousek2014deg}
+\item \citep{allende2013gts}
+\item \citep{tobin-hochstadt2008ts} 
+\item \citep{dart}
+\item \citep{typescript}
+\item \citep{politz2012semantics}
 \end{itemize}

### 2014-11-04 doc/thesis/evaluation.tex
Adding new case studies: http-digest, typical, and mod11

@@
 LuaSocket & 109 & 1 & 13 & 123 & 89\% \\
+\hline
+HTTP Digest & 0 & 0 & 1 & 1 & 0\% \\
+\hline
+Typical & 1 & 0 & 0 & 1 & 100\% \\
+\hline
+Modulo 11 & 7 & 0 & 2 & 9 & 78\% \\
 \hline
@@
 
+\section{HTTP Digest}
+
+The HTTP Digest library \citep{luahttpdigest} implements client side
+HTTP digest authentication for Lua.
+We randonly chose this library as a case study to evaluate Typed Lua
+for rewriting an existing Lua module.
+Table \ref{tab:evalhttpdigest} summarizes the evaluation results for HTTP Digest.
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|c|}
+\hline
+\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\hline
+\multirow{1}{*}{HTTP Digest}
+& http-digest & 0 & 0 & 1 & 1 & 0\% \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Evaluation results for HTTP Digest}
+\label{tab:evalhttpdigest}
+\end{table}
+
+It is difficul to type the interface of the \texttt{http-digest} module
+because it is an extension to the \texttt{http} module from LuaSocket.
+The \texttt{http-digest} module only exports the function
+\texttt{http-digest.request}, which extends the function
+\texttt{http.request} with MD5 authentication.
+Like \texttt{http.request}, the return type of \texttt{http-digest.request}
+also depends on the type of its parameters.
+Its type is either
+\texttt{(string) -> (string, number, \{string : string\})?} or
+\texttt{(url\string_request) -> (number, number, \{string : string\})?}.
+
+Even though we could not precisely type the interface that \texttt{http-digest}
+exports, we could use only static types to rewrite it, and they pointed a bug
+in the code.
+The problem was related to the way the library was loading the MD5 library
+that should be used. 
+This part of the code checks the existence of three different MD5 libraries
+in the system, and uses the first one that is available, or generates an
+error when none is available.
+The code that loads the first option was fine, but the code that loads the
+second and third options were trying to access an undefined global variable.
+
+\section{Typical}
+
+Typical \citep{luatypical} is a library that extends the behavior of the
+function \texttt{type}. 
+We randonly chose this library as a case study to evaluate Typed Lua
+for rewriting an existing Lua module.
+Table \ref{tab:evaltypical} summarizes the evaluation results for Typical.
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|c|}
+\hline
+\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\hline
+\multirow{1}{*}{Typical}
+& typical & 1 & 0 & 0 & 1 & 100\% \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Evaluation results for Typical}
+\label{tab:evaltypical}
+\end{table}
+
+The interface of the \texttt{typical} module is straightforward to type,
+as it contains only the function \texttt{typical.type},
+which has the same type of the function \texttt{type}: \texttt{(value) -> (string)}.
+
+However, we hit some limitations of our type system to rewrite this module.
+First, it uses the function \texttt{getmetatable} to get a table and
+checks whether this table has the field \texttt{\string_\string_type} or not.
+We could not give a precise type to \texttt{getmetatable}, so we used the dynamic
+type \texttt{any} as its return type, and this generates a warning that we could not
+remove.
+Second, the module returns a metatable that extends \texttt{\string_\string_call}
+with \texttt{typical.type}, that is, we can use the module itself as a function,
+though it is a table.
+Our type system still does not support metatables, so we did not extend our version
+of the \texttt{typical} module to support \texttt{\string_\string_call}.
+Third, the module uses \texttt{ipairs} to iterate over an array of functions,
+but our type system also has limited support to \texttt{ipairs}, and generates
+a warning when we try to use the indexed value inside the \texttt{for} body.
+We removed this warning using the numeric \texttt{for} to perform the same loop.
+
+\section{Modulo 11}
+
+Modulo 11 \citep{luamod11} is a library that generates and verifies
+modulo 11 numbers.
+We randonly chose this library as a case study to evaluate Typed Lua
+for rewriting an existing Lua module.
+Table \ref{tab:evalmod11} summarizes the evaluation results for Typical.
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|c|}
+\hline
+\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\hline
+\multirow{1}{*}{Modulo 11}
+& mod11 & 7 & 0 & 2 & 9 & 78\% \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Evaluation results for Modulo 11}
+\label{tab:evalmod11}
+\end{table}
+
+The \texttt{mod11} module was written using an object-oriented idiom that
+our type system does not support, and that is the reason why we could not
+type all the members of its interface.
+More precisely, the original code uses \texttt{setmetatable} to hide
+two attributes, which our type system cannot hide.
+
+In addition, it returns a metatable that extends \texttt{\string_\string_call}
+with the class constructor.
+This allows us to use the module itself to create new instances of a
+Modulo 11 number.
+However, our type system does not support this feature, and we need to
+make explicit calls to the constructor whenever we want to create a
+new instance.
+
+Even though we had these two issues to rewrite the \texttt{mod11} module,
+we could use only static types to rewrite it, and found some interesting
+points.
+The code relies a lot on implicit conversions between strings and numbers,
+and some parts of the code keep on changing the type of local variables.
+These are two practices that may hide bugs.
+
 \section{Related Work}

### 2014-11-05 doc/thesis/evaluation.tex
Adding evaluation data for Lua Standard Libraries 5.3 (beta) and TLC

@@
 \hline
-Lua Standard Libraries & 94 & 5 & 34 & 133 & 71\% \\
+Lua Standard Libraries 5.2 & 94 & 5 & 34 & 133 & 71\% \\
+\hline
+Lua Standard Libraries 5.3 & 87 & 6 & 36 & 129 & 67\% \\
 \hline
@@
 Modulo 11 & 7 & 0 & 2 & 9 & 78\% \\
+\hline
+Typed Lua Compiler & 144 & 0 & 10 & 154 & 94\% \\
 \hline
@@
 \end{center}
-\caption{Evaluation results for Lua Standard Libraries}
+\caption{Evaluation results for Lua Standard Libraries 5.2}
 \label{tab:evallsl}
+\end{table}
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|c|}
+\hline
+\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\hline
+\multirow{9}{*}{Lua Standard Libraries}
+& base & 9 & 1 & 16 & 26 & 35\% \\
+\cline{2-7}
+& coroutine & 1 & 0 & 6 & 7 & 14\% \\
+\cline{2-7}
+& package & 5 & 0 & 3 & 8 & 62\% \\
+\cline{2-7}
+& string & 14 & 0 & 2 & 16 & 88\% \\
+\cline{2-7}
+& utf8 & 6 & 0 & 0 & 6 & 100\% \\
+\cline{2-7}
+& table & 1 & 5 & 1 & 7 & 14\% \\
+\cline{2-7}
+& math & 26 & 0 & 1 & 27 & 96\% \\
+\cline{2-7}
+& io & 15 & 0 & 6 & 21 & 71\% \\
+\cline{2-7}
+& os & 10 & 0 & 1 & 11 & 91\% \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Evaluation results for Lua Standard Libraries 5.3}
+\label{tab:evallsl53}
 \end{table}
@@
 
+\section{Typed Lua Compiler}
+
+The Typed Lua compiler is the last case study that we evaluated, and the
+Table \ref{tab:evaltlc} summarizes the evaluation results.
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|c|}
+\hline
+\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\hline
+\multirow{8}{*}{Typed Lua Compiler}
+& tlast & 46 & 0 & 1 & 47 & 98\% \\
+\cline{2-7}
+& tltype & 65 & 0 & 0 & 65 & 100\% \\
+\cline{2-7}
+& tlst & 26 & 0 & 0 & 26 & 100\% \\
+\cline{2-7}
+& tllexer & 2 & 0 & 9 & 11 & 18\% \\
+\cline{2-7}
+& tlparser & 1 & 0 & 0 & 1 & 100\% \\
+\cline{2-7}
+& tldparser & 1 & 0 & 0 & 1 & 100\% \\
+\cline{2-7}
+& tlchecker & 2 & 0 & 0 & 2 & 100\% \\
+\cline{2-7}
+& tlcode & 1 & 0 & 0 & 1 & 100\% \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Evaluation results for Typed Lua Compiler}
+\label{tab:evaltlc}
+\end{table}
+
+The \texttt{tlast} module implements the abstract syntax tree that
+the compiler uses to perform the type checking.
+We could not precisely type just one function because its return
+type depend on the type of its input parameter.
+
+The \texttt{tltype} module implements the types introduced by
+Typed Lua and the subtyping and consistent-subtyping relationsr;
+it was straightforward to type.
+
+The \texttt{tlst} module implements the environment that the
+compiler uses during the type checking, and it was also straightforward
+to type.
+
+The \texttt{tllexer} module defines common lexical rules for
+the Typed Lua parser and the parser of the description files.
+This module is hard to type because it exports LPeg \citep{lpeg,ierusalimschy2009lpeg}
+patterns, which our type system cannot precisely type because it does
+not support operator overloading.
+
+The \texttt{tlparser} and \texttt{tldparser} implement the Typed Lua
+parser and the parser of the description files, respectively.
+Even though they use LPeg to implement the grammar rules,
+they only export a parsing function that uses LPeg to parse a
+string and return the corresponding AST.
+
+The \texttt{tlchecker} and \texttt{tlcode} modules are also
+straightforward to type.
+The first traverses the AST to perform type checking,
+while the latter traverses the AST to perform code generation.
+
+Even though we could precisely type the interface of most of
+the modules, we had some issues to write mutually recursive
+functions that traverses the AST.
+In this case we had to rely on the dynamic type to defer the
+function declarations.
+
 \section{Related Work}
@@
 \item \citep{typescript}
+\item \citep{bierman2014typescript}
 \item \citep{politz2012semantics}

### 2014-11-07 doc/thesis/evaluation.tex
Splitting the evaluation chapter into two sections: case studies and related work

@@
 We performed some case studies on existing Lua libraries
-to evaluate the design of our type system.
+to evaluate the applicability of our type system.
 For each library, we used Typed Lua to either rewrite its modules
@@
 Typed Lua's description files.
-In this chapter we present our evaluation results, discuss
-some interesting cases, and compare our type system to
-related work.
+In Section \ref{sec:cases} we present our evaluation results
+and discuss some interesting cases.
+
+We also compared our type system to other type systems,
+as a way to evaluate our contributions.
+In Section \ref{sec:relatedwork} we discuss what are the novel
+features introduced by our type system,
+what are the features that it shares with other type systems,
+and whether we can incorporate some interesting features from
+other type systems.
+
+\section{Case Studies}
+\label{sec:cases}
+
+We selected seven case studies to evaluate two aspects of Typed Lua:
+\begin{enumerate}
+\item how precisely it can describe the type of the interface of a module;
+\item if it provides guarantees that the code corresponds to the interface.
+\end{enumerate}
+
+The Lua Standard Library \citep{luamanual} was our first case study.
+We started to think about how we would type its modules at the same time
+that we started to design our type system, as it could give us some
+hints on our type system.
+And it did: optional parameters and overloading on the return type
+are two Lua features that our type system should handle to allow us
+typing some of the functions that the standard library implements.
+
+The second case study that we chose was the MD5 library \citep{lmd5},
+because it is simple and contains just one module.
+We needed a simple case study to introduce Typed Lua's description
+files and \texttt{userdata} declarations.
+These Typed Lua's mechanisms allow programmers to give statically typed
+interfaces to Lua libraries that are written in C.
+
+LuaSocket \citep{luasocket} was the third case study that we used to
+evaluate Typed Lua.
+We chose LuaSocket as a case study because it is the most popular Lua library.
+
+There are three case studies that we randomly selected from the LuaRocks
+repository, they are: HTTP Digest \citep{luahttpdigest},
+Typical \citep{luatypical}, and Mod 11 \citep{luamod11}.
+The first provides client side HTTP digest authentication for Lua.
+The second is an extension to the primitive function \texttt{type}.
+The third is a generator and checker of modulo 11 numbers.
+We randomly selected these libraries to evaluate Typed Lua for
+rewriting existing Lua modules.
+
+The Typed Lua compiler is the last case study that we evaluated.
+We chose it as a case study because it is a large application.
+In addition, bootstrapping the compiler also provides some
+confidence on the type system.
 
@@
 \hline
-Lua Standard Libraries 5.2 & 94 & 5 & 34 & 133 & 71\% \\
+Lua 5.2 Standard Library & 94 & 5 & 34 & 133 & 71\% \\
 \hline
-Lua Standard Libraries 5.3 & 87 & 6 & 36 & 129 & 67\% \\
+Lua 5.3 Standard Library & 87 & 6 & 36 & 129 & 67\% \\
 \hline
@@
 
-Table \ref{tab:evalbycase} sumarizes our evaluation results for each
-case study that we used Typed Lua for typing their members.
-We split the members that we typed into three categories:
+Table \ref{tab:evalbycase} summarizes our evaluation results for each
+of the case studies that we used Typed Lua for typing their members.
+We split the members of each case study into three categories:
 \emph{easy}, \emph{poly}, and \emph{hard}.
-The \emph{easy} category represents the members that we could give
-a precise static type for them.
-The \emph{poly} category represents the members that we believe that
-parametric polymorphism would help give a precise static type for them,
-and we had to rely on the dynamic type to type them because our
-type system does not support parametric polymorphism.
-The \emph{hard} category represents the members that even parametric
-polymorphism would not help give a precise static type for them,
-and we also had to rely on the dynamic type to type them;
+The \emph{easy} category groups the members that we could give
+a precise static type.
+The \emph{poly} category groups the members that we made minimal
+use of the dynamic type as a replacement for the lack of type parameters.
+The \emph{hard} category represents the members that either we
+had to rely on the dynamic type to type them or we used a static
+type that is not precise enough;
 for instance, this category includes functions that require
@@
 are under the \emph{easy} category for each case study.
-This percentage is our evaluation of Typed Lua, as it
-represents how much static typing we could introduce to each one of
-our case studies.
+This percentage gives a lower bound on how much static type safety
+we could add to each one of our case studies.
 
-Before comparing our type system to related work, we will discuss
-each case study in more detail.
-For each case study, we will split the evaluation results according
-to the modules that each one of them include.
-This shall allow us to better discuss the contributions and limitations
-of our type system.
+In the following sections we discuss each case study in more detail.
+For each case study, we split the evaluation results according
+to the modules that each one of them includes.
+This separation by case study allows us to better discuss the
+contributions and limitations of our type system.
 
-\section{Lua Standard Libraries}
+\subsection{Lua Standard Library}
 
-The Lua Standard Libraries \citep{luamanual} were our first case study.
-We started to think about how we would type them at the same time that
-we started to design our type system, as they could give us some hints
-on our type system.
-And they did: optional parameters and overloading on the return type
-are two Lua features that our type system should handle to allow us
-typing some of the functions that the standard libraries provide.
-
-All libraries are separated C modules, and we used Typed Lua's description
-files to give a statically typed interface to each module.
+All of the modules in the standard library are implemented in C, so
+we used Typed Lua to type just the interface of each module.
 The \texttt{debug} module is the only one that we did not include in our
@@
 For instance, we can use its function \texttt{setlocal} to change the value
-of a local variable that is outside of the scope.
-Table \ref{tab:evallsl} sumarizes the evaluation results for the Lua Standard Libraries.
+of a local variable that is not visible in the current scope.
+Table \ref{tab:evallsl} summarizes the evaluation results for the Lua Standard Library.
 
@@
 \hline
-\multirow{9}{*}{Lua Standard Libraries}
+\multirow{9}{*}{Lua Standard Library}
 & base & 9 & 1 & 16 & 26 & 35\% \\
@@
 \end{center}
-\caption{Evaluation results for Lua Standard Libraries 5.2}
+\caption{Evaluation results for Lua 5.2 Standard Library}
 \label{tab:evallsl}
@@
 \end{center}
-\caption{Evaluation results for Lua Standard Libraries 5.3}
+\caption{Evaluation results for Lua 5.3 Standard Library}
 \label{tab:evallsl53}
@@
 The \texttt{base} module was quite hard to type because it
-includes several functions that have dynamic behavior.
+includes several functions that have a dynamic behavior.
 For instance, the function \texttt{next} traverses all the
@@
 We could not type the \texttt{coroutine} module because our
-type system does not include the type \emph{thread}.
+type system cannot describe the computational effects of a program.
 Lua has one-shot delimited continuations \citep{james2011yield}
@@
 We could type all the members of the \texttt{string} module.
-Even though we can use the \texttt{string} functions in object-oriented style,
-we did not use an \emph{userdata} declaration to handle this feature,
-as it would complicate the subtyping rules because the type \texttt{string}
-would become a table type instead of a base type.
-It was simpler and more starightforward to handle this feature as a
+Lua programmers often use the functions of the \texttt{string} module
+in an object-oriented style, and we treat this feature as a
 special case in the implementation.
-For this reason, we did not include the typing of the \texttt{string} methods
-in our evaluation results.
+For this reason, we did not include the typing of the \texttt{string}
+methods in our evaluation results.
 
 The \texttt{table} module is specially difficult to type because
-all of its members require parametric polymorphism.
+almost all of its members require parametric polymorphism.
 All the functions of this module either receive or return a list
@@
 
-The evaluation results show that our type system should include effect
-types, parametric polymorphism, and intersection types, as these
-features would help to increase the static typing of the Lua Standard Libraries.
+The evaluation results show that our type system should include
+parametric polymorphism, intersection types, and effect types,
+as these features would help us increase the static typing of the
+Lua Standard Library.
+Parametric polymorphism would allow us to define generic function and table types.
+Intersection types would allow us to define overloaded function types.
 Effect types would allow us to type coroutines.
-Parametric polymorphism would allow us to define generic function and table types.
-Intersection types would allow us to define function types that have
-different return types according to their input types.
 
-\section{MD5}
+\subsection{MD5}
 
-We chose the MD5 library \citep{lmd5} as a case study because it is
-simple and contains just one module.
-We needed a simple case study to introduce Typed Lua's description
-files and \texttt{userdata} declarations.
-These Typed Lua's mechanims allow programmers to give statically typed
-interfaces to Lua libraries that are written in C.
-Table \ref{tab:evalmd5} sumarizes the evaluation results for MD5.
+The MD5 library is an OpenSSL based message digest library for Lua.
+It contains just the \texttt{md5} module that is written in C,
+so we used Typed Lua's description file to type it.
+Table \ref{tab:evalmd5} summarizes the evaluation results for MD5.
 
@@
 Even tough it was quite straightforward to type the MD5 library,
-we found a little difference between its documentation and its static typing.
-The documentation suggested that the typing of the function \texttt{update}
+we found a little difference between its documentation and its behavior.
+The documentation suggested that the typing of the function \texttt{md5.update}
 should be \texttt{(md5\string_context, string) -> (md5\string_context)},
@@
 its actual type is \texttt{(md5\string_context, string*) -> (md5\string_context)},
-that is, we can pass zero or more strings to \texttt{update}.
+that is, we can pass zero or more strings to \texttt{md5.update}.
 
@@
 In this case study, reading its source code and test script was essential
-to confirm the typing of \texttt{update}.
+to confirm the typing of \texttt{md5.update}.
 
-\section{LuaSocket}
+\subsection{LuaSocket}
 
-LuaSocket \citep{luasocket} is a library that adds network support to Lua,
+LuaSocket is a library that adds network support to Lua,
 and it is split into two parts: a core that is written in C and a set of
@@
 URL manipulation, and LTN12 filters \citep{nehab2008ltn012}.
-We chose LuaSocket as a case study because it is the most popular Lua library.
 We used Typed Lua's description files to type both parts.
@@
 a Lua module instead of rewriting it.
-Table \ref{tab:evalsocket} sumarizes the evaluation results for LuaSocket.
+Table \ref{tab:evalsocket} summarizes the evaluation results for LuaSocket.
 
@@
 
-\section{HTTP Digest}
+\subsection{HTTP Digest}
 
-The HTTP Digest library \citep{luahttpdigest} implements client side
-HTTP digest authentication for Lua.
-We randonly chose this library as a case study to evaluate Typed Lua
-for rewriting an existing Lua module.
+The HTTP Digest library implements client side HTTP digest authentication for Lua.
 Table \ref{tab:evalhttpdigest} summarizes the evaluation results for HTTP Digest.
@@
 
-It is difficul to type the interface of the \texttt{http-digest} module
+It is difficult to type the interface of the \texttt{http-digest} module
 because it is an extension to the \texttt{http} module from LuaSocket.
@@
 
-\section{Typical}
+\subsection{Typical}
 
-Typical \citep{luatypical} is a library that extends the behavior of the
-function \texttt{type}. 
-We randonly chose this library as a case study to evaluate Typed Lua
-for rewriting an existing Lua module.
+Typical is a library that extends the behavior of the function \texttt{type}. 
 Table \ref{tab:evaltypical} summarizes the evaluation results for Typical.
@@
 
-\section{Modulo 11}
+\subsection{Modulo 11}
 
-Modulo 11 \citep{luamod11} is a library that generates and verifies
-modulo 11 numbers.
-We randonly chose this library as a case study to evaluate Typed Lua
-for rewriting an existing Lua module.
+Modulo 11 is a library that generates and verifies modulo 11 numbers.
 Table \ref{tab:evalmod11} summarizes the evaluation results for Typical.
@@
 
-\section{Typed Lua Compiler}
+\subsection{Typed Lua Compiler}
 
@@
 The \texttt{tltype} module implements the types introduced by
-Typed Lua and the subtyping and consistent-subtyping relationsr;
+Typed Lua and the subtyping and consistent-subtyping relations;
 it was straightforward to type.
@@
 \section{Related Work}
+\label{sec:relatedwork}
 

### 2014-11-07 doc/thesis/introduction.tex
Splitting the evaluation chapter into two sections: case studies and related work

@@
 
-Lua is a small scripting language, so designing an optional type
-system for Lua may shed some light on optional type systems for
-scripting languages in general.
-Lua provides mechanisms for organizing a program in modules and
-writing object-oriented programs, but it does not set policy on how
-these features should behave, due to its use as an embedded and
-extension language.
+Lua is a small imperative language with first-class functions
+(with proper lexical scoping) where the only data structure
+mechanism is the \emph{table}, an associative array that can
+represent arrays, records, maps, modules, objects, etc.
+Tables also have syntactic sugar and metaprogramming support
+through operator overloading built into the language.
+Unlike other scripting languages, Lua has very limited coercion
+among different data types.
+We believe that designing an optional type system for Lua may
+shed some light on optional type systems for scripting languages
+in general, as Lua is a small scripting language that shares
+some features with other scripting languages such as JavaScript.
+
+Lua prefers to provide mechanisms instead of fixed policies due
+to its primary use that has always been as an embedded language
+for configuration and extension of other applications.
+This means that even features such as a module system and
+object orientation are a matter of convention instead of built
+into the language.
+The result is a fragmented ecosystem of libraries, and different
+ideas among Lua programmers on how they should use the language
+features, and how they should structure programs.
+
+The lack of standard policies is a challenge for the design of
+an optional type system for Lua.
+For this reason, we are not relying just on the semantics of
+the language to design our type system.
+We also run a mostly automated survey of Lua idioms used in a
+large corpus of Lua libraries, and it also helped us on the
+design of Typed Lua.
+
+So far, Typed Lua is a Lua extension that allows statically typed
+code to coexist and interact with dynamically typed code
+through optional type annotations, and it also adds standard
+constructions for structuring Lua code.
+The Typed Lua compiler warns programmers about type errors,
+but always generates Lua code that runs in unmodified Lua implementations.
+Programmers can enjoy some of the benefits of static types
+even without converting existing Lua modules to Typed Lua:
+they can export a statically typed interface to a dynamically typed module,
+and statically typed users of the module can use the Typed Lua compiler
+to check their use of the module.
 Thus, implementing an optional type system for Lua offers Lua
@@
 without compromising the simplicity and flexibility of dynamic typing.
-
-So far, Typed Lua is a strict superset of Lua that
-provides optional type annotations, compile-time type checking, and
-class-based object-oriented programming through the definition of
-classes, interfaces, and modules.
-In practice, we implement Typed Lua as a programming language that
-extends Lua syntax to add optional type annotations and standard
-constructions for structuring Lua code.
-The compiler uses static types to perform compile-time type
-checking, but also allows Lua code to coexist with Typed Lua code,
-and generates Lua code that runs in unmodified Lua implementations.
 We have an implementation of the Typed Lua compiler that is
@@
 applications evolve to complex systems rather than just scripts.
+We shall see that Typed Lua introduces the refinement of
+tables to support the common way that Lua programmers use
+them to represent both modules and objects.
 
@@
 
-In Chapter \ref{chap:review} we review the literature about static and
-dynamic typing in the same language, which includes the discussion
-about the particularities between optional type systems and gradual
-typing.
+In Chapter \ref{chap:review} we review the literature about static
+and dynamic typing in the same language, we discuss the differences
+between optional type systems and gradual typing, and we also
+present the results of our survey on Lua idioms.
 In Chapter \ref{chap:typedlua} we use code examples to present the
 design of Typed Lua.
-In Chapther \ref{chap:system} we use typing rules to present the
+In Chapter \ref{chap:system} we use typing rules to present the
 formalization of Typed Lua.
 In Chapter \ref{chap:evaluation} we discuss some case studies that
-we used to evaluate our design.
+we used to evaluate our design, and we also compare our design
+to related work.
 In Chapter \ref{chap:conc} we outline our contributions.

### 2014-11-07 doc/thesis/evaluation.tex
Adding the inter category to the evaluation results

@@
 \begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|}
+\begin{tabular}{|l|c|c|c|c|c|c|c|}
 \hline
-\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
 \hline
-Lua 5.2 Standard Library & 94 & 5 & 34 & 133 & 71\% \\
+Lua Standard Library & 87 & 6 & 0 & 36 & 129 & 67\% \\
 \hline
-Lua 5.3 Standard Library & 87 & 6 & 36 & 129 & 67\% \\
+MD5 & 13 & 0 & 0 & 0 & 13 & 100\% \\
 \hline
-MD5 & 13 & 0 & 0 & 13 & 100\% \\
+LuaSocket & 109 & 1 & 0 & 13 & 123 & 89\% \\
 \hline
-LuaSocket & 109 & 1 & 13 & 123 & 89\% \\
+HTTP Digest & 0 & 0 & 0 & 1 & 1 & 0\% \\
 \hline
-HTTP Digest & 0 & 0 & 1 & 1 & 0\% \\
+Typical & 1 & 0 & 0 & 0 & 1 & 100\% \\
 \hline
-Typical & 1 & 0 & 0 & 1 & 100\% \\
+Modulo 11 & 7 & 0 & 0 & 2 & 9 & 78\% \\
 \hline
-Modulo 11 & 7 & 0 & 2 & 9 & 78\% \\
-\hline
-Typed Lua Compiler & 144 & 0 & 10 & 154 & 94\% \\
+Typed Lua Compiler & 144 & 0 & 0 & 10 & 154 & 94\% \\
 \hline
@@
 of the case studies that we used Typed Lua for typing their members.
-We split the members of each case study into three categories:
-\emph{easy}, \emph{poly}, and \emph{hard}.
+We split the members of each case study into four categories:
+\emph{easy}, \emph{poly}, \emph{inter}, and \emph{hard}.
 The \emph{easy} category groups the members that we could give
@@
 use of the dynamic type as a replacement for the lack of type parameters.
-The \emph{hard} category represents the members that either we
-had to rely on the dynamic type to type them or we used a static
-type that is not precise enough;
-for instance, this category includes functions that require
-intersection types to describe their precise static type.
+The \emph{inter} category groups the members that we could give
+a static type that is not precise enough because they require
+intersection types.
+The \emph{hard} category groups the members that we had to rely
+on the dynamic type because our type system cannot give a precise
+static type to them.
 The last column of the table shows the percentage of members that
@@
 \begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|c|}
 \hline
-\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
 \hline
-\multirow{9}{*}{Lua Standard Library}
-& base & 9 & 1 & 16 & 26 & 35\% \\
-\cline{2-7}
-& coroutine & 0 & 0 & 6 & 6 & 0\% \\
-\cline{2-7}
-& package & 5 & 0 & 3 & 8 & 62\% \\
-\cline{2-7}
-& string & 14 & 0 & 0 & 14 & 100\% \\
-\cline{2-7}
-& table & 1 & 4 & 1 & 6 & 17\% \\
-\cline{2-7}
-& math & 28 & 0 & 1 & 29 & 97\% \\
-\cline{2-7}
-& bit32 & 12 & 0 & 0 & 12 & 100\% \\
-\cline{2-7}
-& io & 15 & 0 & 6 & 21 & 71\% \\
-\cline{2-7}
-& os & 10 & 0 & 1 & 11 & 91\% \\
+base & 9 & 1 & 0 & 16 & 26 & 35\% \\
+\hline
+coroutine & 1 & 0 & 0 & 6 & 7 & 14\% \\
+\hline
+package & 5 & 0 & 0 & 3 & 8 & 62\% \\
+\hline
+string & 14 & 0 & 0 & 2 & 16 & 88\% \\
+\hline
+utf8 & 6 & 0 & 0 & 0 & 6 & 100\% \\
+\hline
+table & 1 & 5 & 0 & 1 & 7 & 14\% \\
+\hline
+math & 26 & 0 & 0 & 1 & 27 & 96\% \\
+\hline
+io & 15 & 0 & 0 & 6 & 21 & 71\% \\
+\hline
+os & 10 & 0 & 0 & 1 & 11 & 91\% \\
 \hline
@@
 \end{center}
-\caption{Evaluation results for Lua 5.2 Standard Library}
+\caption{Evaluation results for Lua Standard Library}
 \label{tab:evallsl}
-\end{table}
-
-\begin{table}[!ht]
-\begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|}
-\hline
-\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
-\hline
-\multirow{9}{*}{Lua Standard Library}
-& base & 9 & 1 & 16 & 26 & 35\% \\
-\cline{2-7}
-& coroutine & 1 & 0 & 6 & 7 & 14\% \\
-\cline{2-7}
-& package & 5 & 0 & 3 & 8 & 62\% \\
-\cline{2-7}
-& string & 14 & 0 & 2 & 16 & 88\% \\
-\cline{2-7}
-& utf8 & 6 & 0 & 0 & 6 & 100\% \\
-\cline{2-7}
-& table & 1 & 5 & 1 & 7 & 14\% \\
-\cline{2-7}
-& math & 26 & 0 & 1 & 27 & 96\% \\
-\cline{2-7}
-& io & 15 & 0 & 6 & 21 & 71\% \\
-\cline{2-7}
-& os & 10 & 0 & 1 & 11 & 91\% \\
-\hline
-\end{tabular}
-\end{center}
-\caption{Evaluation results for Lua 5.3 Standard Library}
-\label{tab:evallsl53}
 \end{table}
@@
 \begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|c|}
 \hline
-\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
 \hline
-\multirow{1}{*}{MD5}
-& md5 & 13 & 0 & 0 & 13 & 100\% \\
+md5 & 13 & 0 & 0 & 0 & 13 & 100\% \\
 \hline
@@
 \begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|c|}
 \hline
-\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
 \hline
-\multirow{7}{*}{LuaSocket}
-& socket & 50 & 0 & 10 & 60 & 83\% \\
-\cline{2-7}
-& ftp & 5 & 0 & 1 & 6 & 83\% \\
-\cline{2-7}
-& http & 4 & 0 & 1 & 5 & 80\% \\
-\cline{2-7}
-& smtp & 6 & 0 & 1 & 7 & 86\% \\
-\cline{2-7}
-& mime & 17 & 0 & 0 & 17 & 100\% \\
-\cline{2-7}
-& ltn12 & 19 & 1 & 0 & 20 & 95\% \\
-\cline{2-7}
-& url & 8 & 0 & 0 & 8 & 100\% \\
+socket & 50 & 0 & 0 & 10 & 60 & 83\% \\
+\hline
+ftp & 5 & 0 & 0 & 1 & 6 & 83\% \\
+\hline
+http & 4 & 0 & 0 & 1 & 5 & 80\% \\
+\hline
+smtp & 6 & 0 & 0 & 1 & 7 & 86\% \\
+\hline
+mime & 17 & 0 & 0 & 0 & 17 & 100\% \\
+\hline
+ltn12 & 19 & 1 & 0 & 0 & 20 & 95\% \\
+\hline
+url & 8 & 0 & 0 & 0 & 8 & 100\% \\
 \hline
@@
 \begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|c|}
 \hline
-\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
 \hline
-\multirow{1}{*}{HTTP Digest}
-& http-digest & 0 & 0 & 1 & 1 & 0\% \\
+http-digest & 0 & 0 & 0 & 1 & 1 & 0\% \\
 \hline
@@
 \begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|c|}
 \hline
-\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
 \hline
-\multirow{1}{*}{Typical}
-& typical & 1 & 0 & 0 & 1 & 100\% \\
+typical & 1 & 0 & 0 & 0 & 1 & 100\% \\
 \hline
@@
 \begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|c|}
 \hline
-\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
 \hline
-\multirow{1}{*}{Modulo 11}
-& mod11 & 7 & 0 & 2 & 9 & 78\% \\
+mod11 & 7 & 0 & 0 & 2 & 9 & 78\% \\
 \hline
@@
 \begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|c|}
 \hline
-\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
 \hline
-\multirow{8}{*}{Typed Lua Compiler}
-& tlast & 46 & 0 & 1 & 47 & 98\% \\
-\cline{2-7}
-& tltype & 65 & 0 & 0 & 65 & 100\% \\
-\cline{2-7}
-& tlst & 26 & 0 & 0 & 26 & 100\% \\
-\cline{2-7}
-& tllexer & 2 & 0 & 9 & 11 & 18\% \\
-\cline{2-7}
-& tlparser & 1 & 0 & 0 & 1 & 100\% \\
-\cline{2-7}
-& tldparser & 1 & 0 & 0 & 1 & 100\% \\
-\cline{2-7}
-& tlchecker & 2 & 0 & 0 & 2 & 100\% \\
-\cline{2-7}
-& tlcode & 1 & 0 & 0 & 1 & 100\% \\
+tlast & 46 & 0 & 0 & 1 & 47 & 98\% \\
+\hline
+tltype & 65 & 0 & 0 & 0 & 65 & 100\% \\
+\hline
+tlst & 26 & 0 & 0 & 0 & 26 & 100\% \\
+\hline
+tllexer & 2 & 0 & 0 & 9 & 11 & 18\% \\
+\hline
+tlparser & 1 & 0 & 0 & 0 & 1 & 100\% \\
+\hline
+tldparser & 1 & 0 & 0 & 0 & 1 & 100\% \\
+\hline
+tlchecker & 2 & 0 & 0 & 0 & 2 & 100\% \\
+\hline
+tlcode & 1 & 0 & 0 & 0 & 1 & 100\% \\
 \hline

### 2014-11-10 doc/thesis/system.tex
Writing the examples from chapter 3 in the verbatim abstract syntax

@@
 the other mappings that extisted before.
+
+Função recursiva (another example?).
+
+\begin{verbatim}
+    rec f:(number) -> (number) =
+      fun (x:number)
+        if n == 0 then
+          return 1
+        else
+          return n * f(n - 1)
+    in f(5)
+\end{verbatim}
+
+Dynamic typing and static typing in the same code (another example?).
+
+\begin{verbatim}
+    local a:(number) -> (number) =
+      fun (x:number)
+        if n < 0 then return -n else return n
+    in local d:(any, any) -> (number) =
+         fun (x:any, y:any)
+           skip ; return a(x - y)
+       in d(1, 1)
+\end{verbatim}
+
+Another example:
+
+\begin{verbatim}
+    local x:any = 1 in x = x + 1
+\end{verbatim}
+
+This example is safe, but the following is not:
+
+\begin{verbatim}
+    local x:any, y:any = 1 in x = x + y
+\end{verbatim}
+
+Functions and multiple return.
+
+%\begin{center}
+%\begin{tabular}{llll}
+%\multicolumn{4}{l}{$\mathbf{local} \; m:\Void \rightarrow \Number \times \String =$} \\
+%& \multicolumn{3}{l}{$\mathbf{fun} \; ():\Number \times \String$} \\
+%& & \multicolumn{2}{l}{$\mathbf{skip} \;;\; \mathbf{return} \; 2, ``foo"$} \\
+%\multicolumn{4}{l}{$\mathbf{in}$} \\
+%& \multicolumn{3}{l}{$\mathbf{local} \; s:\Number \times \Number \rightarrow \Number =$} \\
+%& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Number, y:\Number):\Number$} \\
+%& & & \multicolumn{1}{l}{$\mathbf{skip} \;;\; \mathbf{return} \; x + y$} \\
+%& \multicolumn{3}{l}{$\mathbf{in}$} \\
+%& & \multicolumn{2}{l}{$\mathbf{local} \; x:\Number, y:\Number, z:\String = m(), m()$} \\
+%& & \multicolumn{2}{l}{$\mathbf{in} \; s(m(), m())$}
+%\end{tabular}
+%\end{center}
+
+\begin{verbatim}
+  local m:() -> (number, string) =
+    fun () skip ; return 2, "foo"
+  in local s:(number, number) -> (number)
+       fun (x:number, y:number) skip ; return x + y
+     in local x:number, y:number, z:string = m(), m()
+        in s(m(), m())
+\end{verbatim}
+
+Optional parameters.
+
+\begin{verbatim}
+    local m:(string, string|nil) -> (string) =
+      fun (n:string, g:string|nil)
+        g = g or "Hello" ; return g .. n
+    in m("Lua")
+\end{verbatim}
+
+Overloading on the input parameter (another example?).
+
+\begin{verbatim}
+  local o:(string, string|integer) -> (string) =
+    fun (a:string, b:string|integer)
+      if type(b) == "string" then
+        return a .. b
+      else
+        return string["rep"](a, b)
+  in o("foo", 2) 
+\end{verbatim}
+
+Overloading on the return type (another example?).
+
+\begin{verbatim}
+  local d:(integer, integer) -> (integer, integer) | (nil, string) =
+    fun (a:integer, b:integer):(integer, integer) | (nil, string)
+      if b == 0 then
+        return nil, "division by zero"
+      else
+        local r:integer = a \% b
+        in local q:integer = (a - r) / b
+           in skip ; return q, r
+  in local q:X1, r:X2 = d(2, 2)
+     in if q then
+          print(a == b * q + r)
+        else
+          print("ERROR " .. r)
+\end{verbatim}
+
+Table type as hash.
+
+\begin{verbatim}
+    local t:{string:number|nil} = { ["foo"] = 1 } in
+      local x:number|nil = t["foo"] in x = t["bar"]
+\end{verbatim}
+
+Table type as array.
+
+\begin{verbatim}
+    local d:{integer:string|nil} =
+      { "Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat" }
+    in local x:number|nil = d[1] in x = d[8]
+\end{verbatim}
+
+Table type as record (first example).
+
+\begin{verbatim}
+    local p:{"firstname":string, "lastname":string} =
+      { ["lastname"] = "Reed", ["firstname"] = "Lou" }
+    in skip
+\end{verbatim}
+
+Table type as record (second example).
+
+\begin{verbatim}
+    local p:{"firstname":string,
+             "middlename":string|nil,
+             "lastname":string} =
+      { ["firstname"] = "Lewis",
+        ["middlename"] = "Allan",
+        ["lastname"] = "Reed" }
+    in skip
+\end{verbatim}
+
+Table type as record (third example).
+
+\begin{verbatim}
+    local p:{"firstname":string,
+             "middlename":string|nil,
+             "lastname":string} =
+      { ["firstname"] = "Lou",
+        ["lastname"] = "Reed" }
+    in skip
+\end{verbatim}
+
+Refinement of table types (first example).
+
+\begin{verbatim}
+    local p:{} = {}
+    in p["firstname"] <string> = "Lou" ;
+       p["lastname"] <string> = "Reed"
+\end{verbatim}
+
+Refinement of table types (second example).
+
+\begin{verbatim}
+    local p:{} = {}
+    in local b:{} = p
+       in p["firstname"] <string> = "Lou" ;
+          p["lastname"] <string> = "Reed"
+\end{verbatim}
+
+Modules (another example?).
+
+\begin{verbatim}
+    local m:{} = {}
+    in local r:number = 3.14 / 180.0
+       in m["deg"] <(number) -> (number)> =
+          fun (x:number):(number)
+            skip ; return x / r
+        ; m["rad"] <(number) -> (number)> =
+          fun (x:number):(number)
+            skip ; return x * r
+        ; m["pow"] <(number, number) -> (number)> =
+          fun (x:number, y:number):(number)
+            skip ; return x ^ y
+\end{verbatim}
+
+Objects.
+
+\begin{verbatim}
+    local s:{"x":number, "y":number} = { x = 0, y = 0 }
+    in s["new"] <(self, number, number) -> (self)> =
+       fun (self:self, x:number, y:number):(self)
+         local s:{"x":number, "y":number} =
+           setmetatable({}, { __index = self })
+         in s["x"] = x ; s["y"] = y ; return s
+     ; s["move"] <(self, number, number) -> ()> =
+       fun (self:self, x:number, y:number):()
+         self["x"] = self["x"] + x ;
+         self["y"] = self["y"] + y 
+\end{verbatim}

### 2014-11-11 doc/thesis/typedlua.tex
Updating chapter 3 to include the integer type

@@
 In practice, Typed Lua extends the syntax of Lua to introduce
-optional type annotations and class-based object-oriented
-programming through the definition of classes, interfaces, and
-modules.
+optional type annotations.
 In addition, Typed Lua performs local type inference \citep{pierce2000lti}
@@
 Typed Lua also includes a syntactical extension that programmers can use
-to define the types of \emph{userdata}, and uses it to define the type of \emph{thread}. 
+to define the types of \emph{userdata}, and uses it to define the type \emph{thread}. 
 In this section, we present the Typed Lua types that may appear on annotations.
@@
   \textbf{true} \; | \;
-  \textit{Number} \; | \;
+  \textit{Int} \; | \;
+  \textit{Float} \; | \;
   \textit{String}\\
 \textit{basetype} & ::= \; \textbf{boolean} \; | \;
+  \textbf{integer} \; | \;
   \textbf{number} \; | \;
@@
 union types are supertypes of their parts,
-\texttt{boolean}, \texttt{number}, and \texttt{string} are supertypes
-of their respective literal types,
+\texttt{boolean}, \texttt{integer}, \texttt{number}, and \texttt{string}
+are supertypes of their respective literal types,
+\texttt{integer} is subtype of \texttt{number},
 function types are related by contravariance on the input
@@
 \begin{verbatim}
-    local function factorial (n:number):number
+    local function factorial (n:integer):integer
       if n == 0 then
@@
 the local variable \texttt{x}, the compiler uses local type inference to
-assign the type \texttt{number} to \texttt{x}, and this example compiles
+assign the type \texttt{integer} to \texttt{x}, and this example compiles
 without any errors.
 The compiler does not use the literal type \texttt{5},
-instead of the base type \texttt{number}, because this would prevent
-programmers from assigning other numeric values to the variable
+instead of the base type \texttt{integer}, because this would prevent
+programmers from assigning other integer values to the variable
 \texttt{x}.
@@
 \begin{verbatim}
-    local function absolute (n:number):number
+    local function absolute (n:integer):integer
       if n < 0 then
@@
 The function \texttt{distance} receives two parameters of type \texttt{any}
-and returns a value of type \texttt{number}.
+and returns a value of type \texttt{integer}.
 The compiler assigns the dynamic type \texttt{any} to the input
@@
 In this example, Typed Lua cannot guarantee that \texttt{distance} is never
-going to call \texttt{absolute} with a parameter that is not a number,
+going to call \texttt{absolute} with a parameter that is not an integer number,
 because in the semantics of Lua the minus operator can result in a
-value that is not a number.
+value that is not an integer number.
 More precisely, we can call \texttt{absolute} inside \texttt{distance}
 because the expression \texttt{x - y} has type \texttt{any},
-and it is consistent with type \texttt{number}.
+and it is consistent with type \texttt{integer}.
 Still, the dynamic type \texttt{any} may be hiding a value of a type
-that is not a number, making the dynamically typed code crash the
+that is not an integer, making the dynamically typed code crash the
 statically typed code.
@@
 In the previous examples, the type of both \texttt{factorial} and
-\texttt{absolute} is \texttt{(number, value*) -> (number, nil*)},
-and the type of \texttt{distance} is \texttt{(any, any, value*) -> (number, nil*)}.
+\texttt{absolute} is \texttt{(integer, value*) -> (integer, nil*)},
+and the type of \texttt{distance} is \texttt{(any, any, value*) -> (integer, nil*)}.
 If we call \texttt{absolute} with extra arguments, Typed Lua silently ignores
@@
 replace the missing arguments and in the case of \texttt{absolute} generates
-an error, as \texttt{nil} is neither a subtype of \texttt{number} nor consistent
-with \texttt{number}.
+an error, as \texttt{nil} is neither a subtype of \texttt{integer} nor consistent
+with \texttt{integer}.
 
@@
 In the previous examples, the type of both \texttt{factorial} and
-\texttt{absolute} is \texttt{(number) -> (number)},
-and the type of \texttt{distance} is \texttt{(any, any) -> (number)}.
+\texttt{absolute} is \texttt{(integer) -> (integer)},
+and the type of \texttt{distance} is \texttt{(any, any) -> (integer)}.
 If we call \texttt{absolute} with either extra or missing arguments,
@@
     end
-    local function sum (x:number, y:number)
+    local function sum (x:integer, y:integer)
       return x + y
@@
 In the default mode, the type of \texttt{multiple} is
-\texttt{(value*) -> (number, string, nil*)},
-and the type of \texttt{sum} is \texttt{(number, number, value*) -> (number, nil*)}.
+\texttt{(value*) -> (integer, string, nil*)},
+and the type of \texttt{sum} is \texttt{(integer, integer, value*) -> (integer, nil*)}.
 In the right hand side of the multiple assignment, only the first
@@
 so the type of the right hand side is
-\texttt{(number, number, string, nil*)},
+\texttt{(integer, integer, string, nil*)},
 and the type assigned to \texttt{x}, \texttt{y}, and \texttt{z}
-are respectively \texttt{number}, \texttt{number}, and \texttt{string}.
+are respectively \texttt{integer}, \texttt{integer}, and \texttt{string}.
 The same tuple is assigned to the call to \texttt{sum} and it
@@
 
-In the strict mode, the type of \texttt{multiple} is \texttt{() -> (number, string)},
-and the type of \texttt{sum} is \texttt{(number, number) -> (number)}.
+In the strict mode, the type of \texttt{multiple} is \texttt{() -> (integer, string)},
+and the type of \texttt{sum} is \texttt{(integer, integer) -> (integer)}.
 Like in the default mode, only the first value produced by the
@@
 and the type of the right hand side is also
-\texttt{(number, number, string, nil*)},
+\texttt{(integer, integer, string, nil*)},
 because the strict mode affects only the type of function declarations
@@
 Unlike in the default mode, the type of the call to \texttt{sum} is
-\texttt{(number, number, string)}, and does not include the
+\texttt{(integer, integer, string)}, and does not include the
 variadic tail to catch arity mismatch.
@@
 \begin{verbatim}
-    local function overload (s1:string, s2:string|number)
+    local function overload (s1:string, s2:string|integer)
       if type(s2) == "string" then
@@
       else
-        -- string.rep : (string, number, string) -> (string)
+        -- string.rep : (string, integer, string) -> (string)
         return string.rep(s1, s2)
@@
 \begin{verbatim}
-    local function idiv (dividend:number, divisor:number):
-          (number, number)|(nil, string)
+    local function idiv (dividend:integer, divisor:integer):
+          (integer, integer)|(nil, string)
       if divisor == 0 then
@@
 we can annotate the return type of \texttt{idiv} with
-\texttt{(number, number)?} to denote the same union.
+\texttt{(integer, integer)?} to denote the same union.
 The parentheses are always necessary in this case, because
@@
 In our example, Typed Lua assigns the union of tuples
-\texttt{(number, number, nil*)|(nil, string, nil*)} to a type variable $X$.
+\texttt{(integer, integer, nil*)|(nil, string, nil*)} to a type variable $X$.
 Thus, variable \texttt{q} gets the type $X_{1}$
-that is equivalent to \texttt{number|nil},
+that is equivalent to the type \texttt{integer|nil},
 and variable \texttt{r} gets the type $X_{2}$
-that is equivalent to \texttt{number|string}.
+that is equivalent to the type \texttt{integer|string}.
 
@@
 it discriminates against all tuples in the union.
-In our example, $X$ is \texttt{(number, number, nil*)|(nil, string, nil*)}
+In our example, $X$ is \texttt{(integer, integer, nil*)|(nil, string, nil*)}
 outside of the \texttt{if} statement,
-but \texttt{(number, number, nil*)} inside the \texttt{then} block,
+but \texttt{(integer, integer, nil*)} inside the \texttt{then} block,
 and \texttt{(nil, string, nil*)} inside the \texttt{else} block.
-Thus, variable \texttt{q} has type \texttt{number} and variable \texttt{r} also
-has type \texttt{number} inside the \texttt{then} block;
+Thus, variable \texttt{q} has type \texttt{integer} and variable \texttt{r} also
+has type \texttt{integer} inside the \texttt{then} block;
 and variable \texttt{q} has type \texttt{nil} while variable \texttt{r}
@@
 it is written \texttt{\{t\}},
-and maps to the table type \texttt{\{number:t|nil\}}.
+and maps to the table type \texttt{\{integer:t|nil\}}.
 This table type is just a syntactic sugar to the definition of a
-map that maps values of type \texttt{number} to values of type \texttt{t|nil}.
+map that maps values of type \texttt{integer} to values of type \texttt{t|nil}.
 The syntax \ref{syn:table3} defines table types for records with an
@@
 
-Next, we show one example of table type to type a map from strings to numbers:
-\begin{verbatim}
-    local t:{string:number} = { foo = 1 } 
-    local x:number = t.foo               -- compile-time error
-    local y:number? = t.bar              -- y gets nil
-    local z:number = t["bar"] or 0       -- z gets 0 
+Next, we show one example of table type to type a map from strings to integers:
+\begin{verbatim}
+    local t:{string:integer} = { foo = 1 } 
+    local x:integer = t.foo               -- compile-time error
+    local y:integer? = t.bar              -- y gets nil
+    local z:integer = t["bar"] or 0       -- z gets 0 
 \end{verbatim}
@@
 The second line of this example gives a type error because we are
-attempting to assign a value of type \texttt{number|nil} to a
-variable that accepts only values of type \texttt{number}.
+attempting to assign a value of type \texttt{integer|nil} to a
+variable that accepts only values of type \texttt{integer}.
 Although the field \texttt{bar} does not exist in \texttt{t}, the third
@@
 is \texttt{string|nil}.
-An inconvenience aspect of making the types of maps and arrays always
+An inconvenient aspect of making the types of maps and arrays always
 include the type \texttt{nil} is to overload the programmers,
@@
 \begin{verbatim}
-    local album:{1:string, 2:number, 3:string} =
+    local album:{1:string, 2:integer, 3:string} =
       { "Transformer", 1972, "Lou Reed" }
@@
     local interface Element
-      info:number
+      info:integer
       next:Element?
@@
 \begin{verbatim}
-    local function insert (e:Element?, v:number):Element 
+    local function insert (e:Element?, v:integer):Element 
       return { info = v, next = e }
@@
 \begin{verbatim}
-    local Shape = { x = 0, y = 0 }
+    local Shape = { x = 0.0, y = 0.0 }
     const function Shape:move (dx:number, dy:number)
@@
     end
-    local Shape = { x = 0, y = 0 }
+    local Shape = { x = 0.0, y = 0.0 }
     const function Shape:new (x:number, y:number)
@@
     local Circle = setmetatable({}, { __index = Shape })
-    Circle.radius = 0
+    Circle.radius = 0.0
     const function Circle:new (x:number, y:number, radius:value)
       local c = setmetatable(Shape:new(x, y), { __index = self })
-      c.radius = tonumber(radius) or 0
+      c.radius = tonumber(radius) or 0.0
       return c

### 2014-11-12 doc/thesis/introduction.tex
Some english revisions to the introduction

@@
 simplicity and flexibility, because the absence of static types means
-that programmers do not need to bother about abstracting types that
+that programmers do not need to bother with abstracting types that
 should be validated by a type checker.
@@
 
-This simplicity and flexibility allow programmers to write code that
-might require a quite complex type system to statically type check,
+This simplicity and flexibility allows programmers to write code that
+might require a complex type system to statically type check,
 though it may also hide bugs that will be caught only after deployment
@@
 beginning of a project, because their simplicity and flexibility
-allow quick development and make it easier to change code according to
+allows quick development and make it easier to change code according to
 changing requirements.
@@
 This migration usually involves different languages that have
-different syntax and semantics, which usually require a complete
+different syntaxes and semantics, which usually require a complete
 rewrite of existing programs instead of incremental evolution from
@@
 The aim of both approaches is to offer programmers the option
-to use type annotations where static typing is needed and to not use
+to use type annotations where static typing is needed and to avoid
 them where dynamic typing is sufficient.
@@
 run-time semantics.
-While optional type systems do not affect the run-time semantics,
+While optional type systems do not affect run-time semantics,
 gradual typing uses run-time checks to ensure that dynamically typed
@@
 
-Programmers and researchers sometimes use the term gradual typing
-to mean the incremental evolution of dynamically typed code to
+Programmers and researchers sometimes use the term \emph{gradual typing}
+to mean the incremental evolution of dynamically typed code into
 statically typed code.
@@
 together.
-We use the term gradual typing for referring to the work of
+We use the term \emph{gradual typing} to refer to the work of
 \citet{siek2006gradual}.
@@
 an optional type system for Lua that is rich enough to
-preserve some of the Lua idioms that programmers are already used to,
+preserve some of the Lua idioms that programmers are already familiar with,
 but that also includes new constructs that help programmers
@@
 (with proper lexical scoping) where the only data structure
-mechanism is the \emph{table}, an associative array that can
-represent arrays, records, maps, modules, objects, etc.
+mechanism is the \emph{table} --
+an associative array that can represent arrays, records, maps, modules, objects, etc.
 Tables also have syntactic sugar and metaprogramming support
@@
 among different data types.
-We believe that designing an optional type system for Lua may
-shed some light on optional type systems for scripting languages
-in general, as Lua is a small scripting language that shares
-some features with other scripting languages such as JavaScript.
 
 Lua prefers to provide mechanisms instead of fixed policies due
-to its primary use that has always been as an embedded language
-for configuration and extension of other applications.
+to its primary use as an embedded language for configuration and
+extension of other applications.
 This means that even features such as a module system and
-object orientation are a matter of convention instead of built
-into the language.
+object orientation are a matter of convention instead of default
+language constructs.
 The result is a fragmented ecosystem of libraries, and different
 ideas among Lua programmers on how they should use the language
-features, and how they should structure programs.
+features, or how they should structure programs.
 
@@
 an optional type system for Lua.
-For this reason, we are not relying just on the semantics of
+For this reason, we are not relying entirely on the semantics of
 the language to design our type system.
 We also run a mostly automated survey of Lua idioms used in a
-large corpus of Lua libraries, and it also helped us on the
-design of Typed Lua.
+large corpus of Lua libraries, which also has helped in the design of Typed Lua.
 
@@
 code to coexist and interact with dynamically typed code
-through optional type annotations, and it also adds standard
-constructions for structuring Lua code.
+through optional type annotations.
+In addition, it adds default constructs that programmers can use
+to better structure Lua programs.
 The Typed Lua compiler warns programmers about type errors,
@@
 Programmers can enjoy some of the benefits of static types
-even without converting existing Lua modules to Typed Lua:
+even without converting existing Lua modules to Typed Lua --
 they can export a statically typed interface to a dynamically typed module,
@@
 
-Typed Lua intended use is as an application language, and we view
-that policies for organizing a program in modules and writing
+Typed Lua's intended use is as an application language, and
+we believe that policies for organizing a program in modules and writing
 object-oriented programs should be part of the language and
-enforced by its optional type system.
+checked by its optional type system.
 An application language is a programming language that helps
 programmers develop applications from scratch until these
-applications evolve to complex systems rather than just scripts.
+applications develop into complex systems rather than just scripts.
 We shall see that Typed Lua introduces the refinement of
@@
 
-We also believe that Typed Lua help programmers give a more
+We also believe that Typed Lua helps programmers give a more
 formal documentation to already existing Lua code, as static types
@@
 We enforce this point by using Typed Lua to statically type
-the interface of the Lua standard libraries and other commonly used
-Lua libraries, so our compiler can check their usage by Typed Lua
-code.
+the interface of the Lua standard library and other commonly used
+Lua libraries, so our compiler can check Typed Lua code that uses
+these libraries.
 
@@
 expressions.
-Still, Typed Lua needs a small amount of type annotations due
-to the nature of its optional type system.
+Still, Typed Lua only requires a small amount of type annotations
+due to the nature of its optional type system.
 
@@
 
+Finally, we believe that designing an optional type system for Lua may
+shed some light on optional type systems for scripting languages
+in general, as Lua is a small scripting language that shares
+some features with other scripting languages such as JavaScript.
+
 In Chapter \ref{chap:review} we review the literature about static

### 2014-11-13 doc/thesis/system.tex
Updating typing rules
Minor updates to the typing rules sketch

@@
 \mathbf{return} \; el \; | \;
-a\\
+e(el)_{s} \; | \;
+e{:}n(el)_{s}\\
 e ::= & \;\; \mathbf{nil} \; | \;
@@
 {\it string} \; | \;
-{...} \; | \;
+{...}_{e} \; | \;
 f \; | \;
@@
 \# \; e \; | \;
-a \; | \;
-l \; | \;
+e(el)_{e} \; | \;
+e{:}n(el)_{e} \; | \;
+n \; | \;
+e_{1}[e_{2}] \; | \;
 {<}t{>} \; n\\
 l ::= & \;\; n \; | \;
-e_{1}[e_{2}]\\
+e_{1}[e_{2}] \; | \;
+n[e] \; {<}t{>} \; | \;
+n_{1}{:}n_{2} \; {<}t{>}\\
 el ::= & \;\; \mathbf{nothing} \; | \;
@@
 \vec{e}, me\\
-me ::= & \;\; a \; | \;
+me ::= & \;\; e(el) \; | \;
+e{:}n(el) \; | \;
 {...}\\
-a ::= & \;\; e_{1}(\vec{e_{2}}) \; | \;
-e_{1}{:}n(\vec{e_{2}})\\
 f ::= & \;\; \mathbf{fun} \; (){:}r \; s \; ; \; \mathbf{return} \; el \; | \;
@@
 \begin{verbatim}
-    rec f:(number) -> (number) =
-      fun (x:number)
+    rec f:(integer) -> (integer) =
+      fun (x:integer)
         if n == 0 then
@@
 \begin{verbatim}
-    local a:(number) -> (number) =
-      fun (x:number)
+    local a:(integer) -> (integer) =
+      fun (x:integer)
         if n < 0 then return -n else return n
-    in local d:(any, any) -> (number) =
+    in local d:(any, any) -> (integer) =
          fun (x:any, y:any)
@@
 \begin{verbatim}
-  local m:() -> (number, string) =
+  local m:() -> (integer, string) =
     fun () skip ; return 2, "foo"
-  in local s:(number, number) -> (number)
-       fun (x:number, y:number) skip ; return x + y
-     in local x:number, y:number, z:string = m(), m()
+  in local s:(integer, integer) -> (integer)
+       fun (x:integer, y:integer) skip ; return x + y
+     in local x:integer, y:integer, z:string = m(), m()
         in s(m(), m())
@@
 \begin{verbatim}
-    local t:{string:number|nil} = { ["foo"] = 1 } in
-      local x:number|nil = t["foo"] in x = t["bar"]
+    local t:{string:integer|nil} = { ["foo"] = 1 } in
+      local x:integer|nil = t["foo"] in x = t["bar"]
 \end{verbatim}
@@
       { "Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat" }
-    in local x:number|nil = d[1] in x = d[8]
+    in local x:string|nil = d[1] in x = d[8]
 \end{verbatim}
@@
 \begin{verbatim}
-    local s:{"x":number, "y":number} = { x = 0, y = 0 }
+    local s:{"x":number, "y":number} = { x = 0., y = 0.0 }
     in s["new"] <(self, number, number) -> (self)> =

### 2014-11-17 doc/thesis/introduction.tex
Adding text about the typing rules for arithmetic operations

@@
 beginning of a project, because their simplicity and flexibility
-allows quick development and make it easier to change code according to
+allows quick development and makes it easier to change code according to
 changing requirements.
@@
 This migration usually involves different languages that have
-different syntaxes and semantics, which usually require a complete
+different syntaxes and semantics, which usually requires a complete
 rewrite of existing programs instead of incremental evolution from
@@
 programmers develop applications from scratch until these
-applications develop into complex systems rather than just scripts.
+applications evolve into complex systems rather than just scripts.
 We shall see that Typed Lua introduces the refinement of
@@
 
-We also believe that Typed Lua helps programmers give a more
+We also believe that Typed Lua helps programmers give more
 formal documentation to already existing Lua code, as static types

### 2014-11-17 doc/thesis/system.tex
Adding text about the typing rules for arithmetic operations

--- 2014-10-21 doc/thesis/system.tex
@@
 
-In this section we will use a reduced core of Typed Lua to present
-its typing rules in an easier way.
-It is a reduced core of the Lua language with explicit type annotations,
-without syntactic sugar, with only if and while control flow statements,
-and making variable scope explicit.
+In this section we use a reduced core of Typed Lua to present the
+most interesting typing rules of our type system.
+It is also a reduced core of the Lua language with only if and while
+control flow statements, without syntactic sugar, with explicit
+type annotations, with explicit scope of local variables, and
+with explicit declaration of methods.
+We use this reduced core because it simplifies the presentation
+of our type system.
+Appendix \ref{app:rules} presents the full set of typing rules.
 
@@
 \vec{l} = el \; | \;
+m \; | \;
 \mathbf{while} \; e \; \mathbf{do} \; s \; | \;
@@
 {\it string} \; | \;
-{...} \; | \;
+{...}_{e} \; | \;
 f \; | \;
+\{ \} \; | \;
 \{ \; cl \; \} \\
@@
 \# \; e \; | \;
-a \; | \;
-l \; | \;
+e(el)_{e} \; | \;
+e{:}n(el)_{e} \; | \;
+n \; | \;
+e_{1}[e_{2}] \; | \;
 {<}t{>} \; n\\
 l ::= & \;\; n \; | \;
-e_{1}[e_{2}]\\
+e_{1}[e_{2}] \; | \;
+n[e] \; {<}t{>}\\
 el ::= & \;\; \mathbf{nothing} \; | \;
@@
 \vec{e}, me\\
-cl ::= & \;\; \mathbf{nothing} \; | \;
-\vec{c} \; | \;
+cl ::= & \;\; \vec{c} \; | \;
 \vec{e} \; | \;
@@
 \vec{e}, me\\
-me ::= & \;\; a \; | \;
-{...}\\
-a ::= & \;\; e_{1}(\vec{e_{2}}) \; | \;
-e_{1}{:}n(\vec{e_{2}})\\
-f ::= & \;\; \mathbf{fun} \; (){:}r \; s \; ; \; \mathbf{return} \; el \; | \;
-\mathbf{fun} \; ({...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
-& | \; \mathbf{fun} \; (\vec{n{:}t}){:}r \; s \; ; \; \mathbf{return} \; el \; | \;
-\mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
+me ::= & \;\; e(el)_{m} \; | \;
+e{:}n(el)_{m} \; | \;
+{...}_{m}\\
+m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (){:}r \; s \; ; \; \mathbf{return} \; el\\
+& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; ({...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
+& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; s \; ; \; \mathbf{return} \; el\\
+& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t},{...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
+f ::= & \;\; \mathbf{fun} \; (){:}r \; s \; ; \; \mathbf{return} \; el\\
+& | \; \mathbf{fun} \; ({...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
+& | \; \mathbf{fun} \; (\vec{n{:}t}){:}r \; s \; ; \; \mathbf{return} \; el\\
+& | \; \mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
 c ::= & \;\; [e_{1}] = e_{2}\\
@@
 $el$ are expression lists, $c$ are constructor fields, $cl$ are lists of
-constructor fields, $me$ are multiple expressions, $a$ are function
-applications, $f$ are function declarations, $n$ are variable names,
+constructor fields, $me$ are expressions with multiple results,
+$f$ are function declarations, $m$ are method declarations, $n$ are variable names,
 $t$ are first-level types, and $r$ are second-level types.
@@
 $n_{1}{:}t_{1}, ..., n_{n}{:}t_{n}$.
+
+Our reduced code also splits function application, method application, and
+the vararg expression (${...}$) into different categories.
+It uses $e(el)_{s}$ to denote function applications that return no value
+because they appear as statements,
+$e(el)_{e}$ to denote function applications that return only one value,
+and $e(el)_{m}$ to denote function applications that return multiple values.
+It uses the same categories for method applications, but only ${...}_{e}$
+and ${...}_{m}$, as the vararg expression cannot appear as a statement.
 
@@
 the other mappings that extisted before.
+
+Lua has operator overloading, and allows the programmers to redefine
+the behavior of some operations.
+For instance, programmers can use metatables to redefine the
+behavior of arithmetic operations.
+Even though Typed Lua does not support operator overloading yet,
+it includes typing rules that allow programmers to use the
+dynamic type when they are using overloaded operations.
+The following typing rules show how Typed Lua uses the dynamic type
+to handle the overloading of arithmetic operations:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-ARITH5}\\
+\dfrac{\env_{1} \vdash e_{1}:\Any, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t, \env_{3}}
+      {\env_{1} \vdash e_{1} + e_{2}:\Any, \env_{3}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{T-ARITH6}\\
+\dfrac{\env_{1} \vdash e_{1}:t, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:\Any, \env_{3}}
+      {\env_{1} \vdash e_{1} + e_{2}:\Any, \env_{3}}
+\end{array}
+\end{array}
+\]
+
+The rule \textsc{T-ARITH5} allows type checking the following
+example:
+\begin{center}
+\begin{tabular}{l}
+$\mathbf{local} \; x{:}\Any = 1 \; \mathbf{in} \; x = x + 1$
+\end{tabular}
+\end{center}
+
+This example is safe, but the following it is not:
+\begin{center}
+\begin{tabular}{l}
+$\mathbf{local} \; x{:}\Integer, \;y{:}\Any = 1 \; \mathbf{in} \; x = x + y$
+\end{tabular}
+\end{center}
+
+\textit{/* Expression list rules. Where should we describe the
+discarding/replacement of tuple elements? */}
+
+Even though Lua has multiple assignments, the rule \textsc{T-ASSIGNMENT}
+shows that tuples types and consistent-subtyping make straightforward
+to type check them.
+\[
+\begin{array}{c}
+\mylabel{T-ASSIGNMENT}\\
+\begin{array}{c}
+\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
+       r_{1} \lesssim r_{2}}
+      {\env_{1} \vdash \vec{l} = el:\env_{3}}
+\end{array}
+\end{array}
+\]
+
+Multiple assignments and multiple return values can also appear in
+function applications, as the following example shows:
+\begin{center}
+\begin{tabular}{llll}
+\multicolumn{4}{l}{$\mathbf{local} \; m:\Void \rightarrow \Integer \times \String =$} \\
+& \multicolumn{3}{l}{$\mathbf{fun} \; ():\Integer \times \String$} \\
+& & \multicolumn{2}{l}{$\mathbf{skip} \;;\; \mathbf{return} \; 2, ``foo"$} \\
+\multicolumn{4}{l}{$\mathbf{in}$} \\
+& \multicolumn{3}{l}{$\mathbf{local} \; s:\Integer \times \Integer \rightarrow \Integer =$} \\
+& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer, y:\Integer):\Integer$} \\
+& & & \multicolumn{1}{l}{$\mathbf{skip} \;;\; \mathbf{return} \; x + y$} \\
+& \multicolumn{3}{l}{$\mathbf{in} \; s(m()_{e}, m()_{m})_{s}$}
+\end{tabular}
+\end{center}
+
+The $\mathbf{or}$ idiom.
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; x:\String \cup \Nil = \mathbf{nothing} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$\mathbf{local} \; y:\String = x \; \mathbf{or} \; ``Hello"$}
+\end{tabular}
+\end{center}
+
+Overloading on the input parameter.
+\begin{center}
+\begin{tabular}{llll}
+\multicolumn{4}{l}{$\mathbf{local} \; o:\String \times \String \cup \Integer \rightarrow \String =$}\\
+& \multicolumn{3}{l}{$\mathbf{fun} \; (a:\String, b:\String \cup \Integer):\String$}\\
+& & \multicolumn{2}{l}{$\mathbf{if} \; type(b) == ``string" \; \mathbf{then}$}\\
+& & & \multicolumn{1}{l}{$\mathbf{return} \; a \; {..} \; b$}\\
+& & \multicolumn{2}{l}{$\mathbf{else}$}\\
+& & & \multicolumn{1}{l}{$\mathbf{return} \; rep(a, b)_{e}$}\\
+\multicolumn{4}{l}{$\mathbf{in} \; o(``foo", 2)_{s}$}
+\end{tabular}
+\end{center}
+
+We are assuming that the function \emph{rep} is in the environment and has type
+$\String \times \Integer \times \String \cup \Nil \rightarrow \String$.
+
+Overloading on the return type.
+\begin{center}
+\begin{tabular}{lll}
+\multicolumn{3}{l}{$\mathbf{local} \; q:\pi_{1}, r:\pi_{2} = idiv(1, 2) \; \mathbf{in}$}\\
+& \multicolumn{2}{l}{$\mathbf{if} \; type(r) == ``string" \; \mathbf{then}$}\\
+& & \multicolumn{1}{l}{$print(``ERROR: " \; .. \; r)_{s}$}\\
+& \multicolumn{2}{l}{$\mathbf{else}$}\\
+& & \multicolumn{1}{l}{$print(q + r)_{s}$}
+\end{tabular}
+\end{center}
+
+We are assuming that functions \emph{idiv} and \emph{print} are in
+the environment with the respective types
+$\Integer \times \Integer \rightarrow \Integer \times \Integer \sqcup \Nil \times \String$
+and
+$\Value{*} \rightarrow \Void$.
+
+Table type as hash.
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; t:\{\String:\Integer \cup \Nil\}_{closed} = \{ [``foo"] = 1 \} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$\mathbf{local} \; x:\Integer \cup \Nil = t[``foo"] \; \mathbf{in} \; x = t[``bar"]$}
+\end{tabular}
+\end{center}
+
+Table type as array.
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; t:\{\Integer:\String \cup \Nil\}_{closed} = \{ ``a", ``b", ``c" \} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$\mathbf{local} \; x:\String \cup \Nil = t[1] \; \mathbf{in} \; x = t[4]$}
+\end{tabular}
+\end{center}
+
+Table type as record (first example).
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; t:\{``f":\String, ``l":\String\}_{closed} =$}\\
+& \multicolumn{1}{l}{$\{ [``l"] = ``Reed", [``f"] = ``Lou" \}$}\\
+\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
+\end{tabular}
+\end{center}
+
+Table type as record (second example).
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; t:\{``f":\String, ``m":\String \cup \Nil, ``l":\String\}_{closed} =$}\\
+& \multicolumn{1}{l}{$\{ [``f"] = ``Lewis", [``m"] = ``Allan", [``l"] = ``Reed" \}$}\\
+\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
+\end{tabular}
+\end{center}
+
+Table type as record (third example).
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; t:\{``f":\String, ``m":\String \cup \Nil, ``l":\String\}_{closed} =$}\\
+& \multicolumn{1}{l}{$\{ [``f"] = ``Lou", [``l"] = ``Reed" \}$}\\
+\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
+\end{tabular}
+\end{center}
+
+Refinement of table types (first example).
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; t:\{\}_{open} = \{\} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$t[``f"] \; {<}\String{>} = ``Lou";$}\\
+& \multicolumn{1}{l}{$t[``l"] \; {<}\String{>} = ``Reed";$}\\
+\end{tabular}
+\end{center}
+
+Refinement of table types (second example).
+\begin{center}
+\begin{tabular}{lll}
+\multicolumn{3}{l}{$\mathbf{local} \; t:\{\}_{open} = \{\} \; \mathbf{in}$}\\
+& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = p \; \mathbf{in}$}\\
+& & \multicolumn{1}{l}{$t[``f"] \; {<}\String{>} = ``Lou";$}\\
+& & \multicolumn{1}{l}{$t[``l"] \; {<}\String{>} = ``Reed";$}\\
+\end{tabular}
+\end{center}
+
+Objects.
+\begin{center}
+\begin{tabular}{llll}
+\multicolumn{4}{l}{$\mathbf{local} \; s:\{``x:"\Number, ``y":\Number\}_{open} = \{ [``x"] = 0.0, [``y"] = 0.0 \}$}\\
+\multicolumn{4}{l}{$\mathbf{in}$}\\
+& \multicolumn{3}{l}{$\mathbf{fun} \; s{:}new (x:number, y:number):\Self$}\\
+& & \multicolumn{2}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{closed} =$}\\
+& & & \multicolumn{1}{l}{$setmetatable(\{\}, \{ [``\string_\string_index"] = self \}$}\\
+& & \multicolumn{2}{l}{$\mathbf{in} \; s[``x"] = x; \; s[``y"] = y; \; \mathbf{return} \; s$}\\
+; & \multicolumn{3}{l}{$\mathbf{fun} \; s{:}move (x:number, y:number):\Void$}\\
+& & \multicolumn{2}{l}{$self[``x"] = self[``x"] + x;$}\\
+& & \multicolumn{2}{l}{$self[``y"] = self[``y"] + y$}
+\end{tabular}
+\end{center}

### 2014-11-18 doc/thesis/system.tex
Adding text about the typing rules for the or idiom

@@
 
-The $\mathbf{or}$ idiom.
+Typed Lua includes four typing rules to handle the $\mathbf{or}$
+logical operator and its common idioms:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-OR1}\\
+\dfrac{\env_{1} \vdash e_{1}:t, \env_{2} \;\;\;
+       \Nil \not\lesssim t \;\;\;
+       \False \not\lesssim t}
+      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-OR2}\\
+\dfrac{\env_{1} \vdash e_{1}:\Nil, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t, \env_{3}}
+      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{T-OR3}\\
+\dfrac{\env_{1} \vdash e_{1}:\False, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t, \env_{3}}
+      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-OR4}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3}}
+      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(filter(t_{1}, \Nil), \False) \cup t_{2}, \env_{3}}
+\end{array}
+\end{array}
+\]
+
+The rule \textsc{T-OR1} is the rule that implements the short circuit.
+We use the consistent-subtyping in this relation to guarantee that
+the type system avaliates the second expression when the first one
+has the dynamic type, as it can be hiding a false value.
+
+The rules \textsc{T-OR2} and \textsc{T-OR3} guarantee that the final
+result is the type of the second expression, because the first one
+is certainly a false value.
+
+The rule \textsc{T-OR4} is the most general rule, but it is also
+the rule that handles the common $\mathbf{or}$ idioms.
+It uses the predicate \emph{filter} to filter possible false values
+that might be part of the type of the first expression.
+We can use pattern matching to the define the recursive predicate
+\emph{filter} as follows:
+\begin{align*}
+fu(t_{1} \cup t_{2}, t_{3}) & = fu(t_{1}, t_{3}) \cup fu(t_{2}, t_{3})\\
+fu(t_{1} \cup t_{2}, t_{1}) & = fu(t_{2}, t_{1})\\
+fu(t_{1} \cup t_{2}, t_{2}) & = fu(t_{1}, t_{2})\\
+fu(t_{1}, t_{2}) & = t_{1}
+\end{align*}
+
+Using these typing rules our type system can type check the following
+example:
 \begin{center}
@@
 
-Overloading on the input parameter.
+Without the \emph{filter} predicate,
+the expression $x \; \mathbf{or} \; ``Hello"$ would have type
+$\String \cup \Nil \cup ``Hello"$.
+The \emph{filter} predicate removes the type $\Nil$ from the result,
+leaving type $\String \cup ``Hello"$.
+At the end of the evaluation, the expression has type $\String$
+because unions are disjoint and $``Hello" \subtype \String$.
+
+Another common idiom that programmers use in Lua is to overload
+the input parameter of functions, and use the function \texttt{type}
+to execute different actions according to their types.
+\[
+\begin{array}{c}
+\mylabel{T-IF2}\\
+\dfrac{\begin{array}{c}
+       closeall(\env_{1}[n \mapsto \String]) \vdash s_{1}:\env_{2} \\
+       closeall(\env_{1}[n \mapsto filter(\env_{1}(n), \String)) \vdash s_{2}:\env_{3}
+      \end{array}}
+      {\env_{1} \vdash \mathbf{if} \; type(n) == ``string" \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}:closeset(\env_{2}, fav(s_{1}) \cup fav(s_{2}))}
+
+\end{array}
+\]
+
+Using the rule \textsc{T-IF2}, Typed Lua can type check the following example:
 \begin{center}
@@
 $\String \times \Integer \times \String \cup \Nil \rightarrow \String$.
+
+Typed Lua also includes similar rules to handle the tags \texttt{nil},
+\texttt{boolean}, and \texttt{number}.
 

### 2014-11-19 doc/thesis/evaluation.tex
Updating the evaluation data to include the inter category
Adding raw text about related work

@@
 \hline
-Lua Standard Library & 87 & 6 & 0 & 36 & 129 & 67\% \\
+Lua Standard Library & 80 & 6 & 9 & 34 & 129 & 62\% \\
 \hline
@@
 \hline
-LuaSocket & 109 & 1 & 0 & 13 & 123 & 89\% \\
+LuaSocket & 109 & 1 & 2 & 11 & 123 & 89\% \\
 \hline
-HTTP Digest & 0 & 0 & 0 & 1 & 1 & 0\% \\
+HTTP Digest & 0 & 0 & 1 & 0 & 1 & 0\% \\
 \hline
@@
 \hline
-Typed Lua Compiler & 144 & 0 & 0 & 10 & 154 & 94\% \\
+Typed Lua Compiler & 144 & 0 & 1 & 9 & 154 & 94\% \\
 \hline
@@
 \hline
-base & 9 & 1 & 0 & 16 & 26 & 35\% \\
+base & 9 & 1 & 3 & 13 & 26 & 35\% \\
 \hline
@@
 \hline
-string & 14 & 0 & 0 & 2 & 16 & 88\% \\
+string & 12 & 0 & 0 & 4 & 16 & 75\% \\
 \hline
@@
 \hline
-math & 26 & 0 & 0 & 1 & 27 & 96\% \\
+math & 22 & 0 & 4 & 1 & 27 & 81\% \\
 \hline
@@
 \hline
-os & 10 & 0 & 0 & 1 & 11 & 91\% \\
+os & 9 & 0 & 2 & 0 & 11 & 82\% \\
 \hline
@@
 \hline
-ftp & 5 & 0 & 0 & 1 & 6 & 83\% \\
+ftp & 5 & 0 & 1 & 0 & 6 & 83\% \\
 \hline
-http & 4 & 0 & 0 & 1 & 5 & 80\% \\
+http & 4 & 0 & 1 & 0 & 5 & 80\% \\
 \hline
@@
 \hline
-http-digest & 0 & 0 & 0 & 1 & 1 & 0\% \\
+http-digest & 0 & 0 & 1 & 0 & 1 & 0\% \\
 \hline
@@
 \hline
-tlast & 46 & 0 & 0 & 1 & 47 & 98\% \\
+tlast & 46 & 0 & 1 & 0 & 47 & 98\% \\
 \hline
@@
 
-\begin{itemize}
-\item \citep{tidallock}
-\item \citep{bonnaire-sergeant2012typed-clojure}
-\item \citep{vitousek2014deg}
-\item \citep{allende2013gts}
-\item \citep{tobin-hochstadt2008ts} 
-\item \citep{dart}
-\item \citep{typescript}
-\item \citep{bierman2014typescript}
-\item \citep{politz2012semantics}
-\end{itemize}
+Tidal Lock \citep{tidallock} is a prototype of an optional type system for Lua.
+It also includes the refinement of table types.
+In fact, Tidal Lock inspired Typed Lua to also include this feature.
+However, Tidal Lock covers a little subset of Lua.
+Statements include declaration of local variables, multiple assignments,
+function application, and the return statement, that is, they do not
+include any control-flow statements.
+Expressions include primitive literals, table indexing, function application,
+function declaration, and the table constructor, but they do not include
+binary operations.
+
+The table type from Tidal Lock can only represent records, that is,
+it cannot describe hash tables and arrays yet, though we can refine them.
+Tidal Lock also includes field types to describe the type of the fields
+of a table type.
+The field types describe whether a table field is mutable or imutable
+in a table type.
+Field types are the feature that allow the refinement of table types in
+Tidal Lock.
+
+Tidal Lock is similar to Typed Lua because it is also a structural type
+system based on subtyping.
+It also uses local type inference to deduce the type of unannotated expressions,
+and have a dynamic type.
+However, Tidal Lock does not support union types, recursive types, and variadic types.
+It also does not support any object-oriented idiom.
+
+Typed Racket \citep{tobin-hochstadt2008ts} is a statically typed version
+of the Racket language, which is a Scheme dialect.
+The main purpose of Typed Racket is to allow programmers to combine
+untyped modules, which are written in Racket, with typed modules, which are
+written in Typed Racket.
+It also uses local type inference to deduce the type of unannotated expressions.
+
+The type system of Typed Racket introduced \emph{occurrence typing}
+\citep{tobin-hochstadt2010ltu}.
+It is a novel way to use type predicates in control flow statements
+to refine union types.
+The type system is rich and includes basic types, function types,
+union types, recursive types, and structure types.
+Function types can return multiple values.
+Structure types are similar to our interfaces, as they describe
+record types.
+The type system is also structural and based on subtyping.
+It also includes the dynamic type \texttt{Any}, which is the top
+type in the system.
+Typed Racket also supports polymorphic functions and data structures.
+
+Typed Clojure \citep{bonnaire-sergeant2012typed-clojure} is an
+optional type system for Clojure.
+Although Clojure is a Lisp dialect that runs on the Java Virtual Machine,
+Common Language Runtime, and JavaScript, Typed Clojure runs only on
+the Java Virtual Machine.
+Perhaps, this restriction pushed Typed Clojure to support Java classes
+and some Java types such as \texttt{Long}, \texttt{Double}, and \texttt{String}.
+Typed Clojure also provides optional type annotations and uses
+local type inference to deduce the type of unannotated expressions.
+It also assigns the type \texttt{Any} to unannotated function parameters,
+which is the top type in the type system.
+
+The type system of Typed Clojure is rich.
+It includes polymorphic function types, union types, intersection types,
+lists, vectors, maps, sets, and recursive types.
+Function types can also have rest parameters, which are similar
+to our variadic types, but can only appear on the input parameter
+of function types.
+In fact, its function types cannot return multiple results.
+It also uses ocurrence typing to allow control flow statements to
+refine union types.
+The type system is also structural and based on subtyping.
+
+Dart \citep{dart} is a new class-based object-oriented programming
+language.
+It includes optional type annotations and compiles to JavaScript.
+The type system of Dart is nomial and includes base types,
+function types, lists, and maps.
+It also supports generics, and the programmer can define
+generic functions, lists, and maps.
+
+Even though Dart has optional typing and static types by
+default do not affect run-time semantics, it has an
+execution mode that affects run-time.
+The \emph{checked mode} inserts run-time assertions that
+verifies whether static types match run-time tags.
+The \emph{production mode} is the default execution mode
+that does not include any assertions.
+
+TypeScript \citep{typescript} is a JavaScript extension
+that includes optional type annotations and class-based
+object-oriented programming.
+It also uses local type inference to deduce the type
+of unannotated expressions.
+The type system of TypeScript is structural and based
+on subtyping.
+It includes the dynamic type, primitive types, function
+types, array types, tuple types, and recursive types.
+TypeScript also support generics.
+
+Even though TypeScript is unsound by design,
+\citet{bierman2014typescript} shows how to make TypeScript sound.
+They use a reduced core of TypeScript to formalize a
+sound type system for TypeScript, but also to formalize
+its current unsound type system.
+
+\citet{politz2012semantics} proposes semantics and types for objects
+with first-class member names, a well-known feature from scripting languages.
+Their type system uses string patterns to describe the members of
+an object, and define a complex subtyping relation to validate
+these patterns.
+They also provide an implementation of their system to JavaScript.
+
+Gradualtalk \citep{allende2013gts} is a Smalltalk dialect that
+supports gradual typing.
+The type system combines nominal and structural typing.
+It includes function types, union types, structural types,
+nominal types, a self type, and parametric polymorphism.
+The type system also relies on subtyping and consistent-subtyping.
+
+Gradualtalk inserts run-time checks that ensure dynamically
+typed code does not violate statically typed code.
+\citet{allende2013cis} perform a carefull evaluation about
+cast insertion on Gradualtalk.
+They report that usually cast insertions impact on execution
+performance.
+
+Reticulated Python \citep{vitousek2014deg} is a Python compiler
+that supports gradual typing.
+The type system is structural and based on subtyping.
+It includes base types, the dynamic type, list types,
+dictionary types, tuple types, function types, set types,
+object types, class types, and recursive types.
+It includes class and object types to differentiate the
+type of class declarations and instances.
+It also uses local type inference.
+Besides static type checking, Reticulated Python also introduces
+three different approaches for inserting run-time assertions.
+
+Hack \citep{hack} is a new programming language that combines
+static and dynamic type in PHP code.
+Its type system includes generics, nullable types, collections,
+and function types.
+
+The Ruby Type Checker \citep{ren2013rtc} is a library that
+performs type checking during run-time.
+The library provides type annotations that the programmer
+can use on classes and methods.
+Its type system includes nominal types, union types,
+intersection types, method types, parametric polymorphism,
+and type casts.

### 2014-11-20 doc/thesis/system.tex
Fixing #14
Starting to implement the typing rules in Haskell

@@
 
+\textit{/* The following example does not type check */}
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; x:\Nil \cup \False = \mathbf{false} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$\mathbf{local} \; y:\String = x \; \mathbf{or} \; ``Hello"$}
+\end{tabular}
+\end{center}
+
+\textit{/* Should we define the following rules? */}
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-OR4}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+       \Nil \cup \False \subtype t_{1} \;\;\;
+       t_{1} \subtype \Nil \cup \False}
+      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t_{2}, \env_{3}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-OR5}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+       \Nil \subtype t_{1} \;\;\;
+       \False \subtype t_{1}}
+      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(filter(t_{1}, \Nil), \False) \cup t_{2}, \env_{3}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-OR6}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+       \Nil \subtype t_{1} \;\;\;
+       \False \not\subtype t_{1}}
+      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(t_{1}, \Nil) \cup t_{2}, \env_{3}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-OR7}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+       \Nil \not\subtype t_{1} \;\;\;
+       \False \subtype t_{1}}
+      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(t_{1}, \False) \cup t_{2}, \env_{3}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-OR8}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+       \Nil \not\subtype t_{1} \;\;\;
+       \False \not\subtype t_{1}}
+      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t_{1} \cup t_{2}, \env_{3}}
+\end{array}
+\end{array}
+\]
+
+\textit{/* Or who writes this kind of code deserves a type error? */}
+
 Another common idiom that programmers use in Lua is to overload
@@
 & \multicolumn{3}{l}{$\mathbf{fun} \; (a:\String, b:\String \cup \Integer):\String$}\\
-& & \multicolumn{2}{l}{$\mathbf{if} \; type(b) == ``string" \; \mathbf{then}$}\\
-& & & \multicolumn{1}{l}{$\mathbf{return} \; a \; {..} \; b$}\\
-& & \multicolumn{2}{l}{$\mathbf{else}$}\\
-& & & \multicolumn{1}{l}{$\mathbf{return} \; rep(a, b)_{e}$}\\
+& & \multicolumn{2}{l}{$\mathbf{local} \; r:\String = ``" \; \mathbf{in}$}\\
+& & & \multicolumn{1}{l}{$\mathbf{if} \; type(b) == ``string" \; \mathbf{then} \; r = a \;{..}\;b \; \mathbf{else} \; r = rep(a, b)_{e}\; ;$}\\
+& & & \multicolumn{1}{l}{$\mathbf{return} \; r$}\\
 \multicolumn{4}{l}{$\mathbf{in} \; o(``foo", 2)_{s}$}
@@
 Overloading on the return type.
-\begin{center}
-\begin{tabular}{lll}
-\multicolumn{3}{l}{$\mathbf{local} \; q:\pi_{1}, r:\pi_{2} = idiv(1, 2) \; \mathbf{in}$}\\
-& \multicolumn{2}{l}{$\mathbf{if} \; type(r) == ``string" \; \mathbf{then}$}\\
-& & \multicolumn{1}{l}{$print(``ERROR: " \; .. \; r)_{s}$}\\
-& \multicolumn{2}{l}{$\mathbf{else}$}\\
-& & \multicolumn{1}{l}{$print(q + r)_{s}$}
-\end{tabular}
-\end{center}
+\[
+\begin{array}{c}
+\mylabel{T-IF3}\\
+\dfrac{\begin{array}{c}
+       \env_{1}(n) = \pi_{i}\\
+       closeall(\env_{1}[\pi \mapsto fpt(\env_{1}(\pi), \Nil, i)]) \vdash s_{1}:\env_{2} \\
+       closeall(\env_{1}[\pi \mapsto gpt(\env_{1}(\pi), \Nil, i)) \vdash s_{2}:\env_{3}
+      \end{array}}
+      {\env_{1} \vdash \mathbf{if} \; n \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}:closeset(\env_{2}, fav(s_{1}) \cup fav(s_{2}))}
+
+\end{array}
+\]
+
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; q:\pi_{1}, r:\pi_{2} = idiv(1, 2)_{m} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; print(q + r)_{s} \; \mathbf{else} \; print(``ERROR: " \; .. \; r)_{s}$}
+\end{tabular}
+\end{center}
+
+\textit{/* How should we write the typing rule of the local declaration? */}
+
+\textit{/* How this is valid if projection types do not have subtyping rules? */}
 

### 2014-11-20 doc/thesis/haskell/tl.hs

@@ -1,0 +1,117 @@
+
+data TType = TValue
+  | TNil
+  | TBool Bool
+  | TInt Int
+  | TFloat Float
+  | TStr String
+  | TBoolean
+  | TInteger
+  | TNumber
+  | TString
+  | TUnion TType TType
+  deriving (Eq, Show)
+
+data SType = TType TType
+  | TVararg TType
+  | TTuple TType SType
+  deriving (Show)
+
+type Id = (String,TType)
+
+data Stm = SSkip
+  | SSeq Stm Stm
+  | SLocal (Id,[Id]) [Exp] Stm
+  | SAss1 LHS Exp
+  | SAss (LHS,[LHS]) (Exp,[Exp])
+  deriving (Show)
+
+data Exp = ENil
+  | EInt Int
+  | EFloat Float
+  | EStr String
+  | EVar LHS
+  deriving (Show)
+
+data LHS = Var String deriving (Show)
+
+type Env = String -> TType
+
+emptyEnv :: Env
+emptyEnv v = TNil
+
+update :: String -> TType -> Env -> Env
+update v t g = \v' -> if v == v' then t else g v'
+
+showEnv :: Env -> TType
+showEnv g = g ""
+
+subtype1 :: TType -> TType -> Bool
+subtype1 (TBool _) TBoolean = True
+subtype1 (TInt _) TInteger = True
+subtype1 (TInt _) TNumber = True
+subtype1 (TFloat _) TNumber = True
+subtype1 (TStr _) TString = True
+subtype1 TInteger TNumber = True
+subtype1 (TUnion t1 t2) t = (subtype1 t1 t) && (subtype1 t2 t)
+subtype1 t (TUnion t1 t2) = (subtype1 t t1) || (subtype1 t t2)
+subtype1 t TValue = True
+subtype1 t1 t2 = t1 == t2
+
+subtype2 :: SType -> SType -> Bool
+subtype2 (TType t1) (TType t2) = subtype1 t1 t2
+subtype2 (TTuple t1 s1) (TTuple t2 s2) = (subtype1 t1 t2) && (subtype2 s1 s2)
+subtype2 (TVararg t1) (TVararg t2) = subtype1 (TUnion t1 TNil) (TUnion t2 TNil)
+subtype2 (TType t1) (TVararg t2) = subtype1 t1 (TUnion t2 TNil)
[…]

### 2014-11-21 doc/thesis/haskell/tl.hs
Adding function types to the Haskell implementation

@@ -1,7 +1,8 @@
 
 data TType = TValue
   | TNil
-  | TBool Bool
+  | TFalse
+  | TTrue
   | TInt Int
   | TFloat Float
   | TStr String
@@ -10,30 +11,57 @@ data TType = TValue
   | TNumber
   | TString
   | TUnion TType TType
+  | TFunction SType SType
   deriving (Eq, Show)
 
-data SType = TType TType
+data SType = TVoid
+  | TType TType
   | TVararg TType
   | TTuple TType SType
-  deriving (Show)
+  deriving (Eq, Show)
 
 type Id = (String,TType)
 
+type Vararg = (Bool,TType)
+
 data Stm = SSkip
   | SSeq Stm Stm
-  | SLocal (Id,[Id]) [Exp] Stm
-  | SAss1 LHS Exp
-  | SAss (LHS,[LHS]) (Exp,[Exp])
+--  | SLocal (Id,[Id]) [Exp] Stm
+  | SLocal (Id,[Id]) ExpList0 Stm
+--  | SAss (LHS,[LHS]) (Exp,[Exp])
+  | SAss (LHS,[LHS]) ExpList1
+--  | SRet [Exp]
+  | SRet ExpList0
+--  | SApp Exp [Exp]
+  | SApp Exp ExpList0
   deriving (Show)
 
 data Exp = ENil
+  | EFalse
+  | ETrue
   | EInt Int
   | EFloat Float
   | EStr String
+  | EVararg
+  | EFun [Id] Vararg SType Stm
+  | EApp Exp ExpList0
   | EVar LHS
   deriving (Show)
 
+data MExp = MVararg
+  | MApp Exp ExpList0
+  deriving (Show)
+
 data LHS = Var String deriving (Show)
+
+data ExpList0 = ZNothing
+  | OEL ExpList1
+  deriving (Show)
+
+data ExpList1 = OExp Exp
+  | OMExp MExp
+  | LExp Exp ExpList1
+  deriving (Show)
 
 type Env = String -> TType
 
@@ -43,11 +71,26 @@ emptyEnv v = TNil
 update :: String -> TType -> Env -> Env
 update v t g = \v' -> if v == v' then t else g v'
 
+initEnv :: Env
+initEnv = update "..." TString emptyEnv
+
+updateEnv :: [Id] -> Env -> Env
+updateEnv [] g = g
+updateEnv ((n,t):l) g = updateEnv l (update n t g)
+
 showEnv :: Env -> TType
[…]

### 2014-11-23 doc/thesis/system.tex
Starting to write about the refinement of table types

@@
 s_{1} \; ; \; s_{2} \; | \;
-\vec{l} = el \; | \;
+ll = el_{1} \; | \;
 m \; | \;
@@
 \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
-& | \; \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s \; | \;
+& | \; \mathbf{local} \; \vec{n{:}t} = el_{0} \; \mathbf{in} \; s \; | \;
 \mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
-\mathbf{return} \; el \; | \;
-e(el)_{s} \; | \;
+\mathbf{return} \; el\\
+& | \; e(el)_{s} \; | \;
 e{:}n(el)_{s}\\
 e ::= & \;\; \mathbf{nil} \; | \;
-\mathbf{false} \; | \;
-\mathbf{true} \; | \;
-{\it int} \; | \;
-{\it float} \; | \;
-{\it string} \; | \;
+x \; | \;
 {...}_{e} \; | \;
+n \; | \;
+e_{1}[e_{2}] \; | \;
+{<}t{>} \; n \; | \;
 f \; | \;
@@
 e(el)_{e} \; | \;
-e{:}n(el)_{e} \; | \;
-n \; | \;
-e_{1}[e_{2}] \; | \;
-{<}t{>} \; n\\
+e{:}n(el)_{e}\\
+x ::= & \;\; \mathbf{false} \; | \;
+\mathbf{true} \; | \;
+{\it int} \; | \;
+{\it float} \; | \;
+{\it string}\\
 l ::= & \;\; n \; | \;
@@
 n[e] \; {<}t{>}\\
-el ::= & \;\; \mathbf{nothing} \; | \;
-\vec{e} \; | \;
+ll ::= & \;\; \vec{l}\\
+el_{0} ::= & \;\; \mathbf{nothing} \; | \;
+el_{1}\\
+el_{1} ::= & \;\; \vec{e} \; | \;
 me \; | \;
 \vec{e}, me\\
+el ::= & \;\; el_{0}\\
 cl ::= & \;\; \vec{c} \; | \;
-\vec{e} \; | \;
-me \; | \;
-\vec{c}, \vec{e} \; | \;
-\vec{c}, me \; | \;
-\vec{c}, \vec{e}, me \; | \;
-\vec{e}, me\\
+{...}_{m} \; | \;
+\vec{c}, {...}_{m}\\
 me ::= & \;\; e(el)_{m} \; | \;
@@
 
+The most unusual feature of our type system is the refinement of
+table types, which is given by the following rule:
+\[
+\begin{array}{c}
+\mylabel{T-REFINE}\\
+\dfrac{\env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique} \;\;\;
+       \env_{1} \vdash x:t_{1}, \env_{2} \;\;\;
+       \not \exists i \in 1..n \; t_{1} \lesssim k_{i}}
+      {\env_{1} \vdash n[x] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, t_{1}{:}t\}_{open|unique}]}
+\end{array}
+\]
+
+The rule \textsc{T-REFINE} allow us to add new keys to \emph{open}
+or \emph{unique} table types.
+We restricted the keys to literals only because the purpose of
+this rule is to make it easier to add new fields to table types
+that represent records.
+We also do not allow changing the type of a field that is already
+in the table because this could lead to loosen table types that
+have no static type checking. 
+For instance, Typed Lua allows us to write the following code:
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; t:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$t[``x"] \; {<}\String{>} = ``foo";$}\\
+& \multicolumn{1}{l}{$t[``y"] \; {<}\Integer{>} = 1$}\\
+\end{tabular}
+\end{center}
+but it does not allow us to write the following code:
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; t:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$t[``x"] \; {<}\String{>} = ``foo";$}\\
+& \multicolumn{1}{l}{$t[``x"] \; {<}\Integer{>} = 1$}\\
+\end{tabular}
+\end{center}
+
 Lua has operator overloading, and allows the programmers to redefine
@@
 & \multicolumn{3}{l}{$\mathbf{fun} \; ():\Integer \times \String$} \\
-& & \multicolumn{2}{l}{$\mathbf{skip} \;;\; \mathbf{return} \; 2, ``foo"$} \\
+& & \multicolumn{2}{l}{$\mathbf{return} \; 2, ``foo"$} \\
 \multicolumn{4}{l}{$\mathbf{in}$} \\
@@
 & & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer, y:\Integer):\Integer$} \\
-& & & \multicolumn{1}{l}{$\mathbf{skip} \;;\; \mathbf{return} \; x + y$} \\
+& & & \multicolumn{1}{l}{$\mathbf{return} \; x + y$} \\
 & \multicolumn{3}{l}{$\mathbf{in} \; s(m()_{e}, m()_{m})_{s}$}
@@
 
-Refinement of table types (first example).
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; t:\{\}_{open} = \{\} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$t[``f"] \; {<}\String{>} = ``Lou";$}\\
-& \multicolumn{1}{l}{$t[``l"] \; {<}\String{>} = ``Reed";$}\\
-\end{tabular}
-\end{center}
-
 Refinement of table types (second example).

### 2014-11-23 doc/thesis/haskell/tl.hs

@@ -26,13 +26,9 @@ type Vararg = (Bool,TType)
 
 data Stm = SSkip
   | SSeq Stm Stm
---  | SLocal (Id,[Id]) [Exp] Stm
   | SLocal (Id,[Id]) ExpList0 Stm
---  | SAss (LHS,[LHS]) (Exp,[Exp])
   | SAss (LHS,[LHS]) ExpList1
---  | SRet [Exp]
   | SRet ExpList0
---  | SApp Exp [Exp]
   | SApp Exp ExpList0
   deriving (Show)
 
@@ -138,27 +134,13 @@ checkStm SSkip g = TVoid
 checkStm (SSeq s1 s2) g = case checkStm s1 g of
   TVoid -> checkStm s2 g
   _ -> error("seq: ")
---checkStm (SLocal (id,idl) el s) g = if subtype2 (checkExpList el g) (idlist2tuple (id:idl))
---                                    then checkStm s (updateEnv (id:idl) g)
---                                    else error("local: " ++ (show (checkExpList el g)) ++ " </: " ++ (show (idlist2tuple (id:idl))))
 checkStm (SLocal (id,idl) el s) g = if subtype2 (checkExpList0 el g) (idlist2tuple (id:idl))
                                     then checkStm s (updateEnv (id:idl) g)
                                     else error("local: " ++ (show (checkExpList0 el g)) ++ " </: " ++ (show (idlist2tuple (id:idl))))
---checkStm (SAss (l,ll) (e,el)) g = if subtype2 (checkExpList (e:el) g) (checkVarList (l:ll) g)
---                                  then TVoid
---                                  else error("assign: " ++ (show (checkExpList (e:el) g)) ++ " </: " ++ (show (checkVarList (l:ll) g)))
 checkStm (SAss (l,ll) el) g = if subtype2 (checkExpList1 el g) (checkVarList (l:ll) g)
                                   then TVoid
                                   else error("assign: " ++ (show (checkExpList1 el g)) ++ " </: " ++ (show (checkVarList (l:ll) g)))
---checkStm (SRet el) g = checkExpList el g
 checkStm (SRet el) g = checkExpList0 el g
---checkStm (SApp e el) g = case g "strict" of
---  TNil -> case checkExp e g of
---            (TFunction p r) -> if subtype2 (checkExpList el g) p
---                               then TVoid
[…]

### 2014-11-24 doc/thesis/system.tex
Adding raw text about the refinement of table types

@@
 The most unusual feature of our type system is the refinement of
-table types, which is given by the following rule:
+table types:
 \[
@@
 & \multicolumn{1}{l}{$t[``x"] \; {<}\Integer{>} = 1$}\\
+\end{tabular}
+\end{center}
+
+This example also uses the rule \textsc{T-CONSTRUCTOR1} to
+infer the type of the empty table constructor:
+\[
+\begin{array}{c}
+\mylabel{T-CONSTRUCTOR1}\\
+\env \vdash \{\}:\{\}_{unique}, \env
+\end{array}
+\]
+
+Typed Lua also has inference rules to infer table types that we
+can use to initialize hashes, arrays, and records:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-CONSTRUCTOR2}\\
+\dfrac{\env \vdash c:k_{i}{:}v_{i}, \env_{i} \;\;\;
+       n = |\;\vec{c}\;| \;\;\;
+       wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique})}
+      {\env \vdash \{\;\vec{c}\;\}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique}, \env}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-CONSTRUCTOR3}\\
+\dfrac{\env \vdash {...}:t, \env}
+      {\env \vdash \{{...}_{m}\}:\{\Integer{:}t \cup \Nil\}_{unique}, \env}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-CONSTRUCTOR4}\\
+\dfrac{\begin{array}{c}
+       \env \vdash c:k_{i}{:}v_{i}, \env_{i} \;\;\;
+       \env \vdash {...}:t, \env \;\;\;
+       n = |\;\vec{c}\;| \\
+       wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, \Integer{:}t \cup \Nil\}_{unique})
+       \end{array}}
+      {\env \vdash \{\;\vec{c},\;{...}_{m}\;\}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, \Integer{:}v \cup \Nil\}_{unique}, \env}
+\end{array}
+\end{array}
+\]
+
+The rule \textsc{T-CONSTRUCTOR2} type checks the following example:
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; a:\{\String:\Integer \cup \Nil\}_{closed} = \{ [``x"] = 1, [``y"] = 2 \} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$\mathbf{local} \; b:\Integer \cup \Nil = a[``x"] \; \mathbf{in} \; x = a[``z"]$}
+\end{tabular}
+\end{center}
+
+It also uses the rules \textsc{T-FIELD1} and \textsc{T-FIELD2}:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-FIELD1}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+       t_{1} \subtype \Boolean \cup \Number \cup \String}
+      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}:super(t_{2}), \env_{3}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-FIELD2}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+       \Nil \not\subtype t_{1}}
+      {\env_{1} \vdash [e_{1}] = e_{2}: \Any:super(t_{2}), \env_{3}}
+\end{array}
+\end{array}
+\]
+
+The rule \textsc{T-FIELD1} infers fields that keys are subype
+of $\Boolean$, $\Number$, or $\String$.
+The rule \textsc{T-FIELD2} infers fields thaty keys have the
+dynamic type if its type does not include $\Nil$.
+
+Typed Lua handles arrays as hashes that maps $\Integer$ to some
+type $t$.
+We did not include in the abstract syntax tree the simplified
+syntax of table consturctors that initializes arrays because
+we can write it using explicit indices.
+For instance, we write the table constructor
+$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to mean the table constructor
+$\{ ``x", ``y", ``z" \}$.
+We chose to make this simplification because the inference rules
+for the simpflified version does not include anything new.
+They are the same rules that just automatically control the indices.
+For the same reason, we also omitted in the table constructor expressions
+that return multiple results.
+However, we did not omit the vararg expression in the last position
+to show the case that we infer a record with an array part.
+
+The rule \textsc{T-CONSTRUCTOR3} type checks the following example:
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; a:\{\Integer:\String \cup \Nil\}_{closed} = \{ {...}_{m} \} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$\mathbf{local} \; b:\String \cup \Nil = t[1] \; \mathbf{in} \; b = t[5]$}
+\end{tabular}
+\end{center}
+
+We are assuming that $...$ is in the environment as has type $\String$.
+
+The rule \textsc{T-CONSTRUCTOR4} type checks the following example:
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; a:\{``x":\Integer, ``y":\Boolean, \Integer:\String \cup \Nil\}_{closed} =$}\\
+& \multicolumn{1}{l}{$\{ [``x"] = ``foo", [``y"] = \mathbf{true}, {...}_{m} \}$}\\
+\multicolumn{2}{l}{$\mathbf{in} \; a[``x"] = ``bar"$}
+\end{tabular}
+\end{center}
+
+We are also assuming that $...$ is in the environment as has type $\String$.
+
+These rules along \emph{unique} table types also allow us initialize
+record types that have optional fields.
+We can include the optional field in the initialization
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; a:\{``x":\String, ``y":\String \cup \Nil, ``z":\String\}_{closed} =$}\\
+& \multicolumn{1}{l}{$\{ [``x"] = ``foo", [``y"] = ``bar", [``z"] = ``baz" \}$}\\
+\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
+\end{tabular}
+\end{center}
+and we can also ommit it
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; a:\{``x":\String, ``y":\String \cup \Nil, ``z":\String\}_{closed} =$}\\
+& \multicolumn{1}{l}{$\{ [``x"] = ``foo", [``z"] = ``baz" \}$}\\
+\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
+\end{tabular}
+\end{center}
+
+Typed Lua also keeps track whether it is safe or not safe to
+change the type of a table.
+To do that, our type system always promotes an \emph{open}
+table to \emph{closed} and an \emph{unique} table to \emph{open}
+when they appear as expressions.
+The order also matters, as going from \emph{unique} to \emph{open}
+before going from \emph{open} to \emph{closed} closes \emph{unique} tables.
+The rule \textsc{T-IDREAD} along with the predicate \emph{close}
+express this intuition:
+\[
+\begin{array}{c}
+\mylabel{T-IDREAD}\\
+\dfrac{\env_{1}(n_{r}) = t}
+      {\env_{1} \vdash n_{r}:close(t), \env_{1}}
+\end{array}
+\]
+
+The rule \textsc{T-IDREAD} makes this unsafe example not type check:
+\begin{center}
+\begin{tabular}{lll}
+\multicolumn{3}{l}{$\mathbf{local} \; a:\{``x":\String \cup \Nil\}_{closed} = \{ [``x"] = ``foo" \} \; \mathbf{in}$}\\
+& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String \}_{unique} = \{ [``x"] = ``foo" \} \; \mathbf{in}$}\\
+& & \multicolumn{1}{l}{$a = b \;;\; a[``x"] = \mathbf{nil}$}
+\end{tabular}
+\end{center}
+
+The assignment $a = b$ is not valid because the aliasing changed the
+type of $b$ from $\{``x":\String\}_{unique}$ to $\{``x":\String\}_{open}$,
+which is not subtype of $\{``x":\String \cup \Nil\}_{closed}$.
+
+However, sometimes we want to use \emph{unique} tables to initialize
+closed tables, as they allow us to initialize optional fields.
+To do that, Typed Lua includes a cast expression that allow us
+to use the \emph{unique} table type before it becomes \emph{open}.
+The rule \textsc{T-CAST} describes this behavior:
+\[
+\begin{array}{c}
+\mylabel{T-CAST}\\
+\dfrac{t \subtype \env_{1}(n)}
+      {\env_{1} \vdash {<}t{>} \; n:t, \env_{1}[n \mapsto t]}
+\end{array}
+\]
+
+The rule \textsc{T-CAST} allow us to type check the following example:
+\begin{center}
+\begin{tabular}{lll}
+\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{ \} \; \mathbf{in}$}\\
+& \multicolumn{2}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
+& \multicolumn{2}{l}{$a[``y"] \; {<}\String{>} = ``bar";$}\\
+& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{closed} =$}\\
+& & \multicolumn{1}{l}{${<}\{``x":\String, ``y":\String \cup \Nil\}_{open}{>} \; a \; \mathbf{in} \; a[``z"] \; {<}\Integer{>} = 1$}
+\end{tabular}
+\end{center}
+
+We can use $a$ to initialize $b$ because the cast expression converts
+the type of $a$ from $\{``x":\String, ``y":\String\}_{unique}$ to
+$\{``x":\String, ``y":\String \cup \Nil\}_{open}$, which is subtype of
+$\{``x":\String, ``y":\String \cup \Nil\}_{closed}$.
+We can continue to refine the type of $a$ after the aliasing,
+as it still holds an \emph{open} table.
+At the end of this example, $a$ has type
+$\{``x":\String, ``y":\String \cup \Nil, ``z":\Integer\}_{open}$.
+
+Refinement of table types (second example).
+\begin{center}
+\begin{tabular}{lll}
+\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
+& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = a \; \mathbf{in}$}\\
+& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
+& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
+\end{tabular}
+\end{center}
+
+Typed Lua uses the refinement of table types to express an object-oriented
+idioms that is commonly used by Lua programmers.
+\[
+\begin{array}{c}
+\mylabel{T-METHOD3}\\
+\dfrac{\begin{array}{c}
+       \env_{1}(n_{1}) = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{unique|open} \;\;\;
+       \env_{1} \vdash n_{2} : l \;\;\;
+       \not \exists i \in 1..n \; l \lesssim k_{i}\\
+       closeall(\env_{1}[self \mapsto \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{closed}, \vec{n} \mapsto \vec{t}, \ret \mapsto r]) \vdash s:\env_{2}
+       \end{array}}
+      {\begin{array}{c}
+       \env_{1} \vdash \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; s:\\
+       closeset(\env_{1}[n_{1} \mapsto \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}, l{:}\Const \; \Self \times \vec{t} \rightarrow r\}_{unique|closed}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))
+       \end{array}}
+\end{array}
+\]
+
+\[
+\begin{array}{c}
+\mylabel{T-INVOKE1}\\
+\dfrac{\env_{1} \vdash e[n]:p_{1} \rightarrow r, \env_{2} \;\;\;
+       \env_{2} \vdash el:p_{2}, \env_{3} \;\;\;
+       \Self \times p_{2} \lesssim p_{1}}
+      {\env_{1} \vdash e{:}n(el)_{m}:r, \env_{3}}
+\end{array}
+\]
+
+\begin{center}
+\begin{tabular}{llll}
+\multicolumn{4}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{open} = \{ [``x"] = 0.0, [``y"] = 0.0 \}$}\\
+\multicolumn{4}{l}{$\mathbf{in}$}\\
+& \multicolumn{3}{l}{$\mathbf{fun} \; s{:}new (x:number, y:number):\Self$}\\
+& & \multicolumn{2}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{closed} =$}\\
+& & & \multicolumn{1}{l}{$setmetatable(\{\}, \{ [``\string_\string_index"] = self \})$}\\
+& & \multicolumn{2}{l}{$\mathbf{in} \; s[``x"] = x; \; s[``y"] = y; \; \mathbf{return} \; s$}\\
+; & \multicolumn{3}{l}{$\mathbf{fun} \; s{:}move (x:number, y:number):\Void$}\\
+& & \multicolumn{2}{l}{$self[``x"] = self[``x"] + x;$}\\
+& & \multicolumn{2}{l}{$self[``y"] = self[``y"] + y$}
 \end{tabular}
@@
 $\Value{*} \rightarrow \Void$.
-
-Table type as hash.
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; t:\{\String:\Integer \cup \Nil\}_{closed} = \{ [``foo"] = 1 \} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$\mathbf{local} \; x:\Integer \cup \Nil = t[``foo"] \; \mathbf{in} \; x = t[``bar"]$}
-\end{tabular}
-\end{center}
-
-Table type as array.
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; t:\{\Integer:\String \cup \Nil\}_{closed} = \{ ``a", ``b", ``c" \} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$\mathbf{local} \; x:\String \cup \Nil = t[1] \; \mathbf{in} \; x = t[4]$}
-\end{tabular}
-\end{center}
-
-Table type as record (first example).
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; t:\{``f":\String, ``l":\String\}_{closed} =$}\\
-& \multicolumn{1}{l}{$\{ [``l"] = ``Reed", [``f"] = ``Lou" \}$}\\
-\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
-\end{tabular}
-\end{center}
-
-Table type as record (second example).
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; t:\{``f":\String, ``m":\String \cup \Nil, ``l":\String\}_{closed} =$}\\
-& \multicolumn{1}{l}{$\{ [``f"] = ``Lewis", [``m"] = ``Allan", [``l"] = ``Reed" \}$}\\
-\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
-\end{tabular}
-\end{center}
-
-Table type as record (third example).
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; t:\{``f":\String, ``m":\String \cup \Nil, ``l":\String\}_{closed} =$}\\
-& \multicolumn{1}{l}{$\{ [``f"] = ``Lou", [``l"] = ``Reed" \}$}\\
-\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
-\end{tabular}
-\end{center}
-
-Refinement of table types (second example).
-\begin{center}
-\begin{tabular}{lll}
-\multicolumn{3}{l}{$\mathbf{local} \; t:\{\}_{open} = \{\} \; \mathbf{in}$}\\
-& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = p \; \mathbf{in}$}\\
-& & \multicolumn{1}{l}{$t[``f"] \; {<}\String{>} = ``Lou";$}\\
-& & \multicolumn{1}{l}{$t[``l"] \; {<}\String{>} = ``Reed";$}\\
-\end{tabular}
-\end{center}
-
-Objects.
-\begin{center}
-\begin{tabular}{llll}
-\multicolumn{4}{l}{$\mathbf{local} \; s:\{``x:"\Number, ``y":\Number\}_{open} = \{ [``x"] = 0.0, [``y"] = 0.0 \}$}\\
-\multicolumn{4}{l}{$\mathbf{in}$}\\
-& \multicolumn{3}{l}{$\mathbf{fun} \; s{:}new (x:number, y:number):\Self$}\\
-& & \multicolumn{2}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{closed} =$}\\
-& & & \multicolumn{1}{l}{$setmetatable(\{\}, \{ [``\string_\string_index"] = self \}$}\\
-& & \multicolumn{2}{l}{$\mathbf{in} \; s[``x"] = x; \; s[``y"] = y; \; \mathbf{return} \; s$}\\
-; & \multicolumn{3}{l}{$\mathbf{fun} \; s{:}move (x:number, y:number):\Void$}\\
-& & \multicolumn{2}{l}{$self[``x"] = self[``x"] + x;$}\\
-& & \multicolumn{2}{l}{$self[``y"] = self[``y"] + y$}
-\end{tabular}
-\end{center}

### 2014-11-25 doc/thesis/system.tex
Improving the raw text about the refinement of table types

@@
 s_{1} \; ; \; s_{2} \; | \;
-ll = el_{1} \; | \;
+\vec{l} = el \; | \;
 m \; | \;
@@
 \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
-& | \; \mathbf{local} \; \vec{n{:}t} = el_{0} \; \mathbf{in} \; s \; | \;
+& | \; \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s \; | \;
 \mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
-\mathbf{return} \; el\\
-& | \; e(el)_{s} \; | \;
+\mathbf{return} \; el \; | \;
+e(el)_{s} \; | \;
 e{:}n(el)_{s}\\
 e ::= & \;\; \mathbf{nil} \; | \;
-x \; | \;
+c \; | \;
 {...}_{e} \; | \;
@@
 \{ \} \; | \;
-\{ \; cl \; \} \\
+\{ \; \vec{p} \; \} \; | \;
+\{ \; {...}_{m} \; \} \; | \;
+\{ \; \vec{p},{...}_{m} \; \}\\
 & | \; e_{1} + e_{2} \; | \;
@@
 e{:}n(el)_{e}\\
-x ::= & \;\; \mathbf{false} \; | \;
+l ::= & \;\; n \; | \;
+e_{1}[e_{2}] \; | \;
+n[c] \; {<}t{>}\\
+c ::= & \;\; \mathbf{false} \; | \;
 \mathbf{true} \; | \;
@@
 {\it string}\\
-l ::= & \;\; n \; | \;
-e_{1}[e_{2}] \; | \;
-n[e] \; {<}t{>}\\
-ll ::= & \;\; \vec{l}\\
-el_{0} ::= & \;\; \mathbf{nothing} \; | \;
-el_{1}\\
-el_{1} ::= & \;\; \vec{e} \; | \;
+p ::= & \;\; [e_{1}] = e_{2}\\
+el ::= & \;\; \mathbf{nothing} \; | \;
+\vec{e} \; | \;
 me \; | \;
 \vec{e}, me\\
-el ::= & \;\; el_{0}\\
-cl ::= & \;\; \vec{c} \; | \;
-{...}_{m} \; | \;
-\vec{c}, {...}_{m}\\
 me ::= & \;\; e(el)_{m} \; | \;
@@
 {...}_{m}\\
-m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (){:}r \; s \; ; \; \mathbf{return} \; el\\
-& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; ({...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
-& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; s \; ; \; \mathbf{return} \; el\\
-& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t},{...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
-f ::= & \;\; \mathbf{fun} \; (){:}r \; s \; ; \; \mathbf{return} \; el\\
-& | \; \mathbf{fun} \; ({...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
-& | \; \mathbf{fun} \; (\vec{n{:}t}){:}r \; s \; ; \; \mathbf{return} \; el\\
-& | \; \mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
-c ::= & \;\; [e_{1}] = e_{2}\\
+m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (){:}r \; fb \; | \;
+\mathbf{fun} \; n_{1}{:}n_{2} \; ({...}{:}t){:}r \; fb\\
+& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; fb \; | \;
+\mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t},{...}{:}t){:}r \; fb\\
+f ::= & \;\; \mathbf{fun} \; (){:}r \; fb \; | \;
+\mathbf{fun} \; ({...}{:}t){:}r \; fb \; | \;
+\mathbf{fun} \; (\vec{n{:}t}){:}r \; fb \; | \;
+\mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; fb\\
+fb ::= & \;\; s \;;\; \mathbf{return} \; el\\
 n ::= & \;\; {\it name}\\
@@
 The most unusual feature of our type system is the refinement of
-table types:
+table types.
+The rule \textsc{T-REFINE} allow us to add new keys to \emph{open}
+and \emph{unique} table types:
 \[
@@
 \dfrac{\env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique} \;\;\;
-       \env_{1} \vdash x:t_{1}, \env_{2} \;\;\;
+       \env_{1} \vdash c:t_{1}, \env_{2} \;\;\;
        \not \exists i \in 1..n \; t_{1} \lesssim k_{i}}
-      {\env_{1} \vdash n[x] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, t_{1}{:}t\}_{open|unique}]}
-\end{array}
-\]
-
-The rule \textsc{T-REFINE} allow us to add new keys to \emph{open}
-or \emph{unique} table types.
-We restricted the keys to literals only because the purpose of
-this rule is to make it easier to add new fields to table types
-that represent records.
-We also do not allow changing the type of a field that is already
-in the table because this could lead to loosen table types that
-have no static type checking. 
-For instance, Typed Lua allows us to write the following code:
+      {\env_{1} \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, t_{1}{:}t\}_{open|unique}]}
+\end{array}
+\]
+
+We restricted the refinement of table types to include only literal
+keys because its purpose is to make it easier the construction of
+table types that represent records.
+For instance, 
 \begin{center}
 \begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; t:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$t[``x"] \; {<}\String{>} = ``foo";$}\\
-& \multicolumn{1}{l}{$t[``y"] \; {<}\Integer{>} = 1$}\\
-\end{tabular}
-\end{center}
-but it does not allow us to write the following code:
+\multicolumn{2}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
+& \multicolumn{1}{l}{$a[``y"] \; {<}\Integer{>} = 1$}\\
+\end{tabular}
+\end{center}
+uses field assignment to add fields $x$ and $y$ to the local $a$.
+After these field assignments $a$ has type
+$\{``x":\String, ``y":\Integer\}_{unique}$.
+
+We do not allow the refinement of table types to change the type of
+a field that is already present in the table type.
+For instance,
 \begin{center}
 \begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; t:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$t[``x"] \; {<}\String{>} = ``foo";$}\\
-& \multicolumn{1}{l}{$t[``x"] \; {<}\Integer{>} = 1$}\\
-\end{tabular}
-\end{center}
-
-This example also uses the rule \textsc{T-CONSTRUCTOR1} to
-infer the type of the empty table constructor:
+\multicolumn{2}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
+& \multicolumn{1}{l}{$a[``x"] \; {<}\Integer{>} = 1$}\\
+\end{tabular}
+\end{center}
+do not type check, as we are trying to change the type of field
+``x" from $\String$ to $\String \cup \Integer$.
+More precisely, our type system does not allow the table type to
+change towards a type that is not a subtype of the previous type.
+In this example,
+$\{``x":\String \cup \Integer\}_{unique} \not\subtype \{``x":\String\}_{unique}$.
+We have this restriction because this kind of refinement could lead
+to fields that have a too general type.
+
+\textit{/* We need to close table types before assigning them to fields. */}
+\begin{center}
+\begin{tabular}{l}
+$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in} \; a[``x"] \; {<}\{\}_{closed}{>} = \{\}$
+\end{tabular}
+\end{center}
+
+Lua has multiple assignments, so Typed Lua also has to take this into
+consideration while refining table types.
+The rule \textsc{T-ASSINGMENT} shows that tuple types make it easy
+to use the consistent-subtyping relation to check multiple assignments:
+\[
+\begin{array}{c}
+\mylabel{T-ASSIGNMENT}\\
+\begin{array}{c}
+\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
+       r_{1} \lesssim r_{2}}
+      {\env_{1} \vdash \vec{l} = el:\env_{3}}
+\end{array}
+\end{array}
+\]
+
+This rule is so simple because we left the difficult part to the
+type checking of expression lists.
+The list of left-hand values is also a kind of expression list.
+The difficult part is to check whether there is an expression in
+the expression list that messes the environment, and the rule
+\textsc{T-LHSLIST} captures this intuition:
+\[
+\begin{array}{c}
+\mylabel{T-LHSLIST}\\
+\dfrac{\env \vdash l_{k}:t_{k}, \env_{k} \;\;\;
+       merge(\env_{1}, ..., \env_{n}) = \env_{m} \;\;\;
+       n = |\;\vec{l}\;|}
+      {\env \vdash \vec{l}:t_{1} \times ... \times t_{n} \times \Value{*}, \env_{m}}
+\end{array}
+\]
+
+First, this rule uses the same environment $\env$ to type check each
+left-hand side value $l_{k}$, as they can perform different changes
+to the environment.
+Then, it takes each type $t_{k}$ and builds the tuple type for
+the assignment rule.
+It places the type $\Value{*}$ in the end to discard extra values.
+The rule succeeds if the predicate \emph{merge} can
+produce a new environment, which includes all the modifications
+of each environment $\env_{k}$.
+Intuitively, the predicate \emph{merge} fails when an environment
+tries to change the type of a variable towards a type that is not
+a subtype of its previous type.
+
+We can also use multiple assignments to refine table types:
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$a[``x"] \; {<}\String{>}, a[``y"] \; {<}\Integer{>} = ``foo", 1$}\\
+\end{tabular}
+\end{center}
+
+This example type checks because all the environment changes are consistent, and
+$``foo" \times 1 \times \Nil{*} \lesssim \String \times \Integer \times \Value{*}$.
+The type of the expression list includes the type $\Nil{*}$ in the end
+because our type system has typing rules to add this type when
+the type of the expression list does not end with a variadic type.
+Our type system has this behavior to use $\Nil$ in the place of missing
+values, like Lua does.
+This example uses the rule \textsc{T-EXPLIST2}, which is really close
+to the rule \textsc{T-LHSLIST}:
+\[
+\begin{array}{c}
+\mylabel{T-EXPLIST2}\\
+\dfrac{\env \vdash e_{k}:t_{k}, \env_{k} \;\;\;
+       merge(\env_{1}, ..., \env_{n}) = \env_{m} \;\;\;
+       n = |\;\vec{e}\;|}
+      {\env \vdash \vec{e}:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{m}}
+\end{array}
+\]
+
+The next example does not type check because it tries to add
+the same field to $a$:
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$a[``x"] \; {<}\String{>}, a[``x"] \; {<}\Integer{>} = ``foo", 1$}\\
+\end{tabular}
+\end{center}
+
+Typed Lua also keeps track whether it is safe or not safe to
+change the type of a table, but before discussing this subject,
+we will discuss the different table constructors that our
+abstract syntax includes, as they will appear in the next
+examples.
+
+All the examples that we presented until now use the simplest form of
+the table constructor: the empty table $\{\}$.
+Its inference rule is starightforward:
 \[
@@
 
-Typed Lua also has inference rules to infer table types that we
-can use to initialize hashes, arrays, and records:
-\[
-\begin{array}{c}
+Our abstract syntax reduces the more complex uses of the table
+constructor into three forms: $\{\;\vec{p}\;\}$, $\{\;{...}_{m}\;\}$,
+and $\{\;\vec{p},{...}_{m}\;\}$.
+The first one uses a list of expression pairs $[e_{1}] = e_{2}$,
+the second one uses just the vararg expression, and the third one uses both.
+We did not include in the abstract syntax a table constructor
+that uses a list of expressions because we can write it using the first form.
+For instance, we can write the table constructor
+$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to mean the table constructor
+$\{ ``x", ``y", ``z" \}$.
+We chose to not include this form because its inference rule is similar
+to the inference rule of the first form, and just controls the type of
+the ommited indices.
+For the same reason we did not include table constructors with
+multiple expressions that can return a tuple type that does not end
+with a variadic type.
+We did include the vararg expression in the last position to show how
+our type system can infer a table type that has a record part and
+also an array part.
+
+The inference rule \textsc{T-CONSTRUCTOR2} infers a table type for
+the first form:
+\[
 \begin{array}{c}
 \mylabel{T-CONSTRUCTOR2}\\
-\dfrac{\env \vdash c:k_{i}{:}v_{i}, \env_{i} \;\;\;
-       n = |\;\vec{c}\;| \;\;\;
-       wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique})}
-      {\env \vdash \{\;\vec{c}\;\}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique}, \env}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-CONSTRUCTOR3}\\
-\dfrac{\env \vdash {...}:t, \env}
-      {\env \vdash \{{...}_{m}\}:\{\Integer{:}t \cup \Nil\}_{unique}, \env}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-CONSTRUCTOR4}\\
-\dfrac{\begin{array}{c}
-       \env \vdash c:k_{i}{:}v_{i}, \env_{i} \;\;\;
-       \env \vdash {...}:t, \env \;\;\;
-       n = |\;\vec{c}\;| \\
-       wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, \Integer{:}t \cup \Nil\}_{unique})
-       \end{array}}
-      {\env \vdash \{\;\vec{c},\;{...}_{m}\;\}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, \Integer{:}v \cup \Nil\}_{unique}, \env}
-\end{array}
-\end{array}
-\]
-
-The rule \textsc{T-CONSTRUCTOR2} type checks the following example:
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; a:\{\String:\Integer \cup \Nil\}_{closed} = \{ [``x"] = 1, [``y"] = 2 \} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$\mathbf{local} \; b:\Integer \cup \Nil = a[``x"] \; \mathbf{in} \; x = a[``z"]$}
-\end{tabular}
-\end{center}
-
-It also uses the rules \textsc{T-FIELD1} and \textsc{T-FIELD2}:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-FIELD1}\\
+\dfrac{\env_{i} \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
+       n = |\;\vec{p}\;| \;\;\;
+       wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique}) = t}
+      {\env_{1} \vdash \{\;\vec{p}\;\}:t, \env_{n+1}}
+\end{array}
+\]
+
+This inference rule infers the type of each pair, uses these types
+to build a table type, and uses the predicate \emph{wf} to check
+whether the table type is well formed.
+We gave the definition of this predicate in the last section.
+
+The inference rules for expression pairs would be straightforward
+if our table types allowed any type in the keys.
+The rules \textsc{T-PAIR1}, \textsc{T-PAIR2}, and \textsc{T-PAIR3}
+check if the type of the key is a $\Boolean$, a $\Number$,
+or a $\String$, respectively:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-PAIR1}\\
 \dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
        \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
-       t_{1} \subtype \Boolean \cup \Number \cup \String}
-      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}:super(t_{2}), \env_{3}}
+       t_{1} \subtype \Boolean}
+      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}t_{2}, \env_{3}}
 \end{array}
@@
 \begin{array}{c}
-\mylabel{T-FIELD2}\\
+\mylabel{T-PAIR2}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+       t_{1} \subtype \Number}
+      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}t_{2}, \env_{3}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-PAIR3}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+       t_{1} \subtype \String}
+      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}t_{2}, \env_{3}}
+\end{array}
+\end{array}
+\]
+
+The rule \textsc{T-PAIR4} is a fallback for the previous rules,
+but it works only if the type of the key does not include $\Nil$:
+\[
+\begin{array}{c}
+\mylabel{T-PAIR4}\\
 \dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
@@
        \Nil \not\subtype t_{1}}
-      {\env_{1} \vdash [e_{1}] = e_{2}: \Any:super(t_{2}), \env_{3}}
-\end{array}
-\end{array}
-\]
-
-The rule \textsc{T-FIELD1} infers fields that keys are subype
-of $\Boolean$, $\Number$, or $\String$.
-The rule \textsc{T-FIELD2} infers fields thaty keys have the
-dynamic type if its type does not include $\Nil$.
-
-Typed Lua handles arrays as hashes that maps $\Integer$ to some
-type $t$.
-We did not include in the abstract syntax tree the simplified
-syntax of table consturctors that initializes arrays because
-we can write it using explicit indices.
-For instance, we write the table constructor
-$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to mean the table constructor
-$\{ ``x", ``y", ``z" \}$.
-We chose to make this simplification because the inference rules
-for the simpflified version does not include anything new.
-They are the same rules that just automatically control the indices.
-For the same reason, we also omitted in the table constructor expressions
-that return multiple results.
-However, we did not omit the vararg expression in the last position
-to show the case that we infer a record with an array part.
-
-The rule \textsc{T-CONSTRUCTOR3} type checks the following example:
+      {\env_{1} \vdash [e_{1}] = e_{2}: \Any{:}t_{2}, \env_{3}}
+\end{array}
+\]
+
+The rule \textsc{T-CONSTRUCTOR2} along with the rules for expression
+pairs allow Typed Lua to type check the following example:
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; a:\{\String:\Integer \cup \Nil\}_{closed} = \{ [``x"] = 1, [``y"] = 2 \} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$\mathbf{local} \; b:\Integer \cup \Nil = a[``x"] \; \mathbf{in} \; b = a[``z"]$}
+\end{tabular}
+\end{center}
+
+This example type checks because Typed Lua infers the table type
+$\{``x":1, ``y":2\}_{unique}$, which is subtype of
+$\{\String:\Integer \cup \Nil\}_{closed}$.
+Even though the key $``z"$ does not exist in the table $a$,
+the two table indexations type check with type $\Integer \cup \Nil$,
+which is subtype of the type of $b$.
+The rule \textsc{T-INDEX1} describes this behavior:
+\[
+\begin{array}{c}
+\mylabel{T-INDEX1}\\
+\dfrac{\env_{1} \vdash e_{1}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t, \env_{3} \;\;\;
+       \exists i \in 1{..}n \; t \lesssim k_{i}}
+      {\env_{1} \vdash e_{1}[e_{2}]:v_{i}, \env_{3}}
+\end{array}
+\]
+
+Typed Lua handles arrays as hashes that maps integers to some type $t$.
+In Lua, programmers often use the vararg expression to initialize arrays.
+The rules \textsc{T-CONSTRUCTOR3} and \textsc{T-CONSTRUCTOR4} define
+the behavior of the vararg expression for this case:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-CONSTRUCTOR3}\\
+\dfrac{\env_{1}({...}) = t}
+      {\env_{1} \vdash \{{...}_{m}\}:\{\Integer{:}t \cup \Nil\}_{unique}, \env_{1}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-CONSTRUCTOR4}\\
+\dfrac{\begin{array}{c}
+       \env_{i} \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
+       n = |\;\vec{p}\;| \;\;\;
+       \env_{i+1}({...}) = t_{v}\\
+       wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, \Integer{:}t_{v} \cup \Nil\}_{unique}) = t_{t}
+       \end{array}}
+      {\env_{1} \vdash \{\;\vec{p},\;{...}_{m}\;\}:t_{t}, \env_{n+1}}
+\end{array}
+\end{array}
+\]
+
+The rule \textsc{T-CONSTRUCTOR3} type checks the case where we use
+only the vararg expression inside a table constructor to initialize
+an array.
+If we assume that the $...$ is in the environment and has type $\String$,
+the following example type checks through this rule:
 \begin{center}
@@
 
-We are assuming that $...$ is in the environment as has type $\String$.
-
-The rule \textsc{T-CONSTRUCTOR4} type checks the following example:
+The rule \textsc{T-CONSTRUCTOR4} type checks the case where we use
+the table constructor as a record that includes an array part.
+If we assume that the $...$ is in the environment and has type $\String$,
+the following example type checks through this rule:
 \begin{center}
@@
 \end{center}
-
-We are also assuming that $...$ is in the environment as has type $\String$.
-
-These rules along \emph{unique} table types also allow us initialize
-record types that have optional fields.
+but the next example does not type check because the record part
+includes fields that the key types are subtype of the array part:
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; a:\{1:\String, 10:\Boolean, \Integer:\String \cup \Nil\}_{closed} =$}\\
+& \multicolumn{1}{l}{$\{ [1] = ``foo", [10] = \mathbf{true}, {...}_{m} \}$}\\
+\multicolumn{2}{l}{$\mathbf{in} \; a[``x"] = ``bar"$}
+\end{tabular}
+\end{center}
+
+Making the inference rules of the table constructor infer an \emph{unique}
+table type also allow us to use them for initializing record types that
+have optional fields.
 We can include the optional field in the initialization
@@
 \end{center}
-and we can also ommit it
+and we can also omit it in the initialization
 \begin{center}
@@
 
-Typed Lua also keeps track whether it is safe or not safe to
-change the type of a table.
+After discussing the inference rules of the table constructor,
+we can get back to the discussion about the refinement of table types.
+
+Besides allowing the refinement of table types through field assignment,
+Typed Lua also keeps track whether it is safe or not safe to change the
+type of a table.
 To do that, our type system always promotes an \emph{open}
 table to \emph{closed} and an \emph{unique} table to \emph{open}
-when they appear as expressions.
-The order also matters, as going from \emph{unique} to \emph{open}
-before going from \emph{open} to \emph{closed} closes \emph{unique} tables.
-The rule \textsc{T-IDREAD} along with the predicate \emph{close}
-express this intuition:
+when they appear as expressions, or when they enter in a scope that is
+not its declaration scope.
+The order matters, as going from \emph{unique} to \emph{open}
+before going from \emph{open} to \emph{closed} closes \emph{unique} table types.
+
+The rule \textsc{T-IDREAD} uses the predicate \emph{close} to do
+this promotion before we use a local variable:
 \[
@@
 
-The rule \textsc{T-IDREAD} makes this unsafe example not type check:
+This rule does not change the type of the local variable because it
+just need to ensure that an alias will not get the same type.
+For instance, it makes this unsafe example not type check:
 \begin{center}
@@
 
-However, sometimes we want to use \emph{unique} tables to initialize
-closed tables, as they allow us to initialize optional fields.
+However, sometimes we want to use \emph{unique} table types to initialize
+\emph{closed} table types, as they allow us to initialize optional fields.
 To do that, Typed Lua includes a cast expression that allow us
-to use the \emph{unique} table type before it becomes \emph{open}.
+to use an \emph{unique} table type before it becomes \emph{open}.
 The rule \textsc{T-CAST} describes this behavior:
@@
 
-Refinement of table types (second example).
+Even though the rule \textsc{T-CAST} handles the case that we assign
+an alias to a variable, we still have to close this variable due to
+safety.
+The rule \textsc{T-LOCAL} is one example where we handle this case:
+\[
+\begin{array}{c}
+\mylabel{T-LOCAL}\\
+\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
+       r_{1} \lesssim \vec{t} \;\;\;
+       closeset(\env_{2}[\vec{n} \mapsto \vec{t}], fav(\mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s)) \vdash s:\env_{3}}
+      {\env_{1} \vdash \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s:\env_{3} - \{\vec{n} \mapsto \vec{t}\}}
+\end{array}
+\]
+
+This rule prevents the following unsafe example to type check:
 \begin{center}
@@
 
-Typed Lua uses the refinement of table types to express an object-oriented
-idioms that is commonly used by Lua programmers.
+We cannot add the field $``x"$ to $b$ because its type is \emph{closed},
+and thus do not allow chaning the value that is stored in the field $``x"$
+of local $a$.
+
+We also need to make sure to close all \emph{open} table types before we
+type check another scope.
+The rule \textsc{T-FUNCTION3} shows this case:
+\[
+\begin{array}{c}
+\mylabel{T-FUNCTION3}\\
+\dfrac{closeall(\env_{1}[\vec{n} \mapsto \vec{t}, \ret \mapsto r]) \vdash s:\env_{2}}
+      {\env_{1} \vdash \mathbf{fun} \; (\vec{n{:}t}){:}r \; s:\vec{t} \rightarrow r, closeset(\env_{1}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))}
+\end{array}
+\]
+
+This rule prevents the following unsafe example to type check:
+\begin{center}
+\begin{tabular}{llll}
+\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
+& \multicolumn{3}{l}{$\mathbf{local} \; b:\Integer \times \Integer \rightarrow \Integer =$}\\
+& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer, y:\Integer):\Integer$}\\
+& & & \multicolumn{1}{l}{$a[``z"] \; {<}\Integer{>} = 1; \; \mathbf{return} \; x + y$}\\
+& \multicolumn{3}{l}{$\mathbf{in} \; a[``z"] \; {<}\String{>} = ``foo"$}
+\end{tabular}
+\end{center}
+
+Typed Lua uses the refinement of table types to allow programmers to
+build modules and objects.
+It also allows Typed Lua to type check an object-oriented idiom that
+programmers often use it.
 \[
@@
 \end{center}
-
-\textit{/* Expression list rules. Where should we describe the
-discarding/replacement of tuple elements? */}
-
-Even though Lua has multiple assignments, the rule \textsc{T-ASSIGNMENT}
-shows that tuples types and consistent-subtyping make straightforward
-to type check them.
-\[
-\begin{array}{c}
-\mylabel{T-ASSIGNMENT}\\
-\begin{array}{c}
-\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
-       r_{1} \lesssim r_{2}}
-      {\env_{1} \vdash \vec{l} = el:\env_{3}}
-\end{array}
-\end{array}
-\]
 

### 2014-11-27 doc/thesis/system.tex
Minor improvements to the text about typing rules

@@
 any integer value, any floating point value, and any string value.
-We shall see that literal types are important to represet table types
+We shall see that literal types are important to represent table types
 as records.
@@
 and \textsc{S-INT2}, because the first two are reflexive and we
-could achieve the third one trhough transitivity.
+could achieve the third one through transitivity.
 More precisely, we could have defined subsumption rules that
-specify the reflexive and trasitive behavior of subtyping,
+specify the reflexive and transitive behavior of subtyping,
 and use these rules instead of the ones we defined.
 For instance, we would not need \textsc{S-INT2} because we could
-use the transisitive subsumption rule to derive that
+use the transitive subsumption rule to derive that
 \[
@@
 For closed tables, subtyping handles mutable fields invariantly,
-and imutable fields covariantly.
+and immutable fields covariantly.
 The following rules define this behavior.
@@
 \end{align*}
-Second, they do not allow programmers to use the table construtor to
+Second, they do not allow programmers to use the table constructor to
 initialize table types that represent hashes.
@@
 table types to omit fields when $\Nil$ is subtype of the type of
-the value of the ommited key.
+the value of the omitted key.
 The following subtyping rule show this behavior:
@@
 {...}_{e} \; | \;
-n \; | \;
+n_{e} \; | \;
 e_{1}[e_{2}] \; | \;
@@
 e{:}n(el)_{e}\\
-l ::= & \;\; n \; | \;
+l ::= & \;\; n_{l} \; | \;
 e_{1}[e_{2}] \; | \;
@@
 $s$ are statements, $e$ are expressions, $l$ are left-hand values,
-$el$ are expression lists, $c$ are constructor fields, $cl$ are lists of
-constructor fields, $me$ are expressions with multiple results,
-$f$ are function declarations, $m$ are method declarations, $n$ are variable names,
+$el$ are expression lists, $c$ are literal constants, $p$ are expression pairs,
+$me$ are expressions with multiple results, $f$ are function declarations,
+$m$ are method declarations, $n$ are variable names,
 $t$ are first-level types, and $r$ are second-level types.
@@
 
+We also included two kinds of type cast in our core language:
+the expression ${<}t{>} \;n$ and the left-hand value $n[c] \; {<}t{>}$.
+We will show that the first helps allowing safe aliasing between
+table types, while the second allows the refinement of table types.
+Due to the refinement of table types we also split variable names
+into two categories: $n_{e}$ when they appear as expressions and
+$n_{l}$ when they appear as left-hand values.
+
 We will present the typing rules as a deduction system for the typing
 relations $\env_{1} \vdash s : \env_{2}$ and $\env_{1} \vdash e : t, \env_{2}$. 
-We will use the first one for typing statements, and the sencond one
+We will use the first one for typing statements, and the second one
 for typing expressions.
@@
 typing the statement $s$ produces the new type environment $\env_{2}$. 
-The sencond relation means that given the type environment $\env_{1}$,
+The second relation means that given the type environment $\env_{1}$,
 the expression $e$ has type $t$, and produces the new type environment $\env_{2}$.
@@
 variable $n$ that maps to type $t$, and this mapping does not affect
-the other mappings that extisted before.
+the other mappings that existed before.
 
@@
 The rule \textsc{T-ASSINGMENT} shows that tuple types make it easy
-to use the consistent-subtyping relation to check multiple assignments:
+to use the consistent-subtyping relation to type check multiple assignments:
 \[
@@
 The next example does not type check because it tries to add
-the same field to $a$:
+the same field to $a$, but with different types:
 \begin{center}
@@
 the table constructor: the empty table $\{\}$.
-Its inference rule is starightforward:
+Its inference rule is straightforward:
 \[
@@
 For instance, we can write the table constructor
-$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to mean the table constructor
-$\{ ``x", ``y", ``z" \}$.
+$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to express $\{ ``x", ``y", ``z" \}$.
 We chose to not include this form because its inference rule is similar
 to the inference rule of the first form, and just controls the type of
-the ommited indices.
+the omitted indices.
 For the same reason we did not include table constructors with
-multiple expressions that can return a tuple type that does not end
+multiple expressions, which can return a tuple type that does not end
 with a variadic type.
@@
 The rules \textsc{T-PAIR1}, \textsc{T-PAIR2}, and \textsc{T-PAIR3}
-check if the type of the key is a $\Boolean$, a $\Number$,
-or a $\String$, respectively:
+check if the type of the key is a subtype of $\Boolean$, $\Number$,
+or $\String$, respectively:
 \[
@@
 \mylabel{T-IDREAD}\\
-\dfrac{\env_{1}(n_{r}) = t}
-      {\env_{1} \vdash n_{r}:close(t), \env_{1}}
+\dfrac{\env_{1}(n) = t}
+      {\env_{1} \vdash n_{e}:close(t), \env_{1}}
 \end{array}
@@
 & \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String \}_{unique} = \{ [``x"] = ``foo" \} \; \mathbf{in}$}\\
-& & \multicolumn{1}{l}{$a = b \;;\; a[``x"] = \mathbf{nil}$}
+& & \multicolumn{1}{l}{$a = b \;;\; a[``x"] = \mathbf{nil} \;;\; b[``x"] = b[``x"] \;{..}\; ``bar"$}
 \end{tabular}
@@
 We cannot add the field $``x"$ to $b$ because its type is \emph{closed},
-and thus do not allow chaning the value that is stored in the field $``x"$
+and thus do not allow changing the value that is stored in the field $``x"$
 of local $a$.
@@
 
-\[
-\begin{array}{c}
-\mylabel{T-INVOKE1}\\
-\dfrac{\env_{1} \vdash e[n]:p_{1} \rightarrow r, \env_{2} \;\;\;
-       \env_{2} \vdash el:p_{2}, \env_{3} \;\;\;
-       \Self \times p_{2} \lesssim p_{1}}
-      {\env_{1} \vdash e{:}n(el)_{m}:r, \env_{3}}
-\end{array}
-\]
-
+The rule \textsc{T-METHOD3} allow our type system to type check the
+following example:
 \begin{center}
@@
 \multicolumn{4}{l}{$\mathbf{in}$}\\
-& \multicolumn{3}{l}{$\mathbf{fun} \; s{:}new (x:number, y:number):\Self$}\\
+& \multicolumn{3}{l}{$\mathbf{fun} \; s{:}new (x:\Number, y:\Number):\Self$}\\
 & & \multicolumn{2}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{closed} =$}\\
@@
 & & \multicolumn{2}{l}{$\mathbf{in} \; s[``x"] = x; \; s[``y"] = y; \; \mathbf{return} \; s$}\\
-; & \multicolumn{3}{l}{$\mathbf{fun} \; s{:}move (x:number, y:number):\Void$}\\
+; & \multicolumn{3}{l}{$\mathbf{fun} \; s{:}move (x:\Number, y:\Number):\Void$}\\
 & & \multicolumn{2}{l}{$self[``x"] = self[``x"] + x;$}\\
@@
 \end{center}
+
+This example uses the type $\{``x":\Number, ``y":\Number\}_{open}$ to
+initialize the local variable $s$, which represents the type of the class
+that we are defining.
+Then, we use two method definitions to include $new$ and $move$ into
+our class that now has type
+\begin{align*}
+\{ & ``x":\Number, ``y":\Number,\\
+   & ``new":\Self \times \Number \times \Number \rightarrow \Self,\\
+   & ``move":\Self \times \Number \times \Number \rightarrow \Void \}_{open}
+\end{align*}
+
+Inside the definition of the method \emph{new} we use \emph{setmetatable}
+to initialize the local $s$ with the type of $\Self$.
+The rule \textsc{T-SETMETATABLE} express this idea:
+\[
+\begin{array}{c}
+\mylabel{T-SETMETATABLE}\\
+\dfrac{\env_{1} \vdash e:t, \env_{2} \;\;\;
+       t \subtype \{\}_{open}}
+      {\env_{1} \vdash setmetatable(\{\}, \{[``\string_\string_index"] = e\}):close(t), \env_{2}}
+\end{array}
+\]
+
+After we define our class, we can use it to create object instances
+of this class and call its methods.
+The next example assumes the object $o$ is in the environment as
+has the type of the class we just defined:
+\begin{center}
+\begin{tabular}{l}
+$o{:}move(10, 10)_{s}$
+\end{tabular}
+\end{center}
+
+This method call type checks through the rule \textsc{T-INVOKESTM1}.
+This rule uses the inference rule \textsc{T-INVOKE1} and discards the results.
+We defined the rule \textsc{T-INVOKE1} as follows:
+\[
+\begin{array}{c}
+\mylabel{T-INVOKE1}\\
+\dfrac{\env_{1} \vdash e:t, \env_{2} \;\;\;
+       \env_{2} \vdash e[n]:p_{1} \rightarrow r, \env_{3} \;\;\;
+       \env_{3} \vdash el:p_{2}, \env_{4} \;\;\;
+       \Self \times p_{2} \lesssim p_{1}}
+      {\env_{1} \vdash e{:}n(el)_{m}:[\Self \mapsto t]r, \env_{4}}
+\end{array}
+\]
+
+Multiple assignments and multiple return values can also appear in
+function applications, as the following example shows:
+\begin{center}
+\begin{tabular}{llll}
+\multicolumn{4}{l}{$\mathbf{local} \; m:\Void \rightarrow \Integer \times \String =$} \\
+& \multicolumn{3}{l}{$\mathbf{fun} \; ():\Integer \times \String$} \\
+& & \multicolumn{2}{l}{$\mathbf{return} \; 2, ``foo"$} \\
+\multicolumn{4}{l}{$\mathbf{in}$} \\
+& \multicolumn{3}{l}{$\mathbf{local} \; s:\Integer \times \Integer \rightarrow \Integer =$} \\
+& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer, y:\Integer):\Integer$} \\
+& & & \multicolumn{1}{l}{$\mathbf{return} \; x + y$} \\
+& \multicolumn{3}{l}{$\mathbf{in} \; s(m()_{e}, m()_{m})_{s}$}
+\end{tabular}
+\end{center}
+
+This example does not type check because we are trying to call
+$s$ with $\Integer \times \Integer \times \String$,
+but its input parameter has type $\Integer \times \Integer$.
+Lua drops extra values in function calls, and Typed Lua can also
+have this behavior.
+If we change the input type of $s$ to $\Integer \times \Integer \times \Value{*}$,
+this example type checks.
+Typed Lua also has to adjust function calls to use $\Nil$ in the
+place of missing parameters, otherwise it would not be able to
+type check function calls that have optional parameters.
+
+\textit{/* I guess we should have different typing rules for strict and default. */}
+
+Typed Lua includes four typing rules to handle the $\mathbf{or}$
+logical operator and its common idioms:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-OR1}\\
+\dfrac{\env_{1} \vdash e_{1}:t, \env_{2} \;\;\;
+       \Nil \not\lesssim t \;\;\;
+       \False \not\lesssim t}
+      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-OR2}\\
+\dfrac{\env_{1} \vdash e_{1}:\Nil, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t, \env_{3}}
+      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{T-OR3}\\
+\dfrac{\env_{1} \vdash e_{1}:\False, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t, \env_{3}}
+      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-OR4}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3}}
+      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(filter(t_{1}, \Nil), \False) \cup t_{2}, \env_{3}}
+\end{array}
+\end{array}
+\]
+
+The rule \textsc{T-OR1} is the rule that implements the short circuit.
+We use the consistent-subtyping relation in this rule to guarantee that
+the type system checks the second expression when the first one
+has the dynamic type, as it can be hiding a false value.
+
+The rules \textsc{T-OR2} and \textsc{T-OR3} guarantee that the final
+result is the type of the second expression, because the first one
+is certainly a false value.
+
+The rule \textsc{T-OR4} is the most general rule, but it is also
+the rule that handles the common $\mathbf{or}$ idioms.
+It uses the predicate \emph{filter} to filter possible false values
+that might be part of the type of the first expression.
+We can use pattern matching to the define the recursive predicate
+\emph{filter} as follows:
+\begin{align*}
+filter(t_{1} \cup t_{2}, t_{1}) & = filter(t_{2}, t_{1})\\
+filter(t_{1} \cup t_{2}, t_{2}) & = filter(t_{1}, t_{2})\\
+filter(t_{1} \cup t_{2}, t_{3}) & = filter(t_{1}, t_{3}) \cup filter(t_{2}, t_{3})\\
+filter(t_{1}, t_{2}) & = t_{1}
+\end{align*}
+
+Using these typing rules our type system can type check the following
+example:
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; x:\String \cup \Nil = \mathbf{nothing} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$\mathbf{local} \; y:\String = x \; \mathbf{or} \; ``Hello"$}
+\end{tabular}
+\end{center}
+
+Without the \emph{filter} predicate,
+the expression $x \; \mathbf{or} \; ``Hello"$ would have type
+$\String \cup \Nil \cup ``Hello"$.
+The \emph{filter} predicate removes the type $\Nil$ from the result,
+leaving type $\String \cup ``Hello"$.
+At the end of the evaluation, the expression has type $\String$
+because unions are disjoint and $``Hello" \subtype \String$.
+
+\textit{/* The following example does not type check */}
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; x:\Nil \cup \False = \mathbf{false} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$\mathbf{local} \; y:\String = x \; \mathbf{or} \; ``Hello"$}
+\end{tabular}
+\end{center}
+
+\textit{/* Should we define the following rules? */}
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-OR4}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+       \Nil \cup \False \subtype t_{1} \;\;\;
+       t_{1} \subtype \Nil \cup \False}
+      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t_{2}, \env_{3}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-OR5}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+       \Nil \subtype t_{1} \;\;\;
+       \False \subtype t_{1}}
+      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(filter(t_{1}, \Nil), \False) \cup t_{2}, \env_{3}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-OR6}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+       \Nil \subtype t_{1} \;\;\;
+       \False \not\subtype t_{1}}
+      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(t_{1}, \Nil) \cup t_{2}, \env_{3}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-OR7}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+       \Nil \not\subtype t_{1} \;\;\;
+       \False \subtype t_{1}}
+      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(t_{1}, \False) \cup t_{2}, \env_{3}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-OR8}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+       \Nil \not\subtype t_{1} \;\;\;
+       \False \not\subtype t_{1}}
+      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t_{1} \cup t_{2}, \env_{3}}
+\end{array}
+\end{array}
+\]
+
+\textit{/* Or who writes this kind of code deserves a type error? */}
+
+Another common idiom that programmers use in Lua is to overload
+the input parameter of functions, and use the function \texttt{type}
+to execute different actions according to their types.
+\[
+\begin{array}{c}
+\mylabel{T-IF2}\\
+\dfrac{\begin{array}{c}
+       \env_{1}(n) = t\\
+       closeall(\env_{1}[n \mapsto \String]) \vdash s_{1}:\env_{2} \\
+       closeall(\env_{1}[n \mapsto filter(t, \String)) \vdash s_{2}:\env_{3}
+      \end{array}}
+      {\env_{1} \vdash \mathbf{if} \; type(n) == ``string" \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}:closeset(\env_{1}[n \mapsto t], fav(s_{1}) \cup fav(s_{2}))}
+
+\end{array}
+\]
+
+Using the rule \textsc{T-IF2}, Typed Lua can type check the following example:
+\begin{center}
+\begin{tabular}{llll}
+\multicolumn{4}{l}{$\mathbf{local} \; o:\String \times \String \cup \Integer \rightarrow \String =$}\\
+& \multicolumn{3}{l}{$\mathbf{fun} \; (a:\String, b:\String \cup \Integer):\String$}\\
+& & \multicolumn{2}{l}{$\mathbf{local} \; r:\String = ``" \; \mathbf{in}$}\\
+& & & \multicolumn{1}{l}{$\mathbf{if} \; type(b) == ``string" \; \mathbf{then} \; r = a \;{..}\;b \; \mathbf{else} \; r = rep(a, b)_{e}\; ;$}\\
+& & & \multicolumn{1}{l}{$\mathbf{return} \; r$}\\
+\multicolumn{4}{l}{$\mathbf{in} \; o(``foo", 2)_{s}$}
+\end{tabular}
+\end{center}
+
+We are assuming that the function \emph{rep} is in the environment and has type
+$\String \times \Integer \times \String \cup \Nil \rightarrow \String$.
+
+Typed Lua also includes similar rules to handle the tags \texttt{nil},
+\texttt{boolean}, and \texttt{number}.
+There is also a similar rule for handling the type $\Integer$, but
+it works only with Lua 5.3, as it depends on the function \texttt{math.type}.
+This function appears only in Lua 5.3 and it returns the string $``integer"$
+when its input parameter is a number that has an integer representation,
+the string $``float"$ when its input parameter is a number that has a
+floating point representation, or $\Nil$ otherwise.
+
+Lua programmers also overload the return type of functions to denote
+errors, and Typed Lua also handles this case with the rule \textsc{T-IF3}:
+\[
+\begin{array}{c}
+\mylabel{T-IF3}\\
+\dfrac{\begin{array}{c}
+       \env_{1}(n) = \pi_{i}\\
+       closeall(\env_{1}[\pi \mapsto fpt(\env_{1}(\pi), \Nil, i)]) \vdash s_{1}:\env_{2} \\
+       closeall(\env_{1}[\pi \mapsto gpt(\env_{1}(\pi), \Nil, i)) \vdash s_{2}:\env_{3}
+      \end{array}}
+      {\env_{1} \vdash \mathbf{if} \; n \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}:closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2}))}
+
+\end{array}
+\]
+
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; q:\pi_{1}, r:\pi_{2} = idiv(1, 2)_{m} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; print(q + r)_{s} \; \mathbf{else} \; print(``ERROR: " \; .. \; r)_{s}$}
+\end{tabular}
+\end{center}
+
+\textit{/* How should we write the typing rule of the local declaration? */}
+
+\textit{/* How this is valid if projection types do not have subtyping rules? */}
+
+We are assuming that functions \emph{idiv} and \emph{print} are in
+the environment with the respective types
+$\Integer \times \Integer \rightarrow \Integer \times \Integer \sqcup \Nil \times \String$
+and
+$\Value{*} \rightarrow \Void$.
 
@@
 
-Multiple assignments and multiple return values can also appear in
-function applications, as the following example shows:
-\begin{center}
-\begin{tabular}{llll}
-\multicolumn{4}{l}{$\mathbf{local} \; m:\Void \rightarrow \Integer \times \String =$} \\
-& \multicolumn{3}{l}{$\mathbf{fun} \; ():\Integer \times \String$} \\
-& & \multicolumn{2}{l}{$\mathbf{return} \; 2, ``foo"$} \\
-\multicolumn{4}{l}{$\mathbf{in}$} \\
-& \multicolumn{3}{l}{$\mathbf{local} \; s:\Integer \times \Integer \rightarrow \Integer =$} \\
-& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer, y:\Integer):\Integer$} \\
-& & & \multicolumn{1}{l}{$\mathbf{return} \; x + y$} \\
-& \multicolumn{3}{l}{$\mathbf{in} \; s(m()_{e}, m()_{m})_{s}$}
-\end{tabular}
-\end{center}
-
-Typed Lua includes four typing rules to handle the $\mathbf{or}$
-logical operator and its common idioms:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-OR1}\\
-\dfrac{\env_{1} \vdash e_{1}:t, \env_{2} \;\;\;
-       \Nil \not\lesssim t \;\;\;
-       \False \not\lesssim t}
-      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-OR2}\\
-\dfrac{\env_{1} \vdash e_{1}:\Nil, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t, \env_{3}}
-      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{T-OR3}\\
-\dfrac{\env_{1} \vdash e_{1}:\False, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t, \env_{3}}
-      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-OR4}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3}}
-      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(filter(t_{1}, \Nil), \False) \cup t_{2}, \env_{3}}
-\end{array}
-\end{array}
-\]
-
-The rule \textsc{T-OR1} is the rule that implements the short circuit.
-We use the consistent-subtyping in this relation to guarantee that
-the type system avaliates the second expression when the first one
-has the dynamic type, as it can be hiding a false value.
-
-The rules \textsc{T-OR2} and \textsc{T-OR3} guarantee that the final
-result is the type of the second expression, because the first one
-is certainly a false value.
-
-The rule \textsc{T-OR4} is the most general rule, but it is also
-the rule that handles the common $\mathbf{or}$ idioms.
-It uses the predicate \emph{filter} to filter possible false values
-that might be part of the type of the first expression.
-We can use pattern matching to the define the recursive predicate
-\emph{filter} as follows:
-\begin{align*}
-fu(t_{1} \cup t_{2}, t_{3}) & = fu(t_{1}, t_{3}) \cup fu(t_{2}, t_{3})\\
-fu(t_{1} \cup t_{2}, t_{1}) & = fu(t_{2}, t_{1})\\
-fu(t_{1} \cup t_{2}, t_{2}) & = fu(t_{1}, t_{2})\\
-fu(t_{1}, t_{2}) & = t_{1}
-\end{align*}
-
-Using these typing rules our type system can type check the following
-example:
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; x:\String \cup \Nil = \mathbf{nothing} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$\mathbf{local} \; y:\String = x \; \mathbf{or} \; ``Hello"$}
-\end{tabular}
-\end{center}
-
-Without the \emph{filter} predicate,
-the expression $x \; \mathbf{or} \; ``Hello"$ would have type
-$\String \cup \Nil \cup ``Hello"$.
-The \emph{filter} predicate removes the type $\Nil$ from the result,
-leaving type $\String \cup ``Hello"$.
-At the end of the evaluation, the expression has type $\String$
-because unions are disjoint and $``Hello" \subtype \String$.
-
-\textit{/* The following example does not type check */}
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; x:\Nil \cup \False = \mathbf{false} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$\mathbf{local} \; y:\String = x \; \mathbf{or} \; ``Hello"$}
-\end{tabular}
-\end{center}
-
-\textit{/* Should we define the following rules? */}
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-OR4}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
-       \Nil \cup \False \subtype t_{1} \;\;\;
-       t_{1} \subtype \Nil \cup \False}
-      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t_{2}, \env_{3}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-OR5}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
-       \Nil \subtype t_{1} \;\;\;
-       \False \subtype t_{1}}
-      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(filter(t_{1}, \Nil), \False) \cup t_{2}, \env_{3}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-OR6}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
-       \Nil \subtype t_{1} \;\;\;
-       \False \not\subtype t_{1}}
-      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(t_{1}, \Nil) \cup t_{2}, \env_{3}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-OR7}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
-       \Nil \not\subtype t_{1} \;\;\;
-       \False \subtype t_{1}}
-      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(t_{1}, \False) \cup t_{2}, \env_{3}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-OR8}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
-       \Nil \not\subtype t_{1} \;\;\;
-       \False \not\subtype t_{1}}
-      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t_{1} \cup t_{2}, \env_{3}}
-\end{array}
-\end{array}
-\]
-
-\textit{/* Or who writes this kind of code deserves a type error? */}
-
-Another common idiom that programmers use in Lua is to overload
-the input parameter of functions, and use the function \texttt{type}
-to execute different actions according to their types.
-\[
-\begin{array}{c}
-\mylabel{T-IF2}\\
-\dfrac{\begin{array}{c}
-       closeall(\env_{1}[n \mapsto \String]) \vdash s_{1}:\env_{2} \\
-       closeall(\env_{1}[n \mapsto filter(\env_{1}(n), \String)) \vdash s_{2}:\env_{3}
-      \end{array}}
-      {\env_{1} \vdash \mathbf{if} \; type(n) == ``string" \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}:closeset(\env_{2}, fav(s_{1}) \cup fav(s_{2}))}
-
-\end{array}
-\]
-
-Using the rule \textsc{T-IF2}, Typed Lua can type check the following example:
-\begin{center}
-\begin{tabular}{llll}
-\multicolumn{4}{l}{$\mathbf{local} \; o:\String \times \String \cup \Integer \rightarrow \String =$}\\
-& \multicolumn{3}{l}{$\mathbf{fun} \; (a:\String, b:\String \cup \Integer):\String$}\\
-& & \multicolumn{2}{l}{$\mathbf{local} \; r:\String = ``" \; \mathbf{in}$}\\
-& & & \multicolumn{1}{l}{$\mathbf{if} \; type(b) == ``string" \; \mathbf{then} \; r = a \;{..}\;b \; \mathbf{else} \; r = rep(a, b)_{e}\; ;$}\\
-& & & \multicolumn{1}{l}{$\mathbf{return} \; r$}\\
-\multicolumn{4}{l}{$\mathbf{in} \; o(``foo", 2)_{s}$}
-\end{tabular}
-\end{center}
-
-We are assuming that the function \emph{rep} is in the environment and has type
-$\String \times \Integer \times \String \cup \Nil \rightarrow \String$.
-
-Typed Lua also includes similar rules to handle the tags \texttt{nil},
-\texttt{boolean}, and \texttt{number}.
-
-Overloading on the return type.
-\[
-\begin{array}{c}
-\mylabel{T-IF3}\\
-\dfrac{\begin{array}{c}
-       \env_{1}(n) = \pi_{i}\\
-       closeall(\env_{1}[\pi \mapsto fpt(\env_{1}(\pi), \Nil, i)]) \vdash s_{1}:\env_{2} \\
-       closeall(\env_{1}[\pi \mapsto gpt(\env_{1}(\pi), \Nil, i)) \vdash s_{2}:\env_{3}
-      \end{array}}
-      {\env_{1} \vdash \mathbf{if} \; n \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}:closeset(\env_{2}, fav(s_{1}) \cup fav(s_{2}))}
-
-\end{array}
-\]
-
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; q:\pi_{1}, r:\pi_{2} = idiv(1, 2)_{m} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; print(q + r)_{s} \; \mathbf{else} \; print(``ERROR: " \; .. \; r)_{s}$}
-\end{tabular}
-\end{center}
-
-\textit{/* How should we write the typing rule of the local declaration? */}
-
-\textit{/* How this is valid if projection types do not have subtyping rules? */}
-
-We are assuming that functions \emph{idiv} and \emph{print} are in
-the environment with the respective types
-$\Integer \times \Integer \rightarrow \Integer \times \Integer \sqcup \Nil \times \String$
-and
-$\Value{*} \rightarrow \Void$.
+Although this last example is not safe, it shows that optional
+type systems still preserve the flexibility of dynamically
+typed languages along with the benefits of static type checking.

### 2014-11-28 doc/thesis/evaluation.tex
Splitting related work into Lua and Non-Lua

@@
 
+We detected some bugs while doing the bootstrap:
+accessing undeclared global variables.
+
 \section{Related Work}
 \label{sec:relatedwork}
+
+\subsection{Lua Related Work}
+
+Metalua \citep{metalua} is a Lua compiler that supports compile-time
+metaprogramming (CTMP).
+CTMP is a kind of macro system that allows the programmers to interact
+with the compiler \citep{fleutot2007contrasting}. 
+Metalua extends Lua 5.1 syntax to include its macro system,
+and allows programmers to define their own syntax.
+Metalua can provide syntactical support for several object-oriented
+styles, and can also provide syntax for turning simple type
+annotations into run-time assertions.
+
+MoonScript \citep{moonscript} is a programming language that supports
+class-based object-oriented programming, and compiles to idiomatic
+Lua code.
+However, it does not perform compile-time type checking.
+
+LuaInspect \citep{luainspect} is a tool that uses MetaLua to perform
+some code analysis, such as flagging unknown global variables and
+table fields, checking the number of function arguments against
+signatures, and inferring function return values, but it does not
+try to analyze object-oriented code and does not perform compile-time
+type checking.
+
+Tidal Lock \citep{tidallock} is a prototype of another optional type
+system for Lua, which is written in Metalua.
+A remarkable feature of Tidal Lock is the way that it handles tables,
+because it allows the programmer to incrementally build the type of
+a table.
+More precisely, Tidal Lock allows the programmer to create an empty
+table to build its type according to the keys that the programmer
+inserts in it, so the programmer does not need to build the table in
+a single table constructor.
+This feature inspired us to introduce it into Typed Lua,
+but in a simpler way and with different formalization.
+Like Typed Lua, Tidal Lock also has local type inference.
+Unlike Typed Lua, Tidal Lock does not type the common object-oriented
+idioms used by Lua programmers, though it also uses subtyping.
 
@@
 It also does not support any object-oriented idiom.
+
+\subsection{Non-Lua Related Work}
 

### 2014-11-28 doc/thesis/typedlua.tex
Splitting related work into Lua and Non-Lua

@@
 typing of the Lua standard libraries inside Typed Lua.
-
-\section{Related Lua projects}
-
-Metalua \citep{metalua} is a Lua compiler that supports compile-time
-metaprogramming (CTMP).
-CTMP is a kind of macro system that allows the programmers to interact
-with the compiler \citep{fleutot2007contrasting}. 
-Metalua extends Lua 5.1 syntax to include its macro system,
-and allows programmers to define their own syntax.
-Metalua can provide syntactical support for several object-oriented
-styles, and can also provide syntax for turning simple type
-annotations into run-time assertions.
-
-MoonScript \citep{moonscript} is a programming language that supports
-class-based object-oriented programming, and compiles to idiomatic
-Lua code.
-However, it does not perform compile-time type checking.
-
-LuaInspect \citep{luainspect} is a tool that uses MetaLua to perform
-some code analysis, such as flagging unknown global variables and
-table fields, checking the number of function arguments against
-signatures, and inferring function return values, but it does not
-try to analyze object-oriented code and does not perform compile-time
-type checking.
-
-Tidal Lock \citep{tidallock} is a prototype of another optional type
-system for Lua, which is written in Metalua.
-A remarkable feature of Tidal Lock is the way that it handles tables,
-because it allows the programmer to incrementally build the type of
-a table.
-More precisely, Tidal Lock allows the programmer to create an empty
-table to build its type according to the keys that the programmer
-inserts in it, so the programmer does not need to build the table in
-a single table constructor.
-This feature inspired us to introduce it into Typed Lua,
-but in a simpler way and with different formalization.
-Like Typed Lua, Tidal Lock also has local type inference.
-Unlike Typed Lua, Tidal Lock does not type the common object-oriented
-idioms used by Lua programmers, though it also uses subtyping.

### 2014-12-06 doc/thesis/system.tex
Improving the text about subtyping

@@
 
-In this chapter we present the formalization of Typed Lua.
+In this chapter we present the formalization of Typed Lua's type system.
 Besides its practical contribution, Typed Lua also has some interesting
-contributions on the field of optional type systems for scripting
+contributions to the field of optional type systems for scripting
 languages.
-Namely, it is a simple type system that covers several Lua idioms
+They are novel type system features that let Typed Lua cover several Lua idioms
 and features, such as the refinement of tables, multiple return values,
 and optional parameters.
-We will use typing rules to explain how we implemented the features of Typed Lua.
 
@@
 & | \; \mu x.t & \textit{recursive types}\\
-& | \; \pi_{i} & \textit{projection types}\\
+& | \; \pi_{i}^{x} & \textit{projection types}\\
 l ::= & \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} & \\
@@
 \emph{first-level types} and \emph{second-level types}.
-Typed Lua uses first-level types to represent Lua values and
-second-level types to represent expression lists, and it also
-uses them to type multiple assignments and function applications.
+Typed Lua uses first-level types to represent first-class Lua values and
+second-level types to represent tuples of values that appear in 
+multiple assignments and function applications.
 First-level types include literal types, base types, the type $\Nil$,
@@
 Second-level types include the type $\Void$, tuples of first-level
-types optionally ending in a variadic type, and union of these tuples.
+types optionally ending in a variadic type, and unions of these tuples.
 
@@
 any integer value, any floating point value, and any string value.
-We shall see that literal types are important to represent table types
-as records.
+We shall see that literal types are important in our treatment of
+table types as records.
 
@@
 $\Integer$, $\Number$, and $\String$.
-The base types $\Boolean$ and $\String$ represent the values that
-during run-time Lua tags as \texttt{boolean} and \texttt{string},
-respectively;
-the base types $\Integer$ and $\Number$ represent the values
-that during run-time Lua tags as \texttt{number}.
-We introduced two different base types to represent Lua numbers
-because Lua introduced integers in version 5.3.
-
-The subtyping rules for literal types include a reflexive rule,
-and the rules for defining that literal types are subtypes of
-their respective base types;
-the subtyping rules for base types include a reflexive rule,
-and another rule for defining that $\Integer$ is subtype of
-$\Number$.
+The base types $\Boolean$ and $\String$ represent the values that during
+run-time Lua tags as \texttt{boolean} and \texttt{string}, respectively.
+Lua 5.3 introduced two internal representations to the tag \texttt{number}:
+\texttt{integer} for integer numbers and \texttt{float} for real numbers.
+Both \texttt{integer} and \texttt{float} values are subtype of \texttt{number}.
+We introduced the base type $\Number$ to represent either the tag
+\texttt{number} or \texttt{float} values, and the type $\Integer$ to represent
+\texttt{integer} values.
+
+The subtyping rules for literal types and base types include the rules
+for defining that literal types are subtypes of their respective base types,
+and that $\Integer$ is subtype of $\Number$.
 We defined these subtyping rules as follows:
@@
 \begin{array}{c}
-\begin{array}{c}
-\mylabel{S-LITERAL}\\
-\senv \vdash l \subtype l
-\end{array}
-\;
 \begin{array}{c}
@@
 \end{array}
+\;
+\begin{array}{c}
+\mylabel{S-STRING}\\
+\senv \vdash {\it string} \subtype \String
+\end{array}
 \\ \\
@@
 \\ \\
-\begin{array}{c}
-\mylabel{S-STRING}\\
-\senv \vdash {\it string} \subtype \String
-\end{array}
-\begin{array}{c}
-\mylabel{S-BASE}\\
-\senv \vdash b \subtype b
-\end{array}
-\;
 \begin{array}{c}
@@
 
-We could have omitted the rules \textsc{S-LITERAL}, \textsc{S-BASE},
-and \textsc{S-INT2}, because the first two are reflexive and we
-could achieve the third one through transitivity.
-More precisely, we could have defined subsumption rules that
-specify the reflexive and transitive behavior of subtyping,
-and use these rules instead of the ones we defined.
-For instance, we would not need \textsc{S-INT2} because we could
-use the transitive subsumption rule to derive that
+Subtyping is reflexive, thus, we include reflexive rules for
+first-level and second-level types as follows:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-FIRST}\\
+\senv \vdash t \subtype t
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-SECOND}\\
+\senv \vdash s \subtype s
+\end{array}
+\end{array}
+\]
+
+Subtyping is also transitive, and for this reason we could have omitted
+the rule \textsc{S-INT2}.
+More precisely, we could have defined a transitive rule for first-level
+types instead of defining specific rules to transitive cases.
+For instance, a transitive rule would allow us to derive that
 \[
@@
 
-The type $\Nil$ represents the type of values that during run-time
+The type $\Nil$ represents the type of the value that during run-time
 Lua tags as \texttt{nil}.
-It is a separated type because in our type system $\Nil$ is
-not the bottom type.
+In our type system $\Nil$ is not the bottom type, and it is subtype
+only of itself through the reflexive rule.
 Typed Lua has this behavior because in Lua the tag \texttt{nil}
 represents the absence of an useful value.
-Thus, making $\Nil$ the bottom type would severally reduce the
-amount of type safety in our type system.
-The next subtyping rule shows this behavior:
-\[
-\begin{array}{c}
-\mylabel{S-NIL}\\
-\senv \vdash \Nil \subtype \Nil
-\end{array}
-\]
-
-The type $\Value$ is the top type, so any first-level type is
+Thus, making $\Nil$ the bottom type would lead to several expressions
+that would pass the type checker, but that would fail during run-time
+in the presence of a \texttt{nil} value.
+
+The type $\Value$ is the top type, so any first-level type is a
 subtype of $\Value$.
-We shall see that this type along with variadic types help to
-drop extra values on multiple assignments, and function calls,
-thus preserving the semantics of Lua in these cases.
+In the next section we will show that this type along with
+variadic types help the type system drop extra values on
+multiple assignments and function calls, thus preserving the
+semantics of Lua in these cases.
 Following we show its subtyping rule:
@@
 
-The dynamic type $\Any$ is the type that allows programmers to mix
-static and dynamic typing.
-The type $\Any$ is neither the bottom nor the top type, but a
-separated typed that is subtype only of itself:
-\[
-\begin{array}{c}
-\mylabel{S-ANY}\\
-\senv \vdash \Any \subtype \Any
-\end{array}
-\]
+The dynamic type $\Any$ allows programmers to mix static and dynamic typing;
+it is neither the bottom nor the top type,
+but a separate type that is subtype only of itself through the reflexive rule.
 
@@
 differ between subtyping and consistent-subtyping.
-If we set the type $\Any$ as both bottom and top types of our
+If we had set the type $\Any$ as both bottom and top types of our
 subtyping relation, then all programs would type check without errors.
@@
 $\Any$ to any type.
-While the rule \textsc{S-ANY} is neutral and makes transitivity
-not be a problem, the rules \textsc{C-ANY1} and \textsc{C-ANY2}
-are the rules that allow the dynamic type to interact with other
-first-level types, and thus allow dynamically typed code coexist
-with statically typed code.
+The rules \textsc{C-ANY1} and \textsc{C-ANY2} are the rules that
+allow the dynamic type to interact with other first-level types,
+and thus allow dynamically typed code to coexist with statically
+typed code.
 Because of these two rules consistent-subtyping cannot be transitive.
 
-Typed Lua uses the type $\Self$ to represent \emph{objects}.
+Typed Lua uses the type $\Self$ to represent the \emph{receiver}
+in object-oriented method definitions and method calls.
 As we mentioned in Chapter \ref{chap:typedlua}, we need the type
 $\Self$ to prevent programmers from indexing a method without
-calling it with the correct receiver.
-The type $\Self$ is only a subtype of itself:
-\[
-\begin{array}{c}
-\mylabel{S-SELF}\\
-\senv \vdash \Self \subtype \Self
-\end{array}
-\]
+calling it with the correct \emph{receiver}.
+The type $\Self$ is subtype only of itself through the reflexive rule.
 
@@
 different, but fixed types.
-Next we show the subtyping rules for union types:
+The subtyping rules for union types are standard:
 \[
@@
 
-Union types are disjoint, that is, the types in the union are not
-consistent-subtype of any other type in the same union.
+Union types must be disjoint, that is, the types in the union are not
+consistent-subtype of each other.
+In the implementation of Typed Lua we use consitent-subtyping to
+normalize and simplify union types, but we let them free in the
+formalization.
 For instance, the union type $\Boolean \;\cup\; \Any$ results the
@@
 type $\Number \;\cup\; \Nil$ because $1 \lesssim \Number$.
-
-\emph{/* Do we need to show how Typed Lua simplifies union types? */}
 
@@
 expression.
-Next we show their subtyping rules:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-VARARG}\\
-\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2} \cup \Nil}
-      {\senv \vdash t_{1}* \subtype t_{2}*}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TUPLE1}\\
+The subtyping rule for tuple types is standard, but the subtyping
+rules for variadic types are not obvious:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-TUPLE}\\
 \dfrac{\senv \vdash t_{1} \subtype t_{2} \;\;\;
@@
 \begin{array}{c}
-\mylabel{S-TUPLE2}\\
-\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2} \;\;\;
-       \senv \vdash t_{1}* \subtype s_{2}}
-      {\senv \vdash t_{1}* \subtype t_{2} \times s_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TUPLE3}\\
-\dfrac{\senv \vdash t_{1} \subtype t_{2} \cup \Nil \;\;\;
-       \senv \vdash s_{1} \subtype t_{2}*}
-      {\senv \vdash t_{1} \times s_{1} \subtype t_{2}*}
-\end{array}
-\end{array}
-\]
-
-The subtyping rules for variadic and tuple types show that
-these types are covariant.
-The subtyping rule for variadic types is straightforward:
-it shows that a variadic type $t_{1}*$ is subtype of another
-variadic type $t_{2}*$ if $t_{1} \cup \Nil$ is subtype of
-$t_{2} \cup \Nil$.
-The first rule for tuple types shows that a tuple type
-$t_{1} \times s_{1}$ is subtype of another tuple type
-$t_{2} \times s_{2}$ if $t_{1}$ is subtype of $t_{2}$
-and $s_{1}$ is subtype of $s_{2}$.
-A tuple type can optionally end with a variadic type, and
-the rules \textsc{S-TUPLE2} and \textsc{S-TUPLE3} handle this case.
-The rule \textsc{S-TUPLE2} shows that a variadic type is subtype of
+\mylabel{S-VARARG1}\\
+\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2} \cup \Nil}
+      {\senv \vdash t_{1}{*} \subtype t_{2}{*}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-VARARG2}\\
+\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2}}
+      {\senv \vdash t_{1}{*} \subtype t_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-VARARG3}\\
+\dfrac{\senv \vdash t_{1} \subtype t_{2} \cup \Nil}
+      {\senv \vdash t_{1} \subtype t_{2}{*}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-VARARG4}\\
+\dfrac{\senv \vdash t_{1}{*} \subtype t_{2} \;\;\;
+       \senv \vdash t_{1}{*} \subtype s_{2}}
+      {\senv \vdash t_{1}{*} \subtype t_{2} \times s_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-VARARG5}\\
+\dfrac{\senv \vdash t_{1} \subtype t_{2}{*} \;\;\;
+       \senv \vdash s_{1} \subtype t_{2}{*}}
+      {\senv \vdash t_{1} \times s_{1} \subtype t_{2}{*}}
+\end{array}
+\end{array}
+\]
+
+The subtyping rules for tuple and variadic types show that they are covariant.
+While we need just one subtyping rule for tuple types,
+we need five different subtyping rules for variadic types
+to handle all the cases that they can appear.
+The rule \textsc{S-TUPLE1} shows that a tuple type $t_{1} \times s_{1}$
+is subtype of another tuple type $t_{2} \times s_{2}$ if
+$t_{1}$ is subtype of $t_{2}$ and $s_{1}$ is subtype of $s_{2}$.
+The rule \textsc{S-VARARG1} handles the subtyping between two variadic types.
+It shows that $t_{1}{*}$ is subtype of $t_{2}{*}$
+if $t_{1} \cup \Nil$ is subtype of $t_{2} \cup \Nil$.
+This rule explicit includes $\Nil$ in both sides because otherwise
+$\Nil{*}$ would not be a subtype of several other variadic types.
+For instance, $\Nil{*}$ would not be subtype of $\Number{*}$,
+as $\Nil \not\subtype \Number$.
+The rules \textsc{S-VARARG2} and \textsc{S-VARARG3} handles the
+subtyping between a firt-level type and a variadic type.
+The rule \textsc{S-VARARG2} shows that a variadic type $t_{1}{*}$ is
+subtype of a firt-level type $t_{2}$ if $t_{1} \cup \Nil$ is subtype of $t_{2}$.
+The rule \textsc{S-VARARG3} shows that a firt-level type $t_{1}$
+is subtype of a variadic type $t_{2}{*}$ if $t_{1}$ is subtype of $t_{2} \cup \Nil$.
+The rules \textsc{S-VARARG4} and \textsc{S-VARARG5} handle the
+subtyping between a tuple type and a variadic type.
+The rule \textsc{S-VARARG4} shows that a variadic type is subtype of
 a tuple type whenever the type of the value that it generates is
 subtype of all the elements of the tuple.
-The rule \textsc{S-TUPLE3} shows that a tuple type is subtype of a
+The rule \textsc{S-VARARG5} shows that a tuple type is subtype of a
 variadic type whenever all the elements of the tuple are subtype of
 the type of the value that it generates.
-We shall see that these subtyping relations along with the
-types $\Value$ and $\Nil$ allow our type system to reflect
+In the next section we will show that we use these subtyping relations
+along with the types $\Value$ and $\Nil$ to make our type system reflect
 the semantics of Lua on discarding extra parameters and
@@
 
-In our type system function types have the form $p \rightarrow r$,
-where $p$ is the type of the parameters list and $r$ is the
-return type.
-The type of the parameters list can be $\Void$ or a tuple type,
+Function types have the form $p \rightarrow r$, where
+$p$ is the type of the parameter list and $r$ is the return type.
+The type of the parameter list can be $\Void$ or a tuple type,
 while the return type can be $\Void$, a tuple type, or a union of
@@
 The type $\Void$ means an empty tuple.
-Our type system includes union of tuples on the return type because
+Our type system includes unions of tuples on the return type because
 in Lua it is common to overload the return type to denote error,
@@
 The type $\Void$ and union of tuple types are also second-level types.
-The type $\Void$ is subtype only of itself.
-The subtyping rules for union of tuple types are similar to the
-subtyping rules for union of first-level types.
-Next we show the subtyping rules for these second-level types:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-VOID}\\
-\senv \vdash \Void \subtype \Void
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-UNION4}\\
-\dfrac{\senv \vdash s_{1} \subtype s \;\;\;
-       \senv \vdash s_{2} \subtype s}
-      {\senv \vdash s_{1} \sqcup s_{2} \subtype s}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-UNION5}\\
-\dfrac{\senv \vdash s \subtype s_{1}}
-      {\senv \vdash s \subtype s_{1} \sqcup s_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-UNION6}\\
-\dfrac{\senv \vdash s \subtype s_{2}}
-      {\senv \vdash s \subtype s_{1} \sqcup s_{2}}
-\end{array}
-\end{array}
-\]
+The type $\Void$ is subtype only of itself through the reflexive rule.
+The subtyping rules for unions of tuple types are similar to the
+subtyping rules for unions of first-level types.
 
 The subtyping rule for function types is standard, that is, it is
-contravariant on the type of the parameters list and
+contravariant on the type of the parameter list and
 covariant on the return type:
@@
 
-Our type system includes table types to represent Lua tables.
+Table types represent the various forms that Lua tables can take.
 The syntactical form of table types is $\{ k_{i}{:}v_{i}, ..., k_{n}{:}v_{n} \}_{tag}$.
@@
 the type of a table key, and each $v_{i}$ represents the type of the
-value that the table key $k_{i}$ maps to.
-We restricted the type of keys to literal types, base types, and the dynamic type,
-but we can use any first-level type to represent the type of values they map to.
+value that table keys of type $k_{i}$ map to.
+Key types can only be literal types, base types, and the dynamic type.
 We made this restriction to the type of the keys because the statistics
@@
 of the tables are records, lists, and hashes.
-The dynamic type is an option when we need a more loosen table type.
+The dynamic type is an option when we need a loose table type.
+Value types can be any first-level type, and can optionally include
+the $\Const$ attribute to denote immutable values.
 
 We also use the tags \emph{closed}, \emph{open}, and \emph{unique}
-to classify table types, though they have only one syntactical form.
+to classify table types.
 The first tag represents the table types that the type annotations,
@@
 \]
-
-\emph{/* Can we use this well formed rule to allow $k$ to be $t$? */}
 

### 2014-12-07 doc/thesis/system.tex
Improvements to the text about table types

@@
 k ::= & \;\; l \; | \; b \; | \; \Any & \textit{key types}\\
-v ::= & \;\; t \; | \; \Const \; t & \textit{field types}\\ 
+v ::= & \;\; t \; | \; \Const \; t & \textit{value types}\\ 
 \multicolumn{3}{c}{}\\
@@
 tags.
-We will discuss these subtyping rules in the next paragraphs to
-explain why we needed different tags.
-
-The following subtyping rule shows that \emph{closed} table types
-have width subtyping, and depth subtyping on $\Const$ fields;
-it resembles the subtyping rule for record types.
-
+
+The subtyping rule between \emph{closed} table types resembles the
+standard subtyping rule between records:
 \[
@@
        \senv \vdash k_{i}' \subtype k_{j} \;\;\;
-       \senv \vdash v_{j} \subtype_{c} v_{i}'}
+       \senv \vdash v_{j} \subtype v_{i}'}
       {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{closed} \subtype \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
@@
 
-The rule \textsc{S-TABLE1} introduced the relation $\subtype_{c}$.
-It is just an auxiliary relation that defines the subtyping of the
-type of the values stored in table fields.
-We needed an auxiliary relation because the subtyping of the type
-of the values stored in the table fields change according to the
-tag of the table type.
-For closed tables, subtyping handles mutable fields invariantly,
-and immutable fields covariantly.
-The following rules define this behavior.
+The rule \textsc{S-TABLE1} allows \emph{closed} table types to
+have width subtyping, and depth subtyping on $\Const$ fields.
+The subtyping rules for the type of the values stored in a \emph{closed}
+table type describe this behavior:
 \[
@@
        \senv \vdash v_{2} \subtype v_{1}}
-      {\senv \vdash v_{1} \subtype_{c} v_{2}}
-\end{array}
-\\ \\
+      {\senv \vdash v_{1} \subtype v_{2}}
+\end{array}
+\;
 \begin{array}{c}
@@
 \dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash \Const \; v_{1} \subtype_{c} \Const \; v_{2}}
-\end{array}
-\;
+      {\senv \vdash \Const \; v_{1} \subtype \Const \; v_{2}}
+\end{array}
+\\ \\
 \begin{array}{c}
@@
 \dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash v_{1} \subtype_{c} \Const \; v_{2}}
-\end{array}
-\end{array}
-\]
-
-There are two limitations on \emph{closed} table types that led us to
-introduce \emph{open} and \emph{unique} table types.
-First, they do not allow programmers to use the table constructor to
-initialize table types that include optional fields.
+      {\senv \vdash v_{1} \subtype \Const \; v_{2}}
+\end{array}
+\end{array}
+\]
+
+The rule \textsc{S-FIELD1} handles mutable values invariantly,
+while the rule \textsc{S-FIELD2} handles immutable values covariantly.
+It is also safe to promote a mutable value to immutable, and the rule
+\textsc{S-FIELD3} defines this behavior.
+However, it is not safe to promote an immutable value to mutable,
+and that is the reason why we do not have a subtying rule for this case.
+
+There is a limitation on \emph{closed} table types that led us to
+introduce \emph{unique} table types.
+If the table constructor had a \emph{closed} table type, then
+programmers would not be able to use it to initialize a variable with
+a table type that describe a more general type.
 For instance,
-\begin{align*}
-\{ & ``firstname":\String, ``lastname":\String \}_{closed} \not \subtype\\
-\{ & ``firstname":\String,\\
-  &  ``middlename":\String \cup \Nil, ``lastname":\String \}_{closed}
-\end{align*}
-Second, they do not allow programmers to use the table constructor to
-initialize table types that represent hashes.
-For instance,
-\begin{align*}
-\{ & ``x":\Number, ``y":\Number, ``z":\Number \}_{closed} \not \subtype\\
-\{ & \String:\Number \cup \Nil\}_{closed}
-\end{align*}
-
-We introduced \emph{open} table types to allow the refinement
-of table types, and also to allow programmers to omit optional
-fields in the initialization of table types.
-To do that, we defined a subtyping rule that allow \emph{open}
-table types to omit fields when $\Nil$ is subtype of the type of
-the value of the omitted key.
-The following subtyping rule show this behavior:
-\[
-\begin{array}{c}
-\mylabel{S-TABLE2}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
-       \senv \vdash v_{i} \subtype_{c} v_{i}' \;\;\;
-       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
-      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
-                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open}}
-\end{array}
-\]
-
-The rule \textsc{S-TABLE2} introduced the relation $\subtype_{o}$.
-It is an auxiliary relation that defines the subtyping relation
-on optional fields, and we defined it as follows:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-FIELD4}\\
-\dfrac{\senv \vdash \Nil \subtype v}
-      {\senv \vdash \Nil \subtype_{o} v}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FIELD5}\\
-\dfrac{\senv \vdash \Nil \subtype v}
-      {\senv \vdash \Nil \subtype_{o} \Const \; v}
-\end{array}
-\end{array}
-\]
-
-Although \emph{open} table types let programmers omit optional fields,
-they do not let programmers initialize optional fields that are
-mutable, unless they use the same type of the field in the initialization.
-For instance,
-\begin{align*}
-\{ & ``firstname":\String,\\
-   & ``middlename":\String, ``lastname":\String \}_{open} \not \subtype \\
-\{ & ``firstname":\String, \\
-   & ``middlename":\String \cup \Nil, ``lastname":\String \}_{closed}
-\end{align*}
-
-We introduced \emph{unique} table types to solve this problem.
-We use \emph{unique} table types to denote the type of the
-table constructor.
-The following subtyping rule allow programmers to use table constructors
-to initialize and omit optional fields:
-\[
-\begin{array}{c}
-\mylabel{S-TABLE3}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
-       \senv \vdash v_{i} \subtype_{u} v_{i}' \;\;\;
-       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
-      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
-                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open|unique}}
-\end{array}
-\]
-
-The rule \textsc{S-TABLE3} introduced the relation $\subtype_{u}$.
-It is an auxiliary relation that makes \emph{unique} table types
-have width and depth subtyping on all fields, and we defined it
-as follows:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-FIELD6}\\
-\dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash v_{1} \subtype_{u} v_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FIELD7}\\
-\dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash \Const \; v_{1} \subtype_{u} \Const \; v_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FIELD8}\\
-\dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash v_{1} \subtype_{u} \Const \; v_{2}}
-\end{array}
-\end{array}
-\]
-
-Table types can be ambiguous.
-For instance, $\{1:\Number, \Integer:\String\}$ is an ambiguous
-table type because the type of the value stored by key $1$ can be
-$\Number$ or $\String$, as $1 \subtype 1$ and
-$1 \subtype \Integer$.
-Table types can get even more ambiguous when we use the dynamic
-type in the type of table keys.
-For instance, the table type $\{1:\Number, \Integer:\String, \Any:\Boolean\}$
-is ambiguous because the type of the value stored by key $1$ can be
-$\Number$, or $\String$, or $\Boolean$, as $1 \subtype 1$,
-$1 \subtype \Integer$, and $1 \lesssim \Any$.
+\begin{verbatim}
+    local t:{"x":integer, "y":integer?} = { x = 1, y = 2 }
+\end{verbatim}
+whould not type check, as the type of the table constructor would not
+be subtype of the type in the annotation.
+More precisely,
+\[
+\{``x":1, ``y":2\}_{closed} \not\subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
+\]
+Thus, we introduced \emph{unique} table types to represent the type of
+the table constructor and avoid this limitation.
+
+We introduced \emph{open} table types to allow the refinement of table types,
+an idiom very common in Lua.
+
+Even though \emph{open} and \emph{unique} table types have specific
+subtyping rules, we will present them in the next section along with
+the typing rules that formalize the refinement of table types.
+The presentation of these subtyping rules is easier in the next section,
+as they include some unsafe behaviors that the refinement of
+table types keeps track to make them safe.
+
+Any table type has to be \emph{well-formed}.
+Informally, a table type is well-formed if key types do not overlap.
+Formally, a table type is well-formed if it obeys the following rule:
+\[
+\forall i \not\exists j \; i \not= j \wedge k_{i} \lesssim k_{j}
+\]
+
+Well-formed table types avoid ambiguous table types.
+For instance, this rule detects that the table type
+$\{1:\Number, \Integer:\String, \Any:\Boolean\}$ is ambiguous,
+because the type of the value stored by key $1$ can be
+$\Number$, $\String$, or $\Boolean$, as $1 \lesssim 1$,
+$1 \lesssim \Integer$, and $1 \lesssim \Any$.
 Moreover, the type of the value stored by a key of type $\Integer$,
 which is not the literal type $1$, can be $\Number$ or $\Boolean$,
-as $\Integer \subtype \Integer$, and $\Integer \lesssim \Any$.
-
-To avoid ambiguous table types we use the following rule to check
-whether a table type is well formed:
-\[
-\forall i \not\exists j \; i \not= j \wedge k_{i} \lesssim k_{j}
-\]
-
-Our type system also includes recursive types: $\mu x.t$,
-where $t$ is a function or table type with $x$ appearing anywhere
-a function or table type could appear.
-For instance, $\mu x.\{``info":\Number, ``next":x \;\cup\; \Nil\}_{closed}$
-represents the type of single linked list of numbers.
-
-We use the \emph{Amber rule} from the Amber language
-\citep{cardelli1986amber} to define the subtyping rule for
-recursive types:
+as $\Integer \lesssim \Integer$, and $\Integer \lesssim \Any$.
+
+Recursive types have the form $\mu x.t$,
+where $t$ is a first-level type with $x$ appearing anywhere this
+first-level type could appear.
+For instance, $\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}$
+is a type for singly-linked list of integers.
+
+We use the \emph{Amber rule} \citep{cardelli1986amber} to define
+subtyping between recursive types:
 \[
@@
 
-Our type system does not infer recursive types, so we defined two
-extra subtyping rules that handle the cases where they may
-appear inside a table or function type:
+A recursive type may appear inside a first-level type, and our
+type system includes subtyping rules to handle subtyping between
+a first-level type and a recursive type:
 \[
@@
 
-For instance, the rule \textsc{S-AMBERR} allows our type system
-check whether the type of a table constructor is subtype of
-a recursive type or not, as we show in the following example:
+For instance, it is the rule \textsc{S-AMBERR} that allows our
+type system to type check the example that we presented
+in Section \ref{sec:tables}:
+\begin{verbatim}
+    local interface Element
+      info:integer
+      next:Element?
+    end
+    local function insert (e:Element?, v:integer):Element
+      return { info = v, next = e }
+    end
+\end{verbatim}
+
+The interface \texttt{Element} defines the type for singly-linked
+list of integers that we just mentioned, and the function
+\texttt{insert} inserts an element in the begining of the list.
+In this example, the recursive type appears inside the type of
+the table constructor.
+It type checks because the rule \textsc{S-AMBERR} verifies that
+the type of the table constructor is a subtype of \texttt{Element}:
 \begin{align*}
-\{ & ``info":\Number, \\
-   & ``next":\mu x.\{``info":\Number, ``next":x \;\cup\; \Nil\}_{closed} \cup \Nil \}_{unique} \subtype \\
-& \mu x.\{``info":\Number, ``next":x \;\cup\; \Nil\}_{closed}
+\{ & ``info":\Integer, \\
+   & ``next":\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed} \cup \Nil \}_{unique} \subtype \\
+& \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}
 \end{align*}
 
-Our type system also introduces projection types, but they do not
-appear on the subtyping and consistent-subtyping rules.
-They are just a mechanism that the type system uses for handling
-union of tuple types, when they appear on the right-hand side of
-the declaration of local variables.
-That is the reason why we do not need to define subtyping and
-consistent-subtyping rules for them.
-We will see projection types in more detail in the next section.
+Projection types have the form $\pi_{i}^{x}$, where $x$ is a variable
+that stores an union of tuple types in a special type environment,
+and $i$ is an index that indicates how to build a first-level
+union type from $x$.
+For instance, if $x$ is the union of tuple types
+$(\Integer \times \Integer) \sqcup (\Nil \times \String)$,
+then $\pi_{1}^{x}$ represents the type $\Integer \cup \Nil$
+while $\pi_{2}^{x}$ represents the type $\Integer \cup \String$.
+Projection types follow the reflexive subtyping rule.
+In the next section we will show in more detail how our type system
+uses them as a mechanism for handling unions of tuple types,
+when they appear on the right-hand side of the declaration of local variables.
+We will also show how this feature allows our type system to constrain
+the type of a local variable that depends on the type of another local variable.
+
+\textit{/* How should we control invalid indices? */}
 
@@
 In this section we use a reduced core of Typed Lua to present the
-most interesting typing rules of our type system.
-It is also a reduced core of the Lua language with only if and while
-control flow statements, without syntactic sugar, with explicit
-type annotations, with explicit scope of local variables, and
-with explicit declaration of methods.
+most interesting rules of our type system.
+This core limits control flow to if and while statements,
+has explicit type annotations, and explicit scope for variables.
+It also has explicit method declarations and explicit method calls
+instead of treating them as syntactic sugar.
 We use this reduced core because it simplifies the presentation
@@
 
-Our reduced code also splits function application, method application, and
-the vararg expression (${...}$) into different categories.
+Our reduced core also splits function application, method application, and
+the vararg expression (${...}$) into different syntactic categories.
 It uses $e(el)_{s}$ to denote function applications that return no value
@@
       {\env_{1} \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, t_{1}{:}t\}_{open|unique}]}
+\end{array}
+\]
+
+We introduced \emph{open} table types to allow the refinement
+of table types, and also to allow programmers to omit optional
+fields in the initialization of table types.
+To do that, we defined a subtyping rule that allow \emph{open}
+table types to omit fields when $\Nil$ is subtype of the type of
+the value of the omitted key.
+The following subtyping rule show this behavior:
+\[
+\begin{array}{c}
+\mylabel{S-TABLE2}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
+       \senv \vdash v_{i} \subtype_{c} v_{i}' \;\;\;
+       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
+      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
+                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open}}
+\end{array}
+\]
+
+The rule \textsc{S-TABLE2} introduced the relation $\subtype_{o}$.
+It is an auxiliary relation that defines the subtyping relation
+on optional fields, and we defined it as follows:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-FIELD4}\\
+\dfrac{\senv \vdash \Nil \subtype v}
+      {\senv \vdash \Nil \subtype_{o} v}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FIELD5}\\
+\dfrac{\senv \vdash \Nil \subtype v}
+      {\senv \vdash \Nil \subtype_{o} \Const \; v}
+\end{array}
+\end{array}
+\]
+
+Although \emph{open} table types let programmers omit optional fields,
+they do not let programmers initialize optional fields that are
+mutable, unless they use the same type of the field in the initialization.
+For instance,
+\begin{align*}
+\{ & ``firstname":\String,\\
+   & ``middlename":\String, ``lastname":\String \}_{open} \not \subtype \\
+\{ & ``firstname":\String, \\
+   & ``middlename":\String \cup \Nil, ``lastname":\String \}_{closed}
+\end{align*}
+
+We introduced \emph{unique} table types to solve this problem.
+We use \emph{unique} table types to denote the type of the
+table constructor.
+The following subtyping rule allow programmers to use table constructors
+to initialize and omit optional fields:
+\[
+\begin{array}{c}
+\mylabel{S-TABLE3}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
+       \senv \vdash v_{i} \subtype_{u} v_{i}' \;\;\;
+       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
+      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
+                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open|unique}}
+\end{array}
+\]
+
+The rule \textsc{S-TABLE3} introduced the relation $\subtype_{u}$.
+It is an auxiliary relation that makes \emph{unique} table types
+have width and depth subtyping on all fields, and we defined it
+as follows:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-FIELD6}\\
+\dfrac{\senv \vdash v_{1} \subtype v_{2}}
+      {\senv \vdash v_{1} \subtype_{u} v_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FIELD7}\\
+\dfrac{\senv \vdash v_{1} \subtype v_{2}}
+      {\senv \vdash \Const \; v_{1} \subtype_{u} \Const \; v_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FIELD8}\\
+\dfrac{\senv \vdash v_{1} \subtype v_{2}}
+      {\senv \vdash v_{1} \subtype_{u} \Const \; v_{2}}
+\end{array}
 \end{array}

### 2014-12-09 doc/thesis/review.tex
Minor improvements to chapters 2 and 3

@@
 combining static and dynamic typing in the same language.
-After that, we introduce optional type systems and gradual typing.
-We end this section discussing why optional type systems and two
+Then, we introduce optional type systems and gradual typing.
+After that, we discuss why optional type systems and two
 other approaches are often called gradual typing.
+We end this chapter presenting some statistics about the usage of
+some Lua features and idioms that helped us identify how should
+we combine static and dynamic typing in Lua.
 
 \section{A little bit of history}
+\label{sec:history}
 
@@
 checking of dynamically typed languages for detecting programming
-errors while dynamic typing targets the optimization of dynamically
+errors, while dynamic typing targets the optimization of dynamically
 typed code through the elimination of unnecessary run-time checks.
@@
 contracts that check the input and output of higher-order functions;
-and the compiler add blame annotations in the generated code as a way
+and the compiler adds blame annotations in the generated code
 to track assertion failures back to the source of the error.
@@
 \section{Optional Type Systems}
+\label{sec:optional}
 
 Optional type systems \citep{bracha2004pluggable} are an approach for
-plugging static typing in dynamically typed languages through
-optional type annotations that are used to perform compile-time type
-checking, though not influencing the original run-time semantics of
-the language.
-That is, the run-time semantics should still catch type errors
+plugging static typing in dynamically typed languages.
+They use optional type annotations to perform compile-time type checking,
+though they do not influence the original run-time semantics
+of the language.
+This means that the run-time semantics should still catch type errors
 independently of the static type checking.
@@
 \emph{Pluggable type systems} \citep{bracha2004pluggable} generalize
-the idea of optional type systems that Strongtalk put in practice,
-to have different optional type systems that can be layered on top
-of a dynamically typed language without affecting its original run-time
-semantics.
+the idea of optional type systems that Strongtalk put in practice.
+The idea is to have different optional type systems that can be layered
+on top of a dynamically typed language without affecting its original
+run-time semantics.
 Although these systems can be unsound on their interaction with the
@@
 Dart \citep{dart} and TypeScript \citep{typescript} are new
-languages that are designed with an optional type system, and both
-use JavaScript as their code generation target because their main
-purpose is web development.
+languages that are designed with an optional type system.
+Both use JavaScript as their code generation target because
+their main purpose is web development.
 In fact, Dart is a new class-based object-oriented language with
@@
 
-Unfortunately, there is no common formalization for optional type
-systems, so each language ends up implementing its optional type
-system on its own way.
+There is no common formalization for optional type systems, and
+each language ends up implementing its optional type system on
+its own way.
 Strongtalk, Dart, and TypeScript provide an informal description of
@@
 \section{Gradual Typing}
+\label{sec:gradual}
 
@@
 $$
-\begin{array}{rlr}
-T ::= & & \emph{types}\\
+\begin{array}{llr}
+t ::= & & \emph{types}\\
 & \;\; \Number & \emph{base type number}\\
@@
 & | \; ? & \emph{dynamic type}\\
-& | \; T \rightarrow T & \emph{function types}\\
+& | \; t \rightarrow t & \emph{function types}\\
 e ::= & & \emph{expressions}\\
-& \;\; c & \emph{constants}\\
+& \;\; l & \emph{literals}\\
 & | \; x & \emph{variables}\\
-& | \; \lambda x{:}T{.}e & \emph{abstractions}\\
+& | \; \lambda x{:}t{.}e & \emph{abstractions}\\
 & | \; e_{1} e_{2} & \emph{application}
@@
 The central idea of gradual typing is the \emph{consistency}
-relation, written $T_{1} \sim T_{2}$.
+relation, written $t_{1} \sim t_{2}$.
 The consistency relation allows implicit conversions to and from the
@@
 \begin{array}{c}
-T \sim T \mylabel{C-REFL}
+t \sim t \mylabel{C-REFL}
 \end{array}
@@
 \begin{array}{c}
-T \sim \;? \mylabel{C-DYNR}
+t \sim \;? \mylabel{C-DYNR}
 \end{array}
@@
 \begin{array}{c}
-? \sim T \mylabel{C-DYNL}
+? \sim t \mylabel{C-DYNL}
 \end{array}
@@
 \begin{array}{c}
-\dfrac{T_{3} \sim T_{1} \;\;\; T_{2} \sim T_{4}}
-      {T_{1} \rightarrow T_{2} \sim T_{3} \rightarrow T_{4}} \mylabel{C-FUNC}
+\dfrac{t_{3} \sim t_{1} \;\;\; t_{2} \sim t_{4}}
+      {t_{1} \rightarrow t_{2} \sim t_{3} \rightarrow t_{4}} \mylabel{C-FUNC}
 \end{array}
@@
 conversions to and from the dynamic type $?$.
-The rule \textsc{C-FUNC} resembles the subtyping of functions
+The rule \textsc{C-FUNC} resembles the subtyping between function types,
 because it is contravariant on the argument type and covariant on the
@@
 The environment $\env$ is a function from variables to types, and
-the directive $type$ is a function from constants to types.
+the directive $type$ is a function from literal values to types.
 The rule \textsc{T-VAR} uses the environment function $\env$ to get the
 type of a variable $x$.
-The rule \textsc{T-CONST} uses the directive $type$ to get the type of
-a constant $c$.
+The rule \textsc{T-LIT} uses the directive $type$ to get the type of
+a literal $l$.
 The rule \textsc{T-ABS} evaluates the expression $e$ with an environment
-$\env$ that binds the variable $x$ to the type $T_{1}$, and the resulting
-type is the the function type $T_{1} \rightarrow T_{2}$.
+$\env$ that binds the variable $x$ to the type $t_{1}$, and the resulting
+type is the the function type $t_{1} \rightarrow t_{2}$.
 The rule \textsc{T-APP1} handles function calls where the type of the
@@
 \begin{matrix}
-\dfrac{\env(x) = T}
-      {\env \vdash x:T} \mylabel{T-VAR}
+\dfrac{\env(x) = t}
+      {\env \vdash x:t} \mylabel{T-VAR}
 \;\;\;
-\dfrac{type(c) = T}
-      {\env \vdash c:T} \mylabel{T-CONST}
+\dfrac{type(l) = t}
+      {\env \vdash l:t} \mylabel{T-LIT}
 \\ \\
-\dfrac{\env, x:T_{1} \vdash e:T_{2}}
-      {\env \vdash \lambda x:T_{1}.e:T_{1} \rightarrow T_{2}} \mylabel{T-ABS}
+\dfrac{\env, x:t_{1} \vdash e:t_{2}}
+      {\env \vdash \lambda x:t_{1}.e:t_{1} \rightarrow t_{2}} \mylabel{T-ABS}
 \;\;\;
 \dfrac{\env \vdash e_{1}:\;? \;\;\;
-       \env \vdash e_{2}:T}
+       \env \vdash e_{2}:t}
       {\env \vdash e_{1} e_{2}:\;?} \mylabel{T-APP1}
 \\ \\
-\dfrac{\env \vdash e_{1}:T_{1} \rightarrow T_{2} \;\;\;
-       \env \vdash e_{2}:T_{3} \;\;\;
-       T_{3} \sim T_{1}}
-      {\env \vdash e_{1} e_{2}:T_{2}} \mylabel{T-APP2}
+\dfrac{\env \vdash e_{1}:t_{1} \rightarrow t_{2} \;\;\;
+       \env \vdash e_{2}:t_{3} \;\;\;
+       t_{3} \sim t_{1}}
+      {\env \vdash e_{1} e_{2}:t_{2}} \mylabel{T-APP2}
 \end{matrix}
@@
 The subtyping relation $\subtype$ actually is a pitfall on Thatte's
-quasi-static typing, because quasi-static typing sets the dynamic type
+quasi-static typing, because it sets the dynamic type
 as the top and the bottom of the subtying relation:
-$T \subtype \;?$ and $? \subtype T$.
+$t \subtype \;?$ and $? \subtype t$.
 Subtyping is transitive, so we have that
@@
 orthogonal to the subtyping relation, so we can combine them to achieve
-the \emph{consistent-subtyping} relation, written $T_{1} \lesssim T_{2}$.
+the \emph{consistent-subtyping} relation, written $t_{1} \lesssim t_{2}$.
 This relation is essential for designing gradual type systems for
@@
 \begin{array}{c}
-\dfrac{T_{3} \subtype T_{1} \;\;\; T_{2} \subtype T_{4}}
-      {T_{1} \rightarrow T_{2} \subtype T_{3} \rightarrow T_{4}} \mylabel{S-FUN}
+\dfrac{t_{3} \subtype t_{1} \;\;\; t_{2} \subtype t_{4}}
+      {t_{1} \rightarrow t_{2} \subtype t_{3} \rightarrow t_{4}} \mylabel{S-FUN}
 \end{array}
@@
 typed lambda calculus extended with the dynamic type $?$.
-Even though we could have used the reflexive rule $T \subtype T$ to express
+Even though we could have used the reflexive rule $t \subtype t$ to express
 the rules \textsc{S-NUM}, \textsc{S-STR}, and \textsc{S-ANY},
@@
 However, another important feature of gradual typing is the theoretic
-foundation that is provides for inserting run-time checks that
+foundation that it provides for inserting run-time checks that
 prove dynamically typed code does not violate the invariants of
@@
 
-However, \citet{herman2007sgt} showed that there is an efficiency
+\citet{herman2007sgt} showed that there is an efficiency
 concern regarding the run-time checks, because there are two
@@
 \citet{herman2007sgt} use the coercion calculus of
-\citet{henglein1994dts} to solve the problem of space efficiency,
-that is, they use coercions to express casts.
+\citet{henglein1994dts} to express casts as coercions and
+solve the problem of space efficiency.
 Their approach normalizes an arbitrary sequence of coercions to a
@@
 \citet{wadler2009wpc} developed the \emph{blame calculus} as a way
-to attack this issue, and \citet{ahmed2011bfa} extended the
+to handle this issue, and \citet{ahmed2011bfa} extended the
 blame calculus with polymorphism.
@@
 After that, \citet{siek2010blame} proposed a new solution that also
-attacks both problems.
+handles both problems.
 This new solution is based on a concept called \emph{threesome},
@@
 
-ActionScript \citep{moock2007as3} is the first language that
-incorporated a gradual type system to its implementation and
-Perl 6 \citep{tang2007pri} is also being designed with a
-gradual type system, though there is little documentation about
-the gradual type systems of these languages.
-Grace \citep{black2012grace,black2013sg} is a new object-oriented,
-gradually typed, educational language.
-In Grace, modules are gradually typed objects, that is, modules
-may have types and methods as attributes, and can have a state
-\citep{homer2013modules}.
-Gradual Python \citep{reticulated} is an attempt to develop a
-gradual type system for Python that deals with mutable objects
-\citep{siek2013mutable}.
+There are some programming languages that either incorporated
+gradual typing or have research on this subject.
+Reticulated Python \citep{reticulated,vitousek2014deg} is a research
+project that evaluates the costs of gradual typing in Python.
 Gradualtalk \citep{allende2013gts} is a gradually-typed Smalltalk
@@
 objects \citep{allende2013cis}.
+Grace \citep{black2012grace,black2013sg} is a new object-oriented,
+gradually-typed, educational language.
+In Grace, modules are gradually-typed objects, that is, modules
+may have types and methods as attributes, and can have a state
+\citep{homer2013modules}.
+ActionScript \citep{moock2007as3} is one the first languages that
+incorporated gradual typing to its implementation and
+Perl 6 \citep{tang2007pri} is also being designed with gradual typing,
+though there is few documentation about the gradual type systems
+of these languages.
 
 \section{Approaches that are often called Gradual Typing}
+\label{sec:approaches}
 
@@
 BabyJ \citep{anderson2003babyj} and Alore \citep{lehtosalo2011alore}
-are two examples of object-oriented languages that use gradual
-typing to formalize their optional type systems.
-The optional type systems of both BabyJ and Alore are nominal.
+are two examples of object-oriented languages that have an
+optional type system with a formalization that relates to gradual typing,
+though the optional type systems of both BabyJ and Alore are nominal.
 BabyJ uses the relation $\approx$ that is similar to the consistency
@@
 consistent-subtyping relation of \citet{siek2007objects}, but it is
-also written $T_{1} \lesssim T_{2}$.
-The consistent-or-subtype relation holds when $T_{1} \sim T_{2}$
-or $T_{1} <: T_{2}$, where $<:$ is transitive and $\sim$ is not.
+also written $t_{1} \lesssim t_{2}$.
+The consistent-or-subtype relation holds when $t_{1} \sim t_{2}$
+or $t_{1} <: t_{2}$, where $<:$ is transitive and $\sim$ is not.
 Alore also extends its optional type system to include optional
@@
 scripts to typed programs.
-However, note that Typed Racket is a statically typed language
-and type annotations are mandatory.
-So, what makes it gradual is a type system with a dynamic type
+However, Typed Racket is a statically typed language,
+and what makes it gradual is a type system with a dynamic type
 that handles the interaction between Racket and Typed Racket modules.
@@
 the LuaRocks repository at February 1st 2014.
-However, we ignored files that are not compatible with Lua 5.2,
-the latest version of Lua.
+However, we ignored files that were not compatible with Lua 5.2,
+the latest version of Lua at that time.
 We also ignored \emph{machine-generated} files and test files,
@@
 because the use of the \texttt{type} idiom shows that disjoint unions
-will help programmers define data structures that can hold a value of
+would help programmers define data structures that can hold a value of
 several different, but fixed types.
@@
 multiple return values, and multiple assignments.
-Tuple types will require some special attention, because Typed Lua
+Tuple types require some special attention, because Typed Lua
 should be able to adjust tuple types during compile-time, in a
@@
 \texttt{false} plus something else show us that overloading on the
-return type will be useful to the type system.
+return type is also useful to the type system.
 
@@
 
-Based on these observations, Typed Lua should include interfaces and
-classes as syntactic sugar for table types that encode objects.
-Proposing an optional type system for Lua that includes classes may
-sound unusual at first glance, because Lua does not have standard
-policies for object-oriented programming.
-However, Lua provides mechanisms that allow programmers to abstract
-their code in terms of objects, and our statistics confirm that
-an expressive number of programmers are relying on these mechanisms to
-use the object-oriented paradigm in Lua.
-Typed Lua should include interfaces and classes as merely syntactic
-mechanisms that help the compiler type check object-oriented code,
-and they should not influence the semantics of Lua.
+Based on these observations, Typed Lua should include support
+to object-oriented programming.
+Even though Lua does not have standard policies for object-oriented
+programming, it provides mechanisms that allow programmers to
+abstract their code in terms of objects, and our statistics confirm
+that an expressive number of programmers are relying on these mechanisms
+to use the object-oriented paradigm in Lua.
+Typed Lua should include some standard way of defining interfaces and classes
+that the compiler can use to type check object-oriented code,
+but without changing the semantics of Lua.
 
@@
 
-Based on these observations, Typed Lua should also use interfaces as
-syntactic sugar for table types that define modules.
+Based on these observations, Typed Lua should include a way
+for defining table types that represent the type of modules.
 Typed Lua should also support the deprecated style of module
@@
 
-Besides using interfaces for encoding objects and defining modules,
-Typed Lua should also use interfaces for defining the types of
-userdata, as programmers often use userdata in an object-oriented
-style.
+Typed Lua should also include some way for defining the types of
+userdata.
+This feature should also allow programmers to define an userdata
+that can be used in an object-oriented style, as this is another
+common idiom from modules that are written in C.
 

### 2014-12-09 doc/thesis/typedlua.tex
Improvements to the text about table types
Minor improvements to chapters 2 and 3

@@
 \section{Optional type annotations}
+\label{sec:optann}
 
@@
 \begin{align*}
-\textit{type} & ::= \; \textit{primarytype} \; [\texttt{`?'}]\\
-\textit{primarytype} & ::= \; \textit{literaltype} \; | \;
+\textit{type} ::= & \;\; \textit{primarytype} \; [\texttt{`?'}]\\
+\textit{primarytype} ::= & \;\; \textit{literaltype} \; | \;
   \textit{basetype} \; | \;
@@
   \textit{primarytype} \; \texttt{`|'} \; \textit{primarytype}\\
-\textit{literaltype} & ::= \; \textbf{false} \; | \;
+\textit{literaltype} ::= & \;\; \textbf{false} \; | \;
   \textbf{true} \; | \;
@@
   \textit{String}\\
-\textit{basetype} & ::= \; \textbf{boolean} \; | \;
+\textit{basetype} ::= & \;\; \textbf{boolean} \; | \;
   \textbf{integer} \; | \;
@@
   \textbf{string}\\
-\textit{functiontype} & ::= \; \textit{tupletype} \; \texttt{`->'} \; \textit{rettupletype}\\
-\textit{methodtype} & ::= \; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettupletype}\\
-\textit{tabletype} & ::= \; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
-\textit{tupletype} & ::= \; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
-\textit{typelist} & ::= \; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
-\textit{rettupletype} & ::= \; \textit{unionlist} \; [\texttt{`?'}]\\
-\textit{unionlist} & ::= \; \textit{tupletype} \; | \;
-  \textit{unionlist} \; \texttt{`|'} \; \textit{unionlist}\\
-\textit{tabletypebody} & ::= \; \textit{type} \; | \; \textit{fieldtypelist}\\
-\textit{fieldtypelist} & ::= \; [\textbf{const}] \; \textit{fieldtype} \; \{\texttt{`,'} \; [\textbf{const}] \; \textit{fieldtype}\}\\ 
-\textit{fieldtype} & ::= \; \textit{keytype} \; \texttt{`:'} \; \textit{type}\\
-\textit{keytype} & ::= \; \textit{literaltype} \; | \;
-  \textit{basetype} \; | \;
-  \textbf{value} \; | \;
+\textit{functiontype} ::= & \;\; \textit{tupletype} \; \texttt{`->'} \; \textit{rettype}\\
+\textit{tupletype} ::= & \;\; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
+\textit{typelist} ::= & \;\; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
+\textit{rettype} ::= & \;\; \textit{type} \; | \;
+  \textit{uniontuple} \; [\texttt{`?'}]\\
+\textit{uniontuple} ::= & \;\; \textit{tupletype} \; | \;
+  \textit{uniontuple} \; \texttt{`|'} \; \textit{uniontuple}\\
+\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
+\textit{tabletypebody} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type} \; | \;
+  \textit{recordtype}\\
+\textit{keytype} ::= & \;\; \textit{basetype} \; | \;
   \textbf{any}\\
-\textit{rettype} & ::= \; \textit{type} \; | \; \textit{rettupletype}
+\textit{recordtype} ::= & \;\; \textit{recordfield} \; \{\texttt{`,'} \; \textit{recordfield}\} \; [\texttt{`,'} \; \textit{type}]\\
+\textit{recordfield} ::= & \;\; [\textbf{const}] \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}
 \end{align*}
@@
 Second-level types include the type void \texttt{()},
-vararg types, tuple types, and union of tuple types.
+vararg types, tuple types, and unions of tuple types.
 
@@
 value that is not an integer number.
-More precisely, we can call \texttt{absolute} inside \texttt{distance}
+In fact, we can overload the minus operator to return a value that is
+not even a number.
+However, we can call \texttt{absolute} inside \texttt{distance}
 because the expression \texttt{x - y} has type \texttt{any},
@@
 \section{Functions}
+\label{sec:func}
 
@@
 As we can see in Figure \ref{fig:types}, second-level types consist
-of either \emph{tupletype} or \emph{rettupletype}.
+of either \emph{tupletype} or \emph{rettype}.
 A second-level type is the type void \texttt{()},
@@
 \texttt{(integer, integer, string, nil*)},
-and the type assigned to \texttt{x}, \texttt{y}, and \texttt{z}
-are respectively \texttt{integer}, \texttt{integer}, and \texttt{string}.
+and the types assigned to \texttt{x}, \texttt{y}, and \texttt{z}
+are \texttt{integer}, \texttt{integer}, and \texttt{string}, respectively.
 The same tuple is assigned to the call to \texttt{sum} and it
@@
 \section{Unions}
+\label{sec:unions}
 
@@
         local r = dividend % divisor
-        local q = (dividend - r) / divisor
+        local q = (dividend - r) // divisor
         return q, r
@@
 
-Projection types do not appear in the subtyping and consistent-subtyping
-relations because Typed Lua actually replaces them with the union of the
-corresponding component of each tuple.
+Typed Lua replaces projection types with the union of the corresponding
+component of each tuple when it infers that a variable has a projection type.
 If a variable with a projection type appears in a \texttt{type} predicate,
@@
 
-We could also have used \texttt{type(q) == "number"} or
-even \texttt{type(r) == "number"} as the predicate of our example,
+We could also have used \texttt{type(q) == "integer"} or
+even \texttt{type(r) == "integer"} as the predicate of our example,
 as both predicates would produce the same result.
 However, the form that appears in our example is much more succinct and idiomatic.
+Note that the type \texttt{integer} is restricted to Lua 5.3,
+as we use the function \texttt{math.type} to decide whether a number
+is integer or not.
 
@@
 \begin{verbatim}
-    local function limitation (x:number|string, y:number|string)
+    local function limitation (x:number|string)
       if type(x) == "number" then
-        if type(y) == "number" then
-          return x + y
-        end
+        return x + x
       else
-        if type(y) == "string" then
-          return x .. y
-        end
+        return x .. x
       end
@@
 This example shows that we cannot write a function that is
-guaranteed to return a number when we pass two numbers and
-guaranteed to return a string when we pass two strings.
+guaranteed to return a number when we pass a number and
+guaranteed to return a string when we pass a string.
 Even though we could use intersection types to express the type
 of this function as
-\texttt{(number, number) -> (number) \string&
-(string, string) -> (string)},
+\texttt{(number) -> (number) \string&
+(string) -> (string)},
 and therefore solve this limitation,
@@
 \section{Tables and Interfaces}
+\label{sec:tables}
 
@@
 \section{Modules}
+\label{sec:modules}
 
@@
 \begin{verbatim}
-    local m = require "mymath"
-    print(m.pow(2, 3))                   -- 8
-    print(m.pow(2, "foo"))               -- compile-time error
+    local mymath = require "mymath"
+    print(mymath.pow(2, 3))              -- 8
+    print(mymath.pow(2, "foo"))          -- compile-time error
 \end{verbatim}
@@
 The call to \texttt{require} assigns the type of the module \texttt{mymath}
-to \texttt{m}, so Typed Lua can catch misuses of the module.
+to the local \texttt{mymath}, so Typed Lua can catch misuses of the module.
 
@@
 \section{Object-Oriented Programming}
+\label{sec:oop}
 
@@
 
-Typed Lua uses table types along with the self type to
+Typed Lua uses table types along with the \texttt{self} type to
 represent objects, as we show in the following example:
@@
 as it includes an implicit first parameter of type \texttt{self}.
+The double arrow can appear only inside the declaration of
+interfaces and userdata, which we will introduce in the next section.
 The interface definition allows us to export the type of the class \texttt{Shape},
@@
 \section{Description files}
+\label{sec:tld}
 
@@
 \begin{align*}
-\textit{description} & ::= \; \textit{desclist}\\
-\textit{desclist} & ::= \; \textit{descitem} \; \{ \textit{descitem} \}\\
-\textit{descitem} & ::= \; \textit{typedid} \; | \;
+\textit{description} ::= & \; \textit{desclist}\\
+\textit{desclist} ::= & \; \textit{descitem} \; \{ \textit{descitem} \}\\
+\textit{descitem} ::= & \; \textit{typedid} \; | \;
   \textit{interface} \; | \;
-  \textit{userdata}\\
-\textit{typedid} & ::= \; \textit{id} \; \texttt{`:'} \; \textit{type}\\ 
-\textit{interface} & ::= \; \textbf{interface} \; \textit{typedec}\\
-\textit{userdata} & ::= \; \textbf{userdata} \; \textit{typedec}\\
-\textit{typedec} & ::= \; \textit{Name} \; { \; \textit{decitem} \; } \; \textbf{end}\\
-\textit{decitem} & ::= \; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
-\textit{idtype} & ::= \; \textit{type} \; | \; \textit{methodtype}\\
-\textit{idlist} & ::= \; \textit{id} \; \{ \; \texttt{`,'} \; \textit{id} \; \}\\
-\textit{id} & ::= \; [\textbf{const}] \; \textit{Name}
+  \textit{userdata} \; | \;
+  \textit{typealias}\\
+\textit{typedid} ::= & \; \textit{id} \; \texttt{`:'} \; \textit{type}\\ 
+\textit{interface} ::= & \; \textbf{interface} \; \textit{typedec}\\
+\textit{userdata} ::= & \; \textbf{userdata} \; \textit{typedec}\\
+\textit{typedec} ::= & \; \textit{Name} \; { \; \textit{decitem} \; } \; \textbf{end}\\
+\textit{decitem} ::= & \; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
+\textit{idtype} ::= & \; \textit{type} \; | \; \textit{methodtype}\\
+\textit{idlist} ::= & \; \textit{id} \; \{ \; \texttt{`,'} \; \textit{id} \; \}\\
+\textit{id} ::= & \; [\textbf{const}] \; \textit{Name}\\
+\textit{typealias} ::= & \; \textbf{typealias} \; \textit{Name} \; \texttt{`='} \; \textit{type}
 \end{align*}
@@
 along with this table type.
-We can export a type name through the declaration of an interface or
-userdata.
-As we saw, an interface declaration is just a shortcut to a table type.
-An userdata declaration is a shortcut to a table type that includes
-a \emph{brand} that is its name.
+We can export a type name through the declaration of an interface,
+userdata, or type alias.
+As we saw, an interface declaration is just a shortcut to a table type
+that represents a record.
+An userdata declaration is similar to an interface declaration,
+but it also includes its name as the \emph{brand} of the table type.
 The Typed Lua compiler uses this brand to combine structural with
@@
 of each other because they do not share the same brand.
+A type alias declaration assigns a name to a first-level type.
+Even though the type alias declaration may make the interface declaration
+look redundant, it does not have a syntax as nice as the syntax
+of the interface declaration for defining records.
+The type alias declaration is specially useful when we need to create
+names for types that are not records.
+
 The Typed Lua compiler searches for a description file when it cannot
@@
       update : (self, string*) -> (self)
+      version : string
     end
@@
     local z = m.clone("foo")             -- compile-time error
-    print(x:digest() == m.digest(x))
-\end{verbatim}
-
-In this example there is a type error because function \texttt{clone}
+    print(x:digest() == m.digest(y))
+\end{verbatim}
+
+In this example there is a type error because the function \texttt{clone}
 expects a value of type \texttt{md5\string_context} instead of a value
@@
 The description files are the mechanism that we used to include the
-typing of the Lua standard libraries inside Typed Lua.
+typing of the Lua standard library inside Typed Lua.
+In Chapter \ref{chap:evaluation} we discuss the issues that we found
+while typing the Lua standard library.

### 2014-12-10 doc/thesis/typedlua.tex
Adding a section about interfaces and type aliases to chapter 3

@@
 
-\section{Tables and Interfaces}
+\section{Tables}
 \label{sec:tables}
@@
 
-Typed Lua includes \emph{interfaces} as syntactic sugar to table types
-that specify records, as writing table types can be unwieldy when
-records get bigger and types of record fields get more complicated.
-For instance, we can use the following interface as an alias to the
-type of the record we showed in the previous example:
-\begin{verbatim}
-    local interface Person
-      firstname:string
-      lastname:string
-    end
-\end{verbatim}
-
-The declaration of the interface \texttt{Person} is an alias to the
-table type \texttt{\{"firstname":string, "lastname":string\}} in the
-remainder of the current scope.
-Now we can use \texttt{Person} in type declarations:
-\begin{verbatim}
-    local function byebye (person:Person):string
-      return "Goodbye " .. person.firstname ..
-             " " .. person.lastname
-    end
-
-    local user1 = { firstname = "Lewis",
-                    middlename = "Allan",
-                    lastname = "Reed" }
-    local user2 = { firstname = "Lou" }
-    local user3 = { lastname = "Reed",
-                    firstname = "Lou" }
-    local user4 = { "Lou", "Reed" }
-
-    print(greeter(user1)) -- Goodbye Lewis Reed 
-    print(greeter(user2)) -- compile-time error
-    print(greeter(user3)) -- Goodbye Lou Reed
-    print(greeter(user4)) -- compile-time error
-\end{verbatim}
-
-This example shows that our optional type system is structural rather
-than nominal, that is, it checks the structure of types instead of
-their names, and it uses subtyping and consistent-subtyping for
-checking types.
-
-If our record type has optional fields, that is, fields that can be \texttt{nil},
-we need to use an explicit type declaration when declaring a
-variable of this record type, as we show in the following example:
-\begin{verbatim}
-    local interface Person
-      firstname:string
-      middlename:string?
-      lastname:string
-    end
-    local user1:Person = { firstname = "Lewis",
-                           middlename = "Allan",
-                           lastname = "Reed" }
-    local user2:Person = { lastname = "Reed",
-                           firstname = "Lou" }
-\end{verbatim}
-
-We need an explicit type declaration because neither of the types
-that the Typed Lua compiler assigns to the above table constructors
-is a subtype of
-\texttt{\{"firstname":string, "middlename":string|nil, "lastname":string\}},
-the type that \texttt{Person} describes.
-
-We can also use interfaces to define recursive types such as liked lists:
-
-\begin{verbatim}
-    local interface Element
-      info:integer
-      next:Element?
-    end
-\end{verbatim}
-
-We need to explicitly annotate \texttt{next} being of type
-\texttt{Element|nil} because in our type system \texttt{nil}
-is not the bottom type, that is, \texttt{nil} is subtype of itself only.
-Typed Lua has this behavior because in Lua \texttt{nil} represents
-the absence of an useful value.
-Thus, making \texttt{nil} the bottom type would severally reduce the amount
-of type safety in the system.
-
-Now we can use \texttt{Element} to annotate a function that
-inserts an element at the beginning of a list:
-\begin{verbatim}
-    local function insert (e:Element?, v:integer):Element 
-      return { info = v, next = e }
-    end
-\end{verbatim}
-
-We need an explicit type declaration on the return type because
-the Typed Lua compiler cannot infer recursive types, though
-it has subtyping rules to check that the inferred return type
-matches the annotation.
-
-Now we can write a program that declares a list, inserts some
-elements in this list, and then traverses the list printing
-each element of the list:
-\begin{verbatim}
-    local l:Element?
-
-    l = insert(l, 4)
-    l = insert(l, 3)
-    l = insert(l, 2)
-    l = insert(l, 1)
-
-    while l do
-      if l then print(l.info) end
-      l = l.next                  -- compile-time error
-    end
-\end{verbatim}
-
-However, this example has one compile-time error because we
-are attempting to index the variable \texttt{l},
-which can be \texttt{nil}.
-Notice that we would have the same compile-time error while
-printing the list element if we had not narrowed the type of
-\texttt{l}.
-To eliminate this compile-time error we have to write a
-function that narrows the type of \texttt{l} before using it:
-\begin{verbatim}
-    local function get_next (l:Element?)
-      if l then return l.next else return nil end
-    end
-\end{verbatim}
-
-Now we can use \texttt{get\string_next} to rewrite the \texttt{while}
-from the previous example, thus eliminating the compile-time error:
-\begin{verbatim}
-    while l do
-      if l then print(l.info) end
-      l = get_next(l)
-    end
-\end{verbatim}
-
 Lua programmers often build records incrementally, that is,
@@
 support idiomatic definition of Lua modules.
+
+\section{Interfaces and type aliases}
+\label{sec:alias}
+
+Typed Lua includes \emph{interfaces} as syntactic sugar to table types
+that specify records, as writing table types can be unwieldy when
+records get bigger and types of record fields get more complicated.
+For instance, we can use the following interface as an alias to the
+type of the record we showed in the last section:
+\begin{verbatim}
+    local interface Person
+      firstname:string
+      lastname:string
+    end
+\end{verbatim}
+
+The declaration of the interface \texttt{Person} is an alias to the
+table type \texttt{\{"firstname":string, "lastname":string\}} in the
+remainder of the current scope.
+Now we can use \texttt{Person} in type declarations:
+\begin{verbatim}
+    local function byebye (person:Person):string
+      return "Goodbye " .. person.firstname ..
+             " " .. person.lastname
+    end
+
+    local user1 = { firstname = "Lewis",
+                    middlename = "Allan",
+                    lastname = "Reed" }
+    local user2 = { firstname = "Lou" }
+    local user3 = { lastname = "Reed",
+                    firstname = "Lou" }
+    local user4 = { "Lou", "Reed" }
+
+    print(greeter(user1)) -- Goodbye Lewis Reed 
+    print(greeter(user2)) -- compile-time error
+    print(greeter(user3)) -- Goodbye Lou Reed
+    print(greeter(user4)) -- compile-time error
+\end{verbatim}
+
+This example shows that our optional type system is structural rather
+than nominal, that is, it checks the structure of types instead of
+their names, and it uses subtyping and consistent-subtyping for
+checking types.
+
+If our record type has optional fields, that is, fields that can be \texttt{nil},
+we need to use an explicit type declaration when declaring a
+variable of this record type, as we show in the following example:
+\begin{verbatim}
+    local interface Person
+      firstname:string
+      middlename:string?
+      lastname:string
+    end
+    local user1:Person = { firstname = "Lewis",
+                           middlename = "Allan",
+                           lastname = "Reed" }
+    local user2:Person = { lastname = "Reed",
+                           firstname = "Lou" }
+\end{verbatim}
+
+We need an explicit type declaration because neither of the types
+that the Typed Lua compiler assigns to the above table constructors
+is a subtype of
+\texttt{\{"firstname":string, "middlename":string|nil, "lastname":string\}},
+the type that \texttt{Person} describes.
+
+We can also use interfaces to define recursive types such as liked lists:
+
+\begin{verbatim}
+    local interface Element
+      info:integer
+      next:Element?
+    end
+\end{verbatim}
+
+We need to explicitly annotate \texttt{next} being of type
+\texttt{Element|nil} because in our type system \texttt{nil}
+is not the bottom type, that is, \texttt{nil} is subtype of itself only.
+Typed Lua has this behavior because in Lua \texttt{nil} represents
+the absence of an useful value.
+Thus, making \texttt{nil} the bottom type would severally reduce the amount
+of type safety in the system.
+
+Now we can use \texttt{Element} to annotate a function that
+inserts an element at the beginning of a list:
+\begin{verbatim}
+    local function insert (e:Element?, v:integer):Element 
+      return { info = v, next = e }
+    end
+\end{verbatim}
+
+We need an explicit type declaration on the return type because
+the Typed Lua compiler cannot infer recursive types, though
+it has subtyping rules to check that the inferred return type
+matches the annotation.
+
+Now we can write a program that declares a list, inserts some
+elements in this list, and then traverses the list printing
+each element of the list:
+\begin{verbatim}
+    local l:Element?
+
+    l = insert(l, 4)
+    l = insert(l, 3)
+    l = insert(l, 2)
+    l = insert(l, 1)
+
+    while l do
+      if l then print(l.info) end
+      l = l.next                  -- compile-time error
+    end
+\end{verbatim}
+
+However, this example has one compile-time error because we
+are attempting to index the variable \texttt{l},
+which can be \texttt{nil}.
+Notice that we would have the same compile-time error while
+printing the list element if we had not narrowed the type of
+\texttt{l}.
+To eliminate this compile-time error we have to write a
+function that narrows the type of \texttt{l} before using it:
+\begin{verbatim}
+    local function get_next (l:Element?)
+      if l then return l.next else return nil end
+    end
+\end{verbatim}
+
+Now we can use \texttt{get\string_next} to rewrite the \texttt{while}
+from the previous example, thus eliminating the compile-time error:
+\begin{verbatim}
+    while l do
+      if l then print(l.info) end
+      l = get_next(l)
+    end
+\end{verbatim}
+
+Typed Lua also includes a way to create aliases for types that does not
+follow the records syntax.
+However, we limitim aliases to union types, function types, table types,
+and recursive types.
+The reason for this restriction is to avoid the creation of aliases
+to predefined types.
+
+For instance, we can use the \texttt{typealias} construct to define an
+alias to the type of the abstract syntax tree for a language of simple
+arithmetic expressions:
+\begin{verbatim}
+    local typealias Exp = {"tag":"Number", 1:number}
+                        | {"tag":"Add", 1:Exp, 2:Exp}
+                        | {"tag":"Sub", 1:Exp, 2:Exp}
+                        | {"tag":"Mul", 1:Exp, 2:Exp}
+                        | {"tag":"Div", 1:Exp, 2:Exp}
+\end{verbatim}
+
+The type \texttt{Exp} is a recursive type that resembles the
+algebraic data types from functional programming.
+Typed Lua also includes typing rules that let programmers
+deconstruct unions of table types, and it resembles
+pattern matching from functional programming.
+We can use this feature to write an evaluation function for
+our simple language:
+\begin{verbatim}
+    local function eval (e:Exp):number
+      if e.tag == "Number" then return e[1]
+      elseif e.tag == "Add" then return eval(e[1]) + eval(e[2])
+      elseif e.tag == "Sub" then return eval(e[1]) - eval(e[2])
+      elseif e.tag == "Mul" then return eval(e[1]) * eval(e[2])
+      elseif e.tag == "Div" then return eval(e[1]) / eval(e[2])
+      end
+    end
+\end{verbatim}
+
+When Typed Lua finds the pattern \texttt{n.s1 == s2}, it checks
+if the type of the variable \texttt{n} is a union of table types
+and if \texttt{s1} and \texttt{s2} are strings.
+If that is the case, it constrains the type of \texttt{n} inside
+the \texttt{if} to the table type that has the pair \texttt{s1:s2}.
+For instance, in the case of \texttt{e.tag == "Add"}, Typed Lua
+constrains the type of \texttt{e} to \texttt{\{"tag":"Add", 1:Exp, 2:Exp\}}.
+This is safe because Typed Lua restricts this pattern to literal types only.
+
+Even though the type alias declaration may make the interface declaration
+look redundant, it does not have a syntax as nice as the syntax
+of the interface declaration for defining records.
+For this reason worth keeping two different constructs.
+One for records specifically and another to more general types.
 
@@
 A type alias declaration assigns a name to a first-level type.
-Even though the type alias declaration may make the interface declaration
-look redundant, it does not have a syntax as nice as the syntax
-of the interface declaration for defining records.
-The type alias declaration is specially useful when we need to create
-names for types that are not records.
 

### 2014-12-11 doc/thesis/typedlua.tex
Minor improvements the text about type aliases

@@
 \section{Optional type annotations}
-\label{sec:optann}
+\label{sec:annotations}
 
@@
 \section{Functions}
-\label{sec:func}
+\label{sec:functions}
 
@@
 are the foundation for defining modules and objects in Typed Lua.
-In the next section, we show that the refinement of table types
-through field assignment is crucial for making Typed Lua
+In Section \ref{sec:modules} we show that the refinement of table
+types through field assignment is crucial for making Typed Lua
 support idiomatic definition of Lua modules.
+In Section \ref{sec:oop} we show that the refinement of table
+types is also helpful for defining classes and objects.
 
@@
 that specify records, as writing table types can be unwieldy when
-records get bigger and types of record fields get more complicated.
+records get bigger and the types of record fields get more complicated.
 For instance, we can use the following interface as an alias to the
-type of the record we showed in the last section:
+type of the record we showed in the previous section:
 \begin{verbatim}
@@
 follow the records syntax.
-However, we limitim aliases to union types, function types, table types,
+However, we limit aliases to union types, function types, table types,
 and recursive types.
@@
 For instance, we can use the \texttt{typealias} construct to define an
-alias to the type of the abstract syntax tree for a language of simple
+alias to the type of the abstract syntax tree of a language of simple
 arithmetic expressions:
@@
 Typed Lua also includes typing rules that let programmers
-deconstruct unions of table types, and it resembles
+deconstruct unions of table types, which resembles
 pattern matching from functional programming.
@@
 
-When Typed Lua finds the pattern \texttt{n.s1 == s2}, it checks
-if the type of the variable \texttt{n} is a union of table types
-and if \texttt{s1} and \texttt{s2} are strings.
+When Typed Lua finds the pattern \texttt{n[s1] == s2}, it checks
+if the type of the variable \texttt{n} is an union of table types,
+if \texttt{s1} a literal string, and if \texttt{s2} is also a literal string.
 If that is the case, it constrains the type of \texttt{n} inside
 the \texttt{if} to the table type that has the pair \texttt{s1:s2}.
-For instance, in the case of \texttt{e.tag == "Add"}, Typed Lua
-constrains the type of \texttt{e} to \texttt{\{"tag":"Add", 1:Exp, 2:Exp\}}.
+For instance, the expression \texttt{e.tag == "Add"} makes Typed Lua
+constrain the type of \texttt{e} to \texttt{\{"tag":"Add", 1:Exp, 2:Exp\}}.
 This is safe because Typed Lua restricts this pattern to literal types only.
+If we did not add this pattern, we would not be able to traverse the AST,
+as any attempt to index an union is a compile-time error.
 
@@
 of the interface declaration for defining records.
-For this reason worth keeping two different constructs.
-One for records specifically and another to more general types.
+For this reason, it worth having two different constructs.
+One specifically for records and another for more general types.
 
@@
 of each other because they do not share the same brand.
-A type alias declaration assigns a name to a first-level type.
+A type alias declaration is a way to create shortcuts to union types,
+function types, table types, and recursive types.
 

### 2014-12-16 doc/thesis/system.tex
Adding a section about interfaces and type aliases to chapter 3
Splitting the section about types and subtyping into two sections
Improvements to the section about types

@@
 \section{Types}
-
-Our type system uses subtyping \citep{cardelli1984smi,abadi1996to} to order
-types and consistent-subtyping \citep{siek2007objects,siek2013mutable}
-to allow the interaction between statically and dynamically typed code.
-We will explain the subtyping and consistent-subtyping rules along this
-section while we introduce each type from our type system.
-We will focus the discussion on the definition of subtyping because,
-as we mentioned in Chapter \ref{chap:review}, we can combine the
-consistency and subtyping relations to achieve consistent-subtyping.
-The differences between subtyping and consistent-subtyping are the way
-they handle the dynamic type, and the fact that subtyping is transitive,
-but consistent-subtyping is not.
+\label{sec:types}
 
@@
 & | \; t \cup t & \textit{disjoint union types}\\
-& | \; p \rightarrow r & \textit{function types}\\
-& | \; \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{closed|open|unique} & \textit{table types}\\
+& | \; s \rightarrow r & \textit{function types}\\
+& | \; \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed|open|unique} & \textit{table types}\\
 & | \; x & \textit{type variables}\\
@@
 \multicolumn{3}{c}{\textbf{Second-level types}}\\
-s ::= & \;\; t &\\
+s ::= & \;\; \Void & \textit{void type}\\
 & | \; t* & \textit{variadic types}\\
 & | \; t \times s & \textit{tuple types}\\
-p ::= & \;\; \Void \; | \; s & \textit{parameters list type}\\
-r ::= & \;\; \Void \; | \; s \; | \; s \sqcup r & \textit{return types}\\
+r ::= & \;\; s \; | \; s \sqcup r & \textit{return types}\\
 \end{array}
@@
 \emph{first-level types} and \emph{second-level types}.
-Typed Lua uses first-level types to represent first-class Lua values and
-second-level types to represent tuples of values that appear in 
-multiple assignments and function applications.
+First-level types represent first-class Lua values and
+second-level types represent tuples of values that appear in 
+assignments and function applications.
 First-level types include literal types, base types, the type $\Nil$,
@@
 projection types.
-Second-level types include the type $\Void$, tuples of first-level
-types optionally ending in a variadic type, and unions of these tuples.
-
-Literal types represent the type of literal values.
+Second-level types include the type $\Void$, variadic types,
+tuple types, and unions of tuple types.
+
+Literal types $l$ represent the type of literal values.
 They can be the boolean values $\False$ and $\True$,
-any integer value, any floating point value, and any string value.
+an integer value, a floating point value, or a string value.
 We shall see that literal types are important in our treatment of
@@
 
-Our type system also includes the following base types: $\Boolean$,
-$\Integer$, $\Number$, and $\String$.
+Typed Lua includes four base types $b$: $\Boolean$, $\Integer$, $\Number$, and $\String$.
 The base types $\Boolean$ and $\String$ represent the values that during
@@
 \texttt{integer} for integer numbers and \texttt{float} for real numbers.
-Both \texttt{integer} and \texttt{float} values are subtype of \texttt{number}.
+Both \texttt{integer} and \texttt{float} values are a subtype of \texttt{number}.
 We introduced the base type $\Number$ to represent either the tag
@@
 \texttt{integer} values.
+In the next section we will show that $\Integer$ is subtype of $\Number$.
+This allows programmers to use \texttt{integer} values in the
+place of \texttt{float} values.
+
+The type $\Nil$ represents the type of the value that during run-time
+Lua tags as \texttt{nil}.
+
+The type $\Value$ is the top type.
+In Section \ref{sec:rules} we will show that this type,
+along with variadic types, helps the type system to drop extra values
+on assignments and function calls, thus preserving the
+semantics of Lua in these cases.
+
+Typed Lua includes the dynamic type $\Any$ for allowing programmers
+to mix static and dynamic typing.
+
+Typed Lua uses the type $\Self$ to represent the \emph{receiver}
+in object-oriented method definitions and method calls.
+As we mentioned in Chapter \ref{chap:typedlua}, we need the type
+$\Self$ to prevent programs from indexing a method without
+calling it with the correct receiver.
+
+Union types $t_{1} \cup t_{2}$ represent types that can hold a value
+of two different types.
+
+Function types have the form $s \rightarrow r$ to represent Lua functions,
+where $s$ is the type of the parameter list and $r$ is the return type.
+Both $s$ and $r$ are second-level types.
+In the next paragraphs we explain why we need different representations
+for the type of the parameter list and the return type.
+
+Second-level types $s$ are tuples of first-level types that can end
+with either an empty tuple or with a variadic type.
+Typed Lua needs second-level types because tuples are not first-class
+values in Lua, though they may appear on argument passing,
+multiple returns, and multiple assignments.
+The type $\Void$ is the type of an empty tuple.
+A variadic type $t*$ is a generator for a sequence of values
+of type $t \cup \Nil$, and it represents the type of a vararg expression.
+
+Second-level types $r$ include the same second-level types introduced
+by $s$ plus unions of tuple types.
+Typed Lua includes unions of tuples because Lua programs
+usually overload the return type of functions to denote error,
+as we mentioned in Section \ref{sec:statistics}.
+We could not find a nice concrete syntax to represent unions of
+tuples that appear in the type of the parameter list, and
+that is the reason why we restrict them to appear in the return type only.
+The same reason led us to split second-level types into two
+different symbols.
+We use the symbol $\sqcup$ to represent the union between two
+different tuple types.
+We have different unions because $\cup$ represents the union
+between two first-level types, while $\sqcup$ represents the
+union between two tuple types.
+
+Back to first-level types, table types represent the various forms
+that Lua tables can take.
+The syntactical form of table types is $\{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{tag}$,
+where each $k_{i}$ represents the type of a table key,
+and each $v_{i}$ represents the type of the value that table keys of type $k_{i}$ map to.
+Key types can only be literal types, base types, or the dynamic type.
+We made this restriction to the type of the keys because the statistics
+that we discussed in Section \ref{sec:statistics} showed that most
+of the tables are records, lists, and hashes.
+The dynamic type is an option when we need a loose table type.
+Value types can be any first-level type, and can optionally include
+the $\Const$ type to denote immutable values.
+
+We also use the tags \emph{closed}, \emph{open}, and \emph{unique}
+to classify table types.
+The tag \emph{unique} represents tables that we know their contents,
+and that there is no alias to them.
+In particular, the type of the table constructor has this tag.
+The tag \emph{open} represents \emph{unique} table types that
+have at least one alias.
+The tag \emph{closed} represents table types that do not provide
+any guarantees about the types not listed in the table type.
+In particular, type annotations as well as interfaces and userdata
+declarations always describe \emph{closed} table types.
+In the next sections we explain in more detail why we need
+different table types.
+
+Recursive types have the form $\mu x.t$,
+where $t$ is a first-level type that $x$ represents.
+For instance, $\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}$
+is a type for singly-linked lists of integers.
+
+Type Lua includes projection types as a way to project
+unions of tuple types into unions of first-level types.
+Projection types have the form $\pi_{i}^{x}$, where $x$ is a variable
+that stores an union of tuple types in a special type environment,
+and $i$ is an index that indicates how to project a first-level
+union type from $x$.
+For instance, if $x$ is the union of tuple types
+$(\Integer \times \Integer) \sqcup (\Nil \times \String)$,
+then $\pi_{1}^{x}$ represents the type $\Integer \cup \Nil$
+while $\pi_{2}^{x}$ represents the type $\Integer \cup \String$.
+In Section \ref{sec:rules} we show in more detail how our type system
+uses them as a mechanism for handling unions of tuple types,
+when they appear on the right-hand side of the declaration of local variables.
+We also show how this feature allows our type system to constrain
+the type of a local variable that depends on the type of another local variable.
+
+\section{Subtyping}
+\label{sec:subtyping}
+
+Our type system uses subtyping \citep{cardelli1984smi,abadi1996to} to order
+types and consistent-subtyping \citep{siek2007objects,siek2013mutable}
+to allow the interaction between statically and dynamically typed code.
+We will explain the subtyping and consistent-subtyping rules along this
+section while we introduce each type from our type system.
+We will focus the discussion on the definition of subtyping because,
+as we mentioned in Chapter \ref{chap:review}, we can combine the
+consistency and subtyping relations to achieve consistent-subtyping.
+The differences between subtyping and consistent-subtyping are the way
+they handle the dynamic type, and the fact that subtyping is transitive,
+but consistent-subtyping is not.
 
@@
 
-The type $\Nil$ represents the type of the value that during run-time
-Lua tags as \texttt{nil}.
 In our type system $\Nil$ is not the bottom type, and it is subtype
@@
 
-The type $\Value$ is the top type, so any first-level type is a
-subtype of $\Value$.
-In the next section we will show that this type along with
-variadic types help the type system drop extra values on
-multiple assignments and function calls, thus preserving the
-semantics of Lua in these cases.
-Following we show its subtyping rule:
+Any first-level type is a subtype of $\Value$:
 \[
@@
 
-The dynamic type $\Any$ allows programmers to mix static and dynamic typing;
-it is neither the bottom nor the top type,
+The dynamic type $\Any$ is neither the bottom nor the top type,
 but a separate type that is subtype only of itself through the reflexive rule.
@@
 
-Typed Lua uses the type $\Self$ to represent the \emph{receiver}
-in object-oriented method definitions and method calls.
-As we mentioned in Chapter \ref{chap:typedlua}, we need the type
-$\Self$ to prevent programmers from indexing a method without
-calling it with the correct \emph{receiver}.
 The type $\Self$ is subtype only of itself through the reflexive rule.
 
-Union types represent types that can hold a value of several
-different, but fixed types.
 The subtyping rules for union types are standard:
@@
 
-To introduce function types we first need to introduce
-second-level types, as they appear on both sides of function types.
-Second-level types are tuples of first-level types
-that can optionally end with a variadic type.
-Our type system needs second-level types because tuples are
-not first-class values in Lua, though they may appear on
-argument passing, multiple returns, and multiple assignments.
-A variadic type $t*$ is a generator for a sequence of values of
-type $t \cup \Nil$, and it represents the type of the vararg
-expression.
 The subtyping rule for tuple types is standard, but the subtyping
@@
 
-Function types have the form $p \rightarrow r$, where
-$p$ is the type of the parameter list and $r$ is the return type.
-The type of the parameter list can be $\Void$ or a tuple type,
-while the return type can be $\Void$, a tuple type, or a union of
-tuple types.
-The type $\Void$ means an empty tuple.
-Our type system includes unions of tuples on the return type because
-in Lua it is common to overload the return type to denote error,
-as we mentioned in Section \ref{sec:statistics}.
-
-The type $\Void$ and union of tuple types are also second-level types.
 The type $\Void$ is subtype only of itself through the reflexive rule.
@@
 \]
-
-Table types represent the various forms that Lua tables can take.
-The syntactical form of table types is $\{ k_{i}{:}v_{i}, ..., k_{n}{:}v_{n} \}_{tag}$.
-When $n = 0$, then $i = 0$ and it represents the type of an empty table.
-When $n > 0$, then $i$ varies from 1 to $n$, each $k_{i}$ represents
-the type of a table key, and each $v_{i}$ represents the type of the
-value that table keys of type $k_{i}$ map to.
-Key types can only be literal types, base types, and the dynamic type.
-We made this restriction to the type of the keys because the statistics
-that we discussed in Section \ref{sec:statistics} showed that most
-of the tables are records, lists, and hashes.
-The dynamic type is an option when we need a loose table type.
-Value types can be any first-level type, and can optionally include
-the $\Const$ attribute to denote immutable values.
-
-We also use the tags \emph{closed}, \emph{open}, and \emph{unique}
-to classify table types.
-The first tag represents the table types that the type annotations,
-interface, and userdata declarations describe.
-The second tag represents the table types that we can use field assignment
-to add new fields to them.
-The third tag represents the table types of table constructors.
-We have different subtyping rules for table types according to their
-tags.
 
@@
 
-Recursive types have the form $\mu x.t$,
-where $t$ is a first-level type with $x$ appearing anywhere this
-first-level type could appear.
-For instance, $\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}$
-is a type for singly-linked list of integers.
-
 We use the \emph{Amber rule} \citep{cardelli1986amber} to define
@@
 
-Projection types have the form $\pi_{i}^{x}$, where $x$ is a variable
-that stores an union of tuple types in a special type environment,
-and $i$ is an index that indicates how to build a first-level
-union type from $x$.
-For instance, if $x$ is the union of tuple types
-$(\Integer \times \Integer) \sqcup (\Nil \times \String)$,
-then $\pi_{1}^{x}$ represents the type $\Integer \cup \Nil$
-while $\pi_{2}^{x}$ represents the type $\Integer \cup \String$.
 Projection types follow the reflexive subtyping rule.
-In the next section we will show in more detail how our type system
-uses them as a mechanism for handling unions of tuple types,
-when they appear on the right-hand side of the declaration of local variables.
-We will also show how this feature allows our type system to constrain
-the type of a local variable that depends on the type of another local variable.
-
-\textit{/* How should we control invalid indices? */}
 
 \section{Typing rules}
+\label{sec:rules}
 

### 2014-12-17 doc/thesis/system.tex
Improvements to the section about subtyping

@@
 
+Any table type has to be \emph{well-formed}.
+Informally, a table type is well-formed if key types do not overlap.
+In Section \ref{sec:rules} we formalize the definition of well-formed table types.
+We delay the proper formalization of well-formed table types because we use
+consitent-subtyping in this formalization.
+
 Recursive types have the form $\mu x.t$,
@@
 to allow the interaction between statically and dynamically typed code.
-We will explain the subtyping and consistent-subtyping rules along this
-section while we introduce each type from our type system.
-We will focus the discussion on the definition of subtyping because,
+We explain the subtyping and consistent-subtyping rules along this section.
+However, we focus the discussion on the definition of subtyping because,
 as we mentioned in Chapter \ref{chap:review}, we can combine the
@@
 
+We present the subtyping rules as a deduction system for the
+subtyping relation $\senv \vdash t_{1} \subtype t_{2}$.
+The variable $\senv$ is a set of pairs of recursion variables.
+We shall notice that we need this set to record the hypotheses
+that we assume to check subtyping between recursive types.
+
 The subtyping rules for literal types and base types include the rules
@@
 and that $\Integer$ is subtype of $\Number$.
-We defined these subtyping rules as follows:
+We define these subtyping rules as follows:
 \[
@@
 
-The variable $\senv$ is a set of pairs of recursion variables.
-We shall notice that we need this set to record the pairs of valid
-recursive types.
-
-Subtyping is reflexive, thus, we include reflexive rules for
-first-level and second-level types as follows:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-FIRST}\\
-\senv \vdash t \subtype t
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-SECOND}\\
-\senv \vdash s \subtype s
-\end{array}
-\end{array}
-\]
-
-Subtyping is also transitive, and for this reason we could have omitted
+Subtyping is reflexive and transitive;
+due to the transitivity of subtyping, we could have omitted
 the rule \textsc{S-INT2}.
@@
 
-In our type system $\Nil$ is not the bottom type, and it is subtype
-only of itself through the reflexive rule.
+That is the reason why the dynamic type $\Any$ is neither the bottom nor the top type,
+but a separate type that is subtype only of itself through the reflexive rule.
+
+Even though the dynamic type $\Any$ does not interact with subtyping,
+it does interact with consistent-subtyping:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{C-ANY1}\\
+\senv \vdash t \lesssim \Any
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{C-ANY2}\\
+\senv \vdash \Any \lesssim t
+\end{array}
+\end{array}
+\]
+
+If we had set the type $\Any$ as both bottom and top types of our
+subtyping relation, then any type $t_{1}$ would be subtype of
+any other type $t_{2}$.
+The consequence of this is that all programs would type check without errors.
+This would happen due to the transitivity of subtyping, that is,
+we would be able to down-cast any type $t_{1}$ to $\Any$ and then up-cast
+$\Any$ to any other type $t_{2}$.
+The rules \textsc{C-ANY1} and \textsc{C-ANY2} are the rules that
+allow the dynamic type to interact with other first-level types,
+and thus allow dynamically typed code to coexist with statically
+typed code.
+Because of these two rules, consistent-subtyping cannot be transitive.
+Still, these two rules are the only rules that differ between
+subtyping and consistent-subtyping.
+
+Our type system includes the top type $\Value$,
+so any first-level type is a subtype of $\Value$:
+\[
+\begin{array}{c}
+\mylabel{S-VALUE}\\
+\senv \vdash t \subtype \Value
+\end{array}
+\]
+
+Some people might expect that $\Nil$ is the bottom type
+of our type system, but it is not,
+and it is subtype only of itself through the reflexive rule.
+Many programming languages include a bottom type to represent
+an empty value that programmers can use as a default expression
 Typed Lua has this behavior because in Lua the tag \texttt{nil}
@@
 
-Any first-level type is a subtype of $\Value$:
-\[
-\begin{array}{c}
-\mylabel{S-VALUE}\\
-\senv \vdash t \subtype \Value
-\end{array}
-\]
-
-The dynamic type $\Any$ is neither the bottom nor the top type,
-but a separate type that is subtype only of itself through the reflexive rule.
-
-Even though the dynamic type $\Any$ is neutral to subtyping,
-it is not to consistent-subtyping:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{C-ANY1}\\
-\senv \vdash t \lesssim \Any
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-ANY2}\\
-\senv \vdash \Any \lesssim t
-\end{array}
-\end{array}
-\]
-
-The rules that handle the dynamic type are the only rules that
-differ between subtyping and consistent-subtyping.
-If we had set the type $\Any$ as both bottom and top types of our
-subtyping relation, then all programs would type check without errors.
-This would happen due to the transitivity of subtyping, that is,
-we would be able to down-cast any type to $\Any$ and then up-cast
-$\Any$ to any type.
-The rules \textsc{C-ANY1} and \textsc{C-ANY2} are the rules that
-allow the dynamic type to interact with other first-level types,
-and thus allow dynamically typed code to coexist with statically
-typed code.
-Because of these two rules consistent-subtyping cannot be transitive.
-
-The type $\Self$ is subtype only of itself through the reflexive rule.
+Another type that only follows the reflexive rule is the type $\Self$,
+that is, it is subtype only of itself.
 
@@
 
-Union types must be disjoint, that is, the types in the union are not
-consistent-subtype of each other.
 In the implementation of Typed Lua we use consitent-subtyping to
@@
 formalization.
-For instance, the union type $\Boolean \;\cup\; \Any$ results the
-type $\Any$, because $\Boolean \lesssim \Any$;
-and the union type $\Number \;\cup\; \Nil \;\cup\; 1$ results the union
-type $\Number \;\cup\; \Nil$ because $1 \lesssim \Number$.
+For instance, the union type \texttt{boolean|any} results in the
+type \texttt{any}, because \texttt{boolean} is consistent-subtype
+of \texttt{any}.
+Another example is the union type \texttt{number|nil|1} that
+results in the union type \texttt{number|nil}, because
+\texttt{1} is consistent-subtype of \texttt{number}.
+
+The subtyping rule for function types is also standard:
+\[
+\begin{array}{c}
+\mylabel{S-FUNCTION}\\
+\dfrac{\senv \vdash s_{2} \subtype s_{1} \;\;\;
+       \senv \vdash r_{1} \subtype r_{2}}
+      {\senv \vdash s_{1} \rightarrow s_{1} \subtype s_{2} \rightarrow r_{2}}
+\end{array}
+\]
+
+The rule \textsc{S-FUNCTION} shows that subtyping between
+function types is contravariant on the type of the parameter list
+and covariant on the return type.
+In the last section we explained why our type system uses
+second-level types to represent the type of the parameter list
+and the return type.
+Now, we explain their subtyping rules.
+
+The simplest second-level type is the type $\Void$ that
+is subtype only of itself through the reflexive rule.
 
@@
 \dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2}}
-      {\senv \vdash t_{1}{*} \subtype t_{2}}
+      {\senv \vdash t_{1}{*} \subtype t_{2} \times \Void}
 \end{array}
@@
 \dfrac{\senv \vdash t_{1} \subtype t_{2} \cup \Nil}
-      {\senv \vdash t_{1} \subtype t_{2}{*}}
+      {\senv \vdash t_{1} \times \Void \subtype t_{2}{*}}
 \end{array}
@@
 we need five different subtyping rules for variadic types
-to handle all the cases that they can appear.
-The rule \textsc{S-TUPLE1} shows that a tuple type $t_{1} \times s_{1}$
-is subtype of another tuple type $t_{2} \times s_{2}$ if
-$t_{1}$ is subtype of $t_{2}$ and $s_{1}$ is subtype of $s_{2}$.
-The rule \textsc{S-VARARG1} handles the subtyping between two variadic types.
-It shows that $t_{1}{*}$ is subtype of $t_{2}{*}$
+to handle all the cases where they can appear.
+The rule \textsc{S-VARARG1} handles the subtyping between two variadic types,
+while the other rules handle the subtyping between a tuple type and a variadic type.
+The rule \textsc{S-VARARG1} shows that $t_{1}{*}$ is subtype of $t_{2}{*}$
 if $t_{1} \cup \Nil$ is subtype of $t_{2} \cup \Nil$.
@@
 as $\Nil \not\subtype \Number$.
-The rules \textsc{S-VARARG2} and \textsc{S-VARARG3} handles the
-subtyping between a firt-level type and a variadic type.
 The rule \textsc{S-VARARG2} shows that a variadic type $t_{1}{*}$ is
-subtype of a firt-level type $t_{2}$ if $t_{1} \cup \Nil$ is subtype of $t_{2}$.
-The rule \textsc{S-VARARG3} shows that a firt-level type $t_{1}$
+subtype of a tuple type $t_{2} \times \Void$ if $t_{1} \cup \Nil$ is subtype of $t_{2}$.
+The rule \textsc{S-VARARG3} shows that a tuple type $t_{1} \times \Void$
 is subtype of a variadic type $t_{2}{*}$ if $t_{1}$ is subtype of $t_{2} \cup \Nil$.
-The rules \textsc{S-VARARG4} and \textsc{S-VARARG5} handle the
-subtyping between a tuple type and a variadic type.
 The rule \textsc{S-VARARG4} shows that a variadic type is subtype of
@@
 the type of the value that it generates.
-In the next section we will show that we use these subtyping relations
-along with the types $\Value$ and $\Nil$ to make our type system reflect
-the semantics of Lua on discarding extra parameters and
+In the next section we will show that we use these subtyping relations,
+along with the types $\Value$ and $\Nil$, to make our type system reflect
+the semantics of Lua on discarding extra parameters, and
 replacing missing parameters.
 
-The type $\Void$ is subtype only of itself through the reflexive rule.
 The subtyping rules for unions of tuple types are similar to the
-subtyping rules for unions of first-level types.
-
-The subtyping rule for function types is standard, that is, it is
-contravariant on the type of the parameter list and
-covariant on the return type:
-\[
-\begin{array}{c}
-\mylabel{S-FUNCTION}\\
-\dfrac{\senv \vdash p_{2} \subtype p_{1} \;\;\;
-       \senv \vdash r_{1} \subtype r_{2}}
-      {\senv \vdash p_{1} \rightarrow r_{1} \subtype p_{2} \rightarrow r_{2}}
-\end{array}
-\]
-
-The subtyping rule between \emph{closed} table types resembles the
+subtyping rules for unions of first-level types:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-UNION4}\\
+\dfrac{\senv \vdash s_{1} \subtype s \;\;\;
+       \senv \vdash s_{2} \subtype s}
+      {\senv \vdash s_{1} \sqcup s_{2} \subtype s}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-UNION5}\\
+\dfrac{\senv \vdash s \subtype s_{1}}
+      {\senv \vdash s \subtype s_{1} \sqcup s_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-UNION6}\\
+\dfrac{\senv \vdash s \subtype s_{2}}
+      {\senv \vdash s \subtype s_{1} \sqcup s_{2}}
+\end{array}
+\end{array}
+\]
+
+Back to the subtyping rules between first-level types,
+the subtyping rule between \emph{closed} table types resembles the
 standard subtyping rule between records:
@@
 \begin{array}{c}
-\mylabel{S-TABLE1}\\
+\mylabel{S-TABLE}\\
 \dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
@@
        \senv \vdash k_{i}' \subtype k_{j} \;\;\;
-       \senv \vdash v_{j} \subtype v_{i}'}
-      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{closed} \subtype \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
-\end{array}
-\]
-
-The rule \textsc{S-TABLE1} allows \emph{closed} table types to
-have width subtyping, and depth subtyping on $\Const$ fields.
-The subtyping rules for the type of the values stored in a \emph{closed}
-table type describe this behavior:
+       \senv \vdash v_{j} \subtype_{c} v_{i}'}
+      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{closed} \subtype \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}} \; m \ge n
+\end{array}
+\]
+
+The rule \textsc{S-TABLE} introduced the relation $\subtype_{c}$.
+It is just an auxiliary relation that defines the subtyping of the
+type of the values stored in the table fields.
+We need an auxiliary relation because the subtyping of the
+type of the values stored in the table fields changes according to
+the tags of the table types.
+We define the relation $\subtype_{c}$ as follows:
 \[
@@
        \senv \vdash v_{2} \subtype v_{1}}
-      {\senv \vdash v_{1} \subtype v_{2}}
+      {\senv \vdash v_{1} \subtype_{c} v_{2}}
 \end{array}
@@
 \dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash \Const \; v_{1} \subtype \Const \; v_{2}}
+      {\senv \vdash \Const \; v_{1} \subtype_{c} \Const \; v_{2}}
 \end{array}
@@
 \dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash v_{1} \subtype \Const \; v_{2}}
-\end{array}
-\end{array}
-\]
-
-The rule \textsc{S-FIELD1} handles mutable values invariantly,
-while the rule \textsc{S-FIELD2} handles immutable values covariantly.
-It is also safe to promote a mutable value to immutable, and the rule
-\textsc{S-FIELD3} defines this behavior.
-However, it is not safe to promote an immutable value to mutable,
-and that is the reason why we do not have a subtying rule for this case.
+      {\senv \vdash v_{1} \subtype_{c} \Const \; v_{2}}
+\end{array}
+\end{array}
+\]
+
+These rules allow closed table types to have width subtyping
+and depth subtyping on $\Const$ fields.
+The rule \textsc{S-FIELD1} defines that mutable fields are invariant,
+while the rule \textsc{S-FIELD2} defines that immutable fields are covariant.
+The rule \textsc{S-FIELD3} defines that it is safe to promote fields
+from mutable to immutable.
+We do not include a rule that allows promoting fields from immutable
+to mutable because it would not be safe, as this rule would allow
+programmers to change the value that is stored in an immutable field.
 
@@
 programmers would not be able to use it to initialize a variable with
-a table type that describe a more general type.
+a table type that describes a more general type.
 For instance,
@@
 \end{verbatim}
-whould not type check, as the type of the table constructor would not
+would not type check, as the type of the table constructor would not
 be subtype of the type in the annotation.
@@
 \]
+
 Thus, we introduced \emph{unique} table types to represent the type of
@@
 
-We introduced \emph{open} table types to allow the refinement of table types,
+Moreover, we introduced \emph{open} table types to allow the refinement of table types,
 an idiom very common in Lua.
 
-Even though \emph{open} and \emph{unique} table types have specific
-subtyping rules, we will present them in the next section along with
-the typing rules that formalize the refinement of table types.
-The presentation of these subtyping rules is easier in the next section,
+Even though \emph{open} and \emph{unique} table types can interact with
+\emph{closed} table types, they do not follow the subtyping rules.
+Instead, we define a specific relation that handles the interaction
+between the different table types that our type system has.
+We will present this relation in the next section along with the typing
+rules that formalize the refinement of table types.
+The presentation of this relation is easier in the next section,
 as they include some unsafe behaviors that the refinement of
@@
 
-Any table type has to be \emph{well-formed}.
-Informally, a table type is well-formed if key types do not overlap.
-Formally, a table type is well-formed if it obeys the following rule:
-\[
-\forall i \not\exists j \; i \not= j \wedge k_{i} \lesssim k_{j}
-\]
-
-Well-formed table types avoid ambiguous table types.
-For instance, this rule detects that the table type
-$\{1:\Number, \Integer:\String, \Any:\Boolean\}$ is ambiguous,
-because the type of the value stored by key $1$ can be
-$\Number$, $\String$, or $\Boolean$, as $1 \lesssim 1$,
-$1 \lesssim \Integer$, and $1 \lesssim \Any$.
-Moreover, the type of the value stored by a key of type $\Integer$,
-which is not the literal type $1$, can be $\Number$ or $\Boolean$,
-as $\Integer \lesssim \Integer$, and $\Integer \lesssim \Any$.
-
 We use the \emph{Amber rule} \citep{cardelli1986amber} to define
@@
 \begin{array}{c}
+\begin{array}{c}
+\mylabel{S-AMBER}\\
+\dfrac{\senv[x_{1} \subtype x_{2}] \vdash t_{1} \subtype t_{2}}
+      {\senv \vdash \mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}}
+\end{array}
+\;
 \begin{array}{c}
@@
 \end{array}
-\;
-\begin{array}{c}
-\mylabel{S-AMBER}\\
-\dfrac{\senv[x_{1} \subtype x_{2}] \vdash t_{1} \subtype t_{2}}
-      {\senv \vdash \mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}}
-\end{array}
-\end{array}
-\]
-
-The rule \textsc{S-AMBER} needs the rule \textsc{S-ASSUMPTION}
-to show that $\mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}$.
-The rule \textsc{S-ASSUMPTION} checks if the assumption
-$x_{1} \subtype x_{2}$ exists in the set of assumptions $\senv$.
-The rule \textsc{S-AMBER} extends the set of assumptions
-with the assumption $x_{1} \subtype x_{2}$ to check whether
-$t_{1} \subtype t_{2}$ or not.
+\end{array}
+\]
+
+The rule \textsc{S-AMBER} also uses the rule \textsc{S-ASSUMPTION}
+to check whether $\mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}$.
+Both rules use the set of assumptions $\senv$,
+which are pairs of recursion variables.
+The rule \textsc{S-AMBER} extends $\senv$ with the assumption
+$x_{1} \subtype x_{2}$ to check whether $t_{1} \subtype t_{2}$.
+The rule \textsc{S-ASSUMPTION} allows the rule \textsc{S-AMBER}
+to check whether an assumption is valid.
 
@@
 type system includes subtyping rules to handle subtyping between
-a first-level type and a recursive type:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-AMBERL}\\
+recursive types and other first-level types:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-UNFOLDR}\\
+\dfrac{\senv \vdash t_{1} \subtype [x \mapsto \mu x.t_{2}]t_{2}}
+      {\senv \vdash t_{1} \subtype \mu x.t_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-UNFOLDL}\\
 \dfrac{\senv \vdash [x \mapsto \mu x.t_{1}]t_{1} \subtype t_{2}}
@@
 \end{array}
-\;
-\begin{array}{c}
-\mylabel{S-AMBERR}\\
-\dfrac{\senv \vdash t_{1} \subtype [x \mapsto \mu x.t_{2}]t_{2}}
-      {\senv \vdash t_{1} \subtype \mu x.t_{2}}
-\end{array}
-\end{array}
-\]
-
-For instance, it is the rule \textsc{S-AMBERR} that allows our
+\end{array}
+\]
+
+For instance, it is the rule \textsc{S-UNFOLDR} that allows our
 type system to type check the example that we presented
@@
 the table constructor.
-It type checks because the rule \textsc{S-AMBERR} verifies that
+It type checks because the rule \textsc{S-UNFOLDR} verifies that
 the type of the table constructor is a subtype of \texttt{Element}:
@@
 Projection types follow the reflexive subtyping rule.
+A projection type $\pi_{i}^{x}$ is a subtype of another projection type $\pi_{i}^{x}$
+that shares the same union of tuples $x$ and the same index $i$.
 
@@
 whether the table type is well formed.
-We gave the definition of this predicate in the last section.
+Formally, a table type is well-formed if it obeys the following rule:
+\[
+\forall i \not\exists j \; i \not= j \wedge k_{i} \lesssim k_{j}
+\]
+
+Well-formed table types avoid ambiguous table types.
+For instance, this rule detects that the table type
+$\{1:\Number, \Integer:\String, \Any:\Boolean\}$ is ambiguous,
+because the type of the value stored by key $1$ can be
+$\Number$, $\String$, or $\Boolean$, as $1 \lesssim 1$,
+$1 \lesssim \Integer$, and $1 \lesssim \Any$.
+Moreover, the type of the value stored by a key of type $\Integer$,
+which is not the literal type $1$, can be $\Number$ or $\Boolean$,
+as $\Integer \lesssim \Integer$, and $\Integer \lesssim \Any$.
 

### 2014-12-29 doc/thesis/system.tex
Minor improvements to the text about types and subtyping

@@
 tuples that appear in the type of the parameter list, and
-that is the reason why we restrict them to appear in the return type only.
+that is the reason why we restricted them to appear in the return type only.
 The same reason led us to split second-level types into two
@@
 We delay the proper formalization of well-formed table types because we use
-consitent-subtyping in this formalization.
+consistent-subtyping in this formalization.
 
@@
 is a type for singly-linked lists of integers.
+In Section \ref{sec:alias} we mentioned that we can use the following
+interface declaration as an alias to this type:
+\begin{verbatim}
+    local interface Element
+      info:integer
+      next:Element?
+    end
+\end{verbatim}
 
@@
 Subtyping is reflexive and transitive;
-due to the transitivity of subtyping, we could have omitted
-the rule \textsc{S-INT2}.
+due to the transitivity of subtyping, we could have omitted the rule \textsc{S-INT2}.
 More precisely, we could have defined a transitive rule for first-level
@@
 Many programming languages include a bottom type to represent
-an empty value that programmers can use as a default expression
-Typed Lua has this behavior because in Lua the tag \texttt{nil}
+an empty value that programmers can use as a default expression.
+Typed Lua does not have this behavior because in Lua the tag \texttt{nil}
 represents the absence of an useful value.
@@
 
-In the implementation of Typed Lua we use consitent-subtyping to
+In the implementation of Typed Lua we use consistent-subtyping to
 normalize and simplify union types, but we let them free in the
@@
 The subtyping rule for tuple types is standard, but the subtyping
-rules for variadic types are not obvious:
+rules for variadic types are not so obvious:
 \[
@@
 \mylabel{S-VARARG4}\\
-\dfrac{\senv \vdash t_{1}{*} \subtype t_{2} \;\;\;
+\dfrac{\senv \vdash t_{1}{*} \subtype t_{2} \times \Void \;\;\;
        \senv \vdash t_{1}{*} \subtype s_{2}}
@@
 \mylabel{S-VARARG5}\\
-\dfrac{\senv \vdash t_{1} \subtype t_{2}{*} \;\;\;
+\dfrac{\senv \vdash t_{1} \times \Void \subtype t_{2}{*} \;\;\;
        \senv \vdash s_{1} \subtype t_{2}{*}}
@@
 \begin{array}{c}
-\mylabel{S-TABLE}\\
+\mylabel{S-TABLE1}\\
 \dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
@@
 
-The rule \textsc{S-TABLE} introduced the relation $\subtype_{c}$.
+The rule \textsc{S-TABLE1} introduced the relation $\subtype_{c}$.
 It is just an auxiliary relation that defines the subtyping of the
@@
 Even though \emph{open} and \emph{unique} table types can interact with
-\emph{closed} table types, they do not follow the subtyping rules.
-Instead, we define a specific relation that handles the interaction
-between the different table types that our type system has.
-We will present this relation in the next section along with the typing
+\emph{closed} table types, they follow different subtyping rules.
+We will present these rules in the next section along with the typing
 rules that formalize the refinement of table types.
-The presentation of this relation is easier in the next section,
+The presentation of these rules is easier in the next section,
 as they include some unsafe behaviors that the refinement of
@@
 Both rules use the set of assumptions $\senv$,
-which are pairs of recursion variables.
+where each assumption is a pair of recursion variables.
 The rule \textsc{S-AMBER} extends $\senv$ with the assumption
@@
 
-For instance, it is the rule \textsc{S-UNFOLDR} that allows our
-type system to type check the example that we presented
-in Section \ref{sec:tables}:
+The rules \textsc{S-UNFOLDR}, \textsc{S-AMBER}, and \textsc{S-ASSUMPTION}
+allow our type system to type check the function \texttt{insert}
+from Section \ref{sec:alias}:
 \begin{verbatim}
-    local interface Element
-      info:integer
-      next:Element?
-    end
     local function insert (e:Element?, v:integer):Element
@@
 
-The interface \texttt{Element} defines the type for singly-linked
-list of integers that we just mentioned, and the function
-\texttt{insert} inserts an element in the begining of the list.
-In this example, the recursive type appears inside the type of
-the table constructor.
-It type checks because the rule \textsc{S-UNFOLDR} verifies that
+First, a recursive type appears inside the type of the table constructor.
+Thus, the type checker uses the rule \textsc{S-UNFOLDR} to verify whether
 the type of the table constructor is a subtype of \texttt{Element}:
@@
    & ``next":\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed} \cup \Nil \}_{unique} \subtype \\
+& \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}
+\end{align*}
+
+After unfolding, the recursive types appear only as part of the type
+of the value of the key type $``next"$.
+Thus, the type checker uses the rules \textsc{S-AMBER} and \textsc{S-ASSUMPTION}
+to verify whether these recursive types are subtypes between each other:
+\begin{align*}
+& \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed} \subtype \\
 & \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}

### 2015-01-01 doc/thesis/system.tex
Moving the text about subtyping of unique and open table types back to the section about subtyping

@@
 There is a limitation on \emph{closed} table types that led us to
-introduce \emph{unique} table types.
+introduce \emph{open} and \emph{unique} table types.
 If the table constructor had a \emph{closed} table type, then
@@
 
-Thus, we introduced \emph{unique} table types to represent the type of
-the table constructor and avoid this limitation.
-
-Moreover, we introduced \emph{open} table types to allow the refinement of table types,
-an idiom very common in Lua.
-
-Even though \emph{open} and \emph{unique} table types can interact with
-\emph{closed} table types, they follow different subtyping rules.
-We will present these rules in the next section along with the typing
-rules that formalize the refinement of table types.
-The presentation of these rules is easier in the next section,
-as they include some unsafe behaviors that the refinement of
-table types keeps track to make them safe.
+Simply promoting the type of each table value to its supertype would
+not overcome this limitation, as it still would give to the table constructor
+a closed table type without covariant mutable fields.
+Thus, programmers would not be able to use the table constructor to
+initialize a variable with a table type that includes an optional field.
+Using the previous example,
+\begin{align*}
+& \{``x":\Integer, ``y":\Integer\}_{closed} \not\subtype \\
+& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
+\end{align*}
+
+We introduced \emph{unique} table types to avoid this limitation,
+as they represent the type of the tables that we known their contents.
+In particular, this is the case of the table constructor.
+The following subtyping rule defines the subtyping relation among
+\emph{unique} table types and other table types:
+\[
+\begin{array}{c}
+\mylabel{S-TABLE2}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
+       \senv \vdash v_{i} \subtype_{u} v_{i}' \;\;\;
+       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
+      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
+                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open|unique}}
+\end{array}
+\]
+
+Unlike the rule \textsc{S-TABLE1}, the rule \textsc{S-TABLE2} allows
+covariant keys.
+We allow this behavior to \emph{unique} table types because we also
+want to use them as a way to join table fields.
+For instance, we may want to use the table constructor to initialize
+a variable with a table type that describes a hash.
+
+The rule \textsc{S-TABLE2} introduced the auxiliary relations
+$\subtype_{u}$ and $\subtype_{o}$.
+The first allows width and depth subtyping on all fields,
+while the second allows the omission of optional fields.
+We define them as follows:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-FIELD4}\\
+\dfrac{\senv \vdash v_{1} \subtype v_{2}}
+      {\senv \vdash v_{1} \subtype_{u} v_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FIELD5}\\
+\dfrac{\senv \vdash v_{1} \subtype v_{2}}
+      {\senv \vdash \Const \; v_{1} \subtype_{u} \Const \; v_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FIELD6}\\
+\dfrac{\senv \vdash v_{1} \subtype v_{2}}
+      {\senv \vdash v_{1} \subtype_{u} \Const \; v_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-FIELD7}\\
+\dfrac{\senv \vdash \Nil \subtype v}
+      {\senv \vdash \Nil \subtype_{o} v}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FIELD8}\\
+\dfrac{\senv \vdash \Nil \subtype v}
+      {\senv \vdash \Nil \subtype_{o} \Const \; v}
+\end{array}
+\end{array}
+\]
+
+Using \emph{unique} table types to represent the type of the table
+constructor allows our type system to type check the previous example.
+More precisely,
+\[
+\{``x":1, ``y":2\}_{unique} \subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
+\]
+
+Even though \emph{unique} table types overcome our limitation,
+its subtyping rule allows unsafe aliasing.
+The covariance on mutable fields allows programs that create unsafe
+aliases to pass the type checker.
+For instance,
+\begin{verbatim}
+    local t1 = { foo = 5 }
+    local t2:{"foo":integer?} = t1
+    t2.foo = nil
+\end{verbatim}
+should not type check, as the last line is erasing the field \texttt{foo}
+from \texttt{t1}.
+This program would type checks as it is, because our implementation uses
+local type inference to assign static types to unannotated locals.
+In this example, the type of \texttt{t1} would be subtype of the type of \texttt{t2}.
+More precisely,
+\[
+\{``foo":\Integer\}_{unique} \subtype \{``foo":\Integer \cup \Nil\}_{closed}
+\]
+
+In the next section we will show in more detail that this example
+does not type check, because our type system promotes an alias
+expression from \emph{unique} to \emph{closed}, and promotes the
+aliased variable from \emph{unique} to \emph{open}.
+This means that the previous example does not type check because
+the type of the expression \texttt{t1} is not subtype of the type
+of the variable \texttt{t2}.
+More precisely,
+\[
+\{``foo":\Integer\}_{closed} \not\subtype \{``foo":\Integer \cup \Nil\}_{closed}
+\]
+
+Thus, we introduced \emph{open} table types to handle the aliasing of table types.
+We define the subtyping relation among \emph{open} table types
+and \emph{closed} or \emph{open} table types as follows:
+\[
+\begin{array}{c}
+\mylabel{S-TABLE3}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
+       \senv \vdash v_{i} \subtype_{c} v_{i}' \;\;\;
+       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
+      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
+                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open}}
+\end{array}
+\]
+
+The rule \textsc{S-TABLE3} uses the same rules from \textsc{S-TABLE1}
+for handling the type of the values of each key, and it uses the same
+rules from \textsc{S-TABLE2} to allow joining fields plus omitting optional fields.
+
+In the next section we will show in more detail how our type system
+keeps track of \emph{open} and \emph{unique} table types to make them safe.
+Moreover, we will also show that we can use them to allow the refinement
+of table types.
 
@@
 
-We also included two kinds of type cast in our core language:
+We also include two kinds of type cast in our core language:
 the expression ${<}t{>} \;n$ and the left-hand value $n[c] \; {<}t{>}$.
@@
 
-The most unusual feature of our type system is the refinement of
-table types.
+\textit{/* \textbf{begin} TODO list */}
+
+\begin{itemize}
+\item Refinement with \_ENV
+\item Assignment + left-hand side
+\item Locals + inference rules + unique and closed table types
+\end{itemize}
+
+\textit{/* \textbf{end} TODO list */}
+
+We start with the rule that defines the most unusual feature of
+our type system: the refinment of table types.
 The rule \textsc{T-REFINE} allow us to add new keys to \emph{open}
@@
 \mylabel{T-REFINE}\\
-\dfrac{\env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique} \;\;\;
+\dfrac{\begin{array}{c}
+       \env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique}\\
        \env_{1} \vdash c:t_{1}, \env_{2} \;\;\;
-       \not \exists i \in 1..n \; t_{1} \lesssim k_{i}}
+       \not \exists i \in 1..n \; t_{1} \lesssim k_{i} \;\;\;
+       \Nil \not\lesssim t_{1}
+       \end{array}}
       {\env_{1} \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, t_{1}{:}t\}_{open|unique}]}
@@
 
-We introduced \emph{open} table types to allow the refinement
-of table types, and also to allow programmers to omit optional
-fields in the initialization of table types.
-To do that, we defined a subtyping rule that allow \emph{open}
-table types to omit fields when $\Nil$ is subtype of the type of
-the value of the omitted key.
-The following subtyping rule show this behavior:
-\[
-\begin{array}{c}
-\mylabel{S-TABLE2}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
-       \senv \vdash v_{i} \subtype_{c} v_{i}' \;\;\;
-       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
-      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
-                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open}}
-\end{array}
-\]
-
-The rule \textsc{S-TABLE2} introduced the relation $\subtype_{o}$.
-It is an auxiliary relation that defines the subtyping relation
-on optional fields, and we defined it as follows:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-FIELD4}\\
-\dfrac{\senv \vdash \Nil \subtype v}
-      {\senv \vdash \Nil \subtype_{o} v}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FIELD5}\\
-\dfrac{\senv \vdash \Nil \subtype v}
-      {\senv \vdash \Nil \subtype_{o} \Const \; v}
-\end{array}
-\end{array}
-\]
-
-Although \emph{open} table types let programmers omit optional fields,
-they do not let programmers initialize optional fields that are
-mutable, unless they use the same type of the field in the initialization.
+\textit{/* Should we split into two rules instead of using $open|unique$? */}
+
+We restricted the refinement of table types to include only literal
+keys, because its purpose is to make it easier the construction of
+table types that represent records.
+
+We use the refinement of table types to handle the declaration of global variables.
+In Lua, the assignment \texttt{v = 5} is syntactic sugar to \texttt{\string_ENV["v"] = 5}
+when \texttt{v} is not a local variable, where \texttt{\string_ENV} is a table
+that stores the global environment.
+For this reason, Typed Lua treats accesses to global variables as field accesses
+to an open table in the top-level scope.
 For instance,
-\begin{align*}
-\{ & ``firstname":\String,\\
-   & ``middlename":\String, ``lastname":\String \}_{open} \not \subtype \\
-\{ & ``firstname":\String, \\
-   & ``middlename":\String \cup \Nil, ``lastname":\String \}_{closed}
-\end{align*}
-
-We introduced \emph{unique} table types to solve this problem.
-We use \emph{unique} table types to denote the type of the
-table constructor.
-The following subtyping rule allow programmers to use table constructors
-to initialize and omit optional fields:
-\[
-\begin{array}{c}
-\mylabel{S-TABLE3}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
-       \senv \vdash v_{i} \subtype_{u} v_{i}' \;\;\;
-       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
-      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
-                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open|unique}}
-\end{array}
-\]
-
-The rule \textsc{S-TABLE3} introduced the relation $\subtype_{u}$.
-It is an auxiliary relation that makes \emph{unique} table types
-have width and depth subtyping on all fields, and we defined it
-as follows:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-FIELD6}\\
-\dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash v_{1} \subtype_{u} v_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FIELD7}\\
-\dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash \Const \; v_{1} \subtype_{u} \Const \; v_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FIELD8}\\
-\dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash v_{1} \subtype_{u} \Const \; v_{2}}
-\end{array}
-\end{array}
-\]
-
-We restricted the refinement of table types to include only literal
-keys because its purpose is to make it easier the construction of
-table types that represent records.
-For instance,
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
-& \multicolumn{1}{l}{$a[``y"] \; {<}\Integer{>} = 1$}\\
-\end{tabular}
-\end{center}
-uses field assignment to add fields $x$ and $y$ to the local $a$.
-After these field assignments $a$ has type
-$\{``x":\String, ``y":\Integer\}_{unique}$.
+\begin{center}
+\begin{tabular}{l}
+$\string_ENV[``x"] \; {<}\String{>} = ``foo";$\\
+$\string_ENV[``y"] \; {<}\Integer{>} = 1$\\
+\end{tabular}
+\end{center}
+uses field assignment to add fields $``x"$ and $``y"$ to $\string_ENV$.
+We are assuming that $\string_ENV$ is in the environment and has type $\{\}_{open}$.
+Therefore, after these field assignments $\string_ENV$ has type
+$\{``x":\String, ``y":\Integer\}_{open}$.
 
@@
 \begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
-& \multicolumn{1}{l}{$a[``x"] \; {<}\Integer{>} = 1$}\\
-\end{tabular}
-\end{center}
-do not type check, as we are trying to change the type of field
-``x" from $\String$ to $\String \cup \Integer$.
-More precisely, our type system does not allow the table type to
-change towards a type that is not a subtype of the previous type.
-In this example,
-$\{``x":\String \cup \Integer\}_{unique} \not\subtype \{``x":\String\}_{unique}$.
-We have this restriction because this kind of refinement could lead
-to fields that have a too general type.
+\begin{tabular}{l}
+$\string_ENV[``x"] \; {<}\String{>} = ``foo";$\\
+$\string_ENV[``x"] \; {<}\Integer{>} = 1$\\
+\end{tabular}
+\end{center}
+do not type check, as we are trying to add a field that already exists
+in $\string_ENV$.
+We could allow the refinement of existing fields, but this could
+lead to fields that have a too general type, which could decrease
+the ammout of type checking.
 
@@
 \begin{tabular}{l}
-$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in} \; a[``x"] \; {<}\{\}_{closed}{>} = \{\}$
+$\string_ENV[``x"] \; {<}\{\}_{unique}{>} = \{\}$
 \end{tabular}

### 2015-01-02 doc/thesis/system.tex
Adding text about globals in the typing rules section

@@
 
-\textit{/* \textbf{begin} TODO list */}
-
-\begin{itemize}
-\item Refinement with \_ENV
-\item Assignment + left-hand side
-\item Locals + inference rules + unique and closed table types
-\end{itemize}
-
-\textit{/* \textbf{end} TODO list */}
-
 We start with the rule that defines the most unusual feature of
@@
        \not \exists i \in 1..n \; t_{1} \lesssim k_{i} \;\;\;
-       \Nil \not\lesssim t_{1}
+       \Nil \not\lesssim t_{1} \;\;\;
+       t_{2} = close(t)
        \end{array}}
-      {\env_{1} \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, t_{1}{:}t\}_{open|unique}]}
-\end{array}
-\]
-
-\textit{/* Should we split into two rules instead of using $open|unique$? */}
+      {\env_{1} \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, t_{1}{:}t_{2}\}_{open|unique}]}
+\end{array}
+\]
 
@@
 
-We use the refinement of table types to handle the declaration of global variables.
-In Lua, the assignment \texttt{v = 5} is syntactic sugar to \texttt{\string_ENV["v"] = 5}
-when \texttt{v} is not a local variable, where \texttt{\string_ENV} is a table
+We use the refinement of table types to handle global variables.
+In Lua, the assignment \texttt{v = v + 1} translates to
+\texttt{\string_ENV["v"] = \string_ENV["v"] + 1} when \texttt{v}
+is not a local variable, where \texttt{\string_ENV} is a table
 that stores the global environment.
@@
 For instance,
-\begin{center}
-\begin{tabular}{l}
-$\string_ENV[``x"] \; {<}\String{>} = ``foo";$\\
-$\string_ENV[``y"] \; {<}\Integer{>} = 1$\\
-\end{tabular}
-\end{center}
+\[
+\string_ENV[``x"] \; {<}\String{>} = ``foo" \;;\; \string_ENV[``y"] \; {<}\Integer{>} = 1
+\]
 uses field assignment to add fields $``x"$ and $``y"$ to $\string_ENV$.
-We are assuming that $\string_ENV$ is in the environment and has type $\{\}_{open}$.
+In all the examples we assume that $\string_ENV$ is in the environment and
+has type $\{\}_{open}$.
 Therefore, after these field assignments $\string_ENV$ has type
@@
 For instance,
-\begin{center}
-\begin{tabular}{l}
-$\string_ENV[``x"] \; {<}\String{>} = ``foo";$\\
-$\string_ENV[``x"] \; {<}\Integer{>} = 1$\\
-\end{tabular}
-\end{center}
+\[
+\string_ENV[``x"] \; {<}\String{>} = ``foo" \;;\; \string_ENV[``x"] \; {<}\Integer{>} = 1
+\]
 do not type check, as we are trying to add a field that already exists
 in $\string_ENV$.
-We could allow the refinement of existing fields, but this could
-lead to fields that have a too general type, which could decrease
-the ammout of type checking.
-
-\textit{/* We need to close table types before assigning them to fields. */}
+We could have used union types to allow the refinement of existing fields,
+but this could lead to fields that have a too general type, which could
+decrease the ammount of static type checking.
+This means that our type system does not allow the table type to change
+towards a type that is not a subtype of the previous type.
+In this example,
+\[
+\{``x":\String \cup \Integer\}_{open} \not\subtype \{``x":\String\}_{open}
+\]
+
+We also do not allow the refinement of table types to introduce
+\emph{open} or \emph{unique} table fields.
+For instance,
 \begin{center}
@@
 \end{center}
+refines the type of $\string_ENV$ from $\{\}_{open}$ to $\{``x":\{\}_{closed}\}_{open}$,
+though we are trying to refine to $\{``x":\{\}_{unique}\}_{open}$.
+We made this restriction because we still need to prove that the refinement
+of table types is safe.
 
@@
 \dfrac{\env \vdash l_{k}:t_{k}, \env_{k} \;\;\;
-       merge(\env_{1}, ..., \env_{n}) = \env_{m} \;\;\;
+       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
        n = |\;\vec{l}\;|}
@@
 We can also use multiple assignments to refine table types:
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$a[``x"] \; {<}\String{>}, a[``y"] \; {<}\Integer{>} = ``foo", 1$}\\
-\end{tabular}
-\end{center}
+\[
+\string_ENV[``x"] \; {<}\String{>}, \string_ENV[``y"] \; {<}\Integer{>} = ``foo", 1
+\]
 
@@
 The type of the expression list includes the type $\Nil{*}$ in the end
-because our type system has typing rules to add this type when
-the type of the expression list does not end with a variadic type.
+because our type system has typing rules to add this type,
+when the type of the expression list does not end with a variadic type.
 Our type system has this behavior to use $\Nil$ in the place of missing
@@
 \dfrac{\env \vdash e_{k}:t_{k}, \env_{k} \;\;\;
-       merge(\env_{1}, ..., \env_{n}) = \env_{m} \;\;\;
+       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
        n = |\;\vec{e}\;|}
@@
 The next example does not type check because it tries to add
-the same field to $a$, but with different types:
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$a[``x"] \; {<}\String{>}, a[``x"] \; {<}\Integer{>} = ``foo", 1$}\\
-\end{tabular}
-\end{center}
+the same field to $\string_ENV$, but with different types:
+\[
+\string_ENV[``x"] \; {<}\String{>}, \string_ENV[``x"] \; {<}\Integer{>} = ``foo", 1
+\]
 
@@
 
-All the examples that we presented until now use the simplest form of
-the table constructor: the empty table $\{\}$.
+The simplest form of the table constructor is the empty table $\{\}$.
 Its inference rule is straightforward:
@@
        n = |\;\vec{p}\;| \;\;\;
-       wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique}) = t}
+       t = wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique})}
       {\env_{1} \vdash \{\;\vec{p}\;\}:t, \env_{n+1}}
@@
        t_{1} \subtype \Boolean}
-      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}t_{2}, \env_{3}}
+      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
 \end{array}
@@
        t_{1} \subtype \Number}
-      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}t_{2}, \env_{3}}
+      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
 \end{array}
@@
        t_{1} \subtype \String}
-      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}t_{2}, \env_{3}}
+      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
 \end{array}
@@
        \Nil \not\subtype t_{1}}
-      {\env_{1} \vdash [e_{1}] = e_{2}: \Any{:}t_{2}, \env_{3}}
+      {\env_{1} \vdash [e_{1}] = e_{2}: \Any{:}close(t_{2}), \env_{3}}
 \end{array}
@@
 type of a table.
-To do that, our type system always promotes an \emph{open}
-table to \emph{closed} and an \emph{unique} table to \emph{open}
-when they appear as expressions, or when they enter in a scope that is
-not its declaration scope.
-The order matters, as going from \emph{unique} to \emph{open}
-before going from \emph{open} to \emph{closed} closes \emph{unique} table types.
-
-The rule \textsc{T-IDREAD} uses the predicate \emph{close} to do
-this promotion before we use a local variable:
-\[
-\begin{array}{c}
-\mylabel{T-IDREAD}\\
-\dfrac{\env_{1}(n) = t}
-      {\env_{1} \vdash n_{e}:close(t), \env_{1}}
-\end{array}
-\]
-
-This rule does not change the type of the local variable because it
-just need to ensure that an alias will not get the same type.
-For instance, it makes this unsafe example not type check:
+%To do that, our type system always promotes an \emph{open}
+%table to \emph{closed} and an \emph{unique} table to \emph{open}
+%when they appear as expressions, or when they enter in a scope that is
+%not its declaration scope.
+%The order matters, as going from \emph{unique} to \emph{open}
+%before going from \emph{open} to \emph{closed} closes \emph{unique} table types.
+In the last section we mentioned that aliasing \emph{unique} table
+types is unsafe.
+To avoid unsafe aliasing of \emph{unique} table types, the rule \textsc{T-IDREAD1}
+always promotes them to \emph{closed}, and updates the variable in the
+environment to \emph{open}:
+\[
+\begin{array}{c}
+\mylabel{T-IDREAD1}\\
+\dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique}}
+      {\env_{1} \vdash n_{e}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed}, \env_{1}[n \mapsto \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{open}]}
+\end{array}
+\]
+
+This is the rule that makes the unsafe example from last section to
+not type check:
 \begin{center}
 \begin{tabular}{lll}
-\multicolumn{3}{l}{$\mathbf{local} \; a:\{``x":\String \cup \Nil\}_{closed} = \{ [``x"] = ``foo" \} \; \mathbf{in}$}\\
-& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String \}_{unique} = \{ [``x"] = ``foo" \} \; \mathbf{in}$}\\
-& & \multicolumn{1}{l}{$a = b \;;\; a[``x"] = \mathbf{nil} \;;\; b[``x"] = b[``x"] \;{..}\; ``bar"$}
-\end{tabular}
-\end{center}
-
-The assignment $a = b$ is not valid because the aliasing changed the
-type of $b$ from $\{``x":\String\}_{unique}$ to $\{``x":\String\}_{open}$,
-which is not subtype of $\{``x":\String \cup \Nil\}_{closed}$.
+\multicolumn{3}{l}{$\mathbf{local} \; a:\{``foo":\Integer\}_{unique} = \{ [``foo"] = 5 \} \; \mathbf{in}$}\\
+& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``foo":\Integer \cup \Nil \}_{closed} = a \; \mathbf{in}$}\\
+& & \multicolumn{1}{l}{$b[``foo"] = \mathbf{nil}$}
+\end{tabular}
+\end{center}
+
+This examples does not type check because aliasing $a$ produces the
+type $\{``foo":\Integer\}_{closed}$ that is not subtype of
+$\{``foo":\Integer \cup \Nil\}_{closed}$, the type of $b$.
 
@@
 
-Even though the rule \textsc{T-CAST} handles the case that we assign
-an alias to a variable, we still have to close this variable due to
-safety.
-The rule \textsc{T-LOCAL} is one example where we handle this case:
-\[
-\begin{array}{c}
-\mylabel{T-LOCAL}\\
-\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
-       r_{1} \lesssim \vec{t} \;\;\;
-       closeset(\env_{2}[\vec{n} \mapsto \vec{t}], fav(\mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s)) \vdash s:\env_{3}}
-      {\env_{1} \vdash \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s:\env_{3} - \{\vec{n} \mapsto \vec{t}\}}
-\end{array}
-\]
-
-This rule prevents the following unsafe example to type check:
+We should also keep tracking of \emph{open} locals, as they can
+also be unsafe like in the next example:
 \begin{center}
@@
        \not \exists i \in 1..n \; l \lesssim k_{i}\\
-       closeall(\env_{1}[self \mapsto \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{closed}, \vec{n} \mapsto \vec{t}, \ret \mapsto r]) \vdash s:\env_{2}
+       closeall(\env_{1}[self \mapsto \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{closed}, \vec{n} \mapsto \vec{t}, \ret \mapsto r]) \vdash s:\env_{2}\\
+       t_{m} = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}, l{:}\Const \; \Self \times \vec{t} \rightarrow r\}_{unique|open}
        \end{array}}
@@
        \env_{1} \vdash \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; s:\\
-       closeset(\env_{1}[n_{1} \mapsto \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}, l{:}\Const \; \Self \times \vec{t} \rightarrow r\}_{unique|closed}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))
+       closeset(\env_{1}[n_{1} \mapsto t_{m}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))
        \end{array}}
@@
        closeall(\env_{1}[n \mapsto \String]) \vdash s_{1}:\env_{2} \\
-       closeall(\env_{1}[n \mapsto filter(t, \String)) \vdash s_{2}:\env_{3}
+       closeall(\env_{1}[n \mapsto filter(t, \String)) \vdash s_{2}:\env_{3}\\
+       \env_{4} = closeset(\env_{1}[n \mapsto t], fav(s_{1}) \cup fav(s_{2}))
       \end{array}}
-      {\env_{1} \vdash \mathbf{if} \; type(n) == ``string" \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}:closeset(\env_{1}[n \mapsto t], fav(s_{1}) \cup fav(s_{2}))}
+      {\env_{1} \vdash \mathbf{if} \; type(n) == ``string" \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}:\env_{4}}
 
@@
 \textit{/* How should we write the typing rule of the local declaration? */}
-
-\textit{/* How this is valid if projection types do not have subtyping rules? */}
 

### 2015-01-04 doc/thesis/haskell/tl.hs
Loading Lua Standard Library according to Lua version

@@ -12,6 +12,9 @@ data TType = TValue
   | TString
   | TUnion TType TType
   | TFunction SType SType
+  | UTable [(TType,Bool,TType)]
+  | OTable [(TType,Bool,TType)]
+  | CTable [(TType,Bool,TType)]
   deriving (Eq, Show)
 
 data SType = TVoid
@@ -83,6 +86,39 @@ first (TVararg t) = TUnion t TNil
 first (TTuple t s) = t
 first TVoid = error("first(TVoid)")
 
+subtype_closed_fields k b v [] = False
+subtype_closed_fields k2 False v2 ((k1,False,v1):t) = (subtype1 k1 k2 && subtype1 k2 k1 && subtype1 v1 v2 && subtype1 v2 v1) || subtype_closed_fields k2 False v2 t
+subtype_closed_fields k2 True v2 ((k1,True,v1):t) = (subtype1 k1 k2 && subtype1 k2 k1 && subtype1 v1 v2) || subtype_closed_fields k2 True v2 t
+subtype_closed_fields k2 True v2 ((k1,False,v1):t) = (subtype1 k1 k2 && subtype1 k2 k1 && subtype1 v1 v2) || subtype_closed_fields k2 True v2 t
+subtype_closed_fields k2 False v2 ((k1,True,v1):t) = False || subtype_closed_fields k2 False v2 t
+
+subtype_closed_tables [] [] = True
+subtype_closed_tables t [] = True
+subtype_closed_tables [] t = False
+subtype_closed_tables t1 ((k,b,v):t2) = subtype_closed_fields k b v t1 && subtype_closed_tables t1 t2
+
+subtype_unique_fields _ _ v [] = subtype1 TNil v
+subtype_unique_fields k1 b1 v1 ((k2,b2,v2):t) = (subtype1 k1 k2 && subtype1 v1 v2) || subtype_unique_fields k1 b1 v1 t
+
+subtype_unique [] = True
+subtype_unique ((k,b,v):t) = subtype1 TNil v && subtype_unique t
+
+subtype_unique_tables [] [] = True
+subtype_unique_tables t [] = True
+subtype_unique_tables [] t = subtype_unique t
+subtype_unique_tables ((k,b,v):t1) t2 = subtype_unique_fields k b v t2 && subtype_unique_tables t1 t2
+
+subtype_open_fields _ _ v [] = subtype1 TNil v
[…]

### 2015-01-05 doc/thesis/evaluation.tex
Minor improvements to the text about typing Lua Standard Library and MD5

@@
 Typical \citep{luatypical}, and Mod 11 \citep{luamod11}.
-The first provides client side HTTP digest authentication for Lua.
-The second is an extension to the primitive function \texttt{type}.
-The third is a generator and checker of modulo 11 numbers.
+The first one provides client side HTTP digest authentication for Lua.
+The second one is an extension to the primitive function \texttt{type}.
+The third one is a generator and checker of modulo 11 numbers.
 We randomly selected these libraries to evaluate Typed Lua for
@@
 \hline
-Lua Standard Library & 80 & 6 & 9 & 34 & 129 & 62\% \\
+Lua Standard Library & 80 & 6 & 8 & 35 & 129 & 62\% \\
 \hline
@@
 For each case study, we split the evaluation results according
-to the modules that each one of them includes.
+to the modules that each one of them include.
 This separation by case study allows us to better discuss the
@@
 of a local variable that is not visible in the current scope.
-Table \ref{tab:evallsl} summarizes the evaluation results for the Lua Standard Library.
+Table \ref{tab:evallsl} summarizes the evaluation results for the Lua Standard Library
+(version 5.3).
 
@@
 \hline
-base & 9 & 1 & 3 & 13 & 26 & 35\% \\
+base & 9 & 1 & 2 & 14 & 26 & 35\% \\
 \hline
@@
 
-We could give precise static types to most of the modules,
-but some of them are still hard to type with the current type
-system.
-Now, we will discuss why some modules are difficult to type,
+We could give precise static types to several members of
+the Lua Standard Library, though some of them are still
+hard to type with the current type system.
+Now, we discuss why some modules are difficult to type,
 and what kind of limitations they represent on our type system.
@@
 For this reason, it is also difficult to type the function
-\texttt{pairs}.
+\texttt{pairs}, as it uses the function \texttt{next} to
+traverse a table.
 The functions \texttt{getmetatable}, \texttt{rawget}, and
-\texttt{rawset} are also examples of functions that their
+\texttt{rawset} are other examples of functions that their
 behavior depend on the C implementation.
 
-There are also some functions that have a dynamic behavior and
-their return type depend on their input type.
-For instance, the functions \texttt{assert} and \texttt{select}
-can return all the values that they received as their arguments.
+We could not give a precise static type to \texttt{ipairs}
+because our type system does not have parametric polymorphism.
+This function returns an iterator function that traverses a
+list of elements.
 
-We could give a more precise type to \texttt{ipairs} if our
-type system had parametric polymorphism, as this function
-returns an iterator function that operates over a list of
-elements.
-
-Furthermore, the \texttt{base} module also includes two very
-special cases: the functions \texttt{tonumber} and \texttt{collectgarbage}.
+The \texttt{base} module also includes two special cases:
+the functions \texttt{tonumber} and \texttt{collectgarbage}.
 
 The function \texttt{tonumber} is a special case because the
-type of one input parameter depends on the type of another.
-For instance, there are two static types that we can
-assign to the function \texttt{tonumber}:
-\texttt{(value) -> (number?)} and
-\texttt{(string, number) -> (number?)}.
+type of the first parameter depends on the type of the
+second parameter.
+For instance, we can call \texttt{tonumber(1)}, but we cannot
+call \texttt{tonumber(1,2)}.
 More precisely, the first argument of \texttt{tonumber} can be
 a value of any type if it is the only argument, but it must
-be a value of type string if there is a second argument,
-which must be a value of type number.
+be a string if there is a second argument,
+which must be an integer.
 
@@
 which should be a string.
-This function can return one number, one boolean, or two numbers
-depending on the literal string that was given as its first argument.
+For instance, calling \texttt{collectgarbage("collect")} returns an integer number,
+calling \texttt{collectgarbage("count")} returns a floating point number,
+and calling \texttt{collectgarbage("isrunning")} returns a boolean.
 
-We could not type the \texttt{coroutine} module because our
-type system cannot describe the computational effects of a program.
+We could not type most of the members of the \texttt{coroutine} module
+because our type system cannot describe the computational effects of a program.
 Lua has one-shot delimited continuations \citep{james2011yield}
@@
 to represent the type \emph{thread}.
+Still, we could type the function \texttt{coroutine.isyieldable},
+as it has no input parameters, and simply returns a boolean that
+indicates whether the running coroutine is yieldable.
 
@@
 
-We could type all the members of the \texttt{string} module.
 Lua programmers often use the functions of the \texttt{string} module
@@
 methods in our evaluation results.
+
+The \texttt{string} module is not that hard to type, but it includes
+some functions that require dependent types.
+For instance, the type of the arguments that we pass to
+\texttt{string.format} depend on the format string we are using.
+It is fine to call \texttt{string.format("\%d", 1)}, but
+\texttt{string.format("\%d", true)} generates a run-time error.
+
+The \texttt{utf8} module was quite straightforward to type,
+as its members are only operations over strings.
 
@@
 giving a precise type to \texttt{table.concat}, as it operates over lists
-where all elements are strings or numbers, and we can use the table type
-\texttt{\{string|number\}} to express this type on our type system.
+where all elements are strings or numbers.
 
@@
 the calling arity.
-We can call \texttt{table.insert} passing two or three parameters.
-The first parameter is always a list.
-If we call it with two arguments, then the second parameter
-is the value to be inserted at the end of the list.
-If we call it with three arguments, then the third parameter
-is the value to be inserted in the list, and the second
-parameter is the position where it should be inserted.
+More precisely, calling \texttt{table.insert(l, v)} inserts the
+element \texttt{v} at the end of the list \texttt{l},
+while \texttt{table.insert(l, p, v)} inserts the element \texttt{v}
+at the position \texttt{p} of the list \texttt{l},
+but generates a run-time error when \texttt{p} is out of bounds.
 This function also does not follow the semantics of Lua on
@@
 
-Even though the \texttt{math} module is straightforward to type,
-it includes a special case that is hard to type: the function \texttt{math.random}.
-This function is difficult to type because its type also depends
-on the calling arity.
-We can call \texttt{math.random} passing zero, one, or two parameters.
-If we pass no parameter, then it returns a random floating point
-number between \texttt{0} and \texttt{1}.
-If we pass one or two parameters, then all the parameters should be
-integers and it returns an integer number.
-Another problem with this function is that it also does not follow
-the semantics of Lua on discarding extra arguments, and generates
-a run-time error whenever we pass more than three arguments.
+Even though the \texttt{math} module looks straightforward to type,
+it includes several functions that the return type depend on the input type.
+For instance, \texttt{math.abs(1)} returns the integer \texttt{1},
+while \texttt{math.abs(1.1)} returns the floating point \texttt{1.1}.
+
+The \texttt{math} module also includes a special case that is hard to type:
+the function \texttt{math.random}.
+This function is difficult to type because its type depends on the calling arity.
+Calling \texttt{math.random()} returns a random floating point number
+between \texttt{0} and \texttt{1}.
+Calling \texttt{math.random(10)} is equivalent to \texttt{math.random(1,10)},
+and returns an integer between this interval.
+This function also does not follow the semantics of Lua on discarding extra arguments,
+and generates a run-time error whenever we pass more than two arguments.
 There is also a problem on its documentation, as it suggests that
@@
 error instead.
-
-The \texttt{bit32} module was quite straightforward to type,
-as its members are only bitwise operations over numbers.
 
@@
 because its return type depends on whether the file handle that is
-being closed was created with \texttt{popen} or not.
-If it was created with \texttt{popen}, then \texttt{io.close}
-has type \texttt{(file?) -> (boolean?, string, number)};
-otwerwise \texttt{io.close} has type \texttt{(file?) -> (boolean)?}.
+being closed was created with \texttt{popen}.
+Calling \texttt{io.close(f)} returns the tuple \texttt{(boolean?, string, number)}
+when \texttt{f} was created with \texttt{popen},
+and it returns the tuple \texttt{(boolean)?} otherwise.
 The method \texttt{file:close} is difficult to type for the same reason.
@@
 to type because their return types depend on their input types.
-The type of \texttt{io.read} can be either \texttt{() -> (string*)}
-or \texttt{(string|number*) -> (string|number*)}, and
-the type of \texttt{io.lines} can be either \texttt{(string?) -> (string*)}
-or \texttt{(string, string|number*) -> (string|number*)}.
-Returning \texttt{(string|number*)} is also painful to programmers, as
-they need to constrain the value of the return type among three
-different types before cocatenating it, for example.
-The methods \texttt{file:read} and \texttt{file:lines} are difficult
-to type for the same reason.
+For instance, \texttt{io.read("l")} returns a string or \texttt{nil},
+\texttt{io.read("n")} returns a number or \texttt{nil}, and calling
+\texttt{io.read("l", "n")} returns a string or \texttt{nil} and a number or \texttt{nil}.
+The function \texttt{io.lines}, and the methods \texttt{file:read} and
+\texttt{file:lines} are difficult to type for the same reason.
 
@@
 the input type.
-Its type can be either \texttt{() -> (boolean)} or
-\texttt{(string) -> (boolean?, string, number)}.
+More precisely, \texttt{os.execute()} returns a boolean, while
+\texttt{os.execute(c)} returns \texttt{(boolean?, string, number)},
+where \texttt{c} is a string that represents the command to be executed.
 
 The evaluation results show that our type system should include
-parametric polymorphism, intersection types, and effect types,
+intersection types, parametric polymorphism, and effect types,
 as these features would help us increase the static typing of the
 Lua Standard Library.
+Intersection types would allow us to define overloaded function types.
 Parametric polymorphism would allow us to define generic function and table types.
-Intersection types would allow us to define overloaded function types.
 Effect types would allow us to type coroutines.
@@
 we found a little difference between its documentation and its behavior.
-The documentation suggested that the typing of the function \texttt{md5.update}
-should be \texttt{(md5\string_context, string) -> (md5\string_context)},
-while a comment in the source code suggested that it should be
-\texttt{(md5\string_context, string, string*) -> (md5\string_context)}.
-However, while testing it and reading its source code, we noticed that
-its actual type is \texttt{(md5\string_context, string*) -> (md5\string_context)},
+The documentation suggests that the type of the function \texttt{md5.update}
+is \texttt{(md5\string_context, string) -> (md5\string_context)},
+though there is a call to this function in the test script that passes
+an extra string argument.
+Reading the source code, we found that its actual type is
+\texttt{(md5\string_context, string*) -> (md5\string_context)},
 that is, we can pass zero or more strings to \texttt{md5.update}.
 
-This case study shows that it might be not so obvious to identify which
-functions discard extra arguments when we are typing an external library.
-In this case study, reading its source code and test script was essential
-to confirm the typing of \texttt{md5.update}.
+This case study shows that type annotations help programmers maintain the
+documentation updated, as the type checker always validate them.
 

### 2015-01-06 doc/thesis/evaluation.tex
Minor improvements to the text about typing LuaSocket and random case studies

@@
 type of the values that we use to call them.
+For instance, \texttt{socket.skip(0, nil, "hello")} returns the
+tuple \texttt{(nil, "hello")}, \texttt{socket.skip(1, nil, "hello")}
+returns only the string \texttt{"hello"}, and
+\texttt{socket.skip(-1, nil, "hello")} returns the tuple
+\texttt{(-1, nil, "hello")}.
+Passing a negative number to \texttt{socket.skip} can be
+dangerous, as it returns anything that might be in the stack.
+For instance, \texttt{f = socket.skip(-2)} assigns \texttt{socket.skip}
+to \texttt{f}.
+Our type system cannot handle the type of negative numbers, as
+this requires the refinement types from hybrid type checking \citep{flanagan2006htc}.
 
@@
 The function \texttt{ftp.get} downloads data from a given URL,
-and it is difficult to type because its return type depend on
-the type of its parameters.
-Its type is either \texttt{(string) -> (string)?} or
-\texttt{(url\string_argument|url\string_path) -> (number)?}.
-The types \texttt{url\string_argument} and \texttt{url\string_path}
-are table types that describe the URL that should be used to
-download data.
-Both table types include the same fields, except one:
-while the first includes the mandatory field \texttt{argument},
-of type \texttt{string}, the latter includes the mandatory field
-\texttt{path}, also of type \texttt{string}.
+and it is difficult to type because its return type depends on
+the type of its input parameter.
+More precisely, \texttt{ftp.get(url)} returns the tuple
+\texttt{(string) | (nil, string)} if \texttt{url} is a string,
+and it returns the tuple \texttt{(number) | (nil, string)} if
+\texttt{url} is a table.
 
 The function \texttt{http.request} downloads data from a given URL,
-and it is difficult to type because its return type depend on
-the type of its parameters.
-Its type is either
-\texttt{(string, string?) -> (string, number, \{string : string\}, number)?} or
-\texttt{(url\string_request) -> (number, number, \{string : string\}, number)?}.
-The type \texttt{url\string_request} is a table type containing
-details about the downloading URL.
+and it is difficult to type because its return type depends on
+the types of its input parameters.
+More precisely, \texttt{http.request(url, body)} returns the tuple
+\texttt{(string, number, \{string:string\}, number) | (nil, string)}
+if \texttt{url} is a string and \texttt{body} is another string or \texttt{nil},
+but it returns the tuple
+\texttt{(number, number, \{string:string\}, number) | (nil, string)}
+if \texttt{url} is a table and \texttt{body} is \texttt{nil}.
 
@@
 filters to create standard filters.
+
 Even though we could type all the members of the \texttt{mime} module,
@@
 This function is difficult to type because the \texttt{mime} low-level
-filters do not have a standard API.
+filters are polymorphic.
 
@@
 Like \texttt{http.request}, the return type of \texttt{http-digest.request}
-also depends on the type of its parameters.
-Its type is either
-\texttt{(string) -> (string, number, \{string : string\})?} or
-\texttt{(url\string_request) -> (number, number, \{string : string\})?}.
+also depends on the types of its input parameters.
 
@@
 We could not give a precise type to \texttt{getmetatable}, so we used the dynamic
-type \texttt{any} as its return type, and this generates a warning that we could not
-remove.
+type \texttt{any} as its return type, and this generates a warning.
 Second, the module returns a metatable that extends \texttt{\string_\string_call}

### 2015-01-06 doc/thesis/evaluation.tex
Minor improvements to the text about typing TLC and other Lua projects

@@
 
-The Typed Lua compiler is the last case study that we evaluated, and the
-Table \ref{tab:evaltlc} summarizes the evaluation results.
+The Typed Lua compiler is the last case study that we evaluated.
+Table \ref{tab:evaltlc} summarizes its evaluation results.
 
@@
 
-The \texttt{tlast} module implements the abstract syntax tree that
-the compiler uses to perform the type checking.
-We could not precisely type just one function because its return
-type depend on the type of its input parameter.
+The \texttt{tlast} module implements the Abstract Syntax Tree for
+the compiler.
+We could not precisely type just one function, because it has
+an overloaded type that requires intersection types.
 
-The \texttt{tltype} module implements the types introduced by
-Typed Lua and the subtyping and consistent-subtyping relations;
-it was straightforward to type.
+The \texttt{tltype} module implements the types introduced by Typed Lua.
+It also implements the subtyping and consistent-subtyping relations.
+The interface that this module exports was straightforward to type.
 
-The \texttt{tlst} module implements the environment that the
-compiler uses during the type checking, and it was also straightforward
-to type.
+The \texttt{tlst} module implements the symbols table for the compiler.
+The interface that this module exports was also straightforward to type.
 
 The \texttt{tllexer} module defines common lexical rules for
-the Typed Lua parser and the parser of the description files.
+the Typed Lua parser and the description file parser.
 This module is hard to type because it exports LPeg \citep{lpeg,ierusalimschy2009lpeg}
-patterns, which our type system cannot precisely type because it does
-not support operator overloading.
+patterns.
+For instance,
+\begin{verbatim}
+    tllexer.Number = lpeg.C(Hex + Float + Int) / tonumber
+\end{verbatim}
+exports the pattern that matches a number.
+Note that LPeg overloads the definition of arithmetic operators,
+a feature that our type system does not support yet.
+That is the reason why we could not precisely type the LPeg
+patterns that this module exports.
+However, we could type the two error reporting functions that this
+module exports.
 
-The \texttt{tlparser} and \texttt{tldparser} implement the Typed Lua
-parser and the parser of the description files, respectively.
+The \texttt{tlparser} and \texttt{tldparser} modules implement the Typed Lua
+parser and the description file parser, respectively.
 Even though they use LPeg to implement the grammar rules,
-they only export a parsing function that uses LPeg to parse a
-string and return the corresponding AST.
+they only export a parsing function.
+Both use LPeg to parse a string and return the corresponding AST.
 
-The \texttt{tlchecker} and \texttt{tlcode} modules are also
-straightforward to type.
-The first traverses the AST to perform type checking,
+The interface that the \texttt{tlchecker} and \texttt{tlcode}
+modules export are also straightforward to type.
+The former traverses the AST to perform type checking,
 while the latter traverses the AST to perform code generation.
 
-Even though we could precisely type the interface of most of
-the modules, we had some issues to write mutually recursive
-functions that traverses the AST.
-In this case we had to rely on the dynamic type to defer the
-function declarations.
+Even though we could precisely type the interface that most of
+the modules export, we had issues to write mutually recursive functions.
+This kind of functions often appear on compilers construction
+to traverse the data structures that they use.
+However, Typed Lua still does not support mutually recursive functions.
+A way to overcome this limitation was to predeclare these functions
+with an empty body, and then redeclare them with they actual body.
+The first declaration specifies the function type, while the
+second specifies what the function actually does without changing
+any type definition.
 
-We detected some bugs while doing the bootstrap:
-accessing undeclared global variables.
+Traversing the AST would also be problematic if we had not included
+a way to decompose unions of table types, as we mentioned in Section \ref{sec:alias}.
+
+Bootstraping the compiler also helped revealing some bugs.
+We found some accesses to undeclared global variables and also
+to undeclared table fields.
+The compiler also helped pointing the places where we should narrow
+a nilable value before using it.
+In fact, this point appeared in all the case studies that we rewrote
+the Lua code.
+This means that Lua programmers often use values before checking
+whether it is \texttt{nil}.
 
@@
 
-\subsection{Lua Related Work}
+In this section we compare Typed Lua with other projects and type systems.
+We split this section into two parts: the first compares Typed Lua
+with other Lua projects, while the second compares Typed Lua with
+other type systems that are not related to Lua.
+
+\subsection{Other Lua projects}
 
@@
 system for Lua, which is written in Metalua.
-A remarkable feature of Tidal Lock is the way that it handles tables,
-because it allows the programmer to incrementally build the type of
-a table.
-More precisely, Tidal Lock allows the programmer to create an empty
-table to build its type according to the keys that the programmer
-inserts in it, so the programmer does not need to build the table in
-a single table constructor.
-This feature inspired us to introduce it into Typed Lua,
-but in a simpler way and with different formalization.
-Like Typed Lua, Tidal Lock also has local type inference.
-Unlike Typed Lua, Tidal Lock does not type the common object-oriented
-idioms used by Lua programmers, though it also uses subtyping.
-
-Tidal Lock \citep{tidallock} is a prototype of an optional type system for Lua.
-It also includes the refinement of table types.
-In fact, Tidal Lock inspired Typed Lua to also include this feature.
-However, Tidal Lock covers a little subset of Lua.
+Tidal Lock covers a little subset of Lua.
 Statements include declaration of local variables, multiple assignments,
@@
 
+A remarkable feature of Tidal Lock is the refinement of table types.
+This feature inspired us to also include it in Typed Lua,
+but in a simpler way and with different formalization.
+
 The table type from Tidal Lock can only represent records, that is,
@@
 of a table type.
-The field types describe whether a table field is mutable or imutable
+The field types describe if a table field is mutable or imutable
 in a table type.
@@
 
-Tidal Lock is similar to Typed Lua because it is also a structural type
-system based on subtyping.
-It also uses local type inference to deduce the type of unannotated expressions,
-and have a dynamic type.
-However, Tidal Lock does not support union types, recursive types, and variadic types.
-It also does not support any object-oriented idiom.
+Tidal Lock is also a structural type system based on subtyping that has
+local type inference.
+However, it does not support union types, recursive types, and variadic types.
+It also does not type any object-oriented idiom.
 
-\subsection{Non-Lua Related Work}
+\subsection{Other type systems}
 

### 2015-01-07 doc/thesis/evaluation.tex
Minor improvements to the text about other projects

@@
 
-In this section we compare Typed Lua with other projects and type systems.
+In this section we compare Typed Lua with other projects.
 We split this section into two parts: the first compares Typed Lua
 with other Lua projects, while the second compares Typed Lua with
-other type systems that are not related to Lua.
+other projects that are not related to Lua.
 
@@
 of a table type.
-The field types describe if a table field is mutable or imutable
+The field types describe if a table field is mutable or immutable
 in a table type.
@@
 
-\subsection{Other type systems}
+Luacheck \citep{luacheck} is a tool that performs static analysis on Lua code.
+It can flag access to undeclared globals and unused local variables,
+but it does not perform static type checking.
+
+\subsection{Other projects}
 
@@
 
-The type system of Typed Racket introduced \emph{occurrence typing}
+The main feature of Typed Racket's type system is \emph{occurrence typing}
 \citep{tobin-hochstadt2010ltu}.
@@
 to refine union types.
-The type system is rich and includes basic types, function types,
-union types, recursive types, and structure types.
-Function types can return multiple values.
-Structure types are similar to our interfaces, as they describe
-record types.
+Occurrence typing is not sound in the presence of mutation.
+As these kinds of checks are common in other languages, related systems
+have appeared \citep{guha2011tlc,winther2011gtp,pearce2013ccf}.
+
+The type system of Typed Racket also includes function types, recursive
+types, and structure types.
+Its function types also handle multiple return values, and there is
+also a way to describe function types that have optional arguments.
+Its structure types are similar to our interfaces, as they describe record types.
 The type system is also structural and based on subtyping.
-It also includes the dynamic type \texttt{Any}, which is the top
-type in the system.
+It also includes the dynamic type \texttt{Any}, which is the top type in the system.
 Typed Racket also supports polymorphic functions and data structures.
@@
 
-The type system of Typed Clojure is rich.
-It includes polymorphic function types, union types, intersection types,
-lists, vectors, maps, sets, and recursive types.
+The type system of Typed Clojure includes polymorphic function types,
+union types, intersection types, lists, vectors, maps, sets, and recursive types.
 Function types can also have rest parameters, which are similar
@@
 In fact, its function types cannot return multiple results.
-It also uses ocurrence typing to allow control flow statements to
+It also uses occurrence typing to allow control flow statements to
 refine union types.
@@
 It includes optional type annotations and compiles to JavaScript.
-The type system of Dart is nomial and includes base types,
+The type system of Dart is nominal and includes base types,
 function types, lists, and maps.
@@
 of unannotated expressions.
-The type system of TypeScript is structural and based
-on subtyping.
+The type system of TypeScript is structural, based
+on subtyping, and supports generics.
 It includes the dynamic type, primitive types, function
-types, array types, tuple types, and recursive types.
-TypeScript also support generics.
+types, array types, tuple types, recursive types, and
+object types.
+Unlike Typed Lua, TypeScript does not include union types,
+and it uses arrays to represent variadic functions and
+multiple return values.
 
@@
 its current unsound type system.
+
+TeJaS \citep{lerner2013tejas} is a framework for the construction of
+different type systems for JavaScript.
+The authors created a base type system for JavaScript with
+extensible typing rules that allow the experimentation of
+different static analysis.
+They used TeJaS to create a type system that simulates the
+type system of TypeScript.
 
@@
 They report that usually cast insertions impact on execution
-performance.
+performance, so Gradualtalk also has an option that allows
+programmers to turn them off, downgrading Gradualtalk
+to an optional type system.
 
@@
 It includes class and object types to differentiate the
-type of class declarations and instances.
+type of class declarations and instances, respectively.
 It also uses local type inference.
@@
 
-Hack \citep{hack} is a new programming language that combines
-static and dynamic type in PHP code.
-Its type system includes generics, nullable types, collections,
+Hack \citep{hack} is a new programming language that runs on the
+Hip Hop Virtual Machine (HHVM).
+The HHVM is a virtual machine that executes Hack and PHP programs.
+We can view Hack as an extension to PHP that combines static and
+dynamic typing.
+The type system of Hack includes generics, nullable types, collections,
 and function types.
@@
 and type casts.
+
+Grace \citep{black2013sg} is an object-oriented language
+with optional typing.
+Grace is not a dynamically typed language that has been
+extended with an optional type system, but a language
+that has been designed from scratch to have both
+static and dynamic typing.
+\citet{homer2013modules} explores some useful patterns
+that derive from Grace's use of objects as modules
+and its brand of optional structural typing, which
+can also be expressed with Typed Lua's modules as tables.

### 2015-01-08 doc/thesis/abstract.tex
Rewriting the conclusions and adding an abstract

Dynamically typed languages such as Lua avoid static types in favor of
simplicity and flexibility, because the absence of static types means
that programmers do not need to bother with abstracting types that
should be validated by a type checker.
In contrast, statically typed languages provide the early detection of
many bugs, and a better framework for structuring large programs.
These are two advantages of static typing that may lead programmers
to migrate from a dynamically typed to a statically typed language,
when their simple scripts evolve into complex programs.
Optional type systems allow combining dynamic and static typing in
the same language, without influencing its original semantics,
making easier this code evolution from dynamic to static typing.
Designing an optional type system for a dynamically typed language
is challenging, as it should feels natural to programmers that are
already familiar with this language.
In this work we present and formalize the design of Typed Lua,
an optional type system for Lua that introduces novel features
to statically type check some Lua idioms and features.
Even though Lua shares several characteristics with other dynamically
typed languages such as JavaScript, Lua also has several unusual features
that are not present in the type system of these languages.
We discuss how Typed Lua handles features such as functions with flexible
arity that interact with multiple assignment, functions that are overloaded
on the number of return values, and the incremental evolution of record and object types.
We also discuss our design decisions, and present the evaluation results
that we achieved while using Typed Lua to type existing Lua code.

### 2015-01-08 doc/thesis/conclusions.tex
Rewriting the conclusions and adding an abstract

@@
-Our main contribution is the design of a complete optional type system
-for Lua, a procedural scripting language.
-TypeScript is a complete optional type system for JavaScript,
-but it does not include union types, and it uses arrays to represent
-variadic functions and multiple return values.
-Gradualtalk is a complete gradual type system for Smalltalk,
-but Smalltalk is an object-oriented language that does
-not handle multiple return values, multiple assignments, and
-variadic functions.
-Gradualtalk also mixes nominal and structural typing because of
-the object-oriented nature of Smalltalk.
-Typed Racket is a complete static type system for Racket, a functional
-language.
-Typed Racket makes it easier to migrate from untyped to typed code,
-but the migration happens module-by-module and type annotations are
-mandatory on typed modules.
-There are research on the design of optional and gradual type systems
-for JavaScript, Perl, and Python, but they are either experimental or
-limited in their coverage.
-Furthermore, Typed Lua includes table types that use types as keys,
-instead of identifiers, and the fact that Typed Lua also includes
-literal types makes its table types subsume the usual idea of
-record types.
+We presented Typed Lua, an optional type system for Lua.
+Typed Lua is a Lua extension that allows programmers to combine static
+and dynamic typing in Lua code, making easier the evolution of simple scripts
+into large programs.
 
-Although each language is different, we believe that several parts of
-the design of Typed Lua should be relevant to other scripting languages.
-Most notably, JavaScript shares with Lua the use of tests of type tags
-as a way of encoding optional and overloaded parameters, and the
-proliferation of class-based object systems built on top of the
-language's delegation mechanisms.
+Our main contribution is the formalization of a complete optional
+type system for Lua.
+Even though Lua shares several characteristics with other dynamically
+typed languages such as JavaScript, Lua also has several unusual features
+that are not present in the type system of these languages.
+Lua has functions with flexible arity that interact with multiple
+assignment, functions that are overloaded on the number of return values,
+and incremental evolution of record and object types.
 
-Typed Lua should also be a major contribution to the Lua community,
+We introduce the refinement of table types to support the incremental
+evolution of record and object types.
+The refinement of table types plays an important role in our type
+system, as they static type check the idiomatic way that Lua programmers
+use tables to define modules and objects.
+
+We introduce projection types to handle functions that are overloaded
+on the number of return values.
+Projection types are a novel feature of our type system, as they
+allow programmers to narrow the type of a local variable that depends
+on the type of another one.
+
+We use union types and variadic types to allow our type system to
+handle functions with flexible arity.
+While union types can describe optional parameters, variadic types
+describe the type of the vararg expression and the type of functions
+that can receive or return any number of values.
+
+Unlike some of the other optional type systems, we designed Typed Lua
+without deliberated unsound parts.
+However, we still do not have proofs that the novel parts of our type
+system are sound.
+
+To improve our confidence on the design of Typed Lua, we performed several
+case studies to evaluate it.
+Before starting the design of Typed Lua, we realized that we should
+not rely only on the semantics of Lua, as this could lead to a
+cumbersome type system that would not support several idiomatic Lua code.
+For this reason, we performed a mostly automated survey of Lua idioms
+and features to support our design choices.
+
+Even though the evaluation results exposed several limitations of our
+type system, they also gave us confidence that our design still
+introduces a reasonable amount of static type safety to several Lua idioms
+and features.
+We found that the three main limitations of our type system are
+the lack of intersection types, parametric polymorphism, and operator overloading.
+We see a soundness proof along with these limitations as our major targets
+for future work.
+While a soundness proof can consolidate our confidence on our design choices,
+overcoming these limitations can increase the amount of static type safety
+that our type system provides.
+
+We believe that Typed Lua is a major contribution to the Lua community,
 because it offers a framework that programmers can use to document,
@@
 libraries.
+In fact, we already have user feedback from Lua programmers that are
+using Typed Lua in their projects.
+For instance, ZeroBrane Studio is an IDE for Lua development that is
+starting to use Typed Lua to perform static analysis on Lua code.
 
-Finally, Typed Lua can be a base for further research and development
-of static type systems for the Lua language, such as the use of static
-analysis of Lua API calls for checking that a native library is
+Finally, we also believe that Typed Lua can be a base for further research
+and development of static type systems for the Lua language, such as
+the use of static analysis of Lua API calls for checking that a native library is
 exporting the interface to Lua code that it declares, the use of

### 2015-01-10 doc/thesis/system.tex
Minor improvements to the text about typing Lua Standard Library and MD5
Minor improvements to the section about types

@@
 They are novel type system features that let Typed Lua cover several Lua idioms
-and features, such as the refinement of tables, multiple return values,
+and features, such as refinement of tables, multiple return values,
 and optional parameters.
@@
 \begin{array}{rlr}
-\multicolumn{3}{c}{\textbf{First-level types}}\\
-t ::= & \;\; l & \textit{literal types}\\
+t ::= & & \textsc{first-level types:}\\
+& \;\; l & \textit{literal types}\\
 & | \; b & \textit{base types}\\
@@
 & | \; t \cup t & \textit{disjoint union types}\\
-& | \; s \rightarrow r & \textit{function types}\\
+& | \; s \rightarrow s & \textit{function types}\\
 & | \; \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed|open|unique} & \textit{table types}\\
@@
 & | \; \pi_{i}^{x} & \textit{projection types}\\
-l ::= & \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} & \\
-b ::= & \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String & \\
-k ::= & \;\; l \; | \; b \; | \; \Any & \textit{key types}\\
-v ::= & \;\; t \; | \; \Const \; t & \textit{value types}\\ 
-\multicolumn{3}{c}{}\\
-\multicolumn{3}{c}{\textbf{Second-level types}}\\
-s ::= & \;\; \Void & \textit{void type}\\
+%\multicolumn{3}{c}{}\\
+l ::= & & \textsc{{\small literal types:}}\\
+& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
+%\multicolumn{3}{c}{}\\
+b ::= & & \textsc{{\small base types:}}\\
+& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
+%\multicolumn{3}{c}{}\\
+k ::= & & \textsc{{\small key types:}}\\
+& \;\; l \; | \; b \; | \; \Value &\\
+%\multicolumn{3}{c}{}\\
+v ::= & & \textsc{{\small value types:}}\\
+& \;\; t \; | \; \Const \; t &\\ 
+%\multicolumn{3}{c}{}\\
+s ::= & & \textsc{second-level types:}\\
+& \;\; p & \textit{tuple types}\\
+& | \; s \sqcup s & \textit{unions of tuple types}\\
+%\multicolumn{3}{c}{}\\
+p ::= & & \textsc{{\small tuple types:}}\\
+& \;\; \Void & \textit{void type}\\
 & | \; t* & \textit{variadic types}\\
-& | \; t \times s & \textit{tuple types}\\
-r ::= & \;\; s \; | \; s \sqcup r & \textit{return types}\\
+& | \; t \times p & \textit{pair types}
 \end{array}
@@
 
-Figure \ref{fig:typelang} presents the abstract syntax of the
+Figure \ref{fig:typelang} presents the abstract syntax of
 Typed Lua types.
@@
 tuple types, and unions of tuple types.
-
-Literal types $l$ represent the type of literal values.
+Types are ordered by a subtype relationship that we introduce
+in Section \ref{sec:subtyping}, so Lua values may belong to
+several distinct types.
+
+Literal types represent the type of literal values.
 They can be the boolean values $\False$ and $\True$,
 an integer value, a floating point value, or a string value.
-We shall see that literal types are important in our treatment of
+We will see that literal types are important in our treatment of
 table types as records.
 
-Typed Lua includes four base types $b$: $\Boolean$, $\Integer$, $\Number$, and $\String$.
+Typed Lua includes four base types: $\Boolean$, $\Integer$, $\Number$, and $\String$.
 The base types $\Boolean$ and $\String$ represent the values that during
@@
 \texttt{integer} for integer numbers and \texttt{float} for real numbers.
-Both \texttt{integer} and \texttt{float} values are a subtype of \texttt{number}.
-We introduced the base type $\Number$ to represent either the tag
-\texttt{number} or \texttt{float} values, and the type $\Integer$ to represent
-\texttt{integer} values.
+Lua does automatic promotion of \texttt{integer} values to \texttt{float}
+values as needed.
+We introduced the base type $\Number$ to represent \texttt{float} values,
+and the base type $\Integer$ to represent \texttt{integer} values.
 In the next section we will show that $\Integer$ is subtype of $\Number$.
-This allows programmers to use \texttt{integer} values in the
-place of \texttt{float} values.
-
-The type $\Nil$ represents the type of the value that during run-time
-Lua tags as \texttt{nil}.
-
-The type $\Value$ is the top type.
+This allows programmers to keep using \texttt{integer} values where
+\texttt{float} values are expected.
+
+The type $\Nil$ is the type of \texttt{nil}, the value that Lua uses for
+undefined variables, missing parameters, and missing table keys.
+
+The type $\Value$ is the top type, which represents any Lua value.
 In Section \ref{sec:rules} we will show that this type,
@@
 
-Typed Lua includes the dynamic type $\Any$ for allowing programmers
-to mix static and dynamic typing.
-
 Typed Lua uses the type $\Self$ to represent the \emph{receiver}
 in object-oriented method definitions and method calls.
-As we mentioned in Chapter \ref{chap:typedlua}, we need the type
+As we mentioned in Section \ref{sec:oop}, we need the type
 $\Self$ to prevent programs from indexing a method without
@@
 
-Function types have the form $s \rightarrow r$ to represent Lua functions,
-where $s$ is the type of the parameter list and $r$ is the return type.
-Both $s$ and $r$ are second-level types.
-In the next paragraphs we explain why we need different representations
-for the type of the parameter list and the return type.
-
-Second-level types $s$ are tuples of first-level types that can end
-with either an empty tuple or with a variadic type.
+Function types have the form $s \rightarrow s$ and represent Lua functions,
+where $s$ is a second-level type.
+
+Second-level types are tuples of first-level types that can end
+with either an empty tuple or with a variadic type, and
+unions of tuple types.
 Typed Lua needs second-level types because tuples are not first-class
-values in Lua, though they may appear on argument passing,
-multiple returns, and multiple assignments.
+values in Lua, only appearing on argument passing, multiple returns,
+and multiple assignments.
 The type $\Void$ is the type of an empty tuple.
-A variadic type $t*$ is a generator for a sequence of values
-of type $t \cup \Nil$, and it represents the type of a vararg expression.
-
-Second-level types $r$ include the same second-level types introduced
-by $s$ plus unions of tuple types.
-Typed Lua includes unions of tuples because Lua programs
+A variadic type $t*$ represents a sequence of values of type $t \cup \Nil$;
+it is the type of a vararg expression.
+Second-level types include unions of tuples because Lua programs
 usually overload the return type of functions to denote error,
 as we mentioned in Section \ref{sec:statistics}.
-We could not find a nice concrete syntax to represent unions of
-tuples that appear in the type of the parameter list, and
-that is the reason why we restricted them to appear in the return type only.
-The same reason led us to split second-level types into two
-different symbols.
-We use the symbol $\sqcup$ to represent the union between two
-different tuple types.
-We have different unions because $\cup$ represents the union
-between two first-level types, while $\sqcup$ represents the
-union between two tuple types.
+For clarity, we use the symbol $\sqcup$ to represent the union between
+two different tuple types.
+Note that $\cup$ represents the union between two first-level types,
+while $\sqcup$ represents the union between two tuple types.
 
@@
 and each $v_{i}$ represents the type of the value that table keys of type $k_{i}$ map to.
-Key types can only be literal types, base types, or the dynamic type.
+Key types can only be literal types, base types, or the top type.
 We made this restriction to the type of the keys because the statistics
@@
 of the tables are records, lists, and hashes.
-The dynamic type is an option when we need a loose table type.
+The type $\Value$ is an option when we need a loose table type.
+For instance, $\{\Value:\Value\}_{closed}$ represents the type of a
+table that both indices and values can have any type.
 Value types can be any first-level type, and can optionally include
@@
 to classify table types.
-The tag \emph{unique} represents tables that we know their contents,
-and that there is no alias to them.
+The tag \emph{unique} represents tables that have no keys that do not
+inhabit one of the table's key types, and that have no alias.
 In particular, the type of the table constructor has this tag.
@@
 The tag \emph{closed} represents table types that do not provide
-any guarantees about the types not listed in the table type.
-In particular, type annotations as well as interfaces and userdata
-declarations always describe \emph{closed} table types.
+any guarantees about keys with types not listed in the table type.
+In particular, in the concrete syntax, type annotations, interface
+declarations, and userdata declarations always describe \emph{closed} table types.
 In the next sections we explain in more detail why we need
@@
 
-Type Lua includes projection types as a way to project
-unions of tuple types into unions of first-level types.
-Projection types have the form $\pi_{i}^{x}$, where $x$ is a variable
-that stores an union of tuple types in a special type environment,
-and $i$ is an index that indicates how to project a first-level
-union type from $x$.
-For instance, if $x$ is the union of tuple types
-$(\Integer \times \Integer) \sqcup (\Nil \times \String)$,
-then $\pi_{1}^{x}$ represents the type $\Integer \cup \Nil$
-while $\pi_{2}^{x}$ represents the type $\Integer \cup \String$.
+Typed Lua includes projection types as a way to project
+unions of tuple types into unions of first-level types,
+as we mentioned in Section \ref{sec:unions}.
 In Section \ref{sec:rules} we show in more detail how our type system
@@
 the type of a local variable that depends on the type of another local variable.
+
+Typed Lua includes the dynamic type $\Any$ for allowing programmers
+to mix static and dynamic typing.
 
@@
 \mylabel{T-CAST}\\
-\dfrac{t \subtype \env_{1}(n)}
+\dfrac{\env_{1}(n) \subtype t}
       {\env_{1} \vdash {<}t{>} \; n:t, \env_{1}[n \mapsto t]}

### 2015-01-12 doc/thesis/system.tex
Minor improvements to the text about subtyping

@@
 However, we focus the discussion on the definition of subtyping because,
-as we mentioned in Chapter \ref{chap:review}, we can combine the
+as we mentioned in Section \ref{sec:gradual}, we can combine the
 consistency and subtyping relations to achieve consistent-subtyping.
@@
 The variable $\senv$ is a set of pairs of recursion variables.
-We shall notice that we need this set to record the hypotheses
-that we assume to check subtyping between recursive types.
+We need this set to record the hypotheses that we assume when checking
+recursive types.
 
@@
 for defining that literal types are subtypes of their respective base types,
-and that $\Integer$ is subtype of $\Number$.
-We define these subtyping rules as follows:
+and that $\Integer$ is subtype of $\Number$:
 \[
@@
 Subtyping is reflexive and transitive;
-due to the transitivity of subtyping, we could have omitted the rule \textsc{S-INT2}.
+therefore, we could have omitted the rule \textsc{S-INT2}.
 More precisely, we could have defined a transitive rule for first-level
-types instead of defining specific rules to transitive cases.
+types instead of defining specific rules for transitive cases.
 For instance, a transitive rule would allow us to derive that
@@
 way that is close to the implementation, as this way allows us to use
-subtyping to formalize consistent-subtyping.
-
-That is the reason why the dynamic type $\Any$ is neither the bottom nor the top type,
-but a separate type that is subtype only of itself through the reflexive rule.
+subtyping to easily formalize consistent-subtyping.
+
+Our type system includes the top type $\Value$,
+so any first-level type is a subtype of $\Value$:
+\[
+\begin{array}{c}
+\mylabel{S-VALUE}\\
+\senv \vdash t \subtype \Value
+\end{array}
+\]
+
+Many programming languages include a bottom type to represent
+an empty value that programmers can use as a default expression,
+and we could have used the type $\Nil$ for this role.
+However, making $\Nil$ the bottom type would lead to several expressions
+that would pass the type checker, but that would fail during run-time
+in the presence of a \texttt{nil} value.
+Thus, our type system does not have a bottom type, and $\Nil$ is a
+subtype only of itself and of $\Value$.
+
+Another type that is only a subtype of itself and of the type $\Value$
+is the type $\Self$.
+
+The subtyping rules for union types are standard:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-UNION1}\\
+\dfrac{\senv \vdash t_{1} \subtype t \;\;\;
+       \senv \vdash t_{2} \subtype t}
+      {\senv \vdash t_{1} \cup t_{2} \subtype t}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-UNION2}\\
+\dfrac{\senv \vdash t \subtype t_{1}}
+      {\senv \vdash t \subtype t_{1} \cup t_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-UNION3}\\
+\dfrac{\senv \vdash t \subtype t_{2}}
+      {\senv \vdash t \subtype t_{1} \cup t_{2}}
+\end{array}
+\end{array}
+\]
+
+The first rule shows that a union type $t_{1} \cup t_{2}$
+is subtype of $t$ if both $t_{1}$ and $t_{2}$ are subtypes
+of $t$;
+and the other rules show that a type $t$ is subtype
+of a union type $t_{1} \cup t_{2}$ if $t$ is subtype of
+either $t_{1}$ or $t_{2}$.
+
+The subtyping rule for function types is also standard:
+\[
+\begin{array}{c}
+\mylabel{S-FUNCTION}\\
+\dfrac{\senv \vdash s_{2} \subtype s_{1} \;\;\;
+       \senv \vdash r_{1} \subtype r_{2}}
+      {\senv \vdash s_{1} \rightarrow s_{1} \subtype s_{2} \rightarrow r_{2}}
+\end{array}
+\]
+
+The rule \textsc{S-FUNCTION} shows that subtyping between
+function types is contravariant on the type of the parameter list
+and covariant on the return type.
+In the previous section we explained why our type system uses
+second-level types to represent the type of the parameter list
+and the return type.
+Now, we explain their subtyping rules.
+
+The simplest tuple type is the type $\Void$ that is the type of an
+empty tuple and it is subtype only of itself.
+
+The subtyping rule for pair types is standard, but the subtyping rules
+for variadic types are not so obvious:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-PAIR}\\
+\dfrac{\senv \vdash t_{1} \subtype t_{2} \;\;\;
+       \senv \vdash s_{1} \subtype s_{2}}
+      {\senv \vdash t_{1} \times s_{1} \subtype t_{2} \times s_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-VARARG1}\\
+\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2} \cup \Nil}
+      {\senv \vdash t_{1}{*} \subtype t_{2}{*}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-VARARG2}\\
+\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2}}
+      {\senv \vdash t_{1}{*} \subtype t_{2} \times \Void}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-VARARG3}\\
+\dfrac{\senv \vdash t_{1} \subtype t_{2} \cup \Nil}
+      {\senv \vdash t_{1} \times \Void \subtype t_{2}{*}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-VARARG4}\\
+\dfrac{\senv \vdash t_{1}{*} \subtype t_{2} \times \Void \;\;\;
+       \senv \vdash t_{1}{*} \subtype s_{2}}
+      {\senv \vdash t_{1}{*} \subtype t_{2} \times s_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-VARARG5}\\
+\dfrac{\senv \vdash t_{1} \times \Void \subtype t_{2}{*} \;\;\;
+       \senv \vdash s_{1} \subtype t_{2}{*}}
+      {\senv \vdash t_{1} \times s_{1} \subtype t_{2}{*}}
+\end{array}
+\end{array}
+\]
+
+The subtyping rules for pair and variadic types show that they are covariant.
+While we need just one subtyping rule for pair types,
+we need five different subtyping rules for variadic types
+to handle all the cases where they can appear.
+The rule \textsc{S-VARARG1} handles the case that both tuple types end
+with variadic types, while the other rules handle the cases that only
+one tuple type ends with a variadic type.
+Note that the case where both tuple types end with the type $\Void$ does
+not require any special rule.
+The rule \textsc{S-VARARG1} shows that $t_{1}{*}$ is subtype of $t_{2}{*}$
+if $t_{1} \cup \Nil$ is subtype of $t_{2} \cup \Nil$.
+This rule explicit includes $\Nil$ in both sides because otherwise
+$\Nil{*}$ would not be a subtype of several other variadic types.
+For instance, $\Nil{*}$ would not be subtype of $\Number{*}$,
+as $\Nil \not\subtype \Number$.
+In the next section we will show that we use these subtyping relations,
+along with the types $\Value$ and $\Nil$, to make our type system reflect
+the semantics of Lua on discarding extra parameters, and
+replacing missing parameters.
+
+The subtyping rules for unions of tuple types are similar to the
+subtyping rules for unions of first-level types:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-UNION4}\\
+\dfrac{\senv \vdash s_{1} \subtype s \;\;\;
+       \senv \vdash s_{2} \subtype s}
+      {\senv \vdash s_{1} \sqcup s_{2} \subtype s}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-UNION5}\\
+\dfrac{\senv \vdash s \subtype s_{1}}
+      {\senv \vdash s \subtype s_{1} \sqcup s_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-UNION6}\\
+\dfrac{\senv \vdash s \subtype s_{2}}
+      {\senv \vdash s \subtype s_{1} \sqcup s_{2}}
+\end{array}
+\end{array}
+\]
+
+Back to the subtyping rules between first-level types,
+the subtyping rule between \emph{closed} table types resembles the
+standard subtyping rule between records:
+\[
+\begin{array}{c}
+\mylabel{S-TABLE1}\\
+\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
+       \senv \vdash k_{j} \subtype k_{i}' \;\;\;
+       \senv \vdash k_{i}' \subtype k_{j} \;\;\;
+       \senv \vdash v_{j} \subtype_{c} v_{i}'}
+      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{closed} \subtype \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}} \; m \ge n
+\end{array}
+\]
+
+The rule \textsc{S-TABLE1} allows width subtyping between \emph{closed}
+table types, and introduced the auxiliary relation $\subtype_{c}$ to
+handle depth subtyping on the type of the values stored in the table fields.
+We need an auxiliary relation because the subtyping of the
+type of the values stored in the table fields changes according to
+the tags of the table types.
+We define the relation $\subtype_{c}$ as follows:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-FIELD1}\\
+\dfrac{\senv \vdash v_{1} \subtype v_{2} \;\;\;
+       \senv \vdash v_{2} \subtype v_{1}}
+      {\senv \vdash v_{1} \subtype_{c} v_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FIELD2}\\
+\dfrac{\senv \vdash v_{1} \subtype v_{2}}
+      {\senv \vdash \Const \; v_{1} \subtype_{c} \Const \; v_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-FIELD3}\\
+\dfrac{\senv \vdash v_{1} \subtype v_{2}}
+      {\senv \vdash v_{1} \subtype_{c} \Const \; v_{2}}
+\end{array}
+\end{array}
+\]
+
+These rules allow closed table types to have depth subtyping on $\Const$ fields.
+The rule \textsc{S-FIELD1} defines that mutable fields are invariant,
+while the rule \textsc{S-FIELD2} defines that immutable fields are covariant.
+The rule \textsc{S-FIELD3} defines that it is safe to promote fields
+from mutable to immutable.
+We do not include a rule that allows promoting fields from immutable
+to mutable because this would be unsafe due to variance.
+
+There is a limitation on \emph{closed} table types that led us to
+introduce \emph{open} and \emph{unique} table types.
+If the table constructor had a \emph{closed} table type, then
+programmers would not be able to use it to initialize a variable with
+a table type that describes a more general type.
+For instance,
+\begin{verbatim}
+    local t:{"x":integer, "y":integer?} = { x = 1, y = 2 }
+\end{verbatim}
+would not type check, as the type of the table constructor would not
+be subtype of the type in the annotation.
+More precisely,
+\[
+\{``x":1, ``y":2\}_{closed} \not\subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
+\]
+
+Simply promoting the type of each table value to its supertype would
+not overcome this limitation, as it still would give to the table constructor
+a closed table type without covariant mutable fields.
+Thus, programmers would not be able to use the table constructor to
+initialize a variable with a table type that includes an optional field.
+Using the previous example,
+\begin{align*}
+& \{``x":\Integer, ``y":\Integer\}_{closed} \not\subtype \\
+& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
+\end{align*}
+
+We introduced \emph{unique} table types to avoid this limitation,
+as they represent the type of the tables that have no keys that do not
+inhabit one of the table's key types, and that have no alias.
+In particular, this is the case of the table constructor.
+The following subtyping rule defines the subtyping relation among
+\emph{unique} table types and other table types:
+\[
+\begin{array}{c}
+\mylabel{S-TABLE2}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
+       \senv \vdash v_{i} \subtype_{u} v_{i}' \;\;\;
+       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
+      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
+                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open|unique}}
+\end{array}
+\]
+
+Unlike the rule \textsc{S-TABLE1}, the rule \textsc{S-TABLE2} allows
+covariant keys and do not allow width subtyping.
+We allow this behavior to \emph{unique} table types because we also
+want to use them as a way to join table fields.
+For instance, we may want to use the table constructor to initialize
+a variable with a table type that describes a hash.
+
+The rule \textsc{S-TABLE2} introduced the auxiliary relations
+$\subtype_{u}$ and $\subtype_{o}$.
+The first allows depth subtyping on all fields,
+while the second allows the omission of optional fields.
+We define them as follows:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-FIELD4}\\
+\dfrac{\senv \vdash v_{1} \subtype v_{2}}
+      {\senv \vdash v_{1} \subtype_{u} v_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FIELD5}\\
+\dfrac{\senv \vdash v_{1} \subtype v_{2}}
+      {\senv \vdash \Const \; v_{1} \subtype_{u} \Const \; v_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FIELD6}\\
+\dfrac{\senv \vdash v_{1} \subtype v_{2}}
+      {\senv \vdash v_{1} \subtype_{u} \Const \; v_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-FIELD7}\\
+\dfrac{\senv \vdash \Nil \subtype v}
+      {\senv \vdash \Nil \subtype_{o} v}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FIELD8}\\
+\dfrac{\senv \vdash \Nil \subtype v}
+      {\senv \vdash \Nil \subtype_{o} \Const \; v}
+\end{array}
+\end{array}
+\]
+
+We also do not allow width and depth subtyping between \emph{open}
+table types or between \emph{open} and \emph{closed} table types:
+\[
+\begin{array}{c}
+\mylabel{S-TABLE3}\\
+\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
+       \senv \vdash v_{i} \subtype_{c} v_{i}' \;\;\;
+       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
+      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
+                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open}}
+\end{array}
+\]
+
+The rule \textsc{S-TABLE3} uses the same rules from
+\textsc{S-TABLE1} for handling the type of the values of each key, and
+it uses the same rules from \textsc{S-TABLE2} to allow joining fields plus
+omitting optional fields.
+
+In the next section we will show in more detail how our type system
+handles \emph{open} and \emph{unique} table types to allow the refinement of
+table types.
+
+We use the \emph{Amber rule} \citep{cardelli1986amber} to define
+subtyping between recursive types:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-AMBER}\\
+\dfrac{\senv[x_{1} \subtype x_{2}] \vdash t_{1} \subtype t_{2}}
+      {\senv \vdash \mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-ASSUMPTION}\\
+\dfrac{x_{1} \subtype x_{2} \in \senv}
+      {\senv \vdash x_{1} \subtype x_{2}}
+\end{array}
+\end{array}
+\]
+
+The rule \textsc{S-AMBER} also uses the rule \textsc{S-ASSUMPTION}
+to check whether $\mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}$.
+Both rules use the set of assumptions $\senv$,
+where each assumption is a pair of recursion variables.
+The rule \textsc{S-AMBER} extends $\senv$ with the assumption
+$x_{1} \subtype x_{2}$ to check whether $t_{1} \subtype t_{2}$.
+The rule \textsc{S-ASSUMPTION} allows the rule \textsc{S-AMBER}
+to check whether an assumption is valid.
+
+A recursive type may appear inside a first-level type, and our
+type system includes subtyping rules to handle subtyping between
+recursive types and other first-level types:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-UNFOLDR}\\
+\dfrac{\senv \vdash t_{1} \subtype [x \mapsto \mu x.t_{2}]t_{2}}
+      {\senv \vdash t_{1} \subtype \mu x.t_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-UNFOLDL}\\
+\dfrac{\senv \vdash [x \mapsto \mu x.t_{1}]t_{1} \subtype t_{2}}
+      {\senv \vdash \mu x.t_{1} \subtype t_{2}}
+\end{array}
+\end{array}
+\]
+
+As an example, the rule \textsc{S-UNFOLDR} allows our type system to
+type check the function \texttt{insert} from Section \ref{sec:alias}:
+\begin{verbatim}
+    local function insert (e:Element?, v:integer):Element
+      return { info = v, next = e }
+    end
+\end{verbatim}
+that is, the type checker uses the rule \textsc{S-UNFOLDR} to verify whether
+the type of the table constructor is a subtype of \texttt{Element}:
+\begin{align*}
+\{ & ``info":\Integer, \\
+   & ``next":\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed} \cup \Nil \}_{unique} \subtype \\
+& \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}
+\end{align*}
+
+Projection types are subtypes only of themselfs and of $\Value$.
+More precisely, a projection type $\pi_{i}^{x}$ is a subtype of the
+same projection type $\pi_{i}^{x}$, which shares the same union of
+tuples $x$ and the same index $i$, and it is also a subtype of $\Value$.
+
+The dynamic type $\Any$ is neither the bottom nor the top type,
+but a separate type that is subtype only of itself and of $\Value$.
 
 Even though the dynamic type $\Any$ does not interact with subtyping,
-it does interact with consistent-subtyping:
+it does interact with consistent-subtyping.
+We present the consistent-subtyping rules as a deduction system for
+the consistent-subtyping relation $\senv \vdash t_{1} \lesssim t_{2}$.
+Like in the subtyping relation, $\senv$ is also a set of pairs of
+recursion variables.
+We define the consistent-subtyping rules for the dynamic type $\Any$
+as follows:
 \[
@@
 Because of these two rules, consistent-subtyping cannot be transitive.
-Still, these two rules are the only rules that differ between
-subtyping and consistent-subtyping.
-
-Our type system includes the top type $\Value$,
-so any first-level type is a subtype of $\Value$:
-\[
-\begin{array}{c}
-\mylabel{S-VALUE}\\
-\senv \vdash t \subtype \Value
-\end{array}
-\]
-
-Some people might expect that $\Nil$ is the bottom type
-of our type system, but it is not,
-and it is subtype only of itself through the reflexive rule.
-Many programming languages include a bottom type to represent
-an empty value that programmers can use as a default expression.
-Typed Lua does not have this behavior because in Lua the tag \texttt{nil}
-represents the absence of an useful value.
-Thus, making $\Nil$ the bottom type would lead to several expressions
-that would pass the type checker, but that would fail during run-time
-in the presence of a \texttt{nil} value.
-
-Another type that only follows the reflexive rule is the type $\Self$,
-that is, it is subtype only of itself.
-
-The subtyping rules for union types are standard:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-UNION1}\\
-\dfrac{\senv \vdash t_{1} \subtype t \;\;\;
-       \senv \vdash t_{2} \subtype t}
-      {\senv \vdash t_{1} \cup t_{2} \subtype t}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-UNION2}\\
-\dfrac{\senv \vdash t \subtype t_{1}}
-      {\senv \vdash t \subtype t_{1} \cup t_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-UNION3}\\
-\dfrac{\senv \vdash t \subtype t_{2}}
-      {\senv \vdash t \subtype t_{1} \cup t_{2}}
-\end{array}
-\end{array}
-\]
-
-The first rule shows that a union type $t_{1} \cup t_{2}$
-is subtype of $t$ if both $t_{1}$ and $t_{2}$ are subtypes
-of $t$;
-and the other rules show that a type $t$ is subtype
-of a union type $t_{1} \cup t_{2}$ if $t$ is subtype of
-$t_{1}$ or if $t$ is subtype of $t_{2}$.
-
-In the implementation of Typed Lua we use consistent-subtyping to
-normalize and simplify union types, but we let them free in the
+These two rules are the only rules that differ between
+subtyping and consistent-subtyping, if we implement the subtyping rules
+as we do in this section.
+
+In the implementation of Typed Lua we also use consistent-subtyping to
+normalize and simplify union types, though we let them free in the
 formalization.
@@
 \texttt{1} is consistent-subtype of \texttt{number}.
-
-The subtyping rule for function types is also standard:
-\[
-\begin{array}{c}
-\mylabel{S-FUNCTION}\\
-\dfrac{\senv \vdash s_{2} \subtype s_{1} \;\;\;
-       \senv \vdash r_{1} \subtype r_{2}}
-      {\senv \vdash s_{1} \rightarrow s_{1} \subtype s_{2} \rightarrow r_{2}}
-\end{array}
-\]
-
-The rule \textsc{S-FUNCTION} shows that subtyping between
-function types is contravariant on the type of the parameter list
-and covariant on the return type.
-In the last section we explained why our type system uses
-second-level types to represent the type of the parameter list
-and the return type.
-Now, we explain their subtyping rules.
-
-The simplest second-level type is the type $\Void$ that
-is subtype only of itself through the reflexive rule.
-
-The subtyping rule for tuple types is standard, but the subtyping
-rules for variadic types are not so obvious:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-TUPLE}\\
-\dfrac{\senv \vdash t_{1} \subtype t_{2} \;\;\;
-       \senv \vdash s_{1} \subtype s_{2}}
-      {\senv \vdash t_{1} \times s_{1} \subtype t_{2} \times s_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-VARARG1}\\
-\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2} \cup \Nil}
-      {\senv \vdash t_{1}{*} \subtype t_{2}{*}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-VARARG2}\\
-\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2}}
-      {\senv \vdash t_{1}{*} \subtype t_{2} \times \Void}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-VARARG3}\\
-\dfrac{\senv \vdash t_{1} \subtype t_{2} \cup \Nil}
-      {\senv \vdash t_{1} \times \Void \subtype t_{2}{*}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-VARARG4}\\
-\dfrac{\senv \vdash t_{1}{*} \subtype t_{2} \times \Void \;\;\;
-       \senv \vdash t_{1}{*} \subtype s_{2}}
-      {\senv \vdash t_{1}{*} \subtype t_{2} \times s_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-VARARG5}\\
-\dfrac{\senv \vdash t_{1} \times \Void \subtype t_{2}{*} \;\;\;
-       \senv \vdash s_{1} \subtype t_{2}{*}}
-      {\senv \vdash t_{1} \times s_{1} \subtype t_{2}{*}}
-\end{array}
-\end{array}
-\]
-
-The subtyping rules for tuple and variadic types show that they are covariant.
-While we need just one subtyping rule for tuple types,
-we need five different subtyping rules for variadic types
-to handle all the cases where they can appear.
-The rule \textsc{S-VARARG1} handles the subtyping between two variadic types,
-while the other rules handle the subtyping between a tuple type and a variadic type.
-The rule \textsc{S-VARARG1} shows that $t_{1}{*}$ is subtype of $t_{2}{*}$
-if $t_{1} \cup \Nil$ is subtype of $t_{2} \cup \Nil$.
-This rule explicit includes $\Nil$ in both sides because otherwise
-$\Nil{*}$ would not be a subtype of several other variadic types.
-For instance, $\Nil{*}$ would not be subtype of $\Number{*}$,
-as $\Nil \not\subtype \Number$.
-The rule \textsc{S-VARARG2} shows that a variadic type $t_{1}{*}$ is
-subtype of a tuple type $t_{2} \times \Void$ if $t_{1} \cup \Nil$ is subtype of $t_{2}$.
-The rule \textsc{S-VARARG3} shows that a tuple type $t_{1} \times \Void$
-is subtype of a variadic type $t_{2}{*}$ if $t_{1}$ is subtype of $t_{2} \cup \Nil$.
-The rule \textsc{S-VARARG4} shows that a variadic type is subtype of
-a tuple type whenever the type of the value that it generates is
-subtype of all the elements of the tuple.
-The rule \textsc{S-VARARG5} shows that a tuple type is subtype of a
-variadic type whenever all the elements of the tuple are subtype of
-the type of the value that it generates.
-In the next section we will show that we use these subtyping relations,
-along with the types $\Value$ and $\Nil$, to make our type system reflect
-the semantics of Lua on discarding extra parameters, and
-replacing missing parameters.
-
-The subtyping rules for unions of tuple types are similar to the
-subtyping rules for unions of first-level types:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-UNION4}\\
-\dfrac{\senv \vdash s_{1} \subtype s \;\;\;
-       \senv \vdash s_{2} \subtype s}
-      {\senv \vdash s_{1} \sqcup s_{2} \subtype s}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-UNION5}\\
-\dfrac{\senv \vdash s \subtype s_{1}}
-      {\senv \vdash s \subtype s_{1} \sqcup s_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-UNION6}\\
-\dfrac{\senv \vdash s \subtype s_{2}}
-      {\senv \vdash s \subtype s_{1} \sqcup s_{2}}
-\end{array}
-\end{array}
-\]
-
-Back to the subtyping rules between first-level types,
-the subtyping rule between \emph{closed} table types resembles the
-standard subtyping rule between records:
-\[
-\begin{array}{c}
-\mylabel{S-TABLE1}\\
-\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
-       \senv \vdash k_{j} \subtype k_{i}' \;\;\;
-       \senv \vdash k_{i}' \subtype k_{j} \;\;\;
-       \senv \vdash v_{j} \subtype_{c} v_{i}'}
-      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{closed} \subtype \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}} \; m \ge n
-\end{array}
-\]
-
-The rule \textsc{S-TABLE1} introduced the relation $\subtype_{c}$.
-It is just an auxiliary relation that defines the subtyping of the
-type of the values stored in the table fields.
-We need an auxiliary relation because the subtyping of the
-type of the values stored in the table fields changes according to
-the tags of the table types.
-We define the relation $\subtype_{c}$ as follows:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-FIELD1}\\
-\dfrac{\senv \vdash v_{1} \subtype v_{2} \;\;\;
-       \senv \vdash v_{2} \subtype v_{1}}
-      {\senv \vdash v_{1} \subtype_{c} v_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FIELD2}\\
-\dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash \Const \; v_{1} \subtype_{c} \Const \; v_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-FIELD3}\\
-\dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash v_{1} \subtype_{c} \Const \; v_{2}}
-\end{array}
-\end{array}
-\]
-
-These rules allow closed table types to have width subtyping
-and depth subtyping on $\Const$ fields.
-The rule \textsc{S-FIELD1} defines that mutable fields are invariant,
-while the rule \textsc{S-FIELD2} defines that immutable fields are covariant.
-The rule \textsc{S-FIELD3} defines that it is safe to promote fields
-from mutable to immutable.
-We do not include a rule that allows promoting fields from immutable
-to mutable because it would not be safe, as this rule would allow
-programmers to change the value that is stored in an immutable field.
-
-There is a limitation on \emph{closed} table types that led us to
-introduce \emph{open} and \emph{unique} table types.
-If the table constructor had a \emph{closed} table type, then
-programmers would not be able to use it to initialize a variable with
-a table type that describes a more general type.
-For instance,
-\begin{verbatim}
-    local t:{"x":integer, "y":integer?} = { x = 1, y = 2 }
-\end{verbatim}
-would not type check, as the type of the table constructor would not
-be subtype of the type in the annotation.
-More precisely,
-\[
-\{``x":1, ``y":2\}_{closed} \not\subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
-\]
-
-Simply promoting the type of each table value to its supertype would
-not overcome this limitation, as it still would give to the table constructor
-a closed table type without covariant mutable fields.
-Thus, programmers would not be able to use the table constructor to
-initialize a variable with a table type that includes an optional field.
-Using the previous example,
-\begin{align*}
-& \{``x":\Integer, ``y":\Integer\}_{closed} \not\subtype \\
-& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
-\end{align*}
-
-We introduced \emph{unique} table types to avoid this limitation,
-as they represent the type of the tables that we known their contents.
-In particular, this is the case of the table constructor.
-The following subtyping rule defines the subtyping relation among
-\emph{unique} table types and other table types:
-\[
-\begin{array}{c}
-\mylabel{S-TABLE2}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
-       \senv \vdash v_{i} \subtype_{u} v_{i}' \;\;\;
-       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
-      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
-                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open|unique}}
-\end{array}
-\]
-
-Unlike the rule \textsc{S-TABLE1}, the rule \textsc{S-TABLE2} allows
-covariant keys.
-We allow this behavior to \emph{unique} table types because we also
-want to use them as a way to join table fields.
-For instance, we may want to use the table constructor to initialize
-a variable with a table type that describes a hash.
-
-The rule \textsc{S-TABLE2} introduced the auxiliary relations
-$\subtype_{u}$ and $\subtype_{o}$.
-The first allows width and depth subtyping on all fields,
-while the second allows the omission of optional fields.
-We define them as follows:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-FIELD4}\\
-\dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash v_{1} \subtype_{u} v_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FIELD5}\\
-\dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash \Const \; v_{1} \subtype_{u} \Const \; v_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FIELD6}\\
-\dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash v_{1} \subtype_{u} \Const \; v_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-FIELD7}\\
-\dfrac{\senv \vdash \Nil \subtype v}
-      {\senv \vdash \Nil \subtype_{o} v}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FIELD8}\\
-\dfrac{\senv \vdash \Nil \subtype v}
-      {\senv \vdash \Nil \subtype_{o} \Const \; v}
-\end{array}
-\end{array}
-\]
-
-Using \emph{unique} table types to represent the type of the table
-constructor allows our type system to type check the previous example.
-More precisely,
-\[
-\{``x":1, ``y":2\}_{unique} \subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
-\]
-
-Even though \emph{unique} table types overcome our limitation,
-its subtyping rule allows unsafe aliasing.
-The covariance on mutable fields allows programs that create unsafe
-aliases to pass the type checker.
-For instance,
-\begin{verbatim}
-    local t1 = { foo = 5 }
-    local t2:{"foo":integer?} = t1
-    t2.foo = nil
-\end{verbatim}
-should not type check, as the last line is erasing the field \texttt{foo}
-from \texttt{t1}.
-This program would type checks as it is, because our implementation uses
-local type inference to assign static types to unannotated locals.
-In this example, the type of \texttt{t1} would be subtype of the type of \texttt{t2}.
-More precisely,
-\[
-\{``foo":\Integer\}_{unique} \subtype \{``foo":\Integer \cup \Nil\}_{closed}
-\]
-
-In the next section we will show in more detail that this example
-does not type check, because our type system promotes an alias
-expression from \emph{unique} to \emph{closed}, and promotes the
-aliased variable from \emph{unique} to \emph{open}.
-This means that the previous example does not type check because
-the type of the expression \texttt{t1} is not subtype of the type
-of the variable \texttt{t2}.
-More precisely,
-\[
-\{``foo":\Integer\}_{closed} \not\subtype \{``foo":\Integer \cup \Nil\}_{closed}
-\]
-
-Thus, we introduced \emph{open} table types to handle the aliasing of table types.
-We define the subtyping relation among \emph{open} table types
-and \emph{closed} or \emph{open} table types as follows:
-\[
-\begin{array}{c}
-\mylabel{S-TABLE3}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
-       \senv \vdash v_{i} \subtype_{c} v_{i}' \;\;\;
-       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
-      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
-                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open}}
-\end{array}
-\]
-
-The rule \textsc{S-TABLE3} uses the same rules from \textsc{S-TABLE1}
-for handling the type of the values of each key, and it uses the same
-rules from \textsc{S-TABLE2} to allow joining fields plus omitting optional fields.
-
-In the next section we will show in more detail how our type system
-keeps track of \emph{open} and \emph{unique} table types to make them safe.
-Moreover, we will also show that we can use them to allow the refinement
-of table types.
-
-We use the \emph{Amber rule} \citep{cardelli1986amber} to define
-subtyping between recursive types:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-AMBER}\\
-\dfrac{\senv[x_{1} \subtype x_{2}] \vdash t_{1} \subtype t_{2}}
-      {\senv \vdash \mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-ASSUMPTION}\\
-\dfrac{x_{1} \subtype x_{2} \in \senv}
-      {\senv \vdash x_{1} \subtype x_{2}}
-\end{array}
-\end{array}
-\]
-
-The rule \textsc{S-AMBER} also uses the rule \textsc{S-ASSUMPTION}
-to check whether $\mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}$.
-Both rules use the set of assumptions $\senv$,
-where each assumption is a pair of recursion variables.
-The rule \textsc{S-AMBER} extends $\senv$ with the assumption
-$x_{1} \subtype x_{2}$ to check whether $t_{1} \subtype t_{2}$.
-The rule \textsc{S-ASSUMPTION} allows the rule \textsc{S-AMBER}
-to check whether an assumption is valid.
-
-A recursive type may appear inside a first-level type, and our
-type system includes subtyping rules to handle subtyping between
-recursive types and other first-level types:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-UNFOLDR}\\
-\dfrac{\senv \vdash t_{1} \subtype [x \mapsto \mu x.t_{2}]t_{2}}
-      {\senv \vdash t_{1} \subtype \mu x.t_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-UNFOLDL}\\
-\dfrac{\senv \vdash [x \mapsto \mu x.t_{1}]t_{1} \subtype t_{2}}
-      {\senv \vdash \mu x.t_{1} \subtype t_{2}}
-\end{array}
-\end{array}
-\]
-
-The rules \textsc{S-UNFOLDR}, \textsc{S-AMBER}, and \textsc{S-ASSUMPTION}
-allow our type system to type check the function \texttt{insert}
-from Section \ref{sec:alias}:
-\begin{verbatim}
-    local function insert (e:Element?, v:integer):Element
-      return { info = v, next = e }
-    end
-\end{verbatim}
-
-First, a recursive type appears inside the type of the table constructor.
-Thus, the type checker uses the rule \textsc{S-UNFOLDR} to verify whether
-the type of the table constructor is a subtype of \texttt{Element}:
-\begin{align*}
-\{ & ``info":\Integer, \\
-   & ``next":\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed} \cup \Nil \}_{unique} \subtype \\
-& \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}
-\end{align*}
-
-After unfolding, the recursive types appear only as part of the type
-of the value of the key type $``next"$.
-Thus, the type checker uses the rules \textsc{S-AMBER} and \textsc{S-ASSUMPTION}
-to verify whether these recursive types are subtypes between each other:
-\begin{align*}
-& \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed} \subtype \\
-& \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}
-\end{align*}
-
-Projection types follow the reflexive subtyping rule.
-A projection type $\pi_{i}^{x}$ is a subtype of another projection type $\pi_{i}^{x}$
-that shares the same union of tuples $x$ and the same index $i$.
 
@@
 
+Using \emph{unique} table types to represent the type of the table
+constructor allows our type system to type check the previous example.
+More precisely,
+\[
+\{``x":1, ``y":2\}_{unique} \subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
+\]
+
+Even though \emph{unique} table types overcome our limitation.
+The covariance on mutable fields could allow programs that create unsafe
+aliases to pass the type checker.
+For instance,
+\begin{verbatim}
+    local t1 = { foo = 5 }
+    local t2:{"foo":integer?} = t1
+    t2.foo = nil
+\end{verbatim}
+does not type check, as the last line is erasing the field \texttt{foo}
+from \texttt{t1}.
+This program would type checks as it is, because our implementation uses
+local type inference to assign static types to unannotated locals.
+In this example, the type of \texttt{t1} would be subtype of the type of \texttt{t2}.
+More precisely,
+\[
+\{``foo":\Integer\}_{unique} \subtype \{``foo":\Integer \cup \Nil\}_{closed}
+\]
+
+In the next section we will show in more detail that this example
+does not type check, because our type system promotes an alias
+expression from \emph{unique} to \emph{closed}, and promotes the
+aliased variable from \emph{unique} to \emph{open}.
+This means that the previous example does not type check because
+the type of the expression \texttt{t1} is not subtype of the type
+of the variable \texttt{t2}.
+More precisely,
+\[
+\{``foo":\Integer\}_{closed} \not\subtype \{``foo":\Integer \cup \Nil\}_{closed}
+\]
+
+Thus, we introduced \emph{open} table types to handle the aliasing of table types.
+
 The simplest form of the table constructor is the empty table $\{\}$.

### 2015-01-13 doc/thesis/system.tex
Starting the presentation of typing rules with assignments

@@
 $s$ are statements, $e$ are expressions, $l$ are left-hand values,
-$el$ are expression lists, $c$ are literal constants, $p$ are expression pairs,
+$el$ are expression lists, $c$ are literal constants, $p$ are table fields,
 $me$ are expressions with multiple results, $f$ are function declarations,
@@
 
-We will present the typing rules as a deduction system for the typing
-relations $\env_{1} \vdash s : \env_{2}$ and $\env_{1} \vdash e : t, \env_{2}$. 
-We will use the first one for typing statements, and the second one
+We present the typing rules as a deduction system for the typing
+relations $\env_{1}, \penv \vdash s, \env_{2}$ and
+$\env_{1}, \penv \vdash e : t, \env_{2}$.
+We use the first one for typing statements, and the second one
 for typing expressions.
-The first relation means that given the type environment $\env_{1}$,
+The first relation means that given the type environment $\env_{1}$
+and the projection environment $\penv$,
 typing the statement $s$ produces the new type environment $\env_{2}$.
-The second relation means that given the type environment $\env_{1}$,
-the expression $e$ has type $t$, and produces the new type environment $\env_{2}$.
+The second relation means that given the type environment $\env_{1}$
+and the projection environment $\penv$,
+the expression $e$ has type $t$ and produces the new type environment $\env_{2}$.
 We need two typing relations because only expressions have types,
-though both expressions and statements can modify the environment.
-The environment maps variables to types, and we will use $\env_{1}[n \mapsto t]$
-for meaning that we extended the environment $\env_{1}$ with the
-variable $n$ that maps to type $t$, and this mapping does not affect
-the other mappings that existed before.
-
-We start with the rule that defines the most unusual feature of
-our type system: the refinment of table types.
-The rule \textsc{T-REFINE} allow us to add new keys to \emph{open}
-and \emph{unique} table types:
-\[
-\begin{array}{c}
-\mylabel{T-REFINE}\\
-\dfrac{\begin{array}{c}
-       \env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique}\\
-       \env_{1} \vdash c:t_{1}, \env_{2} \;\;\;
-       \not \exists i \in 1..n \; t_{1} \lesssim k_{i} \;\;\;
-       \Nil \not\lesssim t_{1} \;\;\;
-       t_{2} = close(t)
-       \end{array}}
-      {\env_{1} \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, t_{1}{:}t_{2}\}_{open|unique}]}
-\end{array}
-\]
-
-We restricted the refinement of table types to include only literal
-keys, because its purpose is to make it easier the construction of
-table types that represent records.
-
-We use the refinement of table types to handle global variables.
-In Lua, the assignment \texttt{v = v + 1} translates to
-\texttt{\string_ENV["v"] = \string_ENV["v"] + 1} when \texttt{v}
-is not a local variable, where \texttt{\string_ENV} is a table
-that stores the global environment.
-For this reason, Typed Lua treats accesses to global variables as field accesses
-to an open table in the top-level scope.
-For instance,
-\[
-\string_ENV[``x"] \; {<}\String{>} = ``foo" \;;\; \string_ENV[``y"] \; {<}\Integer{>} = 1
-\]
-uses field assignment to add fields $``x"$ and $``y"$ to $\string_ENV$.
-In all the examples we assume that $\string_ENV$ is in the environment and
-has type $\{\}_{open}$.
-Therefore, after these field assignments $\string_ENV$ has type
-$\{``x":\String, ``y":\Integer\}_{open}$.
-
-We do not allow the refinement of table types to change the type of
-a field that is already present in the table type.
-For instance,
-\[
-\string_ENV[``x"] \; {<}\String{>} = ``foo" \;;\; \string_ENV[``x"] \; {<}\Integer{>} = 1
-\]
-do not type check, as we are trying to add a field that already exists
-in $\string_ENV$.
-We could have used union types to allow the refinement of existing fields,
-but this could lead to fields that have a too general type, which could
-decrease the ammount of static type checking.
-This means that our type system does not allow the table type to change
-towards a type that is not a subtype of the previous type.
-In this example,
-\[
-\{``x":\String \cup \Integer\}_{open} \not\subtype \{``x":\String\}_{open}
-\]
-
-We also do not allow the refinement of table types to introduce
-\emph{open} or \emph{unique} table fields.
-For instance,
-\begin{center}
-\begin{tabular}{l}
-$\string_ENV[``x"] \; {<}\{\}_{unique}{>} = \{\}$
-\end{tabular}
-\end{center}
-refines the type of $\string_ENV$ from $\{\}_{open}$ to $\{``x":\{\}_{closed}\}_{open}$,
-though we are trying to refine to $\{``x":\{\}_{unique}\}_{open}$.
-We made this restriction because we still need to prove that the refinement
-of table types is safe.
-
-Lua has multiple assignments, so Typed Lua also has to take this into
-consideration while refining table types.
-The rule \textsc{T-ASSINGMENT} shows that tuple types make it easy
-to use the consistent-subtyping relation to type check multiple assignments:
+though expressions and statements can modify the environment.
+We need two environments because $\env$ maps variables to types
+while $\penv$ maps projection variables to unions of tuple types.
+We use $\env_{1}[n \mapsto t]$ for meaning that we extended the environment
+$\env_{1}$ with the variable $n$ that maps to type $t$.
+We use $\penv[\pi \mapsto r]$ for meaning that we extendend the environment
+$\penv$ with the projection $\pi$ that maps to the union of tuple types $r$.
+
+Lua has multiple assignments, and the rule \textsc{T-ASSIGNMENT} uses
+tuple types along with consitent-subtyping to type check this Lua feature:
 \[
@@
 
-This rule is so simple because we left the difficult part to the
-type checking of expression lists.
-The list of left-hand values is also a kind of expression list.
-The difficult part is to check whether there is an expression in
-the expression list that messes the environment, and the rule
-\textsc{T-LHSLIST} captures this intuition:
+As an example,
+\[
+x, y = 1, ``foo"
+\]
+type checks because
+$1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times \Value{*}$,
+assuming that $x$ and $y$ are in the environment with
+types $\Integer$ and $\String$, respectively.
+Our type system uses $\Value{*}$ to drop extra values
+and $\Nil{*}$ to replace missing values.
+In this example, the rules \textsc{T-LHSLIST} and \textsc{T-EXPLIST2}
+are the rules that handle these features.
+
+We define the rule \textsc{T-LHSLIST} as follows:
 \[
@@
 a subtype of its previous type.
+We will show in this section that the refinment of table types can
+change the environment, and that is the reason why we need to check
+whether the environment is consistent.
+
+The rule \textsc{T-EXPLIST2} is close to the rule \textsc{T-LHSLIST}.
+The difference is that it uses $\Nil{*}$ to replace missing values.
+We define this rule as follows:
+\[
+\begin{array}{c}
+\mylabel{T-EXPLIST2}\\
+\dfrac{\env \vdash e_{k}:t_{k}, \env_{k} \;\;\;
+       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
+       n = |\;\vec{e}\;|}
+      {\env \vdash \vec{e}:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{m}}
+\end{array}
+\]
+
+The typing rules that type check expression lists include $\Nil{*}$
+in the end of the tuple type only when it does not end in a variadic type.
+For instance, the rules \textsc{T-EXPLIST3} and \textsc{T-EXPLIST4}
+handle the case that an expression list is just an expression that
+return multiple values, but only \textsc{T-EXPLIST3} needs to add
+$\Nil{*}$:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-EXPLIST3}\\
+\dfrac{\env \vdash me:t_{1} \times ... \times t_{n} \times \Void, \env_{1}}
+      {\env \vdash me:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{1}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{T-EXPLIST4}\\
+\dfrac{\env \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{1}}
+      {\env \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{1}}
+\end{array}
+\end{array}
+\]
+
+As an example, the rule \textsc{T-EXPLIST3} allows our type system
+to type check the following example:
+\[
+x, y, z = f()_{m}
+\]
+assuming that $x$, $y$, $z$, and $f$ are in the environment with
+types $\Integer$, $\String$, $\String \cup \Nil$, and
+$\Void \rightarrow \Integer \times \String \times \Void$, respectively.
+It type checks because
+\[
+\Integer \times \String \times \Nil{*} \lesssim \Integer \times \String \times \String \cup \Nil \times \Value{*}
+\]
+
+Conversely,
+\[
+x = f()_{m}
+\]
+type checks because
+\[
+\Integer \times \String \times \Nil{*} \lesssim \Integer \times \Value{*}
+\]
+
+Multiple assignments also appear in function and method applications,
+but our type system needs different typing rules for
+expression lists that appear in function and method applications,
+as we also want to catch arity mismatch.
+For instance, assuming that $x$ and $f$ are in the environment with
+types $\Integer \cup \Nil$ and
+$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void \rightarrow \Integer{*}$,
+respectively.
+We want that
+\[
+x = f(``foo")_{m}
+\]
+type checks because
+\[
+``foo" \times \Nil \times \Nil \times \Void \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
+\]
+but we also want that
+\[
+x = f(``foo",1,2,3)_{m}
+\]
+does not type check because
+\[
+``foo" \times 1 \times 2 \times 3 \times \Void \not\lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
+\]
+
+To do that, the rule \textsc{T-APPLY1} uses rules that type check
+expression lists, but that do not include $\Nil{*}$ in the end of
+the tuple type.
+Instead, it uses the predicate $adjust$ to replace any missing
+value with $\Nil$:
+\[
+\begin{array}{c}
+\mylabel{T-APPLY1}\\
+\dfrac{\env_{1} \vdash e:r_{1} \rightarrow r_{2}, \env_{2} \;\;\;
+       \env_{2} \vdash el:r_{3}, \env_{3} \;\;\;
+       r_{4} = adjust(r_{3}, r_{1}) \;\;\;
+       r_{4} \lesssim r_{1}}
+      {\env_{1} \vdash e(el)_{m}:r_{2}, \env_{3}}
+\end{array}
+\]
+
+We can use pattern matching to the define the recursive predicate
+\emph{adjust} as follows:
+\begin{align*}
+adjust(\Void, t{*}) & = \Nil \times \Void\\
+adjust(\Void, t \times s) & = \Nil \times adjust(\Void,s)\\
+adjust(t_{1} \times s_{1}, t_{2} \times s_{2}) & = t_{1} \times adjust(s1,s2)\\
+adjust(s_{1}, s_{2}) & = s_{1}
+\end{align*}
+
+We start with the rule that defines the most unusual feature of
+our type system: the refinment of table types.
+The rule \textsc{T-REFINE} allow us to add new keys to \emph{open}
+and \emph{unique} table types:
+\[
+\begin{array}{c}
+\mylabel{T-REFINE}\\
+\dfrac{\begin{array}{c}
+       \env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique}\\
+       \env_{1} \vdash c:t_{1}, \env_{2} \;\;\;
+       \not \exists i \in 1..n \; t_{1} \lesssim k_{i} \;\;\;
+       \Nil \not\lesssim t_{1} \;\;\;
+       t_{2} = close(t)
+       \end{array}}
+      {\env_{1} \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, t_{1}{:}t_{2}\}_{open|unique}]}
+\end{array}
+\]
+
+We restricted the refinement of table types to include only literal
+keys, because its purpose is to make it easier the construction of
+table types that represent records.
+
+We use the refinement of table types to handle global variables.
+In Lua, the assignment \texttt{v = v + 1} translates to
+\texttt{\string_ENV["v"] = \string_ENV["v"] + 1} when \texttt{v}
+is not a local variable, where \texttt{\string_ENV} is a table
+that stores the global environment.
+For this reason, Typed Lua treats accesses to global variables as field accesses
+to an open table in the top-level scope.
+For instance,
+\[
+\string_ENV[``x"] \; {<}\String{>} = ``foo" \;;\; \string_ENV[``y"] \; {<}\Integer{>} = 1
+\]
+uses field assignment to add fields $``x"$ and $``y"$ to $\string_ENV$.
+In all the examples we assume that $\string_ENV$ is in the environment and
+has type $\{\}_{open}$.
+Therefore, after these field assignments $\string_ENV$ has type
+$\{``x":\String, ``y":\Integer\}_{open}$.
+
+We do not allow the refinement of table types to change the type of
+a field that is already present in the table type.
+For instance,
+\[
+\string_ENV[``x"] \; {<}\String{>} = ``foo" \;;\; \string_ENV[``x"] \; {<}\Integer{>} = 1
+\]
+do not type check, as we are trying to add a field that already exists
+in $\string_ENV$.
+We could have used union types to allow the refinement of existing fields,
+but this could lead to fields that have a too general type, which could
+decrease the ammount of static type checking.
+This means that our type system does not allow the table type to change
+towards a type that is not a subtype of the previous type.
+In this example,
+\[
+\{``x":\String \cup \Integer\}_{open} \not\subtype \{``x":\String\}_{open}
+\]
+
+We also do not allow the refinement of table types to introduce
+\emph{open} or \emph{unique} table fields.
+For instance,
+\begin{center}
+\begin{tabular}{l}
+$\string_ENV[``x"] \; {<}\{\}_{unique}{>} = \{\}$
+\end{tabular}
+\end{center}
+refines the type of $\string_ENV$ from $\{\}_{open}$ to $\{``x":\{\}_{closed}\}_{open}$,
+though we are trying to refine to $\{``x":\{\}_{unique}\}_{open}$.
+We made this restriction because we still need to prove that the refinement
+of table types is safe.
 

### 2015-01-14 doc/thesis/system.tex
Moving the discussion about multiple assignments before the discussion about refinement of table types

@@
 \end{array}
+\]
+
+Using \emph{unique} table types to represent the type of the table
+constructor allows our type system to type check the previous example.
+More precisely,
+\[
+\{``x":1, ``y":2\}_{unique} \subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
 \]
@@
 
-Multiple assignments also appear in function and method applications,
-but our type system needs different typing rules for
-expression lists that appear in function and method applications,
-as we also want to catch arity mismatch.
-For instance, assuming that $x$ and $f$ are in the environment with
-types $\Integer \cup \Nil$ and
-$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void \rightarrow \Integer{*}$,
+Multiple assignments also appear in function applications, and the
+rule \textsc{T-APPLY1} uses the same behavior from multiple assignments
+to drop extra parameters and replace missing parameters:
+\[
+\begin{array}{c}
+\mylabel{T-APPLY1}\\
+\dfrac{\env_{1} \vdash e:r_{1} \rightarrow r_{2}, \env_{2} \;\;\;
+       \env_{2} \vdash el:r_{3}, \env_{3} \;\;\;
+       r_{3} \lesssim r_{1}}
+      {\env_{1} \vdash e(el)_{m}:r_{2}, \env_{3}}
+\end{array}
+\]
+
+Assuming that $x$ and $f$ are local variables that are in the environment
+with types $\Integer \cup \Nil$ and
+$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*} \rightarrow \Integer{*}$,
 respectively.
-We want that
+The program
 \[
@@
 \]
-type checks because
-\[
-``foo" \times \Nil \times \Nil \times \Void \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
-\]
-but we also want that
+type checks through the rule \textsc{T-APPLY1} because
+\[
+``foo" \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
+\]
+and the program
 \[
@@
 \]
-does not type check because
-\[
-``foo" \times 1 \times 2 \times 3 \times \Void \not\lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
-\]
-
-To do that, the rule \textsc{T-APPLY1} uses rules that type check
-expression lists, but that do not include $\Nil{*}$ in the end of
-the tuple type.
-Instead, it uses the predicate $adjust$ to replace any missing
-value with $\Nil$:
+also type checks through the rule \textsc{T-APPLY1} because
+\[
+``foo" \times 1 \times 2 \times 3 \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
+\]
+
+However, we also expect from a type system that it catches arity mismatch.
+For this reason, our type system also has different typing rules for
+expression lists that appear in function applications.
+These rules do not include $\Nil{*}$ when the tuple type does not end
+in a variadic type.
+The rule \textsc{T-APPLY1-STRICT} use these rules to type check
+expression lists, and it also uses the predicate $adjust$ to replace
+any missing value with $\Nil$:
 \[
@@
 
-We start with the rule that defines the most unusual feature of
-our type system: the refinment of table types.
+Now, assuming that $f$ has type
+$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void \rightarrow \Integer{*}$,
+the program
+\[
+x = f(``foo")_{m}
+\]
+type checks through the rule \textsc{T-APPLY1-STRICT} because
+\[
+``foo" \times \Nil \times \Nil \times \Void \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
+\]
+but the program
+\[
+x = f(``foo",1,2,3)_{m}
+\]
+does not type check through the rule \textsc{T-APPLY1-STRICT} because
+\[
+``foo" \times 1 \times 2 \times 3 \times \Void \not\lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
+\]
+
+The same happens with the rules that type check the \textbf{return}
+statement.
+The rule \textsc{T-RETURN} uses $\Nil{*}$ to replace eventual missing
+values, while the rule \textsc{T-RETURN-STRICT} uses the predicate
+\emph{adjust} to replace values that are certainly missing:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-RETURN}\\
+\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
+       \env_{2}(\ret) = r_{2} \;\;\;
+       r_{1} \lesssim r_{2}}
+      {\env_{1} \vdash \mathbf{return} \; el:\env_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-RETURN-STRICT}\\
+\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
+       \env_{2}(\ret) = r_{2} \;\;\;
+       r_{3} = adjust(r_{1}, r_{2}) \;\;\;
+       r_{3} \lesssim r_{2}}
+      {\env_{1} \vdash \mathbf{return} \; el:\env_{2}}
+\end{array}
+\end{array}
+\]
+
+Now, we move the discussion to the rules that type check the
+different table constructors that our abstract syntax includes,
+as they will appear in the next examples:
+
+The simplest form of the table constructor is the empty table $\{\}$.
+Its rule is straightforward:
+\[
+\begin{array}{c}
+\mylabel{T-CONSTRUCTOR1}\\
+\env \vdash \{\}:\{\}_{unique}, \env
+\end{array}
+\]
+
+Our abstract syntax reduces the more complex uses of the table
+constructor into three forms: $\{\;\vec{p}\;\}$, $\{\;{...}_{m}\;\}$,
+and $\{\;\vec{p},{...}_{m}\;\}$.
+The first one uses a list of table fields $[e_{1}] = e_{2}$,
+the second one uses just the vararg expression, and the third one uses both.
+We did not include in the abstract syntax a table constructor
+that uses a list of expressions because we can write it using the first form.
+For instance, we can write the table constructor
+$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to express $\{ ``x", ``y", ``z" \}$.
+We chose to not include the latter form because its type checking rule is similar
+to the type checking rule of the first form, and just controls the type of
+the omitted indices.
+For the same reason we did not include table constructors with
+multiple expressions, which can return a tuple type that does not end
+with a variadic type.
+We did include the vararg expression in the last position to show how
+our type system can type check a table type that has a record part and
+also an array part.
+
+The rule \textsc{T-CONSTRUCTOR2} type checks a table type of the first form:
+\[
+\begin{array}{c}
+\mylabel{T-CONSTRUCTOR2}\\
+\dfrac{\env_{i} \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
+       n = |\;\vec{p}\;| \;\;\;
+       t = wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique})}
+      {\env_{1} \vdash \{\;\vec{p}\;\}:t, \env_{n+1}}
+\end{array}
+\]
+
+The rule \textsc{T-CONSTRUCTOR2} type checks each table field $p_{i}$,
+uses the type of the table fields to build the table type, and
+uses the predicate \emph{wf} to check whether the table type is
+well formed.
+Formally, a table type is well-formed if it obeys the following rule:
+\[
+\forall i \not\exists j \; i \not= j \wedge k_{i} \lesssim k_{j}
+\]
+
+Well-formed table types avoid ambiguous table types.
+For instance, this rule detects that the table type
+$\{1:\Number, \Integer:\String, \Any:\Boolean\}$ is ambiguous,
+because the type of the value stored by key $1$ can be
+$\Number$, $\String$, or $\Boolean$, as $1 \lesssim 1$,
+$1 \lesssim \Integer$, and $1 \lesssim \Any$.
+Moreover, the type of the value stored by a key of type $\Integer$,
+which is not the literal type $1$, can be $\Number$ or $\Boolean$,
+as $\Integer \lesssim \Integer$, and $\Integer \lesssim \Any$.
+
+The type checking of the table fields would be straightforward
+if our table types allowed any first-level type in the keys.
+The rules \textsc{T-FIELD1}, \textsc{T-FIELD2}, and \textsc{T-FIELD3}
+check if the type of the key is a subtype of $\Boolean$, $\Number$,
+or $\String$, respectively:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-FIELD1}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+       t_{1} \subtype \Boolean}
+      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-FIELD2}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+       t_{1} \subtype \Number}
+      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-FIELD3}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+       t_{1} \subtype \String}
+      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
+\end{array}
+\end{array}
+\]
+
+The rule \textsc{T-FIELD4} is a fallback for the previous rules:
+\[
+\begin{array}{c}
+\mylabel{T-FIELD4}\\
+\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t_{2}, \env_{3}}
+      {\env_{1} \vdash [e_{1}] = e_{2}: \Value{:}close(t_{2}), \env_{3}}
+\end{array}
+\]
+
+The rule \textsc{T-CONSTRUCTOR2} along with the rules for table fields
+assign the type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ to
+the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$.
+
+Typed Lua handles arrays as hashes that map integers to some type $t$.
+In Lua, programmers often use the vararg expression to initialize arrays.
+The rules \textsc{T-CONSTRUCTOR3} and \textsc{T-CONSTRUCTOR4} define
+the behavior of the vararg expression for this case:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-CONSTRUCTOR3}\\
+\dfrac{\env_{1}({...}) = t}
+      {\env_{1} \vdash \{{...}_{m}\}:\{\Integer{:}t \cup \Nil\}_{unique}, \env_{1}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-CONSTRUCTOR4}\\
+\dfrac{\begin{array}{c}
+       \env_{i} \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
+       n = |\;\vec{p}\;| \;\;\;
+       \env_{i+1}({...}) = t_{v}\\
+       wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, \Integer{:}t_{v} \cup \Nil\}_{unique}) = t_{t}
+       \end{array}}
+      {\env_{1} \vdash \{\;\vec{p},\;{...}_{m}\;\}:t_{t}, \env_{n+1}}
+\end{array}
+\end{array}
+\]
+
+The rule \textsc{T-CONSTRUCTOR3} type checks the case where we use
+only the vararg expression inside a table constructor to initialize
+an array.
+If we assume that the $...$ is in the environment and has type $\String$,
+the rule \textsc{T-CONSTRUCTOR3} gives the type $\{\Integer:\String \cup \Nil\}_{unique}$
+to the table constructor $\{{...}_{m}\}$.
+
+The rule \textsc{T-CONSTRUCTOR4} type checks the case where we use
+the table constructor as a record that includes an array part.
+If we assume that the $...$ is in the environment and has type $\String$,
+the rule \textsc{T-CONSTRUCTOR4} gives the type
+$\{``x":``foo", ``y":\mathbf{true}, \Integer:\String \cup \Nil\}_{unique}$
+to the table constructor $\{[``x"] = ``foo", [``y"] = \mathbf{true}, {...}_{m}\}$.
+As another example, the rule \textsc{T-CONSTRUCTOR4} fails to type check
+the table constructor $\{ [1] = ``foo", [10] = \mathbf{true}, {...}_{m} \}$,
+because the keys $1$ and $10$ overlap the array part.
+
+After discussing the typing rules of the table constructor,
+we start the discussion of the rules that define the most
+unusual feature of our type system: the refinement of table types.
 The rule \textsc{T-REFINE} allow us to add new keys to \emph{open}
@@
        \env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique}\\
-       \env_{1} \vdash c:t_{1}, \env_{2} \;\;\;
+       \env_{1} \vdash c:k_{n+1}, \env_{2} \;\;\;
        \not \exists i \in 1..n \; t_{1} \lesssim k_{i} \;\;\;
-       \Nil \not\lesssim t_{1} \;\;\;
-       t_{2} = close(t)
+       v_{n+1} = close(t)
        \end{array}}
-      {\env_{1} \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, t_{1}{:}t_{2}\}_{open|unique}]}
+      {\env_{1} \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, k_{n+1}{:}v_{n+1}\}_{open|unique}]}
 \end{array}
@@
 $``foo" \times 1 \times \Nil{*} \lesssim \String \times \Integer \times \Value{*}$.
-The type of the expression list includes the type $\Nil{*}$ in the end
-because our type system has typing rules to add this type,
-when the type of the expression list does not end with a variadic type.
-Our type system has this behavior to use $\Nil$ in the place of missing
-values, like Lua does.
-This example uses the rule \textsc{T-EXPLIST2}, which is really close
-to the rule \textsc{T-LHSLIST}:
-\[
-\begin{array}{c}
-\mylabel{T-EXPLIST2}\\
-\dfrac{\env \vdash e_{k}:t_{k}, \env_{k} \;\;\;
-       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
-       n = |\;\vec{e}\;|}
-      {\env \vdash \vec{e}:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{m}}
-\end{array}
-\]
 
@@
 \]
-
-Typed Lua also keeps track whether it is safe or not safe to
-change the type of a table, but before discussing this subject,
-we will discuss the different table constructors that our
-abstract syntax includes, as they will appear in the next
-examples.
-
-Using \emph{unique} table types to represent the type of the table
-constructor allows our type system to type check the previous example.
-More precisely,
-\[
-\{``x":1, ``y":2\}_{unique} \subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
-\]
-
-Even though \emph{unique} table types overcome our limitation.
-The covariance on mutable fields could allow programs that create unsafe
-aliases to pass the type checker.
-For instance,
-\begin{verbatim}
-    local t1 = { foo = 5 }
-    local t2:{"foo":integer?} = t1
-    t2.foo = nil
-\end{verbatim}
-does not type check, as the last line is erasing the field \texttt{foo}
-from \texttt{t1}.
-This program would type checks as it is, because our implementation uses
-local type inference to assign static types to unannotated locals.
-In this example, the type of \texttt{t1} would be subtype of the type of \texttt{t2}.
-More precisely,
-\[
-\{``foo":\Integer\}_{unique} \subtype \{``foo":\Integer \cup \Nil\}_{closed}
-\]
-
-In the next section we will show in more detail that this example
-does not type check, because our type system promotes an alias
-expression from \emph{unique} to \emph{closed}, and promotes the
-aliased variable from \emph{unique} to \emph{open}.
-This means that the previous example does not type check because
-the type of the expression \texttt{t1} is not subtype of the type
-of the variable \texttt{t2}.
-More precisely,
-\[
-\{``foo":\Integer\}_{closed} \not\subtype \{``foo":\Integer \cup \Nil\}_{closed}
-\]
-
-Thus, we introduced \emph{open} table types to handle the aliasing of table types.
-
-The simplest form of the table constructor is the empty table $\{\}$.
-Its inference rule is straightforward:
-\[
-\begin{array}{c}
-\mylabel{T-CONSTRUCTOR1}\\
-\env \vdash \{\}:\{\}_{unique}, \env
-\end{array}
-\]
-
-Our abstract syntax reduces the more complex uses of the table
-constructor into three forms: $\{\;\vec{p}\;\}$, $\{\;{...}_{m}\;\}$,
-and $\{\;\vec{p},{...}_{m}\;\}$.
-The first one uses a list of expression pairs $[e_{1}] = e_{2}$,
-the second one uses just the vararg expression, and the third one uses both.
-We did not include in the abstract syntax a table constructor
-that uses a list of expressions because we can write it using the first form.
-For instance, we can write the table constructor
-$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to express $\{ ``x", ``y", ``z" \}$.
-We chose to not include this form because its inference rule is similar
-to the inference rule of the first form, and just controls the type of
-the omitted indices.
-For the same reason we did not include table constructors with
-multiple expressions, which can return a tuple type that does not end
-with a variadic type.
-We did include the vararg expression in the last position to show how
-our type system can infer a table type that has a record part and
-also an array part.
-
-The inference rule \textsc{T-CONSTRUCTOR2} infers a table type for
-the first form:
-\[
-\begin{array}{c}
-\mylabel{T-CONSTRUCTOR2}\\
-\dfrac{\env_{i} \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
-       n = |\;\vec{p}\;| \;\;\;
-       t = wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique})}
-      {\env_{1} \vdash \{\;\vec{p}\;\}:t, \env_{n+1}}
-\end{array}
-\]
-
-This inference rule infers the type of each pair, uses these types
-to build a table type, and uses the predicate \emph{wf} to check
-whether the table type is well formed.
-Formally, a table type is well-formed if it obeys the following rule:
-\[
-\forall i \not\exists j \; i \not= j \wedge k_{i} \lesssim k_{j}
-\]
-
-Well-formed table types avoid ambiguous table types.
-For instance, this rule detects that the table type
-$\{1:\Number, \Integer:\String, \Any:\Boolean\}$ is ambiguous,
-because the type of the value stored by key $1$ can be
-$\Number$, $\String$, or $\Boolean$, as $1 \lesssim 1$,
-$1 \lesssim \Integer$, and $1 \lesssim \Any$.
-Moreover, the type of the value stored by a key of type $\Integer$,
-which is not the literal type $1$, can be $\Number$ or $\Boolean$,
-as $\Integer \lesssim \Integer$, and $\Integer \lesssim \Any$.
-
-The inference rules for expression pairs would be straightforward
-if our table types allowed any type in the keys.
-The rules \textsc{T-PAIR1}, \textsc{T-PAIR2}, and \textsc{T-PAIR3}
-check if the type of the key is a subtype of $\Boolean$, $\Number$,
-or $\String$, respectively:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-PAIR1}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
-       t_{1} \subtype \Boolean}
-      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-PAIR2}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
-       t_{1} \subtype \Number}
-      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-PAIR3}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
-       t_{1} \subtype \String}
-      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
-\end{array}
-\end{array}
-\]
-
-The rule \textsc{T-PAIR4} is a fallback for the previous rules,
-but it works only if the type of the key does not include $\Nil$:
-\[
-\begin{array}{c}
-\mylabel{T-PAIR4}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
-       \Nil \not\subtype t_{1}}
-      {\env_{1} \vdash [e_{1}] = e_{2}: \Any{:}close(t_{2}), \env_{3}}
-\end{array}
-\]
-
-The rule \textsc{T-CONSTRUCTOR2} along with the rules for expression
-pairs allow Typed Lua to type check the following example:
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; a:\{\String:\Integer \cup \Nil\}_{closed} = \{ [``x"] = 1, [``y"] = 2 \} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$\mathbf{local} \; b:\Integer \cup \Nil = a[``x"] \; \mathbf{in} \; b = a[``z"]$}
-\end{tabular}
-\end{center}
-
-This example type checks because Typed Lua infers the table type
-$\{``x":1, ``y":2\}_{unique}$, which is subtype of
-$\{\String:\Integer \cup \Nil\}_{closed}$.
-Even though the key $``z"$ does not exist in the table $a$,
-the two table indexations type check with type $\Integer \cup \Nil$,
-which is subtype of the type of $b$.
-The rule \textsc{T-INDEX1} describes this behavior:
-\[
-\begin{array}{c}
-\mylabel{T-INDEX1}\\
-\dfrac{\env_{1} \vdash e_{1}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t, \env_{3} \;\;\;
-       \exists i \in 1{..}n \; t \lesssim k_{i}}
-      {\env_{1} \vdash e_{1}[e_{2}]:v_{i}, \env_{3}}
-\end{array}
-\]
-
-Typed Lua handles arrays as hashes that maps integers to some type $t$.
-In Lua, programmers often use the vararg expression to initialize arrays.
-The rules \textsc{T-CONSTRUCTOR3} and \textsc{T-CONSTRUCTOR4} define
-the behavior of the vararg expression for this case:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-CONSTRUCTOR3}\\
-\dfrac{\env_{1}({...}) = t}
-      {\env_{1} \vdash \{{...}_{m}\}:\{\Integer{:}t \cup \Nil\}_{unique}, \env_{1}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-CONSTRUCTOR4}\\
-\dfrac{\begin{array}{c}
-       \env_{i} \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
-       n = |\;\vec{p}\;| \;\;\;
-       \env_{i+1}({...}) = t_{v}\\
-       wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, \Integer{:}t_{v} \cup \Nil\}_{unique}) = t_{t}
-       \end{array}}
-      {\env_{1} \vdash \{\;\vec{p},\;{...}_{m}\;\}:t_{t}, \env_{n+1}}
-\end{array}
-\end{array}
-\]
-
-The rule \textsc{T-CONSTRUCTOR3} type checks the case where we use
-only the vararg expression inside a table constructor to initialize
-an array.
-If we assume that the $...$ is in the environment and has type $\String$,
-the following example type checks through this rule:
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; a:\{\Integer:\String \cup \Nil\}_{closed} = \{ {...}_{m} \} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$\mathbf{local} \; b:\String \cup \Nil = t[1] \; \mathbf{in} \; b = t[5]$}
-\end{tabular}
-\end{center}
-
-The rule \textsc{T-CONSTRUCTOR4} type checks the case where we use
-the table constructor as a record that includes an array part.
-If we assume that the $...$ is in the environment and has type $\String$,
-the following example type checks through this rule:
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; a:\{``x":\Integer, ``y":\Boolean, \Integer:\String \cup \Nil\}_{closed} =$}\\
-& \multicolumn{1}{l}{$\{ [``x"] = ``foo", [``y"] = \mathbf{true}, {...}_{m} \}$}\\
-\multicolumn{2}{l}{$\mathbf{in} \; a[``x"] = ``bar"$}
-\end{tabular}
-\end{center}
-but the next example does not type check because the record part
-includes fields that the key types are subtype of the array part:
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; a:\{1:\String, 10:\Boolean, \Integer:\String \cup \Nil\}_{closed} =$}\\
-& \multicolumn{1}{l}{$\{ [1] = ``foo", [10] = \mathbf{true}, {...}_{m} \}$}\\
-\multicolumn{2}{l}{$\mathbf{in} \; a[``x"] = ``bar"$}
-\end{tabular}
-\end{center}
-
-Making the inference rules of the table constructor infer an \emph{unique}
-table type also allow us to use them for initializing record types that
-have optional fields.
-We can include the optional field in the initialization
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; a:\{``x":\String, ``y":\String \cup \Nil, ``z":\String\}_{closed} =$}\\
-& \multicolumn{1}{l}{$\{ [``x"] = ``foo", [``y"] = ``bar", [``z"] = ``baz" \}$}\\
-\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
-\end{tabular}
-\end{center}
-and we can also omit it in the initialization
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; a:\{``x":\String, ``y":\String \cup \Nil, ``z":\String\}_{closed} =$}\\
-& \multicolumn{1}{l}{$\{ [``x"] = ``foo", [``z"] = ``baz" \}$}\\
-\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
-\end{tabular}
-\end{center}
-
-After discussing the inference rules of the table constructor,
-we can get back to the discussion about the refinement of table types.
 
@@
 type of a table.
-%To do that, our type system always promotes an \emph{open}
-%table to \emph{closed} and an \emph{unique} table to \emph{open}
-%when they appear as expressions, or when they enter in a scope that is
-%not its declaration scope.
-%The order matters, as going from \emph{unique} to \emph{open}
-%before going from \emph{open} to \emph{closed} closes \emph{unique} table types.
-In the last section we mentioned that aliasing \emph{unique} table
-types is unsafe.
-To avoid unsafe aliasing of \emph{unique} table types, the rule \textsc{T-IDREAD1}
-always promotes them to \emph{closed}, and updates the variable in the
-environment to \emph{open}:
-\[
-\begin{array}{c}
-\mylabel{T-IDREAD1}\\
-\dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique}}
-      {\env_{1} \vdash n_{e}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed}, \env_{1}[n \mapsto \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{open}]}
-\end{array}
-\]
-
-This is the rule that makes the unsafe example from last section to
-not type check:
+For instance,
 \begin{center}
@@
 \end{center}
-
-This examples does not type check because aliasing $a$ produces the
-type $\{``foo":\Integer\}_{closed}$ that is not subtype of
+does not type check because aliasing $a$ produces the type
+$\{``foo":\Integer\}_{closed}$ that is not subtype of
 $\{``foo":\Integer \cup \Nil\}_{closed}$, the type of $b$.
-
-However, sometimes we want to use \emph{unique} table types to initialize
-\emph{closed} table types, as they allow us to initialize optional fields.
+The rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this behavior:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-IDREAD1}\\
+\dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique}}
+      {\env_{1} \vdash n_{e}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed}, \env_{1}[n \mapsto \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{open}]}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-IDREAD2}\\
+\dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{open}}
+      {\env_{1} \vdash n_{e}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed}, \env_{1}}
+\end{array}
+\end{array}
+\]
+
+Aliasing \emph{unique} and \emph{open} table types can be unsafe.
+For this reason, the rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2}
+promote the type of an alias expression from \emph{unique} or \emph{open}
+to \emph{closed}.
+The rule \textsc{T-IDREAD1} also promotes the aliased variable from
+\emph{unique} to \emph{open}, as \emph{unique} table types only
+denote the type of table types that have no alias.
+
+As another example,
+\begin{center}
+\begin{tabular}{lll}
+\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
+& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = a \; \mathbf{in}$}\\
+& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
+& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
+\end{tabular}
+\end{center}
+does not type check because we cannot add the field $``x"$ to $b$,
+as its type is \emph{closed}, and thus does not allow changing the
+value that is stored in the field $``x"$ of local $a$.
+
+We also have different rules for type checking table indexing that
+appear as expressions to avoid promoting an \emph{unique} table to
+\emph{open} in these operations:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-INDEXREAD1}\\
+\dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\} \;\;\;
+       \env_{1} \vdash e_{2}:t, \env_{2} \;\;\;
+       \exists i \in 1{..}n \; t \lesssim k_{i}}
+      {\env_{1} \vdash n_{e}[e_{2}]:v_{i}, \env_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-INDEXREAD1}\\
+\dfrac{\env_{1} \vdash e_{1}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}, \env_{2} \;\;\;
+       \env_{2} \vdash e_{2}:t, \env_{3} \;\;\;
+       \exists i \in 1{..}n \; t \lesssim k_{i}}
+      {\env_{1} \vdash e_{1}[e_{2}]:v_{i}, \env_{3}}
+\end{array}
+\end{array}
+\]
+
+Sometimes we want to use a local variable that holds an \emph{unique}
+table type to initialize a \emph{closed} table type, as \emph{unique}
+table types allow us to initialize optional fields.
 To do that, Typed Lua includes a cast expression that allow us
@@
 \dfrac{\env_{1}(n) \subtype t}
-      {\env_{1} \vdash {<}t{>} \; n:t, \env_{1}[n \mapsto t]}
+      {\env_{1} \vdash {<}t{>} \; n:close(t), \env_{1}[n \mapsto t]}
 \end{array}
@@
 the type of $a$ from $\{``x":\String, ``y":\String\}_{unique}$ to
-$\{``x":\String, ``y":\String \cup \Nil\}_{open}$, which is subtype of
+$\{``x":\String, ``y":\String \cup \Nil\}_{open}$, and results in
 $\{``x":\String, ``y":\String \cup \Nil\}_{closed}$.
@@
 
-We should also keep tracking of \emph{open} locals, as they can
-also be unsafe like in the next example:
-\begin{center}
-\begin{tabular}{lll}
-\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
-& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = a \; \mathbf{in}$}\\
-& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
-& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
-\end{tabular}
-\end{center}
-
-We cannot add the field $``x"$ to $b$ because its type is \emph{closed},
-and thus do not allow changing the value that is stored in the field $``x"$
-of local $a$.
-
-We also need to make sure to close all \emph{open} table types before we
-type check another scope.
+We also need to make sure to close all \emph{unique} and \emph{open}
+table types before we type check another scope.
 The rule \textsc{T-FUNCTION3} shows this case:
@@
 
-Multiple assignments and multiple return values can also appear in
-function applications, as the following example shows:
-\begin{center}
-\begin{tabular}{llll}
-\multicolumn{4}{l}{$\mathbf{local} \; m:\Void \rightarrow \Integer \times \String =$} \\
-& \multicolumn{3}{l}{$\mathbf{fun} \; ():\Integer \times \String$} \\
-& & \multicolumn{2}{l}{$\mathbf{return} \; 2, ``foo"$} \\
-\multicolumn{4}{l}{$\mathbf{in}$} \\
-& \multicolumn{3}{l}{$\mathbf{local} \; s:\Integer \times \Integer \rightarrow \Integer =$} \\
-& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer, y:\Integer):\Integer$} \\
-& & & \multicolumn{1}{l}{$\mathbf{return} \; x + y$} \\
-& \multicolumn{3}{l}{$\mathbf{in} \; s(m()_{e}, m()_{m})_{s}$}
-\end{tabular}
-\end{center}
-
-This example does not type check because we are trying to call
-$s$ with $\Integer \times \Integer \times \String$,
-but its input parameter has type $\Integer \times \Integer$.
-Lua drops extra values in function calls, and Typed Lua can also
-have this behavior.
-If we change the input type of $s$ to $\Integer \times \Integer \times \Value{*}$,
-this example type checks.
-Typed Lua also has to adjust function calls to use $\Nil$ in the
-place of missing parameters, otherwise it would not be able to
-type check function calls that have optional parameters.
-
-\textit{/* I guess we should have different typing rules for strict and default. */}
-
 Typed Lua includes four typing rules to handle the $\mathbf{or}$
@@
 
+\[
+\begin{array}{c}
+\mylabel{T-EXPLIST6}\\
+\dfrac{\env \vdash me:t_{1} \times ... \times t_{n} \times \Void \sqcup t_{1}' \times ... \times t_{n}' \times \Void, \env_{1}}
+      {\env \vdash me:\pi_{1} \times ... \times \pi_{n} \times \Nil{*}, \env_{1}}
+\end{array}
+\]
+
 \begin{center}

### 2015-01-15 doc/thesis/system.tex
Fixing the text about tables subtyping

@@
 The following subtyping rule defines the subtyping relation among
-\emph{unique} table types and other table types:
+\emph{unique} table types and \emph{closed} table types:
 \[
@@
 \mylabel{S-TABLE2}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
-       \senv \vdash v_{i} \subtype_{u} v_{i}' \;\;\;
-       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
+\dfrac{\begin{array}{c}
+       \forall i \in 1..m \; \forall j \in 1..n \;
+       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash v_{i} \subtype_{u} v_{j}' \\
+       \forall j \in 1..n \; \not\exists i \in 1..m \;
+       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
+       \end{array}}
       {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
-                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open|unique}}
-\end{array}
-\]
-
-Unlike the rule \textsc{S-TABLE1}, the rule \textsc{S-TABLE2} allows
-covariant keys and do not allow width subtyping.
-We allow this behavior to \emph{unique} table types because we also
-want to use them as a way to join table fields.
-For instance, we may want to use the table constructor to initialize
+                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
+\end{array}
+\]
+
+The rule \textsc{S-TABLE2} allows width subtyping and covariant keys.
+It allows covariant keys because we also want to use \emph{unique}
+table types as a way to join table fields that inhabit \emph{closed} table types.
+For instance, we want to use the table constructor to initialize
 a variable with a table type that describes a hash.
@@
 
-We also do not allow width and depth subtyping between \emph{open}
-table types or between \emph{open} and \emph{closed} table types:
+Even though we allow width subtyping between \emph{unique} and \emph{closed}
+table types, we do not allow it among \emph{unique} and \emph{open} or
+\emph{unique} table types because it would violate our definition of
+\emph{unique} and \emph{open} table types:
 \[
@@
 \mylabel{S-TABLE3}\\
-\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
-       \senv \vdash v_{i} \subtype_{c} v_{i}' \;\;\;
-       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
+\dfrac{\begin{array}{c}
+       \forall i \in 1..m \\
+       \exists j \in 1..n \;
+       \senv \vdash k_{i} \subtype k_{j}' \land \senv \vdash v_{i} \subtype_{u} v_{j}' \\
+       \forall j \in 1..n \; \not\exists i \in 1..m \;
+       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
+       \end{array}}
+      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
+                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{open|unique}}
+\end{array}
+\]
+
+The rules that handle subtyping among \emph{open} table types and
+\emph{open} or \emph{closed} table types are similar to the rules
+that handle subtyping among \emph{unique} table types and other table types:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-TABLE4}\\
+\dfrac{\begin{array}{c}
+       \forall i \in 1..m \; \forall j \in 1..n \;
+       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash v_{i} \subtype_{c} v_{j}' \\
+       \forall j \in 1..n \; \not\exists i \in 1..m \;
+       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
+       \end{array}}
       {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
-                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open}}
-\end{array}
-\]
-
-The rule \textsc{S-TABLE3} uses the same rules from
-\textsc{S-TABLE1} for handling the type of the values of each key, and
-it uses the same rules from \textsc{S-TABLE2} to allow joining fields plus
-omitting optional fields.
+                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-TABLE5}\\
+\dfrac{\begin{array}{c}
+       \forall i \in 1..m \\
+       \exists j \in 1..n \;
+       \senv \vdash k_{i} \subtype k_{j}' \land \senv \vdash v_{i} \subtype_{c} v_{j}' \\
+       \forall j \in 1..n \; \not\exists i \in 1..m \;
+       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
+       \end{array}}
+      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
+                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{open}}
+\end{array}
+\end{array}
+\]
+
+The rule \textsc{S-TABLE4} allows width subtyping, while
+the rule \textsc{S-TABLE5} does not allow width subtyping.
+Both rules allow joining fields plus omitting optional fields.
+Both rules use $\subtype_{c}$ to allow depth on $\Const$ fields only.
 
@@
 $\{``foo":\Integer \cup \Nil\}_{closed}$, the type of $b$.
-The rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this behavior:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-IDREAD1}\\
-\dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique}}
-      {\env_{1} \vdash n_{e}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed}, \env_{1}[n \mapsto \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{open}]}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-IDREAD2}\\
-\dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{open}}
-      {\env_{1} \vdash n_{e}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed}, \env_{1}}
-\end{array}
+The rule \textsc{T-IDREAD} defines this behavior:
+\[
+\begin{array}{c}
+\mylabel{T-IDREAD}\\
+\dfrac{\env_{1}(n) = t}
+      {\env_{1} \vdash n_{e}:close(t), \env_{1}[n \mapsto open(t)]}
 \end{array}
@@
 Aliasing \emph{unique} and \emph{open} table types can be unsafe.
-For this reason, the rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2}
-promote the type of an alias expression from \emph{unique} or \emph{open}
-to \emph{closed}.
-The rule \textsc{T-IDREAD1} also promotes the aliased variable from
+For this reason, the rule \textsc{T-IDREAD} promote the type of an
+alias expression from \emph{unique} or \emph{open} to \emph{closed},
+and it also promotes the aliased variable from
 \emph{unique} to \emph{open}, as \emph{unique} table types only
@@
 
-We also have different rules for type checking table indexing that
-appear as expressions to avoid promoting an \emph{unique} table to
-\emph{open} in these operations:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-INDEXREAD1}\\
+We also need to close \emph{unique} and \emph{open} tables that
+appear in the left-hand side of assignments.
+For instance,
+\begin{center}
+\begin{tabular}{lll}
+\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
+& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = \{\} \; \mathbf{in}$}\\
+& & \multicolumn{1}{l}{$b = a;$}\\
+& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
+& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
+\end{tabular}
+\end{center}
+does not type check because we cannot add the field $``x"$ to $b$,
+as its type is \emph{closed}, and thus does not allow changing the
+value that is stored in the field $``x"$ of local $a$.
+The rule \textsc{T-IDWRITE} defines this behavior:
+\[
+\begin{array}{c}
+\mylabel{T-IDWRITE}\\
+\dfrac{\env_{1}(n) = t}
+      {\env_{1} \vdash n_{l}:close(t), \env_{1}[n \mapsto close(t)]}
+\end{array}
+\]
+
+We also have different rules for type checking table indexing to avoid
+promoting table types in these operations:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-INDEX1}\\
 \dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\} \;\;\;
@@
        \exists i \in 1{..}n \; t \lesssim k_{i}}
-      {\env_{1} \vdash n_{e}[e_{2}]:v_{i}, \env_{2}}
+      {\env_{1} \vdash n[e_{2}]:v_{i}, \env_{2}}
 \end{array}
@@
 \begin{array}{c}
-\mylabel{T-INDEXREAD1}\\
+\mylabel{T-INDEX2}\\
 \dfrac{\env_{1} \vdash e_{1}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}, \env_{2} \;\;\;
@@
 \mylabel{T-CAST}\\
-\dfrac{\env_{1}(n) \subtype t}
-      {\env_{1} \vdash {<}t{>} \; n:close(t), \env_{1}[n \mapsto t]}
+\dfrac{\env_{1}(n) \subtype t \;\;\;
+       \env_{1}[n \mapsto t] \vdash n_{e}:t_{1}, \env_{2}}
+      {\env_{1} \vdash {<}t{>} \; n:t_{1}, \env_{2}}
 \end{array}

### 2015-01-20 doc/thesis/typedlua.tex
Minor improvements to the text about types and subtyping
Improving the text about optional type annotations, function types, and union types

@@
 Typed Lua is an optional type system for Lua, and its main goal is to
-allow programmers to combine static and dynamic typing in Lua.
-In practice, Typed Lua extends the syntax of Lua to introduce
-optional type annotations.
-In addition, Typed Lua performs local type inference \citep{pierce2000lti}
+provide static type checking for Lua.
+To do that, Typed Lua extends the syntax of Lua 5.3 to introduce
+optional type annotations, and performs local type inference \citep{pierce2000lti}
 to detect more precise types for unannotated expressions.
@@
 
-An important feature of Typed Lua is its backwards compatibility with Lua,
-that is, unannotated Typed Lua code is valid Lua code.
-For this reason, the syntactic extensions introduced by Typed Lua
-include new keywords that are not reserved words.
+Another goal of Typed Lua is to be backwards compatible with Lua.
+This means that any Lua code is valid Typed Lua code.
+To be backwards compatible with Lua, the syntactic extensions introduced
+by Typed Lua does not include new reserved words.
 Appendix \ref{app:syntax} presents the complete syntax of Typed Lua
 in extended BNF.
-
-Another important feature of Typed Lua is its interface between
-statically and dynamically typed modules.
-This means that the programmer can enjoy some of the benefits of
-static types even without converting existing Lua modules to Typed Lua,
-as a dynamically typed module can export a statically typed interface,
-and statically typed users of the module have their use of the module
-checked by the compiler.
 
@@
 typed code does not violate statically typed code during run-time.
-A good reason for not inserting run-time checks at this moment is that they
-can decrease run-time performance \citep{allende2013cis}, and we believe that
-a careful evaluation of run-time checks should be done before inserting them
-on the type system.
+We did not insert run-time checks at this moment because they can decrease
+run-time performance \citep{allende2013cis}.
+We believe that a careful evaluation of run-time checks should be done before
+inserting them on the type system.
+However, this evaluation is out of the scope of this work.
 
 Unlike Dart \citep{dart} and TypeScript \citep{typescript},
-we are designing Typed Lua aiming soundness to make it possible to switch
-Typed Lua from optional typing to gradual typing in the future,
-as a sound type system is a prerequisite to make Typed Lua insert
-run-time checks after static type checking.
+we are designing Typed Lua aiming soundness to make it possible to
+switch Typed Lua from optional typing to gradual typing in the future.
+As we mentioned in the previous chapter,
+a sound type system is a prerequisite to insert run-time checks after
+static type checking, because a sound type system ensures that
+statically typed code will not throw type errors during run-time.
 
@@
 they relate to Lua.
-This examples shall give an informal overview of our optional type system.
-We use typing rules to present the formalization of our optional type system
-in Chapter \ref{chap:system}.
-It worth mentioning that we used our implementation to produce and
-test all the examples that we present along this section.
+These examples shall give an informal overview of our optional type system.
+We will use typing rules to present the formalization of our optional type
+system in the next chapter.
+All the examples that we present along this chapter run in our Typed Lua compiler.
 
@@
 Typed Lua also includes a syntactical extension that programmers can use
-to define the types of \emph{userdata}, and uses it to define the type \emph{thread}. 
-In this section, we present the Typed Lua types that may appear on annotations.
+to define the types of \emph{userdata}.
+We use this syntactical extension to define the type \emph{thread}. 
+In this section we present the Typed Lua types that may appear on annotations.
 We explain all Typed Lua types and syntactical extensions along this chapter.
@@
 \textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
-\textit{tabletypebody} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type} \; | \;
+\textit{tabletypebody} ::= & \;\; \textit{hashtype} \; | \;
   \textit{recordtype}\\
+\textit{hashtype} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type}\\
 \textit{keytype} ::= & \;\; \textit{basetype} \; | \;
-  \textbf{any}\\
+  \textbf{value}\\
 \textit{recordtype} ::= & \;\; \textit{recordfield} \; \{\texttt{`,'} \; \textit{recordfield}\} \; [\texttt{`,'} \; \textit{type}]\\
@@
 
-Figure \ref{fig:types} presents the concrete syntax of the
+Figure \ref{fig:types} presents the concrete syntax of
 Typed Lua types in extended BNF.
@@
 Any first-level type is a subtype of \texttt{value}.
-The rest of the subtyping relationship is standard:
-union types are supertypes of their parts,
-\texttt{boolean}, \texttt{integer}, \texttt{number}, and \texttt{string}
-are supertypes of their respective literal types,
-\texttt{integer} is subtype of \texttt{number},
-function types are related by contravariance on the input
-and covariance on the output, table types have width subtyping,
-with depth subtyping on \texttt{const} fields, tuple and vararg types
-are covariant.
+Union types are supertypes of their parts.
+The base types \texttt{boolean}, \texttt{integer}, \texttt{number},
+and \texttt{string} are supertypes of their respective literal types.
+The base type \texttt{integer} is subtype of \texttt{number}.
+Function types are related by contravariance on the input
+and covariance on the output.
+Table types have width subtyping, with depth subtyping on
+\texttt{const} fields.
+Tuple and vararg types are covariant.
+Unions of tuple types are also supertypes of their parts.
 
@@
 dynamic type \texttt{any} and other types.
-The dynamic type \texttt{any} is a subtype of \texttt{value}, but neither
-a supertype nor a subtype of any other type.
+The dynamic type \texttt{any} is a subtype of \texttt{value}, but it is
+neither a supertype nor a subtype of any other type.
 Our consistent-subtyping relationship follows the standards defined
@@
 In practice, we can pass a value of the dynamic type anytime we want
-a value of some other type, and can pass any value where a
+a value of some other type, and we can pass any value where a
 value of the dynamic type is expected, but the compiler tracks these
@@
 declarations.
-Whenever it is possible to use local type inference, the compiler
-assigns a more specific type to an unannotated declaration.
-When it is not possible to use local type inference, the compiler
-assigns the type \texttt{any} to an unannotated declaration.
-For instance, in function declarations, the compiler does not infer
-a more specific type to unannotated input parameters, because,
-in this case, global type inference often is undecidable in a
-language with subtyping and overloading \citep{wells1999typability}.
-
-In the following example, we use type annotations in a function
-declaration, and we do not use type annotations in a declaration of a
-local variable:
+We use the following example to illustrate how we can annotate a
+function declaration and a variable declaration:
+\begin{verbatim}
+    local function succ (x:integer):integer
+      return x + 1
+    end
+    local x:integer = 7
+    x = succ(x)
+    print(x)      --> 8
+\end{verbatim}
+
+Typed Lua uses local type inference to assign more specific types to
+some unannotated declarations.
+More precisely, Typed Lua can infer the type of local variables and
+the return type of local functions that are not recursive.
+The inference that we implement in Typed Lua is quite simple, as it
+uses only the type of the local expression.
+For local variables, Typed Lua uses the type of the initialization
+expression to assign a more specific type to an unannotated local variable.
+For local functions, Typed Lua uses the type of the returned expression
+to assign a more specific type to an uannotated return type.
+This means that we can rewrite the previous example as follows:
+\begin{verbatim}
+    local function succ (x:integer)
+      return x + 1
+    end
+    local x = 7
+    x = succ(x)
+    print(x)      --> 8
+\end{verbatim}
+
+In this example, the compiler uses local type inference to assign the
+type \texttt{integer} to the local variable \texttt{x} and to the
+return type of the local function \texttt{succ}, making this example
+compile without any warnings.
+Local type inference always use the most general type.
+In this example, the compiler does not use the literal type \texttt{7},
+instead of the base type \texttt{integer}, because this would generate
+a warning when we try to assign other integer value to the variable \texttt{x}.
+Still, programmers can use literal types in type annotations if
+they need a variable that has a very specific type.
+In Section \ref{sec:tables} we will see in more detail that literal
+types are essential to type Lua tables.
+
+Typed Lua assigns the type \texttt{any} to the unannotated declarations
+that it cannot infer a more specific type.
+More precisely, Typed Lua cannot infer more specific types to input
+parameters of function declarations and to the return type of
+recursive functions, as global type inference is undecidable in a
+language with subtyping \citep{wells1999typability}.
+
+We use the following example to illustrate type annotations in the
+declaration of a recursive function:
 \begin{verbatim}
@@
     end
-    local x = 5
-    print(factorial(x))
-\end{verbatim}
-
-Even though we did not add type annotations in the declaration of
-the local variable \texttt{x}, the compiler uses local type inference to
-assign the type \texttt{integer} to \texttt{x}, and this example compiles
-without any errors.
-The compiler does not use the literal type \texttt{5},
-instead of the base type \texttt{integer}, because this would prevent
-programmers from assigning other integer values to the variable
-\texttt{x}.
-The inference that we implement in Typed Lua is quite simple, as it
-uses only the type of the local expression.
-
-Typed Lua allows programmers to combine statically typed code with
-dynamically typed code, as we show in the following example:
+\end{verbatim}
+
+This example compiles without any warnings because we annotated
+the return type of \texttt{factorial}.
+
+We use the following example to illustrate the omission of type
+annotations in the input parameters of a local function,
+and also to show that Typed Lua allows programmers to combine
+statically typed code with dynamically typed code:
 \begin{verbatim}
@@
 The compiler assigns the dynamic type \texttt{any} to the input
-parameters of \texttt{distance} because they do not have type annotations,
+parameters of \texttt{distance} because they do not have type annotations
 and the compiler does not use global type inference, as we mentioned
 previously.
-Even though we did not annotated the return type of \texttt{distance},
+Even though we did not annotate the return type of \texttt{distance},
 the compiler is able to infer its return type because it is local
@@
 In this example, Typed Lua cannot guarantee that \texttt{distance} is never
-going to call \texttt{absolute} with a parameter that is not an integer number,
+going to call \texttt{absolute} with a parameter that is not an integer,
 because in the semantics of Lua the minus operator can result in a
@@
 Still, the dynamic type \texttt{any} may be hiding a value of a type
-that is not an integer, making the dynamically typed code crash the
-statically typed code.
+that is not an integer, making the dynamically typed code break the
+guarantees provided by the statically typed code.
 This is a typical example where run-time checks would ensure safety
@@
 Third, Lua has multiple assignment, and the semantics
-of argument passing is the same of the multiple assignment
-(or vice-versa), that is, calling a function is like doing a
-multiple assignment where the left hand side is the parameter list
-and the right hand side is the argument list.
-
-Typed Lua uses second-level types to encode function types,
+of argument passing is the same of the multiple assignment, that is,
+calling a function is like doing a multiple assignment where the
+left hand side is the parameter list and the right hand side is the
+argument list.
+
+Typed Lua uses second-level types to encode function types
 and to preserve these peculiarities.
@@
 parameters.
-They are just an artifact of the interaction between the type system
-and the semantics of Lua.
-
-As we can see in Figure \ref{fig:types}, second-level types consist
+Second-level types represent tuple types that can appear in multiple assignment.
+Since the semantics of argument passing is the same of multiple assignment,
+second-level types also appear in function types.
+
+\begin{figure}[!ht]
+\textbf{Function types}\\
+\dstart
+\begin{align*}
+\textit{functiontype} ::= & \;\; \textit{tupletype} \; \texttt{`->'} \; \textit{rettype}\\
+\textit{tupletype} ::= & \;\; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
+\textit{typelist} ::= & \;\; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
+\textit{rettype} ::= & \;\; \textit{type} \; | \;
+  \textit{uniontuple} \; [\texttt{`?'}]\\
+\textit{uniontuple} ::= & \;\; \textit{tupletype} \; | \;
+  \textit{uniontuple} \; \texttt{`|'} \; \textit{uniontuple}
+\end{align*}
+\dend
+\caption{The concrete syntax of Typed Lua function types}
+\label{fig:functions}
+\end{figure}
+
+As we can see in Figure \ref{fig:functions}, second-level types consist
 of either \emph{tupletype} or \emph{rettype}.
-A second-level type is the type void \texttt{()},
+A second-level type is either the type void \texttt{()},
 a tuple of first-level types optionally ending in a variadic type,
 or a union of these tuples.
+The type void \texttt{()} represents the type of the empty tuple.
 A variadic type \texttt{t*} is a generator for a sequence of values
 of the union type \texttt{t|nil}.
-Union of tuples may appear only on the return type of functions,
-and together with \emph{projection types} play an important role
-in functions that are overloaded on the return type.
-We will explain union types and projection types in more detail
-in the next section.
-
-To match the semantics of Lua function calls, the \emph{default} operation
-mode of Typed Lua adds a variadic tail to the parts of a function
-type whenever the programmer does not specify one.
-In the previous examples, the type of both \texttt{factorial} and
-\texttt{absolute} is \texttt{(integer, value*) -> (integer, nil*)},
-and the type of \texttt{distance} is \texttt{(any, any, value*) -> (integer, nil*)}.
-If we call \texttt{absolute} with extra arguments, Typed Lua silently ignores
-them, as the type signature lets \texttt{absolute} receive any number of
-extra arguments.
-If we call \texttt{absolute} with fewer arguments, Typed Lua uses \texttt{nil} to
-replace the missing arguments and in the case of \texttt{absolute} generates
-an error, as \texttt{nil} is neither a subtype of \texttt{integer} nor consistent
-with \texttt{integer}.
-
-Typed Lua also provides a \emph{strict} operation mode, and programmers can
-use this mode if they want to catch arity mismatch of function calls.
-In the strict mode, Typed Lua does not add a variadic tail to the
-parts of a function type unless the programmer explicitly declares one.
-In the previous examples, the type of both \texttt{factorial} and
-\texttt{absolute} is \texttt{(integer) -> (integer)},
-and the type of \texttt{distance} is \texttt{(any, any) -> (integer)}.
-If we call \texttt{absolute} with either extra or missing arguments,
-Typed Lua generates errors about arity mismatch.
+Union of tuple types appear in the return type of function types to
+represent overloading on the return type.
+We will explain union types in more detail in the next section.
+We can use only one first-level type \texttt{t} in the return type
+because it is syntactic sugar to the tuple type \texttt{(t)}.
+
+Typed Lua provides two operation modes: the \emph{default} mode and
+the \emph{strict} mode.
+In the default mode, the compiler adds a variadic tail to the
+type of the parameter list and to the return type whenever the
+programmer does not specify one.
+The default mode has this behavior to preserve the semantics
+of function calls in Lua, that is, it discards extra arguments and
+uses \texttt{nil} to replace missing arguments.
+In the strict mode, the compiler does not add a variadic tail
+to the type of the parameter list and to the return type.
+Instead, it preserves the type defined by the programmer to catch
+arity mismatch.
+We will use the following function to illustrate how these two
+operation modes work:
+\begin{verbatim}
+    local function sum (x:integer, y:integer):integer
+      return x + y
+    end
+\end{verbatim}
+
+In the default mode, the type of \texttt{sum} is
+\texttt{(integer, integer, value*) -> (integer, nil*)}.
+The compiler adds \texttt{value*} to the type of the parameter list
+to discard extra arguments.
+For instance, the call \texttt{sum(1, 2, 3)} compiles without any
+warnings because Typed Lua uses \texttt{value*} to drop the extra
+argument \texttt{3}.
+
+In the strict mode, the type of \texttt{sum} is
+\texttt{(integer, integer) -> (integer)}.
+The compiler does not add \texttt{value*} to the type of the parameter
+list to catch arity mismatch.
+For instance, the call \texttt{sum(1, 2, 3)} compiles with a warning
+because we are passing an extra argument \texttt{3} to \texttt{sum}.
+
+Even though these operation modes affect function calls, they do not
+affect other multiple assignment.
+For instance,
+\begin{verbatim}
+    local x:integer, y:integer = 1, 2, 3
+\end{verbatim}
+compiles without any warnings in both operation modes.
+This means that the left-hand side of this local declaration has the
+tuple type \texttt{(integer, integer, value*)}, while the expression
+list in the right-hand side of this local declaration has the
+tuple type \texttt{(1, 2, 3, nil*)} in both operation modes.
 
@@
 in an expression list that is not in tail position.
-The following example shows the interaction between multiple returns
-and expression lists:
-\begin{verbatim}
-    local function multiple ()
+We will use the following function to illustrate the interaction
+between multiple returns and expression lists:
+\begin{verbatim}
+    local function m ():(integer, string)
       return 2, "foo"
     end
-    local function sum (x:integer, y:integer)
-      return x + y
-    end
-    local x, y, z = multiple(), multiple()
-    print(sum(multiple(), multiple()))
-\end{verbatim}
-
-In the default mode, the type of \texttt{multiple} is
-\texttt{(value*) -> (integer, string, nil*)},
-and the type of \texttt{sum} is \texttt{(integer, integer, value*) -> (integer, nil*)}.
-In the right hand side of the multiple assignment, only the first
-value produced by the first call to multiple gets used,
-so the type of the right hand side is
-\texttt{(integer, integer, string, nil*)},
-and the types assigned to \texttt{x}, \texttt{y}, and \texttt{z}
-are \texttt{integer}, \texttt{integer}, and \texttt{string}, respectively.
-The same tuple is assigned to the call to \texttt{sum} and it
-compiles without errors, as the first two components of the tuple
-are consistent with the types of the parameters, and the other
-components are consistent with \texttt{value}.
-
-In the strict mode, the type of \texttt{multiple} is \texttt{() -> (integer, string)},
-and the type of \texttt{sum} is \texttt{(integer, integer) -> (integer)}.
-Like in the default mode, only the first value produced by the
-first call to multiple gets used in the right hand side of the
-multiple assignment;
-and the type of the right hand side is also
-\texttt{(integer, integer, string, nil*)},
-because the strict mode affects only the type of function declarations
-and multiple assignments that appear in function calls.
-Unlike in the default mode, the type of the call to \texttt{sum} is
-\texttt{(integer, integer, string)}, and does not include the
-variadic tail to catch arity mismatch.
-Thus, this example compiles with error because the call to \texttt{sum}
-passes an unexpected extra argument of type \texttt{string}.
+\end{verbatim}
+
+As an example,
+\begin{verbatim}
+    local x:integer, y:integer, z:string = m(), m()
+\end{verbatim}
+compiles without warnings in both operation modes.
+This happens because in the right hand side of the multiple
+assignment, only the first value produced by the first call
+to \texttt{m} gets used, so the type of the right hand side
+is \texttt{(integer, integer, string, nil*)}, which is
+consistent with the type of the left hand side
+\texttt{(integer, integer, string, value*)}.
+
+We can also type variadic functions.
+For instance,
+\begin{verbatim}
+    local function v (...:string):(string*)
+      return ...
+    end
+\end{verbatim}
+has type \texttt{(string*) -> (string*)} in both operation modes.
+The call \texttt{v()} compiles without any warnings in the default and
+strict modes because \texttt{(value*)} and \texttt{()} are consistent
+with \texttt{(string*)}.
+The call \texttt{v("hello", "world")} compiles without any warnings
+in the default and strict modes because \texttt{("hello", "world", nil*)}
+and \texttt{("hello", "world")} are consistent with \texttt{(string*)}.
+The call \texttt{v(...)} compiles withouth any warnings in both
+operation modes because the type of the argument list is \texttt{(string*)},
+assuming that the vararg expression (\texttt{...}) has type \texttt{string}.
 
@@
 
-    print(message("Typed Lua"))         -- Typed Lua
-    print(message("Typed Lua", "Hi "))  -- Hi Typed Lua
+    print(message("Typed Lua"))         --> Typed Lua
+    print(message("Typed Lua", "Hi "))  --> Hi Typed Lua
 \end{verbatim}
@@
 Although the parameter \texttt{greeting} is optional, and has type
-\texttt{string|nil}, the concatenation does not generate a type error
+\texttt{string|nil}, the concatenation does not generate a warning
 because we used the short-circuiting \texttt{or} operator to declare a
@@
 common idiom to assign a default value to an optional parameter. 
-Typed Lua uses a typing rule to encode this idiom: 
+Typed Lua uses the following rule to type this idiom: 
 if the left hand side of \texttt{or} has type \texttt{t|nil} and
@@
 the type of \texttt{x} from \texttt{t|nil} to \texttt{t}
-if it matches the following restrictions: 
+if it matches the following rules: 
 the type of \texttt{e} is a subtype of \texttt{t},
@@
 and it is not being assigned in another function.
-The refinement only affects the type of \texttt{x} in the remainder
-of the current scope.
+The refinement only affects the type of \texttt{x} until the
+next assignment to \texttt{x}.
 In the case of \texttt{greeting}, the assignment refines its type
@@
 This example uses the \texttt{type(x) == "string"} predicate that
-constrains the type of \texttt{x} from \texttt{t|string} to
-\texttt{string} when the predicate is true and \texttt{t} otherwise.
+constrains the type of \texttt{x} from \texttt{string|t} to
+\texttt{string} inside the \texttt{if} body and to \texttt{t}
+inside the \texttt{else} body. 
 This is a simplified form of \emph{flow typing} \citep{guha2011tlc,tobin-hochstadt2010ltu}.
@@
 when it successfully finished its execution,
-and usually returns \texttt{nil} as the first value plus an
-error message or other data that describes the error
-when something failed during its execution.
+and returns \texttt{nil} plus an error message or other data that
+describes the error when something failed during its execution.
 Next, we show an example:
@@
 When Typed Lua finds a union of tuples in the right hand side of
-a declaration, it stores this union in a special type environment
-with a fresh name and assigns \emph{projection types} to the
-variables that appear unannotated in the left hand side of the declaration.
-In our example, Typed Lua assigns the union of tuples
-\texttt{(integer, integer, nil*)|(nil, string, nil*)} to a type variable $X$.
-Thus, variable \texttt{q} gets the type $X_{1}$
-that is equivalent to the type \texttt{integer|nil},
-and variable \texttt{r} gets the type $X_{2}$
-that is equivalent to the type \texttt{integer|string}.
-
-Typed Lua replaces projection types with the union of the corresponding
-component of each tuple when it infers that a variable has a projection type.
+a declaration, it assigns \emph{projection types} to the
+variables that appear unannotated in the left hand side of the
+declaration.
+Projection types do not appear in type annotations, but Typed Lua
+uses them to project unions of tuple types into unions of
+first-level types that have a dependency relation.
+We will discuss projection types in more detail in the next chapter.
+
+So far, Typed Lua replaces projection types with the union of the
+corresponding component of each tuple, when it infers that a variable
+has a projection type.
+In our example, the variables \texttt{q} and \texttt{r} get projection
+types that map to the first and second components of the union of
+tuple types \texttt{(integer, integer, nil*)|(nil, string, nil*)}.
+This means that variables \texttt{q} and \texttt{r} have types
+\texttt{integer|nil} and \texttt{integer|string}, respectively.
+
 If a variable with a projection type appears in a \texttt{type} predicate,
-it discriminates against all tuples in the union.
-In our example, $X$ is \texttt{(integer, integer, nil*)|(nil, string, nil*)}
-outside of the \texttt{if} statement,
+it discriminates against all tuples in the projected union.
+In our example, the projected union has type
+\texttt{(integer, integer, nil*)|(nil, string, nil*)} outside of the
+\texttt{if} statement,
 but \texttt{(integer, integer, nil*)} inside the \texttt{then} block,
 and \texttt{(nil, string, nil*)} inside the \texttt{else} block.
-Thus, variable \texttt{q} has type \texttt{integer} and variable \texttt{r} also
+Thus, variable \texttt{q} has type \texttt{integer|nil} and variable
+\texttt{r} has type \texttt{integer|string} outside of the
+\texttt{if} statement;
+but variable \texttt{q} has type \texttt{integer} and variable \texttt{r} also
 has type \texttt{integer} inside the \texttt{then} block;
@@
 
-We could also have used \texttt{type(q) == "integer"} or
-even \texttt{type(r) == "integer"} as the predicate of our example,
+We could have used \texttt{math.type(q) == "integer"} or
+even \texttt{math.type(r) == "integer"} as the predicate of our example,
 as both predicates would produce the same result.
@@
 Note that the type \texttt{integer} is restricted to Lua 5.3,
-as we use the function \texttt{math.type} to decide whether a number
-is integer or not.
+as we use the function \texttt{math.type} to decide whether a number is integer.
 
@@
 of this function as
-\texttt{(number) -> (number) \string&
-(string) -> (string)},
-and therefore solve this limitation,
+\texttt{(number) -> (number) \string& (string) -> (string)},
 intersection types require more sophisticated flow typing to
@@
 $\texttt{t}_{\texttt{i}}$.
+
+\begin{figure}[!ht]
+\textbf{Table types}\\
+\dstart
+\begin{align*}
+\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
+\textit{tabletypebody} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type} \; | \;
+  \textit{recordtype}\\
+\textit{keytype} ::= & \;\; \textit{basetype} \; | \;
+  \textbf{value}\\
+\textit{recordtype} ::= & \;\; \textit{recordfield} \; \{\texttt{`,'} \; \textit{recordfield}\} \; [\texttt{`,'} \; \textit{type}]\\
+\textit{recordfield} ::= & \;\; [\textbf{const}] \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}
+\end{align*}
+\dend
+\caption{The concrete syntax of Typed Lua table types}
+\label{fig:tables}
+\end{figure}
 
@@
 
-\section{Interfaces and type aliases}
+\section{Type aliases and interfaces}
 \label{sec:alias}
@@
 \end{verbatim}
+
+\begin{figure}[!ht]
+\textbf{Type alias}\\
+\dstart
+\begin{align*}
+\textit{typealias} ::= & \; \textbf{typealias} \; \textit{Name} \; \texttt{`='} \; \textit{type}
+\end{align*}
+\dend
+\caption{The concrete syntax of Typed Lua type aliases}
+\label{fig:aliases}
+\end{figure}
+
+\begin{figure}[!ht]
+\textbf{Interfaces}\\
+\dstart
+\begin{align*}
+\textit{interface} ::= & \; \textbf{interface} \; \textit{typedec}\\
+\textit{typedec} ::= & \; \textit{Name} \; { \; \textit{decitem} \; } \; \textbf{end}\\
+\textit{decitem} ::= & \; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
+\textit{idtype} ::= & \; \textit{type} \; | \; \textit{methodtype}\\
+\textit{idlist} ::= & \; \textit{id} \; \{ \; \texttt{`,'} \; \textit{id} \; \}\\
+\textit{id} ::= & \; [\textbf{const}] \; \textit{Name}\\
+\textit{typealias} ::= & \; \textbf{typealias} \; \textit{Name} \; \texttt{`='} \; \textit{type}
+\end{align*}
+\dend
+\caption{The concrete syntax of Typed Lua interfaces}
+\label{fig:interfaces}
+\end{figure}
 
@@
 We can also use interfaces to define recursive types such as liked lists:
-
 \begin{verbatim}
@@
 \label{sec:tld}
+
+An important feature of Typed Lua is its interface between
+statically and dynamically typed modules.
+This means that the programmer can enjoy some of the benefits of
+static types even without converting existing Lua modules to Typed Lua,
+as a dynamically typed module can export a statically typed interface,
+and statically typed users of the module have their use of the module
+checked by the compiler.
 

### 2015-01-21 doc/thesis/typedlua.tex
Improving the text about table types, aliases, and modules

@@
 \textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
-\textit{tabletypebody} ::= & \;\; \textit{hashtype} \; | \;
+\textit{tabletypebody} ::= & \;\; \textit{maptype} \; | \;
   \textit{recordtype}\\
-\textit{hashtype} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type}\\
+\textit{maptype} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type}\\
 \textit{keytype} ::= & \;\; \textit{basetype} \; | \;
@@
 
-Typed Lua allows the assignment \texttt{x = x or e} to refine
-the type of \texttt{x} from \texttt{t|nil} to \texttt{t}
+Typed Lua allows the assignment \texttt{v = v or e} to refine
+the type of \texttt{v} from \texttt{t|nil} to \texttt{t}
 if it matches the following rules: 
 the type of \texttt{e} is a subtype of \texttt{t},
-and the variable \texttt{x} is local to the current function
+and the variable \texttt{v} is local to the current function
 and it is not being assigned in another function.
-The refinement only affects the type of \texttt{x} until the
-next assignment to \texttt{x}.
+The refinement only affects the type of \texttt{v} until the
+next assignment to \texttt{v}.
 In the case of \texttt{greeting}, the assignment refines its type
@@
 programmers to constrain the type of a local variable inside a condition.
-This example uses the \texttt{type(x) == "string"} predicate that
-constrains the type of \texttt{x} from \texttt{string|t} to
+This example uses the \texttt{type(v) == "string"} predicate that
+constrains the type of \texttt{v} from \texttt{string|t} to
 \texttt{string} inside the \texttt{if} body and to \texttt{t}
@@
 Lua has syntactic sugar for indexing tables as records:
-\texttt{t.k} is syntactic sugar for \texttt{t["k"]}, and means that
-we are accessing the field \texttt{k} from table \texttt{t}. 
-In this section, we show how Typed Lua encodes tables as maps,
+\texttt{t.k} is syntactic sugar for \texttt{t["k"]}.
+In this section, we show how Typed Lua types tables that encode maps,
 arrays, and records.
-
-Table types are the only framework that Typed Lua has to represent
-the different uses that a Lua table has.
-A table type $\texttt{\{k}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
-\texttt{, ..., k}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}$ represents a map
-from values of type $\texttt{k}_{\texttt{i}}$ to values of type
-$\texttt{t}_{\texttt{i}}$.
 
@@
 \textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
-\textit{tabletypebody} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type} \; | \;
+\textit{tabletypebody} ::= & \;\; \textit{maptype} \; | \;
   \textit{recordtype}\\
+\textit{maptype} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type}\\
 \textit{keytype} ::= & \;\; \textit{basetype} \; | \;
@@
 
-Although Typed Lua has just one framework to represent tables,
-it has three different syntaxes for defining the type of a table:
-\begin{align}
-\label{syn:table1}
-\texttt{`\{'} \; \textit{keytype} \; \texttt{`:'} \; \textit{type} \; \texttt{`\}'}\\
-\label{syn:table2}
-\texttt{`\{'} \; \textit{type} \; \texttt{`\}'}\\
-\label{syn:table3}
-\texttt{`\{'} \; \textit{literaltype} \; \texttt{`:'} \; \textit{type} \;
-\{\texttt{`,'} \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}\} \;
-[\texttt{`,'} \; \textit{type}] \; \texttt{`\}'}
-\end{align}
-The syntax \ref{syn:table1} defines table types for maps:
-it is written \texttt{\{k:t\}},
-and maps to the table type \texttt{\{k:t|nil\}}.
-This table type represents a map that maps values of type
-\texttt{k} to values of type \texttt{t|nil},
-as Lua returns \texttt{nil} when we use a non-existing key to index a table.
-The syntax \ref{syn:table2} defines table types for arrays:
-it is written \texttt{\{t\}},
-and maps to the table type \texttt{\{integer:t|nil\}}.
-This table type is just a syntactic sugar to the definition of a
-map that maps values of type \texttt{integer} to values of type \texttt{t|nil}.
-The syntax \ref{syn:table3} defines table types for records with an
-optional array part: it is written
-$\texttt{\{l}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
-\texttt{, ..., l}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}$,
-and maps to the same table type.
-This table type represents a record that maps the literal types
-$\texttt{l}_{\texttt{i}}$\texttt{, ..., }$\texttt{l}_{\texttt{n}}$
-to values of types $\texttt{t}_{\texttt{i}}$\texttt{, ..., }$\texttt{t}_{\texttt{n}}$.
-
+Figure \ref{fig:tables} shows that the concrete syntax of Typed Lua
+table types is restricted to either the type of the empty table,
+maps, arrays, or records with an optional array part.
+We made this design choice due to the results that we obtained
+about the usage of the table constructor, which we discussed
+in Section \ref{sec:statistics}.
+Later in this chapter we will show that we use the syntax of records
+to type modules and objects.
+We will also show that we need the \texttt{const} modifier while
+typing object-oriented code.
+
+The table type \texttt{\{k:t\}} represents the type of a map
+from values of type \texttt{k} to values of type \texttt{t|nil}.
+Table types that represent maps always include the type \texttt{nil},
+because Lua returns \texttt{nil} when we use a non-existing key to
+index a table. 
+The types of the keys are restricted to either base types or the
+type \texttt{value} due to the statistics of the table constructor,
+which we discussed in Section \ref{sec:statistics}.
 Next, we show one example of table type to type a map from strings to integers:
@@
     local t:{string:integer} = { foo = 1 } 
-    local x:integer = t.foo               -- compile-time error
-    local y:integer? = t.bar              -- y gets nil
-    local z:integer = t["bar"] or 0       -- z gets 0 
-\end{verbatim}
-
-The second line of this example gives a type error because we are
+    local x:integer = t.foo           --> compile-time warning
+    local y:integer? = t.bar          --> y gets nil
+    local z:integer = t["bar"] or 0   --> z gets 0 
+\end{verbatim}
+
+The second line of this example raises a warning, because we are
 attempting to assign a value of type \texttt{integer|nil} to a
@@
 Although the field \texttt{bar} does not exist in \texttt{t}, the third
-line of this example does not give a type error because the
+line of this example does not raise a warning, because the
 annotated type matches the type of the values that can be stored in
@@
 
-Next, we show one example of table type to type an array:
+The table type \texttt{\{t\}} represents the type of an array that 
+maps values of type \texttt{integer} to values of type \texttt{t|nil}.
+In other words, Typed Lua handles arrays as syntactic sugar to the
+table type \texttt{\{integer:t\}}.
+Next, we show one example of table type to type an array of strings:
 \begin{verbatim}
@@
                             "Saturday" }
-    local x = days[1]                    -- x gets "Sunday"
-    local y = days[8]                    -- y gets nil
-\end{verbatim}
-
-In this example, the type of both variables \texttt{x} and \texttt{y}
-is \texttt{string|nil}.
+    local i = 5
+    local x = days[1]                 --> x gets "Sunday"
+    local y = days[8]                 --> y gets nil
+    local z = days[i]                 --> z gets "Thursday"
+\end{verbatim}
+
+In this example, the type of \texttt{i} is \texttt{integer},
+while the type of \texttt{x}, \texttt{y}, and \texttt{z} is
+\texttt{string|nil}.
 An inconvenient aspect of making the types of maps and arrays always
@@
 
+The table type
+$\texttt{\{l}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
+\texttt{, ..., l}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}$
+represents the type of a record that maps literal types
+$\texttt{l}_{\texttt{i}}$\texttt{, ..., }$\texttt{l}_{\texttt{n}}$
+to values of types $\texttt{t}_{\texttt{i}}$\texttt{, ..., }$\texttt{t}_{\texttt{n}}$.
+Most programming languages treat records as maps from names
+to types, but we chose to use literal types instead of
+names due to the statistics results that we obtained about
+the usage of the table constructor, which we discussed in
+Section \ref{sec:statistics}.
+This means that we can use the syntax of records to type a
+list that has fixed elements, like the variable \texttt{days} from
+the previous example.
+
 When we know that a list has fixed elements, we can leave the
@@
 a more specific table type to the variable.
-If we remove the annotation in this example, the compiler uses the
+If we remove the annotation in the previous example, the compiler uses the
 syntax of records to infer the following table type to \texttt{days}:
@@
 
-When we use the inferred table type while initializing a table, the
-compiler gives an error message when we try to access an index that
-is out of bounds.
-
-We can also use the style of the inferred type to define heterogeneous
-tuples:
+When we use the syntax of records to type an array, the compiler
+raises a warning when we try to access an index that is out of bounds.
+In the previous example, both \texttt{days[8]} and \texttt{days[i]}
+raise warnings, as both the literal type \texttt{8} and the base type
+\texttt{integer} do not map to any value.
+
+We can also use the syntax of records to type heterogeneous tuples:
 \begin{verbatim}
@@
 
-The next example shows how we can annotate a \texttt{map} function
-that receives another function as its parameter and uses this
-function to alter the elements of a list of numbers, which is
-also a parameter of \texttt{map}:
-\begin{verbatim}
-    local function map (f:(number)->(number), l:{number})
-      for i = 1, #l do
-        l[i] = f(l[i] or 0)
-      end
-    end
-\end{verbatim}
-
-Next, we show one example of table type to type a record:
+Next, we show one example of table type to type a record that maps
+names to strings:
 \begin{verbatim}
@@
 
-Type inference would infer the very same table type that we used to
-annotate the variable \texttt{person}.
+In these two previous examples, local type inference would infer the
+very same table types that we used to annotate the variables
+\texttt{album} and \texttt{person}.
 
 Lua programmers often build records incrementally, that is,
-they usually declares a local variable with an empty table,
+they usually declare a local variable with an empty table,
 and then use assignment to add fields to this table:
@@
     person.lastname = "Reed"
-    print("bye bye " .. person.firstname)  -- bye bye Lou
+    print("bye bye " .. person.firstname)  --> bye bye Lou
 \end{verbatim}
@@
 the programmer to change the type of the value that \texttt{person} points to.
-This type change is safe in this example, but it is not in the following one:
+
+Typed Lua tags a variable that holds a table type as either
+\emph{unique}, \emph{open}, or \emph{closed}.
+If a variable gets is type from a table constructor then it is
+unique, otherwise it is closed.
+Any alias to unique variables makes them open, as we use the
+tag open to keep track of unique variables that are aliased.
+In the previous example, \texttt{person} has an unique type
+because it has no alias.
+Typed Lua also has different subtyping rules that handle these
+different tags.
+We explain these tags and their subtyping rules in more detail
+in Chapter \ref{chap:system}.
+
+Typed Lua uses three rules to decide whether it is sound to
+allow field assignment to change the type of unique and open
+variables:
+the variable must be local to the current block,
+the new type must only add new fields, and
+the variable cannot have been assigned in another function.
+
+The following example shows that it is not sound to allow the
+type of the fields to change:
+\begin{verbatim}
+    local person = { firstname = "Lewis", lastname = "Reed" }
+    person.middlename = "Allan"
+    person.firstname = 1942           --> compile-time warning
+    person.lastname = 2013            --> compile-time warning
+    print("bye bye " .. person.firstname)
+\end{verbatim}
+
+In this example, both third and fourth lines are unsound because
+they are trying to change the type of fields that already exist.
+
+It is also not sound to allow changing the type of an alias:
 \begin{verbatim}
@@
     person.lastname = "Reed"
-    bogus.firstname = true               -- compile-time error
+    bogus.firstname = true            --> compile-time warning
     print("bye bye " .. person.firstname)
@@
 programmer to change the type of the value that is stored in
-\texttt{person.firstname}, and makes the last line crash during
-run-time.
-On the first line, we assign an empty table to \texttt{person}.
-On the second line, we assign the type of \texttt{person} to
+\texttt{person.firstname}, and makes the last line break the
+guarantees provided by the statically typed code.
+In the first line, we assign an empty table to \texttt{person}.
+In the second line, we assign the type of \texttt{person} to
 \texttt{bogus}.
-On the third and fourth lines we change the type of \texttt{person}
-as we did in the previous example.
-On the fifth line we try to change the type of \texttt{bogus}
+In the third and fourth lines we change the type of \texttt{person}.
+In the fifth line we try to change the type of \texttt{bogus}
 from \texttt{\{\}} to \texttt{\{"firstname":boolean\}},
@@
 
+We can create aliases to open variables, but they are always closed.
+Any mutation in the original reference and in the aliased reference
+is not a problem, as the type of the original reference can only
+add new fields.
+For mutable fields this means that the type of the field cannot
+change after it is added to the type of the table.
+In our previous example, the second line changes the type of
+\texttt{person} from unique to open and makes the type of
+\texttt{bogus} closed.
+
 The location of the change also matters, as the next example shows:
@@
     local function spoil ()
-      person.firstname = nil             -- compile-time error
+      person.firstname = nil          --> compile-time warning
     end
@@
 
-In this example, the third line is unsound because it allows the
-programmer to change the type of \texttt{person} outside of the scope
-that it was declared.
+In this example, the third line is unsound because this line allows
+the programmer to change the type of \texttt{person} outside of the
+scope that \texttt{person} was declared.
 The call to \texttt{spoil} erases the field \texttt{firstname} from
-\texttt{person}, and makes the last line crash during run-time.
-
-It is also not sound to allow the type of the fields to change,
-as the next example shows:
-\begin{verbatim}
-    local person = { firstname = "Lewis", lastname = "Reed" }
-    person.middlename = "Allan"
-    person.firstname = 1942              -- compile-time error
-    person.lastname = 2013               -- compile-time error
-    print("bye bye " .. person.firstname)
-\end{verbatim}
-
-In this example, both third and fourth line are unsound because
-they are trying to change the type of fields that already exist.
-
-Typed Lua tags a variable that holds a table type as either
-\emph{open} or \emph{closed}.
-If a variable gets its type from a table constructor then it
-is open, otherwise it is always closed.
-
-Typed Lua uses three restrictions to decide whether it is
-sound to allow field assignment to change the type of an
-open variable or not:
-the variable must be local to the current block,
-the new type must be a subtype of the old one, and
-the variable cannot have been assigned in another function.
-
-We can create aliases to open variables, but they are always closed.
-Any mutation on these aliases is not a problem, as the type of
-the original reference can only change towards a subtype.
-For mutable fields this means that the type of the field cannot
-change after it is added to the type of the table.
-
-Table types along with their refinement through field assignment
-are the foundation for defining modules and objects in Typed Lua.
-In Section \ref{sec:modules} we show that the refinement of table
-types through field assignment is crucial for making Typed Lua
-support idiomatic definition of Lua modules.
-In Section \ref{sec:oop} we show that the refinement of table
-types is also helpful for defining classes and objects.
+\texttt{person}, and makes the last line break the guarantees
+provided by the statically typed code.
 
@@
 
-Typed Lua includes \emph{interfaces} as syntactic sugar to table types
-that specify records, as writing table types can be unwieldy when
-records get bigger and the types of record fields get more complicated.
-For instance, we can use the following interface as an alias to the
-type of the record we showed in the previous section:
-\begin{verbatim}
-    local interface Person
-      firstname:string
-      lastname:string
-    end
-\end{verbatim}
+Typed Lua includes \emph{type aliases} for allowing programmers to
+define their own data types.
+Figure \ref{fig:aliases} shows the concrete syntax of the
+\texttt{typealias} construct.
 
 \begin{figure}[!ht]
-\textbf{Type alias}\\
+\textbf{Type aliases}\\
 \dstart
 \begin{align*}
-\textit{typealias} ::= & \; \textbf{typealias} \; \textit{Name} \; \texttt{`='} \; \textit{type}
+\textit{typealias} ::= & \;\; [\textbf{local}] \; \textbf{typealias} \; \textit{Name} \; \texttt{`='} \; \textit{type}
 \end{align*}
@@
 
+As an example, the following declaration defines the type \texttt{Person}
+as an alias to the table type \texttt{\{"firstname":string, "lastname":string\}}
+in the remainder of the current scope:
+\begin{verbatim}
+    local typealias Person = {"firstname":string,
+                              "lastname":string}
+\end{verbatim}
+
+Typed Lua also includes \emph{interfaces} as syntactic sugar to aliases
+for table types that specify records, as writing table types can be unwieldy
+when records get bigger and the types of record fields get more complicated.
+Figure \ref{fig:interfaces} shows the concrete syntax of the
+\texttt{interface} construct.
+In Section \ref{sec:oop} we will show that we can also use \emph{interface} to
+declare the type of classes, and that \emph{methodtype} is syntactic
+sugar to express the type of methods.
+
 \begin{figure}[!ht]
@@
 \begin{align*}
-\textit{interface} ::= & \; \textbf{interface} \; \textit{typedec}\\
-\textit{typedec} ::= & \; \textit{Name} \; { \; \textit{decitem} \; } \; \textbf{end}\\
-\textit{decitem} ::= & \; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
-\textit{idtype} ::= & \; \textit{type} \; | \; \textit{methodtype}\\
-\textit{idlist} ::= & \; \textit{id} \; \{ \; \texttt{`,'} \; \textit{id} \; \}\\
-\textit{id} ::= & \; [\textbf{const}] \; \textit{Name}\\
-\textit{typealias} ::= & \; \textbf{typealias} \; \textit{Name} \; \texttt{`='} \; \textit{type}
+\textit{interface} ::= & \;\; [\textbf{local}] \; \textbf{interface} \; \textit{typedec}\\
+\textit{typedec} ::= & \;\; \textit{Name} \; { \; \textit{decitem} \; } \; \textbf{end}\\
+\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
+\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
+\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
+\textit{id} ::= & \;\; [\textbf{const}] \; \textit{Name}
 \end{align*}
@@
 
-The declaration of the interface \texttt{Person} is an alias to the
-table type \texttt{\{"firstname":string, "lastname":string\}} in the
-remainder of the current scope.
-Now we can use \texttt{Person} in type declarations:
+As an example, we can use the following \texttt{interface} to
+define the type \texttt{Person} from the previous example:
+\begin{verbatim}
+    local interface Person
+      firstname:string
+      lastname:string
+    end
+\end{verbatim}
+
+After we define the type \texttt{Person}, we can use it in type annotations:
 \begin{verbatim}
@@
 
-    print(greeter(user1)) -- Goodbye Lewis Reed 
-    print(greeter(user2)) -- compile-time error
-    print(greeter(user3)) -- Goodbye Lou Reed
-    print(greeter(user4)) -- compile-time error
+    print(greeter(user1))             --> Goodbye Lewis Reed 
+    print(greeter(user2))             --> compile-time warning
+    print(greeter(user3))             --> Goodbye Lou Reed
+    print(greeter(user4))             --> compile-time warning
 \end{verbatim}
@@
 
-If our record type has optional fields, that is, fields that can be \texttt{nil},
-we need to use an explicit type declaration when declaring a
-variable of this record type, as we show in the following example:
-\begin{verbatim}
-    local interface Person
-      firstname:string
-      middlename:string?
-      lastname:string
-    end
-    local user1:Person = { firstname = "Lewis",
-                           middlename = "Allan",
-                           lastname = "Reed" }
-    local user2:Person = { lastname = "Reed",
-                           firstname = "Lou" }
-\end{verbatim}
-
-We need an explicit type declaration because neither of the types
-that the Typed Lua compiler assigns to the above table constructors
-is a subtype of
-\texttt{\{"firstname":string, "middlename":string|nil, "lastname":string\}},
-the type that \texttt{Person} describes.
-
-We can also use interfaces to define recursive types such as liked lists:
+Even though the interface declaration may look redundant due to
+the type alias declaration, it has a more convenient syntax
+for declaring table types that express records.
+For this reason, it worth having two different constructs;
+one specifically for records and another for more general types.
+
+The \texttt{interface} and \texttt{typealias} constructs also allow
+the declaration of recursive types.
+For instance, the following interface defines a type for singly-linked
+lists of integers:
 \begin{verbatim}
@@
 \end{verbatim}
-
-We need to explicitly annotate \texttt{next} being of type
-\texttt{Element|nil} because in our type system \texttt{nil}
-is not the bottom type, that is, \texttt{nil} is subtype of itself only.
-Typed Lua has this behavior because in Lua \texttt{nil} represents
-the absence of an useful value.
-Thus, making \texttt{nil} the bottom type would severally reduce the amount
-of type safety in the system.
 
@@
     while l do
-      if l then print(l.info) end
-      l = l.next                  -- compile-time error
-    end
-\end{verbatim}
-
-However, this example has one compile-time error because we
-are attempting to index the variable \texttt{l},
-which can be \texttt{nil}.
-Notice that we would have the same compile-time error while
-printing the list element if we had not narrowed the type of
-\texttt{l}.
-To eliminate this compile-time error we have to write a
-function that narrows the type of \texttt{l} before using it:
-\begin{verbatim}
-    local function get_next (l:Element?)
-      if l then return l.next else return nil end
-    end
-\end{verbatim}
-
-Now we can use \texttt{get\string_next} to rewrite the \texttt{while}
-from the previous example, thus eliminating the compile-time error:
-\begin{verbatim}
-    while l do
-      if l then print(l.info) end
-      l = get_next(l)
-    end
-\end{verbatim}
-
-Typed Lua also includes a way to create aliases for types that does not
-follow the records syntax.
-However, we limit aliases to union types, function types, table types,
-and recursive types.
-The reason for this restriction is to avoid the creation of aliases
-to predefined types.
-
-For instance, we can use the \texttt{typealias} construct to define an
-alias to the type of the abstract syntax tree of a language of simple
+      print(l.info)
+      l = l.next
+    end
+\end{verbatim}
+
+Note that the type of \texttt{l} is \texttt{Element} in the
+first line of the \texttt{while} body, but it is restored to
+\texttt{Element|nil} in the second line of the \texttt{while} body
+due to the assignment.
+These rules allow this example to compile without any warnings,
+and we discuss them in more detail in the next chapter.
+
+As another example of recursive type, the following type alias
+defines a type for the abstract syntax tree of a language of simple
 arithmetic expressions:
@@
 algebraic data types from functional programming.
-Typed Lua also includes typing rules that let programmers
-deconstruct unions of table types, which resembles
-pattern matching from functional programming.
+The small set of \texttt{type} predicates introduced by
+Typed Lua also includes a rule that let programmers
+deconstruct unions of table types.
+This specific rule resembles pattern matching from functional programming.
 We can use this feature to write an evaluation function for
@@
 
-When Typed Lua finds the pattern \texttt{n[s1] == s2}, it checks
-if the type of the variable \texttt{n} is an union of table types,
-if \texttt{s1} a literal string, and if \texttt{s2} is also a literal string.
-If that is the case, it constrains the type of \texttt{n} inside
-the \texttt{if} to the table type that has the pair \texttt{s1:s2}.
-For instance, the expression \texttt{e.tag == "Add"} makes Typed Lua
+When Typed Lua finds the predicate $\texttt{v[e}_{\texttt{1}}\texttt{] == e}_{\texttt{2}}$,
+it checks whether the type of the variable \texttt{v} is an union of table types,
+whether the type of $\texttt{e}_{\texttt{1}}$ is a literal type $\texttt{l}_{\texttt{1}}$,
+and whether the type of $\texttt{e}_{\texttt{2}}$ is a literal type $\texttt{l}_{\texttt{2}}$.
+If that is the case, it constrains the type of \texttt{v} inside
+the \texttt{if} to the table type that has the key type
+$\texttt{l}_{\texttt{1}}$ that maps to the type $\texttt{l}_{\texttt{2}}$.
+In our example, the expression \texttt{e.tag == "Add"} makes Typed Lua
 constrain the type of \texttt{e} to \texttt{\{"tag":"Add", 1:Exp, 2:Exp\}}.
-This is safe because Typed Lua restricts this pattern to literal types only.
-If we did not add this pattern, we would not be able to traverse the AST,
-as any attempt to index an union is a compile-time error.
-
-Even though the type alias declaration may make the interface declaration
-look redundant, it does not have a syntax as nice as the syntax
-of the interface declaration for defining records.
-For this reason, it worth having two different constructs.
-One specifically for records and another for more general types.
+If we did not add this special predicate, we would not be able to traverse the AST,
+as any attempt to index an union raises a compile-time warning.
 
@@
 but it provides mechanisms for organizing a program in modules.
-To load a module, Lua first checks whether the module is already
-loaded or not.
-When the module is not loaded, Lua executes its source file as a
-function, and the value returned by this function is the module.
-Thus, Lua modules can be any Lua value.
-However, our survey from Section \ref{sec:statistics} confirms that
+To load a module, Lua first checks whether the module is already loaded.
+When the module is not loaded, Lua executes its source file, and
+the value returned is the module.
+Although programmers can use functions for defining modules,
+our survey from Section \ref{sec:statistics} confirms that
 the convention among Lua programmers is to use tables for defining modules. 
-In this case, the fields of the table are the functions and other
+In this convention, the fields of the table are the functions and other
 values that the module exports.
 
-The simplest way for defining a module in Lua is to declare an
-empty table at the begin of the source file,
-populate this table with the members that the module should export,
-and return this table at the end of the source file. 
-The following example shows how Typed Lua handles this case:
-\begin{verbatim}
-    local mymath = {}
-    local RADIANS_PER_DEGREE = 3.14159 / 180.0
-    mymath.deg = function (x:number)
-      return x / RADIANS_PER_DEGREE
-    end
-    mymath.rad = function (x:number)
-      return x * RADIANS_PER_DEGREE
-    end
-    mymath.pow = function (x:number, y:number)
-      return x ^ y
-    end
-    return mymath
-\end{verbatim}
-
-In this example, Typed Lua uses the refinement of table types to
-allow the programmer to incrementally build the type of the
-module \texttt{mymath}.
-Like in Lua, the variable \texttt{RADIANS\string_PER\string_DEGREE}
-is private because we declared it as local to the module. 
-The \texttt{return} at the end of the source file gives the
-final type of the module \texttt{mymath}:
-\begin{verbatim}
-    { "deg":(number) -> (number),
-      "rad":(number) -> (number),
-      "pow":(number, number) -> (number) }
-\end{verbatim}
-
-Another way of defining modules in Lua is to declare only locals
-in the source file, and return a table constructor that includes
-the members that the module should export at the end of the source file.
-The following example shows how Typed Lua handles this case:
+An idiomatic way for defining modules in Lua is to declare only locals
+in the source file and return a table constructor at the end of the source file.
+The returned constructor includes the members that the module should export.
+The following example illustrates this case in Typed Lua:
 \begin{verbatim}
@@
 
-In this example, the \texttt{return} at the end of the source file
-also gives the final type of the module \texttt{mymath}, which is
-the same of the previous example.
-
-There is still a third way of defining modules in Lua through the
-\texttt{module} function, which has been deprecated.
-In this style, the first line of the module calls \texttt{module}.
-This function installs a fresh table as the global environment for
-the rest of the module, so any assignments to global variables are
-field assignments to this table.
-The \texttt{module} function also sets an \texttt{\string_M} field
-in this table as a circular reference to the table itself,
-so the module can end with \texttt{return \string_M},
-but this explicit return is not necessary.
-
-Even though this style has been deprecated, our survey from
-Section \ref{sec:statistics} indicated that around a third of the
-Lua modules that are in LuaRocks still use this style.
-Thus, Typed Lua also supports this style:
-\begin{verbatim}
-    module "mymath"
+In this example, Typed Lua uses the type of the table constructor
+to allow the programmer to build the type of the module.
+The local variable \texttt{RADIANS\string_PER\string_DEGREE}
+is private because we did not include it in the list of exported members. 
+The \texttt{return} at the end of the source file gives the type of the module:
+\begin{verbatim}
+    { "deg":(number) -> (number),
+      "rad":(number) -> (number),
+      "pow":(number, number) -> (number) }
+\end{verbatim}
+
+Another idiomatic way for defining modules in Lua is to declare an
+empty table at the begin of the source file,
+populate this table with the members that the module should export,
+and return this table at the end of the source file. 
+The following example illustrates this case in Typed Lua:
+\begin{verbatim}
+    local mymath = {}
     local RADIANS_PER_DEGREE = 3.14159 / 180.0
-    deg = function (x:number)
+    mymath.deg = function (x:number)
       return x / RADIANS_PER_DEGREE
     end
-    rad = function (x:number)
+    mymath.rad = function (x:number)
       return x * RADIANS_PER_DEGREE
     end
-    pow = function (x:number, y:number)
+    mymath.pow = function (x:number, y:number)
       return x ^ y
     end
-\end{verbatim}
-
-In this example, Typed Lua treats accesses to global variables as
-field accesses to an open table in the top-level scope,
-and at the end of file this open table gives the final type of the
-module \texttt{mymath}, which is also the same of the previous two
-examples.
-
-After we define the module \texttt{mymath}, regardless of adopted style,
+    return mymath
+\end{verbatim}
+
+In this example, Typed Lua uses the refinement of table types to
+allow the programmer to incrementally build the type of the module.
+The variable \texttt{RADIANS\string_PER\string_DEGREE}
+is private because we declared it as local to the module. 
+The \texttt{return} at the end of the source file gives the
+type of the module, which is the same type of \texttt{mymath}.
+This module has the same type of the module from the previous example.
+
+%There is still a third way for defining modules in Lua through the
+%\texttt{module} function, which has been deprecated.
+%Even though this style has been deprecated, our survey from
+%Section \ref{sec:statistics} indicated that around a third of the
+%Lua modules that are in LuaRocks still use this style.
+%Thus, Typed Lua also supports this style.
+
+After we define the module \texttt{mymath}, regardless of the adopted style,
 users can use it in the standard way:
@@
     local mymath = require "mymath"
-    print(mymath.pow(2, 3))              -- 8
-    print(mymath.pow(2, "foo"))          -- compile-time error
-\end{verbatim}
-
-And now the users of the module \texttt{mymath} have the policies that
-Typed Lua provides.
-The call to \texttt{require} assigns the type of the module \texttt{mymath}
-to the local \texttt{mymath}, so Typed Lua can catch misuses of the module.
-
-The way Typed Lua handles modules using table types is also
+    print(mymath.pow(2, 3))           --> 8
+    print(mymath.pow(2, "foo"))       --> compile-time warning
+\end{verbatim}
+
+In Typed Lua, \texttt{require} is a primitive that statically type checks
+a given module to infer the type of this module.
+This means that the type of its input parameter must be a literal string,
+as Typed Lua uses this literal string to find the source file that
+implements the given module.
+To statically type check a module, Typed Lua follows the same rules
+that Lua follows to load a module.
+Typed Lua first checks whether the module is already statically type checked.
+When the module is not yet statically type checked, Typed Lua
+statically type checks its source file, and the type returned is the
+type of the module.
+Typed Lua raises a compile-time warning when it cannot find the source
+file of a given module.
+
+After we use \texttt{require} to statically type check a module,
+Typed Lua can statically type check the usage of this module.
+In our previous example, the call to \texttt{require} assigns the type
+of the module \texttt{mymath} to the local \texttt{mymath},
+so Typed Lua can catch misuses of the module.
+
+The way that Typed Lua handles modules using table types is also
 relevant for supporting object-oriented programming, as we discuss in

### 2015-01-22 doc/thesis/typedlua.tex
Improvements to the text about OO and TLD

@@
 \textit{interface} ::= & \;\; [\textbf{local}] \; \textbf{interface} \; \textit{typedec}\\
-\textit{typedec} ::= & \;\; \textit{Name} \; { \; \textit{decitem} \; } \; \textbf{end}\\
+\textit{typedec} ::= & \;\; \textit{Name} \; {\textit{decitem}} \; \textbf{end}\\
 \textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
@@
 algebraic data types from functional programming.
-The small set of \texttt{type} predicates introduced by
-Typed Lua also includes a rule that let programmers
-deconstruct unions of table types.
-This specific rule resembles pattern matching from functional programming.
+The small set of \texttt{type} predicates that we mentioned
+in Section \ref{sec:unions} also includes a specific rule
+that is not based on tags.
+This specific rule let programmers discriminate unions of table types
+and resembles the pattern matching from functional programming.
 We can use this feature to write an evaluation function for
@@
 The basic mechanism is the \texttt{:} syntactic sugar for method
-calls and method definitions.
-The Lua interpreter translates \texttt{obj:method(args)} into
-an operation that first evaluates \texttt{obj}, if it is a table
-Lua checks if there is field \texttt{method} in this table,
-and then calls it with the result of evaluating \texttt{obj} as
-the first argument, followed by the result of evaluating the
-argument list in the original expression.
-
-Typed Lua uses table types along with the \texttt{self} type to
-represent objects, as we show in the following example:
+definitions and method calls.
+In the case of method definitions, the Lua compiler translates
+\texttt{function obj:method(args) end} into an operation that
+assigns a function to the field \texttt{method} in \texttt{obj}.
+This function includes a first parameter named \texttt{self} plus
+any other parameters.
+In the case of method calls, the Lua compiler translates
+\texttt{obj:method(args)} into an operation that first evaluates
+\texttt{obj}, uses \texttt{method} to index \texttt{obj},
+and then calls \texttt{obj.method} with the result of evaluating
+\texttt{obj} as the first argument, followed by the result of
+evaluating the argument list in the original expression.
+
+Typed Lua uses table types along with the type \texttt{self} to
+represent objects.
+We use the following example to discuss this feature:
 \begin{verbatim}
     local Shape = { x = 0.0, y = 0.0 }
-    const function Shape:move (dx:number, dy:number)
+    function Shape:move (dx:number, dy:number)
       self.x = self.x + dx
@@
 
-This example introduces some features that we must explain
-before we show the type of \texttt{Shape}.
-In Lua, the \texttt{:} syntactic sugar assigns a function to the
-field with a first parameter named \texttt{self}, plus any other
-parameters.
-Typed Lua does the same, and assigns the type \texttt{self} to the
-implicit parameter \texttt{self}.
-Typed Lua also adds the \texttt{const} annotation because the
-\texttt{const} qualifier is necessary for covariance among object types work.
-Under this circumstances, the type of \texttt{Shape} is an object of
-the following type:
+This example shows that Typed Lua assigns the type \texttt{self}
+to the implicit parameter \texttt{self}.
+The type \texttt{self} represents the type of the \emph{receiver}
+in method definitions and method calls.
+In this example, the refinement of table types allows Typed Lua
+to bind the type \texttt{self} with the same type of \texttt{Shape}:
 \begin{verbatim}
     { "x":number, "y":number,
-      const "move":(self, number, number) -> () }
-\end{verbatim}
-
-While \texttt{:} is syntactic sugar in plain Lua, Typed Lua uses it
+      "move":(self, number, number) -> () }
+\end{verbatim}
+
+While \texttt{:} is syntactic sugar in Lua, Typed Lua uses it
 to check method calls, binding any occurrence of the type \texttt{self}
@@
 Indexing a method and not immediately calling it with the correct
-receiver is a compile-time error:
-\begin{verbatim}
-    Shape.move(Shape, 10, 10)  -- Shape.x = 10 and Shape.y = 10
-    Shape:move(5, 20)          -- Shape.x = 5 and Shape.y = 20
-    local p = Shape.move       -- compile-time error
+receiver is a compile-time warning:
+\begin{verbatim}
+    Shape.move(Shape, 10, 10) --> Shape.x = 10 and Shape.y = 10
+    Shape:move(5, 20)         --> Shape.x = 5 and Shape.y = 20
+    local p = Shape.move      --> compile-time warning
 \end{verbatim}
@@
 Lua looks up in \texttt{t2} for any missing fields of \texttt{t1}.
+As we mentioned in Section \ref{sec:statistics},
 Lua programmers often use this mechanism to simulate classes.
 
-Typed Lua combines this mechanism along with interfaces to allow
-programmers to define classes in a similar fashion to the definition
-of modules, as we show in the following example:
+Typed Lua combines this mechanism along with type aliases and
+interfaces to allow programmers to define classes in a similar
+fashion to the definition of modules.
+We use the following example to discuss how Typed Lua type checks
+object-oriented code:
 \begin{verbatim}
@@
 
+In this example, we use an interface to define the type of the class \texttt{Shape}.
+We use the \texttt{const} annotation in the type of the methods
+because it is necessary for covariance among object types work.
+We use a double arrow instead of a single arrow because it is
+syntactic sugar for defining the type of methods,
+as it includes an implicit first type \texttt{self}.
+The double arrow can appear only inside the declaration of
+interfaces and userdata, which we will introduce in the next section.
+This interface definition exports the following object type to the class \texttt{Shape}:
+\begin{verbatim}
+    { "x":number, "y":number,
+      const "new":(self, number, number) -> (self),
+      const "move":(self, number, number) -> () }
+\end{verbatim}
+
 In Typed Lua, \texttt{setmetatable} is a strict primitive.
@@
 two operands.
-The first operand can be any table \texttt{t1}, but the second operand
-must be a table of the form \texttt{\{ \string_\string_index = t2 \}},
+The first operand can be any table \texttt{t1}, but
+the second operand must be a table of the form
+\texttt{\{ \string_\string_index = t2 \}},
 where \texttt{t2} can be any table.
-
 The reason for being so strict with \texttt{setmetatable} is because
@@
 
-In our example, we use \texttt{setmetatable} to define a constructor
+In the previous example, we use \texttt{setmetatable} to define a constructor
 \texttt{new} to the class \texttt{Shape}.
-The type of \texttt{new} is \texttt{(self, number, number) -> (self)},
-because we use \texttt{setmetatable} to declare \texttt{s} having
-the same type of \texttt{self}.
-
-In the same example, we also declare the interface \texttt{Shape}.
-Notice that we use double arrow instead of single arrow inside the
-definition of the interface \texttt{Shape}.
-The double arrow is syntactic sugar for defining the type of methods,
-as it includes an implicit first parameter of type \texttt{self}.
-The double arrow can appear only inside the declaration of
-interfaces and userdata, which we will introduce in the next section.
-The interface definition allows us to export the type of the class \texttt{Shape},
-which is the same type that is given by the variable \texttt{Shape}:
-\begin{verbatim}
-    { "x":number, "y":number,
-      const "new":(self, number, number) -> (self),
-      const "move":(self, number, number) -> () }
-\end{verbatim}
-
-In Typed Lua, this style of classes definition allow us to use \texttt{require}
-for creating prototype objects that work as classes,
-along with a shortcut to the table type of the class.
-This allows us to use this shortcut on type annotations,
+Note that we use \texttt{setmetatable} inside \texttt{new} to declare
+the local \texttt{s} having the same type of \texttt{self}.
+
+Still in the previous example, we are using the refinement of table
+types to build the type of the variable \texttt{Shape},
+as it should work as our object prototype, matching the type
+that the interface declaration exports.
+This means that the local \texttt{Shape} has type
+\texttt{\{"x":number, "y":number\}} inside the definition of \texttt{new}.
+This also makes Typed Lua bind the type of the local \texttt{Shape}
+to the type \texttt{self} inside the definition of \texttt{new},
+allowing us to access fields \texttt{x} and \texttt{y}.
+After the definition of \texttt{new}, the local \texttt{Shape}
+has type
+\texttt{\{"x":number, "y":number, const "new":(self, number, number) -> (self)\}},
+which is the type that Typed Lua binds to the type \texttt{self}
+inside the definition of \texttt{move}.
+Finally, the local \texttt{Shape} has the same type of the
+interface declaration in the last \texttt{return} statement.
+
+In Typed Lua, the style of classes definition from the previous example
+allow us to use \texttt{require} for creating prototype objects that
+work as classes, along with an alias to the table type of the class.
+This allows us to use the exported alias in type annotations,
 as we show in the following example:
@@
 
-Typed Lua also uses \texttt{setmetatable} to simulate single inheritance,
-as we show in the following example:
-
+Typed Lua also uses \texttt{setmetatable} to simulate single inheritance.
+We use the following example to discuss single inheritance in Typed Lua:
 \begin{verbatim}
@@
 
-A call to \texttt{setmetatable} where \texttt{t1} is a table
-constructor assigned to a fresh local variable, and \texttt{t2}
-is not \texttt{self} gives this variable an open type.
-That is the reason why we can add new methods and override
-existing ones in \texttt{Circle}.
-
-In this example, we also call \texttt{setmetatable} to redefine \texttt{new}.
-Notice how we can call the constructor of \texttt{Shape} inside
+In this example, we use an interface to define the type of the class \texttt{Circle},
+but with a syntactical overhead to manually include new and
+overridden methods.
+This interface definition exports the following object type to the class \texttt{Circle}:
+\begin{verbatim}
+    { "x":number, "y":number, "radius":number,
+      const "new":(self, number, number, value) -> (self),
+      const "move":(self, number, number) -> (),
+      const "radius":() -> (number) }
+\end{verbatim}
+
+In the previous example, we use \texttt{setmetatable} in two different ways.
+In the first case, we use \texttt{setmetatable} to trigger the refinement
+of table types, as we need it to add new methods and override existing ones
+in \texttt{Circle}.
+This is possible because a call to \texttt{setmetatable},
+where \texttt{t1} is a table constructor assigned to a fresh local variable,
+and \texttt{t2} is not \texttt{self} gives this variable an open type.
+In the second case, we use \texttt{setmetatable} to redefine \texttt{new}.
+Note how we can call the constructor of \texttt{Shape} inside
 the overridden constructor.
@@
 
-In the same example, we also declare the interface \texttt{Circle},
-but with a syntactical overhead to manually include the new and
-overridden methods.
-The interface \texttt{Circle} exports the following type:
-\begin{verbatim}
-    { "x":number, "y":number, "radius":number,
-      const "new":(self, number, number, value) -> (self),
-      const "move":(self, number, number) -> (),
-      const "radius":() -> (number) }
-\end{verbatim}
-
-This table type is a subtype of the table type exported by the interface \texttt{Shape},
-and we can use both shortcuts on type annotations, as we show in the following example:
+The type of the class \texttt{Circle} is a subtype of the type of
+the class \texttt{Shape}, and we can use both exported aliases in type
+annotations, as we show in the following example:
 \begin{verbatim}
@@
     local circle4:Shape = circle2
-    print(circle2:area())                -- 314.15926535898
-    print(circle3:area())                -- compile-time error
+    print(circle2:area())             --> 314.15926535898
+    print(circle3:area())             --> compile-time warning
 \end{verbatim}
@@
 
-An important feature of Typed Lua is its interface between
-statically and dynamically typed modules.
-This means that the programmer can enjoy some of the benefits of
+Typed Lua allows programmers to create description files
+for exporting statically typed interfaces to dynamically typed modules.
+This means that programmers can have some of the benefits of
 static types even without converting existing Lua modules to Typed Lua,
@@
 
-The Typed Lua compiler also handles description files.
-The programmer can use description files for two purposes:
-export a statically typed interface for an existing Lua
-module, without converting it to Typed Lua,
-or export a statically typed interface for a Lua module
-that is written in C.
+Furthermore, Typed Lua also allows programmers to create description
+files for exporting statically typed interfaces for a Lua modules
+that are written in C.
 
@@
 \begin{align*}
-\textit{description} ::= & \; \textit{desclist}\\
-\textit{desclist} ::= & \; \textit{descitem} \; \{ \textit{descitem} \}\\
-\textit{descitem} ::= & \; \textit{typedid} \; | \;
+\textit{description} ::= & \;\; \textit{desclist}\\
+\textit{desclist} ::= & \;\; \textit{descitem} \; \{ \textit{descitem} \}\\
+\textit{descitem} ::= & \;\; \textit{typedid} \; | \;
   \textit{interface} \; | \;
@@
   \textit{typealias}\\
-\textit{typedid} ::= & \; \textit{id} \; \texttt{`:'} \; \textit{type}\\ 
-\textit{interface} ::= & \; \textbf{interface} \; \textit{typedec}\\
-\textit{userdata} ::= & \; \textbf{userdata} \; \textit{typedec}\\
-\textit{typedec} ::= & \; \textit{Name} \; { \; \textit{decitem} \; } \; \textbf{end}\\
-\textit{decitem} ::= & \; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
-\textit{idtype} ::= & \; \textit{type} \; | \; \textit{methodtype}\\
-\textit{idlist} ::= & \; \textit{id} \; \{ \; \texttt{`,'} \; \textit{id} \; \}\\
+\textit{typedid} ::= & \;\; [\textbf{const}] \; \textit{id} \; \texttt{`:'} \; \textit{type}\\ 
+\textit{interface} ::= & \;\; \textbf{interface} \; \textit{typedec}\\
+\textit{userdata} ::= & \;\; \textbf{userdata} \; \textit{typedec}\\
+\textit{typedec} ::= & \;\; \textit{Name} \; {\textit{decitem}} \; \textbf{end}\\
+\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
+\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
+\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
 \textit{id} ::= & \; [\textbf{const}] \; \textit{Name}\\
-\textit{typealias} ::= & \; \textbf{typealias} \; \textit{Name} \; \texttt{`='} \; \textit{type}
+\textit{typealias} ::= & \;\; \textbf{typealias} \; \textit{Name} \; \texttt{`='} \; \textit{type}
 \end{align*}
@@
 along with this table type.
-We can export a type name through the declaration of an interface,
-userdata, or type alias.
-As we saw, an interface declaration is just a shortcut to a table type
-that represents a record.
+We can export a type name through the declaration of either an interface,
+an userdata, or a type alias.
+As we mentioned in Section \ref{sec:alias}, a type alias declaration
+creates an alias to a more general type, while an interface declaration
+creates an alias to a table type that represents a record.
 An userdata declaration is similar to an interface declaration,
@@
 The Typed Lua compiler uses this brand to combine structural with
-nominal type checking, so two different userdata are not subtype
-of each other because they do not share the same brand.
-A type alias declaration is a way to create shortcuts to union types,
-function types, table types, and recursive types.
-
-The Typed Lua compiler searches for a description file when it cannot
-not find the respective Typed Lua file that is the argument of
-\texttt{require}.
+nominal type checking, so two userdata that export exactly the same members,
+are not subtype of each other, because they do not share the same brand.
 
@@
 This description file exports the type \texttt{md5\string_context}
-through the \texttt{userdata} declaration, and the table type
-that represents the module.
-Now we can use the Typed Lua compiler to check for type errors on
-our use of the library:
+through an \texttt{userdata} declaration and the table type that
+represents the type of the module.
+Now we can use the Typed Lua compiler to check for type errors in
+our use of the \texttt{lmd5} library:
 \begin{verbatim}
@@
     local y = x:clone()
-    local z = m.clone("foo")             -- compile-time error
-    print(x:digest() == m.digest(y))
-\end{verbatim}
-
-In this example there is a type error because the function \texttt{clone}
-expects a value of type \texttt{md5\string_context} instead of a value
-of type \texttt{string}.
+    local z = m.clone("foo")          --> compile-time warning
+    print(x:digest() == m.digest(y))  --> true
+\end{verbatim}
+
+This example shows that the Typed Lua compiler searches for a
+description file when it cannot not find the respective Typed Lua file
+that is the argument of \texttt{require}.
+In this example, the call to \texttt{require} assigns to the local
+\texttt{m} the table type that the description file of the \texttt{lmd5}
+library exports. 
+Thus, the compiler raises a compile-time warning in the fourth line,
+as the function \texttt{clone} expects a value of type
+\texttt{md5\string_context} instead of a value of type \texttt{string}.
 
@@
 typing of the Lua standard library inside Typed Lua.
-In Chapter \ref{chap:evaluation} we discuss the issues that we found
-while typing the Lua standard library.
+In Chapter \ref{chap:evaluation} we will discuss the issues that we
+found while typing the Lua standard library and other case studies,
+which include the \texttt{lmd5} library.

### 2015-01-23 doc/thesis/system.tex
Adding filter types to the text about types and subtyping

@@
 In this chapter we present the formalization of Typed Lua's type system.
-Besides its practical contribution, Typed Lua also has some interesting
+Besides its practical contributions, Typed Lua also has some interesting
 contributions to the field of optional type systems for scripting
@@
 & | \; \mu x.t & \textit{recursive types}\\
+& | \; \phi(t,t) & \textit{filter types}\\
 & | \; \pi_{i}^{x} & \textit{projection types}\\
@@
 the top type $\Value$, the dynamic type $\Any$, the type $\Self$,
-union types, function types, table types, recursive types, and
-projection types.
-Second-level types include the type $\Void$, variadic types,
-tuple types, and unions of tuple types.
+union types, function types, table types, recursive types,
+filter types, and projection types.
+Second-level types include tuple types and unions of tuple types.
+Tuple types include the type $\Void$, variadic types, and pair types.
 Types are ordered by a subtype relationship that we introduce
@@
 Typed Lua includes four base types: $\Boolean$, $\Integer$, $\Number$, and $\String$.
-The base types $\Boolean$ and $\String$ represent the values that during
-run-time Lua tags as \texttt{boolean} and \texttt{string}, respectively.
+The base types $\Boolean$ and $\String$ represent the values that
+Lua tags as \texttt{boolean} and \texttt{string} during run-time.
 Lua 5.3 introduced two internal representations to the tag \texttt{number}:
@@
 and the base type $\Integer$ to represent \texttt{integer} values.
-In the next section we will show that $\Integer$ is subtype of $\Number$.
+In the next section we will show that $\Integer$ is a subtype of $\Number$.
 This allows programmers to keep using \texttt{integer} values where
@@
 
-Second-level types are tuples of first-level types that can end
-with either an empty tuple or with a variadic type, and
-unions of tuple types.
+Second-level types are either tuple types or unions of tuple types.
+Tuple types are tuples of first-level types that can end with
+either an empty tuple or with a variadic type.
 Typed Lua needs second-level types because tuples are not first-class
@@
 For instance, $\{\Value:\Value\}_{closed}$ represents the type of a
-table that both indices and values can have any type.
+table in which both indices and values can have any type.
 Value types can be any first-level type, and can optionally include
@@
 to classify table types.
-The tag \emph{unique} represents tables that have no keys that do not
-inhabit one of the table's key types, and that have no alias.
+The tag \emph{unique} represents tables with no keys that do not
+inhabit one of the table's key types, and with no alias.
 In particular, the type of the table constructor has this tag.
@@
 
+Typed Lua includes filter types as a way to discriminate the type of local
+variables inside conditions.
+In Section \ref{sec:rules} we show in more detail how our type system
+uses them to formalize the \texttt{type} predicates that we mentioned
+in Section \ref{sec:unions}.
+
 Typed Lua includes projection types as a way to project
-unions of tuple types into unions of first-level types,
-as we mentioned in Section \ref{sec:unions}.
+unions of tuple types into unions of first-level types.
 In Section \ref{sec:rules} we show in more detail how our type system
 uses them as a mechanism for handling unions of tuple types,
-when they appear on the right-hand side of the declaration of local variables.
+when they appear on the right-hand side of the declaration of local variables,
+as we mentioned in Section \ref{sec:unions}.
 We also show how this feature allows our type system to constrain
@@
 to allow the interaction between statically and dynamically typed code.
-We explain the subtyping and consistent-subtyping rules along this section.
+We explain the subtyping and consistent-subtyping rules throughout this section.
 However, we focus the discussion on the definition of subtyping because,
@@
 to differ only in the way they handle the dynamic type.
-Thus, we define the subtyping rules using an algorithmic
-way that is close to the implementation, as this way allows us to use
+Thus, we define the subtyping rules using an algorithmic approach
+that is close to the implementation, as this approach allows us to use
 subtyping to easily formalize consistent-subtyping.
@@
 
-The simplest tuple type is the type $\Void$ that is the type of an
-empty tuple and it is subtype only of itself.
-
-The subtyping rule for pair types is standard, but the subtyping rules
-for variadic types are not so obvious:
-\[
-\begin{array}{c}
+The type $\Void$ is a subtype of itself and of a variadic type:
+\[
+\begin{array}{c}
+\mylabel{S-VOID}\\
+\senv \vdash \Void \subtype t{*}
+\end{array}
+\]
+
+A variadic type $t{*}$ represents a sequence of values of type
+$t \cup \Nil$, and the rule \textsc{S-VOID} handles the case where
+a given sequence is empty.
+
+The subtyping rule for pair types is the standard covariant rule:
+\[
 \begin{array}{c}
@@
 \end{array}
-\;
+\]
+
+The subtyping rules for variadic types are not so obvious:
+\[
+\begin{array}{c}
 \begin{array}{c}
 \mylabel{S-VARARG1}\\
+\senv \vdash \Nil{*} \subtype \Void
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-VARARG2}\\
 \dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2} \cup \Nil}
@@
 \end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-VARARG2}\\
+\;
+\begin{array}{c}
+\mylabel{S-VARARG3}\\
 \dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2}}
@@
 \begin{array}{c}
-\mylabel{S-VARARG3}\\
+\mylabel{S-VARARG4}\\
 \dfrac{\senv \vdash t_{1} \subtype t_{2} \cup \Nil}
@@
 \begin{array}{c}
-\mylabel{S-VARARG4}\\
+\mylabel{S-VARARG5}\\
 \dfrac{\senv \vdash t_{1}{*} \subtype t_{2} \times \Void \;\;\;
@@
 \begin{array}{c}
-\mylabel{S-VARARG5}\\
+\mylabel{S-VARARG6}\\
 \dfrac{\senv \vdash t_{1} \times \Void \subtype t_{2}{*} \;\;\;
@@
 
-The subtyping rules for pair and variadic types show that they are covariant.
-While we need just one subtyping rule for pair types,
-we need five different subtyping rules for variadic types
+We need six different subtyping rules for variadic types
 to handle all the cases where they can appear.
-The rule \textsc{S-VARARG1} handles the case that both tuple types end
-with variadic types, while the other rules handle the cases that only
-one tuple type ends with a variadic type.
-Note that the case where both tuple types end with the type $\Void$ does
-not require any special rule.
-The rule \textsc{S-VARARG1} shows that $t_{1}{*}$ is subtype of $t_{2}{*}$
-if $t_{1} \cup \Nil$ is subtype of $t_{2} \cup \Nil$.
-This rule explicit includes $\Nil$ in both sides because otherwise
+The rule \textsc{S-VARARG1} is a special rule for handling the
+case where we give a sequence of $\Nil$ to the empty tuple.
+The rule \textsc{S-VARARG2} handles the case where both tuple types end
+with variadic types, and shows that $t_{1}{*}$ is a subtype of $t_{2}{*}$
+if $t_{1} \cup \Nil$ is a subtype of $t_{2} \cup \Nil$.
+This rule explicitly includes $\Nil$ in both sides because otherwise
 $\Nil{*}$ would not be a subtype of several other variadic types.
@@
 as $\Nil \not\subtype \Number$.
-In the next section we will show that we use these subtyping relations,
+The other rules handle the cases where only one tuple type ends with a variadic type.
+Note that the case where both tuple types end with the type $\Void$ does
+not require any special rule.
+In the next section we will show that we use these subtyping rules,
 along with the types $\Value$ and $\Nil$, to make our type system reflect
-the semantics of Lua on discarding extra parameters, and
+the semantics of Lua on discarding extra parameters and
 replacing missing parameters.
@@
 The rule \textsc{S-TABLE1} allows width subtyping between \emph{closed}
-table types, and introduced the auxiliary relation $\subtype_{c}$ to
+table types, and introduces the auxiliary relation $\subtype_{c}$ to
 handle depth subtyping on the type of the values stored in the table fields.
@@
 would not type check, as the type of the table constructor would not
-be subtype of the type in the annotation.
+be a subtype of the type in the annotation.
 More precisely,
@@
 
-Projection types are subtypes only of themselfs and of $\Value$.
+Filter types are subtypes only of themselves and of $\Value$.
+More precisey, a filter type $\phi(t_{1},t_{2})$ is a subtype of
+the same filter type $\phi(t_{1},t_{2})$, which shares the same
+types $t_{1}$ and $t_{2}$, and it is also a subtype of $\Value$.
+
+Projection types are subtypes only of themselves and of $\Value$.
 More precisely, a projection type $\pi_{i}^{x}$ is a subtype of the
@@
 the consistent-subtyping relation $\senv \vdash t_{1} \lesssim t_{2}$.
-Like in the subtyping relation, $\senv$ is also a set of pairs of
+As in the subtyping relation, $\senv$ is also a set of pairs of
 recursion variables.
@@
 If we had set the type $\Any$ as both bottom and top types of our
-subtyping relation, then any type $t_{1}$ would be subtype of
+subtyping relation, then any type $t_{1}$ would be a subtype of
 any other type $t_{2}$.
@@
 In the implementation of Typed Lua we also use consistent-subtyping to
-normalize and simplify union types, though we let them free in the
-formalization.
+normalize and simplify union types, though we let union types free in
+the formalization.
 For instance, the union type \texttt{boolean|any} results in the
@@
 e{:}n(el)_{m} \; | \;
-{...}_{m}\\
+{...}_{m} \; | \;
+e^{x}\\
 m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (){:}r \; fb \; | \;

### 2015-01-24 doc/thesis/system.tex
Updating typing rules to use the projections environment

@@
 We introduced \emph{unique} table types to avoid this limitation,
-as they represent the type of the tables that have no keys that do not
-inhabit one of the table's key types, and that have no alias.
+as they represent the type of tables with no keys that do not
+inhabit one of the table's key types, and with no alias.
 In particular, this is the case of the table constructor.
@@
 e{:}n(el)_{m} \; | \;
-{...}_{m} \; | \;
-e^{x}\\
+{...}_{m}\\
 m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (){:}r \; fb \; | \;
@@
 though expressions and statements can modify the environment.
-We need two environments because $\env$ maps variables to types
+We need two environments because $\env$ maps variables to types,
 while $\penv$ maps projection variables to unions of tuple types.
@@
 $\env_{1}$ with the variable $n$ that maps to type $t$.
-We use $\penv[\pi \mapsto r]$ for meaning that we extendend the environment
-$\penv$ with the projection $\pi$ that maps to the union of tuple types $r$.
+We use $\penv[x \mapsto r]$ for meaning that we extendend the environment
+$\penv$ with the projection variable $x$ that maps to the union of tuple types $r$.
 
@@
 \begin{array}{c}
-\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
+\dfrac{\env_{1}, \penv \vdash el:r_{1}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
        r_{1} \lesssim r_{2}}
-      {\env_{1} \vdash \vec{l} = el:\env_{3}}
+      {\env_{1}, \penv \vdash \vec{l} = el,\env_{3}}
 \end{array}
@@
 \mylabel{T-LHSLIST}\\
-\dfrac{\env \vdash l_{k}:t_{k}, \env_{k} \;\;\;
+\dfrac{\env, \penv \vdash l_{k}:t_{k}, \env_{k} \;\;\;
        \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
        n = |\;\vec{l}\;|}
-      {\env \vdash \vec{l}:t_{1} \times ... \times t_{n} \times \Value{*}, \env_{m}}
+      {\env, \penv \vdash \vec{l}:t_{1} \times ... \times t_{n} \times \Value{*}, \env_{m}}
 \end{array}
@@
 left-hand side value $l_{k}$, as they can perform different changes
-to the environment.
+in the environment.
 Then, it takes each type $t_{k}$ and builds the tuple type for
@@
 \mylabel{T-EXPLIST2}\\
-\dfrac{\env \vdash e_{k}:t_{k}, \env_{k} \;\;\;
+\dfrac{\env, \penv \vdash e_{k}:t_{k}, \env_{k} \;\;\;
        \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
        n = |\;\vec{e}\;|}
-      {\env \vdash \vec{e}:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{m}}
+      {\env, \penv \vdash \vec{e}:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{m}}
 \end{array}
@@
 \mylabel{T-EXPLIST3}\\
-\dfrac{\env \vdash me:t_{1} \times ... \times t_{n} \times \Void, \env_{1}}
-      {\env \vdash me:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{1}}
+\dfrac{\env, \penv \vdash me:t_{1} \times ... \times t_{n} \times \Void, \env_{1}}
+      {\env, \penv \vdash me:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{1}}
 \end{array}
@@
 \mylabel{T-EXPLIST4}\\
-\dfrac{\env \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{1}}
-      {\env \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{1}}
+\dfrac{\env, \penv \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{1}}
+      {\env, \penv \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{1}}
 \end{array}
@@
 \]
-assuming that $x$, $y$, $z$, and $f$ are in the environment with
-types $\Integer$, $\String$, $\String \cup \Nil$, and
+
+In this example, we are assuming that $x$, $y$, $z$, and $f$ are in
+the environment with types $\Integer$, $\String$, $\String \cup \Nil$, and
 $\Void \rightarrow \Integer \times \String \times \Void$, respectively.
-It type checks because
+Thus, the example type checks because
 \[
@@
 
-Multiple assignments also appear in function applications, and the
-rule \textsc{T-APPLY1} uses the same behavior from multiple assignments
-to drop extra parameters and replace missing parameters:
+Multiple assignments also appear in function applications.
+The rule \textsc{T-APPLY1} handles the case where function applications
+are expressions that return multiple values:
 \[
@@
 \mylabel{T-APPLY1}\\
-\dfrac{\env_{1} \vdash e:r_{1} \rightarrow r_{2}, \env_{2} \;\;\;
-       \env_{2} \vdash el:r_{3}, \env_{3} \;\;\;
+\dfrac{\env_{1}, \penv \vdash e:r_{1} \rightarrow r_{2}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash el:r_{3}, \env_{3} \;\;\;
        r_{3} \lesssim r_{1}}
-      {\env_{1} \vdash e(el)_{m}:r_{2}, \env_{3}}
-\end{array}
-\]
-
-Assuming that $x$ and $f$ are local variables that are in the environment
-with types $\Integer \cup \Nil$ and
+      {\env_{1}, \penv \vdash e(el)_{m}:r_{2}, \env_{3}}
+\end{array}
+\]
+
+We also use the rule \textsc{T-APPLY1} as the base case for the rules
+that handle the cases where function applications are either statements
+or expressions that return only one value:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-STMAPPLY1}\\
+\dfrac{\env_{1}, \penv \vdash e(el)_{m}:r, \env_{2}}
+      {\env_{1}, \penv \vdash e(el)_{s},\env_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{T-EXPAPPLY1}\\
+\dfrac{\env_{1}, \penv \vdash e(el)_{m}:r, \env_{2}}
+      {\env_{1}, \penv \vdash e(el)_{e}:first(r), \env_{2}}
+\end{array}
+\end{array}
+\]
+
+The rule \textsc{T-STMAPPLY1} discards the returned values,
+while the rule \textsc{T-EXPAPPLY1} uses the predicate \emph{first} to
+ensure that only one value is returned.
+We can use pattern matching to the define the recursive predicate
+\emph{first} as follows:
+\begin{align*}
+first(\Void) & = \Nil\\
+first(t{*}) & = t \cup \Nil\\
+first(t \times s) & = t\\
+first(s_{1} \sqcup s_{2}) & = first(s_{1}) \cup first(s_{2})
+\end{align*}
+
+Assuming that $f$ is a local function in the environment, and that $f$ has type
 $\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*} \rightarrow \Integer{*}$,
-respectively.
-The program
-\[
-x = f(``foo")_{m}
-\]
-type checks through the rule \textsc{T-APPLY1} because
+the function call
+\[
+f(``foo")_{s}
+\]
+type checks through the rules \textsc{T-STMAPPLY1} and \textsc{T-APPLY1}, because
 \[
@@
 \]
-and the program
+and the function call
 \[
@@
 \]
-also type checks through the rule \textsc{T-APPLY1} because
+also type checks through the rules \textsc{T-STMAPPLY1} and \textsc{T-APPLY1}, because
 \[
@@
 
-However, we also expect from a type system that it catches arity mismatch.
-For this reason, our type system also has different typing rules for
-expression lists that appear in function applications.
-These rules do not include $\Nil{*}$ when the tuple type does not end
-in a variadic type.
-The rule \textsc{T-APPLY1-STRICT} use these rules to type check
-expression lists, and it also uses the predicate $adjust$ to replace
-any missing value with $\Nil$:
-\[
-\begin{array}{c}
-\mylabel{T-APPLY1}\\
-\dfrac{\env_{1} \vdash e:r_{1} \rightarrow r_{2}, \env_{2} \;\;\;
-       \env_{2} \vdash el:r_{3}, \env_{3} \;\;\;
-       r_{4} = adjust(r_{3}, r_{1}) \;\;\;
-       r_{4} \lesssim r_{1}}
-      {\env_{1} \vdash e(el)_{m}:r_{2}, \env_{3}}
-\end{array}
-\]
-
-We can use pattern matching to the define the recursive predicate
-\emph{adjust} as follows:
-\begin{align*}
-adjust(\Void, t{*}) & = \Nil \times \Void\\
-adjust(\Void, t \times s) & = \Nil \times adjust(\Void,s)\\
-adjust(t_{1} \times s_{1}, t_{2} \times s_{2}) & = t_{1} \times adjust(s1,s2)\\
-adjust(s_{1}, s_{2}) & = s_{1}
-\end{align*}
-
-Now, assuming that $f$ has type
+Our type system also catches arity mismatch.
+For instance, assuming that $f$ has type
 $\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void \rightarrow \Integer{*}$,
-the program
-\[
-x = f(``foo")_{m}
-\]
-type checks through the rule \textsc{T-APPLY1-STRICT} because
-\[
-``foo" \times \Nil \times \Nil \times \Void \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
-\]
-but the program
+the function call
+\[
+f(``foo")_{s}
+\]
+type checks through the rules \textsc{T-STMAPPLY1} and \textsc{T-APPLY1}, because
+\[
+``foo" \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
+\]
+but the function call
 \[
@@
 \]
-does not type check through the rule \textsc{T-APPLY1-STRICT} because
-\[
-``foo" \times 1 \times 2 \times 3 \times \Void \not\lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
-\]
-
-The same happens with the rules that type check the \textbf{return}
-statement.
-The rule \textsc{T-RETURN} uses $\Nil{*}$ to replace eventual missing
-values, while the rule \textsc{T-RETURN-STRICT} uses the predicate
-\emph{adjust} to replace values that are certainly missing:
-\[
-\begin{array}{c}
+does not type check through the rules \textsc{T-STMAPPLY1} and \textsc{T-APPLY1}, because
+\[
+``foo" \times 1 \times 2 \times 3 \times \Nil{*} \not\lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
+\]
+
+When our type system type checks an expression list,
+it always include $\Nil{*}$ in the end of the type of this expression list,
+if the type does not end in a variadic type.
+Our type system has this behavior because $\Nil{*}$ along with subtyping
+and consistent-subtyping use $\Nil$ to replace missing values.
+This behavior preserves the semantics of Lua on replacing missing values,
+as it is necessary when we ommit optional parameters in a function call,
+like the previous example showed.
+
+Using $\Nil{*}$ in the end of the type of expression lists also allows
+our type system to catch arity mismatch in function calls where
+the type of the function does not include optional parameters.
+For instance, assuming that $f$ has type
+$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Void$,
+the function call
+\[
+f(1)_{s}
+\]
+does not type check through the rules \textsc{T-STMAPPLY1} and \textsc{T-APPLY1}, because
+\[
+1 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Void
+\]
+and the function call
+\[
+x = f(1,2,3)_{m}
+\]
+also does not type check through the rules \textsc{T-STMAPPLY1} and \textsc{T-APPLY1}, because
+\[
+1 \times 2 \times 3 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Void
+\]
+
+The same behavior happens with the rule that type checks the \textbf{return}
+statement:
+\[
 \begin{array}{c}
 \mylabel{T-RETURN}\\
-\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
+\dfrac{\env_{1}, \penv \vdash el:r_{1}, \env_{2} \;\;\;
        \env_{2}(\ret) = r_{2} \;\;\;
        r_{1} \lesssim r_{2}}
-      {\env_{1} \vdash \mathbf{return} \; el:\env_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-RETURN-STRICT}\\
-\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
-       \env_{2}(\ret) = r_{2} \;\;\;
-       r_{3} = adjust(r_{1}, r_{2}) \;\;\;
-       r_{3} \lesssim r_{2}}
-      {\env_{1} \vdash \mathbf{return} \; el:\env_{2}}
-\end{array}
+      {\env_{1}, \penv \vdash \mathbf{return} \; el,\env_{2}}
 \end{array}
@@
 \mylabel{T-CONSTRUCTOR1}\\
-\env \vdash \{\}:\{\}_{unique}, \env
+\env, \penv \vdash \{\}:\{\}_{unique}, \env
 \end{array}
@@
 We did not include in the abstract syntax a table constructor
-that uses a list of expressions because we can write it using the first form.
+that uses a list of expressions, because we can write it using the first form.
 For instance, we can write the table constructor
@@
 We chose to not include the latter form because its type checking rule is similar
-to the type checking rule of the first form, and just controls the type of
+to the type checking rule of the first form, and it just controls the type of
 the omitted indices.
-For the same reason we did not include table constructors with
+For the same reason, we did not include table constructors with
 multiple expressions, which can return a tuple type that does not end
@@
 \mylabel{T-CONSTRUCTOR2}\\
-\dfrac{\env_{i} \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
+\dfrac{\env_{i}, \penv \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
        n = |\;\vec{p}\;| \;\;\;
        t = wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique})}
-      {\env_{1} \vdash \{\;\vec{p}\;\}:t, \env_{n+1}}
+      {\env_{1}, \penv \vdash \{\;\vec{p}\;\}:t, \env_{n+1}}
 \end{array}
@@
 The type checking of the table fields would be straightforward
-if our table types allowed any first-level type in the keys.
+if our table types allowed any first-level type in the type of the keys.
 The rules \textsc{T-FIELD1}, \textsc{T-FIELD2}, and \textsc{T-FIELD3}
@@
 \mylabel{T-FIELD1}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+\dfrac{\env_{1}, \penv \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3} \;\;\;
        t_{1} \subtype \Boolean}
-      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
+      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
 \end{array}
@@
 \mylabel{T-FIELD2}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+\dfrac{\env_{1}, \penv \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3} \;\;\;
        t_{1} \subtype \Number}
-      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
+      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
 \end{array}
@@
 \mylabel{T-FIELD3}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
+\dfrac{\env_{1}, \penv \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3} \;\;\;
        t_{1} \subtype \String}
-      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
-\end{array}
-\end{array}
-\]
+      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
+\end{array}
+\end{array}
+\]
+
+These rules use the predicate \emph{close} in the type of the value
+because our type system do not allow \emph{open} and \emph{unique}
+table types to appear in the type of the values.
+The predicate \emph{close} promotes all table types to \emph{closed}.
+We made this restriction because our type system does not keep track
+of aliases to table fields.
+This means that allowing \emph{open} and \emph{unique} table types
+to appear in the type of the values would allow the creation of
+unsafe aliases.
 
@@
 \mylabel{T-FIELD4}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3}}
-      {\env_{1} \vdash [e_{1}] = e_{2}: \Value{:}close(t_{2}), \env_{3}}
-\end{array}
-\]
-
-The rule \textsc{T-CONSTRUCTOR2} along with the rules for table fields
-assign the type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ to
-the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$.
+\dfrac{\env_{1}, \penv \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3}}
+      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: \Value{:}close(t_{2}), \env_{3}}
+\end{array}
+\]
+
+The rule \textsc{T-FIELD4} shows that our type system uses the type $\Value$
+as the type of the key whenever a key has a type that is neither a
+subtype of $\Boolean$, nor a subtype of $\Number$, nor a subtype of $\String$.
+
+As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
+has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ through the rules
+\textsc{T-FIELD3} and \textsc{T-CONSTRUCTOR2}.
 
@@
 \dfrac{\env_{1}({...}) = t}
-      {\env_{1} \vdash \{{...}_{m}\}:\{\Integer{:}t \cup \Nil\}_{unique}, \env_{1}}
+      {\env_{1}, \penv \vdash \{{...}_{m}\}:\{\Integer{:}t \cup \Nil\}_{unique}, \env_{1}}
 \end{array}
@@
 \dfrac{\begin{array}{c}
-       \env_{i} \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
+       \env_{i}, \penv \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
        n = |\;\vec{p}\;| \;\;\;
@@
        \end{array}}
-      {\env_{1} \vdash \{\;\vec{p},\;{...}_{m}\;\}:t_{t}, \env_{n+1}}
+      {\env_{1}, \penv \vdash \{\;\vec{p},\;{...}_{m}\;\}:t_{t}, \env_{n+1}}
 \end{array}
@@
 an array.
-If we assume that the $...$ is in the environment and has type $\String$,
-the rule \textsc{T-CONSTRUCTOR3} gives the type $\{\Integer:\String \cup \Nil\}_{unique}$
-to the table constructor $\{{...}_{m}\}$.
+If we assume that $...$ is in the environment and has type $\String$,
+the table constructor $\{{...}_{m}\}$ has type $\{\Integer:\String \cup \Nil\}_{unique}$
+through the rule \textsc{T-CONSTRUCTOR3}.
 
@@
 the table constructor as a record that includes an array part.
-If we assume that the $...$ is in the environment and has type $\String$,
-the rule \textsc{T-CONSTRUCTOR4} gives the type
-$\{``x":``foo", ``y":\mathbf{true}, \Integer:\String \cup \Nil\}_{unique}$
-to the table constructor $\{[``x"] = ``foo", [``y"] = \mathbf{true}, {...}_{m}\}$.
-As another example, the rule \textsc{T-CONSTRUCTOR4} fails to type check
+If we assume that $...$ is in the environment and has type $\String$,
+the table constructor $\{[``x"] = ``foo", [``y"] = \mathbf{true}, {...}_{m}\}$
+has type $\{``x":``foo", ``y":\mathbf{true}, \Integer:\String \cup \Nil\}_{unique}$
+through the rule \textsc{T-CONSTRUCTOR4}.
+As another example, the rule \textsc{T-CONSTRUCTOR4} does not type check
 the table constructor $\{ [1] = ``foo", [10] = \mathbf{true}, {...}_{m} \}$,
@@
 unusual feature of our type system: the refinement of table types.
-The rule \textsc{T-REFINE} allow us to add new keys to \emph{open}
+The rule \textsc{T-REFINE} allows adding new keys to \emph{open}
 and \emph{unique} table types:
@@
        \env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique}\\
-       \env_{1} \vdash c:k_{n+1}, \env_{2} \;\;\;
+       \env_{1}, \penv \vdash c:k_{n+1}, \env_{2} \;\;\;
        \not \exists i \in 1..n \; t_{1} \lesssim k_{i} \;\;\;
@@
        \end{array}}
-      {\env_{1} \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, k_{n+1}{:}v_{n+1}\}_{open|unique}]}
+      {\env_{1}, \penv \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, k_{n+1}{:}v_{n+1}\}_{open|unique}]}
 \end{array}
@@
 uses field assignment to add fields $``x"$ and $``y"$ to $\string_ENV$.
-In all the examples we assume that $\string_ENV$ is in the environment and
-has type $\{\}_{open}$.
+In this example and in the next examples we assume that
+$\string_ENV$ is in the environment and has type $\{\}_{open}$.
 Therefore, after these field assignments $\string_ENV$ has type
@@
 though we are trying to refine to $\{``x":\{\}_{unique}\}_{open}$.
-We made this restriction because we still need to prove that the refinement
-of table types is safe.
+As in the type of the table constructor, the refinement of
+table types also does not allow the inclusion of values that
+have \emph{open} and \emph{unique} table types.
 
@@
 
-The next example does not type check because it tries to add
+However, the next example does not type check because it tries to add
 the same field to $\string_ENV$, but with different types:
@@
 Besides allowing the refinement of table types through field assignment,
-Typed Lua also keeps track whether it is safe or not safe to change the
-type of a table.
+Typed Lua also keeps track whether it is safe to change the type of a table.
 For instance,
@@
 \end{center}
-does not type check because aliasing $a$ produces the type
-$\{``foo":\Integer\}_{closed}$ that is not subtype of
+does not type check, as aliasing $a$ produces the type
+$\{``foo":\Integer\}_{closed}$ that is not a subtype of
 $\{``foo":\Integer \cup \Nil\}_{closed}$, the type of $b$.
@@
 \dfrac{\env_{1}(n) = t}
-      {\env_{1} \vdash n_{e}:close(t), \env_{1}[n \mapsto open(t)]}
+      {\env_{1}, \penv \vdash n_{e}:close(t), \env_{1}[n \mapsto open(t)]}
 \end{array}
@@
 \emph{unique} to \emph{open}, as \emph{unique} table types only
-denote the type of table types that have no alias.
+denote the type of tables with no alias.
 
@@
 \end{center}
-does not type check because we cannot add the field $``x"$ to $b$,
-as its type is \emph{closed}, and thus does not allow changing the
-value that is stored in the field $``x"$ of local $a$.
+does not type check, as aliasing $a$ produces the type $\{\}_{closed}$
+that is not a subtype of $\{\}_{open}$, the type of $b$.
+Our type system has this behavior to warn programmers about
+potential unsafe behaviors after this kind of alias.
+In this example, it is unsafe to add the field $``x"$ to $b$,
+as it changes the value that is stored in the field $``x"$ of local $a$.
 
@@
 \dfrac{\env_{1}(n) = t}
-      {\env_{1} \vdash n_{l}:close(t), \env_{1}[n \mapsto close(t)]}
+      {\env_{1}, \penv \vdash n_{l}:close(t), \env_{1}[n \mapsto close(t)]}
 \end{array}
@@
 \dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\} \;\;\;
-       \env_{1} \vdash e_{2}:t, \env_{2} \;\;\;
+       \env_{1}, \penv \vdash e_{2}:t, \env_{2} \;\;\;
        \exists i \in 1{..}n \; t \lesssim k_{i}}
-      {\env_{1} \vdash n[e_{2}]:v_{i}, \env_{2}}
+      {\env_{1}, \penv \vdash n[e_{2}]:v_{i}, \env_{2}}
 \end{array}
@@
 \mylabel{T-INDEX2}\\
-\dfrac{\env_{1} \vdash e_{1}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t, \env_{3} \;\;\;
+\dfrac{\env_{1}, \penv \vdash e_{1}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash e_{2}:t, \env_{3} \;\;\;
        \exists i \in 1{..}n \; t \lesssim k_{i}}
-      {\env_{1} \vdash e_{1}[e_{2}]:v_{i}, \env_{3}}
+      {\env_{1}, \penv \vdash e_{1}[e_{2}]:v_{i}, \env_{3}}
 \end{array}
@@
 \dfrac{\env_{1}(n) \subtype t \;\;\;
-       \env_{1}[n \mapsto t] \vdash n_{e}:t_{1}, \env_{2}}
-      {\env_{1} \vdash {<}t{>} \; n:t_{1}, \env_{2}}
-\end{array}
-\]
-
-The rule \textsc{T-CAST} allow us to type check the following example:
+       \env_{1}[n \mapsto t], \penv \vdash n_{e}:t_{1}, \env_{2}}
+      {\env_{1}, \penv \vdash {<}t{>} \; n:t_{1}, \env_{2}}
+\end{array}
+\]
+
+The rule \textsc{T-CAST} allows our type system to type check the following example:
 \begin{center}
@@
 table types before we type check another scope.
-The rule \textsc{T-FUNCTION3} shows this case:
+The rule \textsc{T-FUNCTION3} illustrates this case:
 \[
@@
 \mylabel{T-FUNCTION3}\\
-\dfrac{closeall(\env_{1}[\vec{n} \mapsto \vec{t}, \ret \mapsto r]) \vdash s:\env_{2}}
-      {\env_{1} \vdash \mathbf{fun} \; (\vec{n{:}t}){:}r \; s:\vec{t} \rightarrow r, closeset(\env_{1}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))}
+\dfrac{closeall(\env_{1}[\vec{n} \mapsto \vec{t}, \ret \mapsto r]), \penv \vdash s, \env_{2}}
+      {\env_{1}, \penv \vdash \mathbf{fun} \; (\vec{n{:}t}){:}r \; s:\vec{t} \rightarrow r, closeset(\env_{1}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))}
 \end{array}
@@
 programmers often use it.
+The rule \textsc{T-METHOD3} illustrates how our type system type checks
+method definitions:
 \[
@@
        \env_{1}(n_{1}) = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{unique|open} \;\;\;
-       \env_{1} \vdash n_{2} : l \;\;\;
+       \env_{1}, \penv \vdash n_{2} : l \;\;\;
        \not \exists i \in 1..n \; l \lesssim k_{i}\\
-       closeall(\env_{1}[self \mapsto \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{closed}, \vec{n} \mapsto \vec{t}, \ret \mapsto r]) \vdash s:\env_{2}\\
+       closeall(\env_{1}[self \mapsto \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{closed}, \vec{n} \mapsto \vec{t}, \ret \mapsto r]), \penv \vdash s, \env_{2}\\
        t_{m} = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}, l{:}\Const \; \Self \times \vec{t} \rightarrow r\}_{unique|open}
@@
       {\begin{array}{c}
-       \env_{1} \vdash \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; s:\\
+       \env_{1}, \penv \vdash \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; s,\\
        closeset(\env_{1}[n_{1} \mapsto t_{m}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))
@@
 The rule \textsc{T-METHOD3} allow our type system to type check the
-following example:
+following example, which is similar to the definition of the class
+\texttt{Shape} from Section \ref{sec:oop}:
 \begin{center}
@@
 \{ & ``x":\Number, ``y":\Number,\\
-   & ``new":\Self \times \Number \times \Number \rightarrow \Self,\\
-   & ``move":\Self \times \Number \times \Number \rightarrow \Void \}_{open}
+   & \Const \; ``new":\Self \times \Number \times \Number \rightarrow \Self,\\
+   & \Const \; ``move":\Self \times \Number \times \Number \rightarrow \Void \}_{open}
 \end{align*}
@@
 to initialize the local $s$ with the type of $\Self$.
-The rule \textsc{T-SETMETATABLE} express this idea:
-\[
-\begin{array}{c}
-\mylabel{T-SETMETATABLE}\\
-\dfrac{\env_{1} \vdash e:t, \env_{2} \;\;\;
+The rule \textsc{T-SETMETATABLE1} express this idea:
+\[
+\begin{array}{c}
+\mylabel{T-SETMETATABLE1}\\
+\dfrac{\env_{1}, \penv \vdash e:t, \env_{2} \;\;\;
        t \subtype \{\}_{open}}
-      {\env_{1} \vdash setmetatable(\{\}, \{[``\string_\string_index"] = e\}):close(t), \env_{2}}
+      {\env_{1}, \penv \vdash setmetatable(\{\}, \{[``\string_\string_index"] = e\}):close(t), \env_{2}}
 \end{array}
@@
 of this class and call its methods.
-The next example assumes the object $o$ is in the environment as
-has the type of the class we just defined:
-\begin{center}
-\begin{tabular}{l}
-$o{:}move(10, 10)_{s}$
-\end{tabular}
-\end{center}
-
-This method call type checks through the rule \textsc{T-INVOKESTM1}.
-This rule uses the inference rule \textsc{T-INVOKE1} and discards the results.
-We defined the rule \textsc{T-INVOKE1} as follows:
+The rule \textsc{T-INVOKE1} handles the case where method calls
+are expressions that return multiple values:
 \[
@@
 \mylabel{T-INVOKE1}\\
-\dfrac{\env_{1} \vdash e:t, \env_{2} \;\;\;
-       \env_{2} \vdash e[n]:p_{1} \rightarrow r, \env_{3} \;\;\;
-       \env_{3} \vdash el:p_{2}, \env_{4} \;\;\;
+\dfrac{\env_{1}, \penv \vdash e:t, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash e[n]:p_{1} \rightarrow r, \env_{3} \;\;\;
+       \env_{3}, \penv \vdash el:p_{2}, \env_{4} \;\;\;
        \Self \times p_{2} \lesssim p_{1}}
-      {\env_{1} \vdash e{:}n(el)_{m}:[\Self \mapsto t]r, \env_{4}}
-\end{array}
-\]
-
-Typed Lua includes four typing rules to handle the $\mathbf{or}$
+      {\env_{1}, \penv \vdash e{:}n(el)_{m}:[\Self \mapsto t]r, \env_{4}}
+\end{array}
+\]
+
+We also use the rule \textsc{T-INVOKE1} as the base case for the rules
+that handle the cases where method calls are either statements
+or expressions that return only one value:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-STMINVOKE1}\\
+\dfrac{\env_{1}, \penv \vdash e{:}n(el)_{m}:r, \env_{2}}
+      {\env_{1}, \penv \vdash e{:}n(el)_{s},\env_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{T-EXPINVOKE1}\\
+\dfrac{\env_{1}, \penv \vdash e{:}n(el)_{m}:r, \env_{2}}
+      {\env_{1}, \penv \vdash e{:}n(el)_{e}:first(r), \env_{2}}
+\end{array}
+\end{array}
+\]
+
+The rule \textsc{T-STMINVOKE1} discards the returned values,
+while the rule \textsc{T-EXPINVOKE1} uses the predicate \emph{first} to
+ensure that only one value is returned.
+
+As an example, assuming that the object $o$ is in the environment and
+has the type of the class from the previous example, the method call
+\[
+o{:}move(10, 10)_{s}
+\]
+type checks through the rules \textsc{T-STMINVOKE1} and \textsc{T-INVOKE1},
+because
+\[
+\Self \times 10 \times 10 \times \Nil{*} \lesssim \Self \times \Number \times \Number \times \Void
+\]
+
+Typed Lua includes five typing rules for handling the $\mathbf{or}$
 logical operator and its common idioms:
@@
 \mylabel{T-OR1}\\
-\dfrac{\env_{1} \vdash e_{1}:t, \env_{2} \;\;\;
+\dfrac{\env_{1}, \penv \vdash e_{1}:t, \env_{2} \;\;\;
        \Nil \not\lesssim t \;\;\;
        \False \not\lesssim t}
-      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{2}}
+      {\env_{1}, \penv \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{2}}
 \end{array}
@@
 \mylabel{T-OR2}\\
-\dfrac{\env_{1} \vdash e_{1}:\Nil, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t, \env_{3}}
-      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
+\dfrac{\env_{1}, \penv \vdash e_{1}:\Nil, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash e_{2}:t, \env_{3}}
+      {\env_{1}, \penv \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
 \end{array}
@@
 \mylabel{T-OR3}\\
-\dfrac{\env_{1} \vdash e_{1}:\False, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t, \env_{3}}
-      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
+\dfrac{\env_{1}, \penv \vdash e_{1}:\False, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash e_{2}:t, \env_{3}}
+      {\env_{1}, \penv \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
 \end{array}
@@
 \mylabel{T-OR4}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3}}
-      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(filter(t_{1}, \Nil), \False) \cup t_{2}, \env_{3}}
-\end{array}
-\end{array}
-\]
-
-The rule \textsc{T-OR1} is the rule that implements the short circuit.
+\dfrac{\env_{1}, \penv \vdash e_{1}:\Nil \cup \False, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash e_{2}:t, \env_{3}}
+      {\env_{1}, \penv \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-OR5}\\
+\dfrac{\env_{1}, \penv \vdash e_{1}:t_{1}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3}}
+      {\env_{1}, \penv \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(filter(t_{1}, \Nil), \False) \cup t_{2}, \env_{3}}
+\end{array}
+\end{array}
+\]
+
+The rule \textsc{T-OR1} is the rule that defines the short circuit.
 We use the consistent-subtyping relation in this rule to guarantee that
@@
 
-The rules \textsc{T-OR2} and \textsc{T-OR3} guarantee that the final
-result is the type of the second expression, because the first one
-is certainly a false value.
-
-The rule \textsc{T-OR4} is the most general rule, but it is also
+The rules \textsc{T-OR2}, \textsc{T-OR3}, and \textsc{T-OR4} guarantee
+that the final result is the type of the second expression, because the
+first one is certainly a false value.
+
+The rule \textsc{T-OR5} is the most general rule, but it is also
 the rule that handles the common $\mathbf{or}$ idioms.
@@
 
-Using these typing rules our type system can type check the following
-example:
+The rule \textsc{T-OR5} allows our type system to type check the following example:
 \begin{center}
@@
 The \emph{filter} predicate removes the type $\Nil$ from the result,
-leaving type $\String \cup ``Hello"$.
-At the end of the evaluation, the expression has type $\String$
-because unions are disjoint and $``Hello" \subtype \String$.
-
-\textit{/* The following example does not type check */}
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; x:\Nil \cup \False = \mathbf{false} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$\mathbf{local} \; y:\String = x \; \mathbf{or} \; ``Hello"$}
-\end{tabular}
-\end{center}
-
-\textit{/* Should we define the following rules? */}
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-OR4}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
-       \Nil \cup \False \subtype t_{1} \;\;\;
-       t_{1} \subtype \Nil \cup \False}
-      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t_{2}, \env_{3}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-OR5}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
-       \Nil \subtype t_{1} \;\;\;
-       \False \subtype t_{1}}
-      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(filter(t_{1}, \Nil), \False) \cup t_{2}, \env_{3}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-OR6}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
-       \Nil \subtype t_{1} \;\;\;
-       \False \not\subtype t_{1}}
-      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(t_{1}, \Nil) \cup t_{2}, \env_{3}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-OR7}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
-       \Nil \not\subtype t_{1} \;\;\;
-       \False \subtype t_{1}}
-      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(t_{1}, \False) \cup t_{2}, \env_{3}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-OR8}\\
-\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
-       \Nil \not\subtype t_{1} \;\;\;
-       \False \not\subtype t_{1}}
-      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t_{1} \cup t_{2}, \env_{3}}
-\end{array}
-\end{array}
-\]
-
-\textit{/* Or who writes this kind of code deserves a type error? */}
+leaving the type $\String \cup ``Hello"$.
+At the end of the type checking, the expression $x \; \mathbf{or} \; ``Hello"$
+has type $\String$ because union types are disjoint and $``Hello" \subtype \String$.
 
@@
 to execute different actions according to their types.
+The rule \textsc{T-IF2} shows the case where our type system
+discriminates the type of a local variable base on the tag \texttt{string}:
 \[
@@
        \env_{1}(n) = t\\
-       closeall(\env_{1}[n \mapsto \String]) \vdash s_{1}:\env_{2} \\
-       closeall(\env_{1}[n \mapsto filter(t, \String)) \vdash s_{2}:\env_{3}\\
+       closeall(\env_{1}[n \mapsto \String]), \penv \vdash s_{1}, \env_{2} \\
+       closeall(\env_{1}[n \mapsto filter(t, \String)), \penv \vdash s_{2}, \env_{3}\\
        \env_{4} = closeset(\env_{1}[n \mapsto t], fav(s_{1}) \cup fav(s_{2}))
       \end{array}}
-      {\env_{1} \vdash \mathbf{if} \; type(n) == ``string" \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}:\env_{4}}
+      {\env_{1}, \penv \vdash \mathbf{if} \; type(n) == ``string" \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
 
@@
 We are assuming that the function \emph{rep} is in the environment and has type
-$\String \times \Integer \times \String \cup \Nil \rightarrow \String$.
+$\String \times \Integer \times \String \cup \Nil \times \Void \rightarrow \String \times \Void$.
 
@@
        \env_{1}(n) = \pi_{i}\\
-       closeall(\env_{1}[\pi \mapsto fpt(\env_{1}(\pi), \Nil, i)]) \vdash s_{1}:\env_{2} \\
-       closeall(\env_{1}[\pi \mapsto gpt(\env_{1}(\pi), \Nil, i)) \vdash s_{2}:\env_{3}
+       closeall(\env_{1}[\pi \mapsto fpt(\env_{1}(\pi), \Nil, i)]), \penv \vdash s_{1}, \env_{2} \\
+       closeall(\env_{1}[\pi \mapsto gpt(\env_{1}(\pi), \Nil, i)), \penv \vdash s_{2}, \env_{3}
       \end{array}}
-      {\env_{1} \vdash \mathbf{if} \; n \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}:closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2}))}
-
+      {\env_{1}, \penv \vdash \mathbf{if} \; n \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2}))}
 \end{array}
@@
 \mylabel{T-EXPLIST6}\\
-\dfrac{\env \vdash me:t_{1} \times ... \times t_{n} \times \Void \sqcup t_{1}' \times ... \times t_{n}' \times \Void, \env_{1}}
-      {\env \vdash me:\pi_{1} \times ... \times \pi_{n} \times \Nil{*}, \env_{1}}
+\dfrac{\env, \penv \vdash me:t_{1} \times ... \times t_{n} \times \Void \sqcup t_{1}' \times ... \times t_{n}' \times \Void, \env_{1}}
+      {\env, \penv \vdash me:\pi_{1} \times ... \times \pi_{n} \times \Nil{*}, \env_{1}}
 \end{array}
@@
 
-\textit{/* How should we write the typing rule of the local declaration? */}
-
 We are assuming that functions \emph{idiv} and \emph{print} are in
 the environment with the respective types
-$\Integer \times \Integer \rightarrow \Integer \times \Integer \sqcup \Nil \times \String$
+$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void$
 and
@@
 \mylabel{T-ARITH5}\\
-\dfrac{\env_{1} \vdash e_{1}:\Any, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:t, \env_{3}}
-      {\env_{1} \vdash e_{1} + e_{2}:\Any, \env_{3}}
+\dfrac{\env_{1}, \penv \vdash e_{1}:\Any, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash e_{2}:t, \env_{3}}
+      {\env_{1}, \penv \vdash e_{1} + e_{2}:\Any, \env_{3}}
 \end{array}
@@
 \mylabel{T-ARITH6}\\
-\dfrac{\env_{1} \vdash e_{1}:t, \env_{2} \;\;\;
-       \env_{2} \vdash e_{2}:\Any, \env_{3}}
-      {\env_{1} \vdash e_{1} + e_{2}:\Any, \env_{3}}
+\dfrac{\env_{1}, \penv \vdash e_{1}:t, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash e_{2}:\Any, \env_{3}}
+      {\env_{1}, \penv \vdash e_{1} + e_{2}:\Any, \env_{3}}
 \end{array}

### 2015-01-27 doc/thesis/system.tex
Improving the text about typing rules

@@
 me \; | \;
-\vec{e}, me\\
+\vec{e}, me \; | \;
+me^{x} \; | \;
+\vec{e}, me^{x}\\
 me ::= & \;\; e(el)_{m} \; | \;
@@
 
-We present the typing rules as a deduction system for the typing
-relations $\env_{1}, \penv \vdash s, \env_{2}$ and
-$\env_{1}, \penv \vdash e : t, \env_{2}$.
-We use the first one for typing statements, and the second one
-for typing expressions.
-The first relation means that given the type environment $\env_{1}$
+We use three different typing relations to present our typing rules
+as a deduction system for these typing relations.
+We use the relation $\env_{1}, \penv \vdash s, \env_{2}$ for typing
+statements.
+This relation means that given the type environment $\env_{1}$
 and the projection environment $\penv$,
 typing the statement $s$ produces the new type environment $\env_{2}$.
-The second relation means that given the type environment $\env_{1}$
+We use the relation $\env_{1}, \penv \vdash e : t, \env_{2}$ for
+typing expressions.
+This relation means that given the type environment $\env_{1}$
 and the projection environment $\penv$,
 the expression $e$ has type $t$ and produces the new type environment $\env_{2}$.
-We need two typing relations because only expressions have types,
-though expressions and statements can modify the environment.
+We use the relation $\env_{1}, \penv \vdash el : r_{1}, \env_{2}, (x,r_{2})$
+for typing expression lists.
+This relation means that given type type environment $\env_{1}$
+and the projection environment $\penv$,
+the expression list $el$ has type $r_{1}$ and produces the new type
+environment $\env_{2}$ and produces the pair $(x,r_{2})$.
+We need three typing relations because statements do not have types
+and expression lists can result in a union of tuple types that
+should be introduced in the projection environment.
+We use $\env_{1}, \penv \vdash el : r, \env_{2}$ when the expression
+list does not result in a union of tuple types.
 We need two environments because $\env$ maps variables to types,
@@
 $\env_{1}$ with the variable $n$ that maps to type $t$.
-We use $\penv[x \mapsto r]$ for meaning that we extendend the environment
+We use $\penv[x \mapsto r]$ for meaning that we extended the environment
 $\penv$ with the projection variable $x$ that maps to the union of tuple types $r$.
@@
 Lua has multiple assignments, and the rule \textsc{T-ASSIGNMENT} uses
-tuple types along with consitent-subtyping to type check this Lua feature:
+tuple types along with consistent-subtyping to type check this Lua feature:
 \[
@@
 a subtype of its previous type.
-We will show in this section that the refinment of table types can
+We will show in this section that the refinement of table types can
 change the environment, and that is the reason why we need to check
@@
 This behavior preserves the semantics of Lua on replacing missing values,
-as it is necessary when we ommit optional parameters in a function call,
+as it is necessary when we omit optional parameters in a function call,
 like the previous example showed.
@@
 but this could lead to fields that have a too general type, which could
-decrease the ammount of static type checking.
+decrease the amount of static type checking.
 This means that our type system does not allow the table type to change
@@
 table types allow us to initialize optional fields.
-To do that, Typed Lua includes a cast expression that allow us
+To do that, Typed Lua includes a cast expression that allows us
 to use an \emph{unique} table type before it becomes \emph{open}.
@@
 
+The rule \textsc{T-FUNCTION3} uses the predicates \emph{closeall}
+and \emph{closeset} to close \emph{open} and \emph{unique} table types.
+The predicate \emph{closeall} closes all \emph{open} and \emph{unique}
+table types before type checking another scope.
+The predicate \emph{closeset} closes a given set of free assigned variables,
+which is given by the predicate \emph{fav}.
 This rule prevents the following unsafe example to type check:
@@
 \multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
-& \multicolumn{3}{l}{$\mathbf{local} \; b:\Integer \times \Integer \rightarrow \Integer =$}\\
+& \multicolumn{3}{l}{$\mathbf{local} \; b:\Integer \times \Integer \times \Void \rightarrow \Integer \times \Void =$}\\
 & & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer, y:\Integer):\Integer$}\\
@@
 
-Typed Lua uses the refinement of table types to allow programmers to
-build modules and objects.
-It also allows Typed Lua to type check an object-oriented idiom that
-programmers often use it.
-The rule \textsc{T-METHOD3} illustrates how our type system type checks
-method definitions:
+Typed Lua also uses the refinement of table types to type check
+common idioms that Lua programmers use for building modules and objects.
+For instance, the rule \textsc{T-METHOD3} illustrates how our type system
+type checks method definitions:
 \[
@@
 
-The rule \textsc{T-METHOD3} allow our type system to type check the
-following example, which is similar to the definition of the class
+The rule \textsc{T-METHOD3} allows our type system to type check the
+following example, which is a translation of the definition of the class
 \texttt{Shape} from Section \ref{sec:oop}:
@@
 \multicolumn{4}{l}{$\mathbf{in}$}\\
-& \multicolumn{3}{l}{$\mathbf{fun} \; s{:}new (x:\Number, y:\Number):\Self$}\\
+& \multicolumn{3}{l}{$\mathbf{fun} \; s{:}new (x:\Number, y:\Number):\Self \times \Void$}\\
 & & \multicolumn{2}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{closed} =$}\\
@@
        \env_{1}(n) = t\\
-       closeall(\env_{1}[n \mapsto \String]), \penv \vdash s_{1}, \env_{2} \\
-       closeall(\env_{1}[n \mapsto filter(t, \String)), \penv \vdash s_{2}, \env_{3}\\
+       closeall(\env_{1}[n \mapsto \phi(t,\String)]), \penv \vdash s_{1}, \env_{2} \\
+       closeall(\env_{1}[n \mapsto \phi(t,filter(t, \String))), \penv \vdash s_{2}, \env_{3}\\
        \env_{4} = closeset(\env_{1}[n \mapsto t], fav(s_{1}) \cup fav(s_{2}))
@@
 
-Using the rule \textsc{T-IF2}, Typed Lua can type check the following example:
+In Typed Lua, \emph{type} is a primitive that triggers a type change.
+In the rule \textsc{T-IF2}, the type of the variable $n$ changes from
+$t$ to $\phi(t,\String)$ inside the $\mathbf{if}$ branch, and it changes
+the type of the variable $n$ from $t$ to $\phi(t,filter(t,\String))$
+inside the $\mathbf{else}$ branch.
+Our type system changes the type of a local variable to a filter type
+because it gives the discriminated type when we use the variable as
+an expression, and it restores the original type when we use the variable
+in an assignment.
+We will use the following example to discuss how our type system uses
+the rule \textsc{T-IF2} to type check this given example:
 \begin{center}
 \begin{tabular}{llll}
-\multicolumn{4}{l}{$\mathbf{local} \; o:\String \times \String \cup \Integer \rightarrow \String =$}\\
-& \multicolumn{3}{l}{$\mathbf{fun} \; (a:\String, b:\String \cup \Integer):\String$}\\
+\multicolumn{4}{l}{$\mathbf{local} \; o:\String \times \String \cup \Integer \times \Void \rightarrow \String \times \Void =$}\\
+& \multicolumn{3}{l}{$\mathbf{fun} \; (a:\String, b:\String \cup \Integer):\String \times \Void$}\\
 & & \multicolumn{2}{l}{$\mathbf{local} \; r:\String = ``" \; \mathbf{in}$}\\
@@
 
-We are assuming that the function \emph{rep} is in the environment and has type
+In this example, we are assuming that the function \emph{rep} is in
+the environment and has type
 $\String \times \Integer \times \String \cup \Nil \times \Void \rightarrow \String \times \Void$.
+Note that rule \textsc{T-IF2} changes the type of $b$ to
+$\phi(\String \cup \Integer, \String)$ inside the $\mathbf{if}$ branch,
+and it changes the type of $b$ to
+$\phi(\String \cup \Integer, \Integer)$ inside the $\mathbf{else}$ branch.
+This example type checks because reading $b$ results in the type
+$\String$ inside the $\mathbf{if}$ branch, and it results in the type
+$\Integer$ inside the $\mathbf{else}$ branch.
+Outside the condition, the local variable $b$ has type $\String \cup \Integer$.
 
@@
 \texttt{boolean}, and \texttt{number}.
-There is also a similar rule for handling the type $\Integer$, but
-it works only with Lua 5.3, as it depends on the function \texttt{math.type}.
+There is also a similar rule for handling the tag \texttt{integer}, but
+it is limited to Lua 5.3, as it depends on the function \texttt{math.type}.
 This function appears only in Lua 5.3 and it returns the string $``integer"$
@@
 
-Lua programmers also overload the return type of functions to denote
-errors, and Typed Lua also handles this case with the rule \textsc{T-IF3}:
-\[
-\begin{array}{c}
-\mylabel{T-IF3}\\
-\dfrac{\begin{array}{c}
-       \env_{1}(n) = \pi_{i}\\
-       closeall(\env_{1}[\pi \mapsto fpt(\env_{1}(\pi), \Nil, i)]), \penv \vdash s_{1}, \env_{2} \\
-       closeall(\env_{1}[\pi \mapsto gpt(\env_{1}(\pi), \Nil, i)), \penv \vdash s_{2}, \env_{3}
-      \end{array}}
-      {\env_{1}, \penv \vdash \mathbf{if} \; n \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2}))}
-\end{array}
-\]
-
-\[
-\begin{array}{c}
-\mylabel{T-EXPLIST6}\\
-\dfrac{\env, \penv \vdash me:t_{1} \times ... \times t_{n} \times \Void \sqcup t_{1}' \times ... \times t_{n}' \times \Void, \env_{1}}
-      {\env, \penv \vdash me:\pi_{1} \times ... \times \pi_{n} \times \Nil{*}, \env_{1}}
-\end{array}
-\]
-
+Lua programmers also overload the return type of functions to denote errors,
+and our type system also handles this idiom.
+For instance,
 \begin{center}
 \begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; q:\pi_{1}, r:\pi_{2} = idiv(1, 2)_{m} \; \mathbf{in}$}\\
+\multicolumn{2}{l}{$\mathbf{local} \; q:\pi_{1}^{x}, r:\pi_{2}^{x} = idiv(1, 2)_{m}^{x} \; \mathbf{in}$}\\
 & \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; print(q + r)_{s} \; \mathbf{else} \; print(``ERROR: " \; .. \; r)_{s}$}
@@
 \end{center}
-
-We are assuming that functions \emph{idiv} and \emph{print} are in
+type checks, assuming that functions \emph{idiv} and \emph{print} are in
 the environment with the respective types
@@
 $\Value{*} \rightarrow \Void$.
+There are three typing rules involved in the type checking of this example,
+which we will explain now.
+
+First, type checking $idiv(1, 2)_{m}^{x}$ results in a tuple of projection
+types and in a pair that the typing rule of the local declaration needs
+for introducing the variable $x$ in the projection environment.
+The rule \textsc{T-EXPLIST6} express this intuition:
+\[
+\begin{array}{c}
+\mylabel{T-EXPLIST6}\\
+\dfrac{\begin{array}{c}
+       \env, \penv \vdash me^{x}:r\\
+       r = t_{1} \times ... \times t_{n} \times \Void \sqcup t_{1}' \times ... \times t_{n}' \times \Void, \env_{1}
+       \end{array}}
+      {\env, \penv \vdash me^{x}:\pi_{1}^{x} \times ... \times \pi_{n}^{x} \times \Nil{*}, \env_{1}, (x,r)}
+\end{array}
+\]
+
+Then, type checking the local declaration checks whether the tuple of projection
+types match the types in the annotations.
+If the types match, the local declaration extends the type environment with the
+new variables, and it also uses the pair $(x,r)$ from the previous rule to
+extend the projection environment before type checking the statements.
+The rule \textsc{T-LOCAL} express this intuition:
+\[
+\begin{array}{c}
+\mylabel{T-LOCAL}\\
+\dfrac{\env_{1}, \penv \vdash el:r_{1}, \env_{2}, (x,r_{2}) \;\;\;
+       r_{1} \lesssim \vec{t} \;\;\;
+       \env_{2}[\vec{n} \mapsto \vec{t}], \penv[x \mapsto r_{2}] \vdash s, \env_{3}}
+      {\env_{1}, \penv \vdash \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s, \env_{3} - \{\vec{n} \mapsto \vec{t}\}}
+\end{array}
+\]
+
+After type checking the statement $s$, the rule \textsc{T-LOCAL} produces a
+new environment without the variables that it introduced before type checking $s$.
+
+Once the projection is in the environment, our type system can allow
+discriminating variables that hold projection types.
+The rule \textsc{T-IF3} illustrates the case of our previous example:
+\[
+\begin{array}{c}
+\mylabel{T-IF3}\\
+\dfrac{\begin{array}{c}
+       \env_{1}(n) = \pi_{i}^{x} \;\;\; \penv(x) = r\\
+       closeall(\env_{1}), \penv[x \mapsto fpt(r,\Nil,i)] \vdash s_{1}, \env_{2} \\
+       closeall(\env_{1}), \penv[x \mapsto gpt(r,\Nil,i)] \vdash s_{2}, \env_{3}
+      \end{array}}
+      {\env_{1}, \penv \vdash \mathbf{if} \; n \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2}))}
+\end{array}
+\]
+
+The rule \textsc{T-IF3} uses the predicates \emph{fpt} and \emph{gpt}
+to filter projections, affecting all variables that bind to the same projection.
+For instance, our previous example type checks through the rule \textsc{T-IF3},
+because it makes the rule \textsc{T-IF3} use the predicate
+\[
+fpt(\Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void, \Nil, 1)
+\]
+to discriminate the projection $x$ to the single tuple
+$\Integer \times \Integer \times \Void$, and the predicate
+\[
+gpt(\Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void, \Nil, 1)
+\]
+to discriminate the projection $x$ to the single tuple
+$\Nil \times \String \times \Void$.
+This means that the type of $q$ ($\pi_{1}^{x}$) and $r$ ($\pi_{2}^{x}$)
+project to $\Integer$ and $\Integer$ inside the $\mathbf{if}$ branch,
+but they project to $\Nil$ and $\String$ inside the $\mathbf{else}$ branch.
+Outside the $\mathbf{if}$ statement, the types of $q$ and $r$ project
+to $\Integer \cup \Nil$ and $\Integer \cup \String$, respectively.
 

### 2015-01-30 doc/thesis/review.tex
Some english revisions to the state of the art

@@
-We begin this chapter presenting a little bit of the history about
+We begin this chapter presenting a little bit of the history behind
 combining static and dynamic typing in the same language.
@@
 We end this chapter presenting some statistics about the usage of
-some Lua features and idioms that helped us identify how should
-we combine static and dynamic typing in Lua.
+some Lua features and idioms that helped us identify how we should
+combine static and dynamic typing in Lua.
 
@@
 in the early eighties, but not for static type checking.
-Instead, programmers may choose to declare types of variables as
+Instead, programmers could choose to declare types of variables as
 optimization hints to the compiler, that is, type declarations are
@@
 run-time checks or simply running the code.
-Soft typing gives this choice to the programmer because it is one of
-its key ideas, that is, type inference and static type checking should
+This means that type inference and static type checking do
 not prevent the programmer from running inconsistent code.
@@
 unnecessary checks of tags.
-The paper describes how to translate dynamically typed code to
+Henglein describes how to translate dynamically typed code into
 statically typed code that uses a \texttt{Dynamic} type.
@@
 Although soft typing and dynamic typing may seem similar, they are not.
-They are not similar because soft typing targets statically type
-checking of dynamically typed languages for detecting programming
-errors, while dynamic typing targets the optimization of dynamically
+Soft typing targets statically type checking of dynamically typed
+languages for detecting programming errors, while
+dynamic typing targets the optimization of dynamically
 typed code through the elimination of unnecessary run-time checks.
@@
 error is in library code, they still may have problems to identify
-if this run-time error is due to a violation of library's contract,
+if this run-time error is due to a violation of library's contract
 or due to a bug, when the library is poorly documented.
@@
 Programmers can choose between using the dynamically typed version
-of BabyJ, when they do not need types at all, and the statically
-typed version of BabyJ, when they need to annotate the code.
-In the statically typed BabyJ, programmers can use the
+of BabyJ when they do not need types at all, and the statically
+typed version of BabyJ when they need to annotate the code.
+In statically typed BabyJ, programmers can use the
 \emph{permissive type} $*$ to annotate the parts of the code that
@@
 run-time semantics.
-Although these systems can be unsound on their interaction with the
+Although these systems can be unsound in their interaction with the
 dynamically typed part of the language or even by design, their
@@
 In fact, Dart is a new class-based object-oriented language with
-optional type annotations and a semantics that resembles the
+optional type annotations and semantics that resembles the
 semantics of Smalltalk, while TypeScript is a strict superset of
@@
 For instance, Dart has covariant arrays, while TypeScript has
-covariant function return types;
-without counting the interaction between statically and dynamically
+covariant function return types,
+besides the interaction between statically and dynamically
 typed code that is also unsound.
@@
 There is no common formalization for optional type systems, and
-each language ends up implementing its optional type system on
+each language ends up implementing its optional type system in
 its own way.
@@
 conversions to and from the dynamic type $?$.
-The rule \textsc{C-FUNC} resembles the subtyping between function types,
+The rule \textsc{C-FUNC} resembles subtyping between function types,
 because it is contravariant on the argument type and covariant on the
@@
 function is statically typed; in this case, the argument type should
-be consistent to the argument type of the function's signature.
+be consistent with the argument type of the function's signature.
 
@@
 \citet{abadi1989dts} and \citet{thatte1990qst} by including a
-dynamic type to a statically typed language.
+dynamic type in a statically typed language.
 However, these three approaches differ in the way they handle the
@@
 
-The subtyping relation $\subtype$ actually is a pitfall on Thatte's
+The subtyping relation $\subtype$ is actually a pitfall on Thatte's
 quasi-static typing, because it sets the dynamic type
@@
 $t \subtype \;?$ and $? \subtype t$.
-Subtyping is transitive, so we have that
+Subtyping is transitive, so we know that
 \[
@@
 the rules \textsc{S-NUM}, \textsc{S-STR}, and \textsc{S-ANY},
-we did not combine these rules in a single one to make explicit the
+we did not combine them into a single rule to make explicit the
 neutrality of the dynamic type $?$ to the subtyping rules.
@@
 handle the casts that involve the dynamic type $?$.
-The consistent-subtyping relation is not transitive, and that is the
-reason why the dynamic type $?$ is not neutral to this relation.
+The consistent-subtyping relation is not transitive, and thus
+the dynamic type $?$ is not neutral to this relation.
 
@@
 consistent-subtyping relation to statically check the interaction
-between statically and dynamically typed code, and not influencing
+between statically and dynamically typed code, without influencing
 the run-time semantics.
@@
 repeated casts to the same function or object.
-\citet{herman2007sgt} use the coercion calculus of
+\citet{herman2007sgt} use the coercion calculus outlined in
 \citet{henglein1994dts} to express casts as coercions and
@@
 because a cast application can be delayed and the error related
-to that cast application can appear far way from the real error.
-\citet{wadler2009wpc} developed the \emph{blame calculus} as a way
-to handle this issue, and \citet{ahmed2011bfa} extended the
+to that cast application can appear considerable distance
+from the real error.
+\citet{wadler2009wpc} developed \emph{blame calculus} as a way
+to handle this issue, and \citet{ahmed2011bfa} extended
 blame calculus with polymorphism.
-The blame calculus is an intermediate language to integrate
+Blame calculus is an intermediate language to integrate
 static and dynamic typing that uses the blame tracking approach
@@
 
-On the one hand, the blame calculus solves the issue regarding
+On the one hand, blame calculus solves the issue regarding
 error reporting;
@@
 by \citet{herman2007sgt}.
-Thus, \citet{siek2009casts} extended the coercion calculus of
+Thus, \citet{siek2009casts} extended the coercion calculus outlined in
 \citet{herman2007sgt} with blame tracking to achieve an
@@
 
-There are some programming languages that either incorporated
-gradual typing or have research on this subject.
+There are some projects that incorporate gradual typing into some
+programming languages.
 Reticulated Python \citep{reticulated,vitousek2014deg} is a research
@@
 Perl 6 \citep{tang2007pri} is also being designed with gradual typing,
-though there is few documentation about the gradual type systems
+though there is little documentation about the gradual type systems
 of these languages.
@@
 
-Hence, optional type annotations for software evolution probably are
-the reasons why optional type systems are commonly called
+Hence, optional type annotations for software evolution are likely
+the reason why optional type systems are commonly called
 gradual type systems.
 Typed Clojure \citep{bonnaire-sergeant2012typed-clojure} is an
-optional type system for Clojure that now is adopting the
+optional type system for Clojure that is now adopting the
 gradual typing slogan.
@@
 Sage \citep{gronski2006sage} is a programming language that
-extended hybrid type checking with a dynamic type to
+extends hybrid type checking with a dynamic type to
 support dynamic and static typing in the same language.
@@
 
-Thus, the inclusion of a dynamic type to hybrid type checking
+Thus, the inclusion of a dynamic type in hybrid type checking
 along with optional type annotations, and the insertion of run-time
-checks probably are the reasons why hybrid type checking is
+checks are likely the reason why hybrid type checking is
 also viewed as a form of gradual typing.
@@
 The approach used by \citet{tobin-hochstadt2008ts} to design and
-implement Typed Racket probably is also called gradual typing
+implement Typed Racket is probably also called gradual typing
 because it allows the programmer to gradually migrate from untyped
@@
 
-Typed Lua should also include some way for defining the types of
+Typed Lua should also include some way to define the types of
 userdata.
-This feature should also allow programmers to define an userdata
+This feature should also allow programmers to define userdata
 that can be used in an object-oriented style, as this is another

### 2015-01-31 doc/thesis/evaluation.tex
Minor improvements to the text about related work

@@
 to evaluate the applicability of our type system.
-For each library, we used Typed Lua to either rewrite its modules
+For each library, we used Typed Lua to either annotate its modules
 or to write statically typed interfaces to its modules through
@@
 
-We also compared our type system to other type systems,
-as a way to evaluate our contributions.
+We also evaluate our contributions through the revision of related work.
 In Section \ref{sec:relatedwork} we discuss what are the novel
-features introduced by our type system,
-what are the features that it shares with other type systems,
-and whether we can incorporate some interesting features from
-other type systems.
+features of our type system, what are the features that it shares
+with other type systems, and whether we can incorporate some
+interesting features from other type systems.
 
@@
 \item how precisely it can describe the type of the interface of a module;
-\item if it provides guarantees that the code corresponds to the interface.
+\item whether it provides guarantees that the code matches the interface.
 \end{enumerate}
@@
 The second case study that we chose was the MD5 library \citep{lmd5},
-because it is simple and contains just one module.
-We needed a simple case study to introduce Typed Lua's description
+because we wanted a simple case study to introduce Typed Lua's description
 files and \texttt{userdata} declarations.
 These Typed Lua's mechanisms allow programmers to give statically typed
-interfaces to Lua libraries that are written in C.
+interfaces to Lua libraries.
 
@@
 We randomly selected these libraries to evaluate Typed Lua for
-rewriting existing Lua modules.
+annotating existing Lua modules.
 
 The Typed Lua compiler is the last case study that we evaluated.
-We chose it as a case study because it is a large application.
-In addition, bootstrapping the compiler also provides some
-confidence on the type system.
+We chose the Typed Lua compiler as a case study because it is a large application,
+and bootstrapping the compiler also provides some confidence in the type system.
 
@@
 
-In this section we compare Typed Lua with other projects.
-We split this section into two parts: the first compares Typed Lua
-with other Lua projects, while the second compares Typed Lua with
-other projects that are not related to Lua.
+In this section we review related work.
+We split this section into two parts: in the first part we review
+other Lua projects, while in the second part we review other
+projects that are not related to Lua.
 
@@
 MoonScript \citep{moonscript} is a programming language that supports
-class-based object-oriented programming, and compiles to idiomatic
-Lua code.
-However, it does not perform compile-time type checking.
+class-based object-oriented programming.
+MoonScript compiles to idiomatic Lua code, but 
+it does not perform compile-time type checking.
 
 LuaInspect \citep{luainspect} is a tool that uses MetaLua to perform
-some code analysis, such as flagging unknown global variables and
-table fields, checking the number of function arguments against
-signatures, and inferring function return values, but it does not
-try to analyze object-oriented code and does not perform compile-time
-type checking.
+some code analysis.
+For instance, it flags unknown global variables and table fields,
+it checks the number of function arguments against signatures, and
+it infers function return values.
+However, it does not try to analyze object-oriented code and
+it does not perform compile-time type checking.
 
@@
 Statements include declaration of local variables, multiple assignments,
-function application, and the return statement, that is, they do not
-include any control-flow statements.
+function application, and the return statement.
+This means that Tidal Lock does not include any control-flow statement.
 Expressions include primitive literals, table indexing, function application,
@@
 
-Tidal Lock is also a structural type system based on subtyping that has
+Tidal Lock is also a structural type system that relies on subtyping and
 local type inference.
@@
 but it does not perform static type checking.
+
+Ravi \citep{ravi} is an experimental Lua dialect.
+Ravi introduces optional static typing for Lua to improve run-time performance.
+To do that, Ravi extends the Lua Virtual Machine to include new
+operations that take into account static type information.
 
@@
 generic functions, lists, and maps.
+Unlike Typed Lua, Dart is unsound by design.
 
@@
 on subtyping, and supports generics.
-It includes the dynamic type, primitive types, function
-types, array types, tuple types, recursive types, and
+It includes the dynamic type, primitive types, union types,
+function types, array types, tuple types, recursive types, and
 object types.
-Unlike Typed Lua, TypeScript does not include union types,
-and it uses arrays to represent variadic functions and
-multiple return values.
+Unlike Typed Lua, TypeScript uses arrays to represent variadic
+functions and multiple return values.
 

### 2015-02-02 doc/thesis/evaluation.tex
Minor improvements to the text about case studies

@@
 \hline
-LuaSocket & 109 & 1 & 2 & 11 & 123 & 89\% \\
+LuaSocket & 110 & 1 & 2 & 10 & 123 & 89\% \\
 \hline
@@
 basic assumptions about Lua code \citep{luamanual}.
-For instance, we can use its function \texttt{setlocal} to change the value
+For instance, we can use the function \texttt{debug.setlocal} to change the value
 of a local variable that is not visible in the current scope.
@@
 the Lua Standard Library, though some of them are still
-hard to type with the current type system.
-Now, we discuss why some modules are difficult to type,
-and what kind of limitations they represent on our type system.
+hard to type with our current type system.
+Now we will discuss why some modules are difficult to type,
+and what kind of limitations they represent in our type system.
 
 The \texttt{base} module was quite hard to type because it
-includes several functions that have a dynamic behavior.
-For instance, the function \texttt{next} traverses all the
-fields of a table, but the order that the indices are
-enumerated is not specified.
-For this reason, it is also difficult to type the function
-\texttt{pairs}, as it uses the function \texttt{next} to
-traverse a table.
-The functions \texttt{getmetatable}, \texttt{rawget}, and
-\texttt{rawset} are other examples of functions that their
-behavior depend on the C implementation.
+includes several functions that rely on reflection.
+For instance, the function \texttt{pairs} traverses all
+keys and values that are stored in a given table.
+As another example, the function \texttt{getmetatable}
+returns the metatable of a given table.
 
-We could not give a precise static type to \texttt{ipairs}
+We could not give a precise static type to \texttt{ipairs},
 because our type system does not have parametric polymorphism.
@@
 
-The \texttt{base} module also includes two special cases:
-the functions \texttt{tonumber} and \texttt{collectgarbage}.
+The \texttt{base} module also includes two overloaded functions:
+\texttt{tonumber} and \texttt{collectgarbage}.
 
-The function \texttt{tonumber} is a special case because the
-type of the first parameter depends on the type of the
-second parameter.
+In the case of \texttt{tonumber}, the type of the first parameter
+depends on the type of the second parameter.
 For instance, we can call \texttt{tonumber(1)}, but we cannot
@@
 
-The function \texttt{collectgarbage} is a special case because
-its return type depends on the value of the first parameter,
-which should be a string.
-For instance, calling \texttt{collectgarbage("collect")} returns an integer number,
-calling \texttt{collectgarbage("count")} returns a floating point number,
+In the case of \texttt{collectgarbage}, the return type changes
+according to an input literal string.
+For instance, calling \texttt{collectgarbage("collect")} returns an integer,
+calling \texttt{collectgarbage("count")} returns a floating point,
 and calling \texttt{collectgarbage("isrunning")} returns a boolean.
@@
 The first stores loaded modules, while the others store module \emph{loaders}.
-They are difficult to type because their types are dynamic and
-depend on the modules a program loads.
+They are difficult to type because their types rely on reflection,
+that is, their types depend on the modules a program loads.
+
+The \texttt{string} module was not that hard to type, but it includes
+some functions that we could not type, as they rely on format
+strings.
+For instance, the type of the arguments that we pass to
+\texttt{string.format} depend on the format string we are using.
+It is fine to call \texttt{string.format("\%d", 1)}, but
+\texttt{string.format("\%d", true)} raises a run-time error.
 
@@
 
-The \texttt{string} module is not that hard to type, but it includes
-some functions that require dependent types.
-For instance, the type of the arguments that we pass to
-\texttt{string.format} depend on the format string we are using.
-It is fine to call \texttt{string.format("\%d", 1)}, but
-\texttt{string.format("\%d", true)} generates a run-time error.
-
-The \texttt{utf8} module was quite straightforward to type,
+The \texttt{utf8} module was straightforward to type,
 as its members are only operations over strings.
 
-The \texttt{table} module is specially difficult to type because
-almost all of its members require parametric polymorphism.
-All the functions of this module either receive or return a list
-of elements, and parametric polymorphism would help us to describe
-them with a generic type.
+The \texttt{table} module was specially difficult to type because
+most of its functions require parametric polymorphism.
+These functions either receive or return a list of elements, and
+parametric polymorphism would help us to describe them with a generic type.
 
@@
 Even though the \texttt{math} module looks straightforward to type,
-it includes several functions that the return type depend on the input type.
+it includes several overloaded functions.
 For instance, \texttt{math.abs(1)} returns the integer \texttt{1},
@@
 
-The \texttt{math} module also includes a special case that is hard to type:
+The \texttt{math} module also includes a function that is hard to type:
 the function \texttt{math.random}.
-This function is difficult to type because its type depends on the calling arity.
-Calling \texttt{math.random()} returns a random floating point number
+It is difficult to type because its type depends on the calling arity.
+Calling \texttt{math.random()} returns a random floating point
 between \texttt{0} and \texttt{1}.
@@
 and generates a run-time error whenever we pass more than two arguments.
-There is also a problem on its documentation, as it suggests that
-the two integers are optional parameters, but, for \texttt{math.random},
-optional parameters behave in a different way.
-Usually, Lua functions replace optional parameters with a default value
-when they are \texttt{nil}, but \texttt{math.random} generates a run-time
-error instead.
 
@@
 The functions \texttt{io.read} and \texttt{io.lines} are difficult
-to type because their return types depend on their input types.
+to type because they are overloaded functions.
 For instance, \texttt{io.read("l")} returns a string or \texttt{nil},
@@
 The function \texttt{os.execute} is the only member of the \texttt{os}
-module that we could not give a precise type.
-This function is difficult to type because its return type depend on
-the input type.
+module that we could not give a precise type, as it is an overloaded
+function.
 More precisely, \texttt{os.execute()} returns a boolean, while
@@
 
-Even tough it was quite straightforward to type the MD5 library,
+Even tough it was straightforward to type the MD5 library,
 we found a little difference between its documentation and its behavior.
@@
 This case study shows that type annotations help programmers maintain the
-documentation updated, as the type checker always validate them.
+documentation updated, as the type checker always validates them.
 
@@
 URL manipulation, and LTN12 filters \citep{nehab2008ltn012}.
-We used Typed Lua's description files to type both parts.
-This is the only case study that we test description files to statically type
-a Lua module instead of rewriting it.
+We used Typed Lua's description files to type both parts, as we also
+wanted to use LuaSocket to test description files to statically type
+the interface of modules that are written in Lua.
 Table \ref{tab:evalsocket} summarizes the evaluation results for LuaSocket.
@@
 \hline
-smtp & 6 & 0 & 0 & 1 & 7 & 86\% \\
+smtp & 7 & 0 & 0 & 0 & 7 & 100\% \\
 \hline
@@
 \texttt{socket.try}, \texttt{socket.protect}, and \texttt{socket.skip}
-that have a dynamic behavior.
-This means that the return type of these functions depend on the
-type of the values that we use to call them.
-For instance, \texttt{socket.skip(0, nil, "hello")} returns the
-tuple \texttt{(nil, "hello")}, \texttt{socket.skip(1, nil, "hello")}
-returns only the string \texttt{"hello"}, and
-\texttt{socket.skip(-1, nil, "hello")} returns the tuple
-\texttt{(-1, nil, "hello")}.
+that are hard to type because they rely on reflection.
+For instance, we can use \texttt{socket.skip} to choose the number of
+values that we want to return.
+As an example, calling \texttt{socket.skip(1, nil, "hello")}
+returns only the string \texttt{"hello"}, because \texttt{1} indicates
+that we do not want to return the first value.
 Passing a negative number to \texttt{socket.skip} can be
 dangerous, as it returns anything that might be in the stack.
-For instance, \texttt{f = socket.skip(-2)} assigns \texttt{socket.skip}
-to \texttt{f}.
+As an example, calling \texttt{socket.skip(-1, nil, "hello")}
+returns the tuple \texttt{(-1, nil, "hello")}, because \texttt{-1} makes
+\texttt{socket.skip} does not skip any values.
+As another example, \texttt{f = socket.skip(-2)} assigns \texttt{socket.skip}
+to \texttt{f}, as \texttt{-2} gets \texttt{socket.skip} from the stack.
 Our type system cannot handle the type of negative numbers, as
-this requires the refinement types from hybrid type checking \citep{flanagan2006htc}.
+this requires more complex types such as the refinement types from
+hybrid type checking \citep{flanagan2006htc}.
 
-We could type most of the members of the modules \texttt{ftp},
-\texttt{http}, and \texttt{smtp}, but we could not precisely type
-the functions \texttt{ftp.get}, \texttt{http.request}, and
-\texttt{smtp.message}.
+We could type most of the members of the modules \texttt{ftp} and
+\texttt{http}, but we could not precisely type
+the functions \texttt{ftp.get} and \texttt{http.request} because
+they are overloaded functions.
 
 The function \texttt{ftp.get} downloads data from a given URL,
-and it is difficult to type because its return type depends on
-the type of its input parameter.
+which can be either a string or a table.
 More precisely, \texttt{ftp.get(url)} returns the tuple
@@
 The function \texttt{http.request} downloads data from a given URL,
-and it is difficult to type because its return type depends on
-the types of its input parameters.
+which can be either a string or a table.
 More precisely, \texttt{http.request(url, body)} returns the tuple
@@
 
-The function \texttt{smtp.message} creates a function that sends
-an SMTP message body, and it is difficult to type because of the
-type of the message body, which is recursive and makes harder
-its initialization.
-
 The modules \texttt{mime} and \texttt{ltn12} have a strong connection.
 The \texttt{mime} module offers low-level and high-level filters
-that apply and remove some text encodings.
+that apply and remove some text encoding.
 The low-level filters are written in C, while the high-level filters
@@
 \texttt{ltn12} module that we could not give a precise type.
-This function is difficult to type because the \texttt{mime} low-level
-filters are polymorphic.
+This function is difficult to type because it is polymorphic.
 
+The modules \texttt{smtp} and \texttt{url} were straightforward to type.
+The \texttt{smtp} module provides functions that send e-mails.
 The \texttt{url} module provides functions that manipulate URLs.
-It is straightforward to type, although it defines a table type for
-parsed URLs that is not precise enough.
-This type is problematic because it is a record type where all
-fields are optional, and makes any table type that does not include
-the optional fields statically type check.
-
-This case study shows that abusing of dynamic typing usually impacts
-the users of a module, as they have to keep on checking each return
-value before using it to prevent API misuses.
 
@@
 \texttt{http.request} with MD5 authentication.
-Like \texttt{http.request}, the return type of \texttt{http-digest.request}
-also depends on the types of its input parameters.
+Like \texttt{http.request}, \texttt{http-digest.request}
+is also an overloaded function.
 
 Even though we could not precisely type the interface that \texttt{http-digest}
-exports, we could use only static types to rewrite it, and they pointed a bug
+exports, we could use only static types to annotate it, and they pointed a bug
 in the code.
@@
 
-However, we hit some limitations of our type system to rewrite this module.
+However, we hit some limitations of our type system while annotating this module.
 First, it uses the function \texttt{getmetatable} to get a table and
@@
 
-Even though we had these two issues to rewrite the \texttt{mod11} module,
-we could use only static types to rewrite it, and found some interesting
+Even though we had these two issues to annotate the \texttt{mod11} module,
+we could use only static types to annotate it, and we found some interesting
 points.
-The code relies a lot on implicit conversions between strings and numbers,
+The code relies on implicit conversions between strings and numbers,
 and some parts of the code keep on changing the type of local variables.
@@
 the Typed Lua parser and the description file parser.
-This module is hard to type because it exports LPeg \citep{lpeg,ierusalimschy2009lpeg}
-patterns.
-For instance,
-\begin{verbatim}
-    tllexer.Number = lpeg.C(Hex + Float + Int) / tonumber
-\end{verbatim}
-exports the pattern that matches a number.
-Note that LPeg overloads the definition of arithmetic operators,
-a feature that our type system does not support yet.
-That is the reason why we could not precisely type the LPeg
-patterns that this module exports.
-However, we could type the two error reporting functions that this
-module exports.
+This module is hard to type because it uses LPeg \citep{lpeg,ierusalimschy2009lpeg}
+patterns, and LPeg uses overloaded arithmetic operators to
+build LPeg patterns. 
+We could not precisely type LPeg patterns because our
+type system still does not support overloading arithmetic
+operators.
+In this module, we could only type the two error reporting functions
+that it exports.
 
@@
 Traversing the AST would also be problematic if we had not included
-a way to decompose unions of table types, as we mentioned in Section \ref{sec:alias}.
+a way to discriminate unions of table types, as we mentioned in Section \ref{sec:alias}.
 
@@
 a nilable value before using it.
-In fact, this point appeared in all the case studies that we rewrote
-the Lua code.
-This means that Lua programmers often use values before checking
-whether it is \texttt{nil}.
+In fact, this point appeared in all the case studies that we used
+Typed Lua to annotate Lua code.
+This means that Lua programmers often use possibly \texttt{nil} values
+before checking whether it is \texttt{nil}.
 

### 2015-02-03 doc/thesis/resumo.tex
Adding a pt_BR abstract

Linguagens dinamicamente tipadas, tais como Lua, não usam tipos estáticos em
favor de simplicidade e flexibilidade, porque a ausência de tipos estáticos
significa que programadores não precisam se preocupar em abstrair tipos que
devem ser validados por um verificador de tipos.
Por outro lado, linguagens estaticamente tipadas ajudam na detecção prévia de
diversos \emph{bugs} e também ajudam na estruturação de programas grandes.
Tais pontos geralmente são vistos como duas vantagens que levam programadores
a migrar de uma linguagem dinamicamente tipada para uma linguagem estaticamente tipada,
quando os pequenos \emph{scripts} deles evoluem para programas complexos.
Sistemas de tipos opcionais nos permitem combinar tipagem dinâmica e estática na
mesma linguagem, sem influenciar a semântica original da linguagem, tornando mais
fácil a evolução de código tipado dinamicamente para código tipado estaticamente.
Desenvolver um sistema de tipos opcional para uma linguagem dinamicamente tipada é
uma tarefa desafiadora, pois ele deve ser o mais natural possível para os programadores
que já estão familiarizados com essa linguagem.
Neste trabalho nós apresentamos e formalizamos Typed Lua, um sistema de tipos opcional
para Lua, o qual introduz novas características para tipar estaticamente alguns idiomas
e características de Lua.
Embora Lua compartilha várias características com outras linguagens dinamicamente
tipadas, em particular JavaScript, Lua também possui várias características não usuais,
as quais não estão presentes nos sistemas de tipos dessas linguagens.
Nós discutimos como Typed Lua tipa estaticamente características tais como
funções com aridade flexível que interagem com atribuições múltiplas,
funções que são sobrecarregadas no número de valores de retorno e a evolução
incremental de estruturas e objetos.
Discutimos também nossas decisões de projeto e apresentamos uma avaliação de
resultados, a qual obtemos ao usar Typed Lua para tipar código Lua existente.

### 2015-02-06 doc/thesis/evaluation.tex
Minor improvements to abstract and adding a chapter to related work

@@
 Typed Lua's description files.
-In Section \ref{sec:cases} we present our evaluation results
-and discuss some interesting cases.
-
-We also evaluate our contributions through the revision of related work.
-In Section \ref{sec:relatedwork} we discuss what are the novel
-features of our type system, what are the features that it shares
-with other type systems, and whether we can incorporate some
-interesting features from other type systems.
-
-\section{Case Studies}
-\label{sec:cases}
+In this chapter we present our evaluation results and discuss some
+interesting cases.
 
@@
 
-\subsection{Lua Standard Library}
+\section{Lua Standard Library}
 
@@
 
-\subsection{MD5}
+\section{MD5}
 
@@
 
-\subsection{LuaSocket}
+\section{LuaSocket}
 
@@
 
-\subsection{HTTP Digest}
+\section{HTTP Digest}
 
@@
 
-\subsection{Typical}
+\section{Typical}
 
@@
 
-\subsection{Modulo 11}
+\section{Modulo 11}
 
@@
 
-\subsection{Typed Lua Compiler}
+\section{Typed Lua Compiler}
 
@@
 before checking whether it is \texttt{nil}.
-
-\section{Related Work}
-\label{sec:relatedwork}
-
-In this section we review related work.
-We split this section into two parts: in the first part we review
-other Lua projects, while in the second part we review other
-projects that are not related to Lua.
-
-\subsection{Other Lua projects}
-
-Metalua \citep{metalua} is a Lua compiler that supports compile-time
-metaprogramming (CTMP).
-CTMP is a kind of macro system that allows the programmers to interact
-with the compiler \citep{fleutot2007contrasting}. 
-Metalua extends Lua 5.1 syntax to include its macro system,
-and allows programmers to define their own syntax.
-Metalua can provide syntactical support for several object-oriented
-styles, and can also provide syntax for turning simple type
-annotations into run-time assertions.
-
-MoonScript \citep{moonscript} is a programming language that supports
-class-based object-oriented programming.
-MoonScript compiles to idiomatic Lua code, but 
-it does not perform compile-time type checking.
-
-LuaInspect \citep{luainspect} is a tool that uses MetaLua to perform
-some code analysis.
-For instance, it flags unknown global variables and table fields,
-it checks the number of function arguments against signatures, and
-it infers function return values.
-However, it does not try to analyze object-oriented code and
-it does not perform compile-time type checking.
-
-Tidal Lock \citep{tidallock} is a prototype of another optional type
-system for Lua, which is written in Metalua.
-Tidal Lock covers a little subset of Lua.
-Statements include declaration of local variables, multiple assignments,
-function application, and the return statement.
-This means that Tidal Lock does not include any control-flow statement.
-Expressions include primitive literals, table indexing, function application,
-function declaration, and the table constructor, but they do not include
-binary operations.
-
-A remarkable feature of Tidal Lock is the refinement of table types.
-This feature inspired us to also include it in Typed Lua,
-but in a simpler way and with different formalization.
-
-The table type from Tidal Lock can only represent records, that is,
-it cannot describe hash tables and arrays yet, though we can refine them.
-Tidal Lock also includes field types to describe the type of the fields
-of a table type.
-The field types describe if a table field is mutable or immutable
-in a table type.
-Field types are the feature that allow the refinement of table types in
-Tidal Lock.
-
-Tidal Lock is also a structural type system that relies on subtyping and
-local type inference.
-However, it does not support union types, recursive types, and variadic types.
-It also does not type any object-oriented idiom.
-
-Luacheck \citep{luacheck} is a tool that performs static analysis on Lua code.
-It can flag access to undeclared globals and unused local variables,
-but it does not perform static type checking.
-
-Ravi \citep{ravi} is an experimental Lua dialect.
-Ravi introduces optional static typing for Lua to improve run-time performance.
-To do that, Ravi extends the Lua Virtual Machine to include new
-operations that take into account static type information.
-
-\subsection{Other projects}
-
-Typed Racket \citep{tobin-hochstadt2008ts} is a statically typed version
-of the Racket language, which is a Scheme dialect.
-The main purpose of Typed Racket is to allow programmers to combine
-untyped modules, which are written in Racket, with typed modules, which are
-written in Typed Racket.
-It also uses local type inference to deduce the type of unannotated expressions.
-
-The main feature of Typed Racket's type system is \emph{occurrence typing}
-\citep{tobin-hochstadt2010ltu}.
-It is a novel way to use type predicates in control flow statements
-to refine union types.
-Occurrence typing is not sound in the presence of mutation.
-As these kinds of checks are common in other languages, related systems
-have appeared \citep{guha2011tlc,winther2011gtp,pearce2013ccf}.
-
-The type system of Typed Racket also includes function types, recursive
-types, and structure types.
-Its function types also handle multiple return values, and there is
-also a way to describe function types that have optional arguments.
-Its structure types are similar to our interfaces, as they describe record types.
-The type system is also structural and based on subtyping.
-It also includes the dynamic type \texttt{Any}, which is the top type in the system.
-Typed Racket also supports polymorphic functions and data structures.
-
-Typed Clojure \citep{bonnaire-sergeant2012typed-clojure} is an
-optional type system for Clojure.
-Although Clojure is a Lisp dialect that runs on the Java Virtual Machine,
-Common Language Runtime, and JavaScript, Typed Clojure runs only on
-the Java Virtual Machine.
-Perhaps, this restriction pushed Typed Clojure to support Java classes
-and some Java types such as \texttt{Long}, \texttt{Double}, and \texttt{String}.
-Typed Clojure also provides optional type annotations and uses
-local type inference to deduce the type of unannotated expressions.
-It also assigns the type \texttt{Any} to unannotated function parameters,
-which is the top type in the type system.
-
-The type system of Typed Clojure includes polymorphic function types,
-union types, intersection types, lists, vectors, maps, sets, and recursive types.
-Function types can also have rest parameters, which are similar
-to our variadic types, but can only appear on the input parameter
-of function types.
-In fact, its function types cannot return multiple results.
-It also uses occurrence typing to allow control flow statements to
-refine union types.
-The type system is also structural and based on subtyping.
-
-Dart \citep{dart} is a new class-based object-oriented programming
-language.
-It includes optional type annotations and compiles to JavaScript.
-The type system of Dart is nominal and includes base types,
-function types, lists, and maps.
-It also supports generics, and the programmer can define
-generic functions, lists, and maps.
-Unlike Typed Lua, Dart is unsound by design.
-
-Even though Dart has optional typing and static types by
-default do not affect run-time semantics, it has an
-execution mode that affects run-time.
-The \emph{checked mode} inserts run-time assertions that
-verifies whether static types match run-time tags.
-The \emph{production mode} is the default execution mode
-that does not include any assertions.
-
-TypeScript \citep{typescript} is a JavaScript extension
-that includes optional type annotations and class-based
-object-oriented programming.
-It also uses local type inference to deduce the type
-of unannotated expressions.
-The type system of TypeScript is structural, based
-on subtyping, and supports generics.
-It includes the dynamic type, primitive types, union types,
-function types, array types, tuple types, recursive types, and
-object types.
-Unlike Typed Lua, TypeScript uses arrays to represent variadic
-functions and multiple return values.
-
-Even though TypeScript is unsound by design,
-\citet{bierman2014typescript} shows how to make TypeScript sound.
-They use a reduced core of TypeScript to formalize a
-sound type system for TypeScript, but also to formalize
-its current unsound type system.
-
-TeJaS \citep{lerner2013tejas} is a framework for the construction of
-different type systems for JavaScript.
-The authors created a base type system for JavaScript with
-extensible typing rules that allow the experimentation of
-different static analysis.
-They used TeJaS to create a type system that simulates the
-type system of TypeScript.
-
-\citet{politz2012semantics} proposes semantics and types for objects
-with first-class member names, a well-known feature from scripting languages.
-Their type system uses string patterns to describe the members of
-an object, and define a complex subtyping relation to validate
-these patterns.
-They also provide an implementation of their system to JavaScript.
-
-Gradualtalk \citep{allende2013gts} is a Smalltalk dialect that
-supports gradual typing.
-The type system combines nominal and structural typing.
-It includes function types, union types, structural types,
-nominal types, a self type, and parametric polymorphism.
-The type system also relies on subtyping and consistent-subtyping.
-
-Gradualtalk inserts run-time checks that ensure dynamically
-typed code does not violate statically typed code.
-\citet{allende2013cis} perform a carefull evaluation about
-cast insertion on Gradualtalk.
-They report that usually cast insertions impact on execution
-performance, so Gradualtalk also has an option that allows
-programmers to turn them off, downgrading Gradualtalk
-to an optional type system.
-
-Reticulated Python \citep{vitousek2014deg} is a Python compiler
-that supports gradual typing.
-The type system is structural and based on subtyping.
-It includes base types, the dynamic type, list types,
-dictionary types, tuple types, function types, set types,
-object types, class types, and recursive types.
-It includes class and object types to differentiate the
-type of class declarations and instances, respectively.
-It also uses local type inference.
-Besides static type checking, Reticulated Python also introduces
-three different approaches for inserting run-time assertions.
-
-Hack \citep{hack} is a new programming language that runs on the
-Hip Hop Virtual Machine (HHVM).
-The HHVM is a virtual machine that executes Hack and PHP programs.
-We can view Hack as an extension to PHP that combines static and
-dynamic typing.
-The type system of Hack includes generics, nullable types, collections,
-and function types.
-
-The Ruby Type Checker \citep{ren2013rtc} is a library that
-performs type checking during run-time.
-The library provides type annotations that the programmer
-can use on classes and methods.
-Its type system includes nominal types, union types,
-intersection types, method types, parametric polymorphism,
-and type casts.
-
-Grace \citep{black2013sg} is an object-oriented language
-with optional typing.
-Grace is not a dynamically typed language that has been
-extended with an optional type system, but a language
-that has been designed from scratch to have both
-static and dynamic typing.
-\citet{homer2013modules} explores some useful patterns
-that derive from Grace's use of objects as modules
-and its brand of optional structural typing, which
-can also be expressed with Typed Lua's modules as tables.

### 2015-02-06 doc/thesis/related.tex
Minor improvements to abstract and adding a chapter to related work

In this chapter we review related work as a way to indentify the novel
features of our type system, what are the features that it shares with
other type systems, and whether we can incorporate some interesting
features from other type systems.
We split this chapter into two parts: in the first part we review
other Lua projects, while in the second part we review other
projects that are not related to Lua.

\section{Other Lua projects}

Metalua \citep{metalua} is a Lua compiler that supports compile-time
metaprogramming (CTMP).
CTMP is a kind of macro system that allows the programmers to interact
with the compiler \citep{fleutot2007contrasting}.
Metalua extends Lua 5.1 syntax to include its macro system,
and allows programmers to define their own syntax.
Metalua can provide syntactical support for several object-oriented
styles, and can also provide syntax for turning simple type
annotations into run-time assertions.

MoonScript \citep{moonscript} is a programming language that supports
class-based object-oriented programming.
MoonScript compiles to idiomatic Lua code, but
it does not perform compile-time type checking.

LuaInspect \citep{luainspect} is a tool that uses MetaLua to perform
some code analysis.
For instance, it flags unknown global variables and table fields,
it checks the number of function arguments against signatures, and
it infers function return values.
However, it does not try to analyze object-oriented code and
it does not perform compile-time type checking.

Tidal Lock \citep{tidallock} is a prototype of another optional type
system for Lua, which is written in Metalua.
Tidal Lock covers a little subset of Lua.
Statements include declaration of local variables, multiple assignments,
function application, and the return statement.
This means that Tidal Lock does not include any control-flow statement.
Expressions include primitive literals, table indexing, function application,
function declaration, and the table constructor, but they do not include
binary operations.

A remarkable feature of Tidal Lock is the refinement of table types.
This feature inspired us to also include it in Typed Lua,
but in a simpler way and with different formalization.

The table type from Tidal Lock can only represent records, that is,
it cannot describe hash tables and arrays yet, though we can refine them.
Tidal Lock also includes field types to describe the type of the fields
of a table type.
The field types describe if a table field is mutable or immutable
in a table type.
Field types are the feature that allow the refinement of table types in
Tidal Lock.

Tidal Lock is also a structural type system that relies on subtyping and
local type inference.
However, it does not support union types, recursive types, and variadic types.
It also does not type any object-oriented idiom.

Luacheck \citep{luacheck} is a tool that performs static analysis on Lua code.
It can flag access to undeclared globals and unused local variables,
but it does not perform static type checking.

Ravi \citep{ravi} is an experimental Lua dialect.
Ravi introduces optional static typing for Lua to improve run-time performance.
To do that, Ravi extends the Lua Virtual Machine to include new
operations that take into account static type information.

\section{Other projects}

Typed Racket \citep{tobin-hochstadt2008ts} is a statically typed version
of the Racket language, which is a Scheme dialect.
The main purpose of Typed Racket is to allow programmers to combine
untyped modules, which are written in Racket, with typed modules, which are
written in Typed Racket.
It also uses local type inference to deduce the type of unannotated expressions.

The main feature of Typed Racket's type system is \emph{occurrence typing}
\citep{tobin-hochstadt2010ltu}.
It is a novel way to use type predicates in control flow statements
to refine union types.
Occurrence typing is not sound in the presence of mutation.
As these kinds of checks are common in other languages, related systems
have appeared \citep{guha2011tlc,winther2011gtp,pearce2013ccf}.

The type system of Typed Racket also includes function types, recursive
types, and structure types.
Its function types also handle multiple return values, and there is
also a way to describe function types that have optional arguments.
Its structure types are similar to our interfaces, as they describe record types.
The type system is also structural and based on subtyping.
It also includes the dynamic type \texttt{Any}, which is the top type in the system.
Typed Racket also supports polymorphic functions and data structures.

Typed Clojure \citep{bonnaire-sergeant2012typed-clojure} is an
optional type system for Clojure.
Although Clojure is a Lisp dialect that runs on the Java Virtual Machine,
Common Language Runtime, and JavaScript, Typed Clojure runs only on
the Java Virtual Machine.
Perhaps, this restriction pushed Typed Clojure to support Java classes
and some Java types such as \texttt{Long}, \texttt{Double}, and \texttt{String}.
Typed Clojure also provides optional type annotations and uses
local type inference to deduce the type of unannotated expressions.
It also assigns the type \texttt{Any} to unannotated function parameters,
which is the top type in the type system.

The type system of Typed Clojure includes polymorphic function types,
union types, intersection types, lists, vectors, maps, sets, and recursive types.
Function types can also have rest parameters, which are similar
to our variadic types, but can only appear on the input parameter
of function types.
In fact, its function types cannot return multiple results.
It also uses occurrence typing to allow control flow statements to
refine union types.
The type system is also structural and based on subtyping.

Dart \citep{dart} is a new class-based object-oriented programming
language.
It includes optional type annotations and compiles to JavaScript.
The type system of Dart is nominal and includes base types,
function types, lists, and maps.
It also supports generics, and the programmer can define
generic functions, lists, and maps.
Unlike Typed Lua, Dart is unsound by design.

Even though Dart has optional typing and static types by
default do not affect run-time semantics, it has an
execution mode that affects run-time.
The \emph{checked mode} inserts run-time assertions that
verifies whether static types match run-time tags.
The \emph{production mode} is the default execution mode
that does not include any assertions.

TypeScript \citep{typescript} is a JavaScript extension
that includes optional type annotations and class-based
object-oriented programming.
It also uses local type inference to deduce the type
of unannotated expressions.
The type system of TypeScript is structural, based
on subtyping, and supports generics.
It includes the dynamic type, primitive types, union types,
function types, array types, tuple types, recursive types, and
object types.
Unlike Typed Lua, TypeScript uses arrays to represent variadic
functions and multiple return values.

Even though TypeScript is unsound by design,
\citet{bierman2014typescript} shows how to make TypeScript sound.
They use a reduced core of TypeScript to formalize a
sound type system for TypeScript, but also to formalize
its current unsound type system.

TeJaS \citep{lerner2013tejas} is a framework for the construction of
different type systems for JavaScript.
The authors created a base type system for JavaScript with
extensible typing rules that allow the experimentation of
different static analysis.
They used TeJaS to create a type system that simulates the
type system of TypeScript.

\citet{politz2012semantics} proposes semantics and types for objects
with first-class member names, a well-known feature from scripting languages.
Their type system uses string patterns to describe the members of
an object, and define a complex subtyping relation to validate
these patterns.
They also provide an implementation of their system to JavaScript.

Gradualtalk \citep{allende2013gts} is a Smalltalk dialect that
supports gradual typing.
The type system combines nominal and structural typing.
It includes function types, union types, structural types,
nominal types, a self type, and parametric polymorphism.
The type system also relies on subtyping and consistent-subtyping.

Gradualtalk inserts run-time checks that ensure dynamically
typed code does not violate statically typed code.
\citet{allende2013cis} perform a carefull evaluation about
cast insertion on Gradualtalk.
They report that usually cast insertions impact on execution
performance, so Gradualtalk also has an option that allows
programmers to turn them off, downgrading Gradualtalk
to an optional type system.

Reticulated Python \citep{vitousek2014deg} is a Python compiler
that supports gradual typing.
The type system is structural and based on subtyping.
It includes base types, the dynamic type, list types,
dictionary types, tuple types, function types, set types,
object types, class types, and recursive types.
It includes class and object types to differentiate the
type of class declarations and instances, respectively.
It also uses local type inference.
Besides static type checking, Reticulated Python also introduces
three different approaches for inserting run-time assertions.

Hack \citep{hack} is a new programming language that runs on the
Hip Hop Virtual Machine (HHVM).
The HHVM is a virtual machine that executes Hack and PHP programs.
We can view Hack as an extension to PHP that combines static and
dynamic typing.
The type system of Hack includes generics, nullable types, collections,
and function types.

The Ruby Type Checker \citep{ren2013rtc} is a library that
performs type checking during run-time.
The library provides type annotations that the programmer
can use on classes and methods.
Its type system includes nominal types, union types,
intersection types, method types, parametric polymorphism,
and type casts.

Grace \citep{black2013sg} is an object-oriented language
with optional typing.
Grace is not a dynamically typed language that has been
extended with an optional type system, but a language
that has been designed from scratch to have both
static and dynamic typing.
\citet{homer2013modules} explores some useful patterns
that derive from Grace's use of objects as modules
and its brand of optional structural typing, which
can also be expressed with Typed Lua's modules as tables.

### 2015-02-09 doc/thesis/system.tex
Minor improvements to the text about case studies
Adding regular table types to the text

@@
 & | \; s \rightarrow s & \textit{function types}\\
-& | \; \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed|open|unique} & \textit{table types}\\
+& | \; \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique|open|closed|regular} & \textit{table types}\\
 & | \; x & \textit{type variables}\\
@@
 The type $\Value$ is an option when we need a loose table type.
-For instance, $\{\Value:\Value\}_{closed}$ represents the type of a
+For instance, $\{\Value:\Value\}_{regular}$ represents the type of a
 table in which both indices and values can have any type.
@@
 
-We also use the tags \emph{closed}, \emph{open}, and \emph{unique}
-to classify table types.
+We also use the tags \emph{unique}, \emph{open}, \emph{closed}, and
+\emph{regular} to classify table types.
 The tag \emph{unique} represents tables with no keys that do not
@@
 have at least one alias.
-The tag \emph{closed} represents table types that do not provide
+The tag \emph{closed} represents \emph{unique} table types
+that may have aliases.
+In particular, the type of a class has this tag.
+The tag \emph{regular} represents table types that do not provide
 any guarantees about keys with types not listed in the table type.
 In particular, in the concrete syntax, type annotations, interface
-declarations, and userdata declarations always describe \emph{closed} table types.
+declarations, and userdata declarations always describe \emph{regular} table types.
 In the next sections we explain in more detail why we need
@@
 where $t$ is a first-level type that $x$ represents.
-For instance, $\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}$
+For instance, $\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{regular}$
 is a type for singly-linked lists of integers.
@@
 Back to the subtyping rules between first-level types,
-the subtyping rule between \emph{closed} table types resembles the
+the subtyping rule among a \emph{closed} or \emph{regular}
+table type and another \emph{regular} table type resembles the
 standard subtyping rule between records:
@@
        \senv \vdash v_{j} \subtype_{c} v_{i}'}
-      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{closed} \subtype \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}} \; m \ge n
-\end{array}
-\]
-
-The rule \textsc{S-TABLE1} allows width subtyping between \emph{closed}
-table types, and introduces the auxiliary relation $\subtype_{c}$ to
-handle depth subtyping on the type of the values stored in the table fields.
+      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{closed|regular} \subtype \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{regular}} \; m \ge n
+\end{array}
+\]
+
+The rule \textsc{S-TABLE1} allows width subtyping and introduces the
+auxiliary relation $\subtype_{c}$ to handle depth subtyping on the
+type of the values stored in the table fields.
 We need an auxiliary relation because the subtyping of the
@@
 
-These rules allow closed table types to have depth subtyping on $\Const$ fields.
+These rules allow depth subtyping on $\Const$ fields.
 The rule \textsc{S-FIELD1} defines that mutable fields are invariant,
@@
 
-There is a limitation on \emph{closed} table types that led us to
+There is a limitation on \emph{regular} table types that led us to
 introduce \emph{open} and \emph{unique} table types.
-If the table constructor had a \emph{closed} table type, then
+If the table constructor had a \emph{regular} table type, then
 programmers would not be able to use it to initialize a variable with
@@
 \[
-\{``x":1, ``y":2\}_{closed} \not\subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
+\{``x":1, ``y":2\}_{regular} \not\subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{regular}
 \]
@@
 not overcome this limitation, as it still would give to the table constructor
-a closed table type without covariant mutable fields.
+a regular table type without covariant mutable fields.
 Thus, programmers would not be able to use the table constructor to
@@
 \begin{align*}
-& \{``x":\Integer, ``y":\Integer\}_{closed} \not\subtype \\
-& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
+& \{``x":\Integer, ``y":\Integer\}_{regular} \not\subtype \\
+& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{regular}
 \end{align*}
@@
 The following subtyping rule defines the subtyping relation among
-\emph{unique} table types and \emph{closed} table types:
+\emph{unique} table types and \emph{regular} table types:
 \[
@@
       {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
-                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
+                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{regular}}
 \end{array}
@@
 It allows covariant keys because we also want to use \emph{unique}
-table types as a way to join table fields that inhabit \emph{closed} table types.
+table types as a way to join table fields that inhabit \emph{regular} table types.
 For instance, we want to use the table constructor to initialize
@@
 \[
-\{``x":1, ``y":2\}_{unique} \subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
-\]
-
-Even though we allow width subtyping between \emph{unique} and \emph{closed}
-table types, we do not allow it among \emph{unique} and \emph{open} or
-\emph{unique} table types because it would violate our definition of
-\emph{unique} and \emph{open} table types:
+\{``x":1, ``y":2\}_{unique} \subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{regular}
+\]
+
+Even though we allow width subtyping between \emph{unique} and \emph{regular}
+table types, we do not allow it among \emph{unique} and other table types
+because it would violate our definition of these other table types:
 \[
@@
       {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
-                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{open|unique}}
-\end{array}
-\]
-
-The rules that handle subtyping among \emph{open} table types and
-\emph{open} or \emph{closed} table types are similar to the rules
-that handle subtyping among \emph{unique} table types and other table types:
-\[
-\begin{array}{c}
+                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{unique|open|closed}}
+\end{array}
+\]
+
+The rule that handles subtyping between \emph{open} and \emph{regular} table
+types allows width subtyping:
+\[
 \begin{array}{c}
@@
       {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
-                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
-\end{array}
-\\ \\
+                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{regular}}
+\end{array}
+\]
+
+However, the rule that handles subtyping among \emph{open} and
+\emph{open} or \emph{closed} table types does not allow width subtyping:
+\[
 \begin{array}{c}
@@
       {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
-                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{open}}
-\end{array}
-\end{array}
-\]
-
-The rule \textsc{S-TABLE4} allows width subtyping, while
-the rule \textsc{S-TABLE5} does not allow width subtyping.
-Both rules allow joining fields plus omitting optional fields.
-Both rules use $\subtype_{c}$ to allow depth on $\Const$ fields only.
+                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{open|closed}}
+\end{array}
+\]
+
+The rules \textsc{S-TABLE4} and \textsc{S-TABLE5} allow joining fields
+plus omitting optional fields.
+Both rules use $\subtype_{c}$ to allow depth subtyping on $\Const$ fields only.
+
+We introduced \emph{closed} table types because we needed a safe way
+to represent the type of classes along with single inheritance.
+The rule that handles subtyping between \emph{closed} table types
+does not allow width subtyping, joining fields, and omitting fields,
+but it allows depth subtyping on $\Const$ fields:
+\[
+\begin{array}{c}
+\mylabel{S-TABLE6}\\
+\dfrac{\forall i \in 1..n \; \exists j \in 1..n \;\;\;
+       \senv \vdash k_{j} \subtype k_{i}' \;\;\;
+       \senv \vdash k_{i}' \subtype k_{j} \;\;\;
+       \senv \vdash v_{j} \subtype_{c} v_{i}'}
+      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed} \subtype \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
+\end{array}
+\]
 
 In the next section we will show in more detail how our type system
-handles \emph{open} and \emph{unique} table types to allow the refinement of
-table types.
+uses \emph{unique}, \emph{open}, and \emph{closed} table types to
+handle the refinement of table types.
 
@@
 \{ & ``info":\Integer, \\
-   & ``next":\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed} \cup \Nil \}_{unique} \subtype \\
-& \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}
+   & ``next":\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{regular} \cup \Nil \}_{unique} \subtype \\
+& \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{regular}
 \end{align*}
@@
 the vararg expression (${...}$) into different syntactic categories.
-It uses $e(el)_{s}$ to denote function applications that return no value
+It uses $e(el)_{s}$ to denote function applications that produce no value
 because they appear as statements,
-$e(el)_{e}$ to denote function applications that return only one value,
-and $e(el)_{m}$ to denote function applications that return multiple values.
-It uses the same categories for method applications, but only ${...}_{e}$
-and ${...}_{m}$, as the vararg expression cannot appear as a statement.
-
-We also include two kinds of type cast in our core language:
+$e(el)_{e}$ to denote function applications that produce only one value,
+even if the function application returns multiple values,
+and $e(el)_{m}$ to denote function applications that produces as many values
+as the function application returns.
+It uses the same categories for method applications.
+For vararg expressions, it only uses ${...}_{e}$ and ${...}_{m}$,
+as the vararg expression cannot appear as a statement.
+
+We also include two kinds of type coercions in our core language:
 the expression ${<}t{>} \;n$ and the left-hand value $n[c] \; {<}t{>}$.
-We will show that the first helps allowing safe aliasing between
-table types, while the second allows the refinement of table types.
-Due to the refinement of table types we also split variable names
-into two categories: $n_{e}$ when they appear as expressions and
-$n_{l}$ when they appear as left-hand values.
-
-We use three different typing relations to present our typing rules
-as a deduction system for these typing relations.
-We use the relation $\env_{1}, \penv \vdash s, \env_{2}$ for typing
-statements.
-This relation means that given the type environment $\env_{1}$
-and the projection environment $\penv$,
-typing the statement $s$ produces the new type environment $\env_{2}$.
-We use the relation $\env_{1}, \penv \vdash e : t, \env_{2}$ for
-typing expressions.
-This relation means that given the type environment $\env_{1}$
-and the projection environment $\penv$,
-the expression $e$ has type $t$ and produces the new type environment $\env_{2}$.
-We use the relation $\env_{1}, \penv \vdash el : r_{1}, \env_{2}, (x,r_{2})$
-for typing expression lists.
-This relation means that given type type environment $\env_{1}$
-and the projection environment $\penv$,
-the expression list $el$ has type $r_{1}$ and produces the new type
-environment $\env_{2}$ and produces the pair $(x,r_{2})$.
-We need three typing relations because statements do not have types
-and expression lists can result in a union of tuple types that
-should be introduced in the projection environment.
-We use $\env_{1}, \penv \vdash el : r, \env_{2}$ when the expression
-list does not result in a union of tuple types.
-We need two environments because $\env$ maps variables to types,
-while $\penv$ maps projection variables to unions of tuple types.
-We use $\env_{1}[n \mapsto t]$ for meaning that we extended the environment
-$\env_{1}$ with the variable $n$ that maps to type $t$.
-We use $\penv[x \mapsto r]$ for meaning that we extended the environment
-$\penv$ with the projection variable $x$ that maps to the union of tuple types $r$.
-
-Lua has multiple assignments, and the rule \textsc{T-ASSIGNMENT} uses
+Both allow the refinement of table types.
+We also split variable names into two categories to have safe aliasing
+of tables in the presence of refinement.
+We use $n_{e}$ when they appear as expressions and $n_{l}$ when they
+appear as left-hand values.
+
+We present the typing rules as a deduction system for the typing relations
+$\env_{1}, \penv \vdash s, \env_{2}$ and $\env_{1}, \penv \vdash e : t, \env_{2}$.
+We use the first one for typing statements and the second one for typing expressions.
+The first relation means that given a type environment $\env_{1}$
+and a projection environment $\penv$, typing a statement $s$ produces
+a new type environment $\env_{2}$.
+The second relation means that given a type environment $\env_{1}$
+and a projection environment $\penv$, typing a expression $e$ has
+type $t$ and produces a new type environment $\env_{2}$.
+We need two typing relations because only expressions have types,
+though expressions and statements can modify a type environment.
+
+Even though we can assign only first-level types to variables,
+our type system should be able to project unions of first-level types
+from unions of second-level types.
+Thus, we need two environments because $\env$ maps variables to first-level types,
+while $\penv$ maps variables to second-level types.
+We will show how our type system uses the environment $\penv$ for
+handling projection types.
+We use $\env_{1}[n \mapsto t]$ to extend the environment $\env_{1}$
+with the variable $n$ that maps to type $t$.
+We use $\penv[x \mapsto r]$ to extend the environment $\penv$ with
+the variable $x$ that maps to the second-level type $r$.
+
+\subsection{Assignment and function application}
+
+Lua has multiple assignment, and the rule \textsc{T-ASSIGNMENT} uses
 tuple types along with consistent-subtyping to type check this Lua feature:
@@
 types $\Integer$ and $\String$, respectively.
-Our type system uses $\Value{*}$ to drop extra values
-and $\Nil{*}$ to replace missing values.
-In this example, the rules \textsc{T-LHSLIST} and \textsc{T-EXPLIST2}
-are the rules that handle these features.
-
+
+Our type system uses the type $\Value{*}$ to drop extra values
+and the type $\Nil{*}$ to replace missing values.
+The rule that type checks left-hand values list introduces the type
+$\Value{*}$ to let the right-hand side produce more values than
+expected in the left-hand side.
+The rules that type check expression list introduce the type
+$\Nil{*}$ to let the right-hand side produce fewer values than
+expected in the left-hand side.
+Our type system includes $\Nil{*}$ in the type of an expression list only
+if its type does not end in another variadic type $t{*}$.
+In the previous example, the rules \textsc{T-LHSLIST} and \textsc{T-EXPLIST2}
+are the rules that introduce these types.
 We define the rule \textsc{T-LHSLIST} as follows:
@@
 \mylabel{T-LHSLIST}\\
-\dfrac{\env, \penv \vdash l_{k}:t_{k}, \env_{k} \;\;\;
+\dfrac{\env, \penv \vdash l_{i}:t_{i}, \env_{i} \;\;\;
        \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
@@
 First, this rule uses the same environment $\env$ to type check each
-left-hand side value $l_{k}$, as they can perform different changes
-in the environment.
-Then, it takes each type $t_{k}$ and builds the tuple type for
-the assignment rule.
-It places the type $\Value{*}$ in the end to discard extra values.
-The rule succeeds if the predicate \emph{merge} can
-produce a new environment, which includes all the modifications
-of each environment $\env_{k}$.
-Intuitively, the predicate \emph{merge} fails when an environment
-tries to change the type of a variable towards a type that is not
-a subtype of its previous type.
-We will show in this section that the refinement of table types can
-change the environment, and that is the reason why we need to check
-whether the environment is consistent.
-
-The rule \textsc{T-EXPLIST2} is close to the rule \textsc{T-LHSLIST}.
-The difference is that it uses $\Nil{*}$ to replace missing values.
-We define this rule as follows:
+left-hand side value $l_{i}$.
+Then, it takes each type $t_{i}$ and builds the tuple type for
+the assignment rule, placing $\Value{*}$ in the end.
+Table refinement can change the type environment while typing each left-hand value.
+Thus, we use a partial auxiliary function \emph{merge} to collect
+all the changes in a new environment $\env_{m}$, if there are no conflicts.
+
+The rule \textsc{T-EXPLIST2} is similar to \textsc{T-LHSLIST}:
 \[
@@
 \mylabel{T-EXPLIST2}\\
-\dfrac{\env, \penv \vdash e_{k}:t_{k}, \env_{k} \;\;\;
+\dfrac{\env, \penv \vdash e_{i}:t_{i}, \env_{i} \;\;\;
        \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
@@
 
-The typing rules that type check expression lists include $\Nil{*}$
-in the end of the tuple type only when it does not end in a variadic type.
-For instance, the rules \textsc{T-EXPLIST3} and \textsc{T-EXPLIST4}
-handle the case that an expression list is just an expression that
-return multiple values, but only \textsc{T-EXPLIST3} needs to add
-$\Nil{*}$:
+We use \textsc{T-EXPLIST2} only when the last expression in the list
+can produce a single value.
+Rules \textsc{T-EXPLIST3} and \textsc{T-EXPLIST4} handle the case where
+an expression list is just an expression that may produce multiple values:
 \[
@@
 \mylabel{T-EXPLIST3}\\
-\dfrac{\env, \penv \vdash me:t_{1} \times ... \times t_{n} \times \Void, \env_{1}}
-      {\env, \penv \vdash me:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{1}}
+\dfrac{\env_{1}, \penv \vdash me:t_{1} \times ... \times t_{n} \times \Void, \env_{2}}
+      {\env_{1}, \penv \vdash me:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{2}}
 \end{array}
@@
 \mylabel{T-EXPLIST4}\\
-\dfrac{\env, \penv \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{1}}
-      {\env, \penv \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{1}}
-\end{array}
-\end{array}
-\]
-
-As an example, the rule \textsc{T-EXPLIST3} allows our type system
-to type check the following example:
+\dfrac{\env_{1}, \penv \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{2}}
+      {\env_{1}, \penv \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{2}}
+\end{array}
+\end{array}
+\]
+
+As an example, these rules allow our type system to type check the following example:
 \[
@@
 the environment with types $\Integer$, $\String$, $\String \cup \Nil$, and
-$\Void \rightarrow \Integer \times \String \times \Void$, respectively.
+either $\Void \rightarrow \Integer \times \String \times \Void$ or
+$\Value{*} \rightarrow \Integer \times \String \times \Nil{*}$,
+depending on whether the programmer is using strict function types.
 Thus, the example type checks because
@@
 
-Multiple assignments also appear in function applications.
+Rules for function applications are similar to the rule for multiple assignment.
 The rule \textsc{T-APPLY1} handles the case where function applications
-are expressions that return multiple values:
+are expressions that produce multiple values:
 \[
@@
 that handle the cases where function applications are either statements
-or expressions that return only one value:
+or expressions that produce only one value:
 \[
@@
 
-The rule \textsc{T-STMAPPLY1} discards the returned values,
-while the rule \textsc{T-EXPAPPLY1} uses the predicate \emph{first} to
-ensure that only one value is returned.
-We can use pattern matching to the define the recursive predicate
-\emph{first} as follows:
+The rule \textsc{T-STMAPPLY1} discards the produced values,
+while the rule \textsc{T-EXPAPPLY1} uses the auxiliary function
+\emph{first} to ensure that only one value is produced.
+We can define \emph{first} inductively as follows:
 \begin{align*}
@@
 first(t{*}) & = t \cup \Nil\\
-first(t \times s) & = t\\
-first(s_{1} \sqcup s_{2}) & = first(s_{1}) \cup first(s_{2})
+first(t \times r) & = t\\
+first(r_{1} \sqcup r_{2}) & = first(r_{1}) \cup first(r_{2})
 \end{align*}
@@
 \[
-f(``foo")_{s}
-\]
-type checks through the rules \textsc{T-STMAPPLY1} and \textsc{T-APPLY1}, because
+f(``foo")_{m}
+\]
+type checks through the rule \textsc{T-APPLY1}, because
 \[
@@
 \[
-x = f(``foo",1,2,3)_{m}
-\]
-also type checks through the rules \textsc{T-STMAPPLY1} and \textsc{T-APPLY1}, because
+f(``foo",1,2,3)_{m}
+\]
+also type checks through the rule \textsc{T-APPLY1}, because
 \[
@@
 \[
-x = f(``foo",1,2,3)_{m}
-\]
-does not type check through the rules \textsc{T-STMAPPLY1} and \textsc{T-APPLY1}, because
+f(``foo",1,2,3)_{m}
+\]
+does not type check through the rule \textsc{T-APPLY1}, because
 \[
@@
 When our type system type checks an expression list,
-it always include $\Nil{*}$ in the end of the type of this expression list,
-if the type does not end in a variadic type.
-Our type system has this behavior because $\Nil{*}$ along with subtyping
-and consistent-subtyping use $\Nil$ to replace missing values.
+it always includes $\Nil{*}$ in the end of the type of this expression list
+if its type does not end in a variadic type.
 This behavior preserves the semantics of Lua on replacing missing values,
-as it is necessary when we omit optional parameters in a function call,
+and it is necessary when we omit optional parameters in a function call,
 like the previous example showed.
@@
 Using $\Nil{*}$ in the end of the type of expression lists also allows
-our type system to catch arity mismatch in function calls where
-the type of the function does not include optional parameters.
+our type system to catch arity mismatch in function calls without optional parameters.
 For instance, assuming that $f$ has type
@@
 \[
-f(1)_{s}
-\]
-does not type check through the rules \textsc{T-STMAPPLY1} and \textsc{T-APPLY1}, because
+f(1)_{m}
+\]
+does not type check through the rule \textsc{T-APPLY1}, because
 \[
@@
 \[
-x = f(1,2,3)_{m}
-\]
-also does not type check through the rules \textsc{T-STMAPPLY1} and \textsc{T-APPLY1}, because
+f(1,2,3)_{m}
+\]
+also does not type check through the rule \textsc{T-APPLY1}, because
 \[
@@
 
-The same behavior happens with the rule that type checks the \textbf{return}
-statement:
-\[
-\begin{array}{c}
-\mylabel{T-RETURN}\\
-\dfrac{\env_{1}, \penv \vdash el:r_{1}, \env_{2} \;\;\;
-       \env_{2}(\ret) = r_{2} \;\;\;
-       r_{1} \lesssim r_{2}}
-      {\env_{1}, \penv \vdash \mathbf{return} \; el,\env_{2}}
-\end{array}
-\]
-
-Now, we move the discussion to the rules that type check the
-different table constructors that our abstract syntax includes,
-as they will appear in the next examples:
-
-The simplest form of the table constructor is the empty table $\{\}$.
-Its rule is straightforward:
+\subsection{Tables and refinement}
+
+The simples expression involving tables in the empty table constructor.
+Its type checking rule is straightforward:
 \[
@@
 We did not include in the abstract syntax a table constructor
-that uses a list of expressions, because we can write it using the first form.
-For instance, we can write the table constructor
+that uses a list of expressions, because we can express it using the first form.
+For instance, we can use the table constructor
 $\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to express $\{ ``x", ``y", ``z" \}$.
-We chose to not include the latter form because its type checking rule is similar
-to the type checking rule of the first form, and it just controls the type of
-the omitted indices.
-For the same reason, we did not include table constructors with
-multiple expressions, which can return a tuple type that does not end
-with a variadic type.
-We did include the vararg expression in the last position to show how
-our type system can type check a table type that has a record part and
-also an array part.
 
@@
 \mylabel{T-CONSTRUCTOR2}\\
-\dfrac{\env_{i}, \penv \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
+\dfrac{\env_{i}, \penv \vdash p_{i}:(k_{i},v_{i}), \env_{i+1} \;\;\;
        n = |\;\vec{p}\;| \;\;\;
-       t = wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique})}
+       t = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique} \;\;\;
+       wf(t)}
       {\env_{1}, \penv \vdash \{\;\vec{p}\;\}:t, \env_{n+1}}
@@
 
-The rule \textsc{T-CONSTRUCTOR2} type checks each table field $p_{i}$,
-uses the type of the table fields to build the table type, and
-uses the predicate \emph{wf} to check whether the table type is
-well formed.
+First, the rule \textsc{T-CONSTRUCTOR2} uses the auxiliary relation
+$\env_{1}, \penv \vdash p : (k,v), \env_{2}$ to type check each table
+field $p_{i}$.
+This auxiliary relation means that given a type environment $\env_{1}$
+and a projection environment $\penv$, typing a table field $p$
+produces the pair $(k,v)$ and a new type environment $\env_{2}$. 
+In the pair $(k,v)$, $k$ is type key type, while $v$ is the value type.
+
+Then, the rule \textsc{T-CONSTRUCTOR2} uses each pair $(k_{i},v_{i})$
+to build the table type that express the type of the given constructor, and
+uses the predicate \emph{wf} to check whether this table type is well-formed.
 Formally, a table type is well-formed if it obeys the following rule:
@@
 
-Well-formed table types avoid ambiguous table types.
+Well-formed table types avoid ambiguity.
 For instance, this rule detects that the table type
@@
 
-The type checking of the table fields would be straightforward
-if our table types allowed any first-level type in the type of the keys.
+The definition of the auxiliary relation that type checks table
+fields would be straightforward if our table types allowed any
+first-level type in the type of the keys.
 The rules \textsc{T-FIELD1}, \textsc{T-FIELD2}, and \textsc{T-FIELD3}
@@
        t_{1} \subtype \Boolean}
-      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
+      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (t_{1},rt(t_{2})), \env_{3}}
 \end{array}
@@
        t_{1} \subtype \Number}
-      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
+      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (t_{1},rt(t_{2})), \env_{3}}
 \end{array}
@@
        t_{1} \subtype \String}
-      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
-\end{array}
-\end{array}
-\]
-
-These rules use the predicate \emph{close} in the type of the value
-because our type system do not allow \emph{open} and \emph{unique}
-table types to appear in the type of the values.
-The predicate \emph{close} promotes all table types to \emph{closed}.
+      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (t_{1},rt(t_{2})), \env_{3}}
+\end{array}
+\end{array}
+\]
+
+These rules use the function \emph{rt} in the type of the value
+because our type system allows only \emph{regular} table types to
+appear in the type of the values.
+The function \emph{rt} promotes all table types to \emph{regular}.
 We made this restriction because our type system does not keep track
 of aliases to table fields.
-This means that allowing \emph{open} and \emph{unique} table types
-to appear in the type of the values would allow the creation of
-unsafe aliases.
+This means that allowing \emph{unique}, \emph{open}, and \emph{closed}
+table types to appear in the type of the values would allow the creation
+of unsafe aliases.
 
@@
        \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3}}
-      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: \Value{:}close(t_{2}), \env_{3}}
+      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (\Value,rt(t_{2})), \env_{3}}
 \end{array}
@@
 As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
-has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ through the rules
+has type $\{``x":1, ``y":\{``z":2\}_{regular}\}_{unique}$ through the rules
 \textsc{T-FIELD3} and \textsc{T-CONSTRUCTOR2}.
@@
 \mylabel{T-CONSTRUCTOR3}\\
-\dfrac{\env_{1}({...}) = t}
-      {\env_{1}, \penv \vdash \{{...}_{m}\}:\{\Integer{:}t \cup \Nil\}_{unique}, \env_{1}}
+\dfrac{\env({...}) = t}
+      {\env, \penv \vdash \{{...}_{m}\}:\{\Integer{:}t \cup \Nil\}_{unique}, \env}
 \end{array}
@@
 \dfrac{\begin{array}{c}
-       \env_{i}, \penv \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
+       \env_{i}, \penv \vdash p_{i}:(k_{i},v_{i}), \env_{i+1} \;\;\;
        n = |\;\vec{p}\;| \;\;\;
        \env_{i+1}({...}) = t_{v}\\
-       wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, \Integer{:}t_{v} \cup \Nil\}_{unique}) = t_{t}
+       t = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, \Integer{:}t_{v} \cup \Nil\}_{unique} \;\;\;
+       wf(t)
        \end{array}}
-      {\env_{1}, \penv \vdash \{\;\vec{p},\;{...}_{m}\;\}:t_{t}, \env_{n+1}}
+      {\env_{1}, \penv \vdash \{\;\vec{p},\;{...}_{m}\;\}:t, \env_{n+1}}
 \end{array}
@@
 the table constructor $\{ [1] = ``foo", [10] = \mathbf{true}, {...}_{m} \}$,
-because the keys $1$ and $10$ overlap the array part.
+because the keys $1$ and $10$ overlap the array part,
+and the resulting table type is not well-formed.
 
@@
 unusual feature of our type system: the refinement of table types.
-The rule \textsc{T-REFINE} allows adding new keys to \emph{open}
-and \emph{unique} table types:
+The first kind of refinement allows programmers to add new
+fields to \emph{unique} or \emph{open} table types through
+field assignment.
+For instance, in Section \ref{sec:tables} we presented the
+following example:
+\begin{verbatim}
+    local person = {}
+    person.firstname = "Lou"
+    person.lastname = "Reed"
+\end{verbatim}
+
+We translate this example to our reduced core as follows:
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; person:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$person[``firstname"] \; {<}\String{>} = ``Lou";$}\\
+& \multicolumn{1}{l}{$person[``lastname"] \; {<}\String{>} = ``Reed"$}
+\end{tabular}
+\end{center}
+
+In this example, we assign the type $\{\}_{unique}$ to the variable
+$person$, then we refine its type to $\{``firstname":\String\}_{unique}$,
+and then we refine its type to $\{``firstname":\String, ``lastname":\String\}_{unique}$.
+Rule \textsc{T-REFINE} type checks this use of refinement:
 \[
@@
        \env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique}\\
-       \env_{1}, \penv \vdash c:k_{n+1}, \env_{2} \;\;\;
-       \not \exists i \in 1..n \; t_{1} \lesssim k_{i} \;\;\;
-       v_{n+1} = close(t)
+       \env_{1}, \penv \vdash c:k, \env_{2} \;\;\;
+       \not \exists i \in 1..n \; k \lesssim k_{i} \;\;\;
+       v = rt(t)
        \end{array}}
-      {\env_{1}, \penv \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, k_{n+1}{:}v_{n+1}\}_{open|unique}]}
+      {\env_{1}, \penv \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, k{:}v\}_{open|unique}]}
 \end{array}
@@
 
-We use the refinement of table types to handle global variables.
+We use the refinement of table types to handle the declaration of
+new global variables.
 In Lua, the assignment \texttt{v = v + 1} translates to
@@
 
-We do not allow the refinement of table types to change the type of
-a field that is already present in the table type.
+We do not allow the refinement of table types to add a field if it is
+already present in the table's type.
 For instance,
@@
 \]
-do not type check, as we are trying to add a field that already exists
-in $\string_ENV$.
-We could have used union types to allow the refinement of existing fields,
-but this could lead to fields that have a too general type, which could
-decrease the amount of static type checking.
-This means that our type system does not allow the table type to change
-towards a type that is not a subtype of the previous type.
-In this example,
-\[
-\{``x":\String \cup \Integer\}_{open} \not\subtype \{``x":\String\}_{open}
-\]
+does not type check, as we are trying to add $``x"$ twice.
 
 We also do not allow the refinement of table types to introduce
-\emph{open} or \emph{unique} table fields.
+fields with table types that are not \emph{regular}.
 For instance,
@@
 \end{center}
-refines the type of $\string_ENV$ from $\{\}_{open}$ to $\{``x":\{\}_{closed}\}_{open}$,
-though we are trying to refine to $\{``x":\{\}_{unique}\}_{open}$.
-As in the type of the table constructor, the refinement of
-table types also does not allow the inclusion of values that
-have \emph{open} and \emph{unique} table types.
-
-We can also use multiple assignments to refine table types:
+refines the type of $\string_ENV$ from $\{\}_{open}$ to $\{``x":\{\}_{regular}\}_{open}$.
+Currently, our type system can only track \emph{unique} and
+\emph{open} table types that are bound to local variables.
+
+We can also use multiple assignment to refine table types:
 \[
@@
 $``foo" \times 1 \times \Nil{*} \lesssim \String \times \Integer \times \Value{*}$.
-
 However, the next example does not type check because it tries to add
@@
 
-Besides allowing the refinement of table types through field assignment,
-Typed Lua also keeps track whether it is safe to change the type of a table.
-For instance,
+Aliasing an \emph{unique} or an \emph{open} table type produces a
+\emph{closed} table type, as defined by rule \textsc{T-IDREAD}:
+\[
+\begin{array}{c}
+\mylabel{T-IDREAD}\\
+\dfrac{\env(n) = t}
+      {\env, \penv \vdash n_{e}:close(t), \env[n \mapsto open(t)]}
+\end{array}
+\]
+
+The following example does not type check, as aliasing $a$ produces
+the type $\{``foo":\Integer\}_{closed}$ that is not a subtype of
+$\{``foo":\Integer \cup \Nil\}_{regular}$, the type of $b$:
 \begin{center}
@@
 \multicolumn{3}{l}{$\mathbf{local} \; a:\{``foo":\Integer\}_{unique} = \{ [``foo"] = 5 \} \; \mathbf{in}$}\\
-& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``foo":\Integer \cup \Nil \}_{closed} = a \; \mathbf{in}$}\\
+& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``foo":\Integer \cup \Nil \}_{regular} = a \; \mathbf{in}$}\\
 & & \multicolumn{1}{l}{$b[``foo"] = \mathbf{nil}$}
@@
 \end{center}
-does not type check, as aliasing $a$ produces the type
-$\{``foo":\Integer\}_{closed}$ that is not a subtype of
-$\{``foo":\Integer \cup \Nil\}_{closed}$, the type of $b$.
-The rule \textsc{T-IDREAD} defines this behavior:
-\[
-\begin{array}{c}
-\mylabel{T-IDREAD}\\
-\dfrac{\env_{1}(n) = t}
-      {\env_{1}, \penv \vdash n_{e}:close(t), \env_{1}[n \mapsto open(t)]}
-\end{array}
-\]
-
-Aliasing \emph{unique} and \emph{open} table types can be unsafe.
-For this reason, the rule \textsc{T-IDREAD} promote the type of an
-alias expression from \emph{unique} or \emph{open} to \emph{closed},
-and it also promotes the aliased variable from
-\emph{unique} to \emph{open}, as \emph{unique} table types only
-denote the type of tables with no alias.
 
@@
 In this example, it is unsafe to add the field $``x"$ to $b$,
-as it changes the value that is stored in the field $``x"$ of local $a$.
+as it changes the value that is stored in the field $``x"$ of $a$.
+
+Aliasing an \emph{unique} table type while keeping the original
+reference \emph{unique} can be unsafe.
+For this reason, the rule \textsc{T-IDREAD} changes the type of
+the original reference from \emph{unique} to \emph{open}.
 
 We also need to close \emph{unique} and \emph{open} tables that
-appear in the left-hand side of assignments.
-For instance,
+appear in the left-hand side of assignments, as defined by
+rule \textsc{T-IDWRITE}:
+\[
+\begin{array}{c}
+\mylabel{T-IDWRITE}\\
+\dfrac{\env(n) = t}
+      {\env, \penv \vdash n_{l}:close(t), \env[n \mapsto close(t)]}
+\end{array}
+\]
+
+As an example,
 \begin{center}
@@
 as its type is \emph{closed}, and thus does not allow changing the
-value that is stored in the field $``x"$ of local $a$.
-The rule \textsc{T-IDWRITE} defines this behavior:
-\[
-\begin{array}{c}
-\mylabel{T-IDWRITE}\\
-\dfrac{\env_{1}(n) = t}
-      {\env_{1}, \penv \vdash n_{l}:close(t), \env_{1}[n \mapsto close(t)]}
-\end{array}
-\]
+value that is stored in the field $``x"$ of $a$.
 
 We also have different rules for type checking table indexing to avoid
-promoting table types in these operations:
+changing table types in these operations:
 \[
@@
 Sometimes we want to use a local variable that holds an \emph{unique}
-table type to initialize a \emph{closed} table type, as \emph{unique}
+table type to initialize a \emph{regular} table type, as \emph{unique}
 table types allow us to initialize optional fields.
-To do that, Typed Lua includes a cast expression that allows us
+To do that, Typed Lua includes a coercion expression that allows us
 to use an \emph{unique} table type before it becomes \emph{open}.
-The rule \textsc{T-CAST} describes this behavior:
-\[
-\begin{array}{c}
-\mylabel{T-CAST}\\
+The rule \textsc{T-COERCION} describes this behavior:
+\[
+\begin{array}{c}
+\mylabel{T-COERCION}\\
 \dfrac{\env_{1}(n) \subtype t \;\;\;
@@
 
-The rule \textsc{T-CAST} allows our type system to type check the following example:
+The rule \textsc{T-COERCION} allows our type system to type check the following example:
 \begin{center}
@@
 & \multicolumn{2}{l}{$a[``y"] \; {<}\String{>} = ``bar";$}\\
-& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{closed} =$}\\
+& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{regular} =$}\\
 & & \multicolumn{1}{l}{${<}\{``x":\String, ``y":\String \cup \Nil\}_{open}{>} \; a \; \mathbf{in} \; a[``z"] \; {<}\Integer{>} = 1$}
@@
 
-We can use $a$ to initialize $b$ because the cast expression converts
+We can use $a$ to initialize $b$ because the coercion expression converts
 the type of $a$ from $\{``x":\String, ``y":\String\}_{unique}$ to
@@
 \mylabel{T-FUNCTION3}\\
-\dfrac{closeall(\env_{1}[\vec{n} \mapsto \vec{t}, \ret \mapsto r]), \penv \vdash s, \env_{2}}
+\dfrac{closeall(\env_{1}[\vec{n} \mapsto \vec{t}]), \penv[\ret \mapsto r] \vdash s, \env_{2}}
       {\env_{1}, \penv \vdash \mathbf{fun} \; (\vec{n{:}t}){:}r \; s:\vec{t} \rightarrow r, closeset(\env_{1}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))}
@@
 and \emph{closeset} to close \emph{open} and \emph{unique} table types.
-The predicate \emph{closeall} closes all \emph{open} and \emph{unique}
+The function \emph{closeall} closes all \emph{open} and \emph{unique}
 table types before type checking another scope.
-The predicate \emph{closeset} closes a given set of free assigned variables,
-which is given by the predicate \emph{fav}.
+The function \emph{closeset} closes a given set of free assigned variables,
+which is given by the function \emph{fav}.
 This rule prevents the following unsafe example to type check:
@@
 
+\subsection{Objects}
+
 Typed Lua also uses the refinement of table types to type check
@@
        \not \exists i \in 1..n \; l \lesssim k_{i}\\
-       closeall(\env_{1}[self \mapsto \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{closed}, \vec{n} \mapsto \vec{t}, \ret \mapsto r]), \penv \vdash s, \env_{2}\\
+       closeall(\env_{1}[self \mapsto \Self, \vec{n} \mapsto \vec{t}]), \penv[\ret \mapsto r] \vdash s, \env_{2}\\
        t_{m} = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}, l{:}\Const \; \Self \times \vec{t} \rightarrow r\}_{unique|open}
@@
 \begin{tabular}{llll}
-\multicolumn{4}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{open} = \{ [``x"] = 0.0, [``y"] = 0.0 \}$}\\
+\multicolumn{4}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{unique} = \{ [``x"] = 0.0, [``y"] = 0.0 \}$}\\
 \multicolumn{4}{l}{$\mathbf{in}$}\\
 & \multicolumn{3}{l}{$\mathbf{fun} \; s{:}new (x:\Number, y:\Number):\Self \times \Void$}\\
-& & \multicolumn{2}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{closed} =$}\\
-& & & \multicolumn{1}{l}{$setmetatable(\{\}, \{ [``\string_\string_index"] = self \})$}\\
-& & \multicolumn{2}{l}{$\mathbf{in} \; s[``x"] = x; \; s[``y"] = y; \; \mathbf{return} \; s$}\\
+& & \multicolumn{2}{l}{$\mathbf{local} \; o:\Self = setmetatable(\{\}, \{ [``\string_\string_index"] = self \})$}\\
+& & \multicolumn{2}{l}{$\mathbf{in} \; o[``x"] = x; \; o[``y"] = y; \; \mathbf{return} \; o$}\\
 ; & \multicolumn{3}{l}{$\mathbf{fun} \; s{:}move (x:\Number, y:\Number):\Void$}\\
 & & \multicolumn{2}{l}{$self[``x"] = self[``x"] + x;$}\\
-& & \multicolumn{2}{l}{$self[``y"] = self[``y"] + y$}
+& & \multicolumn{2}{l}{$self[``y"] = self[``y"] + y;$}\\
+& & \multicolumn{2}{l}{$\mathbf{return} \; \mathbf{nothing}$}
 \end{tabular}
@@
 that we are defining.
-Then, we use two method definitions to include $new$ and $move$ into
-our class that now has type
+Then, it uses two method definitions to include $new$ and $move$ into our class.
+At the end of the example, $s$ has type
 \begin{align*}
 \{ & ``x":\Number, ``y":\Number,\\
-   & \Const \; ``new":\Self \times \Number \times \Number \rightarrow \Self,\\
-   & \Const \; ``move":\Self \times \Number \times \Number \rightarrow \Void \}_{open}
+   & \Const \; ``new":\Self \times \Number \times \Number \times \Void \rightarrow \Self \times \Void,\\
+   & \Const \; ``move":\Self \times \Number \times \Number \times \Void \rightarrow \Void \}_{unique}
 \end{align*}
@@
 Inside the definition of the method \emph{new} we use \emph{setmetatable}
-to initialize the local $s$ with the type of $\Self$.
+to initialize the $s$ with type $\Self$.
 The rule \textsc{T-SETMETATABLE1} express this idea:
@@
 \mylabel{T-SETMETATABLE1}\\
-\dfrac{\env_{1}, \penv \vdash e:t, \env_{2} \;\;\;
-       t \subtype \{\}_{open}}
-      {\env_{1}, \penv \vdash setmetatable(\{\}, \{[``\string_\string_index"] = e\}):close(t), \env_{2}}
+\dfrac{\env(n) = \Self}
+      {\env, \penv \vdash setmetatable(\{\}, \{[``\string_\string_index"] = n\}):\Self, \env}
 \end{array}
@@
 \]
+
+\subsection{Filters and projections}
 
@@
 
+\subsection{Arithmetic operators}
+
 Lua has operator overloading, and allows the programmers to redefine

### 2015-02-11 doc/thesis/system.tex
Improving the text about typing rules

@@
 Filter types are subtypes only of themselves and of $\Value$.
-More precisey, a filter type $\phi(t_{1},t_{2})$ is a subtype of
+More precisely, a filter type $\phi(t_{1},t_{2})$ is a subtype of
 the same filter type $\phi(t_{1},t_{2})$, which shares the same
@@
 We also include two kinds of type coercions in our core language:
-the expression ${<}t{>} \;n$ and the left-hand value $n[c] \; {<}t{>}$.
+the left-hand value $n[c] \; {<}t{>}$ and the expression ${<}t{>} \;n$.
 Both allow the refinement of table types.
@@
 \subsection{Assignment and function application}
+\label{sec:assignment}
 
@@
 \subsection{Tables and refinement}
-
-The simples expression involving tables in the empty table constructor.
+\label{sec:refinement}
+
+The simplest expression involving tables in the empty table constructor.
 Its type checking rule is straightforward:
@@
 \mylabel{T-IDREAD}\\
-\dfrac{\env(n) = t}
-      {\env, \penv \vdash n_{e}:close(t), \env[n \mapsto open(t)]}
-\end{array}
-\]
+\dfrac{\env(n) = t_{1} \;\;\; t_{2} = read(\penv, t_{1})}
+      {\env, \penv \vdash n_{e}:close(t_{2}), \env[n \mapsto open(t_{1})]}
+\end{array}
+\]
+
+In Section \ref{sec:fap} we will explain that \textsc{T-IDREAD} uses
+the auxiliary function \emph{read} because it may be accessing an
+identifier that is bound to a filter or projection type.
+As we mentioned in Section \ref{sec:unions}, our type system includes
+a small set of \texttt{type} predicates that allow programmers to
+discriminate union types.
+In Section \ref{sec:fap} we will also explain that our type system
+uses filter and projection types in the definition of these predicates
+to handle the discrimination of unions of first-level and second-level
+types.
 
@@
 \mylabel{T-IDWRITE}\\
-\dfrac{\env(n) = t}
-      {\env, \penv \vdash n_{l}:close(t), \env[n \mapsto close(t)]}
-\end{array}
-\]
+\dfrac{\env(n) = t_{1} \;\;\; t_{2} = write(t_{1})}
+      {\env, \penv \vdash n_{l}:close(t_{2}), \env[n \mapsto close(t_{2})]}
+\end{array}
+\]
+
+In Section \ref{sec:fap} we will explain that \textsc{T-IDWRITE} uses
+the auxiliary function \emph{write} because it may be accessing an
+identifier that is bound to a filter or projection type.
+As we mentioned in Section \ref{sec:unions}, assignments restore
+discriminated union types to their original types, and
+function \emph{write} also works in this purpose.
 
@@
 We also have different rules for type checking table indexing to avoid
-changing table types in these operations:
+changing table types in these operations, as they cannot create aliases:
 \[
@@
 \mylabel{T-INDEX1}\\
-\dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\} \;\;\;
-       \env_{1}, \penv \vdash e_{2}:t, \env_{2} \;\;\;
-       \exists i \in 1{..}n \; t \lesssim k_{i}}
+\dfrac{\begin{array}{c}
+       \env_{1}(n) = t \;\;\;
+       read(\penv, t) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}\\
+       \env_{1}, \penv \vdash e_{2}:k, \env_{2} \;\;\;
+       \exists i \in 1{..}n \; k \lesssim k_{i}
+       \end{array}}
       {\env_{1}, \penv \vdash n[e_{2}]:v_{i}, \env_{2}}
@@
 \mylabel{T-INDEX2}\\
-\dfrac{\env_{1}, \penv \vdash e_{1}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash e_{2}:t, \env_{3} \;\;\;
-       \exists i \in 1{..}n \; t \lesssim k_{i}}
+\dfrac{\begin{array}{c}
+       \env_{1}, \penv \vdash e_{1}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}, \env_{2}\\
+       \env_{2}, \penv \vdash e_{2}:k, \env_{3} \;\;\;
+       \exists i \in 1{..}n \; k \lesssim k_{i}
+       \end{array}}
       {\env_{1}, \penv \vdash e_{1}[e_{2}]:v_{i}, \env_{3}}
@@
 
-Sometimes we want to use a local variable that holds an \emph{unique}
-table type to initialize a \emph{regular} table type, as \emph{unique}
-table types allow us to initialize optional fields.
-To do that, Typed Lua includes a coercion expression that allows us
-to use an \emph{unique} table type before it becomes \emph{open}.
-The rule \textsc{T-COERCION} describes this behavior:
-\[
-\begin{array}{c}
-\mylabel{T-COERCION}\\
+A second form of refinement happens when we want to use an
+\emph{unique} or \emph{open} table type in a context that expects a
+\emph{closed} or \emph{regular} table type with a different shape.
+This kind of refinement allows programmers to add optional fields
+or merge existing fields.
+To do that, Typed Lua includes a type coercion expression that
+obeys the \textsc{T-COERCE} rule below:
+\[
+\begin{array}{c}
+\mylabel{T-COERCE}\\
 \dfrac{\env_{1}(n) \subtype t \;\;\;
@@
 
-The rule \textsc{T-COERCION} allows our type system to type check the following example:
+The rule \textsc{T-COERCE} allows our type system to type check the following example:
 \begin{center}
@@
 
-We can use $a$ to initialize $b$ because the coercion expression converts
+We can use $a$ to initialize $b$ because the coercion converts
 the type of $a$ from $\{``x":\String, ``y":\String\}_{unique}$ to
 $\{``x":\String, ``y":\String \cup \Nil\}_{open}$, and results in
-$\{``x":\String, ``y":\String \cup \Nil\}_{closed}$.
-We can continue to refine the type of $a$ after the aliasing,
+$\{``x":\String, ``y":\String \cup \Nil\}_{closed}$,
+which is a subtype of
+$\{``x":\String, ``y":\String \cup \Nil\}_{regular}$, the type of $b$.
+We can continue to refine the type of $a$ after aliasing it to $b$,
 as it still holds an \emph{open} table.
@@
 We also need to make sure to close all \emph{unique} and \emph{open}
-table types before we type check another scope.
+table types before we type check a nested scope.
 The rule \textsc{T-FUNCTION3} illustrates this case:
@@
 \dfrac{closeall(\env_{1}[\vec{n} \mapsto \vec{t}]), \penv[\ret \mapsto r] \vdash s, \env_{2}}
-      {\env_{1}, \penv \vdash \mathbf{fun} \; (\vec{n{:}t}){:}r \; s:\vec{t} \rightarrow r, closeset(\env_{1}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))}
-\end{array}
-\]
-
-The rule \textsc{T-FUNCTION3} uses the predicates \emph{closeall}
-and \emph{closeset} to close \emph{open} and \emph{unique} table types.
-The function \emph{closeall} closes all \emph{open} and \emph{unique}
-table types before type checking another scope.
+      {\begin{array}{c}
+       \env_{1}, \penv \vdash \mathbf{fun} \; (\vec{n{:}t}){:}r \; s:\vec{t} \times \Void \rightarrow r,\\
+       openset(closeset(\env_{1}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s)),rv(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))
+       \end{array}}
+\end{array}
+\]
+
+Rule \textsc{T-FUNCTION3} uses some auxiliary functions to change
+the type of variables before type checking a nested scope and
+also to change the type of assigned and referenced variables after
+type checking a nested scope.
+The function \emph{closeall} closes all \emph{unique} and \emph{open}
+table types.
 The function \emph{closeset} closes a given set of free assigned variables,
 which is given by the function \emph{fav}.
-This rule prevents the following unsafe example to type check:
+The function \emph{openset} changes from \emph{unique} to \emph{open}
+a given set of referenced variables, which is given by the function \emph{rv}.
+
+This rule also extends the environment $\penv$, bounding the special
+variable $\ret$ to the return type $r$.
+Rule \textsc{T-RETURN} uses the type that is bound to $\ret$ in
+$\penv$ to type check return statements:
+\[
+\begin{array}{c}
+\mylabel{T-RETURN}\\
+\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
+       \penv(\ret) = r_{2} \;\;\;
+       r_{1} \lesssim r_{2}}
+      {\env_{1} \vdash \mathbf{return} \; el, \env_{2}}
+\end{array}
+\]
+
+As an example, rule \textsc{T-FUNCTION3} prevents the following unsafe example to type check:
 \begin{center}
 \begin{tabular}{llll}
-\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
-& \multicolumn{3}{l}{$\mathbf{local} \; b:\Integer \times \Integer \times \Void \rightarrow \Integer \times \Void =$}\\
-& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer, y:\Integer):\Integer$}\\
-& & & \multicolumn{1}{l}{$a[``z"] \; {<}\Integer{>} = 1; \; \mathbf{return} \; x + y$}\\
-& \multicolumn{3}{l}{$\mathbf{in} \; a[``z"] \; {<}\String{>} = ``foo"$}
+\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique}, b:\{\}_{unique} = \{\}, \{\} \; \mathbf{in}$}\\
+& \multicolumn{3}{l}{$\mathbf{local} \; f:\Integer \times \Void \rightarrow \Integer \times \Void =$}\\
+& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer):\Integer \times \Void$}\\
+& & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
+& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; b[``x"] \; {<}\String{>} = ``foo" \;;\; f(a[``x"])_{s}$}
 \end{tabular}
@@
 
-\subsection{Objects}
-
-Typed Lua also uses the refinement of table types to type check
-common idioms that Lua programmers use for building modules and objects.
-For instance, the rule \textsc{T-METHOD3} illustrates how our type system
+This example does not type check because we cannot add the field
+$``x"$ to $b$, as its type is closed.
+The assignment $b = a$ type checks because, at this point,
+$a$ and $b$ have the same type: $\{\}_{closed}$.
+Their type was closed by \emph{closeall} before type checking
+the function body.
+Their type would be restored to $\{\}_{unique}$ after type checking
+the function body, but that assignment also triggers other two type changes.
+First, the function \emph{fav} includes $b$ in the set of variables
+that should be closed by \emph{closeset}.
+Then, the function \emph{rv} includes $a$ in the set of variables
+that should change from \emph{unique} to \emph{open} by \emph{openset}.
+After declaring $f$, $a$ has type $\{\}_{open}$ and $b$ has type $\{\}_{closed}$,
+so we can refine the type of $a$, but we cannot refine the type of $b$.
+
+\subsection{Classes and objects}
+\label{sec:cao}
+
+Typed Lua also relies on the refinement of table types to type check
+the idioms that Lua programmers use for incrementally building
+classes and objects.
+For instance, rule \textsc{T-METHOD3} illustrates how our type system
 type checks method definitions:
@@
 \dfrac{\begin{array}{c}
-       \env_{1}(n_{1}) = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{unique|open} \;\;\;
-       \env_{1}, \penv \vdash n_{2} : l \;\;\;
+       \env_{1}(n_{1}) = t_{s} \;\;\; t_{s} = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{unique|open}\\
+       \env_{1}, \penv \vdash n_{2} : l, \env_{2} \;\;\;
        \not \exists i \in 1..n \; l \lesssim k_{i}\\
-       closeall(\env_{1}[self \mapsto \Self, \vec{n} \mapsto \vec{t}]), \penv[\ret \mapsto r] \vdash s, \env_{2}\\
-       t_{m} = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}, l{:}\Const \; \Self \times \vec{t} \rightarrow r\}_{unique|open}
+       closeall(\env_{1}[self \mapsto \Self, \vec{n} \mapsto \vec{t}, \self \mapsto t_{s}]),
+       \penv[\ret \mapsto r] \vdash s, \env_{3}\\
+       t_{o} = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}, l{:}\Const \; \Self \times \vec{t} \times \Void \rightarrow r\}_{unique|open}
        \end{array}}
@@
        \env_{1}, \penv \vdash \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; s,\\
-       closeset(\env_{1}[n_{1} \mapsto t_{m}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))
+       openset(closeset(\env_{1}[n_{1} \mapsto t_{o}], fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s)),rv(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))
        \end{array}}
@@
 
-The rule \textsc{T-METHOD3} allows our type system to type check the
+Rule \textsc{T-METHOD3} allows our type system to type check the
 following example, which is a translation of the definition of the class
@@
 \begin{tabular}{llll}
-\multicolumn{4}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{unique} = \{ [``x"] = 0.0, [``y"] = 0.0 \}$}\\
+\multicolumn{4}{l}{$\mathbf{local} \; Shape{:}\{``x"{:}\Number, ``y"{:}\Number\}_{unique} = \{ [``x"] = 0.0, [``y"] = 0.0 \}$}\\
 \multicolumn{4}{l}{$\mathbf{in}$}\\
-& \multicolumn{3}{l}{$\mathbf{fun} \; s{:}new (x:\Number, y:\Number):\Self \times \Void$}\\
+& \multicolumn{3}{l}{$\mathbf{fun} \; Shape{:}new (x:\Number, y:\Number):\Self \times \Void$}\\
 & & \multicolumn{2}{l}{$\mathbf{local} \; o:\Self = setmetatable(\{\}, \{ [``\string_\string_index"] = self \})$}\\
 & & \multicolumn{2}{l}{$\mathbf{in} \; o[``x"] = x; \; o[``y"] = y; \; \mathbf{return} \; o$}\\
-; & \multicolumn{3}{l}{$\mathbf{fun} \; s{:}move (x:\Number, y:\Number):\Void$}\\
+; & \multicolumn{3}{l}{$\mathbf{fun} \; Shape{:}move (x:\Number, y:\Number):\Void$}\\
 & & \multicolumn{2}{l}{$self[``x"] = self[``x"] + x;$}\\
@@
 
-This example uses the type $\{``x":\Number, ``y":\Number\}_{open}$ to
-initialize the local variable $s$, which represents the type of the class
+This example uses the type $\{``x":\Number, ``y":\Number\}_{unique}$ to
+initialize the local variable $Shape$, which represents the class
 that we are defining.
-Then, it uses two method definitions to include $new$ and $move$ into our class.
-At the end of the example, $s$ has type
+Then, it uses two method definitions to include $new$ and $move$ in our class.
+These definitions refine the type of $Shape$, and it becomes the type below:
 \begin{align*}
@@
 
-Inside the definition of the method \emph{new} we use \emph{setmetatable}
-to initialize the $s$ with type $\Self$.
+Inside the definition of the methods \emph{new} and \emph{move},
+we index variables that have type $\Self$.
+This is possible because we include the rule \textsc{T-SELF}:
+\[
+\begin{array}{c}
+\mylabel{T-SELF}\\
+\dfrac{\env_{1}, \penv \vdash e:\Self, \env_{2} \;\;\;
+       \env_{2}(\self) = t}
+      {\env_{1}, \penv \vdash e:t, \env_{2}}
+\end{array}
+\]
+
+When an expression has type $\Self$, our type system uses this
+rule to produce the type that is bound to $\self$ in $\env$.
+The variable $\self$ is a special variable that method definitions
+and method calls use to bind the type of $\Self$ in the environment.
+
+Inside the definition of the method \emph{new}, we use \emph{setmetatable}
+to initialize $o$ with type $\Self$.
 The rule \textsc{T-SETMETATABLE1} express this idea:
@@
 
-After we define our class, we can use it to create object instances
-of this class and call its methods.
-The rule \textsc{T-INVOKE1} handles the case where method calls
-are expressions that return multiple values:
+In Section \ref{sec:oop} we also mentioned that our type system
+uses \emph{setmetatable} to handle single inheritance.
+Rule \textsc{T-SETMETATABLE2} allows refining the type of a new
+class based on the type of an already defined class,
+while rule \textsc{T-SETMETATABLE3} allows overriding a constructor:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-SETMETATABLE2}\\
+\dfrac{\env(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed}}
+      {\env, \penv \vdash setmetatable(\{\}, \{[``\string_\string_index"] = n\}):\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{open}, \env}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-SETMETATABLE3}\\
+\dfrac{\env_{1}, \penv \vdash e = t, \env_{2} \;\;\;
+       t = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed} \;\;\;
+       \env_{1}(n) = \Self \;\;\; \env_{1}(\sigma) \subtype t}
+      {\env_{1}, \penv \vdash setmetatable(e, \{[``\string_\string_index"] = n\}):\Self, \env_{2}[\sigma \mapsto t]}
+\end{array}
+\end{array}
+\]
+
+After we define our class, we can use it to create instances
+of this class and call their methods.
+Rule \textsc{T-INVOKE1} handles the case where method calls
+are expressions that produce multiple values:
 \[
@@
 \mylabel{T-INVOKE1}\\
-\dfrac{\env_{1}, \penv \vdash e:t, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash e[n]:p_{1} \rightarrow r, \env_{3} \;\;\;
-       \env_{3}, \penv \vdash el:p_{2}, \env_{4} \;\;\;
-       \Self \times p_{2} \lesssim p_{1}}
-      {\env_{1}, \penv \vdash e{:}n(el)_{m}:[\Self \mapsto t]r, \env_{4}}
+\dfrac{\begin{array}{c}
+       \env_{1}, \penv \vdash e:t_{s}, \env_{2}\\
+       \env_{2}[\sigma \mapsto t_{s}], \penv \vdash e[n]:\Const \; r_{1} \rightarrow r_{2}, \env_{3}\\
+       \env_{3}[\sigma \mapsto t_{s}], \penv \vdash el:r_{3}, \env_{4} \;\;\;
+       t_{s} \times r_{3} \lesssim r_{1}
+       \end{array}}
+      {\env_{1}, \penv \vdash e{:}n(el)_{m}:[\sigma \mapsto t]r_{2}, \env_{4}}
 \end{array}
@@
 that handle the cases where method calls are either statements
-or expressions that return only one value:
+or expressions that produce only one value:
 \[
@@
 
-The rule \textsc{T-STMINVOKE1} discards the returned values,
+The rule \textsc{T-STMINVOKE1} discards the produced values,
 while the rule \textsc{T-EXPINVOKE1} uses the predicate \emph{first} to
-ensure that only one value is returned.
+ensure that only one value is produced.
 
 As an example, assuming that the object $o$ is in the environment and
-has the type of the class from the previous example, the method call
-\[
-o{:}move(10, 10)_{s}
-\]
-type checks through the rules \textsc{T-STMINVOKE1} and \textsc{T-INVOKE1},
-because
-\[
-\Self \times 10 \times 10 \times \Nil{*} \lesssim \Self \times \Number \times \Number \times \Void
-\]
+has the type of the class $Shape$, the method call
+\[
+o{:}move(10, 10)_{m}
+\]
+type checks through the rule \textsc{T-INVOKE1}, because
+\[
+Shape \times 10 \times 10 \times \Nil{*} \lesssim Shape \times \Number \times \Number \times \Void
+\]
+assuming that $Shape$ is an alias to the type of the class $Shape$.
+Note that rule \textsc{T-INVOKE1} uses the object type to bind
+the type $\Self$ before type checking a method call.
 
 \subsection{Filters and projections}
+\label{sec:fap}
+
+In Section \ref{sec:unions} we mentioned that our type system 
+includes union types to encode three common Lua idioms:
+the use of optional values, the overloading based on the tags of
+input parameters, and the overloading on the return type of functions.
+Now, we will discuss how our type system uses the $\mathbf{or}$
+operator to handle optional values, how it uses filter types to
+discriminate unions of first-level types, and how it uses
+projection types to discriminate overloaded return types,
+which are unions of second-level types.
 
@@
 
-The rule \textsc{T-OR1} is the rule that defines the short circuit.
+Rule \textsc{T-OR1} is the rule that defines the short circuit.
 We use the consistent-subtyping relation in this rule to guarantee that
@@
 
-The rules \textsc{T-OR2}, \textsc{T-OR3}, and \textsc{T-OR4} guarantee
+Rules \textsc{T-OR2}, \textsc{T-OR3}, and \textsc{T-OR4} guarantee
 that the final result is the type of the second expression, because the
@@
 
-The rule \textsc{T-OR5} is the most general rule, but it is also
+Rule \textsc{T-OR5} is the most general rule, but it is also
 the rule that handles the common $\mathbf{or}$ idioms.
-It uses the predicate \emph{filter} to filter possible false values
-that might be part of the type of the first expression.
-We can use pattern matching to the define the recursive predicate
+It uses the auxiliary function \emph{filter} to filter possible false
+values that might be part of the type of the first expression.
+We can use pattern matching to the define the recursive function
 \emph{filter} as follows:
@@
 
-The rule \textsc{T-OR5} allows our type system to type check the following example:
+Rule \textsc{T-OR5} allows our type system to type check the following example:
 \begin{center}
@@
 
-Without the \emph{filter} predicate,
+Without the \emph{filter} function,
 the expression $x \; \mathbf{or} \; ``Hello"$ would have type
 $\String \cup \Nil \cup ``Hello"$.
-The \emph{filter} predicate removes the type $\Nil$ from the result,
+The \emph{filter} function removes the type $\Nil$ from the result,
 leaving the type $\String \cup ``Hello"$.
-At the end of the type checking, the expression $x \; \mathbf{or} \; ``Hello"$
-has type $\String$ because union types are disjoint and $``Hello" \subtype \String$.
+This means that type checking the expression $x \; \mathbf{or} \; ``Hello"$
+results in the type $\String$, because union types are disjoint and
+$``Hello" \lesssim \String$.
 
 Another common idiom that programmers use in Lua is to overload
-the input parameter of functions, and use the function \texttt{type}
+the type of function parameters, and use the function \texttt{type}
 to execute different actions according to their types.
 The rule \textsc{T-IF2} shows the case where our type system
-discriminates the type of a local variable base on the tag \texttt{string}:
+discriminates an union of first-level types based on the tag \texttt{string}:
 \[
@@
 \dfrac{\begin{array}{c}
-       \env_{1}(n) = t\\
+       \env_{1}(n) = t_{1} \cup t_{2}\\
        closeall(\env_{1}[n \mapsto \phi(t,\String)]), \penv \vdash s_{1}, \env_{2} \\
        closeall(\env_{1}[n \mapsto \phi(t,filter(t, \String))), \penv \vdash s_{2}, \env_{3}\\
-       \env_{4} = closeset(\env_{1}[n \mapsto t], fav(s_{1}) \cup fav(s_{2}))
+       \env_{4} = openset(closeset(\env_{1}[n \mapsto t], fav(s_{1}) \cup fav(s_{2})),rv(s_{1}) \cup rv(s_{2}))
       \end{array}}
@@
 In Typed Lua, \emph{type} is a primitive that triggers a type change.
-In the rule \textsc{T-IF2}, the type of the variable $n$ changes from
-$t$ to $\phi(t,\String)$ inside the $\mathbf{if}$ branch, and it changes
-the type of the variable $n$ from $t$ to $\phi(t,filter(t,\String))$
+In the rule \textsc{T-IF2}, the type of a variable $n$ changes from
+$t_{1} \cup t_{2}$ to $\phi(t_{1} \cup t_{2},\String)$ inside the
+$\mathbf{if}$ branch, and it changes the type of a variable $n$ from
+$t_{1} \cup t_{2}$ to $\phi(t,filter(t_{1} \cup t_{2},\String))$
 inside the $\mathbf{else}$ branch.
-Our type system changes the type of a local variable to a filter type
-because it gives the discriminated type when we use the variable as
-an expression, and it restores the original type when we use the variable
+
+Rule \textsc{T-IF2} changes the type of a local variable $n$ to a
+filter type because it gives the discriminated type when we use $n$ as
+an expression, and it restores the original type when we use $n$
 in an assignment.
-We will use the following example to discuss how our type system uses
-the rule \textsc{T-IF2} to type check this given example:
+In Section \ref{sec:refinement} we mentioned that rule \textsc{T-IDREAD}
+uses the function \emph{read} to produce the discriminated type,
+while rule \textsc{T-IDWRITE} uses the function \emph{write} to
+restore the original type.
+Now, we will give the definition of these functions.
+
+We can define \emph{read} as follows:
+\begin{align*}
+read(\penv, \phi(t_{1},t_{2})) & = t_{2}\\
+read(\penv, \pi_{i}^{x}) & = proj(\penv, x, i)\\
+read(\penv, t) & = t
+\end{align*}
+
+The function \emph{read} uses the auxiliary function \emph{proj}
+to project a union of first-level types, based on an union of
+second-level types and an index from a projection type.
+In this section, we will discuss how our type system uses
+projection types to handle overloaded return types.
+
+We can define \emph{write} as follows:
+\begin{align*}
+write(\phi(t_{1},t_{2})) & = t_{1}\\
+write(t) & = t
+\end{align*}
+
+As an example, assuming that \emph{rep} is in the environment and has type
+$\String \times \Integer \times \String \cup \Nil \times \Void \rightarrow \String \times \Void$,
+the following code type checks through rule \textsc{T-IF2}:
 \begin{center}
@@
 
-In this example, we are assuming that the function \emph{rep} is in
-the environment and has type
-$\String \times \Integer \times \String \cup \Nil \times \Void \rightarrow \String \times \Void$.
-Note that rule \textsc{T-IF2} changes the type of $b$ to
-$\phi(\String \cup \Integer, \String)$ inside the $\mathbf{if}$ branch,
-and it changes the type of $b$ to
-$\phi(\String \cup \Integer, \Integer)$ inside the $\mathbf{else}$ branch.
-This example type checks because reading $b$ results in the type
-$\String$ inside the $\mathbf{if}$ branch, and it results in the type
-$\Integer$ inside the $\mathbf{else}$ branch.
-Outside the condition, the local variable $b$ has type $\String \cup \Integer$.
+In this example, the type of $b$ changes to
+$\phi(\String \cup \Integer, \String)$ inside the $\mathbf{if}$ branch, and
+it changes to $\phi(\String \cup \Integer, \Integer)$ inside the $\mathbf{else}$ branch.
+Thus, reading $b$ results in the type $\String$ inside the $\mathbf{if}$ branch,
+and it results in the type $\Integer$ inside the $\mathbf{else}$ branch.
+Outside the condition, $b$ has its original type: $\String \cup \Integer$.
 
@@
 it is limited to Lua 5.3, as it depends on the function \texttt{math.type}.
-This function appears only in Lua 5.3 and it returns the string $``integer"$
+This function appears only in Lua 5.3 and it returns the tag \texttt{integer}
 when its input parameter is a number that has an integer representation,
-the string $``float"$ when its input parameter is a number that has a
-floating point representation, or $\Nil$ otherwise.
+the tag \texttt{float} when its input parameter is a number that has a
+floating point representation, or \texttt{nil} otherwise.
 
 Lua programmers also overload the return type of functions to denote errors,
-and our type system also handles this idiom.
-For instance,
+and our type system uses projection types to handle this idiom.
+
+As an example, assuming that \emph{idiv} and \emph{print} are in the
+environment with the respective types
+$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void$
+and
+$\Value{*} \rightarrow \Void$,
+the following code type checks in our type system:
 \begin{center}
@@
 \end{center}
-type checks, assuming that functions \emph{idiv} and \emph{print} are in
-the environment with the respective types
-$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void$
-and
-$\Value{*} \rightarrow \Void$.
+
 There are three typing rules involved in the type checking of this example,
@@
 
-First, type checking $idiv(1, 2)_{m}^{x}$ results in a tuple of projection
-types and in a pair that the typing rule of the local declaration needs
-for introducing the variable $x$ in the projection environment.
-The rule \textsc{T-EXPLIST6} express this intuition:
+First, that example uses rule \textsc{T-EXPLIST6} for type checking
+$idiv(1, 2)_{m}^{x}$:
 \[
@@
 \dfrac{\begin{array}{c}
-       \env, \penv \vdash me^{x}:r\\
-       r = t_{1} \times ... \times t_{n} \times \Void \sqcup t_{1}' \times ... \times t_{n}' \times \Void, \env_{1}
+       \env_{1}, \penv \vdash me^{x}:r, \env_{2}\\
+       r = t_{1} \times ... \times t_{n} \times \Void \sqcup t_{1}' \times ... \times t_{n}' \times \Void
        \end{array}}
-      {\env, \penv \vdash me^{x}:\pi_{1}^{x} \times ... \times \pi_{n}^{x} \times \Nil{*}, \env_{1}, (x,r)}
-\end{array}
-\]
-
-Then, type checking the local declaration checks whether the tuple of projection
-types match the types in the annotations.
-If the types match, the local declaration extends the type environment with the
-new variables, and it also uses the pair $(x,r)$ from the previous rule to
-extend the projection environment before type checking the statements.
-The rule \textsc{T-LOCAL} express this intuition:
+      {\env_{1}, \penv \vdash me^{x}:\pi_{1}^{x} \times ... \times \pi_{n}^{x} \times \Nil{*}, \env_{2}, (x,r)}
+\end{array}
+\]
+
+Rule \textsc{T-EXPLIST6} uses the auxiliary relation
+$\env_{1}, \penv \vdash el : r_{1}, \env_{2}, (x,r_{2})$.
+This relation means that given a type environment $\env_{1}$ and
+a projection environment $\penv$, an expression list $el$ has type
+$r_{1}$ and produces a new type environment $\env_{2}$ and
+produces a pair $(x,r_{2})$.
+This pair means that the last expression of an expression list $el$
+produces a second-level type $r_{2}$ that should be bound to
+a variable $x$ in the projection environment,
+as the resulting type of this expression is a tuple of projection
+types $\pi_{i}^{x}$.
+
+Our type system uses this pair to allow rule \textsc{T-LOCAL} to bound
+a second-level type $r_{2}$ to a variable $x$ in $\penv$,
+after rule \textsc{T-LOCAL} checks whether the type of an expression
+list $el$ matches the types in the annotations:
 \[
@@
 \dfrac{\env_{1}, \penv \vdash el:r_{1}, \env_{2}, (x,r_{2}) \;\;\;
-       r_{1} \lesssim \vec{t} \;\;\;
+       r_{1} \lesssim \vec{t} \times \Value{*} \;\;\;
        \env_{2}[\vec{n} \mapsto \vec{t}], \penv[x \mapsto r_{2}] \vdash s, \env_{3}}
@@
 
-After type checking the statement $s$, the rule \textsc{T-LOCAL} produces a
+After type checking the statement $s$, rule \textsc{T-LOCAL} produces a
 new environment without the variables that it introduced before type checking $s$.
 
-Once the projection is in the environment, our type system can allow
-discriminating variables that hold projection types.
-The rule \textsc{T-IF3} illustrates the case of our previous example:
+Introducing a variable $x$ in the projection environment allows our
+type system to discriminate projection types $\pi_{i}^{x}$.
+The rule \textsc{T-IF3} shows the case where our type system
+discriminates a projection type based on the tag \texttt{nil}:
 \[
@@
        closeall(\env_{1}), \penv[x \mapsto fpt(r,\Nil,i)] \vdash s_{1}, \env_{2} \\
-       closeall(\env_{1}), \penv[x \mapsto gpt(r,\Nil,i)] \vdash s_{2}, \env_{3}
+       closeall(\env_{1}), \penv[x \mapsto gpt(r,\Nil,i)] \vdash s_{2}, \env_{3} \\
+       \env_{4} = openset(closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2})),rv(s_{1}) \cup rv(s_{2}))
       \end{array}}
-      {\env_{1}, \penv \vdash \mathbf{if} \; n \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2}))}
-\end{array}
-\]
-
-The rule \textsc{T-IF3} uses the predicates \emph{fpt} and \emph{gpt}
-to filter projections, affecting all variables that bind to the same projection.
-For instance, our previous example type checks through the rule \textsc{T-IF3},
-because it makes the rule \textsc{T-IF3} use the predicate
+      {\env_{1}, \penv \vdash \mathbf{if} \; n \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
+\end{array}
+\]
+
+Rule \textsc{T-IF3} uses the auxiliary functions \emph{fpt} and \emph{gpt}
+to filter a projection $x$, affecting all variables that bind to the same projection.
+For instance, our previous example type checks through rule \textsc{T-IF3},
+because it makes the rule \textsc{T-IF3} use the function call
 \[
@@
 to discriminate the projection $x$ to the single tuple
-$\Integer \times \Integer \times \Void$, and the predicate
+$\Integer \times \Integer \times \Void$ inside the $\mathbf{if}$ branch,
+and the function call
 \[
@@
 to discriminate the projection $x$ to the single tuple
-$\Nil \times \String \times \Void$.
-This means that the type of $q$ ($\pi_{1}^{x}$) and $r$ ($\pi_{2}^{x}$)
-project to $\Integer$ and $\Integer$ inside the $\mathbf{if}$ branch,
-but they project to $\Nil$ and $\String$ inside the $\mathbf{else}$ branch.
-Outside the $\mathbf{if}$ statement, the types of $q$ and $r$ project
-to $\Integer \cup \Nil$ and $\Integer \cup \String$, respectively.
-
-\subsection{Arithmetic operators}
-
-Lua has operator overloading, and allows the programmers to redefine
+$\Nil \times \String \times \Void$ inside the $\mathbf{else}$ branch.
+Thus, reading $q$ and $r$ projects $\pi_{1}^{x}$ to $\Integer$ and
+$\pi_{2}^{x}$ to $\Integer$ inside the $\mathbf{if}$ branch,
+but it projects $\pi_{1}^{x}$ to $\Nil$ and $\pi_{2}^{x}$ to $\String$
+inside the $\mathbf{else}$ branch.
+Outside the condition, $q$ and $r$ use the original projection, that is,
+they project to $\Integer \cup \Nil$ and $\Integer \cup \String$, respectively.
+
+\subsection{Arithmetic operations}
+\label{sec:arith}
+
+Lua has operator overloading, and allows programmers to redefine
 the behavior of some operations.

### 2015-02-11 doc/thesis/system.tex
Updating type system appendix to include the projection environment

@@
 have at least one alias.
-The tag \emph{closed} represents \emph{unique} table types
-that may have aliases.
+The tag \emph{closed} represents either \emph{unique} or \emph{open}
+table types that do not allow inserting new keys. 
 In particular, the type of a class has this tag.
@@
 \mylabel{S-FUNCTION}\\
-\dfrac{\senv \vdash s_{2} \subtype s_{1} \;\;\;
-       \senv \vdash r_{1} \subtype r_{2}}
-      {\senv \vdash s_{1} \rightarrow s_{1} \subtype s_{2} \rightarrow r_{2}}
+\dfrac{\senv \vdash s_{3} \subtype s_{1} \;\;\;
+       \senv \vdash s_{2} \subtype s_{4}}
+      {\senv \vdash s_{1} \rightarrow s_{2} \subtype s_{3} \rightarrow s_{4}}
 \end{array}
@@
 \begin{array}{c}
-\mylabel{S-VOID}\\
+\mylabel{S-VOID2}\\
 \senv \vdash \Void \subtype t{*}
@@
 A variadic type $t{*}$ represents a sequence of values of type
-$t \cup \Nil$, and the rule \textsc{S-VOID} handles the case where
+$t \cup \Nil$, and the rule \textsc{S-VOID2} handles the case where
 a given sequence is empty.

### 2015-02-16 doc/thesis/system.tex
Quick fix to #42

@@
 They are novel type system features that let Typed Lua cover several Lua idioms
-and features, such as refinement of tables, multiple return values,
+and features, such as refinement of tables, multiple assignment, multiple return values,
 and optional parameters.
@@
 \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
-& | \; \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s \; | \;
-\mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
-\mathbf{return} \; el \; | \;
-e(el)_{s} \; | \;
-e{:}n(el)_{s}\\
+& | \; \mathbf{local} \; \vec{n{:}T} = el \; \mathbf{in} \; s \; | \;
+\mathbf{local} \; \vec{n} = el \; \mathbf{in} \; s \; | \;
+\mathbf{rec} \; n{:}T = f \; \mathbf{in} \; s\\
+& | \; \mathbf{return} \; el \; | \;
+\lfloor a \rfloor_{0}\\
 e ::= & \;\; \mathbf{nil} \; | \;
-c \; | \;
-{...}_{e} \; | \;
-n_{e} \; | \;
+k \; | \;
+\lfloor {...} \rfloor_{1} \; | \;
+n \; | \;
 e_{1}[e_{2}] \; | \;
-{<}t{>} \; n \; | \;
+{<}T{>} \; n \; | \;
 f \; | \;
-\{ \} \; | \;
 \{ \; \vec{p} \; \} \; | \;
-\{ \; {...}_{m} \; \} \; | \;
-\{ \; \vec{p},{...}_{m} \; \}\\
+\{ \; \vec{p},me \; \}\\
 & | \; e_{1} + e_{2} \; | \;
@@
 \# \; e \; | \;
-e(el)_{e} \; | \;
-e{:}n(el)_{e}\\
+\lfloor a \rfloor_{1}\\
 l ::= & \;\; n_{l} \; | \;
 e_{1}[e_{2}] \; | \;
-n[c] \; {<}t{>}\\
-c ::= & \;\; \mathbf{false} \; | \;
+n[k] \; {<}T{>}\\
+k ::= & \;\; \mathbf{false} \; | \;
 \mathbf{true} \; | \;
@@
 p ::= & \;\; [e_{1}] = e_{2}\\
-el ::= & \;\; \mathbf{nothing} \; | \;
-\vec{e} \; | \;
-me \; | \;
-\vec{e}, me \; | \;
-me^{x} \; | \;
-\vec{e}, me^{x}\\
-me ::= & \;\; e(el)_{m} \; | \;
-e{:}n(el)_{m} \; | \;
-{...}_{m}\\
-m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (){:}r \; fb \; | \;
-\mathbf{fun} \; n_{1}{:}n_{2} \; ({...}{:}t){:}r \; fb\\
-& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; fb \; | \;
-\mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t},{...}{:}t){:}r \; fb\\
-f ::= & \;\; \mathbf{fun} \; (){:}r \; fb \; | \;
-\mathbf{fun} \; ({...}{:}t){:}r \; fb \; | \;
-\mathbf{fun} \; (\vec{n{:}t}){:}r \; fb \; | \;
-\mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; fb\\
+el ::= & \;\; \vec{e} \; | \;
+\vec{e}, me\\
+me ::= & \;\; a \; | \;
+{...}\\
+a ::= & \;\; e(el) \; | \;
+e{:}n(el)\\
+m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}T}){:}S \; fb \; | \;
+\mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}T},{...}{:}T){:}S \; fb\\
+f ::= & \;\; \mathbf{fun} \; (\vec{n{:}T}){:}S \; fb \; | \;
+\mathbf{fun} \; (\vec{n{:}t},{...}{:}T){:}S \; fb\\
 fb ::= & \;\; s \;;\; \mathbf{return} \; el\\

### 2015-02-16 doc/thesis/system.tex
Using capital letters to describe types in subtyping and consistent-subtyping

@@
 \begin{array}{rlr}
-t ::= & & \textsc{first-level types:}\\
-& \;\; l & \textit{literal types}\\
-& | \; b & \textit{base types}\\
+T ::= & & \textsc{first-level types:}\\
+& \;\; L & \textit{literal types}\\
+& | \; B & \textit{base types}\\
 & | \; \Nil & \textit{nil type}\\
@@
 & | \; \Self & \textit{self type}\\
-& | \; t \cup t & \textit{disjoint union types}\\
-& | \; s \rightarrow s & \textit{function types}\\
-& | \; \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique|open|closed|regular} & \textit{table types}\\
+& | \; T_{1} \cup T_{2} & \textit{disjoint union types}\\
+& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
+& | \; \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique|open|fixed|closed} & \textit{table types}\\
 & | \; x & \textit{type variables}\\
-& | \; \mu x.t & \textit{recursive types}\\
-& | \; \phi(t,t) & \textit{filter types}\\
+& | \; \mu x.T & \textit{recursive types}\\
+& | \; \phi(T_{1},T_{2}) & \textit{filter types}\\
 & | \; \pi_{i}^{x} & \textit{projection types}\\
 %\multicolumn{3}{c}{}\\
-l ::= & & \textsc{{\small literal types:}}\\
+L ::= & & \textsc{{\small literal types:}}\\
 & \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
 %\multicolumn{3}{c}{}\\
-b ::= & & \textsc{{\small base types:}}\\
+B ::= & & \textsc{{\small base types:}}\\
 & \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
 %\multicolumn{3}{c}{}\\
-k ::= & & \textsc{{\small key types:}}\\
-& \;\; l \; | \; b \; | \; \Value &\\
+K ::= & & \textsc{{\small key types:}}\\
+& \;\; L \; | \; B \; | \; \Value &\\
 %\multicolumn{3}{c}{}\\
-v ::= & & \textsc{{\small value types:}}\\
-& \;\; t \; | \; \Const \; t &\\ 
+V ::= & & \textsc{{\small value types:}}\\
+& \;\; T \; | \; \Const \; T &\\ 
 %\multicolumn{3}{c}{}\\
-s ::= & & \textsc{second-level types:}\\
-& \;\; p & \textit{tuple types}\\
-& | \; s \sqcup s & \textit{unions of tuple types}\\
+S ::= & & \textsc{second-level types:}\\
+& \;\; P & \textit{tuple types}\\
+& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
 %\multicolumn{3}{c}{}\\
-p ::= & & \textsc{{\small tuple types:}}\\
+P ::= & & \textsc{{\small tuple types:}}\\
 & \;\; \Void & \textit{void type}\\
-& | \; t* & \textit{variadic types}\\
-& | \; t \times p & \textit{pair types}
+& | \; T{*} & \textit{variadic types}\\
+& | \; T \times P & \textit{pair types}
 \end{array}
@@
 
-Union types $t_{1} \cup t_{2}$ represent types that can hold a value
+Union types $T_{1} \cup T_{2}$ represent types that can hold a value
 of two different types.
 
-Function types have the form $s \rightarrow s$ and represent Lua functions,
-where $s$ is a second-level type.
+Function types have the form $S_{1} \rightarrow S_{2}$ and represent Lua functions,
+where $S$ is a second-level type.
 
@@
 The type $\Void$ is the type of an empty tuple.
-A variadic type $t*$ represents a sequence of values of type $t \cup \Nil$;
+A variadic type $T{*}$ represents a sequence of values of type $T \cup \Nil$;
 it is the type of a vararg expression.
@@
 that Lua tables can take.
-The syntactical form of table types is $\{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{tag}$,
-where each $k_{i}$ represents the type of a table key,
-and each $v_{i}$ represents the type of the value that table keys of type $k_{i}$ map to.
+The syntactical form of table types is $\{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n} \}_{tag}$,
+where each $K_{i}$ represents the type of a table key,
+and each $V_{i}$ represents the type of the value that table keys of type $K_{i}$ map to.
 Key types can only be literal types, base types, or the top type.
@@
 The type $\Value$ is an option when we need a loose table type.
-For instance, $\{\Value:\Value\}_{regular}$ represents the type of a
+For instance, $\{\Value:\Value\}_{closed}$ represents the type of a
 table in which both indices and values can have any type.
@@
 
-We also use the tags \emph{unique}, \emph{open}, \emph{closed}, and
-\emph{regular} to classify table types.
+We also use the tags \emph{unique}, \emph{open}, \emph{fixed}, and \emph{closed}
+to classify table types.
 The tag \emph{unique} represents tables with no keys that do not
@@
 have at least one alias.
-The tag \emph{closed} represents either \emph{unique} or \emph{open}
-table types that do not allow inserting new keys. 
+The tag \emph{fixed} represents \emph{unique} table types that we
+do not know how many aliases they have.
 In particular, the type of a class has this tag.
-The tag \emph{regular} represents table types that do not provide
+The tag \emph{closed} represents table types that do not provide
 any guarantees about keys with types not listed in the table type.
 In particular, in the concrete syntax, type annotations, interface
-declarations, and userdata declarations always describe \emph{regular} table types.
+declarations, and userdata declarations always describe \emph{closed} table types.
 In the next sections we explain in more detail why we need
@@
 
-Recursive types have the form $\mu x.t$,
-where $t$ is a first-level type that $x$ represents.
-For instance, $\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{regular}$
+Recursive types have the form $\mu x.T$,
+where $T$ is a first-level type that $x$ represents.
+For instance, $\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}$
 is a type for singly-linked lists of integers.
@@
 variables inside conditions.
-In Section \ref{sec:rules} we show in more detail how our type system
+In Section \ref{sec:fap} we show in more detail how our type system
 uses them to formalize the \texttt{type} predicates that we mentioned
@@
 unions of tuple types into unions of first-level types.
-In Section \ref{sec:rules} we show in more detail how our type system
+In Section \ref{sec:fap} we show in more detail how our type system
 uses them as a mechanism for handling unions of tuple types,
@@
 We present the subtyping rules as a deduction system for the
-subtyping relation $\senv \vdash t_{1} \subtype t_{2}$.
+subtyping relation $\senv \vdash T_{1} \subtype T_{2}$.
 The variable $\senv$ is a set of pairs of recursion variables.
@@
 for defining that literal types are subtypes of their respective base types,
-and that $\Integer$ is subtype of $\Number$:
+and that $\Integer$ is a subtype of $\Number$:
 \[
@@
 \mylabel{S-VALUE}\\
-\senv \vdash t \subtype \Value
+\senv \vdash T \subtype \Value
 \end{array}
@@
 \mylabel{S-UNION1}\\
-\dfrac{\senv \vdash t_{1} \subtype t \;\;\;
-       \senv \vdash t_{2} \subtype t}
-      {\senv \vdash t_{1} \cup t_{2} \subtype t}
+\dfrac{\senv \vdash T_{1} \subtype T \;\;\;
+       \senv \vdash T_{2} \subtype T}
+      {\senv \vdash T_{1} \cup T_{2} \subtype T}
 \end{array}
@@
 \mylabel{S-UNION2}\\
-\dfrac{\senv \vdash t \subtype t_{1}}
-      {\senv \vdash t \subtype t_{1} \cup t_{2}}
+\dfrac{\senv \vdash T \subtype T_{1}}
+      {\senv \vdash T \subtype T_{1} \cup T_{2}}
 \end{array}
@@
 \mylabel{S-UNION3}\\
-\dfrac{\senv \vdash t \subtype t_{2}}
-      {\senv \vdash t \subtype t_{1} \cup t_{2}}
-\end{array}
-\end{array}
-\]
-
-The first rule shows that a union type $t_{1} \cup t_{2}$
-is subtype of $t$ if both $t_{1}$ and $t_{2}$ are subtypes
-of $t$;
-and the other rules show that a type $t$ is subtype
-of a union type $t_{1} \cup t_{2}$ if $t$ is subtype of
-either $t_{1}$ or $t_{2}$.
+\dfrac{\senv \vdash T \subtype T_{2}}
+      {\senv \vdash T \subtype T_{1} \cup T_{2}}
+\end{array}
+\end{array}
+\]
+
+The first rule shows that a union type $T_{1} \cup T_{2}$
+is a subtype of $T$ if both $T_{1}$ and $T_{2}$ are subtypes
+of $T$;
+and the other rules show that a type $T$ is a subtype
+of a union type $T_{1} \cup T_{2}$ if $T$ is a subtype of
+either $T_{1}$ or $T_{2}$.
 
@@
 \mylabel{S-FUNCTION}\\
-\dfrac{\senv \vdash s_{3} \subtype s_{1} \;\;\;
-       \senv \vdash s_{2} \subtype s_{4}}
-      {\senv \vdash s_{1} \rightarrow s_{2} \subtype s_{3} \rightarrow s_{4}}
+\dfrac{\senv \vdash S_{3} \subtype S_{1} \;\;\;
+       \senv \vdash S_{2} \subtype S_{4}}
+      {\senv \vdash S_{1} \rightarrow S_{2} \subtype S_{3} \rightarrow S_{4}}
 \end{array}
@@
 \mylabel{S-VOID2}\\
-\senv \vdash \Void \subtype t{*}
-\end{array}
-\]
-
-A variadic type $t{*}$ represents a sequence of values of type
-$t \cup \Nil$, and the rule \textsc{S-VOID2} handles the case where
+\senv \vdash \Void \subtype T{*}
+\end{array}
+\]
+
+A variadic type $T{*}$ represents a sequence of values of type
+$T \cup \Nil$, and the rule \textsc{S-VOID2} handles the case where
 a given sequence is empty.
@@
 \mylabel{S-PAIR}\\
-\dfrac{\senv \vdash t_{1} \subtype t_{2} \;\;\;
-       \senv \vdash s_{1} \subtype s_{2}}
-      {\senv \vdash t_{1} \times s_{1} \subtype t_{2} \times s_{2}}
-\end{array}
-\]
-
-The subtyping rules for variadic types are not so obvious:
-\[
-\begin{array}{c}
+\dfrac{\senv \vdash T_{1} \subtype T_{2} \;\;\;
+       \senv \vdash P_{1} \subtype P_{2}}
+      {\senv \vdash T_{1} \times P_{1} \subtype T_{2} \times P_{2}}
+\end{array}
+\]
+
+The subtyping rules for variadic types are not so obvious.
+We need six different subtyping rules for variadic types
+to handle all the cases where they can appear.
+
+The rule \textsc{S-VARARG1} is a special rule for handling the
+case where we give a sequence of $\Nil$ to the empty tuple:
+\[
 \begin{array}{c}
@@
 \end{array}
+\]
+
+The rule \textsc{S-VARARG2} handles the case where both tuple types end
+with variadic types:
+\[
+\begin{array}{c}
+\mylabel{S-VARARG2}\\
+\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \cup \Nil}
+      {\senv \vdash T_{1}{*} \subtype T_{2}{*}}
+\end{array}
+\]
+
+This rule shows that $T_{1}{*}$ is a subtype of $T_{2}{*}$
+if $T_{1} \cup \Nil$ is a subtype of $T_{2} \cup \Nil$.
+It explicitly includes $\Nil$ in both sides because otherwise
+$\Nil{*}$ would not be a subtype of several other variadic types.
+For instance, $\Nil{*}$ would not be a subtype of $\Number{*}$,
+as $\Nil \not\subtype \Number$.
+
+The other rules handle the cases where only one tuple type ends with a variadic type:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-VARARG3}\\
+\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2}}
+      {\senv \vdash T_{1}{*} \subtype T_{2} \times \Void}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-VARARG4}\\
+\dfrac{\senv \vdash T_{1} \subtype T_{2} \cup \Nil}
+      {\senv \vdash T_{1} \times \Void \subtype T_{2}{*}}
+\end{array}
 \\ \\
 \begin{array}{c}
-\mylabel{S-VARARG2}\\
-\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2} \cup \Nil}
-      {\senv \vdash t_{1}{*} \subtype t_{2}{*}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-VARARG3}\\
-\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2}}
-      {\senv \vdash t_{1}{*} \subtype t_{2} \times \Void}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-VARARG4}\\
-\dfrac{\senv \vdash t_{1} \subtype t_{2} \cup \Nil}
-      {\senv \vdash t_{1} \times \Void \subtype t_{2}{*}}
+\mylabel{S-VARARG5}\\
+\dfrac{\senv \vdash T_{1}{*} \subtype T_{2} \times \Void \;\;\;
+       \senv \vdash T_{1}{*} \subtype P_{2}}
+      {\senv \vdash T_{1}{*} \subtype T_{2} \times P_{2}}
 \end{array}
@@
 \begin{array}{c}
-\mylabel{S-VARARG5}\\
-\dfrac{\senv \vdash t_{1}{*} \subtype t_{2} \times \Void \;\;\;
-       \senv \vdash t_{1}{*} \subtype s_{2}}
-      {\senv \vdash t_{1}{*} \subtype t_{2} \times s_{2}}
-\end{array}
-\;
-\begin{array}{c}
 \mylabel{S-VARARG6}\\
-\dfrac{\senv \vdash t_{1} \times \Void \subtype t_{2}{*} \;\;\;
-       \senv \vdash s_{1} \subtype t_{2}{*}}
-      {\senv \vdash t_{1} \times s_{1} \subtype t_{2}{*}}
-\end{array}
-\end{array}
-\]
-
-We need six different subtyping rules for variadic types
-to handle all the cases where they can appear.
-The rule \textsc{S-VARARG1} is a special rule for handling the
-case where we give a sequence of $\Nil$ to the empty tuple.
-The rule \textsc{S-VARARG2} handles the case where both tuple types end
-with variadic types, and shows that $t_{1}{*}$ is a subtype of $t_{2}{*}$
-if $t_{1} \cup \Nil$ is a subtype of $t_{2} \cup \Nil$.
-This rule explicitly includes $\Nil$ in both sides because otherwise
-$\Nil{*}$ would not be a subtype of several other variadic types.
-For instance, $\Nil{*}$ would not be subtype of $\Number{*}$,
-as $\Nil \not\subtype \Number$.
-The other rules handle the cases where only one tuple type ends with a variadic type.
+\dfrac{\senv \vdash T_{1} \times \Void \subtype T_{2}{*} \;\;\;
+       \senv \vdash P_{1} \subtype T_{2}{*}}
+      {\senv \vdash T_{1} \times P_{1} \subtype T_{2}{*}}
+\end{array}
+\end{array}
+\]
+
 Note that the case where both tuple types end with the type $\Void$ does
 not require any special rule.
-In the next section we will show that we use these subtyping rules,
+In the next section we will show that we use the subtyping rules for variadic types,
 along with the types $\Value$ and $\Nil$, to make our type system reflect
@@
 \mylabel{S-UNION4}\\
-\dfrac{\senv \vdash s_{1} \subtype s \;\;\;
-       \senv \vdash s_{2} \subtype s}
-      {\senv \vdash s_{1} \sqcup s_{2} \subtype s}
+\dfrac{\senv \vdash S_{1} \subtype S \;\;\;
+       \senv \vdash S_{2} \subtype S}
+      {\senv \vdash S_{1} \sqcup S_{2} \subtype S}
 \end{array}
@@
 \mylabel{S-UNION5}\\
-\dfrac{\senv \vdash s \subtype s_{1}}
-      {\senv \vdash s \subtype s_{1} \sqcup s_{2}}
+\dfrac{\senv \vdash S \subtype S_{1}}
+      {\senv \vdash S \subtype S_{1} \sqcup S_{2}}
 \end{array}
@@
 \mylabel{S-UNION6}\\
-\dfrac{\senv \vdash s \subtype s_{2}}
-      {\senv \vdash s \subtype s_{1} \sqcup s_{2}}
+\dfrac{\senv \vdash S \subtype S_{2}}
+      {\senv \vdash S \subtype S_{1} \sqcup S_{2}}
 \end{array}
@@
 Back to the subtyping rules between first-level types,
-the subtyping rule among a \emph{closed} or \emph{regular}
-table type and another \emph{regular} table type resembles the
+the subtyping rule among a \emph{fixed} or \emph{closed}
+table type and another \emph{closed} table type resembles the
 standard subtyping rule between records:
@@
 \dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
-       \senv \vdash k_{j} \subtype k_{i}' \;\;\;
-       \senv \vdash k_{i}' \subtype k_{j} \;\;\;
-       \senv \vdash v_{j} \subtype_{c} v_{i}'}
-      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{closed|regular} \subtype \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{regular}} \; m \ge n
+       \senv \vdash K_{j} \subtype K_{i}' \;\;\;
+       \senv \vdash K_{i}' \subtype K_{j} \;\;\;
+       \senv \vdash V_{j} \subtype_{c} V_{i}'}
+      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{fixed|closed} \subtype
+                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}} \; m \ge n
 \end{array}
@@
 \mylabel{S-FIELD1}\\
-\dfrac{\senv \vdash v_{1} \subtype v_{2} \;\;\;
-       \senv \vdash v_{2} \subtype v_{1}}
-      {\senv \vdash v_{1} \subtype_{c} v_{2}}
+\dfrac{\senv \vdash V_{1} \subtype V_{2} \;\;\;
+       \senv \vdash V_{2} \subtype V_{1}}
+      {\senv \vdash V_{1} \subtype_{c} V_{2}}
 \end{array}
@@
 \mylabel{S-FIELD2}\\
-\dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash \Const \; v_{1} \subtype_{c} \Const \; v_{2}}
+\dfrac{\senv \vdash V_{1} \subtype V_{2}}
+      {\senv \vdash \Const \; V_{1} \subtype_{c} \Const \; V_{2}}
 \end{array}
@@
 \mylabel{S-FIELD3}\\
-\dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash v_{1} \subtype_{c} \Const \; v_{2}}
+\dfrac{\senv \vdash V_{1} \subtype V_{2}}
+      {\senv \vdash V_{1} \subtype_{c} \Const \; V_{2}}
 \end{array}
@@
 
-There is a limitation on \emph{regular} table types that led us to
+There is a limitation on \emph{closed} table types that led us to
 introduce \emph{open} and \emph{unique} table types.
-If the table constructor had a \emph{regular} table type, then
+If the table constructor had a \emph{closed} table type, then
 programmers would not be able to use it to initialize a variable with
@@
 \[
-\{``x":1, ``y":2\}_{regular} \not\subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{regular}
+\{``x":1, ``y":2\}_{closed} \not\subtype
+\{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
 \]
@@
 \begin{align*}
-& \{``x":\Integer, ``y":\Integer\}_{regular} \not\subtype \\
-& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{regular}
+& \{``x":\Integer, ``y":\Integer\}_{closed} \not\subtype \\
+& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
 \end{align*}
@@
 The following subtyping rule defines the subtyping relation among
-\emph{unique} table types and \emph{regular} table types:
+\emph{unique} table types and \emph{closed} table types:
 \[
@@
        \forall i \in 1..m \; \forall j \in 1..n \;
-       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash v_{i} \subtype_{u} v_{j}' \\
+       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{u} V_{j}' \\
        \forall j \in 1..n \; \not\exists i \in 1..m \;
-       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
+       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
        \end{array}}
-      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
-                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{regular}}
+      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype
+                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
 \end{array}
@@
 It allows covariant keys because we also want to use \emph{unique}
-table types as a way to join table fields that inhabit \emph{regular} table types.
+table types as a way to join table fields that inhabit \emph{closed} table types.
 For instance, we want to use the table constructor to initialize
@@
 \mylabel{S-FIELD4}\\
-\dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash v_{1} \subtype_{u} v_{2}}
+\dfrac{\senv \vdash V_{1} \subtype V_{2}}
+      {\senv \vdash V_{1} \subtype_{u} V_{2}}
 \end{array}
@@
 \mylabel{S-FIELD5}\\
-\dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash \Const \; v_{1} \subtype_{u} \Const \; v_{2}}
+\dfrac{\senv \vdash V_{1} \subtype V_{2}}
+      {\senv \vdash \Const \; V_{1} \subtype_{u} \Const \; V_{2}}
 \end{array}
@@
 \mylabel{S-FIELD6}\\
-\dfrac{\senv \vdash v_{1} \subtype v_{2}}
-      {\senv \vdash v_{1} \subtype_{u} \Const \; v_{2}}
+\dfrac{\senv \vdash V_{1} \subtype V_{2}}
+      {\senv \vdash V_{1} \subtype_{u} \Const \; V_{2}}
 \end{array}
@@
 \mylabel{S-FIELD7}\\
-\dfrac{\senv \vdash \Nil \subtype v}
-      {\senv \vdash \Nil \subtype_{o} v}
+\dfrac{\senv \vdash \Nil \subtype V}
+      {\senv \vdash \Nil \subtype_{o} V}
 \end{array}
@@
 \mylabel{S-FIELD8}\\
-\dfrac{\senv \vdash \Nil \subtype v}
-      {\senv \vdash \Nil \subtype_{o} \Const \; v}
+\dfrac{\senv \vdash \Nil \subtype V}
+      {\senv \vdash \Nil \subtype_{o} \Const \; V}
 \end{array}
@@
 \[
-\{``x":1, ``y":2\}_{unique} \subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{regular}
-\]
-
-Even though we allow width subtyping between \emph{unique} and \emph{regular}
+\{``x":1, ``y":2\}_{unique} \subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
+\]
+
+Even though we allow width subtyping between \emph{unique} and \emph{closed}
 table types, we do not allow it among \emph{unique} and other table types
@@
        \exists j \in 1..n \;
-       \senv \vdash k_{i} \subtype k_{j}' \land \senv \vdash v_{i} \subtype_{u} v_{j}' \\
+       \senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
        \forall j \in 1..n \; \not\exists i \in 1..m \;
-       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
+       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
        \end{array}}
-      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
-                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{unique|open|closed}}
-\end{array}
-\]
-
-The rule that handles subtyping between \emph{open} and \emph{regular} table
+      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype
+                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{unique|open|fixed}}
+\end{array}
+\]
+
+The rule that handles subtyping between \emph{open} and \emph{closed} table
 types allows width subtyping:
@@
        \forall i \in 1..m \; \forall j \in 1..n \;
-       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash v_{i} \subtype_{c} v_{j}' \\
+       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
        \forall j \in 1..n \; \not\exists i \in 1..m \;
-       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
+       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
        \end{array}}
-      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
-                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{regular}}
+      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype
+                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
 \end{array}
@@
 However, the rule that handles subtyping among \emph{open} and
-\emph{open} or \emph{closed} table types does not allow width subtyping:
+\emph{open} or \emph{fixed} table types does not allow width subtyping:
 \[
@@
        \exists j \in 1..n \;
-       \senv \vdash k_{i} \subtype k_{j}' \land \senv \vdash v_{i} \subtype_{c} v_{j}' \\
+       \senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
        \forall j \in 1..n \; \not\exists i \in 1..m \;
-       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
+       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
        \end{array}}
-      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
-                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{open|closed}}
+      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype
+                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{open|fixed}}
 \end{array}
@@
 
-We introduced \emph{closed} table types because we needed a safe way
-to represent the type of classes along with single inheritance.
-The rule that handles subtyping between \emph{closed} table types
+We introduced \emph{fixed} table types because we needed a safe way
+to represent the type of classes that can allow single inheritance
+through the refinement of table types.
+The rule that handles subtyping between \emph{fixed} table types
 does not allow width subtyping, joining fields, and omitting fields,
@@
 \dfrac{\forall i \in 1..n \; \exists j \in 1..n \;\;\;
-       \senv \vdash k_{j} \subtype k_{i}' \;\;\;
-       \senv \vdash k_{i}' \subtype k_{j} \;\;\;
-       \senv \vdash v_{j} \subtype_{c} v_{i}'}
-      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed} \subtype \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
+       \senv \vdash K_{j} \subtype K_{i}' \;\;\;
+       \senv \vdash K_{i}' \subtype K_{j} \;\;\;
+       \senv \vdash V_{j} \subtype_{c} V_{i}'}
+      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{fixed} \subtype
+                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{fixed}}
 \end{array}
@@
 In the next section we will show in more detail how our type system
-uses \emph{unique}, \emph{open}, and \emph{closed} table types to
-handle the refinement of table types.
+uses these tags to handle the refinement of table types and idiomatic
+object-oriented code.
 
@@
 \mylabel{S-AMBER}\\
-\dfrac{\senv[x_{1} \subtype x_{2}] \vdash t_{1} \subtype t_{2}}
-      {\senv \vdash \mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}}
+\dfrac{\senv[x_{1} \subtype x_{2}] \vdash T_{1} \subtype T_{2}}
+      {\senv \vdash \mu x_{1}.T_{1} \subtype \mu x_{2}.T_{2}}
 \end{array}
@@
 The rule \textsc{S-AMBER} also uses the rule \textsc{S-ASSUMPTION}
-to check whether $\mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}$.
+to check whether $\mu x_{1}.T_{1} \subtype \mu x_{2}.T_{2}$.
 Both rules use the set of assumptions $\senv$,
@@
 The rule \textsc{S-AMBER} extends $\senv$ with the assumption
-$x_{1} \subtype x_{2}$ to check whether $t_{1} \subtype t_{2}$.
+$x_{1} \subtype x_{2}$ to check whether $T_{1} \subtype T_{2}$.
 The rule \textsc{S-ASSUMPTION} allows the rule \textsc{S-AMBER}
@@
 \mylabel{S-UNFOLDR}\\
-\dfrac{\senv \vdash t_{1} \subtype [x \mapsto \mu x.t_{2}]t_{2}}
-      {\senv \vdash t_{1} \subtype \mu x.t_{2}}
+\dfrac{\senv \vdash T_{1} \subtype [x \mapsto \mu x.T_{2}]T_{2}}
+      {\senv \vdash T_{1} \subtype \mu x.T_{2}}
 \end{array}
@@
 \mylabel{S-UNFOLDL}\\
-\dfrac{\senv \vdash [x \mapsto \mu x.t_{1}]t_{1} \subtype t_{2}}
-      {\senv \vdash \mu x.t_{1} \subtype t_{2}}
+\dfrac{\senv \vdash [x \mapsto \mu x.T_{1}]T_{1} \subtype T_{2}}
+      {\senv \vdash \mu x.T_{1} \subtype T_{2}}
 \end{array}
@@
 \{ & ``info":\Integer, \\
-   & ``next":\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{regular} \cup \Nil \}_{unique} \subtype \\
-& \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{regular}
+   & ``next":\mu x.\{``info":\Integer,
+                     ``next":x \;\cup\; \Nil\}_{closed} \cup \Nil \}_{unique} \subtype \\
+& \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}
 \end{align*}
@@
 Filter types are subtypes only of themselves and of $\Value$.
-More precisely, a filter type $\phi(t_{1},t_{2})$ is a subtype of
-the same filter type $\phi(t_{1},t_{2})$, which shares the same
-types $t_{1}$ and $t_{2}$, and it is also a subtype of $\Value$.
+More precisely, a filter type $\phi(T_{1},T_{2})$ is a subtype of
+the same filter type $\phi(T_{1},T_{2})$, which shares the same
+types $T_{1}$ and $T_{2}$, and it is also a subtype of $\Value$.
 
@@
 We present the consistent-subtyping rules as a deduction system for
-the consistent-subtyping relation $\senv \vdash t_{1} \lesssim t_{2}$.
+the consistent-subtyping relation $\senv \vdash T_{1} \lesssim T_{2}$.
 As in the subtyping relation, $\senv$ is also a set of pairs of
@@
 \mylabel{C-ANY1}\\
-\senv \vdash t \lesssim \Any
+\senv \vdash T \lesssim \Any
 \end{array}
@@
 \mylabel{C-ANY2}\\
-\senv \vdash \Any \lesssim t
+\senv \vdash \Any \lesssim T
 \end{array}
@@
 If we had set the type $\Any$ as both bottom and top types of our
-subtyping relation, then any type $t_{1}$ would be a subtype of
-any other type $t_{2}$.
+subtyping relation, then any type $T_{1}$ would be a subtype of
+any other type $T_{2}$.
 The consequence of this is that all programs would type check without errors.
 This would happen due to the transitivity of subtyping, that is,
-we would be able to down-cast any type $t_{1}$ to $\Any$ and then up-cast
-$\Any$ to any other type $t_{2}$.
+we would be able to down-cast any type $T_{1}$ to $\Any$ and then up-cast
+$\Any$ to any other type $T_{2}$.
 The rules \textsc{C-ANY1} and \textsc{C-ANY2} are the rules that

### 2015-02-18 doc/thesis/typedlua.tex
Updating syntax documentation
Minor improvements to the text about the informal overview of Typed Lua

@@
 they relate to Lua.
-These examples shall give an informal overview of our optional type system.
+These examples give an informal overview of our optional type system.
 We will use typing rules to present the formalization of our optional type
 system in the next chapter.
-All the examples that we present along this chapter run in our Typed Lua compiler.
+All the examples that we present in this chapter run in our Typed Lua compiler.
 
@@
 In this section we present the Typed Lua types that may appear on annotations.
-We explain all Typed Lua types and syntactical extensions along this chapter.
+We explain all Typed Lua types and syntactical extensions in this chapter.
 
@@
 \begin{verbatim}
-    local function succ (x:integer):integer
-      return x + 1
+    local function succ (n:integer):integer
+      return n + 1
     end
@@
 \begin{verbatim}
-    local function succ (x:integer)
-      return x + 1
+    local function succ (n:integer)
+      return n + 1
     end
@@
 Typed Lua assigns the type \texttt{any} to the unannotated declarations
-that it cannot infer a more specific type.
-More precisely, Typed Lua cannot infer more specific types to input
+that it does not infer a more specific type.
+More precisely, Typed Lua does not infer more specific types to input
 parameters of function declarations and to the return type of
-recursive functions, as global type inference is undecidable in a
-language with subtyping \citep{wells1999typability}.
+recursive functions.
 
@@
 calling a function is like doing a multiple assignment where the
-left hand side is the parameter list and the right hand side is the
+left-hand side is the parameter list and the right-hand side is the
 argument list.
@@
 In the default mode, the compiler adds a variadic tail to the
-type of the parameter list and to the return type whenever the
+type of the parameter list and to the return type if the
 programmer does not specify one.
@@
 arity mismatch.
+
 We will use the following function to illustrate how these two
@@
 compiles without warnings in both operation modes.
-This happens because in the right hand side of the multiple
+This happens because in the right-hand side of the multiple
 assignment, only the first value produced by the first call
-to \texttt{m} gets used, so the type of the right hand side
+to \texttt{m} gets used, so the type of the right-hand side
 is \texttt{(integer, integer, string, nil*)}, which is
-consistent with the type of the left hand side
+consistent with the type of the left-hand side
 \texttt{(integer, integer, string, value*)}.
@@
     end
-
-    print(message("Typed Lua"))         --> Typed Lua
-    print(message("Typed Lua", "Hi "))  --> Hi Typed Lua
 \end{verbatim}
@@
 Typed Lua uses the following rule to type this idiom: 
-if the left hand side of \texttt{or} has type \texttt{t|nil} and
-the right hand side has type \texttt{t} then the \texttt{or} expression has
+if the left-hand side of \texttt{or} has type \texttt{t|nil} and
+the right-hand side has type \texttt{t} then the \texttt{or} expression has
 type \texttt{t}.
@@
 next assignment to \texttt{v}.
-In the case of \texttt{greeting}, the assignment refines its type
+In the case of \texttt{greeting}, the assignment constrains its type
 from \texttt{string|nil} to \texttt{string}.
+
+After we define the function \texttt{message}, we can call it
+with a missing argument in both operation modes:
+\begin{verbatim}
+    print(message("Typed Lua"))         --> Hello Typed Lua
+    print(message("Typed Lua", "Hi "))  --> Hi Typed Lua
+\end{verbatim}
+
+The type of \texttt{message} is
+\texttt{(string, string|nil, value*) -> (string, nil*)}
+in the default mode, and
+\texttt{(string, string|nil) -> (string)}
+in the strict mode.
+The call \texttt{message("Typed Lua")} compiles without
+any warnings in both modes, because the argument list
+\texttt{("Typed Lua", nil*)} is consistent with
+\texttt{(string, string|nil, value*)} in the default mode,
+and it is also consistent with  
+\texttt{(string, string|nil)} in the strict mode.
+The compiler includes \texttt{nil*} in the tail of
+an argument list in both modes, as it is necessary
+to replace any optional parameters that may appear in
+a function declaration.
 
@@
 
-When Typed Lua finds a union of tuples in the right hand side of
+When Typed Lua finds a union of tuples in the right-hand side of
 a declaration, it assigns \emph{projection types} to the
-variables that appear unannotated in the left hand side of the
+variables that appear unannotated in the left-hand side of the
 declaration.
@@
 first-level types that have a dependency relation.
-We will discuss projection types in more detail in the next chapter.
+We will discuss projection types in more detail in Section \ref{sec:fap}.
 
@@
 
-The overloading mechanism of Typed Lua have a limitation:
-the return type cannot depend on the input type.
+The overloading mechanism of Typed Lua has a limitation:
+the return type cannot depend on the input types.
 Next, we show an example:
@@
 Typed Lua tags a variable that holds a table type as either
-\emph{unique}, \emph{open}, or \emph{closed}.
-If a variable gets is type from a table constructor then it is
+\emph{unique}, \emph{open}, \emph{fixed}, or \emph{closed}.
+If a variable gets its type from a table constructor then it is
 unique, otherwise it is closed.
@@
 because it has no alias.
+In Section \ref{sec:oop} we will show that we use the tag
+fixed to describe the type of classes.
 Typed Lua also has different subtyping rules that handle these
@@
 \textit{interface} ::= & \;\; [\textbf{local}] \; \textbf{interface} \; \textit{typedec}\\
-\textit{typedec} ::= & \;\; \textit{Name} \; {\textit{decitem}} \; \textbf{end}\\
+\textit{typedec} ::= & \;\; \textit{Name} \; \{\textit{decitem}\} \; \textbf{end}\\
 \textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
@@
 elements in this list, and then traverses the list printing
-each element of the list:
+each list element:
 \begin{verbatim}
@@
 
-Note that the type of \texttt{l} is \texttt{Element} in the
-first line of the \texttt{while} body, but it is restored to
-\texttt{Element|nil} in the second line of the \texttt{while} body
-due to the assignment.
-These rules allow this example to compile without any warnings,
-and we discuss them in more detail in the next chapter.
+Note that the type of \texttt{l} is \texttt{Element} inside
+the \texttt{while} body, and the assignment \texttt{l = l.next}
+restores the type of \texttt{l} to \texttt{Element|nil}.
+This means that this example compiles without any warnings, because the
+\texttt{while} statement also uses the small set of \texttt{type}
+predicates that we mentioned in Section \ref{sec:unions}.
 
@@
 that is not based on tags.
-This specific rule let programmers discriminate unions of table types
+This specific rule lets programmers discriminate unions of table types
 and resembles the pattern matching from functional programming.
@@
 If that is the case, it constrains the type of \texttt{v} inside
-the \texttt{if} to the table type that has the key type
-$\texttt{l}_{\texttt{1}}$ that maps to the type $\texttt{l}_{\texttt{2}}$.
+the \texttt{if} to the table type that has a key of type
+$\texttt{l}_{\texttt{1}}$ which maps to a value of type $\texttt{l}_{\texttt{2}}$.
 In our example, the expression \texttt{e.tag == "Add"} makes Typed Lua
 constrain the type of \texttt{e} to \texttt{\{"tag":"Add", 1:Exp, 2:Exp\}}.
-If we did not add this special predicate, we would not be able to traverse the AST,
+If we did not add this special predicate, we would not be able to traverse an AST,
 as any attempt to index an union raises a compile-time warning.
@@
 An idiomatic way for defining modules in Lua is to declare only locals
-in the source file and return a table constructor at the end of the source file.
+and return a table constructor at the end of the source file.
 The returned constructor includes the members that the module should export.
@@
 This module has the same type of the module from the previous example.
-
-%There is still a third way for defining modules in Lua through the
-%\texttt{module} function, which has been deprecated.
-%Even though this style has been deprecated, our survey from
-%Section \ref{sec:statistics} indicated that around a third of the
-%Lua modules that are in LuaRocks still use this style.
-%Thus, Typed Lua also supports this style.
 
@@
 
-This example shows that Typed Lua assigns the type \texttt{self}
-to the implicit parameter \texttt{self}.
+Typed Lua assigns the type \texttt{self} to the implicit parameter \texttt{self}.
 The type \texttt{self} represents the type of the \emph{receiver}
 in method definitions and method calls.
-In this example, the refinement of table types allows Typed Lua
-to bind the type \texttt{self} with the same type of \texttt{Shape}:
+In this example, Typed Lua binds the type \texttt{self} to
+the table type \texttt{\{"x":number, "y":number\}} inside \texttt{move}.
+This example also uses the refinement of table types to build the
+type of \texttt{Shape}:
 \begin{verbatim}
@@
 Furthermore, Typed Lua also allows programmers to create description
-files for exporting statically typed interfaces for a Lua modules
+files for exporting statically typed interfaces for Lua modules
 that are written in C.
@@
 \textit{userdata} ::= & \;\; \textbf{userdata} \; \textit{typedec}\\
-\textit{typedec} ::= & \;\; \textit{Name} \; {\textit{decitem}} \; \textbf{end}\\
+\textit{typedec} ::= & \;\; \textit{Name} \; \{\textit{decitem}\} \; \textbf{end}\\
 \textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
@@
 \textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
-\textit{id} ::= & \; [\textbf{const}] \; \textit{Name}\\
+\textit{id} ::= & \;\; [\textbf{const}] \; \textit{Name}\\
 \textit{typealias} ::= & \;\; \textbf{typealias} \; \textit{Name} \; \texttt{`='} \; \textit{type}
@@
 nominal type checking, so two userdata that export exactly the same members,
-are not subtype of each other, because they do not share the same brand.
+but do not have the same name, are not subtype of each other,
+because they do not share the same brand.
 

### 2015-02-19 doc/thesis/evaluation.tex
Typing LuaFileSystem and adding its evaluation results
Changing the evaluation categories to use percentages

@@
 We performed some case studies on existing Lua libraries
-to evaluate the applicability of our type system.
+to evaluate the design of our type system.
 For each library, we used Typed Lua to either annotate its modules
@@
 interesting cases.
-
-We selected seven case studies to evaluate two aspects of Typed Lua:
-\begin{enumerate}
-\item how precisely it can describe the type of the interface of a module;
-\item whether it provides guarantees that the code matches the interface.
-\end{enumerate}
 
@@
 
-LuaSocket \citep{luasocket} was the third case study that we used to
-evaluate Typed Lua.
-We chose LuaSocket as a case study because it is the most popular Lua library.
+LuaSocket \citep{luasocket} and LuaFileSystem \citep{luafilesystem} were
+the third and fourth case studies that we used to evaluate Typed Lua.
+We chose them because they are the most popular Lua libraries.
+We wrote a script that builds the dependency graph of Lua libraries
+that are in the LuaRocks repository, and used this dependency graph
+to identify the most popular Lua libraries.
 
-There are three case studies that we randomly selected from the LuaRocks
+We also randomly selected three case studies from the LuaRocks
 repository, they are: HTTP Digest \citep{luahttpdigest},
 Typical \citep{luatypical}, and Mod 11 \citep{luamod11}.
-The first one provides client side HTTP digest authentication for Lua.
-The second one is an extension to the primitive function \texttt{type}.
-The third one is a generator and checker of modulo 11 numbers.
-We randomly selected these libraries to evaluate Typed Lua for
-annotating existing Lua modules.
+The first provides client side HTTP digest authentication for Lua.
+The second is an extension to the primitive function \texttt{type}.
+The third is a generator and checker of modulo 11 numbers.
+We randomly selected three case studies because we wanted to evaluate
+Typed Lua for annotating existing libraries that are writting in Lua,
+as the previous case studies are mostly libraries that are writting in C.
 
 The Typed Lua compiler is the last case study that we evaluated.
-We chose the Typed Lua compiler as a case study because it is a large application,
-and bootstrapping the compiler also provides some confidence in the type system.
+We chose it as a case study because it is a large application.
+Besides, it is a case study that evaluates the evolution of a script
+to a program.
+
+We used these case studies to evaluate two aspects of Typed Lua:
+\begin{enumerate}
+\item how precisely it can describe the type of the interface of a module;
+\item whether it provides guarantees that the code matches the interface.
+\end{enumerate}
 
@@
 \begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|c|}
+\begin{tabular}{|l|c|c|c|c|c|c|}
 \hline
-\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
 \hline
-Lua Standard Library & 80 & 6 & 8 & 35 & 129 & 62\% \\
+Lua Standard Library & 62\% & 5\% & 6\% & 27\% & 129 \\ % 100%
 \hline
-MD5 & 13 & 0 & 0 & 0 & 13 & 100\% \\
+MD5 & 100\% & 0\% & 0\% & 0\% & 13 \\ % 100%
 \hline
-LuaSocket & 110 & 1 & 2 & 10 & 123 & 89\% \\
+LuaSocket & 89\% & 1\% & 2\% & 8\% & 123 \\ % 100%
 \hline
-HTTP Digest & 0 & 0 & 1 & 0 & 1 & 0\% \\
+LuaFileSystem & 89\% & 0\% & 11\% & 0\% & 19 \\ % 100%
 \hline
-Typical & 1 & 0 & 0 & 0 & 1 & 100\% \\
+HTTP Digest & 0\% & 0\% & 100\% & 0\% & 1 \\ % 100%
 \hline
-Modulo 11 & 7 & 0 & 0 & 2 & 9 & 78\% \\
+Typical & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
 \hline
-Typed Lua Compiler & 144 & 0 & 1 & 9 & 154 & 94\% \\
+Modulo 11 & 78\% & 0\% & 0\% & 22\% & 9 \\ % 100%
+\hline
+Typed Lua Compiler & 93\% & 0\% & 1\% & 6\% & 154 \\ % 100%
 \hline
@@
 We split the members of each case study into four categories:
-\emph{easy}, \emph{poly}, \emph{inter}, and \emph{hard}.
-The \emph{easy} category groups the members that we could give
-a precise static type.
-The \emph{poly} category groups the members that we made minimal
-use of the dynamic type as a replacement for the lack of type parameters.
-The \emph{inter} category groups the members that we could give
-a static type that is not precise enough because they require
-intersection types.
-The \emph{hard} category groups the members that we had to rely
-on the dynamic type because our type system cannot give a precise
-static type to them.
-The last column of the table shows the percentage of members that
-are under the \emph{easy} category for each case study.
-This percentage gives a lower bound on how much static type safety
+\emph{easy}, \emph{poly}, \emph{over}, and \emph{hard}.
+The \emph{easy} category shows the percentage of members that
+we could give a precise static type.
+The \emph{poly} category shows the percentage of members that
+we made minimal use of the dynamic type as a replacement for the
+lack of type parameters.
+The \emph{over} category shows the percentage of members that
+we gave a static type that is not precise enough,
+as these members are overloaded functions that require intersection
+types to describe their precise static types.
+The \emph{hard} category shows the percentage of members that
+we had to rely on the dynamic type because they rely on reflection.
+The last column of the table shows the total number of members
+of each case study.
+The results that we obtained for the \emph{easy} category give
+a lower bound on how much static type safety
 we could add to each one of our case studies.
@@
 For each case study, we split the evaluation results according
-to the modules that each one of them include.
-This separation by case study allows us to better discuss the
-contributions and limitations of our type system.
+to the modules that each one of them include,
+and we discuss the contributions and limitations that these
+evaluation results represent in our type system.
 
@@
 \begin{center}
-\begin{tabular}{|c|c|c|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|}
 \hline
-\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
 \hline
-base & 9 & 1 & 2 & 14 & 26 & 35\% \\
+base & 34\% & 4\% & 8\% & 54\% & 26 \\ % 100%
 \hline
-coroutine & 1 & 0 & 0 & 6 & 7 & 14\% \\
+coroutine & 14\% & 0\% & 0\% & 86\% & 7 \\ % 100%
 \hline
-package & 5 & 0 & 0 & 3 & 8 & 62\% \\
+package & 62\% & 0\% & 0\% & 38\% & 8 \\ % 100%
 \hline
-string & 12 & 0 & 0 & 4 & 16 & 75\% \\
+string & 75\% & 0\% & 0\% & 25\% & 16 \\ % 100%
 \hline
-utf8 & 6 & 0 & 0 & 0 & 6 & 100\% \\
+utf8 & 100\% & 0\% & 0\% & 0\% & 6 \\ % 100%
 \hline
-table & 1 & 5 & 0 & 1 & 7 & 14\% \\
+table & 15\% & 71\% & 0\% & 14\% & 7 \\ % 100%
 \hline
-math & 22 & 0 & 4 & 1 & 27 & 81\% \\
+math & 81\% & 0\% & 15\% & 4\% & 27 \\ % 100%
 \hline
-io & 15 & 0 & 0 & 6 & 21 & 71\% \\
+io & 71\% & 0\% & 0\% & 29\% & 21 \\ % 100%
 \hline
-os & 9 & 0 & 2 & 0 & 11 & 82\% \\
+os & 82\% & 0\% & 18\% & 0\% & 11 \\ % 100%
 \hline
@@
 \begin{center}
-\begin{tabular}{|c|c|c|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|}
 \hline
-\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
 \hline
-md5 & 13 & 0 & 0 & 0 & 13 & 100\% \\
+md5 & 100\% & 0\% & 0\% & 0\% & 13 \\ % 100%
 \hline
@@
 \begin{center}
-\begin{tabular}{|c|c|c|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|}
 \hline
-\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
 \hline
-socket & 50 & 0 & 0 & 10 & 60 & 83\% \\
+socket & 83\% & 0\% & 0\% & 17\% & 60 \\ % 100%
 \hline
-ftp & 5 & 0 & 1 & 0 & 6 & 83\% \\
+ftp & 83\% & 0\% & 17\% & 0\% & 6 \\ % 100%
 \hline
-http & 4 & 0 & 1 & 0 & 5 & 80\% \\
+http & 80\% & 0\% & 20\% & 0\% & 5 \\ % 100%
 \hline
-smtp & 7 & 0 & 0 & 0 & 7 & 100\% \\
+smtp & 100\% & 0\% & 0\% & 0\% & 7 \\ % 100%
 \hline
-mime & 17 & 0 & 0 & 0 & 17 & 100\% \\
+mime & 100\% & 0\% & 0\% & 0\% & 17 \\ % 100%
 \hline
-ltn12 & 19 & 1 & 0 & 0 & 20 & 95\% \\
+ltn12 & 95\% & 5\% & 0\% & 0\% & 20 \\ % 100%
 \hline
-url & 8 & 0 & 0 & 0 & 8 & 100\% \\
+url & 100\% & 0\% & 0\% & 0\% & 8 \\ % 100%
 \hline
@@
 
+\section{LuaFileSystem}
+
+LuaFileSystem is a library that extends the set of functions
+for manipulating file systems in Lua.
+It contains just the \texttt{lfs} module that is written in C,
+so we used Typed Lua's description file to type it.
+Table \ref{tab:evallfs} summarizes the evaluation results for LuaFileSystem.
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|}
+\hline
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
+\hline
+lfs & 89\% & 0\% & 11\% & 0\% & 19 \\ % 100%
+\hline
+\end{tabular}
+\end{center}
+\caption{Evaluation results for LuaFileSystem}
+\label{tab:evallfs}
+\end{table}
+
+Even though we could type most of the functions exported by the
+\texttt{lfs} module, we could not type two overloaded functions
+due to the lack of intersection types in our type system.
+
 \section{HTTP Digest}
@@
 \begin{center}
-\begin{tabular}{|c|c|c|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|}
 \hline
-\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
 \hline
-http-digest & 0 & 0 & 1 & 0 & 1 & 0\% \\
+http-digest & 0\% & 0\% & 100\% & 0\% & 1 \\ % 100%
 \hline
@@
 \begin{center}
-\begin{tabular}{|c|c|c|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|}
 \hline
-\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
 \hline
-typical & 1 & 0 & 0 & 0 & 1 & 100\% \\
+typical & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
 \hline
@@
 \begin{center}
-\begin{tabular}{|c|c|c|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|}
 \hline
-\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
 \hline
-mod11 & 7 & 0 & 0 & 2 & 9 & 78\% \\
+mod11 & 78\% & 0\% & 0\% & 22\% & 9 \\ % 100%
 \hline
@@
 \begin{center}
-\begin{tabular}{|c|c|c|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|}
 \hline
-\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
 \hline
-tlast & 46 & 0 & 1 & 0 & 47 & 98\% \\
+tlast & 98\% & 0\% & 2\% & 0\% & 47 \\ % 100%
 \hline
-tltype & 65 & 0 & 0 & 0 & 65 & 100\% \\
+tltype & 100\% & 0\% & 0\% & 0\% & 65 \\ % 100%
 \hline
-tlst & 26 & 0 & 0 & 0 & 26 & 100\% \\
+tlst & 100\% & 0\% & 0\% & 0\% & 26 \\ % 100%
 \hline
-tllexer & 2 & 0 & 0 & 9 & 11 & 18\% \\
+tllexer & 18\% & 0\% & 0\% & 82\% & 11 \\ % 100%
 \hline
-tlparser & 1 & 0 & 0 & 0 & 1 & 100\% \\
+tlparser & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
 \hline
-tldparser & 1 & 0 & 0 & 0 & 1 & 100\% \\
+tldparser & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
 \hline
-tlchecker & 2 & 0 & 0 & 0 & 2 & 100\% \\
+tlchecker & 100\% & 0\% & 0\% & 0\% & 2 \\ % 100%
 \hline
-tlcode & 1 & 0 & 0 & 0 & 1 & 100\% \\
+tlcode & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
 \hline

### 2015-02-23 doc/thesis/evaluation.tex

@@
 We wrote a script that builds the dependency graph of Lua libraries
-that are in the LuaRocks repository, and used this dependency graph
+that are in the LuaRocks repository, and uses this dependency graph
 to identify the most popular Lua libraries.
@@
 We randomly selected three case studies because we wanted to evaluate
-Typed Lua for annotating existing libraries that are writting in Lua,
-as the previous case studies are mostly libraries that are writting in C.
+Typed Lua for annotating existing libraries that are written in Lua,
+as the previous case studies are mostly libraries that are written in C.
 
@@
 \hline
-Lua Standard Library & 62\% & 5\% & 6\% & 27\% & 129 \\ % 100%
+Lua Standard Library & 62\% & 6\% & 9\% & 23\% & 129 \\ % 100%
 \hline
@@
 For each case study, we split the evaluation results according
-to the modules that each one of them include,
-and we discuss the contributions and limitations that these
-evaluation results represent in our type system.
+to the modules that each one of them include.
+We use these split results to discuss the contributions and
+limitations of our type system.
 
@@
 \hline
-base & 34\% & 4\% & 8\% & 54\% & 26 \\ % 100%
+base & 35\% & 12\% & 8\% & 45\% & 26 \\ % 100%
 \hline
@@
 \hline
-table & 15\% & 71\% & 0\% & 14\% & 7 \\ % 100%
+table & 14\% & 72\% & 14\% & 0\% & 7 \\ % 100%
 \hline
-math & 81\% & 0\% & 15\% & 4\% & 27 \\ % 100%
+math & 81\% & 0\% & 19\% & 0\% & 27 \\ % 100%
 \hline
-io & 71\% & 0\% & 0\% & 29\% & 21 \\ % 100%
+io & 71\% & 0\% & 10\% & 19\% & 21 \\ % 100%
 \hline
@@
 
-We could give precise static types to several members of
-the Lua Standard Library, though some of them are still
-hard to type with our current type system.
-Now we will discuss why some modules are difficult to type,
-and what kind of limitations they represent in our type system.
+The \texttt{base} module was very difficult to type
+because it includes several functions that rely on reflection,
+as the \emph{hard} category shows.
+For instance, the functions \texttt{pairs} and \texttt{getmetatable}
+are in this category.
+While \texttt{pairs} traverses all keys and values that are stored
+in a given table, \texttt{getmetatable} returns the metatable of a
+given table.
 
-The \texttt{base} module was quite hard to type because it
-includes several functions that rely on reflection.
-For instance, the function \texttt{pairs} traverses all
-keys and values that are stored in a given table.
-As another example, the function \texttt{getmetatable}
-returns the metatable of a given table.
+There are some functions in the \texttt{base} module that
+we could not give a precise static type because our type
+system does not have parametric polymorphism,
+as the \emph{poly} category shows.
+This is the case of \texttt{ipairs}, \texttt{rawget}, and
+\texttt{rawset}.
 
-We could not give a precise static type to \texttt{ipairs},
-because our type system does not have parametric polymorphism.
-This function returns an iterator function that traverses a
-list of elements.
-
-The \texttt{base} module also includes two overloaded functions:
-\texttt{tonumber} and \texttt{collectgarbage}.
+The \texttt{base} module also includes some overloaded functions,
+as the \emph{over} category shows.
+We could not type these functions because our type system does
+not include intersection types.
+This is the case of \texttt{tonumber} and \texttt{collectgarbage}.
 
@@
 
-We could not type most of the members of the \texttt{coroutine} module
-because our type system cannot describe the computational effects of a program.
+The \texttt{coroutine} module was also very difficult to type
+because our type system cannot describe the computational effects
+of a program.
+The \emph{hard} category shows the amount of functions that we
+could not give a precise static type for this reason.
 Lua has one-shot delimited continuations \citep{james2011yield}
@@
 to represent the type \emph{thread}.
-Still, we could type the function \texttt{coroutine.isyieldable},
-as it has no input parameters, and simply returns a boolean that
-indicates whether the running coroutine is yieldable.
 
-We could type most of the members of the \texttt{package} module,
-but we could not type the following tables: \texttt{package.loaded},
-\texttt{package.preload}, and \texttt{package.loaders}.
+Still, we could give a precise static type to one function in
+the \texttt{coroutine} module, as the \emph{easy} category shows.
+The function \texttt{coroutine.isyieldable} has no input parameters,
+and it simply returns a boolean that indicates whether the running
+coroutine is yieldable.
+
+We could give precise static types to the constants and functions
+of the \texttt{package} module, but we could not give precise static
+to the following tables:
+\texttt{package.loaded}, \texttt{package.preload}, and \texttt{package.loaders}.
 The first stores loaded modules, while the others store module \emph{loaders}.
@@
 that is, their types depend on the modules a program loads.
+For this reason, they are in the \emph{hard} category.
 
-The \texttt{string} module was not that hard to type, but it includes
-some functions that we could not type, as they rely on format
-strings.
+We could give precise static types to most of the functions of the
+\texttt{string} module, but we could not give precise static types
+to the functions that rely on format strings.
 For instance, the type of the arguments that we pass to
-\texttt{string.format} depend on the format string we are using.
+\texttt{string.format} must match the format string we are using.
 It is fine to call \texttt{string.format("\%d", 1)}, but
 \texttt{string.format("\%d", true)} raises a run-time error.
+These functions that rely on format strings are in the \emph{hard} category.
 
 Lua programmers often use the functions of the \texttt{string} module
-in an object-oriented style, and we treat this feature as a
-special case in the implementation.
-For this reason, we did not include the typing of the \texttt{string}
-methods in our evaluation results.
+in an object-oriented style.
+For instance, \texttt{string.len("foo")} and \texttt{("foo"):len()}
+produce the same result.
+We did not include the typing of the \texttt{string} methods in our
+evaluation results, because we treat this feature as a special case
+in the implementation.
 
@@
 The \texttt{table} module was specially difficult to type because
-most of its functions require parametric polymorphism.
+most of its functions require parametric polymorphism,
+as the \emph{poly} category shows.
 These functions either receive or return a list of elements, and
@@
 However, the lack of parametric polymorphism did not prevent us from
-giving a precise type to \texttt{table.concat}, as it operates over lists
-where all elements are strings or numbers.
+giving a precise type to one function of the \texttt{table} module,
+as the \emph{easy} category shows.
+We could give a precise static type to \texttt{table.concat},
+as it operates over lists where all elements are strings or numbers.
 
-Even if our type system had parametric polymorphism, it would
-be difficult to type \texttt{table.insert}, as its type depends on
+Even if our type system had parametric polymorphism, there is
+still one function of the \texttt{table} module that we could
+not give a precise static type because it is an overloaded function,
+as the \emph{over} category shows.
+This function is \texttt{table.insert}, and its type depends on
 the calling arity.
@@
 Even though the \texttt{math} module looks straightforward to type,
-it includes several overloaded functions.
-For instance, \texttt{math.abs(1)} returns the integer \texttt{1},
+the \emph{over} category shows that it includes several overloaded functions.
+For instance, the functions \texttt{math.abs} and \texttt{math.random}
+are in this category.
+
+In the case of \texttt{math.abs},
+calling \texttt{math.abs(1)} returns the integer \texttt{1},
 while \texttt{math.abs(1.1)} returns the floating point \texttt{1.1}.
+Even though we gave this function the type \texttt{(number) -> (number)},
+it is not precise enough because the return type is always \texttt{number}
+independently of the argument type.
+In other words, the return type should be \texttt{integer} when
+the argument type is also \texttt{integer}, and the return type
+should be \texttt{number} when the argument type is also \texttt{number}.
 
-The \texttt{math} module also includes a function that is hard to type:
-the function \texttt{math.random}.
-It is difficult to type because its type depends on the calling arity.
+In the case of \texttt{math.random}, its type depends on the calling arity.
 Calling \texttt{math.random()} returns a random floating point
@@
 and returns an integer between this interval.
-This function also does not follow the semantics of Lua on discarding extra arguments,
+Like \texttt{table.insert}, this function also does not follow the
+semantics of Lua on discarding extra arguments,
 and generates a run-time error whenever we pass more than two arguments.
@@
 and its methods.
-For this reason, we included the typing of the \texttt{file}
-methods in our evaluation results.
+The evaluation results include both implicit and explicit operations.
 
-We could type most of members of the \texttt{io} module,
-but we could not precisely type the functions \texttt{io.close},
-\texttt{io.read}, and \texttt{io.lines}.
-We also could not type the methods \texttt{file:close},
-\texttt{file:read}, and \texttt{file:lines}.
+We could give precise static types to most of the members of the
+\texttt{io} module, but its categories \emph{over} and \emph{hard}
+show that it includes several members that we could not give a precise
+static type.
+While the function \texttt{io.close} and the method \texttt{file:close}
+are in the \emph{over} category,
+the functions \texttt{io.read} and \texttt{io.lines} are
+in the \emph{hard} category along with the methods
+\texttt{file:read} and \texttt{file:lines}.
 
-The function \texttt{io.close} is difficult to type
-because its return type depends on whether the file handle that is
-being closed was created with \texttt{popen}.
+We could not precisely type \texttt{io.close} because it is
+an overloaded function.
+This means that its return type depends on whether the file handle
+that is being closed was created with \texttt{io.popen}.
 Calling \texttt{io.close(f)} returns the tuple \texttt{(boolean?, string, number)}
-when \texttt{f} was created with \texttt{popen},
-and it returns the tuple \texttt{(boolean)?} otherwise.
-The method \texttt{file:close} is difficult to type for the same reason.
+when \texttt{f} was created with \texttt{io.popen},
+but it returns the tuple \texttt{(boolean)?} otherwise.
+The method \texttt{file:close} has the same issue.
 
-The functions \texttt{io.read} and \texttt{io.lines} are difficult
-to type because they are overloaded functions.
-For instance, \texttt{io.read("l")} returns a string or \texttt{nil},
-\texttt{io.read("n")} returns a number or \texttt{nil}, and calling
+We could not precisely type \texttt{io.read} because its return type
+relies on format strings.
+For instance, calling \texttt{io.read("l")} returns a string or \texttt{nil},
+\texttt{io.read("n")} returns a number or \texttt{nil}, and
 \texttt{io.read("l", "n")} returns a string or \texttt{nil} and a number or \texttt{nil}.
 The function \texttt{io.lines}, and the methods \texttt{file:read} and
-\texttt{file:lines} are difficult to type for the same reason.
+\texttt{file:lines} have the same issue.
 
-The function \texttt{os.execute} is the only member of the \texttt{os}
-module that we could not give a precise type, as it is an overloaded
-function.
-More precisely, \texttt{os.execute()} returns a boolean, while
-\texttt{os.execute(c)} returns \texttt{(boolean?, string, number)},
-where \texttt{c} is a string that represents the command to be executed.
+There are two functions in the \texttt{os} module that we could
+not give a precise static type because they are overloaded functions.
+The functions that are in the \emph{over} category are
+\texttt{os.date} and \texttt{os.execute}.
 
@@
 
-We could type most of the members in the \texttt{socket} module,
-which is the C core.
-However, this module includes some functions such as
-\texttt{socket.try}, \texttt{socket.protect}, and \texttt{socket.skip}
-that are hard to type because they rely on reflection.
-For instance, we can use \texttt{socket.skip} to choose the number of
+We could give precise static types to most of the members
+in the \texttt{socket} module, which is the C core.
+However, this module includes some functions that
+we could not give a precise static type because they
+rely on reflection, as the \emph{hard} category shows.
+For instance, \texttt{socket.skip} is a function that is
+in this category.
+We can use this function to choose the number of
 values that we want to return.
@@
 
-We could type most of the members of the modules \texttt{ftp} and
-\texttt{http}, but we could not precisely type
-the functions \texttt{ftp.get} and \texttt{http.request} because
-they are overloaded functions.
+We could give precise static types to most of the members of the
+modules \texttt{ftp} and \texttt{http}, but we could not precisely type
+two overloaded functions: \texttt{ftp.get} and \texttt{http.request}.
 
@@
 The low-level filters are written in C, while the high-level filters
-use the function \texttt{ltn12.filter.cycle} along with the low level
+use the function \texttt{ltn12.filter.cycle} along with the low-level
 filters to create standard filters.
@@
 
-Even though we could type most of the functions exported by the
-\texttt{lfs} module, we could not type two overloaded functions
+Even though we could precisely type most of the functions exported
+by the \texttt{lfs} module, we could not type two overloaded functions
 due to the lack of intersection types in our type system.
@@
 However, we hit some limitations of our type system while annotating this module.
-First, it uses the function \texttt{getmetatable} to get a table and
-checks whether this table has the field \texttt{\string_\string_type} or not.
+
+First, it uses the \texttt{getmetatable} to get a table and
+checks whether this table has the field \texttt{\string_\string_type}.
 We could not give a precise type to \texttt{getmetatable}, so we used the dynamic
 type \texttt{any} as its return type, and this generates a warning.
-Second, the module returns a metatable that extends \texttt{\string_\string_call}
+
+Second, it returns a metatable that extends \texttt{\string_\string_call}
 with \texttt{typical.type}, that is, we can use the module itself as a function,
@@
 of the \texttt{typical} module to support \texttt{\string_\string_call}.
+
 Third, the module uses \texttt{ipairs} to iterate over an array of functions,
@@
 a warning when we try to use the indexed value inside the \texttt{for} body.
+As we mentioned in this chapter, we use the dynamic type as a replacement for
+the lack of type parameters.
+This means that we get warnings inside an \texttt{ipairs} iteration because
+all iterated elements have the dynamic type.
 We removed this warning using the numeric \texttt{for} to perform the same loop.
@@
 build LPeg patterns. 
-We could not precisely type LPeg patterns because our
+Even though LPeg is the third most popular Lua module,
+we cannot not precisely type LPeg patterns because our
 type system still does not support overloading arithmetic
 operators.
-In this module, we could only type the two error reporting functions
-that it exports.
+In the \texttt{tllexer} module, we could only give precise static types
+to two error reporting functions that it exports.
 
@@
 the modules export, we had issues to write mutually recursive functions.
-This kind of functions often appear on compilers construction
+This kind of functions often appear in compilers construction
 to traverse the data structures that they use.

### 2015-02-23 doc/thesis/related.tex

@@
-In this chapter we review related work as a way to indentify the novel
+In this chapter we review related work as a way to identify the novel
 features of our type system, what are the features that it shares with
@@
 features from other type systems.
-We split this chapter into two parts: in the first part we review
-other Lua projects, while in the second part we review other
+We split this chapter into two sections: in the first section we review
+other Lua projects, while in the second section we review other
 projects that are not related to Lua.
@@
 Tidal Lock covers a little subset of Lua.
-Statements include declaration of local variables, multiple assignments,
+Statements include declaration of local variables, multiple assignment,
 function application, and the return statement.
@@
 
+Sol \citep{sol} is an experimental optional type system for Lua.
+Its type system is similar to ours, as it includes literal types,
+union types, and function types that handle variadic functions.
+However, it does not handle the refinement of tables and it
+includes different types for tables.
+Sol types tables as lists, maps, and objects.
+Its object types handle a specific object-oriented idiom that
+Sol introduces.
+
+Lua Analyzer \citep{luaanalyzer} is an optional type system for Lua
+that is specially designed to work in the Löve Studio,
+an IDE for game developing using the Löve framework.
+It works in Lua 5.1 only, and uses type annotations inside comments.
+It is unsound by design because its dynamic type is both
+top and bottom in the subtyping relation.
+
+Lua Analyzer shares some features with Typed Lua, and also
+has some interesting features that we do not have in Typed Lua.
+It has similar rules for handling the \texttt{or} idiom and
+discriminating union types inside conditions.
+However, these rules are limited to the \texttt{nil} tag only.
+It also includes different types for typing tables.
+It includes regular record types that maps names to types,
+array types, and map types.
+Even though it does not support the refinement of tables,
+it allows the definition of nominal table types that simulate classes.
+This system allows it to type check custom class systems,
+which are common in Lua.
+Function types also support multiple return values and
+variadic functions, but they do not support overloading the
+return type.
+Recently, it included experimental support for type aliases and generics.
+
 Luacheck \citep{luacheck} is a tool that performs static analysis on Lua code.
@@
 operations that take into account static type information.
+Currently, Ravi extends the Lua Virtual Machine to support few types:
+\texttt{int}, \texttt{double}, arrays of integers, and arrays of doubles.
 
@@
 
+Mypy \citep{mypy} is an optional type system for Python.
+The type system of mypy is similar to the type system of
+Reticulated Python, but mypy does not insert run-time checks
+and it has parametric polymorphism.
+In contrast, Reticulated Python can type variadic functions,
+but mypy cannot.
+Recently, Guido van Rossum, Python's author, proposed a
+standard syntax for type annotations in Python \citep{PEP483}
+that is extremely inspired by mypy \citep{PEP484}.
+The main goal of this proposal is to make easier building
+static analysis tools for Python.
+Typing \citep{typing} is a tool that is being developed to
+implement this proposal.
+
 Hack \citep{hack} is a new programming language that runs on the

### 2015-02-23 doc/thesis/review.tex

@@
 For instance, Dart has covariant arrays, while TypeScript has
-covariant function return types,
+covariant parameter types in function signatures,
 besides the interaction between statically and dynamically
@@
 their optional type systems rather than a formal one.
-In the next section we shall see that we can use some features
+In the next section we will show that we can use some features
 of gradual typing \citep{siek2006gradual,siek2007objects} to
@@
 To do that, \citet{siek2006gradual} extended the simply typed
-lambda calculus with the dynamic type $?$, as we shall see in
+lambda calculus with the dynamic type $?$, as we can see in
 Figure \ref{fig:gtlc}.
-In gradual typing, type annotations are optional, so an untyped
+In gradual typing, type annotations are optional, and an untyped
 variable is syntactic sugar for a variable whose declared type is
@@
 \begin{array}{llr}
-t ::= & & \emph{types}\\
-& \;\; \Number & \emph{base type number}\\
-& | \; \String & \emph{base type string}\\
-& | \; ? & \emph{dynamic type}\\
-& | \; t \rightarrow t & \emph{function types}\\
-e ::= & & \emph{expressions}\\
-& \;\; l & \emph{literals}\\
-& | \; x & \emph{variables}\\
-& | \; \lambda x{:}t{.}e & \emph{abstractions}\\
-& | \; e_{1} e_{2} & \emph{application}
+T ::= & & \textsc{types}\\
+& \;\; \Number & \textit{base type number}\\
+& | \; \String & \textit{base type string}\\
+& | \; ? & \textit{dynamic type}\\
+& | \; T \rightarrow T & \textit{function types}\\
+e ::= & & \textsc{expressions}\\
+& \;\; l & \textit{literals}\\
+& | \; x & \textit{variables}\\
+& | \; \lambda x{:}T{.}e & \textit{abstractions}\\
+& | \; e_{1} e_{2} & \textit{application}
 \end{array}
@@
 The central idea of gradual typing is the \emph{consistency}
-relation, written $t_{1} \sim t_{2}$.
+relation, written $T_{1} \sim T_{2}$.
 The consistency relation allows implicit conversions to and from the
@@
 \begin{array}{c}
-t \sim t \mylabel{C-REFL}
+T \sim T \mylabel{C-REFL}
 \end{array}
@@
 \begin{array}{c}
-t \sim \;? \mylabel{C-DYNR}
+T \sim \;? \mylabel{C-DYNR}
 \end{array}
@@
 \begin{array}{c}
-? \sim t \mylabel{C-DYNL}
+? \sim T \mylabel{C-DYNL}
 \end{array}
@@
 \begin{array}{c}
-\dfrac{t_{3} \sim t_{1} \;\;\; t_{2} \sim t_{4}}
-      {t_{1} \rightarrow t_{2} \sim t_{3} \rightarrow t_{4}} \mylabel{C-FUNC}
+\dfrac{T_{3} \sim T_{1} \;\;\; T_{2} \sim T_{4}}
+      {T_{1} \rightarrow T_{2} \sim T_{3} \rightarrow T_{4}} \mylabel{C-FUNC}
 \end{array}
@@
 The rule \textsc{T-ABS} evaluates the expression $e$ with an environment
-$\env$ that binds the variable $x$ to the type $t_{1}$, and the resulting
-type is the the function type $t_{1} \rightarrow t_{2}$.
-The rule \textsc{T-APP1} handles function calls where the type of the
+$\env$ that binds the variable $x$ to the type $T_{1}$, and the resulting
+type is the the function type $T_{1} \rightarrow T_{2}$.
+The rule \textsc{T-APP1} handles function calls where the type of a
 function is dynamically typed; in this case, the argument type may have
 any type and the resulting type has the dynamic type.
-The rule \textsc{T-APP2} handles function calls where the type of the
+The rule \textsc{T-APP2} handles function calls where the type of a
 function is statically typed; in this case, the argument type should
@@
 \begin{matrix}
-\dfrac{\env(x) = t}
-      {\env \vdash x:t} \mylabel{T-VAR}
+\dfrac{\env(x) = T}
+      {\env \vdash x:T} \mylabel{T-VAR}
 \;\;\;
-\dfrac{type(l) = t}
-      {\env \vdash l:t} \mylabel{T-LIT}
+\dfrac{type(l) = T}
+      {\env \vdash l:T} \mylabel{T-LIT}
 \\ \\
-\dfrac{\env, x:t_{1} \vdash e:t_{2}}
-      {\env \vdash \lambda x:t_{1}.e:t_{1} \rightarrow t_{2}} \mylabel{T-ABS}
+\dfrac{\env[x \mapsto T_{1}] \vdash e:T_{2}}
+      {\env \vdash \lambda x:T_{1}.e:T_{1} \rightarrow T_{2}} \mylabel{T-ABS}
 \;\;\;
 \dfrac{\env \vdash e_{1}:\;? \;\;\;
-       \env \vdash e_{2}:t}
+       \env \vdash e_{2}:T}
       {\env \vdash e_{1} e_{2}:\;?} \mylabel{T-APP1}
 \\ \\
-\dfrac{\env \vdash e_{1}:t_{1} \rightarrow t_{2} \;\;\;
-       \env \vdash e_{2}:t_{3} \;\;\;
-       t_{3} \sim t_{1}}
-      {\env \vdash e_{1} e_{2}:t_{2}} \mylabel{T-APP2}
+\dfrac{\env \vdash e_{1}:T_{1} \rightarrow T_{2} \;\;\;
+       \env \vdash e_{2}:T_{3} \;\;\;
+       T_{3} \sim T_{1}}
+      {\env \vdash e_{1} e_{2}:T_{2}} \mylabel{T-APP2}
 \end{matrix}
@@
 as the top and the bottom of the subtying relation:
-$t \subtype \;?$ and $? \subtype t$.
+$T \subtype \;?$ and $? \subtype T$.
 Subtyping is transitive, so we know that
@@
 orthogonal to the subtyping relation, so we can combine them to achieve
-the \emph{consistent-subtyping} relation, written $t_{1} \lesssim t_{2}$.
+the \emph{consistent-subtyping} relation, written $T_{1} \lesssim T_{2}$.
 This relation is essential for designing gradual type systems for
@@
 \begin{array}{c}
-\dfrac{t_{3} \subtype t_{1} \;\;\; t_{2} \subtype t_{4}}
-      {t_{1} \rightarrow t_{2} \subtype t_{3} \rightarrow t_{4}} \mylabel{S-FUN}
+\dfrac{T_{3} \subtype T_{1} \;\;\; T_{2} \subtype T_{4}}
+      {T_{1} \rightarrow T_{2} \subtype T_{3} \rightarrow T_{4}} \mylabel{S-FUN}
 \end{array}
@@
 typed lambda calculus extended with the dynamic type $?$.
-Even though we could have used the reflexive rule $t \subtype t$ to express
+Even though we could have used the reflexive rule $T \subtype T$ to express
 the rules \textsc{S-NUM}, \textsc{S-STR}, and \textsc{S-ANY},
@@
 consistent-subtyping relation of \citet{siek2007objects}, but it is
-also written $t_{1} \lesssim t_{2}$.
-The consistent-or-subtype relation holds when $t_{1} \sim t_{2}$
-or $t_{1} <: t_{2}$, where $<:$ is transitive and $\sim$ is not.
+also written $T_{1} \lesssim T_{2}$.
+The consistent-or-subtype relation holds when $T_{1} \sim T_{2}$
+or $T_{1} <: T_{2}$, where $<:$ is transitive and $\sim$ is not.
 Alore also extends its optional type system to include optional

### 2015-02-26 doc/thesis/system.tex
Fixing the section about typing rules to use capitals

@@
 most interesting rules of our type system.
-This core limits control flow to if and while statements,
-has explicit type annotations, and explicit scope for variables.
-It also has explicit method declarations and explicit method calls
-instead of treating them as syntactic sugar.
-We use this reduced core because it simplifies the presentation
-of our type system.
 Appendix \ref{app:rules} presents the full set of typing rules.
+
+Our core limits control flow to if and while statements;
+it has explicit type annotations, explicit scope for variables,
+explicit method declarations, and explicit method calls.
+Here is a list of features that are not present in our reduced core:
+\begin{itemize}
+\item labels and goto statements are not present because they
+are difficult to handle along with our simplified form of
+\emph{flow typing}, and they are out of the scope for now;
+\item explicit blocks are not present because we are already using
+explicit scope for variables;
+\item other loop structures such as repeat-until, numeric for,
+and generic for are not present because we can use while to
+express them;
+\item fields in the table constructor are reduced to a single
+form because we can use this form to express the missing forms; 
+\item arithmetic operators are reduced to addition and unary minus
+because the other operators have similar typing rules;
+\item relational operators are reduced to equality and less than
+because the other operators have similar typing rules;
+\item bitwise operators are not present because they are similar
+to integer arithmetic operators.
+\end{itemize}
 
@@
 $s$ are statements, $e$ are expressions, $l$ are left-hand values,
-$el$ are expression lists, $c$ are literal constants, $p$ are table fields,
+$el$ are expression lists, $k$ are literal constants, $p$ are table fields,
 $me$ are expressions with multiple results, $f$ are function declarations,
-$m$ are method declarations, $n$ are variable names,
-$t$ are first-level types, and $r$ are second-level types.
-The notation $\vec{n{:}t}$ denotes the non-empty list
-$n_{1}{:}t_{1}, ..., n_{n}{:}t_{n}$.
-
-Our reduced core also splits function application, method application, and
-the vararg expression (${...}$) into different syntactic categories.
-It uses $e(el)_{s}$ to denote function applications that produce no value
-because they appear as statements,
-$e(el)_{e}$ to denote function applications that produce only one value,
-even if the function application returns multiple values,
-and $e(el)_{m}$ to denote function applications that produces as many values
-as the function application returns.
-It uses the same categories for method applications.
-For vararg expressions, it only uses ${...}_{e}$ and ${...}_{m}$,
-as the vararg expression cannot appear as a statement.
+$m$ are method declarations, $a$ are function and method applications,
+$n$ are variable names, $T$ are first-level types, and $S$ are second-level types.
+The notation $\vec{n{:}T}$ denotes the list $n_{1}{:}T_{1}, ..., n_{n}{:}T_{n}$,
+where $n$ can be zero.
+
+Our reduced core includes two statements for declaring local variables.
+One with explicit type annotations, and another without type annotations.
+While we use the former to show how our type system handles the declaration
+of annoatted variables, we use the latter to show how our type system
+handles the declaration of unannotated variables through local type inference
+and also the introduction of projection types.
+
+Our reduced core also includes a truncation operator $\lfloor \rfloor$ for
+function applications, method applications, and the vararg expression.
+We use $\lfloor a \rfloor_{0}$ to denote function and method applications
+that produce no value because they appear as statements.
+We use $\lfloor a \rfloor_{1}$ to denote function and method applications
+that produce only one value, even if the function or method application
+returns mutiple values.
+We use $\lfloor {...} \rfloor_{1}$ to denote a vararg expression that
+produces only one value, as it can appear in the middle of an
+expression list.
 
 We also include two kinds of type coercions in our core language:
-the left-hand value $n[c] \; {<}t{>}$ and the expression ${<}t{>} \;n$.
+the left-hand value $n[k] \; {<}T{>}$ and the expression ${<}T{>} \;n$.
 Both allow the refinement of table types.
@@
 of tables in the presence of refinement.
-We use $n_{e}$ when they appear as expressions and $n_{l}$ when they
+We use $n$ when they appear as expressions and $n_{l}$ when they
 appear as left-hand values.
@@
 We present the typing rules as a deduction system for the typing relations
-$\env_{1}, \penv \vdash s, \env_{2}$ and $\env_{1}, \penv \vdash e : t, \env_{2}$.
+$\env_{1}, \penv \vdash s, \env_{2}$ and $\env_{1}, \penv \vdash e : T, \env_{2}$.
 We use the first one for typing statements and the second one for typing expressions.
@@
 The second relation means that given a type environment $\env_{1}$
-and a projection environment $\penv$, typing a expression $e$ has
-type $t$ and produces a new type environment $\env_{2}$.
+and a projection environment $\penv$, typing an expression $e$ has
+type $T$ and produces a new type environment $\env_{2}$.
 We need two typing relations because only expressions have types,
@@
 handling projection types.
-We use $\env_{1}[n \mapsto t]$ to extend the environment $\env_{1}$
-with the variable $n$ that maps to type $t$.
-We use $\penv[x \mapsto r]$ to extend the environment $\penv$ with
-the variable $x$ that maps to the second-level type $r$.
+We use $\env_{1}[n \mapsto T]$ to extend the environment $\env_{1}$
+with the variable $n$ that maps to type $T$.
+We use $\penv[x \mapsto S]$ to extend the environment $\penv$ with
+the variable $x$ that maps to the second-level type $S$.
 
@@
 \begin{array}{c}
-\dfrac{\env_{1}, \penv \vdash el:r_{1}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
-       r_{1} \lesssim r_{2}}
+\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash \vec{l}:S_{2}, \env_{3} \;\;\;
+       S_{1} \lesssim S_{2}}
       {\env_{1}, \penv \vdash \vec{l} = el,\env_{3}}
@@
 \mylabel{T-LHSLIST}\\
-\dfrac{\env, \penv \vdash l_{i}:t_{i}, \env_{i} \;\;\;
+\dfrac{\env, \penv \vdash l_{i}:T_{i}, \env_{i} \;\;\;
        \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
        n = |\;\vec{l}\;|}
-      {\env, \penv \vdash \vec{l}:t_{1} \times ... \times t_{n} \times \Value{*}, \env_{m}}
+      {\env, \penv \vdash \vec{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
 \end{array}
@@
 left-hand side value $l_{i}$.
-Then, it takes each type $t_{i}$ and builds the tuple type for
+Then, it takes each type $T_{i}$ and builds the tuple type for
 the assignment rule, placing $\Value{*}$ in the end.
@@
 \mylabel{T-EXPLIST2}\\
-\dfrac{\env, \penv \vdash e_{i}:t_{i}, \env_{i} \;\;\;
+\dfrac{\env, \penv \vdash e_{i}:T_{i}, \env_{i} \;\;\;
        \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
        n = |\;\vec{e}\;|}
-      {\env, \penv \vdash \vec{e}:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{m}}
+      {\env, \penv \vdash \vec{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
 \end{array}
@@
 \mylabel{T-EXPLIST3}\\
-\dfrac{\env_{1}, \penv \vdash me:t_{1} \times ... \times t_{n} \times \Void, \env_{2}}
-      {\env_{1}, \penv \vdash me:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{2}}
+\dfrac{\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Void, \env_{2}}
+      {\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{2}}
 \end{array}
@@
 \mylabel{T-EXPLIST4}\\
-\dfrac{\env_{1}, \penv \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{2}}
-      {\env_{1}, \penv \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{2}}
+\dfrac{\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n}{*}, \env_{2}}
+      {\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n}{*}, \env_{2}}
 \end{array}
@@
 \[
-x, y, z = f()_{m}
+x, y, z = f()
 \]
@@
 \[
-x = f()_{m}
+x = f()
 \]
@@
 \mylabel{T-APPLY1}\\
-\dfrac{\env_{1}, \penv \vdash e:r_{1} \rightarrow r_{2}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash el:r_{3}, \env_{3} \;\;\;
-       r_{3} \lesssim r_{1}}
-      {\env_{1}, \penv \vdash e(el)_{m}:r_{2}, \env_{3}}
+\dfrac{\env_{1}, \penv \vdash e:S_{1} \rightarrow S_{2}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash el:S_{3}, \env_{3} \;\;\;
+       S_{3} \lesssim S_{1}}
+      {\env_{1}, \penv \vdash e(el):S_{2}, \env_{3}}
 \end{array}
@@
 \mylabel{T-STMAPPLY1}\\
-\dfrac{\env_{1}, \penv \vdash e(el)_{m}:r, \env_{2}}
-      {\env_{1}, \penv \vdash e(el)_{s},\env_{2}}
+\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
+      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{0},\env_{2}}
 \end{array}
@@
 \mylabel{T-EXPAPPLY1}\\
-\dfrac{\env_{1}, \penv \vdash e(el)_{m}:r, \env_{2}}
-      {\env_{1}, \penv \vdash e(el)_{e}:first(r), \env_{2}}
+\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
+      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{1}:first(S), \env_{2}}
 \end{array}
@@
 first(\Void) & = \Nil\\
-first(t{*}) & = t \cup \Nil\\
-first(t \times r) & = t\\
-first(r_{1} \sqcup r_{2}) & = first(r_{1}) \cup first(r_{2})
+first(T{*}) & = T \cup \Nil\\
+first(T \times P) & = T\\
+first(S_{1} \sqcup S_{2}) & = first(S_{1}) \cup first(S_{2})
 \end{align*}
@@
 \[
-f(``foo")_{m}
+f(``foo")
 \]
@@
 \[
-f(``foo",1,2,3)_{m}
+f(``foo",1,2,3)
 \]
@@
 \[
-f(1)_{m}
+f(1)
 \]
@@
 \[
-f(1,2,3)_{m}
+f(1,2,3)
 \]
@@
 
+Our abstract syntax reduces the syntactic forms of the table constructor
+into two forms: $\{\;\vec{p}\;\}$ and $\{\;\vec{p},me\;\}$.
+The first uses a list of table fields $p_{1}, ..., p_{n}$,
+where $p$ is $[e_{1}] = e_{2}$ and $n$ can be zero.
+The second uses a list of table fields and an expression that can
+produce multiple values.
+We did not include in the abstract syntax a table constructor that uses
+lists of expressions, because we can express this form using lists of
+table fields.
+For instance, we can use the table constructor
+$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to express $\{ ``x", ``y", ``z" \}$.
+
 The simplest expression involving tables in the empty table constructor.
@@
 
-Our abstract syntax reduces the more complex uses of the table
-constructor into three forms: $\{\;\vec{p}\;\}$, $\{\;{...}_{m}\;\}$,
-and $\{\;\vec{p},{...}_{m}\;\}$.
-The first one uses a list of table fields $[e_{1}] = e_{2}$,
-the second one uses just the vararg expression, and the third one uses both.
-We did not include in the abstract syntax a table constructor
-that uses a list of expressions, because we can express it using the first form.
-For instance, we can use the table constructor
-$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to express $\{ ``x", ``y", ``z" \}$.
-
-The rule \textsc{T-CONSTRUCTOR2} type checks a table type of the first form:
+The rule \textsc{T-CONSTRUCTOR2} type checks a table type with a non-empty
+list of table fields:
 \[
@@
 \mylabel{T-CONSTRUCTOR2}\\
-\dfrac{\env_{i}, \penv \vdash p_{i}:(k_{i},v_{i}), \env_{i+1} \;\;\;
+\dfrac{\env_{i}, \penv \vdash p_{i}:(K_{i},V_{i}), \env_{i+1} \;\;\;
        n = |\;\vec{p}\;| \;\;\;
-       t = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique} \;\;\;
-       wf(t)}
-      {\env_{1}, \penv \vdash \{\;\vec{p}\;\}:t, \env_{n+1}}
+       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
+       wf(T)}
+      {\env_{1}, \penv \vdash \{\;\vec{p}\;\}:T, \env_{n+1}}
 \end{array}
@@
 First, the rule \textsc{T-CONSTRUCTOR2} uses the auxiliary relation
-$\env_{1}, \penv \vdash p : (k,v), \env_{2}$ to type check each table
+$\env_{1}, \penv \vdash p : (K,V), \env_{2}$ to type check each table
 field $p_{i}$.
@@
 and a projection environment $\penv$, typing a table field $p$
-produces the pair $(k,v)$ and a new type environment $\env_{2}$. 
-In the pair $(k,v)$, $k$ is type key type, while $v$ is the value type.
-
-Then, the rule \textsc{T-CONSTRUCTOR2} uses each pair $(k_{i},v_{i})$
+produces the pair $(K,V)$ and a new type environment $\env_{2}$. 
+In the pair $(K,V)$, $K$ is the key type, while $V$ is the value type.
+
+Then, the rule \textsc{T-CONSTRUCTOR2} uses each pair $(K_{i},V_{i})$
 to build the table type that express the type of the given constructor, and
@@
 \[
-\forall i \not\exists j \; i \not= j \wedge k_{i} \lesssim k_{j}
+\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
 \]
@@
 \mylabel{T-FIELD1}\\
-\dfrac{\env_{1}, \penv \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3} \;\;\;
-       t_{1} \subtype \Boolean}
-      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (t_{1},rt(t_{2})), \env_{3}}
+\dfrac{\env_{1}, \penv \vdash e_{1}:T_{1}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash e_{2}:T_{2}, \env_{3} \;\;\;
+       T_{1} \subtype \Boolean}
+      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (T_{1},close(T_{2})), \env_{3}}
 \end{array}
@@
 \mylabel{T-FIELD2}\\
-\dfrac{\env_{1}, \penv \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3} \;\;\;
-       t_{1} \subtype \Number}
-      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (t_{1},rt(t_{2})), \env_{3}}
+\dfrac{\env_{1}, \penv \vdash e_{1}:T_{1}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash e_{2}:T_{2}, \env_{3} \;\;\;
+       T_{1} \subtype \Number}
+      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (T_{1},close(T_{2})), \env_{3}}
 \end{array}
@@
 \mylabel{T-FIELD3}\\
-\dfrac{\env_{1}, \penv \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3} \;\;\;
-       t_{1} \subtype \String}
-      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (t_{1},rt(t_{2})), \env_{3}}
-\end{array}
-\end{array}
-\]
-
-These rules use the function \emph{rt} in the type of the value
-because our type system allows only \emph{regular} table types to
+\dfrac{\env_{1}, \penv \vdash e_{1}:T_{1}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash e_{2}:T_{2}, \env_{3} \;\;\;
+       T_{1} \subtype \String}
+      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (T_{1},close(T_{2})), \env_{3}}
+\end{array}
+\end{array}
+\]
+
+These rules use the function \emph{close} in the type of the value
+because our type system allows only \emph{closed} table types to
 appear in the type of the values.
-The function \emph{rt} promotes all table types to \emph{regular}.
+The function \emph{close} promotes all table types to \emph{closed}.
 We made this restriction because our type system does not keep track
 of aliases to table fields.
-This means that allowing \emph{unique}, \emph{open}, and \emph{closed}
+This means that allowing \emph{unique}, \emph{open}, and \emph{fixed}
 table types to appear in the type of the values would allow the creation
@@
 \mylabel{T-FIELD4}\\
-\dfrac{\env_{1}, \penv \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3}}
-      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (\Value,rt(t_{2})), \env_{3}}
+\dfrac{\env_{1}, \penv \vdash e_{1}:T_{1}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash e_{2}:T_{2}, \env_{3}}
+      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (\Value,close(T_{2})), \env_{3}}
 \end{array}
@@
 As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
-has type $\{``x":1, ``y":\{``z":2\}_{regular}\}_{unique}$ through the rules
+has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ through the rules
 \textsc{T-FIELD3} and \textsc{T-CONSTRUCTOR2}.
 
-Typed Lua handles arrays as hashes that map integers to some type $t$.
+Typed Lua handles arrays as hashes that map integers to some type $T$.
 In Lua, programmers often use the vararg expression to initialize arrays.
@@
 \mylabel{T-CONSTRUCTOR3}\\
-\dfrac{\env({...}) = t}
-      {\env, \penv \vdash \{{...}_{m}\}:\{\Integer{:}t \cup \Nil\}_{unique}, \env}
+\dfrac{\env({...}) = T}
+      {\env, \penv \vdash \{{...}\}:\{\Integer{:}T \cup \Nil\}_{unique}, \env}
 \end{array}
@@
 \dfrac{\begin{array}{c}
-       \env_{i}, \penv \vdash p_{i}:(k_{i},v_{i}), \env_{i+1} \;\;\;
+       \env_{i}, \penv \vdash p_{i}:(K_{i},V_{i}), \env_{i+1} \;\;\;
        n = |\;\vec{p}\;| \;\;\;
-       \env_{i+1}({...}) = t_{v}\\
-       t = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, \Integer{:}t_{v} \cup \Nil\}_{unique} \;\;\;
-       wf(t)
+       \env_{i+1}({...}) = T_{v}\\
+       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, \Integer{:}T_{v} \cup \Nil\}_{unique} \;\;\;
+       wf(T)
        \end{array}}
-      {\env_{1}, \penv \vdash \{\;\vec{p},\;{...}_{m}\;\}:t, \env_{n+1}}
+      {\env_{1}, \penv \vdash \{\;\vec{p},\;{...}\;\}:T, \env_{n+1}}
 \end{array}
@@
 If we assume that $...$ is in the environment and has type $\String$,
-the table constructor $\{{...}_{m}\}$ has type $\{\Integer:\String \cup \Nil\}_{unique}$
+the table constructor $\{{...}\}$ has type $\{\Integer:\String \cup \Nil\}_{unique}$
 through the rule \textsc{T-CONSTRUCTOR3}.
@@
 If we assume that $...$ is in the environment and has type $\String$,
-the table constructor $\{[``x"] = ``foo", [``y"] = \mathbf{true}, {...}_{m}\}$
+the table constructor $\{[``x"] = ``foo", [``y"] = \mathbf{true}, {...}\}$
 has type $\{``x":``foo", ``y":\mathbf{true}, \Integer:\String \cup \Nil\}_{unique}$
@@
 As another example, the rule \textsc{T-CONSTRUCTOR4} does not type check
-the table constructor $\{ [1] = ``foo", [10] = \mathbf{true}, {...}_{m} \}$,
+the table constructor $\{ [1] = ``foo", [10] = \mathbf{true}, {...} \}$,
 because the keys $1$ and $10$ overlap the array part,
@@
 \dfrac{\begin{array}{c}
-       \env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique}\\
-       \env_{1}, \penv \vdash c:k, \env_{2} \;\;\;
-       \not \exists i \in 1..n \; k \lesssim k_{i} \;\;\;
-       v = rt(t)
+       \env_{1}(n) = \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n} \}_{open|unique}\\
+       \env_{1}, \penv \vdash k:K, \env_{2} \;\;\;
+       \not \exists i \in 1..n \; K \lesssim K_{i} \;\;\;
+       V = close(T)
        \end{array}}
-      {\env_{1}, \penv \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, k{:}v\}_{open|unique}]}
+      {\env_{1}, \penv \vdash n[k] {<}T{>}:V, \env_{2}[n \mapsto \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, K{:}V\}_{open|unique}]}
 \end{array}
@@
 For this reason, Typed Lua treats accesses to global variables as field accesses
-to an open table in the top-level scope.
+to an \emph{open} table in the top-level scope.
 For instance,
@@
 uses field assignment to add fields $``x"$ and $``y"$ to $\string_ENV$.
-In this example and in the next examples we assume that
+In this example and in the next examples, we assume that
 $\string_ENV$ is in the environment and has type $\{\}_{open}$.
@@
 We also do not allow the refinement of table types to introduce
-fields with table types that are not \emph{regular}.
+fields with table types that are not \emph{closed}.
 For instance,
@@
 \end{center}
-refines the type of $\string_ENV$ from $\{\}_{open}$ to $\{``x":\{\}_{regular}\}_{open}$.
+refines the type of $\string_ENV$ from $\{\}_{open}$ to $\{``x":\{\}_{closed}\}_{open}$.
 Currently, our type system can only track \emph{unique} and
@@
 \mylabel{T-IDREAD}\\
-\dfrac{\env(n) = t_{1} \;\;\; t_{2} = read(\penv, t_{1})}
-      {\env, \penv \vdash n_{e}:close(t_{2}), \env[n \mapsto open(t_{1})]}
+\dfrac{\env(n) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
+      {\env, \penv \vdash n:close(T_{2}), \env[n \mapsto open(T_{1})]}
 \end{array}
@@
 the type $\{``foo":\Integer\}_{closed}$ that is not a subtype of
-$\{``foo":\Integer \cup \Nil\}_{regular}$, the type of $b$:
+$\{``foo":\Integer \cup \Nil\}_{closed}$, the type of $b$:
 \begin{center}
@@
 \multicolumn{3}{l}{$\mathbf{local} \; a:\{``foo":\Integer\}_{unique} = \{ [``foo"] = 5 \} \; \mathbf{in}$}\\
-& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``foo":\Integer \cup \Nil \}_{regular} = a \; \mathbf{in}$}\\
+& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``foo":\Integer \cup \Nil \}_{closed} = a \; \mathbf{in}$}\\
 & & \multicolumn{1}{l}{$b[``foo"] = \mathbf{nil}$}
@@
 \mylabel{T-IDWRITE}\\
-\dfrac{\env(n) = t_{1} \;\;\; t_{2} = write(t_{1})}
-      {\env, \penv \vdash n_{l}:close(t_{2}), \env[n \mapsto close(t_{2})]}
+\dfrac{\env(n) = T_{1} \;\;\; T_{2} = write(T_{1})}
+      {\env, \penv \vdash n_{l}:close(T_{2}), \env[n \mapsto close(T_{2})]}
 \end{array}
@@
 discriminated union types to their original types, and
-function \emph{write} also works in this purpose.
+function \emph{write} works in this purpose.
 
@@
 \dfrac{\begin{array}{c}
-       \env_{1}(n) = t \;\;\;
-       read(\penv, t) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}\\
-       \env_{1}, \penv \vdash e_{2}:k, \env_{2} \;\;\;
-       \exists i \in 1{..}n \; k \lesssim k_{i}
+       \env_{1}(n) = T \;\;\;
+       read(\penv, T) = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}\\
+       \env_{1}, \penv \vdash e_{2}:K, \env_{2} \;\;\;
+       \exists i \in 1{..}n \; K \lesssim K_{i}
        \end{array}}
-      {\env_{1}, \penv \vdash n[e_{2}]:v_{i}, \env_{2}}
+      {\env_{1}, \penv \vdash n[e_{2}]:V_{i}, \env_{2}}
 \end{array}
@@
 \dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash e_{1}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}, \env_{2}\\
-       \env_{2}, \penv \vdash e_{2}:k, \env_{3} \;\;\;
-       \exists i \in 1{..}n \; k \lesssim k_{i}
+       \env_{1}, \penv \vdash e_{1}:\{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}, \env_{2}\\
+       \env_{2}, \penv \vdash e_{2}:K, \env_{3} \;\;\;
+       \exists i \in 1{..}n \; K \lesssim K_{i}
        \end{array}}
-      {\env_{1}, \penv \vdash e_{1}[e_{2}]:v_{i}, \env_{3}}
+      {\env_{1}, \penv \vdash e_{1}[e_{2}]:V_{i}, \env_{3}}
 \end{array}
@@
 \emph{unique} or \emph{open} table type in a context that expects a
-\emph{closed} or \emph{regular} table type with a different shape.
+\emph{fixed} or \emph{closed} table type with a different shape.
 This kind of refinement allows programmers to add optional fields
@@
 \mylabel{T-COERCE}\\
-\dfrac{\env_{1}(n) \subtype t \;\;\;
-       \env_{1}[n \mapsto t], \penv \vdash n_{e}:t_{1}, \env_{2}}
-      {\env_{1}, \penv \vdash {<}t{>} \; n:t_{1}, \env_{2}}
+\dfrac{\env_{1}(n) \subtype T \;\;\;
+       \env_{1}[n \mapsto T], \penv \vdash n:T_{1}, \env_{2}}
+      {\env_{1}, \penv \vdash {<}T{>} \; n:T_{1}, \env_{2}}
 \end{array}
@@
 & \multicolumn{2}{l}{$a[``y"] \; {<}\String{>} = ``bar";$}\\
-& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{regular} =$}\\
+& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{closed} =$}\\
 & & \multicolumn{1}{l}{${<}\{``x":\String, ``y":\String \cup \Nil\}_{open}{>} \; a \; \mathbf{in} \; a[``z"] \; {<}\Integer{>} = 1$}
@@
 which is a subtype of
-$\{``x":\String, ``y":\String \cup \Nil\}_{regular}$, the type of $b$.
+$\{``x":\String, ``y":\String \cup \Nil\}_{closed}$, the type of $b$.
 We can continue to refine the type of $a$ after aliasing it to $b$,
@@
 \mylabel{T-FUNCTION3}\\
-\dfrac{closeall(\env_{1}[\vec{n} \mapsto \vec{t}]), \penv[\ret \mapsto r] \vdash s, \env_{2}}
+\dfrac{closeall(\env_{1}[\vec{n} \mapsto \vec{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}}
       {\begin{array}{c}
-       \env_{1}, \penv \vdash \mathbf{fun} \; (\vec{n{:}t}){:}r \; s:\vec{t} \times \Void \rightarrow r,\\
-       openset(closeset(\env_{1}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s)),rv(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))
+       \env_{1}, \penv \vdash \mathbf{fun} \; (\vec{n{:}T}){:}S \; s:\vec{T} \times \Void \rightarrow S,\\
+       openset(closeset(\env_{1}, fav(\mathbf{fun} \; (\vec{n{:}T}){:}S \; s)),rv(\mathbf{fun} \; (\vec{n{:}T}){:}S \; s))
        \end{array}}
@@
 The function \emph{closeall} closes all \emph{unique} and \emph{open}
-table types.
+table types; it also restores all filter types to their original type.
 The function \emph{closeset} closes a given set of free assigned variables,
-which is given by the function \emph{fav}.
+which is given by the function \emph{fav};
+it also uses this set to restore filter types to their original type.
 The function \emph{openset} changes from \emph{unique} to \emph{open}
@@
 This rule also extends the environment $\penv$, bounding the special
-variable $\ret$ to the return type $r$.
+variable $\ret$ to the return type $S$.
 Rule \textsc{T-RETURN} uses the type that is bound to $\ret$ in
@@
 \mylabel{T-RETURN}\\
-\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
-       \penv(\ret) = r_{2} \;\;\;
-       r_{1} \lesssim r_{2}}
+\dfrac{\env_{1} \vdash el:S_{1}, \env_{2} \;\;\;
+       \penv(\ret) = S_{2} \;\;\;
+       S_{1} \lesssim S_{2}}
       {\env_{1} \vdash \mathbf{return} \; el, \env_{2}}
@@
 & & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
-& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; b[``x"] \; {<}\String{>} = ``foo" \;;\; f(a[``x"])_{s}$}
+& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; b[``x"] \; {<}\String{>} = ``foo" \;;\; f(a[``x"])$}
 \end{tabular}
@@
 \dfrac{\begin{array}{c}
-       \env_{1}(n_{1}) = t_{s} \;\;\; t_{s} = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{unique|open}\\
-       \env_{1}, \penv \vdash n_{2} : l, \env_{2} \;\;\;
-       \not \exists i \in 1..n \; l \lesssim k_{i}\\
-       closeall(\env_{1}[self \mapsto \Self, \vec{n} \mapsto \vec{t}, \self \mapsto t_{s}]),
-       \penv[\ret \mapsto r] \vdash s, \env_{3}\\
-       t_{o} = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}, l{:}\Const \; \Self \times \vec{t} \times \Void \rightarrow r\}_{unique|open}
+       \env_{1}(n_{1}) = T_{s} \;\;\; T_{s} = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique|open}\\
+       \env_{1}, \penv \vdash n_{2} : L, \env_{2} \;\;\;
+       \not \exists i \in 1..n \; L \lesssim K_{i}\\
+       closeall(\env_{1}[self \mapsto \Self, \vec{n} \mapsto \vec{T}, \self \mapsto T_{s}]),
+       \penv[\ret \mapsto S] \vdash s, \env_{3}\\
+       T_{o} = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, L{:}\Const \; \Self \times \vec{T} \times \Void \rightarrow S\}_{unique|open}
        \end{array}}
       {\begin{array}{c}
-       \env_{1}, \penv \vdash \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; s,\\
-       openset(closeset(\env_{1}[n_{1} \mapsto t_{o}], fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s)),rv(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))
+       \env_{1}, \penv \vdash \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}T}){:}S \; s,\\
+       openset(closeset(\env_{1}[n_{1} \mapsto T_{o}], fav(\mathbf{fun} \; (\vec{n{:}T}){:}S \; s)),rv(\mathbf{fun} \; (\vec{n{:}T}){:}S \; s))
        \end{array}}
@@
 \dfrac{\env_{1}, \penv \vdash e:\Self, \env_{2} \;\;\;
-       \env_{2}(\self) = t}
-      {\env_{1}, \penv \vdash e:t, \env_{2}}
+       \env_{2}(\self) = T}
+      {\env_{1}, \penv \vdash e:T, \env_{2}}
 \end{array}
@@
 \mylabel{T-SETMETATABLE2}\\
-\dfrac{\env(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed}}
-      {\env, \penv \vdash setmetatable(\{\}, \{[``\string_\string_index"] = n\}):\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{open}, \env}
+\dfrac{\env(n) = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{fixed}}
+      {\env, \penv \vdash setmetatable(\{\}, \{[``\string_\string_index"] = n\}):\{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{open}, \env}
 \end{array}
@@
 \mylabel{T-SETMETATABLE3}\\
-\dfrac{\env_{1}, \penv \vdash e = t, \env_{2} \;\;\;
-       t = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed} \;\;\;
-       \env_{1}(n) = \Self \;\;\; \env_{1}(\sigma) \subtype t}
-      {\env_{1}, \penv \vdash setmetatable(e, \{[``\string_\string_index"] = n\}):\Self, \env_{2}[\sigma \mapsto t]}
+\dfrac{\env_{1}, \penv \vdash e = T, \env_{2} \;\;\;
+       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{fixed} \;\;\;
+       \env_{1}(n) = \Self \;\;\; \env_{1}(\sigma) \subtype T}
+      {\env_{1}, \penv \vdash setmetatable(e, \{[``\string_\string_index"] = n\}):\Self, \env_{2}[\sigma \mapsto T]}
 \end{array}
@@
 \dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash e:t_{s}, \env_{2}\\
-       \env_{2}[\sigma \mapsto t_{s}], \penv \vdash e[n]:\Const \; r_{1} \rightarrow r_{2}, \env_{3}\\
-       \env_{3}[\sigma \mapsto t_{s}], \penv \vdash el:r_{3}, \env_{4} \;\;\;
-       t_{s} \times r_{3} \lesssim r_{1}
+       \env_{1}, \penv \vdash e:T_{s}, \env_{2}\\
+       \env_{2}[\sigma \mapsto T_{s}], \penv \vdash e[n]:\Const \; S_{1} \rightarrow S_{2}, \env_{3}\\
+       \env_{3}[\sigma \mapsto T_{s}], \penv \vdash el:S_{3}, \env_{4} \;\;\;
+       T_{s} \times S_{3} \lesssim S_{1}
        \end{array}}
-      {\env_{1}, \penv \vdash e{:}n(el)_{m}:[\sigma \mapsto t]r_{2}, \env_{4}}
+      {\env_{1}, \penv \vdash e{:}n(el):[\sigma \mapsto T]S_{2}, \env_{4}}
 \end{array}
@@
 \mylabel{T-STMINVOKE1}\\
-\dfrac{\env_{1}, \penv \vdash e{:}n(el)_{m}:r, \env_{2}}
-      {\env_{1}, \penv \vdash e{:}n(el)_{s},\env_{2}}
+\dfrac{\env_{1}, \penv \vdash e{:}n(el):S, \env_{2}}
+      {\env_{1}, \penv \vdash \lfloor e{:}n(el) \rfloor_{0}, \env_{2}}
 \end{array}
@@
 \mylabel{T-EXPINVOKE1}\\
-\dfrac{\env_{1}, \penv \vdash e{:}n(el)_{m}:r, \env_{2}}
-      {\env_{1}, \penv \vdash e{:}n(el)_{e}:first(r), \env_{2}}
+\dfrac{\env_{1}, \penv \vdash e{:}n(el):S, \env_{2}}
+      {\env_{1}, \penv \vdash \lfloor e{:}n(el) \rfloor_{1}:first(S), \env_{2}}
 \end{array}
@@
 \[
-o{:}move(10, 10)_{m}
+o{:}move(10, 10)
 \]
@@
 
-In Section \ref{sec:unions} we mentioned that our type system 
-includes union types to encode three common Lua idioms:
-the use of optional values, the overloading based on the tags of
-input parameters, and the overloading on the return type of functions.
-Now, we will discuss how our type system uses the $\mathbf{or}$
-operator to handle optional values, how it uses filter types to
-discriminate unions of first-level types, and how it uses
-projection types to discriminate overloaded return types,
-which are unions of second-level types.
-
-Typed Lua includes five typing rules for handling the $\mathbf{or}$
-logical operator and its common idioms:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-OR1}\\
-\dfrac{\env_{1}, \penv \vdash e_{1}:t, \env_{2} \;\;\;
-       \Nil \not\lesssim t \;\;\;
-       \False \not\lesssim t}
-      {\env_{1}, \penv \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-OR2}\\
-\dfrac{\env_{1}, \penv \vdash e_{1}:\Nil, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash e_{2}:t, \env_{3}}
-      {\env_{1}, \penv \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-OR3}\\
-\dfrac{\env_{1}, \penv \vdash e_{1}:\False, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash e_{2}:t, \env_{3}}
-      {\env_{1}, \penv \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-OR4}\\
-\dfrac{\env_{1}, \penv \vdash e_{1}:\Nil \cup \False, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash e_{2}:t, \env_{3}}
-      {\env_{1}, \penv \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-OR5}\\
-\dfrac{\env_{1}, \penv \vdash e_{1}:t_{1}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3}}
-      {\env_{1}, \penv \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(filter(t_{1}, \Nil), \False) \cup t_{2}, \env_{3}}
-\end{array}
-\end{array}
-\]
-
-Rule \textsc{T-OR1} is the rule that defines the short circuit.
-We use the consistent-subtyping relation in this rule to guarantee that
-the type system checks the second expression when the first one
-has the dynamic type, as it can be hiding a false value.
-
-Rules \textsc{T-OR2}, \textsc{T-OR3}, and \textsc{T-OR4} guarantee
-that the final result is the type of the second expression, because the
-first one is certainly a false value.
-
-Rule \textsc{T-OR5} is the most general rule, but it is also
-the rule that handles the common $\mathbf{or}$ idioms.
-It uses the auxiliary function \emph{filter} to filter possible false
-values that might be part of the type of the first expression.
+A common idiom that programmers use in Lua is to use the function
+\texttt{type} to execute different actions according to the type of
+a parameter, and our type system uses filter types to handle this idiom.
+
+As an example, assuming that \emph{rep} is in the environment and has type
+$\String \times \Integer \times \String \cup \Nil \times \Void \rightarrow \String \times \Void$,
+our type system type checks the following program:
+\begin{center}
+\begin{tabular}{llll}
+\multicolumn{4}{l}{$\mathbf{local} \; f:\String \times \String \cup \Integer \times \Void \rightarrow \String \times \Void =$}\\
+& \multicolumn{3}{l}{$\mathbf{fun} \; (a:\String, b:\String \cup \Integer):\String \times \Void$}\\
+& & \multicolumn{2}{l}{$\mathbf{local} \; r:\String = ``" \; \mathbf{in}$}\\
+& & & \multicolumn{1}{l}{$\mathbf{if} \; type(b) == ``string" \; \mathbf{then} \; r = a \;{..}\;b \; \mathbf{else} \; r = rep(a, b)\; ;$}\\
+& & & \multicolumn{1}{l}{$\mathbf{return} \; r$}\\
+\multicolumn{4}{l}{$\mathbf{in} \; \lfloor f(``foo", 2) \rfloor_{0}$}
+\end{tabular}
+\end{center}
+
+This example type checks through rule \textsc{T-IF2}, that is,
+this rule handles the case where our type system discriminates
+an union of first-level types based on the tag \texttt{string}:
+\[
+\begin{array}{c}
+\mylabel{T-IF2}\\
+\dfrac{\begin{array}{c}
+       \env_{1}(n) = T_{1} \cup T_{2}\\
+       closeall(\env_{1})[n \mapsto \phi(T,\String)], \penv \vdash s_{1}, \env_{2} \\
+       closeall(\env_{1})[n \mapsto \phi(T,filter(T, \String)), \penv \vdash s_{2}, \env_{3}\\
+       \env_{4} = openset(closeset(\env_{1}[n \mapsto T], fav(s_{1}) \cup fav(s_{2})),rv(s_{1}) \cup rv(s_{2}))
+      \end{array}}
+      {\env_{1}, \penv \vdash \mathbf{if} \; type(n) == ``string" \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
+\end{array}
+\]
+
+In Typed Lua, \emph{type} is a primitive that triggers a type change.
+In the rule \textsc{T-IF2}, the type of a variable $n$ changes from
+$T_{1} \cup T_{2}$ to $\phi(T_{1} \cup T_{2},\String)$ inside the
+$\mathbf{if}$ branch, and it changes the type of a variable $n$ from
+$T_{1} \cup T_{2}$ to $\phi(T,filter(T_{1} \cup T_{2},\String))$
+inside the $\mathbf{else}$ branch.
+
 We can use pattern matching to the define the recursive function
@@
 \begin{align*}
-filter(t_{1} \cup t_{2}, t_{1}) & = filter(t_{2}, t_{1})\\
-filter(t_{1} \cup t_{2}, t_{2}) & = filter(t_{1}, t_{2})\\
-filter(t_{1} \cup t_{2}, t_{3}) & = filter(t_{1}, t_{3}) \cup filter(t_{2}, t_{3})\\
-filter(t_{1}, t_{2}) & = t_{1}
+filter(T_{1} \cup T_{2}, T_{1}) & = filter(T_{2}, T_{1})\\
+filter(T_{1} \cup T_{2}, T_{2}) & = filter(T_{1}, T_{2})\\
+filter(T_{1} \cup T_{2}, T_{3}) & = filter(T_{1}, T_{3}) \cup filter(T_{2}, T_{3})\\
+filter(T_{1}, T_{2}) & = T_{1}
 \end{align*}
-
-Rule \textsc{T-OR5} allows our type system to type check the following example:
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; x:\String \cup \Nil = \mathbf{nothing} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$\mathbf{local} \; y:\String = x \; \mathbf{or} \; ``Hello"$}
-\end{tabular}
-\end{center}
-
-Without the \emph{filter} function,
-the expression $x \; \mathbf{or} \; ``Hello"$ would have type
-$\String \cup \Nil \cup ``Hello"$.
-The \emph{filter} function removes the type $\Nil$ from the result,
-leaving the type $\String \cup ``Hello"$.
-This means that type checking the expression $x \; \mathbf{or} \; ``Hello"$
-results in the type $\String$, because union types are disjoint and
-$``Hello" \lesssim \String$.
-
-Another common idiom that programmers use in Lua is to overload
-the type of function parameters, and use the function \texttt{type}
-to execute different actions according to their types.
-The rule \textsc{T-IF2} shows the case where our type system
-discriminates an union of first-level types based on the tag \texttt{string}:
-\[
-\begin{array}{c}
-\mylabel{T-IF2}\\
-\dfrac{\begin{array}{c}
-       \env_{1}(n) = t_{1} \cup t_{2}\\
-       closeall(\env_{1}[n \mapsto \phi(t,\String)]), \penv \vdash s_{1}, \env_{2} \\
-       closeall(\env_{1}[n \mapsto \phi(t,filter(t, \String))), \penv \vdash s_{2}, \env_{3}\\
-       \env_{4} = openset(closeset(\env_{1}[n \mapsto t], fav(s_{1}) \cup fav(s_{2})),rv(s_{1}) \cup rv(s_{2}))
-      \end{array}}
-      {\env_{1}, \penv \vdash \mathbf{if} \; type(n) == ``string" \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
-
-\end{array}
-\]
-
-In Typed Lua, \emph{type} is a primitive that triggers a type change.
-In the rule \textsc{T-IF2}, the type of a variable $n$ changes from
-$t_{1} \cup t_{2}$ to $\phi(t_{1} \cup t_{2},\String)$ inside the
-$\mathbf{if}$ branch, and it changes the type of a variable $n$ from
-$t_{1} \cup t_{2}$ to $\phi(t,filter(t_{1} \cup t_{2},\String))$
-inside the $\mathbf{else}$ branch.
 
@@
 \begin{align*}
-read(\penv, \phi(t_{1},t_{2})) & = t_{2}\\
-read(\penv, \pi_{i}^{x}) & = proj(\penv, x, i)\\
-read(\penv, t) & = t
+read(\penv, \phi(T_{1},T_{2})) & = T_{2}\\
+read(\penv, \pi_{i}^{x}) & = proj(\penv(x), i)\\
+read(\penv, T) & = T
 \end{align*}
@@
 \begin{align*}
-write(\phi(t_{1},t_{2})) & = t_{1}\\
-write(t) & = t
+write(\phi(T_{1},T_{2})) & = T_{1}\\
+write(T) & = T
 \end{align*}
 
-As an example, assuming that \emph{rep} is in the environment and has type
-$\String \times \Integer \times \String \cup \Nil \times \Void \rightarrow \String \times \Void$,
-the following code type checks through rule \textsc{T-IF2}:
-\begin{center}
-\begin{tabular}{llll}
-\multicolumn{4}{l}{$\mathbf{local} \; o:\String \times \String \cup \Integer \times \Void \rightarrow \String \times \Void =$}\\
-& \multicolumn{3}{l}{$\mathbf{fun} \; (a:\String, b:\String \cup \Integer):\String \times \Void$}\\
-& & \multicolumn{2}{l}{$\mathbf{local} \; r:\String = ``" \; \mathbf{in}$}\\
-& & & \multicolumn{1}{l}{$\mathbf{if} \; type(b) == ``string" \; \mathbf{then} \; r = a \;{..}\;b \; \mathbf{else} \; r = rep(a, b)_{e}\; ;$}\\
-& & & \multicolumn{1}{l}{$\mathbf{return} \; r$}\\
-\multicolumn{4}{l}{$\mathbf{in} \; o(``foo", 2)_{s}$}
-\end{tabular}
-\end{center}
-
-In this example, the type of $b$ changes to
-$\phi(\String \cup \Integer, \String)$ inside the $\mathbf{if}$ branch, and
-it changes to $\phi(\String \cup \Integer, \Integer)$ inside the $\mathbf{else}$ branch.
+In our previous example, the type of $b$ is
+$\phi(\String \cup \Integer, \String)$ inside the $\mathbf{if}$ branch,
+and it is $\phi(\String \cup \Integer, \Integer)$ inside the $\mathbf{else}$ branch.
 Thus, reading $b$ results in the type $\String$ inside the $\mathbf{if}$ branch,
@@
 \begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; q:\pi_{1}^{x}, r:\pi_{2}^{x} = idiv(1, 2)_{m}^{x} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; print(q + r)_{s} \; \mathbf{else} \; print(``ERROR: " \; .. \; r)_{s}$}
+\multicolumn{2}{l}{$\mathbf{local} \; q, r = idiv(1, 2) \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; \lfloor print(q + r) \rfloor_{0} \; \mathbf{else} \; \lfloor print(``ERROR: " \; .. \; r) \rfloor_{0}$}
 \end{tabular}
@@
 First, that example uses rule \textsc{T-EXPLIST6} for type checking
-$idiv(1, 2)_{m}^{x}$:
+$idiv(1, 2)$:
 \[
@@
 \dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash me^{x}:r, \env_{2}\\
-       r = t_{1} \times ... \times t_{n} \times \Void \sqcup t_{1}' \times ... \times t_{n}' \times \Void
+       \env_{1}, \penv \vdash me:S, \env_{2}\\
+       S = T_{1} \times ... \times T_{n} \times \Void \sqcup T_{1}' \times ... \times T_{n}' \times \Void
        \end{array}}
-      {\env_{1}, \penv \vdash me^{x}:\pi_{1}^{x} \times ... \times \pi_{n}^{x} \times \Nil{*}, \env_{2}, (x,r)}
+      {\env_{1}, \penv \vdash me:\pi_{1}^{x} \times ... \times \pi_{n}^{x} \times \Nil{*}, \env_{2}, (x,S)}
 \end{array}
@@
 Rule \textsc{T-EXPLIST6} uses the auxiliary relation
-$\env_{1}, \penv \vdash el : r_{1}, \env_{2}, (x,r_{2})$.
+$\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$.
 This relation means that given a type environment $\env_{1}$ and
 a projection environment $\penv$, an expression list $el$ has type
-$r_{1}$ and produces a new type environment $\env_{2}$ and
-produces a pair $(x,r_{2})$.
+$S_{1}$ and produces a new type environment $\env_{2}$ and
+produces a pair $(x,S_{2})$.
 This pair means that the last expression of an expression list $el$
-produces a second-level type $r_{2}$ that should be bound to
+produces a second-level type $S_{2}$ that should be bound to
 a variable $x$ in the projection environment,
@@
 
-Our type system uses this pair to allow rule \textsc{T-LOCAL} to bound
-a second-level type $r_{2}$ to a variable $x$ in $\penv$,
-after rule \textsc{T-LOCAL} checks whether the type of an expression
-list $el$ matches the types in the annotations:
+Our type system uses that pair to allow rule \textsc{T-LOCAL} to bound
+a second-level type $S_{2}$ to a variable $x$ in $\penv$:
 \[
@@
 \mylabel{T-LOCAL}\\
-\dfrac{\env_{1}, \penv \vdash el:r_{1}, \env_{2}, (x,r_{2}) \;\;\;
-       r_{1} \lesssim \vec{t} \times \Value{*} \;\;\;
-       \env_{2}[\vec{n} \mapsto \vec{t}], \penv[x \mapsto r_{2}] \vdash s, \env_{3}}
-      {\env_{1}, \penv \vdash \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s, \env_{3} - \{\vec{n} \mapsto \vec{t}\}}
+\dfrac{\begin{array}{c}
+       \env_{1}, \penv \vdash el:T_{1} \times ... \times T_{n}{*}, \env_{2}, (x,S)\\
+       \env_{2}[n_{1} \mapsto T_{1}, ..., n_{n-1} \mapsto T_{n-1}], \penv[x \mapsto S] \vdash s, \env_{3}
+       \end{array}}
+      {\env_{1}, \penv \vdash \mathbf{local} \; \vec{n} = el \; \mathbf{in} \; s, \env_{3} - \{\vec{n} \mapsto \vec{T}\}}
 \end{array}
@@
 \dfrac{\begin{array}{c}
-       \env_{1}(n) = \pi_{i}^{x} \;\;\; \penv(x) = r\\
-       closeall(\env_{1}), \penv[x \mapsto fpt(r,\Nil,i)] \vdash s_{1}, \env_{2} \\
-       closeall(\env_{1}), \penv[x \mapsto gpt(r,\Nil,i)] \vdash s_{2}, \env_{3} \\
+       \env_{1}(n) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
+       closeall(\env_{1}), \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
+       closeall(\env_{1}), \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
        \env_{4} = openset(closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2})),rv(s_{1}) \cup rv(s_{2}))
@@
 \dfrac{\env_{1}, \penv \vdash e_{1}:\Any, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash e_{2}:t, \env_{3}}
+       \env_{2}, \penv \vdash e_{2}:T, \env_{3}}
       {\env_{1}, \penv \vdash e_{1} + e_{2}:\Any, \env_{3}}
@@
 \mylabel{T-ARITH6}\\
-\dfrac{\env_{1}, \penv \vdash e_{1}:t, \env_{2} \;\;\;
+\dfrac{\env_{1}, \penv \vdash e_{1}:T, \env_{2} \;\;\;
        \env_{2}, \penv \vdash e_{2}:\Any, \env_{3}}

### 2015-03-08 doc/thesis/system.tex
Resizing margins and removing several typing rules from the text

@@
 are difficult to handle along with our simplified form of
-\emph{flow typing}, and they are out of the scope for now;
+\emph{flow typing}, and they are out of scope for now;
 \item explicit blocks are not present because we are already using
@@
 form because we can use this form to express the missing forms; 
-\item arithmetic operators are reduced to addition and unary minus
+\item arithmetic operators are reduced to addition only
 because the other operators have similar typing rules;
@@
 because the other operators have similar typing rules;
-\item bitwise operators are not present because they are similar
-to integer arithmetic operators.
+\item bitwise operators are reduced to bitwise and only
+because the other operators have similar typing rules.
 \end{itemize}
@@
 s_{1} \; ; \; s_{2} \; | \;
-\vec{l} = el \; | \;
+\overline{l} = el \; | \;
 m \; | \;
@@
 \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
-& | \; \mathbf{local} \; \vec{n{:}T} = el \; \mathbf{in} \; s \; | \;
-\mathbf{local} \; \vec{n} = el \; \mathbf{in} \; s \; | \;
-\mathbf{rec} \; n{:}T = f \; \mathbf{in} \; s\\
+& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s \; | \;
+\mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s \; | \;
+\mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s\\
 & | \; \mathbf{return} \; el \; | \;
@@
 k \; | \;
-\lfloor {...} \rfloor_{1} \; | \;
-n \; | \;
+id \; | \;
 e_{1}[e_{2}] \; | \;
-{<}T{>} \; n \; | \;
+{<}T{>} \; id \; | \;
 f \; | \;
-\{ \; \vec{p} \; \} \; | \;
-\{ \; \vec{p},me \; \}\\
-& | \; e_{1} + e_{2} \; | \;
-e_{1} \; {..} \; e_{2} \; | \;
-e_{1} == e_{2} \; | \;
+\{ \; \overline{p} \; \} \; | \;
+\{ \; \overline{p},me \; \} \; | \;
+e_{1} + e_{2} \; | \;
+e_{1} \; {..} \; e_{2} \\
+& | \; e_{1} == e_{2} \; | \;
 e_{1} < e_{2} \; | \;
+e_{1} \;\&\; e_{2} \; | \;
 e_{1} \; \mathbf{and} \; e_{2} \; | \;
-e_{1} \; \mathbf{or} \; e_{2} \\
-& | \; \mathbf{not} \; e \; | \;
-- e \; | \;
+e_{1} \; \mathbf{or} \; e_{2} \; | \;
+\mathbf{not} \; e \; | \;
 \# \; e \; | \;
-\lfloor a \rfloor_{1}\\
-l ::= & \;\; n_{l} \; | \;
+\lfloor me \rfloor_{1}\\
+l ::= & \;\; id_{l} \; | \;
 e_{1}[e_{2}] \; | \;
-n[k] \; {<}T{>}\\
+id[k] \; {<}T{>}\\
 k ::= & \;\; \mathbf{false} \; | \;
@@
 p ::= & \;\; [e_{1}] = e_{2}\\
-el ::= & \;\; \vec{e} \; | \;
-\vec{e}, me\\
+el ::= & \;\; \overline{e} \; | \;
+\overline{e}, me\\
 me ::= & \;\; a \; | \;
@@
 e{:}n(el)\\
-m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}T}){:}S \; fb \; | \;
-\mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}T},{...}{:}T){:}S \; fb\\
-f ::= & \;\; \mathbf{fun} \; (\vec{n{:}T}){:}S \; fb \; | \;
-\mathbf{fun} \; (\vec{n{:}t},{...}{:}T){:}S \; fb\\
+m ::= & \;\; \mathbf{fun} \; id_{1}{:}id_{2} \; (\overline{id{:}T}){:}S \; fb \; | \;
+\mathbf{fun} \; id_{1}{:}id_{2} \; (\overline{id{:}T},{...}{:}T){:}S \; fb\\
+f ::= & \;\; \mathbf{fun} \; (\overline{id{:}T}){:}S \; fb \; | \;
+\mathbf{fun} \; (\overline{id{:}T},{...}{:}T){:}S \; fb\\
 fb ::= & \;\; s \;;\; \mathbf{return} \; el\\
-n ::= & \;\; {\it name}\\
+id ::= & \;\; {\it name}\\
 \end{array}
@@
 $m$ are method declarations, $a$ are function and method applications,
-$n$ are variable names, $T$ are first-level types, and $S$ are second-level types.
-The notation $\vec{n{:}T}$ denotes the list $n_{1}{:}T_{1}, ..., n_{n}{:}T_{n}$,
+$id$ are variable names, $T$ are first-level types, and $S$ are second-level types.
+The notation $\overline{id{:}T}$ denotes the list $id_{1}{:}T_{1}, ..., id_{n}{:}T_{n}$,
 where $n$ can be zero.
 
-Our reduced core includes two statements for declaring local variables.
-One with explicit type annotations, and another without type annotations.
+Our reduced core includes two statements for declaring local variables,
+one with and another without type annotations.
 While we use the former to show how our type system handles the declaration
-of annoatted variables, we use the latter to show how our type system
+of annoatated variables, we use the latter to show how our type system
 handles the declaration of unannotated variables through local type inference
@@
 that produce no value because they appear as statements.
-We use $\lfloor a \rfloor_{1}$ to denote function and method applications
-that produce only one value, even if the function or method application
-returns mutiple values.
-We use $\lfloor {...} \rfloor_{1}$ to denote a vararg expression that
-produces only one value, as it can appear in the middle of an
-expression list.
+We use $\lfloor me \rfloor_{1}$ to denote function applications,
+method applications, and vararg expressions that produce only one value,
+even if they return mutiple values.
 
 We also include two kinds of type coercions in our core language:
-the left-hand value $n[k] \; {<}T{>}$ and the expression ${<}T{>} \;n$.
+the left-hand value $id[k] \; {<}T{>}$ and the expression ${<}T{>} \;id$.
 Both allow the refinement of table types.
@@
 of tables in the presence of refinement.
-We use $n$ when they appear as expressions and $n_{l}$ when they
-appear as left-hand values.
+We use $id$ when variable names appear as expressions and $id_{l}$ when
+variable names appear as left-hand values.
 
@@
 The first relation means that given a type environment $\env_{1}$
-and a projection environment $\penv$, typing a statement $s$ produces
-a new type environment $\env_{2}$.
+and a projection environment $\penv$, we can check a statement $s$
+producing a new type environment $\env_{2}$.
 The second relation means that given a type environment $\env_{1}$
-and a projection environment $\penv$, typing an expression $e$ has
+and a projection environment $\penv$, we can check that an expression $e$ has
 type $T$ and produces a new type environment $\env_{2}$.
-We need two typing relations because only expressions have types,
-though expressions and statements can modify a type environment.
 
 Even though we can assign only first-level types to variables,
-our type system should be able to project unions of first-level types
-from unions of second-level types.
-Thus, we need two environments because $\env$ maps variables to first-level types,
-while $\penv$ maps variables to second-level types.
+functions and methods can return unions of second-level types,
+and our type system should be able to project these unions of
+second-level types into unions of first-level types.
+To do that, we use projection types and the environment $\penv$ that
+maps projection variables to second-level types, while the
+environment $\env$ maps variables to first-level types.
 We will show how our type system uses the environment $\penv$ for
-handling projection types.
-We use $\env_{1}[n \mapsto T]$ to extend the environment $\env_{1}$
-with the variable $n$ that maps to type $T$.
+handling projection types in Section \ref{sec:fap}.
+We use $\env_{1}[id \mapsto T]$ to extend the environment $\env_{1}$
+with the variable $id$ that maps to type $T$.
 We use $\penv[x \mapsto S]$ to extend the environment $\penv$ with
-the variable $x$ that maps to the second-level type $S$.
+the projection variable $x$ that maps to the second-level type $S$.
 
@@
 
-Lua has multiple assignment, and the rule \textsc{T-ASSIGNMENT} uses
-tuple types along with consistent-subtyping to type check this Lua feature:
-\[
+Lua has multiple assignment, and our type system uses three
+different kinds of typing rules to type check this feature.
+It uses typing rules that type check the different forms
+of expression lists that can appear in the right-hand side,
+a typing rule that type checks a list of left-hand values,
+and a general rule that uses consistent-subtyping to check
+whether the type of the right-hand side is consistent with
+the type of the left-hand side.
+
+As an example, lets assume that $x$ and $y$ are variables in the
+environment with types $\Integer$ and $\String$, the assignment
+\[
+x, y = 1, ``foo"
+\]
+type checks through the following rules:
+\[
+\begin{array}{c}
 \begin{array}{c}
 \mylabel{T-ASSIGNMENT}\\
-\begin{array}{c}
 \dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash \vec{l}:S_{2}, \env_{3} \;\;\;
+       \env_{2}, \penv \vdash \overline{l}:S_{2}, \env_{3} \;\;\;
        S_{1} \lesssim S_{2}}
-      {\env_{1}, \penv \vdash \vec{l} = el,\env_{3}}
-\end{array}
-\end{array}
-\]
-
-As an example,
-\[
-x, y = 1, ``foo"
-\]
-type checks because
-$1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times \Value{*}$,
-assuming that $x$ and $y$ are in the environment with
-types $\Integer$ and $\String$, respectively.
-
-Our type system uses the type $\Value{*}$ to drop extra values
-and the type $\Nil{*}$ to replace missing values.
-The rule that type checks left-hand values list introduces the type
-$\Value{*}$ to let the right-hand side produce more values than
-expected in the left-hand side.
-The rules that type check expression list introduce the type
-$\Nil{*}$ to let the right-hand side produce fewer values than
-expected in the left-hand side.
-Our type system includes $\Nil{*}$ in the type of an expression list only
-if its type does not end in another variadic type $t{*}$.
-In the previous example, the rules \textsc{T-LHSLIST} and \textsc{T-EXPLIST2}
-are the rules that introduce these types.
-We define the rule \textsc{T-LHSLIST} as follows:
-\[
+      {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-EXPLIST1}\\
+\dfrac{\env, \penv \vdash e_{i}:T_{i}, \env_{i} \;\;\;
+       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
+       n = |\;\overline{e}\;|}
+      {\env, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
+\end{array}
+\\ \\
 \begin{array}{c}
@@
        \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
-       n = |\;\vec{l}\;|}
-      {\env, \penv \vdash \vec{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
-\end{array}
-\]
-
-First, this rule uses the same environment $\env$ to type check each
-left-hand side value $l_{i}$.
-Then, it takes each type $T_{i}$ and builds the tuple type for
-the assignment rule, placing $\Value{*}$ in the end.
-Table refinement can change the type environment while typing each left-hand value.
-Thus, we use a partial auxiliary function \emph{merge} to collect
-all the changes in a new environment $\env_{m}$, if there are no conflicts.
-
-The rule \textsc{T-EXPLIST2} is similar to \textsc{T-LHSLIST}:
-\[
+       n = |\;\overline{l}\;|}
+      {\env, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
+\end{array}
+\end{array}
+\]
+
+The rule \textsc{T-ASSIGNMENT} is the general rule,
+\textsc{T-EXPLIST1} is the rule that type checks an expression list
+where all expressions can only produce a single value, and
+\textsc{T-LHSLIST} is the rule that type checks a list of left-hand values.
+
+Note that our type system uses the type $\Value{*}$ to drop extra values
+and the type $\Nil{*}$ to replace missing values.
+The rule that type checks lists of left-hand values introduces the type
+$\Value{*}$ to let the right-hand side produce more values than
+expected in the left-hand side.
+The rules that type check expression lists introduce the type
+$\Nil{*}$ to let the right-hand side produce fewer values than
+expected in the left-hand side.
+
+In Section \ref{sec:refinement} we will show that table refinement can
+change the type environment while typing an expression or a left-hand value.
+Thus, the rules that type check lists of expressions and lists of
+left-hand values use a partial auxiliary function \emph{merge} to collect
+all environment changes in a new environment $\env_{m}$, if there are no conflicts.
+
+The previous example type checks through these rules because
+\[
+1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times \Value{*}
+\]
+
+As another example, lets assume that $x$, $y$, and $z$ are variables in
+the environment with types $\Integer$, $\String$, and $\String \cup \Nil$,
+the assignment
+\[
+x, y, z = 1, ``foo"
+\]
+type checks because
+\[
+\Integer \times \String \times \Nil{*} \lesssim \Integer \times \String \times \String \cup \Nil \times \Value{*}
+\]
+
+Conversely, the assignment
+\[
+x = 1, ``foo"
+\]
+type checks because
+\[
+\Integer \times \String \times \Nil{*} \lesssim \Integer \times \Value{*}
+\]
+
+Our type system includes $\Nil{*}$ in the type of an expression list only
+if its type does not end in another variadic type $T{*}$.
+For instance, rules \textsc{T-EXPLIST2} and \textsc{T-EXPLIST3} handle
+the case where an expression list is just an expression that may produce
+multiple values.
+The former rule includes the type $\Nil{*}$, but the latter rule does not:
+\[
+\begin{array}{c}
 \begin{array}{c}
 \mylabel{T-EXPLIST2}\\
-\dfrac{\env, \penv \vdash e_{i}:T_{i}, \env_{i} \;\;\;
-       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
-       n = |\;\vec{e}\;|}
-      {\env, \penv \vdash \vec{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
-\end{array}
-\]
-
-We use \textsc{T-EXPLIST2} only when the last expression in the list
-can produce a single value.
-Rules \textsc{T-EXPLIST3} and \textsc{T-EXPLIST4} handle the case where
-an expression list is just an expression that may produce multiple values:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-EXPLIST3}\\
 \dfrac{\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Void, \env_{2}}
@@
 \begin{array}{c}
-\mylabel{T-EXPLIST4}\\
+\mylabel{T-EXPLIST3}\\
 \dfrac{\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n}{*}, \env_{2}}
@@
 \end{array}
-\]
-
-As an example, these rules allow our type system to type check the following example:
-\[
-x, y, z = f()
-\]
-
-In this example, we are assuming that $x$, $y$, $z$, and $f$ are in
-the environment with types $\Integer$, $\String$, $\String \cup \Nil$, and
-either $\Void \rightarrow \Integer \times \String \times \Void$ or
-$\Value{*} \rightarrow \Integer \times \String \times \Nil{*}$,
-depending on whether the programmer is using strict function types.
-Thus, the example type checks because
-\[
-\Integer \times \String \times \Nil{*} \lesssim \Integer \times \String \times \String \cup \Nil \times \Value{*}
-\]
-
-Conversely,
-\[
-x = f()
-\]
-type checks because
-\[
-\Integer \times \String \times \Nil{*} \lesssim \Integer \times \Value{*}
 \]
@@
 Our abstract syntax reduces the syntactic forms of the table constructor
-into two forms: $\{\;\vec{p}\;\}$ and $\{\;\vec{p},me\;\}$.
+into two forms: $\{\;\overline{p}\;\}$ and $\{\;\overline{p},me\;\}$.
 The first uses a list of table fields $p_{1}, ..., p_{n}$,
@@
 
-The rule \textsc{T-CONSTRUCTOR2} type checks a table type with a non-empty
+The rule \textsc{T-CONSTRUCTOR2} type checks a table constructor with a non-empty
 list of table fields:
@@
 This auxiliary relation means that given a type environment $\env_{1}$
-and a projection environment $\penv$, typing a table field $p$
-produces the pair $(K,V)$ and a new type environment $\env_{2}$. 
+and a projection environment $\penv$, we can check a table field $p$
+producing a pair $(K,V)$ and a new type environment $\env_{2}$. 
 In the pair $(K,V)$, $K$ is the key type, while $V$ is the value type.
@@
 Then, the rule \textsc{T-CONSTRUCTOR2} uses each pair $(K_{i},V_{i})$
-to build the table type that express the type of the given constructor, and
+to build the table type that express the type of a given constructor, and
 uses the predicate \emph{wf} to check whether this table type is well-formed.
@@
 
-The definition of the auxiliary relation that type checks table
-fields would be straightforward if our table types allowed any
-first-level type in the type of the keys.
-The rules \textsc{T-FIELD1}, \textsc{T-FIELD2}, and \textsc{T-FIELD3}
-check if the type of the key is a subtype of $\Boolean$, $\Number$,
-or $\String$, respectively:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-FIELD1}\\
-\dfrac{\env_{1}, \penv \vdash e_{1}:T_{1}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash e_{2}:T_{2}, \env_{3} \;\;\;
-       T_{1} \subtype \Boolean}
-      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (T_{1},close(T_{2})), \env_{3}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-FIELD2}\\
-\dfrac{\env_{1}, \penv \vdash e_{1}:T_{1}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash e_{2}:T_{2}, \env_{3} \;\;\;
-       T_{1} \subtype \Number}
-      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (T_{1},close(T_{2})), \env_{3}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-FIELD3}\\
-\dfrac{\env_{1}, \penv \vdash e_{1}:T_{1}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash e_{2}:T_{2}, \env_{3} \;\;\;
-       T_{1} \subtype \String}
-      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (T_{1},close(T_{2})), \env_{3}}
-\end{array}
-\end{array}
-\]
-
-These rules use the function \emph{close} in the type of the value
-because our type system allows only \emph{closed} table types to
-appear in the type of the values.
-The function \emph{close} promotes all table types to \emph{closed}.
-We made this restriction because our type system does not keep track
-of aliases to table fields.
-This means that allowing \emph{unique}, \emph{open}, and \emph{fixed}
-table types to appear in the type of the values would allow the creation
-of unsafe aliases.
-
-The rule \textsc{T-FIELD4} is a fallback for the previous rules:
-\[
-\begin{array}{c}
-\mylabel{T-FIELD4}\\
-\dfrac{\env_{1}, \penv \vdash e_{1}:T_{1}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash e_{2}:T_{2}, \env_{3}}
-      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (\Value,close(T_{2})), \env_{3}}
-\end{array}
-\]
-
-The rule \textsc{T-FIELD4} shows that our type system uses the type $\Value$
-as the type of the key whenever a key has a type that is neither a
-subtype of $\Boolean$, nor a subtype of $\Number$, nor a subtype of $\String$.
-
-As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
-has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ through the rules
-\textsc{T-FIELD3} and \textsc{T-CONSTRUCTOR2}.
-
-Typed Lua handles arrays as hashes that map integers to some type $T$.
-In Lua, programmers often use the vararg expression to initialize arrays.
-The rules \textsc{T-CONSTRUCTOR3} and \textsc{T-CONSTRUCTOR4} define
-the behavior of the vararg expression for this case:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-CONSTRUCTOR3}\\
-\dfrac{\env({...}) = T}
-      {\env, \penv \vdash \{{...}\}:\{\Integer{:}T \cup \Nil\}_{unique}, \env}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-CONSTRUCTOR4}\\
-\dfrac{\begin{array}{c}
-       \env_{i}, \penv \vdash p_{i}:(K_{i},V_{i}), \env_{i+1} \;\;\;
-       n = |\;\vec{p}\;| \;\;\;
-       \env_{i+1}({...}) = T_{v}\\
-       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, \Integer{:}T_{v} \cup \Nil\}_{unique} \;\;\;
-       wf(T)
-       \end{array}}
-      {\env_{1}, \penv \vdash \{\;\vec{p},\;{...}\;\}:T, \env_{n+1}}
-\end{array}
-\end{array}
-\]
-
-The rule \textsc{T-CONSTRUCTOR3} type checks the case where we use
-only the vararg expression inside a table constructor to initialize
-an array.
-If we assume that $...$ is in the environment and has type $\String$,
-the table constructor $\{{...}\}$ has type $\{\Integer:\String \cup \Nil\}_{unique}$
-through the rule \textsc{T-CONSTRUCTOR3}.
-
-The rule \textsc{T-CONSTRUCTOR4} type checks the case where we use
-the table constructor as a record that includes an array part.
-If we assume that $...$ is in the environment and has type $\String$,
-the table constructor $\{[``x"] = ``foo", [``y"] = \mathbf{true}, {...}\}$
-has type $\{``x":``foo", ``y":\mathbf{true}, \Integer:\String \cup \Nil\}_{unique}$
-through the rule \textsc{T-CONSTRUCTOR4}.
-As another example, the rule \textsc{T-CONSTRUCTOR4} does not type check
-the table constructor $\{ [1] = ``foo", [10] = \mathbf{true}, {...} \}$,
-because the keys $1$ and $10$ overlap the array part,
-and the resulting table type is not well-formed.
+Well-formed table types also do not allow \emph{unique} and
+\emph{open} table types to appear in the type of the values.
+We made this restriction because our type system does not keep
+track of aliases to table fields.
+This means that allowing \emph{unique} and \emph{open} table
+types to appear in the type of the value would allow the
+creation of unsafe aliases.
 
@@
 to an \emph{open} table in the top-level scope.
-For instance,
+In the following examples we assume that $\string_ENV$ is in the
+environment and has type $\{\}_{open}$. For instance,
 \[
@@
 uses field assignment to add fields $``x"$ and $``y"$ to $\string_ENV$.
-In this example and in the next examples, we assume that
-$\string_ENV$ is in the environment and has type $\{\}_{open}$.
 Therefore, after these field assignments $\string_ENV$ has type
@@
 
-Aliasing an \emph{unique} or an \emph{open} table type produces a
-\emph{closed} table type, as defined by rule \textsc{T-IDREAD}:
-\[
-\begin{array}{c}
-\mylabel{T-IDREAD}\\
+Aliasing an \emph{unique} or an \emph{open} table type can produce
+either a \emph{closed} or a \emph{fixed} table type, depending on
+the context that we are using a variable.
+Rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this behavior:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-IDREAD1}\\
 \dfrac{\env(n) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
@@
 \end{array}
-\]
-
-In Section \ref{sec:fap} we will explain that \textsc{T-IDREAD} uses
-the auxiliary function \emph{read} because it may be accessing an
-identifier that is bound to a filter or projection type.
+\;
+\begin{array}{c}
+\mylabel{T-IDREAD2}\\
+\dfrac{\env(n) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
+      {\env, \penv \vdash n:fix(T_{2}), \env[n \mapsto fix(T_{1})]}
+\end{array}
+\end{array}
+\]
+
+Both rules use the auxiliary function \emph{read} because they may be
+accessing an identifier that is bound to a filter or projection type.
 As we mentioned in Section \ref{sec:unions}, our type system includes
 a small set of \texttt{type} predicates that allow programmers to
-discriminate union types.
-In Section \ref{sec:fap} we will also explain that our type system
-uses filter and projection types in the definition of these predicates
-to handle the discrimination of unions of first-level and second-level
-types.
+discriminate union types, and our type system uses filter and projection
+types in the definition of these predicates to handle the discrimination
+of unions of first-level and second-level types.
+We can define \emph{read} as follows:
+\begin{align*}
+read(\penv, \phi(T_{1},T_{2})) & = T_{2}\\
+read(\penv, \pi_{i}^{x}) & = proj(\penv(x), i)\\
+read(\penv, T) & = T
+\end{align*}
+
+The function \emph{read} uses the auxiliary function \emph{proj}
+to project a union of first-level types, based on an union of
+second-level types and an index from a projection type.
+In Section \ref{sec:fap} we will discuss how our type system uses
+projection types to handle overloaded return types.
 
@@
 reference \emph{unique} can be unsafe.
-For this reason, the rule \textsc{T-IDREAD} changes the type of
+For this reason, the rule \textsc{T-IDREAD1} changes the type of
 the original reference from \emph{unique} to \emph{open}.
@@
 
-In Section \ref{sec:fap} we will explain that \textsc{T-IDWRITE} uses
-the auxiliary function \emph{write} because it may be accessing an
-identifier that is bound to a filter or projection type.
+This rule uses the auxiliary function \emph{write} because it may be
+accessing an identifier that is bound to a filter or projection type.
 As we mentioned in Section \ref{sec:unions}, assignments restore
@@
 function \emph{write} works in this purpose.
+We can define \emph{write} as follows:
+\begin{align*}
+write(\phi(T_{1},T_{2})) & = T_{1}\\
+write(T) & = T
+\end{align*}
 
@@
 \mylabel{T-FUNCTION3}\\
-\dfrac{closeall(\env_{1}[\vec{n} \mapsto \vec{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}}
+\dfrac{crall(\env_{1}[\vec{n} \mapsto \vec{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}}
       {\begin{array}{c}
        \env_{1}, \penv \vdash \mathbf{fun} \; (\vec{n{:}T}){:}S \; s:\vec{T} \times \Void \rightarrow S,\\
-       openset(closeset(\env_{1}, fav(\mathbf{fun} \; (\vec{n{:}T}){:}S \; s)),rv(\mathbf{fun} \; (\vec{n{:}T}){:}S \; s))
+       openset(crset(\env_{1}, fav(\mathbf{fun} \; (\vec{n{:}T}){:}S \; s)),rv(\mathbf{fun} \; (\vec{n{:}T}){:}S \; s))
        \end{array}}
@@
 type checking a nested scope.
-The function \emph{closeall} closes all \emph{unique} and \emph{open}
-table types; it also restores all filter types to their original type.
-The function \emph{closeset} closes a given set of free assigned variables,
-which is given by the function \emph{fav};
+The function \emph{crall} closes all \emph{unique} and \emph{open}
+table types; it also restores filter types to their original type.
+The function \emph{crset} closes a given set of free assigned variables,
+which is given by the function \emph{fav}, and
 it also uses this set to restore filter types to their original type.
@@
 $a$ and $b$ have the same type: $\{\}_{closed}$.
-Their type was closed by \emph{closeall} before type checking
+Their type was closed by \emph{crall} before type checking
 the function body.
@@
 First, the function \emph{fav} includes $b$ in the set of variables
-that should be closed by \emph{closeset}.
+that should be closed by \emph{crset}.
 Then, the function \emph{rv} includes $a$ in the set of variables
@@
 
-\subsection{Classes and objects}
-\label{sec:cao}
-
-Typed Lua also relies on the refinement of table types to type check
-the idioms that Lua programmers use for incrementally building
-classes and objects.
-For instance, rule \textsc{T-METHOD3} illustrates how our type system
-type checks method definitions:
-\[
-\begin{array}{c}
-\mylabel{T-METHOD3}\\
-\dfrac{\begin{array}{c}
-       \env_{1}(n_{1}) = T_{s} \;\;\; T_{s} = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique|open}\\
-       \env_{1}, \penv \vdash n_{2} : L, \env_{2} \;\;\;
-       \not \exists i \in 1..n \; L \lesssim K_{i}\\
-       closeall(\env_{1}[self \mapsto \Self, \vec{n} \mapsto \vec{T}, \self \mapsto T_{s}]),
-       \penv[\ret \mapsto S] \vdash s, \env_{3}\\
-       T_{o} = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, L{:}\Const \; \Self \times \vec{T} \times \Void \rightarrow S\}_{unique|open}
-       \end{array}}
-      {\begin{array}{c}
-       \env_{1}, \penv \vdash \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}T}){:}S \; s,\\
-       openset(closeset(\env_{1}[n_{1} \mapsto T_{o}], fav(\mathbf{fun} \; (\vec{n{:}T}){:}S \; s)),rv(\mathbf{fun} \; (\vec{n{:}T}){:}S \; s))
-       \end{array}}
-\end{array}
-\]
-
-Rule \textsc{T-METHOD3} allows our type system to type check the
-following example, which is a translation of the definition of the class
-\texttt{Shape} from Section \ref{sec:oop}:
-\begin{center}
-\begin{tabular}{llll}
-\multicolumn{4}{l}{$\mathbf{local} \; Shape{:}\{``x"{:}\Number, ``y"{:}\Number\}_{unique} = \{ [``x"] = 0.0, [``y"] = 0.0 \}$}\\
-\multicolumn{4}{l}{$\mathbf{in}$}\\
-& \multicolumn{3}{l}{$\mathbf{fun} \; Shape{:}new (x:\Number, y:\Number):\Self \times \Void$}\\
-& & \multicolumn{2}{l}{$\mathbf{local} \; o:\Self = setmetatable(\{\}, \{ [``\string_\string_index"] = self \})$}\\
-& & \multicolumn{2}{l}{$\mathbf{in} \; o[``x"] = x; \; o[``y"] = y; \; \mathbf{return} \; o$}\\
-; & \multicolumn{3}{l}{$\mathbf{fun} \; Shape{:}move (x:\Number, y:\Number):\Void$}\\
-& & \multicolumn{2}{l}{$self[``x"] = self[``x"] + x;$}\\
-& & \multicolumn{2}{l}{$self[``y"] = self[``y"] + y;$}\\
-& & \multicolumn{2}{l}{$\mathbf{return} \; \mathbf{nothing}$}
-\end{tabular}
-\end{center}
-
-This example uses the type $\{``x":\Number, ``y":\Number\}_{unique}$ to
-initialize the local variable $Shape$, which represents the class
-that we are defining.
-Then, it uses two method definitions to include $new$ and $move$ in our class.
-These definitions refine the type of $Shape$, and it becomes the type below:
-\begin{align*}
-\{ & ``x":\Number, ``y":\Number,\\
-   & \Const \; ``new":\Self \times \Number \times \Number \times \Void \rightarrow \Self \times \Void,\\
-   & \Const \; ``move":\Self \times \Number \times \Number \times \Void \rightarrow \Void \}_{unique}
-\end{align*}
-
-Inside the definition of the methods \emph{new} and \emph{move},
-we index variables that have type $\Self$.
-This is possible because we include the rule \textsc{T-SELF}:
-\[
-\begin{array}{c}
-\mylabel{T-SELF}\\
-\dfrac{\env_{1}, \penv \vdash e:\Self, \env_{2} \;\;\;
-       \env_{2}(\self) = T}
-      {\env_{1}, \penv \vdash e:T, \env_{2}}
-\end{array}
-\]
-
-When an expression has type $\Self$, our type system uses this
-rule to produce the type that is bound to $\self$ in $\env$.
-The variable $\self$ is a special variable that method definitions
-and method calls use to bind the type of $\Self$ in the environment.
-
-Inside the definition of the method \emph{new}, we use \emph{setmetatable}
-to initialize $o$ with type $\Self$.
-The rule \textsc{T-SETMETATABLE1} express this idea:
-\[
-\begin{array}{c}
-\mylabel{T-SETMETATABLE1}\\
-\dfrac{\env(n) = \Self}
-      {\env, \penv \vdash setmetatable(\{\}, \{[``\string_\string_index"] = n\}):\Self, \env}
-\end{array}
-\]
-
-In Section \ref{sec:oop} we also mentioned that our type system
-uses \emph{setmetatable} to handle single inheritance.
-Rule \textsc{T-SETMETATABLE2} allows refining the type of a new
-class based on the type of an already defined class,
-while rule \textsc{T-SETMETATABLE3} allows overriding a constructor:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-SETMETATABLE2}\\
-\dfrac{\env(n) = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{fixed}}
-      {\env, \penv \vdash setmetatable(\{\}, \{[``\string_\string_index"] = n\}):\{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{open}, \env}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-SETMETATABLE3}\\
-\dfrac{\env_{1}, \penv \vdash e = T, \env_{2} \;\;\;
-       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{fixed} \;\;\;
-       \env_{1}(n) = \Self \;\;\; \env_{1}(\sigma) \subtype T}
-      {\env_{1}, \penv \vdash setmetatable(e, \{[``\string_\string_index"] = n\}):\Self, \env_{2}[\sigma \mapsto T]}
-\end{array}
-\end{array}
-\]
-
-After we define our class, we can use it to create instances
-of this class and call their methods.
-Rule \textsc{T-INVOKE1} handles the case where method calls
-are expressions that produce multiple values:
-\[
-\begin{array}{c}
-\mylabel{T-INVOKE1}\\
-\dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash e:T_{s}, \env_{2}\\
-       \env_{2}[\sigma \mapsto T_{s}], \penv \vdash e[n]:\Const \; S_{1} \rightarrow S_{2}, \env_{3}\\
-       \env_{3}[\sigma \mapsto T_{s}], \penv \vdash el:S_{3}, \env_{4} \;\;\;
-       T_{s} \times S_{3} \lesssim S_{1}
-       \end{array}}
-      {\env_{1}, \penv \vdash e{:}n(el):[\sigma \mapsto T]S_{2}, \env_{4}}
-\end{array}
-\]
-
-We also use the rule \textsc{T-INVOKE1} as the base case for the rules
-that handle the cases where method calls are either statements
-or expressions that produce only one value:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-STMINVOKE1}\\
-\dfrac{\env_{1}, \penv \vdash e{:}n(el):S, \env_{2}}
-      {\env_{1}, \penv \vdash \lfloor e{:}n(el) \rfloor_{0}, \env_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{T-EXPINVOKE1}\\
-\dfrac{\env_{1}, \penv \vdash e{:}n(el):S, \env_{2}}
-      {\env_{1}, \penv \vdash \lfloor e{:}n(el) \rfloor_{1}:first(S), \env_{2}}
-\end{array}
-\end{array}
-\]
-
-The rule \textsc{T-STMINVOKE1} discards the produced values,
-while the rule \textsc{T-EXPINVOKE1} uses the predicate \emph{first} to
-ensure that only one value is produced.
-
-As an example, assuming that the object $o$ is in the environment and
-has the type of the class $Shape$, the method call
-\[
-o{:}move(10, 10)
-\]
-type checks through the rule \textsc{T-INVOKE1}, because
-\[
-Shape \times 10 \times 10 \times \Nil{*} \lesssim Shape \times \Number \times \Number \times \Void
-\]
-assuming that $Shape$ is an alias to the type of the class $Shape$.
-Note that rule \textsc{T-INVOKE1} uses the object type to bind
-the type $\Self$ before type checking a method call.
-
-\subsection{Filters and projections}
+\subsection{Projections}
 \label{sec:fap}
 
-A common idiom that programmers use in Lua is to use the function
-\texttt{type} to execute different actions according to the type of
-a parameter, and our type system uses filter types to handle this idiom.
-
-As an example, assuming that \emph{rep} is in the environment and has type
-$\String \times \Integer \times \String \cup \Nil \times \Void \rightarrow \String \times \Void$,
-our type system type checks the following program:
-\begin{center}
-\begin{tabular}{llll}
-\multicolumn{4}{l}{$\mathbf{local} \; f:\String \times \String \cup \Integer \times \Void \rightarrow \String \times \Void =$}\\
-& \multicolumn{3}{l}{$\mathbf{fun} \; (a:\String, b:\String \cup \Integer):\String \times \Void$}\\
-& & \multicolumn{2}{l}{$\mathbf{local} \; r:\String = ``" \; \mathbf{in}$}\\
-& & & \multicolumn{1}{l}{$\mathbf{if} \; type(b) == ``string" \; \mathbf{then} \; r = a \;{..}\;b \; \mathbf{else} \; r = rep(a, b)\; ;$}\\
-& & & \multicolumn{1}{l}{$\mathbf{return} \; r$}\\
-\multicolumn{4}{l}{$\mathbf{in} \; \lfloor f(``foo", 2) \rfloor_{0}$}
-\end{tabular}
-\end{center}
-
-This example type checks through rule \textsc{T-IF2}, that is,
-this rule handles the case where our type system discriminates
-an union of first-level types based on the tag \texttt{string}:
-\[
-\begin{array}{c}
-\mylabel{T-IF2}\\
-\dfrac{\begin{array}{c}
-       \env_{1}(n) = T_{1} \cup T_{2}\\
-       closeall(\env_{1})[n \mapsto \phi(T,\String)], \penv \vdash s_{1}, \env_{2} \\
-       closeall(\env_{1})[n \mapsto \phi(T,filter(T, \String)), \penv \vdash s_{2}, \env_{3}\\
-       \env_{4} = openset(closeset(\env_{1}[n \mapsto T], fav(s_{1}) \cup fav(s_{2})),rv(s_{1}) \cup rv(s_{2}))
-      \end{array}}
-      {\env_{1}, \penv \vdash \mathbf{if} \; type(n) == ``string" \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
-\end{array}
-\]
-
-In Typed Lua, \emph{type} is a primitive that triggers a type change.
-In the rule \textsc{T-IF2}, the type of a variable $n$ changes from
-$T_{1} \cup T_{2}$ to $\phi(T_{1} \cup T_{2},\String)$ inside the
-$\mathbf{if}$ branch, and it changes the type of a variable $n$ from
-$T_{1} \cup T_{2}$ to $\phi(T,filter(T_{1} \cup T_{2},\String))$
-inside the $\mathbf{else}$ branch.
-
-We can use pattern matching to the define the recursive function
-\emph{filter} as follows:
-\begin{align*}
-filter(T_{1} \cup T_{2}, T_{1}) & = filter(T_{2}, T_{1})\\
-filter(T_{1} \cup T_{2}, T_{2}) & = filter(T_{1}, T_{2})\\
-filter(T_{1} \cup T_{2}, T_{3}) & = filter(T_{1}, T_{3}) \cup filter(T_{2}, T_{3})\\
-filter(T_{1}, T_{2}) & = T_{1}
-\end{align*}
-
-Rule \textsc{T-IF2} changes the type of a local variable $n$ to a
-filter type because it gives the discriminated type when we use $n$ as
-an expression, and it restores the original type when we use $n$
-in an assignment.
-In Section \ref{sec:refinement} we mentioned that rule \textsc{T-IDREAD}
-uses the function \emph{read} to produce the discriminated type,
-while rule \textsc{T-IDWRITE} uses the function \emph{write} to
-restore the original type.
-Now, we will give the definition of these functions.
-
-We can define \emph{read} as follows:
-\begin{align*}
-read(\penv, \phi(T_{1},T_{2})) & = T_{2}\\
-read(\penv, \pi_{i}^{x}) & = proj(\penv(x), i)\\
-read(\penv, T) & = T
-\end{align*}
-
-The function \emph{read} uses the auxiliary function \emph{proj}
-to project a union of first-level types, based on an union of
-second-level types and an index from a projection type.
-In this section, we will discuss how our type system uses
-projection types to handle overloaded return types.
-
-We can define \emph{write} as follows:
-\begin{align*}
-write(\phi(T_{1},T_{2})) & = T_{1}\\
-write(T) & = T
-\end{align*}
-
-In our previous example, the type of $b$ is
-$\phi(\String \cup \Integer, \String)$ inside the $\mathbf{if}$ branch,
-and it is $\phi(\String \cup \Integer, \Integer)$ inside the $\mathbf{else}$ branch.
-Thus, reading $b$ results in the type $\String$ inside the $\mathbf{if}$ branch,
-and it results in the type $\Integer$ inside the $\mathbf{else}$ branch.
-Outside the condition, $b$ has its original type: $\String \cup \Integer$.
-
-Typed Lua also includes similar rules to handle the tags \texttt{nil},
-\texttt{boolean}, and \texttt{number}.
-There is also a similar rule for handling the tag \texttt{integer}, but
-it is limited to Lua 5.3, as it depends on the function \texttt{math.type}.
-This function appears only in Lua 5.3 and it returns the tag \texttt{integer}
-when its input parameter is a number that has an integer representation,
-the tag \texttt{float} when its input parameter is a number that has a
-floating point representation, or \texttt{nil} otherwise.
-
-Lua programmers also overload the return type of functions to denote errors,
+Lua programmers often overload the return type of functions to denote errors,
 and our type system uses projection types to handle this idiom.
 
-As an example, assuming that \emph{idiv} and \emph{print} are in the
-environment with the respective types
-$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void$
-and
-$\Value{*} \rightarrow \Void$,
-the following code type checks in our type system:
+As an example, lets assume that \emph{idiv} and \emph{print} are functions
+in the environment.
+The function \emph{idiv} has type
+$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void$,
+while the function \emph{print} thas type
+$\Value{*} \rightarrow \Void$.
+Our type system uses three typing rules to type check the following program:
 \begin{center}
@@
 
-There are three typing rules involved in the type checking of this example,
-which we will explain now.
-
-First, that example uses rule \textsc{T-EXPLIST6} for type checking
+First, our type system uses rule \textsc{T-EXPLIST6} for type checking
 $idiv(1, 2)$:
@@
 This relation means that given a type environment $\env_{1}$ and
-a projection environment $\penv$, an expression list $el$ has type
-$S_{1}$ and produces a new type environment $\env_{2}$ and
-produces a pair $(x,S_{2})$.
+a projection environment $\penv$, we can check that an expression
+list $el$ has type $S_{1}$ and produces a new type environment $\env_{2}$
+and produces a pair $(x,S_{2})$.
 This pair means that the last expression of an expression list $el$
-produces a second-level type $S_{2}$ that should be bound to
-a variable $x$ in the projection environment,
+produces a union of second-level types $S_{2}$ that sould be bound
+to a variable $x$ in the projection environment,
 as the resulting type of this expression is a tuple of projection
 types $\pi_{i}^{x}$.
-
-Our type system uses that pair to allow rule \textsc{T-LOCAL} to bound
-a second-level type $S_{2}$ to a variable $x$ in $\penv$:
+In our example, we can check that $idiv(1, 2)$ has type
+$\pi_{1}^{x} \times \pi_{2}^{x} \times \Nil{*}$ and produces the pair
+$(x,\Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void)$.
+
+Our type system uses the pair $(x,S_{2})$ to allow rule
+\textsc{T-LOCAL} to bound a union of second-level types $S_{2}$
+to a variable $x$ in $\penv$:
 \[
@@
 \dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash el:T_{1} \times ... \times T_{n}{*}, \env_{2}, (x,S)\\
-       \env_{2}[n_{1} \mapsto T_{1}, ..., n_{n-1} \mapsto T_{n-1}], \penv[x \mapsto S] \vdash s, \env_{3}
+       \env_{1}, \penv \vdash el:S_{1}, \env_{2}, (x,S_{2})\\
+       \env_{2}[id_{1} \mapsto infer(S_{1},1), ..., id_{n} \mapsto infer(S_{1},n)], \penv[x \mapsto S_{2}] \vdash s, \env_{3} \;\;\;
+       n = |\;\overline{id}\;|  
        \end{array}}
-      {\env_{1}, \penv \vdash \mathbf{local} \; \vec{n} = el \; \mathbf{in} \; s, \env_{3} - \{\vec{n} \mapsto \vec{T}\}}
-\end{array}
-\]
-
+      {\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, \env_{3} - \{\overline{id}\}}
+\end{array}
+\]
+
+This rule uses the auxiliary function \emph{infer} to get the
+first-level types of each variable that should be introduced in
+the type environment for type checking $s$.
+In our example, $q$ gets type $\pi_{1}^{x}$ and $r$ gets type
+$\pi_{2}^{x}$.
 After type checking the statement $s$, rule \textsc{T-LOCAL} produces a
@@
 they project to $\Integer \cup \Nil$ and $\Integer \cup \String$, respectively.
-
-\subsection{Arithmetic operations}
-\label{sec:arith}
-
-Lua has operator overloading, and allows programmers to redefine
-the behavior of some operations.
-For instance, programmers can use metatables to redefine the
-behavior of arithmetic operations.
-Even though Typed Lua does not support operator overloading yet,
-it includes typing rules that allow programmers to use the
-dynamic type when they are using overloaded operations.
-The following typing rules show how Typed Lua uses the dynamic type
-to handle the overloading of arithmetic operations:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-ARITH5}\\
-\dfrac{\env_{1}, \penv \vdash e_{1}:\Any, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash e_{2}:T, \env_{3}}
-      {\env_{1}, \penv \vdash e_{1} + e_{2}:\Any, \env_{3}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-ARITH6}\\
-\dfrac{\env_{1}, \penv \vdash e_{1}:T, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash e_{2}:\Any, \env_{3}}
-      {\env_{1}, \penv \vdash e_{1} + e_{2}:\Any, \env_{3}}
-\end{array}
-\end{array}
-\]
-
-The rule \textsc{T-ARITH5} allows type checking the following
-example:
-\begin{center}
-\begin{tabular}{l}
-$\mathbf{local} \; x{:}\Any = 1 \; \mathbf{in} \; x = x + 1$
-\end{tabular}
-\end{center}
-
-This example is safe, but the following it is not:
-\begin{center}
-\begin{tabular}{l}
-$\mathbf{local} \; x{:}\Integer, \;y{:}\Any = 1 \; \mathbf{in} \; x = x + y$
-\end{tabular}
-\end{center}
-
-Although this last example is not safe, it shows that optional
-type systems still preserve the flexibility of dynamically
-typed languages along with the benefits of static type checking.

### 2015-03-09 doc/thesis/system.tex
Fixing typing rules in the appendix

@@
 \mylabel{T-EXPLIST1}\\
-\dfrac{\env, \penv \vdash e_{i}:T_{i}, \env_{i} \;\;\;
-       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
-       n = |\;\overline{e}\;|}
-      {\env, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
+\dfrac{\env_{i}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
+       n = |\;\overline{e}\;| \;\;\;
+       \env_{m} = merge(\env_{1}, ..., \env_{n+1})}
+      {\env_{1}, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
 \end{array}
@@
 \dfrac{\env_{i}, \penv \vdash p_{i}:(K_{i},V_{i}), \env_{i+1} \;\;\;
-       n = |\;\vec{p}\;| \;\;\;
+       n = |\;\overline{p}\;| \;\;\;
        T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
        wf(T)}
-      {\env_{1}, \penv \vdash \{\;\vec{p}\;\}:T, \env_{n+1}}
+      {\env_{1}, \penv \vdash \{\;\overline{p}\;\}:T, \env_{n+1}}
 \end{array}
@@
 \dfrac{\begin{array}{c}
-       \env_{1}(n) = \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n} \}_{open|unique}\\
+       \env_{1}(id) = \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n} \}_{open|unique}\\
        \env_{1}, \penv \vdash k:K, \env_{2} \;\;\;
@@
        \end{array}}
-      {\env_{1}, \penv \vdash n[k] {<}T{>}:V, \env_{2}[n \mapsto \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, K{:}V\}_{open|unique}]}
+      {\env_{1}, \penv \vdash id[k] {<}T{>}:V, \env_{2}[id \mapsto \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, K{:}V\}_{open|unique}]}
 \end{array}
@@
 \mylabel{T-IDREAD1}\\
-\dfrac{\env(n) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
-      {\env, \penv \vdash n:close(T_{2}), \env[n \mapsto open(T_{1})]}
+\dfrac{\env(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
+      {\env, \penv \vdash id:close(T_{2}), \env[id \mapsto open(T_{1})]}
 \end{array}
@@
 \mylabel{T-IDREAD2}\\
-\dfrac{\env(n) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
-      {\env, \penv \vdash n:fix(T_{2}), \env[n \mapsto fix(T_{1})]}
+\dfrac{\env(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
+      {\env, \penv \vdash id:fix(T_{2}), \env[id \mapsto fix(T_{1})]}
 \end{array}
@@
 \mylabel{T-IDWRITE}\\
-\dfrac{\env(n) = T_{1} \;\;\; T_{2} = write(T_{1})}
-      {\env, \penv \vdash n_{l}:close(T_{2}), \env[n \mapsto close(T_{2})]}
+\dfrac{\env(id) = T_{1} \;\;\; T_{2} = write(T_{1})}
+      {\env, \penv \vdash id_{l}:close(T_{2}), \env[id \mapsto close(T_{2})]}
 \end{array}
@@
 \dfrac{\begin{array}{c}
-       \env_{1}(n) = T \;\;\;
+       \env_{1}(id) = T \;\;\;
        read(\penv, T) = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}\\
@@
        \end{array}}
-      {\env_{1}, \penv \vdash n[e_{2}]:V_{i}, \env_{2}}
+      {\env_{1}, \penv \vdash id[e_{2}]:V_{i}, \env_{2}}
 \end{array}
@@
 \mylabel{T-COERCE}\\
-\dfrac{\env_{1}(n) \subtype T \;\;\;
-       \env_{1}[n \mapsto T], \penv \vdash n:T_{1}, \env_{2}}
-      {\env_{1}, \penv \vdash {<}T{>} \; n:T_{1}, \env_{2}}
+\dfrac{\env_{1}(id) \subtype T \;\;\;
+       \env_{1}[id \mapsto T], \penv \vdash id:T_{1}, \env_{2}}
+      {\env_{1}, \penv \vdash {<}T{>} \; id:T_{1}, \env_{2}}
 \end{array}
@@
 table types before we type check a nested scope.
-The rule \textsc{T-FUNCTION3} illustrates this case:
-\[
-\begin{array}{c}
-\mylabel{T-FUNCTION3}\\
-\dfrac{crall(\env_{1}[\vec{n} \mapsto \vec{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}}
-      {\begin{array}{c}
-       \env_{1}, \penv \vdash \mathbf{fun} \; (\vec{n{:}T}){:}S \; s:\vec{T} \times \Void \rightarrow S,\\
-       openset(crset(\env_{1}, fav(\mathbf{fun} \; (\vec{n{:}T}){:}S \; s)),rv(\mathbf{fun} \; (\vec{n{:}T}){:}S \; s))
+The rule \textsc{T-FUNCTION1} illustrates this case:
+\[
+\begin{array}{c}
+\mylabel{T-FUNCTION1}\\
+\dfrac{\begin{array}{c}
+       crall(\env_{1}[\overline{id} \mapsto \overline{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}\\
+       \env_{3} = openset(crset(\env_{1}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)),rv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
        \end{array}}
-\end{array}
-\]
-
-Rule \textsc{T-FUNCTION3} uses some auxiliary functions to change
+      {\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{3}}
+\end{array}
+\]
+
+Rule \textsc{T-FUNCTION1} uses some auxiliary functions to change
 the type of variables before type checking a nested scope and
@@
 Our type system uses the pair $(x,S_{2})$ to allow rule
-\textsc{T-LOCAL} to bound a union of second-level types $S_{2}$
+\textsc{T-LOCAL2} to bound a union of second-level types $S_{2}$
 to a variable $x$ in $\penv$:
@@
 \begin{array}{c}
-\mylabel{T-LOCAL}\\
+\mylabel{T-LOCAL2}\\
 \dfrac{\begin{array}{c}
@@
 \dfrac{\begin{array}{c}
-       \env_{1}(n) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
+       \env_{1}(id) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
        closeall(\env_{1}), \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
@@
       \end{array}}
-      {\env_{1}, \penv \vdash \mathbf{if} \; n \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
+      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
 \end{array}

### 2015-03-09 doc/thesis/system.tex
More improvements to the text about typing rules

@@
 \item relational operators are reduced to equality and less than
-because the other operators have similar typing rules;
+because inequality has similar typing rules to equality and
+other relational operators have similar typing rules to less than;
 \item bitwise operators are reduced to bitwise and only
@@
 one with and another without type annotations.
-While we use the former to show how our type system handles the declaration
-of annoatated variables, we use the latter to show how our type system
+While we use the former to formalize how our type system handles the declaration
+of annotated variables, we use the latter to formalize how our type system
 handles the declaration of unannotated variables through local type inference
@@
 We use $\lfloor a \rfloor_{0}$ to denote function and method applications
-that produce no value because they appear as statements.
+that produce no value, because they appear as statements.
 We use $\lfloor me \rfloor_{1}$ to denote function applications,
@@
 \dfrac{\env_{i}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
-       n = |\;\overline{e}\;| \;\;\;
-       \env_{m} = merge(\env_{1}, ..., \env_{n+1})}
+       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
+       n = |\;\overline{e}\;|}
       {\env_{1}, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
@@
 \mylabel{T-LHSLIST}\\
-\dfrac{\env, \penv \vdash l_{i}:T_{i}, \env_{i} \;\;\;
-       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
+\dfrac{\env_{i}, \penv \vdash l_{i}:T_{i}, \env_{i+1} \;\;\;
+       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
        n = |\;\overline{l}\;|}
-      {\env, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
+      {\env_{1}, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
 \end{array}
@@
 
-The previous example type checks through these rules because
+Our previous example type checks through these rules because
 \[
@@
 \[
-\Integer \times \String \times \Nil{*} \lesssim \Integer \times \String \times \String \cup \Nil \times \Value{*}
-\]
+1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times \String \cup \Nil \times \Value{*}
+\]
+
+Note how $\Nil{*}$ replaces any missing values.
+This example type checks because $\Nil{*}$ produces as many $\Nil$
+values as we need, and $\Nil$ is consistent with $\String \cup \Nil$,
+which is the type of $z$.
 
@@
 \[
-\Integer \times \String \times \Nil{*} \lesssim \Integer \times \Value{*}
-\]
+1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \Value{*}
+\]
+
+Note how $\Value{*}$ discards extra values.
+This example type checks because $\Value{*}$ discards as many extra
+values as we need, and $``foo"$ is consistent with $\Value$.
 
@@
 
-Assuming that $f$ is a local function in the environment, and that $f$ has type
+As an example, assuming that $f$ is a local function in the environment, and that $f$ has type
 $\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*} \rightarrow \Integer{*}$,
@@
 
-When our type system type checks an expression list,
+We just mentioned that when our type system type checks an expression list,
 it always includes $\Nil{*}$ in the end of the type of this expression list
@@
 This means that allowing \emph{unique} and \emph{open} table
-types to appear in the type of the value would allow the
+types to appear in the type of a value would allow the
 creation of unsafe aliases.
-
-After discussing the typing rules of the table constructor,
+As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
+has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$.
+
+After presenting some typing rules of the table constructor,
 we start the discussion of the rules that define the most
@@
 
-We translate this example to our reduced core as follows:
+We can translate this example to our reduced core as follows:
 \begin{center}
 \begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; person:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
+\multicolumn{2}{l}{$\mathbf{local} \; person = \{\} \; \mathbf{in}$}\\
 & \multicolumn{1}{l}{$person[``firstname"] \; {<}\String{>} = ``Lou";$}\\
@@
 In the following examples we assume that $\string_ENV$ is in the
-environment and has type $\{\}_{open}$. For instance,
+environment and has type $\{\}_{open}$.
+
+As an example,
 \[
@@
 the context that we are using a variable.
-Rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this behavior:
+As we mentioned in Sections \ref{sec:oop} and \ref{sec:subtyping},
+we need \emph{fixed} table types to type classes in object-oriented programming.
+In the implementation we fix the aliasing of \emph{unique} and \emph{open}
+table types that appear in a return statement, and in other cases we
+close the aliasing of \emph{unique} and \emph{open} table types.
+However, in the formalization we chose to define this behavior not
+deterministically, as it makes easier the presentation of this behavior.
+Rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this non-deterministic behavior:
 \[
@@
 types in the definition of these predicates to handle the discrimination
-of unions of first-level and second-level types.
+of unions types.
+While filter types discriminate unions of first-level types, projection
+types discriminate unions of second-level types and project unions of
+second-level types into unions of first-level types.
 We can define \emph{read} as follows:
@@
 
-The following example does not type check, as aliasing $a$ produces
-the type $\{``foo":\Integer\}_{closed}$ that is not a subtype of
-$\{``foo":\Integer \cup \Nil\}_{closed}$, the type of $b$:
+As an example,
 \begin{center}
@@
 \end{center}
+does not type check,
+as aliasing $a$ produces the type $\{``foo":\Integer\}_{closed}$
+that is not a subtype of $\{``foo":\Integer \cup \Nil\}_{closed}$,
+the type of $b$.
+Our type system has this behavior to warn programmers about
+potential unsafe behaviors after this kind of alias.
+In this example, it is unsafe to assign $\mathbf{nil}$ to the field
+$`foo"$ of $b$, as it changes the value that is stored in the field
+$``foo"$ of $a$.
 
@@
 that is not a subtype of $\{\}_{open}$, the type of $b$.
-Our type system has this behavior to warn programmers about
-potential unsafe behaviors after this kind of alias.
 In this example, it is unsafe to add the field $``x"$ to $b$,
@@
 
-As an example, rule \textsc{T-FUNCTION3} prevents the following unsafe example to type check:
+As an example, rule \textsc{T-FUNCTION1} prevents the following unsafe example to type check:
 \begin{center}
@@
 \dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash me:S, \env_{2}\\
-       S = T_{1} \times ... \times T_{n} \times \Void \sqcup T_{1}' \times ... \times T_{n}' \times \Void
+       \env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Void \sqcup T_{1}' \times ... \times T_{n}' \times \Void, \env_{2}\\
+       S = T_{1} \times ... \times T_{n} \times \Nil{*} \sqcup T_{1}' \times ... \times T_{n}' \times \Nil{*}
        \end{array}}
@@
 This pair means that the last expression of an expression list $el$
-produces a union of second-level types $S_{2}$ that sould be bound
-to a variable $x$ in the projection environment,
+produces an union of second-level types $S_{2}$ that sould be bound
+to a variable $x$ in the projection environment $\penv$,
 as the resulting type of this expression is a tuple of projection
@@
 $\pi_{1}^{x} \times \pi_{2}^{x} \times \Nil{*}$ and produces the pair
-$(x,\Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void)$.
+$(x,\Integer \times \Integer \times \Nil{*} \sqcup \Nil \times \String \times \Nil{*})$.
 
@@
 \textsc{T-LOCAL2} to bound a union of second-level types $S_{2}$
-to a variable $x$ in $\penv$:
+to a variable $x$ in the projection environment $\penv$:
 \[
@@
 the type environment for type checking $s$.
-In our example, $q$ gets type $\pi_{1}^{x}$ and $r$ gets type
-$\pi_{2}^{x}$.
-After type checking the statement $s$, rule \textsc{T-LOCAL} produces a
-new environment without the variables that it introduced before type checking $s$.
-
-Introducing a variable $x$ in the projection environment allows our
+After type checking the statement $s$, rule \textsc{T-LOCAL2} produces a
+new type environment $\env_{3}$ without the variables that it introduced
+before type checking $s$.
+In our example, we can check that declaring $q$ and $r$
+bounds the projection type $\pi_{1}^{x}$ to $q$ and
+bounds the projection type $\pi_{2}^{x}$ to $r$, where the projection variable $x$ bounds to 
+\[
+\Integer \times \Integer \times \Nil{*} \sqcup \Nil \times \String \times \Nil{*}
+\]
+in the projection environment $\penv$. 
+This means that $q$ uses $\pi_{1}^{x}$ to project its type into the
+union type $\Integer \cup \Nil$, while $r$ uses $\pi_{2}^{x}$ to
+project its type into the union $\Integer \cup \String$.
+
+Introducing a projection variable $x$ in the projection environment allows our
 type system to discriminate projection types $\pi_{i}^{x}$.
@@
 For instance, our previous example type checks through rule \textsc{T-IF3},
-because it makes the rule \textsc{T-IF3} use the function call
-\[
-fpt(\Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void, \Nil, 1)
+because it uses the information provided by the projection type $\pi_{1}^{x}$,
+which is the type of $q$, to make the rule \textsc{T-IF3} use the function call
+\[
+fpt(\Integer \times \Integer \times \Nil{*} \sqcup \Nil \times \String \times \Nil{*}, \Nil, 1)
 \]
 to discriminate the projection $x$ to the single tuple
-$\Integer \times \Integer \times \Void$ inside the $\mathbf{if}$ branch,
+$\Integer \times \Integer \times \Nil{*}$ inside the $\mathbf{if}$ branch,
 and the function call
 \[
-gpt(\Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void, \Nil, 1)
+gpt(\Integer \times \Integer \times \Nil{*} \sqcup \Nil \times \String \times \Nil{*}, \Nil, 1)
 \]
 to discriminate the projection $x$ to the single tuple
-$\Nil \times \String \times \Void$ inside the $\mathbf{else}$ branch.
+$\Nil \times \String \times \Nil{*}$ inside the $\mathbf{else}$ branch.
 Thus, reading $q$ and $r$ projects $\pi_{1}^{x}$ to $\Integer$ and

### 2015-03-11 doc/thesis/system.tex
Even more improvements to the text about typing rules

@@
 In this section we use a reduced core of Typed Lua to present the
-most interesting rules of our type system.
+typing rules that type check multiple assignment, table refinement,
+and overloading on the return type of functions.
+We limit this section to the rules that type check these features
+because these rules are the most interesting rules of our type system.
 Appendix \ref{app:rules} presents the full set of typing rules.
@@
 method applications, and vararg expressions that produce only one value,
-even if they return mutiple values.
+even if they return multiple values.
 
@@
 all environment changes in a new environment $\env_{m}$, if there are no conflicts.
-
-Our previous example type checks through these rules because
+In Section \ref{sec:refinement} we will also show that we can only change the
+environment to add new table fields in a table type, and we cannot change the
+type of a variable or a table field which is already present in a table type.
+
+Back to our previous example, it type checks through these three rules
+that we just mentioned because
 \[
@@
 \mylabel{T-CONSTRUCTOR1}\\
-\env, \penv \vdash \{\}:\{\}_{unique}, \env
+\env_{1}, \penv \vdash \{\}:\{\}_{unique}, \env_{1}
 \end{array}
@@
 \dfrac{\env_{i}, \penv \vdash p_{i}:(K_{i},V_{i}), \env_{i+1} \;\;\;
-       n = |\;\overline{p}\;| \;\;\;
        T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
-       wf(T)}
+       wf(T) \;\;\;
+       n = |\;\overline{p}\;|}
       {\env_{1}, \penv \vdash \{\;\overline{p}\;\}:T, \env_{n+1}}
@@
 \]
+
+As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
+has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$.
 
@@
 producing a pair $(K,V)$ and a new type environment $\env_{2}$. 
-In the pair $(K,V)$, $K$ is the key type, while $V$ is the value type.
+In a pair $(K,V)$, $K$ is the key type, while $V$ is the value type.
 
@@
 creation of unsafe aliases.
-As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
-has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$.
 
@@
 $``foo" \times 1 \times \Nil{*} \lesssim \String \times \Integer \times \Value{*}$.
-However, the next example does not type check because it tries to add
+By consistent we mean that we are only adding new fields.
+Nevertheless, the next example does not type check because it tries to add
 the same field to $\string_ENV$, but with different types:
@@
 \mylabel{T-IDREAD1}\\
-\dfrac{\env(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
-      {\env, \penv \vdash id:close(T_{2}), \env[id \mapsto open(T_{1})]}
+\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
+      {\env_{1}, \penv \vdash id:close(T_{2}), \env_{1}[id \mapsto open(T_{1})]}
 \end{array}
@@
 \mylabel{T-IDREAD2}\\
-\dfrac{\env(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
-      {\env, \penv \vdash id:fix(T_{2}), \env[id \mapsto fix(T_{1})]}
+\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
+      {\env_{1}, \penv \vdash id:fix(T_{2}), \env_{1}[id \mapsto fix(T_{1})]}
 \end{array}
@@
 In this example, it is unsafe to assign $\mathbf{nil}$ to the field
-$`foo"$ of $b$, as it changes the value that is stored in the field
+$``foo"$ of $b$, as it changes the value that is stored in the field
 $``foo"$ of $a$.
@@
 \mylabel{T-IDWRITE}\\
-\dfrac{\env(id) = T_{1} \;\;\; T_{2} = write(T_{1})}
-      {\env, \penv \vdash id_{l}:close(T_{2}), \env[id \mapsto close(T_{2})]}
+\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = write(T_{1})}
+      {\env_{1}, \penv \vdash id_{l}:close(T_{2}), \env_{1}[id \mapsto close(T_{2})]}
 \end{array}
@@
 This rule uses the auxiliary function \emph{write} because it may be
-accessing an identifier that is bound to a filter or projection type.
-As we mentioned in Section \ref{sec:unions}, assignments restore
-discriminated union types to their original types, and
-function \emph{write} works in this purpose.
+accessing an identifier that is bound to a filter type.
+As we mentioned in Sections \ref{sec:unions} and \ref{sec:alias},
+assignments restore discriminated union types to their original types,
+and function \emph{write} works in this purpose.
 We can define \emph{write} as follows:
@@
 does not type check because we cannot add the field $``x"$ to $b$,
-as its type is \emph{closed}, and thus does not allow changing the
-value that is stored in the field $``x"$ of $a$.
-
-We also have different rules for type checking table indexing to avoid
+as its type is \emph{closed} through \textsc{T-IDWRITE}.
+Aliasing $a$ changes the type of $a$ from $\emph{unique}$ to
+$\emph{open}$, and that is the reason why we can add the field
+$``x"$ to the type of $a$.
+Aliasing $a$ also produces the type $\{\}_{closed}$, which is
+the same type that $b$ has in left-hand side of the assignment.
+After the assignment, the type of $b$ is \emph{closed} and thus
+does not allow changing the value that is stored in the field
+$``x"$ of $a$.
+
+Our type system also has different rules for type checking table indexing to avoid
 changing table types in these operations, as they cannot create aliases:
@@
 \]
+
+Note that rule \textsc{T-COERCE} only allows changing the type
+of a variable if the new type is a supertype of the previous type,
+and the resulting type is always \emph{fixed} or \emph{closed}
+to prevent the creation of unsafe aliases.
 
@@
 The function \emph{crall} closes all \emph{unique} and \emph{open}
-table types; it also restores filter types to their original type.
+table types; it also restores filter types to their original types.
 The function \emph{crset} closes a given set of free assigned variables,
 which is given by the function \emph{fav}, and
-it also uses this set to restore filter types to their original type.
+it also uses this set to restore filter types to their original types.
 The function \emph{openset} changes from \emph{unique} to \emph{open}
@@
 $\Integer \times \Integer \times \Void \rightarrow \Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void$,
-while the function \emph{print} thas type
+while the function \emph{print} has type
 $\Value{*} \rightarrow \Void$.
@@
 This pair means that the last expression of an expression list $el$
-produces an union of second-level types $S_{2}$ that sould be bound
+produces an union of second-level types $S_{2}$ that should be bound
 to a variable $x$ in the projection environment $\penv$,
@@
 in the projection environment $\penv$. 
-This means that $q$ uses $\pi_{1}^{x}$ to project its type into the
-union type $\Integer \cup \Nil$, while $r$ uses $\pi_{2}^{x}$ to
-project its type into the union $\Integer \cup \String$.
+After assigning projection types to $q$ and $r$, reading $q$ will
+use the projection type $\pi_{1}^{x}$ to project the type of $q$
+into the union type $\Integer \cup \Nil$, while reading $r$ will
+use the projection type $\pi_{2}^{x}$ to project the type of $r$
+into the union type $\Integer \cup \String$.
 

### 2015-03-13 doc/thesis/system.tex
Adding new styles to formalize the AST

@@
 f \; | \;
-\{ \; \overline{p} \; \} \; | \;
-\{ \; \overline{p},me \; \} \; | \;
-e_{1} + e_{2} \; | \;
-e_{1} \; {..} \; e_{2} \\
-& | \; e_{1} == e_{2} \; | \;
+\{ \; \overline{[e_{1}] = e_{2}} \; \} \; | \;
+\{ \; \overline{[e_{1}] = e_{2}},me \; \} \\
+& | \; e_{1} + e_{2} \; | \;
+e_{1} \; {..} \; e_{2} \; | \;
+e_{1} == e_{2} \; | \;
 e_{1} < e_{2} \; | \;
@@
 e_{1} \; \mathbf{and} \; e_{2} \; | \;
-e_{1} \; \mathbf{or} \; e_{2} \; | \;
-\mathbf{not} \; e \; | \;
+e_{1} \; \mathbf{or} \; e_{2} \\
+& | \; \mathbf{not} \; e \; | \;
 \# \; e \; | \;
@@
 {\it string}\\
-p ::= & \;\; [e_{1}] = e_{2}\\
 el ::= & \;\; \overline{e} \; | \;
@@
 e{:}n(el)\\
-m ::= & \;\; \mathbf{fun} \; id_{1}{:}id_{2} \; (\overline{id{:}T}){:}S \; fb \; | \;
-\mathbf{fun} \; id_{1}{:}id_{2} \; (\overline{id{:}T},{...}{:}T){:}S \; fb\\
-f ::= & \;\; \mathbf{fun} \; (\overline{id{:}T}){:}S \; fb \; | \;
-\mathbf{fun} \; (\overline{id{:}T},{...}{:}T){:}S \; fb\\
-fb ::= & \;\; s \;;\; \mathbf{return} \; el\\
-id ::= & \;\; {\it name}\\
+m ::= & \;\; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el\\
+f ::= & \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el\\
+pl ::= & \;\; \overline{id{:}T} \; | \;
+\overline{id{:}T},{...}{:}T\\
+\end{array}
+$$
+\dend
+\caption{Typed Lua abstract syntax}
+\label{fig:syntax1}
+\end{figure}
+
+\begin{figure}[!ht]
+\textbf{Abstract Syntax}\\
+\dstart
+$$
+\begin{array}{rlr}
+s ::= & & \textsc{statements:}\\
+& \;\; \mathbf{skip} \;
+| \; s_{1} \; ; \; s_{2} \;
+| \; \overline{l} = el & \\
+& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
+| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \\
+& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s \;
+| \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \\
+& | \; \mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s \;
+| \; \mathbf{return} \; el \;
+| \; \lfloor a \rfloor_{0} & \\
+& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
+e ::= & & \textsc{expressions:}\\
+& \;\; \mathbf{nil} \;
+| \; k \;
+| \; id \;
+| \; e_{1}[e_{2}] \;
+| \; {<}T{>} \; id \;
+| \; f & \\
+& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
+| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \\
+& | \; e_{1} + e_{2} \;
+| \; e_{1} \; {..} \; e_{2} \;
+| \; e_{1} == e_{2} \;
+| \; e_{1} < e_{2} & \\
+& | \; e_{1} \;\&\; e_{2} \;
+| \; e_{1} \; \mathbf{and} \; e_{2} \;
+| \; e_{1} \; \mathbf{or} \; e_{2} \;
+| \; \mathbf{not} \; e \;
+| \; \# \; e & \\
+& | \; \lfloor me \rfloor_{1} & \\
+l ::= & & \textsc{left-hand values:}\\
+& \;\; id_{l} \; | \;
+e_{1}[e_{2}] \; | \;
+id[k] \; {<}T{>} & \\
+k ::= & & \textsc{literal constants:}\\
+& \;\; \mathbf{false} \; | \;
+\mathbf{true} \; | \;
+{\it int} \; | \;
+{\it float} \; | \;
+{\it string} & \\
+el ::= & & \textsc{expression lists:}\\
+& \;\; \overline{e} \; | \;
+\overline{e}, me & \\
+me ::= & & \textsc{multiple results:}\\
+& \;\; a \; | \;
+{...} & \\
+a ::= & & \textsc{applications:}\\
+& \;\; e(el) \; | \;
+e{:}n(el) & \\
+f ::= & & \textsc{function declarations:}\\
+& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
+pl ::= & & \textsc{parameter lists:}\\
+& \;\; \overline{id{:}T} \; | \;
+\overline{id{:}T},{...}{:}T & \\
+\end{array}
+$$
+\dend
+\caption{Typed Lua abstract syntax}
+\label{fig:syntax2}
+\end{figure}
+
+\begin{figure}[!ht]
+\textbf{Abstract Syntax}\\
+\dstart
+$$
+\begin{array}{rlr}
+s ::= & & \textsc{statements:}\\
+& \;\; \mathbf{skip} \;
+| \; s_{1} \; ; \; s_{2} & \textit{skip and sequence}\\
+& | \; \overline{l} = el & \textit{multiple assignment}\\
+& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
+| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
+& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s \;
+| \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
+& | \; \mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s & \textit{recursive function} \\
+& | \; \mathbf{return} \; el & \textit{return} \\
+& | \; \lfloor a \rfloor_{0} & \textit{application with no results}\\
+& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \textit{method declaration}\\
+e ::= & & \textsc{expressions:}\\
+& \;\; \mathbf{nil} \;
+| \; k & \textit{nil and other literals}\\
+& | \; id & \textit{variable reading}\\
+& | \; e_{1}[e_{2}] & \textit{table reading}\\
+& | \; {<}T{>} \; id & \textit{type coercion}\\
+& | \; f & \textit{function declaration}\\
+& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
+| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \textit{table constructor}\\
+& | \; e_{1} + e_{2} \;
+| \; e_{1} \; {..} \; e_{2} \;
+| \; e_{1} == e_{2} \;
+| \; e_{1} < e_{2} & \textit{binary operations}\\
+& | \; e_{1} \;\&\; e_{2} \;
+| \; e_{1} \; \mathbf{and} \; e_{2} \;
+| \; e_{1} \; \mathbf{or} \; e_{2} & \textit{binary operations}\\
+& | \; \mathbf{not} \; e \;
+| \; \# \; e & \textit{unary operations} \\
+& | \; \lfloor me \rfloor_{1} & \textit{expressions with one result}\\
+l ::= & & \textsc{left-hand values:}\\
+& \;\; id_{l} & \textit{variable writing}\\
+& | \; e_{1}[e_{2}] & \textit{table writing}\\
+& | \; id[k] \; {<}T{>} & \textit{type coercion}\\
+k ::= & & \textsc{literal constants:}\\
+& \;\; \mathbf{false} \; | \;
+\mathbf{true} \; | \;
+{\it int} \; | \;
+{\it float} \; | \;
+{\it string} & \\
+el ::= & & \textsc{expression lists:}\\
+& \;\; \overline{e} \; | \;
+\overline{e}, me & \\
+me ::= & & \textsc{multiple results:}\\
+& \;\; a & \textit{application}\\
+& | \; {...} & \textit{vararg expression}\\
+a ::= & & \textsc{applications:}\\
+& \;\; e(el) & \textit{function application}\\
+& | \; e{:}n(el) & \textit{method application}\\
+f ::= & & \textsc{function declarations:}\\
+& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
+pl ::= & & \textsc{parameter lists:}\\
+& \;\; \overline{id{:}T} \; | \;
+\overline{id{:}T},{...}{:}T & \\
 \end{array}

### 2015-03-15 doc/thesis/introduction.tex
Minor improvements to introduction, review and begining of typing rules

@@
 The aim of both approaches is to offer programmers the option
-to use type annotations where static typing is needed and to avoid
-them where dynamic typing is sufficient.
+to use type annotations where static typing is needed,
+allowing the incremental migration from dynamic to static typing.
 The difference between these two approaches is the way they treat
@@
 applications evolve into complex systems rather than just scripts.
-We shall see that Typed Lua introduces the refinement of
-tables to support the common way that Lua programmers use
-them to represent both modules and objects.
+We will show that Typed Lua introduces the refinement of
+tables to support the common idioms that Lua programmers use
+to encode both modules and objects.
 
@@
 
-In Chapter \ref{chap:review} we review the literature about static
-and dynamic typing in the same language, we discuss the differences
+This work is split into seven chapters.
+In Chapter \ref{chap:review} we review the literature about blending
+static and dynamic typing in the same language, we discuss the differences
 between optional type systems and gradual typing, and we also
@@
 design of Typed Lua.
-In Chapter \ref{chap:system} we use typing rules to present the
-formalization of Typed Lua.
-In Chapter \ref{chap:evaluation} we discuss some case studies that
-we used to evaluate our design, and we also compare our design
-to related work.
+In Chapter \ref{chap:system} we present our type system.
+In Chapter \ref{chap:evaluation} we discuss the evaluation
+results that we obtained while using Typed Lua to type existing Lua code.
+In Chapter \ref{chap:related} we present some related work.
 In Chapter \ref{chap:conc} we outline our contributions.

### 2015-03-15 doc/thesis/review.tex
Minor improvements to introduction, review and begining of typing rules

@@
 Both use JavaScript as their code generation target because
-their main purpose is web development.
+their main purpose is Web development.
 In fact, Dart is a new class-based object-oriented language with
@@
 \begin{array}{llr}
-T ::= & & \textsc{types}\\
+T ::= & & \textsc{types:}\\
 & \;\; \Number & \textit{base type number}\\
@@
 & | \; T \rightarrow T & \textit{function types}\\
-e ::= & & \textsc{expressions}\\
+e ::= & & \textsc{expressions:}\\
 & \;\; l & \textit{literals}\\
@@
 
-Figure \ref{fig:consistent-subtyping} combines consistency and subtyping to
-compose the consistent-subtyping relation for the simply typed
+Figure \ref{fig:consistent-subtyping} combines the consistency and subtyping
+relations to compose the consistent-subtyping relation for the simply typed
 lambda calculus extended with the dynamic type $?$.
@@
 Blame calculus is an intermediate language to integrate
-static and dynamic typing that uses the blame tracking approach
+static and dynamic typing along with the blame tracking approach
 proposed by \citet{findler2002chf}.
@@
 Perl 6 \citep{tang2007pri} is also being designed with gradual typing,
-though there is little documentation about the gradual type systems
+though there is few documentation about the gradual type systems
 of these languages.

### 2015-03-15 doc/thesis/system.tex
Minor improvements to introduction, review and begining of typing rules

@@
 In this section we use a reduced core of Typed Lua to present the
-typing rules that type check multiple assignment, table refinement,
+most interesting rules of our type system.
+These rules type check multiple assignment, table refinement,
 and overloading on the return type of functions.
-We limit this section to the rules that type check these features
-because these rules are the most interesting rules of our type system.
 Appendix \ref{app:rules} presents the full set of typing rules.
@@
 \begin{itemize}
-\item labels and goto statements are not present because they
-are difficult to handle along with our simplified form of
-\emph{flow typing}, and they are out of scope for now;
-\item explicit blocks are not present because we are already using
-explicit scope for variables;
+\item labels and goto statements (they are difficult to handle along
+with our simplified form of \emph{flow typing}, and they are out of
+scope for now);
+\item explicit blocks (we are already using explicit scope for variables);
 \item other loop structures such as repeat-until, numeric for,
-and generic for are not present because we can use while to
-express them;
-\item fields in the table constructor are reduced to a single
-form because we can use this form to express the missing forms; 
-\item arithmetic operators are reduced to addition only
-because the other operators have similar typing rules;
-\item relational operators are reduced to equality and less than
-because inequality has similar typing rules to equality and
-other relational operators have similar typing rules to less than;
-\item bitwise operators are reduced to bitwise and only
-because the other operators have similar typing rules.
+and generic for (we can use while to express them);
+\item table fields other than $[e_{1}] = e_{2}$
+(we can use this form to express the missing forms); 
+\item arithmetic operators other than $+$
+(other arithmetic operators have similar typing rules);
+\item relational operators other than $==$ and $<$
+(inequality has similar typing rules to $==$ and
+other relational operators have similar typing rules to $<$);
+\item bitwise operators other than $\&$
+(other bitwise operators have similar typing rules).
 \end{itemize}
 
-\begin{figure}[!ht]
-\textbf{Abstract Syntax}\\
-\dstart
-$$
-\begin{array}{rl}
-s ::= & \;\; \mathbf{skip} \; | \;
-s_{1} \; ; \; s_{2} \; | \;
-\overline{l} = el \; | \;
-m \; | \;
-\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
-\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
-& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s \; | \;
-\mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s \; | \;
-\mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s\\
-& | \; \mathbf{return} \; el \; | \;
-\lfloor a \rfloor_{0}\\
-e ::= & \;\; \mathbf{nil} \; | \;
-k \; | \;
-id \; | \;
-e_{1}[e_{2}] \; | \;
-{<}T{>} \; id \; | \;
-f \; | \;
-\{ \; \overline{[e_{1}] = e_{2}} \; \} \; | \;
-\{ \; \overline{[e_{1}] = e_{2}},me \; \} \\
-& | \; e_{1} + e_{2} \; | \;
-e_{1} \; {..} \; e_{2} \; | \;
-e_{1} == e_{2} \; | \;
-e_{1} < e_{2} \; | \;
-e_{1} \;\&\; e_{2} \; | \;
-e_{1} \; \mathbf{and} \; e_{2} \; | \;
-e_{1} \; \mathbf{or} \; e_{2} \\
-& | \; \mathbf{not} \; e \; | \;
-\# \; e \; | \;
-\lfloor me \rfloor_{1}\\
-l ::= & \;\; id_{l} \; | \;
-e_{1}[e_{2}] \; | \;
-id[k] \; {<}T{>}\\
-k ::= & \;\; \mathbf{false} \; | \;
-\mathbf{true} \; | \;
-{\it int} \; | \;
-{\it float} \; | \;
-{\it string}\\
-el ::= & \;\; \overline{e} \; | \;
-\overline{e}, me\\
-me ::= & \;\; a \; | \;
-{...}\\
-a ::= & \;\; e(el) \; | \;
-e{:}n(el)\\
-m ::= & \;\; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el\\
-f ::= & \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el\\
-pl ::= & \;\; \overline{id{:}T} \; | \;
-\overline{id{:}T},{...}{:}T\\
-\end{array}
-$$
-\dend
-\caption{Typed Lua abstract syntax}
-\label{fig:syntax1}
-\end{figure}
-
-\begin{figure}[!ht]
-\textbf{Abstract Syntax}\\
-\dstart
-$$
-\begin{array}{rlr}
-s ::= & & \textsc{statements:}\\
-& \;\; \mathbf{skip} \;
-| \; s_{1} \; ; \; s_{2} \;
-| \; \overline{l} = el & \\
-& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
-| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \\
-& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s \;
-| \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \\
-& | \; \mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s \;
-| \; \mathbf{return} \; el \;
-| \; \lfloor a \rfloor_{0} & \\
-& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
-e ::= & & \textsc{expressions:}\\
-& \;\; \mathbf{nil} \;
-| \; k \;
-| \; id \;
-| \; e_{1}[e_{2}] \;
-| \; {<}T{>} \; id \;
-| \; f & \\
-& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
-| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \\
-& | \; e_{1} + e_{2} \;
-| \; e_{1} \; {..} \; e_{2} \;
-| \; e_{1} == e_{2} \;
-| \; e_{1} < e_{2} & \\
-& | \; e_{1} \;\&\; e_{2} \;
-| \; e_{1} \; \mathbf{and} \; e_{2} \;
-| \; e_{1} \; \mathbf{or} \; e_{2} \;
-| \; \mathbf{not} \; e \;
-| \; \# \; e & \\
-& | \; \lfloor me \rfloor_{1} & \\
-l ::= & & \textsc{left-hand values:}\\
-& \;\; id_{l} \; | \;
-e_{1}[e_{2}] \; | \;
-id[k] \; {<}T{>} & \\
-k ::= & & \textsc{literal constants:}\\
-& \;\; \mathbf{false} \; | \;
-\mathbf{true} \; | \;
-{\it int} \; | \;
-{\it float} \; | \;
-{\it string} & \\
-el ::= & & \textsc{expression lists:}\\
-& \;\; \overline{e} \; | \;
-\overline{e}, me & \\
-me ::= & & \textsc{multiple results:}\\
-& \;\; a \; | \;
-{...} & \\
-a ::= & & \textsc{applications:}\\
-& \;\; e(el) \; | \;
-e{:}n(el) & \\
-f ::= & & \textsc{function declarations:}\\
-& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
-pl ::= & & \textsc{parameter lists:}\\
-& \;\; \overline{id{:}T} \; | \;
-\overline{id{:}T},{...}{:}T & \\
-\end{array}
-$$
-\dend
-\caption{Typed Lua abstract syntax}
-\label{fig:syntax2}
-\end{figure}
+Our reduced core does not lose much expressiveness, as it can express
+any Lua program except those that use labels and goto statements.
 
@@
 $s$ are statements, $e$ are expressions, $l$ are left-hand values,
-$el$ are expression lists, $k$ are literal constants, $p$ are table fields,
-$me$ are expressions with multiple results, $f$ are function declarations,
-$m$ are method declarations, $a$ are function and method applications,
+$k$ are literal constants, $el$ are expression lists,
+$me$ are expressions with multiple results, $a$ are function and method applications,
+$f$ are function declarations, $pl$ are parameter lists,
 $id$ are variable names, $T$ are first-level types, and $S$ are second-level types.
@@
 
-We present the typing rules as a deduction system for the typing relations
-$\env_{1}, \penv \vdash s, \env_{2}$ and $\env_{1}, \penv \vdash e : T, \env_{2}$.
-We use the first one for typing statements and the second one for typing expressions.
-The first relation means that given a type environment $\env_{1}$
-and a projection environment $\penv$, we can check a statement $s$
-producing a new type environment $\env_{2}$.
-The second relation means that given a type environment $\env_{1}$
-and a projection environment $\penv$, we can check that an expression $e$ has
-type $T$ and produces a new type environment $\env_{2}$.
-
 Even though we can assign only first-level types to variables,
@@
 second-level types into unions of first-level types.
-To do that, we use projection types and the environment $\penv$ that
-maps projection variables to second-level types, while the
-environment $\env$ maps variables to first-level types.
-We will show how our type system uses the environment $\penv$ for
-handling projection types in Section \ref{sec:fap}.
+We use two different environments to handle this feature.
+The first environment is the type environment $\env$ that maps
+variables to first-level types.
 We use $\env_{1}[id \mapsto T]$ to extend the environment $\env_{1}$
 with the variable $id$ that maps to type $T$.
-We use $\penv[x \mapsto S]$ to extend the environment $\penv$ with
-the projection variable $x$ that maps to the second-level type $S$.
+The second environment is the projection environment $\penv$ that
+maps projection variables to second-level types.
+We use $\penv[x \mapsto S]$ to extend the environment $\penv$
+with the projection variable $x$ that maps to type $S$.
+In Section \ref{sec:fap} we will show how our type system uses the
+projection environment $\penv$ for handling projection types,
+and also for projecting unions of second-level types into
+unions of first-level types.
+
+We present the typing rules as a deduction system for two typing relations,
+one for typing statements and another for typing expressions.
+
+We use the relation $\env_{1}, \penv \vdash s, \env_{2}$ for typing statements.
+This relation means that given a type environment $\env_{1}$
+and a projection environment $\penv$, we can check that a statement $s$
+produces a new type environment $\env_{2}$.
+
+We use the relation $\env_{1}, \penv \vdash e : T, \env_{2}$ for typing expressions.
+This relation means that given a type environment $\env_{1}$
+and a projection environment $\penv$, we can check that an expression $e$ has
+type $T$ and produces a new type environment $\env_{2}$.
 

### 2015-03-16 doc/thesis/system.tex
Rearranging the explanation of typing rules

@@
 As an example, lets assume that $x$ and $y$ are variables in the
-environment with types $\Integer$ and $\String$, the assignment
+environment with types $\Integer$ and $\String$.
+Let us see how our type system type checks the following assignment
 \[
@@
 \]
-type checks through the following rules:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-ASSIGNMENT}\\
-\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash \overline{l}:S_{2}, \env_{3} \;\;\;
-       S_{1} \lesssim S_{2}}
-      {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
-\end{array}
-\\ \\
+
+First, our type system type checks the expression list in the right-hand
+side of the assignment.
+In our example, the right-hand side of the assignment has type
+$1 \times ``foo" \times \Nil{*}$.
+Note that our type system includes the type $\Nil{*}$ to replace
+missing values.
+The rules that type check expression lists introduce the type
+$\Nil{*}$ to let the right-hand side produce fewer values than
+expected in the left-hand side.
+Our example uses the rule \textsc{T-EXPLIST1} which is the rule
+that type checks an expression list where all expressions can
+only produce a single value:
+\[
 \begin{array}{c}
@@
 \end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-LHSLIST}\\
-\dfrac{\env_{i}, \penv \vdash l_{i}:T_{i}, \env_{i+1} \;\;\;
-       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
-       n = |\;\overline{l}\;|}
-      {\env_{1}, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
-\end{array}
-\end{array}
-\]
-
-The rule \textsc{T-ASSIGNMENT} is the general rule,
-\textsc{T-EXPLIST1} is the rule that type checks an expression list
-where all expressions can only produce a single value, and
-\textsc{T-LHSLIST} is the rule that type checks a list of left-hand values.
-
-Note that our type system uses the type $\Value{*}$ to drop extra values
-and the type $\Nil{*}$ to replace missing values.
-The rule that type checks lists of left-hand values introduces the type
-$\Value{*}$ to let the right-hand side produce more values than
-expected in the left-hand side.
-The rules that type check expression lists introduce the type
-$\Nil{*}$ to let the right-hand side produce fewer values than
-expected in the left-hand side.
+\]
 
@@
 
-Back to our previous example, it type checks through these three rules
-that we just mentioned because
+Then, our type system type checks the list of left-hand values.
+In our example, the left-hand side of the assignment has type 
+$\Integer \times \String \times \Value{*}$.
+Note that our type system uses the type $\Value{*}$ to discard extra values.
+The rule that type checks lists of left-hand values introduces the type
+$\Value{*}$ to let the right-hand side produce more values than
+expected in the left-hand side.
+Our example uses the rule \textsc{T-LHSLIST} which is the rule that type
+checks a list of left-hand values:
+\[
+\begin{array}{c}
+\mylabel{T-LHSLIST}\\
+\dfrac{\env_{i}, \penv \vdash l_{i}:T_{i}, \env_{i+1} \;\;\;
+       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
+       n = |\;\overline{l}\;|}
+      {\env_{1}, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
+\end{array}
+\]
+
+After type checking the right-hand side and the left-hand side of an assignment,
+our type system checks whether their types are consistent.
+The rule \textsc{T-ASSIGNMENT} is the general rule that express this idea:
+\[
+\begin{array}{c}
+\mylabel{T-ASSIGNMENT}\\
+\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash \overline{l}:S_{2}, \env_{3} \;\;\;
+       S_{1} \lesssim S_{2}}
+      {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
+\end{array}
+\]
+
+Back to our example, it type checks through rule \textsc{T-ASSIGNMENT} because
 \[
@@
 As another example, lets assume that $x$, $y$, and $z$ are variables in
-the environment with types $\Integer$, $\String$, and $\String \cup \Nil$,
-the assignment
+the environment with types $\Integer$, $\String$, and $\String \cup \Nil$.
+The assignment
 \[
@@
 \[
-1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times \String \cup \Nil \times \Value{*}
+1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times (\String \cup \Nil) \times \Value{*}
 \]
@@
 that handle the cases where function applications are either statements
-or expressions that produce only one value:
+or expressions that produce only one value.
+The rule \textsc{T-STMAPPLY1} discards the produced values,
+while the rule \textsc{T-EXPAPPLY1} uses the auxiliary function
+\emph{first} to ensure that only one value is produced:
 \[
@@
 
-The rule \textsc{T-STMAPPLY1} discards the produced values,
-while the rule \textsc{T-EXPAPPLY1} uses the auxiliary function
-\emph{first} to ensure that only one value is produced.
 We can define \emph{first} inductively as follows:
@@
 
-As an example, assuming that $f$ is a local function in the environment, and that $f$ has type
-$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*} \rightarrow \Integer{*}$,
-the function call
+As an example, let us assume that $f$ is a local function in the environment,
+and that $f$ has type $\String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Value{*} \rightarrow \Integer{*}$.
+The function call
 \[
@@
 \[
-``foo" \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
+``foo" \times \Nil{*} \lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Value{*}
 \]
@@
 \[
-``foo" \times 1 \times 2 \times 3 \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
+``foo" \times 1 \times 2 \times 3 \times \Nil{*} \lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Value{*}
 \]
@@
 Our type system also catches arity mismatch.
-For instance, assuming that $f$ has type
-$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void \rightarrow \Integer{*}$,
-the function call
+For instance, let us assume that $f$ has type
+$\String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void \rightarrow \Integer{*}$.
+The function call
 \[
@@
 \[
-``foo" \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
+``foo" \times \Nil{*} \lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void
 \]
@@
 \[
-``foo" \times 1 \times 2 \times 3 \times \Nil{*} \not\lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
+``foo" \times 1 \times 2 \times 3 \times \Nil{*} \not\lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void
 \]
@@
 our type system to catch arity mismatch in function calls without optional parameters.
-For instance, assuming that $f$ has type
-$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Void$,
-the function call
+For instance, let us assume that $f$ has type
+$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Void$.
+The function call
 \[
@@
 Our abstract syntax reduces the syntactic forms of the table constructor
-into two forms: $\{\;\overline{p}\;\}$ and $\{\;\overline{p},me\;\}$.
-The first uses a list of table fields $p_{1}, ..., p_{n}$,
-where $p$ is $[e_{1}] = e_{2}$ and $n$ can be zero.
+into two forms: $\{\;\overline{[e_{1}] = e_{2}}\;\}$ and
+$\{\;\overline{[e_{1}] = e_{2}},me\;\}$.
+The first uses a list of table fields $([e_{1}] = e_{2})_{1}, ..., ([e_{1}] = e_{2})_{n}$.
 The second uses a list of table fields and an expression that can
 produce multiple values.
-We did not include in the abstract syntax a table constructor that uses
-lists of expressions, because we can express this form using lists of
-table fields.
-For instance, we can use the table constructor
-$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to express $\{ ``x", ``y", ``z" \}$.
 
@@
 
-The rule \textsc{T-CONSTRUCTOR2} type checks a table constructor with a non-empty
-list of table fields:
-\[
-\begin{array}{c}
-\mylabel{T-CONSTRUCTOR2}\\
-\dfrac{\env_{i}, \penv \vdash p_{i}:(K_{i},V_{i}), \env_{i+1} \;\;\;
-       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
-       wf(T) \;\;\;
-       n = |\;\overline{p}\;|}
-      {\env_{1}, \penv \vdash \{\;\overline{p}\;\}:T, \env_{n+1}}
-\end{array}
-\]
-
-As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
-has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$.
-
-First, the rule \textsc{T-CONSTRUCTOR2} uses the auxiliary relation
-$\env_{1}, \penv \vdash p : (K,V), \env_{2}$ to type check each table
-field $p_{i}$.
+As a more interesting example, let us see how our type system type checks
+the table constructor $\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$.
+
+First, our type system uses the auxiliary relation
+$\env_{1}, \penv \vdash [e_{1}] = e_{2} : (K,V), \env_{2}$ to type check each
+table field.
 This auxiliary relation means that given a type environment $\env_{1}$
-and a projection environment $\penv$, we can check a table field $p$
-producing a pair $(K,V)$ and a new type environment $\env_{2}$. 
-In a pair $(K,V)$, $K$ is the key type, while $V$ is the value type.
-
-Then, the rule \textsc{T-CONSTRUCTOR2} uses each pair $(K_{i},V_{i})$
+and a projection environment $\penv$, checking a table field $[e_{1}] = e_{2}$
+produces a pair $(K,V)$ and a new type environment $\env_{2}$.
+A pair $(K,V)$ means that $e_{1}$ has type $K$ and $e_{2}$ has type $V$,
+where $K$ is the type of the key and $V$ is the type of the value.
+
+After type checking each table field, our type system uses each pair $(K,V)$
 to build the table type that express the type of a given constructor, and
@@
 creation of unsafe aliases.
+The rules that type check table fields use the auxiliary function
+\emph{close} to close the type of the values in a table type.
+
+The rule \textsc{T-CONSTRUCTOR2} uses these steps to type check a
+table constructor with a non-empty list of table fields:
+\[
+\begin{array}{c}
+\mylabel{T-CONSTRUCTOR2}\\
+\dfrac{\begin{array}{c}
+       \env_{i}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
+       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
+       wf(T) \;\;\;
+       n = |\;\overline{[e_{1}] = e_{2}}\;|
+       \end{array}}
+      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{n+1}}
+\end{array}
+\]
+
+Back to our example, the table constructor
+$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ has type
+$\{1:``x", 2:``y", 3:``z"\}$ through rule \textsc{T-CONSTRUCTOR2}.
+
+As another example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
+has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ through rule
+\textsc{T-CONSTRUCTOR2}.
+The inner table is \emph{closed} to prevent the creation of unsafe aliases.
 
@@
 deterministically, as it makes easier the presentation of this behavior.
-Rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this non-deterministic behavior:
+
+As an example,
+\begin{center}
+\begin{tabular}{lll}
+\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
+& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = a \; \mathbf{in}$}\\
+& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
+& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
+\end{tabular}
+\end{center}
+does not type check, as aliasing $a$ produces the type $\{\}_{closed}$
+that is not a subtype of $\{\}_{open}$, the type of $b$.
+Our type system has this behavior to warn programmers about
+potential unsafe behaviors after this kind of alias.
+In this example, it is unsafe to add the field $``x"$ to $b$,
+as it changes the value that is stored in the field $``x"$ of $a$.
+
+Rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this non-deterministic behavior.
+Rule \textsc{T-IDREAD1} uses the auxiliary function \emph{close} to
+produce a \emph{closed} alias.
+It also uses the auxiliary function \emph{open} to change the type of
+the original reference from \emph{unique} to \emph{open},
+because aliasing an \emph{unique} table type while keeping the original
+reference \emph{unique} can be unsafe.
+Rule \textsc{T-IDREAD2} uses the auxiliary function \emph{fix} to
+produce a \emph{fixed} alias.
+It also uses \emph{fix} to change the type of the original reference
+to \emph{fixed}, because a \emph{fixed} table type does not allow
+width subtyping.
+We define these rules as follows:
 \[
@@
 
-As an example,
-\begin{center}
-\begin{tabular}{lll}
-\multicolumn{3}{l}{$\mathbf{local} \; a:\{``foo":\Integer\}_{unique} = \{ [``foo"] = 5 \} \; \mathbf{in}$}\\
-& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``foo":\Integer \cup \Nil \}_{closed} = a \; \mathbf{in}$}\\
-& & \multicolumn{1}{l}{$b[``foo"] = \mathbf{nil}$}
-\end{tabular}
-\end{center}
-does not type check,
-as aliasing $a$ produces the type $\{``foo":\Integer\}_{closed}$
-that is not a subtype of $\{``foo":\Integer \cup \Nil\}_{closed}$,
-the type of $b$.
-Our type system has this behavior to warn programmers about
-potential unsafe behaviors after this kind of alias.
-In this example, it is unsafe to assign $\mathbf{nil}$ to the field
-$``foo"$ of $b$, as it changes the value that is stored in the field
-$``foo"$ of $a$.
-
-As another example,
-\begin{center}
-\begin{tabular}{lll}
-\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
-& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = a \; \mathbf{in}$}\\
-& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
-& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
-\end{tabular}
-\end{center}
-does not type check, as aliasing $a$ produces the type $\{\}_{closed}$
-that is not a subtype of $\{\}_{open}$, the type of $b$.
-In this example, it is unsafe to add the field $``x"$ to $b$,
-as it changes the value that is stored in the field $``x"$ of $a$.
-
-Aliasing an \emph{unique} table type while keeping the original
-reference \emph{unique} can be unsafe.
-For this reason, the rule \textsc{T-IDREAD1} changes the type of
-the original reference from \emph{unique} to \emph{open}.
-
 We also need to close \emph{unique} and \emph{open} tables that
-appear in the left-hand side of assignments, as defined by
-rule \textsc{T-IDWRITE}:
-\[
-\begin{array}{c}
-\mylabel{T-IDWRITE}\\
-\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = write(T_{1})}
-      {\env_{1}, \penv \vdash id_{l}:close(T_{2}), \env_{1}[id \mapsto close(T_{2})]}
-\end{array}
-\]
-
-This rule uses the auxiliary function \emph{write} because it may be
-accessing an identifier that is bound to a filter type.
-As we mentioned in Sections \ref{sec:unions} and \ref{sec:alias},
-assignments restore discriminated union types to their original types,
-and function \emph{write} works in this purpose.
-We can define \emph{write} as follows:
-\begin{align*}
-write(\phi(T_{1},T_{2})) & = T_{1}\\
-write(T) & = T
-\end{align*}
+appear in the left-hand side of assignments, as leaving them
+\emph{unique} and \emph{open} would allow the creation of
+unsafe references.
 
@@
 does not type check because we cannot add the field $``x"$ to $b$,
-as its type is \emph{closed} through \textsc{T-IDWRITE}.
+as its type is \emph{closed}.
 Aliasing $a$ changes the type of $a$ from $\emph{unique}$ to
@@
 
+Rule \textsc{T-IDWRITE} defines this behavior:
+\[
+\begin{array}{c}
+\mylabel{T-IDWRITE}\\
+\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = write(T_{1})}
+      {\env_{1}, \penv \vdash id_{l}:close(T_{2}), \env_{1}[id \mapsto close(T_{2})]}
+\end{array}
+\]
+
+This rule uses the auxiliary function \emph{write} because it may be
+accessing an identifier that is bound to a filter type.
+As we mentioned in Sections \ref{sec:unions} and \ref{sec:alias},
+assignments restore discriminated union types to their original types,
+and function \emph{write} works in this purpose.
+We can define \emph{write} as follows:
+\begin{align*}
+write(\phi(T_{1},T_{2})) & = T_{1}\\
+write(T) & = T
+\end{align*}
+
 Our type system also has different rules for type checking table indexing to avoid
@@
 or merge existing fields.
-To do that, Typed Lua includes a type coercion expression that
-obeys the \textsc{T-COERCE} rule below:
-\[
-\begin{array}{c}
-\mylabel{T-COERCE}\\
-\dfrac{\env_{1}(id) \subtype T \;\;\;
-       \env_{1}[id \mapsto T], \penv \vdash id:T_{1}, \env_{2}}
-      {\env_{1}, \penv \vdash {<}T{>} \; id:T_{1}, \env_{2}}
-\end{array}
-\]
-
-Note that rule \textsc{T-COERCE} only allows changing the type
-of a variable if the new type is a supertype of the previous type,
-and the resulting type is always \emph{fixed} or \emph{closed}
-to prevent the creation of unsafe aliases.
-
-The rule \textsc{T-COERCE} allows our type system to type check the following example:
+To do that, Typed Lua includes a type coercion expression ${<}T{>} \; id$.
+For instance, we can use this type coercion expression to make the following
+example type check:
 \begin{center}
@@
 
+Rule \textsc{T-COERCE} defines the behavior of the type coercion expression:
+\[
+\begin{array}{c}
+\mylabel{T-COERCE}\\
+\dfrac{\env_{1}(id) \subtype T \;\;\;
+       \env_{1}[id \mapsto T], \penv \vdash id:T_{1}, \env_{2}}
+      {\env_{1}, \penv \vdash {<}T{>} \; id:T_{1}, \env_{2}}
+\end{array}
+\]
+
+Note that rule \textsc{T-COERCE} only allows changing the type
+of a variable if the new type is a supertype of the previous type,
+and the resulting type is always \emph{fixed} or \emph{closed}
+to prevent the creation of unsafe aliases.
+
 We also need to make sure to close all \emph{unique} and \emph{open}
 table types before we type check a nested scope.
-The rule \textsc{T-FUNCTION1} illustrates this case:
-\[
-\begin{array}{c}
-\mylabel{T-FUNCTION1}\\
-\dfrac{\begin{array}{c}
-       crall(\env_{1}[\overline{id} \mapsto \overline{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}\\
-       \env_{3} = openset(crset(\env_{1}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)),rv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
-       \end{array}}
-      {\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{3}}
-\end{array}
-\]
-
-Rule \textsc{T-FUNCTION1} uses some auxiliary functions to change
+To do that, our type system uses some auxiliary functions to change
 the type of variables before type checking a nested scope and
@@
 
-This rule also extends the environment $\penv$, bounding the special
-variable $\ret$ to the return type $S$.
-Rule \textsc{T-RETURN} uses the type that is bound to $\ret$ in
-$\penv$ to type check return statements:
-\[
-\begin{array}{c}
-\mylabel{T-RETURN}\\
-\dfrac{\env_{1} \vdash el:S_{1}, \env_{2} \;\;\;
-       \penv(\ret) = S_{2} \;\;\;
-       S_{1} \lesssim S_{2}}
-      {\env_{1} \vdash \mathbf{return} \; el, \env_{2}}
-\end{array}
-\]
-
-As an example, rule \textsc{T-FUNCTION1} prevents the following unsafe example to type check:
+As an example,
 \begin{center}
@@
 \end{center}
-
-This example does not type check because we cannot add the field
+does not type check because we cannot add the field
 $``x"$ to $b$, as its type is closed.
-The assignment $b = a$ type checks because, at this point,
+The assignment $b = a$ type checks because, at that point,
 $a$ and $b$ have the same type: $\{\}_{closed}$.
@@
 
+Rule \textsc{T-FUNCTION1} illustrates this case:
+\[
+\begin{array}{c}
+\mylabel{T-FUNCTION1}\\
+\dfrac{\begin{array}{c}
+       crall(\env_{1}[\overline{id} \mapsto \overline{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}\\
+       \env_{3} = openset(crset(\env_{1}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)),rv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
+       \end{array}}
+      {\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{3}}
+\end{array}
+\]
+
+This rule also extends the environment $\penv$, bounding the special
+variable $\ret$ to the return type $S$.
+Rule \textsc{T-RETURN} uses the type that is bound to $\ret$ in
+$\penv$ to type check return statements:
+\[
+\begin{array}{c}
+\mylabel{T-RETURN}\\
+\dfrac{\env_{1} \vdash el:S_{1}, \env_{2} \;\;\;
+       \penv(\ret) = S_{2} \;\;\;
+       S_{1} \lesssim S_{2}}
+      {\env_{1} \vdash \mathbf{return} \; el, \env_{2}}
+\end{array}
+\]
+
 \subsection{Projections}
@@
 
-As an example, lets assume that \emph{idiv} and \emph{print} are functions
+As an example, let us assume that \emph{idiv} and \emph{print} are functions
 in the environment.
 The function \emph{idiv} has type
-$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void$,
-while the function \emph{print} has type
+\[
+\Integer \times \Integer \times \Void \rightarrow (\Integer \times \Integer \times \Void) \sqcup (\Nil \times \String \times \Void)
+\]
+As we mentioned in Section \ref{sec:unions}, \emph{idiv} performs
+integer division.
+In case of success, it returns two integers: the result and the remainder.
+In case of failure, it returns $\Nil$ plus an error message that indicates
+the error.
+The function \emph{print} is a variadic function of type
 $\Value{*} \rightarrow \Void$.
-Our type system uses three typing rules to type check the following program:
+Let us see how our type system type checks the following program:
 \begin{center}
@@
 
-First, our type system uses rule \textsc{T-EXPLIST6} for type checking
-$idiv(1, 2)$:
-\[
-\begin{array}{c}
-\mylabel{T-EXPLIST6}\\
-\dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Void \sqcup T_{1}' \times ... \times T_{n}' \times \Void, \env_{2}\\
-       S = T_{1} \times ... \times T_{n} \times \Nil{*} \sqcup T_{1}' \times ... \times T_{n}' \times \Nil{*}
-       \end{array}}
-      {\env_{1}, \penv \vdash me:\pi_{1}^{x} \times ... \times \pi_{n}^{x} \times \Nil{*}, \env_{2}, (x,S)}
-\end{array}
-\]
-
-Rule \textsc{T-EXPLIST6} uses the auxiliary relation
-$\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$.
+First, our type system uses the auxiliary relation
+$\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$
+for type checking $idiv(1, 2)$.
 This relation means that given a type environment $\env_{1}$ and
@@
 types $\pi_{i}^{x}$.
-In our example, we can check that $idiv(1, 2)$ has type
+In our example, our type system uses rule \textsc{T-EXPLIST6} for
+type checking $idiv(1, 2)$:
+\[
+\begin{array}{c}
+\mylabel{T-EXPLIST6}\\
+\dfrac{\begin{array}{c}
+       \env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Void \sqcup T_{1}' \times ... \times T_{n}' \times \Void, \env_{2}\\
+       S = T_{1} \times ... \times T_{n} \times \Nil{*} \sqcup T_{1}' \times ... \times T_{n}' \times \Nil{*}
+       \end{array}}
+      {\env_{1}, \penv \vdash me:\pi_{1}^{x} \times ... \times \pi_{n}^{x} \times \Nil{*}, \env_{2}, (x,S)}
+\end{array}
+\]
+
+Note that $idiv(1, 2)$ has type
 $\pi_{1}^{x} \times \pi_{2}^{x} \times \Nil{*}$ and produces the pair
-$(x,\Integer \times \Integer \times \Nil{*} \sqcup \Nil \times \String \times \Nil{*})$.
-
-Our type system uses the pair $(x,S_{2})$ to allow rule
-\textsc{T-LOCAL2} to bound a union of second-level types $S_{2}$
-to a variable $x$ in the projection environment $\penv$:
+\[
+(x,(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})) 
+\]
+
+In the rule that type checks the declaration of unannotated variables,
+our type system uses the pair $(x,S_{2})$ to bound a union of
+second-level types $S_{2}$ to a variable $x$ in the projection
+environment $\penv$.
+In our example, declaring $q$ and $r$
+bounds the projection type $\pi_{1}^{x}$ to $q$ and
+bounds the projection type $\pi_{2}^{x}$ to $r$,
+where the projection variable $x$ bounds to 
+\[
+(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})
+\]
+in the projection environment $\penv$.
+Rule \textsc{T-LOCAL2} captures this intuition:
 \[
@@
 before type checking $s$.
-In our example, we can check that declaring $q$ and $r$
-bounds the projection type $\pi_{1}^{x}$ to $q$ and
-bounds the projection type $\pi_{2}^{x}$ to $r$, where the projection variable $x$ bounds to 
-\[
-\Integer \times \Integer \times \Nil{*} \sqcup \Nil \times \String \times \Nil{*}
-\]
-in the projection environment $\penv$. 
+
 After assigning projection types to $q$ and $r$, reading $q$ will
@@
 into the union type $\Integer \cup \String$.
+However, we want to discriminate these variables to check whether
+the function call returned with success.
 
@@
 \[
-fpt(\Integer \times \Integer \times \Nil{*} \sqcup \Nil \times \String \times \Nil{*}, \Nil, 1)
+fpt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
 \]
@@
 \[
-gpt(\Integer \times \Integer \times \Nil{*} \sqcup \Nil \times \String \times \Nil{*}, \Nil, 1)
+gpt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
 \]

### 2015-03-16 doc/thesis/typedlua.tex
Minor improvements to the first part of Chapter 3

@@
 We believe that a careful evaluation of run-time checks should be done before
-inserting them on the type system.
-However, this evaluation is out of the scope of this work.
+inserting them in the type system.
+However, this evaluation is out of scope of this work.
 
@@
 switch Typed Lua from optional typing to gradual typing in the future.
-As we mentioned in the previous chapter,
-a sound type system is a prerequisite to insert run-time checks after
+A sound type system is a prerequisite to insert run-time checks after
 static type checking, because a sound type system ensures that
@@
 These examples give an informal overview of our optional type system.
-We will use typing rules to present the formalization of our optional type
-system in the next chapter.
+In the next chapter we will use typing rules to present the formalization
+of the most interesting features of our optional type system.
 All the examples that we present in this chapter run in our Typed Lua compiler.
@@
   \textit{uniontuple} \; \texttt{`|'} \; \textit{uniontuple}\\
-\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
+\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'} \; | \;
+  \texttt{`<\{'} \; [\textit{tabletypebody}] \; \texttt{`\}>'}\\
 \textit{tabletypebody} ::= & \;\; \textit{maptype} \; | \;
@@
 Unions of tuple types are also supertypes of their parts.
+We will present the formalization of the subtyping relation
+in Section \ref{sec:subtyping}.
 
@@
 operations, and the programmer can choose to be warned about them.
+We will discuss the formalization of the consistent-subtyping
+relation in Section \ref{sec:subtyping}.
 
@@
 However, we can call \texttt{absolute} inside \texttt{distance}
-because the expression \texttt{x - y} has type \texttt{any},
+because the subtraction expression \texttt{x - y} has type \texttt{any},
 and it is consistent with type \texttt{integer}.
@@
 
-In the default mode, the type of \texttt{sum} is
+In the default mode, \texttt{sum} has type
 \texttt{(integer, integer, value*) -> (integer, nil*)}.
@@
 
-In the strict mode, the type of \texttt{sum} is
+In the strict mode, \texttt{sum} has type
 \texttt{(integer, integer) -> (integer)}.
@@
 tuple type \texttt{(1, 2, 3, nil*)} in both operation modes.
+
+As another example,
+\begin{verbatim}
+    local x:integer, y:integer = sum(2, 2)
+\end{verbatim}
+compiles with warnings in both operation modes.
+This means that the left-hand side of this local declaration has
+the tuple type \texttt{(integer, integer, value*)}, while the
+expression list in the right-hand side of this local declaration
+has the tuple type \texttt{(integer, nil*)} in both operation modes.
+The type \texttt{nil*} generates a value of type \texttt{nil}
+that is not consistent with \texttt{integer}, the type of \texttt{y}.
 
@@
 
-Typed Lua allows the assignment \texttt{v = v or e} to refine
+Typed Lua allows the assignment \texttt{v = v or e} to change
 the type of \texttt{v} from \texttt{t|nil} to \texttt{t}
@@
 and it is not being assigned in another function.
-The refinement only affects the type of \texttt{v} until the
-next assignment to \texttt{v}.
-In the case of \texttt{greeting}, the assignment constrains its type
+The change only affects the type of \texttt{v} in the remainder
+of the current scope.
+Any assignment to \texttt{v} restores its type back to the
+original type.
+In the case of \texttt{greeting}, the assignment change its type
 from \texttt{string|nil} to \texttt{string}.
@@
 any warnings in both modes, because the argument list
-\texttt{("Typed Lua", nil*)} is consistent with
+\texttt{("Typed Lua", nil*)} is consistent with the input type
 \texttt{(string, string|nil, value*)} in the default mode,
-and it is also consistent with  
+and it is also consistent with the input type 
 \texttt{(string, string|nil)} in the strict mode.
@@
 a function declaration.
+In Section \ref{sec:assignment} we will formalize this behavior.
 
@@
 programmers to constrain the type of a local variable inside a condition.
-This example uses the \texttt{type(v) == "string"} predicate that
+This example uses the predicate \texttt{type(v) == "string"} that
 constrains the type of \texttt{v} from \texttt{string|t} to
-\texttt{string} inside the \texttt{if} body and to \texttt{t}
-inside the \texttt{else} body.
+\texttt{string} when the predicate is true and \texttt{t} otherwise.
 This is a simplified form of \emph{flow typing} \citep{guha2011tlc,tobin-hochstadt2010ltu}.
@@
 when it successfully finished its execution,
-and returns \texttt{nil} plus an error message or other data that
+and it returns \texttt{nil} plus an error message or other data that
 describes the error when something failed during its execution.
@@
 \begin{align*}
-\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
+\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'} \; | \;
+  \texttt{`<\{'} \; [\textit{tabletypebody}] \; \texttt{`\}>'}\\
 \textit{tabletypebody} ::= & \;\; \textit{maptype} \; | \;
@@
 in Section \ref{sec:statistics}.
-Later in this chapter we will show that we use the syntax of records
+
+Even though Typed Lua includes two different forms for annotating
+a table, in this section we only discuss the form
+$\texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}$.
+We call this form \emph{closed} table type because it
+represents table types that do not provide
+any guarantees about keys with types not listed in the table type.
+In Sections \ref{sec:modules} and \ref{sec:oop} we will show that
+we use the syntax $\texttt{`<\{'} \; [\textit{tabletypebody}] \; \texttt{`\}>'}$
 to type modules and objects.
+We call this form \emph{fixed} table type because it
+represents tables with no keys that do not inhabit one of the table's
+key types, and can have safe aliases.
 We will also show that we need the \texttt{const} modifier while
 typing object-oriented code.
+In the next chapter we formalize and discuss in more detail the
+different table types that our type system has.
 
@@
 If a variable gets its type from a table constructor then it is
-unique, otherwise it is closed.
-Any alias to unique variables makes them open, as we use the
-tag open to keep track of unique variables that are aliased.
-In the previous example, \texttt{person} has an unique type
+\emph{unique}, otherwise it can be either \emph{fixed} or
+\emph{closed}.
+In this section we discuss the cases that produce \emph{closed}
+table types, while in Sections \ref{sec:modules} and \ref{sec:oop}
+we will discuss the cases that produce \emph{fixed} table types
+to describe the type of modules and classes.
+Any alias to \emph{unique} variables makes them \emph{open}, as we use the
+tag \emph{open} to keep track of \emph{unique} variables that are aliased.
+In the previous example, \texttt{person} has an \emph{unique} type
 because it has no alias.
-In Section \ref{sec:oop} we will show that we use the tag
-fixed to describe the type of classes.
 Typed Lua also has different subtyping rules that handle these
@@
 Typed Lua uses three rules to decide whether it is sound to
-allow field assignment to change the type of unique and open
-variables:
+allow field assignment to change the type of \emph{unique} and
+\emph{open} variables:
 the variable must be local to the current block,
@@
 
-We can create aliases to open variables, but they are always closed.
+We can create aliases to \emph{open} variables, but they are always
+\emph{closed}.
 Any mutation in the original reference and in the aliased reference
@@
 In our previous example, the second line changes the type of
-\texttt{person} from unique to open and makes the type of
-\texttt{bogus} closed.
+\texttt{person} from \emph{unique} to \emph{open} and makes the type of
+\texttt{bogus} \emph{closed}.
 
@@
 As an example, the following declaration defines the type \texttt{Person}
-as an alias to the table type \texttt{\{"firstname":string, "lastname":string\}}
+as an alias to the \emph{closed} table type \texttt{\{"firstname":string, "lastname":string\}}
 in the remainder of the current scope:
@@
 Typed Lua also includes \emph{interfaces} as syntactic sugar to aliases
-for table types that specify records, as writing table types can be unwieldy
+for \emph{closed} table types that specify records, as writing table types can be unwieldy
 when records get bigger and the types of record fields get more complicated.
@@
 \texttt{interface} construct.
-In Section \ref{sec:oop} we will show that we can also use \emph{interface} to
-declare the type of classes, and that \emph{methodtype} is syntactic
-sugar to express the type of methods.
 
@@
 \textit{typedec} ::= & \;\; \textit{Name} \; \{\textit{decitem}\} \; \textbf{end}\\
-\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
-\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
+\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{type}\\
 \textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
@@
 the type alias declaration, it has a more convenient syntax
-for declaring table types that express records.
+for declaring \emph{closed} table types that express records.
 For this reason, it worth having two different constructs;
@@
     local RADIANS_PER_DEGREE = 3.14159 / 180.0
-    local function deg (x:number)
+    local function deg (x:number):number
       return x / RADIANS_PER_DEGREE
     end
-    local function rad (x:number)
+    local function rad (x:number):number
       return x * RADIANS_PER_DEGREE
     end
-    local function pow (x:number, y:number)
+    local function pow (x:number, y:number):number
       return x ^ y
@@
 \begin{verbatim}
-    { "deg":(number) -> (number),
+    <{ "deg":(number) -> (number),
        "rad":(number) -> (number),
-      "pow":(number, number) -> (number) }
-\end{verbatim}
+       "pow":(number, number) -> (number) }>
+\end{verbatim}
+
+Typed Lua always fixes \emph{unique} and \emph{open} table types
+that appear in a return statement.
+Typed Lua has this behavior to ensure that modules and classes
+always implement \emph{fixed} table types.
+In the next section we will show that we build classes in a
+similar way that we use to build modules, and we will also show
+that classes should have \emph{fixed} table types to allow safe
+inheritance.
 
@@
     local RADIANS_PER_DEGREE = 3.14159 / 180.0
-    mymath.deg = function (x:number)
+    mymath.deg = function (x:number):number
       return x / RADIANS_PER_DEGREE
     end
-    mymath.rad = function (x:number)
+    mymath.rad = function (x:number):number
       return x * RADIANS_PER_DEGREE
     end
-    mymath.pow = function (x:number, y:number)
+    mymath.pow = function (x:number, y:number):number
       return x ^ y
@@
 In this example, Typed Lua uses the refinement of table types to
-allow the programmer to incrementally build the type of the module.
+allow the programmer to incrementally build the type of the module
+\texttt{mymath}.
 The variable \texttt{RADIANS\string_PER\string_DEGREE}
@@
 The \texttt{return} at the end of the source file gives the
-type of the module, which is the same type of \texttt{mymath}.
+type of the module, which is the same type of the variable \texttt{mymath}.
 This module has the same type of the module from the previous example.
@@
 as Typed Lua uses this literal string to find the source file that
-implements the given module.
+implements a given module.
 To statically type check a module, Typed Lua follows the same rules
@@
 
-The way that Typed Lua handles modules using table types is also
-relevant for supporting object-oriented programming, as we discuss in
-the next section.
+The way that Typed Lua handles modules using \emph{fixed} table types
+is also relevant for supporting object-oriented programming,
+as we discuss in the next section.
 
@@
 \label{sec:oop}
+
+\begin{figure}[!ht]
+\textbf{Classes}\\
+\dstart
+\begin{align*}
+\textit{interface} ::= & \;\; [\textbf{local}] \; \textbf{interface} \; \textit{typedec}\\
+\textit{typedec} ::= & \;\; \textit{Name} \; \{\textit{decitem}\} \; \textbf{end}\\
+\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
+\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
+\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
+\textit{id} ::= & \;\; [\textbf{const}] \; \textit{Name}\\
+\textit{methodtype} ::= & \;\; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettype}
+\end{align*}
+\dend
+\caption{The concrete syntax of Typed Lua interfaces}
+\label{fig:classes}
+\end{figure}
 

### 2015-03-17 doc/thesis/typedlua.tex
Minor improvements to the last part of Chapter 3

@@
 
-\begin{figure}[!ht]
-\textbf{Classes}\\
-\dstart
-\begin{align*}
-\textit{interface} ::= & \;\; [\textbf{local}] \; \textbf{interface} \; \textit{typedec}\\
-\textit{typedec} ::= & \;\; \textit{Name} \; \{\textit{decitem}\} \; \textbf{end}\\
-\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
-\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
-\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
-\textit{id} ::= & \;\; [\textbf{const}] \; \textit{Name}\\
-\textit{methodtype} ::= & \;\; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettype}
-\end{align*}
-\dend
-\caption{The concrete syntax of Typed Lua interfaces}
-\label{fig:classes}
-\end{figure}
-
 Lua provides minimal support for object-oriented programming.
@@
 
-Typed Lua uses table types along with the type \texttt{self} to
-represent objects.
+Typed Lua uses table types along with the type
+\texttt{self} to represent objects.
 We use the following example to discuss this feature:
@@
 Lua programmers often use this mechanism to simulate classes.
-
-Typed Lua combines this mechanism along with type aliases and
-interfaces to allow programmers to define classes in a similar
-fashion to the definition of modules.
-We use the following example to discuss how Typed Lua type checks
-object-oriented code:
-\begin{verbatim}
-    interface Shape
-      x, y:number
-      const new:(number, number) => (self)
-      const move:(number, number) => ()
-    end
+We will use the following example to discuss this feature:
+\begin{verbatim}
     local Shape = { x = 0.0, y = 0.0 }
-    const function Shape:new (x:number, y:number)
-      local s = setmetatable({}, { __index = self })
+    const function Shape:new (x:number, y:number):self
+      local s:self = setmetatable({}, { __index = self })
       s.x = x
@@
     end
-    const function Shape:move (dx:number, dy:number)
+    const function Shape:move (dx:number, dy:number):()
       self.x = self.x + dx
@@
 
-In this example, we use an interface to define the type of the class \texttt{Shape}.
-We use the \texttt{const} annotation in the type of the methods
-because it is necessary for covariance among object types work.
-We use a double arrow instead of a single arrow because it is
-syntactic sugar for defining the type of methods,
-as it includes an implicit first type \texttt{self}.
-The double arrow can appear only inside the declaration of
-interfaces and userdata, which we will introduce in the next section.
-This interface definition exports the following object type to the class \texttt{Shape}:
-\begin{verbatim}
-    { "x":number, "y":number,
-      const "new":(self, number, number) -> (self),
-      const "move":(self, number, number) -> () }
-\end{verbatim}
-
-In Typed Lua, \texttt{setmetatable} is a strict primitive.
-Typed Lua only allows calling \texttt{setmetatable} with
-two operands.
-The first operand can be any table \texttt{t1}, but
-the second operand must be a table of the form
-\texttt{\{ \string_\string_index = t2 \}},
-where \texttt{t2} can be any table.
+In Typed Lua, \texttt{setmetatable} is a strict primitive
+that obeys three typing rules.
 The reason for being so strict with \texttt{setmetatable} is because
@@
 as prototype objects.
-In a \texttt{setmetatable} expression, if the type of \texttt{t2}
-is a subtype of the type of \texttt{t1}, it changes \texttt{t1} to
-have the same type of \texttt{t2}.
-
-In the previous example, we use \texttt{setmetatable} to define a constructor
-\texttt{new} to the class \texttt{Shape}.
-Note that we use \texttt{setmetatable} inside \texttt{new} to declare
-the local \texttt{s} having the same type of \texttt{self}.
-
-Still in the previous example, we are using the refinement of table
-types to build the type of the variable \texttt{Shape},
-as it should work as our object prototype, matching the type
-that the interface declaration exports.
+
+The first \texttt{setmetatable} rule appears in our previous example.
+In a \texttt{setmetatable} expression
+\texttt{setmetatable(\{\},\{\string_\string_index = id\})},
+if \texttt{id} has type \texttt{self} then the expression also also has type \texttt{self}.
+We will explain the other two rules while we explain how
+Typed Lua type checks single inheritance.
+
+In our example, we are using the refinement of table types to
+build the type of the variable \texttt{Shape},
+as it should work as our class.
 This means that the local \texttt{Shape} has type
@@
 inside the definition of \texttt{move}.
-Finally, the local \texttt{Shape} has the same type of the
-interface declaration in the last \texttt{return} statement.
-
-In Typed Lua, the style of classes definition from the previous example
+We use the \texttt{const} annotation in the type of the methods
+because it is necessary for covariance among object types work.
+Finally, the last \texttt{return} statement fixes the type of
+\texttt{Shape} because \emph{fixed} table types do not allow width
+subtyping, and this behavior allows Typed Lua to use the refinment
+of table types to type check single inheritance, as we will see in
+this section.
+
+A return statement that appears in the top-level also exports
+a type alias to be used in type annotations.
+For instance, our example exports the following type alias:
+\begin{verbatim}
+    typealias Shape =
+      <{ "x":number, "y":number,
+         const "new":(self, number, number) -> (self),
+         const "move":(self, number, number) -> () }>
+\end{verbatim}
+
+The style of classes definition from the previous example
 allow us to use \texttt{require} for creating prototype objects that
@@
 
+Typed Lua also includes a \texttt{class} construct that programmers
+can use to document the type of the classes they export..
+Figure \ref{fig:classes} shows the concrete syntax of the
+\texttt{class} construct.
+
+\begin{figure}[!ht]
+\textbf{Classes}\\
+\dstart
+\begin{align*}
+\textit{class} ::= & \;\; [\textbf{local}] \; \textbf{class} \; \textit{typedec}\\
+\textit{typedec} ::= & \;\; \textit{Name} \; \{\textit{decitem}\} \; \textbf{end}\\
+\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
+\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
+\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
+\textit{id} ::= & \;\; [\textbf{const}] \; \textit{Name}\\
+\textit{methodtype} ::= & \;\; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettype}
+\end{align*}
+\dend
+\caption{The concrete syntax of Typed Lua classes}
+\label{fig:classes}
+\end{figure}
+
+As an example, we can use the following \texttt{class} declaration to
+document the type \texttt{Shape} from the previous example:
+\begin{verbatim}
+    class Shape
+      x, y:number
+      const new:(number, number) => (self)
+      const move:(number, number) => ()
+    end
+\end{verbatim}
+
+We use a double arrow instead of a single arrow because it is
+syntactic sugar for defining the type of methods,
+as it includes an implicit first type \texttt{self}.
+The double arrow can appear only inside the declaration of
+classes and userdata, which we will introduce in the next section.
+
 Typed Lua also uses \texttt{setmetatable} to simulate single inheritance.
@@
     local Shape = require "shape"
-    interface Circle
+    local Circle = setmetatable({}, { __index = Shape })
+    Circle.radius = 0.0
+    const function Circle:new (x:number, y:number, radius:value):self
+      local c:self = setmetatable(Shape:new(x, y), { __index = self })
+      c.radius = tonumber(radius) or 0.0
+      return c
+    end
+    const function Circle:area ():number
+      return math.pi * self.radius * self.radius
+    end
+    return Circle
+\end{verbatim}
+
+In this example, we use the other two \texttt{setmetatable} rules.
+The second rule appears to trigger the refinement of table types,
+as we need this rule to add new methods and also to override
+existing methods in \texttt{Circle}.
+In a \texttt{setmetatable} expression
+\texttt{setmetatable(\{\},\{\string_\string_index = id\})},
+if \texttt{id} has a \emph{fixed} table type then it is safe to
+produce an equivalent \emph{open} table type, as \emph{fixed}
+table types do not allow hiding fields.
+The third rule appears to redefine the constructor \texttt{new}.
+In a \texttt{setmetatable} expression
+\texttt{setmetatable(e,\{\string_\string_index = id\})},
+if \texttt{e} has a \emph{fixed} table type \texttt{t}, \texttt{id} has
+type \texttt{self}, and the type bound to \texttt{self} is a subtype of
+\texttt{t}, then the expression has type \texttt{self}.
+Note how we can use this rule to call the constructor of \texttt{Shape}
+inside the overridden constructor.
+A limitation of this class system is that the overridden constructor
+must be a subtype of the original constructor,
+so the type of \texttt{radius} has to be quite permissive.
+
+Like in the preivous example, the last return statement in the
+top-level exports a \emph{fixed} table type, which we could
+document through the following class declaration:
+\begin{verbatim}
+    class Circle
       x, y, radius:number
@@
     end
-    local Circle = setmetatable({}, { __index = Shape })
-    Circle.radius = 0.0
-    const function Circle:new (x:number, y:number, radius:value)
-      local c = setmetatable(Shape:new(x, y), { __index = self })
-      c.radius = tonumber(radius) or 0.0
-      return c
-    end
-    const function Circle:area ()
-      return math.pi * self.radius * self.radius
-    end
-    return Circle
-\end{verbatim}
-
-In this example, we use an interface to define the type of the class \texttt{Circle},
-but with a syntactical overhead to manually include new and
-overridden methods.
-This interface definition exports the following object type to the class \texttt{Circle}:
-\begin{verbatim}
-    { "x":number, "y":number, "radius":number,
+\end{verbatim}
+
+Note that this class definition introduces a syntactical overhead
+to manually include new and overriden methods.
+This class definition is equivalent to the alias that exports the
+following \emph{fixed} table type to the class \texttt{Circle}:
+\begin{verbatim}
+    <{ "x":number, "y":number, "radius":number,
        const "new":(self, number, number, value) -> (self),
        const "move":(self, number, number) -> (),
-      const "radius":() -> (number) }
-\end{verbatim}
-
-In the previous example, we use \texttt{setmetatable} in two different ways.
-In the first case, we use \texttt{setmetatable} to trigger the refinement
-of table types, as we need it to add new methods and override existing ones
-in \texttt{Circle}.
-This is possible because a call to \texttt{setmetatable},
-where \texttt{t1} is a table constructor assigned to a fresh local variable,
-and \texttt{t2} is not \texttt{self} gives this variable an open type.
-In the second case, we use \texttt{setmetatable} to redefine \texttt{new}.
-Note how we can call the constructor of \texttt{Shape} inside
-the overridden constructor.
-A limitation of this class system is that the overridden constructor
-must be a subtype of the original constructor,
-so the type of \texttt{radius} has to be quite permissive.
+       const "radius":() -> (number) }>
+\end{verbatim}
 
@@
 In all examples, if we erase all type and \texttt{const} annotations,
-and type declarations, they become valid Lua code, which run with the
-same semantics as Typed Lua code.
+they become valid Lua code, which run with the same semantics as Typed Lua code.
 

### 2015-03-17 doc/thesis/system.tex
Minor improvements to Chapter 4

@@
 
-In this chapter we present the formalization of Typed Lua's type system.
+In this chapter we present the abstract syntax of Typed Lua types,
+the subtyping rules, and the most interesting typing rules.
 Besides its practical contributions, Typed Lua also has some interesting
@@
 They are novel type system features that let Typed Lua cover several Lua idioms
-and features, such as refinement of tables, multiple assignment, multiple return values,
-and optional parameters.
+and features, such as refinement of tables, multiple assignment, and multiple return values.
 
@@
 
-Union types $T_{1} \cup T_{2}$ represent types that can hold a value
+Union types $T_{1} \cup T_{2}$ represent data types that can hold a value
 of two different types.
@@
 variables inside conditions.
-In Section \ref{sec:fap} we show in more detail how our type system
-uses them to formalize the \texttt{type} predicates that we mentioned
-in Section \ref{sec:unions}.
+Our type system uses filter types to formalize the \texttt{type} predicates
+that we mentioned in Section \ref{sec:unions}.
+This means that \texttt{type} predicates use filter types of the form
+$\phi(T_{1},T_{2})$ to discriminate local variables that are bound to
+union types.
+In a filter type $\phi(T_{1},T_{2})$, $T_{1}$ is the original type and
+$T_{2}$ is the discriminated type.
 
@@
 unions of tuple types into unions of first-level types.
-In Section \ref{sec:fap} we show in more detail how our type system
+In Section \ref{sec:fap} we will show in more detail how our type system
 uses them as a mechanism for handling unions of tuple types,
-when they appear on the right-hand side of the declaration of local variables,
+when they appear in the right-hand side of the declaration of local variables,
 as we mentioned in Section \ref{sec:unions}.
@@
 not overcome this limitation, as it still would give to the table constructor
-a regular table type without covariant mutable fields.
+a \emph{closed} table type without covariant mutable fields.
 Thus, programmers would not be able to use the table constructor to
@@
 In the next section we will show in more detail how our type system
-uses these tags to handle the refinement of table types and idiomatic
-object-oriented code.
+uses these tags to handle the refinement of table types.
 
@@
 $id$ are variable names, $T$ are first-level types, and $S$ are second-level types.
-The notation $\overline{id{:}T}$ denotes the list $id_{1}{:}T_{1}, ..., id_{n}{:}T_{n}$,
-where $n$ can be zero.
+The notation $\overline{id{:}T}$ denotes the list $id_{1}{:}T_{1}, ..., id_{n}{:}T_{n}$.
 
@@
 environment with types $\Integer$ and $\String$.
-Let us see how our type system type checks the following assignment
+Let us see how our type system type checks the following assignment:
 \[
@@
 expected in the left-hand side.
-Our example uses the rule \textsc{T-EXPLIST1} which is the rule
-that type checks an expression list where all expressions can
-only produce a single value:
+Our example uses the rule \textsc{T-EXPLIST1} to type check
+the right-hand side of the assignment.
+The rule \textsc{T-EXPLIST1} is the rule that type checks an
+expression list where all expressions can only produce a single value:
 \[
@@
 \mylabel{T-EXPLIST1}\\
-\dfrac{\env_{i}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
+\dfrac{\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
        \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
@@
 
-Then, our type system type checks the list of left-hand values.
+After type checking the right-hand side, our type system type checks the list
+of left-hand values.
 In our example, the left-hand side of the assignment has type 
@@
 expected in the left-hand side.
-Our example uses the rule \textsc{T-LHSLIST} which is the rule that type
+Our example uses the rule \textsc{T-LHSLIST} to type
 checks a list of left-hand values:
@@
 \mylabel{T-LHSLIST}\\
-\dfrac{\env_{i}, \penv \vdash l_{i}:T_{i}, \env_{i+1} \;\;\;
+\dfrac{\env_{1}, \penv \vdash l_{i}:T_{i}, \env_{i+1} \;\;\;
        \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
@@
 that handle the cases where function applications are either statements
-or expressions that produce only one value.
+that produce no value or expressions that produce only one value.
 The rule \textsc{T-STMAPPLY1} discards the produced values,
@@
 Our type system also catches arity mismatch.
+To do that, we end the input type of a funcion with type $\Void$
+instead of $\Value{*}$.
 For instance, let us assume that $f$ has type
@@
 In the implementation we fix the aliasing of \emph{unique} and \emph{open}
-table types that appear in a return statement, and in other cases we
+table types that appear in a top-level return statement, and in other cases we
 close the aliasing of \emph{unique} and \emph{open} table types.
@@
 In case of success, it returns two integers: the result and the remainder.
-In case of failure, it returns $\Nil$ plus an error message that indicates
+In case of failure, it returns $\Nil$ plus an error message that describes
 the error.
@@
 in the projection environment $\penv$.
-Rule \textsc{T-LOCAL2} captures this intuition:
+Rule \textsc{T-LOCAL2} illustrates this intuition:
 \[
@@
 into the union type $\Integer \cup \String$.
-However, we want to discriminate these variables to check whether
+Now, we may want to discriminate these variables to check whether
 the function call returned with success.

### 2015-03-18 doc/thesis/evaluation.tex
Minor improvements to the evaluation results

@@
 \hline
-Lua Standard Library & 62\% & 6\% & 9\% & 23\% & 129 \\ % 100%
+Lua Standard Library & 64\% & 5\% & 8\% & 23\% & 129 \\ % 100%
 \hline
@@
 of the case studies that we used Typed Lua for typing their members.
+An exported member is any Lua value that a module might export.
 We split the members of each case study into four categories:
@@
 The \emph{hard} category shows the percentage of members that
-we had to rely on the dynamic type because they rely on reflection.
+we could not give a precise static type.
 The last column of the table shows the total number of members
@@
 \hline
-base & 35\% & 12\% & 8\% & 45\% & 26 \\ % 100%
+base & 35\% & 4\% & 8\% & 53\% & 26 \\ % 100%
 \hline
@@
 \hline
-io & 71\% & 0\% & 10\% & 19\% & 21 \\ % 100%
+io & 81\% & 0\% & 0\% & 19\% & 21 \\ % 100%
 \hline
@@
 as the \emph{poly} category shows.
-This is the case of \texttt{ipairs}, \texttt{rawget}, and
-\texttt{rawset}.
+This is the case of \texttt{ipairs}.
 
@@
 
-In the case of \texttt{tonumber}, the type of the first parameter
-depends on the type of the second parameter.
+In the case of \texttt{tonumber}, it has two different types:
+\texttt{(value) -> (number)} and \texttt{(string, integer) -> (number)}.
+This means that the type of the first parameter depends on the
+type of the second parameter.
 For instance, we can call \texttt{tonumber(1)}, but we cannot
 call \texttt{tonumber(1,2)}.
-More precisely, the first argument of \texttt{tonumber} can be
+Note that the first argument of \texttt{tonumber} can be
 a value of any type if it is the only argument, but it must
@@
 could use to describe control transfers with continuations.
-However, for now, coroutines are out of the scope of our type
-system, and we use an empty \emph{userdata} declaration
-to represent the type \emph{thread}.
+However, for now, coroutines are out of scope of our type
+system, and we use an empty \texttt{userdata} declaration
+to represent the type \texttt{thread}.
 
@@
 of the \texttt{package} module, but we could not give precise static
-to the following tables:
+types to the following tables that it exports:
 \texttt{package.loaded}, \texttt{package.preload}, and \texttt{package.loaders}.
@@
 These functions that rely on format strings are in the \emph{hard} category.
-
-Lua programmers often use the functions of the \texttt{string} module
-in an object-oriented style.
-For instance, \texttt{string.len("foo")} and \texttt{("foo"):len()}
-produce the same result.
-We did not include the typing of the \texttt{string} methods in our
-evaluation results, because we treat this feature as a special case
-in the implementation.
 
@@
 the calling arity.
-More precisely, calling \texttt{table.insert(l, v)} inserts the
+That is, calling \texttt{table.insert(l, v)} inserts the
 element \texttt{v} at the end of the list \texttt{l},
@@
 at the position \texttt{p} of the list \texttt{l},
-but generates a run-time error when \texttt{p} is out of bounds.
+and generates a run-time error when \texttt{p} is out of bounds.
 This function also does not follow the semantics of Lua on
@@
 In the case of \texttt{math.abs},
-calling \texttt{math.abs(1)} returns the integer \texttt{1},
-while \texttt{math.abs(1.1)} returns the floating point \texttt{1.1}.
-Even though we gave this function the type \texttt{(number) -> (number)},
+it has two different types: \texttt{(integer) -> (integer)} and
+\texttt{(number) -> (number)}.
+Even though we gave this function the more general type \texttt{(number) -> (number)},
 it is not precise enough because the return type is always \texttt{number}
@@
 
-In the case of \texttt{math.random}, its type depends on the calling arity.
+In the case of \texttt{math.random}, it also has two different types:
+\texttt{() -> (number)} and \texttt{(integer, integer?) -> (integer)}.
+This means that the type of \texttt{math.random} depends on the calling arity.
 Calling \texttt{math.random()} returns a random floating point
@@
 while the explicit operations are methods of a file descriptor.
-We used an \emph{userdata} declaration to introduce the type
+We used an \texttt{userdata} declaration to introduce the type
 \texttt{file} for representing the type of a file descriptor
@@
 We could give precise static types to most of the members of the
-\texttt{io} module, but its categories \emph{over} and \emph{hard}
-show that it includes several members that we could not give a precise
-static type.
-While the function \texttt{io.close} and the method \texttt{file:close}
-are in the \emph{over} category,
-the functions \texttt{io.read} and \texttt{io.lines} are
+\texttt{io} module, but the \emph{hard} category shows that it
+includes some members that we could not give a precise static type.
+The functions \texttt{io.read} and \texttt{io.lines} are
 in the \emph{hard} category along with the methods
 \texttt{file:read} and \texttt{file:lines}.
-
-We could not precisely type \texttt{io.close} because it is
-an overloaded function.
-This means that its return type depends on whether the file handle
-that is being closed was created with \texttt{io.popen}.
-Calling \texttt{io.close(f)} returns the tuple \texttt{(boolean?, string, number)}
-when \texttt{f} was created with \texttt{io.popen},
-but it returns the tuple \texttt{(boolean)?} otherwise.
-The method \texttt{file:close} has the same issue.
 
@@
 The \texttt{mime} module offers low-level and high-level filters
-that apply and remove some text encoding.
+that apply and remove some text encodings.
 The low-level filters are written in C, while the high-level filters
@@
 It contains just the \texttt{lfs} module that is written in C,
-so we used Typed Lua's description file to type it.
+so we used Typed Lua's description files to type it.
 Table \ref{tab:evallfs} summarizes the evaluation results for LuaFileSystem.
@@
 Even though we could not precisely type the interface that \texttt{http-digest}
-exports, we could use only static types to annotate it, and they pointed a bug
-in the code.
+exports, we could use only static types to annotate this module,
+and they pointed a bug in the code.
 The problem was related to the way the library was loading the MD5 library
@@
 the lack of type parameters.
-This means that we get warnings inside an \texttt{ipairs} iteration because
+This means that we get warnings inside an \texttt{ipairs} iteration, because
 all iterated elements have the dynamic type.
@@
 
-The interface that the \texttt{tlchecker} and \texttt{tlcode}
-modules export are also straightforward to type.
+We could type the interfaces that modules \texttt{tlchecker} and
+\texttt{tlcode} export.
 The former traverses the AST to perform type checking,
@@
 A way to overcome this limitation was to predeclare these functions
-with an empty body, and then redeclare them with they actual body.
+with an empty body, and then redeclare them with their actual body.
 The first declaration specifies the function type, while the
@@
 a way to discriminate unions of table types, as we mentioned in Section \ref{sec:alias}.
+Without a way to discriminate a union of table types, any attempt
+to index this union of table types would generate a warning.
 

### 2015-03-18 doc/thesis/conclusions.tex
Very minor improvements to related work and conclusions

@@
 Before starting the design of Typed Lua, we realized that we should
-not rely only on the semantics of Lua, as this could lead to a
+not rely on the semantics of Lua only, as this could lead to a
 cumbersome type system that would not support several idiomatic Lua code.
@@
 For instance, ZeroBrane Studio is an IDE for Lua development that is
-starting to use Typed Lua to perform static analysis on Lua code.
+starting to use Typed Lua to perform static analysis in Lua code.
 

### 2015-03-18 doc/thesis/related.tex
Very minor improvements to related work and conclusions

@@
 Currently, Ravi extends the Lua Virtual Machine to support few types:
-\texttt{int}, \texttt{double}, arrays of integers, and arrays of doubles.
+\texttt{integer}, \texttt{number}, arrays of integers, and arrays of numbers.
 
@@
 \citet{allende2013cis} perform a carefull evaluation about
-cast insertion on Gradualtalk.
+cast insertion in Gradualtalk.
 They report that usually cast insertions impact on execution

### 2015-03-19 doc/thesis/typedlua.tex
Removing the notation for fixed table types

@@
   \textit{uniontuple} \; \texttt{`|'} \; \textit{uniontuple}\\
-\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'} \; | \;
-  \texttt{`<\{'} \; [\textit{tabletypebody}] \; \texttt{`\}>'}\\
+\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
 \textit{tabletypebody} ::= & \;\; \textit{maptype} \; | \;
@@
 they need a variable that has a very specific type.
-In Section \ref{sec:tables} we will see in more detail that literal
-types are essential to type Lua tables.
+In Section \ref{sec:tables} we will see in more detail that literal types
+are essential to type Lua tables.
 
@@
 
+Typed Lua always includes \texttt{nil*} in the end of the type
+of an expression list that does not end in a varidic type.
+This behavior preserves the semantics of Lua on replacing missing values,
+and it is necessary when we omit optional parameters in a function call.
+We will discuss optional parameters in the next section.
+
 We can also type variadic functions.
@@
 
-Typed Lua does not allow assignments to variables that hold a projection type,
-unless the union has been discriminated down to a single value.
+Typed Lua does not allow assignments to variables that hold a projection type.
 Unrestricted assignment to these variables would be unsound,
@@
 
-This example shows that we cannot write a function that is
+This example means that we cannot write a function that is
 guaranteed to return a number when we pass a number and
@@
 \begin{align*}
-\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'} \; | \;
-  \texttt{`<\{'} \; [\textit{tabletypebody}] \; \texttt{`\}>'}\\
+\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
 \textit{tabletypebody} ::= & \;\; \textit{maptype} \; | \;
@@
 in Section \ref{sec:statistics}.
-
-Even though Typed Lua includes two different forms for annotating
-a table, in this section we only discuss the form
-$\texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}$.
-We call this form \emph{closed} table type because it
-represents table types that do not provide
-any guarantees about keys with types not listed in the table type.
-In Sections \ref{sec:modules} and \ref{sec:oop} we will show that
-we use the syntax $\texttt{`<\{'} \; [\textit{tabletypebody}] \; \texttt{`\}>'}$
+Later in this chapter we will show that we use the syntax of records
 to type modules and objects.
-We call this form \emph{fixed} table type because it
-represents tables with no keys that do not inhabit one of the table's
-key types, and can have safe aliases.
 We will also show that we need the \texttt{const} modifier while
 typing object-oriented code.
-In the next chapter we formalize and discuss in more detail the
-different table types that our type system has.
 
@@
 include the type \texttt{nil} is to overload the programmers,
-as they need to use \texttt{or} or \texttt{if} to narrow the type of
-the elements they are accessing.
+as they need to use the logical \texttt{or} operator or the \texttt{if}
+statement to narrow the type of the elements they are accessing.
 
@@
 syntax of records to infer the following table type to \texttt{days}:
-\texttt{\{1:string, 2:string, 3:string, 4:string, 5:string, 6:string, 7:string\}}.
+\begin{verbatim}
+    { 1:string, 2:string, 3:string, 4:string,
+      5:string, 6:string, 7:string }
+\end{verbatim}
 
@@
 raises a warning when we try to access an index that is out of bounds.
-In the previous example, both \texttt{days[8]} and \texttt{days[i]}
-raise warnings, as both the literal type \texttt{8} and the base type
-\texttt{integer} do not map to any value.
+In the previous example, the expressions \texttt{days[8]} and \texttt{days[i]}
+would raise warnings if we had used the records syntax,
+as both the literal type \texttt{8} and the base type
+\texttt{integer} would not map to any value.
 
@@
 If a variable gets its type from a table constructor then it is
-\emph{unique}, otherwise it can be either \emph{fixed} or
-\emph{closed}.
-In this section we discuss the cases that produce \emph{closed}
-table types, while in Sections \ref{sec:modules} and \ref{sec:oop}
-we will discuss the cases that produce \emph{fixed} table types
-to describe the type of modules and classes.
+\emph{unique}, otherwise it is \emph{closed}.
 Any alias to \emph{unique} variables makes them \emph{open}, as we use the
@@
 because it has no alias.
+In Section \ref{sec:oop} we will show that we use the tag
+\emph{fixed} to describe the type of classes.
 Typed Lua also has different subtyping rules that handle these
@@
 
-We can create aliases to \emph{open} variables, but they are always
-\emph{closed}.
+We can create aliases to \emph{open} variables, but they are always \emph{closed}.
 Any mutation in the original reference and in the aliased reference
@@
 add new fields.
-For mutable fields this means that the type of the field cannot
-change after it is added to the type of the table.
+For mutable fields this means that the type of a field cannot
+change after it is added to the type of a table.
 In our previous example, the second line changes the type of
@@
 provided by the statically typed code.
+
+In Section \ref{sec:refinement} we present the formalization of the
+rules that type check the refinement of table types.
 
@@
 As an example, the following declaration defines the type \texttt{Person}
-as an alias to the \emph{closed} table type \texttt{\{"firstname":string, "lastname":string\}}
+as an alias to the table type \texttt{\{"firstname":string, "lastname":string\}}
 in the remainder of the current scope:
 \begin{verbatim}
-    local typealias Person = {"firstname":string,
-                              "lastname":string}
+    local typealias Person = {"firstname":string, "lastname":string}
 \end{verbatim}
@@
 Typed Lua also includes \emph{interfaces} as syntactic sugar to aliases
-for \emph{closed} table types that specify records, as writing table types can be unwieldy
+for table types that specify records, as writing table types can be unwieldy
 when records get bigger and the types of record fields get more complicated.
@@
 \texttt{interface} construct.
+In Section \ref{sec:oop} we will show that we can also use \emph{interface} to
+document the type of objects, and that \emph{methodtype} is syntactic
+sugar to express the type of methods.
 
@@
 \textit{typedec} ::= & \;\; \textit{Name} \; \{\textit{decitem}\} \; \textbf{end}\\
-\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{type}\\
+\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
+\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
 \textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
@@
 the type alias declaration, it has a more convenient syntax
-for declaring \emph{closed} table types that express records.
-For this reason, it worth having two different constructs;
+for declaring table types that express records.
+For this reason, it worth having two different constructs,
 one specifically for records and another for more general types.
@@
 
-We need an explicit type declaration on the return type because
+We need an explicit type declaration in the return type because
 the Typed Lua compiler cannot infer recursive types, though
@@
 \begin{verbatim}
-    <{ "deg":(number) -> (number),
+    { "deg":(number) -> (number),
       "rad":(number) -> (number),
-       "pow":(number, number) -> (number) }>
+      "pow":(number, number) -> (number) }
 \end{verbatim}
@@
 Typed Lua always fixes \emph{unique} and \emph{open} table types
-that appear in a return statement.
-Typed Lua has this behavior to ensure that modules and classes
-always implement \emph{fixed} table types.
+that appear in a top-level return statement.
+This means that modules have \emph{fixed} table types in Typed Lua.
+Typed Lua has this behavior to ensure that classes always implement
+\emph{fixed} table types.
 In the next section we will show that we build classes in a
@@
 that classes should have \emph{fixed} table types to allow safe
-inheritance.
+single inheritance.
 
@@
 In this example, Typed Lua uses the refinement of table types to
-allow the programmer to incrementally build the type of the module
-\texttt{mymath}.
+incrementally build the type of the module \texttt{mymath}.
 The variable \texttt{RADIANS\string_PER\string_DEGREE}
@@
 as Typed Lua uses this literal string to find the source file that
-implements a given module.
+implements the given module.
 To statically type check a module, Typed Lua follows the same rules
@@
 
-The way that Typed Lua handles modules using \emph{fixed} table types
-is also relevant for supporting object-oriented programming,
-as we discuss in the next section.
+The way that Typed Lua handles modules using \emph{fixed} table types is
+also relevant for supporting object-oriented programming, as we discuss in
+the next section.
 
@@
 In the case of method calls, the Lua compiler translates
-\texttt{obj:method(args)} into an operation that first evaluates
+\texttt{obj:method(args)} into an operation that evaluates
 \texttt{obj}, uses \texttt{method} to index \texttt{obj},
@@
 
-Typed Lua uses table types along with the type
+Typed Lua uses \emph{closed} table types along with the type
 \texttt{self} to represent objects.
@@
 In this example, Typed Lua binds the type \texttt{self} to
-the table type \texttt{\{"x":number, "y":number\}} inside \texttt{move}.
+the \emph{closed} table type \texttt{\{"x":number, "y":number\}} inside \texttt{move}.
 This example also uses the refinement of table types to build the
@@
 \texttt{setmetatable(\{\},\{\string_\string_index = id\})},
-if \texttt{id} has type \texttt{self} then the expression also also has type \texttt{self}.
+if \texttt{id} has type \texttt{self} then the expression also has type \texttt{self}.
 We will explain the other two rules while we explain how
@@
 because it is necessary for covariance among object types work.
-Finally, the last \texttt{return} statement fixes the type of
-\texttt{Shape} because \emph{fixed} table types do not allow width
+Finally, the top-level \texttt{return} statement fixes the type of
+\texttt{Shape}, because \emph{fixed} table types do not allow width
 subtyping, and this behavior allows Typed Lua to use the refinment
@@
 
-A return statement that appears in the top-level also exports
-a type alias to be used in type annotations.
-For instance, our example exports the following type alias:
-\begin{verbatim}
-    typealias Shape =
-      <{ "x":number, "y":number,
-         const "new":(self, number, number) -> (self),
-         const "move":(self, number, number) -> () }>
-\end{verbatim}
-
-The style of classes definition from the previous example
-allow us to use \texttt{require} for creating prototype objects that
-work as classes, along with an alias to the table type of the class.
-This allows us to use the exported alias in type annotations,
-as we show in the following example:
-\begin{verbatim}
-    local Shape = require "shape"
-    local shape1 = Shape:new(0, 5)
-    local shape2:Shape = Shape:new(10, 10)
-\end{verbatim}
-
-Typed Lua also includes a \texttt{class} construct that programmers
-can use to document the type of the classes they export..
-Figure \ref{fig:classes} shows the concrete syntax of the
-\texttt{class} construct.
-
-\begin{figure}[!ht]
-\textbf{Classes}\\
-\dstart
-\begin{align*}
-\textit{class} ::= & \;\; [\textbf{local}] \; \textbf{class} \; \textit{typedec}\\
-\textit{typedec} ::= & \;\; \textit{Name} \; \{\textit{decitem}\} \; \textbf{end}\\
-\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
-\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
-\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
-\textit{id} ::= & \;\; [\textbf{const}] \; \textit{Name}\\
-\textit{methodtype} ::= & \;\; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettype}
-\end{align*}
-\dend
-\caption{The concrete syntax of Typed Lua classes}
-\label{fig:classes}
-\end{figure}
-
-As an example, we can use the following \texttt{class} declaration to
-document the type \texttt{Shape} from the previous example:
-\begin{verbatim}
-    class Shape
+A return statement that appears in the top-level also exports an
+interface which we can use in type annotations.
+For instance, our previous example exports the following interface:
+\begin{verbatim}
+    interface Shape
       x, y:number
@@
 The double arrow can appear only inside the declaration of
-classes and userdata, which we will introduce in the next section.
-
-Typed Lua also uses \texttt{setmetatable} to simulate single inheritance.
+interfaces and userdata, which we will introduce in the next section.
+
+The style of classes definition from the previous example
+allows us to use \texttt{require} for creating prototype objects that
+work as classes, along with an alias to the object type.
+This allows us to use the exported alias in type annotations,
+as we show in the following example:
+\begin{verbatim}
+    local Shape = require "shape"
+    local shape1 = Shape:new(0, 5)
+    local shape2:Shape = Shape:new(10, 10)
+\end{verbatim}
+
+Note that \texttt{Shape} has a \emph{fixed} table type, because it
+describes the type of the class \texttt{Shape}, but \texttt{shape1}
+and \texttt{shape2} have \emph{closed} table types, because they
+describe the type of objects from the class \texttt{Shape}.
+Now, we will see that we need \emph{fixed} table types to allow
+\texttt{setmetatable} to simulate single inheritance.
 We use the following example to discuss single inheritance in Typed Lua:
@@
 existing methods in \texttt{Circle}.
+The third rule appears to redefine the constructor \texttt{new}.
+
 In a \texttt{setmetatable} expression
@@
 table types do not allow hiding fields.
-The third rule appears to redefine the constructor \texttt{new}.
+In our example, it is safe to use the type of \texttt{Shape}
+to initialize \texttt{Circle}, openning the type of \texttt{Circle}
+to allow the refinement of table types to build the type of
+the class \texttt{Circle}.
+
 In a \texttt{setmetatable} expression
-\texttt{setmetatable(e,\{\string_\string_index = id\})},
-if \texttt{e} has a \emph{fixed} table type \texttt{t}, \texttt{id} has
-type \texttt{self}, and the type bound to \texttt{self} is a subtype of
-\texttt{t}, then the expression has type \texttt{self}.
+\texttt{setmetatable(e:(el),\{\string_\string_index = id\})},
+if the expression \texttt{e:(el)} has a \emph{closed} table type \texttt{t},
+\texttt{id} has type \texttt{self}, and the type bound to \texttt{self}
+is a subtype of \texttt{t}, then the expression has type \texttt{self}.
 Note how we can use this rule to call the constructor of \texttt{Shape}
@@
 so the type of \texttt{radius} has to be quite permissive.
-
-Like in the preivous example, the last return statement in the
-top-level exports a \emph{fixed} table type, which we could
-document through the following class declaration:
-\begin{verbatim}
-    class Circle
+Also note that this is the only form of refinement that allows
+changing the type of a table field, if the new type is a subtype
+of the previous type.
+
+Like in the example that we introduced the class \texttt{Shape},
+in the example that we introduced the class \texttt{Circle},
+the top-level return statement exports an
+interface which we can use in type annotations:
+\begin{verbatim}
+    interface Circle
       x, y, radius:number
@@
 
-Note that this class definition introduces a syntactical overhead
-to manually include new and overriden methods.
-This class definition is equivalent to the alias that exports the
-following \emph{fixed} table type to the class \texttt{Circle}:
-\begin{verbatim}
-    <{ "x":number, "y":number, "radius":number,
-       const "new":(self, number, number, value) -> (self),
-       const "move":(self, number, number) -> (),
-       const "radius":() -> (number) }>
-\end{verbatim}
-
-The type of the class \texttt{Circle} is a subtype of the type of
-the class \texttt{Shape}, and we can use both exported aliases in type
-annotations, as we show in the following example:
+Even though the class \texttt{Circle} is not a subtype of the
+class \texttt{Shape}, because \emph{fixed} table types do not
+have width subtyping, the objects that have the exported type
+\texttt{Circle} are subtypes of the objects that have the
+exported type \texttt{Shape}, because \emph{closed} table types
+have width subtyping.
+We discuss \emph{fixed} and \emph{closed} table types in more
+detail in the next chapter.
+
+We can use both exported aliases in type annotations, as we
+show in the following example:
 \begin{verbatim}
@@
 
+\newpage
 \section{Description files}
@@
 \begin{verbatim}
+
     userdata md5_context
@@
 
-This example shows that the Typed Lua compiler searches for a
+The Typed Lua compiler searches for a
 description file when it cannot not find the respective Typed Lua file

### 2015-03-19 doc/thesis/system.tex
Fixing typos

@@
 Our type system also catches arity mismatch.
-To do that, we end the input type of a funcion with type $\Void$
+To do that, we end the input type of a function with type $\Void$
 instead of $\Value{*}$.
@@
 close the aliasing of \emph{unique} and \emph{open} table types.
-However, in the formalization we chose to define this behavior not
-deterministically, as it makes easier the presentation of this behavior.
+However, in the formalization we chose to define this behavior in
+a not deterministic way, as it makes easier the presentation of this behavior.
 

### 2015-03-19 doc/thesis/typedlua.tex
Fixing typos

@@
 For local functions, Typed Lua uses the type of the returned expression
-to assign a more specific type to an uannotated return type.
+to assign a more specific type to an unannotated return type.
 This means that we can rewrite the previous example as follows:
@@
 and \texttt{("hello", "world")} are consistent with \texttt{(string*)}.
-The call \texttt{v(...)} compiles withouth any warnings in both
+The call \texttt{v(...)} compiles without any warnings in both
 operation modes because the type of the argument list is \texttt{(string*)},
@@
 \texttt{Shape}, because \emph{fixed} table types do not allow width
-subtyping, and this behavior allows Typed Lua to use the refinment
+subtyping, and this behavior allows Typed Lua to use the refinement
 of table types to type check single inheritance, as we will see in
@@
 In our example, it is safe to use the type of \texttt{Shape}
-to initialize \texttt{Circle}, openning the type of \texttt{Circle}
+to initialize \texttt{Circle}, opening the type of \texttt{Circle}
 to allow the refinement of table types to build the type of

### 2015-03-22 doc/thesis/conclusions.tex
Improving conclusions

--- 2014-04-07 doc/thesis/conclusions.tex
@@
-Our main contribution is the design of a complete optional type system
-for Lua, a procedural scripting language.
-TypeScript is a complete optional type system for JavaScript,
-but it does not include union types, and it uses arrays to represent
-variadic functions and multiple return values.
-Gradualtalk is a complete gradual type system for Smalltalk,
-but Smalltalk is an object-oriented language that does
-not handle multiple return values, multiple assignments, and
-variadic functions.
-Gradualtalk also mixes nominal and structural typing because of
-the object-oriented nature of Smalltalk.
-Typed Racket is a complete static type system for Racket, a functional
-language.
-Typed Racket makes it easier to migrate from untyped to typed code,
-but the migration happens module-by-module and type annotations are
-mandatory on typed modules.
-There are research on the design of optional and gradual type systems
-for JavaScript, Perl, and Python, but they are either experimental or
-limited in their coverage.
-Furthermore, Typed Lua includes table types that use types as keys,
-instead of identifiers, and the fact that Typed Lua also includes
-literal types makes its table types subsume the usual idea of
-record types.
+In this work we presented Typed Lua, an optional type system for Lua.
+Typed Lua is a Lua extension that allows programmers to combine
+static and dynamic typing in Lua code, making easier the evolution
+of simple scripts into large programs.
 
-Although each language is different, we believe that several parts of
-the design of Typed Lua should be relevant to other scripting languages.
-Most notably, JavaScript shares with Lua the use of tests of type tags
-as a way of encoding optional and overloaded parameters, and the
-proliferation of class-based object systems built on top of the
-language's delegation mechanisms.
+Our main contribution is the formalization of a complete optional type
+system that introduces several novel type system features to statically
+type check some Lua idioms and features.
+Even though Lua shares several features with other dynamically
+typed languages such as JavaScript, Lua also has several unusual features.
+These unusual features include tables (or associative arrays) as the solely data
+structure mechanism, functions with multiple return values, and functions
+with flexible arity that interact with multiple assignment.
+We highlight the following novel features of our type system:
+\begin{itemize}
+\item the refinement of table types support the incremental evolution
+of record and object types, playing an important role to statically
+type check the idiomatic way that Lua programmers use tables to
+define modules and objects;
+\item projection types handle functions that are overloaded on the
+number of return values, allowing programmers to narrow the type of
+a local variable that depends on the type of another one;
+\item union types and variadic types help our type system handling
+functions with flexible arity, that is, union types are helpful to
+describe optional parameters while variadic types are helpful to
+describe the type of the vararg expression and the type of functions
+that can receive or return any number of values.
+\end{itemize}
 
-Typed Lua should also be a major contribution to the Lua community,
+A key feature in optional type systems is usability.
+This means that optional type systems should not change the idioms
+that programmers are already familiar with.
+Instead, optional type system should fit existing idioms to
+statically type check them.
+In other words, design a simple type system that only relies on
+the semantics and run-time tags of a dynamically typed language
+can overload programmers, forcing them to code in a different way.
+Thus, the most challenging aspect of optional type systems is
+to design a complex type system that feels natural to the programmers.
+
+Before starting the design of Typed Lua, we took usability into consideration.
+We realized that we should not rely on the semantics of Lua only,
+as this could lead to a cumbersome type system that would not support
+several idiomatic Lua code.
+For this reason, we performed a mostly automated survey of Lua idioms
+and features to inform our design choices.
+
+After designing and implementing Typed Lua, we performed several
+case studies to evaluate the usability of our type system.
+Our evaluation results showed that our type system can statically
+type check several Lua idioms and features, though the evaluation
+results also exposed several limitations of our type system.
+We found that the three main limitations of our type system are
+the lack of intersection types, parametric polymorphism, and operator overloading.
+Overcoming these limitations is our major target for future work,
+as it will allow us to statically type check more programs.
+
+Unlike other optional type systems, we designed Typed Lua without
+deliberated unsound parts.
+However, we still do not have proofs that the novel features of
+our type system are sound.
+We see a soundness proof as another major future work, as it is
+necessary to use static types for code optimization.
+
+Finally, we believe that Typed Lua is a major contribution to the Lua community,
 because it offers a framework that programmers can use to document,
@@
 libraries.
-
-Finally, Typed Lua can be a base for further research and development
-of static type systems for the Lua language, such as the use of static
-analysis of Lua API calls for checking that a native library is
-exporting the interface to Lua code that it declares, the use of
-effect types for typing Lua coroutines, and the use of static types
-for code optimization.
+In fact, we already have user feedback from Lua programmers that are
+using Typed Lua in their projects.
+For instance, ZeroBrane Studio is an IDE for Lua development that is
+starting to use Typed Lua to perform static analysis in Lua code.

### 2015-03-22 doc/thesis/system.tex
Improving conclusions

@@
 | \; k & \textit{nil and other literals}\\
-& | \; id & \textit{variable reading}\\
-& | \; e_{1}[e_{2}] & \textit{table reading}\\
+& | \; id & \textit{variable access}\\
+& | \; e_{1}[e_{2}] & \textit{table access}\\
 & | \; {<}T{>} \; id & \textit{type coercion}\\
@@
 l ::= & & \textsc{left-hand values:}\\
-& \;\; id_{l} & \textit{variable writing}\\
-& | \; e_{1}[e_{2}] & \textit{table writing}\\
+& \;\; id_{l} & \textit{variable assignment}\\
+& | \; e_{1}[e_{2}] & \textit{table assignment}\\
 & | \; id[k] \; {<}T{>} & \textit{type coercion}\\
@@
 \dfrac{\begin{array}{c}
-       \env_{i}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
+       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
        T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
        wf(T) \;\;\;
-       n = |\;\overline{[e_{1}] = e_{2}}\;|
+       n = |\;\overline{[e_{1}] = e_{2}}\;| \\
+       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) 
        \end{array}}
-      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{n+1}}
+      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{m}}
 \end{array}
@@
 \begin{verbatim}
+
     local person = {}

### 2015-03-22 doc/thesis/conclusions.tex
Adding a summary of the evalution results in the conclusions

@@
 In this work we presented Typed Lua, an optional type system for Lua.
-Typed Lua is a Lua extension that allows programmers to combine
-static and dynamic typing in Lua code, making easier the evolution
+We implemented Typed Lua as a Lua extension that allows programmers to
+combine static and dynamic typing in Lua code, making easier the evolution
 of simple scripts into large programs.
@@
 system that introduces several novel type system features to statically
-type check some Lua idioms and features.
+type check Lua programs.
 Even though Lua shares several features with other dynamically
 typed languages such as JavaScript, Lua also has several unusual features.
-These unusual features include tables (or associative arrays) as the solely data
-structure mechanism, functions with multiple return values, and functions
-with flexible arity that interact with multiple assignment.
+These unusual features include tables (or associative arrays) as the sole
+mechanism for structured data, besides functions with multiple return values
+and flexible arity that interact with multiple assignment.
 We highlight the following novel features of our type system:
 \begin{itemize}
-\item the refinement of table types support the incremental evolution
-of record and object types, playing an important role to statically
-type check the idiomatic way that Lua programmers use tables to
-define modules and objects;
+\item type refinement allows the incremental evolution of record and
+object types, playing an important role in statically type checking
+the idiomatic way in which Lua programmers use tables to define modules
+and objects;
 \item projection types handle functions that are overloaded on the
-number of return values, allowing programmers to narrow the type of
-a local variable that depends on the type of another one;
-\item union types and variadic types help our type system handling
-functions with flexible arity, that is, union types are helpful to
-describe optional parameters while variadic types are helpful to
-describe the type of the vararg expression and the type of functions
+number and types of return values, allowing programmers to narrow the
+types of a set of variables by narrowing the type of a single component
+of this set;
+\item union types and variadic types help our type system handle
+functions with flexible arity, that is, union types are helpful in
+describing optional parameters while variadic types are helpful in
+describing the type of the vararg expression and the type of functions
 that can receive or return any number of values.
@@
 that programmers are already familiar with.
-Instead, optional type system should fit existing idioms to
+Instead, optional type systems should fit existing idioms to
 statically type check them.
-In other words, design a simple type system that only relies on
-the semantics and run-time tags of a dynamically typed language
-can overload programmers, forcing them to code in a different way.
-Thus, the most challenging aspect of optional type systems is
-to design a complex type system that feels natural to the programmers.
+Designing a too simple type system can overload programmers by forcing
+them to change the way they program in the language to fit the type system,
+while designing a too complex type system can overload programmers with
+types and error messages that are hard to understand, even if type inference
+removes the necessity of annotating the program with these complex types.
+The most challenging aspect of designing optional type systems is to find
+the right amount of complexity for a type system that feels natural to the programmers.
 
-Before starting the design of Typed Lua, we took usability into consideration.
+Usability has been a concern in the design of Typed Lua since the beginning.
 We realized that we should not rely on the semantics of Lua only,
 as this could lead to a cumbersome type system that would not support
-several idiomatic Lua code.
+several Lua idioms.
 For this reason, we performed a mostly automated survey of Lua idioms
@@
 After designing and implementing Typed Lua, we performed several
-case studies to evaluate the usability of our type system.
+case studies to evaluate how successful we were in our goal of
+providing an usable type system.
+We evaluated 29 modules from 8 different case studies,
+and we could give precise static types to 83\% of the 449
+members that these modules export.
+In the median, we could give precise static types to 89\%
+of the members from each module.
 Our evaluation results showed that our type system can statically
@@
 Unlike other optional type systems, we designed Typed Lua without
-deliberated unsound parts.
+deliberate unsound parts.
 However, we still do not have proofs that the novel features of

### 2015-03-22 doc/thesis/system.tex
Minor fixes in the text
Adding some missing functions to the section about typing rules

@@
 \dend
-\caption{Typed Lua abstract syntax}
+\caption{The abstract syntax of Typed Lua}
 \label{fig:syntax}
@@
 projection types to handle overloaded return types.
+We can define \emph{proj} as follows:
+\begin{align*}
+proj(T_{1} \times ... \times T_{n}{*}, i) & = \left\{
+\begin{array}{ll}
+T_{i} & \text{if $i < n$}\\
+T_{i} \cup \Nil & \text{if $i = n$}
+\end{array} \right.
+\\
+proj(S_{1} \sqcup S_{2}, i) & = proj(S_{1}, i) \cup proj(S_{2}, i)
+\end{align*}
 
@@
 This rule uses the auxiliary function \emph{infer} to get the
-first-level types of each variable that should be introduced in
+most general first-level types of each variable that should be introduced in
 the type environment for type checking $s$.
@@
 before type checking $s$.
+We can define \emph{infer} as follows:
+\begin{align*}
+infer(T_{1} \times ... \times T_{n}{*}, i) & = \left\{
+\begin{array}{ll}
+general(T_{i}) & \text{if $i < n$}\\
+general(T_{n} \cup \Nil) & \text{if $i >= n$}
+\end{array} \right.
+\\ \\
+general(\False) & = \Boolean\\
+general(\True) & = \Boolean\\
+general({\it int}) & = \Integer\\
+general({\it float}) & = \Number\\
+general({\it string}) & = \String\\
+general(T_{1} \cup T_{2}) & = general(T_{1}) \cup general(T_{2})\\
+general(S_{1} \rightarrow S_{2}) & = general2(S_{1}) \rightarrow general2(S_{2})\\
+general(\{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{tag}) & = \{K_{1}{:}general(V_{1}), ..., K_{n}{:}general(V_{n})\}_{tag}\\
+general(\mu x.T) & = \mu x.general(T)\\
+general(T) & = T\\
+\\
+general2(\Void) & = \Void\\
+general2(T{*}) & = general(T){*}\\
+general2(T \times P) & = general(T) \times general2(P)\\
+general2(S_{1} \sqcup S_{2}) & = general2(S_{1}) \sqcup general2(S_{2})
+\end{align*}
 

### 2015-03-23 doc/thesis/typedlua.tex
Final touches before sending to juries

@@
 syntactic sugar for defining the type of methods,
-as it includes an implicit first type \texttt{self}.
+as it includes an implicit first input type \texttt{self}.
 The double arrow can appear only inside the declaration of
@@
 In a \texttt{setmetatable} expression
-\texttt{setmetatable(e:(el),\{\string_\string_index = id\})},
-if the expression \texttt{e:(el)} has a \emph{closed} table type \texttt{t},
+\texttt{setmetatable(e,\{\string_\string_index = id\})},
+if the expression \texttt{e} has a \emph{closed} table type \texttt{t},
 \texttt{id} has type \texttt{self}, and the type bound to \texttt{self}
@@
 must be a subtype of the original constructor,
-so the type of \texttt{radius} has to be quite permissive.
+so the type of the input parameter \texttt{radius} has to be quite permissive.
 Also note that this is the only form of refinement that allows

### 2015-04-22 doc/thesis_puc/abstract.tex
Adding abstract and appendix

--- 2015-01-08 doc/thesis/abstract.tex
@@
 when their simple scripts evolve into complex programs.
+
 Optional type systems allow combining dynamic and static typing in
@@
 already familiar with this language.
+
 In this work we present and formalize the design of Typed Lua,
@@
 that are not present in the type system of these languages.
-We discuss how Typed Lua handles features such as functions with flexible
-arity that interact with multiple assignment, functions that are overloaded
-on the number of return values, and the incremental evolution of record and object types.
-We also discuss our design decisions, and present the evaluation results
-that we achieved while using Typed Lua to type existing Lua code.
+These features include functions with flexible arity, multiple assignment,
+functions that are overloaded on the number of return values, and the
+incremental evolution of record and object types.
+We discuss how Typed Lua handles these features and our design decisions.
+Finally, we present the evaluation results that we achieved while using
+Typed Lua to type existing Lua code.

### 2015-04-22 doc/thesis_puc/resumo.tex
Adding abstract and appendix

--- 2015-02-03 doc/thesis/resumo.tex
@@
 quando os pequenos \emph{scripts} deles evoluem para programas complexos.
+
 Sistemas de tipos opcionais nos permitem combinar tipagem dinâmica e estática na
@@
 que já estão familiarizados com essa linguagem.
+
 Neste trabalho nós apresentamos e formalizamos Typed Lua, um sistema de tipos opcional
@@
 e características de Lua.
-Embora Lua compartilha várias características com outras linguagens dinamicamente
+Embora Lua compartilhe várias características com outras linguagens dinamicamente
 tipadas, em particular JavaScript, Lua também possui várias características não usuais,
 as quais não estão presentes nos sistemas de tipos dessas linguagens.
-Nós discutimos como Typed Lua tipa estaticamente características tais como
-funções com aridade flexível que interagem com atribuições múltiplas,
-funções que são sobrecarregadas no número de valores de retorno e a evolução
-incremental de estruturas e objetos.
-Discutimos também nossas decisões de projeto e apresentamos uma avaliação de
-resultados, a qual obtemos ao usar Typed Lua para tipar código Lua existente.
+Essas características incluem funções com aridade flexível, atribuições múltiplas,
+funções que são sobrecarregadas no número de valores de retorno e
+a evolução incremental de registros e objetos.
+Nós discutimos como Typed Lua tipa estaticamente essas características e
+também discutimos nossas decisões de projeto.
+Finalmente, apresentamos uma avaliação de resultados,
+a qual obtemos ao usar Typed Lua para tipar código Lua existente.

### 2015-04-22 doc/thesis_puc/abstract.tex
Minor fixes to abstracts

@@
 Optional type systems allow combining dynamic and static typing in
-the same language, without influencing its original semantics,
+the same language, without affecting its original semantics,
 making easier this code evolution from dynamic to static typing.
 Designing an optional type system for a dynamically typed language
-is challenging, as it should feels natural to programmers that are
+is challenging, as it should feel natural to programmers that are
 already familiar with this language.

### 2015-04-22 doc/thesis_puc/resumo.tex
Minor fixes to abstracts

@@
 Sistemas de tipos opcionais nos permitem combinar tipagem dinâmica e estática na
-mesma linguagem, sem influenciar a semântica original da linguagem, tornando mais
+mesma linguagem, sem afetar a semântica original da linguagem, tornando mais
 fácil a evolução de código tipado dinamicamente para código tipado estaticamente.
@@
 Finalmente, apresentamos uma avaliação de resultados,
-a qual obtemos ao usar Typed Lua para tipar código Lua existente.
+a qual obtivemos ao usar Typed Lua para tipar código Lua existente.

### 2015-04-22 doc/thesis_puc/evaluation.tex
Adding all chapters to PUC-Rio template

--- 2015-03-18 doc/thesis/evaluation.tex
@@
 
-The Lua Standard Library \citep{luamanual} was our first case study.
+The Lua Standard Library \cite{luamanual} was our first case study.
 We started to think about how we would type its modules at the same time
@@
 
-The second case study that we chose was the MD5 library \citep{lmd5},
+The second case study that we chose was the MD5 library \cite{lmd5},
 because we wanted a simple case study to introduce Typed Lua's description
@@
 
-LuaSocket \citep{luasocket} and LuaFileSystem \citep{luafilesystem} were
+LuaSocket \cite{luasocket} and LuaFileSystem \cite{luafilesystem} were
 the third and fourth case studies that we used to evaluate Typed Lua.
@@
 We also randomly selected three case studies from the LuaRocks
-repository, they are: HTTP Digest \citep{luahttpdigest},
-Typical \citep{luatypical}, and Mod 11 \citep{luamod11}.
+repository, they are: HTTP Digest \cite{luahttpdigest},
+Typical \cite{luatypical}, and Mod 11 \cite{luamod11}.
 The first provides client side HTTP digest authentication for Lua.
@@
 evaluation results, because it provides several functions that violate
-basic assumptions about Lua code \citep{luamanual}.
+basic assumptions about Lua code \cite{luamanual}.
 For instance, we can use the function \texttt{debug.setlocal} to change the value
@@
 could not give a precise static type for this reason.
-Lua has one-shot delimited continuations \citep{james2011yield}
-in the form of \emph{coroutines} \citep{moura2009rc}, and
-effect systems \citep{nielson1999type} are an approach that we
+Lua has one-shot delimited continuations \cite{james2011yield}
+in the form of \emph{coroutines} \cite{moura2009rc}, and
+effect systems \cite{nielson1999type} are an approach that we
 could use to describe control transfers with continuations.
@@
 support for SMTP, HTTP, and FTP client protocols, MIME encoding,
-URL manipulation, and LTN12 filters \citep{nehab2008ltn012}.
+URL manipulation, and LTN12 filters \cite{nehab2008ltn012}.
 We used Typed Lua's description files to type both parts, as we also
@@
 this requires more complex types such as the refinement types from
-hybrid type checking \citep{flanagan2006htc}.
+hybrid type checking \cite{flanagan2006htc}.
 
@@
 the Typed Lua parser and the description file parser.
-This module is hard to type because it uses LPeg \citep{lpeg,ierusalimschy2009lpeg}
+This module is hard to type because it uses LPeg \cite{lpeg,ierusalimschy2009lpeg}
 patterns, and LPeg uses overloaded arithmetic operators to

### 2015-04-22 doc/thesis_puc/related.tex
Adding all chapters to PUC-Rio template

--- 2015-03-18 doc/thesis/related.tex
@@
 
-Metalua \citep{metalua} is a Lua compiler that supports compile-time
+Metalua \cite{metalua} is a Lua compiler that supports compile-time
 metaprogramming (CTMP).
 CTMP is a kind of macro system that allows the programmers to interact
-with the compiler \citep{fleutot2007contrasting}.
+with the compiler \cite{fleutot2007contrasting}.
 Metalua extends Lua 5.1 syntax to include its macro system,
@@
 
-MoonScript \citep{moonscript} is a programming language that supports
+MoonScript \cite{moonscript} is a programming language that supports
 class-based object-oriented programming.
@@
 
-LuaInspect \citep{luainspect} is a tool that uses MetaLua to perform
+LuaInspect \cite{luainspect} is a tool that uses MetaLua to perform
 some code analysis.
@@
 
-Tidal Lock \citep{tidallock} is a prototype of another optional type
+Tidal Lock \cite{tidallock} is a prototype of another optional type
 system for Lua, which is written in Metalua.
@@
 
-Sol \citep{sol} is an experimental optional type system for Lua.
+Sol \cite{sol} is an experimental optional type system for Lua.
 Its type system is similar to ours, as it includes literal types,
@@
 
-Lua Analyzer \citep{luaanalyzer} is an optional type system for Lua
+Lua Analyzer \cite{luaanalyzer} is an optional type system for Lua
 that is specially designed to work in the Löve Studio,
@@
 
-Luacheck \citep{luacheck} is a tool that performs static analysis on Lua code.
+Luacheck \cite{luacheck} is a tool that performs static analysis on Lua code.
 It can flag access to undeclared globals and unused local variables,
@@
 
-Ravi \citep{ravi} is an experimental Lua dialect.
+Ravi \cite{ravi} is an experimental Lua dialect.
 Ravi introduces optional static typing for Lua to improve run-time performance.
@@
 
-Typed Racket \citep{tobin-hochstadt2008ts} is a statically typed version
+Typed Racket \cite{tobin-hochstadt2008ts} is a statically typed version
 of the Racket language, which is a Scheme dialect.
@@
 The main feature of Typed Racket's type system is \emph{occurrence typing}
-\citep{tobin-hochstadt2010ltu}.
+\cite{tobin-hochstadt2010ltu}.
 It is a novel way to use type predicates in control flow statements
@@
 As these kinds of checks are common in other languages, related systems
-have appeared \citep{guha2011tlc,winther2011gtp,pearce2013ccf}.
+have appeared \cite{guha2011tlc,winther2011gtp,pearce2013ccf}.
 
@@
 
-Typed Clojure \citep{bonnaire-sergeant2012typed-clojure} is an
+Typed Clojure \cite{bonnaire-sergeant2012typed-clojure} is an
 optional type system for Clojure.
@@
 
-Dart \citep{dart} is a new class-based object-oriented programming
+Dart \cite{dart} is a new class-based object-oriented programming
 language.
@@
 
-TypeScript \citep{typescript} is a JavaScript extension
+TypeScript \cite{typescript} is a JavaScript extension
 that includes optional type annotations and class-based
@@
 Even though TypeScript is unsound by design,
-\citet{bierman2014typescript} shows how to make TypeScript sound.
+Bierman and associates \cite{bierman2014typescript} shows how to
+make TypeScript sound.
 They use a reduced core of TypeScript to formalize a
@@
 
-TeJaS \citep{lerner2013tejas} is a framework for the construction of
+TeJaS \cite{lerner2013tejas} is a framework for the construction of
 different type systems for JavaScript.
@@
 
-\citet{politz2012semantics} proposes semantics and types for objects
-with first-class member names, a well-known feature from scripting languages.
+Politz and associates \cite{politz2012semantics} proposes semantics
+and types for objects with first-class member names, a well-known
+feature from scripting languages.
 Their type system uses string patterns to describe the members of
@@
 
-Gradualtalk \citep{allende2013gts} is a Smalltalk dialect that
+Gradualtalk \cite{allende2013gts} is a Smalltalk dialect that
 supports gradual typing.
@@
 typed code does not violate statically typed code.
-\citet{allende2013cis} perform a carefull evaluation about
-cast insertion in Gradualtalk.
+Allende and associates \cite{allende2013cis} perform a careful
+evaluation about cast insertion in Gradualtalk.
 They report that usually cast insertions impact on execution
@@
 
-Reticulated Python \citep{vitousek2014deg} is a Python compiler
+Reticulated Python \cite{vitousek2014deg} is a Python compiler
 that supports gradual typing.
@@
 
-Mypy \citep{mypy} is an optional type system for Python.
+Mypy \cite{mypy} is an optional type system for Python.
 The type system of mypy is similar to the type system of
@@
 Recently, Guido van Rossum, Python's author, proposed a
-standard syntax for type annotations in Python \citep{PEP483}
-that is extremely inspired by mypy \citep{PEP484}.
+standard syntax for type annotations in Python \cite{PEP483}
+that is extremely inspired by mypy \cite{PEP484}.
 The main goal of this proposal is to make easier building
 static analysis tools for Python.
-Typing \citep{typing} is a tool that is being developed to
+Typing \cite{typing} is a tool that is being developed to
 implement this proposal.
 
-Hack \citep{hack} is a new programming language that runs on the
+Hack \cite{hack} is a new programming language that runs on the
 Hip Hop Virtual Machine (HHVM).
@@
 
-The Ruby Type Checker \citep{ren2013rtc} is a library that
+The Ruby Type Checker \cite{ren2013rtc} is a library that
 performs type checking during run-time.
@@
 
-Grace \citep{black2013sg} is an object-oriented language
+Grace \cite{black2013sg} is an object-oriented language
 with optional typing.
@@
 static and dynamic typing.
-\citet{homer2013modules} explores some useful patterns
-that derive from Grace's use of objects as modules
+Homer and associates \cite{homer2013modules} explores some
+useful patterns that derive from Grace's use of objects as modules
 and its brand of optional structural typing, which

### 2015-04-22 doc/thesis_puc/review.tex
Adding all chapters to PUC-Rio template

--- 2015-03-15 doc/thesis/review.tex
@@
 
-Common LISP \citep{steele1982ocl} introduced optional type annotations
+Common LISP \cite{steele1982ocl} introduced optional type annotations
 in the early eighties, but not for static type checking.
@@
 
-\citet{abadi1989dts} extended the simply typed lambda calculus with the
-\texttt{Dynamic} type and the \texttt{dynamic} and \texttt{typecase}
-constructs, with the aim to safely integrate dynamic code in
-statically typed languages.
+Abadi and associates \cite{abadi1989dts} extended the simply typed
+lambda calculus with the \texttt{Dynamic} type and the \texttt{dynamic}
+and \texttt{typecase} constructs, with the aim to safely integrate dynamic
+code in statically typed languages.
 The \texttt{Dynamic} type is a pair \texttt{(v,T)} where \texttt{v} is a
@@
 
-The \emph{quasi-static} type system proposed by \citet{thatte1990qst}
+The \emph{quasi-static} type system proposed by Thatte \cite{thatte1990qst}
 performs implicit coercions and run-time checks to replace the
 \texttt{dynamic} and \texttt{typecase} constructs that were proposed by
-\citet{abadi1989dts}.
+Abadi and associates \cite{abadi1989dts}.
 To do that, quasi-static typing relies on subtyping with a top type
@@
 
-\emph{Soft typing} \citep{cartwright1991soft} is another approach
+\emph{Soft typing} \cite{cartwright1991soft} is another approach
 to combine static and dynamic typing in the same language.
@@
 
-\emph{Dynamic typing} \citep{henglein1994dts} is an approach
+\emph{Dynamic typing} \cite{henglein1994dts} is an approach
 that optimizes code from dynamically typed languages by eliminating
@@
 
-\citet{findler2002chf} proposed contracts for higher-order functions
-and blame annotations for run-time checks.
+Findler and Felleisen \cite{findler2002chf} proposed contracts for
+higher-order functions and blame annotations for run-time checks.
 Contracts perform dynamic type checking instead of static type checking,
@@
 
-BabyJ \citep{anderson2003babyj} is an object-oriented language
+BabyJ \cite{anderson2003babyj} is an object-oriented language
 without inheritance that allows programmers to incrementally annotate
@@
 
-\citet{ou2004dtd} specified a language that combines static types
-with dependent types.
+Ou and associates \cite{ou2004dtd} specified a language that combines
+static types with dependent types.
 To ensure safety, the compiler automatically inserts coercions
@@
 
-Optional type systems \citep{bracha2004pluggable} are an approach for
+Optional type systems \cite{bracha2004pluggable} are an approach for
 plugging static typing in dynamically typed languages.
@@
 same semantic rules and the type system serves only for discarding
-programs that may have undesired behaviors \citep{bracha2004pluggable}.
+programs that may have undesired behaviors \cite{bracha2004pluggable}.
 
-Strongtalk \citep{bracha1993strongtalk,bracha1996strongtalk} is
+Strongtalk \cite{bracha1993strongtalk,bracha1996strongtalk} is
 a version of Smalltalk that comes with an optional type system.
@@
 
-\emph{Pluggable type systems} \citep{bracha2004pluggable} generalize
+\emph{Pluggable type systems} \cite{bracha2004pluggable} generalize
 the idea of optional type systems that Strongtalk put in practice.
@@
 
-Dart \citep{dart} and TypeScript \citep{typescript} are new
+Dart \cite{dart} and TypeScript \cite{typescript} are new
 languages that are designed with an optional type system.
@@
 In the next section we will show that we can use some features
-of gradual typing \citep{siek2006gradual,siek2007objects} to
+of gradual typing \cite{siek2006gradual,siek2007objects} to
 formalize optional type systems.
@@
 
-The main goal of gradual typing \citep{siek2006gradual} is to allow
+The main goal of gradual typing \cite{siek2006gradual} is to allow
 programmers to choose between static and dynamic typing in the same
 language.
-To do that, \citet{siek2006gradual} extended the simply typed
-lambda calculus with the dynamic type $?$, as we can see in
+To do that, Siek and Taha \cite{siek2006gradual} extended the simply
+typed lambda calculus with the dynamic type $?$, as we can see in
 Figure \ref{fig:gtlc}.
@@
 dynamic type, and disallows conversions between inconsistent types
-\citep{siek2006gradual}.
+\cite{siek2006gradual}.
 For instance, $\Number \sim \;?$, $? \sim \Number$,
@@
 Gradual typing is similar to the approaches proposed by
-\citet{abadi1989dts} and \citet{thatte1990qst} by including a
-dynamic type in a statically typed language.
+Abadi and associates \cite{abadi1989dts} and Thatte \cite{thatte1990qst}
+by including a dynamic type in a statically typed language.
 However, these three approaches differ in the way they handle the
 dynamic type.
-While \citet{siek2006gradual} rely on the consistency relation,
-\citet{abadi1989dts} rely on type equality with explicit projections
-and injections, and \citet{thatte1990qst} relies on subtyping.
+While Siek and Taha \cite{siek2006gradual} rely on the consistency relation,
+Abadi and associates \cite{abadi1989dts} rely on type equality with explicit
+projections and injections, and Thatte \cite{thatte1990qst} relies on subtyping.
 
@@
 
-Later, \citet{siek2007objects} reported that the consistency relation is
-orthogonal to the subtyping relation, so we can combine them to achieve
+Later, Siek and Taha \cite{siek2007objects} reported that the consistency relation
+is orthogonal to the subtyping relation, so we can combine them to achieve
 the \emph{consistent-subtyping} relation, written $T_{1} \lesssim T_{2}$.
@@
 statically typed code, thus preserving type safety.
-To do that, \citet{siek2006gradual,siek2007objects} defined the
-run-time semantics of gradual typing as a translation to an
+To do that, Siek and Taha \cite{siek2006gradual,siek2007objects}
+defined the run-time semantics of gradual typing as a translation to an
 intermediate language with explicit casts at the frontiers between
@@
 The semantics of these casts is based on the higher-order contracts
-proposed by \citet{findler2002chf}.
+proposed by Findler and Felleisen \cite{findler2002chf}.
 
-\citet{herman2007sgt} showed that there is an efficiency
-concern regarding the run-time checks, because there are two
-ways that casts can lead to unbounded space consumption.
+Herman and associates \cite{herman2007sgt} showed that there is an
+efficiency concern regarding the run-time checks, because there are
+two ways that casts can lead to unbounded space consumption.
 The first affects tail recursion while the second appears when
@@
 repeated casts to the same function or object.
-\citet{herman2007sgt} use the coercion calculus outlined in
-\citet{henglein1994dts} to express casts as coercions and
-solve the problem of space efficiency.
+Herman and associates \cite{herman2007sgt} use the coercion calculus
+outlined by Henglein \cite{henglein1994dts} to express casts
+as coercions and solve the problem of space efficiency.
 Their approach normalizes an arbitrary sequence of coercions to a
@@
 from the real error.
-\citet{wadler2009wpc} developed \emph{blame calculus} as a way
-to handle this issue, and \citet{ahmed2011bfa} extended
-blame calculus with polymorphism.
+Wadler and Findler \cite{wadler2009wpc} developed \emph{blame calculus}
+as a way to handle this issue, and Ahmed and associates \cite{ahmed2011bfa}
+extended blame calculus with polymorphism.
 Blame calculus is an intermediate language to integrate
 static and dynamic typing along with the blame tracking approach
-proposed by \citet{findler2002chf}.
+proposed by Findler and Felleisen \cite{findler2002chf}.
 
@@
 on the other hand, it has the space efficiency problem reported
-by \citet{herman2007sgt}.
-Thus, \citet{siek2009casts} extended the coercion calculus outlined in
-\citet{herman2007sgt} with blame tracking to achieve an
-implementation of the blame calculus that is space efficient.
-After that, \citet{siek2010blame} proposed a new solution that also
-handles both problems.
+by Herman and associates \cite{herman2007sgt}.
+Thus, Siek and associates \cite{siek2009casts} extended the coercion
+calculus outlined by Herman and associates \cite{herman2007sgt} with
+blame tracking to achieve an implementation of the blame calculus that
+is space efficient.
+After that, Siek and Wadler \cite{siek2010blame} proposed a new solution
+that also handles both problems.
 This new solution is based on a concept called \emph{threesome},
@@
 programming languages.
-Reticulated Python \citep{reticulated,vitousek2014deg} is a research
+Reticulated Python \cite{reticulated,vitousek2014deg} is a research
 project that evaluates the costs of gradual typing in Python.
-Gradualtalk \citep{allende2013gts} is a gradually-typed Smalltalk
+Gradualtalk \cite{allende2013gts} is a gradually-typed Smalltalk
 that introduces a new cast insertion strategy for gradually-typed
-objects \citep{allende2013cis}.
-Grace \citep{black2012grace,black2013sg} is a new object-oriented,
+objects \cite{allende2013cis}.
+Grace \cite{black2012grace,black2013sg} is a new object-oriented,
 gradually-typed, educational language.
@@
 may have types and methods as attributes, and can have a state
-\citep{homer2013modules}.
-ActionScript \citep{moock2007as3} is one the first languages that
+\cite{homer2013modules}.
+ActionScript \cite{moock2007as3} is one the first languages that
 incorporated gradual typing to its implementation and
-Perl 6 \citep{tang2007pri} is also being designed with gradual typing,
+Perl 6 \cite{tang2007pri} is also being designed with gradual typing,
 though there is few documentation about the gradual type systems
@@
 checks.
-BabyJ \citep{anderson2003babyj} and Alore \citep{lehtosalo2011alore}
+BabyJ \cite{anderson2003babyj} and Alore \cite{lehtosalo2011alore}
 are two examples of object-oriented languages that have an
@@
 The consistent-or-subtype relation is different from the
-consistent-subtyping relation of \citet{siek2007objects}, but it is
-also written $T_{1} \lesssim T_{2}$.
+consistent-subtyping relation of Siek and Taha \cite{siek2007objects},
+but it is also written $T_{1} \lesssim T_{2}$.
 The consistent-or-subtype relation holds when $T_{1} \sim T_{2}$
@@
 gradual type systems.
-Typed Clojure \citep{bonnaire-sergeant2012typed-clojure} is an
+Typed Clojure \cite{bonnaire-sergeant2012typed-clojure} is an
 optional type system for Clojure that is now adopting the
@@
 
-\citet{flanagan2006htc} introduced \emph{hybrid type checking},
+Flanagan \cite{flanagan2006htc} introduced \emph{hybrid type checking},
 an approach that combines static types and \emph{refinement} types.
@@
 
-Sage \citep{gronski2006sage} is a programming language that
+Sage \cite{gronski2006sage} is a programming language that
 extends hybrid type checking with a dynamic type to
@@
 
-\citet{tobin-hochstadt2006ims} proposed another approach for gradually
-migrating from dynamically typed to statically typed code,
-and they coined the term \emph{from scripts to programs} for
-referring to this kind of interlanguage migration.
+Tobin-Hochstadt and Felleisen \cite{tobin-hochstadt2006ims} proposed
+another approach for gradually migrating from dynamically typed to
+statically typed code, and they coined the term
+\emph{from scripts to programs} for referring to this kind of
+interlanguage migration.
 In their approach, the migration from dynamically typed to
 statically typed code happens module-by-module, so they designed
-and implemented Typed Racket \citep{tobin-hochstadt2008ts} for
+and implemented Typed Racket \cite{tobin-hochstadt2008ts} for
 this purpose.
@@
 
-The approach used by \citet{tobin-hochstadt2008ts} to design and
-implement Typed Racket is probably also called gradual typing
+The approach used by Tobin-Hochstadt and Felleisen \cite{tobin-hochstadt2008ts}
+to design and implement Typed Racket is probably also called gradual typing
 because it allows the programmer to gradually migrate from untyped
@@
 We used the LuaRocks repository to build our statistics database;
-LuaRocks \citep{hisham2013luarocks} is a package manager for Lua
+LuaRocks \cite{hisham2013luarocks} is a package manager for Lua
 modules.

### 2015-04-22 doc/thesis_puc/system.tex
Adding all chapters to PUC-Rio template

--- 2015-03-22 doc/thesis/system.tex
@@
 
-Our type system uses subtyping \citep{cardelli1984smi,abadi1996to} to order
-types and consistent-subtyping \citep{siek2007objects,siek2013mutable}
+Our type system uses subtyping \cite{cardelli1984smi,abadi1996to} to order
+types and consistent-subtyping \cite{siek2007objects,siek2013mutable}
 to allow the interaction between statically and dynamically typed code.
@@
 
-We use the \emph{Amber rule} \citep{cardelli1986amber} to define
+We use the \emph{Amber rule} \cite{cardelli1986amber} to define
 subtyping between recursive types:
@@
 Our example uses the rule \textsc{T-LHSLIST} to type
-checks a list of left-hand values:
+check a list of left-hand values:
 \[
@@
 \begin{align*}
-proj(T_{1} \times ... \times T_{n}{*}, i) & = \left\{
-\begin{array}{ll}
-T_{i} & \text{if $i < n$}\\
-T_{i} \cup \Nil & \text{if $i = n$}
-\end{array} \right.
-\\
+proj(T_{1} \times ... \times T_{n} \times T{*}, i) & =  T_{i} \;\;\; \text{if $i <= n$}\\
 proj(S_{1} \sqcup S_{2}, i) & = proj(S_{1}, i) \cup proj(S_{2}, i)

### 2015-04-22 doc/thesis_puc/typedlua.tex
Adding all chapters to PUC-Rio template

--- 2015-03-23 doc/thesis/typedlua.tex
@@
 To do that, Typed Lua extends the syntax of Lua 5.3 to introduce
-optional type annotations, and performs local type inference \citep{pierce2000lti}
+optional type annotations, and performs local type inference \cite{pierce2000lti}
 to detect more precise types for unannotated expressions.
@@
 We use the consistent-subtyping relation of gradual typing
-\citep{siek2007objects,siek2013mutable} to formalize Typed Lua,
+\cite{siek2007objects,siek2013mutable} to formalize Typed Lua,
 though it does not insert run-time checks in the gradual typing style.
@@
 We did not insert run-time checks at this moment because they can decrease
-run-time performance \citep{allende2013cis}.
+run-time performance \cite{allende2013cis}.
 We believe that a careful evaluation of run-time checks should be done before
@@
 
-Unlike Dart \citep{dart} and TypeScript \citep{typescript},
+Unlike Dart \cite{dart} and TypeScript \cite{typescript},
 we are designing Typed Lua aiming soundness to make it possible to
@@
 Our consistent-subtyping relationship follows the standards defined
-by the gradual typing of objects \citep{siek2007objects,siek2013mutable}.
+by the gradual typing of objects \cite{siek2007objects,siek2013mutable}.
 In practice, we can pass a value of the dynamic type anytime we want
@@
 \texttt{string} when the predicate is true and \texttt{t} otherwise.
-This is a simplified form of \emph{flow typing} \citep{guha2011tlc,tobin-hochstadt2010ltu}.
+This is a simplified form of \emph{flow typing} \cite{guha2011tlc,tobin-hochstadt2010ltu}.
 As with \texttt{or}, the variable must be local to the function
@@
 
-\newpage
 \section{Description files}
@@
 
-The following example shows the description file for \texttt{lmd5} \citep{lmd5},
+The following example shows the description file for \texttt{lmd5} \cite{lmd5},
 a MD5 digest library for Lua that is written in C:

### 2015-04-24 doc/thesis_puc/review.tex
Adding some terms to be defined in the glossary

@@
 They use optional type annotations to perform compile-time type checking,
-though they do not influence the original run-time semantics
+though they do not affect the original run-time semantics
 of the language.
@@
 
-Gradual typing is similar to the approaches proposed by
-Abadi and associates \cite{abadi1989dts} and Thatte \cite{thatte1990qst}
-by including a dynamic type in a statically typed language.
+Gradual typing \cite{siek2006gradual} is similar to two previous
+approaches \cite{abadi1989dts,thatte1990qst}, because they also include
+a dynamic type in a statically typed language.
 However, these three approaches differ in the way they handle the
@@
 Abadi and associates \cite{abadi1989dts} rely on type equality with explicit
-projections and injections, and Thatte \cite{thatte1990qst} relies on subtyping.
+projections plus injections, and Thatte \cite{thatte1990qst} relies on subtyping.
 
@@
 When we combine consistency and subtyping, we are making subtyping handle
-what casts are safe among static types, and we are making consistency
+which casts are safe among static types, and we are making consistency
 handle the casts that involve the dynamic type $?$.
@@
 consistent-subtyping relation to statically check the interaction
-between statically and dynamically typed code, without influencing
+between statically and dynamically typed code, without affecting
 the run-time semantics.
@@
 In Grace, modules are gradually-typed objects, that is, modules
-may have types and methods as attributes, and can have a state
-\cite{homer2013modules}.
+may have types with methods as attributes, and they can also have
+a state \cite{homer2013modules}.
 ActionScript \cite{moock2007as3} is one the first languages that

### 2015-05-06 doc/thesis_puc/evaluation.tex
Fixing typos

@@
 returns the tuple \texttt{(-1, nil, "hello")}, because \texttt{-1} makes
-\texttt{socket.skip} does not skip any values.
+\texttt{socket.skip} not skip any values.
 As another example, \texttt{f = socket.skip(-2)} assigns \texttt{socket.skip}
@@
 Even though LPeg is the third most popular Lua module,
-we cannot not precisely type LPeg patterns because our
+we cannot precisely type LPeg patterns because our
 type system still does not support overloading arithmetic

### 2015-05-06 doc/thesis_puc/review.tex
Fixing typos

@@
 
-Abadi and associates \cite{abadi1989dts} extended the simply typed
+Abadi et al. \cite{abadi1989dts} extended the simply typed
 lambda calculus with the \texttt{Dynamic} type and the \texttt{dynamic}
@@
 \texttt{dynamic} and \texttt{typecase} constructs that were proposed by
-Abadi and associates \cite{abadi1989dts}.
+Abadi et al. \cite{abadi1989dts}.
 To do that, quasi-static typing relies on subtyping with a top type
@@
 
-Ou and associates \cite{ou2004dtd} specified a language that combines
+Ou et al. \cite{ou2004dtd} specified a language that combines
 static types with dependent types.
@@
 the dynamic type $?$, that is, $\lambda x.e$ is equivalent to $\lambda x{:}?.e$.
-Under these circumstances, we view gradual typing as a way to add
+Under these circumstances, we can view gradual typing as a way to add
 a dynamic type to statically typed languages.
@@
 The consistency relation is both reflexive and symmetric, but
-it is neither commutative nor transitive.
+it is not transitive.
 
@@
 While Siek and Taha \cite{siek2006gradual} rely on the consistency relation,
-Abadi and associates \cite{abadi1989dts} rely on type equality with explicit
+Abadi et al. \cite{abadi1989dts} rely on type equality with explicit
 projections plus injections, and Thatte \cite{thatte1990qst} relies on subtyping.
@@
 
-Herman and associates \cite{herman2007sgt} showed that there is an
+Herman et al. \cite{herman2007sgt} showed that there is an
 efficiency concern regarding the run-time checks, because there are
@@
 repeated casts to the same function or object.
-Herman and associates \cite{herman2007sgt} use the coercion calculus
+Herman et al. \cite{herman2007sgt} use the coercion calculus
 outlined by Henglein \cite{henglein1994dts} to express casts
@@
 Wadler and Findler \cite{wadler2009wpc} developed \emph{blame calculus}
-as a way to handle this issue, and Ahmed and associates \cite{ahmed2011bfa}
+as a way to handle this issue, and Ahmed et al. \cite{ahmed2011bfa}
 extended blame calculus with polymorphism.
@@
 on the other hand, it has the space efficiency problem reported
-by Herman and associates \cite{herman2007sgt}.
-Thus, Siek and associates \cite{siek2009casts} extended the coercion
-calculus outlined by Herman and associates \cite{herman2007sgt} with
+by Herman et al. \cite{herman2007sgt}.
+Thus, Siek et al. \cite{siek2009casts} extended the coercion
+calculus outlined by Herman et al. \cite{herman2007sgt} with
 blame tracking to achieve an implementation of the blame calculus that
@@
 7\% of the reads use non-literal expressions to index a table.
-It also worth mentioning that 45\% of the reads are actually
+It is worth mentioning that 45\% of the reads are actually
 function calls.
@@
 
-We also measured how many files have code that iterate over tables to
+We also measured how many files have code that iterates over tables to
 observe how frequently iteration is used.
-We observed that 23\% of the files have code that iterate over keys
+We observed that 23\% of the files have code that iterates over keys
 of any value, that is, the call to \texttt{pairs} appears at least
 once in these files (the median is twice per file);
-21\% of the files have code that iterate over integer keys, that is,
+21\% of the files have code that iterates over integer keys, that is,
 the call to \texttt{ipairs} appears at least once in these files

### 2015-05-06 doc/thesis_puc/system.tex
Fixing typos

@@
 \begin{array}{c}
-\mylabel{S-NUMBER}\\
+\mylabel{S-INTEGER}\\
 \senv \vdash \Integer \subtype \Number
@@
 s ::= & & \textsc{statements:}\\
-& \;\; \mathbf{skip} \;
-| \; s_{1} \; ; \; s_{2} & \textit{skip and sequence}\\
+& \;\; \mathbf{skip} & \textit{skip}\\
+& | \; s_{1} \; ; \; s_{2} & \textit{sequence}\\
 & | \; \overline{l} = el & \textit{multiple assignment}\\
@@
 | \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
-& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s \;
-| \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
+& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
+& | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
 & | \; \mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s & \textit{recursive function} \\
@@
 e ::= & & \textsc{expressions:}\\
-& \;\; \mathbf{nil} \;
-| \; k & \textit{nil and other literals}\\
+& \;\; \mathbf{nil} & \textit{nil}\\
+& | \; k & \textit{other literals}\\
 & | \; id & \textit{variable access}\\
@@
 \end{array}
-\;
+\\ \\
 \begin{array}{c}
 \mylabel{T-EXPLIST3}\\
-\dfrac{\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n}{*}, \env_{2}}
-      {\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n}{*}, \env_{2}}
+\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n}{*}, \env_{1}
 \end{array}

### 2015-05-06 doc/thesis_puc/typedlua.tex
Fixing typos

@@
 To be backwards compatible with Lua, the syntactic extensions introduced
-by Typed Lua does not include new reserved words.
+by Typed Lua do not include new reserved words.
 Appendix \ref{app:syntax} presents the complete syntax of Typed Lua
@@
 compile without any warnings.
-Local type inference always use the most general type.
+Local type inference always uses the most general type.
 In this example, the compiler does not use the literal type \texttt{7},
@@
     local function distance (x, y)
-      return abs(x - y)
+      return absolute(x - y)
     end
@@
 
-Lua has first-class functions, but they have some peculiarities.
+Lua has first-class functions, and they have some peculiarities.
 First, the number of arguments of a function call does
@@
 Even though these operation modes affect function calls, they do not
-affect other multiple assignment.
+affect other multiple assignments.
 For instance,
@@
 Typed Lua always includes \texttt{nil*} in the end of the type
-of an expression list that does not end in a varidic type.
+of an expression list that does not end in a variadic type.
 This behavior preserves the semantics of Lua on replacing missing values,
@@
 This means that we can use the syntax of records to type a
-list that has fixed elements, like the variable \texttt{days} from
-the previous example.
+list that has a fixed number of elements,
+like the variable \texttt{days} from the previous example.
 
@@
 
-    print(greeter(user1))             --> Goodbye Lewis Reed 
-    print(greeter(user2))             --> compile-time warning
-    print(greeter(user3))             --> Goodbye Lou Reed
-    print(greeter(user4))             --> compile-time warning
+    print(byebye(user1))             --> Goodbye Lewis Reed 
+    print(byebye(user2))             --> compile-time warning
+    print(byebye(user3))             --> Goodbye Lou Reed
+    print(byebye(user4))             --> compile-time warning
 \end{verbatim}
@@
 for declaring table types that express records.
-For this reason, it worth having two different constructs,
+For this reason, it is worth having two different constructs,
 one specifically for records and another for more general types.
@@
 Now we can write a program that declares a list, inserts some
-elements in this list, and then traverses the list printing
-each list element:
+elements in it, and then traverses the list printing
+each element:
 \begin{verbatim}
@@
 The Typed Lua compiler searches for a
-description file when it cannot not find the respective Typed Lua file
+description file when it cannot find the respective Typed Lua file
 that is the argument of \texttt{require}.

### 2015-05-09 doc/papers/paper/paper.tex
Adding starting files for a new paper

--- 2015-05-06 doc/thesis_puc/system.tex
@@
+\documentclass{sigplanconf}
+
+\usepackage[utf8]{inputenc}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{url}
+\usepackage{color}
+\usepackage{multirow}
+
+\newcommand{\Value}{\mathbf{value}}
+\newcommand{\Any}{\mathbf{any}}
+\newcommand{\Nil}{\mathbf{nil}}
+\newcommand{\Self}{\mathbf{self}}
+\newcommand{\False}{\mathbf{false}}
+\newcommand{\True}{\mathbf{true}}
+\newcommand{\Boolean}{\mathbf{boolean}}
+\newcommand{\Integer}{\mathbf{integer}}
+\newcommand{\Number}{\mathbf{number}}
+\newcommand{\String}{\mathbf{string}}
+\newcommand{\Void}{\mathbf{void}}
+\newcommand{\Const}{\mathbf{const}}
+
+\newcommand{\mylabel}[1]{\; (\textsc{#1})}
+\newcommand{\env}{\Gamma}
+\newcommand{\penv}{\Pi}
+\newcommand{\senv}{\Sigma}
+\newcommand{\subtype}{<:}
+\newcommand{\ret}{\rho}
+\newcommand{\self}{\sigma}
+
+\def\dstart{\hbox to \hsize{\vrule depth 4pt\hrulefill\vrule depth 4pt}}
+\def\dend{\hbox to \hsize{\vrule height 4pt\hrulefill\vrule height 4pt}}
+
+\begin{document}
+
+\special{papersize=8.5in,11in}
+\setlength{\pdfpageheight}{\paperheight}
+\setlength{\pdfpagewidth}{\paperwidth}
+
+\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country}
+\copyrightyear{20yy}
+\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
+\doi{nnnnnnn.nnnnnnn}
+
+\titlebanner{banner above paper title}
+\preprintfooter{short description of paper}
+
+\title{Optional typing for Lua}
+
+\authorinfo{André Murbach Maidl}
+           {PUC-Rio}
+           {amaidl@inf.puc-rio.br}
+\authorinfo{Fabio Mascarenhas}
+           {UFRJ}
+           {fabiom@dcc.ufrj.br}
+\authorinfo{Roberto Ierusalimschy}
+           {PUC-Rio}
+           {roberto@inf.puc-rio.br}
+
+\maketitle
+
+\begin{abstract}
+This is the text of the abstract.
+\end{abstract}
+
+\category{CR-number}{subcategory}{third-level}
+
+\terms
+term1, term2
+
+\keywords
+keyword1, keyword2
+
+\section{Introduction}
+
+Dynamically typed languages such as Lua avoid static types in favor of
+simplicity and flexibility, because the absence of static types means
+that programmers do not need to bother with abstracting types that
+should be validated by a type checker.
+Instead, dynamically typed languages use run-time \emph{type tags}
+to classify the values they compute, so their implementation can use
+these tags to perform run-time (or dynamic) type checking
+\cite{pierce2002tpl}.
+
+This simplicity and flexibility allows programmers to write code that
+might require a complex type system to statically type check,
+though it may also hide bugs that will be caught only after deployment
+if programmers do not properly test their code.
+In contrast, static type checking helps programmers detect many
+bugs during the development phase.
+Static types also provide a conceptual framework that helps
+programmers define modules and interfaces that can be combined to
+structure the development of programs.
+
+Thus, early error detection and better program structure are two
+advantages of static type checking that can lead programmers to
+migrate their code from a dynamically typed to a statically
+typed language, when their simple scripts evolve into complex programs
+\cite{tobin-hochstadt2006ims}.
+Dynamically typed languages certainly help programmers during the
+beginning of a project, because their simplicity and flexibility
+allows quick development and makes it easier to change code according to
+changing requirements.
+However, programmers tend to migrate from dynamically typed to
+statically typed code as soon as the project has consolidated its
+requirements, because the robustness of static types helps
+programmers link requirements to abstractions.
+This migration usually involves different languages that have
+different syntaxes and semantics, which usually requires a complete
+rewrite of existing programs instead of incremental evolution from
+dynamic to static types.
+
+Ideally, programming languages should offer programmers the
+option to choose between static and dynamic typing:
+\emph{optional type systems} \cite{bracha2004pluggable} and
+\emph{gradual typing} \cite{siek2006gradual} are two similar
+approaches for blending static and dynamic typing in the same
+language.
+The aim of both approaches is to offer programmers the option
+to use type annotations where static typing is needed,
+allowing the incremental migration from dynamic to static typing.
+The difference between these two approaches is the way they treat
+run-time semantics.
+While optional type systems do not affect run-time semantics,
+gradual typing uses run-time checks to ensure that dynamically typed
+code does not violate the invariants of statically typed code.
+
+Programmers and researchers sometimes use the term \emph{gradual typing}
+to mean the incremental evolution of dynamically typed code into
+statically typed code.
+For this reason, gradual typing may also refer to optional type
+systems and other approaches that blend static and dynamic typing to
+help programmers incrementally migrate from dynamic to static typing
+without having to switch to a different language, though all these
+approaches differ in the way they handle static and dynamic typing
+together.
+We use the term \emph{gradual typing} to refer to the work of
+Siek and Taha \cite{siek2006gradual}.
+
+In this work we present the design and evaluation of Typed Lua:
+an optional type system for Lua that is rich enough to
+preserve some of the Lua idioms that programmers are already familiar with,
+but that also includes new constructs that help programmers
+structure Lua programs.
+
+Lua is a small imperative language with first-class functions
+(with proper lexical scoping) where the only data structure
+mechanism is the \emph{table} --
+an associative array that can represent arrays, records, maps, modules, objects, etc.
+Tables also have syntactic sugar and metaprogramming support
+through operator overloading built into the language.
+Unlike other scripting languages, Lua has very limited coercion
+among different data types.
+
+Lua prefers to provide mechanisms instead of fixed policies due
+to its primary use as an embedded language for configuration and
+extension of other applications.
+This means that even features such as a module system and
+object orientation are a matter of convention instead of default
+language constructs.
+The result is a fragmented ecosystem of libraries, and different
+ideas among Lua programmers on how they should use the language
+features, or how they should structure programs.
+
+The lack of standard policies is a challenge for the design of
+an optional type system for Lua.
+For this reason, we are not relying entirely on the semantics of
+the language to design our type system.
+We also run a mostly automated survey of Lua idioms used in a
+large corpus of Lua libraries, which also has helped in the design of Typed Lua.
+
+So far, Typed Lua is a Lua extension that allows statically typed
+code to coexist and interact with dynamically typed code
+through optional type annotations.
+In addition, it adds default constructs that programmers can use
+to better structure Lua programs.
+The Typed Lua compiler warns programmers about type errors,
+but always generates Lua code that runs in unmodified Lua implementations.
+Programmers can enjoy some of the benefits of static types
+even without converting existing Lua modules to Typed Lua --
+they can export a statically typed interface to a dynamically typed module,
+and statically typed users of the module can use the Typed Lua compiler
+to check their use of the module.
+Thus, implementing an optional type system for Lua offers Lua
+programmers one way to obtain most of the advantages of static typing
+without compromising the simplicity and flexibility of dynamic typing.
+We have an implementation of the Typed Lua compiler that is
+available online\footnote{https://github.com/andremm/typedlua}.
+
+Typed Lua's intended use is as an application language, and
+we believe that policies for organizing a program in modules and writing
+object-oriented programs should be part of the language and
+checked by its optional type system.
+An application language is a programming language that helps
+programmers develop applications from scratch until these
+applications evolve into complex systems rather than just scripts.
+We will show that Typed Lua introduces the refinement of
+tables to support the common idioms that Lua programmers use
+to encode both modules and objects.
+
+We also believe that Typed Lua helps programmers give more
+formal documentation to already existing Lua code, as static types
+are also a useful source of documentation in languages that provide
+type annotations, because type annotations are always validated by
+the type checker and therefore never get outdated.
+Thus, programmers can use Typed Lua to define axioms about the
+interfaces and types of dynamically typed modules.
+We enforce this point by using Typed Lua to statically type
+the interface of the Lua standard library and other commonly used
+Lua libraries, so our compiler can check Typed Lua code that uses
+these libraries.
+
+Typed Lua performs a very limited form of local type inference
+\cite{pierce2000lti}, as static typing does not necessarily mean
+that programmers need to insert type annotations in the code.
+Several statically typed languages such as Haskell provide some
+amount of type inference that automatically deduces the types of
+expressions.
+Still, Typed Lua only requires a small amount of type annotations
+due to the nature of its optional type system.
+
+Typed Lua does not deal with code optimization, although another
+important advantage of static types is that they help the compiler
+perform optimizations and generate more efficient code.
+However, we believe that the formalization of our optional type
+system is precise enough to aid optimization in some Lua implementations.
+
+We use some of the ideas of gradual typing to formalize Typed Lua.
+Even though Typed Lua is an optional type system and thus does not
+include run-time checks between dynamic and static regions of the
+code, we believe that using the foundations of gradual typing to
+formalize our optional type system will allow us to include run-time
+checks in the future.
+
+Finally, we believe that designing an optional type system for Lua may
+shed some light on optional type systems for scripting languages
+in general, as Lua is a small scripting language that shares
+some features with other scripting languages such as JavaScript.
+
+%This work is split into seven chapters.
+%In Chapter \ref{chap:review} we review the literature about blending
+%static and dynamic typing in the same language, we discuss the differences
+%between optional type systems and gradual typing, and we also
+%present the results of our survey on Lua idioms.
+%In Chapter \ref{chap:typedlua} we use code examples to present the
+%design of Typed Lua.
+%In Chapter \ref{chap:system} we present our type system.
+%In Chapter \ref{chap:evaluation} we discuss the evaluation
+%results that we obtained while using Typed Lua to type existing Lua code.
+%In Chapter \ref{chap:related} we present some related work.
+%In Chapter \ref{chap:conc} we outline our contributions.
+
+\section{Statistics about the usage of Lua}
+
+In this section we present statistics about the usage of Lua
+features and idioms.
+We collected statistics about how programmers use tables, functions,
+dynamic type checking, object-oriented programming, and modules.
+We shall see that these statistics informed important design decisions
+on our optional type system.
+
+We used the LuaRocks repository to build our statistics database;
+LuaRocks \cite{hisham2013luarocks} is a package manager for Lua
+modules.
+We downloaded the 3928 \texttt{.lua} files that were available in
+the LuaRocks repository at February 1st 2014.
+However, we ignored files that were not compatible with Lua 5.2,
+the latest version of Lua at that time.
+We also ignored \emph{machine-generated} files and test files,
+because these files may not represent idiomatic Lua code,
+and might skew our statistics towards non-typical uses of Lua.
+This left 2598 \texttt{.lua} files from 262 different projects for
+our statistics database;
+we parsed these files and processed their abstract syntax tree
+to gather the statistics that we show in this section.
+
+To verify how programmers use tables, we measured how they
+initialize, index, and iterate tables.
+We present these statistics in the next three paragraphs to discuss
+their influence on our type system.
+
+The table constructor appears 23185 times.
+In 36\% of the occurrences it is a constructor that initializes a
+record (e.g., \texttt{\{ x = 120, y = 121 \}});
+in 29\% of the occurrences it is a constructor that initializes a
+list (e.g., \texttt{\{ "one", "two", "three", "four" \}});
+in 8\% of the occurrences it is a constructor that initializes a
+record with a list part;
+and in less than 1\% of the occurrences (4 times) it is a constructor
+that uses only the booleans \texttt{true} and \texttt{false} as indexes.
+At all, in 73\% of the occurrences it is a constructor that uses
+only literal keys;
+in 26\% of the occurrences it is the empty constructor;
+in 1\% of the occurrences it is a constructor with non-literal keys
+only, that is, a constructor that uses variables and function calls
+to create the indexes of a table;
+and in less than 1\% of the occurrences (19 times) it is a constructor
+that mixes literal keys and non-literal keys.
+
+The indexing of tables appears 130448 times:
+86\% of them are for reading a table field while
+14\% of them are for writing into a table field.
+We can classify the indexing operations that are reads as follows:
+89\% of the reads use a literal string key,
+4\% of the reads use a literal number key,
+and less than 1\% of the reads (10 times) use a literal boolean key.
+At all, 93\% of the reads use literals to index a table while
+7\% of the reads use non-literal expressions to index a table.
+It is worth mentioning that 45\% of the reads are actually
+function calls.
+More precisely, 25\% of the reads use literals to call a function,
+20\% of the reads use literals to call a method, that is,
+a function call that uses the colon syntactic sugar, 
+and less than 1\% of the reads (195 times) use non-literal expressions
+to call a function.
+We can also classify the indexing operations that are writes as follows: 
+69\% of the writes use a literal string key,
+2\% of the writes use a literal number key,
+and less than 1\% of the writes (1 time) uses a literal boolean key.
+At all, 71\% of the writes use literals to index a table while
+29\% of the writes use non-literal expressions to index a table.
+
+We also measured how many files have code that iterates over tables to
+observe how frequently iteration is used.
+We observed that 23\% of the files have code that iterates over keys
+of any value, that is, the call to \texttt{pairs} appears at least
+once in these files (the median is twice per file);
+21\% of the files have code that iterates over integer keys, that is,
+the call to \texttt{ipairs} appears at least once in these files
+(the median is also twice per file);
+and 10\% of the files have code that use the numeric \texttt{for}
+along with the length operator (the median is once per file).
+
+The numbers about table initialization, indexing, and iteration
+show us that tables are mostly used to represent records, lists,
+and associative arrays.
+Therefore, Typed Lua should include a table type for handling
+these uses of Lua tables.
+Even though the statistics show that programmers initialize tables
+more often than they use the empty constructor to
+dynamically initialize tables, the statistics of the empty
+constructor are still expressive and indicate that Typed Lua should
+also include a way to handle this style of defining table types.
+
+We found a total of 24858 function declarations in our database
+(the median is six per file).
+Next, we discuss how frequently programmers use dynamic type
+checking and multiple return values inside these functions.
+
+We observed that 9\% of the functions perform dynamic type checking
+on their input parameters, that is, these functions use \texttt{type}
+to inspect the tags of Lua values (the median is once per function).
+We randomly selected 20 functions to sample how programmers are
+using \texttt{type}, and we got the following data:
+50\% of these functions use \texttt{type} for asserting the tags of
+their input parameters, that is, they raise an error when the tag of a
+certain parameter does not match the expected tag, and
+50\% of these functions use \texttt{type} for overloading, that is,
+they execute different code according to the inspected tag.
+
+These numbers show us that Typed Lua should include union types,
+because the use of the \texttt{type} idiom shows that disjoint unions
+would help programmers define data structures that can hold a value of
+several different, but fixed types.
+Typed Lua should also use \texttt{type} as a mechanism for decomposing
+unions, though it may be restricted to base types only.
+
+We observed that 10\% of the functions explicitly return multiple
+values.
+We also observed that 5\% of the functions return \texttt{nil} plus
+something else, for signaling an unexpected behavior;
+and 1\% of the functions return \texttt{false} plus something else,
+also for signaling an unexpected behavior.
+
+Typed Lua should include function types to represent Lua functions,
+and tuple types to represent the signatures of Lua functions,
+multiple return values, and multiple assignments.
+Tuple types require some special attention, because Typed Lua
+should be able to adjust tuple types during compile-time, in a
+similar way to what Lua does with function calls and multiple
+assignments during run-time.
+In addition, the number of functions that return \texttt{nil} and
+\texttt{false} plus something else show us that overloading on the
+return type is also useful to the type system.
+
+We also measured how frequently programmers use the object-oriented
+paradigm in Lua.
+We observed that 23\% of the function declarations are actually
+method declarations.
+More precisely, 14\% of them use the colon syntactic sugar while
+9\% of them use \texttt{self} as their first parameter.
+We also observed that 63\% of the projects extend tables with
+metatables, that is, they call \texttt{setmetatable} at least once,
+and 27\% of the projects access the metatable of a given table,
+that is, they call \texttt{getmetatable} at least once.
+In fact, 45\% of the projects extend tables with metatables and
+declare methods:
+13\% using the colon syntactic sugar, 14\% using \texttt{self}, and
+18\% using both.
+
+Based on these observations, Typed Lua should include support
+to object-oriented programming.
+Even though Lua does not have standard policies for object-oriented
+programming, it provides mechanisms that allow programmers to
+abstract their code in terms of objects, and our statistics confirm
+that an expressive number of programmers are relying on these mechanisms
+to use the object-oriented paradigm in Lua.
+Typed Lua should include some standard way of defining interfaces and classes
+that the compiler can use to type check object-oriented code,
+but without changing the semantics of Lua.
+
+We also measured how programmers are defining modules.
+We observed that 38\% of the files use the current way of defining
+modules, that is, these files return a table that contains the
+exported members of the module at the end of the file;
+22\% of the files still use the deprecated way of defining modules,
+that is, these files call the function \texttt{module};
+and 1\% of the files use both ways.
+At all, 61\% of the files are modules while 39\% of the files are
+plain scripts.
+The number of plain scripts is high considering the origin of
+our database.
+However, we did not ignore sample scripts, which usually serve to
+help the users of a given module on how to use this module, and
+that is the reason why we have a high number of plain scripts.
+
+Based on these observations, Typed Lua should include a way
+for defining table types that represent the type of modules.
+Typed Lua should also support the deprecated style of module
+definition, using global names as exported members of the module.
+
+Typed Lua should also include some way to define the types of
+userdata.
+This feature should also allow programmers to define userdata
+that can be used in an object-oriented style, as this is another
+common idiom from modules that are written in C.
+
+The last statistics that we collected were about variadic functions
+and vararg expressions.
+We observed that 8\% of the functions are variadic, that is,
+their last parameter is the vararg expression.
+We also observed that 5\% of the initialization of lists
+(or 2\% of the occurrences of the table constructor) use solely the
+vararg expression.
+Typed Lua should include a \emph{vararg type} to handle variadic
+functions and vararg expressions.
+
+\section{Typing Rules}
+
 In the previous chapter we presented an informal overview of Typed Lua.
@@
 
-\section{Types}
-\label{sec:types}
-
-\begin{figure}[!ht]
+\subsection{Types}
+
+\begin{figure*}[!ht]
 \textbf{Type Language}\\
@@
 \label{fig:typelang}
-\end{figure}
+\end{figure*}
 
@@
 
-\section{Subtyping}
-\label{sec:subtyping}
+\subsection{Subtyping}
 
@@
 \mylabel{S-TABLE1}\\
-\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
+\dfrac{\begin{array}{c}
+       \forall i \in 1..n \; \exists j \in 1..m \\
        \senv \vdash K_{j} \subtype K_{i}' \;\;\;
        \senv \vdash K_{i}' \subtype K_{j} \;\;\;
-       \senv \vdash V_{j} \subtype_{c} V_{i}'}
+       \senv \vdash V_{j} \subtype_{c} V_{i}'
+       \end{array}}
       {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{fixed|closed} \subtype
@@
 \mylabel{S-TABLE6}\\
-\dfrac{\forall i \in 1..n \; \exists j \in 1..n \;\;\;
+\dfrac{\begin{array}{c}
+       \forall i \in 1..n \; \exists j \in 1..n \\
        \senv \vdash K_{j} \subtype K_{i}' \;\;\;
        \senv \vdash K_{i}' \subtype K_{j} \;\;\;
-       \senv \vdash V_{j} \subtype_{c} V_{i}'}
+       \senv \vdash V_{j} \subtype_{c} V_{i}'
+       \end{array}}
       {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{fixed} \subtype
@@
 \begin{align*}
-\{ & ``info":\Integer, \\
-   & ``next":\mu x.\{``info":\Integer,
-                     ``next":x \;\cup\; \Nil\}_{closed} \cup \Nil \}_{unique} \subtype \\
+\{ & ``info":\Integer, ``next":\mu x.\{ ``info":\Integer, \\
+   & ``next":x \;\cup\; \Nil\}_{closed} \cup \Nil \}_{unique} \subtype \\
 & \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}
@@
 
-\section{Typing rules}
-\label{sec:rules}
+\subsection{Typing rules}
 
@@
 
-\begin{figure}[!ht]
+\begin{figure*}[!ht]
 \textbf{Abstract Syntax}\\
@@
 \label{fig:syntax}
-\end{figure}
+\end{figure*}
 
@@
 
-\subsection{Assignment and function application}
-\label{sec:assignment}
+\subsubsection{Assignment and function application}
 
@@
 \mylabel{T-EXPLIST2}\\
-\dfrac{\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Void, \env_{2}}
-      {\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{2}}
+\dfrac{\begin{array}{c}
+       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \\
+       \env_{1}, \penv \vdash me:T_{n+1} \times ... \times \Void, \env_{n+2} \\
+       n = |\;\overline{e}\;| \;\;\;
+       \env_{m} = merge(\env_{1}, ..., \env_{n+2})
+       \end{array}}
+      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n+1} \times ... \times \Nil{*}, \env_{m}}
 \end{array}
@@
 \mylabel{T-EXPLIST3}\\
-\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n}{*}, \env_{1}
+\dfrac{\begin{array}{c}
+       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \\
+       \env_{1}, \penv \vdash me:T_{n+1} \times ... \times T{*}, \env_{n+2} \\
+       n = |\;\overline{e}\;| \;\;\;
+       \env_{m} = merge(\env_{1}, ..., \env_{n+2})
+       \end{array}}
+      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n+1} \times ... \times T{*}, \env_{m}}
 \end{array}
@@
 
-\subsection{Tables and refinement}
-\label{sec:refinement}
+\subsubsection{Tables and refinement}
 
@@
 
-\subsection{Projections}
-\label{sec:fap}
+\subsubsection{Projections}
 
@@
 they project to $\Integer \cup \Nil$ and $\Integer \cup \String$, respectively.
+
+\section{Evaluation}
+
+We performed some case studies on existing Lua libraries
+to evaluate the design of our type system.
+For each library, we used Typed Lua to either annotate its modules
+or to write statically typed interfaces to its modules through
+Typed Lua's description files.
+In this chapter we present our evaluation results and discuss some
+interesting cases.
+
+The Lua Standard Library \cite{luamanual} was our first case study.
+We started to think about how we would type its modules at the same time
+that we started to design our type system, as it could give us some
+hints on our type system.
+And it did: optional parameters and overloading on the return type
+are two Lua features that our type system should handle to allow us
+typing some of the functions that the standard library implements.
+
+The second case study that we chose was the MD5 library \cite{lmd5},
+because we wanted a simple case study to introduce Typed Lua's description
+files and \texttt{userdata} declarations.
+These Typed Lua's mechanisms allow programmers to give statically typed
+interfaces to Lua libraries.
+
+LuaSocket \cite{luasocket} and LuaFileSystem \cite{luafilesystem} were
+the third and fourth case studies that we used to evaluate Typed Lua.
+We chose them because they are the most popular Lua libraries.
+We wrote a script that builds the dependency graph of Lua libraries
+that are in the LuaRocks repository, and uses this dependency graph
+to identify the most popular Lua libraries.
+
+We also randomly selected three case studies from the LuaRocks
+repository, they are: HTTP Digest \cite{luahttpdigest},
+Typical \cite{luatypical}, and Mod 11 \cite{luamod11}.
+The first provides client side HTTP digest authentication for Lua.
+The second is an extension to the primitive function \texttt{type}.
+The third is a generator and checker of modulo 11 numbers.
+We randomly selected three case studies because we wanted to evaluate
+Typed Lua for annotating existing libraries that are written in Lua,
+as the previous case studies are mostly libraries that are written in C.
+
+The Typed Lua compiler is the last case study that we evaluated.
+We chose it as a case study because it is a large application.
+Besides, it is a case study that evaluates the evolution of a script
+to a program.
+
+We used these case studies to evaluate two aspects of Typed Lua:
+\begin{enumerate}
+\item how precisely it can describe the type of the interface of a module;
+\item whether it provides guarantees that the code matches the interface.
+\end{enumerate}
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|c|}
+\hline
+\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
+\hline
+Lua Standard Library & 64\% & 5\% & 8\% & 23\% & 129 \\ % 100%
+\hline
+MD5 & 100\% & 0\% & 0\% & 0\% & 13 \\ % 100%
+\hline
+LuaSocket & 89\% & 1\% & 2\% & 8\% & 123 \\ % 100%
+\hline
+LuaFileSystem & 89\% & 0\% & 11\% & 0\% & 19 \\ % 100%
+\hline
+HTTP Digest & 0\% & 0\% & 100\% & 0\% & 1 \\ % 100%
+\hline
+Typical & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
+\hline
+Modulo 11 & 78\% & 0\% & 0\% & 22\% & 9 \\ % 100%
+\hline
+Typed Lua Compiler & 93\% & 0\% & 1\% & 6\% & 154 \\ % 100%
+\hline
+\end{tabular}
+\end{center}
+\caption{Evaluation results for each case study}
+\label{tab:evalbycase}
+\end{table}
+
+Table \ref{tab:evalbycase} summarizes our evaluation results for each
+of the case studies that we used Typed Lua for typing their members.
+An exported member is any Lua value that a module might export.
+We split the members of each case study into four categories:
+\emph{easy}, \emph{poly}, \emph{over}, and \emph{hard}.
+The \emph{easy} category shows the percentage of members that
+we could give a precise static type.
+The \emph{poly} category shows the percentage of members that
+we made minimal use of the dynamic type as a replacement for the
+lack of type parameters.
+The \emph{over} category shows the percentage of members that
+we gave a static type that is not precise enough,
+as these members are overloaded functions that require intersection
+types to describe their precise static types.
+The \emph{hard} category shows the percentage of members that
+we could not give a precise static type.
+The last column of the table shows the total number of members
+of each case study.
+The results that we obtained for the \emph{easy} category give
+a lower bound on how much static type safety
+we could add to each one of our case studies.
+
+In the following sections we discuss each case study in more detail.
+For each case study, we split the evaluation results according
+to the modules that each one of them include.
+We use these split results to discuss the contributions and
+limitations of our type system.
+
+\subsection{Lua Standard Library}
+
+All of the modules in the standard library are implemented in C, so
+we used Typed Lua to type just the interface of each module.
+The \texttt{debug} module is the only one that we did not include in our
+evaluation results, because it provides several functions that violate
+basic assumptions about Lua code \cite{luamanual}.
+For instance, we can use the function \texttt{debug.setlocal} to change the value
+of a local variable that is not visible in the current scope.
+Table \ref{tab:evallsl} summarizes the evaluation results for the Lua Standard Library
+(version 5.3).
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|}
+\hline
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
+\hline
+base & 35\% & 4\% & 8\% & 53\% & 26 \\ % 100%
+\hline
+coroutine & 14\% & 0\% & 0\% & 86\% & 7 \\ % 100%
+\hline
+package & 62\% & 0\% & 0\% & 38\% & 8 \\ % 100%
+\hline
+string & 75\% & 0\% & 0\% & 25\% & 16 \\ % 100%
+\hline
+utf8 & 100\% & 0\% & 0\% & 0\% & 6 \\ % 100%
+\hline
+table & 14\% & 72\% & 14\% & 0\% & 7 \\ % 100%
+\hline
+math & 81\% & 0\% & 19\% & 0\% & 27 \\ % 100%
+\hline
+io & 81\% & 0\% & 0\% & 19\% & 21 \\ % 100%
+\hline
+os & 82\% & 0\% & 18\% & 0\% & 11 \\ % 100%
+\hline
+\end{tabular}
+\end{center}
+\caption{Evaluation results for Lua Standard Library}
+\label{tab:evallsl}
+\end{table}
+
+The \texttt{base} module was very difficult to type
+because it includes several functions that rely on reflection,
+as the \emph{hard} category shows.
+For instance, the functions \texttt{pairs} and \texttt{getmetatable}
+are in this category.
+While \texttt{pairs} traverses all keys and values that are stored
+in a given table, \texttt{getmetatable} returns the metatable of a
+given table.
+
+There are some functions in the \texttt{base} module that
+we could not give a precise static type because our type
+system does not have parametric polymorphism,
+as the \emph{poly} category shows.
+This is the case of \texttt{ipairs}.
+
+The \texttt{base} module also includes some overloaded functions,
+as the \emph{over} category shows.
+We could not type these functions because our type system does
+not include intersection types.
+This is the case of \texttt{tonumber} and \texttt{collectgarbage}.
+
+In the case of \texttt{tonumber}, it has two different types:
+\texttt{(value) -> (number)} and \texttt{(string, integer) -> (number)}.
+This means that the type of the first parameter depends on the
+type of the second parameter.
+For instance, we can call \texttt{tonumber(1)}, but we cannot
+call \texttt{tonumber(1,2)}.
+Note that the first argument of \texttt{tonumber} can be
+a value of any type if it is the only argument, but it must
+be a string if there is a second argument,
+which must be an integer.
+
+In the case of \texttt{collectgarbage}, the return type changes
+according to an input literal string.
+For instance, calling \texttt{collectgarbage("collect")} returns an integer,
+calling \texttt{collectgarbage("count")} returns a floating point,
+and calling \texttt{collectgarbage("isrunning")} returns a boolean.
+
+The \texttt{coroutine} module was also very difficult to type
+because our type system cannot describe the computational effects
+of a program.
+The \emph{hard} category shows the amount of functions that we
+could not give a precise static type for this reason.
+Lua has one-shot delimited continuations \cite{james2011yield}
+in the form of \emph{coroutines} \cite{moura2009rc}, and
+effect systems \cite{nielson1999type} are an approach that we
+could use to describe control transfers with continuations.
+However, for now, coroutines are out of scope of our type
+system, and we use an empty \texttt{userdata} declaration
+to represent the type \texttt{thread}.
+
+Still, we could give a precise static type to one function in
+the \texttt{coroutine} module, as the \emph{easy} category shows.
+The function \texttt{coroutine.isyieldable} has no input parameters,
+and it simply returns a boolean that indicates whether the running
+coroutine is yieldable.
+
+We could give precise static types to the constants and functions
+of the \texttt{package} module, but we could not give precise static
+types to the following tables that it exports:
+\texttt{package.loaded}, \texttt{package.preload}, and \texttt{package.loaders}.
+The first stores loaded modules, while the others store module \emph{loaders}.
+They are difficult to type because their types rely on reflection,
+that is, their types depend on the modules a program loads.
+For this reason, they are in the \emph{hard} category.
+
+We could give precise static types to most of the functions of the
+\texttt{string} module, but we could not give precise static types
+to the functions that rely on format strings.
+For instance, the type of the arguments that we pass to
+\texttt{string.format} must match the format string we are using.
+It is fine to call \texttt{string.format("\%d", 1)}, but
+\texttt{string.format("\%d", true)} raises a run-time error.
+These functions that rely on format strings are in the \emph{hard} category.
+
+The \texttt{utf8} module was straightforward to type,
+as its members are only operations over strings.
+
+The \texttt{table} module was specially difficult to type because
+most of its functions require parametric polymorphism,
+as the \emph{poly} category shows.
+These functions either receive or return a list of elements, and
+parametric polymorphism would help us to describe them with a generic type.
+
+However, the lack of parametric polymorphism did not prevent us from
+giving a precise type to one function of the \texttt{table} module,
+as the \emph{easy} category shows.
+We could give a precise static type to \texttt{table.concat},
+as it operates over lists where all elements are strings or numbers.
+
+Even if our type system had parametric polymorphism, there is
+still one function of the \texttt{table} module that we could
+not give a precise static type because it is an overloaded function,
+as the \emph{over} category shows.
+This function is \texttt{table.insert}, and its type depends on
+the calling arity.
+That is, calling \texttt{table.insert(l, v)} inserts the
+element \texttt{v} at the end of the list \texttt{l},
+while \texttt{table.insert(l, p, v)} inserts the element \texttt{v}
+at the position \texttt{p} of the list \texttt{l},
+and generates a run-time error when \texttt{p} is out of bounds.
+This function also does not follow the semantics of Lua on
+discarding extra arguments, and generates a run-time error whenever
+we pass more than three arguments, even if the first three arguments
+match its signature.
+
+Even though the \texttt{math} module looks straightforward to type,
+the \emph{over} category shows that it includes several overloaded functions.
+For instance, the functions \texttt{math.abs} and \texttt{math.random}
+are in this category.
+
+In the case of \texttt{math.abs},
+it has two different types: \texttt{(integer) -> (integer)} and
+\texttt{(number) -> (number)}.
+Even though we gave this function the more general type \texttt{(number) -> (number)},
+it is not precise enough because the return type is always \texttt{number}
+independently of the argument type.
+In other words, the return type should be \texttt{integer} when
+the argument type is also \texttt{integer}, and the return type
+should be \texttt{number} when the argument type is also \texttt{number}.
+
+In the case of \texttt{math.random}, it also has two different types:
+\texttt{() -> (number)} and \texttt{(integer, integer?) -> (integer)}.
+This means that the type of \texttt{math.random} depends on the calling arity.
+Calling \texttt{math.random()} returns a random floating point
+between \texttt{0} and \texttt{1}.
+Calling \texttt{math.random(10)} is equivalent to \texttt{math.random(1,10)},
+and returns an integer between this interval.
+Like \texttt{table.insert}, this function also does not follow the
+semantics of Lua on discarding extra arguments,
+and generates a run-time error whenever we pass more than two arguments.
+
+The \texttt{io} module provides operations for manipulating files,
+and these operations can use implicit or explicit file descriptors.
+The implicit operations are functions in the \texttt{io} table,
+while the explicit operations are methods of a file descriptor.
+We used an \texttt{userdata} declaration to introduce the type
+\texttt{file} for representing the type of a file descriptor
+and its methods.
+The evaluation results include both implicit and explicit operations.
+
+We could give precise static types to most of the members of the
+\texttt{io} module, but the \emph{hard} category shows that it
+includes some members that we could not give a precise static type.
+The functions \texttt{io.read} and \texttt{io.lines} are
+in the \emph{hard} category along with the methods
+\texttt{file:read} and \texttt{file:lines}.
+
+We could not precisely type \texttt{io.read} because its return type
+relies on format strings.
+For instance, calling \texttt{io.read("l")} returns a string or \texttt{nil},
+\texttt{io.read("n")} returns a number or \texttt{nil}, and
+\texttt{io.read("l", "n")} returns a string or \texttt{nil} and a number or \texttt{nil}.
+The function \texttt{io.lines}, and the methods \texttt{file:read} and
+\texttt{file:lines} have the same issue.
+
+There are two functions in the \texttt{os} module that we could
+not give a precise static type because they are overloaded functions.
+The functions that are in the \emph{over} category are
+\texttt{os.date} and \texttt{os.execute}.
+
+The evaluation results show that our type system should include
+intersection types, parametric polymorphism, and effect types,
+as these features would help us increase the static typing of the
+Lua Standard Library.
+Intersection types would allow us to define overloaded function types.
+Parametric polymorphism would allow us to define generic function and table types.
+Effect types would allow us to type coroutines.
+
+\subsection{MD5}
+
+The MD5 library is an OpenSSL based message digest library for Lua.
+It contains just the \texttt{md5} module that is written in C,
+so we used Typed Lua's description file to type it.
+Table \ref{tab:evalmd5} summarizes the evaluation results for MD5.
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|}
+\hline
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
+\hline
+md5 & 100\% & 0\% & 0\% & 0\% & 13 \\ % 100%
+\hline
+\end{tabular}
+\end{center}
+\caption{Evaluation results for MD5}
+\label{tab:evalmd5}
+\end{table}
+
+Even tough it was straightforward to type the MD5 library,
+we found a little difference between its documentation and its behavior.
+The documentation suggests that the type of the function \texttt{md5.update}
+is \texttt{(md5\string_context, string) -> (md5\string_context)},
+though there is a call to this function in the test script that passes
+an extra string argument.
+Reading the source code, we found that its actual type is
+\texttt{(md5\string_context, string*) -> (md5\string_context)},
+that is, we can pass zero or more strings to \texttt{md5.update}.
+
+This case study shows that type annotations help programmers maintain the
+documentation updated, as the type checker always validates them.
+
+\subsection{LuaSocket}
+
+LuaSocket is a library that adds network support to Lua,
+and it is split into two parts: a core that is written in C and a set of
+Lua modules.
+The C core provides TCP and UDP support, while the Lua modules provide
+support for SMTP, HTTP, and FTP client protocols, MIME encoding,
+URL manipulation, and LTN12 filters \cite{nehab2008ltn012}.
+We used Typed Lua's description files to type both parts, as we also
+wanted to use LuaSocket to test description files to statically type
+the interface of modules that are written in Lua.
+Table \ref{tab:evalsocket} summarizes the evaluation results for LuaSocket.
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|}
+\hline
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
+\hline
+socket & 83\% & 0\% & 0\% & 17\% & 60 \\ % 100%
+\hline
+ftp & 83\% & 0\% & 17\% & 0\% & 6 \\ % 100%
+\hline
+http & 80\% & 0\% & 20\% & 0\% & 5 \\ % 100%
+\hline
+smtp & 100\% & 0\% & 0\% & 0\% & 7 \\ % 100%
+\hline
+mime & 100\% & 0\% & 0\% & 0\% & 17 \\ % 100%
+\hline
+ltn12 & 95\% & 5\% & 0\% & 0\% & 20 \\ % 100%
+\hline
+url & 100\% & 0\% & 0\% & 0\% & 8 \\ % 100%
+\hline
+\end{tabular}
+\end{center}
+\caption{Evaluation results for LuaSocket}
+\label{tab:evalsocket}
+\end{table}
+
+We could give precise static types to most of the members
+in the \texttt{socket} module, which is the C core.
+However, this module includes some functions that
+we could not give a precise static type because they
+rely on reflection, as the \emph{hard} category shows.
+For instance, \texttt{socket.skip} is a function that is
+in this category.
+We can use this function to choose the number of
+values that we want to return.
+As an example, calling \texttt{socket.skip(1, nil, "hello")}
+returns only the string \texttt{"hello"}, because \texttt{1} indicates
+that we do not want to return the first value.
+Passing a negative number to \texttt{socket.skip} can be
+dangerous, as it returns anything that might be in the stack.
+As an example, calling \texttt{socket.skip(-1, nil, "hello")}
+returns the tuple \texttt{(-1, nil, "hello")}, because \texttt{-1} makes
+\texttt{socket.skip} not skip any values.
+As another example, \texttt{f = socket.skip(-2)} assigns \texttt{socket.skip}
+to \texttt{f}, as \texttt{-2} gets \texttt{socket.skip} from the stack.
+Our type system cannot handle the type of negative numbers, as
+this requires more complex types such as the refinement types from
+hybrid type checking \cite{flanagan2006htc}.
+
+We could give precise static types to most of the members of the
+modules \texttt{ftp} and \texttt{http}, but we could not precisely type
+two overloaded functions: \texttt{ftp.get} and \texttt{http.request}.
+
+The function \texttt{ftp.get} downloads data from a given URL,
+which can be either a string or a table.
+More precisely, \texttt{ftp.get(url)} returns the tuple
+\texttt{(string) | (nil, string)} if \texttt{url} is a string,
+and it returns the tuple \texttt{(number) | (nil, string)} if
+\texttt{url} is a table.
+
+The function \texttt{http.request} downloads data from a given URL,
+which can be either a string or a table.
+More precisely, \texttt{http.request(url, body)} returns the tuple
+\texttt{(string, number, \{string:string\}, number) | (nil, string)}
+if \texttt{url} is a string and \texttt{body} is another string or \texttt{nil},
+but it returns the tuple
+\texttt{(number, number, \{string:string\}, number) | (nil, string)}
+if \texttt{url} is a table and \texttt{body} is \texttt{nil}.
+
+The modules \texttt{mime} and \texttt{ltn12} have a strong connection.
+The \texttt{mime} module offers low-level and high-level filters
+that apply and remove some text encodings.
+The low-level filters are written in C, while the high-level filters
+use the function \texttt{ltn12.filter.cycle} along with the low-level
+filters to create standard filters.
+
+Even though we could type all the members of the \texttt{mime} module,
+the function \texttt{ltn12.filter.cycle} is the only member of the
+\texttt{ltn12} module that we could not give a precise type.
+This function is difficult to type because it is polymorphic.
+
+The modules \texttt{smtp} and \texttt{url} were straightforward to type.
+The \texttt{smtp} module provides functions that send e-mails.
+The \texttt{url} module provides functions that manipulate URLs.
+
+\subsection{LuaFileSystem}
+
+LuaFileSystem is a library that extends the set of functions
+for manipulating file systems in Lua.
+It contains just the \texttt{lfs} module that is written in C,
+so we used Typed Lua's description files to type it.
+Table \ref{tab:evallfs} summarizes the evaluation results for LuaFileSystem.
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|}
+\hline
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
+\hline
+lfs & 89\% & 0\% & 11\% & 0\% & 19 \\ % 100%
+\hline
+\end{tabular}
+\end{center}
+\caption{Evaluation results for LuaFileSystem}
+\label{tab:evallfs}
+\end{table}
+
+Even though we could precisely type most of the functions exported
+by the \texttt{lfs} module, we could not type two overloaded functions
+due to the lack of intersection types in our type system.
+
+\subsection{HTTP Digest}
+
+The HTTP Digest library implements client side HTTP digest authentication for Lua.
+Table \ref{tab:evalhttpdigest} summarizes the evaluation results for HTTP Digest.
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|}
+\hline
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
+\hline
+http-digest & 0\% & 0\% & 100\% & 0\% & 1 \\ % 100%
+\hline
+\end{tabular}
+\end{center}
+\caption{Evaluation results for HTTP Digest}
+\label{tab:evalhttpdigest}
+\end{table}
+
+It is difficult to type the interface of the \texttt{http-digest} module
+because it is an extension to the \texttt{http} module from LuaSocket.
+The \texttt{http-digest} module only exports the function
+\texttt{http-digest.request}, which extends the function
+\texttt{http.request} with MD5 authentication.
+Like \texttt{http.request}, \texttt{http-digest.request}
+is also an overloaded function.
+
+Even though we could not precisely type the interface that \texttt{http-digest}
+exports, we could use only static types to annotate this module,
+and they pointed a bug in the code.
+The problem was related to the way the library was loading the MD5 library
+that should be used. 
+This part of the code checks the existence of three different MD5 libraries
+in the system, and uses the first one that is available, or generates an
+error when none is available.
+The code that loads the first option was fine, but the code that loads the
+second and third options were trying to access an undefined global variable.
+
+\subsection{Typical}
+
+Typical is a library that extends the behavior of the function \texttt{type}. 
+Table \ref{tab:evaltypical} summarizes the evaluation results for Typical.
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|}
+\hline
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
+\hline
+typical & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
+\hline
+\end{tabular}
+\end{center}
+\caption{Evaluation results for Typical}
+\label{tab:evaltypical}
+\end{table}
+
+The interface of the \texttt{typical} module is straightforward to type,
+as it contains only the function \texttt{typical.type},
+which has the same type of the function \texttt{type}: \texttt{(value) -> (string)}.
+
+However, we hit some limitations of our type system while annotating this module.
+
+First, it uses the \texttt{getmetatable} to get a table and
+checks whether this table has the field \texttt{\string_\string_type}.
+We could not give a precise type to \texttt{getmetatable}, so we used the dynamic
+type \texttt{any} as its return type, and this generates a warning.
+
+Second, it returns a metatable that extends \texttt{\string_\string_call}
+with \texttt{typical.type}, that is, we can use the module itself as a function,
+though it is a table.
+Our type system still does not support metatables, so we did not extend our version
+of the \texttt{typical} module to support \texttt{\string_\string_call}.
+
+Third, the module uses \texttt{ipairs} to iterate over an array of functions,
+but our type system also has limited support to \texttt{ipairs}, and generates
+a warning when we try to use the indexed value inside the \texttt{for} body.
+As we mentioned in this chapter, we use the dynamic type as a replacement for
+the lack of type parameters.
+This means that we get warnings inside an \texttt{ipairs} iteration, because
+all iterated elements have the dynamic type.
+We removed this warning using the numeric \texttt{for} to perform the same loop.
+
+\subsection{Modulo 11}
+
+Modulo 11 is a library that generates and verifies modulo 11 numbers.
+Table \ref{tab:evalmod11} summarizes the evaluation results for Typical.
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|}
+\hline
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
+\hline
+mod11 & 78\% & 0\% & 0\% & 22\% & 9 \\ % 100%
+\hline
+\end{tabular}
+\end{center}
+\caption{Evaluation results for Modulo 11}
+\label{tab:evalmod11}
+\end{table}
+
+The \texttt{mod11} module was written using an object-oriented idiom that
+our type system does not support, and that is the reason why we could not
+type all the members of its interface.
+More precisely, the original code uses \texttt{setmetatable} to hide
+two attributes, which our type system cannot hide.
+
+In addition, it returns a metatable that extends \texttt{\string_\string_call}
+with the class constructor.
+This allows us to use the module itself to create new instances of a
+Modulo 11 number.
+However, our type system does not support this feature, and we need to
+make explicit calls to the constructor whenever we want to create a
+new instance.
+
+Even though we had these two issues to annotate the \texttt{mod11} module,
+we could use only static types to annotate it, and we found some interesting
+points.
+The code relies on implicit conversions between strings and numbers,
+and some parts of the code keep on changing the type of local variables.
+These are two practices that may hide bugs.
+
+\subsection{Typed Lua Compiler}
+
+The Typed Lua compiler is the last case study that we evaluated.
+Table \ref{tab:evaltlc} summarizes its evaluation results.
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|}
+\hline
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
+\hline
+tlast & 98\% & 0\% & 2\% & 0\% & 47 \\ % 100%
+\hline
+tltype & 100\% & 0\% & 0\% & 0\% & 65 \\ % 100%
+\hline
+tlst & 100\% & 0\% & 0\% & 0\% & 26 \\ % 100%
+\hline
+tllexer & 18\% & 0\% & 0\% & 82\% & 11 \\ % 100%
+\hline
+tlparser & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
+\hline
+tldparser & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
+\hline
+tlchecker & 100\% & 0\% & 0\% & 0\% & 2 \\ % 100%
+\hline
+tlcode & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
+\hline
+\end{tabular}
+\end{center}
+\caption{Evaluation results for Typed Lua Compiler}
+\label{tab:evaltlc}
+\end{table}
+
+The \texttt{tlast} module implements the Abstract Syntax Tree for
+the compiler.
+We could not precisely type just one function, because it has
+an overloaded type that requires intersection types.
+
+The \texttt{tltype} module implements the types introduced by Typed Lua.
+It also implements the subtyping and consistent-subtyping relations.
+The interface that this module exports was straightforward to type.
+
+The \texttt{tlst} module implements the symbols table for the compiler.
+The interface that this module exports was also straightforward to type.
+
+The \texttt{tllexer} module defines common lexical rules for
+the Typed Lua parser and the description file parser.
+This module is hard to type because it uses LPeg \cite{lpeg,ierusalimschy2009lpeg}
+patterns, and LPeg uses overloaded arithmetic operators to
+build LPeg patterns. 
+Even though LPeg is the third most popular Lua module,
+we cannot precisely type LPeg patterns because our
+type system still does not support overloading arithmetic
+operators.
+In the \texttt{tllexer} module, we could only give precise static types
+to two error reporting functions that it exports.
+
+The \texttt{tlparser} and \texttt{tldparser} modules implement the Typed Lua
+parser and the description file parser, respectively.
+Even though they use LPeg to implement the grammar rules,
+they only export a parsing function.
+Both use LPeg to parse a string and return the corresponding AST.
+
+We could type the interfaces that modules \texttt{tlchecker} and
+\texttt{tlcode} export.
+The former traverses the AST to perform type checking,
+while the latter traverses the AST to perform code generation.
+
+Even though we could precisely type the interface that most of
+the modules export, we had issues to write mutually recursive functions.
+This kind of functions often appear in compilers construction
+to traverse the data structures that they use.
+However, Typed Lua still does not support mutually recursive functions.
+A way to overcome this limitation was to predeclare these functions
+with an empty body, and then redeclare them with their actual body.
+The first declaration specifies the function type, while the
+second specifies what the function actually does without changing
+any type definition.
+
+Traversing the AST would also be problematic if we had not included
+a way to discriminate unions of table types, as we mentioned in Section \ref{sec:alias}.
+Without a way to discriminate a union of table types, any attempt
+to index this union of table types would generate a warning.
+
+Bootstraping the compiler also helped revealing some bugs.
+We found some accesses to undeclared global variables and also
+to undeclared table fields.
+The compiler also helped pointing the places where we should narrow
+a nilable value before using it.
+In fact, this point appeared in all the case studies that we used
+Typed Lua to annotate Lua code.
+This means that Lua programmers often use possibly \texttt{nil} values
+before checking whether it is \texttt{nil}.
+
+\section{Related Work}
+
+In this chapter we review related work, and we split it into two sections:
+in the first section we review other Lua projects,
+while in the second section we review other projects that are not related to Lua.
+
+\subsection{Other Lua projects}
+
+Metalua \cite{metalua} is a Lua compiler that supports compile-time
+metaprogramming (CTMP).
+CTMP is a kind of macro system that allows the programmers to interact
+with the compiler \cite{fleutot2007contrasting}.
+Metalua extends Lua 5.1 syntax to include its macro system,
+and allows programmers to define their own syntax.
+Metalua can provide syntactical support for several object-oriented
+styles, and can also provide syntax for turning simple type
+annotations into run-time assertions.
+
+MoonScript \cite{moonscript} is a programming language that supports
+class-based object-oriented programming.
+MoonScript compiles to idiomatic Lua code, but
+it does not perform compile-time type checking.
+
+LuaInspect \cite{luainspect} is a tool that uses MetaLua to perform
+some code analysis.
+For instance, it flags unknown global variables and table fields,
+it checks the number of function arguments against signatures, and
+it infers function return values.
+However, it does not try to analyze object-oriented code and
+it does not perform compile-time type checking.
+
+Tidal Lock \cite{tidallock} is a prototype of another optional type
+system for Lua, which is written in Metalua.
+Tidal Lock covers a little subset of Lua.
+Statements include declaration of local variables, multiple assignment,
+function application, and the return statement.
+This means that Tidal Lock does not include any control-flow statement.
+Expressions include primitive literals, table indexing, function application,
+function declaration, and the table constructor, but they do not include
+binary operations.
+
+A remarkable feature of Tidal Lock is the refinement of table types.
+This feature inspired us to also include it in Typed Lua,
+but in a simpler way and with different formalization.
+
+The table type from Tidal Lock can only represent records, that is,
+it cannot describe hash tables and arrays yet, though we can refine them.
+Tidal Lock also includes field types to describe the type of the fields
+of a table type.
+The field types describe if a table field is mutable or immutable
+in a table type.
+Field types are the feature that allow the refinement of table types in
+Tidal Lock.
+
+Tidal Lock is also a structural type system that relies on subtyping and
+local type inference.
+However, it does not support union types, recursive types, and variadic types.
+It also does not type any object-oriented idiom.
+
+Sol \cite{sol} is an experimental optional type system for Lua.
+Its type system is similar to ours, as it includes literal types,
+union types, and function types that handle variadic functions.
+However, it does not handle the refinement of tables and it
+includes different types for tables.
+Sol types tables as lists, maps, and objects.
+Its object types handle a specific object-oriented idiom that
+Sol introduces.
+
+Lua Analyzer \cite{luaanalyzer} is an optional type system for Lua
+that is specially designed to work in the Löve Studio,
+an IDE for game developing using the Löve framework.
+It works in Lua 5.1 only, and uses type annotations inside comments.
+It is unsound by design because its dynamic type is both
+top and bottom in the subtyping relation.
+
+Lua Analyzer shares some features with Typed Lua, and also
+has some interesting features that we do not have in Typed Lua.
+It has similar rules for handling the \texttt{or} idiom and
+discriminating union types inside conditions.
+However, these rules are limited to the \texttt{nil} tag only.
+It also includes different types for typing tables.
+It includes regular record types that maps names to types,
+array types, and map types.
+Even though it does not support the refinement of tables,
+it allows the definition of nominal table types that simulate classes.
+This system allows it to type check custom class systems,
+which are common in Lua.
+Function types also support multiple return values and
+variadic functions, but they do not support overloading the
+return type.
+Recently, it included experimental support for type aliases and generics.
+
+Luacheck \cite{luacheck} is a tool that performs static analysis on Lua code.
+It can flag access to undeclared globals and unused local variables,
+but it does not perform static type checking.
+
+Ravi \cite{ravi} is an experimental Lua dialect.
+Ravi introduces optional static typing for Lua to improve run-time performance.
+To do that, Ravi extends the Lua Virtual Machine to include new
+operations that take into account static type information.
+Currently, Ravi extends the Lua Virtual Machine to support few types:
+\texttt{integer}, \texttt{number}, arrays of integers, and arrays of numbers.
+
+\subsection{Other projects}
+
+Typed Racket \cite{tobin-hochstadt2008ts} is a statically typed version
+of the Racket language, which is a Scheme dialect.
+The main purpose of Typed Racket is to allow programmers to combine
+untyped modules, which are written in Racket, with typed modules, which are
+written in Typed Racket.
+It also uses local type inference to deduce the type of unannotated expressions.
+
+The main feature of Typed Racket's type system is \emph{occurrence typing}
+\cite{tobin-hochstadt2010ltu}.
+It is a novel way to use type predicates in control flow statements
+to refine union types.
+Occurrence typing is not sound in the presence of mutation.
+As these kinds of checks are common in other languages, related systems
+have appeared \cite{guha2011tlc,winther2011gtp,pearce2013ccf}.
+
+The type system of Typed Racket also includes function types, recursive
+types, and structure types.
+Its function types also handle multiple return values, and there is
+also a way to describe function types that have optional arguments.
+Its structure types are similar to our interfaces, as they describe record types.
+The type system is also structural and based on subtyping.
+It also includes the dynamic type \texttt{Any}, which is the top type in the system.
+Typed Racket also supports polymorphic functions and data structures.
+
+Typed Clojure \cite{bonnaire-sergeant2012typed-clojure} is an
+optional type system for Clojure.
+Although Clojure is a Lisp dialect that runs on the Java Virtual Machine,
+Common Language Runtime, and JavaScript, Typed Clojure runs only on
+the Java Virtual Machine.
+Perhaps, this restriction pushed Typed Clojure to support Java classes
+and some Java types such as \texttt{Long}, \texttt{Double}, and \texttt{String}.
+Typed Clojure also provides optional type annotations and uses
+local type inference to deduce the type of unannotated expressions.
+It also assigns the type \texttt{Any} to unannotated function parameters,
+which is the top type in the type system.
+
+The type system of Typed Clojure includes polymorphic function types,
+union types, intersection types, lists, vectors, maps, sets, and recursive types.
+Function types can also have rest parameters, which are similar
+to our variadic types, but can only appear on the input parameter
+of function types.
+In fact, its function types cannot return multiple results.
+It also uses occurrence typing to allow control flow statements to
+refine union types.
+The type system is also structural and based on subtyping.
+
+Dart \cite{dart} is a new class-based object-oriented programming
+language.
+It includes optional type annotations and compiles to JavaScript.
+The type system of Dart is nominal and includes base types,
+function types, lists, and maps.
+It also supports generics, and the programmer can define
+generic functions, lists, and maps.
+Unlike Typed Lua, Dart is unsound by design.
+
+Even though Dart has optional typing and static types by
+default do not affect run-time semantics, it has an
+execution mode that affects run-time.
+The \emph{checked mode} inserts run-time assertions that
+verifies whether static types match run-time tags.
+The \emph{production mode} is the default execution mode
+that does not include any assertions.
+
+TypeScript \cite{typescript} is a JavaScript extension
+that includes optional type annotations and class-based
+object-oriented programming.
+It also uses local type inference to deduce the type
+of unannotated expressions.
+The type system of TypeScript is structural, based
+on subtyping, and supports generics.
+It includes the dynamic type, primitive types, union types,
+function types, array types, tuple types, recursive types, and
+object types.
+Unlike Typed Lua, TypeScript uses arrays to represent variadic
+functions and multiple return values.
+
+Even though TypeScript is unsound by design,
+Bierman et al. \cite{bierman2014typescript} shows how to
+make TypeScript sound.
+They use a reduced core of TypeScript to formalize a
+sound type system for TypeScript, but also to formalize
+its current unsound type system.
+
+TeJaS \cite{lerner2013tejas} is a framework for the construction of
+different type systems for JavaScript.
+The authors created a base type system for JavaScript with
+extensible typing rules that allow the experimentation of
+different static analysis.
+They used TeJaS to create a type system that simulates the
+type system of TypeScript.
+
+Politz et al. \cite{politz2012semantics} proposes semantics
+and types for objects with first-class member names, a well-known
+feature from scripting languages.
+Their type system uses string patterns to describe the members of
+an object, and define a complex subtyping relation to validate
+these patterns.
+They also provide an implementation of their system to JavaScript.
+
+Gradualtalk \cite{allende2013gts} is a Smalltalk dialect that
+supports gradual typing.
+The type system combines nominal and structural typing.
+It includes function types, union types, structural types,
+nominal types, a self type, and parametric polymorphism.
+The type system also relies on subtyping and consistent-subtyping.
+
+Gradualtalk inserts run-time checks that ensure dynamically
+typed code does not violate statically typed code.
+Allende et al. \cite{allende2013cis} perform a careful
+evaluation about cast insertion in Gradualtalk.
+They report that usually cast insertions impact on execution
+performance, so Gradualtalk also has an option that allows
+programmers to turn them off, downgrading Gradualtalk
+to an optional type system.
+
+Reticulated Python \cite{vitousek2014deg} is a Python compiler
+that supports gradual typing.
+The type system is structural and based on subtyping.
+It includes base types, the dynamic type, list types,
+dictionary types, tuple types, function types, set types,
+object types, class types, and recursive types.
+It includes class and object types to differentiate the
+type of class declarations and instances, respectively.
+It also uses local type inference.
+Besides static type checking, Reticulated Python also introduces
+three different approaches for inserting run-time assertions.
+
+Mypy \cite{mypy} is an optional type system for Python.
+The type system of mypy is similar to the type system of
+Reticulated Python, but mypy does not insert run-time checks
+and it has parametric polymorphism.
+In contrast, Reticulated Python can type variadic functions,
+but mypy cannot.
+Recently, Guido van Rossum, Python's author, proposed a
+standard syntax for type annotations in Python \cite{PEP483}
+that is extremely inspired by mypy \cite{PEP484}.
+The main goal of this proposal is to make easier building
+static analysis tools for Python.
+Typing \cite{typing} is a tool that is being developed to
+implement this proposal.
+
+Hack \cite{hack} is a new programming language that runs on the
+Hip Hop Virtual Machine (HHVM).
+The HHVM is a virtual machine that executes Hack and PHP programs.
+We can view Hack as an extension to PHP that combines static and
+dynamic typing.
+The type system of Hack includes generics, nullable types, collections,
+and function types.
+
+The Ruby Type Checker \cite{ren2013rtc} is a library that
+performs type checking during run-time.
+The library provides type annotations that the programmer
+can use on classes and methods.
+Its type system includes nominal types, union types,
+intersection types, method types, parametric polymorphism,
+and type casts.
+
+Grace \cite{black2013sg} is an object-oriented language
+with optional typing.
+Grace is not a dynamically typed language that has been
+extended with an optional type system, but a language
+that has been designed from scratch to have both
+static and dynamic typing.
+Homer et al. \cite{homer2013modules} explores some
+useful patterns that derive from Grace's use of objects as modules
+and its brand of optional structural typing, which
+can also be expressed with Typed Lua's modules as tables.
+
+\section{Conclusion}
+
+In this work we presented Typed Lua, an optional type system for Lua.
+We implemented Typed Lua as a Lua extension that allows programmers to
+combine static and dynamic typing in Lua code, making easier the evolution
+of simple scripts into large programs.
+
+Our main contribution is the formalization of a complete optional type
+system that introduces several novel type system features to statically
+type check Lua programs.
+Even though Lua shares several features with other dynamically
+typed languages such as JavaScript, Lua also has several unusual features.
+These unusual features include tables (or associative arrays) as the sole
+mechanism for structured data, besides functions with multiple return values
+and flexible arity that interact with multiple assignment.
+We highlight the following novel features of our type system:
+\begin{itemize}
+\item type refinement allows the incremental evolution of record and
+object types, playing an important role in statically type checking
+the idiomatic way in which Lua programmers use tables to define modules
+and objects;
+\item projection types handle functions that are overloaded on the
+number and types of return values, allowing programmers to narrow the
+types of a set of variables by narrowing the type of a single component
+of this set;
+\item union types and variadic types help our type system handle
+functions with flexible arity, that is, union types are helpful in
+describing optional parameters while variadic types are helpful in
+describing the type of the vararg expression and the type of functions
+that can receive or return any number of values.
+\end{itemize}
+
+A key feature in optional type systems is usability.
+This means that optional type systems should not change the idioms
+that programmers are already familiar with.
+Instead, optional type systems should fit existing idioms to
+statically type check them.
+Designing a too simple type system can overload programmers by forcing
+them to change the way they program in the language to fit the type system,
+while designing a too complex type system can overload programmers with
+types and error messages that are hard to understand, even if type inference
+removes the necessity of annotating the program with these complex types.
+The most challenging aspect of designing optional type systems is to find
+the right amount of complexity for a type system that feels natural to the programmers.
+
+Usability has been a concern in the design of Typed Lua since the beginning.
+We realized that we should not rely on the semantics of Lua only,
+as this could lead to a cumbersome type system that would not support
+several Lua idioms.
+For this reason, we performed a mostly automated survey of Lua idioms
+and features to inform our design choices.
+
+After designing and implementing Typed Lua, we performed several
+case studies to evaluate how successful we were in our goal of
+providing an usable type system.
+We evaluated 29 modules from 8 different case studies,
+and we could give precise static types to 83\% of the 449
+members that these modules export.
+For half of the modules, we could give precise static types to
+at least 89\% of the members from each module.
+Our evaluation results showed that our type system can statically
+type check several Lua idioms and features, though the evaluation
+results also exposed several limitations of our type system.
+We found that the three main limitations of our type system are
+the lack of intersection types, parametric polymorphism, and operator overloading.
+Overcoming these limitations is our major target for future work,
+as it will allow us to statically type check more programs.
+
+Unlike other optional type systems, we designed Typed Lua without
+deliberate unsound parts.
+However, we still do not have proofs that the novel features of
+our type system are sound.
+We see a soundness proof as another major future work, as it is
+necessary to use static types for code optimization.
+
+Finally, we believe that Typed Lua is a major contribution to the Lua community,
+because it offers a framework that programmers can use to document,
+test, and better structure their applications.
+For libraries where a full conversion to static type checking should
+prove unfeasible or too much work, the community can use Typed Lua
+just to document the external interfaces of the libraries,
+giving the benefits of static type checking to the users of these
+libraries.
+In fact, we already have user feedback from Lua programmers that are
+using Typed Lua in their projects.
+For instance, ZeroBrane Studio is an IDE for Lua development that is
+starting to use Typed Lua to perform static analysis in Lua code.
+
+\bibliographystyle{abbrvnat}
+\bibliography{paper}
+
+\end{document}

### 2015-05-11 doc/thesis/introduction.tex
Renaming thesis directory

@@
 these tags to perform run-time (or dynamic) type checking
-\citep{pierce2002tpl}.
+\cite{pierce2002tpl}.
 
@@
 typed language, when their simple scripts evolve into complex programs
-\citep{tobin-hochstadt2006ims}.
+\cite{tobin-hochstadt2006ims}.
 Dynamically typed languages certainly help programmers during the
@@
 option to choose between static and dynamic typing:
-\emph{optional type systems} \citep{bracha2004pluggable} and
-\emph{gradual typing} \citep{siek2006gradual} are two similar
+\emph{optional type systems} \cite{bracha2004pluggable} and
+\emph{gradual typing} \cite{siek2006gradual} are two similar
 approaches for blending static and dynamic typing in the same
@@
 We use the term \emph{gradual typing} to refer to the work of
-\citet{siek2006gradual}.
+Siek and Taha \cite{siek2006gradual}.
 
@@
 Typed Lua performs a very limited form of local type inference
-\citep{pierce2000lti}, as static typing does not necessarily mean
+\cite{pierce2000lti}, as static typing does not necessarily mean
 that programmers need to insert type annotations in the code.

### 2015-05-11 doc/thesis/system.tex
Final touches before sending to juries
Renaming thesis directory

@@
 our type system checks whether their types are consistent.
-The rule \textsc{T-ASSIGNMENT} is the general rule that express this idea:
+The rule \textsc{T-ASSIGNMENT} is the general rule that expresses this idea:
 \[
@@
 
-The simplest expression involving tables in the empty table constructor.
+The simplest expression involving tables is the empty table constructor.
 Its type checking rule is straightforward:

### 2015-05-11 doc/thesis/typedlua.tex
Renaming thesis directory

@@
       else
-        -- string.rep : (string, integer, string) -> (string)
+        -- string.rep : (string, integer, string?) -> (string)
         return string.rep(s1, s2)
@@
 We use the \texttt{const} annotation in the type of the methods
-because it is necessary for covariance among object types work.
+because it is necessary for covariance among object types to work.
 Finally, the top-level \texttt{return} statement fixes the type of

### 2015-05-11 doc/thesis/glossary.tex
Defining the glossary terms to be defined

\begin{description}
\item[bottom type] TO DO.
\item[closed table type] TO DO.
\item[coercion] TO DO.
\item[consistency] TO DO.
\item[consistent-subtyping] TO DO.
\item[contravariant] TO DO.
\item[covariant] TO DO.
\item[depth subtyping] TO DO.
\item[dynamic type] TO DO.
\item[filter type] TO DO.
\item[fixed table type] TO DO.
\item[flow typing] TO DO.
\item[free assigned variable] TO DO.
\item[gradual type system] TO DO.
\item[gradual typing] TO DO.
\item[invariant] TO DO.
\item[metatable] TO DO.
\item[nominal type system] TO DO.
\item[open table type] TO DO.
\item[optional type system] TO DO.
\item[projection environment] TO DO.
\item[projection type] TO DO.
\item[prototype object] TO DO.
\item[self-like delegation] TO DO.
\item[sound type system] TO DO.
\item[structural type system] TO DO.
\item[table refinement] TO DO.
\item[top type] TO DO.
\item[type environment] TO DO.
\item[type tag] TO DO.
\item[unique table type] TO DO.
\item[unsound type system] TO DO.
\item[userdata] TO DO.
\item[vararg expression] TO DO.
\item[variadic function] TO DO.
\item[variance] TO DO.
\item[width subtyping] TO DO.
\end{description}

### 2015-05-15 doc/thesis/glossary.tex
Adding glossary first draft

@@
 \begin{description}
-\item[bottom type] TO DO.
-\item[closed table type] TO DO.
-\item[coercion] TO DO.
-\item[consistency] TO DO.
-\item[consistent-subtyping] TO DO.
-\item[contravariant] TO DO.
-\item[covariant] TO DO.
-\item[depth subtyping] TO DO.
-\item[dynamic type] TO DO.
-\item[filter type] TO DO.
-\item[fixed table type] TO DO.
-\item[flow typing] TO DO.
-\item[free assigned variable] TO DO.
-\item[gradual type system] TO DO.
-\item[gradual typing] TO DO.
-\item[invariant] TO DO.
-\item[metatable] TO DO.
-\item[nominal type system] TO DO.
-\item[open table type] TO DO.
-\item[optional type system] TO DO.
-\item[projection environment] TO DO.
-\item[projection type] TO DO.
-\item[prototype object] TO DO.
-\item[self-like delegation] TO DO.
-\item[sound type system] TO DO.
-\item[structural type system] TO DO.
-\item[table refinement] TO DO.
-\item[top type] TO DO.
-\item[type environment] TO DO.
-\item[type tag] TO DO.
-\item[unique table type] TO DO.
-\item[unsound type system] TO DO.
-\item[userdata] TO DO.
-\item[vararg expression] TO DO.
-\item[variadic function] TO DO.
-\item[variance] TO DO.
-\item[width subtyping] TO DO.
+\item[bottom type] It is a type that is subtype of all types.
+
+\item[closed table type] It is a table type that does not allow adding fields.
+
+\item[coercion] It is a relation that allows converting values from one type to values
+of another type without error.
+
+\item[consistency] It is a relation that gradual typing uses to control the insertion of
+run-time checks between dynamically typed and statically typed code.
+
+\item[consistent-subtyping] It is a relation that combines consistency and subtyping.
+
+\item[contravariant] Subtyping is contravariant when it reverses the subtyping order,
+that is, it orders types from a more generic type to a more specific type.
+
+\item[covariant] Subtyping is covariant when it preserves the subtyping order,
+that is, it orders types from a more specific type to a more general type.
+
+\item[depth subtyping] It allows changing the type of individual fields of a table type.
+
+\item[dynamic type] It is a type that allows combining dynamic and static typing in the same code.
+It is neither the bottom nor the top type in the subtyping relation, but
+a subtype only of itself.
+Gradual typing uses the dynamic type along with the consistency relation to
+identify where run-time casts should be inserted to prevent that dynamically
+typed code violates statically typed code.
+
+\item[filter type] It is a type that allows Typed Lua to discriminate the type of local variables
+inside control flow statements.
+
+\item[fixed table type] It is a table type that does not allow adding fields. It also does not allow
+width subtyping to make single inheritance safe.
+
+\item[flow typing] It is a combination of static typing and flow analysis.
+
+\item[free assigned variable] It is a free variable that appears in an assignment.
+
+\item[gradual type system] It is a type system that uses the consistency relation to check the interaction
+among the dynamic type and other types.
+
+\item[gradual typing] It is an approach that uses a gradual type system to allow static and dynamic
+typing in the same code, but inserting run-time checks between statistically
+typed and dynamically typed code.
+
+\item[invariant] Subtyping is invariant when it does not allow ordering types, that is,
+it is a way that express type equality through subtyping.
+
+\item[metatable] It is a Lua table that allows changing its behavior.
+
+\item[nominal type system] It is a type system that uses name declarations to define the equivalence among types.
+
+\item[open table type] It is a table type that allows adding fields, and usually refers to table types that
+have aliases.
+
+\item[optional type system] It is a type system that allows combining static and dynamic typing in the same
+language, but without affecting the run-time semantics.
+
+\item[projection environment] It is a type environment that Typed Lua uses to assign projection variables to
+second-level types.
+
+\item[projection type] It is a type that allows Typed Lua to discriminate the type of local variables
+that have a dependency relation.
+
+\item[prototype object] It is an object that works like a class, that is, it is a prototype for creating
+new objects of a given class.
+
+\item[self-like delegation] It is a technique to implement inheritance in dynamically typed languages.
+
+\item[sound type system] It is a type system that does not type check all programs that contain a type error.
+
+\item[structural type system] It is a type system that uses the type structure to define the equivalence among types.
+
+\item[table refinement] It is an operation that allows changing a table type to include new fields or
+to specialize existing fields.
+
+\item[top type] It is a type that is supertype of all types.
+
+\item[type environment] It is an environment that assigns variable names to types.
+
+\item[type tag] In dynamically typed languages a type tag describes the type of a value during
+run-time.
+
+\item[unique table type] It is a table type that allows adding fields, and it is also the table type
+that describe the type of the table constructor.
+
+\item[unsound type system] It is a type system that type checks certain programs that contain type errors.
+
+\item[userdata] It is a way to define new types in an application or library that is written in C.
+
+\item[vararg expression] It is a Lua expression that can result in an arbitrary number of values.
+
+\item[variadic function] It is a Lua function that can receive an arbitrary number of arguments.
+
+\item[variance] It is the way types are ordered.
+
+\item[width subtyping] It allows adding more fields to a table type.
+
 \end{description}

### 2015-05-15 doc/thesis/system.tex
Removing rules that only handle expression with multiple results

@@
 \dfrac{\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
-       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
+       \env_{f} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
        n = |\;\overline{e}\;|}
-      {\env_{1}, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
+      {\env_{1}, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{f}}
 \end{array}
@@
 \dfrac{\env_{1}, \penv \vdash l_{i}:T_{i}, \env_{i+1} \;\;\;
-       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
+       \env_{f} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
        n = |\;\overline{l}\;|}
-      {\env_{1}, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
+      {\env_{1}, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{f}}
 \end{array}
@@
 if its type does not end in another variadic type $T{*}$.
-For instance, rules \textsc{T-EXPLIST2} and \textsc{T-EXPLIST3} handle
-the case where an expression list is just an expression that may produce
+For instance, the rules \textsc{T-EXPLIST2} and \textsc{T-EXPLIST3} handle
+the case where an expression list ends with an expression that may produce
 multiple values.
@@
 \mylabel{T-EXPLIST2}\\
-\dfrac{\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Void, \env_{2}}
-      {\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{2}}
+\dfrac{\begin{array}{c}
+       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
+       \env_{1}, \penv \vdash me:T_{n+1} \times ... \times T_{n+m} \times \Void, \env_{n+2}\\
+       \env_{f} = merge(\env_{1}, ..., \env_{n+2}) \;\;\;
+       n = |\;\overline{e}\;|
+       \end{array}}
+      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n+m} \times \Nil{*}, \env_{f}}
 \end{array}
@@
 \mylabel{T-EXPLIST3}\\
-\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n}{*}, \env_{1}
+\dfrac{\begin{array}{c}
+       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
+       \env_{1}, \penv \vdash me:T_{n+1} \times ... \times T_{n+m}{*}, \env_{n+2}\\
+       \env_{f} = merge(\env_{1}, ..., \env_{n+2}) \;\;\;
+       n = |\;\overline{e}\;|
+       \end{array}}
+      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n+m}{*}, \env_{f}}
 \end{array}
@@
        n = |\;\overline{[e_{1}] = e_{2}}\;| \\
-       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) 
+       \env_{f} = merge(\env_{1}, ..., \env_{n+1})
        \end{array}}
-      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{m}}
+      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{f}}
 \end{array}
@@
 types $\pi_{i}^{x}$.
-In our example, our type system uses rule \textsc{T-EXPLIST6} for
+In our example, our type system uses rule \textsc{T-EXPLIST4} for
 type checking $idiv(1, 2)$:
@@
 \begin{array}{c}
-\mylabel{T-EXPLIST6}\\
+\mylabel{T-EXPLIST4}\\
 \dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Void \sqcup T_{1}' \times ... \times T_{n}' \times \Void, \env_{2}\\
-       S = T_{1} \times ... \times T_{n} \times \Nil{*} \sqcup T_{1}' \times ... \times T_{n}' \times \Nil{*}
+       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
+       \env_{1}, \penv \vdash me:S, \env_{n+2}\\
+       S = T_{n+1} \times ... \times T_{n+m} \times \Void \sqcup T_{n+1}' \times ... \times T_{n+m}' \times \Void\\
+       \env_{f} = merge(\env_{1}, ..., \env_{n+2}) \;\;\;
+       n = |\;\overline{e}\;|
        \end{array}}
-      {\env_{1}, \penv \vdash me:\pi_{1}^{x} \times ... \times \pi_{n}^{x} \times \Nil{*}, \env_{2}, (x,S)}
+      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times \Nil{*}, \env_{f}, (x,S)}
 \end{array}

### 2015-05-20 doc/thesis/typedlua.tex
Improving the text about recursive functions and the differences between any and value

@@
 
+Before we start discussing examples of Typed Lua code, it is worth
+mentioning that there is a subtle difference between the dynamic
+type \texttt{any} and the top type \texttt{value}.
+Although both types mean that they accept a value of any other type,
+the type \texttt{value} is not a good option for handling the
+interaction between dynamically typed and statically typed code.
+Gradual typing uses the dynamic type \texttt{any} to identify
+where it should insert run-time checks for asserting that dynamically
+typed code does not violate statically typed code.
+Typed Lua also uses the dynamic type \texttt{any} in this sense,
+though it is an optional type system.
+More precisely, we use \texttt{any} instead of \texttt{value} to allow
+programmers blending dynamic and static typing because we use the
+consistent-subtyping relation to formalize our optional type system,
+as it is a first step to switch Typed Lua from optional typing to
+gradual typing in the future.
+
 Typed Lua allows optional type annotations in variable and function
@@
 to assign a more specific type to an unannotated return type.
-This means that we can rewrite the previous example as follows:
+
+This means that we can rewrite the previous example to use local type
+inference for inferring the return type of \texttt{succ} and also for
+inferring the type of \texttt{x}:
 \begin{verbatim}
@@
 
-Typed Lua assigns the type \texttt{any} to the unannotated declarations
+Typed Lua assigns the dynamic type \texttt{any} to the unannotated declarations
 that it does not infer a more specific type.
-More precisely, Typed Lua does not infer more specific types to input
+More precisely, Typed Lua does not infer more specific types to the input
 parameters of function declarations and to the return type of
 recursive functions.
+The Typed Lua compiler cannot infer them because it performs type checking in
+a single step that simulates the program execution.
+We could have split type checking into two steps to try solving this
+limitation, but it would have implications in the mechanisms
+that Typed Lua uses to handle the discrimination of union types
+and the refinement of table types.
+We will discuss these features in Section \ref{sec:unions} and
+Section \ref{sec:tables}, respectivelly.
 
@@
 the return type of \texttt{factorial}.
+Local type inference cannot use the type of the returned
+expression when it includes a function call to the function that is
+being type checked, as its return type is still unknown to the type checker.
+For this reason, we need to annotate the return type of
+recursive local functions to inform the type cheker what type
+it should use while type checking their body.
 
@@
 between the interaction of dynamically typed and statically typed code.
+
+Even though Typed Lua can type check recursive functions when we
+annotate their return type, it has some limitations for type checking
+mutually recursive functions, even if we annotate their return type.
+We will use the following example for discussing this limitation and
+also for presenting an alternative solution:
+\begin{verbatim}
+    local even, odd
+
+    function even (n:integer):boolean
+      if n == 0 then return true
+      elseif n > 0 then return odd(n - 1)
+      else return odd(n + 1)
+      end
+    end
+
+    function odd (n:integer):boolean
+      if n == 0 then return false
+      elseif n > 0 then return even(n - 1)
+      else return even(n + 1)
+      end
+    end
+\end{verbatim}
+
+This example shows an attempt of annotating a common idiom that Lua
+programmers use for defining mutually recursive functions, but it
+generates compile-time warnings.
+The problem is related to the fact that Lua does not split a program
+into declarations and statements, and also to the fact that Typed Lua
+performs type checking in a single step.
+This means that forwarding the declaration of a local variable assigns
+the type \texttt{any} to it, making the compiler warn the programmers
+about function calls to this variable, as they can assign any value
+to a variable that has the dynamic type.
+In this example, the Typed Lua compiler generates warnings when we
+call \texttt{even} and \texttt{odd}.
+
+One way to overcome these warnings is to predeclare these functions
+with an empty body before the actual declaration.
+Even though this is a verbose solution and the Typed Lua compiler
+has an option to suppress warnings related to the dynamic type
+\texttt{any}, it ensures that further assignments to forwarded
+local variables will not change their type.
+Next we show how we can apply this alteranative solution to the
+previous example, making it compile without any warnings:
+\begin{verbatim}
+    local function even (n:integer):boolean return true end
+    local function odd (n:integer):boolean return false end
+
+    function even (n:integer):boolean
+      if n == 0 then return true
+      elseif n > 0 then return odd(n - 1)
+      else return odd(n + 1)
+      end
+    end
+
+    function odd (n:integer):boolean
+      if n == 0 then return false
+      elseif n > 0 then return even(n - 1)
+      else return even(n + 1)
+      end
+    end
+\end{verbatim}
 
@@
 
-As we can see in Figure \ref{fig:functions}, second-level types consist
-of either \emph{tupletype} or \emph{rettype}.
+Figure \ref{fig:functions} shows that Typed Lua uses second-level
+types in the definition of function types.
 A second-level type is either the type void \texttt{()},

### 2015-05-22 doc/thesis/typedlua.tex
Improving the text about operation modes and evaluation categories

@@
 affect other multiple assignments.
-For instance,
+This means that both operation modes discard extra arguments in
+multiple assignments and they also use \texttt{nil} to replace missing
+arguments in multiple assignments.
+We made this design decision to avoid being to restrictive, that is,
+we did not want to give unnecessary warnings to programmers,
+as discarded values do not have any implications during run-time.
+
+As an example,
 \begin{verbatim}
@@
 \end{verbatim}
-compiles without any warnings in both operation modes.
+compiles without any compile-time warnings in both operation modes.
 This means that the left-hand side of this local declaration has the
 tuple type \texttt{(integer, integer, value*)}, while the expression
-list in the right-hand side of this local declaration has the
-tuple type \texttt{(1, 2, 3, nil*)} in both operation modes.
+list in the right-hand side of this local declaration has the tuple
+type \texttt{(1, 2, 3, nil*)} in both operation modes.
+The literal type \texttt{3} is consistent with \texttt{value},
+and the type \texttt{nil*} generates a value of type \texttt{nil}
+that is also consistent with \texttt{value}.
 
@@
 \end{verbatim}
-compiles with warnings in both operation modes.
+compiles with one compile-time warning in both of the operation modes.
 This means that the left-hand side of this local declaration has
@@
 \end{verbatim}
-compiles without warnings in both operation modes.
+compiles without compile-time warnings in both operation modes.
 This happens because in the right-hand side of the multiple
@@
 has type \texttt{(string*) -> (string*)} in both operation modes.
-The call \texttt{v()} compiles without any warnings in the default and
-strict modes because \texttt{(value*)} and \texttt{()} are consistent
-with \texttt{(string*)}.
+The function call \texttt{v()} compiles without any compile-time warnings in
+the default and strict modes because \texttt{(value*)} and \texttt{()}
+are consistent with \texttt{(string*)}.
 The call \texttt{v("hello", "world")} compiles without any warnings
@@
 and \texttt{("hello", "world")} are consistent with \texttt{(string*)}.
-The call \texttt{v(...)} compiles without any warnings in both
+Calling \texttt{v(...)} compiles without any warnings in both
 operation modes because the type of the argument list is \texttt{(string*)},
@@
 
-The type of \texttt{message} is
+The type of the function \texttt{message} is
 \texttt{(string, string|nil, value*) -> (string, nil*)}
@@
 in the strict mode.
-The call \texttt{message("Typed Lua")} compiles without
-any warnings in both modes, because the argument list
-\texttt{("Typed Lua", nil*)} is consistent with the input type
-\texttt{(string, string|nil, value*)} in the default mode,
-and it is also consistent with the input type
+The function call \texttt{message("Typed Lua")} compiles without
+any compile-time warnings in both modes, because the argument list
+\texttt{("Typed Lua")} has type \texttt{("Typed Lua", nil*)} that
+is consistent with the input type \texttt{(string, string|nil, value*)}
+in the default mode, and that is also consistent with the input type
 \texttt{(string, string|nil)} in the strict mode.
@@
 So far, Typed Lua replaces projection types with the union of the
-corresponding component of each tuple, when it infers that a variable
+corresponding component of each tuple, when it infers that a local variable
 has a projection type.
@@
 If a variable with a projection type appears in a \texttt{type} predicate,
-it discriminates against all tuples in the projected union.
+it discriminates all tuples in the projected union.
 In our example, the projected union has type
@@
 guaranteed to return a string when we pass a string.
-Even though we could use intersection types to express the type
-of this function as
-\texttt{(number) -> (number) \string& (string) -> (string)},
+Even though we could have used an intersection type
+\texttt{(number) -> (number) \string& (string) -> (string)}
+to express the type of this function,
 intersection types require more sophisticated flow typing to

### 2015-05-22 doc/thesis/typedlua.tex
Improving the text and examples about table types

@@
 in Section \ref{sec:statistics}.
+More precisely, those results indicated that programmers seldom
+define a table constructor that is neither an empty table nor
+a table that contains only literal keys.
+This means that our design can type check most of the usages of
+the table constructor.
 Later in this chapter we will show that we use the syntax of records
@@
 which we discussed in Section \ref{sec:statistics}.
+More precisely, those results indicated that programmers seldom
+use non-literal keys in the definition of a table constructor.
+For typing maps, base types can type check most of the cases where
+programmers use a table constructor to initialize a map,
+while the type \texttt{value} is still an option to allow programmers
+to type check uncommon table constructors.
+This means that this restriction to key types in maps has no
+impact in the usability of Typed Lua.
 Next, we show one example of table type to type a map from strings to integers:
@@
 Typed Lua also has different subtyping rules that handle these
-different tags.
-We explain these tags and their subtyping rules in more detail
-in Chapter \ref{chap:system}.
+different tags, which we explain in Section \ref{sec:subtyping}.
 
@@
 
-The following example shows that it is not sound to allow the
-type of the fields to change:
+Before discussing each one of these rules, it is worth discussing
+why Typed Lua needs \emph{unique} and \emph{open} table types.
+Even though both \emph{unique} and \emph{open} table types
+allow the refinement of table types, we need both tags to
+increase usability and to avoid some unsafe behaviors.
+
+If we did not have \emph{unique} table types and \emph{open}
+table types had the same behavior that they have now,
+we would not be able to type check some safe constructions,
+like in the following example:
+\begin{verbatim}
+    local t:{"x":integer, "y":integer?} = { x = 1, y = 2 }
+\end{verbatim}
+
+Typed Lua type checks this example because the table constructor
+has an \emph{unique} table type \texttt{\{"x":1, "y":2\}}
+that is consistent with the type in the annotation.
+However, if the table constructor had an \emph{open} table type,
+it would not type check because it does not allow us to convert
+a table field from \texttt{integer} to \texttt{integer?}.
+We need this kind of behavior to avoid unsafe constructions
+while aliasing \emph{unique} table types, like in the follwoing
+example:
+\begin{verbatim}
+    local t1 = { x = 1 }
+    local t2:{"x":integer} = t1
+    local t3:{"x":integer?} = t1      --> compile-time warning
+    t3.x = nil
+\end{verbatim}
+
+Typed Lua generates a compile-time warning in this example
+because aliasing \texttt{t1} changes its type from \emph{unique} to \emph{open}.
+First, Typed Lua assigns an \emph{unique} table type to \texttt{t1},
+but then it changes its type to \emph{open} before aliasing
+\texttt{t1} to \texttt{t2}.
+Even though it is safe aliasing \texttt{t1} to \texttt{t2},
+it is not safe aliasing \texttt{t1} to \texttt{t3}, as it allows
+removing the value that is stored in the field \texttt{x} from
+\texttt{t1} through an assignment to field \texttt{x} from \texttt{t3}.
+Typed Lua uses \emph{open} table types to prevent this kind of
+unsafe behavior, as it would be allowed by \emph{unique} table
+types due to their loose subtyping rules.
+
+Back to the rules that Typed Lua uses to allow the refinement
+of table types, the following example shows that it is not sound
+to allow the type of the fields to change:
 \begin{verbatim}
@@
 \begin{verbatim}
-    local typealias Person = {"firstname":string, "lastname":string}
+    local typealias Person = { "firstname":string,
+                               "lastname":string }
 \end{verbatim}
@@
 This means that modules have \emph{fixed} table types in Typed Lua.
-Typed Lua has this behavior to ensure that classes always implement
-\emph{fixed} table types.
+This behavior ensures that classes always have \emph{fixed} table types.
 In the next section we will show that we build classes in a
@@
 After we define the module \texttt{mymath}, regardless of the adopted style,
-users can use it in the standard way:
+users can use it in the standard way, but with the static type checking
+provided by Typed Lua.
+Next we show an example:
 \begin{verbatim}
@@
 In Typed Lua, \texttt{require} is a primitive that statically type checks
-a given module to infer the type of this module.
+a given module to infer its type.
 This means that the type of its input parameter must be a literal string,
@@
 Lua has a mechanism for self-like (or JavaScript-like) delegation of
-missing fields in a table.
+missing table fields.
 After the call \texttt{setmetatable(t1, \{ \string_\string_index = t2 \})},
@@
     Circle.radius = 0.0
-    const function Circle:new (x:number, y:number, radius:value):self
-      local c:self = setmetatable(Shape:new(x, y), { __index = self })
+    const function Circle:new (x:number,
+                               y:number,
+                               radius:value):self
+      local c:self = setmetatable(Shape:new(x, y),
+                                  { __index = self })
       c.radius = tonumber(radius) or 0.0

### 2015-05-22 doc/paper/typedlua.tex
Final version published in DYLA

@@
-\documentclass[preprint]{sig-alternate}
+\documentclass{sig-alternate}
 
@@
 \newcommand{\Const}{\mathbf{const}}
+\newcommand{\Self}{\mathbf{self}}
 
@@
 
-\conferenceinfo{Dyla}{'14, Edinburgh, UK}
+\permission{Permission to make digital or hard copies of all or part of this work for
+   personal or classroom use is granted without fee provided that copies are
+   not made or distributed for profit or commercial advantage and that copies
+   bear this notice and the full citation on the first page. Copyrights for
+   components of this work owned by others than the author(s) must be
+   honored. Abstracting with credit is permitted. To copy otherwise, or
+   republish, to post on servers or to redistribute to lists, requires prior
+   specific permission and/or a fee. Request permissions from
+   Permissions@acm.org.}
+\conferenceinfo{Dyla}{'14, June 09-11 2014, Edinburgh, United Kingdom} 
+\CopyrightYear{is held by the owner/author(s). Publication rights licensed to
+ACM.}
+\crdata{978-1-4503-2916-3/14/06}
 
@@
 optionally-typed extension of the Lua scripting language.
+Lua is an imperative scripting language with first
+class functions and lightweight metaprogramming mechanisms.
 The design of Typed Lua's type system has a novel combination
@@
 programmers are used to, while bringing static type
-safety to them.
+safety to them. We show how the major features of the
+type system type these idioms with some
+examples, and discuss some of the design issues we faced.
 \end{abstract}
+
+\category{D.3.1}{Programming Languages}{Formal Definitions and Theory}[Semantics]
+\category{F.3.3}{Logics and Meanings of Programs}{Studies of Program Constructs}[Type structure]
 
@@
 and Typed Lua uses them to type multiple assignment and function
-application. Types are ordered by a subtype relationship, where
-any first-class type is a subtype of $\Top$.
+application.
+
+Types are ordered by a subtype relationship, where
+any first-class type is a subtype of $\Top$. The rest of the 
+subtype relationship is standard: union types are supertypes
+of their parts, $\Number$, $\Boolean$, and $\String$ are
+supertypes of their respective literal types, function
+types are related by contravariance on the input part and
+covariance in the output part, table types have width
+subtyping, with depth subtyping on $\Const$ fields,
+tuple and vararg types are covariant.
 
@@
 & | \; \Any & \textit{dynamic type}\\
+& | \; \Self & \textit{self type}\\
 & | \; T \cup T & \textit{disjoint union types}\\
@@
 \multicolumn{3}{c}{\textbf{Second-class types}} \\
-S ::= &  \multicolumn{2}{l}{\; V \; | \; V \cup S} \\
+S ::= &  \multicolumn{2}{l}{\; \Void \; | \; V \; | \; V \cup S} \\
 V ::= & \;\; T & \\
@@
 
-Typed Lua always adds a
-variadic tail to the parts of a function type if none is specified, to match
-the semantics of Lua function calls. In the examples above, the types
-of {\tt factorial} and {\tt abs} are actually $\Number \times \Top *
- \rightarrow \Number \times \Nil *$, and the type of {\tt distance} is
-$\Any \times \Any \times \Top * \rightarrow \Number \times \Nil *$.
-
-If we call {\tt abs} with extra arguments, Typed Lua silently ignores
-them, as the type signature lets {\tt abs} receive any number of extra
-arguments. If we call {\tt abs} in the right side of an assignment to
-more than one lvalue, Typed Lua checks if the first lvalue has a type
-consistent with $\Number$, and any other lvalues need to have a type
-consistent with $\Nil$.
+In its default mode of operation, Typed Lua always adds a
+variadic tail to the parts of a function type if none is 
+specified, to match the semantics of Lua function calls. In the 
+examples above, the types
+of {\tt factorial} and {\tt abs} are actually
+$\Number \times \Top * \rightarrow \Number \times \Nil *$,
+and the type of {\tt distance} is $\Any \times \Any \times \Top * 
+\rightarrow \Number \times \Nil *$.
+
+If we call {\tt abs} with extra arguments, Typed Lua silently
+ignores them, as the type signature lets {\tt abs} receive any 
+number of extra arguments. If we call {\tt abs} in the right side 
+of an assignment to more than one lvalue, Typed Lua checks if the 
+first lvalue has a type consistent with $\Number$, and any other 
+lvalues need to have a type consistent with $\Nil$.
+
+There is an optional stricter mode of operation where
+Typed Lua does not give variadic tails to the parts
+of a function type unless the programmer explictly
+declares it, and so will also check all function calls
+for arity mismatch.
 
 A variadic type can only appear in the tail position of a tuple,
-because Lua takes only the first value of any expression that appears
-in an expression list that is not in tail position. The following example
-shows the interaction between multiple returns and expression lists:
+because Lua takes only the first value of any expression that 
+appears
+in an expression list that is not in tail position. The following 
+example
+shows the interaction between multiple returns and expression 
+lists:
 
@@
 
-Typed Lua lets assignment change the type of a local variable in
-cases that respect a simple heuristic: the new type must be a
-subtype of the previous one, the variable must be local
-to the current function, and the variable must not have been
-assigned to in another function. The new type only applies for
-the remainder of the current scope. In the case of {\tt
+Typed Lua lets the assignment $x = x \ \mathbf{or} \ e$ change
+the type of $x$ from $t \cup \Nil$ to $t$,
+as long as the type of $e$ is a subtype of $t$, $x$ is local
+to the current function, and it is not assigned in
+another function. The change only affects the type of $x$
+in the remainder of the current scope. In the case of {\tt
 greeting}, the assignment on line three changes its type to
@@
 over a local variable in a condition, constrain the type of that
-variable. The function above uses the {\tt type(X) == "string"}
-predicate which constrains the type of $X$ from $T \cup \String$ to
-$\String$ when the predicate is true and $T$ otherwise. This is
-a simplified form of {\em flow typing}~\cite{guha:flow}.
+variable. The function above uses the {\tt type(x) == "string"}
+predicate which constrains the type of $x$ from $T \cup \String$ to
+$\String$ when the predicate is true and $T$ otherwise.
+This is a simplified form of {\em flow 
+typing}~\cite{guha:flow,tobin:occur}. As with {\tt or}, the
+variable must be local to the function, and cannot be assigned
+to in another function.
 
@@
 passed a number and guaranteed to return a string if passed
-a string, for example.
+a string, for example. While intersection types provide
+a way to express the type of such a function as $\Number 
+\rightarrow \Number \cap \String \rightarrow \String$,
+more sophisticated flow typing is needed to actually check
+that a function has this type, and we are still working on this
+problem.
 
@@
 Another syntax defines table types for arrays:
-it is written \texttt{\{ t \}}, and maps to the table type
-$\{\Number:t\}$. A third syntax sugar defines
+it is written \texttt{\{ t \}}, and is equivalent to
+the table type
+$\{\Number:t\}$. A third syntax defines
 table types for records:
@@
 as a possible value.
+
+While this runtime check is an instance where the semantics
+of a Typed Lua program deviates from the semantics of
+plain Lua, the alternatives would be to make all
+arrays and maps have an element type that includes {\tt nil},
+and either make the programmer narrow the type with {\tt or}
+or an {\tt if} statement, making using the elements more
+incovenient, or make {\tt nil} a subtype of every type,
+reducing the amount of type safety in the system. Notice 
+that this does not change the semantics of dynamically
+typed programs, as the runtime checks are only added when
+the table has a strict static type.
 
@@
 always closed. The type of an open variable may change
-by field assignment, subject to two restrictions: the
+by field assignment, subject to three restrictions: the
 variable must be local to the current block, the new type
@@
 cannot have been assigned to in another function.
+
+An variable with an open table type may be aliased, but
+these aliases are not open. Any mutation on these aliases
+is not a problem, as the type of the original reference
+can only change towards a subtype. For mutable fields
+this means that the type of the field cannot change
+once it is added to the type of the table.
 
@@
 
-We can use recursive table types to represent objects,
+We can use table types and the {\em self type}
+to represent objects,
 and Typed Lua has syntactic sugar to make defining these
@@
 syntactic sugar for having a first parameter named
-{\tt self} with the same type as the interface. The
+{\tt self} with type $\Self$. The
 {\tt const} qualifier is necessary for covariance
@@
 
-Type-checking a method call is straightforward, but
-there is still the matter of how to construct a value with
+While {\tt :} is syntactic sugar in plain Lua, Typed Lua
+uses it to type-check method calls, to bind any ocurrence
+of the $\Self$ type in the type of the method to the
+receiver. Indexing a method but not immediately calling
+it with the correct receiver is a compile-time error.
+
+There is still the matter of how to construct a value with
 the object type above. The following example shows one
@@
 from Lua, and assigns a function to the field with a first
-parameter named {\tt self}, plus any other parameters.
-
-Lua has a mechanism for Self-like delegation of missing
-fields in a table. After we do {\tt setmetatable(t1,
-{ \_\_index = t2 })}, Lua looks up in {\tt t2} any missing
+parameter named {\tt self}, plus any other parameters. Typed
+Lua adds the {\tt const} annotation, and gives type $\Self$
+to the implicit parameter {\tt self}.
+
+Lua has a mechanism for Self-like (or JavaScript-like)
+ delegation of missing
+fields in a table. After {\tt setmetatable(t1,
+\{ \_\_index = t2 \})}, Lua looks up in {\tt t2} any missing
 fields of {\tt t1}. Lua programmers often use this mechanism
@@
 to {\tt Shape} in the type annotation, as a shortcut
-to the recursive table type that Typed Lua has assigned
+to the table type that Typed Lua has assigned
 to this variable.
@@
 Typed Lua assigns the type
-$\mathbf{Shape} \times \Number \times \Number \times
-\Top* \rightarrow \mathbf{Shape} \times \Nil*$ to
+$\Self \times \Number \times \Number \times
+\Top* \rightarrow \Self \times \Nil*$ to
 {\tt new}. In a {\tt setmetatable} expression,
-if the type $t_1$ of the first operand is open and a supertype
+if the type $t_1$ of the first operand is a supertype
 of the type $t_2$ of the second operand's {\_\_index} field,
-it changes the type of the first operand to $t_1$, and
-it remains open.
-
-As the result of {\tt setmetatable} can be an open
-table type, we can simulate single inheritance, and
-override methods:
+it changes the type of the first operand to $t_2$, 
+so the local variable {\tt s} 
+has the same type as {\tt self}.
+
+We can also use {\tt setmetatable} to simulate single
+inheritance, as {\tt setmetatable} on a table constructor
+assigned to a fresh local variable gives this variable
+an open type, which lets us add new methods and override
+existing ones:
 
@@
 two examples above are valid Lua code, with the same
-semantics as the Typed Lua Code.
+semantics as the Typed Lua code.
 
@@
 
-\citep{tobin-hochstadt2006ims} shows how programs
+\citet{tobin-hochstadt2006ims} shows how programs
 in the untyped lambda calculus can be incrementally
@@
 language. This approach has been realized in the
-Typed Scheme (later Typed Racket) language~\citet{tobin-hochstadt2008ts}.
+Typed Scheme (later Typed Racket) language~\citep{tobin-hochstadt2008ts,tobin:occur}.
 
@@
 the variable. As these types of checks are common in
-other languages, related systems have
+other languages, but ocurrence typing is not sound
+in the presence of mutation, related systems have
 appeared~\citep{guha:flow,winther:flow,pearce:whiley}.

### 2015-05-22 doc/dyla14/sig-alternate.tex
Rename old paper folder

% This is "sig-alternate.tex" V2.0 May 2012
% This file should be compiled with V2.5 of "sig-alternate.cls" May 2012
%
% This example file demonstrates the use of the 'sig-alternate.cls'
% V2.5 LaTeX2e document class file. It is for those submitting
% articles to ACM Conference Proceedings WHO DO NOT WISH TO
% STRICTLY ADHERE TO THE SIGS (PUBS-BOARD-ENDORSED) STYLE.
% The 'sig-alternate.cls' file will produce a similar-looking,
% albeit, 'tighter' paper resulting in, invariably, fewer pages.
%
% ----------------------------------------------------------------------------------------------------------------
% This .tex file (and associated .cls V2.5) produces:
%       1) The Permission Statement
%       2) The Conference (location) Info information
%       3) The Copyright Line with ACM data
%       4) NO page numbers
%
% as against the acm_proc_article-sp.cls file which
% DOES NOT produce 1) thru' 3) above.
%
% Using 'sig-alternate.cls' you have control, however, from within
% the source .tex file, over both the CopyrightYear
% (defaulted to 200X) and the ACM Copyright Data
% (defaulted to X-XXXXX-XX-X/XX/XX).
% e.g.
% \CopyrightYear{2007} will cause 2007 to appear in the copyright line.
% \crdata{0-12345-67-8/90/12} will cause 0-12345-67-8/90/12 to appear in the copyright line.
%
% ---------------------------------------------------------------------------------------------------------------
% This .tex source is an example which *does* use
% the .bib file (from which the .bbl file % is produced).
% REMEMBER HOWEVER: After having produced the .bbl file,
% and prior to final submission, you *NEED* to 'insert'
% your .bbl file into your source .tex file so as to provide
% ONE 'self-contained' source file.
%
% ================= IF YOU HAVE QUESTIONS =======================
% Questions regarding the SIGS styles, SIGS policies and
% procedures, Conferences etc. should be sent to
% Adrienne Griscti (griscti@acm.org)
%
% Technical questions _only_ to
% Gerald Murray (murray@hq.acm.org)
% ===============================================================
%
% For tracking purposes - this is V2.0 - May 2012

\documentclass{sig-alternate}

\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

\title{Alternate {\ttlit ACM} SIG Proceedings Paper in LaTeX
Format\titlenote{(Produces the permission block, and
copyright information). For use with
SIG-ALTERNATE.CLS. Supported by ACM.}}
\subtitle{[Extended Abstract]
\titlenote{A full version of this paper is available as
\textit{Author's Guide to Preparing ACM SIG Proceedings Using
\LaTeX$2_\epsilon$\ and BibTeX} at
\texttt{www.acm.org/eaddress.htm}}}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

\numberofauthors{8} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
Ben Trovato\titlenote{Dr.~Trovato insisted his name be first.}\\
       \affaddr{Institute for Clarity in Documentation}\\
       \affaddr{1932 Wallamaloo Lane}\\
       \affaddr{Wallamaloo, New Zealand}\\
       \email{trovato@corporation.com}
% 2nd. author
\alignauthor
G.K.M. Tobin\titlenote{The secretary disavows
any knowledge of this author's actions.}\\
       \affaddr{Institute for Clarity in Documentation}\\
       \affaddr{P.O. Box 1212}\\
       \affaddr{Dublin, Ohio 43017-6221}\\
       \email{webmaster@marysville-ohio.com}
% 3rd. author
\alignauthor Lars Th{\o}rv{\"a}ld\titlenote{This author is the
one who did all the really hard work.}\\
       \affaddr{The Th{\o}rv{\"a}ld Group}\\
       \affaddr{1 Th{\o}rv{\"a}ld Circle}\\
       \affaddr{Hekla, Iceland}\\
       \email{larst@affiliation.org}
\and  % use '\and' if you need 'another row' of author names
% 4th. author
\alignauthor Lawrence P. Leipuner\\
       \affaddr{Brookhaven Laboratories}\\
       \affaddr{Brookhaven National Lab}\\
       \affaddr{P.O. Box 5000}\\
       \email{lleipuner@researchlabs.org}
% 5th. author
\alignauthor Sean Fogarty\\
       \affaddr{NASA Ames Research Center}\\
       \affaddr{Moffett Field}\\
       \affaddr{California 94035}\\
       \email{fogartys@amesres.org}
% 6th. author
\alignauthor Charles Palmer\\
       \affaddr{Palmer Research Laboratories}\\
       \affaddr{8600 Datapoint Drive}\\
       \affaddr{San Antonio, Texas 78229}\\
       \email{cpalmer@prl.com}
}
% There's nothing stopping you putting the seventh, eighth, etc.
% author on the opening page (as the 'third row') but we ask,
% for aesthetic reasons that you place these 'additional authors'
% in the \additional authors block, viz.
\additionalauthors{Additional authors: John Smith (The Th{\o}rv{\"a}ld Group,
email: {\texttt{jsmith@affiliation.org}}) and Julius P.~Kumquat
(The Kumquat Consortium, email: {\texttt{jpkumquat@consortium.net}}).}
\date{30 July 1999}
% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

\maketitle
\begin{abstract}
This paper provides a sample of a \LaTeX\ document which conforms,
somewhat loosely, to the formatting guidelines for
ACM SIG Proceedings. It is an {\em alternate} style which produces
a {\em tighter-looking} paper and was designed in response to
concerns expressed, by authors, over page-budgets.
It complements the document \textit{Author's (Alternate) Guide to
Preparing ACM SIG Proceedings Using \LaTeX$2_\epsilon$\ and Bib\TeX}.
This source file has been written with the intention of being
compiled under \LaTeX$2_\epsilon$\ and BibTeX.

The developers have tried to include every imaginable sort
of ``bells and whistles", such as a subtitle, footnotes on
title, subtitle and authors, as well as in the text, and
every optional component (e.g. Acknowledgments, Additional
Authors, Appendices), not to mention examples of
equations, theorems, tables and figures.

To make best use of this sample document, run it through \LaTeX\
and BibTeX, and compare this source code with the printed
output produced by the dvi file. A compiled PDF version
is available on the web page to help you with the
`look and feel'.
\end{abstract}

% A category with the (minimum) three required fields
\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

\terms{Theory}

\keywords{ACM proceedings, \LaTeX, text tagging}

\section{Introduction}
The \textit{proceedings} are the records of a conference.
ACM seeks to give these conference by-products a uniform,
high-quality appearance.  To do this, ACM has some rigid
requirements for the format of the proceedings documents: there
is a specified format (balanced  double columns), a specified
set of fonts (Arial or Helvetica and Times Roman) in
certain specified sizes (for instance, 9 point for body copy),
a specified live area (18 $\times$ 23.5 cm [7" $\times$ 9.25"]) centered on
the page, specified size of margins (1.9 cm [0.75"]) top, (2.54 cm [1"]) bottom
and (1.9 cm [.75"]) left and right; specified column width
(8.45 cm [3.33"]) and gutter size (.83 cm [.33"]).

The good news is, with only a handful of manual
settings\footnote{Two of these, the {\texttt{\char'134 numberofauthors}}
and {\texttt{\char'134 alignauthor}} commands, you have
already used; another, {\texttt{\char'134 balancecolumns}}, will
be used in your very last run of \LaTeX\ to ensure
balanced column heights on the last page.}, the \LaTeX\ document
class file handles all of this for you.

The remainder of this document is concerned with showing, in
the context of an ``actual'' document, the \LaTeX\ commands
specifically available for denoting the structure of a
proceedings paper, rather than with giving rigorous descriptions
or explanations of such commands.

\section{The {\secit Body} of The Paper}
Typically, the body of a paper is organized
into a hierarchical structure, with numbered or unnumbered
headings for sections, subsections, sub-subsections, and even
smaller sections.  The command \texttt{{\char'134}section} that
precedes this paragraph is part of such a
hierarchy.\footnote{This is the second footnote.  It
starts a series of three footnotes that add nothing
informational, but just give an idea of how footnotes work
and look. It is a wordy one, just so you see
how a longish one plays out.} \LaTeX\ handles the numbering
and placement of these headings for you, when you use
the appropriate heading commands around the titles
of the headings.  If you want a sub-subsection or
smaller part to be unnumbered in your output, simply append an
asterisk to the command name.  Examples of both
numbered and unnumbered headings will appear throughout the
balance of this sample document.

Because the entire article is contained in
the \textbf{document} environment, you can indicate the
start of a new paragraph with a blank line in your
input file; that is why this sentence forms a separate paragraph.

\subsection{Type Changes and {\subsecit Special} Characters}
We have already seen several typeface changes in this sample.  You
can indicate italicized words or phrases in your text with
the command \texttt{{\char'134}textit}; emboldening with the
command \texttt{{\char'134}textbf}
and typewriter-style (for instance, for computer code) with
\texttt{{\char'134}texttt}.  But remember, you do not
have to indicate typestyle changes when such changes are
part of the \textit{structural} elements of your
article; for instance, the heading of this subsection will
be in a sans serif\footnote{A third footnote, here.
Let's make this a rather short one to
see how it looks.} typeface, but that is handled by the
document class file. Take care with the use
of\footnote{A fourth, and last, footnote.}
the curly braces in typeface changes; they mark
the beginning and end of
the text that is to be in the different typeface.

You can use whatever symbols, accented characters, or
non-English characters you need anywhere in your document;
you can find a complete list of what is
available in the \textit{\LaTeX\
User's Guide}\cite{Lamport:LaTeX}.

\subsection{Math Equations}
You may want to display math equations in three distinct styles:
inline, numbered or non-numbered display.  Each of
the three are discussed in the next sections.

\subsubsection{Inline (In-text) Equations}
A formula that appears in the running text is called an
inline or in-text formula.  It is produced by the
\textbf{math} environment, which can be
invoked with the usual \texttt{{\char'134}begin. . .{\char'134}end}
construction or with the short form \texttt{\$. . .\$}. You
can use any of the symbols and structures,
from $\alpha$ to $\omega$, available in
\LaTeX\cite{Lamport:LaTeX}; this section will simply show a
few examples of in-text equations in context. Notice how
this equation: \begin{math}\lim_{n\rightarrow \infty}x=0\end{math},
set here in in-line math style, looks slightly different when
set in display style.  (See next section).

\subsubsection{Display Equations}
A numbered display equation -- one set off by vertical space
from the text and centered horizontally -- is produced
by the \textbf{equation} environment. An unnumbered display
equation is produced by the \textbf{displaymath} environment.

Again, in either environment, you can use any of the symbols
and structures available in \LaTeX; this section will just
give a couple of examples of display equations in context.
First, consider the equation, shown as an inline equation above:
\begin{equation}\lim_{n\rightarrow \infty}x=0\end{equation}
Notice how it is formatted somewhat differently in
the \textbf{displaymath}
environment.  Now, we'll enter an unnumbered equation:
\begin{displaymath}\sum_{i=0}^{\infty} x + 1\end{displaymath}
and follow it with another numbered equation:
\begin{equation}\sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f\end{equation}
just to demonstrate \LaTeX's able handling of numbering.

\subsection{Citations}
Citations to articles \cite{bowman:reasoning,
clark:pct, braams:babel, herlihy:methodology},
conference proceedings \cite{clark:pct} or
books \cite{salas:calculus, Lamport:LaTeX} listed
in the Bibliography section of your
article will occur throughout the text of your article.
You should use BibTeX to automatically produce this bibliography;
you simply need to insert one of several citation commands with
a key of the item cited in the proper location in
the \texttt{.tex} file \cite{Lamport:LaTeX}.
The key is a short reference you invent to uniquely
identify each work; in this sample document, the key is
the first author's surname and a
word from the title.  This identifying key is included
with each item in the \texttt{.bib} file for your article.

The details of the construction of the \texttt{.bib} file
are beyond the scope of this sample document, but more
information can be found in the \textit{Author's Guide},
and exhaustive details in the \textit{\LaTeX\ User's
Guide}\cite{Lamport:LaTeX}.

This article shows only the plainest form
of the citation command, using \texttt{{\char'134}cite}.
This is what is stipulated in the SIGS style specifications.
No other citation format is endorsed or supported.

\subsection{Tables}
Because tables cannot be split across pages, the best
placement for them is typically the top of the page
nearest their initial cite.  To
ensure this proper ``floating'' placement of tables, use the
environment \textbf{table} to enclose the table's contents and
the table caption.  The contents of the table itself must go
in the \textbf{tabular} environment, to
be aligned properly in rows and columns, with the desired
horizontal and vertical rules.  Again, detailed instructions
on \textbf{tabular} material
is found in the \textit{\LaTeX\ User's Guide}.

Immediately following this sentence is the point at which
Table 1 is included in the input file; compare the
placement of the table here with the table in the printed
dvi output of this document.

\begin{table}
\centering
\caption{Frequency of Special Characters}
\begin{tabular}{|c|c|l|} \hline
Non-English or Math&Frequency&Comments\\ \hline
\O & 1 in 1,000& For Swedish names\\ \hline
$\pi$ & 1 in 5& Common in math\\ \hline
\$ & 4 in 5 & Used in business\\ \hline
$\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
\hline\end{tabular}
\end{table}

To set a wider table, which takes up the whole width of
the page's live area, use the environment
\textbf{table*} to enclose the table's contents and
the table caption.  As with a single-column table, this wide
table will ``float" to a location deemed more desirable.
Immediately following this sentence is the point at which
Table 2 is included in the input file; again, it is
instructive to compare the placement of the
table here with the table in the printed dvi
output of this document.

\begin{table*}
\centering
\caption{Some Typical Commands}
\begin{tabular}{|c|c|l|} \hline
Command&A Number&Comments\\ \hline
\texttt{{\char'134}alignauthor} & 100& Author alignment\\ \hline
\texttt{{\char'134}numberofauthors}& 200& Author enumeration\\ \hline
\texttt{{\char'134}table}& 300 & For tables\\ \hline
\texttt{{\char'134}table*}& 400& For wider tables\\ \hline\end{tabular}
\end{table*}
% end the environment with {table*}, NOTE not {table}!

\subsection{Figures}
Like tables, figures cannot be split across pages; the
best placement for them
is typically the top or the bottom of the page nearest
their initial cite.  To ensure this proper ``floating'' placement
of figures, use the environment
\textbf{figure} to enclose the figure and its caption.

This sample document contains examples of \textbf{.eps}
and \textbf{.ps} files to be displayable with \LaTeX.  More
details on each of these is found in the \textit{Author's Guide}.

\begin{figure}
\centering
\epsfig{file=fly.eps}
\caption{A sample black and white graphic (.eps format).}
\end{figure}

\begin{figure}
\centering
\epsfig{file=fly.eps, height=1in, width=1in}
\caption{A sample black and white graphic (.eps format)
that has been resized with the \texttt{epsfig} command.}
\end{figure}

As was the case with tables, you may want a figure
that spans two columns.  To do this, and still to
ensure proper ``floating'' placement of tables, use the environment
\textbf{figure*} to enclose the figure and its caption.
and don't forget to end the environment with
{figure*}, not {figure}!

\begin{figure*}
\centering
\epsfig{file=flies.eps}
\caption{A sample black and white graphic (.eps format)
that needs to span two columns of text.}
\end{figure*}

Note that either {\textbf{.ps}} or {\textbf{.eps}} formats are
used; use
the \texttt{{\char'134}epsfig} or \texttt{{\char'134}psfig}
commands as appropriate for the different file types.

\begin{figure}
\centering
\psfig{file=rosette.ps, height=1in, width=1in,}
\caption{A sample black and white graphic (.ps format) that has
been resized with the \texttt{psfig} command.}
\vskip -6pt
\end{figure}

\subsection{Theorem-like Constructs}
Other common constructs that may occur in your article are
the forms for logical constructs like theorems, axioms,
corollaries and proofs.  There are
two forms, one produced by the
command \texttt{{\char'134}newtheorem} and the
other by the command \texttt{{\char'134}newdef}; perhaps
the clearest and easiest way to distinguish them is
to compare the two in the output of this sample document:

This uses the \textbf{theorem} environment, created by
the\linebreak\texttt{{\char'134}newtheorem} command:
\newtheorem{theorem}{Theorem}
\begin{theorem}
Let $f$ be continuous on $[a,b]$.  If $G$ is
an antiderivative for $f$ on $[a,b]$, then
\begin{displaymath}\int^b_af(t)dt = G(b) - G(a).\end{displaymath}
\end{theorem}

The other uses the \textbf{definition} environment, created
by the \texttt{{\char'134}newdef} command:
\newdef{definition}{Definition}
\begin{definition}
If $z$ is irrational, then by $e^z$ we mean the
unique number which has
logarithm $z$: \begin{displaymath}{\log e^z = z}\end{displaymath}
\end{definition}

Two lists of constructs that use one of these
forms is given in the
\textit{Author's  Guidelines}.

There is one other similar construct environment, which is
already set up
for you; i.e. you must \textit{not} use
a \texttt{{\char'134}newdef} command to
create it: the \textbf{proof} environment.  Here
is a example of its use:
\begin{proof}
Suppose on the contrary there exists a real number $L$ such that
\begin{displaymath}
\lim_{x\rightarrow\infty} \frac{f(x)}{g(x)} = L.
\end{displaymath}
Then
\begin{displaymath}
l=\lim_{x\rightarrow c} f(x)
= \lim_{x\rightarrow c}
\left[ g{x} \cdot \frac{f(x)}{g(x)} \right ]
= \lim_{x\rightarrow c} g(x) \cdot \lim_{x\rightarrow c}
\frac{f(x)}{g(x)} = 0\cdot L = 0,
\end{displaymath}
which contradicts our assumption that $l\neq 0$.
\end{proof}

Complete rules about using these environments and using the
two different creation commands are in the
\textit{Author's Guide}; please consult it for more
detailed instructions.  If you need to use another construct,
not listed therein, which you want to have the same
formatting as the Theorem
or the Definition\cite{salas:calculus} shown above,
use the \texttt{{\char'134}newtheorem} or the
\texttt{{\char'134}newdef} command,
respectively, to create it.

\subsection*{A {\secit Caveat} for the \TeX\ Expert}
Because you have just been given permission to
use the \texttt{{\char'134}newdef} command to create a
new form, you might think you can
use \TeX's \texttt{{\char'134}def} to create a
new command: \textit{Please refrain from doing this!}
Remember that your \LaTeX\ source code is primarily intended
to create camera-ready copy, but may be converted
to other forms -- e.g. HTML. If you inadvertently omit
some or all of the \texttt{{\char'134}def}s recompilation will
be, to say the least, problematic.

\section{Conclusions}
This paragraph will end the body of this sample document.
Remember that you might still have Acknowledgments or
Appendices; brief samples of these
follow.  There is still the Bibliography to deal with; and
we will make a disclaimer about that here: with the exception
of the reference to the \LaTeX\ book, the citations in
this paper are to articles which have nothing to
do with the present subject and are used as
examples only.
%\end{document}  % This is where a 'short' article might terminate

%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}
This section is optional; it is a location for you
to acknowledge grants, funding, editing assistance and
what have you.  In the present case, for example, the
authors would like to thank Gerald Murray of ACM for
his help in codifying this \textit{Author's Guide}
and the \textbf{.cls} and \textbf{.tex} files that it describes.

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%
%APPENDICES are optional
%\balancecolumns
\appendix
%Appendix A
\section{Headings in Appendices}
The rules about hierarchical headings discussed above for
the body of the article are different in the appendices.
In the \textbf{appendix} environment, the command
\textbf{section} is used to
indicate the start of each Appendix, with alphabetic order
designation (i.e. the first is A, the second B, etc.) and
a title (if you include one).  So, if you need
hierarchical structure
\textit{within} an Appendix, start with \textbf{subsection} as the
highest level. Here is an outline of the body of this
document in Appendix-appropriate form:
\subsection{Introduction}
\subsection{The Body of the Paper}
\subsubsection{Type Changes and  Special Characters}
\subsubsection{Math Equations}
\paragraph{Inline (In-text) Equations}
\paragraph{Display Equations}
\subsubsection{Citations}
\subsubsection{Tables}
\subsubsection{Figures}
\subsubsection{Theorem-like Constructs}
\subsubsection*{A Caveat for the \TeX\ Expert}
\subsection{Conclusions}
\subsection{Acknowledgments}
\subsection{Additional Authors}
This section is inserted by \LaTeX; you do not insert it.
You just add the names and information in the
\texttt{{\char'134}additionalauthors} command at the start
of the document.
\subsection{References}
Generated by bibtex from your ~.bib file.  Run latex,
then bibtex, then latex twice (to resolve references)
to create the ~.bbl file.  Insert that ~.bbl file into
the .tex source file and comment out
the command \texttt{{\char'134}thebibliography}.
% This next section command marks the start of
% Appendix B, and does not continue the present hierarchy
\section{More Help for the Hardy}
The sig-alternate.cls file itself is chock-full of succinct
and helpful comments.  If you consider yourself a moderately
experienced to expert user of \LaTeX, you may find reading
it useful but please remember not to change it.
%\balancecolumns % GM June 2007
% That's all folks!
\end{document}

### 2015-05-23 doc/thesis/review.tex
Adding a statistics summary and a brief review about gradual guarantee

@@
 
-\citet{abadi1989dts} extended the simply typed lambda calculus with the
-\texttt{Dynamic} type and the \texttt{dynamic} and \texttt{typecase}
-constructs, with the aim to safely integrate dynamic code in
-statically typed languages.
+Abadi et al. \cite{abadi1989dts} extended the simply typed
+lambda calculus with the \texttt{Dynamic} type and the \texttt{dynamic}
+and \texttt{typecase} constructs, with the aim to safely integrate dynamic
+code in statically typed languages.
 The \texttt{Dynamic} type is a pair \texttt{(v,T)} where \texttt{v} is a
@@
 
-The \emph{quasi-static} type system proposed by \citet{thatte1990qst}
+The \emph{quasi-static} type system proposed by Thatte \cite{thatte1990qst}
 performs implicit coercions and run-time checks to replace the
 \texttt{dynamic} and \texttt{typecase} constructs that were proposed by
-\citet{abadi1989dts}.
+Abadi et al. \cite{abadi1989dts}.
 To do that, quasi-static typing relies on subtyping with a top type
@@
 
-\citet{ou2004dtd} specified a language that combines static types
-with dependent types.
+Ou et al. \cite{ou2004dtd} specified a language that combines
+static types with dependent types.
 To ensure safety, the compiler automatically inserts coercions
@@
 
-Optional type systems \citep{bracha2004pluggable} are an approach for
+Optional type systems \cite{bracha2004pluggable} are an approach for
 plugging static typing in dynamically typed languages.
 They use optional type annotations to perform compile-time type checking,
-though they do not influence the original run-time semantics
+though they do not affect the original run-time semantics
 of the language.
@@
 
-Gradual typing is similar to the approaches proposed by
-\citet{abadi1989dts} and \citet{thatte1990qst} by including a
-dynamic type in a statically typed language.
+Gradual typing \cite{siek2006gradual} is similar to two previous
+approaches \cite{abadi1989dts,thatte1990qst}, because they also include
+a dynamic type in a statically typed language.
 However, these three approaches differ in the way they handle the
 dynamic type.
-While \citet{siek2006gradual} rely on the consistency relation,
-\citet{abadi1989dts} rely on type equality with explicit projections
-and injections, and \citet{thatte1990qst} relies on subtyping.
+While Siek and Taha \cite{siek2006gradual} rely on the consistency relation,
+Abadi et al. \cite{abadi1989dts} rely on type equality with explicit
+projections plus injections, and Thatte \cite{thatte1990qst} relies on subtyping.
 
@@
 The semantics of these casts is based on the higher-order contracts
-proposed by \citet{findler2002chf}.
+proposed by Findler and Felleisen \cite{findler2002chf}.
 
-\citet{herman2007sgt} showed that there is an efficiency
-concern regarding the run-time checks, because there are two
-ways that casts can lead to unbounded space consumption.
+Herman et al. \cite{herman2007sgt} showed that there is an
+efficiency concern regarding the run-time checks, because there are
+two ways that casts can lead to unbounded space consumption.
 The first affects tail recursion while the second appears when
@@
 repeated casts to the same function or object.
-\citet{herman2007sgt} use the coercion calculus outlined in
-\citet{henglein1994dts} to express casts as coercions and
-solve the problem of space efficiency.
+Herman et al. \cite{herman2007sgt} use the coercion calculus
+outlined by Henglein \cite{henglein1994dts} to express casts
+as coercions and solve the problem of space efficiency.
 Their approach normalizes an arbitrary sequence of coercions to a
@@
 from the real error.
-\citet{wadler2009wpc} developed \emph{blame calculus} as a way
-to handle this issue, and \citet{ahmed2011bfa} extended
-blame calculus with polymorphism.
+Wadler and Findler \cite{wadler2009wpc} developed \emph{blame calculus}
+as a way to handle this issue, and Ahmed et al. \cite{ahmed2011bfa}
+extended blame calculus with polymorphism.
 Blame calculus is an intermediate language to integrate
 static and dynamic typing along with the blame tracking approach
-proposed by \citet{findler2002chf}.
+proposed by Findler and Felleisen \cite{findler2002chf}.
 
@@
 on the other hand, it has the space efficiency problem reported
-by \citet{herman2007sgt}.
-Thus, \citet{siek2009casts} extended the coercion calculus outlined in
-\citet{herman2007sgt} with blame tracking to achieve an
-implementation of the blame calculus that is space efficient.
-After that, \citet{siek2010blame} proposed a new solution that also
-handles both problems.
+by Herman et al. \cite{herman2007sgt}.
+Thus, Siek et al. \cite{siek2009casts} extended the coercion
+calculus outlined by Herman et al. \cite{herman2007sgt} with
+blame tracking to achieve an implementation of the blame calculus that
+is space efficient.
+After that, Siek and Wadler \cite{siek2010blame} proposed a new solution
+that also handles both problems.
 This new solution is based on a concept called \emph{threesome},
@@
 programming languages.
-Reticulated Python \citep{reticulated,vitousek2014deg} is a research
+Reticulated Python \cite{reticulated,vitousek2014deg} is a research
 project that evaluates the costs of gradual typing in Python.
-Gradualtalk \citep{allende2013gts} is a gradually-typed Smalltalk
+Gradualtalk \cite{allende2013gts} is a gradually-typed Smalltalk
 that introduces a new cast insertion strategy for gradually-typed
-objects \citep{allende2013cis}.
-Grace \citep{black2012grace,black2013sg} is a new object-oriented,
+objects \cite{allende2013cis}.
+Grace \cite{black2012grace,black2013sg} is a new object-oriented,
 gradually-typed, educational language.
 In Grace, modules are gradually-typed objects, that is, modules
-may have types and methods as attributes, and can have a state
-\citep{homer2013modules}.
-ActionScript \citep{moock2007as3} is one the first languages that
+may have types with methods as attributes, and they can also have
+a state \cite{homer2013modules}.
+ActionScript \cite{moock2007as3} is one the first languages that
 incorporated gradual typing to its implementation and
-Perl 6 \citep{tang2007pri} is also being designed with gradual typing,
+Perl 6 \cite{tang2007pri} is also being designed with gradual typing,
 though there is few documentation about the gradual type systems
@@
 typed, but unlike optional type systems, gradual typing changes
-the run-time semantics to preserve type safety, and it is a way to
-add a dynamic type to statically typed languages.
+the run-time semantics to preserve type safety.
 More precisely, programming languages that include a gradual type
-system implement the semantics of statically typed languages, so
+system can implement the semantics of statically typed languages, so
 the gradual type system inserts casts in the translated code to
@@
 programming languages that include an optional type system still
-implement the semantics of dynamically typed languages, so all
+need implement the semantics of dynamically typed languages, so all
 the type checking also belongs to the semantics of each operation.
@@
 that handles the interaction between Racket and Typed Racket modules.
+
+Recently, Siek et al. \cite{siek2015refined} described a formal
+criteria on what is gradual typing: the \emph{gradual guarantee}.
+Besides allowing static and dynamic typing in the same code
+along with type soundness, the gradual guarantee states that
+removing type annotations from a gradually typed program that is
+well typed must continue well typed.
+The other direction must be also valid, that is, adding correct type
+annotations to a gradually typed program that is well typed must
+continue well typed.
+In other words, the gradual guarantee states that any changes to
+the annotations does not change the static or the dynamic behavior
+of a program \cite{siek2015refined}.
+The authors prove the gradual guarantee and discuss whether
+some previous projects match this criteria.
 
@@
 functions and vararg expressions.
+
+We end this chapter presenting a summary of the statistics that
+we just presented:
+\begin{itemize}
+\item table constructor
+\begin{itemize}
+\item only use literal keys: 73\%
+\item empty: 26\%
+\item only use non-literal keys: 1\%
+\end{itemize}
+\item table indexing
+\begin{itemize}
+\item reads: 86\%
+\begin{itemize}
+\item use literals: 93\%
+\item use non-literals: 7\%
+\end{itemize}
+\item writes: 14\%
+\begin{itemize}
+\item use literals: 71\%
+\item use non-literals: 29\%
+\end{itemize}
+\end{itemize}
+\item iteration over tables
+\begin{itemize}
+\item files that call \texttt{pairs}: 23\%
+\item files that call \texttt{ipairs}: 21\%
+\item files that use the numeric for along with the length operator: 10\%
+\end{itemize}
+\item function declarations
+\begin{itemize}
+\item perform dynamic type checking in their input parameters: 9\%
+\begin{itemize}
+\item asserting input parameters: 50\%
+\item overloading input parameters: 50\%
+\end{itemize}
+\item return multiple values: 10\%
+\begin{itemize}
+\item return \texttt{nil} plus something else: 50\%
+\item return \texttt{false} plus something else: 10\%
+\item return other multiple values: 40\%
+\end{itemize}
+\item are method declarations: 23\%
+\item are variadic: 8\%
+\end{itemize}
+\item object-oriented programming
+\begin{itemize}
+\item projects that extend metatables and declare methods: 45\%
+\end{itemize}
+\item modules
+\begin{itemize}
+\item files that define modules: 61\%
+\item files that are plain scripts: 39\%
+\end{itemize}
+\end{itemize}

### 2015-05-23 doc/thesis/typedlua.tex
Adding a statistics summary and a brief review about gradual guarantee

@@
 We will discuss these features in Section \ref{sec:unions} and
-Section \ref{sec:tables}, respectivelly.
+Section \ref{sec:tables}, respectively.
 
@@
 For this reason, we need to annotate the return type of
-recursive local functions to inform the type cheker what type
+recursive local functions to inform the type checker what type
 it should use while type checking their body.
@@
 local variables will not change their type.
-Next we show how we can apply this alteranative solution to the
+Next we show how we can apply this alternative solution to the
 previous example, making it compile without any warnings:
@@
 We need this kind of behavior to avoid unsafe constructions
-while aliasing \emph{unique} table types, like in the follwoing
+while aliasing \emph{unique} table types, like in the following
 example:

### 2015-05-23 doc/thesis/glossary.tex
Some improvements to glossary

@@
 
-\item[closed table type] It is a table type that does not allow adding fields.
+\item[closed table type] It is the type of table annotations in Typed Lua.
+A closed table type does not provide any guarantees that about keys
+with types not listed in the table type, and it also does not
+allow adding fields to the table type.
 
@@
 
-\item[consistency] It is a relation that gradual typing uses to control the insertion of
-run-time checks between dynamically typed and statically typed code.
+\item[consistency] Gradual typing uses the consistency relation instead of type equality
+to check the interaction among the dynamic type and other types.
+This relation allows us combining dynamic and static typing in the
+same language, but still catching static type errors.
+The consistency relation is reflexive and symmetric, but it is not transitive.
 
-\item[consistent-subtyping] It is a relation that combines consistency and subtyping.
+\item[consistent-subtyping] It is a relation that combines consistency and subtyping, allowing
+the definition of gradual type systems for object-oriented languages.
+The consistent-subtyping relation is reflexive and symmetric, but it is not transitive.
 
 \item[contravariant] Subtyping is contravariant when it reverses the subtyping order,
-that is, it orders types from a more generic type to a more specific type.
+that is, a subtyping rule is contravariant when it orders
+types from more generic types (supertypes) to more specific types (subtypes).
 
 \item[covariant] Subtyping is covariant when it preserves the subtyping order,
-that is, it orders types from a more specific type to a more general type.
+that is, a subtyping rule is covariant when it orders
+types from more specific types (subtypes) to more general types (supertypes).
 
-\item[depth subtyping] It allows changing the type of individual fields of a table type.
+\item[depth subtyping] It allows the supertype to change the type of individual fields in the subtype.
 
@@
 
-\item[filter type] It is a type that allows Typed Lua to discriminate the type of local variables
-inside control flow statements.
+\item[filter type] It is a type that allows Typed Lua to discriminate and restore the type of
+local variables inside control flow statements.
 
-\item[fixed table type] It is a table type that does not allow adding fields. It also does not allow
-width subtyping to make single inheritance safe.
+\item[fixed table type] It is a table type that does not allow width subtyping to make single
+inheritance safe in Typed Lua, and it also does not allow adding fields to
+the table type.
 
-\item[flow typing] It is a combination of static typing and flow analysis.
+\item[flow typing] It is a combination of static typing and flow analysis to allow variables
+to have different types at different parts of the program.
 
@@
 
-\item[gradual type system] It is a type system that uses the consistency relation to check the interaction
-among the dynamic type and other types.
+\item[gradual type system] It is a type system that uses the consistency relation instead of type equality
+to check the interaction among the dynamic type and other types.
 
@@
 typed and dynamically typed code.
+This run-time checks ensure that dynamically typed code does not violate
+statically typed code.
 
 \item[invariant] Subtyping is invariant when it does not allow ordering types, that is,
-it is a way that express type equality through subtyping.
+it is a way to define type equality through subtyping.
 
-\item[metatable] It is a Lua table that allows changing its behavior.
+\item[metatable] It is a Lua table that allows changing the behavior of other tables
+it is attached to.
 
-\item[nominal type system] It is a type system that uses name declarations to define the equivalence among types.
+\item[nominal type system] It is a type system that uses the name of the types to check the
+compatibility among them.
 
-\item[open table type] It is a table type that allows adding fields, and usually refers to table types that
-have aliases.
+\item[open table type] It is the type of the tables with keys that do not inhabit one of
+the table's key types, have at least one alias, and allows adding
+fields to the table type.
 
@@
 
-\item[projection environment] It is a type environment that Typed Lua uses to assign projection variables to
+\item[projection environment] It is an environment that Typed Lua uses to assign projection variables to
 second-level types.
@@
 \item[projection type] It is a type that allows Typed Lua to discriminate the type of local variables
-that have a dependency relation.
+that have a dependency relation, as they include a projection variable and an
+index that allow Typed Lua to project unions of second-level types into
+unions of first-level types.
 
-\item[prototype object] It is an object that works like a class, that is, it is a prototype for creating
-new objects of a given class.
+\item[prototype object] It is an object that works like a class, that is, it is an object from
+which other objects inherit its attributes.
 
-\item[self-like delegation] It is a technique to implement inheritance in dynamically typed languages.
+\item[self-like delegation] It is a technique to implement inheritance in dynamically typed languages
+through prototype objects.
 
@@
 
-\item[structural type system] It is a type system that uses the type structure to define the equivalence among types.
+\item[structural type system] It is a type system that uses the structure of types to check the compatibility among them.
 
@@
 
-\item[type environment] It is an environment that assigns variable names to types.
+\item[type environment] It is an environment that Typed Lua uses to assign variable names to first-level types.
 
@@
 
-\item[unique table type] It is a table type that allows adding fields, and it is also the table type
-that describe the type of the table constructor.
+\item[unique table type] It is the type of the tables with keys that do not inhabit one of
+the table's key types, does not have any alias, and allow adding
+fields to the table type.
+It is also the table type that describes the type of the table constructor.
 
@@
 
-\item[userdata] It is a way to define new types in an application or library that is written in C.
+\item[userdata] It is a Lua data type that allows Lua variables to hold values from applications
+or libraries that are written in C.
 
@@
 
-\item[width subtyping] It allows adding more fields to a table type.
+\item[width subtyping] It allows the subtype to have fields that do not exist in the supertype.
 

### 2015-05-27 doc/thesis/typedlua.tex
Fixing the text for bibliographic data in the PUC template

@@
     local Shape = { x = 0.0, y = 0.0 }
+
     const function Shape:new (x:number, y:number):self
@@
     end
+
     const function Shape:move (dx:number, dy:number):()
@@
     end
+
     return Shape
@@
     local Shape = require "shape"
+
     local Circle = setmetatable({}, { __index = Shape })
+
     Circle.radius = 0.0
+
     const function Circle:new (x:number,
@@
     end
+
     const function Circle:area ():number
@@
     end
+
     return Circle

### 2015-06-01 doc/thesis/typedlua.tex
Minor improvements in the thesis text

@@
 original type.
-In the case of \texttt{greeting}, the assignment change its type
+In the case of \texttt{greeting}, the assignment changes its type
 from \texttt{string|nil} to \texttt{string}.
@@
 $\texttt{l}_{\texttt{1}}$ which maps to a value of type $\texttt{l}_{\texttt{2}}$.
-In our example, the expression \texttt{e.tag == "Add"} makes Typed Lua
-constrain the type of \texttt{e} to \texttt{\{"tag":"Add", 1:Exp, 2:Exp\}}.
+In our example, the expression \texttt{e.tag == "Add"} makes the Typed Lua
+compiler constrain the type of \texttt{e} to \texttt{\{"tag":"Add", 1:Exp, 2:Exp\}}.
 If we did not add this special predicate, we would not be able to traverse an AST,

### 2015-06-04 doc/papers/paper/paper.tex

@@
 some features with other scripting languages such as JavaScript.
-
-%This work is split into seven chapters.
-%In Chapter \ref{chap:review} we review the literature about blending
-%static and dynamic typing in the same language, we discuss the differences
-%between optional type systems and gradual typing, and we also
-%present the results of our survey on Lua idioms.
-%In Chapter \ref{chap:typedlua} we use code examples to present the
-%design of Typed Lua.
-%In Chapter \ref{chap:system} we present our type system.
-%In Chapter \ref{chap:evaluation} we discuss the evaluation
-%results that we obtained while using Typed Lua to type existing Lua code.
-%In Chapter \ref{chap:related} we present some related work.
-%In Chapter \ref{chap:conc} we outline our contributions.
 
@@
 
-\subsubsection{Assignment and function application}
-
-Lua has multiple assignment, and our type system uses three
-different kinds of typing rules to type check this feature.
-It uses typing rules that type check the different forms
-of expression lists that can appear in the right-hand side,
-a typing rule that type checks a list of left-hand values,
-and a general rule that uses consistent-subtyping to check
-whether the type of the right-hand side is consistent with
-the type of the left-hand side.
-
-As an example, lets assume that $x$ and $y$ are variables in the
-environment with types $\Integer$ and $\String$.
-Let us see how our type system type checks the following assignment:
-\[
-x, y = 1, ``foo"
-\]
-
-First, our type system type checks the expression list in the right-hand
-side of the assignment.
-In our example, the right-hand side of the assignment has type
-$1 \times ``foo" \times \Nil{*}$.
-Note that our type system includes the type $\Nil{*}$ to replace
-missing values.
-The rules that type check expression lists introduce the type
-$\Nil{*}$ to let the right-hand side produce fewer values than
-expected in the left-hand side.
-Our example uses the rule \textsc{T-EXPLIST1} to type check
-the right-hand side of the assignment.
-The rule \textsc{T-EXPLIST1} is the rule that type checks an
-expression list where all expressions can only produce a single value:
-\[
-\begin{array}{c}
-\mylabel{T-EXPLIST1}\\
-\dfrac{\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
-       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
-       n = |\;\overline{e}\;|}
-      {\env_{1}, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
-\end{array}
-\]
-
-In Section \ref{sec:refinement} we will show that table refinement can
-change the type environment while typing an expression or a left-hand value.
-Thus, the rules that type check lists of expressions and lists of
-left-hand values use a partial auxiliary function \emph{merge} to collect
-all environment changes in a new environment $\env_{m}$, if there are no conflicts.
-In Section \ref{sec:refinement} we will also show that we can only change the
-environment to add new table fields in a table type, and we cannot change the
-type of a variable or a table field which is already present in a table type.
-
-After type checking the right-hand side, our type system type checks the list
-of left-hand values.
-In our example, the left-hand side of the assignment has type 
-$\Integer \times \String \times \Value{*}$.
-Note that our type system uses the type $\Value{*}$ to discard extra values.
-The rule that type checks lists of left-hand values introduces the type
-$\Value{*}$ to let the right-hand side produce more values than
-expected in the left-hand side.
-Our example uses the rule \textsc{T-LHSLIST} to type
-check a list of left-hand values:
-\[
-\begin{array}{c}
-\mylabel{T-LHSLIST}\\
-\dfrac{\env_{1}, \penv \vdash l_{i}:T_{i}, \env_{i+1} \;\;\;
-       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
-       n = |\;\overline{l}\;|}
-      {\env_{1}, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
-\end{array}
-\]
-
-After type checking the right-hand side and the left-hand side of an assignment,
-our type system checks whether their types are consistent.
-The rule \textsc{T-ASSIGNMENT} is the general rule that express this idea:
-\[
-\begin{array}{c}
-\mylabel{T-ASSIGNMENT}\\
-\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash \overline{l}:S_{2}, \env_{3} \;\;\;
-       S_{1} \lesssim S_{2}}
-      {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
-\end{array}
-\]
-
-Back to our example, it type checks through rule \textsc{T-ASSIGNMENT} because
-\[
-1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times \Value{*}
-\]
-
-As another example, lets assume that $x$, $y$, and $z$ are variables in
-the environment with types $\Integer$, $\String$, and $\String \cup \Nil$.
-The assignment
-\[
-x, y, z = 1, ``foo"
-\]
-type checks because
-\[
-1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times (\String \cup \Nil) \times \Value{*}
-\]
-
-Note how $\Nil{*}$ replaces any missing values.
-This example type checks because $\Nil{*}$ produces as many $\Nil$
-values as we need, and $\Nil$ is consistent with $\String \cup \Nil$,
-which is the type of $z$.
-
-Conversely, the assignment
-\[
-x = 1, ``foo"
-\]
-type checks because
-\[
-1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \Value{*}
-\]
-
-Note how $\Value{*}$ discards extra values.
-This example type checks because $\Value{*}$ discards as many extra
-values as we need, and $``foo"$ is consistent with $\Value$.
-
-Our type system includes $\Nil{*}$ in the type of an expression list only
-if its type does not end in another variadic type $T{*}$.
-For instance, rules \textsc{T-EXPLIST2} and \textsc{T-EXPLIST3} handle
-the case where an expression list is just an expression that may produce
-multiple values.
-The former rule includes the type $\Nil{*}$, but the latter rule does not:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-EXPLIST2}\\
-\dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \\
-       \env_{1}, \penv \vdash me:T_{n+1} \times ... \times \Void, \env_{n+2} \\
-       n = |\;\overline{e}\;| \;\;\;
-       \env_{m} = merge(\env_{1}, ..., \env_{n+2})
-       \end{array}}
-      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n+1} \times ... \times \Nil{*}, \env_{m}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-EXPLIST3}\\
-\dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \\
-       \env_{1}, \penv \vdash me:T_{n+1} \times ... \times T{*}, \env_{n+2} \\
-       n = |\;\overline{e}\;| \;\;\;
-       \env_{m} = merge(\env_{1}, ..., \env_{n+2})
-       \end{array}}
-      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n+1} \times ... \times T{*}, \env_{m}}
-\end{array}
-\end{array}
-\]
-
-Rules for function applications are similar to the rule for multiple assignment.
-The rule \textsc{T-APPLY1} handles the case where function applications
-are expressions that produce multiple values:
-\[
-\begin{array}{c}
-\mylabel{T-APPLY1}\\
-\dfrac{\env_{1}, \penv \vdash e:S_{1} \rightarrow S_{2}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash el:S_{3}, \env_{3} \;\;\;
-       S_{3} \lesssim S_{1}}
-      {\env_{1}, \penv \vdash e(el):S_{2}, \env_{3}}
-\end{array}
-\]
-
-We also use the rule \textsc{T-APPLY1} as the base case for the rules
-that handle the cases where function applications are either statements
-that produce no value or expressions that produce only one value.
-The rule \textsc{T-STMAPPLY1} discards the produced values,
-while the rule \textsc{T-EXPAPPLY1} uses the auxiliary function
-\emph{first} to ensure that only one value is produced:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-STMAPPLY1}\\
-\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
-      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{0},\env_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{T-EXPAPPLY1}\\
-\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
-      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{1}:first(S), \env_{2}}
-\end{array}
-\end{array}
-\]
-
-We can define \emph{first} inductively as follows:
-\begin{align*}
-first(\Void) & = \Nil\\
-first(T{*}) & = T \cup \Nil\\
-first(T \times P) & = T\\
-first(S_{1} \sqcup S_{2}) & = first(S_{1}) \cup first(S_{2})
-\end{align*}
-
-As an example, let us assume that $f$ is a local function in the environment,
-and that $f$ has type $\String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Value{*} \rightarrow \Integer{*}$.
-The function call
-\[
-f(``foo")
-\]
-type checks through the rule \textsc{T-APPLY1}, because
-\[
-``foo" \times \Nil{*} \lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Value{*}
-\]
-and the function call
-\[
-f(``foo",1,2,3)
-\]
-also type checks through the rule \textsc{T-APPLY1}, because
-\[
-``foo" \times 1 \times 2 \times 3 \times \Nil{*} \lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Value{*}
-\]
-
-Our type system also catches arity mismatch.
-To do that, we end the input type of a function with type $\Void$
-instead of $\Value{*}$.
-For instance, let us assume that $f$ has type
-$\String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void \rightarrow \Integer{*}$.
-The function call
-\[
-f(``foo")
-\]
-type checks through the rule \textsc{T-APPLY1}, because
-\[
-``foo" \times \Nil{*} \lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void
-\]
-but the function call
-\[
-f(``foo",1,2,3)
-\]
-does not type check through the rule \textsc{T-APPLY1}, because
-\[
-``foo" \times 1 \times 2 \times 3 \times \Nil{*} \not\lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void
-\]
-
-We just mentioned that when our type system type checks an expression list,
-it always includes $\Nil{*}$ in the end of the type of this expression list
-if its type does not end in a variadic type.
-This behavior preserves the semantics of Lua on replacing missing values,
-and it is necessary when we omit optional parameters in a function call,
-like the previous example showed.
-
-Using $\Nil{*}$ in the end of the type of expression lists also allows
-our type system to catch arity mismatch in function calls without optional parameters.
-For instance, let us assume that $f$ has type
-$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Void$.
-The function call
-\[
-f(1)
-\]
-does not type check through the rule \textsc{T-APPLY1}, because
-\[
-1 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Void
-\]
-and the function call
-\[
-f(1,2,3)
-\]
-also does not type check through the rule \textsc{T-APPLY1}, because
-\[
-1 \times 2 \times 3 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Void
-\]
-
 \subsubsection{Tables and refinement}
@@
 
-\section{Evaluation}
-
-We performed some case studies on existing Lua libraries
-to evaluate the design of our type system.
-For each library, we used Typed Lua to either annotate its modules
-or to write statically typed interfaces to its modules through
-Typed Lua's description files.
-In this chapter we present our evaluation results and discuss some
-interesting cases.
-
-The Lua Standard Library \cite{luamanual} was our first case study.
-We started to think about how we would type its modules at the same time
-that we started to design our type system, as it could give us some
-hints on our type system.
-And it did: optional parameters and overloading on the return type
-are two Lua features that our type system should handle to allow us
-typing some of the functions that the standard library implements.
-
-The second case study that we chose was the MD5 library \cite{lmd5},
-because we wanted a simple case study to introduce Typed Lua's description
-files and \texttt{userdata} declarations.
-These Typed Lua's mechanisms allow programmers to give statically typed
-interfaces to Lua libraries.
-
-LuaSocket \cite{luasocket} and LuaFileSystem \cite{luafilesystem} were
-the third and fourth case studies that we used to evaluate Typed Lua.
-We chose them because they are the most popular Lua libraries.
-We wrote a script that builds the dependency graph of Lua libraries
-that are in the LuaRocks repository, and uses this dependency graph
-to identify the most popular Lua libraries.
-
-We also randomly selected three case studies from the LuaRocks
-repository, they are: HTTP Digest \cite{luahttpdigest},
-Typical \cite{luatypical}, and Mod 11 \cite{luamod11}.
-The first provides client side HTTP digest authentication for Lua.
-The second is an extension to the primitive function \texttt{type}.
-The third is a generator and checker of modulo 11 numbers.
-We randomly selected three case studies because we wanted to evaluate
-Typed Lua for annotating existing libraries that are written in Lua,
-as the previous case studies are mostly libraries that are written in C.
-
-The Typed Lua compiler is the last case study that we evaluated.
-We chose it as a case study because it is a large application.
-Besides, it is a case study that evaluates the evolution of a script
-to a program.
-
-We used these case studies to evaluate two aspects of Typed Lua:
-\begin{enumerate}
-\item how precisely it can describe the type of the interface of a module;
-\item whether it provides guarantees that the code matches the interface.
-\end{enumerate}
-
-\begin{table}[!ht]
-\begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|}
-\hline
-\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
-\hline
-Lua Standard Library & 64\% & 5\% & 8\% & 23\% & 129 \\ % 100%
-\hline
-MD5 & 100\% & 0\% & 0\% & 0\% & 13 \\ % 100%
-\hline
-LuaSocket & 89\% & 1\% & 2\% & 8\% & 123 \\ % 100%
-\hline
-LuaFileSystem & 89\% & 0\% & 11\% & 0\% & 19 \\ % 100%
-\hline
-HTTP Digest & 0\% & 0\% & 100\% & 0\% & 1 \\ % 100%
-\hline
-Typical & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
-\hline
-Modulo 11 & 78\% & 0\% & 0\% & 22\% & 9 \\ % 100%
-\hline
-Typed Lua Compiler & 93\% & 0\% & 1\% & 6\% & 154 \\ % 100%
-\hline
-\end{tabular}
-\end{center}
-\caption{Evaluation results for each case study}
-\label{tab:evalbycase}
-\end{table}
-
-Table \ref{tab:evalbycase} summarizes our evaluation results for each
-of the case studies that we used Typed Lua for typing their members.
-An exported member is any Lua value that a module might export.
-We split the members of each case study into four categories:
-\emph{easy}, \emph{poly}, \emph{over}, and \emph{hard}.
-The \emph{easy} category shows the percentage of members that
-we could give a precise static type.
-The \emph{poly} category shows the percentage of members that
-we made minimal use of the dynamic type as a replacement for the
-lack of type parameters.
-The \emph{over} category shows the percentage of members that
-we gave a static type that is not precise enough,
-as these members are overloaded functions that require intersection
-types to describe their precise static types.
-The \emph{hard} category shows the percentage of members that
-we could not give a precise static type.
-The last column of the table shows the total number of members
-of each case study.
-The results that we obtained for the \emph{easy} category give
-a lower bound on how much static type safety
-we could add to each one of our case studies.
-
-In the following sections we discuss each case study in more detail.
-For each case study, we split the evaluation results according
-to the modules that each one of them include.
-We use these split results to discuss the contributions and
-limitations of our type system.
-
-\subsection{Lua Standard Library}
-
-All of the modules in the standard library are implemented in C, so
-we used Typed Lua to type just the interface of each module.
-The \texttt{debug} module is the only one that we did not include in our
-evaluation results, because it provides several functions that violate
-basic assumptions about Lua code \cite{luamanual}.
-For instance, we can use the function \texttt{debug.setlocal} to change the value
-of a local variable that is not visible in the current scope.
-Table \ref{tab:evallsl} summarizes the evaluation results for the Lua Standard Library
-(version 5.3).
-
-\begin{table}[!ht]
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
-\hline
-\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
-\hline
-base & 35\% & 4\% & 8\% & 53\% & 26 \\ % 100%
-\hline
-coroutine & 14\% & 0\% & 0\% & 86\% & 7 \\ % 100%
-\hline
-package & 62\% & 0\% & 0\% & 38\% & 8 \\ % 100%
-\hline
-string & 75\% & 0\% & 0\% & 25\% & 16 \\ % 100%
-\hline
-utf8 & 100\% & 0\% & 0\% & 0\% & 6 \\ % 100%
-\hline
-table & 14\% & 72\% & 14\% & 0\% & 7 \\ % 100%
-\hline
-math & 81\% & 0\% & 19\% & 0\% & 27 \\ % 100%
-\hline
-io & 81\% & 0\% & 0\% & 19\% & 21 \\ % 100%
-\hline
-os & 82\% & 0\% & 18\% & 0\% & 11 \\ % 100%
-\hline
-\end{tabular}
-\end{center}
-\caption{Evaluation results for Lua Standard Library}
-\label{tab:evallsl}
-\end{table}
-
-The \texttt{base} module was very difficult to type
-because it includes several functions that rely on reflection,
-as the \emph{hard} category shows.
-For instance, the functions \texttt{pairs} and \texttt{getmetatable}
-are in this category.
-While \texttt{pairs} traverses all keys and values that are stored
-in a given table, \texttt{getmetatable} returns the metatable of a
-given table.
-
-There are some functions in the \texttt{base} module that
-we could not give a precise static type because our type
-system does not have parametric polymorphism,
-as the \emph{poly} category shows.
-This is the case of \texttt{ipairs}.
-
-The \texttt{base} module also includes some overloaded functions,
-as the \emph{over} category shows.
-We could not type these functions because our type system does
-not include intersection types.
-This is the case of \texttt{tonumber} and \texttt{collectgarbage}.
-
-In the case of \texttt{tonumber}, it has two different types:
-\texttt{(value) -> (number)} and \texttt{(string, integer) -> (number)}.
-This means that the type of the first parameter depends on the
-type of the second parameter.
-For instance, we can call \texttt{tonumber(1)}, but we cannot
-call \texttt{tonumber(1,2)}.
-Note that the first argument of \texttt{tonumber} can be
-a value of any type if it is the only argument, but it must
-be a string if there is a second argument,
-which must be an integer.
-
-In the case of \texttt{collectgarbage}, the return type changes
-according to an input literal string.
-For instance, calling \texttt{collectgarbage("collect")} returns an integer,
-calling \texttt{collectgarbage("count")} returns a floating point,
-and calling \texttt{collectgarbage("isrunning")} returns a boolean.
-
-The \texttt{coroutine} module was also very difficult to type
-because our type system cannot describe the computational effects
-of a program.
-The \emph{hard} category shows the amount of functions that we
-could not give a precise static type for this reason.
-Lua has one-shot delimited continuations \cite{james2011yield}
-in the form of \emph{coroutines} \cite{moura2009rc}, and
-effect systems \cite{nielson1999type} are an approach that we
-could use to describe control transfers with continuations.
-However, for now, coroutines are out of scope of our type
-system, and we use an empty \texttt{userdata} declaration
-to represent the type \texttt{thread}.
-
-Still, we could give a precise static type to one function in
-the \texttt{coroutine} module, as the \emph{easy} category shows.
-The function \texttt{coroutine.isyieldable} has no input parameters,
-and it simply returns a boolean that indicates whether the running
-coroutine is yieldable.
-
-We could give precise static types to the constants and functions
-of the \texttt{package} module, but we could not give precise static
-types to the following tables that it exports:
-\texttt{package.loaded}, \texttt{package.preload}, and \texttt{package.loaders}.
-The first stores loaded modules, while the others store module \emph{loaders}.
-They are difficult to type because their types rely on reflection,
-that is, their types depend on the modules a program loads.
-For this reason, they are in the \emph{hard} category.
-
-We could give precise static types to most of the functions of the
-\texttt{string} module, but we could not give precise static types
-to the functions that rely on format strings.
-For instance, the type of the arguments that we pass to
-\texttt{string.format} must match the format string we are using.
-It is fine to call \texttt{string.format("\%d", 1)}, but
-\texttt{string.format("\%d", true)} raises a run-time error.
-These functions that rely on format strings are in the \emph{hard} category.
-
-The \texttt{utf8} module was straightforward to type,
-as its members are only operations over strings.
-
-The \texttt{table} module was specially difficult to type because
-most of its functions require parametric polymorphism,
-as the \emph{poly} category shows.
-These functions either receive or return a list of elements, and
-parametric polymorphism would help us to describe them with a generic type.
-
-However, the lack of parametric polymorphism did not prevent us from
-giving a precise type to one function of the \texttt{table} module,
-as the \emph{easy} category shows.
-We could give a precise static type to \texttt{table.concat},
-as it operates over lists where all elements are strings or numbers.
-
-Even if our type system had parametric polymorphism, there is
-still one function of the \texttt{table} module that we could
-not give a precise static type because it is an overloaded function,
-as the \emph{over} category shows.
-This function is \texttt{table.insert}, and its type depends on
-the calling arity.
-That is, calling \texttt{table.insert(l, v)} inserts the
-element \texttt{v} at the end of the list \texttt{l},
-while \texttt{table.insert(l, p, v)} inserts the element \texttt{v}
-at the position \texttt{p} of the list \texttt{l},
-and generates a run-time error when \texttt{p} is out of bounds.
-This function also does not follow the semantics of Lua on
-discarding extra arguments, and generates a run-time error whenever
-we pass more than three arguments, even if the first three arguments
-match its signature.
-
-Even though the \texttt{math} module looks straightforward to type,
-the \emph{over} category shows that it includes several overloaded functions.
-For instance, the functions \texttt{math.abs} and \texttt{math.random}
-are in this category.
-
-In the case of \texttt{math.abs},
-it has two different types: \texttt{(integer) -> (integer)} and
-\texttt{(number) -> (number)}.
-Even though we gave this function the more general type \texttt{(number) -> (number)},
-it is not precise enough because the return type is always \texttt{number}
-independently of the argument type.
-In other words, the return type should be \texttt{integer} when
-the argument type is also \texttt{integer}, and the return type
-should be \texttt{number} when the argument type is also \texttt{number}.
-
-In the case of \texttt{math.random}, it also has two different types:
-\texttt{() -> (number)} and \texttt{(integer, integer?) -> (integer)}.
-This means that the type of \texttt{math.random} depends on the calling arity.
-Calling \texttt{math.random()} returns a random floating point
-between \texttt{0} and \texttt{1}.
-Calling \texttt{math.random(10)} is equivalent to \texttt{math.random(1,10)},
-and returns an integer between this interval.
-Like \texttt{table.insert}, this function also does not follow the
-semantics of Lua on discarding extra arguments,
-and generates a run-time error whenever we pass more than two arguments.
-
-The \texttt{io} module provides operations for manipulating files,
-and these operations can use implicit or explicit file descriptors.
-The implicit operations are functions in the \texttt{io} table,
-while the explicit operations are methods of a file descriptor.
-We used an \texttt{userdata} declaration to introduce the type
-\texttt{file} for representing the type of a file descriptor
-and its methods.
-The evaluation results include both implicit and explicit operations.
-
-We could give precise static types to most of the members of the
-\texttt{io} module, but the \emph{hard} category shows that it
-includes some members that we could not give a precise static type.
-The functions \texttt{io.read} and \texttt{io.lines} are
-in the \emph{hard} category along with the methods
-\texttt{file:read} and \texttt{file:lines}.
-
-We could not precisely type \texttt{io.read} because its return type
-relies on format strings.
-For instance, calling \texttt{io.read("l")} returns a string or \texttt{nil},
-\texttt{io.read("n")} returns a number or \texttt{nil}, and
-\texttt{io.read("l", "n")} returns a string or \texttt{nil} and a number or \texttt{nil}.
-The function \texttt{io.lines}, and the methods \texttt{file:read} and
-\texttt{file:lines} have the same issue.
-
-There are two functions in the \texttt{os} module that we could
-not give a precise static type because they are overloaded functions.
-The functions that are in the \emph{over} category are
-\texttt{os.date} and \texttt{os.execute}.
-
-The evaluation results show that our type system should include
-intersection types, parametric polymorphism, and effect types,
-as these features would help us increase the static typing of the
-Lua Standard Library.
-Intersection types would allow us to define overloaded function types.
-Parametric polymorphism would allow us to define generic function and table types.
-Effect types would allow us to type coroutines.
-
-\subsection{MD5}
-
-The MD5 library is an OpenSSL based message digest library for Lua.
-It contains just the \texttt{md5} module that is written in C,
-so we used Typed Lua's description file to type it.
-Table \ref{tab:evalmd5} summarizes the evaluation results for MD5.
-
-\begin{table}[!ht]
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
-\hline
-\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
-\hline
-md5 & 100\% & 0\% & 0\% & 0\% & 13 \\ % 100%
-\hline
-\end{tabular}
-\end{center}
-\caption{Evaluation results for MD5}
-\label{tab:evalmd5}
-\end{table}
-
-Even tough it was straightforward to type the MD5 library,
-we found a little difference between its documentation and its behavior.
-The documentation suggests that the type of the function \texttt{md5.update}
-is \texttt{(md5\string_context, string) -> (md5\string_context)},
-though there is a call to this function in the test script that passes
-an extra string argument.
-Reading the source code, we found that its actual type is
-\texttt{(md5\string_context, string*) -> (md5\string_context)},
-that is, we can pass zero or more strings to \texttt{md5.update}.
-
-This case study shows that type annotations help programmers maintain the
-documentation updated, as the type checker always validates them.
-
-\subsection{LuaSocket}
-
-LuaSocket is a library that adds network support to Lua,
-and it is split into two parts: a core that is written in C and a set of
-Lua modules.
-The C core provides TCP and UDP support, while the Lua modules provide
-support for SMTP, HTTP, and FTP client protocols, MIME encoding,
-URL manipulation, and LTN12 filters \cite{nehab2008ltn012}.
-We used Typed Lua's description files to type both parts, as we also
-wanted to use LuaSocket to test description files to statically type
-the interface of modules that are written in Lua.
-Table \ref{tab:evalsocket} summarizes the evaluation results for LuaSocket.
-
-\begin{table}[!ht]
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
-\hline
-\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
-\hline
-socket & 83\% & 0\% & 0\% & 17\% & 60 \\ % 100%
-\hline
-ftp & 83\% & 0\% & 17\% & 0\% & 6 \\ % 100%
-\hline
-http & 80\% & 0\% & 20\% & 0\% & 5 \\ % 100%
-\hline
-smtp & 100\% & 0\% & 0\% & 0\% & 7 \\ % 100%
-\hline
-mime & 100\% & 0\% & 0\% & 0\% & 17 \\ % 100%
-\hline
-ltn12 & 95\% & 5\% & 0\% & 0\% & 20 \\ % 100%
-\hline
-url & 100\% & 0\% & 0\% & 0\% & 8 \\ % 100%
-\hline
-\end{tabular}
-\end{center}
-\caption{Evaluation results for LuaSocket}
-\label{tab:evalsocket}
-\end{table}
-
-We could give precise static types to most of the members
-in the \texttt{socket} module, which is the C core.
-However, this module includes some functions that
-we could not give a precise static type because they
-rely on reflection, as the \emph{hard} category shows.
-For instance, \texttt{socket.skip} is a function that is
-in this category.
-We can use this function to choose the number of
-values that we want to return.
-As an example, calling \texttt{socket.skip(1, nil, "hello")}
-returns only the string \texttt{"hello"}, because \texttt{1} indicates
-that we do not want to return the first value.
-Passing a negative number to \texttt{socket.skip} can be
-dangerous, as it returns anything that might be in the stack.
-As an example, calling \texttt{socket.skip(-1, nil, "hello")}
-returns the tuple \texttt{(-1, nil, "hello")}, because \texttt{-1} makes
-\texttt{socket.skip} not skip any values.
-As another example, \texttt{f = socket.skip(-2)} assigns \texttt{socket.skip}
-to \texttt{f}, as \texttt{-2} gets \texttt{socket.skip} from the stack.
-Our type system cannot handle the type of negative numbers, as
-this requires more complex types such as the refinement types from
-hybrid type checking \cite{flanagan2006htc}.
-
-We could give precise static types to most of the members of the
-modules \texttt{ftp} and \texttt{http}, but we could not precisely type
-two overloaded functions: \texttt{ftp.get} and \texttt{http.request}.
-
-The function \texttt{ftp.get} downloads data from a given URL,
-which can be either a string or a table.
-More precisely, \texttt{ftp.get(url)} returns the tuple
-\texttt{(string) | (nil, string)} if \texttt{url} is a string,
-and it returns the tuple \texttt{(number) | (nil, string)} if
-\texttt{url} is a table.
-
-The function \texttt{http.request} downloads data from a given URL,
-which can be either a string or a table.
-More precisely, \texttt{http.request(url, body)} returns the tuple
-\texttt{(string, number, \{string:string\}, number) | (nil, string)}
-if \texttt{url} is a string and \texttt{body} is another string or \texttt{nil},
-but it returns the tuple
-\texttt{(number, number, \{string:string\}, number) | (nil, string)}
-if \texttt{url} is a table and \texttt{body} is \texttt{nil}.
-
-The modules \texttt{mime} and \texttt{ltn12} have a strong connection.
-The \texttt{mime} module offers low-level and high-level filters
-that apply and remove some text encodings.
-The low-level filters are written in C, while the high-level filters
-use the function \texttt{ltn12.filter.cycle} along with the low-level
-filters to create standard filters.
-
-Even though we could type all the members of the \texttt{mime} module,
-the function \texttt{ltn12.filter.cycle} is the only member of the
-\texttt{ltn12} module that we could not give a precise type.
-This function is difficult to type because it is polymorphic.
-
-The modules \texttt{smtp} and \texttt{url} were straightforward to type.
-The \texttt{smtp} module provides functions that send e-mails.
-The \texttt{url} module provides functions that manipulate URLs.
-
-\subsection{LuaFileSystem}
-
-LuaFileSystem is a library that extends the set of functions
-for manipulating file systems in Lua.
-It contains just the \texttt{lfs} module that is written in C,
-so we used Typed Lua's description files to type it.
-Table \ref{tab:evallfs} summarizes the evaluation results for LuaFileSystem.
-
-\begin{table}[!ht]
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
-\hline
-\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
-\hline
-lfs & 89\% & 0\% & 11\% & 0\% & 19 \\ % 100%
-\hline
-\end{tabular}
-\end{center}
-\caption{Evaluation results for LuaFileSystem}
-\label{tab:evallfs}
-\end{table}
-
-Even though we could precisely type most of the functions exported
-by the \texttt{lfs} module, we could not type two overloaded functions
-due to the lack of intersection types in our type system.
-
-\subsection{HTTP Digest}
-
-The HTTP Digest library implements client side HTTP digest authentication for Lua.
-Table \ref{tab:evalhttpdigest} summarizes the evaluation results for HTTP Digest.
-
-\begin{table}[!ht]
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
-\hline
-\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
-\hline
-http-digest & 0\% & 0\% & 100\% & 0\% & 1 \\ % 100%
-\hline
-\end{tabular}
-\end{center}
-\caption{Evaluation results for HTTP Digest}
-\label{tab:evalhttpdigest}
-\end{table}
-
-It is difficult to type the interface of the \texttt{http-digest} module
-because it is an extension to the \texttt{http} module from LuaSocket.
-The \texttt{http-digest} module only exports the function
-\texttt{http-digest.request}, which extends the function
-\texttt{http.request} with MD5 authentication.
-Like \texttt{http.request}, \texttt{http-digest.request}
-is also an overloaded function.
-
-Even though we could not precisely type the interface that \texttt{http-digest}
-exports, we could use only static types to annotate this module,
-and they pointed a bug in the code.
-The problem was related to the way the library was loading the MD5 library
-that should be used. 
-This part of the code checks the existence of three different MD5 libraries
-in the system, and uses the first one that is available, or generates an
-error when none is available.
-The code that loads the first option was fine, but the code that loads the
-second and third options were trying to access an undefined global variable.
-
-\subsection{Typical}
-
-Typical is a library that extends the behavior of the function \texttt{type}. 
-Table \ref{tab:evaltypical} summarizes the evaluation results for Typical.
-
-\begin{table}[!ht]
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
-\hline
-\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
-\hline
-typical & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
-\hline
-\end{tabular}
-\end{center}
-\caption{Evaluation results for Typical}
-\label{tab:evaltypical}
-\end{table}
-
-The interface of the \texttt{typical} module is straightforward to type,
-as it contains only the function \texttt{typical.type},
-which has the same type of the function \texttt{type}: \texttt{(value) -> (string)}.
-
-However, we hit some limitations of our type system while annotating this module.
-
-First, it uses the \texttt{getmetatable} to get a table and
-checks whether this table has the field \texttt{\string_\string_type}.
-We could not give a precise type to \texttt{getmetatable}, so we used the dynamic
-type \texttt{any} as its return type, and this generates a warning.
-
-Second, it returns a metatable that extends \texttt{\string_\string_call}
-with \texttt{typical.type}, that is, we can use the module itself as a function,
-though it is a table.
-Our type system still does not support metatables, so we did not extend our version
-of the \texttt{typical} module to support \texttt{\string_\string_call}.
-
-Third, the module uses \texttt{ipairs} to iterate over an array of functions,
-but our type system also has limited support to \texttt{ipairs}, and generates
-a warning when we try to use the indexed value inside the \texttt{for} body.
-As we mentioned in this chapter, we use the dynamic type as a replacement for
-the lack of type parameters.
-This means that we get warnings inside an \texttt{ipairs} iteration, because
-all iterated elements have the dynamic type.
-We removed this warning using the numeric \texttt{for} to perform the same loop.
-
-\subsection{Modulo 11}
-
-Modulo 11 is a library that generates and verifies modulo 11 numbers.
-Table \ref{tab:evalmod11} summarizes the evaluation results for Typical.
-
-\begin{table}[!ht]
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
-\hline
-\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
-\hline
-mod11 & 78\% & 0\% & 0\% & 22\% & 9 \\ % 100%
-\hline
-\end{tabular}
-\end{center}
-\caption{Evaluation results for Modulo 11}
-\label{tab:evalmod11}
-\end{table}
-
-The \texttt{mod11} module was written using an object-oriented idiom that
-our type system does not support, and that is the reason why we could not
-type all the members of its interface.
-More precisely, the original code uses \texttt{setmetatable} to hide
-two attributes, which our type system cannot hide.
-
-In addition, it returns a metatable that extends \texttt{\string_\string_call}
-with the class constructor.
-This allows us to use the module itself to create new instances of a
-Modulo 11 number.
-However, our type system does not support this feature, and we need to
-make explicit calls to the constructor whenever we want to create a
-new instance.
-
-Even though we had these two issues to annotate the \texttt{mod11} module,
-we could use only static types to annotate it, and we found some interesting
-points.
-The code relies on implicit conversions between strings and numbers,
-and some parts of the code keep on changing the type of local variables.
-These are two practices that may hide bugs.
-
-\subsection{Typed Lua Compiler}
-
-The Typed Lua compiler is the last case study that we evaluated.
-Table \ref{tab:evaltlc} summarizes its evaluation results.
-
-\begin{table}[!ht]
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
-\hline
-\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
-\hline
-tlast & 98\% & 0\% & 2\% & 0\% & 47 \\ % 100%
-\hline
-tltype & 100\% & 0\% & 0\% & 0\% & 65 \\ % 100%
-\hline
-tlst & 100\% & 0\% & 0\% & 0\% & 26 \\ % 100%
-\hline
-tllexer & 18\% & 0\% & 0\% & 82\% & 11 \\ % 100%
-\hline
-tlparser & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
-\hline
-tldparser & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
-\hline
-tlchecker & 100\% & 0\% & 0\% & 0\% & 2 \\ % 100%
-\hline
-tlcode & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
-\hline
-\end{tabular}
-\end{center}
-\caption{Evaluation results for Typed Lua Compiler}
-\label{tab:evaltlc}
-\end{table}
-
-The \texttt{tlast} module implements the Abstract Syntax Tree for
-the compiler.
-We could not precisely type just one function, because it has
-an overloaded type that requires intersection types.
-
-The \texttt{tltype} module implements the types introduced by Typed Lua.
-It also implements the subtyping and consistent-subtyping relations.
-The interface that this module exports was straightforward to type.
-
-The \texttt{tlst} module implements the symbols table for the compiler.
-The interface that this module exports was also straightforward to type.
-
-The \texttt{tllexer} module defines common lexical rules for
-the Typed Lua parser and the description file parser.
-This module is hard to type because it uses LPeg \cite{lpeg,ierusalimschy2009lpeg}
-patterns, and LPeg uses overloaded arithmetic operators to
-build LPeg patterns. 
-Even though LPeg is the third most popular Lua module,
-we cannot precisely type LPeg patterns because our
-type system still does not support overloading arithmetic
-operators.
-In the \texttt{tllexer} module, we could only give precise static types
-to two error reporting functions that it exports.
-
-The \texttt{tlparser} and \texttt{tldparser} modules implement the Typed Lua
-parser and the description file parser, respectively.
-Even though they use LPeg to implement the grammar rules,
-they only export a parsing function.
-Both use LPeg to parse a string and return the corresponding AST.
-
-We could type the interfaces that modules \texttt{tlchecker} and
-\texttt{tlcode} export.
-The former traverses the AST to perform type checking,
-while the latter traverses the AST to perform code generation.
-
-Even though we could precisely type the interface that most of
-the modules export, we had issues to write mutually recursive functions.
-This kind of functions often appear in compilers construction
-to traverse the data structures that they use.
-However, Typed Lua still does not support mutually recursive functions.
-A way to overcome this limitation was to predeclare these functions
-with an empty body, and then redeclare them with their actual body.
-The first declaration specifies the function type, while the
-second specifies what the function actually does without changing
-any type definition.
-
-Traversing the AST would also be problematic if we had not included
-a way to discriminate unions of table types, as we mentioned in Section \ref{sec:alias}.
-Without a way to discriminate a union of table types, any attempt
-to index this union of table types would generate a warning.
-
-Bootstraping the compiler also helped revealing some bugs.
-We found some accesses to undeclared global variables and also
-to undeclared table fields.
-The compiler also helped pointing the places where we should narrow
-a nilable value before using it.
-In fact, this point appeared in all the case studies that we used
-Typed Lua to annotate Lua code.
-This means that Lua programmers often use possibly \texttt{nil} values
-before checking whether it is \texttt{nil}.
-
 \section{Related Work}

### 2015-06-04 doc/thesis/evaluation.tex

@@
 \emph{easy}, \emph{poly}, \emph{over}, and \emph{hard}.
+In the next four paragraphs we explain each category in more detail.
+The last column of the table shows the total number of members
+of each case study.
+
 The \emph{easy} category shows the percentage of members that
 we could give a precise static type.
+For instance, the function \texttt{string.len} from the Lua
+standard library is in this category because we could use
+Typed Lua to describe its type: \texttt{(string) -> (integer)}.
+This function returns the length of a given string.
+Note that the results that we obtained for this category give
+a lower bound on how much static type safety
+we could add to each one of our case studies.
+
 The \emph{poly} category shows the percentage of members that
@@
 lack of type parameters.
+For instance, the function \texttt{table.sort} from the Lua
+standard library is in this category because it is a generic function.
+This function sorts a given list of elements, which is a generic list.
+However, we had to assign to this function the type
+\texttt{(\{any\}, nil|(any, any) -> (boolean)) -> ()} because
+Typed Lua does not have parametric polymorphism.
+
 The \emph{over} category shows the percentage of members that
@@
 types to describe their precise static types.
+For instance, the function \texttt{math.abs} from the Lua
+standard library is in this category because it has two types:
+\texttt{(integer) -> (integer)} and \texttt{(number) -> (number)}.
+This function returns the absolute value of a given number,
+which can be either integer or float.
+Even though we gave this function the more general type \texttt{(number) -> (number)},
+it is not precise enough because the return type is always \texttt{number}
+independently of the argument type.
+In other words, the return type should be \texttt{integer} when
+the argument type is also \texttt{integer}, and the return type
+should be \texttt{number} when the argument type is also \texttt{number}.
+However, we cannot give such a precise type to this function because
+Typed Lua does not support overloaded functions.
+
 The \emph{hard} category shows the percentage of members that
 we could not give a precise static type.
-The last column of the table shows the total number of members
-of each case study.
-The results that we obtained for the \emph{easy} category give
-a lower bound on how much static type safety
-we could add to each one of our case studies.
+For instance, the function \texttt{assert} from the Lua
+standard library is in this category because it can return any number
+of values of any type, being difficult to type this behavior.
+Still, we gave this function the type \texttt{(value*) -> (any*)}.
 
@@
 the \emph{over} category shows that it includes several overloaded functions.
-For instance, the functions \texttt{math.abs} and \texttt{math.random}
-are in this category.
-
-In the case of \texttt{math.abs},
-it has two different types: \texttt{(integer) -> (integer)} and
-\texttt{(number) -> (number)}.
-Even though we gave this function the more general type \texttt{(number) -> (number)},
-it is not precise enough because the return type is always \texttt{number}
-independently of the argument type.
-In other words, the return type should be \texttt{integer} when
-the argument type is also \texttt{integer}, and the return type
-should be \texttt{number} when the argument type is also \texttt{number}.
-
-In the case of \texttt{math.random}, it also has two different types:
+For instance, the function \texttt{math.random} is in this category
+because it has two different types:
 \texttt{() -> (number)} and \texttt{(integer, integer?) -> (integer)}.
@@
 we found a little difference between its documentation and its behavior.
-The documentation suggests that the type of the function \texttt{md5.update}
+The documentation suggests that the type of \texttt{md5.update}
 is \texttt{(md5\string_context, string) -> (md5\string_context)},
@@
 returns the tuple \texttt{(-1, nil, "hello")}, because \texttt{-1} makes
-\texttt{socket.skip} does not skip any values.
-As another example, \texttt{f = socket.skip(-2)} assigns \texttt{socket.skip}
+\texttt{socket.skip} not skip any values.
+As another example, the code \texttt{f = socket.skip(-2)} assigns \texttt{socket.skip}
 to \texttt{f}, as \texttt{-2} gets \texttt{socket.skip} from the stack.
@@
 which can be either a string or a table.
-More precisely, \texttt{http.request(url, body)} returns the tuple
+More precisely, \texttt{http.request(url, body)} returns the tuple type
 \texttt{(string, number, \{string:string\}, number) | (nil, string)}
 if \texttt{url} is a string and \texttt{body} is another string or \texttt{nil},
-but it returns the tuple
+but it returns the tuple type
 \texttt{(number, number, \{string:string\}, number) | (nil, string)}

### 2015-06-04 doc/thesis/glossary.tex

@@
 \item[closed table type] It is the type of table annotations in Typed Lua.
-A closed table type does not provide any guarantees that about keys
-with types not listed in the table type, and it also does not
-allow adding fields to the table type.
+A closed table type does not provide any guarantees about keys
+with types not listed in the table type.
+It also does not allow table refinement to add fields to the table type.
 
@@
 
-\item[consistency] Gradual typing uses the consistency relation instead of type equality
-to check the interaction among the dynamic type and other types.
+\item[consistency] Gradual typing uses the consistency relation to check the interaction
+among the dynamic type and other types.
 This relation allows us combining dynamic and static typing in the
 same language, but still catching static type errors.
-The consistency relation is reflexive and symmetric, but it is not transitive.
+The consistency relation is reflexive and symmetric, but it is not transitive
+to prevent that casts from a static type to the dynamic type can be
+combined with casts from the dynamic type to another static type.
 
@@
 the definition of gradual type systems for object-oriented languages.
-The consistent-subtyping relation is reflexive and symmetric, but it is not transitive.
+Like the consistency relation, it is reflexive and symmetric, but it is not transitive.
 
@@
 
-\item[depth subtyping] It allows the supertype to change the type of individual fields in the subtype.
+\item[depth subtyping] It allows the supertype to vary the type of individual fields in the subtype.
 
@@
 Gradual typing uses the dynamic type along with the consistency relation to
-identify where run-time casts should be inserted to prevent that dynamically
-typed code violates statically typed code.
+identify the parts of the code where run-time casts should be inserted to
+prevent that dynamically typed code violates statically typed code.
 
-\item[filter type] It is a type that allows Typed Lua to discriminate and restore the type of
-local variables inside control flow statements.
+\item[filter type] It is a type that allows Typed Lua to discriminate the type of local variables
+inside control flow statements, as these variables are bound to unions of
+first-level types.
 
-\item[fixed table type] It is a table type that does not allow width subtyping to make single
-inheritance safe in Typed Lua, and it also does not allow adding fields to
-the table type.
+\item[fixed table type] It is the type of classes in Typed Lua.
+It is a table type that does not allow width subtyping to make single
+inheritance safe.
+It also does not allow table refinement to add fields to the table type.
 
@@
 \item[gradual type system] It is a type system that uses the consistency relation instead of type equality
-to check the interaction among the dynamic type and other types.
+to perform static type checking.
 
 \item[gradual typing] It is an approach that uses a gradual type system to allow static and dynamic
-typing in the same code, but inserting run-time checks between statistically
+typing in the same code, but inserting run-time checks between statically
 typed and dynamically typed code.
-This run-time checks ensure that dynamically typed code does not violate
+These run-time checks ensure that dynamically typed code does not violate
 statically typed code.
+More precisely, gradual typing allows programmers to change between
+dynamic and static typing without changing the dynamic or the static behavior
+of the program.
 
@@
 \item[open table type] It is the type of the tables with keys that do not inhabit one of
-the table's key types, have at least one alias, and allows adding
-fields to the table type.
+the table's key types, and have at least one alias.
+It also allows table refinement to add fields to the table type.
 
@@
 language, but without affecting the run-time semantics.
+This means that the programmer has the option to use or not use its
+static analysis to check for static type errors, as these errors
+will be caught by the run-time semantics anyway.
 
-\item[projection environment] It is an environment that Typed Lua uses to assign projection variables to
-second-level types.
+\item[projection environment] It is an environment that Typed Lua uses to handle unions of second-level
+types that are bound to projection types.
 
 \item[projection type] It is a type that allows Typed Lua to discriminate the type of local variables
-that have a dependency relation, as they include a projection variable and an
-index that allow Typed Lua to project unions of second-level types into
-unions of first-level types.
+that have a dependency relation, as they project an union of second-level types
+into unions of first-level types.
+These unions of second-level types can describe the type of several local variables.
+A projection type uses an index to define which components of a union of second-level
+types should be used to project the union of first-level types of a local variable.
+When programmers discriminate a projection type, they are discriminating the
+union of second-level types that is bound to this projection type, affecting
+all variables that bind to the same projection type, which also bind to the
+same union of second-level types.
 
@@
 through prototype objects. 
+In this technique, when an object tries to access an attribute that is not
+present, it looks for this attribute in its parent object.
 
@@
 
-\item[table refinement] It is an operation that allows changing a table type to include new fields or
-to specialize existing fields.
+\item[table refinement] It is an operation from Typed Lua that allows programmers to change a table type
+to include new fields or to specialize existing fields.
 
@@
 
-\item[type tag] In dynamically typed languages a type tag describes the type of a value during
-run-time.
+\item[type tag] In dynamically typed languages, a type tag describes the type of a value during run-time.
 
-\item[unique table type] It is the type of the tables with keys that do not inhabit one of
-the table's key types, does not have any alias, and allow adding
-fields to the table type.
-It is also the table type that describes the type of the table constructor.
+\item[unique table type] It is the table type that describes the type of a table constructor.
+It is the type of the tables with keys that do not inhabit one of
+the table's key types, and that does not have any alias.
+It also allows table refinement to add fields to the table type.
 

### 2015-06-04 doc/thesis/related.tex
Fixing typos

@@
-In this chapter we review related work as a way to identify the novel
-features of our type system, what are the features that it shares with
-other type systems, and whether we can incorporate some interesting
-features from other type systems.
-We split this chapter into two sections: in the first section we review
-other Lua projects, while in the second section we review other
-projects that are not related to Lua.
+In this chapter we review related work, and we split it into two sections:
+in the first section we review other Lua projects,
+while in the second section we review other projects that are not related to Lua.
 
@@
 Even though TypeScript is unsound by design,
-\citet{bierman2014typescript} shows how to make TypeScript sound.
+Bierman et al. \cite{bierman2014typescript} shows how to
+make TypeScript sound.
 They use a reduced core of TypeScript to formalize a
@@
 
-\citet{politz2012semantics} proposes semantics and types for objects
-with first-class member names, a well-known feature from scripting languages.
+Politz et al. \cite{politz2012semantics} proposes semantics
+and types for objects with first-class member names, a well-known
+feature from scripting languages.
 Their type system uses string patterns to describe the members of
@@
 typed code does not violate statically typed code.
-\citet{allende2013cis} perform a carefull evaluation about
-cast insertion in Gradualtalk.
+Allende et al. \cite{allende2013cis} perform a careful
+evaluation about cast insertion in Gradualtalk.
 They report that usually cast insertions impact on execution
@@
 static and dynamic typing.
-\citet{homer2013modules} explores some useful patterns
-that derive from Grace's use of objects as modules
+Homer et al. \cite{homer2013modules} explores some
+useful patterns that derive from Grace's use of objects as modules
 and its brand of optional structural typing, which

### 2015-06-04 doc/thesis/system.tex

@@
 
-Back to our example, the table constructor
+Back to our example, the constructor
 $\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ has type
-$\{1:``x", 2:``y", 3:``z"\}$ through rule \textsc{T-CONSTRUCTOR2}.
-
-As another example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
+$\{1:``x", 2:``y", 3:``z"\}_{unique}$ through rule \textsc{T-CONSTRUCTOR2}.
+
+As another example, the constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
 has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ through rule
@@
 \begin{verbatim}
-
     local person = {}
@@
 \end{array} \right.
-\\ \\
+\\ 
 general(\False) & = \Boolean\\
@@
 general(T) & = T\\
-\\
+\end{align*}
+
+\begin{align*}
 general2(\Void) & = \Void\\

### 2015-06-05 doc/papers/paper/paper.tex

@@
 
-\title{Optional typing for Lua}
+\title{Formalization of Typed Lua}
 
@@
 \begin{abstract}
-This is the text of the abstract.
+Dynamically typed languages such as Lua avoid static types in order
+to have a simpler and more flexible language, because their programmers
+to not need to bother with abstracting types that need to be validated
+by a type checker. In contrast, statically typed languages provide
+earlier detection of many defects, and a better framework for structuring
+large programs. The advantages of static types often lead programmers to
+migrate from a dynamically typed to a statically typed language when
+their simple scripts evolve into complex programs.
+
+Optional type systems are one way to have both static and dynamic
+typing in the same language while keeping its original semantics,
+making evolving a program from dynamic to static typing a matter of
+describing the implied types that it it using and adding annotations
+to make those types explicit. Designing an optional type system for
+an existing dynamically typed language is challenging, as its types
+should feel natural to programmers that are already familiar with
+this language.
+
+In this work, we give a formal description of Typed Lua, an
+optional type system for Lua, with a focus on its novel type system
+features: incremental evolution of record and object types via
+imperative update that does not sacrifice width subtyping,
+and combining flow typing with multiple assignment and with functions that
+return multiple values. While our type system is tailored to the
+features and idioms of Lua, its features can be adapted to other
+imperative scripting languages.
 \end{abstract}

### 2015-06-05 doc/papers/paper/paper.tex
simplify introduction of dls paper

@@
 
-Dynamically typed languages such as Lua avoid static types in favor of
-simplicity and flexibility, because the absence of static types means
-that programmers do not need to bother with abstracting types that
-should be validated by a type checker.
-Instead, dynamically typed languages use run-time \emph{type tags}
-to classify the values they compute, so their implementation can use
-these tags to perform run-time (or dynamic) type checking
-\cite{pierce2002tpl}.
-
-This simplicity and flexibility allows programmers to write code that
-might require a complex type system to statically type check,
-though it may also hide bugs that will be caught only after deployment
-if programmers do not properly test their code.
-In contrast, static type checking helps programmers detect many
-bugs during the development phase.
-Static types also provide a conceptual framework that helps
-programmers define modules and interfaces that can be combined to
-structure the development of programs.
-
-Thus, early error detection and better program structure are two
-advantages of static type checking that can lead programmers to
-migrate their code from a dynamically typed to a statically
-typed language, when their simple scripts evolve into complex programs
-\cite{tobin-hochstadt2006ims}.
-Dynamically typed languages certainly help programmers during the
-beginning of a project, because their simplicity and flexibility
-allows quick development and makes it easier to change code according to
-changing requirements.
-However, programmers tend to migrate from dynamically typed to
-statically typed code as soon as the project has consolidated its
-requirements, because the robustness of static types helps
-programmers link requirements to abstractions.
-This migration usually involves different languages that have
-different syntaxes and semantics, which usually requires a complete
-rewrite of existing programs instead of incremental evolution from
-dynamic to static types.
+Dynamically typed languages forgo static type checking
+in favor of using run-time {\em type tags} to classify
+the values they compute, so its operations can use these
+tags to perform run-time (or dynamic) checks and signal
+errors in case of invalid operands~\cite{pierce2002tpl}.
+The lack of static types allows programmers to write code
+that might have required a complex type system to statically
+type, at the cost of hiding defects that will only be caught
+after deployment if the programmers do not properly
+test their code.
+
+In contrast, statically typed languages help programmers
+detect defects during development, and also provide
+a conceptual framework that helps programmers define modules
+and interfaces that can be combined to structure the development
+of large programs.
+
+The early error detection and better tools for structuring
+programs are two advantages of statically typed languages that
+lead programmers to migrate their code from a dynamically
+typed to a statically typed language when their scripts
+evolve into complex programs~\cite{tobin-hochstadt2006ims}.
+This migration from dynamic to static usually involves
+different languages, which distinct syntax and semantics,
+requiring a complexe rewrite of existing programs instead of
+incremental evolution.
 
 Ideally, programming languages should offer programmers the
-option to choose between static and dynamic typing:
-\emph{optional type systems} \cite{bracha2004pluggable} and
-\emph{gradual typing} \cite{siek2006gradual} are two similar
+option to choose between static and dynamic typing.
+\emph{Optional type systems}~\cite{bracha2004pluggable} and
+\emph{gradual typing}~\cite{siek2006gradual} are two similar
 approaches for blending static and dynamic typing in the same
-language.
-The aim of both approaches is to offer programmers the option
-to use type annotations where static typing is needed,
-allowing the incremental migration from dynamic to static typing.
-The difference between these two approaches is the way they treat
-run-time semantics.
-While optional type systems do not affect run-time semantics,
-gradual typing uses run-time checks to ensure that dynamically typed
-code does not violate the invariants of statically typed code.
-
-Programmers and researchers sometimes use the term \emph{gradual typing}
-to mean the incremental evolution of dynamically typed code into
-statically typed code.
-For this reason, gradual typing may also refer to optional type
-systems and other approaches that blend static and dynamic typing to
-help programmers incrementally migrate from dynamic to static typing
-without having to switch to a different language, though all these
-approaches differ in the way they handle static and dynamic typing
-together.
-We use the term \emph{gradual typing} to refer to the work of
-Siek and Taha \cite{siek2006gradual}.
-
-In this work we present the design and evaluation of Typed Lua:
-an optional type system for Lua that is rich enough to
-preserve some of the Lua idioms that programmers are already familiar with,
-but that also includes new constructs that help programmers
-structure Lua programs.
-
-Lua is a small imperative language with first-class functions
-(with proper lexical scoping) where the only data structure
+language. Their aim is to offer programmers the option
+of adding type annotations where static typing is needed,
+allowing the incremental migration from dynamic to static
+typing. The difference between these two approaches is the
+way they treat the runtime semantics of the language:
+optional type systems preserve the original semantics
+of a dynamically typed language, at the cost of runtime
+safety, while gradual typing changes the semantics to
+include more extensive runtime checking to ensure that
+dynamically typed portions of the program do not violate
+the type safety of statically typed portions.
+
+Lua is a small imperative language with lexically-scoped first-class functions where the only data structuring
 mechanism is the \emph{table} --
-an associative array that can represent arrays, records, maps, modules, objects, etc.
-Tables also have syntactic sugar and metaprogramming support
-through operator overloading built into the language.
-Unlike other scripting languages, Lua has very limited coercion
-among different data types.
+an associative array that has syntactic and runtime support
+to efficiently represent arrays, records, maps, modules,
+objects, etc. Unlike other scripting languages, Lua has very limited coercion among different data types.
 
@@
 This means that even features such as a module system and
-object orientation are a matter of convention instead of default
-language constructs.
+object orientation are a matter of convention instead of
+built-in language constructs.
 The result is a fragmented ecosystem of libraries, and different
 ideas among Lua programmers on how they should use the language
-features, or how they should structure programs.
+features or on how they should structure programs.
+
+Typed Lua is an optional type system for
+Lua that is rich enough to preserve some of the idioms
+that Lua programmers are already familiar with, while
+helping programmers structure large Lua programs and
+catch errors at compile-time. An informal
+presentation of the design of Typed Lua has been
+published~\cite{dylatypedlua}.
+
+In this paper, we give a formal basis to an updated version of that design, with a focus on novel type system features
+that Typed Lua uses to type some tricky Lua idioms.
+One of these idioms is the use of imperative updates to
+build records and objects one field at a time; Typed Lua
+lets record and object types evolve incrementally along
+with these updates while not sacrificing width subtyping.
+Another idiom combines overloading on the return values
+of a function, multiple return values, multiple assignment,
+and tag checks to discriminate a several returned values
+by checking just one of them; Typed Lua deconstructs unions
+of tuples in a way that the dependencies between the types
+of each value in the tuple can be tracked.
 
 The lack of standard policies is a challenge for the design of
-an optional type system for Lua.
-For this reason, we are not relying entirely on the semantics of
-the language to design our type system.
-We also run a mostly automated survey of Lua idioms used in a
-large corpus of Lua libraries, which also has helped in the design of Typed Lua.
-
-So far, Typed Lua is a Lua extension that allows statically typed
-code to coexist and interact with dynamically typed code
-through optional type annotations.
-In addition, it adds default constructs that programmers can use
-to better structure Lua programs.
-The Typed Lua compiler warns programmers about type errors,
-but always generates Lua code that runs in unmodified Lua implementations.
-Programmers can enjoy some of the benefits of static types
-even without converting existing Lua modules to Typed Lua --
-they can export a statically typed interface to a dynamically typed module,
-and statically typed users of the module can use the Typed Lua compiler
-to check their use of the module.
-Thus, implementing an optional type system for Lua offers Lua
-programmers one way to obtain most of the advantages of static typing
-without compromising the simplicity and flexibility of dynamic typing.
-We have an implementation of the Typed Lua compiler that is
-available online\footnote{https://github.com/andremm/typedlua}.
-
-Typed Lua's intended use is as an application language, and
-we believe that policies for organizing a program in modules and writing
-object-oriented programs should be part of the language and
-checked by its optional type system.
-An application language is a programming language that helps
-programmers develop applications from scratch until these
-applications evolve into complex systems rather than just scripts.
-We will show that Typed Lua introduces the refinement of
-tables to support the common idioms that Lua programmers use
-to encode both modules and objects.
-
-We also believe that Typed Lua helps programmers give more
-formal documentation to already existing Lua code, as static types
-are also a useful source of documentation in languages that provide
-type annotations, because type annotations are always validated by
-the type checker and therefore never get outdated.
-Thus, programmers can use Typed Lua to define axioms about the
-interfaces and types of dynamically typed modules.
-We enforce this point by using Typed Lua to statically type
-the interface of the Lua standard library and other commonly used
-Lua libraries, so our compiler can check Typed Lua code that uses
-these libraries.
-
-Typed Lua performs a very limited form of local type inference
-\cite{pierce2000lti}, as static typing does not necessarily mean
-that programmers need to insert type annotations in the code.
-Several statically typed languages such as Haskell provide some
-amount of type inference that automatically deduces the types of
-expressions.
-Still, Typed Lua only requires a small amount of type annotations
-due to the nature of its optional type system.
-
-Typed Lua does not deal with code optimization, although another
-important advantage of static types is that they help the compiler
-perform optimizations and generate more efficient code.
-However, we believe that the formalization of our optional type
-system is precise enough to aid optimization in some Lua implementations.
-
-We use some of the ideas of gradual typing to formalize Typed Lua.
-Even though Typed Lua is an optional type system and thus does not
-include run-time checks between dynamic and static regions of the
-code, we believe that using the foundations of gradual typing to
-formalize our optional type system will allow us to include run-time
-checks in the future.
-
-Finally, we believe that designing an optional type system for Lua may
-shed some light on optional type systems for scripting languages
-in general, as Lua is a small scripting language that shares
-some features with other scripting languages such as JavaScript.
+an optional type system for Lua. For this reason, the design
+of Typed Lua has been informed by a mostly automated survey
+of Lua idioms used by a large corpus of Lua libraries.
+This paper also presents the methodology of this survey and
+a summary of its results.
+
+The rest of the paper is organized as follows: Section~2
+presents the methodology and results of our survey of Lua
+idioms; Section 3 presents the types of Typed Lua's
+type system, and their subtyping and consistent-subtyping
+relationd; Section 4 presents the typing rules for
+evolution of table types; Section 5 presents the typing rules
+for destructuring unions of tuples into projection types
+and flow typing on these projections; Section 6 reviews
+related work; Section 7 presents our conclusions, and
+outlines our plans for future work.
 

### 2015-06-06 doc/papers/paper/paper.tex
Removing dangling pointers to thesis chapters

@@
 \section{Introduction}
+\label{sec:intro}
 
@@
 evolve into complex programs~\cite{tobin-hochstadt2006ims}.
-This migration from dynamic to static usually involves
-different languages, which distinct syntax and semantics,
+This migration from dynamic to static typing usually involves
+different languages that have distinct syntax and semantics,
 requiring a complexe rewrite of existing programs instead of
@@
 presentation of the design of Typed Lua has been
-published~\cite{dylatypedlua}.
+published~\cite{maidl2014tl}.
 
@@
 
-The rest of the paper is organized as follows: Section~2
-presents the methodology and results of our survey of Lua
-idioms; Section 3 presents the types of Typed Lua's
-type system, and their subtyping and consistent-subtyping
-relationd; Section 4 presents the typing rules for
-evolution of table types; Section 5 presents the typing rules
-for destructuring unions of tuples into projection types
-and flow typing on these projections; Section 6 reviews
-related work; Section 7 presents our conclusions, and
-outlines our plans for future work.
+The rest of the paper is organized as follows:
+Section \ref{sec:statistics} presents the methodology and results of
+our survey of Lua idioms;
+Section \ref{sec:types} presents the types of Typed Lua's
+type system, and their subtyping and consistent-subtyping relations;
+Section \ref{sec:tables} presents the typing rules for evolution of table types;
+Section \ref{sec:projections} presents the typing rules for
+destructuring unions of tuples into projection types and
+flow typing on these projections;
+Section \ref{sec:related} reviews related work;
+Section \ref{sec:conclusion} presents our conclusions,
+and outlines our plans for future work.
 
 \section{Statistics about the usage of Lua}
+\label{sec:statistics}
 
@@
 
-\section{Typing Rules}
-
-In the previous chapter we presented an informal overview of Typed Lua.
-We showed that programmers can use Typed Lua to combine static and dynamic
-typing in the same code, and it allows them to incrementally migrate from
-dynamic to static typing.
-This is a benefit to programmers that use dynamically typed languages
-to build large applications, as static types detect many bugs
-during the development phase, and also provide better documentation.
-
-In this chapter we present the abstract syntax of Typed Lua types,
-the subtyping rules, and the most interesting typing rules.
-Besides its practical contributions, Typed Lua also has some interesting
-contributions to the field of optional type systems for scripting
-languages.
-They are novel type system features that let Typed Lua cover several Lua idioms
-and features, such as refinement of tables, multiple assignment, and multiple return values.
+\section{Types}
+\label{sec:types}
+
+In this section we present the abstract syntax of Typed Lua types
+and their subtyping rules.
+
+%In the previous chapter we presented an informal overview of Typed Lua.
+%We showed that programmers can use Typed Lua to combine static and dynamic
+%typing in the same code, and it allows them to incrementally migrate from
+%dynamic to static typing.
+%This is a benefit to programmers that use dynamically typed languages
+%to build large applications, as static types detect many bugs
+%during the development phase, and also provide better documentation.
+
+%In this chapter we present the abstract syntax of Typed Lua types,
+%the subtyping rules, and the most interesting typing rules.
+%Besides its practical contributions, Typed Lua also has some interesting
+%contributions to the field of optional type systems for scripting
+%languages.
+%They are novel type system features that let Typed Lua cover several Lua idioms
+%and features, such as refinement of tables, multiple assignment, and multiple return values.
 
@@
 Types are ordered by a subtype relationship that we introduce
-in Section \ref{sec:subtyping}, so Lua values may belong to
-several distinct types.
+in the next section, so Lua values may belong to several distinct types.
 
@@
 The type $\Value$ is the top type, which represents any Lua value.
-In Section \ref{sec:rules} we will show that this type,
-along with variadic types, helps the type system to drop extra values
-on assignments and function calls, thus preserving the
+This type along with variadic types help the type system to drop
+extra values on assignments and function calls, thus preserving the
 semantics of Lua in these cases.
@@
 in object-oriented method definitions and method calls.
-As we mentioned in Section \ref{sec:oop}, we need the type
-$\Self$ to prevent programs from indexing a method without
+As we mentioned in our informal overview of Typed Lua \cite{maidl2014tl},
+we need the type $\Self$ to prevent programs from indexing a method without
 calling it with the correct receiver.
@@
 Informally, a table type is well-formed if key types do not overlap.
-In Section \ref{sec:rules} we formalize the definition of well-formed table types.
+In Section \ref{sec:tables} we formalize the definition of well-formed table types.
 We delay the proper formalization of well-formed table types because we use
@@
 where $T$ is a first-level type that $x$ represents.
-For instance, $\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}$
+As an example,
+\[
+\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}
+\]
 is a type for singly-linked lists of integers.
-In Section \ref{sec:alias} we mentioned that we can use the following
-interface declaration as an alias to this type:
+In Typed Lua, we can use the following interface declaration
+as an alias to this type:
 \begin{verbatim}
@@
 Our type system uses filter types to formalize the \texttt{type} predicates
-that we mentioned in Section \ref{sec:unions}.
+that we introduced in our previous design of Typed Lua \cite{maidl2014tl}.
 This means that \texttt{type} predicates use filter types of the form
@@
 unions of tuple types into unions of first-level types.
-In Section \ref{sec:fap} we will show in more detail how our type system
+In Section \ref{sec:projections} we will show in more detail how our type system
 uses them as a mechanism for handling unions of tuple types,
 when they appear in the right-hand side of the declaration of local variables,
-as we mentioned in Section \ref{sec:unions}.
+as we introduced in our previous design of Typed Lua \cite{maidl2014tl}.
 We also show how this feature allows our type system to constrain
@@
 to mix static and dynamic typing.
+There is a subtle difference between the dynamic type $\Any$ and the top type $\Value$.
+Although both types mean that they accept a value of any other type,
+the type $\Value$ is not a good option for handling the
+interaction between dynamically typed and statically typed code.
+Gradual typing uses the dynamic type $\Any$ to identify
+where it should insert run-time checks for asserting that dynamically
+typed code does not violate statically typed code.
+Typed Lua also uses the dynamic type $\Any$ in this sense,
+though it is an optional type system.
+More precisely, we use $\Any$ instead of $\Value$ to allow
+programmers blending dynamic and static typing because we use the
+consistent-subtyping relation to formalize our optional type system,
+as it is a first step to switch Typed Lua from optional typing to
+gradual typing in the future.
 
@@
 We explain the subtyping and consistent-subtyping rules throughout this section.
-However, we focus the discussion on the definition of subtyping because,
-as we mentioned in Section \ref{sec:gradual}, we can combine the
-consistency and subtyping relations to achieve consistent-subtyping.
-The differences between subtyping and consistent-subtyping are the way
+However, we focus the discussion on the definition of subtyping because
+we can combine the consistency and subtyping relations to achieve
+consistent-subtyping \cite{siek2007objects,siek2013mutable}.
+The differences between consistent-subtyping and subtyping are the way
 they handle the dynamic type, and the fact that subtyping is transitive,
@@
 \begin{array}{c}
-\mylabel{S-VOID2}\\
+\mylabel{S-VOID}\\
 \senv \vdash \Void \subtype T{*}
@@
 A variadic type $T{*}$ represents a sequence of values of type
-$T \cup \Nil$, and the rule \textsc{S-VOID2} handles the case where
+$T \cup \Nil$, and the rule \textsc{S-VOID} handles the case where
 a given sequence is empty.
@@
 not require any special rule.
-In the next section we will show that we use the subtyping rules for variadic types,
-along with the types $\Value$ and $\Nil$, to make our type system reflect
-the semantics of Lua on discarding extra parameters and
-replacing missing parameters.
+We use the subtyping rules for variadic types along with the types
+$\Value$ and $\Nil$ to make our type system reflect the semantics of
+Lua on discarding extra parameters and replacing missing parameters.
 
@@
        \end{array}}
-      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{fixed|closed} \subtype
-                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}} \; m \ge n
+      {\begin{array}{c}
+       \senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{fixed|closed} \subtype\\
+                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}
+       \end{array}} \; m \ge n
 \end{array}
@@
 \begin{verbatim}
-    local t:{"x":integer, "y":integer?} = { x = 1, y = 2 }
+    local t:{ "x":integer,
+              "y":integer? } = { x = 1, y = 2 }
 \end{verbatim}
@@
 More precisely,
-\[
-\{``x":1, ``y":2\}_{closed} \not\subtype
-\{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
-\]
+\begin{align*}
+& \{``x":1, ``y":2\}_{closed} \not\subtype \\
+& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
+\end{align*}
 
@@
 More precisely,
-\[
-\{``x":1, ``y":2\}_{unique} \subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
-\]
+\begin{align*}
+& \{``x":1, ``y":2\}_{unique} \subtype \\
+& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
+\end{align*}
 
@@
        \end{array}}
-      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype
-                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{unique|open|fixed}}
+      {\begin{array}{c}
+       \senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype\\
+                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{unique|open|fixed}
+       \end{array}}
 \end{array}
@@
        \end{array}}
-      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype
-                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{open|fixed}}
+      {\begin{array}{c}
+       \senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype\\
+                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{open|fixed}
+       \end{array}}
 \end{array}
@@
 
-In the next section we will show in more detail how our type system
+In Section \ref{sec:tables} we will show in more detail how our type system
 uses these tags to handle the refinement of table types.
@@
 As an example, the rule \textsc{S-UNFOLDR} allows our type system to
-type check the function \texttt{insert} from Section \ref{sec:alias}:
+type check a function that inserts an element in the begining of
+a singly-linked list:
 \begin{verbatim}
-    local function insert (e:Element?, v:integer):Element
+    local function insert (e:Element?,
+                           v:integer):Element
       return { info = v, next = e }
@@
 
-\subsection{Typing rules}
-
-In this section we use a reduced core of Typed Lua to present the
-most interesting rules of our type system.
-These rules type check multiple assignment, table refinement,
-and overloading on the return type of functions.
-Appendix \ref{app:rules} presents the full set of typing rules.
+\section{Typing rules}
+\label{sec:rules}
+
+In this section we use a reduced core of Typed Lua to present
+the typing rules that type check the novel features of our type system.
+These rules type check table refinement and
+overloading on the return type of functions.
 
@@
 with the projection variable $x$ that maps to type $S$.
-In Section \ref{sec:fap} we will show how our type system uses the
+In Section \ref{sec:projections} we will show how our type system uses the
 projection environment $\penv$ for handling projection types,
@@
 
-\subsubsection{Tables and refinement}
+\subsection{Tables and refinement}
+\label{sec:tables}
 
@@
 $\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ has type
-$\{1:``x", 2:``y", 3:``z"\}$ through rule \textsc{T-CONSTRUCTOR2}.
+$\{1:``x", 2:``y", 3:``z"\}_{unique}$ through rule \textsc{T-CONSTRUCTOR2}.
 
@@
 field assignment.
-For instance, in Section \ref{sec:tables} we presented the
-following example:
+As an example, we can translate
 \begin{verbatim}
-
     local person = {}
@@
 \end{verbatim}
-
-We can translate this example to our reduced core as follows:
+to our reduced core as follows:
 \begin{center}
@@
 $person$, then we refine its type to $\{``firstname":\String\}_{unique}$,
-and then we refine its type to $\{``firstname":\String, ``lastname":\String\}_{unique}$.
+and then to $\{``firstname":\String, ``lastname":\String\}_{unique}$.
 Rule \textsc{T-REFINE} type checks this use of refinement:
@@
        \end{array}}
-      {\env_{1}, \penv \vdash id[k] {<}T{>}:V, \env_{2}[id \mapsto \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, K{:}V\}_{open|unique}]}
+      {\begin{array}{l}
+       \env_{1}, \penv \vdash id[k] {<}T{>}:V, \\
+       \env_{2}[id \mapsto \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, K{:}V\}_{open|unique}]
+       \end{array}}
 \end{array}
@@
 As an example,
-\[
-\string_ENV[``x"] \; {<}\String{>} = ``foo" \;;\; \string_ENV[``y"] \; {<}\Integer{>} = 1
-\]
+\begin{align*}
+& \string_ENV[``x"] \; {<}\String{>} = ``foo" \;; \\
+& \string_ENV[``y"] \; {<}\Integer{>} = 1
+\end{align*}
 uses field assignment to add fields $``x"$ and $``y"$ to $\string_ENV$.
@@
 For instance,
-\[
-\string_ENV[``x"] \; {<}\String{>} = ``foo" \;;\; \string_ENV[``x"] \; {<}\Integer{>} = 1
-\]
+\begin{align*}
+& \string_ENV[``x"] \; {<}\String{>} = ``foo" \;; \\
+& \string_ENV[``x"] \; {<}\Integer{>} = 1
+\end{align*}
 does not type check, as we are trying to add $``x"$ twice.
@@
 the context that we are using a variable.
-As we mentioned in Sections \ref{sec:oop} and \ref{sec:subtyping},
-we need \emph{fixed} table types to type classes in object-oriented programming.
+We need \emph{fixed} table types to type classes in object-oriented programming.
 In the implementation we fix the aliasing of \emph{unique} and \emph{open}
@@
 
-Rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this non-deterministic behavior.
+The rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this non-deterministic behavior.
 Rule \textsc{T-IDREAD1} uses the auxiliary function \emph{close} to
@@
 accessing an identifier that is bound to a filter or projection type.
-As we mentioned in Section \ref{sec:unions}, our type system includes
-a small set of \texttt{type} predicates that allow programmers to
-discriminate union types, and our type system uses filter and projection
-types in the definition of these predicates to handle the discrimination
+As we presented in our previous design of Typed Lua \cite{maidl2014tl},
+our type system includes a small set of \texttt{type} predicates that allow
+programmers to discriminate union types, and our type system uses filter and
+projection types in the definition of these predicates to handle the discrimination
 of unions types.
@@
 second-level types and an index from a projection type.
-In Section \ref{sec:fap} we will discuss how our type system uses
+In Section \ref{sec:projections} we will discuss how our type system uses
 projection types to handle overloaded return types.
@@
 accessing an identifier that is bound to a filter type.
-As we mentioned in Sections \ref{sec:unions} and \ref{sec:alias},
-assignments restore discriminated union types to their original types,
+In Typed Lua, assignments restore discriminated union types to their original types,
 and function \emph{write} works in this purpose.
@@
 \begin{center}
-\begin{tabular}{lll}
-\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{ \} \; \mathbf{in}$}\\
-& \multicolumn{2}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
-& \multicolumn{2}{l}{$a[``y"] \; {<}\String{>} = ``bar";$}\\
-& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{closed} =$}\\
-& & \multicolumn{1}{l}{${<}\{``x":\String, ``y":\String \cup \Nil\}_{open}{>} \; a \; \mathbf{in} \; a[``z"] \; {<}\Integer{>} = 1$}
+\begin{tabular}{llll}
+\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{ \} \; \mathbf{in}$}\\
+& \multicolumn{3}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
+& \multicolumn{3}{l}{$a[``y"] \; {<}\String{>} = ``bar";$}\\
+& \multicolumn{3}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{closed} =$}\\
+& & \multicolumn{2}{l}{${<}\{``x":\String, ``y":\String \cup \Nil\}_{open}{>} \; a \; \mathbf{in}$}\\
+& & & \multicolumn{1}{l}{$a[``z"] \; {<}\Integer{>} = 1$}
 \end{tabular}
@@
 & & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
-& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; b[``x"] \; {<}\String{>} = ``foo" \;;\; f(a[``x"])$}
+& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;$}\\
+& & \multicolumn{2}{l}{$b[``x"] \; {<}\String{>} = ``foo" \;;\; f(a[``x"])$}
 \end{tabular}
@@
        crall(\env_{1}[\overline{id} \mapsto \overline{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}\\
-       \env_{3} = openset(crset(\env_{1}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)),rv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
+       \env_{3} = crset(\env_{1}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))\\
+       \env_{4} = openset(\env_{3}, rv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
        \end{array}}
-      {\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{3}}
+      {\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{4}}
 \end{array}
@@
 
-\subsubsection{Projections}
+\subsection{Projections}
+\label{sec:projections}
 
@@
 in the environment.
-The function \emph{idiv} has type
-\[
-\Integer \times \Integer \times \Void \rightarrow (\Integer \times \Integer \times \Void) \sqcup (\Nil \times \String \times \Void)
-\]
-As we mentioned in Section \ref{sec:unions}, \emph{idiv} performs
-integer division.
+The function \emph{idiv} performs integer division and has type
+\begin{align*}
+& \Integer \times \Integer \times \Void \rightarrow \\
+& (\Integer \times \Integer \times \Void) \sqcup (\Nil \times \String \times \Void)
+\end{align*}
 In case of success, it returns two integers: the result and the remainder.
@@
 \begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; q, r = idiv(1, 2) \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; \lfloor print(q + r) \rfloor_{0} \; \mathbf{else} \; \lfloor print(``ERROR: " \; .. \; r) \rfloor_{0}$}
+\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(1, 2) \, \mathbf{in}$}\\
+& \multicolumn{1}{l}{$\mathbf{if} \, q \, \mathbf{then} \, \lfloor print(q + r) \rfloor_{0} \, \mathbf{else} \, \lfloor print(``ERROR: " \, .. \, r) \rfloor_{0}$}
 \end{tabular}
@@
 $\pi_{1}^{x} \times \pi_{2}^{x} \times \Nil{*}$ and produces the pair
-\[
-(x,(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})) 
-\]
+$(x,(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}))$.
 
@@
        \env_{1}, \penv \vdash el:S_{1}, \env_{2}, (x,S_{2})\\
-       \env_{2}[id_{1} \mapsto infer(S_{1},1), ..., id_{n} \mapsto infer(S_{1},n)], \penv[x \mapsto S_{2}] \vdash s, \env_{3} \;\;\;
+       \env_{3} = \env_{2}[id_{1} \mapsto infer(S_{1},1), ..., id_{n} \mapsto infer(S_{1},n)]\\
+       \env_{3}, \penv[x \mapsto S_{2}] \vdash s, \env_{4} \;\;\;
        n = |\;\overline{id}\;|  
        \end{array}}
-      {\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, \env_{3} - \{\overline{id}\}}
+      {\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, \env_{4} - \{\overline{id}\}}
 \end{array}
@@
 \end{array} \right.
-\\ \\
+\end{align*}
+
+\begin{align*}
 general(\False) & = \Boolean\\
@@
 general(S_{1} \rightarrow S_{2}) & = general2(S_{1}) \rightarrow general2(S_{2})\\
-general(\{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{tag}) & = \{K_{1}{:}general(V_{1}), ..., K_{n}{:}general(V_{n})\}_{tag}\\
+general(\{K_{1}{:}V_{1}, ..., \\
+K_{n}{:}V_{n}\}_{tag}) & = \{K_{1}{:}general(V_{1}), ..., \\
+& \;\;\;\;\;\; K_{n}{:}general(V_{n})\}_{tag} \\
 general(\mu x.T) & = \mu x.general(T)\\
@@
        closeall(\env_{1}), \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
-       \env_{4} = openset(closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2})),rv(s_{1}) \cup rv(s_{2}))
+       \env_{4} = closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2})) \\
+       \env_{5} = openset(\env_{4}, rv(s_{1}) \cup rv(s_{2}))
       \end{array}}
-      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
+      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{5}}
 \end{array}
@@
 \section{Related Work}
+\label{sec:related}
 
@@
 \section{Conclusion}
+\label{sec:conclusion}
 

### 2015-06-09 doc/papers/paper/paper.tex
rewrite of section 2 for DLS paper

@@
 
-\section{Statistics about the usage of Lua}
+\section{A Survey of Lua Idioms}
 \label{sec:statistics}
 
-In this section we present statistics about the usage of Lua
-features and idioms.
-We collected statistics about how programmers use tables, functions,
-dynamic type checking, object-oriented programming, and modules.
-We shall see that these statistics informed important design decisions
-on our optional type system.
-
-We used the LuaRocks repository to build our statistics database;
-LuaRocks \cite{hisham2013luarocks} is a package manager for Lua
-modules.
-We downloaded the 3928 \texttt{.lua} files that were available in
-the LuaRocks repository at February 1st 2014.
-However, we ignored files that were not compatible with Lua 5.2,
-the latest version of Lua at that time.
-We also ignored \emph{machine-generated} files and test files,
-because these files may not represent idiomatic Lua code,
-and might skew our statistics towards non-typical uses of Lua.
-This left 2598 \texttt{.lua} files from 262 different projects for
-our statistics database;
-we parsed these files and processed their abstract syntax tree
-to gather the statistics that we show in this section.
-
-To verify how programmers use tables, we measured how they
-initialize, index, and iterate tables.
-We present these statistics in the next three paragraphs to discuss
-their influence on our type system.
-
-The table constructor appears 23185 times.
-In 36\% of the occurrences it is a constructor that initializes a
-record (e.g., \texttt{\{ x = 120, y = 121 \}});
-in 29\% of the occurrences it is a constructor that initializes a
-list (e.g., \texttt{\{ "one", "two", "three", "four" \}});
-in 8\% of the occurrences it is a constructor that initializes a
-record with a list part;
-and in less than 1\% of the occurrences (4 times) it is a constructor
-that uses only the booleans \texttt{true} and \texttt{false} as indexes.
-At all, in 73\% of the occurrences it is a constructor that uses
-only literal keys;
-in 26\% of the occurrences it is the empty constructor;
-in 1\% of the occurrences it is a constructor with non-literal keys
-only, that is, a constructor that uses variables and function calls
-to create the indexes of a table;
-and in less than 1\% of the occurrences (19 times) it is a constructor
-that mixes literal keys and non-literal keys.
-
-The indexing of tables appears 130448 times:
-86\% of them are for reading a table field while
-14\% of them are for writing into a table field.
-We can classify the indexing operations that are reads as follows:
-89\% of the reads use a literal string key,
-4\% of the reads use a literal number key,
-and less than 1\% of the reads (10 times) use a literal boolean key.
-At all, 93\% of the reads use literals to index a table while
-7\% of the reads use non-literal expressions to index a table.
-It is worth mentioning that 45\% of the reads are actually
-function calls.
-More precisely, 25\% of the reads use literals to call a function,
-20\% of the reads use literals to call a method, that is,
-a function call that uses the colon syntactic sugar, 
-and less than 1\% of the reads (195 times) use non-literal expressions
-to call a function.
-We can also classify the indexing operations that are writes as follows: 
-69\% of the writes use a literal string key,
-2\% of the writes use a literal number key,
-and less than 1\% of the writes (1 time) uses a literal boolean key.
-At all, 71\% of the writes use literals to index a table while
-29\% of the writes use non-literal expressions to index a table.
-
-We also measured how many files have code that iterates over tables to
-observe how frequently iteration is used.
-We observed that 23\% of the files have code that iterates over keys
-of any value, that is, the call to \texttt{pairs} appears at least
-once in these files (the median is twice per file);
-21\% of the files have code that iterates over integer keys, that is,
-the call to \texttt{ipairs} appears at least once in these files
-(the median is also twice per file);
-and 10\% of the files have code that use the numeric \texttt{for}
-along with the length operator (the median is once per file).
-
-The numbers about table initialization, indexing, and iteration
-show us that tables are mostly used to represent records, lists,
-and associative arrays.
-Therefore, Typed Lua should include a table type for handling
-these uses of Lua tables.
-Even though the statistics show that programmers initialize tables
-more often than they use the empty constructor to
-dynamically initialize tables, the statistics of the empty
-constructor are still expressive and indicate that Typed Lua should
-also include a way to handle this style of defining table types.
-
-We found a total of 24858 function declarations in our database
-(the median is six per file).
-Next, we discuss how frequently programmers use dynamic type
-checking and multiple return values inside these functions.
-
-We observed that 9\% of the functions perform dynamic type checking
-on their input parameters, that is, these functions use \texttt{type}
-to inspect the tags of Lua values (the median is once per function).
-We randomly selected 20 functions to sample how programmers are
-using \texttt{type}, and we got the following data:
-50\% of these functions use \texttt{type} for asserting the tags of
-their input parameters, that is, they raise an error when the tag of a
-certain parameter does not match the expected tag, and
-50\% of these functions use \texttt{type} for overloading, that is,
-they execute different code according to the inspected tag.
-
-These numbers show us that Typed Lua should include union types,
-because the use of the \texttt{type} idiom shows that disjoint unions
-would help programmers define data structures that can hold a value of
-several different, but fixed types.
-Typed Lua should also use \texttt{type} as a mechanism for decomposing
-unions, though it may be restricted to base types only.
-
-We observed that 10\% of the functions explicitly return multiple
-values.
-We also observed that 5\% of the functions return \texttt{nil} plus
-something else, for signaling an unexpected behavior;
-and 1\% of the functions return \texttt{false} plus something else,
-also for signaling an unexpected behavior.
-
-Typed Lua should include function types to represent Lua functions,
-and tuple types to represent the signatures of Lua functions,
-multiple return values, and multiple assignments.
-Tuple types require some special attention, because Typed Lua
-should be able to adjust tuple types during compile-time, in a
-similar way to what Lua does with function calls and multiple
-assignments during run-time.
-In addition, the number of functions that return \texttt{nil} and
-\texttt{false} plus something else show us that overloading on the
-return type is also useful to the type system.
-
-We also measured how frequently programmers use the object-oriented
-paradigm in Lua.
-We observed that 23\% of the function declarations are actually
-method declarations.
-More precisely, 14\% of them use the colon syntactic sugar while
-9\% of them use \texttt{self} as their first parameter.
-We also observed that 63\% of the projects extend tables with
-metatables, that is, they call \texttt{setmetatable} at least once,
-and 27\% of the projects access the metatable of a given table,
-that is, they call \texttt{getmetatable} at least once.
-In fact, 45\% of the projects extend tables with metatables and
-declare methods:
-13\% using the colon syntactic sugar, 14\% using \texttt{self}, and
-18\% using both.
-
-Based on these observations, Typed Lua should include support
-to object-oriented programming.
-Even though Lua does not have standard policies for object-oriented
-programming, it provides mechanisms that allow programmers to
-abstract their code in terms of objects, and our statistics confirm
-that an expressive number of programmers are relying on these mechanisms
-to use the object-oriented paradigm in Lua.
-Typed Lua should include some standard way of defining interfaces and classes
-that the compiler can use to type check object-oriented code,
-but without changing the semantics of Lua.
-
-We also measured how programmers are defining modules.
-We observed that 38\% of the files use the current way of defining
-modules, that is, these files return a table that contains the
-exported members of the module at the end of the file;
-22\% of the files still use the deprecated way of defining modules,
-that is, these files call the function \texttt{module};
-and 1\% of the files use both ways.
-At all, 61\% of the files are modules while 39\% of the files are
-plain scripts.
-The number of plain scripts is high considering the origin of
-our database.
-However, we did not ignore sample scripts, which usually serve to
-help the users of a given module on how to use this module, and
-that is the reason why we have a high number of plain scripts.
-
-Based on these observations, Typed Lua should include a way
-for defining table types that represent the type of modules.
-Typed Lua should also support the deprecated style of module
-definition, using global names as exported members of the module.
-
-Typed Lua should also include some way to define the types of
-userdata.
-This feature should also allow programmers to define userdata
-that can be used in an object-oriented style, as this is another
-common idiom from modules that are written in C.
-
-The last statistics that we collected were about variadic functions
-and vararg expressions.
-We observed that 8\% of the functions are variadic, that is,
-their last parameter is the vararg expression.
-We also observed that 5\% of the initialization of lists
-(or 2\% of the occurrences of the table constructor) use solely the
-vararg expression.
-Typed Lua should include a \emph{vararg type} to handle variadic
-functions and vararg expressions.
+In order to find out how Lua is used in practice, to inform
+the design of Typed Lua, we did a survey of existing Lua
+libraries. Our corpus is the repository for the LuaRocks
+package manager~\cite{hisham2013luarocks}, downloaded on
+February 1st, 2014. We did not survey all of the Lua
+scripts in the repository: scripts that were not compatible
+with Lua 5.2 (the version of Lua at that time) were ignored,
+and we also ignored machine-generated scripts and test scripts,
+as those could skew the result towards non-idiomatic uses
+of Lua. This left 2,598 scripts out of a total of 3,928, from
+262 different projects. The scripts were parsed, and their
+abstract syntax trees analysed to collect the data that we
+show in the rest of this section.
+
+Tables are the core data structure of Lua programs, and
+the language has features that let tables be used as
+tuples, lists, maps, records, abstract data types, classes,
+objects, modules, etc. We performed several analyses to
+quantify the different ways that Lua programs use tables.
+
+Our first analysis surveyed table constructors (expressions
+that create and initialize a table) to find out how a table
+begins its life. Of all of the 23,185 table constructors of
+our corpus, 36\% of them create a record (a table with string
+keys such as \texttt{\{ x = 120, y = 121 \}}), 29\% of them create
+a list or a tuple (a table with consecutive integer keys such as \texttt{\{ "one", "two", "three", "four" \}}), 26\% of them create an empty table, and 8\% of them create a table with
+both a record part and a list part. The remaining 1\% of them create a table with non-literal keys.
+
+We also analysed expressions that access a table (130,448 instances), either to read a value out of it (86\% of such expressions) or to write a 
+value into it (14\% of such expressions). Most of the expressions
+that read a value out of a table use a literal key,
+either a literal string (89\% of the reads) or a literal
+number (4\% of the reads). Lua has syntactic sugar that
+turns field accesses using ``dot'' notation into table
+accesses through literal string keys.
+
+In 45\% of the expressions that read a value out of a table,
+its value is immediately called. These calls are split almost
+evenly between function calls and method calls (25\% of reads
+are function calls, 20\% of reads are method calls).
+These results show that the use of tables as a namespacing
+mechanism for modules, as objects, and as records, 
+is prevalent.
+
+Expressions that write into a table also mostly use
+literal keys (69\% of them use a literal string, 2\%
+of them use a literal number), although a large 29\%
+of writes use non-literal keys, in contrast to only 7\% of
+reads.
+
+In order to gauge how frequently tables are used as
+collections, we looked for the presence of code that
+iterates over tables, and found out that 23\% of the
+scripts in our corpus iterate over a map at least once,
+and 31\% of the scripts iterate over a list
+at least once. \emph{TODO: Não sei se esse número está correto,
+apenas somei as duas porcentagens, mas pode haver interseção.}
+
+Besides measuring the number of method calls, we also
+measured other kinds of expressions and statements to gauge
+how frequently Lua programmers use the object-oriented paradigm.
+Our corpus has 24,858 function declarations, and 23\% of these
+are method declarations. Of the 262 projects in our corpus,
+63\% use {\em metatables}, a metaprogramming mechanism that
+lets the program extend the behavior of tables and is mostly
+used for implementing prototype-based inheritance.
+
+Typed Lua's table types, described in the next section,
+reflect the way Lua programmers use Lua tables in practice,
+according to the results above. A table type can express
+records, tuples, maps, lists, objects, or a combination of
+those.
+
+Lua modules are tables containing the modules' exports.
+The current idiomatic way to define a module is to populate
+an initially empty table with the members the module wishes
+to export, and return this table at the end of the script
+that defines the module. Around two-thirds of the modules
+in our corpus use this idiom. The other third uses a
+deprecated idiom, where a call to the {\tt module} function
+at the top of the script installs a new global environment,
+and the exported members are global variables assigned to
+in the rest of the module.
+
+The global environment is a also a table, so the two idioms
+are equivalent in terms of how Typed Lua deals with module
+definitions. Modules in Typed Lua make extensive use of Typed
+Lua's rules for evolution of table types
+(Section~\ref{sec:tables}).
+
+In order to gauge how common is the use of dynamic type
+checking as a way to define overloaded functions, we
+measured how many of the functions inspect the
+tags of their input parameters, and found out that 9\%
+of the functions in our corpus do this, split evenly
+between using the tag in an assertion, as a form of
+type checking, and using the tag to decide which code
+to execute, as a form of overloading. While Typed Lua
+supports both union types and flow typing, general
+support for overloaded functions still has to improve.
+
+A Lua idiom that is specially problematic for static type
+systems is overloading on the number and types and
+multiple return values. Around 6\% of functions use
+this idiom as a way to signal errors: in case of errors,
+the function returns either {\tt nil} or {\tt false} plus
+either an error message or error object instead of its usual
+return values or throwing an exception. The caller then
+tests the first returned value to check if an error occured
+or not. Standard union types and flow typing cannot deal
+with this, and Typed Lua introduces {\em projection types},
+a way of tracking dependencies between types in a union of
+tuples after a destructuring assignment.
 

### 2015-06-09 doc/papers/paper/paper.tex
rewrite of related work section for DLS paper

@@
 
-In this chapter we review related work, and we split it into two sections:
-in the first section we review other Lua projects,
-while in the second section we review other projects that are not related to Lua.
-
-\subsection{Other Lua projects}
-
-Metalua \cite{metalua} is a Lua compiler that supports compile-time
-metaprogramming (CTMP).
-CTMP is a kind of macro system that allows the programmers to interact
-with the compiler \cite{fleutot2007contrasting}.
-Metalua extends Lua 5.1 syntax to include its macro system,
-and allows programmers to define their own syntax.
-Metalua can provide syntactical support for several object-oriented
-styles, and can also provide syntax for turning simple type
-annotations into run-time assertions.
-
-MoonScript \cite{moonscript} is a programming language that supports
-class-based object-oriented programming.
-MoonScript compiles to idiomatic Lua code, but
-it does not perform compile-time type checking.
-
-LuaInspect \cite{luainspect} is a tool that uses MetaLua to perform
-some code analysis.
-For instance, it flags unknown global variables and table fields,
-it checks the number of function arguments against signatures, and
-it infers function return values.
-However, it does not try to analyze object-oriented code and
-it does not perform compile-time type checking.
-
-Tidal Lock \cite{tidallock} is a prototype of another optional type
-system for Lua, which is written in Metalua.
-Tidal Lock covers a little subset of Lua.
-Statements include declaration of local variables, multiple assignment,
-function application, and the return statement.
-This means that Tidal Lock does not include any control-flow statement.
-Expressions include primitive literals, table indexing, function application,
-function declaration, and the table constructor, but they do not include
-binary operations.
-
-A remarkable feature of Tidal Lock is the refinement of table types.
-This feature inspired us to also include it in Typed Lua,
-but in a simpler way and with different formalization.
-
-The table type from Tidal Lock can only represent records, that is,
-it cannot describe hash tables and arrays yet, though we can refine them.
-Tidal Lock also includes field types to describe the type of the fields
-of a table type.
-The field types describe if a table field is mutable or immutable
-in a table type.
-Field types are the feature that allow the refinement of table types in
-Tidal Lock.
-
-Tidal Lock is also a structural type system that relies on subtyping and
-local type inference.
-However, it does not support union types, recursive types, and variadic types.
-It also does not type any object-oriented idiom.
-
-Sol \cite{sol} is an experimental optional type system for Lua.
-Its type system is similar to ours, as it includes literal types,
-union types, and function types that handle variadic functions.
-However, it does not handle the refinement of tables and it
-includes different types for tables.
-Sol types tables as lists, maps, and objects.
-Its object types handle a specific object-oriented idiom that
-Sol introduces.
-
-Lua Analyzer \cite{luaanalyzer} is an optional type system for Lua
-that is specially designed to work in the Löve Studio,
-an IDE for game developing using the Löve framework.
-It works in Lua 5.1 only, and uses type annotations inside comments.
-It is unsound by design because its dynamic type is both
-top and bottom in the subtyping relation.
-
-Lua Analyzer shares some features with Typed Lua, and also
-has some interesting features that we do not have in Typed Lua.
-It has similar rules for handling the \texttt{or} idiom and
-discriminating union types inside conditions.
-However, these rules are limited to the \texttt{nil} tag only.
-It also includes different types for typing tables.
-It includes regular record types that maps names to types,
-array types, and map types.
-Even though it does not support the refinement of tables,
-it allows the definition of nominal table types that simulate classes.
-This system allows it to type check custom class systems,
-which are common in Lua.
-Function types also support multiple return values and
-variadic functions, but they do not support overloading the
-return type.
-Recently, it included experimental support for type aliases and generics.
-
-Luacheck \cite{luacheck} is a tool that performs static analysis on Lua code.
-It can flag access to undeclared globals and unused local variables,
-but it does not perform static type checking.
-
-Ravi \cite{ravi} is an experimental Lua dialect.
-Ravi introduces optional static typing for Lua to improve run-time performance.
-To do that, Ravi extends the Lua Virtual Machine to include new
-operations that take into account static type information.
-Currently, Ravi extends the Lua Virtual Machine to support few types:
-\texttt{integer}, \texttt{number}, arrays of integers, and arrays of numbers.
-
-\subsection{Other projects}
-
-Typed Racket \cite{tobin-hochstadt2008ts} is a statically typed version
-of the Racket language, which is a Scheme dialect.
-The main purpose of Typed Racket is to allow programmers to combine
-untyped modules, which are written in Racket, with typed modules, which are
-written in Typed Racket.
-It also uses local type inference to deduce the type of unannotated expressions.
-
-The main feature of Typed Racket's type system is \emph{occurrence typing}
-\cite{tobin-hochstadt2010ltu}.
-It is a novel way to use type predicates in control flow statements
-to refine union types.
-Occurrence typing is not sound in the presence of mutation.
-As these kinds of checks are common in other languages, related systems
-have appeared \cite{guha2011tlc,winther2011gtp,pearce2013ccf}.
-
-The type system of Typed Racket also includes function types, recursive
-types, and structure types.
-Its function types also handle multiple return values, and there is
-also a way to describe function types that have optional arguments.
-Its structure types are similar to our interfaces, as they describe record types.
-The type system is also structural and based on subtyping.
-It also includes the dynamic type \texttt{Any}, which is the top type in the system.
-Typed Racket also supports polymorphic functions and data structures.
-
-Typed Clojure \cite{bonnaire-sergeant2012typed-clojure} is an
+Tidal Lock~\cite{tidallock} is a prototype of another optional type system for Lua. It covers just a small subset of Lua
+in its current form. Its most remarkable feature is
+how it structures its table types to support a form of type
+evolution through imperative assignment. Typed Lua uses
+the same general idea of letting the type of a table evolve
+through assignment, but the structure of both the table types
+and the typing rules that support this are completely different.
+
+Sol~\cite{sol} is another experimental optional type system
+for Lua. While it has some similarities to Typed Lua, it
+has more limited table types: Sol tables can only be lists,
+maps, and objects that follow a specific object-oriented
+idiom that Sol introduces. There is no evolution of table
+types.
+
+Lua Analyzer~\cite{luaanalyzer} is another optional type
+system for Lua that is specially designed to work with
+Löve Studio, an IDE for game development using Lua.
+It is unsound by design, and primarily for supporting
+IDE autocompletion. There is no evolution of table types.
+
+Ravi~\cite{ravi} is an experimental Lua dialect
+with an LLVM-based JIT compiler. It has a basic
+optional type system to improve performance
+of numerical code. This type system only supports
+numbers (integers and floating point) and arrays.
+
+Typed Racket~\cite{tobin-hochstadt2008ts} is a statically typed version of the Racket language, which is a Scheme dialect.
+The main purpose of Typed Racket is to allow programmers to combine untyped modules, which are written in Racket, with typed modules, which are written in Typed Racket. Typed Racket tracks
+values that cross the boundary between the typed and
+untyped parts to be able to correctly assign blame to type
+errors that occur in the typed parts, so is a gradual type
+system.
+
+Typed Racket introduced {\em occurrence typing}, a form of
+flow typing~\cite{tobin-hochstadt2010ltu}, where type
+predicates are used to refining union types.
+As this use of type predicates is common in other
+dynamically-typed languages, related systems
+have appeared~\cite{guha2011tlc,winther2011gtp,pearce2013ccf}.
+
+Typed Clojure~\cite{bonnaire-sergeant2012typed-clojure} is an
 optional type system for Clojure.
-Although Clojure is a Lisp dialect that runs on the Java Virtual Machine,
-Common Language Runtime, and JavaScript, Typed Clojure runs only on
-the Java Virtual Machine.
-Perhaps, this restriction pushed Typed Clojure to support Java classes
-and some Java types such as \texttt{Long}, \texttt{Double}, and \texttt{String}.
-Typed Clojure also provides optional type annotations and uses
-local type inference to deduce the type of unannotated expressions.
-It also assigns the type \texttt{Any} to unannotated function parameters,
-which is the top type in the type system.
-
 The type system of Typed Clojure includes polymorphic function types,
 union types, intersection types, lists, vectors, maps, sets, and recursive types.
-Function types can also have rest parameters, which are similar
-to our variadic types, but can only appear on the input parameter
-of function types.
-In fact, its function types cannot return multiple results.
-It also uses occurrence typing to allow control flow statements to
-refine union types.
-The type system is also structural and based on subtyping.
-
-Dart \cite{dart} is a new class-based object-oriented programming
-language.
-It includes optional type annotations and compiles to JavaScript.
-The type system of Dart is nominal and includes base types,
-function types, lists, and maps.
-It also supports generics, and the programmer can define
-generic functions, lists, and maps.
-Unlike Typed Lua, Dart is unsound by design.
-
-Even though Dart has optional typing and static types by
-default do not affect run-time semantics, it has an
-execution mode that affects run-time.
-The \emph{checked mode} inserts run-time assertions that
-verifies whether static types match run-time tags.
-The \emph{production mode} is the default execution mode
-that does not include any assertions.
-
-TypeScript \cite{typescript} is a JavaScript extension
+Function types can have {\em rest} parameters, which are similar
+to our variadic types, but Typed Clojure functions cannot return
+multiple results. Typed Clojure also uses occurrence typing
+
+Dart~\cite{dart} is class-based object-oriented programming
+language with an optional type system. Dart's type system
+is nominal, and includes polymorphic types. The type
+system is unsound by design, with instances of polymorphic
+types being covariant in their type arguments.
+
+TypeScript~\cite{typescript} is a JavaScript extension
 that includes optional type annotations and class-based
 object-oriented programming.
-It also uses local type inference to deduce the type
-of unannotated expressions.
 The type system of TypeScript is structural, based
-on subtyping, and supports generics.
-It includes the dynamic type, primitive types, union types,
-function types, array types, tuple types, recursive types, and
-object types.
-Unlike Typed Lua, TypeScript uses arrays to represent variadic
-functions and multiple return values.
-
-Even though TypeScript is unsound by design,
-Bierman et al. \cite{bierman2014typescript} shows how to
-make TypeScript sound.
-They use a reduced core of TypeScript to formalize a
-sound type system for TypeScript, but also to formalize
-its current unsound type system.
-
-TeJaS \cite{lerner2013tejas} is a framework for the construction of
-different type systems for JavaScript.
-The authors created a base type system for JavaScript with
-extensible typing rules that allow the experimentation of
-different static analysis.
-They used TeJaS to create a type system that simulates the
+on subtyping, and has polymorphic types, union types,
+function types, array types, tuple types, recursive types,
+record types, and
+object types. It is unsound by design, although there is a
+sound subset that forms the basis of a gradual type
+system~\cite{bierman2014typescript}. Although JavaScript
+also has the same idiom of building a record through
+imperative assignment to an initially empty record,
+TypeScript does not have evolution of its record and
+object types; the fact that {\tt undefined} is a bottom 
+type in its subtyping relation still lets programmers use
+the idiom, at the cost of type safety.
+
+TeJaS~\cite{lerner2013tejas} is a framework for the
+construction of different type systems for JavaScript.
+It has a base type system with extensible typing rules.
+As an example, the paper shows how TeJaS can simulate the
 type system of TypeScript.
 
-Politz et al. \cite{politz2012semantics} proposes semantics
+Politz et al.~\cite{politz2012semantics} proposes semantics
 and types for objects with first-class member names, a well-known
-feature from scripting languages.
-Their type system uses string patterns to describe the members of
+feature from scripting languages. Their type system uses string 
+patterns to describe the members of
 an object, and define a complex subtyping relation to validate
@@
 
-Gradualtalk \cite{allende2013gts} is a Smalltalk dialect that
-supports gradual typing.
-The type system combines nominal and structural typing.
-It includes function types, union types, structural types,
-nominal types, a self type, and parametric polymorphism.
-The type system also relies on subtyping and consistent-subtyping.
-
-Gradualtalk inserts run-time checks that ensure dynamically
+Gradualtalk~\cite{allende2013gts} is a Smalltalk dialect that
+with a gradual type system. Its type system combines
+ nominal and structural typing.
+It includes function types, union types, structural object types,
+nominal object types types, a self type, and parametric polymorphism. Like Typed Lua, Gradualtalk formalizes the
+interaction between typed and untyped code with consistent-subtyping.
+
+Gradualtalk inserts runtime checks that ensure dynamically
 typed code does not violate statically typed code.
-Allende et al. \cite{allende2013cis} perform a careful
-evaluation about cast insertion in Gradualtalk.
-They report that usually cast insertions impact on execution
-performance, so Gradualtalk also has an option that allows
-programmers to turn them off, downgrading Gradualtalk
-to an optional type system.
-
-Reticulated Python \cite{vitousek2014deg} is a Python compiler
-that supports gradual typing.
-The type system is structural and based on subtyping.
-It includes base types, the dynamic type, list types,
+Allende et al.~\cite{allende2013cis} performs a careful
+evaluation of the performance impact of these checks, and
+conclude that the performance hit is large enough that
+Gradualtalk needs to have an option that turns off these
+checks, downgrading Gradualtalk into an optional type
+system.
+
+Reticulated Python~\cite{vitousek2014deg} is a
+gradual type system for Python. It is structural, based on subtyping, and includes list types,
 dictionary types, tuple types, function types, set types,
 object types, class types, and recursive types.
-It includes class and object types to differentiate the
-type of class declarations and instances, respectively.
-It also uses local type inference.
 Besides static type checking, Reticulated Python also introduces
-three different approaches for inserting run-time assertions.
-
-Mypy \cite{mypy} is an optional type system for Python.
+three different approaches for inserting runtime assertions
+that preserve the gradual typing guarantee.
+
+Mypy~\cite{mypy} is an optional type system for Python.
 The type system of mypy is similar to the type system of
-Reticulated Python, but mypy does not insert run-time checks
-and it has parametric polymorphism.
-In contrast, Reticulated Python can type variadic functions,
-but mypy cannot.
-Recently, Guido van Rossum, Python's author, proposed a
-standard syntax for type annotations in Python \cite{PEP483}
-that is extremely inspired by mypy \cite{PEP484}.
-The main goal of this proposal is to make easier building
+Reticulated Python, but mypy has parametric polymorphism and
+does not insert runtime checks.
+
+Guido van Rossum, Python's designer, proposed a
+standard syntax for type annotations in Python
+that is inspired by mypy~\cite{PEP484}.
+The goal of this proposal is to make it easier to build
 static analysis tools for Python.
-Typing \cite{typing} is a tool that is being developed to
+Typing~\cite{typing} is a tool that is being developed to
 implement this proposal.
 
-Hack \cite{hack} is a new programming language that runs on the
-Hip Hop Virtual Machine (HHVM).
-The HHVM is a virtual machine that executes Hack and PHP programs.
-We can view Hack as an extension to PHP that combines static and
-dynamic typing.
-The type system of Hack includes generics, nullable types, collections,
-and function types.
-
-The Ruby Type Checker \cite{ren2013rtc} is a library that
-performs type checking during run-time.
-The library provides type annotations that the programmer
-can use on classes and methods.
-Its type system includes nominal types, union types,
-intersection types, method types, parametric polymorphism,
-and type casts.
-
-Grace \cite{black2013sg} is an object-oriented language
-with optional typing.
-Grace is not a dynamically typed language that has been
-extended with an optional type system, but a language
-that has been designed from scratch to have both
-static and dynamic typing.
-Homer et al. \cite{homer2013modules} explores some
-useful patterns that derive from Grace's use of objects as modules
-and its brand of optional structural typing, which
-can also be expressed with Typed Lua's modules as tables.
+Hack~\cite{hack} is an optionally-typed dialect of PHP
+that runs on the Hip Hop Virtual Machine (HHVM).
+It has nominal object types, parametric polymorphism,
+nullable types, array types, function types, and tuple
+types.
+
+The Ruby Type Checker~\cite{ren2013rtc} is a library that
+augments the regular dynamic type checking of the Ruby
+implementation with extra runtime checks generated from type
+annotations that the programmer can use on classes and
+methods. It has nominal object types, union types, intersection
+types, method types, and parametric polymorphism.
+
+Grace~\cite{black2013sg} is an object-oriented language
+with optional typing. Like Dart, it is a language
+that has been designed from scratch to have an
+optional type system, instead of an existing dynamically
+typed language that has been retrofitted with a type
+system. Unlike Dart, the type system of Grace is not
+deliberately unsound.
 

### 2015-06-09 doc/papers/paper/paper.tex
another rewrite of related work section of DLS paper, for consolidation and simplification

@@
 
-Typed Clojure~\cite{bonnaire-sergeant2012typed-clojure} is an
-optional type system for Clojure.
-The type system of Typed Clojure includes polymorphic function types,
-union types, intersection types, lists, vectors, maps, sets, and recursive types.
-Function types can have {\em rest} parameters, which are similar
-to our variadic types, but Typed Clojure functions cannot return
-multiple results. Typed Clojure also uses occurrence typing
-
-Dart~\cite{dart} is class-based object-oriented programming
-language with an optional type system. Dart's type system
-is nominal, and includes polymorphic types. The type
-system is unsound by design, with instances of polymorphic
-types being covariant in their type arguments.
-
-TypeScript~\cite{typescript} is a JavaScript extension
-that includes optional type annotations and class-based
-object-oriented programming.
-The type system of TypeScript is structural, based
-on subtyping, and has polymorphic types, union types,
-function types, array types, tuple types, recursive types,
-record types, and
-object types. It is unsound by design, although there is a
-sound subset that forms the basis of a gradual type
-system~\cite{bierman2014typescript}. Although JavaScript
-also has the same idiom of building a record through
-imperative assignment to an initially empty record,
-TypeScript does not have evolution of its record and
-object types; the fact that {\tt undefined} is a bottom 
-type in its subtyping relation still lets programmers use
-the idiom, at the cost of type safety.
-
-TeJaS~\cite{lerner2013tejas} is a framework for the
-construction of different type systems for JavaScript.
-It has a base type system with extensible typing rules.
-As an example, the paper shows how TeJaS can simulate the
-type system of TypeScript.
-
-Politz et al.~\cite{politz2012semantics} proposes semantics
-and types for objects with first-class member names, a well-known
-feature from scripting languages. Their type system uses string 
-patterns to describe the members of
-an object, and define a complex subtyping relation to validate
-these patterns.
-They also provide an implementation of their system to JavaScript.
-
-Gradualtalk~\cite{allende2013gts} is a Smalltalk dialect that
+Gradualtalk~\cite{allende2013gts} is a Smalltalk dialect 
 with a gradual type system. Its type system combines
@@
 nominal object types types, a self type, and parametric polymorphism. Like Typed Lua, Gradualtalk formalizes the
-interaction between typed and untyped code with consistent-subtyping.
-
-Gradualtalk inserts runtime checks that ensure dynamically
-typed code does not violate statically typed code.
-Allende et al.~\cite{allende2013cis} performs a careful
-evaluation of the performance impact of these checks, and
-conclude that the performance hit is large enough that
-Gradualtalk needs to have an option that turns off these
-checks, downgrading Gradualtalk into an optional type
-system.
+interaction between typed and untyped code with consistent-subtyping. Due to the performance impact of
+the runtime checks that ensure the gradual typing guarantees,
+Gradualtalk can be downgraded into an optional type system
+through a configuration switch~\cite{allende2013cis}.
 
@@
 
-Mypy~\cite{mypy} is an optional type system for Python.
-The type system of mypy is similar to the type system of
-Reticulated Python, but mypy has parametric polymorphism and
-does not insert runtime checks.
-
-Guido van Rossum, Python's designer, proposed a
-standard syntax for type annotations in Python
-that is inspired by mypy~\cite{PEP484}.
-The goal of this proposal is to make it easier to build
-static analysis tools for Python.
-Typing~\cite{typing} is a tool that is being developed to
-implement this proposal.
-
-Hack~\cite{hack} is an optionally-typed dialect of PHP
-that runs on the Hip Hop Virtual Machine (HHVM).
-It has nominal object types, parametric polymorphism,
-nullable types, array types, function types, and tuple
-types.
-
-The Ruby Type Checker~\cite{ren2013rtc} is a library that
-augments the regular dynamic type checking of the Ruby
-implementation with extra runtime checks generated from type
-annotations that the programmer can use on classes and
-methods. It has nominal object types, union types, intersection
-types, method types, and parametric polymorphism.
-
-Grace~\cite{black2013sg} is an object-oriented language
-with optional typing. Like Dart, it is a language
-that has been designed from scratch to have an
-optional type system, instead of an existing dynamically
-typed language that has been retrofitted with a type
-system. Unlike Dart, the type system of Grace is not
-deliberately unsound.
+Several dynamically typed languages now have optional
+type systems: Clojure~\cite{bonnaire-sergeant2012typed-clojure},
+JavaScript~\cite{typescript}, Python~\cite{mypy}, 
+and PHP~\cite{hack}. While Lua has some similarities to
+all of these languages, none of these optional type
+systems have the features described in this paper.
+Most of these languages do not have idioms
+that inspired these features; JavaScript has the
+idiom of adding fields to an initially empty
+record through assignment, but TypeScript sidesteps
+the issue by having allowing an empty record to
+have any record or object type, as the type of
+missing fields is the bottom type of its type system.
+
+Dart~\cite{dart} and Grace~\cite{black2013sg} are
+two languages that have been designed from scratch
+to have an optional type system, instead of being
+existing dynamically typed languages with a retrofitted 
+type system.
 

### 2015-06-10 doc/papers/paper/paper.tex
rewrote conclusion of DLS paper, now down to 13 pages

@@
 
-In this work we presented Typed Lua, an optional type system for Lua.
-We implemented Typed Lua as a Lua extension that allows programmers to
-combine static and dynamic typing in Lua code, making easier the evolution
-of simple scripts into large programs.
-
-Our main contribution is the formalization of a complete optional type
-system that introduces several novel type system features to statically
-type check Lua programs.
-Even though Lua shares several features with other dynamically
-typed languages such as JavaScript, Lua also has several unusual features.
-These unusual features include tables (or associative arrays) as the sole
-mechanism for structured data, besides functions with multiple return values
-and flexible arity that interact with multiple assignment.
-We highlight the following novel features of our type system:
-\begin{itemize}
-\item type refinement allows the incremental evolution of record and
-object types, playing an important role in statically type checking
-the idiomatic way in which Lua programmers use tables to define modules
-and objects;
-\item projection types handle functions that are overloaded on the
-number and types of return values, allowing programmers to narrow the
-types of a set of variables by narrowing the type of a single component
-of this set;
-\item union types and variadic types help our type system handle
-functions with flexible arity, that is, union types are helpful in
-describing optional parameters while variadic types are helpful in
-describing the type of the vararg expression and the type of functions
-that can receive or return any number of values.
-\end{itemize}
+We have presented a formalization of the Typed Lua,
+an optional type system for Lua, with a focus on rules
+for two novel features of the system that type unusual
+idioms present in Lua programs.
+
+In the first idiom, records and objects are
+defined through assignment to an initially empty
+table. In Typed Lua's type system the type of a
+table can change, either by assignment
+or by using the table in a context that expects
+a different (but compatible) type than the one it
+has. The system tracks aliasing of table references
+to be able to do this type evolution in a type-safe
+way.
+
+In the second idiom, a function that can return different
+kinds of sequences of return values (for example, one
+sequence for its usual path, another for error conditions)
+is modeled by having a union of tuple types as its return
+type. A destructuring assignment on a union of tuple types
+decomposes the tuple in a way that the dependencies between
+the types of each member of the union can be tracked.
+Narrowing the type of one of the tuple's components with
+a type predicate can then narrow the types of the other
+components.
 
 A key feature in optional type systems is usability.
-This means that optional type systems should not change the idioms
-that programmers are already familiar with.
-Instead, optional type systems should fit existing idioms to
-statically type check them.
-Designing a too simple type system can overload programmers by forcing
-them to change the way they program in the language to fit the type system,
-while designing a too complex type system can overload programmers with
-types and error messages that are hard to understand, even if type inference
-removes the necessity of annotating the program with these complex types.
-The most challenging aspect of designing optional type systems is to find
-the right amount of complexity for a type system that feels natural to the programmers.
-
-Usability has been a concern in the design of Typed Lua since the beginning.
-We realized that we should not rely on the semantics of Lua only,
-as this could lead to a cumbersome type system that would not support
-several Lua idioms.
-For this reason, we performed a mostly automated survey of Lua idioms
-and features to inform our design choices.
-
-After designing and implementing Typed Lua, we performed several
+Optional type systems retrofitted to an existing
+language should fit the language's idioms, adding
+static type safety to them. If the system is too
+simple it may require too much of a change in the way
+the programmers use the language. On the other hand,
+if it is too complex it may overload the programmers
+with types and error messages that are hard to
+understand. The most challenging aspect of designing optional type systems is to find the right amount of complexity for a type system that feels natural to the programmers.
+
+Usability has been a concern in the design of Typed Lua since the beginning. We realized that a design based solely on what
+is possible by the semantics of Lua could lead to a 
+complex type system. For this reason, we surveyed a large
+corpus of Lua code to identify important idioms that
+our type system should support.
+
+We performed several
 case studies to evaluate how successful we were in our goal of
@@
 at least 89\% of the members from each module.
-Our evaluation results showed that our type system can statically
+
+Our evaluation results show that our type system can statically
 type check several Lua idioms and features, though the evaluation
-results also exposed several limitations of our type system.
+also exposed several limitations of our type system.
 We found that the three main limitations of our type system are
-the lack of intersection types, parametric polymorphism, and operator overloading.
+the lack of full support for overloaded functions, the lack
+of parametric polymorphism, and operator overloading.
 Overcoming these limitations is our major target for future work,
@@
 However, we still do not have proofs that the novel features of
-our type system are sound.
-We see a soundness proof as another major future work, as it is
-necessary to use static types for code optimization.
-
-Finally, we believe that Typed Lua is a major contribution to the Lua community,
-because it offers a framework that programmers can use to document,
-test, and better structure their applications.
-For libraries where a full conversion to static type checking should
-prove unfeasible or too much work, the community can use Typed Lua
-just to document the external interfaces of the libraries,
-giving the benefits of static type checking to the users of these
-libraries.
-In fact, we already have user feedback from Lua programmers that are
-using Typed Lua in their projects.
-For instance, ZeroBrane Studio is an IDE for Lua development that is
-starting to use Typed Lua to perform static analysis in Lua code.
+our type system are sound. We are working on a generalization
+of the typing rules for evolution of table types where tagged
+types can be attached to arbitrary references instead of
+just local variables, and expect that a proof of soundness
+will come out of this effort.
+
+Typed Lua has a working implementation that
+Lua programmers can already use as a framework to
+document, test, and structure their applications,
+and we already have user feedback from Lua programmers
+that are using Typed Lua in their projects.
+One example is ZeroBrane Studio, a Lua IDE that is
+integrating Typed Lua.
+
+Even applications where a full conversion to static
+type checking is unfeasible in the current state
+of the type system, or too much work, can use Typed
+Lua to document the external interfaces of its
+libraries, giving the benefits of static typing
+checking to the users of those libraries.
 

### 2015-06-10 doc/papers/paper/paper.tex
Fixing the statistics about list iteration

@@
 type system, and their subtyping and consistent-subtyping relations;
-Section \ref{sec:tables} presents the typing rules for evolution of table types;
-Section \ref{sec:projections} presents the typing rules for
-destructuring unions of tuples into projection types and
-flow typing on these projections;
+Section \ref{sec:rules} presents the typing rules for evolution of table types,
+and the typing rules for destructuring unions of tuples into projection types,
+allowing flow typing on these projections;
 Section \ref{sec:related} reviews related work;
@@
 scripts in our corpus iterate over a map at least once,
-and 31\% of the scripts iterate over a list
-at least once. \emph{TODO: Não sei se esse número está correto,
-apenas somei as duas porcentagens, mas pode haver interseção.}
+and 27\% of the scripts iterate over a list at least once.
 

### 2015-06-10 doc/papers/paper/paper.tex
rewrite of section on projection types

@@
 
-\subsection{Projections}
+\subsection{Projection Types}
 \label{sec:projections}
 
-Lua programmers often overload the return type of functions to denote errors,
-and our type system uses projection types to handle this idiom.
-
-As an example, let us assume that \emph{idiv} and \emph{print} are functions
-in the environment.
-The function \emph{idiv} performs integer division and has type
-\begin{align*}
-& \Integer \times \Integer \times \Void \rightarrow \\
-& (\Integer \times \Integer \times \Void) \sqcup (\Nil \times \String \times \Void)
-\end{align*}
-In case of success, it returns two integers: the result and the remainder.
-In case of failure, it returns $\Nil$ plus an error message that describes
-the error.
-The function \emph{print} is a variadic function of type
-$\Value{*} \rightarrow \Void$.
-Let us see how our type system type checks the following program:
+As we have seen in Section~\ref{sec:statistics}, Lua
+programmers often overload the return type of functions
+to denote errors, returning a {\tt nil} result and
+an error message in case of an error instead of the
+usual return values.
+
+As an example, consider the integer division function
+below, which returns the quotient and rest of integer
+division, or a division by zero error:
+\begin{verbatim}
+  local function idiv(x : integer, y : integer)
+    if divisor == 0 then
+      return nil, "division by zero"
+    else
+      return x // y, x % y
+    end 
+  end
+\end{verbatim}
+
+Typed Lua infers the return type of this function as
+$(\Integer \times \Integer \times \Nil*) \sqcup (\Nil \times \String \times \Nil{*})$. The idiom for using this such
+a function is to test the first returned value:
+\begin{verbatim}
+  local q, r = idiv(a, b)
+  if q then
+    -- assume q and r are integers
+  else 
+    -- assume q is nil r is a string
+  end
+\end{verbatim}
+
+A standard way of decomposing the union in the assignment
+would let the type of {\tt q} be $\Integer \times \Nil$ and
+the type of {\tt r} be $\Integer \times \String$. Flow typing
+would narrow the type of {\tt q} to $\Integer$ in the {\tt then}
+branch and to $\Nil$ in the {\tt else} branch, but would
+leave the type of {\tt r} unchanged.
+
+{\em Projection types}
+are a general way to make the dependency between the types
+of {\tt q} and {\tt r} survive the decomposition of the tuple,
+so flow typing can narrow all of the components by testing
+just one of them.
+
+Intuitively, a projection type $\pi_{i}^{x}$ denotes the union of
+the i-th components of all the tuples in the union of tuple types
+denoted by the label $x$. Unions of tuple types referenced by projections are kept in their own {\em projection environment} $\Pi$. A fresh label is generated whenever the last term
+in an expression list has a union of tuple types as its type.
+
+Before introducing the typing rules for projections, let us translate the example above into the abstract syntax of Figure~\ref{fig:syntax}:
 \begin{center}
 	\begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(1, 2) \, \mathbf{in}$}\\
-& \multicolumn{1}{l}{$\mathbf{if} \, q \, \mathbf{then} \, \lfloor print(q + r) \rfloor_{0} \, \mathbf{else} \, \lfloor print(``ERROR: " \, .. \, r) \rfloor_{0}$}
+		\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(1, 2) \, \mathbf{in} \, \mathbf{if} \, q \, \mathbf{then} \, s_1 \, \mathbf{else} \, s_2$}
 	\end{tabular}
@@
 
-First, our type system uses the auxiliary relation
-$\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$
-for type checking $idiv(1, 2)$.
-This relation means that given a type environment $\env_{1}$ and
-a projection environment $\penv$, we can check that an expression
-list $el$ has type $S_{1}$ and produces a new type environment $\env_{2}$
-and produces a pair $(x,S_{2})$.
-This pair means that the last expression of an expression list $el$
-produces an union of second-level types $S_{2}$ that should be bound
-to a variable $x$ in the projection environment $\penv$,
-as the resulting type of this expression is a tuple of projection
-types $\pi_{i}^{x}$.
-In our example, our type system uses rule \textsc{T-EXPLIST4} for
-type checking $idiv(1, 2)$:
-\[
-\begin{array}{c}
-\mylabel{T-EXPLIST4}\\
+Notice that the local variables $q$ and $r$ are declared without 
+an explicit type. The rule for typing this {\bf local} statement,
+{\sc T-LOCAL-UN}, introduces a new label in the projection
+environment used for typing the {\bf if} statement:
+\[
+\begin{array}{c}
+\mylabel{T-LOCAL-UN}\\
 \dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
-       \env_{1}, \penv \vdash me:S, \env_{n+2}\\
-       S = T_{n+1} \times ... \times T_{n+m} \times \Void \sqcup T_{n+1}' \times ... \times T_{n+m}' \times \Void\\
-       \env_{f} = merge(\env_{1}, ..., \env_{n+2}) \;\;\;
-       n = |\;\overline{e}\;|
-       \end{array}}
-      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times \Nil{*}, \env_{f}, (x,S)}
-\end{array}
-\]
-
-Note that $idiv(1, 2)$ has type
-$\pi_{1}^{x} \times \pi_{2}^{x} \times \Nil{*}$ and produces the pair
-$(x,(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}))$.
-
-In the rule that type checks the declaration of unannotated variables,
-our type system uses the pair $(x,S_{2})$ to bound a union of
-second-level types $S_{2}$ to a variable $x$ in the projection
-environment $\penv$.
-In our example, declaring $q$ and $r$
-bounds the projection type $\pi_{1}^{x}$ to $q$ and
-bounds the projection type $\pi_{2}^{x}$ to $r$,
-where the projection variable $x$ bounds to 
-\[
-(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})
-\]
-in the projection environment $\penv$.
-Rule \textsc{T-LOCAL2} illustrates this intuition:
-\[
-\begin{array}{c}
-\mylabel{T-LOCAL2}\\
-\dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash el:S_{1}, \env_{2}, (x,S_{2})\\
-       \env_{3} = \env_{2}[id_{1} \mapsto infer(S_{1},1), ..., id_{n} \mapsto infer(S_{1},n)]\\
-       \env_{3}, \penv[x \mapsto S_{2}] \vdash s, \env_{4} \;\;\;
+	\env_{1}, \penv \vdash el:T_1 \times \ldots T_n \times P, \env_{2}, (x,S)\\
+	\env_{3} = \env_{2}[id_{1} \mapsto T_1, \ldots, id_{n} \mapsto T_n]\\
+	\env_{3}, \penv[x \mapsto S] \vdash s, \env_{4} \;\;\;
 	n = |\;\overline{id}\;|  
@@
 
-This rule uses the auxiliary function \emph{infer} to get the
-most general first-level types of each variable that should be introduced in
-the type environment for type checking $s$.
-After type checking the statement $s$, rule \textsc{T-LOCAL2} produces a
-new type environment $\env_{3}$ without the variables that it introduced
-before type checking $s$.
-We can define \emph{infer} as follows:
-\begin{align*}
-infer(T_{1} \times ... \times T_{n}{*}, i) & = \left\{
-\begin{array}{ll}
-general(T_{i}) & \text{if $i < n$}\\
-general(T_{n} \cup \Nil) & \text{if $i >= n$}
-\end{array} \right.
-\end{align*}
-
-\begin{align*}
-general(\False) & = \Boolean\\
-general(\True) & = \Boolean\\
-general({\it int}) & = \Integer\\
-general({\it float}) & = \Number\\
-general({\it string}) & = \String\\
-general(T_{1} \cup T_{2}) & = general(T_{1}) \cup general(T_{2})\\
-general(S_{1} \rightarrow S_{2}) & = general2(S_{1}) \rightarrow general2(S_{2})\\
-general(\{K_{1}{:}V_{1}, ..., \\
-K_{n}{:}V_{n}\}_{tag}) & = \{K_{1}{:}general(V_{1}), ..., \\
-& \;\;\;\;\;\; K_{n}{:}general(V_{n})\}_{tag} \\
-general(\mu x.T) & = \mu x.general(T)\\
-general(T) & = T\\
-\\
-general2(\Void) & = \Void\\
-general2(T{*}) & = general(T){*}\\
-general2(T \times P) & = general(T) \times general2(P)\\
-general2(S_{1} \sqcup S_{2}) & = general2(S_{1}) \sqcup general2(S_{2})
-\end{align*}
-
-After assigning projection types to $q$ and $r$, reading $q$ will
-use the projection type $\pi_{1}^{x}$ to project the type of $q$
-into the union type $\Integer \cup \Nil$, while reading $r$ will
-use the projection type $\pi_{2}^{x}$ to project the type of $r$
-into the union type $\Integer \cup \String$.
-Now, we may want to discriminate these variables to check whether
-the function call returned with success.
-
-Introducing a projection variable $x$ in the projection environment allows our
-type system to discriminate projection types $\pi_{i}^{x}$.
-The rule \textsc{T-IF3} shows the case where our type system
-discriminates a projection type based on the tag \texttt{nil}:
-\[
-\begin{array}{c}
-\mylabel{T-IF3}\\
+This first sequent above the bar uses the auxiliary typing relation $\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$.
+This rule does the work of generating a fresh projection label,
+and projection types associated with it, if the last term of
+the expression list has a union of tuple types as its type,
+as rule {\sc T-EXPLIST-PROJ} below shows:
+\[
+\begin{array}{c}
+\mylabel{T-EXPLIST-PROJ}\\
+\dfrac{\begin{array}{c}
+	\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
+	\env_{1}, \penv \vdash me:S, \env_{n+2} \;\;\; n=|\overline{e}|\\
+	S = S_1 \sqcup S_2 \;\;\;m = max(|S_1|, |S_2|)-1 \;\;\; x \notin \Pi\\
+	\env_{f} = merge(\env_{1}, ..., \env_{n+2})
+	\end{array}}
+{\env_{1}, \penv \vdash \overline{e},me:}\\ {T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times last(S), \env_{f}, (x,S)}
+\end{array}
+\]
+
+In our example, the two rules combined will bind $x$ to
+$(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})$ in the projection environment $\Pi$,
+and $q$ and $r$ respectively to $\pi_{1}^{x}$ and $\pi_{2}^{x}$
+in the type environment used for typing the {\bf if} statement.
+
+At this point, using $q$ or $r$ as an r-value will project
+their types into either $\Integer \cup \Nil$ or $\Integer \cup \String$. But the more interesting case is inside the
+clauses of the {\bf if} statement. The particular uses the rule {\sc T-IF-FNIL-PROJ} below, which uses flow typing:
+\[
+\begin{array}{c}
+\mylabel{T-IF-FNIL-PROJ}\\
 \dfrac{\begin{array}{c}
@@
 
-Rule \textsc{T-IF3} uses the auxiliary functions \emph{fpt} and \emph{gpt}
-to filter a projection $x$, affecting all variables that bind to the same projection.
-For instance, our previous example type checks through rule \textsc{T-IF3},
-because it uses the information provided by the projection type $\pi_{1}^{x}$,
-which is the type of $q$, to make the rule \textsc{T-IF3} use the function call
-\[
-fpt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
-\]
-to discriminate the projection $x$ to the single tuple
-$\Integer \times \Integer \times \Nil{*}$ inside the $\mathbf{if}$ branch,
-and the function call
-\[
-gpt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
-\]
-to discriminate the projection $x$ to the single tuple
-$\Nil \times \String \times \Nil{*}$ inside the $\mathbf{else}$ branch.
+Rule \textsc{T-IF-FNIL-PROJ} uses the auxiliary functions \emph{fpt} and \emph{gpt}
+to filter a projection $x$. Function {\em fpt} filters out $\Nil$ from the i-th
+component of each tuple, removing the entire tuple if the i-th
+component does not exist, or is $\Nil$ itself. Function {\em gpt}
+filters the i-th component to $\Nil$ if the type includes
+$\Nil$, removing the entire tuple if the i-th component does
+not exist, or does not include $\Nil$.
+
+In our example, $x$ would be bound to $\Integer \times \Integer \times \Nil{*}$ in the {\bf then} branch, and would be bound
+to $\Nil \times \String \times \Nil{*}$ inside the {\bf else}
+branch.
+
 Thus, reading $q$ and $r$ projects $\pi_{1}^{x}$ to $\Integer$ and

### 2015-06-10 doc/papers/paper/paper.tex
finishing rewrite of section on projection types for dls paper

@@
 their types into either $\Integer \cup \Nil$ or $\Integer \cup \String$. But the more interesting case is inside the
-clauses of the {\bf if} statement. The particular uses the rule {\sc T-IF-FNIL-PROJ} below, which uses flow typing:
+branches of the {\bf if} statement. The particular uses the rule {\sc T-IF-FNIL-PROJ} below, which uses flow typing:
 \[
@@
 to $\Nil \times \String \times \Nil{*}$ inside the {\bf else}
-branch.
-
-Thus, reading $q$ and $r$ projects $\pi_{1}^{x}$ to $\Integer$ and
-$\pi_{2}^{x}$ to $\Integer$ inside the $\mathbf{if}$ branch,
-but it projects $\pi_{1}^{x}$ to $\Nil$ and $\pi_{2}^{x}$ to $\String$
-inside the $\mathbf{else}$ branch.
-Outside the condition, $q$ and $r$ use the original projection, that is,
-they project to $\Integer \cup \Nil$ and $\Integer \cup \String$, respectively.
+branch. Now $q$ and $r$ project to $\Integer$ and $\Integer$
+in the {\bf then} branch, and to $\Nil$ and $\String$ in the
+{\bf else} branch.
+
+A projection type in l-value position does not trigger
+the projection, effectively forbidding assignment to a
+variable that has a projection type. Allowing assignment
+would be unsafe, as it could break the dependency among the
+components of the union, as the following example shows:
+\begin{verbatim}
+  local q, r = idiv(a, b)
+  r = "foo"
+  if q then
+    -- wrongly assume q and r are integers
+  end
+\end{verbatim}
 

### 2015-06-10 doc/papers/paper/paper.tex
began rewrite and simplification of Section 3 for DLS paper

@@
 
-\section{Types}
+\section{Types and Subtyping}
 \label{sec:types}
-
-In this section we present the abstract syntax of Typed Lua types
-and their subtyping rules.
-
-%In the previous chapter we presented an informal overview of Typed Lua.
-%We showed that programmers can use Typed Lua to combine static and dynamic
-%typing in the same code, and it allows them to incrementally migrate from
-%dynamic to static typing.
-%This is a benefit to programmers that use dynamically typed languages
-%to build large applications, as static types detect many bugs
-%during the development phase, and also provide better documentation.
-
-%In this chapter we present the abstract syntax of Typed Lua types,
-%the subtyping rules, and the most interesting typing rules.
-%Besides its practical contributions, Typed Lua also has some interesting
-%contributions to the field of optional type systems for scripting
-%languages.
-%They are novel type system features that let Typed Lua cover several Lua idioms
-%and features, such as refinement of tables, multiple assignment, and multiple return values.
-
-\subsection{Types}
 
@@
 T ::= & & \textsc{first-level types:}\\
+& \;\; F & \textit{first-class types}\\
+& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
+& | \; \pi_{i}^{x} & \textit{projection types}\\
+F ::= & & \textsc{{\small first-class types:}}\\
 & \;\; L & \textit{literal types}\\
@@
 & | \; \Self & \textit{self type}\\
-& | \; T_{1} \cup T_{2} & \textit{disjoint union types}\\
+& | \; F_{1} \cup F_{2} & \textit{union types}\\
 & | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
@@
 & | \; x & \textit{type variables}\\
-& | \; \mu x.T & \textit{recursive types}\\
-& | \; \phi(T_{1},T_{2}) & \textit{filter types}\\
-& | \; \pi_{i}^{x} & \textit{projection types}\\
+& | \; \mu x.F & \textit{recursive types}\\
 %\multicolumn{3}{c}{}\\
@@
 V ::= & & \textsc{{\small value types:}}\\
-& \;\; T \; | \; \Const \; T &\\ 
+& \;\; F \; | \; \Const \; F &\\ 
 %\multicolumn{3}{c}{}\\
@@
 & \;\; \Void & \textit{void type}\\
-& | \; T{*} & \textit{variadic types}\\
-& | \; T \times P & \textit{pair types}
+& | \; F{*} & \textit{variadic types}\\
+& | \; F \times P & \textit{pair types}
 \end{array}
@@
 Figure \ref{fig:typelang} presents the abstract syntax of
-Typed Lua types.
-Typed Lua splits types into two categories:
+Typed Lua types. They are split into two categories:
 \emph{first-level types} and \emph{second-level types}.
@@
 assignments and function applications.
-First-level types include literal types, base types, the type $\Nil$,
-the top type $\Value$, the dynamic type $\Any$, the type $\Self$,
-union types, function types, table types, recursive types,
-filter types, and projection types.
-Second-level types include tuple types and unions of tuple types.
-Tuple types include the type $\Void$, variadic types, and pair types.
-Types are ordered by a subtype relationship that we introduce
-in the next section, so Lua values may belong to several distinct types.
-
-Literal types represent the type of literal values.
-They can be the boolean values $\False$ and $\True$,
-an integer value, a floating point value, or a string value.
-We will see that literal types are important in our treatment of
-table types as records.
-
-Typed Lua includes four base types: $\Boolean$, $\Integer$, $\Number$, and $\String$.
-The base types $\Boolean$ and $\String$ represent the values that
-Lua tags as \texttt{boolean} and \texttt{string} during run-time.
-Lua 5.3 introduced two internal representations to the tag \texttt{number}:
-\texttt{integer} for integer numbers and \texttt{float} for real numbers.
-Lua does automatic promotion of \texttt{integer} values to \texttt{float}
-values as needed.
-We introduced the base type $\Number$ to represent \texttt{float} values,
-and the base type $\Integer$ to represent \texttt{integer} values.
-In the next section we will show that $\Integer$ is a subtype of $\Number$.
-This allows programmers to keep using \texttt{integer} values where
-\texttt{float} values are expected.
-
-The type $\Nil$ is the type of \texttt{nil}, the value that Lua uses for
+
+First-level types are
+themselves split into {\em first-class types}, which can be
+used in type annotations and most type constructors, and
+{\em filter types} and {\em projection types}, special
+types that only appear in support of flow typing.
+Typed Lua includes projection types as a way to project
+unions of tuple types into unions of first-level types,
+and we detailt their use in Section~\ref{sec:projections} .
+
+Types are ordered by a subtype relationship, so Lua values may belong to several distinct types. The details of the subtyping
+relationship are given later in this section.
+
+Literal types represent the type of literal boolean, number
+and string values. Their main application is the formation in record and tuple types.
+
+The base types represent values that Lua tags as booleans,
+strings, or numbers during runtime. Prior to version Lua 5.3 all Lua numbers were double-precision floating point numbers.
+Lua 5.3 introduced the possiblity of having 64-bit integers
+as numbers, with automatic promotion of integer values to
+floating-point representation as needed. As automatic
+promotion makes integers essentialy subtypes of floating-point numbers, Typed Lua uses $\Number$ as the type of floating-point numbers, and $\Integer$ as the type of integers.
+
+The type $\Nil$ is the type of {\tt nil}, the value that Lua uses for
 undefined variables, missing parameters, and missing table keys.
+Many programming languages include a bottom type to represent
+an empty value that programmers can use as a default expression.
+We could have used the type $\Nil$ for this role, but
+this would lead to reduced static type safety. Typed Lua's
+type system does not have a bottom type.
 
 The type $\Value$ is the top type, which represents any Lua value.
-This type along with variadic types help the type system to drop
-extra values on assignments and function calls, thus preserving the
-semantics of Lua in these cases.
-
-Typed Lua uses the type $\Self$ to represent the \emph{receiver}
-in object-oriented method definitions and method calls.
-As we mentioned in our informal overview of Typed Lua \cite{maidl2014tl},
-we need the type $\Self$ to prevent programs from indexing a method without
-calling it with the correct receiver.
-
-Union types $T_{1} \cup T_{2}$ represent data types that can hold a value
-of two different types.
-
-Function types have the form $S_{1} \rightarrow S_{2}$ and represent Lua functions,
-where $S$ is a second-level type.
-
-Second-level types are either tuple types or unions of tuple types.
-Tuple types are tuples of first-level types that can end with
-either an empty tuple or with a variadic type.
-Typed Lua needs second-level types because tuples are not first-class
-values in Lua, only appearing on argument passing, multiple returns,
-and multiple assignments.
-The type $\Void$ is the type of an empty tuple.
-A variadic type $T{*}$ represents a sequence of values of type $T \cup \Nil$;
-it is the type of a vararg expression.
-Second-level types include unions of tuples because Lua programs
-usually overload the return type of functions to denote error,
-as we mentioned in Section \ref{sec:statistics}.
-For clarity, we use the symbol $\sqcup$ to represent the union between
-two different tuple types.
-Note that $\cup$ represents the union between two first-level types,
-while $\sqcup$ represents the union between two tuple types.
-
-Back to first-level types, table types represent the various forms
+Its main purpose is to be able to easily express function types
+that silently ignore extra arguments, as well as easily ignoring extra r-values in multiple assignment, preserving the semantics of Lua for these operations, as they do not sacrifice type safety.
+
+The type $\Self$ is the receiver in object-oriented method
+definitions and method calls, and lets Typed Lua have objects
+as regular table types with methods as regular function types
+with an explicit receiver parameter.
+
+Table types represent the various forms
 that Lua tables can take.
@@
 
-Recursive types have the form $\mu x.T$,
-where $T$ is a first-level type that $x$ represents.
-As an example,
-\[
-\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}
-\]
-is a type for singly-linked lists of integers.
-In Typed Lua, we can use the following interface declaration
-as an alias to this type:
-\begin{verbatim}
-    local interface Element
-      info:integer
-      next:Element?
-    end
-\end{verbatim}
-
-Typed Lua includes filter types as a way to discriminate the type of local
-variables inside conditions.
-Our type system uses filter types to formalize the \texttt{type} predicates
-that we introduced in our previous design of Typed Lua \cite{maidl2014tl}.
-This means that \texttt{type} predicates use filter types of the form
-$\phi(T_{1},T_{2})$ to discriminate local variables that are bound to
-union types.
-In a filter type $\phi(T_{1},T_{2})$, $T_{1}$ is the original type and
-$T_{2}$ is the discriminated type.
-
-Typed Lua includes projection types as a way to project
-unions of tuple types into unions of first-level types.
-In Section \ref{sec:projections} we will show in more detail how our type system
-uses them as a mechanism for handling unions of tuple types,
-when they appear in the right-hand side of the declaration of local variables,
-as we introduced in our previous design of Typed Lua \cite{maidl2014tl}.
-We also show how this feature allows our type system to constrain
-the type of a local variable that depends on the type of another local variable.
-
-Typed Lua includes the dynamic type $\Any$ for allowing programmers
-to mix static and dynamic typing.
-There is a subtle difference between the dynamic type $\Any$ and the top type $\Value$.
-Although both types mean that they accept a value of any other type,
-the type $\Value$ is not a good option for handling the
-interaction between dynamically typed and statically typed code.
-Gradual typing uses the dynamic type $\Any$ to identify
-where it should insert run-time checks for asserting that dynamically
-typed code does not violate statically typed code.
-Typed Lua also uses the dynamic type $\Any$ in this sense,
-though it is an optional type system.
-More precisely, we use $\Any$ instead of $\Value$ to allow
-programmers blending dynamic and static typing because we use the
-consistent-subtyping relation to formalize our optional type system,
-as it is a first step to switch Typed Lua from optional typing to
-gradual typing in the future.
+The type $\Any$ is the dynamic type, and allows
+interaction between statically and dynamically typed
+portions of a program. It is neither a top nor a bottom
+type, although it can work as either, as a value of
+any type can be coerced to $\Any$ and vice-versa.
+Gradual type systems augment these coercions with
+information that can be used to blame the dynamically
+typed of the program for type errors that may occur in the
+statically typed part. Currently Typed Lua only 
+warns the programmer that coercions to and from $\Any$
+may be unsafe.
+
+Second-level types are either tuple types or unions of tuple types, and represent sequences of values that appear as arguments, return values, and the left and right sides of
+assignments. Tuple types are heterogeneous lists of first-class
+types that end either with $\Void$ or with a {\em variadic type}.
+A variadic type $T{*}$ represents an unbounded sequence of values of type $T \cup \Nil$.
+
+For clarity, we use the symbol $\sqcup$ to represent the union between two different tuple types, and reserve $\cup$ for the
+union of two first-class types.
 
@@
 \]
-
-Many programming languages include a bottom type to represent
-an empty value that programmers can use as a default expression,
-and we could have used the type $\Nil$ for this role.
-However, making $\Nil$ the bottom type would lead to several expressions
-that would pass the type checker, but that would fail during run-time
-in the presence of a \texttt{nil} value.
-Thus, our type system does not have a bottom type, and $\Nil$ is a
-subtype only of itself and of $\Value$.
 

### 2015-06-11 doc/papers/paper/paper.tex
rewrite of the section on subtyping for DLS paper

@@
 Section \ref{sec:types} presents the types of Typed Lua's
-type system, and their subtyping and consistent-subtyping relations;
+type system; Section~\ref{sec:subtyping} presents
+their subtyping and consistent-subtyping relations;
 Section \ref{sec:rules} presents the typing rules for evolution of table types,
@@
 
-\section{Types and Subtyping}
+\section{Types}
 \label{sec:types}
@@
 types that end either with $\Void$ or with a {\em variadic type}.
-A variadic type $T{*}$ represents an unbounded sequence of values of type $T \cup \Nil$.
+A variadic type $F{*}$ represents an unbounded sequence of values of type $F \cup \Nil$.
 
@@
 
-\subsection{Subtyping}
-
-Our type system uses subtyping \cite{cardelli1984smi,abadi1996to} to order
-types and consistent-subtyping \cite{siek2007objects,siek2013mutable}
-to allow the interaction between statically and dynamically typed code.
-We explain the subtyping and consistent-subtyping rules throughout this section.
-However, we focus the discussion on the definition of subtyping because
-we can combine the consistency and subtyping relations to achieve
-consistent-subtyping \cite{siek2007objects,siek2013mutable}.
-The differences between consistent-subtyping and subtyping are the way
-they handle the dynamic type, and the fact that subtyping is transitive,
-but consistent-subtyping is not.
+\section{Subtyping and Consistent-Subtyping}
+\label{sec:subtyping}
+
+Typed Lua's type system uses subtyping to order
+types and  consistent-subtyping~\cite{siek2007objects,siek2013mutable}
+to model compatibility between types in a way that allows
+the interaction between statically and dynamically typed
+parts of the program without collapsing the subtyping
+relation.
+
+The consistent-subtyping relation is defined in terms
+of subtyping, so we focus the discussion in this section
+on characterising the subtyping relation.
 
 We present the subtyping rules as a deduction system for the
-subtyping relation $\senv \vdash T_{1} \subtype T_{2}$.
-The variable $\senv$ is a set of pairs of recursion variables.
-We need this set to record the hypotheses that we assume when checking
-recursive types.
-
-The subtyping rules for literal types and base types include the rules
-for defining that literal types are subtypes of their respective base types,
-and that $\Integer$ is a subtype of $\Number$:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-FALSE}\\
-\senv \vdash \False \subtype \Boolean
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-TRUE}\\
-\senv \vdash \True \subtype \Boolean
+subtyping relation $\senv \vdash t_{1} \subtype t_{2}$.
+The variable $\senv$ is a set of subtyping assumptions used
+by the subtyping relation between recursive types.
+
+Most of the subtyping rules are standard; we will focus
+on the interesting cases, involving second-level types and
+table types.
+
+Tuple types are covariant:
+\[
+\begin{array}{c}
+\mylabel{S-PAIR}\\
+\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
+       \senv \vdash P_{1} \subtype P_{2}}
+      {\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
+\end{array}
+\]
+
+The semantics of Lua lets the right-hand side of an assignment
+produce more values than l-values in the left-hand side,
+with extra values silently dropped. It also lets the
+right-hand side produce fewer values than l-values in the
+left-hand side, using {\tt nil} to plug the holes. The
+same behavior applies for lists of arguments vs. parameters.
+
+Typed Lua preserves this behavior in assignment, but allows
+both functions with strict arity (which use $\Void$) and
+functions with flexible arity (which use variadic types).
+
+The types of expression lists always end in a variadic type
+that is $\Nil{*}$ if the number of values the list produces
+is statically known. Rule \textsc{S-VARNIL-VOID} makes lets
+an expression list with a statically known number of
+values be used as the arguments or return values of a function
+with a strict arity:
+\[
+\begin{array}{c}
+\mylabel{S-VARNIL-VOID}\\
+\senv \vdash \Nil{*} \subtype \Void
+\end{array}
+\]
+
+The other rules let assignment and functions with flexible
+arity reflect the semantics of Lua in a type-safe way. We
+need rule {\sc S-VARIADIC} because just having variadic
+types be covariant would be too restrictive (for example,
+$\Nil{*}$ would not be a subtype of $\Number{*}$):
+\[
+\begin{array}{c}
+\mylabel{S-VARIADIC}\\
+\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \cup \Nil}
+      {\senv \vdash F_{1}{*} \subtype F_{2}{*}}
+\end{array}
+\]
+
+Rules {\sc S-VAR-EXPAND} and {\sc S-VAR-DROP} state the
+conditions when variadic tuples if different length are
+compatible:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-VAR-EXPAND}\\
+\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \;\;\;
+       \senv \vdash F_{1}{*} \subtype P_{2}}
+      {\senv \vdash F_{1}{*} \subtype F_{2} \times P_{2}}
 \end{array}
@@
 \begin{array}{c}
-\mylabel{S-STRING}\\
-\senv \vdash {\it string} \subtype \String
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-INT1}\\
-\senv \vdash {\it int} \subtype \Integer
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-INT2}\\
-\senv \vdash {\it int} \subtype \Number
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-FLOAT}\\
-\senv \vdash {\it float} \subtype \Number
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-INTEGER}\\
-\senv \vdash \Integer \subtype \Number
-\end{array}
-\end{array}
-\]
-
-Subtyping is reflexive and transitive;
-therefore, we could have omitted the rule \textsc{S-INT2}.
-More precisely, we could have defined a transitive rule for first-level
-types instead of defining specific rules for transitive cases.
-For instance, a transitive rule would allow us to derive that
-\[
-\dfrac{\senv \vdash 1 \subtype \Integer \;\;\;
-       \senv \vdash \Integer \subtype \Number}
-      {\senv \vdash 1 \subtype \Number}
-\]
-
-However, we are using the subtyping rules as the template for defining
-the consistent-subtyping rules, and consistent-subtyping is not
-transitive.
-More precisely, we want the subtyping and consistent-subtyping rules
-to differ only in the way they handle the dynamic type.
-Thus, we define the subtyping rules using an algorithmic approach
-that is close to the implementation, as this approach allows us to use
-subtyping to easily formalize consistent-subtyping.
-
-Our type system includes the top type $\Value$,
-so any first-level type is a subtype of $\Value$:
-\[
-\begin{array}{c}
-\mylabel{S-VALUE}\\
-\senv \vdash T \subtype \Value
-\end{array}
-\]
-
-Another type that is only a subtype of itself and of the type $\Value$
-is the type $\Self$.
-
-The subtyping rules for union types are standard:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-UNION1}\\
-\dfrac{\senv \vdash T_{1} \subtype T \;\;\;
-       \senv \vdash T_{2} \subtype T}
-      {\senv \vdash T_{1} \cup T_{2} \subtype T}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-UNION2}\\
-\dfrac{\senv \vdash T \subtype T_{1}}
-      {\senv \vdash T \subtype T_{1} \cup T_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-UNION3}\\
-\dfrac{\senv \vdash T \subtype T_{2}}
-      {\senv \vdash T \subtype T_{1} \cup T_{2}}
-\end{array}
-\end{array}
-\]
-
-The first rule shows that a union type $T_{1} \cup T_{2}$
-is a subtype of $T$ if both $T_{1}$ and $T_{2}$ are subtypes
-of $T$;
-and the other rules show that a type $T$ is a subtype
-of a union type $T_{1} \cup T_{2}$ if $T$ is a subtype of
-either $T_{1}$ or $T_{2}$.
-
-The subtyping rule for function types is also standard:
-\[
-\begin{array}{c}
-\mylabel{S-FUNCTION}\\
-\dfrac{\senv \vdash S_{3} \subtype S_{1} \;\;\;
-       \senv \vdash S_{2} \subtype S_{4}}
-      {\senv \vdash S_{1} \rightarrow S_{2} \subtype S_{3} \rightarrow S_{4}}
-\end{array}
-\]
-
-The rule \textsc{S-FUNCTION} shows that subtyping between
-function types is contravariant on the type of the parameter list
-and covariant on the return type.
-In the previous section we explained why our type system uses
-second-level types to represent the type of the parameter list
-and the return type.
-Now, we explain their subtyping rules.
-
-The type $\Void$ is a subtype of itself and of a variadic type:
-\[
-\begin{array}{c}
-\mylabel{S-VOID}\\
-\senv \vdash \Void \subtype T{*}
-\end{array}
-\]
-
-A variadic type $T{*}$ represents a sequence of values of type
-$T \cup \Nil$, and the rule \textsc{S-VOID} handles the case where
-a given sequence is empty.
-
-The subtyping rule for pair types is the standard covariant rule:
-\[
-\begin{array}{c}
-\mylabel{S-PAIR}\\
-\dfrac{\senv \vdash T_{1} \subtype T_{2} \;\;\;
-       \senv \vdash P_{1} \subtype P_{2}}
-      {\senv \vdash T_{1} \times P_{1} \subtype T_{2} \times P_{2}}
-\end{array}
-\]
-
-The subtyping rules for variadic types are not so obvious.
-We need six different subtyping rules for variadic types
-to handle all the cases where they can appear.
-
-The rule \textsc{S-VARARG1} is a special rule for handling the
-case where we give a sequence of $\Nil$ to the empty tuple:
-\[
-\begin{array}{c}
-\mylabel{S-VARARG1}\\
-\senv \vdash \Nil{*} \subtype \Void
-\end{array}
-\]
-
-The rule \textsc{S-VARARG2} handles the case where both tuple types end
-with variadic types:
-\[
-\begin{array}{c}
-\mylabel{S-VARARG2}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \cup \Nil}
-      {\senv \vdash T_{1}{*} \subtype T_{2}{*}}
-\end{array}
-\]
-
-This rule shows that $T_{1}{*}$ is a subtype of $T_{2}{*}$
-if $T_{1} \cup \Nil$ is a subtype of $T_{2} \cup \Nil$.
-It explicitly includes $\Nil$ in both sides because otherwise
-$\Nil{*}$ would not be a subtype of several other variadic types.
-For instance, $\Nil{*}$ would not be a subtype of $\Number{*}$,
-as $\Nil \not\subtype \Number$.
-
-The other rules handle the cases where only one tuple type ends with a variadic type:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-VARARG3}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2}}
-      {\senv \vdash T_{1}{*} \subtype T_{2} \times \Void}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-VARARG4}\\
-\dfrac{\senv \vdash T_{1} \subtype T_{2} \cup \Nil}
-      {\senv \vdash T_{1} \times \Void \subtype T_{2}{*}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-VARARG5}\\
-\dfrac{\senv \vdash T_{1}{*} \subtype T_{2} \times \Void \;\;\;
-       \senv \vdash T_{1}{*} \subtype P_{2}}
-      {\senv \vdash T_{1}{*} \subtype T_{2} \times P_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-VARARG6}\\
-\dfrac{\senv \vdash T_{1} \times \Void \subtype T_{2}{*} \;\;\;
-       \senv \vdash P_{1} \subtype T_{2}{*}}
-      {\senv \vdash T_{1} \times P_{1} \subtype T_{2}{*}}
-\end{array}
-\end{array}
-\]
-
-Note that the case where both tuple types end with the type $\Void$ does
-not require any special rule.
-We use the subtyping rules for variadic types along with the types
-$\Value$ and $\Nil$ to make our type system reflect the semantics of
-Lua on discarding extra parameters and replacing missing parameters.
-
-The subtyping rules for unions of tuple types are similar to the
-subtyping rules for unions of first-level types:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-UNION4}\\
-\dfrac{\senv \vdash S_{1} \subtype S \;\;\;
-       \senv \vdash S_{2} \subtype S}
-      {\senv \vdash S_{1} \sqcup S_{2} \subtype S}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-UNION5}\\
-\dfrac{\senv \vdash S \subtype S_{1}}
-      {\senv \vdash S \subtype S_{1} \sqcup S_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-UNION6}\\
-\dfrac{\senv \vdash S \subtype S_{2}}
-      {\senv \vdash S \subtype S_{1} \sqcup S_{2}}
-\end{array}
-\end{array}
-\]
-
-Back to the subtyping rules between first-level types,
-the subtyping rule among a \emph{fixed} or \emph{closed}
-table type and another \emph{closed} table type resembles the
-standard subtyping rule between records:
-\[
-\begin{array}{c}
-\mylabel{S-TABLE1}\\
+\mylabel{S-VAR-DROP}\\
+\dfrac{\senv \vdash F_{1} \subtype F_{2} \cup \Nil \;\;\;
+       \senv \vdash P_{1} \subtype F_{2}{*}}
+      {\senv \vdash F_{1} \times P_{1} \subtype F_{2}{*}}
+\end{array}
+\end{array}
+\]
+
+We could have the axioms $\senv \vdash F{*} \subtype F \cup \Nil \times F{*}$ and $\senv \vdash F \cup \Nil \times F{*}  \subtype F{*}$ instead of the two rules above, which can be derived
+from these axioms by transitivity, but having the rules makes
+the subtyping relation easier to implement.
+
+We will now focus on the subtyping relation among table
+types. Subtyping between a {\em fixed} or {\em closed}
+table type and a {\em closed} table type resembles the
+standard subtyping rule between mutable records, with
+width subtyping plus depth subtyping from a mutable
+or immutable field to an immutable field:
+\[
+\begin{array}{c}
+\mylabel{S-TABLE-FC-C}\\
 \dfrac{\begin{array}{c}
@@
 
-The rule \textsc{S-TABLE1} allows width subtyping and introduces the
-auxiliary relation $\subtype_{c}$ to handle depth subtyping on the
-type of the values stored in the table fields.
-We need an auxiliary relation because the subtyping of the
-type of the values stored in the table fields changes according to
-the tags of the table types.
-We define the relation $\subtype_{c}$ as follows:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-FIELD1}\\
+Rule \textsc{S-TABLE-FC-C} uses the auxiliary relation $\subtype_{c}$ for handling depth subtyping:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-FIELD-C-MM}\\
 \dfrac{\senv \vdash V_{1} \subtype V_{2} \;\;\;
@@
 \begin{array}{c}
-\mylabel{S-FIELD2}\\
+\mylabel{S-FIELD-C-II}\\
 \dfrac{\senv \vdash V_{1} \subtype V_{2}}
@@
 \begin{array}{c}
-\mylabel{S-FIELD3}\\
+\mylabel{S-FIELD-C-MI}\\
 \dfrac{\senv \vdash V_{1} \subtype V_{2}}
@@
 
-These rules allow depth subtyping on $\Const$ fields.
-The rule \textsc{S-FIELD1} defines that mutable fields are invariant,
-while the rule \textsc{S-FIELD2} defines that immutable fields are covariant.
-The rule \textsc{S-FIELD3} defines that it is safe to promote fields
-from mutable to immutable.
-We do not include a rule that allows promoting fields from immutable
-to mutable because this would be unsafe due to variance.
-
-There is a limitation on \emph{closed} table types that led us to
-introduce \emph{open} and \emph{unique} table types.
-If the table constructor had a \emph{closed} table type, then
-programmers would not be able to use it to initialize a variable with
-a table type that describes a more general type.
-For instance,
-\begin{verbatim}
-    local t:{ "x":integer,
-              "y":integer? } = { x = 1, y = 2 }
-\end{verbatim}
-would not type check, as the type of the table constructor would not
-be a subtype of the type in the annotation.
-More precisely,
-\begin{align*}
-& \{``x":1, ``y":2\}_{closed} \not\subtype \\
-& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
-\end{align*}
-
-Simply promoting the type of each table value to its supertype would
-not overcome this limitation, as it still would give to the table constructor
-a \emph{closed} table type without covariant mutable fields.
-Thus, programmers would not be able to use the table constructor to
-initialize a variable with a table type that includes an optional field.
-Using the previous example,
-\begin{align*}
-& \{``x":\Integer, ``y":\Integer\}_{closed} \not\subtype \\
-& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
-\end{align*}
-
-We introduced \emph{unique} table types to avoid this limitation,
-as they represent the type of tables with no keys that do not
-inhabit one of the table's key types, and with no alias.
-In particular, this is the case of the table constructor.
-The following subtyping rule defines the subtyping relation among
-\emph{unique} table types and \emph{closed} table types:
-\[
-\begin{array}{c}
-\mylabel{S-TABLE2}\\
+The subtyping relation between {\em unique} and {\em closed}
+table types is more relaxed, as the typing rules of Section~\ref{sec:tables} guarantee that a {\em unique}
+table type can only appear on the left side of a subtyping
+judgment if the reference to the table with the {\em unique}
+tag is being destroyed:
+\[
+\begin{array}{c}
+\mylabel{S-TABLE-U-C}\\
 \dfrac{\begin{array}{c}
@@
 
-The rule \textsc{S-TABLE2} allows width subtyping and covariant keys.
-It allows covariant keys because we also want to use \emph{unique}
-table types as a way to join table fields that inhabit \emph{closed} table types.
-For instance, we want to use the table constructor to initialize
-a variable with a table type that describes a hash.
-
-The rule \textsc{S-TABLE2} introduced the auxiliary relations
-$\subtype_{u}$ and $\subtype_{o}$.
-The first allows depth subtyping on all fields,
-while the second allows the omission of optional fields.
-We define them as follows:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-FIELD4}\\
+Rule {\sc S-TABLE-U-C} allows with subtyping, depth subtyping
+on any field, not only immutable fields, as well as the addition
+of new fields and the merging of existing fields, as long as the types of the values of new and merged fields include $\Nil$.
+This rule uses the auxiliary relations $\subtype_{u}$ and $\subtype_{o}$, which respectively allow depth subtyping on all fields, and the addition of new fields:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-FIELD-U-MM}\\
 \dfrac{\senv \vdash V_{1} \subtype V_{2}}
@@
 \end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FIELD-U-IM}\\
+\dfrac{\senv \vdash V_{1} \subtype V_{2}}
+{\senv \vdash \Const \; V_{1} \subtype_{u} V_{2}}
+\end{array}
 \\ \\
 \begin{array}{c}
-\mylabel{S-FIELD5}\\
+\mylabel{S-FIELD-U-II}\\
 \dfrac{\senv \vdash V_{1} \subtype V_{2}}
@@
 \begin{array}{c}
-\mylabel{S-FIELD6}\\
+\mylabel{S-FIELD-U-MI}\\
 \dfrac{\senv \vdash V_{1} \subtype V_{2}}
@@
 \begin{array}{c}
-\mylabel{S-FIELD7}\\
+\mylabel{S-FIELD-O-M}\\
 \dfrac{\senv \vdash \Nil \subtype V}
@@
 \begin{array}{c}
-\mylabel{S-FIELD8}\\
+\mylabel{S-FIELD-O-I}\\
 \dfrac{\senv \vdash \Nil \subtype V}
@@
 
-Using \emph{unique} table types to represent the type of the table
-constructor allows our type system to type check the previous example.
-More precisely,
-\begin{align*}
-& \{``x":1, ``y":2\}_{unique} \subtype \\
-& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
-\end{align*}
-
-Even though we allow width subtyping between \emph{unique} and \emph{closed}
-table types, we do not allow it among \emph{unique} and other table types
-because it would violate our definition of these other table types:
-\[
-\begin{array}{c}
-\mylabel{S-TABLE3}\\
+The rationale for rule {\sc S-TABLE-U-C} derives from the
+fact that reading a non-existing key just yield $\Nil$
+instead of being an error in Lua semantics, plus the fact
+that the type system guarantees that a table with a
+{\em unique} type has no keys with a value that does not
+belong to one of its key types, and does not have any
+aliases. A table with a type $\{1{:}\String, 2{:}\String,
+``x"{:}\Integer, ``y"{:}\Value\}_{unique}$ can be safely recast as a
+table with type 
+$\{\Number{:}\String \cup \Nil, ``x"{:}\Number, ``z"{:}
+\Number \cup \Nil\}_{closed}$ as long as it cannot be
+accessed anymore with the original type.
+
+Subtyping from an {\em open} table type to a {\em closed}
+table type also allows width subtyping, merging fields,
+and adding new fields, but forbids depth subtyping between
+mutable fields or an immutable and a mutable field, as
+the typing rules guarantee that the {\em open} reference
+is destroyed, but the {\em open} table may already have {\em closed} aliases:
+\[
+\begin{array}{c}
+\mylabel{S-TABLE-O-C}\\
+\dfrac{\begin{array}{c}
+	\forall i \in 1..m \; \forall j \in 1..n \;
+	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
+	\forall j \in 1..n \; \not\exists i \in 1..m \;
+	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
+	\end{array}}
+{\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype
+	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
+\end{array}
+\]
+
+The subtyping relation from {\em unique} to {\em unique},
+{\em open} or {\em fixed} table types is still flexible, but forbids standard width subtyping, as all these table types keep the guarantee that the values of all keys in tables of these types belong to one of their key types.
+Merging fields and adding new fields are still allowed:
+\[
+\begin{array}{c}
+\mylabel{S-TABLE-U-UOF}\\
 \dfrac{\begin{array}{c}
@@
 
-The rule that handles subtyping between \emph{open} and \emph{closed} table
-types allows width subtyping:
-\[
-\begin{array}{c}
-\mylabel{S-TABLE4}\\
-\dfrac{\begin{array}{c}
-       \forall i \in 1..m \; \forall j \in 1..n \;
-       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
-       \forall j \in 1..n \; \not\exists i \in 1..m \;
-       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
-       \end{array}}
-      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype
-                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
-\end{array}
-\]
-
-However, the rule that handles subtyping among \emph{open} and
-\emph{open} or \emph{fixed} table types does not allow width subtyping:
-\[
-\begin{array}{c}
-\mylabel{S-TABLE5}\\
+The subtyping relation between {\em open} and {\em open} or
+{\em fixed} table types mixes the depth subtyping restrictions of rule {\sc S-TABLE-O-C} with the width subtyping restrictions of
+rule {\sc S-TABLE-U-UOF}, but still allow merging fields and
+adding new fields:
+\[
+\begin{array}{c}
+\mylabel{S-TABLE-O-OF}\\
 \dfrac{\begin{array}{c}
@@
 
-The rules \textsc{S-TABLE4} and \textsc{S-TABLE5} allow joining fields
-plus omitting optional fields.
-Both rules use $\subtype_{c}$ to allow depth subtyping on $\Const$ fields only.
-
-We introduced \emph{fixed} table types because we needed a safe way
-to represent the type of classes that can allow single inheritance
-through the refinement of table types.
-The rule that handles subtyping between \emph{fixed} table types
-does not allow width subtyping, joining fields, and omitting fields,
-but it allows depth subtyping on $\Const$ fields:
-\[
-\begin{array}{c}
-\mylabel{S-TABLE6}\\
+Finally, the subtyping relation between {\em fixed} table
+types is the more restrictive, and does not allow width
+subtyping, merging fields, or adding fields, just standard
+depth subtyping:
+\[
+\begin{array}{c}
+\mylabel{S-TABLE-F-F}\\
 \dfrac{\begin{array}{c}
@@
 
-In Section \ref{sec:tables} we will show in more detail how our type system
-uses these tags to handle the refinement of table types.
-
-We use the \emph{Amber rule} \cite{cardelli1986amber} to define
-subtyping between recursive types:
+We will revisit the subtyping rules among table types on
+Section~\ref{sec:tables}.
+
+Subtyping between two recursive types uses the \emph{Amber rule}~\cite{cardelli1986amber}:
 \[
@@
 
-The rule \textsc{S-AMBER} also uses the rule \textsc{S-ASSUMPTION}
-to check whether $\mu x_{1}.T_{1} \subtype \mu x_{2}.T_{2}$.
-Both rules use the set of assumptions $\senv$,
-where each assumption is a pair of recursion variables.
-The rule \textsc{S-AMBER} extends $\senv$ with the assumption
-$x_{1} \subtype x_{2}$ to check whether $T_{1} \subtype T_{2}$.
-The rule \textsc{S-ASSUMPTION} allows the rule \textsc{S-AMBER}
-to check whether an assumption is valid.
-
-A recursive type may appear inside a first-level type, and our
-type system includes subtyping rules to handle subtyping between
-recursive types and other first-level types:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-UNFOLDR}\\
+Subtyping between a recursive type and a non-recursive type, or
+vice-versa, uses a pair of rules that unfold the recursive type:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-UNFOLD-R}\\
 \dfrac{\senv \vdash T_{1} \subtype [x \mapsto \mu x.T_{2}]T_{2}}
@@
 \begin{array}{c}
-\mylabel{S-UNFOLDL}\\
+\mylabel{S-UNFOLD-L}\\
 \dfrac{\senv \vdash [x \mapsto \mu x.T_{1}]T_{1} \subtype T_{2}}
@@
 
-As an example, the rule \textsc{S-UNFOLDR} allows our type system to
-type check a function that inserts an element in the begining of
-a singly-linked list:
-\begin{verbatim}
-    local function insert (e:Element?,
-                           v:integer):Element
-      return { info = v, next = e }
-    end
-\end{verbatim}
-that is, the type checker uses the rule \textsc{S-UNFOLDR} to verify whether
-the type of the table constructor is a subtype of \texttt{Element}:
-\begin{align*}
-\{ & ``info":\Integer, ``next":\mu x.\{ ``info":\Integer, \\
-   & ``next":x \;\cup\; \Nil\}_{closed} \cup \Nil \}_{unique} \subtype \\
-& \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}
-\end{align*}
-
-Filter types are subtypes only of themselves and of $\Value$.
-More precisely, a filter type $\phi(T_{1},T_{2})$ is a subtype of
-the same filter type $\phi(T_{1},T_{2})$, which shares the same
-types $T_{1}$ and $T_{2}$, and it is also a subtype of $\Value$.
-
-Projection types are subtypes only of themselves and of $\Value$.
-More precisely, a projection type $\pi_{i}^{x}$ is a subtype of the
-same projection type $\pi_{i}^{x}$, which shares the same union of
-tuples $x$ and the same index $i$, and it is also a subtype of $\Value$.
-
-The dynamic type $\Any$ is neither the bottom nor the top type,
-but a separate type that is subtype only of itself and of $\Value$.
-
-Even though the dynamic type $\Any$ does not interact with subtyping,
-it does interact with consistent-subtyping.
-We present the consistent-subtyping rules as a deduction system for
-the consistent-subtyping relation $\senv \vdash T_{1} \lesssim T_{2}$.
-As in the subtyping relation, $\senv$ is also a set of pairs of
-recursion variables.
-We define the consistent-subtyping rules for the dynamic type $\Any$
-as follows:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{C-ANY1}\\
-\senv \vdash T \lesssim \Any
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{C-ANY2}\\
-\senv \vdash \Any \lesssim T
-\end{array}
-\end{array}
-\]
-
-If we had set the type $\Any$ as both bottom and top types of our
-subtyping relation, then any type $T_{1}$ would be a subtype of
-any other type $T_{2}$.
-The consequence of this is that all programs would type check without errors.
-This would happen due to the transitivity of subtyping, that is,
-we would be able to down-cast any type $T_{1}$ to $\Any$ and then up-cast
-$\Any$ to any other type $T_{2}$.
-The rules \textsc{C-ANY1} and \textsc{C-ANY2} are the rules that
-allow the dynamic type to interact with other first-level types,
-and thus allow dynamically typed code to coexist with statically
-typed code.
-Because of these two rules, consistent-subtyping cannot be transitive.
-These two rules are the only rules that differ between
-subtyping and consistent-subtyping, if we implement the subtyping rules
-as we do in this section.
-
-In the implementation of Typed Lua we also use consistent-subtyping to
-normalize and simplify union types, though we let union types free in
-the formalization.
-For instance, the union type \texttt{boolean|any} results in the
-type \texttt{any}, because \texttt{boolean} is consistent-subtype
-of \texttt{any}.
-Another example is the union type \texttt{number|nil|1} that
-results in the union type \texttt{number|nil}, because
-\texttt{1} is consistent-subtype of \texttt{number}.
+The dynamic type $\Any$ is only a subtype of itself and of
+$\Value$, and no type except $\Any$ is a subtype of $\Any$.
+The type $\Any$ interacts with other types through the
+{\em consistency} and {\em consistent-subtyping} 
+relations~\cite{siek2007objects}. We use the symbol $\sim$ for
+the consistency, and $\lesssim$ for consistent-subtyping.
+
+The type $\Any$ is trivially consistent with all types, and all types
+are consistent with $\Any$. A type is also consistent with
+itself. Two union types $F_1 \cup F_2$ and $F_3 \cup F_4$ are
+consistent if either $F_1 \sim F_3$ and $F_2 \sim F_4$ or
+$F_1 \sim F_4$ and $F_2 \sim F_3$. Two function types
+$S_1 \rightarrow S_2$ and $S_3 \rightarrow S_4$ are consistent
+if and only if $S_1 \rightarrow S_3$ and $S_2 \rightarrow S_4$.
+Two table types are consistent if and only if they have the same key types, and the value types of each key type are consistent.
+Two recursive types are consistent if and only if their contents are consistent assuming that their type variables are consistent.
+Unions of tuple types follow the same rules as unions of first-class types.
+Variadic types are consistent if and only if their base types are consistent.
+Finally, pair types $F_1 \times P_1$ and $F_2 \times P_2$ are
+consistent if and only if $F_1 \sim F_2$ and $P_1 \sim P_2$.
+
+In a language with subtyping, the consistency relation is
+too restrictive. We need to combine it with subtyping into a
+{\em consistent-subtyping} relation. The definition is
+straightforward: a type is a consistent-subtype
+of another if the first type is a subtype of a type
+that is consistent with the other, or if the first type
+is consistent with a type that is a subtype of the other.
+
+In the implementation of Typed Lua, we also use consistent-subtyping to normalize and simplify union types.
+For instance, the union type $\Boolean \cup \Any$ simplifies
+to $\Any$, because $\Boolean$ is a consistent-subtype
+of $\Any$. Another example is the union type
+$\Number \cup \Nil \cup 1$, that simplifies to
+$\Number \cup \Nil$, because $1$ is a consistent-subtype of $\Number$.
 

### 2015-06-11 doc/papers/paper/paper.tex
more changes to section 2 of the DLS paper, reworked the description of the different table types

@@
 
-Table types represent the various forms
-that Lua tables can take.
-The syntactical form of table types is $\{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n} \}_{tag}$,
-where each $K_{i}$ represents the type of a table key,
-and each $V_{i}$ represents the type of the value that table keys of type $K_{i}$ map to.
-Key types can only be literal types, base types, or the top type.
-We made this restriction to the type of the keys because the statistics
-that we discussed in Section \ref{sec:statistics} showed that most
-of the tables are records, lists, and hashes.
-The type $\Value$ is an option when we need a loose table type.
-For instance, $\{\Value:\Value\}_{closed}$ represents the type of a
-table in which both indices and values can have any type.
-Value types can be any first-level type, and can optionally include
-the $\Const$ type to denote immutable values.
-
-We also use the tags \emph{unique}, \emph{open}, \emph{fixed}, and \emph{closed}
-to classify table types.
-The tag \emph{unique} represents tables with no keys that do not
-inhabit one of the table's key types, and with no alias.
-In particular, the type of the table constructor has this tag.
-The tag \emph{open} represents \emph{unique} table types that
-have at least one alias.
-The tag \emph{fixed} represents \emph{unique} table types that we
-do not know how many aliases they have.
-In particular, the type of a class has this tag.
-The tag \emph{closed} represents table types that do not provide
-any guarantees about keys with types not listed in the table type.
-In particular, in the concrete syntax, type annotations, interface
-declarations, and userdata declarations always describe \emph{closed} table types.
-In the next sections we explain in more detail why we need
-different table types.
+Table types represent some of the forms that Lua
+tables can take. What a particular table type denotes
+depends on its {\em tag}: the {\em unique}, {\em open}
+and {\em fixed} tags all denote tables where it is
+guaranteed that every key belongs to one of the {\em key
+types} $K_i$. We restrict key types to literal types,
+base types, and the top type to make the system simpler
+for the user, as our survey of Lua programs showed that
+this restriction reflects the way Lua tables are used.
+Fields are assignable by default, but can be marked
+as non-assignable with the {\bf const} qualifier. Having
+a field be {\bf const} is not a guarantee that its
+value cannot change, as the table may have aliases with
+a non-{\bf const} type for that field.
+
+The difference between {\em unique}, {\em open} and
+{\em fixed} tables is the guarantees that the type system
+give about the reference to a value of that type. A reference
+to an {\em unique} table is guaranteed to have no aliases,
+so allows the greater flexibility in reinterpreting its type;
+a reference to an {\em open} table is guaranteed to have
+only {\em closed} aliases (we will explain the the meaning of the {\em closed} tag shortly), which still lets the type system
+reinterpret the type of a reference, but with more
+restrictions; a {\em fixed} table can have any number of {\em fixed} or {\em closed} aliases, so its type cannot change
+anymore.
+
+Table types with the {\em fixed} tag are suitable for typing tables that are used as maps, lists, or even a mixture of those. The guarantee that there are no keys with a type that is not
+one of the key types allows type-safe iteration on these
+tables.
+
+This guarantee forbids width subtyping, which is a necessity
+for object-oriented programming, so we have the {\em closed}
+tag, which removes the guarantee and allows width subtyping
+between other table types and {\em closed} table types. A {\em closed} table type is akin to an object-oriented {\em interface}.
 
 Any table type has to be \emph{well-formed}.
-Informally, a table type is well-formed if key types do not overlap.
-In Section \ref{sec:tables} we formalize the definition of well-formed table types.
-We delay the proper formalization of well-formed table types because we use
-consistent-subtyping in this formalization.
+Informally, a table type is well-formed if key types do not overlap. A formal definition of this relation depends on
+the consistent-subtyping relation which we present in the
+next section, so we delay the definition until Section~\ref{sec:tables}, where the typing rules for
+table types are explained.
 

### 2015-06-11 doc/papers/paper/paper.tex
rewrite of the beginning of section on typing rules, renaming of typing rules, DLS paper

@@
 \label{sec:rules}
-
-In this section we use a reduced core of Typed Lua to present
-the typing rules that type check the novel features of our type system.
-These rules type check table refinement and
-overloading on the return type of functions.
-
-Our core limits control flow to if and while statements;
-it has explicit type annotations, explicit scope for variables,
-explicit method declarations, and explicit method calls.
-Here is a list of features that are not present in our reduced core:
-\begin{itemize}
-\item labels and goto statements (they are difficult to handle along
-with our simplified form of \emph{flow typing}, and they are out of
-scope for now);
-\item explicit blocks (we are already using explicit scope for variables);
-\item other loop structures such as repeat-until, numeric for,
-and generic for (we can use while to express them);
-\item table fields other than $[e_{1}] = e_{2}$
-(we can use this form to express the missing forms); 
-\item arithmetic operators other than $+$
-(other arithmetic operators have similar typing rules);
-\item relational operators other than $==$ and $<$
-(inequality has similar typing rules to $==$ and
-other relational operators have similar typing rules to $<$);
-\item bitwise operators other than $\&$
-(other bitwise operators have similar typing rules).
-\end{itemize}
-
-Our reduced core does not lose much expressiveness, as it can express
-any Lua program except those that use labels and goto statements.
 
@@
 	\dend
-\caption{The abstract syntax of Typed Lua}
+	\caption{The abstract syntax of Core Typed Lua}
 	\label{fig:syntax}
@@
 
-Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
-It splits the syntactic categories as follows:
-$s$ are statements, $e$ are expressions, $l$ are left-hand values,
-$k$ are literal constants, $el$ are expression lists,
-$me$ are expressions with multiple results, $a$ are function and method applications,
-$f$ are function declarations, $pl$ are parameter lists,
-$id$ are variable names, $T$ are first-level types, and $S$ are second-level types.
-The notation $\overline{id{:}T}$ denotes the list $id_{1}{:}T_{1}, ..., id_{n}{:}T_{n}$.
-
-Our reduced core includes two statements for declaring local variables,
-one with and another without type annotations.
-While we use the former to formalize how our type system handles the declaration
-of annotated variables, we use the latter to formalize how our type system
-handles the declaration of unannotated variables through local type inference
-and also the introduction of projection types.
-
-Our reduced core also includes a truncation operator $\lfloor \rfloor$ for
-function applications, method applications, and the vararg expression.
-We use $\lfloor a \rfloor_{0}$ to denote function and method applications
-that produce no value, because they appear as statements.
-We use $\lfloor me \rfloor_{1}$ to denote function applications,
-method applications, and vararg expressions that produce only one value,
-even if they return multiple values.
-
-We also include two kinds of type coercions in our core language:
-the left-hand value $id[k] \; {<}T{>}$ and the expression ${<}T{>} \;id$.
-Both allow the refinement of table types.
-We also split variable names into two categories to have safe aliasing
-of tables in the presence of refinement.
-We use $id$ when variable names appear as expressions and $id_{l}$ when
-variable names appear as left-hand values.
-
-Even though we can assign only first-level types to variables,
-functions and methods can return unions of second-level types,
-and our type system should be able to project these unions of
-second-level types into unions of first-level types.
-We use two different environments to handle this feature.
-The first environment is the type environment $\env$ that maps
-variables to first-level types.
-We use $\env_{1}[id \mapsto T]$ to extend the environment $\env_{1}$
-with the variable $id$ that maps to type $T$.
-The second environment is the projection environment $\penv$ that
-maps projection variables to second-level types.
-We use $\penv[x \mapsto S]$ to extend the environment $\penv$
-with the projection variable $x$ that maps to type $S$.
-In Section \ref{sec:projections} we will show how our type system uses the
-projection environment $\penv$ for handling projection types,
-and also for projecting unions of second-level types into
-unions of first-level types.
-
-We present the typing rules as a deduction system for two typing relations,
-one for typing statements and another for typing expressions.
-
-We use the relation $\env_{1}, \penv \vdash s, \env_{2}$ for typing statements.
-This relation means that given a type environment $\env_{1}$
-and a projection environment $\penv$, we can check that a statement $s$
-produces a new type environment $\env_{2}$.
-
-We use the relation $\env_{1}, \penv \vdash e : T, \env_{2}$ for typing expressions.
-This relation means that given a type environment $\env_{1}$
-and a projection environment $\penv$, we can check that an expression $e$ has
-type $T$ and produces a new type environment $\env_{2}$.
-
-\subsection{Tables and refinement}
+Figure \ref{fig:syntax} presents the abstract syntax of 
+Core Typed Lua, a large subset of concrete syntax
+that the Typed Lua compiler handles. We use this abstract
+syntax in the presentation of the typing rules
+for tables and evolution of table types (Section~\ref{sec:tables})
+and for functions with overloaded multiple returns and
+projection types (Section~\ref{sec:projections}).
+Core Typed Lua simplifies the presentation of typing rules
+without sacrificing expressiveness.
+
+The main differences between Core Typed Lua and the concrete
+syntax are:
+\begin{itemize}
+\item lack of labels and goto statements, as they are incompatible
+with the {\em flow typing} of Typed Lua's type system;
+\item local variable declarations have an explicit block,
+instead of being just statements;
+\item no {\bf elseif} clause in {\bf if} statements;
+\item no {\bf repeat} and {\bf for} loops, just {\bf while};
+\item the only arithmetic operator is $+$, the only relational
+operators are $==$ and $<$, and the only bitwise operator
+is $\&$, as all other operators have typing rules identical to
+one of these;
+\item type coercion for evolution of table types is explicit,
+with two coercion terms, the l-value $id[k] \; {<}T{>}$ for adding a field to 
+a table through assignment, and the r-value ${<}T{>} \;id$ for reinterpreting
+a table type to a compatible type that is expected by the context;
+\item all function declarations are anonymous, so there is a
+statement for declaring recursive functions;
+\item function or method applications that throw away all
+results or all but the first result are explicitly annotated
+as such, instead this being implicit in their syntactic context;
+\item identifiers that appear in l-value position are explicitly
+annotated as such with an $l$ subscript.
+\end{itemize}
+
+We present the typing rules as a deduction system for two typing relations, one for typing statements and another for typing expressions.
+Typing judgments for statements have the form $\env_{1}, \penv \vdash s, \env_{2}$, meaning that
+given a type environment $\env_{1}$ and a {\em projection environment} $\penv$ we can prove that a statement $s$ produces a new type environment $\env_{2}$. Typing judgements for
+expressions have the form $\env_{1}, \penv \vdash e : T, \env_{2}$, meaning that given a type environment $\env_{1}$
+and a projection environment $\penv$ we can prove that an expression $e$ has type $T$ and produces a new type environment $\env_{2}$.
+
+The type environment is standard, and maps
+identifiers to first-level types. Some typing rules of
+Section~\ref{sec:tables} can change the type of an identifier,
+so the typing relation has both an input and an output type
+environment. The {\em projection environment} $\Pi$ maps projection labels to second-level types, and we explain
+its use in Section~\ref{sec:projections}.
+
+\subsection{Tables and Evolution of Table Types}
 \label{sec:tables}
@@
 \begin{array}{c}
-\mylabel{T-CONSTRUCTOR1}\\
+\mylabel{T-CONSTRUCTOR-EMPTY}\\
 \env_{1}, \penv \vdash \{\}:\{\}_{unique}, \env_{1}
@@
 
-The rule \textsc{T-CONSTRUCTOR2} uses these steps to type check a
+The rule \textsc{T-CONSTRUCTOR-FIELDS} uses these steps to type check a
 table constructor with a non-empty list of table fields:
@@
 \begin{array}{c}
-\mylabel{T-CONSTRUCTOR2}\\
+\mylabel{T-CONSTRUCTOR-FIELDS}\\
 \dfrac{\begin{array}{c}
@@
 $\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ has type
-$\{1:``x", 2:``y", 3:``z"\}_{unique}$ through rule \textsc{T-CONSTRUCTOR2}.
+$\{1:``x", 2:``y", 3:``z"\}_{unique}$ through rule \textsc{T-CONSTRUCTOR-FIELDS}.
 
@@
 has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ through rule
-\textsc{T-CONSTRUCTOR2}.
+\textsc{T-CONSTRUCTOR-FIEDLS}.
 The inner table is \emph{closed} to prevent the creation of unsafe aliases.
@@
 and then to $\{``firstname":\String, ``lastname":\String\}_{unique}$.
-Rule \textsc{T-REFINE} type checks this use of refinement:
-\[
-\begin{array}{c}
-\mylabel{T-REFINE}\\
+Rule \textsc{T-NEWFIELD} type checks this use of refinement:
+\[
+\begin{array}{c}
+\mylabel{T-NEWFIELD}\\
 \dfrac{\begin{array}{c}
@@
 
-The rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this non-deterministic behavior.
-Rule \textsc{T-IDREAD1} uses the auxiliary function \emph{close} to
+The rules \textsc{T-ID-R-CLOSE} and \textsc{T-ID-R-FIX} define this non-deterministic behavior.
+Rule \textsc{T-ID-R-CLOSE} uses the auxiliary function \emph{close} to
 produce a \emph{closed} alias.
@@
 reference \emph{unique} can be unsafe.
-Rule \textsc{T-IDREAD2} uses the auxiliary function \emph{fix} to
+Rule \textsc{T-ID-R-FIX} uses the auxiliary function \emph{fix} to
 produce a \emph{fixed} alias.
@@
 \begin{array}{c}
-\mylabel{T-IDREAD1}\\
+\mylabel{T-ID-R-CLOSE}\\
 \dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
@@
 \begin{array}{c}
-\mylabel{T-IDREAD2}\\
+\mylabel{T-ID-R-FIX}\\
 \dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
@@
 
-Rule \textsc{T-IDWRITE} defines this behavior:
-\[
-\begin{array}{c}
-\mylabel{T-IDWRITE}\\
+Rule \textsc{T-ID-L} defines this behavior:
+\[
+\begin{array}{c}
+\mylabel{T-ID-L}\\
 \dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = write(T_{1})}
@@
 \begin{array}{c}
-\mylabel{T-INDEX1}\\
+\mylabel{T-INDEX-ID}\\
 \dfrac{\begin{array}{c}
@@
 \begin{array}{c}
-\mylabel{T-INDEX2}\\
+\mylabel{T-INDEX-EXP}\\
 \dfrac{\begin{array}{c}
@@
 
-Rule \textsc{T-FUNCTION1} illustrates this case:
-\[
-\begin{array}{c}
-\mylabel{T-FUNCTION1}\\
+Rule \textsc{T-FUNCTION} illustrates this case:
+\[
+\begin{array}{c}
+\mylabel{T-FUNCTION}\\
 \dfrac{\begin{array}{c}

### 2015-06-12 doc/papers/paper/paper.tex
DLS paper: rewrite of the explanation on what happens with closed-over variables, simplification of some rules

@@
 
-We also need to make sure to close all \emph{unique} and \emph{open}
-table types before we type check a nested scope.
-To do that, our type system uses some auxiliary functions to change
-the type of variables before type checking a nested scope and
-also to change the type of assigned and referenced variables after
-type checking a nested scope.
-The function \emph{crall} closes all \emph{unique} and \emph{open}
-table types; it also restores filter types to their original types.
-The function \emph{crset} closes a given set of free assigned variables,
-which is given by the function \emph{fav}, and
-it also uses this set to restore filter types to their original types.
-The function \emph{openset} changes from \emph{unique} to \emph{open}
-a given set of referenced variables, which is given by the function \emph{rv}.
-
-As an example,
+Lua has first-class functions, so another way of aliasing
+an {\em unique} and {\em open} references is by closing
+over them in a function body. In the environment used for
+typing the function body, all {\em unique} and {\em open}
+tables are {\em closed} with the auxiliary function $closeall$.
+This is not enough, as assignment
+to a closed-over variable can be unsafe. In the following
+example, assume that the $g$ receives a parameter of type
+$\{ ``x": \String \cup \Nil\}$:
+\begin{center}
+	\begin{tabular}{llll}
+		\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{open}
+			= \{\} \; \mathbf{in}$}\\
+		& \multicolumn{3}{l}{$\mathbf{local} \; f:\Void \rightarrow \Void =$}\\
+		& & \multicolumn{2}{l}{$\mathbf{fun} \; ():\Void$}\\
+		& & & \multicolumn{1}{l}{$a = \{[``x"]=5\} $}\\
+		& \multicolumn{3}{l}{$\mathbf{in} \; g({<}\{``x":
+			 \String \cup \Nil\}{>} \; a)$}
+	\end{tabular}
+\end{center}
+
+The type of $a$ on line 3 is $\{\}_{closed}$, so the assignment
+would type-check. But then the function call on the last line
+would also type-check, as $\{\}_{open}$ can be coerced to
+$\{ ``x": \String \cup \Nil\}$. A solution is to change the
+type of $a$ from {\em unique} to {\em closed} after the function
+declaration. In order to do that, we use the auxiliary function
+$fav$ ({\em free assigned-to variables}) to collect this
+set of identifiers, and then function $closeset$ changes
+the type environment to so any of these variables that
+has an {\em unique} or {\em open} table type becomes {\em closed}.
+
+Closed-over variables that are not assigned still may
+change their type if they have a {\em unique} table type
+and are aliased inside the function body. We use the auxiliary
+function $frv$ ({\em free referenced variables}) to collect
+this set, and then function $openset$ changes any variables
+with a {\em unique} table type in the environment to {\em open}.
+
+Rule {\sc T-FUNCTION} types non-variadic function declarations,
+and shows how these auxiliary functions work together:
+\[
+\begin{array}{c}
+\mylabel{T-FUNCTION}\\
+\dfrac{\begin{array}{c}
+	closeall(\env_{1})[\overline{id \mapsto T}], \penv[\ret \mapsto S] \vdash s, \env_{2}\\
+	\env_{3} = openset(\env_{1}, frv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)) \\
+	\env_{4} = closeset(\env_{3}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
+	\end{array}}
+{\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{3}}
+\end{array}
+%\begin{array}{c}
+%\mylabel{T-FUNCTION}\\
+%\dfrac{\begin{array}{c}
+%	crall(\env_{1}[\overline{id} \mapsto \overline{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}\\
+%	\env_{3} = crset(\env_{1}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))\\
+%	\env_{4} = openset(\env_{3}, rv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
+%	\end{array}}
+%{\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{4}}
+%\end{array}
+\]
+
+Notice that the environment $\env_{2}$ resulting from
+typing the function body is just ignored.
+
+The following example typechecks, and uses all the parts of {\sc T-FUNCTION}:
 \begin{center}
@@
 & & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
-& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;$}\\
-& & \multicolumn{2}{l}{$b[``x"] \; {<}\String{>} = ``foo" \;;\; f(a[``x"])$}
+& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; f(a[``x"])$}\\
+%& & \multicolumn{2}{l}{$b[``x"] \; {<}\String{>} = ``foo" \;;\;
+%	 f(a[``x"])$}
 \end{tabular}
 \end{center}
-does not type check because we cannot add the field
-$``x"$ to $b$, as its type is closed.
-The assignment $b = a$ type checks because, at that point,
-$a$ and $b$ have the same type: $\{\}_{closed}$.
-Their type was closed by \emph{crall} before type checking
-the function body.
-Their type would be restored to $\{\}_{unique}$ after type checking
-the function body, but that assignment also triggers other two type changes.
-First, the function \emph{fav} includes $b$ in the set of variables
-that should be closed by \emph{crset}.
-Then, the function \emph{rv} includes $a$ in the set of variables
-that should change from \emph{unique} to \emph{open} by \emph{openset}.
-After declaring $f$, $a$ has type $\{\}_{open}$ and $b$ has type $\{\}_{closed}$,
-so we can refine the type of $a$, but we cannot refine the type of $b$.
-
-Rule \textsc{T-FUNCTION} illustrates this case:
-\[
-\begin{array}{c}
-\mylabel{T-FUNCTION}\\
-\dfrac{\begin{array}{c}
-       crall(\env_{1}[\overline{id} \mapsto \overline{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}\\
-       \env_{3} = crset(\env_{1}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))\\
-       \env_{4} = openset(\env_{3}, rv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
-       \end{array}}
-      {\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{4}}
-\end{array}
-\]
-
-This rule also extends the environment $\penv$, bounding the special
-variable $\ret$ to the return type $S$.
-Rule \textsc{T-RETURN} uses the type that is bound to $\ret$ in
-$\penv$ to type check return statements:
-\[
-\begin{array}{c}
-\mylabel{T-RETURN}\\
-\dfrac{\env_{1} \vdash el:S_{1}, \env_{2} \;\;\;
-       \penv(\ret) = S_{2} \;\;\;
-       S_{1} \lesssim S_{2}}
-      {\env_{1} \vdash \mathbf{return} \; el, \env_{2}}
-\end{array}
-\]
+
+The assignment $b=a$ typechecks because, at that point, 
+both $b$ and $a$ have type $\{\}_{closed}$ due to $closeall$.
+The set $fav$ is $\{b\}$, and the set $frv$ is $\{a\}$.
+This causes the type of $b$ to also be $\{\}_{closed}$
+in the body of the inner {\bf local} statement, due to $closeset$, and the type of $a$ to be $\{\}_{open}$ due
+to $openset$. Being $\{\}_{open}$, adding a new field
+is allowed, changing the type of $a$ to $\{``x": \Integer\}_{open}$ after the semicolon. After the call to
+$f$, the variable $b$ is an alias to $a$ but this is not
+an issue, as $\{\}_{closed}$ is a supertype of $\{``x": \Integer\}_{open}$.
 
@@
 	\end{array}}
-{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, \env_{4} - \{\overline{id}\}}
-\end{array}
-\]
-
-This first sequent above the bar uses the auxiliary typing relation $\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$.
+{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, (\env_{4}-\{\overline{id}\})[\overline{id \mapsto \env_2(id)}]}
+\end{array}
+\]
+
+The first judgement above the bar uses the auxiliary typing relation $\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$.
 This rule does the work of generating a fresh projection label,
@@
 \[
+%\begin{array}{c}
+%\mylabel{T-IF-FNIL-PROJ}\\
+%\dfrac{\begin{array}{c}
+%       \env_{1}(id) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
+%       closeall(\env_{1}), \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
+%       closeall(\env_{1}), \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
+%       \env_{4} = closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2})) \\
+%       \env_{5} = openset(\env_{4}, rv(s_{1}) \cup rv(s_{2}))
+%      \end{array}}
+%      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{5}}
+%\end{array}
 \begin{array}{c}
@@
        \env_{1}(id) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
-       closeall(\env_{1}), \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
-       closeall(\env_{1}), \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
-       \env_{4} = closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2})) \\
-       \env_{5} = openset(\env_{4}, rv(s_{1}) \cup rv(s_{2}))
+       \env_{1}, \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
+       \env_{1}, \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
+       \env_{4} = join(\env_{2}, \env_{3}) 
       \end{array}}
-      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{5}}
+      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
 \end{array}
@@
 
-A projection type in l-value position does not trigger
-the projection, effectively forbidding assignment to a
-variable that has a projection type. Allowing assignment
+Assignment to an l-value with a projection type is forbidden. Allowing assignment
 would be unsafe, as it could break the dependency among the

### 2015-06-12 doc/papers/paper/paper.tex
DLS paper: simplify rule T-CONSTRUCTOR-FIELDS
DLS: put examples in introduction to make motivation clearer

@@
 that Typed Lua uses to type some tricky Lua idioms.
-One of these idioms is the use of imperative updates to
-build records and objects one field at a time; Typed Lua
-lets record and object types evolve incrementally along
-with these updates while not sacrificing width subtyping.
+One of these idioms is the use of assignment to
+build records and objects one field at a time, as in
+the following example, where the {\tt newshape} function
+is a constructor for objects with {\tt x} and {\tt y}
+fields and a {\tt move} method:
+\begin{verbatim}
+  function new_shape(x, y)
+    local shape = {}
+    shape.x = x
+    shape.y = y
+    function shape:move(dx, dy)
+      self.x = self.x + dx
+      self.y = self.y + dy
+      return self
+    end
+    return shape
+  end
+\end{verbatim}
+
+The constructor function builds the object piece-by-piece,
+from an initially empty table. Typed Lua
+lets record and object types evolve incrementally 
+via assignment while not sacrificing width subtyping.
+
 Another idiom combines overloading on the return values
-of a function, multiple return values, multiple assignment,
+of a function, multiple return values, multiple assignment
+to decompose multiple return values,
 and tag checks to discriminate a several returned values
-by checking just one of them; Typed Lua deconstructs unions
-of tuples in a way that the dependencies between the types
-of each value in the tuple can be tracked.
+by checking just one of them, as in the following example:
+\begin{verbatim}
+  function idiv(x, y)
+    if y == 0 then
+      return nil, "division by zero"
+    else
+      return x // y, x % y
+    end 
+  end
+
+  local q, r = idiv(a, b)
+  if q then
+    assert(a == q*b + r)
+  else
+    error(r)
+  end
+\end{verbatim}
+
+In the {\tt then} branch it is guaranteed that both {\tt q}
+and {\tt r} are integers, and in the {\bf else} branch it is
+guaranteed that {\tt q} is {\tt nil} and {\tt r} is a string.
+In Typed Lua, the return type of {\tt idiv} can be a union of
+tuples describing the possible return values. Typed Lua deconstructs unions of tuples in a way that the dependencies between the types of each value in the tuple can be tracked.
 
@@
 \dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
-       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
-       wf(T) \;\;\;
-       n = |\;\overline{[e_{1}] = e_{2}}\;| \\
-       \env_{f} = merge(\env_{1}, ..., \env_{n+1})
+       \overline{\env_1, \penv \vdash e_2: V, \env_v}
+        \;\;\;
+       \overline{\env_v, \penv \vdash e_1: K, \env_f}
+        \\
+       T = \{\overline{K{:}V}\}_{unique} \;\;\;
+       \env_{2} = merge(\overline{\env_{f}}) \;\;\; wf(T)
        \end{array}}
-      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{f}}
-\end{array}
+      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{2}}
+\end{array}
+%\dfrac{\begin{array}{c}
+%	\env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
+%	T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
+%	wf(T) \;\;\;
+%	n = |\;\overline{[e_{1}] = e_{2}}\;| \\
+%	\env_{f} = merge(\env_{1}, ..., \env_{n+1})
+%	\end{array}}
+%{\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{f}}
+%\end{array}
 \]
@@
 
-As an example, consider the integer division function
-below, which returns the quotient and rest of integer
-division, or a division by zero error:
+As an example, consider again the integer division function
+from Section~1, which returns the quotient and rest of integer
+division, or a division by zero error, now statically typed:
 \begin{verbatim}
-  local function idiv(x : integer, y : integer)
+  local function idiv(x : integer, y : integer):
+        (integer, integer) | (nil, string)
     if divisor == 0 then
@@
 
-Typed Lua infers the return type of this function as
+The return type of this function is
 $(\Integer \times \Integer \times \Nil*) \sqcup (\Nil \times \String \times \Nil{*})$. The idiom for using this such

### 2015-06-12 doc/papers/paper/paper.tex
DLS: remove void type, still needs to come back and explain a little better how nil* works

@@
 \newcommand{\String}{\mathbf{string}}
-\newcommand{\Void}{\mathbf{void}}
+\newcommand{\Void}{\Nil{*}}
 \newcommand{\Const}{\mathbf{const}}
@@
 P ::= & & \textsc{{\small tuple types:}}\\
-& \;\; \Void & \textit{void type}\\
+%& \;\; \Void & \textit{void type}\\
 & | \; F{*} & \textit{variadic types}\\
@@
 assignments. Tuple types are heterogeneous lists of first-class
-types that end either with $\Void$ or with a {\em variadic type}.
+types that end with a {\em variadic type}.
 A variadic type $F{*}$ represents an unbounded sequence of values of type $F \cup \Nil$.
@@
 Typed Lua preserves this behavior in assignment, but allows
-both functions with strict arity (which use $\Void$) and
-functions with flexible arity (which use variadic types).
+both functions with strict arity, which use $\Void$, and
+functions with flexible arity, which use other variadic types,
+and can use $\Value{*}$ if they want to just ignore extra
+arguments.
 
@@
 that is $\Nil{*}$ if the number of values the list produces
-is statically known. Rule \textsc{S-VARNIL-VOID} makes lets
-an expression list with a statically known number of
-values be used as the arguments or return values of a function
-with a strict arity:
-\[
-\begin{array}{c}
-\mylabel{S-VARNIL-VOID}\\
-\senv \vdash \Nil{*} \subtype \Void
-\end{array}
-\]
+is statically known.
 

### 2015-06-12 doc/papers/paper/paper.tex
DLS paper: change shape to point

@@
 build records and objects one field at a time, as in
-the following example, where the {\tt newshape} function
+the following example, where the {\tt newpoint} function
 is a constructor for objects with {\tt x} and {\tt y}
@@
 \begin{verbatim}
-  function new_shape(x, y)
-    local shape = {}
-    shape.x = x
-    shape.y = y
-    function shape:move(dx, dy)
+  function new_point(x, y)
+    local point = {}
+    point.x = x
+    point.y = y
+    function point:move(dx, dy)
       self.x = self.x + dx
@@
     end
-    return shape
+    return point
   end

### 2015-06-12 doc/papers/paper/paper.tex
DLS paper: finished removal of void type

@@
 types that end with a {\em variadic type}.
-A variadic type $F{*}$ represents an unbounded sequence of values of type $F \cup \Nil$.
+A variadic type $F{*}$ represents an unbounded sequence of values of type $F \cup \Nil$. Lua pads sequences with {\tt nil} to
+have as many values as needed, and in Typed Lua sequences
+where the number of values is statically known have a $\Nil{*}$
+tail to reflect this semantics.
 
@@
 
-Tuple types are covariant:
+Tuple types are covariant, so the subtyping rule for pair types is straightforward:
 \[
@@
 
-Typed Lua preserves this behavior in assignment, but allows
-both functions with strict arity, which use $\Void$, and
-functions with flexible arity, which use other variadic types,
-and can use $\Value{*}$ if they want to just ignore extra
-arguments.
-
-The types of expression lists always end in a variadic type
-that is $\Nil{*}$ if the number of values the list produces
-is statically known.
-
-The other rules let assignment and functions with flexible
-arity reflect the semantics of Lua in a type-safe way. We
-need rule {\sc S-VARIADIC} because just having variadic
-types be covariant would be too restrictive (for example,
-$\Nil{*}$ would not be a subtype of $\Number{*}$):
+Typed Lua preserves this behavior in assignment by synthesizing
+tuple types that have a $\Value{*}$ tail for the left side
+of an assignment. Non-variadic functions have a stricter
+behavior: their parameter lists have a $\Nil{*}$ tail, so
+a function can be passed any number of extra arguments
+as long as they are literal {\tt nil}, but the type system will
+reject extra arguments of any other type. Missing arguments
+are also fine if the corresponding parameters have {\tt nil}
+as a subtype: this is the Lua idiom for optional parameters.
+
+Functions that want to keep the Lua behavior of just ignoring
+any extra arguments can be variadic with type $\Value{*}$ as
+the tail of their parameter tuple.
+
+The remaining subtyping rules for second-level types
+reflect this semantics. Rule {\sc S-VARIADIC} states
+that variadic types are covariant (just having $F_1{*} \subtype F_2{*}$ iff $F_1 \subtype F_2$ would be too restrictive,
+as $\Nil{*}$ would not be a subtype of $\Number{*}$):
 \[
@@
 Rules {\sc S-VAR-EXPAND} and {\sc S-VAR-DROP} state the
-conditions when variadic tuples if different length are
+conditions when variadic tuples of different length are
 compatible:

### 2015-06-12 doc/papers/paper/paper.tex
DLS paper: rewrote explanatin of typing rules for table constructors

@@
 
-Our abstract syntax reduces the syntactic forms of the table constructor
-into two forms: $\{\;\overline{[e_{1}] = e_{2}}\;\}$ and
-$\{\;\overline{[e_{1}] = e_{2}},me\;\}$.
-The first uses a list of table fields $([e_{1}] = e_{2})_{1}, ..., ([e_{1}] = e_{2})_{n}$.
-The second uses a list of table fields and an expression that can
-produce multiple values.
-
-The simplest expression involving tables in the empty table constructor.
-Its type checking rule is straightforward:
-\[
-\begin{array}{c}
-\mylabel{T-CONSTRUCTOR-EMPTY}\\
-\env_{1}, \penv \vdash \{\}:\{\}_{unique}, \env_{1}
-\end{array}
-\]
-
-As a more interesting example, let us see how our type system type checks
-the table constructor $\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$.
-
-First, our type system uses the auxiliary relation
-$\env_{1}, \penv \vdash [e_{1}] = e_{2} : (K,V), \env_{2}$ to type check each
-table field.
-This auxiliary relation means that given a type environment $\env_{1}$
-and a projection environment $\penv$, checking a table field $[e_{1}] = e_{2}$
-produces a pair $(K,V)$ and a new type environment $\env_{2}$.
-A pair $(K,V)$ means that $e_{1}$ has type $K$ and $e_{2}$ has type $V$,
-where $K$ is the type of the key and $V$ is the type of the value.
-
-After type checking each table field, our type system uses each pair $(K,V)$
-to build the table type that express the type of a given constructor, and
-uses the predicate \emph{wf} to check whether this table type is well-formed.
-Formally, a table type is well-formed if it obeys the following rule:
-\[
-\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
-\]
-
-Well-formed table types avoid ambiguity.
-For instance, this rule detects that the table type
-$\{1:\Number, \Integer:\String, \Any:\Boolean\}$ is ambiguous,
-because the type of the value stored by key $1$ can be
-$\Number$, $\String$, or $\Boolean$, as $1 \lesssim 1$,
-$1 \lesssim \Integer$, and $1 \lesssim \Any$.
-Moreover, the type of the value stored by a key of type $\Integer$,
-which is not the literal type $1$, can be $\Number$ or $\Boolean$,
-as $\Integer \lesssim \Integer$, and $\Integer \lesssim \Any$.
-
-Well-formed table types also do not allow \emph{unique} and
-\emph{open} table types to appear in the type of the values.
-We made this restriction because our type system does not keep
-track of aliases to table fields.
-This means that allowing \emph{unique} and \emph{open} table
-types to appear in the type of a value would allow the
-creation of unsafe aliases.
-The rules that type check table fields use the auxiliary function
-\emph{close} to close the type of the values in a table type.
-
-The rule \textsc{T-CONSTRUCTOR-FIELDS} uses these steps to type check a
-table constructor with a non-empty list of table fields:
+The simplest expression involving tables is the empty
+table constructor $\{\}$. The empty table constructor always
+has type $\{\}_{unique}$. A more interesting example is $\{ [``x"] = 1, [``y"] = 2 \}$, which has type $\{ ``x":1 , ``y": 2 \}_{unique}$. The subtyping rule for {\em unique} tables lets
+us assign this table to a variable with a more general
+type such as $\{ ``x":\Number , ``y": \Number \}_{unique}$ or
+even $\{ \String:\Number \cup \Nil \}_{unique}$. We give
+the typing rule for table constructors of this form below: 
 \[
@@
 \dfrac{\begin{array}{c}
-       \overline{\env_1, \penv \vdash e_2: V, \env_v}
+	\overline{\env_1, \penv \vdash e_2: fix(V), \env_v}
 	\;\;\;
@@
 
-Back to our example, the table constructor
-$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ has type
-$\{1:``x", 2:``y", 3:``z"\}_{unique}$ through rule \textsc{T-CONSTRUCTOR-FIELDS}.
-
-As another example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
-has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ through rule
-\textsc{T-CONSTRUCTOR-FIEDLS}.
-The inner table is \emph{closed} to prevent the creation of unsafe aliases.
+Due to the system for changing the table type of variables that
+we will explain in this section, any expression can potentially
+change the type environment. Lua semantics does not dictate
+an order of evaluation among the pairs of field initializers,
+so all of them are typed in the same input type environment.
+The $merge$ auxiliary function then combines their output
+environments, diverging in case of conflicting changes.
+
+The rule also checks if the resulting type is {\em well-formed}
+with the predicate $wf$. Formally, we define this predicate
+for table types with the following formula:
+\[
+\begin{array}{c}
+\forall i \; ((\nexists j \; i \not= j \,\wedge\, K_{i} \lesssim K_{j}) \,\wedge\, wf(V_{i}) \,\wedge\,\\ \lnot isunique(V_{i}) \,\wedge\, \lnot isopen(V_{i}))
+\end{array}
+\]
+
+Well-formed table types avoid ambiguity.
+For instance, this rule detects that the table type
+$\{1:\Number, \Integer:\String, \Any:\Boolean\}$ is ambiguous,
+because indexing it with a key of type $\Integer$ could
+yield either a number, a string or a boolean at runtime.
+
+Well-formed table types also do not allow \emph{unique} and
+\emph{open} table types to appear in the type of the values,
+and they must be well-formed. Our type system currently cannot
+track aliases of table fields, so this restriction is
+necessary.
+
+This restriction is the reason rule {\sc T-CONSTRUCTOR-FIELDS}
+uses the auxiliary function $fix$ to change any {\em open}
+and {\em unique} table types used in the field initializer
+to {\em fixed}. As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
+has type $\{``x":1, ``y":\{``z":2\}_{fixed}\}_{unique}$.
 

### 2015-06-12 doc/papers/paper/paper.tex
DLS paper: some fixes in intro

@@
 \begin{abstract}
-Dynamically typed languages such as Lua avoid static types in order
-to have a simpler and more flexible language, because their programmers
-to not need to bother with abstracting types that need to be validated
-by a type checker. In contrast, statically typed languages provide
-earlier detection of many defects, and a better framework for structuring
-large programs. The advantages of static types often lead programmers to
-migrate from a dynamically typed to a statically typed language when
-their simple scripts evolve into complex programs.
-
-Optional type systems are one way to have both static and dynamic
-typing in the same language while keeping its original semantics,
-making evolving a program from dynamic to static typing a matter of
-describing the implied types that it it using and adding annotations
-to make those types explicit. Designing an optional type system for
-an existing dynamically typed language is challenging, as its types
-should feel natural to programmers that are already familiar with
+Programmers often migrate from a dynamically typed
+to a statically typed language when their simple scripts
+evolve into complex programs. Optional type systems are
+one way of having both static and dynamic typing
+in the same language, while keeping its dynamically typed
+semantics. This makes evolving a program from dynamic to
+static typing a matter of describing the implied types that
+it it using and adding annotations to make those types explicit.
+Designing an optional type system for
+an existing dynamically typed language is challenging, as its types should feel natural to programmers that are already familiar with
 this language.
@@
 In this work, we give a formal description of Typed Lua, an
-optional type system for Lua, with a focus on its novel type system
-features: incremental evolution of record and object types via
-imperative update that does not sacrifice width subtyping,
-and combining flow typing with multiple assignment and with functions that
-return multiple values. While our type system is tailored to the
-features and idioms of Lua, its features can be adapted to other
-imperative scripting languages.
+optional type system for Lua, with a focus on two of its novel type system features: incremental evolution of record and object types via imperative update that does not sacrifice width subtyping, and combining flow typing with multiple assignment and with functions that return multiple values.
+While our type system is tailored to the features and idioms of Lua, its features can be adapted to other imperative scripting languages.
 \end{abstract}
@@
 \emph{Optional type systems}~\cite{bracha2004pluggable} and
-\emph{gradual typing}~\cite{siek2006gradual} are two similar
+\emph{gradual typing}~\cite{siek2006gradual} are two 
 approaches for blending static and dynamic typing in the same
@@
 safety, while gradual typing changes the semantics to
-include more extensive runtime checking to ensure that
+include more extensive runtime checking. These extra checks
+ensure that
 dynamically typed portions of the program do not violate
@@
 Lua is a small imperative language with lexically-scoped first-class functions where the only data structuring
-mechanism is the \emph{table} --
-an associative array that has syntactic and runtime support
-to efficiently represent arrays, records, maps, modules,
-objects, etc. Unlike other scripting languages, Lua has very limited coercion among different data types.
-
-Lua prefers to provide mechanisms instead of fixed policies due
+mechanism is the \emph{table}, an associative array. Lua prefers to provide mechanisms instead of fixed policies due
 to its primary use as an embedded language for configuration and
@@
 This means that even features such as a module system and
-object orientation are a matter of convention instead of
-built-in language constructs.
+object orientation are a matter of convention on how
+programs use the syntactic sugar and metaprogramming
+facilities of tables, instead of
+built-in language constructs. Tables can efficiently represent arrays, records, maps, modules,
+objects, etc.
+
 The result is a fragmented ecosystem of libraries, and different
 ideas among Lua programmers on how they should use the language
-features or on how they should structure programs.
-
-Typed Lua is an optional type system for
+features or on how they should structure programs. One example
+of this fragmentation is the proliferation of object systems.
+
+Typed Lua~\cite{maidl2014tl} is an optional type system for
 Lua that is rich enough to preserve some of the idioms
@@
 helping programmers structure large Lua programs and
-catch errors at compile-time. An informal
-presentation of the design of Typed Lua has been
-published~\cite{maidl2014tl}.
-
-In this paper, we give a formal basis to an updated version of that design, with a focus on novel type system features
-that Typed Lua uses to type some tricky Lua idioms.
-One of these idioms is the use of assignment to
-build records and objects one field at a time, as in
-the following example, where the {\tt newpoint} function
+catch errors at compile-time.
+
+In this paper, we give a formal basis to an updated version of Typed Lua's original design, with a focus on two novel type system
+features that Typed Lua uses to type some tricky Lua idioms.
+The lack of standard policies is a challenge for the design of
+an optional type system for Lua. For this reason, the design
+of Typed Lua has been informed by a mostly automated survey
+of Lua idioms used by a large corpus of Lua libraries.
+This paper also presents the methodology of this survey and
+a summary of its results.
+
+The first idioms that we focus on this paper is the use of assignment to build records and objects one field at a time, as in
+the Lua fragment in Figure~\ref{fig:newpoint}, where the {\tt newpoint} function
 is a constructor for objects with {\tt x} and {\tt y}
-fields and a {\tt move} method:
+fields and a {\tt move} method.
+
+\begin{figure}[t]
+\label{fig:newpoint}
 \begin{verbatim}
@@
 \end{verbatim}
-
-The constructor function builds the object piece-by-piece,
+\caption{Building an object piece-by-piece}
+\end{figure}
+
+The {\tt newpoint} function builds the object piece-by-piece,
 from an initially empty table. Typed Lua
@@
 
-Another idiom combines overloading on the return values
-of a function, multiple return values, multiple assignment
+The other idiom that we focus on combines overloading on the return values of a function, multiple return values, multiple assignment
 to decompose multiple return values,
-and tag checks to discriminate a several returned values
-by checking just one of them, as in the following example:
+and tag checks to discriminate several returned values
+by checking just one of them. The Lua fragment in Figure~\ref{fig:over} shows a function with overloaded multiple
+returns, and how a caller of this function can discriminate
+which list of multiple values the function returned.
+
+\begin{figure}[t]
+\label{fig:over}
 \begin{verbatim}
@@
 \end{verbatim}
+\caption{Overloading on multiple returns}
+\end{figure}
 
@@
 
-The lack of standard policies is a challenge for the design of
-an optional type system for Lua. For this reason, the design
-of Typed Lua has been informed by a mostly automated survey
-of Lua idioms used by a large corpus of Lua libraries.
-This paper also presents the methodology of this survey and
-a summary of its results.
-
-The rest of the paper is organized as follows:
+The remainder of the paper is organized as follows:
 Section \ref{sec:statistics} presents the methodology and results of
@@
 their subtyping and consistent-subtyping relations;
-Section \ref{sec:rules} presents the typing rules for evolution of table types,
-and the typing rules for destructuring unions of tuples into projection types,
-allowing flow typing on these projections;
+Section \ref{sec:rules} presents the typing rules
+for the two idioms that we focus on this paper;
 Section \ref{sec:related} reviews related work;
 Section \ref{sec:conclusion} presents our conclusions,
-and outlines our plans for future work.
+and outlines plans for future work.
 
@@
 our corpus, 36\% of them create a record (a table with string
-keys such as \texttt{\{ x = 120, y = 121 \}}), 29\% of them create
-a list or a tuple (a table with consecutive integer keys such as \texttt{\{ "one", "two", "three", "four" \}}), 26\% of them create an empty table, and 8\% of them create a table with
+keys such as \texttt{\{x = 1, y = 2\}}), 29\% of them create
+a list or a tuple (a table with consecutive integer keys such as \texttt{\{"one", "two", "three", "four"\}}), 26\% of them create an empty table, and 8\% of them create a table with
 both a record part and a list part. The remaining 1\% of them create a table with non-literal keys.
@@
 \begin{figure*}[!ht]
-\textbf{Type Language}\\
-\dstart
 $$
@@
 & | \; F{*} & \textit{variadic types}\\
-& | \; F \times P & \textit{pair types}
+& | \; T \times P & \textit{pair types}
 \end{array}
 $$
-\dend
-\caption{The abstract syntax of Typed Lua types}
+\caption{Type Language}
 \label{fig:typelang}
@@
 for the user, as our survey of Lua programs showed that
-this restriction reflects the way Lua tables are used.
+this restriction reflects the way Lua tables are commonly used.
+
 Fields are assignable by default, but can be marked
@@
 value cannot change, as the table may have aliases with
-a non-{\bf const} type for that field.
+a non-{\bf const} type for that field. Non-assignable
+fields enable {\em depth subtyping} between table types.
 
@@
 {\em fixed} tables is the guarantees that the type system
-give about the reference to a value of that type. A reference
+gives about the reference to a value of that type. A reference
 to an {\em unique} table is guaranteed to have no aliases,
-so allows the greater flexibility in reinterpreting its type;
+so it allows greater flexibility in reinterpreting its type;
 a reference to an {\em open} table is guaranteed to have
-only {\em closed} aliases (we will explain the the meaning of the {\em closed} tag shortly), which still lets the type system
+only {\em closed} aliases (we will explain the meaning of the {\em closed} tag shortly), which still lets the type system
 reinterpret the type of a reference, but with more
@@
 
-We present the subtyping rules as a deduction system for the
+The subtyping rules are a deduction system for the
 subtyping relation $\senv \vdash t_{1} \subtype t_{2}$.
@@
 by the subtyping relation between recursive types. 
-
-Most of the subtyping rules are standard; we will focus
-on the interesting cases, involving second-level types and
-table types.
-
+Most of the subtyping rules are standard, so we will
+focus on the interesting cases.
+
+\begin{figure}
+\label{fig:subtuples}
+\[
+\begin{array}{c}
+
+\]
+\caption{Subtyping for second-level types}	
+\end{figure}
+
+Figure~\ref{fig:subtuples} shows the subtyping rules for
+second-level types. 
 Tuple types are covariant, so the subtyping rule for pair types is straightforward:
-\[
-\begin{array}{c}
-\mylabel{S-PAIR}\\
-\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
-       \senv \vdash P_{1} \subtype P_{2}}
-      {\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
-\end{array}
-\]
+
+%\[
+%\begin{array}{c}
+%\mylabel{S-PAIR}\\
+%\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
+%       \senv \vdash P_{1} \subtype P_{2}}
+%      {\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
+%\end{array}
+%\]
 

### 2015-06-12 doc/papers/paper/paper.tex
DLS paper: reformatting

@@
 
-In the {\tt then} branch it is guaranteed that both {\tt q}
+In the {\bf then} branch it is guaranteed that both {\tt q}
 and {\tt r} are integers, and in the {\bf else} branch it is
@@
 
-\begin{figure}
+\begin{figure}[t]
 \label{fig:subtuples}
@@
 \begin{array}{c}
-
+\begin{array}{c}
+\mylabel{S-Pair}\\
+\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
+	\senv \vdash P_{1} \subtype P_{2}}
+{\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
+\end{array}
+\begin{array}{c}
+\mylabel{S-Variadic}\\
+\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \cup \Nil}
+{\senv \vdash F_{1}{*} \subtype F_{2}{*}}
+\end{array}\\\\
+\begin{array}{c}
+\mylabel{S-VarExpand}\\
+\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \;\;\;
+	\senv \vdash F_{1}{*} \subtype P_{2}}
+{\senv \vdash F_{1}{*} \subtype F_{2} \times P_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-VarDrop}\\
+\dfrac{\senv \vdash F_{1} \subtype F_{2} \cup \Nil \;\;\;
+	\senv \vdash P_{1} \subtype F_{2}{*}}
+{\senv \vdash F_{1} \times P_{1} \subtype F_{2}{*}}
+\end{array}
+\end{array}
 \]
@@
 second-level types. 
-Tuple types are covariant, so the subtyping rule for pair types is straightforward:
-
-%\[
-%\begin{array}{c}
-%\mylabel{S-PAIR}\\
-%\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
-%       \senv \vdash P_{1} \subtype P_{2}}
-%      {\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
-%\end{array}
-%\]
+Tuple types are covariant, so the subtyping rule {\sc S-Pair} for pair types  is straightforward.
 
@@
 The remaining subtyping rules for second-level types
-reflect this semantics. Rule {\sc S-VARIADIC} states
+reflect this semantics. Rule {\sc S-Variadic} states
 that variadic types are covariant (just having $F_1{*} \subtype F_2{*}$ iff $F_1 \subtype F_2$ would be too restrictive,
-as $\Nil{*}$ would not be a subtype of $\Number{*}$):
-\[
-\begin{array}{c}
-\mylabel{S-VARIADIC}\\
-\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \cup \Nil}
-      {\senv \vdash F_{1}{*} \subtype F_{2}{*}}
-\end{array}
-\]
-
-Rules {\sc S-VAR-EXPAND} and {\sc S-VAR-DROP} state the
+as $\Nil{*}$ would not be a subtype of $\Number{*}$).
+Rules {\sc S-VarExpand} and {\sc S-VarDrop} state the
 conditions when variadic tuples of different length are
-compatible:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-VAR-EXPAND}\\
-\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \;\;\;
-       \senv \vdash F_{1}{*} \subtype P_{2}}
-      {\senv \vdash F_{1}{*} \subtype F_{2} \times P_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-VAR-DROP}\\
-\dfrac{\senv \vdash F_{1} \subtype F_{2} \cup \Nil \;\;\;
-       \senv \vdash P_{1} \subtype F_{2}{*}}
-      {\senv \vdash F_{1} \times P_{1} \subtype F_{2}{*}}
-\end{array}
-\end{array}
-\]
+compatible.
 
@@
 
-We will now focus on the subtyping relation among table
-types. Subtyping between a {\em fixed} or {\em closed}
-table type and a {\em closed} table type resembles the
-standard subtyping rule between mutable records, with
-width subtyping plus depth subtyping from a mutable
-or immutable field to an immutable field:
+\begin{figure}[t]
+\label{fig:subtable}
 \[
 \begin{array}{c}
-\mylabel{S-TABLE-FC-C}\\
+\begin{array}{c}
+\mylabel{S-TableFCtoC}\\
 \dfrac{\begin{array}{c}
@@
 	\end{array}} \; m \ge n
-\end{array}
-\]
-
-Rule \textsc{S-TABLE-FC-C} uses the auxiliary relation $\subtype_{c}$ for handling depth subtyping:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-FIELD-C-MM}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2} \;\;\;
-       \senv \vdash V_{2} \subtype V_{1}}
-      {\senv \vdash V_{1} \subtype_{c} V_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FIELD-C-II}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2}}
-      {\senv \vdash \Const \; V_{1} \subtype_{c} \Const \; V_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-FIELD-C-MI}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2}}
-      {\senv \vdash V_{1} \subtype_{c} \Const \; V_{2}}
-\end{array}
-\end{array}
-\]
-
-The subtyping relation between {\em unique} and {\em closed}
-table types is more relaxed, as the typing rules of Section~\ref{sec:tables} guarantee that a {\em unique}
-table type can only appear on the left side of a subtyping
-judgment if the reference to the table with the {\em unique}
-tag is being destroyed:
-\[
-\begin{array}{c}
-\mylabel{S-TABLE-U-C}\\
+\end{array}\\\\
+\begin{array}{c}
+\mylabel{S-TableUtoC}\\
 \dfrac{\begin{array}{c}
@@
 	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
-\end{array}
-\]
-
-Rule {\sc S-TABLE-U-C} allows with subtyping, depth subtyping
-on any field, not only immutable fields, as well as the addition
-of new fields and the merging of existing fields, as long as the types of the values of new and merged fields include $\Nil$.
-This rule uses the auxiliary relations $\subtype_{u}$ and $\subtype_{o}$, which respectively allow depth subtyping on all fields, and the addition of new fields:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-FIELD-U-MM}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2}}
-      {\senv \vdash V_{1} \subtype_{u} V_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FIELD-U-IM}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2}}
-{\senv \vdash \Const \; V_{1} \subtype_{u} V_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-FIELD-U-II}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2}}
-      {\senv \vdash \Const \; V_{1} \subtype_{u} \Const \; V_{2}}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FIELD-U-MI}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2}}
-      {\senv \vdash V_{1} \subtype_{u} \Const \; V_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-FIELD-O-M}\\
-\dfrac{\senv \vdash \Nil \subtype V}
-      {\senv \vdash \Nil \subtype_{o} V}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-FIELD-O-I}\\
-\dfrac{\senv \vdash \Nil \subtype V}
-      {\senv \vdash \Nil \subtype_{o} \Const \; V}
-\end{array}
-\end{array}
-\]
-
-The rationale for rule {\sc S-TABLE-U-C} derives from the
-fact that reading a non-existing key just yield $\Nil$
-instead of being an error in Lua semantics, plus the fact
-that the type system guarantees that a table with a
-{\em unique} type has no keys with a value that does not
-belong to one of its key types, and does not have any
-aliases. A table with a type $\{1{:}\String, 2{:}\String,
-``x"{:}\Integer, ``y"{:}\Value\}_{unique}$ can be safely recast as a
-table with type 
-$\{\Number{:}\String \cup \Nil, ``x"{:}\Number, ``z"{:}
-\Number \cup \Nil\}_{closed}$ as long as it cannot be
-accessed anymore with the original type.
-
-Subtyping from an {\em open} table type to a {\em closed}
-table type also allows width subtyping, merging fields,
-and adding new fields, but forbids depth subtyping between
-mutable fields or an immutable and a mutable field, as
-the typing rules guarantee that the {\em open} reference
-is destroyed, but the {\em open} table may already have {\em closed} aliases:
-\[
-\begin{array}{c}
-\mylabel{S-TABLE-O-C}\\
+\end{array}\\\\
+\begin{array}{c}
+\mylabel{S-TableOtoC}\\
 \dfrac{\begin{array}{c}
@@
 	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
-\end{array}
-\]
-
-The subtyping relation from {\em unique} to {\em unique},
-{\em open} or {\em fixed} table types is still flexible, but forbids standard width subtyping, as all these table types keep the guarantee that the values of all keys in tables of these types belong to one of their key types.
-Merging fields and adding new fields are still allowed:
-\[
-\begin{array}{c}
-\mylabel{S-TABLE-U-UOF}\\
+\end{array}\\\\
+\begin{array}{c}
+\mylabel{S-TableUtoUOF}\\
 \dfrac{\begin{array}{c}
@@
 	\end{array}}
-\end{array}
-\]
-
-The subtyping relation between {\em open} and {\em open} or
-{\em fixed} table types mixes the depth subtyping restrictions of rule {\sc S-TABLE-O-C} with the width subtyping restrictions of
-rule {\sc S-TABLE-U-UOF}, but still allow merging fields and
-adding new fields:
-\[
-\begin{array}{c}
-\mylabel{S-TABLE-O-OF}\\
+\end{array}\\\\
+\begin{array}{c}
+\mylabel{S-TableOtoOF}\\
 \dfrac{\begin{array}{c}
@@
 	\end{array}}
-\end{array}
-\]
-
-Finally, the subtyping relation between {\em fixed} table
-types is the more restrictive, and does not allow width
-subtyping, merging fields, or adding fields, just standard
-depth subtyping:
-\[
-\begin{array}{c}
-\mylabel{S-TABLE-F-F}\\
+\end{array}\\\\
+\begin{array}{c}
+\mylabel{S-TableFtoF}\\
 \dfrac{\begin{array}{c}
@@
 \end{array}
+\end{array}
 \]
+\caption{Subtyping for table types}
+\end{figure}
+
+\begin{figure}[t]
+\label{fig:subfield}
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{S-FieldClsMtoM}\\
+\dfrac{\senv \vdash V_{1} \subtype V_{2} \;\;\;
+	\senv \vdash V_{2} \subtype V_{1}}
+{\senv \vdash V_{1} \subtype_{c} V_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FieldClsItoI}\\
+\dfrac{\senv \vdash V_{1} \subtype V_{2}}
+{\senv \vdash \Const \; V_{1} \subtype_{c} \Const \; V_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-FieldClsMtoI}\\
+\dfrac{\senv \vdash V_{1} \subtype V_{2}}
+{\senv \vdash V_{1} \subtype_{c} \Const \; V_{2}}
+\end{array}\\\\
+\begin{array}{c}
+\mylabel{S-FieldUnqMtoM}\\
+\dfrac{\senv \vdash V_{1} \subtype V_{2}}
+{\senv \vdash V_{1} \subtype_{u} V_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FieldUnqItoM}\\
+\dfrac{\senv \vdash V_{1} \subtype V_{2}}
+{\senv \vdash \Const \; V_{1} \subtype_{u} V_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-FieldUnqItoI}\\
+\dfrac{\senv \vdash V_{1} \subtype V_{2}}
+{\senv \vdash \Const \; V_{1} \subtype_{u} \Const \; V_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FieldUnqMtoI}\\
+\dfrac{\senv \vdash V_{1} \subtype V_{2}}
+{\senv \vdash V_{1} \subtype_{u} \Const \; V_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-FieldOpnNtoM}\\
+\dfrac{\senv \vdash \Nil \subtype V}
+{\senv \vdash \Nil \subtype_{o} V}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{S-FieldOpnNtoI}\\
+\dfrac{\senv \vdash \Nil \subtype V}
+{\senv \vdash \Nil \subtype_{o} \Const \; V}
+\end{array}
+\end{array}
+\]
+\caption{Subtyping for fields}
+\end{figure}
+
+Figure~\ref{fig:subtable} shows the subtyping rules for
+table types. These rules use the auxiliary relations
+$\subtype_{c}$, $\subtype_{u}$, and $\subtype_{o}$
+shown in Figure~\ref{fig:subfield} to handle depth
+subtyping where applicable.
+
+Rule {\sc S-TableFCtoC} for subtyping
+between a {\em fixed} or {\em closed}
+table type and a {\em closed} table type resembles the
+standard subtyping rule between mutable records, with
+width subtyping plus depth subtyping from a mutable
+or immutable field to an immutable field.
+
+Rule {\sc S-TableUtoC} for subtyping between {\em unique} and {\em closed} table types is more permissive, as the typing rules of Section~\ref{sec:tables} guarantee that a {\em unique}
+table type can only appear on the left side of a subtyping
+judgment if the reference to the table with the {\em unique}
+tag is being destroyed.
+This rule allows with subtyping, depth subtyping on any field, not only immutable fields, as well as the addition of new fields and the merging of existing fields, as long as the types of the values of new and merged fields include $\Nil$.
+
+The rationale for the permissiveness of rule {\sc S-TableUtoC} derives from the fact that reading a non-existing key just yields $\Nil$ instead of being an error in Lua semantics, plus the fact
+that the type system guarantees that a table with a
+{\em unique} type has no keys with a value that does not
+belong to one of its key types, and does not have any
+aliases. With these guarantess, a table with type $\{1{:}\String, 2{:}\String,
+``x"{:}\Integer, ``y"{:}\Value\}_{unique}$ can be safely recast as a table with type 
+$\{\Number{:}\String \cup \Nil, ``x"{:}\Number, ``z"{:}
+\Number \cup \Nil\}_{closed}$ as long as it becomes
+inaccessible with the original type, which Section~\ref{sec:tables} will show that is the case
+for any type judgement where this subtyping rule is used.
+
+Rule {\sc S-TableOtoC} for subtyping from an {\em open} table type to a {\em closed} table type also allows width subtyping, merging fields, and adding new fields, but forbids depth subtyping between mutable fields or an immutable and a mutable field, as the typing rules guarantee that the {\em open} reference
+is destroyed, but the {\em open} table may already have {\em closed} aliases.
+
+Rule {\sc S-TableUtoUOF} for subtyping from {\em unique} to {\em unique}, {\em open} or {\em fixed} table types is still permissive, but forbids standard width subtyping, as all these table types keep the guarantee that the values of all keys in tables of these types belong to one of their key types.
+Merging fields and adding new fields are still allowed.
+
+Rule {\sc S-TableOtoOF} for subtyping from {\em open} to {\em open} or {\em fixed} table types mixes the depth subtyping restrictions of rule {\sc S-TableOtoC} with the width subtyping restrictions of rule {\sc S-TableUtoUOF}, but still allows merging fields and adding new fields.
+
+Finally, rule {\sc S-TableFtoF} for subtyping between {\em fixed} table types is the most restrictive, and does not allow width
+subtyping, merging fields, or adding fields, just standard
+depth subtyping.
 
@@
 
-Subtyping between two recursive types uses the \emph{Amber rule}~\cite{cardelli1986amber}:
+\begin{figure}[t]
+\label{fig:subrec}
 \[
@@
 \begin{array}{c}
-\mylabel{S-AMBER}\\
+\mylabel{S-Amber}\\
 \dfrac{\senv[x_{1} \subtype x_{2}] \vdash T_{1} \subtype T_{2}}
@@
 \begin{array}{c}
-\mylabel{S-ASSUMPTION}\\
+\mylabel{S-Assumption}\\
 \dfrac{x_{1} \subtype x_{2} \in \senv}
 {\senv \vdash x_{1} \subtype x_{2}}
-\end{array}
-\end{array}
-\]
-
-Subtyping between a recursive type and a non-recursive type, or
-vice-versa, uses a pair of rules that unfold the recursive type:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{S-UNFOLD-R}\\
+\end{array}\\\\
+\begin{array}{c}
+\mylabel{S-UnfoldRight}\\
 \dfrac{\senv \vdash T_{1} \subtype [x \mapsto \mu x.T_{2}]T_{2}}
@@
 \begin{array}{c}
-\mylabel{S-UNFOLD-L}\\
+\mylabel{S-UnfoldLeft}\\
 \dfrac{\senv \vdash [x \mapsto \mu x.T_{1}]T_{1} \subtype T_{2}}
@@
 \]
+\caption{Subtyping for recursive types}
+\end{figure}
+
+Figure~\ref{fig:subrec} shows the subtyping rules for
+recursive types. Rule {\sc S-Amber} is the \emph{Amber rule}~\cite{cardelli1986amber} for subtyping between
+two recursive types. Rules {\sc S-UnfoldRight} and {\sc S-UnfoldLeft} unfold the recursive type for establishing
+subtyping between a non-recursive and a recursive type,
+or vice-versa.
 
@@
 \begin{figure*}[!ht]
-	\textbf{Abstract Syntax}\\
-	\dstart
 	$$
@@
 	$$
-	\dend
-	\caption{The abstract syntax of Core Typed Lua}
+	\caption{Abstract Syntax}
 	\label{fig:syntax}
@@
 
-As an example, consider again the integer division function
-from Section~1, which returns the quotient and rest of integer
-division, or a division by zero error, now statically typed:
-\begin{verbatim}
-  local function idiv(x : integer, y : integer):
-        (integer, integer) | (nil, string)
-    if divisor == 0 then
-      return nil, "division by zero"
-    else
-      return x // y, x % y
-    end 
-  end
-\end{verbatim}
-
-The return type of this function is
-$(\Integer \times \Integer \times \Nil*) \sqcup (\Nil \times \String \times \Nil{*})$. The idiom for using this such
-a function is to test the first returned value:
-\begin{verbatim}
-  local q, r = idiv(a, b)
-  if q then
-    -- assume q and r are integers
-  else 
-    -- assume q is nil r is a string
-  end
-\end{verbatim}
+As an example, consider again the {\tt idiv} function
+from Figure~\ref{fig:over}, which returns the quotient and rest of integer division, or a division by zero error.
+In Typed Lua, we can give a return type $(\Integer \times \Integer \times \Nil*) \sqcup (\Nil \times \String \times \Nil{*})$ to that function.
+
+As is also shown in Figure~\ref{fig:over}, the idiom for using such
+a function is to test the first returned value. We
+can translate that code to the abstract syntax of
+Figure~\ref{fig:syntax}, leaving placeholders for
+the body of the {\bf then} and {\bf else} branches:
+\begin{center}
+	\begin{tabular}{ll}
+		\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(a, b) \, \mathbf{in} \, \mathbf{if} \, q \, \mathbf{then} \, s_1 \, \mathbf{else} \, s_2$}
+	\end{tabular}
+\end{center}
 
@@
 
-Before introducing the typing rules for projections, let us translate the example above into the abstract syntax of Figure~\ref{fig:syntax}:
-\begin{center}
-	\begin{tabular}{ll}
-		\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(1, 2) \, \mathbf{in} \, \mathbf{if} \, q \, \mathbf{then} \, s_1 \, \mathbf{else} \, s_2$}
-	\end{tabular}
-\end{center}
-
-Notice that the local variables $q$ and $r$ are declared without 
+Notice that in the above example the local variables $q$ and $r$ are declared without 
 an explicit type. The rule for typing this {\bf local} statement,
-{\sc T-LOCAL-UN}, introduces a new label in the projection
+{\sc T-LocalProj}, introduces a new label in the projection
 environment used for typing the {\bf if} statement:
@@
 \begin{array}{c}
-\mylabel{T-LOCAL-UN}\\
+\mylabel{T-LocalProj}\\
 \dfrac{\begin{array}{c}
@@
 the expression list has a union of tuple types as its type,
-as rule {\sc T-EXPLIST-PROJ} below shows:
+as rule {\sc T-ExpListProj} below shows:
 \[
 \begin{array}{c}
-\mylabel{T-EXPLIST-PROJ}\\
+\mylabel{T-ExpListProj}\\
 \dfrac{\begin{array}{c}
@@
 
-In our example, the two rules combined will bind $x$ to
+In the above example, the combination of these two rules will bind $x$ to
 $(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})$ in the projection environment $\Pi$,
@@
 their types into either $\Integer \cup \Nil$ or $\Integer \cup \String$. But the more interesting case is inside the
-branches of the {\bf if} statement. The particular uses the rule {\sc T-IF-FNIL-PROJ} below, which uses flow typing:
+branches of the {\bf if} statement. The particular uses the rule {\sc T-IfFilterNilProj} below, which uses flow typing:
 \[
 %\begin{array}{c}
-%\mylabel{T-IF-FNIL-PROJ}\\
+%\mylabel{T-IfFilterNilProj}\\
 %\dfrac{\begin{array}{c}
@@
 \begin{array}{c}
-\mylabel{T-IF-FNIL-PROJ}\\
+\mylabel{T-IfFilterNilProj}\\
 \dfrac{\begin{array}{c}
@@
 
-Rule \textsc{T-IF-FNIL-PROJ} uses the auxiliary functions \emph{fpt} and \emph{gpt}
+Rule \textsc{T-IfFilterNilProj} uses the auxiliary functions \emph{fpt} and \emph{gpt}
 to filter a projection $x$. Function {\em fpt} filters out $\Nil$ from the i-th

### 2015-06-13 doc/papers/paper/paper.tex
DLS paper: add preprint option for page numbers
DLS paper: reformat section on projection types, fix conclusions

@@
-\documentclass{sigplanconf}
+\documentclass[preprint]{sigplanconf}
 
@@
 
-\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country}
+\conferenceinfo{DLS '15}{Month d--d, 20yy, City, ST, Country}
 \copyrightyear{20yy}
@@
 
-\titlebanner{banner above paper title}
-\preprintfooter{short description of paper}
-
-\title{Formalization of Typed Lua}
+\titlebanner{Draft}
+\preprintfooter{A Formalization of Typed Lua}
+
+\title{A Formalization of Typed Lua}
 
@@
 
+A key feature in optional type systems is usability.
+Optional type systems retrofitted to an existing
+language should fit the language's idioms, adding
+static type safety to them. If the system is too
+simple it may require too much of a change in the way
+programmers use the language. On the other hand,
+if it is too complex it may overload the programmers
+with types and error messages that are hard to
+understand. The most challenging aspect of designing optional type systems is to find the right amount of complexity for a type system that feels natural to the programmers.
+
 Lua is a small imperative language with lexically-scoped first-class functions where the only data structuring
@@
 
-\begin{figure*}[!ht]
+\begin{figure*}[t]
 $$
@@
 & | \; B & \textit{base types}\\
+& | \; \mathbf{void} & \textit{void type}\\
 & | \; \Nil & \textit{nil type}\\
@@
 
-The type $\Nil$ is the type of {\tt nil}, the value that Lua uses for
-undefined variables, missing parameters, and missing table keys.
+The type {\bf void} is a type with no values, used by the
+type system as a way to detect branches that are unreacheable
+due to flow typing. This type is neutral with regards to
+the type union operator: ${\bf void} \cup T = T \cup {\bf void} = T$ for any type $T$.
+
+The type $\Nil$ is the type of {\tt nil}, the value that Lua uses for undefined variables, missing parameters, and missing table keys.
 Many programming languages include a bottom type to represent
@@
 
+\begin{figure*}[t]
+\label{fig:projtypes}
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-LocalProj}\\
+\dfrac{\begin{array}{c}
+	\env_{1}, \penv \vdash el:T_1 \times \ldots T_n \times P, \env_{2}, (x,S)\\
+	\env_{3} = \env_{2}[id_{1} \mapsto T_1, \ldots, id_{n} \mapsto T_n]\\
+	\env_{3}, \penv[x \mapsto S] \vdash s, \env_{4} \;\;\;
+	n = |\;\overline{id}\;|  
+	\end{array}}
+{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, (\env_{4}-\{\overline{id}\})[\overline{id \mapsto \env_2(id)}]}
+\end{array}
+\begin{array}{c}
+\mylabel{T-ExpListProj}\\
+\dfrac{\begin{array}{c}
+	\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
+	\env_{1}, \penv \vdash me:S, \env_{n+2} \;\;\; n=|\overline{e}|\\
+	S = S_1 \sqcup S_2 \;\;\;m = max(|S_1|, |S_2|)-1 \;\;\; x \notin \Pi\\
+	\env_{f} = merge(\env_{1}, ..., \env_{n+2})
+	\end{array}}
+{\env_{1}, \penv \vdash \overline{e},me:}\\ {T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times last(S), \env_{f}, (x,S)}
+\end{array}\\\\
+\begin{array}{c}
+\mylabel{T-IfFilterProjThen}\\
+\dfrac{\begin{array}{c}
+	\env_{1}(id) = \pi_{i}^{x} \\ S_t = fonilpt(\Pi(x), i)\\ 
+	finil(proj(\Pi(x), i)) = {\bf void} \\
+	\env_{1}, \penv[x \mapsto S_t] \vdash s_{1}, \env_{2} 
+	\end{array}}
+{\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{2}}
+\end{array}
+\begin{array}{c}
+\mylabel{T-IfFilterProjElse}\\
+\dfrac{\begin{array}{c}
+	\env_{1}(id) = \pi_{i}^{x} \\
+	 S_e = finilpt(\Pi(x), i)\\
+	fonil(proj(\Pi(x), i)) = {\bf void} \\
+	\env_{1}, \penv[x \mapsto S_e] \vdash s_{2}, \env_{2} 
+	\end{array}}
+{\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{2}}
+\end{array}
+\begin{array}{c}
+\mylabel{T-IfFilterProjBoth}\\
+\dfrac{\begin{array}{c}
+	\env_{1}(id) = \pi_{i}^{x} \\ S_t = fonilpt(\Pi(x), i) \; S_e = finilpt(\Pi(x), i)\\
+	\env_{1}, \penv[x \mapsto S_t] \vdash s_{1}, \env_{2} \\
+	\env_{1}, \penv[x \mapsto S_e] \vdash s_{2}, \env_{3} \\
+	\env_{4} = join(\env_{2}, \env_{3}) 
+	\end{array}}
+{\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
+\end{array}
+\end{array}
+\]
+\caption{Rules for Projection Types}
+\end{figure*}
+
 As we have seen in Section~\ref{sec:statistics}, Lua
@@
 in an expression list has a union of tuple types as its type.
-
-Notice that in the above example the local variables $q$ and $r$ are declared without 
-an explicit type. The rule for typing this {\bf local} statement,
+Figure~\ref{fig:projtypes} presents the rules for introducing and filtering projection types.
+
+Notice that in the above example the local variables $q$ and $r$ are declared without an explicit type. 
+ The rule for typing this {\bf local} statement,
 {\sc T-LocalProj}, introduces a new label in the projection
-environment used for typing the {\bf if} statement:
-\[
-\begin{array}{c}
-\mylabel{T-LocalProj}\\
-\dfrac{\begin{array}{c}
-	\env_{1}, \penv \vdash el:T_1 \times \ldots T_n \times P, \env_{2}, (x,S)\\
-	\env_{3} = \env_{2}[id_{1} \mapsto T_1, \ldots, id_{n} \mapsto T_n]\\
-	\env_{3}, \penv[x \mapsto S] \vdash s, \env_{4} \;\;\;
-	n = |\;\overline{id}\;|  
-	\end{array}}
-{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, (\env_{4}-\{\overline{id}\})[\overline{id \mapsto \env_2(id)}]}
-\end{array}
-\]
-
-The first judgement above the bar uses the auxiliary typing relation $\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$.
-This rule does the work of generating a fresh projection label,
+environment used for typing the {\bf if} statement.
+The first judgement above the bar of rule {\sc T-LocalProj} uses the auxiliary typing relation $\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$.
+
+Rule {\sc T-ExpListProj} is the typing rule for this auxiliary
+relation. It does the work of generating a fresh projection label,
 and projection types associated with it, if the last term of
-the expression list has a union of tuple types as its type,
-as rule {\sc T-ExpListProj} below shows:
-\[
-\begin{array}{c}
-\mylabel{T-ExpListProj}\\
-\dfrac{\begin{array}{c}
-	\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
-	\env_{1}, \penv \vdash me:S, \env_{n+2} \;\;\; n=|\overline{e}|\\
-	S = S_1 \sqcup S_2 \;\;\;m = max(|S_1|, |S_2|)-1 \;\;\; x \notin \Pi\\
-	\env_{f} = merge(\env_{1}, ..., \env_{n+2})
-	\end{array}}
-{\env_{1}, \penv \vdash \overline{e},me:}\\ {T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times last(S), \env_{f}, (x,S)}
-\end{array}
-\]
-
-In the above example, the combination of these two rules will bind $x$ to
+the expression list has a union of tuple types as its type.
+
+In the above example, the combination of these two rules binds $x$ to
 $(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})$ in the projection environment $\Pi$,
@@
 At this point, using $q$ or $r$ as an r-value will project
-their types into either $\Integer \cup \Nil$ or $\Integer \cup \String$. But the more interesting case is inside the
-branches of the {\bf if} statement. The particular uses the rule {\sc T-IfFilterNilProj} below, which uses flow typing:
-\[
-%\begin{array}{c}
-%\mylabel{T-IfFilterNilProj}\\
-%\dfrac{\begin{array}{c}
-%       \env_{1}(id) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
-%       closeall(\env_{1}), \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
-%       closeall(\env_{1}), \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
-%       \env_{4} = closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2})) \\
-%       \env_{5} = openset(\env_{4}, rv(s_{1}) \cup rv(s_{2}))
-%      \end{array}}
-%      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{5}}
-%\end{array}
-\begin{array}{c}
-\mylabel{T-IfFilterNilProj}\\
-\dfrac{\begin{array}{c}
-       \env_{1}(id) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
-       \env_{1}, \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
-       \env_{1}, \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
-       \env_{4} = join(\env_{2}, \env_{3}) 
-      \end{array}}
-      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
-\end{array}
-\]
-
-Rule \textsc{T-IfFilterNilProj} uses the auxiliary functions \emph{fpt} and \emph{gpt}
-to filter a projection $x$. Function {\em fpt} filters out $\Nil$ from the i-th
-component of each tuple, removing the entire tuple if the i-th
-component does not exist, or is $\Nil$ itself. Function {\em gpt}
-filters the i-th component to $\Nil$ if the type includes
-$\Nil$, removing the entire tuple if the i-th component does
-not exist, or does not include $\Nil$.
+their types into either $\Integer \cup \Nil$ or $\Integer \cup \String$. The more interesting cases are inside the
+branches of the {\bf if} statement, where rule {\sc T-IfFilterNilProj} filters the whole union of tuples referenced
+by the projection type.
+
+Rule \textsc{T-IfFilterNilProj} uses the auxiliary functions \emph{fonilpt} ({\em filter-out nil in projection type}) and \emph{finilpt} ({\em filter-in nil in projection type}) to
+filter the union referenced by $\pi_{i}^{x}$, respectively
+filtering $\Nil$ out at every i-th component (removing
+the entire tuple if it is $\Nil$) and filtering $\Nil$ in at every i-th component (removing the entire tuple if it is not $\Nil$).
 
@@
 
+Rules {\sc T-IfFilterProjThen} and {\sc T-IfFilterProjElse}
+respectively cover the case where the {\bf else} branch is unreachable and the case where the {\bf then} branch is unreacheable, because either the projected type of $\pi_i^x$
+is not a supertype of $\Nil$ or is $\Nil$.
+
 Assignment to an l-value with a projection type is forbidden. Allowing assignment
-would be unsafe, as it could break the dependency among the
+would be unsound, as it could break the dependency among the
 components of the union, as the following example shows:
-\begin{verbatim}
-  local q, r = idiv(a, b)
-  r = "foo"
-  if q then
-    -- wrongly assume q and r are integers
-  end
-\end{verbatim}
+\begin{center}
+	\begin{tabular}{ll}
+		\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(a, b) \, \mathbf{in}$} \\
+		&  $r = ``foo" ;\; \mathbf{if} \, q \, \mathbf{then} \, s_1 \, \mathbf{else} \, s_2$
+	\end{tabular}
+\end{center}
+
+The projected type of $r$ outside of the {\bf if} statement
+is $\Integer \cup \String$, so the assignment seems correct,
+but inside $s_1$ the projected type of $r$ would still be
+$\Integer$.
 
@@
 
-\section{Conclusion}
+\section{Conclusions}
 \label{sec:conclusion}
 
-We have presented a formalization of the Typed Lua,
+We have presented a formalization of Typed Lua,
 an optional type system for Lua, with a focus on rules
@@
 or by using the table in a context that expects
-a different (but compatible) type than the one it
-has. The system tracks aliasing of table references
-to be able to do this type evolution in a type-safe
+a different (but compatible) type from its current one.
+The system tracks aliasing of table references
+in order to do this type evolution in a type-safe
 way.
@@
 In the second idiom, a function that can return different
-kinds of sequences of return values (for example, one
-sequence for its usual path, another for error conditions)
+kinds of tuples of return values (for example, one
+tuple for its usual path, another for error conditions)
 is modeled by having a union of tuple types as its return
@@
 Narrowing the type of one of the tuple's components with
-a type predicate can then narrow the types of the other
+a type predicate can narrow the types of the other
 components.
-
-A key feature in optional type systems is usability.
-Optional type systems retrofitted to an existing
-language should fit the language's idioms, adding
-static type safety to them. If the system is too
-simple it may require too much of a change in the way
-the programmers use the language. On the other hand,
-if it is too complex it may overload the programmers
-with types and error messages that are hard to
-understand. The most challenging aspect of designing optional type systems is to find the right amount of complexity for a type system that feels natural to the programmers.
 
@@
 and we could give precise static types to 83\% of the 449
-members that these modules export.
+values that these modules export.
 For half of the modules, we could give precise static types to
-at least 89\% of the members from each module.
+at least 89\% of the exports from each module.
 
@@
 type check several Lua idioms and features, though the evaluation
-also exposed several limitations of our type system.
-We found that the three main limitations of our type system are
+also exposed several limitations.
+We found that the three main limitations are
 the lack of full support for overloaded functions, the lack
 of parametric polymorphism, and operator overloading.
-Overcoming these limitations is our major target for future work,
+Overcoming these limitations is one major target for future work,
 as it will allow us to statically type check more programs.
 
-Unlike other optional type systems, we designed Typed Lua without
-deliberate unsound parts.
+Unlike some optional type systems, such as the type systems
+of the Typescript and Dart languages, we designed Typed Lua without deliberately unsound parts.
 However, we still do not have proofs that the novel features of
@@
 
-Typed Lua has a working implementation that
+Typed Lua has a working implementation\footnote{Available
+at \url{http://github.com/andremm/typelua}.} that
 Lua programmers can already use as a framework to
-document, test, and structure their applications,
-and we already have user feedback from Lua programmers
-that are using Typed Lua in their projects.
-One example is ZeroBrane Studio, a Lua IDE that is
-integrating Typed Lua.
-
+document, test, and structure their applications.
 Even applications where a full conversion to static
@@
 
+We already have feedback from Lua projects
+that are evaluating the use of Typed Lua, such
+as the LuaRocks package manager, which is a key
+piece of infrastructure for the Lua library ecosystem,
+and ZeroBrane Studio, an IDE for Lua programs that is itself
+written in Lua.
+
 \bibliographystyle{abbrvnat}

### 2015-06-13 doc/papers/paper/paper.tex
DLS: some fixes to the section on projection types

@@
 	\env_{1}, \penv \vdash me:S, \env_{n+2} \;\;\; n=|\overline{e}|\\
-	S = S_1 \sqcup S_2 \;\;\;m = max(|S_1|, |S_2|)-1 \;\;\; x \notin \Pi\\
+	S = S_1 \sqcup S_2 \;\;\;m = max(|S_1|, |S_2|) \;\;\; x \notin \Pi\\
 	\env_{f} = merge(\env_{1}, ..., \env_{n+2})
 	\end{array}}
-{\env_{1}, \penv \vdash \overline{e},me:}\\ {T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times last(S), \env_{f}, (x,S)}
+{\env_{1}, \penv \vdash \overline{e},me:}\\ {T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x}{*}, \env_{f}, (x,S)}
 \end{array}\\\\
@@
 {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
+\end{array}\\\\
+%\begin{array}{rcll}
+%last(F_1{*} \sqcup F_2{*}) & = & F_1{*} & \mathrm{if} \; F_2{*} \subtype F_1{*}\\
+%last(F_1{*} \sqcup F_2{*}) & = & F_2{*} & \mathrm{if} \; F_1{*} \subtype F_2{*}\\
+%last(F_1{*} \sqcup F_2{*}) & = & (F_1 \cup F_2){*} & \mathrm{otherwise}\\
+%last(F{*} \sqcup T \times P) & = & last(\Nil{*} \sqcup P) \\
+%last(T \times P \sqcup F{*}) & = & last(P \sqcup \Nil{*}) \\
+%last(P \sqcup S) & = & last(P \sqcup last(S)) \\
+%last(S \sqcup P) & = & last(last(S) \sqcup P) \\
+%last(S_1 \sqcup S_2) & = & last(last(S_1) \sqcup last(S_2))
+%\end{array}\\\\
+\begin{array}{rcll}
+proj(S_1 \sqcup S_2, i) & = & proj(S_1, i) \cup proj(S_2, i) \\
+proj(F{*}, i) & = & F \cup \Nil \\
+proj(T \times P, 1) & = & T \\
+proj(T \times P, i) & = & proj(P, i-1)
+\end{array}\\\\
+\begin{array}{rcll}
+fonilpt(P_1 \sqcup P_2) & = & P_2 & \mathrm{if} \; fonil(proj(P_1, i)) = {\bf void} \\
+fonilpt(P_1 \sqcup P_2) & = & P_1 & \mathrm{if} \; fonil(proj(P_2, i)) = {\bf void} \\
+fonilpt(P_1 \sqcup P_2) & = & P_1 \sqcup P_2 & \mathrm{otherwise}\\
+fonilpt(P \sqcup S) & = & fonilpt(S) & \mathrm{if} \; fonil(proj(P, i)) = {\bf void} \\
+fonilpt(P \sqcup S) & = & P \sqcup fonilpt(S) & \mathrm{otherwise}\\
+fonilpt(S \sqcup P_2) & = & fonilpt(S) & \mathrm{if} \; fonil(proj(P, i)) = {\bf void} \\
+fonilpt(S \sqcup P) & = & fonilpt(S) \sqcup P & \mathrm{otherwise}\\
+fonilpt(S_1 \sqcup S_2) & = & fonilpt(S_1) \sqcup fonilpt(S_2)
+\end{array}\\\\
+\begin{array}{rcll}
+fonil(F_1 \cup F_2) & = & fonil(F_1) \cup fonil(F_2) \\
+fonil(F) & = & {\bf void} & \mathrm{if} \; F \subtype \Nil \; \mathrm{and} \; \Nil \subtype F\\
+fonil(F) & = & F & \mathrm{otherwise}
 \end{array}
@@
 the expression list has a union of tuple types as its type.
+The rule generates as many projection types as the the length
+of the longest tuple.
 
@@
 the entire tuple if it is $\Nil$) and filtering $\Nil$ in at every i-th component (removing the entire tuple if it is not $\Nil$).
+These functions respectively use the functions $fonil$ and $finil$ to filter regular unions, plus $proj$ to test projected components.
+The definitions of $finilpt$ and $finil$ are analogous to the definitions of $finilpt$ and $finil$.
 
@@
 
-Assignment to an l-value with a projection type is forbidden. Allowing assignment
-would be unsound, as it could break the dependency among the
-components of the union, as the following example shows:
+Assignment to an l-value with a projection type is forbidden. Allowing assignment is unsound, as it can break the dependency among the components of the union. In the example below,
+the projected type of $r$ outside of the {\bf if} statement
+is $\Integer \cup \String$, so the assignment seems correct,
+but inside $s_1$ the projected type of $r$ would still be
+$\Integer$:
 \begin{center}
@@
 \end{center}
-
-The projected type of $r$ outside of the {\bf if} statement
-is $\Integer \cup \String$, so the assignment seems correct,
-but inside $s_1$ the projected type of $r$ would still be
-$\Integer$.
 

### 2015-06-13 doc/papers/paper/paper.tex
DLS: several changes

@@
 \begin{figure}[t]
-\label{fig:newpoint}
 \begin{verbatim}
@@
 \caption{Building an object piece-by-piece}
+\label{fig:newpoint}
 \end{figure}
@@
 \begin{figure}[t]
-\label{fig:over}
 \begin{verbatim}
@@
 \caption{Overloading on multiple returns}
+\label{fig:over}
 \end{figure}
@@
 & | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
-& | \; \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique|open|fixed|closed} & \textit{table types}\\
+& | \; \{\overline{K:V}\}_{unique|open|fixed|closed} & \textit{table types}\\
 & | \; x & \textit{type variables}\\
@@
 \begin{figure}[t]
-\label{fig:subtuples}
 \[
@@
 \caption{Subtyping for second-level types}	
+\label{fig:subtuples}
 \end{figure}
@@
 \begin{figure}[t]
-\label{fig:subtable}
 \[
@@
 \dfrac{\begin{array}{c}
-	\forall i \in 1..n \; \exists j \in 1..m \\
-	\senv \vdash K_{j} \subtype K_{i}' \;\;\;
-	\senv \vdash K_{i}' \subtype K_{j} \;\;\;
-	\senv \vdash V_{j} \subtype_{c} V_{i}'
+	\forall j \; \exists i \;\;\;
+	\senv \vdash K_{i} \subtype K_{j}' \;\
+	\senv \vdash K_{j}' \subtype K_{i} \;
+	\senv \vdash V_{i} \subtype_{c} V_{j}'
 	\end{array}}
 {\begin{array}{c}
-	\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{fixed|closed} \subtype\\
-	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}
-	\end{array}} \; m \ge n
+	\senv \vdash \{\overline{K:V}\}_{fixed|closed} \subtype
+	\{\overline{K^\prime:V^\prime}\}_{closed}
+	\end{array}}
 \end{array}\\\\
@@
 \dfrac{\begin{array}{c}
-	\forall i \in 1..m \; \forall j \in 1..n \;
+	\forall i \; \forall j \;\;\;
 	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{u} V_{j}' \\
-	\forall j \in 1..n \; \not\exists i \in 1..m \;
+	\forall j  \; \nexists i \;\;\;
 	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
 	\end{array}}
-{\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype
-	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
+{\senv \vdash \{\overline{K:V}\}_{unique} \subtype
+	\{\overline{K^\prime:V^\prime}\}_{closed}}
 \end{array}\\\\
@@
 \dfrac{\begin{array}{c}
-	\forall i \in 1..m \; \forall j \in 1..n \;
+	\forall i \; \forall j \;\;\;
 	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
-	\forall j \in 1..n \; \not\exists i \in 1..m \;
+	\forall j \; \nexists i \;\;\;
 	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
 	\end{array}}
-{\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype
-	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
+{\senv \vdash \{\overline{K:V}\}_{open} \subtype
+	\{\overline{K^\prime:V^\prime}\}_{closed}}
 \end{array}\\\\
@@
 \dfrac{\begin{array}{c}
-	\forall i \in 1..m \\
-	\exists j \in 1..n \;
+	\forall i \; \exists j \;\;\;
 	\senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
-	\forall j \in 1..n \; \not\exists i \in 1..m \;
+	\forall j  \; \nexists i \;\;\;
 	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
@@
 {\begin{array}{c}
-	\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype\\
-	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{unique|open|fixed}
+	\senv \vdash \{\overline{K:V}\}_{unique} \subtype
+	\{\overline{K^\prime:V^\prime}\}_{unique|open|fixed}
 	\end{array}}
@@
 \dfrac{\begin{array}{c}
-	\forall i \in 1..m \\
-	\exists j \in 1..n \;
+	\forall i \; \exists j \;\;\;
 	\senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
-	\forall j \in 1..n \; \not\exists i \in 1..m \;
+	\forall j \; \nexists i \;\;\;
 	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
@@
 {\begin{array}{c}
-	\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype\\
-	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{open|fixed}
+	\senv \vdash \{\overline{K:V}\}_{open} \subtype
+	\{\overline{K^\prime:V^\prime}\}_{open|fixed}
 	\end{array}}
@@
 \dfrac{\begin{array}{c}
-	\forall i \in 1..n \; \exists j \in 1..n \\
-	\senv \vdash K_{j} \subtype K_{i}' \;\;\;
-	\senv \vdash K_{i}' \subtype K_{j} \;\;\;
-	\senv \vdash V_{j} \subtype_{c} V_{i}'
-	\end{array}}
-{\senv \vdash \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{fixed} \subtype
-	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{fixed}}
+	\forall j \; \exists i \;\;\;
+	\senv \vdash K_{i} \subtype K_{j}' \;
+	\senv \vdash K_{j}' \subtype K_{i} \;
+	\senv \vdash V_{i} \subtype_{c} V_{j}'\\
+	\forall i \; \exists j \;\;\;
+	\senv \vdash K_{i} \subtype K_{j}' \;
+	\senv \vdash K_{j}' \subtype K_{i} \;
+	\senv \vdash V_{i} \subtype_{c} V_{j}'\\
+	\end{array}}
+{\senv \vdash \{\overline{K:V}\}_{fixed} \subtype
+	\{\overline{K^\prime:V^\prime}\}_{fixed}}
 \end{array}
@@
 \caption{Subtyping for table types}
+\label{fig:tablesub}
 \end{figure}
@@
 \begin{figure}[t]
-\label{fig:subfield}
 \[
@@
 \caption{Subtyping for fields}
+\label{fig:subfield}
 \end{figure}
 
-Figure~\ref{fig:subtable} shows the subtyping rules for
+Figure~\ref{fig:tablesub} shows the subtyping rules for
 table types. These rules use the auxiliary relations
@@
 \begin{figure}[t]
-\label{fig:subrec}
 \[
@@
 \caption{Subtyping for recursive types}
+\label{fig:subrec}
 \end{figure}
 
-Figure~\ref{fig:subrec} shows the subtyping rules for
-recursive types. Rule {\sc S-Amber} is the \emph{Amber rule}~\cite{cardelli1986amber} for subtyping between
-two recursive types. Rules {\sc S-UnfoldRight} and {\sc S-UnfoldLeft} unfold the recursive type for establishing
-subtyping between a non-recursive and a recursive type,
-or vice-versa.
-
-The dynamic type $\Any$ is only a subtype of itself and of
-$\Value$, and no type except $\Any$ is a subtype of $\Any$.
-The type $\Any$ interacts with other types through the
-{\em consistency} and {\em consistent-subtyping} 
-relations~\cite{siek2007objects}. We use the symbol $\sim$ for
-the consistency, and $\lesssim$ for consistent-subtyping.
-
-The type $\Any$ is trivially consistent with all types, and all types
-are consistent with $\Any$. A type is also consistent with
-itself. Two union types $F_1 \cup F_2$ and $F_3 \cup F_4$ are
-consistent if either $F_1 \sim F_3$ and $F_2 \sim F_4$ or
-$F_1 \sim F_4$ and $F_2 \sim F_3$. Two function types
-$S_1 \rightarrow S_2$ and $S_3 \rightarrow S_4$ are consistent
-if and only if $S_1 \rightarrow S_3$ and $S_2 \rightarrow S_4$.
-Two table types are consistent if and only if they have the same key types, and the value types of each key type are consistent.
-Two recursive types are consistent if and only if their contents are consistent assuming that their type variables are consistent.
-Unions of tuple types follow the same rules as unions of first-class types.
-Variadic types are consistent if and only if their base types are consistent.
-Finally, pair types $F_1 \times P_1$ and $F_2 \times P_2$ are
-consistent if and only if $F_1 \sim F_2$ and $P_1 \sim P_2$.
-
-In a language with subtyping, the consistency relation is
-too restrictive. We need to combine it with subtyping into a
-{\em consistent-subtyping} relation. The definition is
-straightforward: a type is a consistent-subtype
-of another if the first type is a subtype of a type
-that is consistent with the other, or if the first type
-is consistent with a type that is a subtype of the other.
-
-In the implementation of Typed Lua, we also use consistent-subtyping to normalize and simplify union types.
-For instance, the union type $\Boolean \cup \Any$ simplifies
-to $\Any$, because $\Boolean$ is a consistent-subtype
-of $\Any$. Another example is the union type
-$\Number \cup \Nil \cup 1$, that simplifies to
-$\Number \cup \Nil$, because $1$ is a consistent-subtype of $\Number$.
-
-\section{Typing rules}
-\label{sec:rules}
-
-\begin{figure*}[!ht]
+\begin{figure*}[t]
 	$$
@@
 	| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
-	& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
+	& | \; \mathbf{local} \; \overline{id{:}F} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
 	& | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
-	& | \; \mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s & \textit{recursive function} \\
+	& | \; \mathbf{rec} \; id{:}F = f \; \mathbf{in} \; s & \textit{recursive function} \\
 	& | \; \mathbf{return} \; el & \textit{return} \\
@@
 	& | \; e_{1}[e_{2}] & \textit{table access}\\
-	& | \; {<}T{>} \; id & \textit{type coercion}\\
+	& | \; {<}F{>} \; id & \textit{type coercion}\\
 	& | \; f & \textit{function declaration}\\
@@
 	& | \; e_{1}[e_{2}] & \textit{table assignment}\\
-	& | \; id[k] \; {<}T{>} & \textit{type coercion}\\
+	& | \; id[k] \; {<}V{>} & \textit{type coercion}\\
 	k ::= & & \textsc{literal constants:}\\
@@
 	pl ::= & & \textsc{parameter lists:}\\
-	& \;\; \overline{id{:}T} \; | \;
-	\overline{id{:}T},{...}{:}T & \\
+	& \;\; \overline{id{:}F} \; | \;
+	\overline{id{:}F},{...}{:}F & \\
 	\end{array}
@@
 	\label{fig:syntax}
+\end{figure*}
+
+Figure~\ref{fig:subrec} shows the subtyping rules for
+recursive types. Rule {\sc S-Amber} is the \emph{Amber rule}~\cite{cardelli1986amber} for subtyping between
+two recursive types. Rules {\sc S-UnfoldRight} and {\sc S-UnfoldLeft} unfold the recursive type for establishing
+subtyping between a non-recursive and a recursive type,
+or vice-versa.
+
+The dynamic type $\Any$ is only a subtype of itself and of
+$\Value$, and no type except $\Any$ is a subtype of $\Any$.
+The type $\Any$ interacts with other types through the
+{\em consistency} and {\em consistent-subtyping} 
+relations~\cite{siek2007objects}. We use the symbol $\sim$ for
+the consistency, and $\lesssim$ for consistent-subtyping.
+
+The type $\Any$ is trivially consistent with all types, and all types
+are consistent with $\Any$. A type is also consistent with
+itself. Two union types $F_1 \cup F_2$ and $F_3 \cup F_4$ are
+consistent if either $F_1 \sim F_3$ and $F_2 \sim F_4$ or
+$F_1 \sim F_4$ and $F_2 \sim F_3$. Two function types
+$S_1 \rightarrow S_2$ and $S_3 \rightarrow S_4$ are consistent
+if and only if $S_1 \rightarrow S_3$ and $S_2 \rightarrow S_4$.
+Two table types are consistent if and only if they have the same key types, and the value types of each key type are consistent.
+Two recursive types are consistent if and only if their contents are consistent assuming that their type variables are consistent.
+Unions of tuple types follow the same rules as unions of first-class types.
+Variadic types are consistent if and only if their base types are consistent.
+Finally, pair types $F_1 \times P_1$ and $F_2 \times P_2$ are
+consistent if and only if $F_1 \sim F_2$ and $P_1 \sim P_2$.
+
+In a language with subtyping, the consistency relation is
+too restrictive. We need to combine it with subtyping into a
+{\em consistent-subtyping} relation. The definition is
+straightforward: a type is a consistent-subtype
+of another if the first type is a subtype of a type
+that is consistent with the other, or if the first type
+is consistent with a type that is a subtype of the other.
+
+In the implementation of Typed Lua, we also use consistent-subtyping to normalize and simplify union types.
+For instance, the union type $\Boolean \cup \Any$ simplifies
+to $\Any$, because $\Boolean$ is a consistent-subtype
+of $\Any$. Another example is the union type
+$\Number \cup \Nil \cup 1$, that simplifies to
+$\Number \cup \Nil$, because $1$ is a consistent-subtype of $\Number$.
+
+\section{Typing rules}
+\label{sec:rules}
+
+\begin{figure*}[t]
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-Constructor}\\
+\dfrac{\begin{array}{c}
+\overline{\env_1, \penv \vdash e_2: fix(V), \env_v}
+	\;\;\;
+	\overline{\env_v, \penv \vdash e_1: K, \env_f}
+	\\
+	T = \{\overline{K{:}V}\}_{unique} \;\;\;
+	\env_{2} = merge(\overline{\env_{f}}) \;\;\; wf(T)
+	\end{array}}
+{\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{2}}
+\end{array}
+\begin{array}{c}
+\mylabel{T-NewFieldUnique}\\
+\dfrac{\begin{array}{c}
+	\env_{1}(id) = \{ \overline{K:V} \}_{unique}\\
+	\env_{1}, \penv \vdash k:K^{new}, \env_{2} \;\;\;
+	\overline{K^{new} \not\lesssim K} \;\;\;
+	V^{new} = fix(V)
+	\end{array}}
+{\begin{array}{l}
+	\env_{1}, \penv \vdash id[k] {<}V{>}:V^new,
+	\env_{2}[id \mapsto \{ \overline{K{:}V}, K^{new}{:}V^{new}\}_{unique}]
+	\end{array}}
+\end{array}\\\\
+\begin{array}{c}
+\mylabel{T-IdReadClose}\\
+\dfrac{\env(id) = F}
+{\env, \penv \vdash id:close(F), \env[id \mapsto open(F)]}
+\end{array}
+\begin{array}{c}
+\mylabel{T-IdReadFix}\\
+\dfrac{\env(id) = F}
+{\env, \penv \vdash id:fix(F), \env[id \mapsto fix(F)]}
+\end{array}
+\end{array}
+\]
+\caption{Rules for Table Types}
+\label{fig:tablerules}
 \end{figure*}
@@
 \item type coercion for evolution of table types is explicit,
-with two coercion terms, the l-value $id[k] \; {<}T{>}$ for adding a field to 
-a table through assignment, and the r-value ${<}T{>} \;id$ for reinterpreting
+with two coercion terms, the l-value $id[k] \; {<}V{>}$ for adding a field to 
+a table through assignment, and the r-value ${<}F{>} \;id$ for reinterpreting
 a table type to a compatible type that is expected by the context;
@@
 
+\begin{figure*}
+	\[
+	\begin{array}{rcl}
+	wf(\{\overline{K:V}\}_{unique|open|fixed|closed}) & = & \forall i \; ((\nexists j \; i \not= j \,\wedge\, K_{i} \lesssim K_{j}) \,\wedge\, wf(V_{i}) \,\wedge\\
+	& & \;\;\;\; \lnot tag(V_{i},unique) \,\wedge\, \lnot tag(V_{i},open))\\
+	wf(F_1 \cup F_2) & = & wf(F_1) \,\wedge\, wf(F_2) \\
+	wf(S_1 \rightarrow S_2) & = & wf(S_1) \,\wedge\, wf(S_2) \\
+	wf(\mu x.F) & = & wf(F) \\
+	wf({\bf const}\; F) & = & wf(F) \\
+	wf(\phi(F_1,F_2)) & = & wf(F_1)\,\wedge\,wf(F_2)\\
+	wf(S_1 \sqcup S_2) & = & wf(S_1) \,\wedge\, S_2\\
+	wf(F{*}) &=& wf(F) \\
+	wf(T \times P) &=& wf(T) \,\wedge\,wf(P)\\
+	wf(T) & = & \top \;\;\;\mathrm{for\; all\; other\; cases}\\\\
+	tag(F_1 \cup F_2, t) & = & tag(F_1, t) \,\vee\, tag(F_2,t) \\
+	tag(\{\overline{K:V}\}_t, t) & = & \top\\
+	\multicolumn{3}{c}{
+		\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{other\;recusive\;cases\;are\;similar}}\\
+	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{other\;base\;cases\;are\;}\bot}\\\\
+	fix(\{\overline{K:V}\}_{unique|open}) & = &
+	\{\overline{K:V}\}_{fixed}\\
+	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{other\;cases\;by\;straightforward\;structural\;recursion}}\\\\
+	merge(\overline{\env}) & = & reduce(\overline{\env}, merge2)\\
+	merge2(\env_1,\env_2) & = & \{\overline{(id,merget(\env_1(id),\env_2(id))}\}\\
+	merget(T_1, T_2) & = & T_1 \;\;\; \mathrm{if\;} T_2 \subtype T_1\\
+	merget(T_1, T_2) & = & T_2 \;\;\; \mathrm{if\;} T_1 \subtype T_2\\
+	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;\forall i\;V^l_i\subtype_u V^r_i \,\vee\,V^r_i\subtype_u V^l_i \;\mathrm{and\;the\;right\; side\; is\;}wf}\\
+	merget(\{\overline{K:V^l},\overline{K^{\prime}:V^\prime}\}_{unique},
+	\{\overline{K:V^r},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{unique}) &=&
+	\{\overline{K:sup_u(V^l,V^r)},\overline{K^\prime:V^\prime},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{unique}
+	\\
+	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;\forall i\;V^l_i\subtype_c V^r_i \,\vee\,V^r_i\subtype_c V^l_i
+		\;\mathrm{and\;the\;right\; side\; is\;}wf}\\
+	merget(\{\overline{K:V^l},\overline{K^{\prime}:V^{\prime}}\}_{unique|open},
+	\{\overline{K:V^r},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{unique|open}) &=&
+	\{\overline{K:sup_c(V^l,V^r)},\overline{K^\prime:V^\prime},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{open}
+	\\
+	merget(T_1, T_2) & = & \bot \;\;\; \mathrm{otherwise}\\\\
+	join(\env_1,\env_2) & = & \{\overline{(id,joint(\env_1(id),\env_2(id))}\}\\
+	joint(T_1, T_2) & = & T_1 \;\;\; \mathrm{if\;} T_2 \subtype T_1\\
+	joint(T_1, T_2) & = & T_2 \;\;\; \mathrm{if\;} T_1 \subtype T_2\\
+	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;\forall i\;V^l_i\subtype_u V^r_i \,\vee\,V^r_i\subtype_u V^l_i
+		\;\mathrm{and\;the\;right\; side\; is\;}wf}\\
+	joint(\{\overline{K:V^l},\overline{K^{\prime}:V^{\prime}}\}_{unique},
+	\{\overline{K:V^r},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{unique}) &=&
+	\{\overline{K:sup_u(V^l,V^r)},\overline{K^\prime:V^\prime\cup\Nil},\overline{K^{\prime\prime}:V^{\prime\prime}\cup\Nil}\}_{unique}\\
+	joint(T_1, T_2) & = & \bot \;\;\; \mathrm{otherwise}\\\\
+	sup_u(V_1, V_2) & = & V_2 \;\;\; \mathrm{if}\; V_1 \subtype_u V_2\\
+	sup_u(V_1, V_2) & = & V_1 \;\;\; \mathrm{if}\; V_2 \subtype_u V_1\\
+	\end{array}
+	\]
+	\caption{Auxiliary Functions}
+\label{fig:funcs}
+\end{figure*}
+
 The simplest expression involving tables is the empty
@@
 type such as $\{ ``x":\Number , ``y": \Number \}_{unique}$ or
-even $\{ \String:\Number \cup \Nil \}_{unique}$. We give
-the typing rule for table constructors of this form below: 
-\[
-\begin{array}{c}
-\mylabel{T-CONSTRUCTOR-FIELDS}\\
-\dfrac{\begin{array}{c}
-	\overline{\env_1, \penv \vdash e_2: fix(V), \env_v}
-	\;\;\;
-	\overline{\env_v, \penv \vdash e_1: K, \env_f}
-	\\
-	T = \{\overline{K{:}V}\}_{unique} \;\;\;
-	\env_{2} = merge(\overline{\env_{f}}) \;\;\; wf(T)
-	\end{array}}
-{\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{2}}
-\end{array}
-%\dfrac{\begin{array}{c}
-%	\env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
-%	T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
-%	wf(T) \;\;\;
-%	n = |\;\overline{[e_{1}] = e_{2}}\;| \\
-%	\env_{f} = merge(\env_{1}, ..., \env_{n+1})
-%	\end{array}}
-%{\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{f}}
-%\end{array}
-\]
-
+even $\{ \String:\Number \cup \Nil \}_{unique}$.
+
+Figure~\ref{fig:tablerules} presents the typing rules for table types.
+Rule {\sc T-Constructor} is the rule for constructors
+like the examples above, that
+do not end with an expression that potentially returns
+multiple values.
 Due to the system for changing the table type of variables that
@@
 so all of them are typed in the same input type environment.
-The $merge$ auxiliary function then combines their output
+
+The $merge$ auxiliary function then combines output
 environments, diverging in case of conflicting changes.
-
-The rule also checks if the resulting type is {\em well-formed}
-with the predicate $wf$. Formally, we define this predicate
-for table types with the following formula:
-\[
-\begin{array}{c}
-\forall i \; ((\nexists j \; i \not= j \,\wedge\, K_{i} \lesssim K_{j}) \,\wedge\, wf(V_{i}) \,\wedge\,\\ \lnot isunique(V_{i}) \,\wedge\, \lnot isopen(V_{i}))
-\end{array}
-\]
-
-Well-formed table types avoid ambiguity.
-For instance, this rule detects that the table type
-$\{1:\Number, \Integer:\String, \Any:\Boolean\}$ is ambiguous,
-because indexing it with a key of type $\Integer$ could
+Fields can always be added to {\em unique} and {\em open}
+tables in one enviroment but not the other. Two environments
+can add a field with the same key, as long as they have
+the same type or one is a supertype of the other, respecting
+any restrictions with regards to depth subtyping that the table may have.
+
+Rule {\sc T-Constructor} also checks if the resulting type is {\em well-formed} with the predicate $wf$. Checking for well-formedness avoids ambiguous table types like
+$\{1:\Number, \Integer:\String, \Any:\Boolean\}$,
+where indexing it with a key of type $\Integer$ could
 yield either a number, a string or a boolean at runtime.
@@
 Well-formed table types also do not allow \emph{unique} and
-\emph{open} table types to appear in the type of the values,
-and they must be well-formed. Our type system currently cannot
+\emph{open} table types to appear in the type of the values.
+Our type system currently cannot
 track aliases of table fields, so this restriction is
 necessary.
-
-This restriction is the reason rule {\sc T-CONSTRUCTOR-FIELDS}
-uses the auxiliary function $fix$ to change any {\em open}
-and {\em unique} table types used in the field initializer
-to {\em fixed}. As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
+This limitation of the type system is the reason the {\sc T-Constructor} rule uses the auxiliary function $fix$ to change any {\em open} and {\em unique} table types used in the field initializer to {\em fixed}.
+As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
 has type $\{``x":1, ``y":\{``z":2\}_{fixed}\}_{unique}$.
 
-After presenting some typing rules of the table constructor,
-we start the discussion of the rules that define the most
-unusual feature of our type system: the refinement of table types.
-The first kind of refinement allows programmers to add new
-fields to \emph{unique} or \emph{open} table types through
-field assignment.
-As an example, we can translate
-\begin{verbatim}
-    local person = {}
-    person.firstname = "Lou"
-    person.lastname = "Reed"
-\end{verbatim}
-to our reduced core as follows:
+There are two ways the type of a {\em unique} and {\em open} tables can change in Typed
+Lua. The first way is by assignment to a non-existing
+field. Let us go back to the example shown in Figure~\ref{fig:newpoint}. Assuming that the type of the
+$x$ and $y$ is $\Number$, we can translate the first three lines
+of function {\tt new\_point} to the abstract syntax of Figure~\ref{fig:syntax}:
 \begin{center}
 \begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; person = \{\} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$person[``firstname"] \; {<}\String{>} = ``Lou";$}\\
-& \multicolumn{1}{l}{$person[``lastname"] \; {<}\String{>} = ``Reed"$}
+\multicolumn{2}{l}{$\mathbf{local} \; point: \{\}_{unique} = \{\} \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$point[``x"] \; {<}\Number{>} = x;$}\\
+& \multicolumn{1}{l}{$point[``y"] \; {<}\Number{>} = y$}
 \end{tabular}
@@
 
-In this example, we assign the type $\{\}_{unique}$ to the variable
-$person$, then we refine its type to $\{``firstname":\String\}_{unique}$,
-and then to $\{``firstname":\String, ``lastname":\String\}_{unique}$.
-Rule \textsc{T-NEWFIELD} type checks this use of refinement:
-\[
-\begin{array}{c}
-\mylabel{T-NEWFIELD}\\
-\dfrac{\begin{array}{c}
-       \env_{1}(id) = \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n} \}_{open|unique}\\
-       \env_{1}, \penv \vdash k:K, \env_{2} \;\;\;
-       \not \exists i \in 1..n \; K \lesssim K_{i} \;\;\;
-       V = close(T)
-       \end{array}}
-      {\begin{array}{l}
-       \env_{1}, \penv \vdash id[k] {<}T{>}:V, \\
-       \env_{2}[id \mapsto \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, K{:}V\}_{open|unique}]
-       \end{array}}
-\end{array}
-\]
-
-We restricted the refinement of table types to include only literal
-keys, because its purpose is to make it easier the construction of
-table types that represent records.
-
-We use the refinement of table types to handle the declaration of
-new global variables.
-In Lua, the assignment \texttt{v = v + 1} translates to
-\texttt{\string_ENV["v"] = \string_ENV["v"] + 1} when \texttt{v}
-is not a local variable, where \texttt{\string_ENV} is a table
-that stores the global environment.
-For this reason, Typed Lua treats accesses to global variables as field accesses
-to an \emph{open} table in the top-level scope.
-In the following examples we assume that $\string_ENV$ is in the
-environment and has type $\{\}_{open}$.
-
-As an example,
-\begin{align*}
-& \string_ENV[``x"] \; {<}\String{>} = ``foo" \;; \\
-& \string_ENV[``y"] \; {<}\Integer{>} = 1
-\end{align*}
-uses field assignment to add fields $``x"$ and $``y"$ to $\string_ENV$.
-Therefore, after these field assignments $\string_ENV$ has type
-$\{``x":\String, ``y":\Integer\}_{open}$.
-
-We do not allow the refinement of table types to add a field if it is
-already present in the table's type.
-For instance,
-\begin{align*}
-& \string_ENV[``x"] \; {<}\String{>} = ``foo" \;; \\
-& \string_ENV[``x"] \; {<}\Integer{>} = 1
-\end{align*}
-does not type check, as we are trying to add $``x"$ twice.
-
-We also do not allow the refinement of table types to introduce
-fields with table types that are not \emph{closed}.
-For instance,
+The expressions on the left-side of the assignment are type
+coercion l-values, and their type-checking is responsible
+for changing the type of $point$. The type annotations in
+these expressions are not
+required in actual Typed Lua code, as the compiler infers the types from the corresponding expression on the right side.
+
+Rule {\sc T-NewFieldUnique} typechecks coercion l-values on
+{\em unique} table types, guaranteeing that the new type
+is still well-formed. In the example above, the left
+side of the first assignment changes the type of $point$
+to $\{``x":\Number\}_{unique}$. The rest of the assignment
+typechecks, as the right-hand side has type $\Number$.
+The second assignment changes the type of $point$ to
+$\{``x":\Number, ``y": \Number\}_{unique}$.
+
+We could add both fields with a single assignment statement,
+as in the following example:
 \begin{center}
-\begin{tabular}{l}
-$\string_ENV[``x"] \; {<}\{\}_{unique}{>} = \{\}$
+	\begin{tabular}{ll}
+		\multicolumn{2}{l}{$\mathbf{local} \; point: \{\}_{unique} = \{\} \; \mathbf{in}$}\\
+		& \multicolumn{1}{l}{$point[``x"] \; {<}\Number{>},point[``y"] \; {<}\Number{>} = x,y;$}
 	\end{tabular}
 \end{center}
-refines the type of $\string_ENV$ from $\{\}_{open}$ to $\{``x":\{\}_{closed}\}_{open}$.
-Currently, our type system can only track \emph{unique} and
-\emph{open} table types that are bound to local variables.
-
-We can also use multiple assignment to refine table types:
-\[
-\string_ENV[``x"] \; {<}\String{>}, \string_ENV[``y"] \; {<}\Integer{>} = ``foo", 1
-\]
-
-This example type checks because all the environment changes are consistent, and
-$``foo" \times 1 \times \Nil{*} \lesssim \String \times \Integer \times \Value{*}$.
-By consistent we mean that we are only adding new fields.
-Nevertheless, the next example does not type check because it tries to add
-the same field to $\string_ENV$, but with different types:
-\[
-\string_ENV[``x"] \; {<}\String{>}, \string_ENV[``x"] \; {<}\Integer{>} = ``foo", 1
-\]
-
-Aliasing an \emph{unique} or an \emph{open} table type can produce
-either a \emph{closed} or a \emph{fixed} table type, depending on
-the context that we are using a variable.
-We need \emph{fixed} table types to type classes in object-oriented programming.
-In the implementation we fix the aliasing of \emph{unique} and \emph{open}
-table types that appear in a top-level return statement, and in other cases we
-close the aliasing of \emph{unique} and \emph{open} table types.
-However, in the formalization we chose to define this behavior in
-a not deterministic way, as it makes easier the presentation of this behavior.
+
+In this case, the first coercion expression would still change the type of $point$ to $\{``x":\Number\}_{unique}$, but the
+second would change it to $\{``y":\Number\}_{unique}$.
+Merging the two yields $\{``x":\Number, ``y": \Number\}_{unique}$
+through the third case of function $merget$ defined in Figure~\ref{fig:funcs}.
+
+Open tables have a rule {\sc T-NewFieldOpen} that is not shown
+in Figure~\ref{fig:tablerules} due to space. The only thing that
+changes from one rule to the other is the tag in the type of $id$.
+
+The type system needs to track aliasing of table references,
+as the flexibility of {\em unique} and {\em open} tables
+depend on respectively having no aliases and having only {\em closed} aliases. The system conservatively assumes that
+a table has been aliased it its identifier appears in r-value position, with a couple of exceptions which we will explain shortly.
+
+Rule {\sc T-Id} is responsible for tracking
+aliasing. Its auxiliary function $close$ is similar to
+function $fix$ defined in Figure~\ref{fig:funcs},
+except that it changes the tag of {\em unique} or
+{\em open} tables to {\em closed}. Its purpose
+is to construct a {\em closed} type for the alias,
+if necessary. Its auxiliary function $open$ is similar
+to $fix$ and $close$, except that it changes the type of
+{\em unique} tables to {\em open}. Its purpose
+is to track the fact that a $id$ now has an alias,
+if necessary.
 
@@
 as it changes the value that is stored in the field $``x"$ of $a$.
-
-The rules \textsc{T-ID-R-CLOSE} and \textsc{T-ID-R-FIX} define this non-deterministic behavior.
-Rule \textsc{T-ID-R-CLOSE} uses the auxiliary function \emph{close} to
-produce a \emph{closed} alias.
-It also uses the auxiliary function \emph{open} to change the type of
-the original reference from \emph{unique} to \emph{open},
-because aliasing an \emph{unique} table type while keeping the original
-reference \emph{unique} can be unsafe.
-Rule \textsc{T-ID-R-FIX} uses the auxiliary function \emph{fix} to
-produce a \emph{fixed} alias.
-It also uses \emph{fix} to change the type of the original reference
-to \emph{fixed}, because a \emph{fixed} table type does not allow
-width subtyping.
-We define these rules as follows:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-ID-R-CLOSE}\\
-\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
-      {\env_{1}, \penv \vdash id:close(T_{2}), \env_{1}[id \mapsto open(T_{1})]}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-ID-R-FIX}\\
-\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
-      {\env_{1}, \penv \vdash id:fix(T_{2}), \env_{1}[id \mapsto fix(T_{1})]}
-\end{array}
-\end{array}
-\]
-
-Both rules use the auxiliary function \emph{read} because they may be
-accessing an identifier that is bound to a filter or projection type.
-As we presented in our previous design of Typed Lua \cite{maidl2014tl},
-our type system includes a small set of \texttt{type} predicates that allow
-programmers to discriminate union types, and our type system uses filter and
-projection types in the definition of these predicates to handle the discrimination
-of unions types.
-While filter types discriminate unions of first-level types, projection
-types discriminate unions of second-level types and project unions of
-second-level types into unions of first-level types.
-We can define \emph{read} as follows:
-\begin{align*}
-read(\penv, \phi(T_{1},T_{2})) & = T_{2}\\
-read(\penv, \pi_{i}^{x}) & = proj(\penv(x), i)\\
-read(\penv, T) & = T
-\end{align*}
-
-The function \emph{read} uses the auxiliary function \emph{proj}
-to project a union of first-level types, based on an union of
-second-level types and an index from a projection type.
-In Section \ref{sec:projections} we will discuss how our type system uses
-projection types to handle overloaded return types.
-We can define \emph{proj} as follows:
-\begin{align*}
-proj(T_{1} \times ... \times T_{n} \times T{*}, i) & =  T_{i} \;\;\; \text{if $i <= n$}\\
-proj(S_{1} \sqcup S_{2}, i) & = proj(S_{1}, i) \cup proj(S_{2}, i)
-\end{align*}
 

### 2015-06-14 doc/papers/paper/paper.tex
DLS paper: finished revision, "release candidate" for submission

@@
 
-\conferenceinfo{DLS '15}{Month d--d, 20yy, City, ST, Country}
-\copyrightyear{20yy}
+\conferenceinfo{DLS '15}{October 25-30, 2015, Pittsburgh, PA, USA}
+\copyrightyear{2015}
 \copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
@@
 In this work, we give a formal description of Typed Lua, an
-optional type system for Lua, with a focus on two of its novel type system features: incremental evolution of record and object types via imperative update that does not sacrifice width subtyping, and combining flow typing with multiple assignment and with functions that return multiple values.
+optional type system for Lua, with a focus on two of its novel type system features: incremental evolution of imperative record and object types that is both lightweight and type-safe, and
+the combination of flow typing, functions that return multiple values, and multiple assignment.
 While our type system is tailored to the features and idioms of Lua, its features can be adapted to other imperative scripting languages.
@@
 
-\category{CR-number}{subcategory}{third-level}
+\category{D.3.1}{Programming Languages}{Formal Definitions and Theory}[Type Systems]
 
 \terms
-term1, term2
+Languages, Design
 
 \keywords
-keyword1, keyword2
+Optional Type Systems, Gradual Typing, Lua
 
@@
 different languages that have distinct syntax and semantics,
-requiring a complexe rewrite of existing programs instead of
+requiring a complex rewrite of existing programs instead of
 incremental evolution.
@@
 The first idioms that we focus on this paper is the use of assignment to build records and objects one field at a time, as in
-the Lua fragment in Figure~\ref{fig:newpoint}, where the {\tt newpoint} function
+the Lua fragment in Figure~\ref{fig:newpoint}, where the {\tt new\_point} function
 is a constructor for objects with {\tt x} and {\tt y}
@@
 
-The {\tt newpoint} function builds the object piece-by-piece,
+The {\tt new\_point} function builds the object piece-by-piece,
 from an initially empty table. Typed Lua
@@
 262 different projects. The scripts were parsed, and their
-abstract syntax trees analysed to collect the data that we
+abstract syntax trees analyzed to collect the data that we
 show in the rest of this section.
@@
 
-We also analysed expressions that access a table (130,448 instances), either to read a value out of it (86\% of such expressions) or to write a 
+We also analyzed expressions that access a table (130,448 instances), either to read a value out of it (86\% of such expressions) or to write a 
 value into it (14\% of such expressions). Most of the expressions
@@
 return values or throwing an exception. The caller then
-tests the first returned value to check if an error occured
+tests the first returned value to check if an error occurred
 or not. Standard union types and flow typing cannot deal
@@
 \begin{array}{rlr}
-T ::= & & \textsc{first-level types:}\\
-& \;\; F & \textit{first-class types}\\
-& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
-& | \; \pi_{i}^{x} & \textit{projection types}\\
-F ::= & & \textsc{{\small first-class types:}}\\
+F ::= & & \textsc{first-level types:}\\
 & \;\; L & \textit{literal types}\\
 & | \; B & \textit{base types}\\
-& | \; \mathbf{void} & \textit{void type}\\
 & | \; \Nil & \textit{nil type}\\
@@
 & | \; \{\overline{K:V}\}_{unique|open|fixed|closed} & \textit{table types}\\
-& | \; x & \textit{type variables}\\
+& |\; x & \textit{type variables}\\
 & | \; \mu x.F & \textit{recursive types}\\
@@
 B ::= & & \textsc{{\small base types:}}\\
-& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
+& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\\\
 %\multicolumn{3}{c}{}\\
-K ::= & & \textsc{{\small key types:}}\\
-& \;\; L \; | \; B \; | \; \Value &\\
-%\multicolumn{3}{c}{}\\
-V ::= & & \textsc{{\small value types:}}\\
-& \;\; F \; | \; \Const \; F &\\ 
 %\multicolumn{3}{c}{}\\
@@
 & | \; F{*} & \textit{variadic types}\\
-& | \; T \times P & \textit{pair types}
+& | \; F \times P & \textit{pair types}\\\\
 \end{array}
@@
 
-Figure \ref{fig:typelang} presents the abstract syntax of
-Typed Lua types. They are split into two categories:
+\begin{figure*}[t]
+$$
+\begin{array}{rlr}
+T ::= & & \textsc{{\small expression types:}}\\
+& \;\; V & \\
+& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
+& | \; \pi_{i}^{x} & \textit{projection types}\\
+E ::= & & \textsc{{\small expression list types:}}\\
+& \;\; T{*} & \\
+& | \; T \times E & \\
+K ::= & & \textsc{{\small key types:}}\\
+& \;\; L \; | \; B \; | \; \Value &\\
+%\multicolumn{3}{c}{}\\
+V ::= & & \textsc{{\small value types:}}\\
+& \;\; F \; | \; \Const \; F &\\
+R ::= & & \textsc{\small{filter result type:}}\\
+& \;\; \mathbf{void} & \textit{void type}\\
+& |\; F
+\end{array}
+$$
+\caption{Special Types}
+\label{fig:special}
+\end{figure*}
+
+Figure~\ref{fig:typelang} presents the abstract syntax of
+the core Typed Lua types. They are split into two categories:
 \emph{first-level types} and \emph{second-level types}.
 First-level types represent first-class Lua values and
-second-level types represent tuples of values that appear in 
-assignments and function applications.
-
-First-level types are
-themselves split into {\em first-class types}, which can be
-used in type annotations and most type constructors, and
-{\em filter types} and {\em projection types}, special
-types that only appear in support of flow typing.
-Typed Lua includes projection types as a way to project
-unions of tuple types into unions of first-level types,
-and we detailt their use in Section~\ref{sec:projections} .
+second-level types represent tuples of values that appear
+as parameters and return values of functions.
 
 Types are ordered by a subtype relationship, so Lua values may belong to several distinct types. The details of the subtyping
-relationship are given later in this section.
+relationship are given later in Section~\ref{sec:subtyping}.
 
@@
 strings, or numbers during runtime. Prior to version Lua 5.3 all Lua numbers were double-precision floating point numbers.
-Lua 5.3 introduced the possiblity of having 64-bit integers
+Lua 5.3 introduced the possibility of having 64-bit integers
 as numbers, with automatic promotion of integer values to
 floating-point representation as needed. As automatic
-promotion makes integers essentialy subtypes of floating-point numbers, Typed Lua uses $\Number$ as the type of floating-point numbers, and $\Integer$ as the type of integers.
+promotion makes integers essentially subtypes of floating-point numbers, Typed Lua uses $\Number$ as the type of floating-point numbers, and $\Integer$ as the type of integers.
 
 The type {\bf void} is a type with no values, used by the
-type system as a way to detect branches that are unreacheable
+type system as a way to detect branches that are unreachable
 due to flow typing. This type is neutral with regards to
@@
 Second-level types are either tuple types or unions of tuple types, and represent sequences of values that appear as arguments, return values, and the left and right sides of
-assignments. Tuple types are heterogeneous lists of first-class
+assignments. Tuple types are heterogeneous lists of first-level
 types that end with a {\em variadic type}.
@@
 For clarity, we use the symbol $\sqcup$ to represent the union between two different tuple types, and reserve $\cup$ for the
-union of two first-class types.
+union of two first-level types.
+
+Figure~\ref{fig:special} presents some types that, while
+they may not be present in type annotations, play some
+special role in the type system. {\em Expression types}
+denote the type of expressions in the type system, which,
+in addition to first-level types, can also be
+{\em filter types}, which are the result of flow typing,
+and {\em projection types}, which are the result of
+decomposing a union of tuples. {\em Expression list types}
+denote the type of lists of expressions. {\em Key types}
+and {\em value types} have also been mentioned above. Finally, {\em filter result types}
+are also used in flow typing; the {\bf void} type is
+a type with no values.
 
@@
 of subtyping, so we focus the discussion in this section
-on characterising the subtyping relation.
+on characterizing the subtyping relation.
 
@@
 Figure~\ref{fig:subtuples} shows the subtyping rules for
-second-level types. 
-Tuple types are covariant, so the subtyping rule {\sc S-Pair} for pair types  is straightforward.
+second-level types. Expression list types have similar
+subtyping rules. Tuple types are covariant, so the subtyping rule {\sc S-Pair} for pair types is straightforward.
 
@@
 \mylabel{S-FieldClsMtoM}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2} \;\;\;
-	\senv \vdash V_{2} \subtype V_{1}}
-{\senv \vdash V_{1} \subtype_{c} V_{2}}
+\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
+	\senv \vdash F_{2} \subtype F_{1}}
+{\senv \vdash F_{1} \subtype_{c} F_{2}}
 \end{array}
@@
 \mylabel{S-FieldClsItoI}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2}}
-{\senv \vdash \Const \; V_{1} \subtype_{c} \Const \; V_{2}}
+\dfrac{\senv \vdash F_{1} \subtype F_{2}}
+{\senv \vdash \Const \; F_{1} \subtype_{c} \Const \; F_{2}}
 \end{array}
@@
 \mylabel{S-FieldClsMtoI}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2}}
-{\senv \vdash V_{1} \subtype_{c} \Const \; V_{2}}
+\dfrac{\senv \vdash F_{1} \subtype F_{2}}
+{\senv \vdash F_{1} \subtype_{c} \Const \; F_{2}}
 \end{array}\\\\
@@
 \mylabel{S-FieldUnqMtoM}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2}}
-{\senv \vdash V_{1} \subtype_{u} V_{2}}
+\dfrac{\senv \vdash F_{1} \subtype F_{2}}
+{\senv \vdash F_{1} \subtype_{u} F_{2}}
 \end{array}
@@
 \mylabel{S-FieldUnqItoM}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2}}
-{\senv \vdash \Const \; V_{1} \subtype_{u} V_{2}}
+\dfrac{\senv \vdash F_{1} \subtype F_{2}}
+{\senv \vdash \Const \; F_{1} \subtype_{u} F_{2}}
 \end{array}
@@
 \mylabel{S-FieldUnqItoI}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2}}
-{\senv \vdash \Const \; V_{1} \subtype_{u} \Const \; V_{2}}
+\dfrac{\senv \vdash F_{1} \subtype F_{2}}
+{\senv \vdash \Const \; F_{1} \subtype_{u} \Const \; F_{2}}
 \end{array}
@@
 \mylabel{S-FieldUnqMtoI}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2}}
-{\senv \vdash V_{1} \subtype_{u} \Const \; V_{2}}
+\dfrac{\senv \vdash F_{1} \subtype F_{2}}
+{\senv \vdash F_{1} \subtype_{u} \Const \; F_{2}}
 \end{array}
@@
 \mylabel{S-FieldOpnNtoM}\\
-\dfrac{\senv \vdash \Nil \subtype V}
-{\senv \vdash \Nil \subtype_{o} V}
+\dfrac{\senv \vdash \Nil \subtype F}
+{\senv \vdash \Nil \subtype_{o} F}
 \end{array}
@@
 \mylabel{S-FieldOpnNtoI}\\
-\dfrac{\senv \vdash \Nil \subtype V}
-{\senv \vdash \Nil \subtype_{o} \Const \; V}
+\dfrac{\senv \vdash \Nil \subtype F}
+{\senv \vdash \Nil \subtype_{o} \Const \; F}
 \end{array}
@@
 belong to one of its key types, and does not have any
-aliases. With these guarantess, a table with type $\{1{:}\String, 2{:}\String,
+aliases. With these guarantees, a table with type $\{1{:}\String, 2{:}\String,
 ``x"{:}\Integer, ``y"{:}\Value\}_{unique}$ can be safely recast as a table with type 
@@
 inaccessible with the original type, which Section~\ref{sec:tables} will show that is the case
-for any type judgement where this subtyping rule is used.
+for any type judgment where this subtyping rule is used.
 
@@
 \mylabel{S-Amber}\\
-\dfrac{\senv[x_{1} \subtype x_{2}] \vdash T_{1} \subtype T_{2}}
-{\senv \vdash \mu x_{1}.T_{1} \subtype \mu x_{2}.T_{2}}
+\dfrac{\senv[x_{1} \subtype x_{2}] \vdash F_{1} \subtype F_{2}}
+{\senv \vdash \mu x_{1}.F_{1} \subtype \mu x_{2}.F_{2}}
 \end{array}
@@
 \mylabel{S-UnfoldRight}\\
-\dfrac{\senv \vdash T_{1} \subtype [x \mapsto \mu x.T_{2}]T_{2}}
-{\senv \vdash T_{1} \subtype \mu x.T_{2}}
+\dfrac{\senv \vdash F_{1} \subtype [x \mapsto \mu x.F_{2}]F_{2}}
+{\senv \vdash F_{1} \subtype \mu x.F_{2}}
 \end{array}
@@
 \mylabel{S-UnfoldLeft}\\
-\dfrac{\senv \vdash [x \mapsto \mu x.T_{1}]T_{1} \subtype T_{2}}
-{\senv \vdash \mu x.T_{1} \subtype T_{2}}
+\dfrac{\senv \vdash [x \mapsto \mu x.F_{1}]F_{1} \subtype F_{2}}
+{\senv \vdash \mu x.F_{1} \subtype F_{2}}
 \end{array}
@@
 Two recursive types are consistent if and only if their contents are consistent assuming that their type variables are consistent.
-Unions of tuple types follow the same rules as unions of first-class types.
+Unions of tuple types follow the same rules as unions of first-level types.
 Variadic types are consistent if and only if their base types are consistent.
@@
 \dfrac{\begin{array}{c}
-\overline{\env_1, \penv \vdash e_2: fix(V), \env_v}
+\overline{\env_1, \penv \vdash e_2: F, \env_v}
 	\;\;\;
@@
 	\\
-	T = \{\overline{K{:}V}\}_{unique} \;\;\;
+	T = \{\overline{K{:}fix(F)}\}_{unique} \;\;\;
 	\env_{2} = merge(\overline{\env_{f}}) \;\;\; wf(T)
@@
 {\begin{array}{l}
-	\env_{1}, \penv \vdash id[k] {<}V{>}:V^new,
+	\env_{1}, \penv \vdash id[k] {<}V{>}:V^{new},
 	\env_{2}[id \mapsto \{ \overline{K{:}V}, K^{new}{:}V^{new}\}_{unique}]
@@
 \begin{array}{c}
-\mylabel{T-IdReadClose}\\
+\mylabel{T-IdRead}\\
 \dfrac{\env(id) = F}
@@
 \begin{array}{c}
-\mylabel{T-IdReadFix}\\
+\mylabel{T-IdWrite}\\
 \dfrac{\env(id) = F}
-{\env, \penv \vdash id:fix(F), \env[id \mapsto fix(F)]}
+{\env, \penv \vdash id_{l}:F, \env}
+\end{array}
+\begin{array}{c}
+\mylabel{T-IndexId}\\
+\dfrac{\begin{array}{c}
+	\env_{1}(id) = \{\overline{K:V}\} \;\;\;
+	\env_{1}, \penv \vdash e_{2}:K^e, \env_{2} \;\;\;
+	 K^e \lesssim K_{i}
+	\end{array}}
+{\env_{1}, \penv \vdash id[e_{2}]:V_{i}, \env_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{T-IndexExp}\\
+\dfrac{\begin{array}{c}
+	\env_{1}, \penv \vdash e_{1}:\{\overline{K:V}\}, \env_{2}\\
+	\env_{2}, \penv \vdash e_{2}:K^e, \env_{3} \;\;\;
+	 K^e \lesssim K_{i}
+	\end{array}}
+{\env_{1}, \penv \vdash e_{1}[e_{2}]:V_{i}, \env_{3}}
+\end{array}
+\begin{array}{c}
+\mylabel{T-CoerceClosed}\\
+\dfrac{\env_{1}(id) \subtype F \;\;\; tag(F,closed)}
+{\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto reopen(F)]}
+\end{array}
+\begin{array}{c}
+\mylabel{T-CoerceFixed}\\
+\dfrac{\env_{1}(id) \subtype F \;\;\; tag(F,fixed)}
+{\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto F]}
+\end{array}\\
+\begin{array}{c}
+\mylabel{T-Function}\\
+\dfrac{\begin{array}{c}
+	closeall(\env_{1})[\overline{id \mapsto F}], \penv[\ret \mapsto S] \vdash s, \env_{2}\\
+	\env_{3} = openset(\env_{1}, frv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)) \\
+	\env_{4} = closeset(\env_{3}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
+	\end{array}}
+{\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}F}){:}S \; s:F_1 \times \ldots F_n \times \Void \rightarrow S, \env_{3}}
 \end{array}
@@
 Typing judgments for statements have the form $\env_{1}, \penv \vdash s, \env_{2}$, meaning that
-given a type environment $\env_{1}$ and a {\em projection environment} $\penv$ we can prove that a statement $s$ produces a new type environment $\env_{2}$. Typing judgements for
+given a type environment $\env_{1}$ and a {\em projection environment} $\penv$ we can prove that a statement $s$ produces a new type environment $\env_{2}$. Typing judgments for
 expressions have the form $\env_{1}, \penv \vdash e : T, \env_{2}$, meaning that given a type environment $\env_{1}$
@@
 	& & \;\;\;\; \lnot tag(V_{i},unique) \,\wedge\, \lnot tag(V_{i},open))\\
+	wf({\bf const}\; F) & = & wf(F) \\
 	wf(F_1 \cup F_2) & = & wf(F_1) \,\wedge\, wf(F_2) \\
+	wf(\mu x.F) & = & wf(F) \\
 	wf(S_1 \rightarrow S_2) & = & wf(S_1) \,\wedge\, wf(S_2) \\
-	wf(\mu x.F) & = & wf(F) \\
-	wf({\bf const}\; F) & = & wf(F) \\
-	wf(\phi(F_1,F_2)) & = & wf(F_1)\,\wedge\,wf(F_2)\\
 	wf(S_1 \sqcup S_2) & = & wf(S_1) \,\wedge\, S_2\\
 	wf(F{*}) &=& wf(F) \\
-	wf(T \times P) &=& wf(T) \,\wedge\,wf(P)\\
-	wf(T) & = & \top \;\;\;\mathrm{for\; all\; other\; cases}\\\\
+	wf(F \times P) &=& wf(F) \,\wedge\,wf(P)\\
+	wf(F) & = & \top \;\;\;\mathrm{for\; all\; other\; cases}\\\\
 	tag(F_1 \cup F_2, t) & = & tag(F_1, t) \,\vee\, tag(F_2,t) \\
@@
 	\multicolumn{3}{c}{
-		\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{other\;recusive\;cases\;are\;similar}}\\
-	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{other\;base\;cases\;are\;}\bot}\\\\
+		\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{other\;recusive\;cases\;are\;similar,\;other\;base\;cases\;are\;}\bot}\\
+\\\\
 	fix(\{\overline{K:V}\}_{unique|open}) & = &
 	\{\overline{K:V}\}_{fixed}\\
-	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{other\;cases\;by\;straightforward\;structural\;recursion}}\\\\
+	\multicolumn{3}{c}{\;\;\;\;\;\;\;\mathrm{other\;cases\;by\;straightforward\;structural\;recursion}}\\\\
 	merge(\overline{\env}) & = & reduce(\overline{\env}, merge2)\\
@@
 	sup_u(V_1, V_2) & = & V_2 \;\;\; \mathrm{if}\; V_1 \subtype_u V_2\\
-	sup_u(V_1, V_2) & = & V_1 \;\;\; \mathrm{if}\; V_2 \subtype_u V_1\\
+	sup_u(V_1, V_2) & = & V_1 \;\;\; \mathrm{if}\; V_2 \subtype_u V_1\\\\
+	reopen(\{\overline{K:V}\}_{closed}) & = & \{\overline{K:V}\}_{open} \\
+	reopen(F) & = & F
 	\end{array}
@@
 Fields can always be added to {\em unique} and {\em open}
-tables in one enviroment but not the other. Two environments
+tables in one environment but not the other. Two environments
 can add a field with the same key, as long as they have
@@
 
-Rule {\sc T-Id} is responsible for tracking
+Rule {\sc T-IdRead} is responsible for tracking
 aliasing. Its auxiliary function $close$ is similar to
@@
 
-As an example,
+Consider the following example, where we make an alias
+of a {\em unique} table:
 \begin{center}
@@
 \multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
-& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = a \; \mathbf{in}$}\\
-& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
-& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
+& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{closed} = a \; \mathbf{in}a[``x"] \; {<}\String{>} = ``foo";$}
 \end{tabular}
 \end{center}
-does not type check, as aliasing $a$ produces the type $\{\}_{closed}$
-that is not a subtype of $\{\}_{open}$, the type of $b$.
-Our type system has this behavior to warn programmers about
-potential unsafe behaviors after this kind of alias.
-In this example, it is unsafe to add the field $``x"$ to $b$,
-as it changes the value that is stored in the field $``x"$ of $a$.
-
-We also need to close \emph{unique} and \emph{open} tables that
-appear in the left-hand side of assignments, as leaving them
-\emph{unique} and \emph{open} would allow the creation of
-unsafe references.
-
-As an example,
-\begin{center}
-\begin{tabular}{lll}
-\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
-& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = \{\} \; \mathbf{in}$}\\
-& & \multicolumn{1}{l}{$b = a;$}\\
-& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
-& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
-\end{tabular}
-\end{center}
-does not type check because we cannot add the field $``x"$ to $b$,
-as its type is \emph{closed}.
-Aliasing $a$ changes the type of $a$ from $\emph{unique}$ to
-$\emph{open}$, and that is the reason why we can add the field
-$``x"$ to the type of $a$.
-Aliasing $a$ also produces the type $\{\}_{closed}$, which is
-the same type that $b$ has in left-hand side of the assignment.
-After the assignment, the type of $b$ is \emph{closed} and thus
-does not allow changing the value that is stored in the field
-$``x"$ of $a$.
-
-Rule \textsc{T-ID-L} defines this behavior:
-\[
-\begin{array}{c}
-\mylabel{T-ID-L}\\
-\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = write(T_{1})}
-      {\env_{1}, \penv \vdash id_{l}:close(T_{2}), \env_{1}[id \mapsto close(T_{2})]}
-\end{array}
-\]
-
-This rule uses the auxiliary function \emph{write} because it may be
-accessing an identifier that is bound to a filter type.
-In Typed Lua, assignments restore discriminated union types to their original types,
-and function \emph{write} works in this purpose.
-We can define \emph{write} as follows:
-\begin{align*}
-write(\phi(T_{1},T_{2})) & = T_{1}\\
-write(T) & = T
-\end{align*}
-
-Our type system also has different rules for type checking table indexing to avoid
-changing table types in these operations, as they cannot create aliases:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-INDEX-ID}\\
-\dfrac{\begin{array}{c}
-       \env_{1}(id) = T \;\;\;
-       read(\penv, T) = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}\\
-       \env_{1}, \penv \vdash e_{2}:K, \env_{2} \;\;\;
-       \exists i \in 1{..}n \; K \lesssim K_{i}
-       \end{array}}
-      {\env_{1}, \penv \vdash id[e_{2}]:V_{i}, \env_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-INDEX-EXP}\\
-\dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash e_{1}:\{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}, \env_{2}\\
-       \env_{2}, \penv \vdash e_{2}:K, \env_{3} \;\;\;
-       \exists i \in 1{..}n \; K \lesssim K_{i}
-       \end{array}}
-      {\env_{1}, \penv \vdash e_{1}[e_{2}]:V_{i}, \env_{3}}
-\end{array}
-\end{array}
-\]
-
-A second form of refinement happens when we want to use an
-\emph{unique} or \emph{open} table type in a context that expects a
-\emph{fixed} or \emph{closed} table type with a different shape.
-This kind of refinement allows programmers to add optional fields
-or merge existing fields.
-To do that, Typed Lua includes a type coercion expression ${<}T{>} \; id$.
-For instance, we can use this type coercion expression to make the following
-example type check:
+
+This example typechecks; after the initialization of $b$
+the type of $a$ is {\em open} instead of {\em unique}, but
+this still lets the assignment add a new field to $a$.
+If we changed the tag of the type annotation of $b$ to
+any other tag the example would not typecheck anymore,
+as the type of expression $a$ is $\{\}_{closed}$.
+
+If we did not change the tag on the type produced by $a$
+we could break type safety, as we could declare $b$ with
+tag {\em unique} or {\em open} and add fields that would
+conflict with the type of the fields on $a$, but at
+runtime both reference the same table.
+
+It is not a problem for an {\em unique} or {\em open}
+table to appear as an l-value, because {\sc T-IdRead} is
+sufficient to forbid the creation of an alias to another
+{\em unique} or {\em open} table. For this reason,
+identifiers that appear as l-values have their own rule
+{\sc T-IdWrite}.
+
+Another use of the identifier that does not create an
+alias is when indexing the table. Rule {\sc T-IndexId}
+covers this case. For comparison, rule {\sc T-Index}
+applies for indexing expressions where the expression
+denoting the table is not an identifier.
+
+A second way of changing a table type is by a 
+{\em coercion expression}. Coercion expressions let
+a {\em unique} or {\em open} table be used in a context
+that expects a table type with a different shape.
+Coercion expressions always create an alias, so
+the type produced by the coercion is always
+{\em closed} or {\em fixed}. If the coercion is to a
+{\em closed} table type the type of the table changes
+to an {\em open} type with the same shape, but if
+the coercion is to a {\em fixed} type the table has
+to assume the same type.
+
+Rules {\sc T-CoerceClosed} and {\sc T-CoerceFixed} type
+coercion expressions ${<}F{>} \; id$.
+As with coercion l-values, the Typed Lua compiler automatically
+infers the type $F$, so to the programmer a coercion
+expression looks just like an ordinary use of the table.
+Notice that the type of $id$ after {\sc T-CoerceClosed} will be an {\em open} type
+only if $F$ is a plain {\em closed} table type.
+Coercion is only possible towards a supertype, although
+we have seen in Section~\ref{sec:subtyping} that the lack
+of aliases makes the subtyping relation of {\em unique}
+very flexible.
+
+In the example below, we use a coercion expression to
+widen the type of a field in a {\em unique}
+table, and then add another field, to show that the
+table is now {\em open}:
 \begin{center}
@@
 & \multicolumn{3}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{closed} =$}\\
-& & \multicolumn{2}{l}{${<}\{``x":\String, ``y":\String \cup \Nil\}_{open}{>} \; a \; \mathbf{in}$}\\
+& & \multicolumn{2}{l}{${<}\{``x":\String, ``y":\String \cup \Nil\}_{closed}{>} \; a \; \mathbf{in}$}\\
 & & & \multicolumn{1}{l}{$a[``z"] \; {<}\Integer{>} = 1$}
@@
 
-We can use $a$ to initialize $b$ because the coercion converts
-the type of $a$ from $\{``x":\String, ``y":\String\}_{unique}$ to
-$\{``x":\String, ``y":\String \cup \Nil\}_{open}$, and results in
-$\{``x":\String, ``y":\String \cup \Nil\}_{closed}$,
-which is a subtype of
-$\{``x":\String, ``y":\String \cup \Nil\}_{closed}$, the type of $b$.
-We can continue to refine the type of $a$ after aliasing it to $b$,
-as it still holds an \emph{open} table.
-At the end of this example, $a$ has type
-$\{``x":\String, ``y":\String \cup \Nil, ``z":\Integer\}_{open}$.
-
-Rule \textsc{T-COERCE} defines the behavior of the type coercion expression:
-\[
-\begin{array}{c}
-\mylabel{T-COERCE}\\
-\dfrac{\env_{1}(id) \subtype T \;\;\;
-       \env_{1}[id \mapsto T], \penv \vdash id:T_{1}, \env_{2}}
-      {\env_{1}, \penv \vdash {<}T{>} \; id:T_{1}, \env_{2}}
-\end{array}
-\]
-
-Note that rule \textsc{T-COERCE} only allows changing the type
-of a variable if the new type is a supertype of the previous type,
-and the resulting type is always \emph{fixed} or \emph{closed}
-to prevent the creation of unsafe aliases.
+The type $a$ has by the time of the inner {\bf local} block
+is $\{``x":\String, ``y":\String\}_{unique}$. This type is
+a subtype of the $\{``x":\String, ``y":\String\cup\Nil\}_{closed}$, as {\em unique} tables
+have depth subtyping even in non-{\bf const} fields.
+The coercion expression and the initialization of $b$
+typecheck, and the type of $a$ changes to
+$\{``x":\String, ``y":\String\cup\Nil\}_{open}$.
+As the type is {\em open}, we can continue adding fields
+to it, but are now restricted on future coercions. For
+example, we will not be able to widen the type of $``x"$
+to $\String \cup \Nil$, as that could break type safety
+(by writing $\Nil$ to ``x" using reference $a$, and reading
+``x" using the alias $b$).
 
@@
 		& & & \multicolumn{1}{l}{$a = \{[``x"]=5\} $}\\
-		& \multicolumn{3}{l}{$\mathbf{in} \; g({<}\{``x":
-			 \String \cup \Nil\}{>} \; a)$}
+		& \multicolumn{3}{l}{$\mathbf{in} \; \lfloor g({<}\{``x":
+			 \String \cup \Nil\}{>} \; a) \rfloor_{0}$}
 	\end{tabular}
@@
 
-The type of $a$ on line 3 is $\{\}_{closed}$, so the assignment
+The type of $a$ on line 4 is $\{\}_{closed}$, so the assignment
 would type-check. But then the function call on the last line
@@
 
-Rule {\sc T-FUNCTION} types non-variadic function declarations,
-and shows how these auxiliary functions work together:
-\[
-\begin{array}{c}
-\mylabel{T-FUNCTION}\\
-\dfrac{\begin{array}{c}
-	closeall(\env_{1})[\overline{id \mapsto T}], \penv[\ret \mapsto S] \vdash s, \env_{2}\\
-	\env_{3} = openset(\env_{1}, frv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)) \\
-	\env_{4} = closeset(\env_{3}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
-	\end{array}}
-{\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{3}}
-\end{array}
-%\begin{array}{c}
-%\mylabel{T-FUNCTION}\\
-%\dfrac{\begin{array}{c}
-%	crall(\env_{1}[\overline{id} \mapsto \overline{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}\\
-%	\env_{3} = crset(\env_{1}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))\\
-%	\env_{4} = openset(\env_{3}, rv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
-%	\end{array}}
-%{\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{4}}
-%\end{array}
-\]
-
+Rule {\sc T-Function} types non-variadic function declarations,
+and shows how these auxiliary functions work together.
 Notice that the environment $\env_{2}$ resulting from
-typing the function body is just ignored.
-
-The following example typechecks, and uses all the parts of {\sc T-FUNCTION}:
+typing the function body is just ignored. The rule for
+declaring variadic functions and recursive functions are
+similar. Functions $closeall$, $closeset$, $openset$, $fav$,
+and $frv$ are straightforward, and were omitted due to lack of space.
+
+The following example typechecks, and uses all the parts of {\sc T-Function}:
 \begin{center}
@@
 & & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
-& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; f(a[``x"])$}\\
-%& & \multicolumn{2}{l}{$b[``x"] \; {<}\String{>} = ``foo" \;;\;
-%	 f(a[``x"])$}
+& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; \lfloor f(a[``x"]) \rfloor_0$}\\
 \end{tabular}
@@
 proj(F{*}, i) & = & F \cup \Nil \\
-proj(T \times P, 1) & = & T \\
-proj(T \times P, i) & = & proj(P, i-1)
+proj(F \times P, 1) & = & F \\
+proj(F \times P, i) & = & proj(P, i-1)
 \end{array}\\\\
@@
 \begin{array}{rcll}
-fonil(F_1 \cup F_2) & = & fonil(F_1) \cup fonil(F_2) \\
+fonil(F_1 \cup F_2) & = & fonil(F_1) & \mathrm{if}\;fonil(F_2) = {\bf void}\\
+fonil(F_1 \cup F_2) & = & fonil(F_2) & \mathrm{if}\;fonil(F_1) = {\bf void}\\
+fonil(F_1 \cup F_2) & = & fonil(F_1) \cup fonil(F_2) & \mathrm{otherwise}\\
+%\multicolumn{4}{c}{\mathrm{the\;next\;two\;cases\;apply\;if\;}F
+%	\;\mathrm{is\;not\;a\;union}}\\
 fonil(F) & = & {\bf void} & \mathrm{if} \; F \subtype \Nil \; \mathrm{and} \; \Nil \subtype F\\
@@
 environment used for typing the {\bf if} statement.
-The first judgement above the bar of rule {\sc T-LocalProj} uses the auxiliary typing relation $\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$.
+The first judgment above the bar of rule {\sc T-LocalProj} uses the auxiliary typing relation $\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$.
 
@@
 
-Rule \textsc{T-IfFilterNilProj} uses the auxiliary functions \emph{fonilpt} ({\em filter-out nil in projection type}) and \emph{finilpt} ({\em filter-in nil in projection type}) to
+Rule \textsc{T-IfFilterProjBoth} uses the auxiliary functions \emph{fonilpt} ({\em filter-out nil in projection type}) and \emph{finilpt} ({\em filter-in nil in projection type}) to
 filter the union referenced by $\pi_{i}^{x}$, respectively
@@
 
+Rule {\sc T-IfFilterProjBoth} also uses the auxiliary function $join$ defined in Figure~\ref{fig:funcs} to try to combine the environments that result from typing the two branches.
+In a $join$ only one environment is guaranteed to be from code that has been executed.
+
 In our example, $x$ would be bound to $\Integer \times \Integer \times \Nil{*}$ in the {\bf then} branch, and would be bound
@@
 Rules {\sc T-IfFilterProjThen} and {\sc T-IfFilterProjElse}
-respectively cover the case where the {\bf else} branch is unreachable and the case where the {\bf then} branch is unreacheable, because either the projected type of $\pi_i^x$
+respectively cover the case where the {\bf else} branch is unreachable and the case where the {\bf then} branch is unreachable, because either the projected type of $\pi_i^x$
 is not a supertype of $\Nil$ or is $\Nil$.
@@
 Unlike some optional type systems, such as the type systems
-of the Typescript and Dart languages, we designed Typed Lua without deliberately unsound parts.
+of the TypeScript and Dart languages, we designed Typed Lua without deliberately unsound parts.
 However, we still do not have proofs that the novel features of

### 2015-06-14 doc/papers/paper/paper.tex
DLS paper: killed an overfull hbox

@@
 Typed Lua's type system uses subtyping to order
-types and  consistent-subtyping~\cite{siek2007objects,siek2013mutable}
+its types and  consistent-subtyping~\cite{siek2007objects,siek2013mutable}
 to model compatibility between types in a way that allows

### 2015-06-14 doc/papers/paper/paper.tex
DLS paper: note that not all rules are shown (RC2)
DLS: formatting (RC3)
DLS paper: fixing typos

@@
 static typing a matter of describing the implied types that
-it it using and adding annotations to make those types explicit.
+it is using and adding annotations to make those types explicit.
 Designing an optional type system for
@@
 
-Typed Lua~\cite{maidl2014tl} is an optional type system for
+Typed Lua~\cite{maidl2014tl,maidl2015tl} is an optional type system for
 Lua that is rich enough to preserve some of the idioms
@@
 tag is being destroyed.
-This rule allows with subtyping, depth subtyping on any field, not only immutable fields, as well as the addition of new fields and the merging of existing fields, as long as the types of the values of new and merged fields include $\Nil$.
+This rule allows width subtyping, depth subtyping on any field, not only immutable fields, as well as the addition of new fields and the merging of existing fields, as long as the types of the values of new and merged fields include $\Nil$.
 
 The rationale for the permissiveness of rule {\sc S-TableUtoC} derives from the fact that reading a non-existing key just yields $\Nil$ instead of being an error in Lua semantics, plus the fact
-that the type system guarantees that a table with a
+that the type system guarantees that a table with an
 {\em unique} type has no keys with a value that does not
@@
 $S_1 \rightarrow S_2$ and $S_3 \rightarrow S_4$ are consistent
-if and only if $S_1 \rightarrow S_3$ and $S_2 \rightarrow S_4$.
+if and only if $S_1 \sim S_3$ and $S_2 \sim S_4$.
 Two table types are consistent if and only if they have the same key types, and the value types of each key type are consistent.
@@
 and a projection environment $\penv$ we can prove that an expression $e$ has type $T$ and produces a new type environment $\env_{2}$.
+Only the most relevant rules for the features that we are focusing on are shown.
 
@@
 \begin{figure*}
+	{\small
 	\[
@@
 	wf(S_1 \rightarrow S_2) & = & wf(S_1) \,\wedge\, wf(S_2) \\
-	wf(S_1 \sqcup S_2) & = & wf(S_1) \,\wedge\, S_2\\
+	wf(S_1 \sqcup S_2) & = & wf(S_1) \,\wedge\, wf(S_2)\\
 	wf(F{*}) &=& wf(F) \\
@@
 	\end{array}
-	\]
+	\]}
 	\caption{Auxiliary Functions}
@@
 Open tables have a rule {\sc T-NewFieldOpen} that is not shown
-in Figure~\ref{fig:tablerules} due to space. The only thing that
+in Figure~\ref{fig:tablerules} due to brevity. The only thing that
 changes from one rule to the other is the tag in the type of $id$.
@@
 depend on respectively having no aliases and having only {\em closed} aliases. The system conservatively assumes that
-a table has been aliased it its identifier appears in r-value position, with a couple of exceptions which we will explain shortly.
+a table has been aliased if its identifier appears in r-value position, with a couple of exceptions which we will explain shortly.
 
@@
 \multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
-& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{closed} = a \; \mathbf{in}a[``x"] \; {<}\String{>} = ``foo";$}
+& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{closed} = a \; \mathbf{in}$}\\
+& & $\;a[``x"] \; {<}\String{>} = ``foo";$
 \end{tabular}
@@
 to $\String \cup \Nil$, as that could break type safety
-(by writing $\Nil$ to ``x" using reference $a$, and reading
-``x" using the alias $b$).
+(by writing $\Nil$ to $``x"$ using reference $a$, and reading
+$``x"$ using the alias $b$).
 
@@
 Closed-over variables that are not assigned still may
-change their type if they have a {\em unique} table type
+change their type if they have an {\em unique} table type
 and are aliased inside the function body. We use the auxiliary
@@
 this set, and then function $openset$ changes any variables
-with a {\em unique} table type in the environment to {\em open}.
+with an {\em unique} table type in the environment to {\em open}.
 
@@
 similar. Functions $closeall$, $closeset$, $openset$, $fav$,
-and $frv$ are straightforward, and were omitted due to lack of space.
+and $frv$ are straightforward, and were omitted for brevity.
 
@@
 to $openset$. Being $\{\}_{open}$, adding a new field
-is allowed, changing the type of $a$ to $\{``x": \Integer\}_{open}$ after the semicolon. After the call to
+is allowed, changing the type of $a$ to $\{``x": \Integer\}_{open}$ after the last semicolon. After the call to
 $f$, the variable $b$ is an alias to $a$ but this is not
@@
 A standard way of decomposing the union in the assignment
-would let the type of {\tt q} be $\Integer \times \Nil$ and
-the type of {\tt r} be $\Integer \times \String$. Flow typing
-would narrow the type of {\tt q} to $\Integer$ in the {\tt then}
-branch and to $\Nil$ in the {\tt else} branch, but would
-leave the type of {\tt r} unchanged.
+would let the type of $q$be $\Integer \times \Nil$ and
+the type of $r$ be $\Integer \times \String$. Flow typing
+would narrow the type of $q$ to $\Integer$ in the {\bf then}
+branch and to $\Nil$ in the {\bf else} branch, but would
+leave the type of $r$ unchanged.
 
@@
 are a general way to make the dependency between the types
-of {\tt q} and {\tt r} survive the decomposition of the tuple,
+of $q$ and $r$ survive the decomposition of the tuple,
 so flow typing can narrow all of the components by testing
@@
 Several dynamically typed languages now have optional
-type systems: Clojure~\cite{bonnaire-sergeant2012typed-clojure},
+type systems: Clojure~\cite{bonnaire-sergeant2015typed-clojure},
 JavaScript~\cite{typescript}, Python~\cite{mypy}, 
@@
 also exposed several limitations.
-We found that the three main limitations are
-the lack of full support for overloaded functions, the lack
-of parametric polymorphism, and operator overloading.
+We found that the three main limitations are the lack of
+full support for overloaded functions, parametric polymorphism,
+and operator overloading.
 Overcoming these limitations is one major target for future work,
@@
 
-Typed Lua has a working implementation\footnote{Available
-at \url{http://github.com/andremm/typelua}.} that
-Lua programmers can already use as a framework to
+Typed Lua has a working implementation\footnote{Available at
+\url{https://github.com/andremm/typedlua}.}
+that Lua programmers can already use as a framework to
 document, test, and structure their applications.
@@
 type checking is unfeasible in the current state
-of the type system, or too much work, can use Typed
-Lua to document the external interfaces of its
-libraries, giving the benefits of static typing
+of the type system, or too much work,
+Lua programmers can still use Typed Lua to document the
+external interfaces of their libraries, giving the benefits of static typing
 checking to the users of those libraries.

### 2015-06-14 doc/papers/paper/paper.tex
DLS paper: some last-minute fixes and additions

@@
 & | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
-& | \; \{\overline{K:V}\}_{unique|open|fixed|closed} & \textit{table types}\\
+& | \; \{\overline{F:V}\}_{unique|open|fixed|closed} & \textit{table types}\\
 & |\; x & \textit{type variables}\\
@@
 & | \; T \times E & \\
-K ::= & & \textsc{{\small key types:}}\\
-& \;\; L \; | \; B \; | \; \Value &\\
+%K ::= & & \textsc{{\small key types:}}\\
+%& \;\; L \; | \; B \; | \; \Value &\\
 %\multicolumn{3}{c}{}\\
@@
 Literal types represent the type of literal boolean, number
-and string values. Their main application is the formation in record and tuple types.
+and string values. Having just one value each, their main application is in constructing table types with fields where it is guaranteed that the value of that field will not be {\tt nil}.
 
@@
 guaranteed that every key belongs to one of the {\em key
-types} $K_i$. We restrict key types to literal types,
-base types, and the top type to make the system simpler
-for the user, as our survey of Lua programs showed that
-this restriction reflects the way Lua tables are commonly used.
+types} $F_i$. 
+%We restrict key types to literal types,
+%base types, and the top type to make the system simpler
+%for the user, as our survey of Lua programs showed that
+%this restriction reflects the way Lua tables are commonly used.
 
@@
 	\forall j \; \exists i \;\;\;
-	\senv \vdash K_{i} \subtype K_{j}' \;\
-	\senv \vdash K_{j}' \subtype K_{i} \;
+	\senv \vdash F_{i} \subtype F_{j}' \;\
+	\senv \vdash F_{j}' \subtype F_{i} \;
 	\senv \vdash V_{i} \subtype_{c} V_{j}'
@@
 {\begin{array}{c}
-	\senv \vdash \{\overline{K:V}\}_{fixed|closed} \subtype
-	\{\overline{K^\prime:V^\prime}\}_{closed}
+	\senv \vdash \{\overline{F:V}\}_{fixed|closed} \subtype
+	\{\overline{F^\prime:V^\prime}\}_{closed}
 	\end{array}}
@@
 	\forall i \; \forall j \;\;\;
-	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{u} V_{j}' \\
+	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash V_{i} \subtype_{u} V_{j}' \\
 	\forall j  \; \nexists i \;\;\;
-	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
-	\end{array}}
-{\senv \vdash \{\overline{K:V}\}_{unique} \subtype
-	\{\overline{K^\prime:V^\prime}\}_{closed}}
+	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
+	\end{array}}
+{\senv \vdash \{\overline{F:V}\}_{unique} \subtype
+	\{\overline{F^\prime:V^\prime}\}_{closed}}
 \end{array}\\\\
@@
 	\forall i \; \forall j \;\;\;
-	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
+	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
 	\forall j \; \nexists i \;\;\;
-	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
-	\end{array}}
-{\senv \vdash \{\overline{K:V}\}_{open} \subtype
-	\{\overline{K^\prime:V^\prime}\}_{closed}}
+	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
+	\end{array}}
+{\senv \vdash \{\overline{F:V}\}_{open} \subtype
+	\{\overline{F^\prime:V^\prime}\}_{closed}}
 \end{array}\\\\
@@
 	\forall i \; \exists j \;\;\;
-	\senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
+	\senv \vdash F_{i} \subtype F_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
 	\forall j  \; \nexists i \;\;\;
-	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
+	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
 	\end{array}}
 {\begin{array}{c}
-	\senv \vdash \{\overline{K:V}\}_{unique} \subtype
-	\{\overline{K^\prime:V^\prime}\}_{unique|open|fixed}
+	\senv \vdash \{\overline{F:V}\}_{unique} \subtype
+	\{\overline{F^\prime:V^\prime}\}_{unique|open|fixed}
 	\end{array}}
@@
 	\forall i \; \exists j \;\;\;
-	\senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
+	\senv \vdash F_{i} \subtype F_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
 	\forall j \; \nexists i \;\;\;
-	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
+	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
 	\end{array}}
 {\begin{array}{c}
-	\senv \vdash \{\overline{K:V}\}_{open} \subtype
-	\{\overline{K^\prime:V^\prime}\}_{open|fixed}
+	\senv \vdash \{\overline{F:V}\}_{open} \subtype
+	\{\overline{F^\prime:V^\prime}\}_{open|fixed}
 	\end{array}}
@@
 	\forall j \; \exists i \;\;\;
-	\senv \vdash K_{i} \subtype K_{j}' \;
-	\senv \vdash K_{j}' \subtype K_{i} \;
+	\senv \vdash F_{i} \subtype F_{j}' \;
+	\senv \vdash F_{j}' \subtype F_{i} \;
 	\senv \vdash V_{i} \subtype_{c} V_{j}'\\
 	\forall i \; \exists j \;\;\;
-	\senv \vdash K_{i} \subtype K_{j}' \;
-	\senv \vdash K_{j}' \subtype K_{i} \;
+	\senv \vdash F_{i} \subtype F_{j}' \;
+	\senv \vdash F_{j}' \subtype F_{i} \;
 	\senv \vdash V_{i} \subtype_{c} V_{j}'\\
 	\end{array}}
-{\senv \vdash \{\overline{K:V}\}_{fixed} \subtype
-	\{\overline{K^\prime:V^\prime}\}_{fixed}}
+{\senv \vdash \{\overline{F:V}\}_{fixed} \subtype
+	\{\overline{F^\prime:V^\prime}\}_{fixed}}
 \end{array}
@@
 	& \;\; id_{l} & \textit{variable assignment}\\
-	& | \; e_{1}[e_{2}] & \textit{table assignment}\\
+	& | \; e_{1}[e_{2}]_{l} & \textit{table assignment}\\
 	& | \; id[k] \; {<}V{>} & \textit{type coercion}\\
@@
 \begin{figure*}[t]
+{\small
 \[
@@
 \dfrac{\begin{array}{c}
-\overline{\env_1, \penv \vdash e_2: F, \env_v}
+	\overline{\env_v, \penv \vdash e_1: F_1, \env_f}
 	\;\;\;
-	\overline{\env_v, \penv \vdash e_1: K, \env_f}
-	\\
-	T = \{\overline{K{:}fix(F)}\}_{unique} \;\;\;
+\overline{\env_1, \penv \vdash e_2: F_2, \env_v}	\\
+	T = \{\overline{F_1{:}vt(F_1,F_2)}\}_{unique} \;\;\;
 	\env_{2} = merge(\overline{\env_{f}}) \;\;\; wf(T)
@@
 \dfrac{\begin{array}{c}
-	\env_{1}(id) = \{ \overline{K:V} \}_{unique}\\
-	\env_{1}, \penv \vdash k:K^{new}, \env_{2} \;\;\;
-	\overline{K^{new} \not\lesssim K} \;\;\;
-	V^{new} = fix(V)
+	\env_{1}(id) = \{ \overline{F:V} \}_{unique}\\
+	\env_{1}, \penv \vdash k:F^{new}, \env_{2} \;\;\;
+	\overline{F^{new} \not\lesssim F} \;\;\;
+	V^{new} = vt(F^{new}, V)
 	\end{array}}
@@
 	\env_{1}, \penv \vdash id[k] {<}V{>}:V^{new},
-	\env_{2}[id \mapsto \{ \overline{K{:}V}, K^{new}{:}V^{new}\}_{unique}]
+	\env_{2}[id \mapsto \{ \overline{F{:}V}, F^{new}{:}V^{new}\}_{unique}]
 	\end{array}}
@@
 \begin{array}{c}
-\mylabel{T-IndexId}\\
+\mylabel{T-IndexIdRead}\\
 \dfrac{\begin{array}{c}
-	\env_{1}(id) = \{\overline{K:V}\} \;\;\;
-	\env_{1}, \penv \vdash e_{2}:K^e, \env_{2} \;\;\;
-	 K^e \lesssim K_{i}
-	\end{array}}
-{\env_{1}, \penv \vdash id[e_{2}]:V_{i}, \env_{2}}
+	\env_{1}(id) = \{\overline{F:V}\} \;\;\;
+	\env_{1}, \penv \vdash e_{2}:F^e, \env_{2} \;\;\;
+	 F^e \lesssim F_{i}
+	\end{array}}
+{\env_{1}, \penv \vdash id[e_{2}]:rconst(V_{i}), \env_{2}}
 \end{array}
@@
 \begin{array}{c}
-\mylabel{T-IndexExp}\\
+\mylabel{T-IndexExpRead}\\
 \dfrac{\begin{array}{c}
-	\env_{1}, \penv \vdash e_{1}:\{\overline{K:V}\}, \env_{2}\\
-	\env_{2}, \penv \vdash e_{2}:K^e, \env_{3} \;\;\;
-	 K^e \lesssim K_{i}
-	\end{array}}
-{\env_{1}, \penv \vdash e_{1}[e_{2}]:V_{i}, \env_{3}}
+	\env_{1}, \penv \vdash e_{1}:\{\overline{F:V}\}, \env_{2}\\
+	\env_{2}, \penv \vdash e_{2}:F^e, \env_{3} \;\;\;
+	 F^e \lesssim F_{i}
+	\end{array}}
+{\env_{1}, \penv \vdash e_{1}[e_{2}]:rconst(V_{i}), \env_{3}}
 \end{array}
@@
 {\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto F]}
-\end{array}\\
+\end{array}\\\\
 \begin{array}{c}
@@
 \end{array}
+\begin{array}{c}
+\mylabel{T-Local}\\
+\dfrac{\begin{array}{c}
+	\env_{1}, \penv \vdash el:E, \env_{2} \\
+	E \lesssim F_1 \times \ldots \times F_n \times \Value{*} \\
+	\env_{2}[\overline{id \mapsto F}], \penv \vdash s, \env_{3}
+	\end{array}}
+{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id{:}F} = el \; \mathbf{in} \; s, (\env_{3} - \{\overline{id}\})[\overline{id \mapsto \env_2(id)}]}
+\end{array}
 \end{array}
 \]
-\caption{Rules for Table Types}
+}
+\caption{Typing Rules}
 \label{fig:tablerules}
@@
 	\begin{array}{rcl}
-	wf(\{\overline{K:V}\}_{unique|open|fixed|closed}) & = & \forall i \; ((\nexists j \; i \not= j \,\wedge\, K_{i} \lesssim K_{j}) \,\wedge\, wf(V_{i}) \,\wedge\\
+	wf(\{\overline{F:V}\}_{unique|open|fixed|closed}) & = & \forall i \; ((\nexists j \; i \not= j \,\wedge\, F_{i} \lesssim F_{j}) \,\wedge\, wf(V_{i}) \,\wedge\\
 	& & \;\;\;\; \lnot tag(V_{i},unique) \,\wedge\, \lnot tag(V_{i},open))\\
@@
 	tag(F_1 \cup F_2, t) & = & tag(F_1, t) \,\vee\, tag(F_2,t) \\
-	tag(\{\overline{K:V}\}_t, t) & = & \top\\
+	tag(\{\overline{F:V}\}_t, t) & = & \top\\
 	\multicolumn{3}{c}{
@@
 \\\\
-	fix(\{\overline{K:V}\}_{unique|open}) & = &
-	\{\overline{K:V}\}_{fixed}\\
+    vt(L, V) & = & fix(V) \\
+    vt(F_1, F_2) & = & nil(fix(F_2))\\
+    vt(F_1, {\bf const}\;F_2) & = & {\bf const}\;nil(fix(F_2))\\\\
+    nil(F) & = & F \;\;\; \mathrm{if}\;\Nil \lesssim F\\
+    nil(F) & = & F \cup \Nil \;\;\; \mathrm{otherwise}\\\\
+	fix(\{\overline{F:V}\}_{unique|open}) & = &
+	\{\overline{F:V}\}_{fixed}\\
 	\multicolumn{3}{c}{\;\;\;\;\;\;\;\mathrm{other\;cases\;by\;straightforward\;structural\;recursion}}\\\\
@@
 	merge2(\env_1,\env_2) & = & \{\overline{(id,merget(\env_1(id),\env_2(id))}\}\\
-	merget(T_1, T_2) & = & T_1 \;\;\; \mathrm{if\;} T_2 \subtype T_1\\
-	merget(T_1, T_2) & = & T_2 \;\;\; \mathrm{if\;} T_1 \subtype T_2\\
-	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;\forall i\;V^l_i\subtype_u V^r_i \,\vee\,V^r_i\subtype_u V^l_i \;\mathrm{and\;the\;right\; side\; is\;}wf}\\
-	merget(\{\overline{K:V^l},\overline{K^{\prime}:V^\prime}\}_{unique},
-	\{\overline{K:V^r},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{unique}) &=&
-	\{\overline{K:sup_u(V^l,V^r)},\overline{K^\prime:V^\prime},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{unique}
+	merget(T_1, T_2) & = & T_1 \;\;\; \mathrm{if\;} T_2 \lesssim T_1\\
+	merget(T_1, T_2) & = & T_2 \;\;\; \mathrm{if\;} T_1 \lesssim T_2\\
+	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;
+		\overline{V^l \lesssim_u V_r \,\vee\, V^r \lesssim_u V_l}\;\mathrm{and\;the\;right\; side\; is\;}wf}\\
+	merget(\{\overline{F:V^l},\overline{F^{\prime}:V^\prime}\}_{unique},
+	\{\overline{F:V^r},\overline{F^{\prime\prime}:V^{\prime\prime}}\}_{unique}) &=&
+	\{\overline{F:sup_u(V^l,V^r)},\overline{F^\prime:V^\prime},\overline{F^{\prime\prime}:V^{\prime\prime}}\}_{unique}
 	\\
-	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;\forall i\;V^l_i\subtype_c V^r_i \,\vee\,V^r_i\subtype_c V^l_i
-		\;\mathrm{and\;the\;right\; side\; is\;}wf}\\
-	merget(\{\overline{K:V^l},\overline{K^{\prime}:V^{\prime}}\}_{unique|open},
-	\{\overline{K:V^r},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{unique|open}) &=&
-	\{\overline{K:sup_c(V^l,V^r)},\overline{K^\prime:V^\prime},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{open}
+	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;
+		\overline{V^l \lesssim_c V_r \,\vee\, V^r \lesssim_c V_l}\;\mathrm{and\;the\;right\; side\; is\;}wf}\\
+	merget(\{\overline{F:V^l},\overline{F^{\prime}:V^{\prime}}\}_{unique|open},
+	\{\overline{F:V^r},\overline{F^{\prime\prime}:V^{\prime\prime}}\}_{unique|open}) &=&
+	\{\overline{F:sup_c(V^l,V^r)},\overline{F^\prime:V^\prime},\overline{F^{\prime\prime}:V^{\prime\prime}}\}_{open}
 	\\
@@
 	join(\env_1,\env_2) & = & \{\overline{(id,joint(\env_1(id),\env_2(id))}\}\\
-	joint(T_1, T_2) & = & T_1 \;\;\; \mathrm{if\;} T_2 \subtype T_1\\
-	joint(T_1, T_2) & = & T_2 \;\;\; \mathrm{if\;} T_1 \subtype T_2\\
-	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;\forall i\;V^l_i\subtype_u V^r_i \,\vee\,V^r_i\subtype_u V^l_i
-		\;\mathrm{and\;the\;right\; side\; is\;}wf}\\
-	joint(\{\overline{K:V^l},\overline{K^{\prime}:V^{\prime}}\}_{unique},
-	\{\overline{K:V^r},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{unique}) &=&
-	\{\overline{K:sup_u(V^l,V^r)},\overline{K^\prime:V^\prime\cup\Nil},\overline{K^{\prime\prime}:V^{\prime\prime}\cup\Nil}\}_{unique}\\
+	joint(T_1, T_2) & = & T_1 \;\;\; \mathrm{if\;} T_2 \lesssim T_1\\
+	joint(T_1, T_2) & = & T_2 \;\;\; \mathrm{if\;} T_1 \lesssim T_2\\
+	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;
+		\overline{V^l \lesssim_u V_r \,\vee\, V^r \lesssim_u V_l}\;\mathrm{and\;the\;right\; side\; is\;}wf}\\
+	joint(\{\overline{F:V^l},\overline{F^{\prime}:V^{\prime}}\}_{unique},
+	\{\overline{F:V^r},\overline{F^{\prime\prime}:V^{\prime\prime}}\}_{unique}) &=&
+	\{\overline{F:sup_u(V^l,V^r)},\overline{F^\prime:nil(V^\prime)},\overline{F^{\prime\prime}:nil(V^{\prime\prime})}\}_{unique}\\
 	joint(T_1, T_2) & = & \bot \;\;\; \mathrm{otherwise}\\\\
-	sup_u(V_1, V_2) & = & V_2 \;\;\; \mathrm{if}\; V_1 \subtype_u V_2\\
-	sup_u(V_1, V_2) & = & V_1 \;\;\; \mathrm{if}\; V_2 \subtype_u V_1\\\\
-	reopen(\{\overline{K:V}\}_{closed}) & = & \{\overline{K:V}\}_{open} \\
+	sup_u(V_1, V_2) & = & V_2 \;\;\; \mathrm{if}\; V_1 \lesssim_u V_2\\
+	sup_u(V_1, V_2) & = & V_1 \;\;\; \mathrm{if}\; V_2 \lesssim_u V_1\\\\
+	reopen(\{\overline{F:V}\}_{closed}) & = & \{\overline{F:V}\}_{open} \\
 	reopen(F) & = & F
@@
 track aliases of table fields, so this restriction is
-necessary.
-This limitation of the type system is the reason the {\sc T-Constructor} rule uses the auxiliary function $fix$ to change any {\em open} and {\em unique} table types used in the field initializer to {\em fixed}.
-As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
+necessary. This limitation of the type system is the reason the {\sc T-Constructor} rule uses the auxiliary function $fix$
+(in the definition of $vt$)
+ to change any {\em open} and {\em unique} table types used in the field initializer to {\em fixed}. As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
 has type $\{``x":1, ``y":\{``z":2\}_{fixed}\}_{unique}$.
@@
 changes from one rule to the other is the tag in the type of $id$.
+There are analogous rules for adding methods to a {\em unique}
+and {\em open tables} as a side-effect of typechecking a method declaration.
 
@@
 \multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
-& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{closed} = a \; \mathbf{in}$}\\
-& & $\;a[``x"] \; {<}\String{>} = ``foo";$
+& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{closed} = a \; \mathbf{in}\;\;a[``x"] \; {<}\String{>} = ``foo";$}
 \end{tabular}
@@
 Another use of the identifier that does not create an
-alias is when indexing the table. Rule {\sc T-IndexId}
-covers this case. For comparison, rule {\sc T-Index}
+alias is when indexing the table. Rule {\sc T-IndexIdRead}
+covers this case (function $rconst$ just strips the {\bf const} qualifier from the value type, if present). For comparison, rule {\sc T-IndexExpRead}
 applies for indexing expressions where the expression
-denoting the table is not an identifier.
+denoting the table is not an identifier. The rules for
+indexing expressions in l-value position are similar,
+except to make sure that the field is not {\bf const} .
 
@@
 \begin{figure*}[t]
+{\small
 \label{fig:projtypes}
@@
 proj(S_1 \sqcup S_2, i) & = & proj(S_1, i) \cup proj(S_2, i) \\
-proj(F{*}, i) & = & F \cup \Nil \\
+proj(F{*}, i) & = & nil(F) \\
 proj(F \times P, 1) & = & F \\
@@
 \end{array}
-\]
+\]}
 \caption{Rules for Projection Types}

### 2015-06-14 doc/papers/paper/paper.tex
DLS paper: new version (RC5)

@@
 T ::= & & \textsc{{\small expression types:}}\\
-& \;\; V & \\
+& \;\; F & \\
 & | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
@@
 
-The type {\bf void} is a type with no values, used by the
-type system as a way to detect branches that are unreachable
-due to flow typing. This type is neutral with regards to
-the type union operator: ${\bf void} \cup T = T \cup {\bf void} = T$ for any type $T$.
-
 The type $\Nil$ is the type of {\tt nil}, the value that Lua uses for undefined variables, missing parameters, and missing table keys.
@@
 The subtyping rules are a deduction system for the
-subtyping relation $\senv \vdash t_{1} \subtype t_{2}$.
+subtyping relation $\subtype$ between two types of the
+same kind.
 The variable $\senv$ is a set of subtyping assumptions used
@@
 The type environment is standard, and maps
-identifiers to first-level types. Some typing rules of
+identifiers to expression types. Some typing rules of
 Section~\ref{sec:tables} can change the type of an identifier,
@@
     vt(F_1, {\bf const}\;F_2) & = & {\bf const}\;nil(fix(F_2))\\\\
-    nil(F) & = & F \;\;\; \mathrm{if}\;\Nil \lesssim F\\
-    nil(F) & = & F \cup \Nil \;\;\; \mathrm{otherwise}\\\\
+    nil(T) & = & T \;\;\; \mathrm{if}\;\Nil \lesssim T\\
+    nil(T) & = & T \cup \Nil \;\;\; \mathrm{otherwise}\\\\
 	fix(\{\overline{F:V}\}_{unique|open}) & = &
@@
 to a closed-over variable can be unsafe. In the following
-example, assume that the $g$ receives a parameter of type
-$\{ ``x": \String \cup \Nil\}$:
+example, assume that $g$ receives a parameter of type
+$\{ ``x": \String \cup \Nil\}_{closed}$:
 \begin{center}
@@
 		& \multicolumn{3}{l}{$\mathbf{in} \; \lfloor g({<}\{``x":
-			 \String \cup \Nil\}{>} \; a) \rfloor_{0}$}
+			 \String \cup \Nil\}_{closed}{>} \; a) \rfloor_{0}$}
 	\end{tabular}
@@
 \dfrac{\begin{array}{c}
-	\env_{1}, \penv \vdash el:T_1 \times \ldots T_n \times P, \env_{2}, (x,S)\\
-	\env_{3} = \env_{2}[id_{1} \mapsto T_1, \ldots, id_{n} \mapsto T_n]\\
+	\env_{1}, \penv \vdash el:E, \env_{2}, (x,S)\\
+	\env_{3} = \env_{2}[id_{1} \mapsto proj(E, 1), \ldots, id_{n} \mapsto proj(E, n)]\\
 	\env_{3}, \penv[x \mapsto S] \vdash s, \env_{4} \;\;\;
@@
 proj(F \times P, 1) & = & F \\
-proj(F \times P, i) & = & proj(P, i-1)
+proj(F \times P, i) & = & proj(P, i-1)\\
+proj(E{*}, i) & = & nil(E) \\
+proj(T \times E, 1) & = & T \\
+proj(T \times E, i) & = & proj(E, i-1)
 \end{array}\\\\
@@
 A standard way of decomposing the union in the assignment
-would let the type of $q$be $\Integer \times \Nil$ and
+would let the type of $q$ be $\Integer \times \Nil$ and
 the type of $r$ be $\Integer \times \String$. Flow typing
@@
 At this point, using $q$ or $r$ as an r-value will project
-their types into either $\Integer \cup \Nil$ or $\Integer \cup \String$. The more interesting cases are inside the
+their types into either $\Integer \cup \Nil$ or $\Integer \cup \String$ via rule {\sc T-IdReadProj}, which just uses auxiliary function $proj$. The more interesting cases are inside the
 branches of the {\bf if} statement, where rule {\sc T-IfFilterNilProj} filters the whole union of tuples referenced
@@
 
-Assignment to an l-value with a projection type is forbidden. Allowing assignment is unsound, as it can break the dependency among the components of the union. In the example below,
+Assignment to an l-value with a projection type is forbidden
+(there is no rule {\sc T-IdWriteProj}). Allowing assignment is unsound, as it can break the dependency among the components of the union. In the example below,
 the projected type of $r$ outside of the {\bf if} statement

### 2015-06-14 doc/papers/paper/paper.tex
DLS paper: small fix to text on projection type rules

@@
 environment used for typing the {\bf if} statement.
-The first judgment above the bar of rule {\sc T-LocalProj} uses the auxiliary typing relation $\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$.
-
+
+The first judgment above the bar of rule {\sc T-LocalProj} uses the auxiliary typing relation $\env_{1}, \penv \vdash el : E, \env_{2}, (x,S)$.
 Rule {\sc T-ExpListProj} is the typing rule for this auxiliary
@@
 
-In the above example, the combination of these two rules binds $x$ to
+In the $idiv$ example above, the combination of these two rules binds $x$ to
 $(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})$ in the projection environment $\Pi$,

### 2015-06-17 doc/thesis/system.tex
Back porting the section about types from paper to thesis

@@
 
-\begin{figure}[!ht]
+Typed Lua includes types that can appear in annotations and some
+special types that cannot appear in annotations, though they
+play special role in type checking some Lua idioms and handling flow typing.
+Special types cannot appear in annotations because they do not have
+a corresponding concrete syntax, so we also isolated them in the
+formalization to make the implementation easier.
+First we present the types that compose the type language of our
+type system, and then we present the special types, indicating
+why they are necessary.
+
+\begin{figure*}[!ht]
 \textbf{Type Language}\\
@@
 \begin{array}{rlr}
-T ::= & & \textsc{first-level types:}\\
+F ::= & & \textsc{first-level types:}\\
 & \;\; L & \textit{literal types}\\
@@
 & | \; \Self & \textit{self type}\\
-& | \; T_{1} \cup T_{2} & \textit{disjoint union types}\\
+& | \; F_{1} \cup F_{2} & \textit{union types}\\
 & | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
-& | \; \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique|open|fixed|closed} & \textit{table types}\\
-& | \; x & \textit{type variables}\\
-& | \; \mu x.T & \textit{recursive types}\\
-& | \; \phi(T_{1},T_{2}) & \textit{filter types}\\
-& | \; \pi_{i}^{x} & \textit{projection types}\\
+& | \; \{\overline{F{:}V}\}_{unique|open|fixed|closed} & \textit{table types}\\
+& |\; x & \textit{type variables}\\
+& | \; \mu x.F & \textit{recursive types}\\
 %\multicolumn{3}{c}{}\\
@@
 %\multicolumn{3}{c}{}\\
-K ::= & & \textsc{{\small key types:}}\\
-& \;\; L \; | \; B \; | \; \Value &\\
-%\multicolumn{3}{c}{}\\
 V ::= & & \textsc{{\small value types:}}\\
-& \;\; T \; | \; \Const \; T &\\ 
+& \;\; F \; | \; \Const \; F &\\
 %\multicolumn{3}{c}{}\\
@@
 P ::= & & \textsc{{\small tuple types:}}\\
-& \;\; \Void & \textit{void type}\\
-& | \; T{*} & \textit{variadic types}\\
-& | \; T \times P & \textit{pair types}
+& \;\; F{*} & \textit{variadic types}\\
+& | \; F \times P & \textit{pair types}
 \end{array}
@@
 First-level types represent first-class Lua values and
-second-level types represent tuples of values that appear in 
-assignments and function applications.
+second-level types represent tuples of values that appear in the
+input and output of functions.
 First-level types include literal types, base types, the type $\Nil$,
 the top type $\Value$, the dynamic type $\Any$, the type $\Self$,
-union types, function types, table types, recursive types,
-filter types, and projection types.
+union types, function types, table types, and recursive types.
 Second-level types include tuple types and unions of tuple types.
-Tuple types include the type $\Void$, variadic types, and pair types.
+Tuple types include variadic types and pair types.
 Types are ordered by a subtype relationship that we introduce
@@
 
+Typed Lua includes the dynamic type $\Any$ for allowing programmers
+to mix static and dynamic typing.
+
 Typed Lua uses the type $\Self$ to represent the \emph{receiver}
@@
 
-Union types $T_{1} \cup T_{2}$ represent data types that can hold a value
+Union types $F_{1} \cup F_{2}$ represent data types that can hold a value
 of two different types.
@@
 Second-level types are either tuple types or unions of tuple types.
-Tuple types are tuples of first-level types that can end with
-either an empty tuple or with a variadic type.
+Tuple types are tuples of first-level types that end with a variadic type.
 Typed Lua needs second-level types because tuples are not first-class
@@
 and multiple assignments.
-The type $\Void$ is the type of an empty tuple.
-A variadic type $T{*}$ represents a sequence of values of type $T \cup \Nil$;
+A variadic type $F{*}$ represents a sequence of values of type $F \cup \Nil$;
 it is the type of a vararg expression.
@@
 that Lua tables can take.
-The syntactical form of table types is $\{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n} \}_{tag}$,
-where each $K_{i}$ represents the type of a table key,
-and each $V_{i}$ represents the type of the value that table keys of type $K_{i}$ map to.
-Key types can only be literal types, base types, or the top type.
-We made this restriction to the type of the keys because the statistics
-that we discussed in Section \ref{sec:statistics} showed that most
-of the tables are records, lists, and hashes.
-The type $\Value$ is an option when we need a loose table type.
-For instance, $\{\Value:\Value\}_{closed}$ represents the type of a
-table in which both indices and values can have any type.
-Value types can be any first-level type, and can optionally include
-the $\Const$ type to denote immutable values.
+The syntactical form of table types is $\{\overline{F{:}V}\}_{unique|open|fixed|closed}$,
+where the notation $\overline{F{:}V}$ denotes the list $F_{1}{:}V_{1}, ..., F_{n}{:}V_{n}$.
+Each $V_{i}$ represents the type of the value that table keys of type $F_{i}$ map to.
+%Key types can only be literal types, base types, or the top type.
+%We made this restriction to the type of the keys because the statistics
+%that we discussed in Section \ref{sec:statistics} showed that most
+%of the tables are records, lists, and hashes.
+%The type $\Value$ is an option when we need a loose table type.
+%For instance, $\{\Value:\Value\}_{closed}$ represents the type of a
+%table in which both indices and values can have any type.
+Value types represent mutable values by default, but we can use the
+$\Const$ type to make them represent immutable values.
 
@@
 
-Recursive types have the form $\mu x.T$,
-where $T$ is a first-level type that $x$ represents.
+Recursive types have the form $\mu x.F$,
+where $F$ is a first-level type that $x$ represents.
 For instance, $\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}$
@@
 
+With the recursive types we finish the dicussion about Typed Lua types,
+and we begin the discussion about the special types.
+
+\begin{figure*}[!ht]
+\textbf{Special Types}\\
+\dstart
+$$
+\begin{array}{rlr}
+T ::= & & \textsc{expression types:}\\
+& \;\; F & \textit{first-level types}\\
+& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
+& | \; \pi_{i}^{x} & \textit{projection types}\\
+%\multicolumn{3}{c}{}\\
+E ::= & & \textsc{expression list types:}\\
+& \;\; T{*} & \textit{variadic types}\\
+& | \; T \times E & \textit{pair types}\\
+%\multicolumn{3}{c}{}\\
+R ::= & & \textsc{filter result types:}\\
+& \;\; \mathbf{void} & \textit{void type}\\
+& |\; F & \textit{first-level types}
+\end{array}
+$$
+\dend
+\caption{The special types used by Typed Lua}
+\label{fig:specialtypes}
+\end{figure*}
+
+Figure \ref{fig:specialtypes} presents the special types that
+Typed Lua includes for typing some Lua idioms and flow typing.
+Typed Lua splits special types into three categories:
+\emph{expression types}, \emph{expression list types}, and
+\emph{filter result types}.
+Expression types represent the type of expressions in the type system,
+which can be any first-level type, a filter type, or a projection type.
+Expression list types represent the type of lists of expressions,
+which are tuples of expression types that end with a variadic type.
+Filter result types are either the type $\Void$ or a first-level type.
+In Section \ref{sec:fap} we will show that Typed Lua uses filter types
+in flow typing, being the type $\Void$ a type with no values, used by the
+type system as a way to detect branches that are unreachable due to flow typing.
+
 Typed Lua includes filter types as a way to discriminate the type of local
@@
 the type of a local variable that depends on the type of another local variable.
-
-Typed Lua includes the dynamic type $\Any$ for allowing programmers
-to mix static and dynamic typing.
 

### 2015-06-17 doc/papers/paper/paper.tex

@@
 {\em fixed} tables is the guarantees that the type system
-gives about the reference to a value of that type. A reference
-to an {\em unique} table is guaranteed to have no aliases,
+gives about the reference to a value of that type. A
+reference to an {\em unique} table is guaranteed to
+point to a table that has no other references to it,
 so it allows greater flexibility in reinterpreting its type;
 a reference to an {\em open} table is guaranteed to have
-only {\em closed} aliases (we will explain the meaning of the {\em closed} tag shortly), which still lets the type system
+only {\em closed} references pointing to the
+same table (we will explain the meaning of the {\em closed} tag shortly), which still lets the type system
 reinterpret the type of a reference, but with more
-restrictions; a {\em fixed} table can have any number of {\em fixed} or {\em closed} aliases, so its type cannot change
+restrictions; a reference to a {\em fixed} table can have any number of
+{\em fixed} or {\em closed} references point to it, 
+so its type cannot change
 anymore.
@@
 \emph{open} table types to appear in the type of the values.
-Our type system currently cannot
-track aliases of table fields, so this restriction is
+At the moment, local variables are the only kind of reference that the system
+can track, so this restriction is
 necessary. This limitation of the type system is the reason the {\sc T-Constructor} rule uses the auxiliary function $fix$
@@
 
-The type system needs to track aliasing of table references,
+The type system needs to track aliasing of tables,
 as the flexibility of {\em unique} and {\em open} tables
@@
 tables are {\em closed} with the auxiliary function $closeall$.
-This is not enough, as assignment
-to a closed-over variable can be unsafe. In the following
+Assignment to a closed-over table can be unsafe, because the {\em reference} to it is aliased as well. In the following
 example, assume that $g$ receives a parameter of type
@@
 		& \multicolumn{3}{l}{$\mathbf{local} \; f:\Void \rightarrow \Void =$}\\
-		& & \multicolumn{2}{l}{$\mathbf{fun} \; ():\Void$}\\
-		& & & \multicolumn{1}{l}{$a = \{[``x"]=5\} $}\\
+		& & \multicolumn{2}{l}{$\mathbf{fun} \; ():\Void\;\;\;
+		 a = \{[``x"]=5\} $}\\
 		& \multicolumn{3}{l}{$\mathbf{in} \; \lfloor g({<}\{``x":
@@
 a different (but compatible) type from its current one.
-The system tracks aliasing of table references
+The system tracks aliasing of tables
 in order to do this type evolution in a type-safe

### 2015-06-18 doc/thesis/system.tex
Back porting the section about subtyping from paper to thesis

@@
 We present the subtyping rules as a deduction system for the
-subtyping relation $\senv \vdash T_{1} \subtype T_{2}$.
+subtyping relation $\senv \vdash X \subtype Y$, where $X$ and $Y$
+are two types of the same kind.
 The variable $\senv$ is a set of pairs of recursion variables.
@@
 \mylabel{S-VALUE}\\
-\senv \vdash T \subtype \Value
+\senv \vdash F \subtype \Value
 \end{array}
@@
 \mylabel{S-UNION1}\\
-\dfrac{\senv \vdash T_{1} \subtype T \;\;\;
-       \senv \vdash T_{2} \subtype T}
-      {\senv \vdash T_{1} \cup T_{2} \subtype T}
+\dfrac{\senv \vdash F_{1} \subtype F \;\;\;
+       \senv \vdash F_{2} \subtype F}
+      {\senv \vdash F_{1} \cup F_{2} \subtype F}
 \end{array}
@@
 \mylabel{S-UNION2}\\
-\dfrac{\senv \vdash T \subtype T_{1}}
-      {\senv \vdash T \subtype T_{1} \cup T_{2}}
+\dfrac{\senv \vdash F \subtype F_{1}}
+      {\senv \vdash F \subtype F_{1} \cup F_{2}}
 \end{array}
@@
 \mylabel{S-UNION3}\\
-\dfrac{\senv \vdash T \subtype T_{2}}
-      {\senv \vdash T \subtype T_{1} \cup T_{2}}
-\end{array}
-\end{array}
-\]
-
-The first rule shows that a union type $T_{1} \cup T_{2}$
-is a subtype of $T$ if both $T_{1}$ and $T_{2}$ are subtypes
-of $T$;
-and the other rules show that a type $T$ is a subtype
-of a union type $T_{1} \cup T_{2}$ if $T$ is a subtype of
-either $T_{1}$ or $T_{2}$.
+\dfrac{\senv \vdash F \subtype F_{2}}
+      {\senv \vdash F \subtype F_{1} \cup F_{2}}
+\end{array}
+\end{array}
+\]
+
+The first rule shows that a union type $F_{1} \cup F_{2}$
+is a subtype of $F$ if both $F_{1}$ and $F_{2}$ are subtypes
+of $F$;
+and the other rules show that a type $F$ is a subtype
+of a union type $F_{1} \cup F_{2}$ if $F$ is a subtype of
+either $F_{1}$ or $F_{2}$.
 
@@
 
-The type $\Void$ is a subtype of itself and of a variadic type:
-\[
-\begin{array}{c}
-\mylabel{S-VOID2}\\
-\senv \vdash \Void \subtype T{*}
-\end{array}
-\]
-
-A variadic type $T{*}$ represents a sequence of values of type
-$T \cup \Nil$, and the rule \textsc{S-VOID2} handles the case where
-a given sequence is empty.
-
 The subtyping rule for pair types is the standard covariant rule:
@@
 \mylabel{S-PAIR}\\
-\dfrac{\senv \vdash T_{1} \subtype T_{2} \;\;\;
+\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
        \senv \vdash P_{1} \subtype P_{2}}
-      {\senv \vdash T_{1} \times P_{1} \subtype T_{2} \times P_{2}}
+      {\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
 \end{array}
@@
 The subtyping rules for variadic types are not so obvious.
-We need six different subtyping rules for variadic types
+We need three different subtyping rules for variadic types
 to handle all the cases where they can appear.
 
-The rule \textsc{S-VARARG1} is a special rule for handling the
-case where we give a sequence of $\Nil$ to the empty tuple:
-\[
-\begin{array}{c}
-\mylabel{S-VARARG1}\\
-\senv \vdash \Nil{*} \subtype \Void
-\end{array}
-\]
-
-The rule \textsc{S-VARARG2} handles the case where both tuple types end
+The rule \textsc{S-VARARG1} handles the case where both tuple types end
 with variadic types:
@@
 \mylabel{S-VARARG2}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \cup \Nil}
-      {\senv \vdash T_{1}{*} \subtype T_{2}{*}}
-\end{array}
-\]
-
-This rule shows that $T_{1}{*}$ is a subtype of $T_{2}{*}$
-if $T_{1} \cup \Nil$ is a subtype of $T_{2} \cup \Nil$.
+\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \cup \Nil}
+      {\senv \vdash F_{1}{*} \subtype F_{2}{*}}
+\end{array}
+\]
+
+This rule shows that $F_{1}{*}$ is a subtype of $F_{2}{*}$
+if $F_{1} \cup \Nil$ is a subtype of $F_{2} \cup \Nil$.
 It explicitly includes $\Nil$ in both sides because otherwise
@@
 \begin{array}{c}
+\mylabel{S-VARARG2}\\
+\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \;\;\;
+       \senv \vdash F_{1}{*} \subtype P_{2}}
+      {\senv \vdash F_{1}{*} \subtype F_{2} \times P_{2}}
+\end{array}
+\;
+\begin{array}{c}
 \mylabel{S-VARARG3}\\
-\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2}}
-      {\senv \vdash T_{1}{*} \subtype T_{2} \times \Void}
-\end{array}
-\;
-\begin{array}{c}
-\mylabel{S-VARARG4}\\
-\dfrac{\senv \vdash T_{1} \subtype T_{2} \cup \Nil}
-      {\senv \vdash T_{1} \times \Void \subtype T_{2}{*}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-VARARG5}\\
-\dfrac{\senv \vdash T_{1}{*} \subtype T_{2} \times \Void \;\;\;
-       \senv \vdash T_{1}{*} \subtype P_{2}}
-      {\senv \vdash T_{1}{*} \subtype T_{2} \times P_{2}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{S-VARARG6}\\
-\dfrac{\senv \vdash T_{1} \times \Void \subtype T_{2}{*} \;\;\;
-       \senv \vdash P_{1} \subtype T_{2}{*}}
-      {\senv \vdash T_{1} \times P_{1} \subtype T_{2}{*}}
-\end{array}
-\end{array}
-\]
-
-Note that the case where both tuple types end with the type $\Void$ does
-not require any special rule.
+\dfrac{\senv \vdash F_{1} \subtype F_{2} \cup \Nil \;\;\;
+       \senv \vdash P_{1} \subtype F_{2}{*}}
+      {\senv \vdash F_{1} \times P_{1} \subtype F_{2}{*}}
+\end{array}
+\end{array}
+\]
+
+These rules state the conditions when variadic tuples of different
+length are compatible.
 In the next section we will show that we use the subtyping rules for variadic types,
@@
 \mylabel{S-TABLE1}\\
-\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
-       \senv \vdash K_{j} \subtype K_{i}' \;\;\;
-       \senv \vdash K_{i}' \subtype K_{j} \;\;\;
+\dfrac{\forall i \; \exists j \;\;\;
+       \senv \vdash F_{j} \subtype F_{i}' \;\;\;
+       \senv \vdash F_{i}' \subtype F_{j} \;\;\;
        \senv \vdash V_{j} \subtype_{c} V_{i}'}
-      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{fixed|closed} \subtype
-                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}} \; m \ge n
+      {\senv \vdash \{\overline{F{:}V}\}_{fixed|closed} \subtype
+                    \{\overline{F'{:}V'}\}_{closed}}
 \end{array}
@@
 \mylabel{S-FIELD1}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2} \;\;\;
-       \senv \vdash V_{2} \subtype V_{1}}
-      {\senv \vdash V_{1} \subtype_{c} V_{2}}
+\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
+       \senv \vdash F_{2} \subtype F_{1}}
+      {\senv \vdash F_{1} \subtype_{c} F_{2}}
 \end{array}
@@
 \mylabel{S-FIELD2}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2}}
-      {\senv \vdash \Const \; V_{1} \subtype_{c} \Const \; V_{2}}
+\dfrac{\senv \vdash F_{1} \subtype F_{2}}
+      {\senv \vdash \Const \; F_{1} \subtype_{c} \Const \; F_{2}}
 \end{array}
@@
 \mylabel{S-FIELD3}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2}}
-      {\senv \vdash V_{1} \subtype_{c} \Const \; V_{2}}
+\dfrac{\senv \vdash F_{1} \subtype F_{2}}
+      {\senv \vdash F_{1} \subtype_{c} \Const \; F_{2}}
 \end{array}
@@
 \dfrac{\begin{array}{c}
-       \forall i \in 1..m \; \forall j \in 1..n \;
-       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{u} V_{j}' \\
-       \forall j \in 1..n \; \not\exists i \in 1..m \;
-       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
+       \forall i \; \forall j \;\;\;
+       \senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash V_{i} \subtype_{u} V_{j}'\\
+       \forall j \; \nexists i \;\;\;
+       \senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
        \end{array}}
-      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype
-                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
+      {\senv \vdash \{\overline{F{:}V}\}_{unique} \subtype
+                    \{\overline{F'{:}V'}\}_{closed}}
 \end{array}
@@
 \mylabel{S-FIELD4}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2}}
-      {\senv \vdash V_{1} \subtype_{u} V_{2}}
+\dfrac{\senv \vdash F_{1} \subtype F_{2}}
+      {\senv \vdash F_{1} \subtype_{u} F_{2}}
 \end{array}
@@
 \mylabel{S-FIELD5}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2}}
-      {\senv \vdash \Const \; V_{1} \subtype_{u} \Const \; V_{2}}
+\dfrac{\senv \vdash F_{1} \subtype F_{2}}
+      {\senv \vdash \Const \; F_{1} \subtype_{u} \Const \; F_{2}}
+\end{array}
+\\ \\
+\begin{array}{c}
+\mylabel{S-FIELD6}\\
+\dfrac{\senv \vdash F_{1} \subtype F_{2}}
+      {\senv \vdash \Const \; F_{1} \subtype_{u} F_{2}}
 \end{array}
@@
 \begin{array}{c}
-\mylabel{S-FIELD6}\\
-\dfrac{\senv \vdash V_{1} \subtype V_{2}}
-      {\senv \vdash V_{1} \subtype_{u} \Const \; V_{2}}
+\mylabel{S-FIELD7}\\
+\dfrac{\senv \vdash F_{1} \subtype F_{2}}
+      {\senv \vdash F_{1} \subtype_{u} \Const \; F_{2}}
 \end{array}
@@
 \begin{array}{c}
-\mylabel{S-FIELD7}\\
-\dfrac{\senv \vdash \Nil \subtype V}
-      {\senv \vdash \Nil \subtype_{o} V}
+\mylabel{S-FIELD8}\\
+\dfrac{\senv \vdash \Nil \subtype F}
+      {\senv \vdash \Nil \subtype_{o} F}
 \end{array}
@@
 \begin{array}{c}
-\mylabel{S-FIELD8}\\
-\dfrac{\senv \vdash \Nil \subtype V}
-      {\senv \vdash \Nil \subtype_{o} \Const \; V}
+\mylabel{S-FIELD9}\\
+\dfrac{\senv \vdash \Nil \subtype F}
+      {\senv \vdash \Nil \subtype_{o} \Const \; F}
 \end{array}
@@
 \dfrac{\begin{array}{c}
-       \forall i \in 1..m \\
-       \exists j \in 1..n \;
-       \senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
-       \forall j \in 1..n \; \not\exists i \in 1..m \;
-       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
+       \forall i \; \exists j \;\;\;
+       \senv \vdash F_{i} \subtype F_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
+       \forall j \; \nexists i \;\;\;
+       \senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
        \end{array}}
-      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype
-                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{unique|open|fixed}}
+      {\senv \vdash \{\overline{F{:}V}\}_{unique} \subtype
+                    \{\overline{F'{:}V'}\}_{unique|open|fixed}}
 \end{array}
@@
 \dfrac{\begin{array}{c}
-       \forall i \in 1..m \; \forall j \in 1..n \;
-       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
-       \forall j \in 1..n \; \not\exists i \in 1..m \;
-       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
+       \forall i \; \forall j \;\;\;
+       \senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
+       \forall j \; \nexists i \;\;\;
+       \senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
        \end{array}}
-      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype
-                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
+      {\senv \vdash \{\overline{F{:}V}\}_{open} \subtype
+                    \{\overline{F'{:}V'}\}_{closed}}
 \end{array}
@@
 \dfrac{\begin{array}{c}
-       \forall i \in 1..m \\
-       \exists j \in 1..n \;
-       \senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
-       \forall j \in 1..n \; \not\exists i \in 1..m \;
-       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
+       \forall i \; \exists j \;\;\;
+       \senv \vdash F_{i} \subtype F_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
+       \forall j \; \nexists i \;\;\;
+       \senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
        \end{array}}
-      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype
-                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{open|fixed}}
+      {\senv \vdash \{\overline{F{:}V}\}_{open} \subtype
+                    \{\overline{F'{:}V'}\}_{open|fixed}}
 \end{array}
@@
 \mylabel{S-TABLE6}\\
-\dfrac{\forall i \in 1..n \; \exists j \in 1..n \;\;\;
-       \senv \vdash K_{j} \subtype K_{i}' \;\;\;
-       \senv \vdash K_{i}' \subtype K_{j} \;\;\;
-       \senv \vdash V_{j} \subtype_{c} V_{i}'}
-      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{fixed} \subtype
-                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{fixed}}
+\dfrac{\begin{array}{c}
+       \forall i \; \exists j \;\;\;
+       \senv \vdash F_{i} \subtype F_{j}' \;\;\;
+       \senv \vdash F_{j}' \subtype F_{i} \;\;\;
+       \senv \vdash V_{i} \subtype_{c} V_{j}' \\
+       \forall j \; \exists i \;\;\;
+       \senv \vdash F_{i} \subtype F_{j}' \;\;\;
+       \senv \vdash F_{j}' \subtype F_{i} \;\;\;
+       \senv \vdash V_{i} \subtype_{c} V_{j}' \\
+       \end{array}}
+      {\senv \vdash \{\overline{F{:}V}\}_{fixed} \subtype
+                    \{\overline{F'{:}V'}\}_{fixed}}
 \end{array}
@@
 \mylabel{S-AMBER}\\
-\dfrac{\senv[x_{1} \subtype x_{2}] \vdash T_{1} \subtype T_{2}}
-      {\senv \vdash \mu x_{1}.T_{1} \subtype \mu x_{2}.T_{2}}
+\dfrac{\senv[x_{1} \subtype x_{2}] \vdash F_{1} \subtype F_{2}}
+      {\senv \vdash \mu x_{1}.F_{1} \subtype \mu x_{2}.F_{2}}
 \end{array}
@@
 The rule \textsc{S-AMBER} also uses the rule \textsc{S-ASSUMPTION}
-to check whether $\mu x_{1}.T_{1} \subtype \mu x_{2}.T_{2}$.
+to check whether $\mu x_{1}.F_{1} \subtype \mu x_{2}.F_{2}$.
 Both rules use the set of assumptions $\senv$,
@@
 The rule \textsc{S-AMBER} extends $\senv$ with the assumption
-$x_{1} \subtype x_{2}$ to check whether $T_{1} \subtype T_{2}$.
+$x_{1} \subtype x_{2}$ to check whether $F_{1} \subtype F_{2}$.
 The rule \textsc{S-ASSUMPTION} allows the rule \textsc{S-AMBER}
@@
 \mylabel{S-UNFOLDR}\\
-\dfrac{\senv \vdash T_{1} \subtype [x \mapsto \mu x.T_{2}]T_{2}}
-      {\senv \vdash T_{1} \subtype \mu x.T_{2}}
+\dfrac{\senv \vdash F_{1} \subtype [x \mapsto \mu x.F_{2}]F_{2}}
+      {\senv \vdash F_{1} \subtype \mu x.F_{2}}
 \end{array}
@@
 \mylabel{S-UNFOLDL}\\
-\dfrac{\senv \vdash [x \mapsto \mu x.T_{1}]T_{1} \subtype T_{2}}
-      {\senv \vdash \mu x.T_{1} \subtype T_{2}}
+\dfrac{\senv \vdash [x \mapsto \mu x.F_{1}]F_{1} \subtype F_{2}}
+      {\senv \vdash \mu x.F_{1} \subtype F_{2}}
 \end{array}
@@
 Filter types are subtypes only of themselves and of $\Value$.
-More precisely, a filter type $\phi(T_{1},T_{2})$ is a subtype of
-the same filter type $\phi(T_{1},T_{2})$, which shares the same
-types $T_{1}$ and $T_{2}$, and it is also a subtype of $\Value$.
+More precisely, a filter type $\phi(F_{1},F_{2})$ is a subtype of
+the same filter type $\phi(F_{1},F_{2})$, which shares the same
+types $F_{1}$ and $F_{2}$, and it is also a subtype of $\Value$.
 
@@
 We present the consistent-subtyping rules as a deduction system for
-the consistent-subtyping relation $\senv \vdash T_{1} \lesssim T_{2}$.
+the consistent-subtyping relation $\senv \vdash X \lesssim Y$,
+where $X$ and $Y$ are two types of the same kind.
 As in the subtyping relation, $\senv$ is also a set of pairs of
@@
 \mylabel{C-ANY1}\\
-\senv \vdash T \lesssim \Any
+\senv \vdash F \lesssim \Any
 \end{array}
@@
 \mylabel{C-ANY2}\\
-\senv \vdash \Any \lesssim T
+\senv \vdash \Any \lesssim F
 \end{array}
@@
 If we had set the type $\Any$ as both bottom and top types of our
-subtyping relation, then any type $T_{1}$ would be a subtype of
-any other type $T_{2}$.
+subtyping relation, then any type $F_{1}$ would be a subtype of
+any other type $F_{2}$.
 The consequence of this is that all programs would type check without errors.
 This would happen due to the transitivity of subtyping, that is,
-we would be able to down-cast any type $T_{1}$ to $\Any$ and then up-cast
-$\Any$ to any other type $T_{2}$.
+we would be able to down-cast any type $F_{1}$ to $\Any$ and then up-cast
+$\Any$ to any other type $F_{2}$.
 The rules \textsc{C-ANY1} and \textsc{C-ANY2} are the rules that

### 2015-06-18 doc/thesis/system.tex
Back porting abstract syntax from paper to thesis

@@
 | \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
-& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
+& | \; \mathbf{local} \; \overline{id{:}F} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
 & | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
-& | \; \mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s & \textit{recursive function} \\
+& | \; \mathbf{rec} \; id{:}F = e \; \mathbf{in} \; s & \textit{recursive declaration} \\
 & | \; \mathbf{return} \; el & \textit{return} \\
@@
 & | \; e_{1}[e_{2}] & \textit{table access}\\
-& | \; {<}T{>} \; id & \textit{type coercion}\\
+& | \; {<}F{>} \; id & \textit{type coercion}\\
 & | \; f & \textit{function declaration}\\
@@
 & \;\; id_{l} & \textit{variable assignment}\\
-& | \; e_{1}[e_{2}] & \textit{table assignment}\\
-& | \; id[k] \; {<}T{>} & \textit{type coercion}\\
+& | \; e_{1}[e_{2}]_{l} & \textit{table assignment}\\
+& | \; id[e] \; {<}V{>} & \textit{type coercion}\\
 k ::= & & \textsc{literal constants:}\\
@@
 pl ::= & & \textsc{parameter lists:}\\
-& \;\; \overline{id{:}T} \; | \;
-\overline{id{:}T},{...}{:}T & \\
+& \;\; \overline{id{:}F} \; | \;
+\overline{id{:}F},{...}{:}F & \\
 \end{array}
@@
 $f$ are function declarations, $pl$ are parameter lists,
-$id$ are variable names, $T$ are first-level types, and $S$ are second-level types.
-The notation $\overline{id{:}T}$ denotes the list $id_{1}{:}T_{1}, ..., id_{n}{:}T_{n}$.
+$id$ are variable names, $F$ are first-level types, and $S$ are second-level types.
+The notation $\overline{id{:}F}$ denotes the list $id_{1}{:}F_{1}, ..., id_{n}{:}F_{n}$.
 
@@
 We also include two kinds of type coercions in our core language:
-the left-hand value $id[k] \; {<}T{>}$ and the expression ${<}T{>} \;id$.
+the left-hand value $id[e] \; {<}V{>}$ and the expression ${<}F{>} \;id$.
 Both allow the refinement of table types.
@@
 The first environment is the type environment $\env$ that maps
-variables to first-level types.
+variables to expression types, as the type of an expression can
+be a first-level type, a filter type, or a projection type.
 We use $\env_{1}[id \mapsto T]$ to extend the environment $\env_{1}$
@@
 left-hand values use a partial auxiliary function \emph{merge} to collect
-all environment changes in a new environment $\env_{m}$, if there are no conflicts.
+all environment changes in a new environment $\env_{f}$, if there are no conflicts.
 In Section \ref{sec:refinement} we will also show that we can only change the
@@
 \mylabel{T-ASSIGNMENT}\\
-\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash \overline{l}:S_{2}, \env_{3} \;\;\;
-       S_{1} \lesssim S_{2}}
+\dfrac{\env_{1}, \penv \vdash el:E_{1}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash \overline{l}:E_{2}, \env_{3} \;\;\;
+       E_{1} \lesssim E_{2}}
       {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
@@
 
-Our type system includes $\Nil{*}$ in the type of an expression list only
-if its type does not end in another variadic type $T{*}$.
-For instance, the rules \textsc{T-EXPLIST2} and \textsc{T-EXPLIST3} handle
-the case where an expression list ends with an expression that may produce
-multiple values.
-The former rule includes the type $\Nil{*}$, but the latter rule does not:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-EXPLIST2}\\
-\dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
-       \env_{1}, \penv \vdash me:T_{n+1} \times ... \times T_{n+m} \times \Void, \env_{n+2}\\
-       \env_{f} = merge(\env_{1}, ..., \env_{n+2}) \;\;\;
-       n = |\;\overline{e}\;|
-       \end{array}}
-      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n+m} \times \Nil{*}, \env_{f}}
-\end{array}
-\\ \\
-\begin{array}{c}
-\mylabel{T-EXPLIST3}\\
-\dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
-       \env_{1}, \penv \vdash me:T_{n+1} \times ... \times T_{n+m}{*}, \env_{n+2}\\
-       \env_{f} = merge(\env_{1}, ..., \env_{n+2}) \;\;\;
-       n = |\;\overline{e}\;|
-       \end{array}}
-      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n+m}{*}, \env_{f}}
-\end{array}
-\end{array}
-\]
-
 Rules for function applications are similar to the rule for multiple assignment.
@@
 \mylabel{T-APPLY1}\\
-\dfrac{\env_{1}, \penv \vdash e:S_{1} \rightarrow S_{2}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash el:S_{3}, \env_{3} \;\;\;
-       S_{3} \lesssim S_{1}}
-      {\env_{1}, \penv \vdash e(el):S_{2}, \env_{3}}
+\dfrac{\env_{1}, \penv \vdash e:E_{1} \rightarrow E_{2}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash el:E_{3}, \env_{3} \;\;\;
+       E_{3} \lesssim E_{1}}
+      {\env_{1}, \penv \vdash e(el):E_{2}, \env_{3}}
 \end{array}
@@
 \mylabel{T-STMAPPLY1}\\
-\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
+\dfrac{\env_{1}, \penv \vdash e(el):E, \env_{2}}
       {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{0},\env_{2}}
@@
 \mylabel{T-EXPAPPLY1}\\
-\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
-      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{1}:first(S), \env_{2}}
+\dfrac{\env_{1}, \penv \vdash e(el):E, \env_{2}}
+      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{1}:first(E), \env_{2}}
 \end{array}
@@
 \begin{align*}
-first(\Void) & = \Nil\\
-first(T{*}) & = T \cup \Nil\\
-first(T \times P) & = T\\
+first(F{*}) & = F \cup \Nil\\
+first(F \times P) & = F\\
 first(S_{1} \sqcup S_{2}) & = first(S_{1}) \cup first(S_{2})
@@
 Our type system also catches arity mismatch.
-To do that, we end the input type of a function with type $\Void$
+To do that, we end the input type of a function with type $\Nil{*}$
 instead of $\Value{*}$.
 For instance, let us assume that $f$ has type
-$\String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void \rightarrow \Integer{*}$.
+$\String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Nil{*} \rightarrow \Integer{*}$.
 The function call
@@
 \[
-``foo" \times \Nil{*} \lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void
+``foo" \times \Nil{*} \lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Nil{*}
 \]
@@
 \[
-``foo" \times 1 \times 2 \times 3 \times \Nil{*} \not\lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void
+``foo" \times 1 \times 2 \times 3 \times \Nil{*} \not\lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Nil{*}
 \]
@@
 For instance, let us assume that $f$ has type
-$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Void$.
+$\Integer \times \Integer \times \Nil{*} \rightarrow \Integer \times \Nil{*}$.
 The function call
@@
 \[
-1 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Void
+1 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Nil{*}
 \]
@@
 \[
-1 \times 2 \times 3 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Void
+1 \times 2 \times 3 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Nil{*}
 \]

### 2015-06-19 doc/thesis/system.tex
Back porting table refinement from paper to thesis

@@
 Each $V_{i}$ represents the type of the value that table keys of type $F_{i}$ map to.
-%Key types can only be literal types, base types, or the top type.
-%We made this restriction to the type of the keys because the statistics
-%that we discussed in Section \ref{sec:statistics} showed that most
-%of the tables are records, lists, and hashes.
-%The type $\Value$ is an option when we need a loose table type.
-%For instance, $\{\Value:\Value\}_{closed}$ represents the type of a
-%table in which both indices and values can have any type.
 Value types represent mutable values by default, but we can use the
@@
 
-The simplest expression involving tables is the empty table constructor.
-Its type checking rule is straightforward:
-\[
-\begin{array}{c}
-\mylabel{T-CONSTRUCTOR1}\\
-\env_{1}, \penv \vdash \{\}:\{\}_{unique}, \env_{1}
-\end{array}
-\]
+The simplest expression involving tables is the empty table constructor $\{\}$;
+it always has type $\{\}_{unique}$.
 
@@
 First, our type system uses the auxiliary relation
-$\env_{1}, \penv \vdash [e_{1}] = e_{2} : (K,V), \env_{2}$ to type check each
+$\env_{1}, \penv \vdash [e_{1}] = e_{2} : (F,V), \env_{2}$ to type check each
 table field.
@@
 and a projection environment $\penv$, checking a table field $[e_{1}] = e_{2}$
-produces a pair $(K,V)$ and a new type environment $\env_{2}$.
-A pair $(K,V)$ means that $e_{1}$ has type $K$ and $e_{2}$ has type $V$,
-where $K$ is the type of the key and $V$ is the type of the value.
-
-After type checking each table field, our type system uses each pair $(K,V)$
+produces a pair $(F,V)$ and a new type environment $\env_{2}$.
+A pair $(F,V)$ means that $e_{1}$ has type $F$ and $e_{2}$ has type $V$,
+where $F$ is the type of the key and $V$ is the type of the value.
+
+After type checking each table field, our type system uses each pair $(F,V)$
 to build the table type that express the type of a given constructor, and
@@
 \[
-\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
+\begin{array}{rcl}
+wf(\{\overline{F:V}\}_{unique|open|fixed|closed}) & = & \forall i \; ((\nexists j \; i \not= j \,\wedge\, F_{i} \lesssim F_{j}) \,\wedge\, wf(V_{i}) \,\wedge\\
+& & \;\;\;\; \lnot tag(V_{i},unique) \,\wedge\, \lnot tag(V_{i},open))\\
+wf({\bf const}\; F) & = & wf(F) \\
+wf(F_1 \cup F_2) & = & wf(F_1) \,\wedge\, wf(F_2) \\
+wf(\mu x.F) & = & wf(F) \\
+wf(S_1 \rightarrow S_2) & = & wf(S_1) \,\wedge\, wf(S_2) \\
+wf(S_1 \sqcup S_2) & = & wf(S_1) \,\wedge\, wf(S_2)\\
+wf(F{*}) &=& wf(F) \\
+wf(F \times P) &=& wf(F) \,\wedge\,wf(P)\\
+wf(F) & = & \top \;\;\;\mathrm{for\; all\; other\; cases}\\\\
+tag(F_1 \cup F_2, t) & = & tag(F_1, t) \,\vee\, tag(F_2,t) \\
+tag(\{\overline{F:V}\}_t, t) & = & \top\\
+\end{array}
 \]
@@
 The rules that type check table fields use the auxiliary function
-\emph{close} to close the type of the values in a table type.
+\emph{fix} to change any \emph{unique} and \emph{open} table types
+used in the field initializer to \emph{fixed}.
 
@@
 \dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
-       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
+       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(F_{i},V_{i}), \env_{i+1} \\
+       T = \{F_{1}{:}V_{1}, ..., F_{n}{:}V_{n}\}_{unique} \;\;\;
        wf(T) \;\;\;
@@
 As another example, the constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
-has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ through rule
+has type $\{``x":1, ``y":\{``z":2\}_{fixed}\}_{unique}$ through rule
 \textsc{T-CONSTRUCTOR2}.
-The inner table is \emph{closed} to prevent the creation of unsafe aliases.
+The inner table is \emph{fixed} to prevent the creation of unsafe aliases.
 
@@
 and then we refine its type to $\{``firstname":\String, ``lastname":\String\}_{unique}$.
-Rule \textsc{T-REFINE} type checks this use of refinement:
-\[
-\begin{array}{c}
-\mylabel{T-REFINE}\\
+Rule \textsc{T-REFINE1} type checks this use of refinement:
+\[
+\begin{array}{c}
+\mylabel{T-REFINE1}\\
 \dfrac{\begin{array}{c}
-       \env_{1}(id) = \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n} \}_{open|unique}\\
-       \env_{1}, \penv \vdash k:K, \env_{2} \;\;\;
-       \not \exists i \in 1..n \; K \lesssim K_{i} \;\;\;
-       V = close(T)
+       \env_{1}(id) = \{\overline{F{:}V}\}_{unique}\\
+       \env_{1}, \penv \vdash e:F_{new}, \env_{2} \;\;\;
+       \nexists i \in 1..n \; F_{new} \lesssim F_{i} \;\;\;
+       V_{new} = vt(F_{new},V) \;\;\; n = |\overline{F{:}V}|
        \end{array}}
-      {\env_{1}, \penv \vdash id[k] {<}T{>}:V, \env_{2}[id \mapsto \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, K{:}V\}_{open|unique}]}
-\end{array}
-\]
-
-We restricted the refinement of table types to include only literal
-keys, because its purpose is to make it easier the construction of
-table types that represent records.
+      {\env_{1}, \penv \vdash id[e] {<}V{>}:V_{new}, \env_{2}[id \mapsto \{\overline{F{:}V}, F_{new}{:}V_{new}\}_{unique}]}
+\end{array}
+\]
+
+The rule \textsc{T-REFINE} uses the auxiliary function \emph{vt}
+to adjust the type of the value according to the type of the key.
+More precisely, \emph{vt} includes the type $\Nil$ in the type of
+the value when the type of the key is not a literal type.
+It is also this function that uses \emph{fix} to prevent
+\emph{unique} and \emph{open} table types to appear in the type
+of the values.
 
@@
 We also do not allow the refinement of table types to introduce
-fields with table types that are not \emph{closed}.
+fields with table types that are neither \emph{fixed} nor \emph{closed}.
 For instance,
@@
 \end{center}
-refines the type of $\string_ENV$ from $\{\}_{open}$ to $\{``x":\{\}_{closed}\}_{open}$.
+refines the type of $\string_ENV$ from $\{\}_{open}$ to $\{``x":\{\}_{fixed}\}_{open}$.
 Currently, our type system can only track \emph{unique} and
@@
 By consistent we mean that we are only adding new fields.
+More precisely, the first coercion expression refines the type of $\string_ENV$
+to $\{``x":\String\}_{open}$, while
+the second coercion expression refines the type of $\string_ENV$
+to $\{``y":\Integer\}_{open}$.
+Merging the two yields $\{``x":\String, ``y":\Integer\}_{open}$.
 Nevertheless, the next example does not type check because it tries to add
@@
 \mylabel{T-IDREAD1}\\
-\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
-      {\env_{1}, \penv \vdash id:close(T_{2}), \env_{1}[id \mapsto open(T_{1})]}
+\dfrac{\env_{1}(id) = F}
+      {\env_{1}, \penv \vdash id:close(F), \env_{1}[id \mapsto open(F)]}
 \end{array}
@@
 \mylabel{T-IDREAD2}\\
-\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
-      {\env_{1}, \penv \vdash id:fix(T_{2}), \env_{1}[id \mapsto fix(T_{1})]}
-\end{array}
-\end{array}
-\]
-
-Both rules use the auxiliary function \emph{read} because they may be
-accessing an identifier that is bound to a filter or projection type.
-As we mentioned in Section \ref{sec:unions}, our type system includes
-a small set of \texttt{type} predicates that allow programmers to
-discriminate union types, and our type system uses filter and projection
-types in the definition of these predicates to handle the discrimination
-of unions types.
-While filter types discriminate unions of first-level types, projection
-types discriminate unions of second-level types and project unions of
-second-level types into unions of first-level types.
-We can define \emph{read} as follows:
-\begin{align*}
-read(\penv, \phi(T_{1},T_{2})) & = T_{2}\\
-read(\penv, \pi_{i}^{x}) & = proj(\penv(x), i)\\
-read(\penv, T) & = T
-\end{align*}
-
-The function \emph{read} uses the auxiliary function \emph{proj}
-to project a union of first-level types, based on an union of
-second-level types and an index from a projection type.
-In Section \ref{sec:fap} we will discuss how our type system uses
-projection types to handle overloaded return types.
-We can define \emph{proj} as follows:
-\begin{align*}
-proj(T_{1} \times ... \times T_{n} \times T{*}, i) & =  T_{i} \;\;\; \text{if $i <= n$}\\
-proj(S_{1} \sqcup S_{2}, i) & = proj(S_{1}, i) \cup proj(S_{2}, i)
-\end{align*}
-
-We also need to close \emph{unique} and \emph{open} tables that
-appear in the left-hand side of assignments, as leaving them
-\emph{unique} and \emph{open} would allow the creation of
-unsafe references.
-
-As an example,
-\begin{center}
-\begin{tabular}{lll}
-\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
-& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = \{\} \; \mathbf{in}$}\\
-& & \multicolumn{1}{l}{$b = a;$}\\
-& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
-& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
-\end{tabular}
-\end{center}
-does not type check because we cannot add the field $``x"$ to $b$,
-as its type is \emph{closed}.
-Aliasing $a$ changes the type of $a$ from $\emph{unique}$ to
-$\emph{open}$, and that is the reason why we can add the field
-$``x"$ to the type of $a$.
-Aliasing $a$ also produces the type $\{\}_{closed}$, which is
-the same type that $b$ has in left-hand side of the assignment.
-After the assignment, the type of $b$ is \emph{closed} and thus
-does not allow changing the value that is stored in the field
-$``x"$ of $a$.
-
-Rule \textsc{T-IDWRITE} defines this behavior:
+\dfrac{\env_{1}(id) = F}
+      {\env_{1}, \penv \vdash id:fix(F), \env_{1}[id \mapsto fix(F)]}
+\end{array}
+\end{array}
+\]
+
+We do not need to close \emph{unique} and \emph{open} tables that
+appear in the left-hand side of assignments, because \textsc{T-IDREAD}
+is sufficient to forbid the creation of an alias to another \emph{unique}
+or \emph{open} table.
+For this reason, identifiers that appear in the left-hand side
+of assignements have their own rule \textsc{T-IDWRITE}:
 \[
@@
 \mylabel{T-IDWRITE}\\
-\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = write(T_{1})}
-      {\env_{1}, \penv \vdash id_{l}:close(T_{2}), \env_{1}[id \mapsto close(T_{2})]}
-\end{array}
-\]
-
-This rule uses the auxiliary function \emph{write} because it may be
-accessing an identifier that is bound to a filter type.
-As we mentioned in Sections \ref{sec:unions} and \ref{sec:alias},
-assignments restore discriminated union types to their original types,
-and function \emph{write} works in this purpose.
-We can define \emph{write} as follows:
-\begin{align*}
-write(\phi(T_{1},T_{2})) & = T_{1}\\
-write(T) & = T
-\end{align*}
+\dfrac{\env_{1}(id) = F}
+      {\env_{1}, \penv \vdash id_{l}:F, \env_{1}}
+\end{array}
+\]
 
 Our type system also has different rules for type checking table indexing to avoid
-changing table types in these operations, as they cannot create aliases:
-\[
-\begin{array}{c}
-\begin{array}{c}
-\mylabel{T-INDEX1}\\
+changing table types in these operations, as they cannot create aliases.
+These rules also use the auxiliary function \emph{rconst} to strip the
+$\Const$ type from the type of the value, if present.
+We define these rules as follows:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-INDEXREAD1}\\
 \dfrac{\begin{array}{c}
-       \env_{1}(id) = T \;\;\;
-       read(\penv, T) = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}\\
-       \env_{1}, \penv \vdash e_{2}:K, \env_{2} \;\;\;
-       \exists i \in 1{..}n \; K \lesssim K_{i}
+       \env_{1}(id) = \{\overline{F{:}V}\} \;\;\;
+       \env_{1}, \penv \vdash e_{2}:F, \env_{2} \;\;\;
+       \exists i \in 1{..}n \; F \lesssim F_{i} \;\;\;
+       n = |\overline{F{:}V}|
        \end{array}}
-      {\env_{1}, \penv \vdash id[e_{2}]:V_{i}, \env_{2}}
+      {\env_{1}, \penv \vdash id[e_{2}]:rconst(V_{i}), \env_{2}}
 \end{array}
@@
 \begin{array}{c}
-\mylabel{T-INDEX2}\\
+\mylabel{T-INDEXREAD2}\\
 \dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash e_{1}:\{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}, \env_{2}\\
-       \env_{2}, \penv \vdash e_{2}:K, \env_{3} \;\;\;
-       \exists i \in 1{..}n \; K \lesssim K_{i}
+       \env_{1}, \penv \vdash e_{1}:\{\overline{F{:}V}\}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash e_{2}:F, \env_{3} \;\;\;
+       \exists i \in 1{..}n \; F \lesssim F_{i} \;\;\;
+       n = |\overline{F{:}V}|
        \end{array}}
-      {\env_{1}, \penv \vdash e_{1}[e_{2}]:V_{i}, \env_{3}}
-\end{array}
-\end{array}
-\]
+      {\env_{1}, \penv \vdash e_{1}[e_{2}]:rconst(V_{i}), \env_{3}}
+\end{array}
+\end{array}
+\]
+
+Rule \textsc{T-INDEXREAD1} defines the case where using an identifier
+to index a table does not create an alias, while rule \textsc{T-INDEXREAD2}
+defines the case for indexing expressions where the expression
+denoting the table is not an identifier.
+The rules for indexing left-hand values are similar to these rules,
+except that they ensure that the field is not $\Const$.
 
@@
 or merge existing fields.
-To do that, Typed Lua includes a type coercion expression ${<}T{>} \; id$.
+To do that, Typed Lua includes a type coercion expression ${<}F{>} \; id$.
 For instance, we can use this type coercion expression to make the following
@@
 
-Rule \textsc{T-COERCE} defines the behavior of the type coercion expression:
-\[
-\begin{array}{c}
-\mylabel{T-COERCE}\\
-\dfrac{\env_{1}(id) \subtype T \;\;\;
-       \env_{1}[id \mapsto T], \penv \vdash id:T_{1}, \env_{2}}
-      {\env_{1}, \penv \vdash {<}T{>} \; id:T_{1}, \env_{2}}
-\end{array}
-\]
-
-Note that rule \textsc{T-COERCE} only allows changing the type
+Rules \textsc{T-COERCE1} and \textsc{T-COERCE2} define the behavior of the
+type coercion expression:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-COERCE1}\\
+\dfrac{\env_{1}(id) \subtype F \;\;\; tag(F,closed)}
+      {\env_{1}, \penv \vdash {<}F{>} \; id:F, \env_{1}[id \mapsto reopen(F)]}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{T-COERCE2}\\
+\dfrac{\env_{1}(id) \subtype F \;\;\; tag(F,fixed)}
+      {\env_{1}, \penv \vdash {<}F{>} \; id:F, \env_{1}[id \mapsto F]}
+\end{array}
+\end{array}
+\]
+
+Note that the coercion rules only allows changing the type
 of a variable if the new type is a supertype of the previous type,
@@
 type checking a nested scope.
-The function \emph{crall} closes all \emph{unique} and \emph{open}
-table types; it also restores filter types to their original types.
-The function \emph{crset} closes a given set of free assigned variables,
-which is given by the function \emph{fav}, and
-it also uses this set to restore filter types to their original types.
+The function \emph{closeall} closes all \emph{unique} and \emph{open} table types.
+The function \emph{closeset} closes a given set of free assigned variables,
+which is given by the function \emph{fav}.
 The function \emph{openset} changes from \emph{unique} to \emph{open}
-a given set of referenced variables, which is given by the function \emph{rv}.
+a given set of referenced variables, which is given by the function \emph{frv}.
 
@@
 \multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique}, b:\{\}_{unique} = \{\}, \{\} \; \mathbf{in}$}\\
-& \multicolumn{3}{l}{$\mathbf{local} \; f:\Integer \times \Void \rightarrow \Integer \times \Void =$}\\
-& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer):\Integer \times \Void$}\\
+& \multicolumn{3}{l}{$\mathbf{local} \; f:\Integer \times \Nil{*} \rightarrow \Integer \times \Nil{*} =$}\\
+& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer):\Integer \times \Nil{*}$}\\
 & & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
@@
 $a$ and $b$ have the same type: $\{\}_{closed}$.
-Their type was closed by \emph{crall} before type checking
+Their type was closed by \emph{closeall} before type checking
 the function body.
@@
 First, the function \emph{fav} includes $b$ in the set of variables
-that should be closed by \emph{crset}.
-Then, the function \emph{rv} includes $a$ in the set of variables
+that should be closed by \emph{closeset}.
+Then, the function \emph{frv} includes $a$ in the set of variables
 that should change from \emph{unique} to \emph{open} by \emph{openset}.
@@
 \dfrac{\begin{array}{c}
-       crall(\env_{1}[\overline{id} \mapsto \overline{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}\\
-       \env_{3} = openset(crset(\env_{1}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)),rv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
+       closeall(\env_{1})[\overline{id \mapsto F}], \penv[\ret \mapsto S] \vdash s, \env_{2} \\
+       \env_{3} = openset(\env_{1}, frv(\mathbf{fun} \; (\overline{id{:}F}){:}S \; s)) \\
+       \env_{4} = closeset(\env_{3}, fav(\mathbf{fun} \; (\overline{id{:}F}){:}S \; s))
        \end{array}}
-      {\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{3}}
+      {\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}F}){:}S \; s:F_{1} \times ... \times F_{n} \times \Nil{*} \rightarrow S, \env_{4}}
 \end{array}

### 2015-06-19 doc/thesis/system.tex
Back porting projections from paper to thesis

@@
 & & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
-& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; b[``x"] \; {<}\String{>} = ``foo" \;;\; f(a[``x"])$}
+& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; b[``x"] \; {<}\String{>} = ``foo" \;;\; \lfloor f(a[``x"]) \rfloor_{0}$}
 \end{tabular}
@@
 \[
-\Integer \times \Integer \times \Void \rightarrow (\Integer \times \Integer \times \Void) \sqcup (\Nil \times \String \times \Void)
+\Integer \times \Integer \times \Nil{*} \rightarrow (\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})
 \]
@@
 The function \emph{print} is a variadic function of type
-$\Value{*} \rightarrow \Void$.
+$\Value{*} \rightarrow \Nil{*}$.
 Let us see how our type system type checks the following program:
@@
 First, our type system uses the auxiliary relation
-$\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$
+$\env_{1}, \penv \vdash el : E, \env_{2}, (x,S)$
 for type checking $idiv(1, 2)$.
@@
 a projection environment $\penv$, we can check that an expression
-list $el$ has type $S_{1}$ and produces a new type environment $\env_{2}$
-and produces a pair $(x,S_{2})$.
+list $el$ has type $E$ and produces a new type environment $\env_{2}$
+and produces a pair $(x,S)$.
 This pair means that the last expression of an expression list $el$
-produces an union of second-level types $S_{2}$ that should be bound
-to a variable $x$ in the projection environment $\penv$,
+produces an union of second-level types $S$ that should be bound
+to a fresh variable $x$ in the projection environment $\penv$,
 as the resulting type of this expression is a tuple of projection
@@
        \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
-       \env_{1}, \penv \vdash me:S, \env_{n+2}\\
-       S = T_{n+1} \times ... \times T_{n+m} \times \Void \sqcup T_{n+1}' \times ... \times T_{n+m}' \times \Void\\
-       \env_{f} = merge(\env_{1}, ..., \env_{n+2}) \;\;\;
-       n = |\;\overline{e}\;|
+       \env_{1}, \penv \vdash me:S, \env_{n+2} \;\;\;
+       n = |\;\overline{e}\;| \\
+       S = T_{n+1} \times ... \times T_{n+m} \times \Nil{*} \sqcup T_{n+1}' \times ... \times T_{n+m}' \times \Nil{*} \;\;\;
+       x \notin \penv \\
+       \env_{f} = merge(\env_{1}, ..., \env_{n+2})
        \end{array}}
@@
 In the rule that type checks the declaration of unannotated variables,
-our type system uses the pair $(x,S_{2})$ to bound a union of
-second-level types $S_{2}$ to a variable $x$ in the projection
+our type system uses the pair $(x,S)$ to bound a union of
+second-level types $S$ to a variable $x$ in the projection
 environment $\penv$.
@@
 \dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash el:S_{1}, \env_{2}, (x,S_{2})\\
-       \env_{2}[id_{1} \mapsto infer(S_{1},1), ..., id_{n} \mapsto infer(S_{1},n)], \penv[x \mapsto S_{2}] \vdash s, \env_{3} \;\;\;
+       \env_{1}, \penv \vdash el:E, \env_{2}, (x,S)\\
+       \env_{3} = \env_{2}[id_{1} \mapsto infer(E,1), ..., id_{n} \mapsto infer(E,n)] \\
+       \env_{3}, \penv[x \mapsto S] \vdash s, \env_{4} \;\;\;
        n = |\;\overline{id}\;|  
        \end{array}}
-      {\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, \env_{3} - \{\overline{id}\}}
+      {\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, (\env_{4} - \{\overline{id}\})[\overline{id \mapsto \env_{2}(id)}]}
 \end{array}
@@
 before type checking $s$.
-We can define \emph{infer} as follows:
-\begin{align*}
-infer(T_{1} \times ... \times T_{n}{*}, i) & = \left\{
-\begin{array}{ll}
-general(T_{i}) & \text{if $i < n$}\\
-general(T_{n} \cup \Nil) & \text{if $i >= n$}
-\end{array} \right.
-\\ 
-general(\False) & = \Boolean\\
-general(\True) & = \Boolean\\
-general({\it int}) & = \Integer\\
-general({\it float}) & = \Number\\
-general({\it string}) & = \String\\
-general(T_{1} \cup T_{2}) & = general(T_{1}) \cup general(T_{2})\\
-general(S_{1} \rightarrow S_{2}) & = general2(S_{1}) \rightarrow general2(S_{2})\\
-general(\{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{tag}) & = \{K_{1}{:}general(V_{1}), ..., K_{n}{:}general(V_{n})\}_{tag}\\
-general(\mu x.T) & = \mu x.general(T)\\
-general(T) & = T\\
-\end{align*}
-
-\begin{align*}
-general2(\Void) & = \Void\\
-general2(T{*}) & = general(T){*}\\
-general2(T \times P) & = general(T) \times general2(P)\\
-general2(S_{1} \sqcup S_{2}) & = general2(S_{1}) \sqcup general2(S_{2})
-\end{align*}
 
@@
 \dfrac{\begin{array}{c}
-       \env_{1}(id) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
-       closeall(\env_{1}), \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
-       closeall(\env_{1}), \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
-       \env_{4} = openset(closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2})),rv(s_{1}) \cup rv(s_{2}))
+       \env_{1}(id) = \pi_{i}^{x}\\
+       S_{t} = fonilpt(\penv(x), i) \;\;\;
+       S_{e} = finilpt(\penv(x), i) \\
+       \env_{1}, \penv[x \mapsto S_{t}] \vdash s_{1}, \env_{2} \;\;\;
+       \env_{1}, \penv[x \mapsto S_{e}] \vdash s_{2}, \env_{3}\\
+       \env_{4} = join(\env_{2}, \env_{3})
       \end{array}}
@@
 
-Rule \textsc{T-IF3} uses the auxiliary functions \emph{fpt} and \emph{gpt}
+Rule \textsc{T-IF3} uses the auxiliary functions \emph{fonilpt} and \emph{finilpt}
 to filter a projection $x$, affecting all variables that bind to the same projection.
+The former function filters out the tuples that contain $\Nil$ in the $i$-th component,
+while the latter function filters out the tuples that do not contain $\Nil$ in the $i$-th component.
 For instance, our previous example type checks through rule \textsc{T-IF3},
@@
 \[
-fpt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
+fonilpt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
 \]
@@
 \[
-gpt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
+finilpt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
 \]
@@
 they project to $\Integer \cup \Nil$ and $\Integer \cup \String$, respectively.
+
+Typed Lua does not allow assignments to left-hand values that are bound
+to a projection type.
+This kind of assignment would be unsound, because it could break the
+dependency relation that the components of each tuple of the union have.
+For instance, the following example does not type check:
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{2}{l}{$\mathbf{local} \; q, r = idiv(1, 2) \; \mathbf{in}$}\\
+& \multicolumn{1}{l}{$r = ``foo";$}\\
+& \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; \lfloor print(q + r) \rfloor_{0} \; \mathbf{else} \; \lfloor print(``ERROR: " \; .. \; r) \rfloor_{0}$}
+\end{tabular}
+\end{center}
+
+In this example, the projected type of $r$ outsided of the $\mathbf{if}$
+statement is $\Integer \cup \String$, so the assignment looks fine.
+However, the projected type of $r$ inside the $\mathbf{if}$ branch is $\Integer$,
+not matching the $\String$ value that $r$ has after the assignment.

### 2015-06-28 doc/thesis/typedlua.tex
Updating strict versus default

@@
 and union types.
-Second-level types include the type void \texttt{()},
-vararg types, tuple types, and unions of tuple types.
+Second-level types include vararg types, tuple types, and unions of tuple types.
 
@@
 types in the definition of function types.
-A second-level type is either the type void \texttt{()},
-a tuple of first-level types optionally ending in a variadic type,
-or a union of these tuples.
-The type void \texttt{()} represents the type of the empty tuple.
+A second-level type is either a tuple of first-level types optionally
+ending in a variadic type or a union of these tuples.
 A variadic type \texttt{t*} is a generator for a sequence of values
 of the union type \texttt{t|nil}.
+The empty tuple \texttt{()} is syntactic sugar to \texttt{(nil*)},
+as we will show that Typed Lua always ends a tuple type in a variadic type.
 Union of tuple types appear in the return type of function types to
@@
 the \emph{strict} mode.
-In the default mode, the compiler adds a variadic tail to the
-type of the parameter list and to the return type if the
-programmer does not specify one.
-The default mode has this behavior to preserve the semantics
-of function calls in Lua, that is, it discards extra arguments and
-uses \texttt{nil} to replace missing arguments.
-In the strict mode, the compiler does not add a variadic tail
-to the type of the parameter list and to the return type.
-Instead, it preserves the type defined by the programmer to catch
-arity mismatch.
+In the default mode, the compiler adds \texttt{value*} to the type
+of the parameter list and \texttt{nil*} to the return type when
+the programmer does not specify a variadic tail. 
+It has this behavior to preserve the semantics
+of function calls in Lua, that is, it uses \texttt{value*} to discard
+extra arguments and \texttt{nil*} to replace missing arguments.
+In the strict mode, the compiler adds \texttt{nil*} to both
+parts of the function type.
+Even though it still uses \texttt{nil*} to replace missing values,
+it also uses \texttt{nil*} instead of \texttt{value*} to catch arity mismatch.
 
@@
 In the strict mode, \texttt{sum} has type
-\texttt{(integer, integer) -> (integer)}.
-The compiler does not add \texttt{value*} to the type of the parameter
+\texttt{(integer, integer, nil*) -> (integer, nil*)}.
+The compiler adds \texttt{nil*} to the type of the parameter
 list to catch arity mismatch.
@@
 The function call \texttt{v()} compiles without any compile-time warnings in
-the default and strict modes because \texttt{(value*)} and \texttt{()}
-are consistent with \texttt{(string*)}.
-The call \texttt{v("hello", "world")} compiles without any warnings
-in the default and strict modes because \texttt{("hello", "world", nil*)}
-and \texttt{("hello", "world")} are consistent with \texttt{(string*)}.
+both modes because the type of the argument list is \texttt{(nil*)},
+which is consistent with \texttt{(string*)}.
+The call \texttt{v("hello", "world")} also compiles without any warnings
+in both modes because the type of the argument list is \texttt{("hello", "world", nil*)},
+which is consistent with \texttt{(string*)}.
 Calling \texttt{v(...)} compiles without any warnings in both
@@
 The type of the function \texttt{message} is
-\texttt{(string, string|nil, value*) -> (string, nil*)}
-in the default mode, and
-\texttt{(string, string|nil) -> (string)}
+\texttt{(string, string|nil, value*) -> (string, nil*)} in the
+default mode, and
+\texttt{(string, string|nil, nil*) -> (string, nil*)}
 in the strict mode.
-The function call \texttt{message("Typed Lua")} compiles without
+The call \texttt{message("Typed Lua")} compiles without
 any compile-time warnings in both modes, because the argument list
@@
 in the default mode, and that is also consistent with the input type
-\texttt{(string, string|nil)} in the strict mode.
+\texttt{(string, string|nil, nil*)} in the strict mode.
 The compiler includes \texttt{nil*} in the tail of

### 2015-06-29 doc/thesis/system.tex
Starting to update typing rules after backport
Updating strict versus default
Minor fixes to chapter 4

@@
 We present the subtyping rules as a deduction system for the
-subtyping relation $\senv \vdash X \subtype Y$, where $X$ and $Y$
-are two types of the same kind.
+subtyping relation $\senv \vdash T_{1} \subtype T_{2}$,
+where $T_{1}$ and $T_{2}$ are two types of the same kind.
 The variable $\senv$ is a set of pairs of recursion variables.
@@
 
-The rule \textsc{S-VARARG1} handles the case where both tuple types end
-with variadic types:
-\[
-\begin{array}{c}
-\mylabel{S-VARARG2}\\
+The rule \textsc{S-VARARG1} handles subtyping between two
+variadic types:
+\[
+\begin{array}{c}
+\mylabel{S-VARARG1}\\
 \dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \cup \Nil}
@@
 
-The other rules handle the cases where only one tuple type ends with a variadic type:
+The other rules handle subtyping between a varidic type and a pair type:
 \[
@@
 
-These rules state the conditions when variadic tuples of different
+These rules state the conditions when tuple types of different
 length are compatible.
@@
 a variable with a table type that describes a hash.
+More precisely, this rule states that it is safe to recast the table type
+$\{``x":\Integer, ``y":\Integer, ``z":\Integer\}_{unique}$ to
+$\{\String:\Integer \cup \Nil\}_{closed}$, as long as the
+new type becomes inacessible with the original type.
 
@@
 \[
-\{``x":1, ``y":2\}_{unique} \subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
+\{``x":1, ``y":2\}_{unique} \subtype
+\{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
 \]
@@
 We present the consistent-subtyping rules as a deduction system for
-the consistent-subtyping relation $\senv \vdash X \lesssim Y$,
-where $X$ and $Y$ are two types of the same kind.
+the consistent-subtyping relation $\senv \vdash T_{1} \lesssim T_{2}$,
+where $T_{1}$ and $T_{2}$ are two types of the same kind.
 As in the subtyping relation, $\senv$ is also a set of pairs of
@@
 
-\subsection{Assignment and function application}
+\subsection*{Assignment and function application}
 \label{sec:assignment}
@@
 
-\subsection{Tables and refinement}
+\subsection*{Tables and refinement}
 \label{sec:refinement}
@@
 
-The rule \textsc{T-REFINE} uses the auxiliary function \emph{vt}
+The rule \textsc{T-REFINE1} uses the auxiliary function \emph{vt}
 to adjust the type of the value according to the type of the key.
@@
 
-\subsection{Projections}
+\subsection*{Projections}
 \label{sec:fap}

### 2015-07-01 doc/thesis/system.tex
Minor fixes to the section about table refinement

@@
 
-First, our type system uses the auxiliary relation
+First, Typed Lua uses the auxiliary relation
 $\env_{1}, \penv \vdash [e_{1}] = e_{2} : (F,V), \env_{2}$ to type check each
@@
 creation of unsafe aliases.
-The rules that type check table fields use the auxiliary function
-\emph{fix} to change any \emph{unique} and \emph{open} table types
+Due to this restriction, the rule that type check table fields
+use the auxiliary function \emph{fix} (in the definition of \emph{vt})
+to change any \emph{unique} and \emph{open} table types
 used in the field initializer to \emph{fixed}.
+Rule \textsc{T-FIELD} defines this behavior:
+\[
+\begin{array}{c}
+\mylabel{T-FIELD}\\
+\dfrac{\env_{1}, \penv \vdash e_{2}:V, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash e_{1}:F, \env_{3}}
+      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (F,vt(V)), \env_{3}}
+\end{array}
+\]
+
+The rule \textsc{T-FIELD} uses the auxiliary function \emph{vt}
+to adjust the type of the value according to the type of the key.
+More precisely, \emph{vt} includes the type $\Nil$ in the type of
+the value when the type of the key is not a literal type.
+It also uses \emph{fix} to prevent \emph{unique} and \emph{open}
+table types to appear in the type of the values.
+We define these functions as follows:
+\begin{align*}
+vt(L, V) & = fix(V) \\
+vt(F_1, F_2) & = nil(fix(F_2))\\
+vt(F_1, {\bf const}\;F_2) & = {\bf const}\;nil(fix(F_2))\\
+\\
+nil(T) & = T \;\;\; \mathrm{if}\;\Nil \lesssim T\\
+nil(T) & = T \cup \Nil \;\;\; \mathrm{otherwise}\\
+\\
+fix(F_{1} \cup F_{2}) & = fix(F_{1}) \cup fix(F_{2})\\
+fix(\{\overline{F{:}V}\}_{unique|open}) & = \{\overline{F{:}V}\}_{fixed} \\
+fix(F) & = F
+\end{align*}
 
 The rule \textsc{T-CONSTRUCTOR2} uses these steps to type check a
-table constructor with a non-empty list of table fields:
+table constructor with a list of table fields that do not end with
+an expression that potentially returns multiple values:
 \[
@@
 $\{1:``x", 2:``y", 3:``z"\}_{unique}$ through rule \textsc{T-CONSTRUCTOR2}.
+The subtyping rule for \emph{unique} table types allows us assigning
+this table to a variable with a more general type such as
+$\{1:\String, 2:\String, 3:\String\}_{closed}$ or even
+$\{\Integer:\String \cup \Nil\}_{closed}$.
 
@@
 
-The rule \textsc{T-REFINE1} uses the auxiliary function \emph{vt}
-to adjust the type of the value according to the type of the key.
-More precisely, \emph{vt} includes the type $\Nil$ in the type of
-the value when the type of the key is not a literal type.
-It is also this function that uses \emph{fix} to prevent
-\emph{unique} and \emph{open} table types to appear in the type
-of the values.
+The rule for refining \emph{open} table types is similar,
+changing only the tag in the type of $id$:
+\[
+\begin{array}{c}
+\mylabel{T-REFINE2}\\
+\dfrac{\begin{array}{c}
+       \env_{1}(id) = \{\overline{F{:}V}\}_{open}\\
+       \env_{1}, \penv \vdash e:F_{new}, \env_{2} \;\;\;
+       \nexists i \in 1..n \; F_{new} \lesssim F_{i} \;\;\;
+       V_{new} = vt(F_{new},V) \;\;\; n = |\overline{F{:}V}|
+       \end{array}}
+      {\env_{1}, \penv \vdash id[e] {<}V{>}:V_{new}, \env_{2}[id \mapsto \{\overline{F{:}V}, F_{new}{:}V_{new}\}_{open}]}
+\end{array}
+\]
+
+Our type system also includes analogous rules for adding methods
+to \emph{unique} and \emph{open} tables as a side-effect of
+type checking a method declaration, but we will not discuss them
+in this section for brevity.
 
@@
 to prevent the creation of unsafe aliases.
+If the coercion is to a \emph{closed} table type the type of the
+table changes to an \emph{open} table type with the same shape,
+but if the coercion is to a \emph{fixed} table type the table
+has to assume the same type.
 
@@
 
-Rule \textsc{T-FUNCTION1} illustrates this case:
+Rule \textsc{T-FUNCTION1} types non-variadic function declarations,
+and it illustrates this case:
 \[
@@
 \]
+
+%The rules for declaring variadic functions and recursive functions
+%are similar to \textsc{T-FUNCTION1}, and we will not
+%show them in this section for brevity.
 

### 2015-07-02 doc/thesis/system.tex
Minor fixes to the section about projection types

@@
 
-%The rules for declaring variadic functions and recursive functions
-%are similar to \textsc{T-FUNCTION1}, and we will not
-%show them in this section for brevity.
+The rules for declaring variadic functions and recursive functions
+are similar to \textsc{T-FUNCTION1}, and we will not
+show them in this section for brevity.
 
@@
 Lua programmers often overload the return type of functions to denote errors,
-and our type system uses projection types to handle this idiom.
+returning \texttt{nil} and an error message in case of error instead of
+the usual return values, and our type system uses projection types to handle this idiom.
 
@@
 into the union type $\Integer \cup \String$.
-Now, we may want to discriminate these variables to check whether
+Our type system defines this behavior through rule \textsc{T-IDREAD4},
+which uses the auxiliary function \emph{proj} to project an
+union of second-level types into an union of first-level types:
+\[
+\begin{array}{c}
+\mylabel{T-IDREAD4}\\
+\dfrac{\env_{1}(id) = \pi_{i}^{x}}
+      {\env_{1}, \penv \vdash id:proj(\penv(x), i), \env_{1}}
+\end{array}
+\]
+
+Now, we may want to discriminate $q$ and $r$ to check whether
 the function call returned with success.
-
 Introducing a projection variable $x$ in the projection environment allows our
-type system to discriminate projection types $\pi_{i}^{x}$.
+type system to discriminate projection types $\pi_{i}^{x}$,
+as they are a general way to not compromise the depency between
+the types of $q$ and $r$ after discriminating one of them,
+so flow typing can narrow the type of both variables by testing
+just one of them because the projection types of both variables
+bound to the same projection variable.
 The rule \textsc{T-IF3} shows the case where our type system

### 2015-07-06 doc/thesis/evaluation.tex
More improvements to the description of evaluation categories

@@
 The \emph{poly} category shows the percentage of members that
-we made minimal use of the dynamic type as a replacement for the
+we made minimal use of the dynamic type, as a replacement for the
 lack of type parameters.
@@
 standard library is in this category because it is a generic function.
-This function sorts a given list of elements, which is a generic list.
+It sorts a given list of elements, which is a generic list.
 However, we had to assign to this function the type
 \texttt{(\{any\}, nil|(any, any) -> (boolean)) -> ()} because
-Typed Lua does not have parametric polymorphism.
+Typed Lua does not support parametric polymorphism.
+A better type for it would be
+\texttt{(\{<T>\}, nil|(<T>, <T>) -> (boolean)) -> ()}.
 
 The \emph{over} category shows the percentage of members that
-we gave a static type that is not precise enough,
-as these members are overloaded functions that require intersection
-types to describe their precise static types.
+require intersection types to describe their precise static types,
+as they are overloaded functions.
 For instance, the function \texttt{math.abs} from the Lua
@@
 The \emph{hard} category shows the percentage of members that
-we could not give a precise static type.
-For instance, the function \texttt{assert} from the Lua
-standard library is in this category because it can return any number
-of values of any type, being difficult to type this behavior.
-Still, we gave this function the type \texttt{(value*) -> (any*)}.
+do not fit in one of the previous categories, as they are difficult to type.
+For instance, the function \texttt{string.format} from the Lua
+standard library is in this category because it relies on format
+strings, which are difficult to type.
+Still, we gave this function the type \texttt{(string, value*) -> (string)}.
 

### 2015-07-07 doc/thesis/review.tex
Moving statistics summary into a table

@@
 
-We end this chapter presenting a summary of the statistics that
-we just presented:
-\begin{itemize}
-\item table constructor
-\begin{itemize}
-\item only use literal keys: 73\%
-\item empty: 26\%
-\item only use non-literal keys: 1\%
-\end{itemize}
-\item table indexing
-\begin{itemize}
-\item reads: 86\%
-\begin{itemize}
-\item use literals: 93\%
-\item use non-literals: 7\%
-\end{itemize}
-\item writes: 14\%
-\begin{itemize}
-\item use literals: 71\%
-\item use non-literals: 29\%
-\end{itemize}
-\end{itemize}
-\item iteration over tables
-\begin{itemize}
-\item files that call \texttt{pairs}: 23\%
-\item files that call \texttt{ipairs}: 21\%
-\item files that use the numeric for along with the length operator: 10\%
-\end{itemize}
-\item function declarations
-\begin{itemize}
-\item perform dynamic type checking in their input parameters: 9\%
-\begin{itemize}
-\item asserting input parameters: 50\%
-\item overloading input parameters: 50\%
-\end{itemize}
-\item return multiple values: 10\%
-\begin{itemize}
-\item return \texttt{nil} plus something else: 50\%
-\item return \texttt{false} plus something else: 10\%
-\item return other multiple values: 40\%
-\end{itemize}
-\item are method declarations: 23\%
-\item are variadic: 8\%
-\end{itemize}
-\item object-oriented programming
-\begin{itemize}
-\item projects that extend metatables and declare methods: 45\%
-\end{itemize}
-\item modules
-\begin{itemize}
-\item files that define modules: 61\%
-\item files that are plain scripts: 39\%
-\end{itemize}
-\end{itemize}
+Table \ref{tab:statistics} summarizes the statistics that we presented in this section.
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{|l|p{6cm}|c|}
+\hline
+\multirow{5}{*}{table constructor}
+& create a record & 36\% \\
+\cline{2-3}
+& create a list & 29\% \\
+\cline{2-3}
+& create an empty table & 26\% \\
+\cline{2-3}
+& create a table with a record part and a list part & 8\% \\
+\cline{2-3}
+& create a table with non-literal keys & 1\% \\
+\hline
+\multirow{4}{*}{table access}
+& reading with literal keys & 80\% \\
+\cline{2-3}
+& writing with literal keys & 10\% \\
+\cline{2-3}
+& reading with non-literal keys & 6\% \\
+\cline{2-3}
+& writing with non-literal keys & 4\% \\
+\hline
+\multirow{2}{*}{iteration over tables} &
+files that iterate over a list & 27\% \\
+\cline{2-3}
+& files that iterate over a map & 23\% \\
+\hline
+\multirow{4}{*}{function declarations}
+& inspect the tags of their input parameters & 9\% \\
+\cline{2-3}
+& return multiple values to signal errors & 6\% \\
+\cline{2-3}
+& are variadic & 8\% \\
+\cline{2-3}
+& are method declarations & 23\% \\
+\hline
+\multirow{1}{*}{object-oriented programming}
+& projects that use metatables and declare methods & 45\% \\
+\hline
+\multirow{2}{*}{modules}
+& files that are modules & 61\% \\
+\cline{2-3}
+& files that are plain scripts & 39\% \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Summary of the statistics about the usage of Lua}
+\label{tab:statistics}
+\end{table}

### 2015-07-24 doc/thesis/glossary.tex
Fixing glossary definitions and adding page referenes to them

--- 2015-05-15 doc/thesis/glossary.tex
@@
 \begin{description}
-\item[bottom type] It is a type that is subtype of all types.
+\item[bottom type] A type that is subtype of all types.
 
-\item[closed table type] It is a table type that does not allow adding fields.
+\item[closed table type] A table type that does not provide any guarantees
+about keys with types not listed in the table type.
+See complete definition in page \pageref{def:tabletype}.
 
-\item[coercion] It is a relation that allows converting values from one type to values
-of another type without error.
+\item[coercion] A relation that allows converting values from one type to
+values of another type without error.
 
-\item[consistency] It is a relation that gradual typing uses to control the insertion of
-run-time checks between dynamically typed and statically typed code.
+\item[consistency] A relation used by gradual typing to check the interaction
+between the dynamic type and other types.
+See complete definition in page \pageref{def:consistency}.
 
-\item[consistent-subtyping] It is a relation that combines consistency and subtyping.
+\item[consistent-subtyping] A relation that combines consistency and subtyping.
+See complete definition in page \pageref{def:consistent-subtyping}.
 
-\item[contravariant] Subtyping is contravariant when it reverses the subtyping order,
-that is, it orders types from a more generic type to a more specific type.
+\item[contravariant] A part of the type constructor is contravariant when
+it reverses the subtyping order.
 
-\item[covariant] Subtyping is covariant when it preserves the subtyping order,
-that is, it orders types from a more specific type to a more general type.
+\item[covariant] A part of the type constructor is covariant when
+it preserves the subtyping order.
 
-\item[depth subtyping] It allows changing the type of individual fields of a table type.
+\item[depth subtyping] An operation that allows variance in the type of
+record fields.
 
-\item[dynamic type] It is a type that allows combining dynamic and static typing in the same code.
-It is neither the bottom nor the top type in the subtyping relation, but
-a subtype only of itself.
-Gradual typing uses the dynamic type along with the consistency relation to
-identify where run-time casts should be inserted to prevent that dynamically
-typed code violates statically typed code.
+\item[dynamic type] A type used by gradual typing to denote unknown values.
+See complete definition in page \pageref{def:dynamictype}.
 
-\item[filter type] It is a type that allows Typed Lua to discriminate the type of local variables
-inside control flow statements.
+\item[filter type] A type used by Typed Lua to discriminate the type of
+local variables inside control flow statements.
+See complete definition in page \pageref{def:filtertype}.
 
-\item[fixed table type] It is a table type that does not allow adding fields. It also does not allow
-width subtyping to make single inheritance safe.
+\item[fixed table type] A table type which guarantees that there are no
+keys with a type that is not one of its key types, and that can have
+any number of \emph{fixed} or \emph{closed} references point to it.
+See complete definition in page \pageref{def:tabletype}.
 
-\item[flow typing] It is a combination of static typing and flow analysis.
+\item[flow typing] An approach that combines static typing and flow analysis to
+allow variables to have different types at different parts of the program.
 
-\item[free assigned variable] It is a free variable that appears in an assignment.
+\item[free assigned variable] A free variable that appears in an assignment.
 
-\item[gradual type system] It is a type system that uses the consistency relation to check the interaction
-among the dynamic type and other types.
+\item[gradual type system] A type system that uses the consistency relation
+instead of type equality to perform static type checking.
+See complete definition in page \pageref{sec:gradual}.
 
-\item[gradual typing] It is an approach that uses a gradual type system to allow static and dynamic
-typing in the same code, but inserting run-time checks between statistically
-typed and dynamically typed code.
+\item[gradual typing] An approach that uses a gradual type system to allow
+static and dynamic typing in the same code, but inserting run-time checks
+between statically typed and dynamically typed code.
+See complete definition in page \pageref{sec:gradual}.
 
-\item[invariant] Subtyping is invariant when it does not allow ordering types, that is,
-it is a way that express type equality through subtyping.
+\item[invariant] A part of the type constructor is invariant when it forbids variance.
+It is also a way to define type equality through subtyping.
 
-\item[metatable] It is a Lua table that allows changing its behavior.
+\item[metatable] A Lua table that allows changing the behavior of other tables
+it is attached to.
 
-\item[nominal type system] It is a type system that uses name declarations to define the equivalence among types.
+\item[nominal type system] A type system that uses the type names to check the
+compatibility among them.
 
-\item[open table type] It is a table type that allows adding fields, and usually refers to table types that
-have aliases.
+\item[open table type] A table type which guarantees that there are no
+keys with a type that is not one of its key types, and that only have
+\emph{closed} references point to it.
+See complete definition in page \pageref{def:tabletype}.
 
-\item[optional type system] It is a type system that allows combining static and dynamic typing in the same
-language, but without affecting the run-time semantics.
+\item[optional type system] A type system that allows combining static and
+dynamic typing in the same language, but without affecting the run-time semantics.
+See complete definition in page \pageref{sec:optional}.
 
-\item[projection environment] It is a type environment that Typed Lua uses to assign projection variables to
-second-level types.
+\item[projection environment] An environment used by Typed Lua to handle unions of
+second-level types that are bound to projection types.
 
-\item[projection type] It is a type that allows Typed Lua to discriminate the type of local variables
-that have a dependency relation.
+\item[projection type] A type used by Typed Lua to discriminate the type of local
+variables that have a dependency relation.
+See complete definition in page \pageref{def:projectiontype}.
 
-\item[prototype object] It is an object that works like a class, that is, it is a prototype for creating
-new objects of a given class.
+\item[prototype object] An object that works like a class, that is, it is an object from
+which other objects inherit its attributes.
 
-\item[self-like delegation] It is a technique to implement inheritance in dynamically typed languages.
+\item[self-like delegation] A technique to implement inheritance in dynamically typed
+languages through prototype objects.
 
-\item[sound type system] It is a type system that does not type check all programs that contain a type error.
+\item[sound type system] A type system that does not type check all programs that contain a type error.
 
-\item[structural type system] It is a type system that uses the type structure to define the equivalence among types.
+\item[structural type system] A type system that uses type structures to check the compatibility among them.
 
-\item[table refinement] It is an operation that allows changing a table type to include new fields or
-to specialize existing fields.
+\item[table refinement] An operation from Typed Lua that allows programmers to change a table type
+to include new fields or to specialize existing fields.
+See complete definition in page \pageref{sec:refinement}.
 
-\item[top type] It is a type that is supertype of all types.
+\item[top type] A type that is supertype of all types.
 
-\item[type environment] It is an environment that assigns variable names to types.
+\item[type environment] An environment used by Typed Lua to assign variable names to first-level types.
 
-\item[type tag] In dynamically typed languages a type tag describes the type of a value during
-run-time.
+\item[type tag] A tag that describes the type of a value during run-time in dynamically
+typed languages.
 
-\item[unique table type] It is a table type that allows adding fields, and it is also the table type
-that describe the type of the table constructor.
+\item[unique table type] A table type which guarantees that there are no
+keys with a type that is not one of its key types, and that do not have
+any references point to it.
+See complete definition in page \pageref{def:tabletype}.
 
-\item[unsound type system] It is a type system that type checks certain programs that contain type errors.
+\item[unsound type system] A type system that type checks certain programs that contain type errors.
 
-\item[userdata] It is a way to define new types in an application or library that is written in C.
+\item[userdata] A Lua data type that allows Lua to hold values from applications
+or libraries that are written in C.
 
-\item[vararg expression] It is a Lua expression that can result in an arbitrary number of values.
+\item[vararg expression] A Lua expression that can result in an arbitrary number of values.
 
-\item[variadic function] It is a Lua function that can receive an arbitrary number of arguments.
+\item[variadic function] A function that can receive an arbitrary number of arguments.
 
-\item[variance] It is the way types are ordered.
+\item[variance] A way to define the subtyping order between the components
+of a type constructor.
 
-\item[width subtyping] It allows adding more fields to a table type.
+\item[width subtyping] An operation that allows adding fields to a record.
 

### 2015-07-24 doc/thesis/review.tex
Fixing glossary definitions and adding page referenes to them

@@
 
+\label{def:dynamictype}
 The main goal of gradual typing \cite{siek2006gradual} is to allow
@@
 
+\label{def:consistency}
 The central idea of gradual typing is the \emph{consistency}
@@
 
+\label{def:consistent-subtyping}
 Later, Siek and Taha \cite{siek2007objects} reported that the consistency relation

### 2015-07-24 doc/thesis/system.tex
Fixing glossary definitions and adding page referenes to them

@@
 
+\label{def:tabletype}
 We also use the tags \emph{unique}, \emph{open}, \emph{fixed}, and \emph{closed}
@@
 
+\label{def:filtertype}
 Typed Lua includes filter types as a way to discriminate the type of local
@@
 
+\label{def:projectiontype}
 Typed Lua includes projection types as a way to project
@@
 \begin{array}{c}
-\mylabel{S-PAIR}\\
+\mylabel{S-PAIR1}\\
 \dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
@@
 
-Filter types are subtypes only of themselves and of $\Value$.
+Filter types are subtypes only of themselves.
 More precisely, a filter type $\phi(F_{1},F_{2})$ is a subtype of
 the same filter type $\phi(F_{1},F_{2})$, which shares the same
-types $F_{1}$ and $F_{2}$, and it is also a subtype of $\Value$.
-
-Projection types are subtypes only of themselves and of $\Value$.
+types $F_{1}$ and $F_{2}$.
+
+Projection types are subtypes only of themselves.
 More precisely, a projection type $\pi_{i}^{x}$ is a subtype of the
 same projection type $\pi_{i}^{x}$, which shares the same union of
-tuples $x$ and the same index $i$, and it is also a subtype of $\Value$.
+tuples $x$ and the same index $i$.
+
+The subtyping rules for expression list types are similar to the
+subtyping rules for tuple types.
 
@@
 
-\section{Typing rules}
+\section{Type checking}
 \label{sec:rules}

### 2015-07-24 doc/thesis/system.tex
Fixing rules to use the new syntactic categories

@@
        \env_{2}, \penv \vdash e_{1}:F, \env_{3}}
-      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (F,vt(V)), \env_{3}}
+      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (F,vt(F,V)), \env_{3}}
 \end{array}
@@
 
-The rule \textsc{T-CONSTRUCTOR2} uses these steps to type check a
+The rule \textsc{T-CONSTRUCTOR1} uses these steps to type check a
 table constructor with a list of table fields that do not end with
@@
 \begin{array}{c}
-\mylabel{T-CONSTRUCTOR2}\\
+\mylabel{T-CONSTRUCTOR1}\\
 \dfrac{\begin{array}{c}
@@
 $\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ has type
-$\{1:``x", 2:``y", 3:``z"\}_{unique}$ through rule \textsc{T-CONSTRUCTOR2}.
+$\{1:``x", 2:``y", 3:``z"\}_{unique}$ through rule \textsc{T-CONSTRUCTOR1}.
 The subtyping rule for \emph{unique} table types allows us assigning
@@
 has type $\{``x":1, ``y":\{``z":2\}_{fixed}\}_{unique}$ through rule
-\textsc{T-CONSTRUCTOR2}.
+\textsc{T-CONSTRUCTOR1}.
 The inner table is \emph{fixed} to prevent the creation of unsafe aliases.
@@
 For this reason, identifiers that appear in the left-hand side
-of assignements have their own rule \textsc{T-IDWRITE}:
-\[
-\begin{array}{c}
-\mylabel{T-IDWRITE}\\
+of assignements have their own rule \textsc{T-IDWRITE1}:
+\[
+\begin{array}{c}
+\mylabel{T-IDWRITE1}\\
 \dfrac{\env_{1}(id) = F}

### 2015-07-27 doc/thesis/typedlua.tex
Fixing text to mention that tables are tagged and not variables

@@
 
-Typed Lua tags a variable that holds a table type as either
-\emph{unique}, \emph{open}, \emph{fixed}, or \emph{closed}.
-If a variable gets its type from a table constructor then it is
-\emph{unique}, otherwise it is \emph{closed}.
-Any alias to \emph{unique} variables makes them \emph{open}, as we use the
-tag \emph{open} to keep track of \emph{unique} variables that are aliased.
+Typed Lua tags table types as either \emph{unique}, \emph{open},
+\emph{fixed}, or \emph{closed} to identify which operations are safe.
+The table constructor describes \emph{unique} table types,
+as they are tables that we know all the fields and that there is
+no alias to it.
+Any alias to \emph{unique} tables makes them \emph{open},
+as we use this tag to keep track of \emph{unique} aliasing.
 In the previous example, \texttt{person} has an \emph{unique} type
@@
 \emph{fixed} to describe the type of classes.
+Type annotations define \emph{closed} table types, as this tag
+describes table types that do not provide any guarantees about keys
+that do not appear in the table type.
 Typed Lua also has different subtyping rules that handle these
@@
 allow field assignment to change the type of \emph{unique} and
-\emph{open} variables:
-the variable must be local to the current block,
+\emph{open} tables:
+the table must assigned to a local variable to the current block,
 the new type must only add new fields, and
@@
 
-Before discussing each one of these rules, it is worth discussing
-why Typed Lua needs \emph{unique} and \emph{open} table types.
 Even though both \emph{unique} and \emph{open} table types
@@
 increase usability and to avoid some unsafe behaviors.
-
 If we did not have \emph{unique} table types and \emph{open}
@@
 
-We can create aliases to \emph{open} variables, but they are always \emph{closed}.
+We can create aliases to \emph{open} tables, but they are always \emph{closed}.
 Any mutation in the original reference and in the aliased reference

### 2015-07-29 doc/thesis/system.tex
Several fixes to the text about typing rules

@@
 Each $V_{i}$ represents the type of the value that table keys of type $F_{i}$ map to.
-Value types represent mutable values by default, but we can use the
-$\Const$ type to make them represent immutable values.
+Value types represent mutable fields by default, but we can use the
+$\Const$ type to make them represent immutable fields.
+Making a field $\Const$ does not guarantee that its value cannot change,
+as the table may have aliases with a non-$\Const$ type for that field.
+Typed Lua needs immutable fields to enable depth subtyping between table types.
 
@@
 to classify table types.
-The tag \emph{unique} represents tables with no keys that do not
-inhabit one of the table's key types, and with no alias.
-In particular, the type of the table constructor has this tag.
-The tag \emph{open} represents \emph{unique} table types that
-have at least one alias.
-The tag \emph{fixed} represents \emph{unique} table types that we
-do not know how many aliases they have.
-In particular, the type of a class has this tag.
 The tag \emph{closed} represents table types that do not provide
@@
 declarations, and userdata declarations always describe \emph{closed} table types.
-In the next sections we explain in more detail why we need
-different table types.
+The tags \emph{unique}, \emph{open}, and \emph{fixed} represent
+tables with no keys that do not inhabit one of the table's key types,
+but with different guarantees about the reference to a value of
+that type.
+A reference to an \emph{unique} table is guaranteed to point
+to a table that has no other references to it.
+In particular, the type of the table constructor has this tag,
+as it allows greater flexibility in reinterpreting its type.
+A reference to an \emph{open} table is guaranteed to have only
+\emph{closed} references pointing to the same table,
+a guarantee that still lets the type system reinterpret the type
+of a reference, but with more restrictions.
+A reference to a \emph{fixed} table can have any number of
+\emph{fixed} or \emph{closed} references point to it,
+so its type cannot change anymore.
+In particular, the type of a class has this tag in our type system.
+
+A \emph{fixed} table type guarantees that there are
+no keys with a type that is not one of its key types.
+Even though this guarantee allows type-safe iteration on
+\emph{fixed} table types, it forbids width subtyping that
+is necessary for object-oriented programming, so \emph{closed}
+table types remove this guarantee to allow width subytping
+between other table types and \emph{closed} table types.
+This means that objects have \emph{closed} table types,
+while their classes have \emph{fixed} table types.
 
@@
 
-With the recursive types we finish the dicussion about Typed Lua types,
-and we begin the discussion about the special types.
+With recursive types we finish the dicussion about Typed Lua types,
+and we begin the discussion about special types.
 
@@
 This means that \texttt{type} predicates use filter types of the form
-$\phi(T_{1},T_{2})$ to discriminate local variables that are bound to
+$\phi(F_{1},F_{2})$ to discriminate local variables that are bound to
 union types.
-In a filter type $\phi(T_{1},T_{2})$, $T_{1}$ is the original type and
-$T_{2}$ is the discriminated type.
+In a filter type $\phi(F_{1},T_{2})$, $F_{1}$ is the original type and
+$F_{2}$ is the discriminated type.
 
@@
 \mylabel{T-EXPLIST1}\\
-\dfrac{\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
+\dfrac{\env_{1}, \penv \vdash e_{i}:F_{i}, \env_{i+1} \;\;\;
        \env_{f} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
        n = |\;\overline{e}\;|}
-      {\env_{1}, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{f}}
+      {\env_{1}, \penv \vdash \overline{e}:F_{1} \times ... \times F_{n} \times \Nil{*}, \env_{f}}
 \end{array}
@@
 \mylabel{T-LHSLIST}\\
-\dfrac{\env_{1}, \penv \vdash l_{i}:T_{i}, \env_{i+1} \;\;\;
+\dfrac{\env_{1}, \penv \vdash l_{i}:F_{i}, \env_{i+1} \;\;\;
        \env_{f} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
        n = |\;\overline{l}\;|}
-      {\env_{1}, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{f}}
+      {\env_{1}, \penv \vdash \overline{l}:F_{1} \times ... \times F_{n} \times \Value{*}, \env_{f}}
 \end{array}
@@
 \mylabel{T-ASSIGNMENT}\\
-\dfrac{\env_{1}, \penv \vdash el:E_{1}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash \overline{l}:E_{2}, \env_{3} \;\;\;
-       E_{1} \lesssim E_{2}}
+\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash \overline{l}:S_{2}, \env_{3} \;\;\;
+       S_{1} \lesssim S_{2}}
       {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
@@
 \mylabel{T-APPLY1}\\
-\dfrac{\env_{1}, \penv \vdash e:E_{1} \rightarrow E_{2}, \env_{2} \;\;\;
-       \env_{2}, \penv \vdash el:E_{3}, \env_{3} \;\;\;
-       E_{3} \lesssim E_{1}}
-      {\env_{1}, \penv \vdash e(el):E_{2}, \env_{3}}
+\dfrac{\env_{1}, \penv \vdash e:S_{1} \rightarrow S_{2}, \env_{2} \;\;\;
+       \env_{2}, \penv \vdash el:S_{3}, \env_{3} \;\;\;
+       S_{3} \lesssim S_{1}}
+      {\env_{1}, \penv \vdash e(el):S_{2}, \env_{3}}
 \end{array}
@@
 while the rule \textsc{T-EXPAPPLY1} uses the auxiliary function
-\emph{first} to ensure that only one value is produced:
+\emph{proj} to ensure that only the first value is produced:
 \[
@@
 \mylabel{T-STMAPPLY1}\\
-\dfrac{\env_{1}, \penv \vdash e(el):E, \env_{2}}
+\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
       {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{0},\env_{2}}
@@
 \mylabel{T-EXPAPPLY1}\\
-\dfrac{\env_{1}, \penv \vdash e(el):E, \env_{2}}
-      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{1}:first(E), \env_{2}}
-\end{array}
-\end{array}
-\]
-
-We can define \emph{first} inductively as follows:
+\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
+      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{1}:proj(S,1), \env_{2}}
+\end{array}
+\end{array}
+\]
+
+We can define \emph{proj} inductively as follows:
 \begin{align*}
-first(F{*}) & = F \cup \Nil\\
-first(F \times P) & = F\\
-first(S_{1} \sqcup S_{2}) & = first(S_{1}) \cup first(S_{2})
+proj(S_1 \sqcup S_2, i) & = proj(S_1, i) \cup proj(S_2, i) \\
+proj(F{*}, i) & = nil(F) \\
+proj(F \times P, 1) & = F \\
+proj(F \times P, i) & = proj(P, i-1)\\
+proj(E{*}, i) & = nil(E) \\
+proj(T \times E, 1) & = T \\
+proj(T \times E, i) & = proj(E, i-1) \\
+nil(T) & = \left\{
+\begin{array}{ll}
+T & \text{if $\Nil \lesssim T$} \\
+T \cup \Nil & \text{otherwise}
+\end{array} \right.
 \end{align*}
@@
 A pair $(F,V)$ means that $e_{1}$ has type $F$ and $e_{2}$ has type $V$,
-where $F$ is the type of the key and $V$ is the type of the value.
+where $F$ is the type of the key and $V$ is the type of the field value.
 
@@
 uses the predicate \emph{wf} to check whether this table type is well-formed.
+The predicate \emph{wf} also uses the auxiliary predicate \emph{tag} to
+forbid \emph{unique} and \emph{open} fields.
 Formally, a table type is well-formed if it obeys the following rule:
@@
 wf(F \times P) &=& wf(F) \,\wedge\,wf(P)\\
-wf(F) & = & \top \;\;\;\mathrm{for\; all\; other\; cases}\\\\
-tag(F_1 \cup F_2, t) & = & tag(F_1, t) \,\vee\, tag(F_2,t) \\
-tag(\{\overline{F:V}\}_t, t) & = & \top\\
+wf(F) & = & \top \;\;\;\mathrm{for\; all\; other\; cases}
 \end{array}
@@
 Well-formed table types also do not allow \emph{unique} and
-\emph{open} table types to appear in the type of the values.
+\emph{open} table types to appear in the type of the field values.
 We made this restriction because our type system does not keep
@@
 It also uses \emph{fix} to prevent \emph{unique} and \emph{open}
-table types to appear in the type of the values.
+table types to appear in the type of the field values.
 We define these functions as follows:
@@
 \\
-nil(T) & = T \;\;\; \mathrm{if}\;\Nil \lesssim T\\
-nil(T) & = T \cup \Nil \;\;\; \mathrm{otherwise}\\
-\\
 fix(F_{1} \cup F_{2}) & = fix(F_{1}) \cup fix(F_{2})\\
@@
 \dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(F_{i},V_{i}), \env_{i+1} \\
-       T = \{F_{1}{:}V_{1}, ..., F_{n}{:}V_{n}\}_{unique} \;\;\;
+       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(F_{i},V_{i}), \env_{i+1} \;\;\;
+       T = \{F_{1}{:}V_{1}, ..., F_{n}{:}V_{n}\}_{unique} \\
        wf(T) \;\;\;
-       n = |\;\overline{[e_{1}] = e_{2}}\;| \\
+       n = |\;\overline{[e_{1}] = e_{2}}\;| \;\;\;
        \env_{f} = merge(\env_{1}, ..., \env_{n+1})
@@
 new global variables.
-In Lua, the assignment \texttt{v = v + 1} translates to
+In Lua, the assignment \texttt{v = v + 1} translates to the statement
 \texttt{\string_ENV["v"] = \string_ENV["v"] + 1} when \texttt{v}
@@
 The rules for declaring variadic functions and recursive functions
-are similar to \textsc{T-FUNCTION1}, and we will not
-show them in this section for brevity.
+are similar to \textsc{T-FUNCTION1}, and we did not discuss them in
+this section for brevity.
 
@@
 in the environment.
-The function \emph{idiv} has type
+As we mentioned in Section \ref{sec:unions}, \emph{idiv} performs
+integer division and has type
 \[
@@
 \]
-As we mentioned in Section \ref{sec:unions}, \emph{idiv} performs
-integer division.
 In case of success, it returns two integers: the result and the remainder.
@@
 $\Value{*} \rightarrow \Nil{*}$.
+Let us also assume that $a$ and $b$ are local variables in the environment,
+and that both have type $\Integer$.
 Let us see how our type system type checks the following program:
@@
 \begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; q, r = idiv(1, 2) \; \mathbf{in}$}\\
+\multicolumn{2}{l}{$\mathbf{local} \; q, r = idiv(a, b) \; \mathbf{in}$}\\
 & \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; \lfloor print(q + r) \rfloor_{0} \; \mathbf{else} \; \lfloor print(``ERROR: " \; .. \; r) \rfloor_{0}$}
@@
 $\env_{1}, \penv \vdash el : E, \env_{2}, (x,S)$
-for type checking $idiv(1, 2)$.
+for type checking $idiv(a, b)$.
 This relation means that given a type environment $\env_{1}$ and
@@
 types $\pi_{i}^{x}$.
-In our example, our type system uses rule \textsc{T-EXPLIST4} for
-type checking $idiv(1, 2)$:
-\[
-\begin{array}{c}
-\mylabel{T-EXPLIST4}\\
+In our example, our type system uses rule \textsc{T-EXPLIST3} for
+type checking $idiv(a, b)$:
+\[
+\begin{array}{c}
+\mylabel{T-EXPLIST3}\\
 \dfrac{\begin{array}{c}
-       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
-       \env_{1}, \penv \vdash me:S, \env_{n+2} \;\;\;
-       n = |\;\overline{e}\;| \\
-       S = T_{n+1} \times ... \times T_{n+m} \times \Nil{*} \sqcup T_{n+1}' \times ... \times T_{n+m}' \times \Nil{*} \;\;\;
-       x \notin \penv \\
-       \env_{f} = merge(\env_{1}, ..., \env_{n+2})
+       \env_{1}, \penv \vdash e_{i}:F_{i}, \env_{i+1} \;\;\;
+       \env_{1}, \penv \vdash me:S, \env_{n+2}\\
+       S = F_{n+1} \times ... \times F_{n+m} \times \Nil{*} \sqcup F_{n+1}' \times ... \times F_{n+m}' \times \Nil{*} \\
+       \env_{f} = merge(\env_{1}, ..., \env_{n+2}) \;\;\;
+       n = |\;\overline{e}\;|
        \end{array}}
-      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times \Nil{*}, \env_{f}, (x,S)}
-\end{array}
-\]
-
-Note that $idiv(1, 2)$ has type
+      {\env_{1}, \penv \vdash \overline{e},me:F_{1} \times ... \times F_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times \Nil{*}, \env_{f}, (x,S)}
+\end{array}
+\]
+
+Note that $idiv(a, b)$ has type
 $\pi_{1}^{x} \times \pi_{2}^{x} \times \Nil{*}$ and produces the pair
@@
 \begin{tabular}{ll}
-\multicolumn{2}{l}{$\mathbf{local} \; q, r = idiv(1, 2) \; \mathbf{in}$}\\
+\multicolumn{2}{l}{$\mathbf{local} \; q, r = idiv(a, b) \; \mathbf{in}$}\\
 & \multicolumn{1}{l}{$r = ``foo";$}\\

### 2015-07-30 doc/thesis/system.tex
More fixes to the text about typing rules

@@
 
-With recursive types we finish the dicussion about Typed Lua types,
+With recursive types we finish the discussion about Typed Lua types,
 and we begin the discussion about special types.
@@
 $\{\String:\Integer \cup \Nil\}_{closed}$, as long as the
-new type becomes inacessible with the original type.
+new type becomes inaccessible with the original type.
 
@@
 We do not need to close \emph{unique} and \emph{open} tables that
-appear in the left-hand side of assignments, because \textsc{T-IDREAD}
-is sufficient to forbid the creation of an alias to another \emph{unique}
-or \emph{open} table.
+appear in the left-hand side of assignments, because \textsc{T-IDREAD1}
+and \textsc{T-IDREAD2} are sufficient to forbid the creation of an alias
+to another \emph{unique} or \emph{open} table.
 For this reason, identifiers that appear in the left-hand side
-of assignements have their own rule \textsc{T-IDWRITE1}:
+of assignments have their own rule \textsc{T-IDWRITE1}:
 \[
@@
 This rule uses the auxiliary function \emph{infer} to get the
-most general first-level types of each variable that should be introduced in
+most general types of each variable that should be introduced in
 the type environment for type checking $s$.
@@
 before type checking $s$.
+We can define \emph{infer} as follows:
+\begin{align*}
+infer(T_{1} \times ... \times T_{n}{*}, i) & = \left\{
+\begin{array}{ll}
+general(T_{i}) & \text{if $i < n$}\\
+general(nil(T_{n})) & \text{if $i >= n$}
+\end{array} \right.
+\end{align*}
+
+\begin{align*}
+general(\False) & = \Boolean\\
+general(\True) & = \Boolean\\
+general({\it int}) & = \Integer\\
+general({\it float}) & = \Number\\
+general({\it string}) & = \String\\
+general(F_{1} \cup F_{2}) & = general(F_{1}) \cup general(F_{2})\\
+general(S_{1} \rightarrow S_{2}) & = general2(S_{1}) \rightarrow general2(S_{2})\\
+general(\{F_{1}{:}V_{1}, ..., F_{n}{:}V_{n}\}_{tag}) & = \{F_{1}{:}general(V_{1}), ..., F_{n}{:}general(V_{n})\}_{tag}\\
+general(\mu x.F) & = \mu x.general(F)\\
+general(T) & = T\\
+\\
+general2(F{*}) & = general(F){*}\\
+general2(F \times P) & = general(F) \times general2(P)\\
+general2(S_{1} \sqcup S_{2}) & = general2(S_{1}) \sqcup general2(S_{2})
+\end{align*}
 
@@
 type system to discriminate projection types $\pi_{i}^{x}$,
-as they are a general way to not compromise the depency between
+as they are a general way to not compromise the dependency between
 the types of $q$ and $r$ after discriminating one of them,
@@
 bound to the same projection variable.
-The rule \textsc{T-IF3} shows the case where our type system
-discriminates a projection type based on the tag \texttt{nil}:
-\[
-\begin{array}{c}
-\mylabel{T-IF3}\\
+
+The rule \textsc{T-IF5} shows the case where our type system
+discriminates a projection type based on the tag \texttt{nil}.
+It uses the auxiliary functions \emph{fopt} and \emph{fipt}
+to filter a projection $x$, affecting all variables that bind to the same projection.
+More precisely, the former function filters out the tuples that contain a type $\Nil$ in the $i$-th component,
+while the latter function filters out the tuples that do not contain $\Nil$ in the $i$-th component.
+We define \textsc{T-IF5} as follows:
+\[
+\begin{array}{c}
+\mylabel{T-IF5}\\
 \dfrac{\begin{array}{c}
-       \env_{1}(id) = \pi_{i}^{x}\\
-       S_{t} = fonilpt(\penv(x), i) \;\;\;
-       S_{e} = finilpt(\penv(x), i) \\
-       \env_{1}, \penv[x \mapsto S_{t}] \vdash s_{1}, \env_{2} \;\;\;
+       \env_{1}(id) = \pi_{i}^{x} \\
+       S_{t} = fopt(\penv(x), \Nil, i) \;\;\;
+       S_{e} = fipt(\penv(x), \Nil, i) \\
+       \env_{1}, \penv[x \mapsto S_{t}] \vdash s_{1}, \env_{2}\\
        \env_{1}, \penv[x \mapsto S_{e}] \vdash s_{2}, \env_{3}\\
@@
 
-Rule \textsc{T-IF3} uses the auxiliary functions \emph{fonilpt} and \emph{finilpt}
-to filter a projection $x$, affecting all variables that bind to the same projection.
-The former function filters out the tuples that contain $\Nil$ in the $i$-th component,
-while the latter function filters out the tuples that do not contain $\Nil$ in the $i$-th component.
-For instance, our previous example type checks through rule \textsc{T-IF3},
+As an example, our previous example type checks through rule \textsc{T-IF5},
 because it uses the information provided by the projection type $\pi_{1}^{x}$,
-which is the type of $q$, to make the rule \textsc{T-IF3} use the function call
-\[
-fonilpt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
+which is the type of $q$, to make the rule \textsc{T-IF5} use the function call
+\[
+fopt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
 \]
@@
 \[
-finilpt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
+fipt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
 \]
@@
 they project to $\Integer \cup \Nil$ and $\Integer \cup \String$, respectively.
+
+Our type system also includes rules that check whether a branch is unreachable.
+Rules \textsc{T-IF6} and \textsc{T-IF7} respectively cover the case where the
+$\mathbf{else}$ branch is unreachable and the case where the $\mathbf{then}$
+branch is unreachable, because either the projected type of $\pi_{i}^{x}$ is
+not a supertype of $\Nil$ or it is $\Nil$.
+We define these rules as follows:
+\[
+\begin{array}{c}
+\begin{array}{c}
+\mylabel{T-IF6}\\
+\dfrac{\begin{array}{c}
+       \env_{1}(id) = \pi_{i}^{x} \\
+       S_{t} = fopt(\penv(x), \Nil, i) \\
+       fit(proj(\penv(x), i), \Nil) = \Void \\
+       \env_{1}, \penv[x \mapsto S_{t}] \vdash s_{1}, \env_{2}
+      \end{array}}
+      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{2}}
+\end{array}
+\;
+\begin{array}{c}
+\mylabel{T-IF7}\\
+\dfrac{\begin{array}{c}
+       \env_{1}(id) = \pi_{i}^{x} \\
+       S_{e} = fipt(\penv(x), \Nil, i) \\
+       fot(proj(\penv(x), i), \Nil) = \Void \\
+       \env_{1}, \penv[x \mapsto S_{e}] \vdash s_{2}, \env_{2}
+      \end{array}}
+      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{2}}
+\end{array}
+\end{array}
+\]
 
@@
 
-In this example, the projected type of $r$ outsided of the $\mathbf{if}$
+In this example, the projected type of $r$ outside of the $\mathbf{if}$
 statement is $\Integer \cup \String$, so the assignment looks fine.

### 2015-07-31 doc/thesis/glossary.tex
Minor fixes to thesis text

@@
 
-\item[depth subtyping] An operation that allows variance in the type of
+\item[depth subtyping] An operation over records that allows variance in the type of
 record fields.
@@
 
-\item[type environment] An environment used by Typed Lua to assign variable names to first-level types.
+\item[type environment] An environment that binds variable names to types.
 
@@
 
-\item[variance] A way to define the subtyping order between the components
+\item[variance] A property that defines the subtyping order between the components
 of a type constructor.
 
-\item[width subtyping] An operation that allows adding fields to a record.
+\item[width subtyping] An operation over records that allows the subtype to include
+fields that do not exist in the supertype.
 

### 2015-07-31 doc/thesis/system.tex
Minor fixes to thesis text

@@
 
-In Section \ref{sec:refinement} we will show that table refinement can
+Later, in this section we will show that table refinement can
 change the type environment while typing an expression or a left-hand value.
@@
 all environment changes in a new environment $\env_{f}$, if there are no conflicts.
-In Section \ref{sec:refinement} we will also show that we can only change the
+We will also show that we can only change the
 environment to add new table fields in a table type, and we cannot change the
@@
 
-Note that the coercion rules only allows changing the type
+Note that the coercion rules only allow changing the type
 of a variable if the new type is a supertype of the previous type,
@@
 After type checking the statement $s$, rule \textsc{T-LOCAL2} produces a
-new type environment $\env_{3}$ without the variables that it introduced
+new type environment $\env_{4}$ without the variables that it introduced
 before type checking $s$.
@@
 to filter a projection $x$, affecting all variables that bind to the same projection.
-More precisely, the former function filters out the tuples that contain a type $\Nil$ in the $i$-th component,
-while the latter function filters out the tuples that do not contain $\Nil$ in the $i$-th component.
+More precisely, the former function filters out the tuples that contain a type $F$ in the $i$-th component,
+while the latter function filters out the tuples that do not contain $F$ in the $i$-th component.
 We define \textsc{T-IF5} as follows:
@@
 
-As an example, our previous example type checks through rule \textsc{T-IF5},
+Our previous example type checks through rule \textsc{T-IF5},
 because it uses the information provided by the projection type $\pi_{1}^{x}$,

### 2015-08-02 doc/thesis/ack.tex
Adding acks and short resume

First of all I would like to thank three indispensable people in my life:
my wife, my mother, and my father.
Izabella is my inexhaustible source of inspiration, and she is
always pushing me forward.
My mother is a very brave person, and she inspires me to never give up.
My father is a very kind person, and he inspires me to try being more sensitive.
I am thankful to them not only because the support they gave me while
writing this thesis, but also because they are the reason of my life.

This work would have not happened without all the help from my advisors.
Roberto has a singular view of computer science that pushed me to
pursuit the best I could do, while Fabio gave me the confidence I
needed about the work we were doing.
I am glad that both of them had patience to guide me through this
work, and for all the nice meetings we had that led me to learn
a lot of new concepts and ideas.

It is hard to complete a thesis without having some leisure moments,
so I thank my friend and roommate André Ramiro for all the crazy and
fun moments that we had, which helped me to keep focused during working hours.

I also thank my friends at LabLua for the nice work environment,
specially Hisham and Pablo for all the support during the
hardest moments.

It may look easy moving to Rio, as it is a sunny place with lots
of things to do, but it may not be as easy as it looks for
someone that grew up in a cloudy and rainy town.
Thus, I thank my cousin Rebecca and my aunt Nazira for all the help
and affection they gave me during my adaptation in Rio, which
were fundamental to keep my goals in mind.

I thank the guys from Rio de Janeiro Fixed Gear (RJFG) for all
the fun we had during the night rides that I participated,
and for helping me to discover Rio.

I am also most grateful to the professors Ana Lúcia de Moura,
Anamaria Martins Moreira, Edward Hermann Haeusler, and
Roberto da Silva Bigonha for their careful review that helped
to improve this work.

Finally, I would like to thank CAPES, Google Summer of Code, and PUC-Rio
for partially funding this work.

### 2015-08-13 doc/thesis/glossary.tex
Fixing some typos and some glossary definitions

@@
 
-\item[contravariant] A part of the type constructor is contravariant when
-it reverses the subtyping order.
+\item[contravariant] A part of a type constructor is contravariant when it
+reverses the subtyping order, that is, the part $T_{1}$ of a type being
+a subtype of the corresponding part $T_{2}$ of another type implies that
+$T_{2} \subtype T_{1}$.
 
-\item[covariant] A part of the type constructor is covariant when
-it preserves the subtyping order.
+\item[covariant] A part of a type constructor is covariant when it
+preserves the subtyping order, that is, the part $T_{1}$ of a type being
+a subtype of the corresponding part $T_{2}$ of another type implies that
+$T_{1} \subtype T_{2}$.
 
-\item[depth subtyping] An operation over records that allows variance in the type of
-record fields.
+\item[depth subtyping] A subtyping relation over records that allows variance
+in the type of record fields.
 
@@
 keys with a type that is not one of its key types, and that can have
-any number of \emph{fixed} or \emph{closed} references point to it.
+any number of \emph{fixed} or \emph{closed} references.
 See complete definition in page \pageref{def:tabletype}.
@@
 
-\item[gradual type system] A type system that uses the consistency relation
-instead of type equality to perform static type checking.
+\item[gradual type system] A type system that uses either the consistency relation
+or the consistent-subtyping relation instead of type equality to perform
+static type checking.
 See complete definition in page \pageref{sec:gradual}.
@@
 
-\item[invariant] A part of the type constructor is invariant when it forbids variance.
+\item[invariant] A part of a type constructor is invariant when it forbids variance,
+that is, the part $T_{1}$ of a type being a subtype of the corresponding part $T_{2}$ of
+another type implies that $T_{1} \subtype T_{2}$ and $T_{2} \subtype T_{1}$.
 It is also a way to define type equality through subtyping.
@@
 keys with a type that is not one of its key types, and that only have
-\emph{closed} references point to it.
+\emph{closed} references.
 See complete definition in page \pageref{def:tabletype}.
@@
 \item[unique table type] A table type which guarantees that there are no
-keys with a type that is not one of its key types, and that do not have
-any references point to it.
+keys with a type that is not one of its key types, and that has only one reference.
 See complete definition in page \pageref{def:tabletype}.
@@
 
-\item[variadic function] A function that can receive an arbitrary number of arguments.
+\item[variadic function] A function that can use an arbitrary number of arguments.
 
@@
 
-\item[width subtyping] An operation over records that allows the subtype to include
-fields that do not exist in the supertype.
+\item[width subtyping] A subtyping relation over records that allows the subtype to
+include fields that do not exist in the supertype.
 

### 2015-08-13 doc/thesis/typedlua.tex
Fixing some typos and some glossary definitions

@@
 
-Typed Lua provides two operation modes: the \emph{default} mode and
+Typed Lua provides two modes of operation: the \emph{default} mode and
 the \emph{strict} mode.
@@
 We will use the following function to illustrate how these two
-operation modes work:
+modes work:
 \begin{verbatim}
@@
 
-Even though these operation modes affect function calls, they do not
+Even though these modes of operation affect function calls, they do not
 affect other multiple assignments.
-This means that both operation modes discard extra arguments in
+This means that both modes discard extra arguments in
 multiple assignments and they also use \texttt{nil} to replace missing
 arguments in multiple assignments.
-We made this design decision to avoid being to restrictive, that is,
+We made this design decision to avoid being too restrictive, that is,
 we did not want to give unnecessary warnings to programmers,
@@
 \end{verbatim}
-compiles without any compile-time warnings in both operation modes.
+compiles without any compile-time warnings in both modes of operation.
 This means that the left-hand side of this local declaration has the
@@
 list in the right-hand side of this local declaration has the tuple
-type \texttt{(1, 2, 3, nil*)} in both operation modes.
+type \texttt{(1, 2, 3, nil*)} in both modes.
 The literal type \texttt{3} is consistent with \texttt{value},
@@
 \end{verbatim}
-compiles with one compile-time warning in both of the operation modes.
+compiles with one compile-time warning in both modes of operation.
 This means that the left-hand side of this local declaration has
@@
 expression list in the right-hand side of this local declaration
-has the tuple type \texttt{(integer, nil*)} in both operation modes.
+has the tuple type \texttt{(integer, nil*)} in both modes.
 The type \texttt{nil*} generates a value of type \texttt{nil}
@@
 \end{verbatim}
-compiles without compile-time warnings in both operation modes.
+compiles without compile-time warnings in both modes of operation.
 This happens because in the right-hand side of the multiple
@@
 \end{verbatim}
-has type \texttt{(string*) -> (string*)} in both operation modes.
+has type \texttt{(string*) -> (string*)} in both modes of operation.
 The function call \texttt{v()} compiles without any compile-time warnings in
@@
 After we define the function \texttt{message}, we can call it
-with a missing argument in both operation modes:
+with a missing argument in both modes of operation:
 \begin{verbatim}

### 2015-08-13 doc/thesis/evaluation.tex
Improving table formats

@@
 \begin{tabular}{|l|c|c|c|c|c|c|}
+\cline{2-5}
+\multicolumn{1}{c}{} & \multicolumn{4}{|c|}{percentage of members} & \multicolumn{1}{c}{} \\
 \hline
-\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
+\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{\# members} \\
 \hline
@@
 \begin{tabular}{|c|c|c|c|c|c|}
+\cline{2-5}
+\multicolumn{1}{c}{} & \multicolumn{4}{|c|}{percentage of members} & \multicolumn{1}{c}{} \\
 \hline
-\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
+\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{\# members} \\
 \hline

### 2015-08-13 doc/thesis/review.tex
Improving table formats

@@
 \begin{center}
-\begin{tabular}{|l|p{6cm}|c|}
+\begin{tabular}{|p{6cm}|p{6cm}|c|}
 \hline
-\multirow{5}{*}{table constructor}
+\multirow{5}{*}{\minitab[l]{table constructor \\ (\% per static occurrences)}}
 & create a record & 36\% \\
@@
 \hline
-\multirow{4}{*}{table access}
+\multirow{4}{*}{\minitab[l]{table access \\ (\% per static occurrences)}}
 & reading with literal keys & 80\% \\
@@
 \hline
-\multirow{2}{*}{iteration over tables} &
+\multirow{2}{*}{\minitab[l]{iteration over tables \\ (\% per files)}} &
 files that iterate over a list & 27\% \\
@@
 \hline
-\multirow{4}{*}{function declarations}
+\multirow{4}{*}{\minitab[l]{function declarations \\ (\% per static occurrences)}}
 & inspect the tags of their input parameters & 9\% \\
@@
 \hline
-\multirow{1}{*}{object-oriented programming}
+\multirow{1}{*}{\minitab[l]{object-oriented programming \\ (\% per projects)}}
 & projects that use metatables and declare methods & 45\% \\
 \hline
-\multirow{2}{*}{modules}
+\multirow{2}{*}{\minitab[l]{modules \\ (\% per files)}}
 & files that are modules & 61\% \\

### 2015-08-17 doc/thesis/conclusions.tex
Improving the informal description about table tags

@@
 members that these modules export.
-In the median, we could give precise static types to 89\%
-of the members from each module.
+For half of the modules, we could give precise static types to
+at least 89\% of the members from each module.
 Our evaluation results showed that our type system can statically
@@
 For instance, ZeroBrane Studio is an IDE for Lua development that is
-starting to use Typed Lua to perform static analysis in Lua code.
+evaluating the use of Typed Lua to perform static analysis in Lua code.

### 2015-08-17 doc/thesis/typedlua.tex
Improving the informal description about table tags

@@
 \emph{fixed}, or \emph{closed} to identify which operations are safe.
-The table constructor describes \emph{unique} table types,
-as they are tables that we know all the fields and that there is
-no alias to it.
-Any alias to \emph{unique} tables makes them \emph{open},
-as we use this tag to keep track of \emph{unique} aliasing.
-In the previous example, \texttt{person} has an \emph{unique} type
-because it has no alias.
-In Section \ref{sec:oop} we will show that we use the tag
-\emph{fixed} to describe the type of classes.
 Type annotations define \emph{closed} table types, as this tag
@@
 that do not appear in the table type.
+In contrast, \emph{unique}, \emph{open}, and \emph{fixed} table types
+guarantee that a table does not contain any keys which are
+not listed in its table type, though they provide different guarantees
+over their references.
+An \emph{unique} table type guarantees that it has only one reference.
+In particular, this is the case of table constructors, so they
+always have \emph{unique} table types.
+In the previous example, \texttt{person} has an \emph{unique} type
+because it has no alias.
+Any alias to \emph{unique} tables makes them \emph{open},
+as we use this tag to keep track of \emph{unique} aliasing.
+This means that \emph{open} table types can only have \emph{closed}
+aliases, and it is guaranteed that only one reference remains \emph{open}.
+Finally, \emph{fixed} table types can have any number of \emph{closed}
+and \emph{fixed} aliases, so its type cannot change.
+In Section \ref{sec:oop} we will show that we use \emph{fixed} table
+types to describe the type of classes.
 Typed Lua also has different subtyping rules that handle these
@@
 
-Typed Lua uses three rules to decide whether it is sound to
-allow field assignment to change the type of \emph{unique} and
-\emph{open} tables:
-the table must assigned to a local variable to the current block,
+Typed Lua uses three rules to check whether field assignment can
+change the type of \emph{unique} and \emph{open} tables:
+the table must be assigned to a local variable to the current block,
 the new type must only add new fields, and
@@
 Back to the rules that Typed Lua uses to allow the refinement
-of table types, the following example shows that it is not sound
-to allow the type of the fields to change:
+of table types, the following example shows that it is forbidden
+to change the type of existing fields:
 \begin{verbatim}
@@
 
-In this example, both third and fourth lines are unsound because
+In this example, both third and fourth lines are not valid because
 they are trying to change the type of fields that already exist.
-
-It is also not sound to allow changing the type of an alias:
+Even though it would be sound to allow changing the type of fields
+in \emph{unique} table types because they cannot have any alias,
+this could still lead to a weak static type checking approach.
+In other words, allowing the type of the fields to evolve from a
+base type to an union type or even to the dynamic type could hide
+some type errors, so our design choice was to keep it as precise
+as we could.
+Moreover, it would be unsound to allow changing the type of fields
+in \emph{open} table types because they can have \emph{closed} aliases,
+so it made more sense to have the same rule to both table types.
+
+The following example shows that it is not allowed to change
+the type of an alias:
 \begin{verbatim}
@@
     local Shape = { x = 0.0, y = 0.0 }
+
     function Shape:move (dx:number, dy:number)
@@
 
-    const function Circle:new (x:number,
-                               y:number,
+    const function Circle:new (x:number, y:number,
                                radius:value):self

### 2015-08-18 doc/papers/paper/paper.tex
small fixes

@@
 	& | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
-	& | \; \mathbf{rec} \; id{:}F = f \; \mathbf{in} \; s & \textit{recursive function} \\
+	& | \; \mathbf{rec} \; id{:}F = e \; \mathbf{in} \; s & \textit{recursive declaration} \\
 	& | \; \mathbf{return} \; el & \textit{return} \\
@@
 	& | \; e_{1}[e_{2}]_{l} & \textit{table assignment}\\
-	& | \; id[k] \; {<}V{>} & \textit{type coercion}\\
+	& | \; id[e] \; {<}V{>} & \textit{type coercion}\\
 	k ::= & & \textsc{literal constants:}\\
@@
 	\env_{1}(id) = \{ \overline{F:V} \}_{unique}\\
-	\env_{1}, \penv \vdash k:F^{new}, \env_{2} \;\;\;
+	\env_{1}, \penv \vdash e:F^{new}, \env_{2} \;\;\;
 	\overline{F^{new} \not\lesssim F} \;\;\;
@@
 {\begin{array}{l}
-	\env_{1}, \penv \vdash id[k] {<}V{>}:V^{new},
+	\env_{1}, \penv \vdash id[e] {<}V{>}:V^{new},
 	\env_{2}[id \mapsto \{ \overline{F{:}V}, F^{new}{:}V^{new}\}_{unique}]
@@
 \item all function declarations are anonymous, so there is a
-statement for declaring recursive functions;
+statement for recursive declarations;
 \item function or method applications that throw away all

### 2015-08-18 doc/papers/dls15/paper.tex
change point size, squeeze figures

--- 2015-08-18 doc/papers/paper/paper.tex
@@
-\documentclass[preprint]{sigplanconf}
+\documentclass[10pt]{sigplanconf}
 
@@
 \begin{figure}[t]
+{\footnotesize
 \begin{verbatim}
@@
 \end{verbatim}
+}
 \caption{Building an object piece-by-piece}
@@
 \end{verbatim}
+}
 \caption{Overloading on multiple returns}
@@
 
-\begin{figure*}[t]
+\begin{figure}[t]
+{\scriptsize
 $$
@@
 L ::= & & \textsc{{\small literal types:}}\\
-& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
+&  \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
 %\multicolumn{3}{c}{}\\
 B ::= & & \textsc{{\small base types:}}\\
-& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\\\
+& \Boolean \ |\  \Integer \ |\ \Number\ |\ \String &\\\\
 %\multicolumn{3}{c}{}\\
@@
 $$
+}
 \caption{Type Language}
 \label{fig:typelang}
-\end{figure*}
-
-\begin{figure*}[t]
+\end{figure}
+
+\begin{figure}[t]
+{\footnotesize
 $$
@@
 $$
+}
 \caption{Special Types}
 \label{fig:special}
-\end{figure*}
+\end{figure}
 
@@
 \begin{figure}[t]
+{\footnotesize
 \[
@@
 \]
+}
 \caption{Subtyping for second-level types}	
@@
 \]
+}
 \caption{Subtyping for table types}
@@
 \]
+}
 \caption{Subtyping for fields}
@@
 \]
+}
 \caption{Subtyping for recursive types}
@@
 
-\begin{figure*}[t]
+\begin{figure}[t]
+{\scriptsize
 	$$
@@
 	& | \; \mathbf{return} \; el & \textit{return} \\
-	& | \; \lfloor a \rfloor_{0} & \textit{application with no results}\\
+	& | \; \lfloor a \rfloor_{0} & \textit{function call}\\
 	& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \textit{method declaration}\\
@@
 	| \; \# \; e & \textit{unary operations} \\
-	& | \; \lfloor me \rfloor_{1} & \textit{expressions with one result}\\
-	l ::= & & \textsc{left-hand values:}\\
+	& | \; \lfloor me \rfloor_{1} & \textit{one result}\\
+	l ::= & & \textsc{lvalues:}\\
 	& \;\; id_{l} & \textit{variable assignment}\\
@@
 	& | \; id[e] \; {<}V{>} & \textit{type coercion}\\
-	k ::= & & \textsc{literal constants:}\\
+	k ::= & & \textsc{literals:}\\
 	& \;\; \mathbf{false} \; | \;
@@
 	& | \; {...} & \textit{vararg expression}\\
-	a ::= & & \textsc{applications:}\\
-	& \;\; e(el) & \textit{function application}\\
-	& | \; e{:}n(el) & \textit{method application}\\
-	f ::= & & \textsc{function declarations:}\\
+	a ::= & & \textsc{calls:}\\
+	& \;\; e(el) & \textit{function call}\\
+	& | \; e{:}n(el) & \textit{method call}\\
+	f ::= & & \textsc{functions:}\\
 	& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
@@
 	$$
+}
 	\caption{Abstract Syntax}
 	\label{fig:syntax}
-\end{figure*}
+\end{figure}
 
@@
 \begin{figure*}[t]
-{\small
+{\footnotesize
 \[
@@
 
-\begin{figure*}
-	{\small
+\begin{figure}
+	{\footnotesize
 	\[
 	\begin{array}{rcl}
-	wf(\{\overline{F:V}\}_{unique|open|fixed|closed}) & = & \forall i \; ((\nexists j \; i \not= j \,\wedge\, F_{i} \lesssim F_{j}) \,\wedge\, wf(V_{i}) \,\wedge\\
-	& & \;\;\;\; \lnot tag(V_{i},unique) \,\wedge\, \lnot tag(V_{i},open))\\
+	wf(\{\overline{F:V}\}_{u|o|f|c}) & = &
+	 \forall i \; ((\nexists j \; i \not= j \,\wedge\, F_{i} \lesssim F_{j}) \,\wedge \\
+	 & &  wf(V_{i}) \,\wedge\, \lnot tag(V_{i},unique) \,\wedge\\
+	 & &  \lnot tag(V_{i},open))\\
 	wf({\bf const}\; F) & = & wf(F) \\
@@
 	\multicolumn{3}{c}{
-		\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{other\;recusive\;cases\;are\;similar,\;other\;base\;cases\;are\;}\bot}\\
+		\;\;\;\;\mathrm{other\;recusive\;cases\;are\;similar,\;other\;base\;cases\;are\;}\bot}\\
 \\\\
@@
 	\{\overline{F:V}\}_{fixed}\\
-	\multicolumn{3}{c}{\;\;\;\;\;\;\;\mathrm{other\;cases\;by\;straightforward\;structural\;recursion}}\\\\
+	\multicolumn{3}{c}{\mathrm{other\;cases\;by\;straightforward\;structural\;recursion}}
+	\end{array}
+	\]}
+	\caption{Auxiliary Functions}
+\label{fig:funcs}
+\end{figure}
+
+\begin{figure*}
+{\footnotesize
+\[
+\begin{array}{rcl}
 	merge(\overline{\env}) & = & reduce(\overline{\env}, merge2)\\
@@
 	joint(T_1, T_2) & = & \bot \;\;\; \mathrm{otherwise}\\\\
-	sup_u(V_1, V_2) & = & V_2 \;\;\; \mathrm{if}\; V_1 \lesssim_u V_2\\
-	sup_u(V_1, V_2) & = & V_1 \;\;\; \mathrm{if}\; V_2 \lesssim_u V_1\\\\
-	reopen(\{\overline{F:V}\}_{closed}) & = & \{\overline{F:V}\}_{open} \\
-	reopen(F) & = & F
-	\end{array}
+	sup_u(V_1, V_2) \,=\, V_2 \;\;\; \mathrm{if}\; V_1 \lesssim_u V_2 & &
+	   sup_u(V_1, V_2) \,=\, V_1 \;\;\; \mathrm{if}\; V_2 \lesssim_u V_1 \\
+	\\
+	reopen(\{\overline{F:V}\}_{closed}) \,=\, \{\overline{F:V}\}_{open} & &
+	  reopen(F) \,=\, F\end{array}
 \]}
-	\caption{Auxiliary Functions}
-\label{fig:funcs}
+\caption{Merge and Join Functions}
+\label{fig:mergejoin}
 \end{figure*}
@@
 of function {\tt new\_point} to the abstract syntax of Figure~\ref{fig:syntax}:
+{\small
 \begin{center}
@@
 \end{center}
+}
 
@@
 as in the following example:
+{\small
 \begin{center}
@@
 \end{center}
-
+}
 In this case, the first coercion expression would still change the type of $point$ to $\{``x":\Number\}_{unique}$, but the
@@
 of a {\em unique} table:
+{\small
 \begin{center}
@@
 table is now {\em open}:
+{\small
 \begin{center}
@@
 $\{ ``x": \String \cup \Nil\}_{closed}$:
+{\small
 \begin{center}
@@
 The following example typechecks, and uses all the parts of {\sc T-Function}:
+{\small
 \begin{center}
@@
 \begin{figure*}[t]
-{\small
+{\footnotesize
 \label{fig:projtypes}
@@
 the body of the {\bf then} and {\bf else} branches:
+{\small
 \begin{center}
@@
 $\Integer$:
+{\small
 \begin{center}

### 2015-08-18 doc/papers/dls15/paper.tex
squeeze the paper into 13 pages after changing text to 10-point

@@
 
+%\usepackage[compact]{titlesec}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{multirow}
+
+\setlength{\textfloatsep}{10pt}
 
@@
     local point = {}
-    point.x = x
-    point.y = y
+    point.x, point.y = x, y
     function point:move(dx, dy)
-      self.x = self.x + dx
-      self.y = self.y + dy
-      return self
+      self.x, self.y = self.x + dx, self.y + dy
     end
@@
   function idiv(x, y)
-    if y == 0 then
-      return nil, "division by zero"
-    else
-      return x // y, x % y
+    if y == 0 then return nil, "division by zero"
+    else return x // y, x % y end 
   end
-  end
 
   local q, r = idiv(a, b)
-  if q then
-    assert(a == q*b + r)
-  else
-    error(r)
-  end
+  if q then assert(a == q*b + r)
+  else error(r) end
 \end{verbatim}
@@
 
-The rationale for the permissiveness of rule {\sc S-TableUtoC} derives from the fact that reading a non-existing key just yields $\Nil$ instead of being an error in Lua semantics, plus the fact
+The rationale for the permissiveness of {\sc S-TableUtoC} derives from the fact that reading a non-existing key just yields $\Nil$ instead of being an error in Lua semantics, plus the fact
 that the type system guarantees that a table with an
@@
 
-In the implementation of Typed Lua, we also use consistent-subtyping to normalize and simplify union types.
+The implementation of Typed Lua also uses consistent-subtyping to normalize and simplify union types.
 For instance, the union type $\Boolean \cup \Any$ simplifies
@@
 us assign this table to a variable with a more general
-type such as $\{ ``x":\Number , ``y": \Number \}_{unique}$ or
+type: $\{ ``x":\Number , ``y": \Number \}_{unique}$, or
 even $\{ \String:\Number \cup \Nil \}_{unique}$.
@@
 
-Rule {\sc T-Constructor} also checks if the resulting type is {\em well-formed} with the predicate $wf$. Checking for well-formedness avoids ambiguous table types like
+Rule {\sc T-Constructor} also checks if the resulting type is {\em well-formed} with the predicate $wf$.
+Checking for well-formedness avoids ambiguous table types.
+An example of ambiguous table type is
 $\{1:\Number, \Integer:\String, \Any:\Boolean\}$,
@@
 of function {\tt new\_point} to the abstract syntax of Figure~\ref{fig:syntax}:
-{\small
+{\footnotesize
 \begin{center}
@@
 \multicolumn{2}{l}{$\mathbf{local} \; point: \{\}_{unique} = \{\} \; \mathbf{in}$}\\
-& \multicolumn{1}{l}{$point[``x"] \; {<}\Number{>} = x;$}\\
-& \multicolumn{1}{l}{$point[``y"] \; {<}\Number{>} = y$}
+& \multicolumn{1}{l}{$point[``x"] \; {<}\Number{>} = x;\;
+point[``y"] \; {<}\Number{>} = y$}
 \end{tabular}
@@
 as in the following example:
-{\small
+{\footnotesize
 \begin{center}
@@
 of a {\em unique} table:
-{\small
+{\footnotesize
 \begin{center}
@@
 table is now {\em open}:
-{\small
+{\footnotesize
 \begin{center}
@@
 \multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{ \} \; \mathbf{in}$}\\
-& \multicolumn{3}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
-& \multicolumn{3}{l}{$a[``y"] \; {<}\String{>} = ``bar";$}\\
+& \multicolumn{3}{l}{$a[``x"] \; {<}\String{>} = ``foo";\;
+a[``y"] \; {<}\String{>} = ``bar";$}\\
 & \multicolumn{3}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{closed} =$}\\
@@
 $\{ ``x": \String \cup \Nil\}_{closed}$:
-{\small
+{\footnotesize
 \begin{center}
@@
 			= \{\} \; \mathbf{in}$}\\
-		& \multicolumn{3}{l}{$\mathbf{local} \; f:\Void \rightarrow \Void =$}\\
-		& & \multicolumn{2}{l}{$\mathbf{fun} \; ():\Void\;\;\;
+		& \multicolumn{3}{l}{$\mathbf{local} \; f:\Void \rightarrow \Void =\;\mathbf{fun} \; ():\Void\;\;\;
 		 a = \{[``x"]=5\} $}\\
-		& \multicolumn{3}{l}{$\mathbf{in} \; \lfloor g({<}\{``x":
+		& & \multicolumn{2}{l}{$\mathbf{in} \; \lfloor g({<}\{``x":
 			 \String \cup \Nil\}_{closed}{>} \; a) \rfloor_{0}$}
@@
 The following example typechecks, and uses all the parts of {\sc T-Function}:
-{\small
+{\footnotesize
 \begin{center}
@@
 {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
-\end{array}\\\\
-%\begin{array}{rcll}
-%last(F_1{*} \sqcup F_2{*}) & = & F_1{*} & \mathrm{if} \; F_2{*} \subtype F_1{*}\\
-%last(F_1{*} \sqcup F_2{*}) & = & F_2{*} & \mathrm{if} \; F_1{*} \subtype F_2{*}\\
-%last(F_1{*} \sqcup F_2{*}) & = & (F_1 \cup F_2){*} & \mathrm{otherwise}\\
-%last(F{*} \sqcup T \times P) & = & last(\Nil{*} \sqcup P) \\
-%last(T \times P \sqcup F{*}) & = & last(P \sqcup \Nil{*}) \\
-%last(P \sqcup S) & = & last(P \sqcup last(S)) \\
-%last(S \sqcup P) & = & last(last(S) \sqcup P) \\
-%last(S_1 \sqcup S_2) & = & last(last(S_1) \sqcup last(S_2))
-%\end{array}\\\\
-\begin{array}{rcll}
-proj(S_1 \sqcup S_2, i) & = & proj(S_1, i) \cup proj(S_2, i) \\
-proj(F{*}, i) & = & nil(F) \\
-proj(F \times P, 1) & = & F \\
-proj(F \times P, i) & = & proj(P, i-1)\\
-proj(E{*}, i) & = & nil(E) \\
-proj(T \times E, 1) & = & T \\
-proj(T \times E, i) & = & proj(E, i-1)
-\end{array}\\\\
-\begin{array}{rcll}
-fonilpt(P_1 \sqcup P_2) & = & P_2 & \mathrm{if} \; fonil(proj(P_1, i)) = {\bf void} \\
-fonilpt(P_1 \sqcup P_2) & = & P_1 & \mathrm{if} \; fonil(proj(P_2, i)) = {\bf void} \\
-fonilpt(P_1 \sqcup P_2) & = & P_1 \sqcup P_2 & \mathrm{otherwise}\\
-fonilpt(P \sqcup S) & = & fonilpt(S) & \mathrm{if} \; fonil(proj(P, i)) = {\bf void} \\
-fonilpt(P \sqcup S) & = & P \sqcup fonilpt(S) & \mathrm{otherwise}\\
-fonilpt(S \sqcup P_2) & = & fonilpt(S) & \mathrm{if} \; fonil(proj(P, i)) = {\bf void} \\
-fonilpt(S \sqcup P) & = & fonilpt(S) \sqcup P & \mathrm{otherwise}\\
-fonilpt(S_1 \sqcup S_2) & = & fonilpt(S_1) \sqcup fonilpt(S_2)
-\end{array}\\\\
-\begin{array}{rcll}
-fonil(F_1 \cup F_2) & = & fonil(F_1) & \mathrm{if}\;fonil(F_2) = {\bf void}\\
-fonil(F_1 \cup F_2) & = & fonil(F_2) & \mathrm{if}\;fonil(F_1) = {\bf void}\\
-fonil(F_1 \cup F_2) & = & fonil(F_1) \cup fonil(F_2) & \mathrm{otherwise}\\
-%\multicolumn{4}{c}{\mathrm{the\;next\;two\;cases\;apply\;if\;}F
-%	\;\mathrm{is\;not\;a\;union}}\\
-fonil(F) & = & {\bf void} & \mathrm{if} \; F \subtype \Nil \; \mathrm{and} \; \Nil \subtype F\\
-fonil(F) & = & F & \mathrm{otherwise}
 \end{array}
@@
 \end{figure*}
+
+\begin{figure}[t]
+{\scriptsize
+\label{fig:projtypes}
+\[
+\begin{array}{c}
+\begin{array}{rcl}
+fonilpt(P_1 \sqcup P_2) & = & P_2 \;\; \mathrm{if} \; fonil(proj(P_1, i)) = {\bf void} \\
+fonilpt(P_1 \sqcup P_2) & = & P_1 \;\; \mathrm{if} \; fonil(proj(P_2, i)) = {\bf void} \\
+fonilpt(P_1 \sqcup P_2) & = & P_1 \sqcup P_2 \;\; \mathrm{otherwise}\\
+fonilpt(P \sqcup S) & = & fonilpt(S) \;\; \mathrm{if} \; fonil(proj(P, i)) = {\bf void} \\
+fonilpt(P \sqcup S) & = & P \sqcup fonilpt(S) \;\; \mathrm{otherwise}\\
+fonilpt(S \sqcup P_2) & = & fonilpt(S) \;\; \mathrm{if} \; fonil(proj(P, i)) = {\bf void} \\
+fonilpt(S \sqcup P) & = & fonilpt(S) \sqcup P \;\; \mathrm{otherwise}\\
+fonilpt(S_1 \sqcup S_2) & = & fonilpt(S_1) \sqcup fonilpt(S_2)
+\end{array}\\\\
+\begin{array}{rcl}
+fonil(F_1 \cup F_2) & = & fonil(F_1) \;\; \mathrm{if}\;fonil(F_2) = {\bf void}\\
+fonil(F_1 \cup F_2) & = & fonil(F_2) \;\; \mathrm{if}\;fonil(F_1) = {\bf void}\\
+fonil(F_1 \cup F_2) & = & fonil(F_1) \cup fonil(F_2) \;\; \mathrm{otherwise}\\
+%\multicolumn{4}{c}{\mathrm{the\;next\;two\;cases\;apply\;if\;}F
+%	\;\mathrm{is\;not\;a\;union}}\\
+fonil(F) & = & {\bf void} \;\; \mathrm{if} \; F \subtype \Nil \; \mathrm{and} \; \Nil \subtype F\\
+fonil(F) & = & F \;\; \mathrm{otherwise}
+\end{array}
+\end{array}
+\]}
+\caption{Auxiliary Functions for Projection Types}
+\end{figure}
 
@@
 the body of the {\bf then} and {\bf else} branches:
-{\small
+{\footnotesize
 \begin{center}
@@
 $\Integer$:
-{\small
+{\footnotesize
 \begin{center}
 	\begin{tabular}{ll}
-		\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(a, b) \, \mathbf{in}$} \\
-		&  $r = ``foo" ;\; \mathbf{if} \, q \, \mathbf{then} \, s_1 \, \mathbf{else} \, s_2$
+		\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(a, b) \, \mathbf{in}\; r = ``foo" ;\; \mathbf{if} \, q \, \mathbf{then} \, s_1 \, \mathbf{else} \, s_2$}
 	\end{tabular}

### 2015-08-18 doc/papers/dls15/paper.tex
preparing dls paper for camera-ready version

@@
 
-%\usepackage[compact]{titlesec}
+\usepackage[compact]{titlesec}
 \usepackage[utf8]{inputenc}
@@
 
-\setlength{\textfloatsep}{10pt}
+\setlength{\textfloatsep}{8pt}
 
@@
 In this work, we give a formal description of Typed Lua, an
-optional type system for Lua, with a focus on two of its novel type system features: incremental evolution of imperative record and object types that is both lightweight and type-safe, and
-the combination of flow typing, functions that return multiple values, and multiple assignment.
+optional type system for Lua, with a focus on two of its novel type system features: incremental evolution of imperative record and object types that is both lightweight and type-safe, and {\em projection types}, a combination of flow typing, functions that return multiple values, and multiple assignment.
 While our type system is tailored to the features and idioms of Lua, its features can be adapted to other imperative scripting languages.
@@
 allowing the incremental migration from dynamic to static
-typing. The difference between these two approaches is the
+typing.
+
+The difference between these two approaches is the
 way they treat the runtime semantics of the language:
-optional type systems preserve the original semantics
-of a dynamically typed language, at the cost of runtime
-safety, while gradual typing changes the semantics to
-include more extensive runtime checking. These extra checks
-ensure that
-dynamically typed portions of the program do not violate
-the type safety of statically typed portions.
+optional type systems have the {\em trace preservation}
+property~\cite{strongscript}, where type annotations
+do not influence runtime behavior, and the original
+semantics of the dynamically typed language is preserved,
+ while gradual typing changes the semantics to
+include more extensive runtime checking in the
+interface between typed and untyped code. 
+In an optionally-typed language, dynamically typed code
+can lead to type errors in statically typed code, even
+if the optional type system is sound. In a gradually-typed
+language, the extra runtime checks ensure that 
+this does not occur.
+ The decision
+between the two approaches is not binary, and languages
+can have both optional and gradual types~\cite{liketypes,strongscript}.
 
@@
 lets record and object types evolve incrementally 
-via assignment while not sacrificing width subtyping.
+via assignment while not sacrificing width subtyping,
+where we can form a supertype of a table type by ``forgetting'' fields.
 
@@
 In Typed Lua, the return type of {\tt idiv} can be a union of
-tuples describing the possible return values. Typed Lua deconstructs unions of tuples in a way that the dependencies between the types of each value in the tuple can be tracked.
+tuples describing the possible return values. Typed Lua deconstructs unions of tuples into {\em projection types}, and these types track
+the dependencies between the types of each value in the tuple.
 
@@
 types} $F_i$.
+
 %We restrict key types to literal types,
@@
 Fields are assignable by default, but can be marked
-as non-assignable with the {\bf const} qualifier. Having
-a field be {\bf const} is not a guarantee that its
+as non-assignable with the {\bf const} qualifier in its key type.
+Having a field be {\bf const} is not a guarantee that its
 value cannot change, as the table may have aliases with
 a non-{\bf const} type for that field. Non-assignable
-fields enable {\em depth subtyping} between table types.
+fields enable {\em depth subtyping} between table types: these
+fields are covariant on the types of their values, while
+assignable fields are always invariant.
 
@@
 and a projection environment $\penv$ we can prove that an expression $e$ has type $T$ and produces a new type environment $\env_{2}$.
-Only the most relevant rules for the features that we are focusing on are shown.
+Only the most relevant rules are shown. The first author's Ph. D. thesis has the full set of rules~\cite{maidl2015tl}.
 
@@
 
-Ravi~\cite{ravi} is an experimental Lua dialect
-with an LLVM-based JIT compiler. It has a basic
-optional type system to improve performance
-of numerical code. This type system only supports
-numbers (integers and floating point) and arrays.
-
 Typed Racket~\cite{tobin-hochstadt2008ts} is a statically typed version of the Racket language, which is a Scheme dialect.
@@
 type systems: Clojure~\cite{bonnaire-sergeant2015typed-clojure},
-JavaScript~\cite{typescript}, Python~\cite{mypy}, 
-and PHP~\cite{hack}. While Lua has some similarities to
+JavaScript, Python~\cite{mypy}, 
+and PHP. While Lua has some similarities to
 all of these languages, none of these optional type
@@
 
-Dart~\cite{dart} and Grace~\cite{black2013sg} are
+Grace~\cite{black2013sg} and Dart are
 two languages that have been designed from scratch
@@
 For half of the modules, we could give precise static types to
-at least 89\% of the exports from each module.
+at least 89\% of the exports from each module. The methodology,
+detailed results and analysis of this evaluation is part
+of the first author's Ph. D. thesis~\cite{maidl2015tl}.
 
@@
 
+\acks
+
+The authors thank the anonymous reviewers for their feedback.
+Work on Typed Lua has been supported by grants from Google
+Summer of Code and CNPq, the Brazilian Research Council.
+
 \bibliographystyle{abbrvnat}

### 2015-08-18 doc/papers/dls15/paper.tex
fix a few typos in rules

@@
 \mylabel{T-CoerceClosed}\\
-\dfrac{\env_{1}(id) \subtype F \;\;\; tag(F,closed)}
+\dfrac{\env(id) \subtype F \;\;\; tag(F,closed)}
 {\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto reopen(F)]}
@@
 \mylabel{T-CoerceFixed}\\
-\dfrac{\env_{1}(id) \subtype F \;\;\; tag(F,fixed)}
+\dfrac{\env(id) \subtype F \;\;\; tag(F,fixed)}
 {\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto F]}
@@
 	closeall(\env_{1})[\overline{id \mapsto F}], \penv[\ret \mapsto S] \vdash s, \env_{2}\\
-	\env_{3} = openset(\env_{1}, frv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)) \\
-	\env_{4} = closeset(\env_{3}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
-	\end{array}}
-{\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}F}){:}S \; s:F_1 \times \ldots F_n \times \Void \rightarrow S, \env_{3}}
+	\env_{3} = openset(\env_{1}, frv(\mathbf{fun} \; (\overline{id{:}F}){:}S \; s)) \\
+	\env_{4} = closeset(\env_{3}, fav(\mathbf{fun} \; (\overline{id{:}F}){:}S \; s))
+	\end{array}}
+{\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}F}){:}S \; s:F_1 \times \ldots F_n \times \Void \rightarrow S, \env_{4}}
 \end{array}
@@
 \dfrac{\begin{array}{c}
-	\env_{1}, \penv \vdash el:E, \env_{2} \\
-	E \lesssim F_1 \times \ldots \times F_n \times \Value{*} \\
+	\env_{1}, \penv \vdash el:P, \env_{2} \\
+	P \lesssim F_1 \times \ldots \times F_n \times \Value{*} \\
 	\env_{2}[\overline{id \mapsto F}], \penv \vdash s, \env_{3}
@@
 Checking for well-formedness avoids ambiguous table types.
-An example of ambiguous table type is
+An example of ambiguous table type is the type
 $\{1:\Number, \Integer:\String, \Any:\Boolean\}$,
@@
 \dfrac{\begin{array}{c}
-	\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
+	\env_{1}, \penv \vdash e_{i}:F_{i}, \env_{i+1} \;\;\;
 	\env_{1}, \penv \vdash me:S, \env_{n+2} \;\;\; n=|\overline{e}|\\
@@
 	\end{array}}
-{\env_{1}, \penv \vdash \overline{e},me:}\\ {T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x}{*}, \env_{f}, (x,S)}
+{\env_{1}, \penv \vdash \overline{e},me:}\\ {F_{1} \times ... \times F_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x}{*}, \env_{f}, (x,S)}
 \end{array}\\\\

### 2015-08-20 doc/papers/dls15/paper.tex
Fixing some typos for DSL'15 camera ready

@@
 %\multicolumn{3}{c}{}\\
-L ::= & & \textsc{{\small literal types:}}\\
+L ::= & & \textsc{{\tiny literal types:}}\\
 &  \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
 %\multicolumn{3}{c}{}\\
-B ::= & & \textsc{{\small base types:}}\\
+B ::= & & \textsc{{\tiny base types:}}\\
 & \Boolean \ |\  \Integer \ |\ \Number\ |\ \String &\\\\
@@
 %\multicolumn{3}{c}{}\\
+V ::= & & \textsc{{\tiny value types:}}\\
+& \;\; F \; | \; \Const \; F &\\
 S ::= & & \textsc{second-level types:}\\
@@
 %\multicolumn{3}{c}{}\\
-P ::= & & \textsc{{\small tuple types:}}\\
+P ::= & & \textsc{{\tiny tuple types:}}\\
 %& \;\; \Void & \textit{void type}\\
@@
 \begin{array}{rlr}
-T ::= & & \textsc{{\small expression types:}}\\
-& \;\; F & \\
+T ::= & & \textsc{expression types:}\\
+& \;\; F & \textit{first-level types}\\
 & | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
 & | \; \pi_{i}^{x} & \textit{projection types}\\
-E ::= & & \textsc{{\small expression list types:}}\\
-& \;\; T{*} & \\
-& | \; T \times E & \\
+E ::= & & \textsc{expression list types:}\\
+& \;\; T{*} & \textit{variadic types}\\
+& | \; T \times E & \textit{pair types}\\
 %K ::= & & \textsc{{\small key types:}}\\
@@
 %\multicolumn{3}{c}{}\\
-V ::= & & \textsc{{\small value types:}}\\
-& \;\; F \; | \; \Const \; F &\\
-R ::= & & \textsc{\small{filter result type:}}\\
+R ::= & & \textsc{filter result type:}\\
 & \;\; \mathbf{void} & \textit{void type}\\
-& |\; F
+& |\; F & \textit{first-level types}
 \end{array}
@@
 information that can be used to blame the dynamically
-typed of the program for type errors that may occur in the
+typed part of the program for type errors that may occur in the
 statically typed part. Currently Typed Lua only
@@
 decomposing a union of tuples. {\em Expression list types}
-denote the type of lists of expressions. {\em Key types}
-and {\em value types} have also been mentioned above. Finally, {\em filter result types}
+denote the type of lists of expressions.
+Finally, {\em filter result types}
 are also used in flow typing; the {\bf void} type is
@@
 
-Rule {\sc S-TableUtoC} for subtyping between {\em unique} and {\em closed} table types is more permissive, as the typing rules of Section~\ref{sec:tables} guarantee that a {\em unique}
+Rule {\sc S-TableUtoC} for subtyping between {\em unique} and {\em closed} table types is more permissive, as the typing rules of Section~\ref{sec:tables} guarantee that an {\em unique}
 table type can only appear on the left side of a subtyping
@@
 
-We will revisit the subtyping rules among table types on
+We will revisit the subtyping rules among table types in
 Section~\ref{sec:tables}.
@@
 relations~\cite{siek2007objects}. We use the symbol $\sim$ for
-the consistency, and $\lesssim$ for consistent-subtyping.
+consistency, and $\lesssim$ for consistent-subtyping.
 
@@
 $S_1 \rightarrow S_2$ and $S_3 \rightarrow S_4$ are consistent
-if and only if $S_1 \sim S_3$ and $S_2 \sim S_4$.
+if and only if $S_3 \sim S_1$ and $S_2 \sim S_4$.
 Two table types are consistent if and only if they have the same key types, and the value types of each key type are consistent.
@@
 }
-\caption{Typing Rules}
+\caption{Typing Rules for Table Types}
 \label{fig:tablerules}
@@
 \item type coercion for evolution of table types is explicit,
-with two coercion terms, the l-value $id[k] \; {<}V{>}$ for adding a field to 
+with two coercion terms, the l-value $id[e] \; {<}V{>}$ for adding a field to 
 a table through assignment, and the r-value ${<}F{>} \;id$ for reinterpreting
@@
 	\]}
-	\caption{Auxiliary Functions}
+	\caption{Auxiliary Functions for Table Types}
 \label{fig:funcs}
@@
 can track, so this restriction is
-necessary. This limitation of the type system is the reason the {\sc T-Constructor} rule uses the auxiliary function $fix$
+necessary. This limitation of the type system is the reason why the {\sc T-Constructor} rule uses the auxiliary function $fix$
 (in the definition of $vt$)
@@
 
-There are two ways the type of a {\em unique} and {\em open} tables can change in Typed
+There are two ways the type of {\em unique} and {\em open} tables can change in Typed
 Lua. The first way is by assignment to a non-existing
@@
 changes from one rule to the other is the tag in the type of $id$.
-There are analogous rules for adding methods to a {\em unique}
+There are analogous rules for adding methods to {\em unique}
 and {\em open tables} as a side-effect of typechecking a method declaration.
@@
 {\em unique} tables to {\em open}. Its purpose
-is to track the fact that a $id$ now has an alias,
+is to track the fact that an $id$ now has an alias,
 if necessary.
@@
 Consider the following example, where we make an alias
-of a {\em unique} table:
+of an {\em unique} table:
 {\footnotesize
@@
 tag {\em unique} or {\em open} and add fields that would
-conflict with the type of the fields on $a$, but at
+conflict with the type of the fields in $a$, but at
 runtime both reference the same table.
@@
 {\em coercion expression}. Coercion expressions let
-a {\em unique} or {\em open} table be used in a context
+an {\em unique} or {\em open} table be used in a context
 that expects a table type with a different shape.
@@
 In the example below, we use a coercion expression to
-widen the type of a field in a {\em unique}
+widen the type of a field in an {\em unique}
 table, and then add another field, to show that the
@@
 
-The type $a$ has by the time of the inner {\bf local} block
+The type of $a$ by the time of the inner {\bf local} block
 is $\{``x":\String, ``y":\String\}_{unique}$. This type is
-a subtype of the $\{``x":\String, ``y":\String\cup\Nil\}_{closed}$, as {\em unique} tables
+a subtype of $\{``x":\String, ``y":\String\cup\Nil\}_{closed}$, as {\em unique} tables
 have depth subtyping even in non-{\bf const} fields.
@@
 Lua has first-class functions, so another way of aliasing
-an {\em unique} and {\em open} references is by closing
+{\em unique} and {\em open} references is by closing
 over them in a function body. In the environment used for
@@
 
-The type of $a$ on line 4 is $\{\}_{closed}$, so the assignment
+The type of $a$ on line 2 is $\{\}_{closed}$, so the assignment
 would type-check. But then the function call on the last line
@@
 $\{ ``x": \String \cup \Nil\}$. A solution is to change the
-type of $a$ from {\em unique} to {\em closed} after the function
+type of $a$ from {\em open} to {\em closed} after the function
 declaration. In order to do that, we use the auxiliary function
@@
 \begin{array}{rcl}
-fonilpt(P_1 \sqcup P_2) & = & P_2 \;\; \mathrm{if} \; fonil(proj(P_1, i)) = {\bf void} \\
-fonilpt(P_1 \sqcup P_2) & = & P_1 \;\; \mathrm{if} \; fonil(proj(P_2, i)) = {\bf void} \\
-fonilpt(P_1 \sqcup P_2) & = & P_1 \sqcup P_2 \;\; \mathrm{otherwise}\\
-fonilpt(P \sqcup S) & = & fonilpt(S) \;\; \mathrm{if} \; fonil(proj(P, i)) = {\bf void} \\
-fonilpt(P \sqcup S) & = & P \sqcup fonilpt(S) \;\; \mathrm{otherwise}\\
-fonilpt(S \sqcup P_2) & = & fonilpt(S) \;\; \mathrm{if} \; fonil(proj(P, i)) = {\bf void} \\
-fonilpt(S \sqcup P) & = & fonilpt(S) \sqcup P \;\; \mathrm{otherwise}\\
-fonilpt(S_1 \sqcup S_2) & = & fonilpt(S_1) \sqcup fonilpt(S_2)
+fonilpt(P_1 \sqcup P_2, i) & = & P_2 \;\; \mathrm{if} \; fonil(proj(P_1, i)) = {\bf void} \\
+fonilpt(P_1 \sqcup P_2, i) & = & P_1 \;\; \mathrm{if} \; fonil(proj(P_2, i)) = {\bf void} \\
+fonilpt(P_1 \sqcup P_2, i) & = & P_1 \sqcup P_2 \;\; \mathrm{otherwise}\\
+fonilpt(P \sqcup S, i) & = & fonilpt(S,i) \;\; \mathrm{if} \; fonil(proj(P, i)) = {\bf void} \\
+fonilpt(P \sqcup S, i) & = & P \sqcup fonilpt(S,i) \;\; \mathrm{otherwise}\\
+fonilpt(S \sqcup P_2, i) & = & fonilpt(S,i) \;\; \mathrm{if} \; fonil(proj(P, i)) = {\bf void} \\
+fonilpt(S \sqcup P, i) & = & fonilpt(S,i) \sqcup P \;\; \mathrm{otherwise}\\
+fonilpt(S_1 \sqcup S_2, i) & = & fonilpt(S_1,i) \sqcup fonilpt(S_2,i)
 \end{array}\\\\
@@
 denoted by the label $x$. Unions of tuple types referenced by projections are kept in their own {\em projection environment} $\Pi$. A fresh label is generated whenever the last term
-in an expression list has a union of tuple types as its type.
+in an expression list has an union of tuple types as its type.
 Figure~\ref{fig:projtypes} presents the rules for introducing and filtering projection types.
@@
 These functions respectively use the functions $fonil$ and $finil$ to filter regular unions, plus $proj$ to test projected components.
-The definitions of $finilpt$ and $finil$ are analogous to the definitions of $finilpt$ and $finil$.
+The definitions of $finilpt$ and $finil$ are analogous to the definitions of $fonilpt$ and $fonil$.
 
@@
 tuple for its usual path, another for error conditions)
-is modeled by having a union of tuple types as its return
+is modeled by having an union of tuple types as its return
 type. A destructuring assignment on a union of tuple types
@@
 Work on Typed Lua has been supported by grants from Google
-Summer of Code and CNPq, the Brazilian Research Council.
+Summer of Code as well as CAPES and CNPq, two Brazilian Research Councils.
 

### 2015-08-21 doc/papers/dls15/paper.tex
camera-ready version for dls 2015

@@
 
+\toappear{}
+
 \special{papersize=8.5in,11in}
@@
 
-\conferenceinfo{DLS '15}{October 25-30, 2015, Pittsburgh, PA, USA}
+\conferenceinfo{DLS '15}{October 27th, 2015, Pittsburgh, PA, USA}
 \copyrightyear{2015}
@@
            {UFRJ}
-           {fabiom@dcc.ufrj.br}
+           {mascarenhas@acm.org}
 \authorinfo{Roberto Ierusalimschy}
@@
 B ::= & & \textsc{{\tiny base types:}}\\
-& \Boolean \ |\  \Integer \ |\ \Number\ |\ \String &\\\\
+& \Boolean \, |\,  \Integer \, |\, \Number\, |\, \String &\\\\
 %\multicolumn{3}{c}{}\\
@@
 
-In our example, $x$ would be bound to $\Integer \times \Integer \times \Nil{*}$ in the {\bf then} branch, and would be bound
+In our example, variable $x$ would be bound to $\Integer \times \Integer \times \Nil{*}$ inside the {\bf then} branch, and bound
 to $\Nil \times \String \times \Nil{*}$ inside the {\bf else}

### 2015-08-21 doc/papers/dls15/paper.tex
add country to affiliation

@@
 \authorinfo{André Murbach Maidl}
-           {PUC-Rio}
+           {PUC-Rio, Brazil}
            {amaidl@inf.puc-rio.br}
 \authorinfo{Fabio Mascarenhas}
-           {UFRJ}
+           {UFRJ, Brazil}
            {mascarenhas@acm.org}
 \authorinfo{Roberto Ierusalimschy}
-           {PUC-Rio}
+           {PUC-Rio, Brazil}
            {roberto@inf.puc-rio.br}

### 2015-08-26 doc/thesis/resumo.tex
Fixing formatting required by CCPGP

@@
 e características de Lua.
-Embora Lua compartilha várias características com outras linguagens dinamicamente
+Embora Lua compartilhe várias características com outras linguagens dinamicamente
 tipadas, em particular JavaScript, Lua também possui várias características não usuais,
 as quais não estão presentes nos sistemas de tipos dessas linguagens.
-Nós discutimos como Typed Lua tipa estaticamente características tais como
-funções com aridade flexível que interagem com atribuições múltiplas,
-funções que são sobrecarregadas no número de valores de retorno e a evolução
-incremental de estruturas e objetos.
-Discutimos também nossas decisões de projeto e apresentamos uma avaliação de
-resultados, a qual obtemos ao usar Typed Lua para tipar código Lua existente.
+Essas características incluem funções com aridade flexível, atribuições múltiplas,
+funções que são sobrecarregadas no número de valores de retorno e
+a evolução incremental de registros e objetos.
+Nós discutimos como Typed Lua tipa estaticamente essas características e
+também discutimos nossas decisões de projeto.
+Finalmente, apresentamos uma avaliação de resultados,
+a qual obtivemos ao usar Typed Lua para tipar código Lua existente.
